### 2020-04-15 write-up/covid-writeup.tex
renamed

\documentclass[12pt]{article}
\usepackage{graphicx}
\usepackage{setspace}
\DeclareGraphicsExtensions{.eps, .ps}
\usepackage{soul,color,hyperref}
\usepackage{amsmath, amsthm, amsfonts}
\usepackage[margin=1in]{geometry}

\numberwithin{equation}{section}
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{example}{Example}
\newtheorem{remark}{Remark}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{criterion}[theorem]{Criterion}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}

\def\pr{\text{pr}}

\begin{document}
\title{The Hypothesis of Testing: Paradoxes arising out of reported coronavirus case-counts}
% \author{Walter Dempsey\thanks{Department of Biostatistics, University of Michigan, 1415 Washington Heights, U.S.A. E-mail: wdem@umich.edu}}
\maketitle

\begin{abstract}
Many statisticians, epidemiologists, economists and data scientists have registered their serious reservations regarding the reported coronavirus case-counts. Comparing countries and states using those case-counts seem inappropriate when every nation/state have adopted different testing strategies and protocols. Estimating prevalence of COVID-19 based on these data is a hopeless exercise and several groups have recently argued for estimating the number of truly infected cases by using mortality rates.
% In a lucid article on FiveThirtyEight.com Nate Silver walks us through the problem by a simple numerical exercise in terms of the basic deterministic models of an epidemic using hypothetical scenarios involving R0 and the effective R.
In this note, we aim to (a)	we posit a conceptual mathematical framework to characterize both sampling bias and misclassification/imperfection of the test simultaneously on the estimation of the prevalence rate, (b)	review current testing strategies in some of the countries where we have testing data, and (c) provide guidelines for testing strategy/disease surveillance that may help track the pulse of the epidemic, to identify disease free areas and identify disease outbreaks.
% Also large-scale anti-body testing.
\end{abstract}

\section{Introduction}

The World Health Organization has declared the coronavirus disease 2019 (COVID-19) a public health emergency.  As of March 28th, 2020, a total of 622,450 cases have been confirmed worldwide.  As of that afternoon, the New York Times reports at least 113,031 people across the United States have tested positive for the virus, and at least 1,895 patients with the virus have died.  Aggressive policies have been put in place across the US with at least 50\% of the US population officially urged to stay home via state-wide executive actions.

Despite these necessary steps, the data landscape for understanding COVID-19 remains limited.  Public databases maintained by Johns Hopkins University (\url{https://bit.ly/2UqFSuA}) and the New York Times (\url{https://bit.ly/2vUHfrK}) provide incoming county-level information of confirmed cases and deaths.
Statisticians, epidemiologists, economists, and data scientists the world over have been using this granular data to

The WHO reports, however, only 103,945 total tests performed in the US as of March 19th.  Due to limited testing capacity, local and state health departments have focused on testing only those from high-risk populations, resulting in low data quality due to bias in sampling from the overall population.  The public health community requires auxiliary sources of information to improve national and local health policy decisions.  Certain survey efforts are underway, but may take time to yield fruit.

The goal of this paper is to express reservations at the use of case-counts as a proxy for prevalence and disease trajectory as well as its use as direct input into estimation of standard epidemiological models for inference and forecasting.  The two critical features are selection bias due to varied testing strategies and measurement error due to imperfect tests (i.e., false positives and negatives).  We demonstrate via theoretical analysis
More testing equals more data but does not imply more information.  The key issue is selection bias and measurement-error.

Given population sizes are quite heterogeneous,
the current data are of limited use to understand prevalence and even trajectory of the disease.

A critical question is to understand the mismatch between  there are alternative data streams that may be leveraged to understand the handling the COVID-19 pandemic in the US.

\subsection{Related work}

There has been an abundance..

However, most complain about testing \emph{capacity} (cite Nate Silver), there is clearly issues of data quality.

\subsection{Outline}

\begin{itemize}
	\item Selection bias + MEM = problems! Scales with population size so cross country comparisons are meaningless.
	\item What about cross-time same spot?  That's equally difficult!
	\item Present a compartmental model that accounts for these time-varying factors.  Discuss identifiability (statistical issue) and how to proceed
	\item Review testing and building selection models (sensitivity analysis)
	\item Robustness (what directions are the models most resilient) and the notion of forecasts in bubbles
	\item Testing, disease surveillance and second-wave testing goals.
\end{itemize}

\section{}

The naive data analyst may think
\begin{center}
\emph{As we increase testing capacity, we will learn more about disease prevalence.}
\end{center}

Case-count is considered a proxy for disease prevalence and contagion spread.

Let $N$ denote the population size.  At a fixed moment in time, let $Y_j \in \{ 0,1\}$ for $j=1,\ldots, N$ denote COVID status.  As in survey sampling methodology, we treat these as fixed, unknown population quantities. For now, we ignore the dynamic nature of the viral outbreak; so either individual $j$ is COVID-19 positive and $Y_j=1$ or is COVID-19 negative and $Y_j=0$. A testing indicator, $I_j \in \{0,1\}$ is an indicator that the individual was selected for testing ($I = 1$) or not ($I=0$).

A primary question is the prevalence o
We are interested in the population average $\bar Y = \frac{1}{N} \sum_{j=1}^N Y_j$. Suppose we observe a sample of $n$ tests.  A natural candidate for prevalence is the proportion of positive tests $\bar y = \frac{1}{n} \sum_{i=1}^n y_i$.

Based on Meng (2019), we can express the error between $\bar y$ and the true proportion as
$$
\bar y_n - \bar Y =  \rho_{I, Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_Y
$$
Under random sampling $E [ \rho_{I,Y} ]$ = 0 and there is no bias.  Under selective testing,

Case-count
Figure~\ref{} presents testing per capita across reporting countries.  We see th at

Two countries with the same testing strategy (i.e., $\rho_{I,Y}$ equal) can yield wildly different estimates due to pop size.

\subsection{Imperfect tests and }

Meng's analysis is critical when selection bias is present.  For the current crisis, the analysis should be extended in two directions.  First, tests are imperfect. Measurement-error.

Let $P_j$ be an indicator that equals 1 when we observe reverse of the true outcome and equals 0 when we observe the true outcome. We suppose this is a stochastic variable that satisfies $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.  If individual $j$ is selected (i.e., $I_j = 1$) then the  $Y_j^\star = Y_j(1-P_j) + (1-Y_j) P_j$.

$$
\bar y_n - \bar Y = \sqrt{\frac{1-f}{f}} \left[ \rho_{I,Y} \times \mu_Y (1-\mu_Y) + \rho_{I,PZ} \times \sigma_{PZ} + \sqrt{\frac{f}{1-f}}  \left( FP - (FP+FN) \bar Y \right) \right] .
$$
where $\sigma_{PZ} = \sqrt{2 FP} \sqrt{\bar Y \left[ 1 - \frac{FP+FN}{FP} \bar Y \right]}$.

This is quite complex but he main point is that the sign of the new terms can be either positive or negative.  Moreover, there is an interaction between...
This means that raw application is, of course, biased but the expected bias directionality is not guaranteed.  There's a clear interaction between selection bias, measurement-error, prevalence, and data quantity!

To see this

The attentive data analyst will recognize the estimator $\bar y_n$ will be biased even for simple random samples and suggest the alternative $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$. We again wish to express the error $\tilde  y_n - \bar Y$ in statistical terms.  Let $Z_j = 1 - 2Y_j$; this converts the binary $0/1$ variable into a sign indicator, equal to $1$ when the outcome is.  In the appendix, we show that the error now can be expressed as
$$
\tilde y_n - \bar Y = (1+FP+FN) \rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y} + \rho_{I,PZ} \times \sqrt{\frac{1-f}{f}} \times \sigma_{PZ}.
$$
The first term is the same as before but increased by $(1 + FP + FN)$ to account for the additional uncertainty due to measurement-error.  The second term is the interaction between selection bias and measurement error.  Interestingly we can show that the sign is reversed, i.e., $sgn(\rho_{I,PZ}) = 1 - sgn(\rho_{I,Y})$, leading to an absolute \emph{reduction} in the error.

For this adjusted estimate, we can see that the directionality is maintained.  Therefore, the benefit is if the expected correlation is positive, we can feel like the estimate is likely an over-estimate and vice versa.  You pay a penalty in terms of variance and therefore the MSE may not be much smaller (CHECK!)

\subsection{Testing strategies and population size: a tango}

Data analysts have argued whether

\subsection{Regrettable rates}

The data analyst, now wary of estimating prevalence and total counts, pauses and thinks.  They retur  n shortly thereafter, a bit agitated but still  \emph{Ok, perhaps total counts is a lost cause. Certainly, however, we can estimate the rate of growth.  All I want to know is when we hit the point at which the curve flattens and number of deaths decrease.  That can't be too hard, surely!}

The parameter $R_0$ in the ubiquitous SIR (susceptible-)

Ratio estimator $\bar y_2/ \bar y_1$.

Taylor series expansion:

$$
\begin{aligned}
\frac{\bar y_2}{\bar y_1} - \frac{\bar Y_2}{\bar Y_1}
&= \frac{\bar y_2}{\bar Y_1  \left(1 + \rho_{I,Y_1} \sqrt{\frac{1-f}{f}} CV (Y_1) \right) } - \frac{\bar Y_2}{\bar Y_1}  \\
&\approx \frac{\bar y_2}{\bar Y_1} \left(1 - \rho_{I,Y_1} \sqrt{\frac{1-f}{f}} CV (Y_1) \right) - \frac{\bar Y_2}{\bar Y_1} \\
&\approx \frac{1}{\bar Y_1} \left[ \rho_{I, Y_2} \sqrt{\frac{1-f}{f}} \sigma_{Y_2} - \rho_{I, Y_1} \sqrt{\frac{1-f}{f}} \sigma_{Y_1}
\left( \frac{E_{J_2} [ I_{J_2} Y_{2,J_2} ]}{E_{J_2} [ I_{J_2} ]} + \frac{E_{J_2} [ I_{J_2} ] E_{J_2} [ Y_{2,J_2} ]}{E_{J_2} [ I_{J_2} ]} - \frac{E_{J_2} [ I_{J_2} ] E_{J_2} [ Y_{2,J_2} ]}{E_{J_2} [ I_{J_2} ]} \right) \right] \\
&\approx \frac{1}{\bar Y_1} \left[ \rho_{I, Y_2} \sqrt{\frac{1-f}{f}} \sigma_{Y_2} - \rho_{I, Y_1} \sqrt{\frac{1-f}{f}} \sigma_{Y_1}
\left( \rho_{I, Y_2} \sqrt{\frac{1-f}{f}} \sigma_{Y_2} + \mu_{Y_2} \right) \right]
\end{aligned}
$$

We talk about wanting to know if we are at the peak.  Let's define that time as when $\bar Y_2 / \bar Y_1 = 0$.  Then $\mu_{Y_1} = \mu_{Y_2} = \mu$ and $\sigma_{Y_1} = \sigma_{Y_2} = \sigma$. If selection at each time is independent and the selection process is stationary, we have bias of
$$
\rho \sqrt{\frac{1-f}{f}} \sigma \sqrt{\frac{1-\mu}{\mu}} \left[ \sqrt{\frac{1-\mu}{\mu}} - \rho \sqrt{\frac{1-f}{f}} \right ]
$$

If we collect $f = 0.01$ and the peak occurs at $\mu = 0.03$ with selection bias of $\rho = 0.05$. $1/\rho^2 \times (\mu/(1-\mu))^4$

Then we can be even further off than the prevalence by a factor of the odds ratio!

Suppose we collected $f=1/2$ and peak occurs at $\mu = 0.10$ and $\rho = 0.005$, then $9-\rho 3 = 9 - 0.005 \times 3 = 8.985$ and

\subsection{Testing increase changes data quality}

Many (cites) hve

\section{Compartmental model}

(A) Heterogeneity in $R_0$
(B) Selection bias
(C)

Above we took a general design perspective.  Here we take a model-assisted perspective.

Susceptible individuals can become exposed to the disease.

Selection bias occurs at 2 phases. Phase 1 is self-selection to seek

Phase 2 is the decision to receive testing.  This is performed by the

The result is we have multiple types of people: (A) Those disease-free who seek treatment due to high risk.

We must then layer on measurement-error.  Tests can

The SIR model with parameters.

\subsection{Stratified sampling}

Suppose we stratify such that weighting is uninmportant.

\emph{Increasing testing leads to lower correlation in}

\subsection{Heterogeneiety  SIR with measurement-error and }

Here we explain a general state-space process model as compared to the measurement-process.  While related, the distinction drives our key results.

Observer controls testing, FP and FN.  Patients

Their behavior may depend on forecasts and therefore the notion of a counterfactual is not well defined.  For example, suppose an highly prominent model....

The transition from susceptible to infected is a complex function.  It is based on personal choices;  the second choice
$B$

\emph{Non-identifiable} and rely on assumptions.  As per Cox's statement.

This makes models \emph{fragile} and rely on strong data-generating assumptions.  For a disease

\section{Decision-making: data versus information.}

Above we point to flaws in using observed data to reason about the

\subsection{Interventions on network effects}

While we think of the intervention as reducing risk COVID-19; in reality the interventio nis

Classical contact pretends the graph is fixed.  Here we assume the intervention is to reduce degree and remove edges.  Contact tracing is indeed the way we can

How does $R_0$ relate to connectivity?  This accounts for the heterogeneity. Epi knows this, but I think this point is salient.

\section{That which does not break us, makes us stronger (but potentially not smarter)}

A slightly altered version of Nietzsche's famous quote has become a mantra for post-pandemic thinking: \emph{That which which does not break us, makes us stronger}.  While potentially true via viral resistance, it is not clear that governments have yet to learn lessons on pandemic response.

During the current wave, understanding prevalence is key.  It helps us determine long-term risk for a community and make targeted interventions.  This will allow
When prevalence is below a certain threshold, we can return to daily life.

Once we ``return to normal''\footnote{or at least the new normal}, it is clear that testing strategies should focus on early detection, heading off future outbreaks.
\emph{Anyone who wishes to go back to work}.  While laudable, without complete compliance, we may be riddles with selection bias and measurement-error.  This ignores even the practical and ethical quandaries of how to .

What do we do when we are faced with?  We design experiments with our objectives in mind.

(A) Prevalence: Targeted shutdowns
(B) Risk Detection: Contact tracing and altering

Use prior results to show that ratio estimator under SRS is a good predictor of potential outbreak.

``Disease free''

\section{Conclusion}

Interventions in this area are designed experiments.  Unlike Fischer's null, here the objectives are settings, statisticians

\appendix

\section{Setup and calculation}

I focused this note on the very simply setting of understanding the error when both selection bias and measurement-error are present.  I worked within Meng's derivation as it provides a guide for how to think this error in our current context.  The conclusion so far is that the mathematics will be quite cumbersome, but my hope is that we can potentially draw out some interesting conclusions.  In particular, my hope is to provide theoretical confirmation that the interaction of selection bias and measurement-error does not lead to ``attenuation'' of the effect as in a simple measurement-error context. This is a very very early draft but I wanted to share current ``state-of-affairs'' to see whether you thought further investigation along this line of thinking was merited.

\begin{itemize}
\item Finite-population of size $N$
\item Binary outcome $Y_j \in \{0,1\}$ for $j=1,\ldots, N$.  As in survey methodology, we treat these as fixed, unknown population quantities. For now, we ignore the dynamic nature of the viral outbreak. So you are either COVID positive ($Y=1$) or negative ($Y=0$)
\item Selection indicator $I_j \in \{0,1\}$.  This is an indicator that the individual was selected for testing ($I = 1$) or not ($I=0$).
\item Measurement-error: Let $P_j$ be an indicator that equals 1 when we observe reverse of the true outcome and equals 0 when we observe the true outcome. We suppose this is a stochastic variable that satisfies $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the probability of a false-negative and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the probability of a false-positive.
\item If we select individual $j$ (i.e., $I_j = 1$) then we observe $Y_j^\star = Y_j(1-P_j) + (1-Y_j) P_j$
\item  We are interested in the population average $\bar Y = \frac{1}{N} \sum_{j=1}^N Y_j$
\item Consider the mean estimator
$$
\bar y_n = \frac{1}{n} \sum_{j=1}^n Y_j^\star = \frac{\sum_{i=1}^N  I_j Y_j^\star }{\sum_{j=1}^N  I_j } = \frac{\sum_{i=1}^N  I_j \left[ Y_j (1-P_j) + (1-Y_j) P_j \right]}{\sum_{j=1}^N  I_j }
$$
\item We wish to express the error $\bar y_n - \bar Y$ in statistical terms
\item For any set of numbers $\{ A_1, \ldots, A_N \}$ we can view it as the support of a random variable $A_J$ induced by the random index $J$ defined on $\{1,\ldots, N\}$.  When $J$ is uniformly distributed $E_J (A_J) = \sum_{j=1}^N A_j / N \equiv \bar A_N$.
\item Then
$$
\begin{aligned}
\bar y_n  - \bar Y_N &= \frac{E_J \left[ I_J \left[ Y_J (1-P_J) + (1-Y_J) P_J \right] \right]}{E_J [ I_J ] } - E_J[Y_J] \\
&= \frac{E_J \left[ I_J P_J (1-2Y_J) \right]}{E_J [ I_J ] } + \left( \frac{E_J [I_J Y_J]}{E_J [ I_J ] } - \frac{E_J[Y_J] E_J[I_J]}{E_J[I_J]} \right) \\
\end{aligned}
$$
\item The term in parentheses can be re-written as
$$
\begin{aligned}
\frac{E_J [I_J Y_J]- E_J[Y_J] E_J[I_J]}{E_J[I_J]} &=  \frac{E_J [I_J Y_J]- E_J[Y_J] E_J[I_J]}{\sqrt{V_J(I_J) V_J(Y_J)}} \frac{\sqrt{V_J(I_J)}}{E_J[I_J]} \times \sqrt{V_J(Y_J)} \\
&= \rho_{I,Y} \times \sqrt{\frac{(1-f)}{f}} \times \sigma_Y
\end{aligned}
$$
which agrees with the decomposition by Meng (2019).
\item For the other term, first define $Z_j = 1 - 2 Y_j $. Then $Z_j = 1$ if $Y_j = 0$ and $Z_j = -1$ if $Y_j = 1$. Then the other term can be re-written as
$$
\begin{aligned}
\frac{E_J \left[ I_J P_J (1-2Y_J) \right]}{E_J [ I_J ] } &= \left( \frac{E_J \left[ I_J P_J Z_J \right]}{E_J [ I_J ] } -  \frac{E_J \left[ P_J Z_J \right] E_J[ I_J]}{E_J [ I_J ] } \right) +  \frac{E_J \left[ P_J Z_J \right] E_J[ I_J]}{E_J [ I_J ] } \\
\end{aligned}
$$

\item The term in parentheses can be re-expressed using Meng's techniques as:
$$
\rho_{I, PZ} \times \sqrt{\frac{1-f}{f}} \times \sigma_{PZ}
$$
where now the ``data defect'' and ``problem difficulty'' are with respect to $PZ$ rather than $Y$.
\item The final term is equal to
$$
\begin{aligned}
E_J [P_J Z_J ] &= E_J [ E_J [ P_J Z_J \mid Y_J ] ] \\
&= \pr (P = 1 \mid Y = 0) (1-\bar Y) - \pr(P=1 \mid Y = 1) \bar Y \\
&= FP - (FP + FN) \cdot \bar Y
\end{aligned}
$$
\item Combining these yields:
$$
\bar y_n - \bar Y = \sqrt{\frac{1-f}{f}} \left[\rho_{I,Y} \sigma_Y + \rho_{I, PZ} \sigma_{PZ} + \sqrt{\frac{f}{1-f}} \left( FP - (FP+FN) \mu_Y \right) \right]
$$
\item We know for binary outcome $Y$ that $\sigma_Y = \sqrt{\mu_Y (1-\mu_Y)}$.
\item We can try and understand $\sigma_{PZ}$:
$$
\begin{aligned}
V_J(P_J Z_J) &= E_J[(P_J Z_J)^2] - E[P] E[Z] \\
&= E[P] - E[P] (1 - 2 \mu_Y) = 2 \mu_Y E_J [ P_J ] \\
&= 2 \mu_Y \left( FP (1-\mu_Y) + FN \mu_Y \right) \\
\Rightarrow \sigma_{PZ} &= \sqrt{ 2 \mu_Y \left( FP (1-\mu_Y) + FN \mu_Y \right) }
\end{aligned}
$$
\item Then the formula for the error is given by:
$$
\sqrt{\frac{1-f}{f}} \left[\rho_{I,Y} \sqrt{\mu_Y (1-\mu_Y)} + \rho_{I, PZ} \sqrt{ 2 \mu_Y \left( FP (1-\mu_Y) + FN \mu_Y \right )} + \sqrt{\frac{f}{1-f}} \left( FP - (FP+FN) \mu_Y \right) \right]
$$
\item Let's consider a simple case first and assume no false positive results, i.e., set $FP=0$.  Then $\sigma_{PZ} = \mu_Y \sqrt{2 FN}$.  Then the math greatly simplifies:
$$
\bar y_n - \bar Y = \mu_Y \sqrt{\frac{1-f}{f}} \left[\rho_{I,Y} (1-\mu_Y) + \sqrt{FN} \left( \sqrt{2} \rho_{I, PZ} - \sqrt{\frac{f}{1-f}} \sqrt{FN} \right)  \right]
$$
\item We see an interesting trade-off in the second term.  The term in parentheses is positive iff
$$
\rho^2_{I,PZ}  \geq \frac{f}{1-f} \frac{FN}{2}
$$
\item Thus the sign of the additional error depends on selection bias ($\rho_{I, PZ}$), false negative rates (FN), and sample fraction $f$.  Moreover, the scale of the additional term depends on the false negative rate and the infection rate $\mu_Y = \bar Y$.
\item Let's consider the other simple case and assume no false negative results, i.e., set $FN=0$.  Then $\sigma_{PZ} = \sqrt{2 FP \mu_Y (1-\mu_Y)}$.  Then the math greatly simplifies:
$$
\bar y_n - \bar Y = \sqrt{1-\mu_Y} \sqrt{\frac{1-f}{f}} \left[\rho_{I,Y} \mu_Y \sqrt{1-\mu_Y} +  \sqrt{FP} \left( \sqrt{2 \mu_Y} \rho_{I, PZ} - \sqrt{\frac{f}{1-f}} \sqrt{FP} \sqrt{1-\mu_Y} \right)  \right]
$$
\item We see an interesting trade-off in the second term.  The term in parentheses is positive iff
$$
\rho^2_{I,PZ}  \geq \frac{f}{1-f} \frac{FP}{2} \times \underbrace{\frac{1-\mu_Y}{\mu_Y}}_{\text{Inverse Odds Ratio}}
$$
\item Thus the sign of the additional error depends on selection bias ($\rho_{I, PZ}$), false positive rates (FP), sample fraction $f$, AND the odds ratio.  Moreover, the scale of the additional term depends on the false positive rate and one minus the infection rate $1-\mu_Y = 1-\bar Y$.

\item Let's compute the MSE

\item If we adjust $\bar y_n$ by taking $\bar y_n + (FP+FN)\bar y_n - FP$; then
$$
\sqrt{\frac{1-f}{f}} \left[ \rho_{I, Y} \sigma_{Y} (1+FP+FN)  + \rho_{I, PZ} \sqrt{ 2 \mu_y \left( FP (1-\mu_y) + FN \mu_Y \right)} \right]
$$
\item In FN = 0; we have
$$
\sqrt{\frac{1-f}{f}} \sqrt{\mu_Y(1-\mu_Y)} \left[ \rho_{I, Y} (1+FP) + \rho_{I, PZ} \sqrt{ 2 FP } \right]
$$
\item So the false positives actually offset some of the positive correlation.
\end{itemize}

$$
\rho_{I,PZ} = \frac{C(I, PZ)}{\sqrt{V()}}
$$

\section{Quotes}

I prefer to think of a statistical sensibility rather than statistical thinking. It’s “less than an agenda but more than an attitude.”  It allows for methodological preference while avoiding dogma. Paired with data analytic humility and I think you have proper “data science”

“Routine statistical questions are less common than questionable statistical routines...” McCullagh (2005).

If an issue can be addressed nonparametrically then it will often be better to tackle it parametrically; however, if it cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically.” (p.96)

A test of meaningfulness of a possible model for a data-generating process is whether it can be used directly to simulate data.” (p.104).  In our current setting, this most certainly related to simulation while accounting for measruement error.

\end{itemize}

\end{document}

### 2020-04-17 write-up/covid-writeup.tex
covid update

@@
 
+Prevalence is important in the long-run because the
+
 Based on Meng (2019), we can express the error between $\bar y$ and the true proportion as
@@
 
+\emph{Increasing testing leads to lower correlation only if we see the gap $\Delta$ decrease.  If we just see $f$ }
+
 \section{Compartmental model}
@@
 
-\subsection{Stratified sampling}
+\subsection{Ration-based testing allocation}
 
-Suppose we stratify such that weighting is uninmportant.
+We assume that the percentage of tests is rationed according to severity of symptoms.
+
+Note that a person who displays severe symptoms may be COVID negative.
 

### 2020-04-20 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 
-\subsection{Imperfect tests and }
+\subsection{Imperfect testing}
 
-Meng's analysis is critical when selection bias is present.  For the current crisis, the analysis should be extended in two directions.  First, tests are imperfect. Measurement-error.
+Tests are imperfect.
+COVID-19 testing is no exception.
+State-level case-count data is often used to signify disease prevalence per state.
+When assessing prevalence in a community, sensitivity (true positive rate) and specificity (true negative rate) must be taken into account.
+Here we investigate the interplay between imperfect testing and selection bias.
+The standard story of ``measurement-error leads to parameter attenuation'' does not pan out.
+Instead, the two sources of error become entangled, and the biases can become magnified, muted, or even switch signs.
 
-Let $P_j$ be an indicator that equals 1 when we observe reverse of the true outcome and equals 0 when we observe the true outcome.
-We suppose this is a stochastic variable that satisfies $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.
-If individual $j$ is selected (i.e., $I_j = 1$) then the  $Y_j^\star = Y_j(1-P_j) + (1-Y_j) P_j$.
+First we require some additional notation.
+Let $P_j$ be an indicator of measurement error, equal to $1$ when we mismeasure the outcome and $0$ when we observe the true outcome.
+We suppose this is a stochastic variable that satisfies $\pr(P_j = 1 \mid Y_j = 1, I_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0, I_j = 1) =: FP$ is the false-positive rate.
+If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^\star = Y_j(1-P_j) + (1-Y_j) P_j$.
 
@@
 There's a clear interaction between selection bias, measurement-error, prevalence, and data quantity!
+We are not the first to note this, but  (twitter guy) for a longer discussion of additional myths related
 
-To see this
+ critical when selection bias is present.  For the current crisis, the analysis should be extended in two directions.  First, tests are imperfect. Measurement-error.
+
+In addition, testing quality is likely to be a dynamic process.
+The US is keen to increase testing.
+Commentators have requested expedited FDA approval for variations on the currently available tests to speed up production.
+This is likely to increase testing inaccuracies.
+The bias and MSE depend on interaction among specificity,
 
@@
 
+\begin{itemize}
+\item Simulation study showing that $R_t$ and prevalence over time changes the effective sample size....
+\end{itemize}
+
 \subsection{Ration-based testing allocation}
@@
 This makes models \emph{fragile} and rely on strong data-generating assumptions.  For a disease
+
+\subsection{Stratified sampling: improving precision in low-prevalence environments}
+
+Sampling network is hard.  Sampling high-risk strata is easy.  Simple example.
+DTR and network connectivity is useful for understanding spread and where "testing should go".
 

### 2020-04-21 write-up/covid-writeup.tex
updated for fixed math

@@
 \usepackage{graphicx}
+\usepackage{wrapfig}
 \usepackage{setspace}
@@
 \def\pr{\text{pr}}
+\def\sgn{\text{sgn}}
 
@@
 
+For the current crisis, the analysis should be extended in two directions.  First, imperfect testing . Measurement-error.
+
 \subsection{Imperfect testing}
@@
 COVID-19 testing is no exception.
-State-level case-count data is often used to signify disease prevalence per state.
-When assessing prevalence in a community, sensitivity (true positive rate) and specificity (true negative rate) must be taken into account.
+Data analysis must account for test inaccuracy to distinguish signal and noise.
 Here we investigate the interplay between imperfect testing and selection bias.
-The standard story of ``measurement-error leads to parameter attenuation'' does not pan out.
-Instead, the two sources of error become entangled, and the biases can become magnified, muted, or even switch signs.
+When discussing testing inaccuracies, the standard assumption is measurement-error leads to parameter attenuation.
+When paired with selection bias, however, the two sources of error become entangled, and the biases can become magnified, muted, or even switch signs.
 
@@
 If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^\star = Y_j(1-P_j) + (1-Y_j) P_j$.
+Suppose disease prevalence was estimated as the fraction who tested positive for COVID-19, i.e., $\bar y_n^\star = \frac{\sum_{i=1}^N I_j Y_j^\star}{\sum_{i=1}^N I_j}$.
+We can again investigate the error compared to the true prevalence $\bar Y$ in statistical terms:
+$$
+\bar y_n^\star - \bar Y = \sqrt{\frac{1-f}{f}} \left[ \rho_{I,Y} \times \sigma_Y + \rho_{I,PZ} \times \sigma_{PZ} + \sqrt{\frac{f}{1-f}}  \left( FP - (FP+FN) \bar Y \right) \right] .
+$$
+where $Z = 1-2Y$.
+The first term represents the perfect testing regime, the second term represents the interaction between imperfect testing and selection bias, while the third time represents the bias due to imperfect testing.
+From here on, we refer to $\rho_{I,Y}$ as the \emph{true data quality}, and $\rho_{I,PZ}$ as the \emph{observed data quality} that accounts for both selection bias and measurement-error.
+We show the sign of $\rho_{I,PZ}$ is the opposite of the sign of $\rho_{I,Y}$, and so the second term adjusts the true data quality.
 
+We start by considering the first two terms and assess whether the sign of the bias can reverse due to the interaction of measurement-error and selection bias.
+To do this, we require the sampling rates differential.
+Let $f_1 := P_J(I_J = 1 \mid Y_J = 1)$ and $f_0 := P_J(I_J = 1 \mid Y_J = 1)$ be the sampling rates.
+Then if $\Delta = f_1 - f_0$ is the sampling rate differential, we have
 $$
-\bar y_n - \bar Y = \sqrt{\frac{1-f}{f}} \left[ \rho_{I,Y} \times \mu_Y (1-\mu_Y) + \rho_{I,PZ} \times \sigma_{PZ} + \sqrt{\frac{f}{1-f}}  \left( FP - (FP+FN) \bar Y \right) \right] .
+\rho_{I,Y} \times \sigma_Y + \rho_{I,PZ} \times \sigma_{PZ} =
+\rho_{I,Y} \times \sigma_Y \left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right].
 $$
-where $\sigma_{PZ} = \sqrt{2 FP} \sqrt{\bar Y \left[ 1 - \frac{FP+FN}{FP} \bar Y \right]}$.
+The final term in brackets is the \emph{measurement-error adjustment to data quality} which is a complex function of sampling rate differential, the odds ratio, and the ratio of measurement-error interaction with prevalence and sampling rates interaction with prevalence.
+Note $\sgn(\Delta) = \sgn(\rho_{I,Y})$ so the measurement-error adjustment either shrinks the data quality measure toward zero or reverses its sign.
 
-This is quite complex but he main point is that the sign of the new terms can be either positive or negative.  Moreover, there is an interaction between...
-This means that raw application is, of course, biased but the expected bias directionality is not guaranteed.
-There's a clear interaction between selection bias, measurement-error, prevalence, and data quantity!
-We are not the first to note this, but  (twitter guy) for a longer discussion of additional myths related
+While prior investigations have notes the interaction between measurement-error and selection bias (cites), the interaction with the sample size relative to the population,i.e., $f$, has largely been ignored.
+The above statistical decomposition clarifies the importance of this quantity~$f$.
+In particular, note that the statistical error also includes a bias term due to measurement-error and this term increases as the sampled fraction $f$ increases.
+Therefore, how the first two terms interact with the final term depends on the fraction of the population sampled.
+This interaction is complex, but implies that whether the estimate $\bar y^\star_n$ is an overestimate or underestimate is a complicated question due to the relation amongst these three pieces.
 
- critical when selection bias is present.  For the current crisis, the analysis should be extended in two directions.  First, tests are imperfect. Measurement-error.
-
-In addition, testing quality is likely to be a dynamic process.
-The US is keen to increase testing.
-Commentators have requested expedited FDA approval for variations on the currently available tests to speed up production.
-This is likely to increase testing inaccuracies.
-The bias and MSE depend on interaction among specificity,
-
-The attentive data analyst will recognize the estimator $\bar y_n$ will be biased even for simple random samples and suggest the alternative $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$.
+The attentive data analyst will recognize the estimator $\bar y_n$ is biased even for simple random samples and, if sensitivity and specificity were known a priori, may suggest the alternative estimator $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ which is unbiased under simple random sampling.
 We again wish to express the error $\tilde  y_n - \bar Y$ in statistical terms.
-Let $Z_j = 1 - 2Y_j$; this converts the binary $0/1$ variable into a sign indicator, equal to $1$ when the outcome is.
 In the appendix, we show that the error now can be expressed as
 $$
-\tilde y_n - \bar Y = (1+FP+FN) \rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y} + \rho_{I,PZ} \times \sqrt{\frac{1-f}{f}} \times \sigma_{PZ}.
+\tilde y_n - \bar Y = \rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
+\left[ 1 + FP + FN - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right].
 $$
+
+Above we ignore an important consideration, namely that testing quality is likely a dynamic process.
+In the US, states are rightly trying to increase testing capacity.
+Commentators have requested expedited FDA approval for variations on the currently available tests to speed up production.
+This is likely to increase testing inaccuracies, impacting both false negative and false positive rates.
+Increased capacity may reduce
+
+\newpage
+
+\begin{wrapfigure}{r}{0.5\textwidth}
+\centering
+\includegraphics[width = 0.4\textwidth]{../methods/figs/mem_heatmap_article.png}
+\caption{Measurement-error data quality adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.15$ and $FN=0.10$.
+Color scaled so blue = $-6$, white = 0, and red = $6$.}
+\end{wrapfigure}
+
+For fixed prevalence,
+
 The first term is the same as before but increased by $(1 + FP + FN)$ to account for the additional uncertainty due to measurement-error.
@@
 
+{\bf How many days of dropping to guarantee it has decreased}
+
 \subsection{Testing increase changes data quality}
@@
 \emph{Increasing testing leads to lower correlation only if we see the gap $\Delta$ decrease.  If we just see $f$ }
+
+\subsection{Clinical trials}
+
+Random sampling negates effect modifiers.
+Random treatment assignment negates unobserved confounders.
+Show simple example where $E[U]$ is high due to sampling bias and therefore you will think that the effect is significant!
+But it's because of sampling bias.
+Requires poststratification.
 
@@
 
-\section{Setup and calculation}
+\section{Imperfect testing: derivation}
 
-I focused this note on the very simply setting of understanding the error when both selection bias and measurement-error are present.
-I worked within Meng's derivation as it provides a guide for how to think this error in our current context.
-The conclusion so far is that the mathematics will be quite cumbersome, but my hope is that we can potentially draw out some interesting conclusions.
-In particular, my hope is to provide theoretical confirmation that the interaction of selection bias and measurement-error does not lead to ``attenuation'' of the effect as in a simple measurement-error context.
-This is a very very early draft but I wanted to share current ``state-of-affairs'' to see whether you thought further investigation along this line of thinking was merited.
-
-\begin{itemize}
-\item Finite-population of size $N$
-\item Binary outcome $Y_j \in \{0,1\}$ for $j=1,\ldots, N$.
-As in survey methodology, we treat these as fixed, unknown population quantities.
-For now, we ignore the dynamic nature of the viral outbreak.
-So you are either COVID positive ($Y=1$) or negative ($Y=0$)
-\item Selection indicator $I_j \in \{0,1\}$.  This is an indicator that the individual was selected for testing ($I = 1$) or not ($I=0$).
-\item Measurement-error: Let $P_j$ be an indicator that equals 1 when we observe reverse of the true outcome and equals 0 when we observe the true outcome.
-We suppose this is a stochastic variable that satisfies $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the probability of a false-negative and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the probability of a false-positive.
-\item If we select individual $j$ (i.e., $I_j = 1$) then we observe $Y_j^\star = Y_j(1-P_j) + (1-Y_j) P_j$
-\item  We are interested in the population average $\bar Y = \frac{1}{N} \sum_{j=1}^N Y_j$
-\item Consider the mean estimator
+We considered the mean estimator
 $$
-\bar y_n = \frac{1}{n} \sum_{j=1}^n Y_j^\star = \frac{\sum_{i=1}^N  I_j Y_j^\star }{\sum_{j=1}^N  I_j } = \frac{\sum_{i=1}^N  I_j \left[ Y_j (1-P_j) + (1-Y_j) P_j \right]}{\sum_{j=1}^N  I_j }
+\bar y_n^\star = \frac{\sum_{j=1}^N Y_j^\star I_j}{\sum_{j=1}^N I_j} = \frac{\sum_{i=1}^N  I_j Y_j^\star }{\sum_{j=1}^N  I_j } = \frac{\sum_{i=1}^N  I_j \left[ Y_j (1-P_j) + (1-Y_j) P_j \right]}{\sum_{j=1}^N  I_j }
 $$
-\item We wish to express the error $\bar y_n - \bar Y$ in statistical terms
-\item For any set of numbers $\{ A_1, \ldots, A_N \}$ we can view it as the support of a random variable $A_J$ induced by the random index $J$ defined on $\{1,\ldots, N\}$.
+For any set of numbers $\{ A_1, \ldots, A_N \}$ we can view it as the support of a random variable $A_J$ induced by the random index $J$ defined on $\{1,\ldots, N\}$.
 When $J$ is uniformly distributed $E_J (A_J) = \sum_{j=1}^N A_j / N \equiv \bar A_N$.
-\item Then
+Then
 $$
 \begin{aligned}
-\bar y_n  - \bar Y_N &= \frac{E_J \left[ I_J \left[ Y_J (1-P_J) + (1-Y_J) P_J \right] \right]}{E_J [ I_J ] } - E_J[Y_J] \\
+\bar y_n^\star  - \bar Y_N &= \frac{E_J \left[ I_J \left[ Y_J (1-P_J) + (1-Y_J) P_J \right] \right]}{E_J [ I_J ] } - E_J[Y_J] \\
 &= \frac{E_J \left[ I_J P_J (1-2Y_J) \right]}{E_J [ I_J ] } + \left( \frac{E_J [I_J Y_J]}{E_J [ I_J ] } - \frac{E_J[Y_J] E_J[I_J]}{E_J[I_J]} \right) \\
@@
 $$
-\item The term in parentheses can be re-written as
+The term in parentheses can be re-written as
 $$
@@
 $$
-which agrees with the decomposition by Meng (2019).
-\item For the other term, first define $Z_j = 1 - 2 Y_j $. Then $Z_j = 1$ if $Y_j = 0$ and $Z_j = -1$ if $Y_j = 1$. Then the other term can be re-written as
+which agrees with Meng's (2019) decomposition. For the other term, first we define $Z_j := 1 - 2 Y_j $. Then $Z_j = 1$ if $Y_j = 0$ and $Z_j = -1$ if $Y_j = 1$. Then the term can be re-written as
 $$
@@
 $$
-
-\item The term in parentheses can be re-expressed using Meng's techniques as:
+The term in parentheses can be re-expressed using the previous technique as:
 $$
@@
 $$
-where now the ``data defect'' and ``problem difficulty'' are with respect to $PZ$ rather than $Y$.
-\item The final term is equal to
+where now the ``data defect'' and ``problem difficulty'' are with respect to $PZ$ rather than $Y$. The final term is equal to
 $$
@@
 $$
-\item Combining these yields:
+Combining these yields:
 $$
-\bar y_n - \bar Y = \sqrt{\frac{1-f}{f}} \left[\rho_{I,Y} \sigma_Y + \rho_{I, PZ} \sigma_{PZ} + \sqrt{\frac{f}{1-f}} \left( FP - (FP+FN) \mu_Y \right) \right]
+\bar y_n^\star - \bar Y = \sqrt{\frac{1-f}{f}} \left[\rho_{I,Y} \sigma_Y + \rho_{I, PZ} \sigma_{PZ} + \sqrt{\frac{f}{1-f}} \left( FP - (FP+FN) \bar Y \right) \right]
 $$
-\item We know for binary outcome $Y$ that $\sigma_Y = \sqrt{\mu_Y (1-\mu_Y)}$.
-\item We can try and understand $\sigma_{PZ}$:
+For the binary outcome $Y$, we have $\sigma_Y = \sqrt{\bar Y (1-\bar Y)}$. Moreover,
 $$
@@
 V_J(P_J Z_J) &= E_J[(P_J Z_J)^2] - E[P] E[Z] \\
-&= E[P] - E[P] (1 - 2 \mu_Y) = 2 \mu_Y E_J [ P_J ] \\
-&= 2 \mu_Y \left( FP (1-\mu_Y) + FN \mu_Y \right) \\
-\Rightarrow \sigma_{PZ} &= \sqrt{ 2 \mu_Y \left( FP (1-\mu_Y) + FN \mu_Y \right) }
+&= E[P] - E[P] (1 - 2 \bar Y) = 2 \bar Y E_J [ P_J ] \\
+&= 2 \bar Y \left( FP (1-\bar Y) + FN \bar Y \right) \\
+\Rightarrow \sigma_{PZ} &= \sqrt{ 2 \bar Y \left( FP (1-\bar Y) + FN \cdot  \bar Y \right) }
 \end{aligned}
 $$
-\item Then the formula for the error is given by:
+Then the formula for the error is given by:
 $$
-\sqrt{\frac{1-f}{f}} \left[\rho_{I,Y} \sqrt{\mu_Y (1-\mu_Y)} + \rho_{I, PZ} \sqrt{ 2 \mu_Y \left( FP (1-\mu_Y) + FN \mu_Y \right )} + \sqrt{\frac{f}{1-f}} \left( FP - (FP+FN) \mu_Y \right) \right]
+\sqrt{\frac{1-f}{f}} \left[\rho_{I,Y} \sqrt{\bar Y (1-\bar Y)} + \rho_{I, PZ} \sqrt{ 2 \bar Y \left( FP (1-\bar Y) + FN \cdot \bar Y \right )} + \sqrt{\frac{f}{1-f}} \left( FP - (FP+FN) \bar Y \right) \right]
 $$
-\item Let's consider a simple case first and assume no false positive results, i.e., set $FP=0$.  Then $\sigma_{PZ} = \mu_Y \sqrt{2 FN}$.  Then the math greatly simplifies:
+
+\subsection{Simple cases}
+
+We consider two simple cases here.  First, we assume no false positive results, i.e., set $FP=0$.  Then $\sigma_{PZ} = \bar Y \sqrt{2 FN}$.  Then the math simplifies:
 $$
-\bar y_n - \bar Y = \mu_Y \sqrt{\frac{1-f}{f}} \left[\rho_{I,Y} (1-\mu_Y) + \sqrt{FN} \left( \sqrt{2} \rho_{I, PZ} - \sqrt{\frac{f}{1-f}} \sqrt{FN} \right)  \right]
+\bar y_n^\star - \bar Y = \bar Y \sqrt{\frac{1-f}{f}} \left[\rho_{I,Y} \sqrt{\frac{1-\bar Y}{\bar Y}} + \sqrt{FN} \left( \sqrt{2} \rho_{I, PZ} - \sqrt{\frac{f}{1-f}} \sqrt{FN} \right)  \right]
 $$
-\item We see an interesting trade-off in the second term.  The term in parentheses is positive iff
+The sign of the error therefore depends on true data quality ($\rho_{I, Y}$), odds ratio, observed data quality~$\rho_{I,PZ}$, false negative rates (FN), and sample fraction $f$.
+
+Second, we assume no false negative results, i.e., set $FN=0$.  Then $\sigma_{PZ} = \sqrt{2 FP \bar Y (1-\bar Y)}$.  Then the math simplifies:
 $$
-\rho^2_{I,PZ}  \geq \frac{f}{1-f} \frac{FN}{2}
+\bar y_n^\star - \bar Y = \sqrt{\bar Y (1-\bar Y)} \sqrt{\frac{1-f}{f}} \left[\rho_{I,Y} +  \sqrt{FP} \left( \sqrt{2} \rho_{I, PZ} + \sqrt{\frac{f}{1-f}} \sqrt{FP} \sqrt{1-\bar Y} \right)  \right]
 $$
-\item Thus the sign of the additional error depends on selection bias ($\rho_{I, PZ}$), false negative rates (FN), and sample fraction $f$.
-Moreover, the scale of the additional term depends on the false negative rate and the infection rate $\mu_Y = \bar Y$.
-\item Let's consider the other simple case and assume no false negative results, i.e., set $FN=0$.  Then $\sigma_{PZ} = \sqrt{2 FP \mu_Y (1-\mu_Y)}$.  Then the math greatly simplifies:
+The sign of the error therefore depends on true data quality ($\rho_{I, Y}$),  observed data quality~$\rho_{I,PZ}$, false positive rates (FP), one minus prevalence ($1-\bar Y$), and sample fraction $f$.
+
+\subsection{Adjusted estimator}
+
+Define $\bar y_n^{\star \star} = \bar y_n^\star + FN \bar y_n^\star - FP(1-\bar y_n^\star)$; then the error is
 $$
-\bar y_n - \bar Y = \sqrt{1-\mu_Y} \sqrt{\frac{1-f}{f}} \left[\rho_{I,Y} \mu_Y \sqrt{1-\mu_Y} +  \sqrt{FP} \left( \sqrt{2 \mu_Y} \rho_{I, PZ} - \sqrt{\frac{f}{1-f}} \sqrt{FP} \sqrt{1-\mu_Y} \right)  \right]
+\bar y_n^{\star \star} - \bar Y = \sqrt{\frac{1-f}{f}} \left[ \rho_{I, Y} \sigma_{Y} (1+FP+FN)  + \rho_{I, PZ} \sqrt{ 2 \bar Y \left( FP (1-\bar Y) + FN \bar Y \right)} \right]
 $$
-\item We see an interesting trade-off in the second term.  The term in parentheses is positive iff
+In the case when false negative rate is zero, we have
 $$
-\rho^2_{I,PZ}  \geq \frac{f}{1-f} \frac{FP}{2} \times \underbrace{\frac{1-\mu_Y}{\mu_Y}}_{\text{Inverse Odds Ratio}}
+\bar y_n^{\star \star} - \bar Y = \sqrt{\frac{1-f}{f}} \sqrt{\bar Y(1-\bar Y)} \left[ \rho_{I, Y} (1+FP) + \rho_{I, PZ} \sqrt{ 2 FP } \right]
 $$
-\item Thus the sign of the additional error depends on selection bias ($\rho_{I, PZ}$), false positive rates (FP), sample fraction $f$, AND the odds ratio.
-Moreover, the scale of the additional term depends on the false positive rate and one minus the infection rate $1-\mu_Y = 1-\bar Y$.
+In the case when false positive rate is zero, we have
+$$
+\bar y_n^{\star \star} - \bar Y = \sqrt{\frac{1-f}{f}} \sqrt{\bar Y(1-\bar Y)} \left[ \rho_{I, Y} (1+FN) + \rho_{I, PZ} \sqrt{ 2 FN } \times \sqrt{\frac{\bar Y}{1-\bar Y}} \right]
+$$
 
-\item Let's compute the MSE
-
-\item If we adjust $\bar y_n$ by taking $\bar y_n + (FP+FN)\bar y_n - FP$; then
-$$
-\sqrt{\frac{1-f}{f}} \left[ \rho_{I, Y} \sigma_{Y} (1+FP+FN)  + \rho_{I, PZ} \sqrt{ 2 \mu_y \left( FP (1-\mu_y) + FN \mu_Y \right)} \right]
-$$
-\item In FN = 0; we have
-$$
-\sqrt{\frac{1-f}{f}} \sqrt{\mu_Y(1-\mu_Y)} \left[ \rho_{I, Y} (1+FP) + \rho_{I, PZ} \sqrt{ 2 FP } \right]
-$$
-\item So the false positives actually offset some of the positive correlation.
-\end{itemize}
+\subsection{Estimate of observed data quality}
 
 $$
-\rho_{I,PZ} = \frac{C(I, PZ)}{\sqrt{V()}}
+\begin{aligned}
+\rho_{I,PZ} &= \frac{C(I, PZ)}{\sqrt{V(PZ) V(I)}} \\
+&= \frac{C(I, PZ)}{\sqrt{V(Y) V(I)}} \sqrt{\frac{V(Y)}{V(PZ)}} \\
+&= \rho_{I,Y} \frac{C(I,PZ)}{C(I,Y)} \sqrt{ \frac{(1-\bar Y)}{2 ( FP (1-\bar Y) + FN \cdot \bar Y)} }
+\end{aligned}
+$$
+
+$$
+\begin{aligned}
+C(I, PZ) &= E[ I P Z ] - E[I] E[PZ] \\
+&=  [FP f_0 - (FP f_0 + FN f_1) \bar Y] - f [ FP - (FP+FN) \bar Y ] \\
+&=  - FP \Delta \bar Y + FP \bar Y^2 \Delta - FN \bar Y^2 \Delta \\
+&=  - \Delta \bar Y (FP \cdot (1-\bar Y) + FN \cdot \bar Y) \\
+\end{aligned}
+$$
+where $f = f_1 \bar Y + f_0 (1-\bar Y)$ so $f_0 - f = -\Delta \bar Y$ and $f_1 - f = \Delta (1-\bar Y)$.
+$$
+\begin{aligned}
+C(I, Y) &= E[ I Y ] - f \bar Y \\
+&=  f_1 \bar Y + f_0 (1-\bar Y) - f \bar Y \\
+&=  f_0 (1-\bar Y) + \Delta (1-\bar Y) \bar Y \\
+&= (1-\bar Y) (f_0 + \Delta \bar Y)
+\end{aligned}
+$$
+Combining yields
+$$
+\begin{aligned}
+\rho_{I,PZ} &= \rho_{I,Y} \times \frac{- \Delta \bar Y (FP \cdot (1-\bar Y) + FN \cdot \bar Y) }{(1-\bar Y) (f_0 + \Delta \bar Y)} \times \sqrt{ \frac{(1-\bar Y)}{2 ( FP (1-\bar Y) + FN \cdot \bar Y)} } \\
+&= - \rho_{I, Y} \times \Delta \times \sqrt{\frac{\bar Y}{1-\bar Y}} \frac{\sqrt{FP(1-\bar Y) + FN \cdot \bar Y}}{f_0 (1-\bar Y) + f_1 \bar Y} \times \sqrt{\frac{\bar Y}{2}}
+\end{aligned}
+$$
+We can then re-write $\rho_{I,Y} \sigma_Y + \rho_{I,PZ} \sigma_{PZ}$ as
+$$
+\rho_{I,Y} \sigma_Y \left( 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right)
+$$
+Suppose $f_1 = M \cdot f_0$.  Then $\Delta = f_1 - f_0 = f_0 (M-1)$ and $f_0 (1-\bar Y) + f_1 \bar Y = f_0 ( (1-\bar Y) + M \bar Y )$ and we can re-write above as
+$$
+\rho_{I,Y} \sigma_Y \left( 1 - (M-1) \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{(1-\bar Y) + M \bar Y} \right).
 $$
@@
 
-\end{itemize}
-
 \end{document}

### 2020-04-21 methods/mem_heatmap.R

@@ -1,0 +1,32 @@
+FP = 0.1; FN = 0.1
+odds_ratio = 1
+M = 10
+
+memadj <- function(odds_ratio, M, FP, FN) {
+  prevalence = odds_ratio/(1+odds_ratio)
+  return ( 1 + FP + FN - (M-1) * odds_ratio * (FP*(1-prevalence) + FN *(prevalence))/((1-prevalence) + M * prevalence))
+}
+
+seq_oddsratio = exp(seq(log(0.01), log(50), length.out = 100))
+seq_M = exp(seq(log(0.5),log(20), length.out = 100))
+result = matrix(nrow = length(seq_oddsratio), ncol = length(seq_M))
+FP = 0.15
+FN = 0.10
+
+for(i in 1:length(seq_oddsratio)) {
+  for (j in 1:length(seq_M)) {
+    odds = seq_oddsratio[i]
+    M = seq_M[j]
+    result[i,j] = memadj(odds, M, FP, FN)
+  }
+}
+
+result = data.frame(result)
+row.names(result) = round(seq_oddsratio,3)
+colnames(result) = round(seq_M,2)
+
+library("gplots")
+heatmap.2(as.matrix(result), col = bluered(100), dendrogram='none',
+          trace = "none", density.info = "none", Rowv=FALSE,
+          Colv=FALSE, key.xlab = "Relative Sampling Frequency", key.ylab = "Odds Ratio")
+

### 2020-04-28 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
-\documentclass[12pt]{article}
+\documentclass[11pt]{article}
 \usepackage{graphicx}
@@
 \usepackage{amsmath, amsthm, amsfonts}
-\usepackage[margin=1in]{geometry}
+\usepackage[margin=1.5in]{geometry}
 
@@
 \section{Introduction}
-
 The World Health Organization has declared the coronavirus disease 2019 (COVID-19) a public health emergency.
-As of March 28th, 2020, a total of 622,450 cases have been confirmed worldwide.
-As of that afternoon, the New York Times reports at least 113,031 people across the United States have tested positive for the virus, and at least 1,895 patients with the virus have died.
-Aggressive policies have been put in place across the US with at least 50\% of the US population officially urged to stay home via state-wide executive actions.
+As of April 27th, 2020, a total of 2,993,00 cases have been confirmed worldwide.
+As of that afternoon, the New York Times reports at least 965,214 people across the United States have tested positive for the virus, and at least 49,465 patients with the virus have died.
+Aggressive policies had been put in place across the US with at least 50\% of the US population officially urged to stay home via state-wide executive actions.
 
@@
 Statisticians, epidemiologists, economists, and data scientists the world over have been using this granular data to
+
+The CDC reports 413,867 total tests performed in the US as of April 25th.
+
+Due to limited testing capacity, local and state health departments have focused on testing only those from high-risk populations, resulting in low data quality due to bias in sampling from the overall population.
+
+A critical question is whether there are alternative data streams that may be leveraged to understand the handling of the COVID-19 pandemic in the US.
+
+Despite these necessary steps, the data landscape for understanding COVID-19 remains limited.
+Public databases maintained by Johns Hopkins University (\url{https://bit.ly/2UqFSuA}) and the New York Times (\url{https://bit.ly/2vUHfrK}) provide incoming county-level information of confirmed cases and deaths.
 
@@
 
+Measurement-error discussions.
+Selection bias . Proponents of this critique suggest well-designed studies.
+
+Stanford study was Ionides.  They however, forgot about Critique #2: measurement
+
 \subsection{Outline}
 
+This article discusses the relationship between three statistical concepts: selection bias, measurement-error, and the long-forgotten population size.
+We clarify mathematically why the situation is much more complex than it first appears.
+Through simple mathematical argument, we demonstrate three important, yet often forgotten, tenants of
+
 \begin{itemize}
-	\item Selection bias + MEM = problems! Scales with population size so cross country comparisons are meaningless.
-	\item What about cross-time same spot?  That's equally difficult!
+	\item Unadjusted prevalence rates have a bias due to imperfect testing; bias is a complex function of measurement-error and
+	\item Adjusted prevalence rates have an additional term in the error decomposition that highlights the interplay among measurement error, selection bias, and prevalence!
+	\item All errors in these rates with population size so cross country comparisons are meaningless.
+Even if we scale by population size, the errors don't go away.
+So comparing two countries of different sizes based on case-counts per million is difficult
+	\item When testing increases are associated with increases in false positive/negative rates, the effective sample size may decrease.
+	\item Comparison of daily trends in rate of positive tests is equally problematic.  When testing rates and selection protocols are held constant, increases are underestimated and
 	\item Present a compartmental model that accounts for these time-varying factors.  Discuss identifiability (statistical issue) and how to proceed
@@
 \tilde y_n - \bar Y = \rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
-\left[ 1 + FP + FN - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right].
+\times \underbrace{\left[ 1 + FP + FN - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right]}_{D_M}.
+$$
+
+The updated formula accounts for measurement-error to re
+
+This results in an updates Z-score:
+$$
+Z_{n,N}
 $$
@@
 Increased capacity may reduce
-
-\newpage
 
@@
 
+The effective sample size can then be bounded by
+$$
+n_{eff} = \frac{f}{1-f} \times E_{I} \left[ \rho_{I,Y}^2 D_M^2 \right]
+$$
+
 \subsection{Testing strategies and population size: a tango}
@@
 Data analysts have argued whether
+
+Increasing testing under test rationing lowers correlation.  Example in Michigan and try and find CA and NY numbers.  Then show that this may cause higher FPs (cite antibody studies).
+
+Plot Michigan tests per million and positives.  Next to NY.  Both as of April moved to prioritized sampling for symptomatic individuals.
+
+\subsection{Cross-country comparisons}
+
+So we can't estimate prevalence.  Can we not get a decent comparison?
+$$
+y_1 - y_2 = \frac{N_1}{n_1} ( \bar y_1 - Y_{1}) + \left( \frac{N_1}{n_1} Y_{1} - \frac{N_2}{n_2} Y_2  \right).
+$$
+OK then we should compare relative counts?
+The question is then ``What are you comparing at a populatoin level?''
+$$
+y_1/N_1 - y_2/ N_2 = \left[ f_1 \sigma_{Y_1} \rho_{I_1, Y_1} \sqrt{\frac{1-f_1}{f_1}} D_{M_1} - f_2 \sigma_{Y_2} \rho_{I_1, Y_1} \sqrt{\frac{1-f_1}{f_1}}  \right] + \left[ f_1 \bar Y_1 - f_2 \bar Y_2 \right]
+$$
+Is this a quantity of genuine interest?
+Under SRS, this makes sense if $f_1 = f_2$.
+Otherwise, you are comparing apples and oranges.
+With selection bias, it's not clear that agreement in this quantity is even desirable.
+
+$$
+\frac{n}{N} \times \rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y} \times D_M
+$$
+
+\subsection{Ration-based testing allocation}
+
+We assume that the percentage of tests is rationed according to severity of symptoms.
+
+Note that a person who displays severe symptoms may be COVID negative.  This gives us a sense of the data quality.
+
+\emph{Increasing testing leads to lower correlation in}
+
+\subsection{Estimation of Data Quality}
+
+We estimate $\rho$ using old and new methods.  Current SRS in US are weak.  So we just use my back of the envelope calculation. Tests per million as proxy since everyone is doing rationing. Cluster.
 
@@
 \frac{\bar y_2}{\bar y_1} - \frac{\bar Y_2}{\bar Y_1}
-&= \frac{\bar y_2}{\bar Y_1  \left(1 + \rho_{I,Y_1} \sqrt{\frac{1-f}{f}} CV (Y_1) \right) } - \frac{\bar Y_2}{\bar Y_1}  \\
-&\approx \frac{\bar y_2}{\bar Y_1} \left(1 - \rho_{I,Y_1} \sqrt{\frac{1-f}{f}} CV (Y_1) \right) - \frac{\bar Y_2}{\bar Y_1} \\
-&\approx \frac{1}{\bar Y_1} \left[ \rho_{I, Y_2} \sqrt{\frac{1-f}{f}} \sigma_{Y_2} - \rho_{I, Y_1} \sqrt{\frac{1-f}{f}} \sigma_{Y_1}
-\left( \frac{E_{J_2} [ I_{J_2} Y_{2,J_2} ]}{E_{J_2} [ I_{J_2} ]} + \frac{E_{J_2} [ I_{J_2} ] E_{J_2} [ Y_{2,J_2} ]}{E_{J_2} [ I_{J_2} ]} - \frac{E_{J_2} [ I_{J_2} ] E_{J_2} [ Y_{2,J_2} ]}{E_{J_2} [ I_{J_2} ]} \right) \right] \\
-&\approx \frac{1}{\bar Y_1} \left[ \rho_{I, Y_2} \sqrt{\frac{1-f}{f}} \sigma_{Y_2} - \rho_{I, Y_1} \sqrt{\frac{1-f}{f}} \sigma_{Y_1}
-\left( \rho_{I, Y_2} \sqrt{\frac{1-f}{f}} \sigma_{Y_2} + \mu_{Y_2} \right) \right]
+&= \frac{\bar Y_2}{\bar Y_1} \left[ \rho_{I_2, Y_2} \sqrt{\frac{1-f_2}{f_2}} CV(Y_2)  - \rho_{I_1,Y_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \left[ \rho_{I_2, Y_2} \sqrt{\frac{1-f_2}{f_2}} CV(Y_2) + 1 \right] \right]
 \end{aligned}
 $$
+
+Supposing the sampling rate is constant, and
 
@@
 
+What is the effective sample size for these tests?
+
+\subsubsection{Rate comparisons: a compounding mess}
+
 \subsection{Testing increase changes data quality}
@@
 
-\section{Compartmental model}
-
-(A) Heterogeneity in $R_0$
-(B) Selection bias
-(C)
-
-Above we took a general design perspective.  Here we take a model-assisted perspective.
-
-Susceptible individuals can become exposed to the disease.
-
-Selection bias occurs at 2 phases. Phase 1 is self-selection to seek
-
-Phase 2 is the decision to receive testing.  This is performed by the
-
-The result is we have multiple types of people: (A) Those disease-free who seek treatment due to high risk.
-
-We must then layer on measurement-error.  Tests can
-
-The SIR model with parameters.
-
-\begin{itemize}
-\item Simulation study showing that $R_t$ and prevalence over time changes the effective sample size....
-\end{itemize}
-
-\subsection{Ration-based testing allocation}
-
-We assume that the percentage of tests is rationed according to severity of symptoms.
-
-Note that a person who displays severe symptoms may be COVID negative.
-
-\emph{Increasing testing leads to lower correlation in}
-
-\subsection{Heterogeneiety  SIR with measurement-error and }
+\subsection{SI(E)R with measurement-error and }
 
@@
 Above we point to flaws in using observed data to reason about the
-
-\subsection{Interventions on network effects}
-
-While we think of the intervention as reducing risk COVID-19; in reality the interventio nis
-
-Classical contact pretends the graph is fixed.  Here we assume the intervention is to reduce degree and remove edges.  Contact tracing is indeed the way we can
-
-How does $R_0$ relate to connectivity?  This accounts for the heterogeneity. Epi knows this, but I think this point is salient.
 
@@
 
+\subsection{Ratio estimator}
+
+$$
+\begin{aligned}
+\frac{\bar y_2}{\bar y_1} - \frac{\bar Y_2}{\bar Y_1}
+&= \frac{\bar y_2}{\bar Y_1  \left(1 + \rho_{I_1,Y_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \right) } - \frac{\bar Y_2}{\bar Y_1}  \\
+&\approx \frac{\bar y_2}{\bar Y_1} \left(1 - \rho_{I_1,Y_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \right) - \frac{\bar Y_2}{\bar Y_1} \\
+&= \frac{1}{\bar Y_1}  \rho_{I_2, Y_2} \sqrt{\frac{1-f_2}{f_2}} \sigma_{Y_2} - \frac{\bar y_2}{\bar Y_1} \cdot \rho_{I_1,Y_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1).
+\end{aligned}
+$$
+Focusing on the second term, we have
+$$
+\begin{aligned}
+\frac{\bar y_2}{\bar Y_1} \cdot \rho_{I_1,Y_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) &= \frac{1}{\bar Y_1} \cdot \rho_{I_1,Y_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \left[ \bar y_2 - \bar Y_2 + \bar Y_2 \right] \\
+&= \frac{1}{\bar Y_1} \cdot \rho_{I_1,Y_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \left[ \rho_{I_2, Y_2} \sqrt{\frac{1-f_2}{f_2}} \sigma_{Y_2} + \bar Y_2 \right] \\
+&= \frac{\bar Y_2}{\bar Y_1} \cdot \rho_{I_1,Y_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \left[ \rho_{I_2, Y_2} \sqrt{\frac{1-f_2}{f_2}} CV(Y_2) + 1 \right] \\
+\end{aligned}
+$$
+Combining these we have
+$$
+\frac{\bar Y_2}{\bar Y_1} \left[ \rho_{I_2, Y_2} \sqrt{\frac{1-f_2}{f_2}} CV(Y_2)  - \rho_{I_1,Y_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \left[ \rho_{I_2, Y_2} \sqrt{\frac{1-f_2}{f_2}} CV(Y_2) + 1 \right] \right]
+$$
+Under the assumption that the data quality and the sampling fraction are constant across time, this can be simplified
+$$
+\rho \times \sqrt{\frac{1-f}{f}} \times \frac{\bar Y_2}{\bar Y_1} \left[ \bar Y_1 - \bar Y_2 - (1-\bar Y_1)(1-\bar Y_2) \rho \sqrt{\frac{1-f}{f}} \right]
+$$
+
 \section{Quotes}

### 2020-04-29 write-up/covid-writeup.tex
covid update

@@
 \begin{abstract}
-Many statisticians, epidemiologists, economists and data scientists have registered their serious reservations regarding the reported coronavirus case-counts.
+Many statisticians, epidemiologists, economists and data scientists have registered serious reservations regarding the reported coronavirus case-counts.
+Limited testing capacity across many states has been widely identified as a key driver of suppressed coronavirus case-counts.
+Calls for increased testing capacity are well-justified and increasingly frequent.
+While expanded testing is laudable, the impact of selection bias does not disappear until .
+Moreover, tests are imperfect.
+The rate of false positives and false negatives interacts in complex ways with selection bias.
+This note attempts to clarify this interaction.
+We demonstrate pitfalls and paradoxes that can arise when considering case-count data in the presence of selection bias and measurement-error.
+The discussion guides a sequence of suggestions in current case-count visualizations and reporting.
+unadjusted prevalence estimates Measurement-error do not  misinterpret  that does not cover   testing, if not on  increases  election bias and measurement-error are often cited as key drivers of case count data.
+It is known, for example, that false negatives deflate
+
+However, most researchers forget about the importance of population sizes
+
+In this note, we demonstrate through simple calculations how selection bias and theoretical
+.
+
 Comparing countries and states using those case-counts seem inappropriate when every nation/state have adopted different testing strategies and protocols.
-Estimating prevalence of COVID-19 based on these data is a hopeless exercise and several groups have recently argued for estimating the number of truly infected cases by using mortality rates.
+Here, we show the paradoxes that can persist  Estimating prevalence of COVID-19 based on these data is a hopeless exercise and several groups have recently argued for estimating the number of truly infected cases by using mortality rates.
 % In a lucid article on FiveThirtyEight.com Nate Silver walks us through the problem by a simple numerical exercise in terms of the basic deterministic models of an epidemic using hypothetical scenarios involving R0 and the effective R.
@@
 
-\subsection{SI(E)R with measurement-error and }
+\subsection{Death counts}
+
+\subsection{Stratified setting}
+
+\section{Brief discussion of model-based approaches}
 
@@
 
-This makes models \emph{fragile} and rely on strong data-generating assumptions.  For a disease
+This makes models \emph{fragile} and rely on strong data-generating assumptions.  In these settings data integration
 
-\subsection{Stratified sampling: improving precision in low-prevalence environments}
+\section{Stratified sampling: improving precision in low-prevalence environments}
 
@@
 
-Interventions in this area are designed experiments.  Unlike Fischer's null, here the objectives are settings, statisticians
+Interventions in this area are designed experiments.  Unlike Fischer's null, here the objectives are settings, statisticians.
+
+Account for data quality in models by incorporating auxiliary information.  Perform advanced sensitivity analysis.  Forecasts should be called coutnerfactual forecasts.
 

### 2020-04-29 methods/mem_heatmap.R

@@ -1,6 +1,10 @@
-FP = 0.1; FN = 0.1
-odds_ratio = 1
-M = 10
+memunadj <- function(rho, f, odds_ratio, M, FP, FN) {
+  prevalence = odds_ratio/(1+odds_ratio)
+  D_M = 1 - (M-1) * odds_ratio * (FP*(1-prevalence) + FN *(prevalence))/((1-prevalence) + M * prevalence)
+  term1 = rho*D_M*sqrt(prevalence*(1-prevalence)) * sqrt((1-f)/f)
+  term2 = FP-(FP+FN)*prevalence
+  return (term1 + term2)
+}
 
 memadj <- function(odds_ratio, M, FP, FN) {
   prevalence = odds_ratio/(1+odds_ratio)
@@ -10,14 +14,19 @@ memadj <- function(odds_ratio, M, FP, FN) {
 seq_oddsratio = exp(seq(log(0.01), log(50), length.out = 100))
 seq_M = exp(seq(log(0.5),log(20), length.out = 100))
 result = matrix(nrow = length(seq_oddsratio), ncol = length(seq_M))
-FP = 0.15
-FN = 0.10
+result_unadj = matrix(nrow = length(seq_oddsratio), ncol = length(seq_M))
+FP = 0.01
+FN = 0.15
+rho = 0.01
+f = 0.01
+
 
 for(i in 1:length(seq_oddsratio)) {
   for (j in 1:length(seq_M)) {
     odds = seq_oddsratio[i]
     M = seq_M[j]
     result[i,j] = memadj(odds, M, FP, FN)
+    result_unadj[i,j] = memunadj(rho, f, odds, M, FP, FN)
   }
 }
 

### 2020-05-04 write-up/covid-writeup.tex
covid update

@@
 \def\sgn{\text{sgn}}
+\def\I{\bf I}
 
@@
 Calls for increased testing capacity are well-justified and increasingly frequent.
-While expanded testing is laudable, the impact of selection bias does not disappear until .
+While expanded testing is a laudable goal, selection bias will impact estimates of disease prevalence and the effective reproduction number until the entire population is sampled.
 Moreover, tests are imperfect.
-The rate of false positives and false negatives interacts in complex ways with selection bias.
+False positive/negative rates interact in complex ways with selection bias.
 This note attempts to clarify this interaction.
-We demonstrate pitfalls and paradoxes that can arise when considering case-count data in the presence of selection bias and measurement-error.
-The discussion guides a sequence of suggestions in current case-count visualizations and reporting.
-unadjusted prevalence estimates Measurement-error do not  misinterpret  that does not cover   testing, if not on  increases  election bias and measurement-error are often cited as key drivers of case count data.
-It is known, for example, that false negatives deflate
-
-However, most researchers forget about the importance of population sizes
-
-In this note, we demonstrate through simple calculations how selection bias and theoretical
-.
-
-Comparing countries and states using those case-counts seem inappropriate when every nation/state have adopted different testing strategies and protocols.
-Here, we show the paradoxes that can persist  Estimating prevalence of COVID-19 based on these data is a hopeless exercise and several groups have recently argued for estimating the number of truly infected cases by using mortality rates.
-% In a lucid article on FiveThirtyEight.com Nate Silver walks us through the problem by a simple numerical exercise in terms of the basic deterministic models of an epidemic using hypothetical scenarios involving R0 and the effective R.
-In this note, we aim to (a)	we posit a conceptual mathematical framework to characterize both sampling bias and misclassification/imperfection of the test simultaneously on the estimation of the prevalence rate, (b)	review current testing strategies in some of the countries where we have testing data, and (c) provide guidelines for testing strategy/disease surveillance that may help track the pulse of the epidemic, to identify disease free areas and identify disease outbreaks.
-% Also large-scale anti-body testing.
+Through simple calculations, we demonstrate pitfalls and paradoxes that can arise when considering case-count data in the presence of selection bias and measurement-error.
+The discussion guides a series of suggestions to improve current case-count reporting.
 \end{abstract}
@@
 Public databases maintained by Johns Hopkins University (\url{https://bit.ly/2UqFSuA}) and the New York Times (\url{https://bit.ly/2vUHfrK}) provide incoming county-level information of confirmed cases and deaths.
-Statisticians, epidemiologists, economists, and data scientists the world over have been using this granular data to
-
-The CDC reports 413,867 total tests performed in the US as of April 25th.
-
-Due to limited testing capacity, local and state health departments have focused on testing only those from high-risk populations, resulting in low data quality due to bias in sampling from the overall population.
-
-A critical question is whether there are alternative data streams that may be leveraged to understand the handling of the COVID-19 pandemic in the US.
-
-Despite these necessary steps, the data landscape for understanding COVID-19 remains limited.
-Public databases maintained by Johns Hopkins University (\url{https://bit.ly/2UqFSuA}) and the New York Times (\url{https://bit.ly/2vUHfrK}) provide incoming county-level information of confirmed cases and deaths.
-
-The WHO reports, however, only 103,945 total tests performed in the US as of March 19th.
-Due to limited testing capacity, local and state health departments have focused on testing only those from high-risk populations, resulting in low data quality due to bias in sampling from the overall population.
-The public health community requires auxiliary sources of information to improve national and local health policy decisions.
-Certain survey efforts are underway, but may take time to yield fruit.
+Statisticians, epidemiologists, economists, and data scientists the world over have been using this granular data to build model-based forecasts of future COVID-19 case-counts, deaths, and hospitalizations.
+Variations on the SIR models have been used to draw inferences and build forecasts.
+In most cases, parameter inference uses observed case-count and/or observed COVID-19 related deaths to infer latent trajectories of the pandemic (cites).
 
 The goal of this paper is to express reservations at the use of case-counts as a proxy for prevalence and disease trajectory as well as its use as direct input into estimation of standard epidemiological models for inference and forecasting.
-The two critical features are selection bias due to varied testing strategies and measurement error due to imperfect tests (i.e., false positives and negatives).
-We demonstrate via theoretical analysis
-More testing equals more data but does not imply more information.  The key issue is selection bias and measurement-error.
+The reason is simple: current models ignore selection bias and measurement-error.
 
-Given population sizes are quite heterogeneous,
-the current data are of limited use to understand prevalence and even trajectory of the disease.
+Selection bias enters due to differences in testing strategies across countries and states.
+In the US, for example, limited testing capacity has caused local and state health departments to focus on testing only  high-risk populations.
+Testing requires the individual to self-select into testing.
+Demands for testing increases, while laudable, ignore the issues of self-selection and measurement-error.
+While increasing testing capacity increases data quantity, there is no guarantee for increased data quality.
+We will show that aggressive pushes for ramped up testing capacity that are tied to decreases in data quality may have a deleterious effect in our ability to estimate quantities of interest such as prevalence and effective reproduction number.
 
-A critical question is to understand the mismatch between  there are alternative data streams that may be leveraged to understand the handling the COVID-19 pandemic in the US.
+In this paper, we will demonstrate the complex interaction between these two fundamental concepts.
+Measurement-error is often associated with parameter attenuation; howeve,r we show that the combination with selection effects can cause bias in parameter estimates to change sign and increase/decrease in magnitude.
+Moreover, the bias may depend on the sampling fraction, prevalence, and population size.
+Without a complete understanding of these interations, we are destined to misinterpret case-counts and come to erroneous conclusions.
+We will show, through simple calculations, how we can misguide ourselves and the need for more precision and care in presenting COVD-19 statistical models to the broader research and non-academic communities.
 
@@
 
-Stanford study was Ionides.  They however, forgot about Critique #2: measurement
+Stanford study was Ionides.  They however, forgot about Critique \#2: measurement
 
@@
 We clarify mathematically why the situation is much more complex than it first appears.
-Through simple mathematical argument, we demonstrate three important, yet often forgotten, tenants of
+Through simple mathematical argument, we demonstrate five important, yet often forgotten, tenants of statistical science.
+First we show that unadjusted prevalence rates are unsurprisingly biased when tests are imperfect.
+What is surprising is that the direction and magnitude of the bias can vary substantially and interacts with the sampling fraction.
+Next we show that data quality for adjusted prevalence rates incorporate an additional term in the error decomposition that highlights the interplay among measurement error, selection bias, and prevalence.
+Third, we show that daily trends in rate of positive tests is equally problematic even if testing rates and selection protocols are held constant.
+We show that the rate of change in COVID-19 observed case-counts are often an underestimate of the true rate.
+This implies use of observed rates will lead a data analyst to be overly optimistic in the early stages that the disease is not an issue, and also overly optimistic in the decline of the prevalence after the peak.
+Fourth, we show that estimates of the reproducility rate are also impacted.
 
-\begin{itemize}
-	\item Unadjusted prevalence rates have a bias due to imperfect testing; bias is a complex function of measurement-error and
-	\item Adjusted prevalence rates have an additional term in the error decomposition that highlights the interplay among measurement error, selection bias, and prevalence!
-	\item All errors in these rates with population size so cross country comparisons are meaningless.
-Even if we scale by population size, the errors don't go away.
-So comparing two countries of different sizes based on case-counts per million is difficult
-	\item When testing increases are associated with increases in false positive/negative rates, the effective sample size may decrease.
-	\item Comparison of daily trends in rate of positive tests is equally problematic.  When testing rates and selection protocols are held constant, increases are underestimated and
-	\item Present a compartmental model that accounts for these time-varying factors.  Discuss identifiability (statistical issue) and how to proceed
-	\item Review testing and building selection models (sensitivity analysis)
-	\item Robustness (what directions are the models most resilient) and the notion of forecasts in bubbles
-	\item Testing, disease surveillance and second-wave testing goals.
-\end{itemize}
+The above analyses are focused on the issues with using observed case-counts to understand the trajectory for a single country or state.
+The media and organizations are also interested in comparisons of case-counts across countries.
+To account for population size, case-counts are scaled by the country population size.
+We show that even after this adjustment, cross-country comparisons can be plagued by estimation error scaling by population size.
+We end the paper with a discussion of the benefits of randomized testing, where auxiliary information is most relevant, and the distinction between data quality and decision making in the current pandemic.
 
-\section{}
+\section{Analysis of case-count data}
 
-The naive data analyst may think
-\begin{center}
-\emph{As we increase testing capacity, we will learn more about disease prevalence.}
-\end{center}
+We start with some simple notation.
+Let $N$ denote the population size.
+For state-level analysis $N$ is the state's total population, while for country-level analysis $N$ is the country's total population.
+At a fixed moment in time, let $Y_j$ denote COVID-19 status for the $j$th individual in the population, $j=1,\ldots, N$.
+Here, like in survey methodology (cite), we treat COVID-19 status as a fixed but unknown population quantity of interest.
+For simplicity, we start by ignoring the dynamic nature of the viral outbreak as well as the fact that individuals can recover from the disease and assume either individual $j$ is COVID-19 positive and $Y_j=1$ or is COVID-19 negative and $Y_j=0$.
+We also let $I_j \in \{0,1\}$ be an indicator that the individual was selected for testing ($I = 1$) or not ($I=0$).
 
-Case-count is considered a proxy for disease prevalence and contagion spread.
+To start, we assume the primary questions of interest are estimates of the overall number of COVID-19 cases and/or disease prevalence.
+That is, we are interested in either the population total $Y = \sum_{j=1}^N Y_j$ or the population average $\bar Y = Y/N$.
+Suppose that $n$ tests are performed and we observe the values $y_1, \ldots, y_n \in \{0,1\}$.
+Then a natural candidate for prevalence is the proportion of positive tests $\bar y = \frac{1}{n} \sum_{i=1}^n y_i$, and a natural candidate for overall cases is $N \times \bar y$.
+Under simple random sampling (SRS) or any other epsem design, the above are unbiased estimators of the population-level quantities of interest.
+Under SRS, the variance of the estimator can be expressed as $\frac{1}{N-1} \times \frac{1-f}{f} \times \sigma_Y^2$ where $f = n/N$ is the sampling fraction and $\sigma_Y^2 = \frac{1}{N} \sum_{i=1}^N (Y_i - \bar Y)^2 = \bar Y (1- \bar Y)$.
 
-Let $N$ denote the population size.
-At a fixed moment in time, let $Y_j \in \{ 0,1\}$ for $j=1,\ldots, N$ denote COVID status.
-As in survey sampling methodology, we treat these as fixed, unknown population quantities.
-For now, we ignore the dynamic nature of the viral outbreak; so either individual $j$ is COVID-19 positive and $Y_j=1$ or is COVID-19 negative and $Y_j=0$.
-A testing indicator, $I_j \in \{0,1\}$ is an indicator that the individual was selected for testing ($I = 1$) or not ($I=0$).
+When the sampling mechanism is non-random, selection effects may cause bias in the above estimates.
+To better understand this issue, Meng (2019) provided the following intuitive and powerful statistical decomposition of the error between $\bar y$ and the true proportion $\bar Y$
+$$
+\bar y_n - \bar Y =  \rho_{I, Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_Y.
+$$
+The first term represents \emph{data quality}, the second \emph{data quantity}, and the third \emph{problem difficulty}.
+The term $\rho_{I,Y}$ is the empirical correlation between the population values~$\{ Y_j \}_{j=1}^N$ and the selection values $\{ I_j \}_{j=1}^N$.
+Under simple random sampling, $E_{\I} [ \rho_{I,Y} ] = 0$, where the expectation is with respect to the selection mechanism~$\I$, so there is no bias.
+The SRS variance formula above shows that $E_{\I} [ \rho_{I,Y}^2 ]  = 1/(N-1)$.
+The key issue with selective testing is that $E_{\I} [ \rho_{I,Y} ] \neq 0$.
+Meng identified this as the fundamental issue that can lead to paradoxes in the analysis of big data.
 
-A primary question is the prevalence o
-We are interested in the population average $\bar Y = \frac{1}{N} \sum_{j=1}^N Y_j$.
-Suppose we observe a sample of $n$ tests.
-A natural candidate for prevalence is the proportion of positive tests $\bar y = \frac{1}{n} \sum_{i=1}^n y_i$.
+Here we highlight the core insights relevant to the COVID-19 crisis.  First, comparing the mean-squared error under selection mechanism $\I$ and SRS, we see that
+$$
+\frac{E_{\I} \left[ (\bar y_n - \bar Y)^2 \right]}{\sqrt{V_{SRS} (\bar Y)}} = (N-1) E_{\I} \left[ \rho_{I,Y}^2 \right].
+$$
+This points to a troubling and paradoxical situation: the error relative to SRS increases as a function of population size.
+Meng terms this the ``Law of Large Populations'' (LLP).
+The points to a critical issue in the current media practices in communication of case-count data: two countries with the same testing strategy (i.e., $\rho_{I,Y}$ equal) can yield wildly different estimates due to population size.
+Large countries like the US may have similar true prevalence rates as smaller countries like the UK.
+Even under similar testing strategies, the mean-squared error in the prevalence rate for the US will have substantially more variation.
+For example, the MSE will be almost 5 times that in the US compared to the UK.
+Therefore, strong opinions built from case-count records may be substantially distorted.
 
-Prevalence is important in the long-run because the
+Second, there have been calls for increased testing.
+While important, many conflate increased testing capacity with increased quality of testing.
+We will discuss this point further in Section XX.
+In the COVID-19 crisis, for example, New York has performed a total of $n = 872,481$ tests from April 20th to 30th.
+In Figure XX, we present the trajectory of testing per day and positive cases per day.
+The New York population is $N=8,399,000$; removing those individuals who have been already tested leaves a population of $7,XXX,XXX$.
+This leads to a fraction sampled of $f = 0.1$.
+A recent study attempted to estimate prevalence in New York.
+After appropriate adjustments (see Appendix XX for details), a plausible range for prevalence seems to be between $2\%$ and $12\%$.
+Suppose that individuals who are COVID-19 positive are $1.5$ times more likely to get tested than those individuals who are COVID-19 negative (a modest relative rate).
+The question is ``What is the sample size from a SRS that would yield equivalent MSE in the estimated prevalence?'' Using the above MSE result, Meng showed the effective sample size is equal to $n_{eff} \leq \frac{1}{E_I [ \rho_{I,Y}^2 ]} \frac{f}{1-f}$ and that ranges from $208$ to $38$ across the range of plausible values of prevalence.
+This is the sad state of affairs, saying that the effective sample size even in best case scenarios is not better than a small random sample from the population.
 
-Based on Meng (2019), we can express the error between $\bar y$ and the true proportion as
-$$
-\bar y_n - \bar Y =  \rho_{I, Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_Y
-$$
-Under random sampling $E [ \rho_{I,Y} ]$ = 0 and there is no bias.  Under selective testing,
-
-Case-count
-Figure~\ref{} presents testing per capita across reporting countries.  We see th at
-
-Two countries with the same testing strategy (i.e., $\rho_{I,Y}$ equal) can yield wildly different estimates due to pop size.
-
-For the current crisis, the analysis should be extended in two directions.  First, imperfect testing . Measurement-error.
+The remainder of this paper aims to build upon these fundamental insights by extending the decomposition in two directions: accounting for  measurement-error and the temporal nature of the pandemic.
+We then discuss testing strategies, effective sampling methods, and data versus decision-making.
 
@@
 COVID-19 testing is no exception.
-Data analysis must account for test inaccuracy to distinguish signal and noise.
 Here we investigate the interplay between imperfect testing and selection bias.
@@
 Let $P_j$ be an indicator of measurement error, equal to $1$ when we mismeasure the outcome and $0$ when we observe the true outcome.
-We suppose this is a stochastic variable that satisfies $\pr(P_j = 1 \mid Y_j = 1, I_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0, I_j = 1) =: FP$ is the false-positive rate.
+We suppose this is a stochastic variable that satisfies $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.
 If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^\star = Y_j(1-P_j) + (1-Y_j) P_j$.
@@
 To do this, we require the sampling rates differential.
-Let $f_1 := P_J(I_J = 1 \mid Y_J = 1)$ and $f_0 := P_J(I_J = 1 \mid Y_J = 1)$ be the sampling rates.
-Then if $\Delta = f_1 - f_0$ is the sampling rate differential, we have
+Let $f_1 := \pr (I_J = 1 \mid Y_J = 1)$ and $f_0 := \pr(I_J = 1 \mid Y_J = 1)$ be the sampling rates.
+Then $\Delta = f_1 - f_0$ is the sampling rate differential and we have
 $$
@@
 The final term in brackets is the \emph{measurement-error adjustment to data quality} which is a complex function of sampling rate differential, the odds ratio, and the ratio of measurement-error interaction with prevalence and sampling rates interaction with prevalence.
-Note $\sgn(\Delta) = \sgn(\rho_{I,Y})$ so the measurement-error adjustment either shrinks the data quality measure toward zero or reverses its sign.
+Note $\sgn(\Delta) = \sgn(\rho_{I,Y})$ by equation~\ref{eq:binaryrho} so the measurement-error adjustment either shrinks the data quality measure toward zero or reverses its sign.
 
-While prior investigations have notes the interaction between measurement-error and selection bias (cites), the interaction with the sample size relative to the population,i.e., $f$, has largely been ignored.
+While prior investigations have noted the interaction between measurement-error and selection bias (cites), the interaction with the sample size relative to the population, i.e., $f$, has largely been ignored.
 The above statistical decomposition clarifies the importance of this quantity~$f$.
@@
 This interaction is complex, but implies that whether the estimate $\bar y^\star_n$ is an overestimate or underestimate is a complicated question due to the relation amongst these three pieces.
+
+In the current setting, supposing the ratio of conditional selection rates $f_1/f_0$ is equal to 1.5.
+The false negative rate ranges from 15\% to 40\%, while the false positive rate is low (around 1\%).
+Then under plausible prevalence rates that we discussed before, the relative MSE ranges from 1.4 to 0.12.
+So that means in some cases we have a huge increase in MSE and in other settings we have a huge decrease in MSE.
+What drives this is the false negative rate interaction with prevalence and sampling rates.
+Therefore, whether we are better or worse off is a very difficult question to answer.
 
@@
 $$
-
-The updated formula accounts for measurement-error to re
-
-This results in an updates Z-score:
-$$
-Z_{n,N}
-$$
-
-Above we ignore an important consideration, namely that testing quality is likely a dynamic process.
-In the US, states are rightly trying to increase testing capacity.
-Commentators have requested expedited FDA approval for variations on the currently available tests to speed up production.
-This is likely to increase testing inaccuracies, impacting both false negative and false positive rates.
-Increased capacity may reduce
-
-\begin{wrapfigure}{r}{0.5\textwidth}
-\centering
-\includegraphics[width = 0.4\textwidth]{../methods/figs/mem_heatmap_article.png}
-\caption{Measurement-error data quality adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.15$ and $FN=0.10$.
-Color scaled so blue = $-6$, white = 0, and red = $6$.}
-\end{wrapfigure}
-
-For fixed prevalence,
-
 The first term is the same as before but increased by $(1 + FP + FN)$ to account for the additional uncertainty due to measurement-error.
@@
 
-The effective sample size can then be bounded by
+\begin{wrapfigure}{r}{0.5\textwidth}
+\centering
+\includegraphics[width = 0.4\textwidth]{../methods/figs/mem_heatmap_article.png}
+\caption{Measurement-error data quality adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.15$ and $FN=0.10$.
+Color scaled so blue = $-6$, white = 0, and red = $6$.}
+\vspace{-0.7cm}
+\end{wrapfigure}
+
+Under the same numbers, $D_M$ is approximately 1.7.  The effective sample size can then be bounded by
 $$
-n_{eff} = \frac{f}{1-f} \times E_{I} \left[ \rho_{I,Y}^2 D_M^2 \right]
+n_{eff} = \frac{f}{1-f} \times \frac{1}{E_{I} \left[ \rho_{I,Y}^2 D_M^2 \right]}.
 $$
+This lowers the ranges by a factor of almost $3$, leading to effective sample sizes from 69 to 13.
 
-\subsection{Testing strategies and population size: a tango}
+The above argument shows that the effective sample size is affected in a complex manner by the interaction between selection bias and measurement error.
+Here, we discuss the impact of the interaction on the effective sample size when testing capacity is tied to measurement-error.
+In the current pandemic, there have been well-justified calls for increases in testing capacity.
+These increases, however, may come at the cost of increases in false-positive and false-negative rates.
+In terms of effective sample size, if the data quality term remains constant, then testing increases can actually hurt effective sample size.
+Consider the simplest case where $f$ increases from $0.05$ to $0.1$; suppose this required including tests such that the false negative rate jumps from $0.05$ to $0.15$.
+Then we would think effective sample size should increase by a factor of $3.4$ but the actual increase is by a factor of only $1.8$!
+Think about what this is saying, this is saying that sampling $15\%$ of the population rather than $5\%$ will only increase the effective sample size by a factor $1.8$ if the selection-bias persists and the false negative rate climbs to $15\%$.
+It has been reported that false negative rates could increase more than that (cite NYT) and the false positive rate could go up slightly as well.
+In a worst case scenario, if $FN \to 0.3$ and $FP \to 0.05$ then the factor is only 1.3.
 
-Data analysts have argued whether
+\subsection{Regrettable rates}
+\label{section:rates}
 
-Increasing testing under test rationing lowers correlation.  Example in Michigan and try and find CA and NY numbers.  Then show that this may cause higher FPs (cite antibody studies).
+The data analyst, now frustrated with me but at least wary of estimating prevalence and total counts, pauses and thinks.
+They return shortly thereafter, a bit agitated but persistent: \emph{Ok, perhaps total counts is a lost cause.
+Certainly, however, we can estimate the rate of growth.
+All I want to know is when we hit the point at which the curve flattens and number of deaths decrease.
+That can't be too hard, surely!}
 
-Plot Michigan tests per million and positives.  Next to NY.  Both as of April moved to prioritized sampling for symptomatic individuals.
+Unfortunately, ratio estimators do not cancel errors as now both the numerator and the denominator are uncertain.
+This means selection bias and measurement error can have paradoxical effects.
+Here we consider the ratio estimators for the relative change in the prevalence rate.
+We let $\bar Y_1$ and $\bar Y_2$ denotes the prevalence at time step $1$ and $2$ respectively (i.e., prevalence on two consecutive days).
+Then the ratio estimator is given by $r = \bar y_1 / \bar y_2$.
+We assume both numerator and denominator are the prevalence estimates adjusted for measurement-error.
+If the sample size at each time were equal (i.e., $n_1 = n_2 = n$) then this analysis would be equivalent to comparing the increase in observed case-counts (i.e., $\bar y_1/\bar y_2 = y_1/y_2$).
+We wish to express the error $\frac{\bar y_2}{\bar y_1} - \frac{\bar Y_2}{\bar Y_1}$ in statistical terms.
+Using a Taylor series approximation, we show that the error can be expressed approximately as
+$$
+\begin{aligned}
+\frac{\bar Y_2}{\bar Y_1} \left[ \rho_{I_2, Y_2} D_{M_2} \sqrt{\frac{1-f_2}{f_2}} CV(Y_2)  - \rho_{I_1,Y_1} D_{M_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \left[ \rho_{I_2, Y_2} D_{M_2} \sqrt{\frac{1-f_2}{f_2}} CV(Y_2) + 1 \right] \right]
+\end{aligned}
+$$
+where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement-error adjustment, and $CV(Y_j) = \sigma_{Y_j}/\bar Y_j$ is the coefficient of variation at time step $j$.
+We see that the magnitude of the error depends on the population-level ratio $\bar Y_2 / \bar Y_1$ so a large decrease in the prevalence rate will have a relatively small error as compared to a large increase in the prevalence rate.
+There are three terms in the brackets.
+The two terms $\rho_{I_2, Y_2} D_{M_2} \sqrt{\frac{1-f_2}{f_2}} CV(Y_2) - \rho_{I_1, Y_1} D_{M_1} \sqrt{\frac{1-f_1}{f_1}} CV(Y_1)$  represent the ``cancellation'' that the naive analyst is hoping will occur.
+We can see that when the data quality, sampling fraction, and measurement-error adjustments are help constant this different simplifies greatly to $\rho_{I,Y} D_M \sqrt{\frac{1-f}{f}} ( \bar Y_1 - \bar Y_2)$.
+The final component $\bar Y_1 - \bar Y_2$ represents impact of the prevalence trajectory on the statistical error.
+When prevalence is increasing the component is negative, and when prevalence is decreasing the component is positive.
+At the peak this term is zero, representing when the naive analysts intuition is closest to correct.
+Regardless, we always observe the final term which is a cross term that depends on the statistical error in both stages.
+
+Figure XX shows a trajectory of bias under several prevalence and data quality scenarios.
+We see that in each case, the rate of change in prevalence is underestimated at most times, except for large decreases which are overestimated.
+This points to a very sad realization.
+At the beginning of the pandemic, we are underestimating the rate of change based on observed case counts.
+Not only that, but the error is at its largest when the increase is large.
+This leads us to not take the virus seriously early on based on the observed data!
+Moreover, after the peak, we see that the bias is also negative other than for large decreases in the rate of change in prevalence.
+This implies our estimates based on observed data trick us into thinking that the virus is ``going away'' faster than it actually is!
+
+A final question is whether an observed drop in prevalence can be interpreted as a ``real'' decrease in prevalence.  To this end, one would typically construct a Z-score and perform a
+
+Moreover, the question is ``Can we trust the observed data to let us know if we have reached the peak?''
+
+Based on this analysis, seeing a falling rate for XX time steps in a row would be needed to feel confidence that the true prevalence rate is on the decline.
+
+\subsubsection{Estimation of effective reproduction number}
+Many well-respected epidemiologists argue that tracking the effective reproduction number $R_t$ is the only way to manage through the crisis (cites).
+We show here how these estimates are also impacted by selection bias and measurement error.
+For simplicity, we again assume the estimates are adjusted for known false positive and false negative rates.
+Luis and Ribeiro (2008) show that under a Poisson likelihood, a simple relation between the trajectory of \emph{true new cases} and the effective reproduction number can be derived (see Appendix XX for discussion).
+In particular, under an SIR model, the number of case counts is Poisson distributed with rate $K_{t-1} \exp \left( \frac{1}{\gamma} (R_t - 1) \right)$ where $\gamma$ is the serial interval, which is approximately $7$ days for COVID-19 (Sanche, 2020).
+Using this, a method of moments estimator of the effective reproduction number at time $t$ can be given by
+$$
+R_t = 1 + \frac{1}{\gamma} \log \left( \frac{K_t}{K_{t-1}} \right).
+$$
+Of course, we do not observe $K_t$ and $K_{t-1}$ but noisy proxies $y_t$ and $y_{t-1}$, i.e., the observed case-count on day $t$ and $t-1$ respectively.
+The hope is that $y_t/y_{t-1}$ is a good proxy for $K_t/K_{t-1}$.
+Under SRS with the same sample size at each step, this is true for sufficiently large sample size.
+However, similar to Section~\ref{section:rates}, the proxy has error in both numerator and denominator that may cause issues.
+We again wish with express the statistical error in the $1+\frac{1}{\gamma} \log \left( \frac{y_t}{y_{t-1}} \right) - R_t$ in useful terms.
+Based on Section~\ref{section:rates}, we can re-arrange those terms to show that the error is  given by
+$$
+\frac{1}{\gamma}\log \bigg[ &1 + \rho_{I_t, K_t} D_{M_t} \sqrt{\frac{1-f}{f}} CV(K_t)  \\
+&- \rho_{I_{t-1},K_{t-1}} \sqrt{\frac{1-f}{f}} CV (K_{t-1}) \left[ \rho_{I_t, Y_t} \sqrt{\frac{1-f}{f}} CV(K_{t-1}) + 1 \right] \right].
+$$
+The coefficient of variation $CV(K_t) = (1- \bar Y_t)$ is the fraction of the uninfected population that remain uninfected.
+So here we see the exact same trade-off as with the rates but on the logarithmic scale.
+For small values $\log(1+x) = x$ so the biases are similar.
+However, the error is no longer scaled by the rate of change in prevalence $\bar Y_2/\bar Y_1$.
+Instead the error is scaled by the serial interval $\gamma^{-1} = 1/7$.
+
+The above decomposition of the statistical shows us that
+
+The question of interest is ``Is the effective reproduction number below one?'' When this is true, we can feel like that the .
+Therefore, we find the effective reproduction number suffers a similar story. And is consistently underestimated for in initial phases and then we are overconfident in the decline.
+
+\subsection{Clinical trials}
+
+In
+
+Random sampling negates effect modifiers.
+Random treatment assignment negates unobserved confounders.
+Show simple example where $E[U]$ is high due to sampling bias and therefore you will think that the effect is significant!
+But it's because of sampling bias.
+Requires poststratification.
 
@@
 
-\subsection{Ration-based testing allocation}
+\section{Testing strategies and population size: a tango}
 
+Data analysts have argued whether
+
+Increasing testing under test rationing lowers correlation.  Example in Michigan and try and find CA and NY numbers.  Then show that this may cause higher FPs (cite antibody studies).
+
+Plot Michigan tests per million and positives.  Next to NY.  Both as of April moved to prioritized sampling for symptomatic individuals.
+
+\subsection{US testing: a ration system}
+
+There is a lot of hidden information in the selection indicator $I_j$.
+Up until recently, an individual had to first \emph{want} to be tested, then call their primary health care provider who decided whether they met thresholds for testing
+
+. Currently, we are unaware of reliable information on the number of cases where
+
+In the UU
 We assume that the percentage of tests is rationed according to severity of symptoms.
@@
 
+The CDC reports 413,867 total tests performed in the US as of April 25th.
+
 \subsection{Estimation of Data Quality}
@@
 We estimate $\rho$ using old and new methods.  Current SRS in US are weak.  So we just use my back of the envelope calculation. Tests per million as proxy since everyone is doing rationing. Cluster.
-
-\subsection{Regrettable rates}
-
-The data analyst, now wary of estimating prevalence and total counts, pauses and thinks.
-They retur  n shortly thereafter, a bit agitated but still  \emph{Ok, perhaps total counts is a lost cause.
-Certainly, however, we can estimate the rate of growth.
-All I want to know is when we hit the point at which the curve flattens and number of deaths decrease.
-That can't be too hard, surely!}
-
-The parameter $R_0$ in the ubiquitous SIR (susceptible-)
-
-Ratio estimator $\bar y_2/ \bar y_1$.
-
-Taylor series expansion:
-
-$$
-\begin{aligned}
-\frac{\bar y_2}{\bar y_1} - \frac{\bar Y_2}{\bar Y_1}
-&= \frac{\bar Y_2}{\bar Y_1} \left[ \rho_{I_2, Y_2} \sqrt{\frac{1-f_2}{f_2}} CV(Y_2)  - \rho_{I_1,Y_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \left[ \rho_{I_2, Y_2} \sqrt{\frac{1-f_2}{f_2}} CV(Y_2) + 1 \right] \right]
-\end{aligned}
-$$
-
-Supposing the sampling rate is constant, and
-
-We talk about wanting to know if we are at the peak.
-Let's define that time as when $\bar Y_2 / \bar Y_1 = 0$.
-Then $\mu_{Y_1} = \mu_{Y_2} = \mu$ and $\sigma_{Y_1} = \sigma_{Y_2} = \sigma$.
-If selection at each time is independent and the selection process is stationary, we have bias of
-$$
-\rho \sqrt{\frac{1-f}{f}} \sigma \sqrt{\frac{1-\mu}{\mu}} \left[ \sqrt{\frac{1-\mu}{\mu}} - \rho \sqrt{\frac{1-f}{f}} \right ]
-$$
-
-If we collect $f = 0.01$ and the peak occurs at $\mu = 0.03$ with selection bias of $\rho = 0.05$. $1/\rho^2 \times (\mu/(1-\mu))^4$
-
-Then we can be even further off than the prevalence by a factor of the odds ratio!
-
-Suppose we collected $f=1/2$ and peak occurs at $\mu = 0.10$ and $\rho = 0.005$, then $9-\rho 3 = 9 - 0.005 \times 3 = 8.985$ and
-
-{\bf How many days of dropping to guarantee it has decreased}
-
-What is the effective sample size for these tests?
-
-\subsubsection{Rate comparisons: a compounding mess}
-
 \subsection{Testing increase changes data quality}
@@
 \emph{Increasing testing leads to lower correlation only if we see the gap $\Delta$ decrease.  If we just see $f$ }
-
-\subsection{Clinical trials}
-
-Random sampling negates effect modifiers.
-Random treatment assignment negates unobserved confounders.
-Show simple example where $E[U]$ is high due to sampling bias and therefore you will think that the effect is significant!
-But it's because of sampling bias.
-Requires poststratification.
-
-\subsection{Death counts}
 
@@
 DTR and network connectivity is useful for understanding spread and where "testing should go".
+
+Huge benefits when the rates are very different.
 
@@
 
+Use of these bounds as E-scores.  How bad can it go?
+
 Interventions in this area are designed experiments.  Unlike Fischer's null, here the objectives are settings, statisticians.
@@
 Account for data quality in models by incorporating auxiliary information.  Perform advanced sensitivity analysis.  Forecasts should be called coutnerfactual forecasts.
+
+Poststratifiers and systematic measurement error.
+
+A critical question is whether there are alternative data streams that may be leveraged to understand the handling of the COVID-19 pandemic in the US.
 
@@
 
+\section{Derivation for the effective sample size}
+
+\begin{equation} \label{eq:binaryrho}
+\rho_{I,Y} = \Delta \sqrt{\frac{\bar Y (1 - \bar Y)}{f (1-f)} }
+\end{equation}
+where $\Delta = P_J (I_J = 1 \mid Y_J = 1) - P(I_J = 1 \mid Y_J = 0)$.  Suppose that $\Delta = 10\%$ and $\bar Y = 10\%$ then $\rho_{I,Y}$ is given by
+We expect the correlation to be more on the order
+
 \section{Quotes}

### 2020-05-04 methods/mem_heatmap.R

@@ -18,7 +18,7 @@ result_unadj = matrix(nrow = length(seq_oddsratio), ncol = length(seq_M))
 FP = 0.01
 FN = 0.15
 rho = 0.01
-f = 0.01
+f = 0.001
 
 
 for(i in 1:length(seq_oddsratio)) {
@@ -34,8 +34,15 @@ result = data.frame(result)
 row.names(result) = round(seq_oddsratio,3)
 colnames(result) = round(seq_M,2)
 
+result_unadj = data.frame(result_unadj)
+row.names(result_unadj) = round(seq_oddsratio,3)
+colnames(result_unadj) = round(seq_M,2)
+
 library("gplots")
 heatmap.2(as.matrix(result), col = bluered(100), dendrogram='none',
           trace = "none", density.info = "none", Rowv=FALSE,
           Colv=FALSE, key.xlab = "Relative Sampling Frequency", key.ylab = "Odds Ratio")
 
+heatmap.2(as.matrix(result_unadj), col = bluered(100), dendrogram='none',
+          trace = "none", density.info = "none", Rowv=FALSE,
+          Colv=FALSE, key.xlab = "Relative Sampling Frequency", key.ylab = "Odds Ratio")

### 2020-05-05 write-up/covid-writeup.tex
update r and tex files

@@
 
+Let ${\bf u} = (u_1,u_2) \in \mathbb{R}^2$ and $g({\bf u}) = \frac{u_2}{u_1}$, i.e., a differentiable function $g:\mathbb{R}^2 \to \mathbb{R}$.
+Centering a Taylor series expansion of second-order around coordinates $(U_2, U_1) \in \mathbb{R}^2$ yields
+$$
+\begin{aligned}
+g({\bf u}) =& g(U_1, U_2) - \frac{U_2}{U_1^2} (u_1 - U_1) + \frac{1}{U_1} (u_2 - U_2) \\
+&+ \frac{1}{2} \left[ \frac{2 U_2}{U_1^3} (u_1 - U_1)^2 + 0 \times (u_2 - U_2)^2 - 2 \times (u_1 - U_1) (u_2 - U_2) \frac{1}{U_1^2} \right]
+\end{aligned}
+$$
+Plugging in $(\bar y_1, \bar y_2)$ for $(u_1, u_2)$ and $(\bar Y_1, \bar Y_2)$ for $(U_1, U_2)$ yields
+$$
+\begin{aligned}
+\frac{\bar y_2}{\bar y_1} - \frac{\bar Y_2}{\bar Y_1} =& - \frac{\bar Y_2}{\bar Y_1^2} (\bar y_1 - \bar Y_1) + \frac{1}{\bar Y_1} (\bar y_2 - \bar Y_2) \\
+&+ \frac{\bar Y_2}{\bar Y_1^3} (\bar y_1 - \bar Y_1)^2 -  (\bar y_1 - \bar Y_1) (\bar y_2 - \bar Y_2) \frac{1}{\bar Y_1^2} \\
+&= \frac{\bar Y_2}{\bar Y_1} \bigg[  \rho_{I_2,Y_2} \sqrt{\frac{1-f_2}{f_2}} CV (Y_2)  -\rho_{I_1,Y_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \\
+&+ \rho^2_{I_1,Y_1} \frac{1-f_1}{f_1} CV^2 (Y_1) -  \rho_{I_1,Y_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \times \rho_{I_2,Y_2} \sqrt{\frac{1-f_2}{f_2}} CV (Y_2)   \bigg] \\
+&= \frac{\bar Y_2}{\bar Y_1} \bigg[ \rho_{I_2,Y_2} \sqrt{\frac{1-f_2}{f_2}} CV (Y_2)  -\rho_{I_1,Y_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \bigg] \left[ 1 - \rho_{I_1,Y_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \right]
+\end{aligned}
+$$
+where the second equality is obtained by plugging in the statistical decomposition of the error for both time points and the coefficient of variation being defined as $CV(Y) := \sigma_Y/\mu_Y$.
+\newpage
 $$

### 2020-05-05 methods/regrettable_rates.R

@@ -1,0 +1,109 @@
+
+bary1 = 0.05
+bary2 = 0.02
+rho = 0.01; f = 0.001
+bary2/bary1
+
+bary1 - bary2 - (1-bary1)*(1-bary2) * rho * sqrt((1-f)/f)
+
+
+f = 0.1
+bary = seq(0.02, 0.14, by = 0.01)
+M = 1.5
+f0 = f / (M*bary +(1-bary))
+f1 = M*f0
+Delta = f1-f0
+
+rho = Delta * sqrt( bary*(1-bary)/ (f *(1-f)))
+
+f/(1-f) * 1/(rho^2)
+
+M = 2
+FN = 0.3
+FP = 0.01
+rho = 0.05
+
+temp = rho*sqrt(bary*(1-bary)) * (1 - (M-1) * (FP * (1-bary) + FN* bary) / ( (1-bary) + M* bary) * bary / (1-bary)) + sqrt(f/(1-f))*(FP-(FP+FN)*bary)
+
+temp2 = rho*sqrt(bary*(1-bary)) * (1 + FP + FN - (M-1) * (FP * (1-bary) + FN* bary) / ( (1-bary) + M* bary) * bary / (1-bary))
+
+(temp*sqrt((1-f)/f) /   (rho*sqrt((1-f)/f)*sqrt(bary*(1-bary))))^2
+
+(temp2*sqrt((1-f)/f) /   (rho*sqrt((1-f)/f)*sqrt(bary*(1-bary))))^2
+
+
+## Change f
+
+f = 0.05
+bary = seq(0.02, 0.14, by = 0.01)
+M = 1.5
+FN = 0.05; FP = 0.01; rho = 0.05
+
+term1 = f/(1-f) * 1/(1 + FP + FN - (M-1) * (FP * (1-bary) + FN* bary) / ( (1-bary) + M* bary) * bary / (1-bary))^2 * 1/rho^2
+
+f = 0.1
+bary = seq(0.02, 0.14, by = 0.01)
+M = 1.5
+FN = 0.2; FP = 0.01
+
+term2 = f/(1-f) * 1/(1 + FP + FN - (M-1) * (FP * (1-bary) + FN* bary) / ( (1-bary) + M* bary) * bary / (1-bary))^2 * 1/rho^2
+
+f = 0.1
+bary = seq(0.02, 0.14, by = 0.01)
+M = 1.5
+FN = 0.15; FP = 0.01
+
[…]

### 2020-05-05 write-up/covid-writeup.tex
removed excess derivations

@@
 where the second equality is obtained by plugging in the statistical decomposition of the error for both time points and the coefficient of variation being defined as $CV(Y) := \sigma_Y/\mu_Y$.
-\newpage
-$$
-\begin{aligned}
-\frac{\bar y_2}{\bar y_1} - \frac{\bar Y_2}{\bar Y_1}
-&= \frac{\bar y_2}{\bar Y_1  \left(1 + \rho_{I_1,Y_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \right) } - \frac{\bar Y_2}{\bar Y_1}  \\
-&\approx \frac{\bar y_2}{\bar Y_1} \left(1 - \rho_{I_1,Y_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \right) - \frac{\bar Y_2}{\bar Y_1} \\
-&= \frac{1}{\bar Y_1}  \rho_{I_2, Y_2} \sqrt{\frac{1-f_2}{f_2}} \sigma_{Y_2} - \frac{\bar y_2}{\bar Y_1} \cdot \rho_{I_1,Y_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1).
-\end{aligned}
-$$
-Focusing on the second term, we have
-$$
-\begin{aligned}
-\frac{\bar y_2}{\bar Y_1} \cdot \rho_{I_1,Y_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) &= \frac{1}{\bar Y_1} \cdot \rho_{I_1,Y_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \left[ \bar y_2 - \bar Y_2 + \bar Y_2 \right] \\
-&= \frac{1}{\bar Y_1} \cdot \rho_{I_1,Y_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \left[ \rho_{I_2, Y_2} \sqrt{\frac{1-f_2}{f_2}} \sigma_{Y_2} + \bar Y_2 \right] \\
-&= \frac{\bar Y_2}{\bar Y_1} \cdot \rho_{I_1,Y_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \left[ \rho_{I_2, Y_2} \sqrt{\frac{1-f_2}{f_2}} CV(Y_2) + 1 \right] \\
-\end{aligned}
-$$
-Combining these we have
-$$
-\frac{\bar Y_2}{\bar Y_1} \left[ \rho_{I_2, Y_2} \sqrt{\frac{1-f_2}{f_2}} CV(Y_2)  - \rho_{I_1,Y_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \left[ \rho_{I_2, Y_2} \sqrt{\frac{1-f_2}{f_2}} CV(Y_2) + 1 \right] \right]
-$$
-Under the assumption that the data quality and the sampling fraction are constant across time, this can be simplified
-$$
-\rho \times \sqrt{\frac{1-f}{f}} \times \frac{\bar Y_2}{\bar Y_1} \left[ \bar Y_1 - \bar Y_2 - (1-\bar Y_1)(1-\bar Y_2) \rho \sqrt{\frac{1-f}{f}} \right]
-$$
 

### 2020-05-05 write-up/covid-writeup.tex
deleted model section

@@
 \begin{aligned}
-\frac{\bar Y_2}{\bar Y_1} \left[ \rho_{I_2, Y_2} D_{M_2} \sqrt{\frac{1-f_2}{f_2}} CV(Y_2)  - \rho_{I_1,Y_1} D_{M_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \left[ \rho_{I_2, Y_2} D_{M_2} \sqrt{\frac{1-f_2}{f_2}} CV(Y_2) + 1 \right] \right]
+\frac{\bar Y_2}{\bar Y_1} &\times \bigg[ \rho_{I_2,Y_2} D_{M_2} \sqrt{\frac{1-f_2}{f_2}} CV (Y_2)  -\rho_{I_1,Y_1} D_{M_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \bigg] \\
+&\times \left[ 1 - \rho_{I_1,Y_1} D_{M_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \right]
 \end{aligned}
@@
 where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement-error adjustment, and $CV(Y_j) = \sigma_{Y_j}/\bar Y_j$ is the coefficient of variation at time step $j$.
-We see that the magnitude of the error depends on the population-level ratio $\bar Y_2 / \bar Y_1$ so a large decrease in the prevalence rate will have a relatively small error as compared to a large increase in the prevalence rate.
-There are three terms in the brackets.
-The two terms $\rho_{I_2, Y_2} D_{M_2} \sqrt{\frac{1-f_2}{f_2}} CV(Y_2) - \rho_{I_1, Y_1} D_{M_1} \sqrt{\frac{1-f_1}{f_1}} CV(Y_1)$  represent the ``cancellation'' that the naive analyst is hoping will occur.
-We can see that when the data quality, sampling fraction, and measurement-error adjustments are help constant this different simplifies greatly to $\rho_{I,Y} D_M \sqrt{\frac{1-f}{f}} ( \bar Y_1 - \bar Y_2)$.
-The final component $\bar Y_1 - \bar Y_2$ represents impact of the prevalence trajectory on the statistical error.
-When prevalence is increasing the component is negative, and when prevalence is decreasing the component is positive.
-At the peak this term is zero, representing when the naive analysts intuition is closest to correct.
-Regardless, we always observe the final term which is a cross term that depends on the statistical error in both stages.
+The magnitude of the error depends on the population-level ratio $\bar Y_2 / \bar Y_1$ so a large decrease in the prevalence rate will have a relatively small error as compared to a large increase in the prevalence rate.
+The second term represents the ``cancellation'' that the naive analyst is hoping will occur.
+The analyst's intuition only holds when the difference is equal to zero.
+This occurs at times when data quality, sampling fraction, measurement-error, and prevalence are all held constant.
+In the case of New York, for example, recent weeks it seems plausible that $\Delta$ (i.e., selection bias) has not changed drastically.
+Figure XX shows the rate of testing remains constant and the total cases relative to the population is small enough such that $f_1 \equiv f_2$ is reasonable.
+Finally, the measurement-error adjustment is likely to be constant as no new testing procedures have been reported (as far as this author knows).
+We will show how such a setting leads to over-estimation prior to and under-estimation after peak prevalence.
+The final component is an adjustment to this differential error based on the statistical error at the first time point.
 
-Figure XX shows a trajectory of bias under several prevalence and data quality scenarios.
-We see that in each case, the rate of change in prevalence is underestimated at most times, except for large decreases which are overestimated.
-This points to a very sad realization.
-At the beginning of the pandemic, we are underestimating the rate of change based on observed case counts.
-Not only that, but the error is at its largest when the increase is large.
-This leads us to not take the virus seriously early on based on the observed data!
-Moreover, after the peak, we see that the bias is also negative other than for large decreases in the rate of change in prevalence.
-This implies our estimates based on observed data trick us into thinking that the virus is ``going away'' faster than it actually is!
+Figure XXa shows a trajectory of bias under several prevalence and data quality scenarios.
+We generate data from an SIR model with infection rate $\beta = XX$, total population = $100,000$.
+We estimate the bias under a range of values for the relative sampling fraction $f_1/f_0$ and a sampling fraction of $f = 0.05$.
+Figure XXb shows the trajectory of $\bar Y_2 / \bar Y_1$.
+We see that in each case, the rate of change in prevalence is overestimated prior to the peak and then underestimated afterwards.
+From a decision making perspective, such biases have impacts.
+First, the over-estimation may give lawmakers and governors more leverage in proposing aggressive actions that reduce prevalence.
+Of course, the analysis supports the argument that the estimates based on available data are overly pessimistic.
+What both sides miss is that the direction of bias is non-constant over time.
+After we hit the peak, estimates will under-estimate prevalence.
+Lawmakers and governors will be put under pressure to relax social distancing measures.
+This will drive a sequence of cycle of panic and .
 
-A final question is whether an observed drop in prevalence can be interpreted as a ``real'' decrease in prevalence.  To this end, one would typically construct a Z-score and perform a
-
-Moreover, the question is ``Can we trust the observed data to let us know if we have reached the peak?''
-
-Based on this analysis, seeing a falling rate for XX time steps in a row would be needed to feel confidence that the true prevalence rate is on the decline.
+Moreover, the question is ``Can we trust the observed data to let us know if we have reached the peak?''  It appears that the peak time is the easiest to estimate in terms of having minimal error using availability data.
+Note, however, that the standard errors on such estimates will be incredibly wide as the effective sample size is quite small compared to SRS at each time.
+\textcolor{red}{Short description of effective sample size}.
 
@@
 Luis and Ribeiro (2008) show that under a Poisson likelihood, a simple relation between the trajectory of \emph{true new cases} and the effective reproduction number can be derived (see Appendix XX for discussion).
-In particular, under an SIR model, the number of case counts is Poisson distributed with rate $K_{t-1} \exp \left( \frac{1}{\gamma} (R_t - 1) \right)$ where $\gamma$ is the serial interval, which is approximately $7$ days for COVID-19 (Sanche, 2020).
+In particular, under an SIR model, the number of case counts on day $t$ is Poisson distributed with rate $K_{t-1} \exp \left( \frac{1}{\gamma} (R_t - 1) \right)$ where $K_{t-1}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval which is approximately $7$ days for COVID-19 (Sanche, 2020).
 Using this, a method of moments estimator of the effective reproduction number at time $t$ can be given by
@@
 The hope is that $y_t/y_{t-1}$ is a good proxy for $K_t/K_{t-1}$.
+Let $\hat R_t = 1 + \frac{1}{\gamma} \log \left( y_t / y_{t-1} \right)$ be the estimated effective reproduction number on day $t$.
 Under SRS with the same sample size at each step, this is true for sufficiently large sample size.
 However, similar to Section~\ref{section:rates}, the proxy has error in both numerator and denominator that may cause issues.
-We again wish with express the statistical error in the $1+\frac{1}{\gamma} \log \left( \frac{y_t}{y_{t-1}} \right) - R_t$ in useful terms.
-Based on Section~\ref{section:rates}, we can re-arrange those terms to show that the error is  given by
+We again wish to express the statistical error of $\hat R_t - R_t$ in useful terms.
+We can re-arrange the error decomposition from Section~\ref{section:rates} to show that this error is given by
 $$
-\frac{1}{\gamma}\log \bigg[ &1 + \rho_{I_t, K_t} D_{M_t} \sqrt{\frac{1-f}{f}} CV(K_t)  \\
-&- \rho_{I_{t-1},K_{t-1}} \sqrt{\frac{1-f}{f}} CV (K_{t-1}) \left[ \rho_{I_t, Y_t} \sqrt{\frac{1-f}{f}} CV(K_{t-1}) + 1 \right] \right].
+\begin{aligned}
+\frac{1}{\gamma}\log &\bigg( 1 + \bigg[ \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t)  -\rho_{I_{t-1},K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t) \bigg] \\
+&\times \left[ 1 - \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t) \right] \bigg)
+\end{aligned}
 $$
-The coefficient of variation $CV(K_t) = (1- \bar Y_t)$ is the fraction of the uninfected population that remain uninfected.
 So here we see the exact same trade-off as with the rates but on the logarithmic scale.
@@
 Instead the error is scaled by the serial interval $\gamma^{-1} = 1/7$.
+Moreover, the error depends not on aggregate prevalence $\bar Y_t$ but fraction of new cases $\bar K_t$.
 
-The above decomposition of the statistical shows us that
-
-The question of interest is ``Is the effective reproduction number below one?'' When this is true, we can feel like that the .
-Therefore, we find the effective reproduction number suffers a similar story. And is consistently underestimated for in initial phases and then we are overconfident in the decline.
+In Figure XXc, we show the error in this quantity.  The trajectory follows the same as before; however, the errors are....
 
@@
 
-In
+Here we show how selection bias can even creep into clinical trial analysis.
+The key concern is whether clinical trials on COVID-19 recruit from the pool of individuals who have tested positive for COVID-19 or whether they sample randomly from the population, test, and then recruit from this subset of tested individuals.
+To see this issue, suppose we have an outcome $Y$, a treatment $A$ and an \emph{unobserved} $U$.
+Suppose treatment is assigned at random, i.e., $A =1$ with probability $50\%$ and $A=0$ with probability $50\%$.
+Suppose the conditional mean of the outcome satisfies  $E(Y \mid U, A ) = \beta_0 + \beta_1 A + \beta_2 U + \beta_3 A U$.
+Typically, we are interested in the \emph{causal effect} of $A$ on $Y$.
+In counterfactual language,
+$$
+E( Y(A=1) - Y(A=0) ) = E_U ( E_A ( Y(A=1) - Y(A=0) \mid U=u ))
+= \beta_1 + \beta_3 E(U).
+$$
+Even in a randomized experiment, the question is what the correct value for $E(U)$ is.
+For COVID-19, the value of interest is the expected value of $U$ in the population of COVID-19 positive individuals.
+Due to selection bias, however, the bias in this marginal treatment effect compared to the true marginal effect of interest is $\beta_1 \cdot E_{\I} [ \rho_{I,U} ] \times \frac{f}{1-f} \times \sigma_U$.
+Selection bias then leads to the estimated effect being biased if $\beta_1 \neq 0$.
+Directionality of the bias will depend on the relationship between $U$ and $Y$.
+If $U$ is a measure of disease severity, then it could be expected to see $\beta_1 > 0$ and therefore treatment effects to be overly optimistic as most individuals recruited are potentially higher on the severity index than the population average.
 
-Random sampling negates effect modifiers.
-Random treatment assignment negates unobserved confounders.
-Show simple example where $E[U]$ is high due to sampling bias and therefore you will think that the effect is significant!
-But it's because of sampling bias.
-Requires poststratification.
+Randomization of treatment assignment in clinical trials negates unobserved confounders.
+This, however, does not negate effect modifiers.
+Therefore, for marginal treatment effects to be interpretable, there must be a well-defined population.
+As our main interest is in causal effects on the population of COVID-19 positive.
+Randomization within the clinical trial design yields internal validity, but we also need external validity to estimate the marginal effect correctly.
 
 \subsection{Cross-country comparisons}
+
+So far we have focused on understanding the disease in a single location and the limitations of using case-count data to understand population quantities of interest.  Many in the
 
@@
 With selection bias, it's not clear that agreement in this quantity is even desirable.
+
+A final question is whether an observed drop in prevalence can be interpreted as a ``real'' decrease in prevalence.  To this end, one would typically construct a Z-score and perform a
 
@@
 \subsection{Stratified setting}
-
-\section{Brief discussion of model-based approaches}
-
-Here we explain a general state-space process model as compared to the measurement-process.  While related, the distinction drives our key results.
-
-Observer controls testing, FP and FN.  Patients
-
-Their behavior may depend on forecasts and therefore the notion of a counterfactual is not well defined.  For example, suppose an highly prominent model....
-
-The transition from susceptible to infected is a complex function.  It is based on personal choices;  the second choice
-$B$
-
-\emph{Non-identifiable} and rely on assumptions.  As per Cox's statement.
-
-This makes models \emph{fragile} and rely on strong data-generating assumptions.  In these settings data integration
 
@@
 where the second equality is obtained by plugging in the statistical decomposition of the error for both time points and the coefficient of variation being defined as $CV(Y) := \sigma_Y/\mu_Y$.
+Under measurement-error, the extra terms $D_{2}$ and $D_1$ can be inserted in the correct locations.
 

### 2020-05-06 write-up/covid-writeup.tex
styles needed

@@
-\documentclass[11pt]{article}
+ \documentclass[11pt]{amsart}
+
+\RequirePackage[OT1]{fontenc}
+
+\usepackage{biblatex}
+%\bibliographystyle{apalike}
+
+\usepackage{tchdr}
+\boldshortcuts
+
 \usepackage{graphicx}
@@
 
-\numberwithin{equation}{section}
-\theoremstyle{plain}
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{example}{Example}
-\newtheorem{remark}{Remark}
-\newtheorem{corollary}[theorem]{Corollary}
-\newtheorem{criterion}[theorem]{Criterion}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{exercise}[theorem]{Exercise}
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{proposition}[theorem]{Proposition}
+% \numberwithin{equation}{section}
+% \theoremstyle{plain}
+% \newtheorem{theorem}{Theorem}[section]
+% \newtheorem{example}{Example}
+% \newtheorem{remark}{Remark}
+% \newtheorem{corollary}[theorem]{Corollary}
+% \newtheorem{criterion}[theorem]{Criterion}
+% \newtheorem{definition}[theorem]{Definition}
+% \newtheorem{exercise}[theorem]{Exercise}
+% \newtheorem{lemma}[theorem]{Lemma}
+% \newtheorem{proposition}[theorem]{Proposition}
 
@@
 
-\subsection{Regrettable rates}
+\subsection{Regrettable rates: complex biases resulting from self-selection}
 \label{section:rates}
@@
 Here we consider the ratio estimators for the relative change in the prevalence rate.
-We let $\bar Y_1$ and $\bar Y_2$ denotes the prevalence at time step $1$ and $2$ respectively (i.e., prevalence on two consecutive days).
-Then the ratio estimator is given by $r = \bar y_1 / \bar y_2$.
+We let $\bar Y_{t-1}$ and $\bar Y_{t}$ denotes the prevalence at time step $t-1$ and $t$ respectively (i.e., prevalence on two consecutive days).
+Then the ratio estimator is given by $r = \bar y_t / \bar y_{t-1}$.
 We assume both numerator and denominator are the prevalence estimates adjusted for measurement-error.
@@
 
-\subsection{Cross-country comparisons}
+\section{Cross-population comparisons}
 
-So far we have focused on understanding the disease in a single location and the limitations of using case-count data to understand population quantities of interest.  Many in the
+So far we have focused on understanding the limitations of using case-count data to understand population quantities of interest for a \emph{single} population (i.e., United States, Germany, France or Michigan, New York, London in isolation).
+Many are interested in cross-population comparisons.
+Many pieces in the media as well as academic articles have plots of case-count over time aligned by time of first known case.
+Others have claimed such comparisons unfair due to unequal population sizes and instead plotted case-count per million.
+In this section, we focus on the statistical issues relevant to such comparisons.
+We do not make any comments on the multi-causal nature of success/failure in countries.
 
-So we can't estimate prevalence.  Can we not get a decent comparison?
+\subsection{Prevalence comparisons}
+
+At a fixed time, suppose prevalence estimates $\bar y_1$ and $\bar y_2$ are observed for two populations. We can express the error $\bar y_1 - \bar y_2$ in statistical terms as
+$$
+(\bar Y_1 - \bar Y_2) + \left[ \rho_{I_1, Y_1} \times \sqrt{\frac{1-f_1}{f_1}} \times \sigma_{Y_1}  \right]
+$$
+
+We could instead think about the number of new cases $K_1$ and $K_2$ within each population.  Then
 $$
@@
 $$
+
+\subsection{Rate comparisons}
 

### 2020-05-06 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 We assume both numerator and denominator are the prevalence estimates adjusted for measurement-error.
-If the sample size at each time were equal (i.e., $n_1 = n_2 = n$) then this analysis would be equivalent to comparing the increase in observed case-counts (i.e., $\bar y_1/\bar y_2 = y_1/y_2$).
-We wish to express the error $\frac{\bar y_2}{\bar y_1} - \frac{\bar Y_2}{\bar Y_1}$ in statistical terms.
+If the sample size at each time were equal (i.e., $n_1 = n_2 = n$) then this analysis would be equivalent to comparing the increase in observed case-counts (i.e., $\bar y_t/\bar y_{t-1} = y_t/y_{t-1}$).
+We wish to express the error $\frac{\bar y_t}{\bar y_{t-1}} - \frac{\bar Y_{t}}{\bar Y_{t-1}}$ in statistical terms.
 Using a Taylor series approximation, we show that the error can be expressed approximately as
@@
 \begin{aligned}
-\frac{\bar Y_2}{\bar Y_1} &\times \bigg[ \rho_{I_2,Y_2} D_{M_2} \sqrt{\frac{1-f_2}{f_2}} CV (Y_2)  -\rho_{I_1,Y_1} D_{M_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \bigg] \\
-&\times \left[ 1 - \rho_{I_1,Y_1} D_{M_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \right]
+\frac{\bar Y_t}{\bar Y_{t-1}} &\times \bigg[ \rho_{I_t,Y_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (Y_t)  -\rho_{I_{t-1},Y_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \bigg] \\
+&\times \left[ 1 - \rho_{I_{t-1},Y_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \right]
 \end{aligned}
@@
 where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement-error adjustment, and $CV(Y_j) = \sigma_{Y_j}/\bar Y_j$ is the coefficient of variation at time step $j$.
-The magnitude of the error depends on the population-level ratio $\bar Y_2 / \bar Y_1$ so a large decrease in the prevalence rate will have a relatively small error as compared to a large increase in the prevalence rate.
+The magnitude of the error depends on the population-level ratio $\bar Y_{t-1} / \bar Y_{t-1}$ so a large decrease in the prevalence rate will have a relatively small error as compared to a large increase in the prevalence rate.
 The second term represents the ``cancellation'' that the naive analyst is hoping will occur.
@@
 In the case of New York, for example, recent weeks it seems plausible that $\Delta$ (i.e., selection bias) has not changed drastically.
-Figure XX shows the rate of testing remains constant and the total cases relative to the population is small enough such that $f_1 \equiv f_2$ is reasonable.
+Figure XX shows the rate of testing remains constant and the total cases relative to the population is small enough such that $f_t \equiv f_{t-1}$ is reasonable.
 Finally, the measurement-error adjustment is likely to be constant as no new testing procedures have been reported (as far as this author knows).
@@
 We estimate the bias under a range of values for the relative sampling fraction $f_1/f_0$ and a sampling fraction of $f = 0.05$.
-Figure XXb shows the trajectory of $\bar Y_2 / \bar Y_1$.
+Figure XXb shows the trajectory of $\bar Y_t / \bar Y_{t-1}$.
 We see that in each case, the rate of change in prevalence is overestimated prior to the peak and then underestimated afterwards.
@@
 Luis and Ribeiro (2008) show that under a Poisson likelihood, a simple relation between the trajectory of \emph{true new cases} and the effective reproduction number can be derived (see Appendix XX for discussion).
-In particular, under an SIR model, the number of case counts on day $t$ is Poisson distributed with rate $K_{t-1} \exp \left( \frac{1}{\gamma} (R_t - 1) \right)$ where $K_{t-1}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval which is approximately $7$ days for COVID-19 (Sanche, 2020).
+In particular, under an SIR model, the number of case counts on day $t$ is Poisson distributed with rate $K_{t-1} \exp \left( \frac{1}{\gamma} (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval which is approximately $7$ days for COVID-19 (Sanche, 2020).
 Using this, a method of moments estimator of the effective reproduction number at time $t$ can be given by
@@
 
-So far we have focused on understanding the limitations of using case-count data to understand population quantities of interest for a \emph{single} population (i.e., United States, Germany, France or Michigan, New York, London in isolation).
+So far we have focused on understanding the limitations of using case-count data to understand population quantities of interest for a \emph{single} population (i.e., United States, Germany, France or Michigan, New York, London in isolation) at a single moment and over time.
 Many are interested in cross-population comparisons.
@@
 
-At a fixed time, suppose prevalence estimates $\bar y_1$ and $\bar y_2$ are observed for two populations. We can express the error $\bar y_1 - \bar y_2$ in statistical terms as
+At a fixed time, suppose prevalence estimates $\bar y_1$ and $\bar y_2$ are observed for two populations. We can express $\bar y_1 - \bar y_2$ in statistical terms as
 $$
-(\bar Y_1 - \bar Y_2) + \left[ \rho_{I_1, Y_1} \times \sqrt{\frac{1-f_1}{f_1}} \times \sigma_{Y_1}  \right]
+(\bar Y_1 - \bar Y_2) + \left[ \rho_{I_1, Y_1} \times \sqrt{\frac{1-f_1}{f_1}} \times \sigma_{Y_1}  - \rho_{I_2, Y_2} \times \sqrt{\frac{1-f_2}{f_2}} \times \sigma_{Y_2} \right].
 $$
+The first term is the difference in population prevalence (i.e., the quantity of interest).
+If random sampling were performed in each population, $E_{\I} [ \rho_{I_j, Y_j}] = 0$ and so the estimate would be unbiased.
+However, the second term represents the complex error that results from selection bias.
+The particular test of interest is $H_0: \bar Y_1 = \bar Y_2$.  Under the null, $\sigma_{Y_1} = \sigma_{Y_2}$.  In classical statistics, we would compute a Z-score:
+$$
+\frac{\bar y_1 - \bar y_2}{ V_{SRS} (\bar y_1 - \bar y_2) } =
+\frac{\rho_{I_1, Y_1} \times \sqrt{\frac{1-f_1}{f_1}}  - \rho_{I_2, Y_2} \times \sqrt{\frac{1-f_2}{f_2}} }{ \sqrt{\frac{1}{N_1 -1} \times \frac{1-f_1}{f_1} + \frac{1}{N_2 - 1} \times \frac{1-f_2}{f_2}} } \overset{\star}{=}
+\sqrt{\frac{(N_1-1)(N_2 -1)}{N_1 + N_2}} \left[ \rho_{I_1, Y_1} - \rho_{I_2, Y_2} \right]
+$$
+Where the last equality holds under $f_1 = f_2$, i.e., same sampling rate relative to population size.
+This tells us that the exact error in the comparison of sample means, as an estimate of the difference in population means, scales as a function of the difference in data quality $\rho_{I_1, Y_1} - \rho_{I_2, Y_2}$ times a square root of a population adjustment.
+For highly unequal population sizes (e.g., $N_1 \gg N_2$), the adjustment is approximated well by the square root of the smaller population size minus one (e.g.
+$\sqrt{N_2 -1}$).
+
+While problematic, this points to a potential setting in which comparisons are still valid: when sampling fraction~$f_1 = f_2$ are equal and data quality are equal.
+The issue is how to assess data quality equivalence among populations.
+\textcolor{red}{Discuss the measures in the literature.}
+
+\subsection{Case count comparisons}
 

### 2020-05-06 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 $$
-\frac{\bar y_1 - \bar y_2}{ V_{SRS} (\bar y_1 - \bar y_2) } =
-\frac{\rho_{I_1, Y_1} \times \sqrt{\frac{1-f_1}{f_1}}  - \rho_{I_2, Y_2} \times \sqrt{\frac{1-f_2}{f_2}} }{ \sqrt{\frac{1}{N_1 -1} \times \frac{1-f_1}{f_1} + \frac{1}{N_2 - 1} \times \frac{1-f_2}{f_2}} } \overset{\star}{=}
+\begin{aligned}
+\frac{\bar y_1 - \bar y_2}{ V_{SRS} (\bar y_1 - \bar y_2) } &=
+\frac{\rho_{I_1, Y_1} \times \sqrt{\frac{1-f_1}{f_1}}  - \rho_{I_2, Y_2} \times \sqrt{\frac{1-f_2}{f_2}} }{ \sqrt{\frac{1}{N_1 -1} \times \frac{1-f_1}{f_1} + \frac{1}{N_2 - 1} \times \frac{1-f_2}{f_2}} }  \\
+&\overset{\star}{=}
 \sqrt{\frac{(N_1-1)(N_2 -1)}{N_1 + N_2}} \left[ \rho_{I_1, Y_1} - \rho_{I_2, Y_2} \right]
+\end{aligned}
 $$
@@
 
-While problematic, this points to a potential setting in which comparisons are still valid: when sampling fraction~$f_1 = f_2$ are equal and data quality are equal.
+\textcolor{red}{Take, for example, a comparison of the US and Canada.
+The two population sizes are approximately 328 million and 38 million people.
+Then the population adjustment is $5.84$ million.
+As of May 5th, Canada has performed a total of 24.92 tests per 1,000 people while the US has performed a total of 22.01 tests per 1,000 people (cite: https://ourworldindata.org/grapher/full-list-cumulative-total-tests-per-thousand)} so the data quantities are approximately equal to $f = 23/1000 = 0.023$.
+Then, under a prevalence of 10\%, one would need the differential $\Delta_1-\Delta_2$ to be much smaller than $11.68 \times 10^{-6}$ in order for the selection bias to not impact.
+So only under the very strong assumptions of equal data quantity and quality will the comparisons be valid and the $Z$-score can be treated as in the classical setting.
 The issue is how to assess data quality equivalence among populations.
-\textcolor{red}{Discuss the measures in the literature.}
+\textcolor{red}{Discuss the measures in the literature.}  If this does not hold, then the width of the CIs need to be multipled by
+
+\textcolor{red}{What is adjustment to CIs}.   Inserting the effective sample sizes can give us a better sense for a appropriate $Z$-score.  For example, using $n_eff = 20$ for each. And then
+$$
+Z_n = \frac{\bar y_1 - \bar y_2}{ \frac{1-f}{f} \sigma_y \sqrt{ \frac{1}{n_1 - 1} + \frac{1}{n_2 - 1}}}.
+$$
+So we see that we cannot reject the null. \textcolor{red}{We can do the same for different rates as well.}  This is not perfect but more appropriate.
 
@@
 
-\section{Testing strategies and population size: a tango}
+An alternative is the comparison on the rates of case-count change over time.
+Here, for simplicity, we focus on comparing the estimated effective reproductive rate.
+We assume the two time-series are aligned so that $t=0$ is the time of first .
+This negates the alignment issue and is common in the plots.
 
-Data analysts have argued whether
+While the above issues on $Z$-scores and effective sample size are important, here we highlight a separate issue.
+Namely, the interplay of bias when two trajectories when the peaks are slightly off.
+We show that the biases can make us think the gap between two similar trajectories is larger than we would like!!!!
 
-Increasing testing under test rationing lowers correlation.  Example in Michigan and try and find CA and NY numbers.  Then show that this may cause higher FPs (cite antibody studies).
-
-Plot Michigan tests per million and positives.  Next to NY.  Both as of April moved to prioritized sampling for symptomatic individuals.
-
-\subsection{US testing: a ration system}
-
-There is a lot of hidden information in the selection indicator $I_j$.
-Up until recently, an individual had to first \emph{want} to be tested, then call their primary health care provider who decided whether they met thresholds for testing
-
-. Currently, we are unaware of reliable information on the number of cases where
-
-In the UU
-We assume that the percentage of tests is rationed according to severity of symptoms.
-
-Note that a person who displays severe symptoms may be COVID negative.  This gives us a sense of the data quality.
-
-\emph{Increasing testing leads to lower correlation in}
-
-The CDC reports 413,867 total tests performed in the US as of April 25th.
+$$
+R_{t1} - R_{t2}
+$$
 
@@
 
+Comparison of states with similar $f$ and similar testing inclusion protocols.  Not perfect.
+
 We estimate $\rho$ using old and new methods.  Current SRS in US are weak.  So we just use my back of the envelope calculation. Tests per million as proxy since everyone is doing rationing. Cluster.
-\subsection{Testing increase changes data quality}
-
-Many (cites) hve
-
-\emph{Increasing testing leads to lower correlation only if we see the gap $\Delta$ decrease.  If we just see $f$ }
 
@@
 
-\section{Conclusion}
+\section{Discussion}
+
+In current crisis, data quality is tied to data quantity due to rationing system. \emph{Testing }
+Increasing testing under test rationing lowers correlation.  Example in Michigan and try and find CA and NY numbers.  Then show that this may cause higher FPs (cite antibody studies).
+
+Plot Michigan tests per million and positives.  Next to NY.  Both as of April moved to prioritized sampling for symptomatic individuals.
+
+There is a lot of hidden information in the selection indicator $I_j$.
+Up until recently, an individual had to first \emph{want} to be tested, then call their primary health care provider who decided whether they met thresholds for testing
+
+. Currently, we are unaware of reliable information on the number of cases where
+
+In the UU
+We assume that the percentage of tests is rationed according to severity of symptoms.
+
+Note that a person who displays severe symptoms may be COVID negative.  This gives us a sense of the data quality.
+
+\emph{Increasing testing leads to lower correlation only if we see the gap $\Delta$ decrease.  If we just see $f$ }
+
+The CDC reports 413,867 total tests performed in the US as of April 25th.
 

### 2020-05-07 write-up/covid-writeup.tex
yes

@@
 $$
-(\bar Y_1 - \bar Y_2) + \left[ \rho_{I_1, Y_1} \times \sqrt{\frac{1-f_1}{f_1}} \times \sigma_{Y_1}  - \rho_{I_2, Y_2} \times \sqrt{\frac{1-f_2}{f_2}} \times \sigma_{Y_2} \right].
+(\bar Y_1 - \bar Y_2) + \left[ \rho_{I_1, Y_1} \times D_{M_1} \times \sqrt{\frac{1-f_1}{f_1}} \times \sigma_{Y_1}  - \rho_{I_2, Y_2} \times D_{M_2} \times \sqrt{\frac{1-f_2}{f_2}} \times \sigma_{Y_2} \right].
 $$
@@
 \frac{\bar y_1 - \bar y_2}{ V_{SRS} (\bar y_1 - \bar y_2) } &=
-\frac{\rho_{I_1, Y_1} \times \sqrt{\frac{1-f_1}{f_1}}  - \rho_{I_2, Y_2} \times \sqrt{\frac{1-f_2}{f_2}} }{ \sqrt{\frac{1}{N_1 -1} \times \frac{1-f_1}{f_1} + \frac{1}{N_2 - 1} \times \frac{1-f_2}{f_2}} }  \\
+\frac{\rho_{I_1, Y_1} \times D_{M_1} \times \sqrt{\frac{1-f_1}{f_1}}  - \rho_{I_2, Y_2} \times D_{M_2} \times \sqrt{\frac{1-f_2}{f_2}} }{ \sqrt{\frac{1}{N_1 -1} \times \frac{1-f_1}{f_1} + \frac{1}{N_2 - 1} \times \frac{1-f_2}{f_2}} }  \\
 &\overset{\star}{=}
@@
 We show that the biases can make us think the gap between two similar trajectories is larger than we would like!!!!
-
 $$
-R_{t1} - R_{t2}
+R_{t1} - R_{t2} = \frac{1}{\gamma} \log \left( \frac{1 + e_{1t}}{1+e_{2t}} \right)
 $$
+where $e_{jt}$ is the error associated given in Section XX.
+Recall that the error allowed for over-estimation prior to the peak and then under-estimation post-peak.
+Here, these errors can mingle in interesting ways.
+Consider two countries (A and B) in which the peak occurs 2 weeks prior for country A then country B but other.
+Figure XX presents 1 week prior to the peak of A up to one week after the peak for country B.
+Data for each country is simulated from an SIR model as described in Section XX.
+We see that the errors interact in complex ways.
+In week 1, the two countries appear the same; then in week 2, the gap between country A and B enlarges significantly as country A seems to see rapid improvement; then in week 3, country B improves and actually catches up to country A; then in month 4, country A sees rapid improvement while country B appears worse.
+We are not claiming this is always the case, and indeed for small enough selection bias and/or large gaps in case-counts such complex behavior will not occur; however, this is simpy to point out how observed information can tell a more complex story than the truth (i.e., country A recovers 1 month prior to country B).
 

### 2020-05-07 methods/sir_example.R

@@ -1,0 +1,103 @@
+## Load deSolve package
+library(deSolve)
+
+## Create an SIR function
+sir <- function(time, state, parameters) {
+  
+  with(as.list(c(state, parameters)), {
+    
+    dS <- -beta * S * I
+    dI <-  beta * S * I - gamma * I
+    dR <-                 gamma * I
+    
+    return(list(c(dS, dI, dR)))
+  })
+}
+
+### Set parameters
+## Proportion in each compartment: Susceptible 0.999999, Infected 0.000001, Recovered 0
+init       <- c(S = 1-1e-6, I = 1e-6, R = 0.0)
+## beta: infection parameter; gamma: recovery parameter
+# parameters <- c(beta = 1.4247, gamma = 0.14286)
+parameters1 <- c(beta = 1.2, gamma = 0.15)
+parameters2 <- c(beta = 0.9, gamma = 0.15)
+## Time frame
+times      <- seq(0, 50, by = 1)
+
+## Solve using ode (General Solver for Ordinary Differential Equations)
+out1 <- ode(y = init, times = times, func = sir, parms = parameters1)
+out2 <- ode(y = init, times = times, func = sir, parms = parameters2)
+## change to data frame
+out1 <- as.data.frame(out1)
+out2 <- as.data.frame(out2)
+## Delete time variable
+out1$time <- NULL
+out2$time <- NULL
+## Show data
+png(filename = "C:/Users/wdem/Documents/GitHub/covid-umich/methods/figs/sir.png",
+    width = 480, height = 480, units = "px", pointsize = 12)
+par(mar = c(1.5,2,1,1)+0.1)
+plot(times, out1$I, type = "l", axes = FALSE, bty = "n")
+lines(times, out2$I, col = "red")
+axis(side = 1, labels = FALSE)
+axis(side = 2, labels = FALSE)
+mtext("Fraction Infected", side = 2, line = 1, cex = 0.75)
+dev.off()
+
+## Setup for bias plots
+f = 0.02
+M = 4.75
+days = 1:365
+bary1 = -diff(out1$S)
+bary2 = -diff(out2$S)
+
+bias_fn <- function(M, bary, f, FP, FN) {
+  const = (bary*M + (1-bary))
[…]

### 2020-05-07 write-up/covid-writeup.tex
SIR updates

@@
 \begin{document}
-\title{The Hypothesis of Testing: Paradoxes arising out of reported coronavirus case-counts}
+\title[The Hypothesis of Testing]{The Hypothesis of Testing: Paradoxes arising out of reported coronavirus case-counts}
 % \author{Walter Dempsey\thanks{Department of Biostatistics, University of Michigan, 1415 Washington Heights, U.S.A. E-mail: wdem@umich.edu}}
@@
 
-\begin{wrapfigure}{r}{0.5\textwidth}
-\centering
-\includegraphics[width = 0.4\textwidth]{../methods/figs/mem_heatmap_article.png}
-\caption{Measurement-error data quality adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.15$ and $FN=0.10$.
-Color scaled so blue = $-6$, white = 0, and red = $6$.}
-\vspace{-0.7cm}
-\end{wrapfigure}
-
 Under the same numbers, $D_M$ is approximately 1.7.  The effective sample size can then be bounded by
@@
 In a worst case scenario, if $FN \to 0.3$ and $FP \to 0.05$ then the factor is only 1.3.
+
+\begin{wrapfigure}{r}{0.5\textwidth}
+\centering
+\includegraphics[width = 0.4\textwidth]{../methods/figs/mem_heatmap_article.png}
+\caption{Measurement-error data quality adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.15$ and $FN=0.10$.
+Color scaled so blue = $-6$, white = 0, and red = $6$.}
+\vspace{-0.3cm}
+\end{wrapfigure}
 
@@
 In Figure XXc, we show the error in this quantity.  The trajectory follows the same as before; however, the errors are....
+
+\begin{figure}
+\centering
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{../methods/figs/sir_ratio.png}
+  \caption{A subfigure}
+  \label{fig:sub1}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{../methods/figs/sir_rt.png}
+  \caption{A subfigure}
+  \label{fig:sub2}
+\end{subfigure}
+\caption{A figure with two subfigures}
+\label{fig:test}
+\end{figure}
 
@@
 
+\begin{figure}
+\centering
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{../methods/figs/sir.png}
+  \caption{A subfigure}
+  \label{fig:sub1}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{../methods/figs/sir_rt_comparison.png}
+  \caption{A subfigure}
+  \label{fig:sub2}
+\end{subfigure}
+\caption{A figure with two subfigures}
+\label{fig:test}
+\end{figure}
+
 \subsection{Estimation of Data Quality}
@@
 Comparison of states with similar $f$ and similar testing inclusion protocols.  Not perfect.
+
+\[
+\rho = .004 \Rightarrow \Delta = 0.035 \times \sqrt{\frac{f \cdot (1-f)}{\bar Y \cdot (1 - \bar Y) }} = 0.00224 \Rightarrow M =
+\]
 

### 2020-05-07 methods/sir_example.R

@@ -35,7 +35,7 @@ out1$time <- NULL
 out2$time <- NULL
 ## Show data
 png(filename = "C:/Users/wdem/Documents/GitHub/covid-umich/methods/figs/sir.png",
-    width = 480, height = 480, units = "px", pointsize = 12)
+    width = 6.5, height = 4, units = "in", res = 1200, pointsize = 12)
 par(mar = c(1.5,2,1,1)+0.1)
 plot(times, out1$I, type = "l", axes = FALSE, bty = "n")
 lines(times, out2$I, col = "red")
@@ -46,18 +46,20 @@ dev.off()
 
 ## Setup for bias plots
 f = 0.02
-M = 4.75
+M = 4
 days = 1:365
+FP = 0.01; FN = 0.10
 bary1 = -diff(out1$S)
 bary2 = -diff(out2$S)
 
-bias_fn <- function(M, bary, f, FP, FN) {
+bias_fn <- function(M, bary, f, FP = 0.01, FN = 0.15) {
   const = (bary*M + (1-bary))
   f0 = f/const
   f1 = M*f0
   Delta = f1 - f0
+  D_M = 1 + FP + FN - Delta * bary/(1-bary) * (FP*(1-bary) + FN * bary)/ (f0 * (1-bary) + f1 * bary)
   rho = Delta*sqrt(bary*(1-bary))/sqrt(f*(1-f))
-  error = rho*sqrt((1-f)/f) * sqrt(1-bary)/sqrt(bary)
+  error = rho*sqrt((1-f)/f) * sqrt(1-bary)/sqrt(bary)*D_M
   lag_error = error[1:(length(error)-1)]
   cur_error = error[2:(length(error))]
   
@@ -89,15 +91,51 @@ lines(bias2$r+bias2$bias, col = "red", lty = 2)
 
 ymax = max(bias1$Rt+bias1$logbias, bias1$Rt, bias2$Rt+bias2$logbias, bias2$Rt, na.rm = TRUE)
 ymin = min(bias1$Rt+bias1$logbias, bias1$Rt, bias2$Rt+bias2$logbias, bias2$Rt, na.rm = TRUE)
-png(filename = "C:/Users/wdem/Documents/GitHub/covid-umich/methods/figs/sir_rt.png",
-    width = 480, height = 480, units = "px", pointsize = 12)
-par(mar = c(1.5,3,1,1)+0.1)
[…]

### 2020-05-08 write-up/covid-writeup.tex
draft so close

@@
 Color scaled so blue = $-6$, white = 0, and red = $6$.}
+\label{fig:heatmap}
 \vspace{-0.3cm}
@@
   \includegraphics[width=.9\linewidth]{../methods/figs/sir_ratio.png}
-  \caption{A subfigure}
+  \caption{Ratio estimator}
   \label{fig:sub1}
@@
   \includegraphics[width=.9\linewidth]{../methods/figs/sir_rt.png}
-  \caption{A subfigure}
+  \caption{Effective reproductive rate estimator}
   \label{fig:sub2}
 \end{subfigure}
-\caption{A figure with two subfigures}
+\caption{Potential bias due to self-selection and measurement-error from a deterministic SIR model with $\beta = 1.4$ and $\gamma = 0.2$.
+Here we assume $f = 0.02$, $FP = 0.01$, $FN = 0.15$, and use a range of relative sampling fraction $M = f_1/f_0$.}
 \label{fig:test}
@@
   \includegraphics[width=.9\linewidth]{../methods/figs/sir.png}
-  \caption{A subfigure}
+  \caption{Fraction of new cases in population}
   \label{fig:sub1}
@@
   \includegraphics[width=.9\linewidth]{../methods/figs/sir_rt_comparison.png}
-  \caption{A subfigure}
+  \caption{Effective reproduction rate estimators}
   \label{fig:sub2}
 \end{subfigure}
-\caption{A figure with two subfigures}
+\caption{Left: fraction infected in two SIR models with $\beta = 1.4$ and $0.9$ respectively and $\gamma = 0.2$ with same initial conditions.
+Right: comparison of $\hat R_t$ across time with $FN = 0.2$, $FP = 0.01$, and $M = 4$.}
 \label{fig:test}
@@
 
+In prior sections, we have discussed the impact of selection bias and measurement-error on a range of outcomes of interest in COVID-19 research.
+Here, we try and address the empirical question of how to estimate the data quality $E_{\I} \left[ \rho_{I,Y} \right]$ in the current context.
+In Meng (2019), estimation of the data quality relied on observing the true outcome.
+In election data, for example, we observe the voting distribution and can use this observation as ground truth.
+We are not so lucky in the current crisis.
+Here, we propose a simple procedure that uses survey samples as noisy but unbiased estimates of the true prevalence.
+Our goal is less inferential and more to provide a sensitivity analysis that can aid in understanding the amount of information in observational COVID case-count data.
+
+On April 23rd, Andrew Cuomo announced results from a study in New York state.
+It found 13.9\% of 3,000 people tested across the state had signs of the virus.
+The study did not report sensitivity and specificity; therefore, we take reported measures from the Santa Clara study (cite) \footnote{We recognize the debate regarding its conclusions.
+Here we are only using this study for its information regarding measurement-error rates.}.
+The reported specificity is $82.8\%$ (95\% CI: 76.0-88.4\%) and sensitivity is $99.5\%$ (95\% CI: 99.2-99.7\%).
+This corresponds to a false negative rate of $17.2\%$ (11.6-24.0\%) and false positive rate of $0.5\%$ (0.3 - 0.8\%).
+Correcting for these rates we have an estimated prevalence of $15.9\%$ with a range of 14.8\% to 17.0\% using the ranges of the false positive and negative rates.
+
+Figure XXa plots the number of new total and positive tests per day.
+Figure XXb plots the fraction of tests that were positiver per day.
+In both cases, we see smoothed these plots as the spiky pattern is likely due to testing backlogs.
+Given the report was on the 23rd, we suppose for simplicity that the study was run several days prior on the 20th of April.
+On the 20th, there were $19,654$ tests performed and $617,555$ tests had been performed up to to and including the 19th.
+The population of New York state is $19,378,102$.
+Subtracting off the number of individuals who had already been tested yields a sampling fraction of $f = 0.001$.
+The smoothed prevalence estimate on the 20th is 28.3\%.
+Adjusting for measurement-error yields an estimate of $32.5\%$.
+
+Here, we assume the New York survey is a simple random sample from the state's population.
 Comparison of states with similar $f$ and similar testing inclusion protocols.  Not perfect.
@@
 \[
-\rho = .004 \Rightarrow \Delta = 0.035 \times \sqrt{\frac{f \cdot (1-f)}{\bar Y \cdot (1 - \bar Y) }} = 0.00224 \Rightarrow M =
+\rho D_M = \sqrt{\frac{f}{1-f}} \frac{\text{0.166}}{\sigma_Y} = 1.43 \times 10^{-2} \Rightarrow \Delta = 1.06 \times 10^{-3} \Rightarrow M = 2.29.
 \]
+The range for the CIs is (2.05, 2.54).
+We can also perform sensitivity as a function of the false negative and positive rates.
+Under these measurement-error ranges, we saw that the impact on the effective reproductive rate estimator was much smaller.
+However, for $FP = 0.05$ and $FN = 0.005$ we have $M = 4.31$.
+While nothing above is definitive, it points to the interplay of selection bias and measurement-error.
+The outlined procedure gives the community tools to understand how sensitivity and specificity impact our estimators and therefore how much we should trust conclusions based on these assumptions.
+Under the Santa Clara study estimates of these quantities, while we expect impact, the bias is much smaller than under other estimates.
 
-We estimate $\rho$ using old and new methods.  Current SRS in US are weak.  So we just use my back of the envelope calculation. Tests per million as proxy since everyone is doing rationing. Cluster.
-
-\subsection{Stratified setting}
+\begin{figure}
+\centering
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{../methods/figs/NY_casecount.png}
+  \caption{NY new cases per day}
+  \label{fig:sub1}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{../methods/figs/NY_fracpos.png}
+  \caption{NY fraction of new tests positive per day}
+  \label{fig:sub2}
+\end{subfigure}
+\caption{New York COVID-19 reported data }
+\label{fig:test}
+\end{figure}
 
@@
 
-Sampling network is hard.  Sampling high-risk strata is easy.  Simple example.
-DTR and network connectivity is useful for understanding spread and where "testing should go".
+This article argues why random sampling is a powerful tool and the limitations of data obtained through a self-selection process.
+Few COVID-19 studies (cite) have been released that aim to build representative samples.
+One difficulty in the current crisis is the low prevalence relative to the population in many areas.
+Take the recent Santa Clara study, for example.
+The key issue with low prevalence is that if the test's specificity is not lower than the prevalence, then the observed data is consistent with zero recovered individuals.
 
-Huge benefits when the rates are very different.
+In such settings, sampling at random is doomed; however, there are solutions to such settings.
+Stratify the population into risk categories (based on population density, occupation, age, and other important factors) and perform Neyman allocation (i.e., sample strata $h$ with $P_h \sqrt{\bar Y_h \times (1-\bar Y_h)}$ where $P_h$ is the strata frequency in the population and $\bar Y_h$ the prevalence in this.
+Of course, $P_h$ and $\bar Y_h$ are not known a priori; however, if we can get decent esitmates.
+\textcolor{red}{Huge benefits when the rates are very different.}
+
+One could go further.
+\textcolor{explain network stuff}.
+However, network information is difficult to obtain.
+Geographic location, occupation, age and basic demographic information are much easier to obtain and are a good proxy for the latent network connectivity.
 
@@
 
-Above we point to flaws in using observed data to reason about the
+Having read through the above technical discussions, one may come to the incorrect conclusion that because we are wary of trusting the observational data, we must wait for better data sources in order to act (Ionides cite).
+This conclusions mistakes data and information.
+Lack of high quality data means we should be skeptical regarding conclusions drawn from these data sources.
+There are whole fields dedicated to careful analysis in such settings (e.g., causal inference from observational studies).
+Being a data skeptic, does not mean one must wait for hgih quality data to act.
+One must wait for high quality information.
+Yet, in teh current crisis, we have just that.
+Indeed, epidemiologists and public health officials have a strong understanding of the basic facts of pandemics and disease spread.
+Social distancing and
+
+Moreover, the risks of interventions to por risk   the data and therefore
+Of course, there are tradeoffs and long-term stay-at-home orders may have negative economic consequences.
+This is beyond the scope of the current article.
+While not the main thrust of this article, we are well aware that discussions of data quality can sometimes be misinterpreted as claims that we do not have the requisite information to act immediately and decisively.
+It is that point that we are refuting.
+It is simply not the case.
 
@@
 
-A slightly altered version of Nietzsche's famous quote has become a mantra for post-pandemic thinking: \emph{That which which does not break us, makes us stronger}.
-While potentially true via viral resistance, it is not clear that governments have yet to learn lessons on pandemic response.
+A slightly altered version of Nietzsche's famous quote has become a mantra for post-pandemic thinking: \emph{that which which does not break us, makes us stronger}.
+While a potentially true statement with respect to viral resistance, it is not clear that governments have yet to learn lessons on pandemic response.
+Governments are pushing increased testing capacity and contact tracing.
+Contact tracing is incredibly useful for identifying carriers early and preventing further spreading.
+I would argue that, after the current wave, understanding prevalence is also key.
+We know that the probability of an outbreak is a function of current prevalence and network connectivity.
+Within the SIR framework, there is a critical threshold for  prevalence below which the chance of a full-scale outbreak is near-zero and above the likelihood jumps rapidly (cite).
+Therefore, knowing prevalence will help governments determine long-term community-level risk and allow for more targeted interventions -- shutting down only certain locations when necessary.
+Once we ``return to normal''\footnote{or at least the new normal}, it is clear that testing strategies should include a focus on early detection to help head off future outbreaks.
 
-During the current wave, understanding prevalence is key.  It helps us determine long-term risk for a community and make targeted interventions.  This will allow
-When prevalence is below a certain threshold, we can return to daily life.
-
-Once we ``return to normal''\footnote{or at least the new normal}, it is clear that testing strategies should focus on early detection, heading off future outbreaks.
-\emph{Anyone who wishes to go back to work}.
-While laudable, without complete compliance, we may be riddles with selection bias and measurement-error.
-This ignores even the practical and ethical quandaries of how to .
-
-What do we do when we are faced with?  We design experiments with our objectives in mind.
-
-(A) Prevalence: Targeted shutdowns
-(B) Risk Detection: Contact tracing and altering
-
-Use prior results to show that ratio estimator under SRS is a good predictor of potential outbreak.
-
-``Disease free''
+Governments implicitly argue that increased testing capacity will perform this task.
+Without complete compliance, however, our understanding of future outbreaks may be plagued by self-selection bias, compounded by changing sensitivity and specificity rates.
+Random testing removes these effect modifiers, giving governments more information to fight the disease.
+This is not to say random testing should replace current testing practices; only that the random testing should supplement and will provide valuable information even in small numbers.
 
@@
 
-In current crisis, data quality is tied to data quantity due to rationing system. \emph{Testing }
-Increasing testing under test rationing lowers correlation.  Example in Michigan and try and find CA and NY numbers.  Then show that this may cause higher FPs (cite antibody studies).
+There is nothing routine about COVID-19.  The resulting statistical questions are
+, ``routine statistical questions are less common than questionable statistical routines.''
 
-Plot Michigan tests per million and positives.  Next to NY.  Both as of April moved to prioritized sampling for symptomatic individuals.
+In current crisis, data quality is often tied to data quantity due to rationing system.
+Increasing testing under test rationing lowers correlation.
+There is a lot of hidden information in the selection indicator $I_j$.
+Up until recently, an individual had to first \emph{want} to be tested, then call their primary health care provider who decided whether they met thresholds for testing.
+Currently, we are unaware of reliable information on the number of cases where.
 
-There is a lot of hidden information in the selection indicator $I_j$.
-Up until recently, an individual had to first \emph{want} to be tested, then call their primary health care provider who decided whether they met thresholds for testing
+We show how one can build sensitivity analyses to understand the potential bias in the observational data.
+These biases depend significantly on the interaction between self-selection and measurement-error.
 
-. Currently, we are unaware of reliable information on the number of cases where
-
-In the UU
-We assume that the percentage of tests is rationed according to severity of symptoms.
-
-Note that a person who displays severe symptoms may be COVID negative.  This gives us a sense of the data quality.
-
-\emph{Increasing testing leads to lower correlation only if we see the gap $\Delta$ decrease.  If we just see $f$ }
+Interventions in this area are designed experiments.  Unlike Fischer's null, here the objectives are settings, statisticians.
 
 The CDC reports 413,867 total tests performed in the US as of April 25th.
-
-Use of these bounds as E-scores.  How bad can it go?
-
-Interventions in this area are designed experiments.  Unlike Fischer's null, here the objectives are settings, statisticians.
 
@@
 A critical question is whether there are alternative data streams that may be leveraged to understand the handling of the COVID-19 pandemic in the US.
+
+Extensions of the SIR models can try and capture these issues; however, without strong assumptions on the selection mechanism, the estimates are often not nonparametrically identifiable.
+As Cox and Hinckley (XXXX, p.
+96) identified, when an issue ``cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically.'' In the absence of randomized experiments, the best route forward are detailed sensitivity analyses and humility in conclusions drawn from such data sources.
 
@@
 
-\section{Quotes}
+% \section{Quotes}
 
-I prefer to think of a statistical sensibility rather than statistical thinking.
+% I prefer to think of a statistical sensibility rather than statistical thinking.
 It’s “less than an agenda but more than an attitude.”  It allows for methodological preference while avoiding dogma.
@@
 
-“Routine statistical questions are less common than questionable statistical routines...” McCullagh (2005).
-
-If an issue can be addressed nonparametrically then it will often be better to tackle it parametrically; however, if it cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically.” (p.96)
-
-A test of meaningfulness of a possible model for a data-generating process is whether it can be used directly to simulate data.” (p.104).
+% A test of meaningfulness of a possible model for a data-generating process is whether it can be used directly to simulate data.” (p.104).
 In our current setting, this most certainly related to simulation while accounting for measruement error.

### 2020-05-08 methods/ny_calc.R

@@ -1,0 +1,27 @@
+FP = 0.01; FN = 0.005
+bary0 = 0.139
+bary = bary0*(1+FN) - FP*(1-bary0)
+f= 0.001
+true_error = (0.325-bary)/sqrt(bary*(1-bary)) * sqrt(f/(1-f))
+
+error <- function(FP, FN, bary, f, true_error) {
+  delta_error <- function(Delta) {
+    odds = bary / (1-bary)
+    num = FP*(1-bary) + FN*bary
+    f0 = f - Delta * bary
+    f1 = f0 + Delta
+    den = f0 *(1-bary) +  f1 * bary
+    answer = Delta * sqrt(bary*(1-bary))/sqrt(f*(1-f)) * (1+FP+FN - Delta * odds * num/den )
+    return(abs(answer-true_error))
+  }
+  return(delta_error)
+}
+
+temp = error(FP,FN,bary,f, true_error)
+temp(0.1)
+
+output = optim(Delta, temp, method = "Brent", lower = 0, upper = f/bary)
+
+f0 = f-output$par * bary
+f1 = f0 + output$par
+f1/f0

### 2020-05-08 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
- \documentclass[11pt]{amsart}
+\documentclass[aoas]{amsart}
 
@@
 
+\usepackage[foot]{amsaddr}
 \usepackage{biblatex}
@@
 \usepackage{amsmath, amsthm, amsfonts}
-\usepackage[margin=1.5in]{geometry}
+% \usepackage[margin=1.5in]{geometry}
 
@@
 \begin{document}
-\title[The Hypothesis of Testing]{The Hypothesis of Testing: Paradoxes arising out of reported coronavirus case-counts}
-% \author{Walter Dempsey\thanks{Department of Biostatistics, University of Michigan, 1415 Washington Heights, U.S.A. E-mail: wdem@umich.edu}}
-\maketitle
+\title[The Hypothesis of Testing]{The Hypothesis of Testing: Paradoxes arising out of reported coronavirus case-counts} %
+
+\author{Walter Dempsey}
+\address{Department of Biostatistics, University of Michigan, Ann Arbor, MI 48109}
+% \email{wdem@umich.edu}
 
@@
 False positive/negative rates interact in complex ways with selection bias.
-This note attempts to clarify this interaction.
+In this paper, we attempt to clarify this interaction.
 Through simple calculations, we demonstrate pitfalls and paradoxes that can arise when considering case-count data in the presence of selection bias and measurement-error.
@@
 \end{abstract}
+
+\maketitle
 
@@
 
-Despite these necessary steps, the data landscape for understanding COVID-19 remains limited.
-Public databases maintained by Johns Hopkins University (\url{https://bit.ly/2UqFSuA}) and the New York Times (\url{https://bit.ly/2vUHfrK}) provide incoming county-level information of confirmed cases and deaths.
-Statisticians, epidemiologists, economists, and data scientists the world over have been using this granular data to build model-based forecasts of future COVID-19 case-counts, deaths, and hospitalizations.
-Variations on the SIR models have been used to draw inferences and build forecasts.
-In most cases, parameter inference uses observed case-count and/or observed COVID-19 related deaths to infer latent trajectories of the pandemic (cites).
+Despite these steps, the data landscape for understanding COVID-19 remains limited.
+Public databases maintained by \href{https://bit.ly/2UqFSuA}{Johns Hopkins University} (cite) and the \href{https://bit.ly/2vUHfrK}{New York Times} (cite) provide incoming county-level information of confirmed cases and deaths.
+Statisticians, epidemiologists, economists, and data scientists have all used this granular data to forecast COVID-19 case-counts, deaths, and hospitalizations (cite).
+Variations on the SIR models (cite) have been used to draw inferences about infection rates and intensity.
+Parameter inference most often uses observed case-count and/or observed COVID-19 related deaths to infer latent trajectories of the pandemic (cites).
 
-The goal of this paper is to express reservations at the use of case-counts as a proxy for prevalence and disease trajectory as well as its use as direct input into estimation of standard epidemiological models for inference and forecasting.
-The reason is simple: current models ignore selection bias and measurement-error.
-
+The goal of this paper is to express reservations at the use of case-counts as a proxy for disease prevalence and intensity as well as its use as direct input into estimation of standard epidemiological models for inference and forecasting.
+The reason is simple: current models do not account simultaneously for selection bias and measurement-error.
 Selection bias enters due to differences in testing strategies across countries and states.
 In the US, for example, limited testing capacity has caused local and state health departments to focus on testing only high-risk populations.
-Testing requires the individual to self-select into testing.
-Demands for testing increases, while laudable, ignore the issues of self-selection and measurement-error.
+Moreover, testing requires the individual to self-select into receiving the test.
+Demands for increased testing capacity, while laudable, ignore the issues of self-selection and measurement-error.
 While increasing testing capacity increases data quantity, there is no guarantee for increased data quality.
-We will show that aggressive pushes for ramped up testing capacity that are tied to decreases in data quality may have a deleterious effect in our ability to estimate quantities of interest such as prevalence and effective reproduction number.
+We will show that aggressive pushes for ramped up testing capacity that are tied to decreases in data quality may have a very mitigated impact  in our ability to estimate quantities of interest such as prevalence and effective reproduction number.
 
 In this paper, we will demonstrate the complex interaction between these two fundamental concepts.
-Measurement-error is often associated with parameter attenuation; howeve,r we show that the combination with selection effects can cause bias in parameter estimates to change sign and increase/decrease in magnitude.
+Measurement-error is often associated with parameter attenuation; we show, however, that the combination with selection effects can cause bias in parameter estimates to change sign and increase/decrease in magnitude.
 Moreover, the bias may depend on the sampling fraction, prevalence, and population size.
 Without a complete understanding of these interations, we are destined to misinterpret case-counts and come to erroneous conclusions.
-We will show, through simple calculations, how we can misguide ourselves and the need for more precision and care in presenting COVD-19 statistical models to the broader research and non-academic communities.
+We will show, through simple calculations, why there is a strong need for more precision and care in presenting COVID-19 analyses to the broader research and non-academic communities.
 
@@
 Third, we show that daily trends in rate of positive tests is equally problematic even if testing rates and selection protocols are held constant.
-We show that the rate of change in COVID-19 observed case-counts are often an underestimate of the true rate.
-This implies use of observed rates will lead a data analyst to be overly optimistic in the early stages that the disease is not an issue, and also overly optimistic in the decline of the prevalence after the peak.
-Fourth, we show that estimates of the reproducility rate are also impacted.
+We show that the rate of change in COVID-19 observed case-counts often first overestimate and then underestimate the true rate of change.
+This implies use of observed rates will lead a data analyst to be overly pessimistic in the early stages leading up to the peak time, and overly optimistic in its immediate aftermath.
+We show that estimates of the effective reproduction number are similarly impacted.
+Fourth, we show that cross-country comparisons are difficult at best with population size, sampling fraction, and data quality all interacting to impact null hypothesis testing.
+Fifth, we discuss a simple sensitivity analysis tool for estimating data quality and therefore understanding the potential impact on case-count analyses.
 
@@
 
+\begin{wrapfigure}{r}{0.5\textwidth}
+\centering
+\includegraphics[width = 0.4\textwidth]{../methods/figs/mem_heatmap_article.png}
+\caption{Measurement-error data quality adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.15$ and $FN=0.10$.
+Color scaled so blue = $-6$, white = 0, and red = $6$.}
+\label{fig:heatmap}
+\vspace{-0.3cm}
+\end{wrapfigure}
+
 The above argument shows that the effective sample size is affected in a complex manner by the interaction between selection bias and measurement error.
@@
 In a worst case scenario, if $FN \to 0.3$ and $FP \to 0.05$ then the factor is only 1.3.
-
-\begin{wrapfigure}{r}{0.5\textwidth}
-\centering
-\includegraphics[width = 0.4\textwidth]{../methods/figs/mem_heatmap_article.png}
-\caption{Measurement-error data quality adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.15$ and $FN=0.10$.
-Color scaled so blue = $-6$, white = 0, and red = $6$.}
-\label{fig:heatmap}
-\vspace{-0.3cm}
-\end{wrapfigure}
 
@@
 
+Social capital and models.
+
 \appendix

### 2020-05-10 write-up/covid-writeup.tex
covid draft almost done

@@
 Without a complete understanding of these interations, we are destined to misinterpret case-counts and come to erroneous conclusions.
-We will show, through simple calculations, why there is a strong need for more precision and care in presenting COVID-19 analyses to the broader research and non-academic communities.
+We will demonstrate why there is a strong need for more precision and care in presenting COVID-19 analyses to the broader research and non-academic communities.
 
@@
 We clarify mathematically why the situation is much more complex than it first appears.
-Through simple mathematical argument, we demonstrate five important, yet often forgotten, tenants of statistical science.
-First we show that unadjusted prevalence rates are unsurprisingly biased when tests are imperfect.
+Through simple mathematical arguments, we clarify five important issues related to COVID-19 case-count analysis.
+First, we show that unadjusted prevalence rates are (unsurprisingly) biased when tests are imperfect.
 What is surprising is that the direction and magnitude of the bias can vary substantially and interacts with the sampling fraction.
-Next we show that data quality for adjusted prevalence rates incorporate an additional term in the error decomposition that highlights the interplay among measurement error, selection bias, and prevalence.
+Next we show that data quality for adjusted prevalence rates includes an additional term in the error decomposition that captures the interplay among measurement error, selection bias, and prevalence.
 Third, we show that daily trends in rate of positive tests is equally problematic even if testing rates and selection protocols are held constant.
-We show that the rate of change in COVID-19 observed case-counts often first overestimate and then underestimate the true rate of change.
-This implies use of observed rates will lead a data analyst to be overly pessimistic in the early stages leading up to the peak time, and overly optimistic in its immediate aftermath.
-We show that estimates of the effective reproduction number are similarly impacted.
+We show that the rate of change in COVID-19 observed case-counts overestimates the true rate of change prior to the peak time and underestimates it immediately after.
+This implies a data scientist analyzing observed rates will be overly pessimistic in the early stages of the pandemic, and overly optimistic subsequent to the peak times.
+We show that estimates of the effective reproduction number may be similarly impacted.
 Fourth, we show that cross-country comparisons are difficult at best with population size, sampling fraction, and data quality all interacting to impact null hypothesis testing.
-Fifth, we discuss a simple sensitivity analysis tool for estimating data quality and therefore understanding the potential impact on case-count analyses.
-
-The above analyses are focused on the issues with using observed case-counts to understand the trajectory for a single country or state.
-The media and organizations are also interested in comparisons of case-counts across countries.
-To account for population size, case-counts are scaled by the country population size.
-We show that even after this adjustment, cross-country comparisons can be plagued by estimation error scaling by population size.
-We end the paper with a discussion of the benefits of randomized testing, where auxiliary information is most relevant, and the distinction between data quality and decision making in the current pandemic.
+Fifth, we discuss a sensitivity analysis method for estimating data quality and its use in understanding the potential biases in COVID-19 case-count analyses.
+We then end the paper with a discussion of the benefits of randomized testing, where auxiliary information is most relevant, and the distinction between data quality and decision making in the current pandemic.
 
@@
 At a fixed moment in time, let $Y_j$ denote COVID-19 status for the $j$th individual in the population, $j=1,\ldots, N$.
-Here, like in survey methodology (cite), we treat COVID-19 status as a fixed but unknown population quantity of interest.
+Here, like in survey methodology (cite), we treat COVID-19 status as a fixed but unknown quantity of interest.
 For simplicity, we start by ignoring the dynamic nature of the viral outbreak as well as the fact that individuals can recover from the disease and assume either individual $j$ is COVID-19 positive and $Y_j=1$ or is COVID-19 negative and $Y_j=0$.
@@
 Then a natural candidate for prevalence is the proportion of positive tests $\bar y = \frac{1}{n} \sum_{i=1}^n y_i$, and a natural candidate for overall cases is $N \times \bar y$.
-Under simple random sampling (SRS) or any other epsem design, the above are unbiased estimators of the population-level quantities of interest.
+Under simple random sampling (SRS) or any other epsem\footnote{equal probability of selection method} design, the above are unbiased estimators of the population-level quantities of interest.
 Under SRS, the variance of the estimator can be expressed as $\frac{1}{N-1} \times \frac{1-f}{f} \times \sigma_Y^2$ where $f = n/N$ is the sampling fraction and $\sigma_Y^2 = \frac{1}{N} \sum_{i=1}^N (Y_i - \bar Y)^2 = \bar Y (1- \bar Y)$.
@@
 Meng terms this the ``Law of Large Populations'' (LLP).
-The points to a critical issue in the current media practices in communication of case-count data: two countries with the same testing strategy (i.e., $\rho_{I,Y}$ equal) can yield wildly different estimates due to population size.
+This points to a critical issue in the current media practices in communication of case-count data: two countries with the same testing strategy (i.e., $E_{\I} [\rho_{I,Y} ]$ equal) can yield wildly different estimates due to population size.
 Large countries like the US may have similar true prevalence rates as smaller countries like the UK.
-Even under similar testing strategies, the mean-squared error in the prevalence rate for the US will have substantially more variation.
-For example, the MSE will be almost 5 times that in the US compared to the UK.
-Therefore, strong opinions built from case-count records may be substantially distorted.
+Even under similar testing strategies, the mean-squared error in the prevalence rate for the US will have substantially more variation in comparison to SRS.
+Comparing the US to the UK, for example, the MSE will increase by a factor of almost five.
+Thus conclusions drawn from observed case-count records may be not just wrong, but very wrong.
+
+\begin{wrapfigure}{r}{0.5\textwidth}
+\centering
+\includegraphics[width = 0.4\textwidth]{../methods/figs/US-casecount.png}
+\caption{US total tests and total positive tests per day}
+\label{fig:ustests}
+\vspace{-0.3cm}
+\end{wrapfigure}
 
@@
 While important, many conflate increased testing capacity with increased quality of testing.
-We will discuss this point further in Section XX.
-In the COVID-19 crisis, for example, New York has performed a total of $n = 872,481$ tests from April 20th to 30th.
-In Figure XX, we present the trajectory of testing per day and positive cases per day.
-The New York population is $N=8,399,000$; removing those individuals who have been already tested leaves a population of $7,XXX,XXX$.
-This leads to a fraction sampled of $f = 0.1$.
-A recent study attempted to estimate prevalence in New York.
-After appropriate adjustments (see Appendix XX for details), a plausible range for prevalence seems to be between $2\%$ and $12\%$.
-Suppose that individuals who are COVID-19 positive are $1.5$ times more likely to get tested than those individuals who are COVID-19 negative (a modest relative rate).
-The question is ``What is the sample size from a SRS that would yield equivalent MSE in the estimated prevalence?'' Using the above MSE result, Meng showed the effective sample size is equal to $n_{eff} \leq \frac{1}{E_I [ \rho_{I,Y}^2 ]} \frac{f}{1-f}$ and that ranges from $208$ to $38$ across the range of plausible values of prevalence.
-This is the sad state of affairs, saying that the effective sample size even in best case scenarios is not better than a small random sample from the population.
+We will discuss this point further in Section \ref{section:testingcomments}.
+As of May 8th, the US has performed $8,412,095$ tests in total.
+Figure~\ref{fig:ustests} shows the trajectory of testing per day and positive cases per day in the US.
+The US population is roughly $328$ million, meaning the fraction sampled is $f = 0.026$.
+After smoothing, the empirical prevalence is $9.1\%$ on May 8th.
+Supposing COVID-19 positive individuals are $2$ times more likely to get tested than those individuals who are COVID-19 negative, the question is ``What is the sample size from a SRS that would yield equivalent MSE in the estimated prevalence?'' Using the above MSE result, Meng showed the effective sample size is equal to $n_{eff} \leq \frac{1}{E_I [ \rho_{I,Y}^2 ]} \frac{f}{1-f}$.
+Under these parameters, the effective sample size is $14$.
+Recent proposals (cite) have suggested increasing testing capacity which may reduce the relative sampling rate.
+Supposing it drops to $1.2$ then the effective sample size will increase to $299$.
+See Appendix~\ref{section:effss} for the mathematical details and calculations in additional scenarios.
+The above calculations point to a sad state of affairs, saying that the effective sample size even in best case scenarios is not better than a small random sample from the population.
 
 The remainder of this paper aims to build upon these fundamental insights by extending the decomposition in two directions: accounting for  measurement-error and the temporal nature of the pandemic.
-We then discuss testing strategies, effective sampling methods, and data versus decision-making.
+We then discuss effective sampling methods and data versus decision-making.
 
@@
 Here we investigate the interplay between imperfect testing and selection bias.
-When discussing testing inaccuracies, the standard assumption is measurement-error leads to parameter attenuation.
-When paired with selection bias, however, the two sources of error become entangled, and the biases can become magnified, muted, or even switch signs.
+In discussions of inaccurate testing, the standard assumption is measurement-error leads to parameter attenuation.
+When paired with selection bias, however, the two sources of error become entangled, and resulting biases can become magnified, muted, or even switch signs.
 
@@
 From here on, we refer to $\rho_{I,Y}$ as the \emph{true data quality}, and $\rho_{I,PZ}$ as the \emph{observed data quality} that accounts for both selection bias and measurement-error.
-We show the sign of $\rho_{I,PZ}$ is the opposite of the sign of $\rho_{I,Y}$, and so the second term adjusts the true data quality.
+In the appendix, we show the sign of $\rho_{I,PZ}$ is the opposite of the sign of $\rho_{I,Y}$, which implies that the observed data quality adds bias in the opposite direction from the true data quality.
 
 We start by considering the first two terms and assess whether the sign of the bias can reverse due to the interaction of measurement-error and selection bias.
-To do this, we require the sampling rates differential.
+To do this, we define the sampling rates differential.
 Let $f_1 := \pr (I_J = 1 \mid Y_J = 1)$ and $f_0 := \pr(I_J = 1 \mid Y_J = 1)$ be the sampling rates.
-Then $\Delta = f_1 - f_0$ is the sampling rate differential and we have
+Then $\Delta = f_1 - f_0$ is the sampling rate differential.
+Using these terms, we can re-express the first two terms as
 $$
@@
 
-In the current setting, supposing the ratio of conditional selection rates $f_1/f_0$ is equal to 1.5.
-The false negative rate ranges from 15\% to 40\%, while the false positive rate is low (around 1\%).
-Then under plausible prevalence rates that we discussed before, the relative MSE ranges from 1.4 to 0.12.
-So that means in some cases we have a huge increase in MSE and in other settings we have a huge decrease in MSE.
+Consider again the current COVID-19 pandemic.
+We continue to assume the ratio of conditional selection rates $f_1/f_0$ is equal to 1.5.
+In Section~\ref{section:est_dq}, we discuss recent research suggesting a false negative rate around 17.2\% and a false positive rate around 0.05\%.
+Under these rates and the current US prevalence rates, the ratio of the MSE to MSE under no measurement error is 0.436; if we switch false negative rates down to 5\% and increase the false positive rate to 5\% then the relative MSE is 2.89.
+This is merely to demonstrate that in some cases we see a huge increase in MSE and in other settings we have a huge decrease in MSE.
 What drives this is the false negative rate interaction with prevalence and sampling rates.
-Therefore, whether we are better or worse off is a very difficult question to answer.
+Therefore, whether we are better or worse off with respect to the MSE is a very difficult question to answer.
 
@@
 The second term is the interaction between selection bias and measurement error.
-Interestingly we can show that the sign is reversed, i.e., $sgn(\rho_{I,PZ}) = 1 - sgn(\rho_{I,Y})$, leading to an absolute \emph{reduction} in the error.
+Figure~\ref{fig:heatmap} presents the measurement-error data quality adjustment as a function of the relative frequency and odds ratio.
+We see that the adjustment can be both positive and negative as well as greater than or less than zero.
 
-For this adjusted estimate, we can see that the directionality is maintained.
-Therefore, the benefit is if the expected correlation is positive, we can feel like the estimate is likely an over-estimate and vice versa.
-You pay a penalty in terms of variance and therefore the MSE may not be much smaller (CHECK!)
-
-Under the same numbers, $D_M$ is approximately 1.7.  The effective sample size can then be bounded by
+We next investigate the impact of measurement-error on the effective sample size calculation.  The effective sample size can now be bounded by
 $$
@@
 $$
-This lowers the ranges by a factor of almost $3$, leading to effective sample sizes from 69 to 13.
+In Section \ref{section:est_dq}, we discuss recent studies that estimate the COVID-19 false positive rate of $0.5$\% and false negative rate $17.2$\%.
+Supposing COVID-19 positive individuals are $2$ times more likely to get tested than those individuals who are COVID-19 negative, the effective sample size drops from $14$ to $10$.
+If the relative sampling fraction drops to $1.2$ then the effective sample size drops from $299$ to $216$.
 
@@
 Here, we discuss the impact of the interaction on the effective sample size when testing capacity is tied to measurement-error.
-In the current pandemic, there have been well-justified calls for increases in testing capacity.
+In the current COVID-19 pandemic, there have been well-justified calls for increases in testing capacity.
 These increases, however, may come at the cost of increases in false-positive and false-negative rates.
 In terms of effective sample size, if the data quality term remains constant, then testing increases can actually hurt effective sample size.
-Consider the simplest case where $f$ increases from $0.05$ to $0.1$; suppose this required including tests such that the false negative rate jumps from $0.05$ to $0.15$.
-Then we would think effective sample size should increase by a factor of $3.4$ but the actual increase is by a factor of only $1.8$!
-Think about what this is saying, this is saying that sampling $15\%$ of the population rather than $5\%$ will only increase the effective sample size by a factor $1.8$ if the selection-bias persists and the false negative rate climbs to $15\%$.
-It has been reported that false negative rates could increase more than that (cite NYT) and the false positive rate could go up slightly as well.
-In a worst case scenario, if $FN \to 0.3$ and $FP \to 0.05$ then the factor is only 1.3.
+Consider the simplest case where $f$ increases from $0.05$ to $0.1$; suppose this increase was associated with  false positive rate increasing from $0.5$\% to $5$\% and the false negative rate rising from $5\%$ to $20$\%.
+Then the standard analysis would suggest a reduction in the MSE by 47\% but the actual decrease is by 26\%.
+This implies that sampling $15\%$ of the population rather than $5\%$ will only increase the effective sample size by a factor $2.9$ rather than the expected $4$ fold increase that is expected.
+In a worst case scenario, if the false negative rate climbs to $0.30$ and the false positive rate to $0.10$ then the factor effective sample size increases by only a factor of $2.3$.
 
@@
 
-The data analyst, now frustrated with me but at least wary of estimating prevalence and total counts, pauses and thinks.
-They return shortly thereafter, a bit agitated but persistent: \emph{Ok, perhaps total counts is a lost cause.
+The data analyst, now least wary of estimating prevalence and total counts, pauses and thinks.
+They return shortly thereafter, with a follow-up consideration: \emph{Ok, perhaps estimating prevalence is difficult.
 Certainly, however, we can estimate the rate of growth.
@@
 We wish to express the error $\frac{\bar y_t}{\bar y_{t-1}} - \frac{\bar Y_{t}}{\bar Y_{t-1}}$ in statistical terms.
-Using a Taylor series approximation, we show that the error can be expressed approximately as
+Using a Taylor series approximation, we show in the appendix that the error can be expressed approximately as
 $$
@@
 where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement-error adjustment, and $CV(Y_j) = \sigma_{Y_j}/\bar Y_j$ is the coefficient of variation at time step $j$.
-The magnitude of the error depends on the population-level ratio $\bar Y_{t-1} / \bar Y_{t-1}$ so a large decrease in the prevalence rate will have a relatively small error as compared to a large increase in the prevalence rate.
+The magnitude of the error depends on the population-level ratio $\bar Y_{t} / \bar Y_{t-1}$ so a large decrease in the prevalence rate will have a relatively small error as compared to a large increase in the prevalence rate.
 The second term represents the ``cancellation'' that the naive analyst is hoping will occur.
 The analyst's intuition only holds when the difference is equal to zero.
-This occurs at times when data quality, sampling fraction, measurement-error, and prevalence are all held constant.
-In the case of New York, for example, recent weeks it seems plausible that $\Delta$ (i.e., selection bias) has not changed drastically.
-Figure XX shows the rate of testing remains constant and the total cases relative to the population is small enough such that $f_t \equiv f_{t-1}$ is reasonable.
-Finally, the measurement-error adjustment is likely to be constant as no new testing procedures have been reported (as far as this author knows).
-We will show how such a setting leads to over-estimation prior to and under-estimation after peak prevalence.
+This occurs at times when data quality, sampling fraction, measurement-error, and prevalence are all constant across time.
+
+Consider the state of New York.
+Figure~\ref{ny-covid-tests} shows the number of tests and positive tests per day in New York state.
+Figure~\ref{ny-covid-frac} presents the fraction of tests that were positive per day in New York state.
+For cancellation, we would need the expected data qualities to be uncorrelated and equal.
+Testing appears relatively stable over a week which suggests assuming $\Delta_t$ (i.e., differential in the selection rate) varies slowly over small windows of time is plausible.
+Note, however, that constant $\Delta$ does not imply constant data quality (see Appendix \ref{section:effss} for details).
+Figure~\ref{ny-covid-tests} shows the rate of testing remains relatively constant in the last few weeks, specifically relative to the total NY population, so $f_t \approx f_{t-1}$ is reasonable.
+Finally, the measurement-error adjustment is likely to be constant as no new testing procedures have been reported.
+This would not be true in states like Michigan, where testing criteria have been expanding every few weeks.
 The final component is an adjustment to this differential error based on the statistical error at the first time point.
 
-Figure XXa shows a trajectory of bias under several prevalence and data quality scenarios.
-We generate data from an SIR model with infection rate $\beta = XX$, total population = $100,000$.
-We estimate the bias under a range of values for the relative sampling fraction $f_1/f_0$ and a sampling fraction of $f = 0.05$.
-Figure XXb shows the trajectory of $\bar Y_t / \bar Y_{t-1}$.
-We see that in each case, the rate of change in prevalence is overestimated prior to the peak and then underestimated afterwards.
-From a decision making perspective, such biases have impacts.
+Figure \ref{fig:ratio-bias} shows a trajectory of the ratio and potential bias under an SIR model for the epidemic dynamis, with state evolutoin given by the following ODE
+$$
+\begin{aligned}
+\frac{\partial S_t}{\partial t} &= - \beta S_t I_t N^{-1}; \quad
+\frac{\partial I_t}{\partial t} = \beta S_t I_t N^{-1} - \gamma I_t; \quad
+\frac{\partial R_t}{\partial t} = - \gamma I_t \\
+\end{aligned}
+$$
+where $S_t$ is the number of susceptible individuals at time $t$, $I_t$ is the number of infected individuals at time $t$, $N$ is the total poulation size, and $R_t$ is the number of removed (recovered or deceased) individuals at time $t$.
+Figure~\ref{fig:fracpop} presents the fraction of population newly infected individuals at each time step for $\beta = 1.4$ and $\gamma = 0.2$.
+Figure \ref{fig:ratio-bias} shows the trajectory of $\bar Y_t / \bar Y_{t-1}$.
+We see that in each case, the rate of change in prevalence is likely to be  overestimated prior to the peak and then underestimated afterwards.
+From a decision making perspective, such biases have potential impact.
 First, the over-estimation may give lawmakers and governors more leverage in proposing aggressive actions that reduce prevalence.
@@
 What both sides miss is that the direction of bias is non-constant over time.
-After we hit the peak, estimates will under-estimate prevalence.
-Lawmakers and governors will be put under pressure to relax social distancing measures.
-This will drive a sequence of cycle of panic and .
-
+After we hit the peak, estimates will likely under-estimate prevalence.
+Such bias puts pressure on lawmakers and governors to prematurely relax social distancing measures.
 Moreover, the question is ``Can we trust the observed data to let us know if we have reached the peak?''  It appears that the peak time is the easiest to estimate in terms of having minimal error using availability data.
-Note, however, that the standard errors on such estimates will be incredibly wide as the effective sample size is quite small compared to SRS at each time.
-\textcolor{red}{Short description of effective sample size}.
+Note, however, that the standard errors on such estimates will be incredibly wide due to the effective sample size small based on our prior discussions.
 
@@
 For simplicity, we again assume the estimates are adjusted for known false positive and false negative rates.
-Luis and Ribeiro (2008) show that under a Poisson likelihood, a simple relation between the trajectory of \emph{true new cases} and the effective reproduction number can be derived (see Appendix XX for discussion).
+Luis and Ribeiro (2008) show that under a Poisson likelihood, a simple relation between the trajectory of \emph{true new cases} and the effective reproduction number can be derived.
 In particular, under an SIR model, the number of case counts on day $t$ is Poisson distributed with rate $K_{t-1} \exp \left( \frac{1}{\gamma} (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval which is approximately $7$ days for COVID-19 (Sanche, 2020).
@@
 Moreover, the error depends not on aggregate prevalence $\bar Y_t$ but fraction of new cases $\bar K_t$.
-
-In Figure XXc, we show the error in this quantity.  The trajectory follows the same as before; however, the errors are....
+Figure~\ref{fig:r0-bias} displays the bias for the effective reproductive number and how it is impacted by the relative sampling fraction.
+Results are similar to the ratio estimate, but the impact on the underestimation is more severe due to the logarithmic transformation.
 
@@
   \caption{Ratio estimator}
-  \label{fig:sub1}
+  \label{fig:ratio-bias}
 \end{subfigure}%
@@
   \caption{Effective reproductive rate estimator}
-  \label{fig:sub2}
+  \label{fig:r0-bias}
 \end{subfigure}
@@
 Here we assume $f = 0.02$, $FP = 0.01$, $FN = 0.15$, and use a range of relative sampling fraction $M = f_1/f_0$.}
-\label{fig:test}
+\label{fig:rates}
 \end{figure}
@@
 The key concern is whether clinical trials on COVID-19 recruit from the pool of individuals who have tested positive for COVID-19 or whether they sample randomly from the population, test, and then recruit from this subset of tested individuals.
-To see this issue, suppose we have an outcome $Y$, a treatment $A$ and an \emph{unobserved} $U$.
+To see this issue, suppose we have an outcome $Y$, a treatment $A$ and an \emph{unobserved} variable $U$.
 Suppose treatment is assigned at random, i.e., $A =1$ with probability $50\%$ and $A=0$ with probability $50\%$.
@@
 For COVID-19, the value of interest is the expected value of $U$ in the population of COVID-19 positive individuals.
-Due to selection bias, however, the bias in this marginal treatment effect compared to the true marginal effect of interest is $\beta_1 \cdot E_{\I} [ \rho_{I,U} ] \times \frac{f}{1-f} \times \sigma_U$.
+Due to selection bias, however, the bias in this marginal treatment effect compared to the true marginal effect of interest is $\beta_1 \cdot E_{\I} [ \rho_{I,U} ] \times \sqrt{ \frac{f}{1-f} } \times \sigma_U$.
 Selection bias then leads to the estimated effect being biased if $\beta_1 \neq 0$.
 Directionality of the bias will depend on the relationship between $U$ and $Y$.
-If $U$ is a measure of disease severity, then it could be expected to see $\beta_1 > 0$ and therefore treatment effects to be overly optimistic as most individuals recruited are potentially higher on the severity index than the population average.
+If $U$ is a measure of disease severity, then it could be expected to see $\beta_1 > 0$ and $E_{\I} [ \rho_{I, U} ] >0$.
+In such situations, the treatment effect estimates will tend to be overly optimistic as most individuals recruited are potentially higher on the severity index than the population average.
+Not only that, but for fixed data quality, the error compared to SRS in terms of effect estimation depends scales with the population size (i.e., the law of large populations).
 
@@
 Therefore, for marginal treatment effects to be interpretable, there must be a well-defined population.
-As our main interest is in causal effects on the population of COVID-19 positive.
-Randomization within the clinical trial design yields internal validity, but we also need external validity to estimate the marginal effect correctly.
+As often our main interest is in causal effects on the population of COVID-19 positive.
+Randomization within the clinical trial design yields internal validity, but we also need external validity to estimate the correct marginal effect of interest.
 
@@
 So only under the very strong assumptions of equal data quantity and quality will the comparisons be valid and the $Z$-score can be treated as in the classical setting.
-The issue is how to assess data quality equivalence among populations.
-\textcolor{red}{Discuss the measures in the literature.}  If this does not hold, then the width of the CIs need to be multipled by
 
@@
   \caption{Fraction of new cases in population}
-  \label{fig:sub1}
+  \label{fig:fracpop}
 \end{subfigure}%
@@
   \caption{Effective reproduction rate estimators}
-  \label{fig:sub2}
+  \label{fig:eff}
 \end{subfigure}
@@
 \subsection{Estimation of Data Quality}
+\label{section:est_dq}
 
@@
   \caption{NY new cases per day}
-  \label{fig:sub1}
+  \label{fig:ny-covid-test}
 \end{subfigure}%
@@
   \caption{NY fraction of new tests positive per day}
-  \label{fig:sub2}
+  \label{fig:ny-covid-frac}
 \end{subfigure}
 \caption{New York COVID-19 reported data }
-\label{fig:test}
+\label{fig:ny-covid}
 \end{figure}
@@
 \section{That which does not break us, makes us stronger (but potentially not smarter)}
+\label{section:testingcomments}
 
@@
 \section{Derivation for the effective sample size}
+\label{section:effss}
 
+For binary outcomes, we have
 \begin{equation} \label{eq:binaryrho}
@@
 \end{equation}
-where $\Delta = P_J (I_J = 1 \mid Y_J = 1) - P(I_J = 1 \mid Y_J = 0)$.  Suppose that $\Delta = 10\%$ and $\bar Y = 10\%$ then $\rho_{I,Y}$ is given by
-We expect the correlation to be more on the order
+where $\Delta = P_J (I_J = 1 \mid Y_J = 1) - P(I_J = 1 \mid Y_J = 0) = f_1 - f_0$.  Suppose that $M = f_1/f_0$; then
+$f_0 = f / (\bar Y \cdot (M-1) + 1)$.  For $\bar Y = 0.091$, $M = 2$ and $f = 0.026$ we have $f_0 = 0.024$ and $f_1 = 0.048$ which implies $\Delta = 0.024$.
+Then $\rho_{I,Y} = 0.043$.  Then $E_{\I} [ \rho_{I,Y}^2 ] \leq (E_{\I} [\rho_{I,Y} ]^2 = 0.0019$.  Then
+$$
+n_{eff} \leq \frac{f}{1-f} \frac{1}{0.0019} = 14.39.
+$$
+Using this computation, we compute effective sample size under a range of values $\bary$, $M$ with $f = 0.026$ (i.e., current sampling fraction) and $f =1/7$ (sample individuals roughly every 7 days).
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{rrrrrr}
+& \multicolumn{5}{c}{$M$} \\ \cline{2-6}
+$\bar y$ & 1.2 & 1.4 & 1.6 & 1.8 & 2 \\
+  \hline
+0.016 & 1598.00 & 402.00 & 180.00 & 102.00 & 66.00 \\
+  0.036 & 731.00 & 185.00 & 84.00 & 48.00 & 31.00 \\
+  0.056 & 484.00 & 124.00 & 56.00 & 32.00 & 21.00 \\
+  0.076 & 367.00 & 94.00 & 43.00 & 25.00 & 16.00 \\
+  0.096 & 299.00 & 78.00 & 36.00 & 21.00 & 14.00 \\
+   \hline
+\end{tabular}
+\end{table}
+
+We also present the same plot under $FP = 0.005$ and $FN = 0.172$ to show the impact of measurement-error on effective sample size.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{rrrrrr}
+  & \multicolumn{5}{c}{$M$} \\ \cline{2-6}
+$\bar y$ & 1.2 & 1.4 & 1.6 & 1.8 & 2 \\
+  \hline
+0.016 & 1154.00 & 290.00 & 130.00 & 73.00 & 47.00 \\
+  0.036 & 528.00 & 134.00 & 60.00 & 34.00 & 22.00 \\
+  0.056 & 349.00 & 89.00 & 41.00 & 23.00 & 15.00 \\
+  0.076 & 265.00 & 68.00 & 31.00 & 18.00 & 12.00 \\
+  0.096 & 216.00 & 56.00 & 26.00 & 15.00 & 10.00 \\
+   \hline
+\end{tabular}
+\end{table}
 

### 2020-05-10 methods/effective_ss.R

@@ -1,0 +1,91 @@
+bary = 0.091
+f_current = 0.026
+f_proposal = 1/7
+M = 1.5
+FP = 0.005; FN = 0.172
+
+eff_ss <- function(bary, f, M, FP=0.0, FN=0.0) {
+  f0 = f/(bary*(M-1)+1)
+  f1 = M*f0
+  Delta = f1 - f0  
+  rho = Delta*sqrt(bary*(1-bary))/sqrt(f*(1-f))
+  num = FP*(1-bary) + FN*bary
+  den = f0*(1-bary) + f1*bary
+  D_m = 1+FP+FN - Delta * bary/(1-bary) * num/den
+  neff = f/(1-f) * 1/(rho^2 * D_m^2)
+  return(neff)
+}  
+
+f = 0.1
+seq_bary = seq(0.016, 0.096, 0.02)
+seq_M = seq(1.2, 2, 0.2)
+result = matrix(nrow = length(seq_bary), ncol = length(seq_M))
+
+for(i in 1:length(seq_bary)) {
+  for(j in 1:length(seq_M)) {
+    result[i,j] = eff_ss(seq_bary[i], f_current, seq_M[j], 0, 0)    
+  }
+}
+
+library("xtable")
+colnames(result) = seq_M
+rownames(result) = seq_bary
+xtable(round(result,0))
+
+result2 = matrix(nrow = length(seq_bary), ncol = length(seq_M))
+
+for(i in 1:length(seq_bary)) {
+  for(j in 1:length(seq_M)) {
+    result2[i,j] = eff_ss(seq_bary[i], f_current, seq_M[j], FP, FN)    
+  }
+}
+
+colnames(result2) = seq_M
+rownames(result2) = seq_bary
+xtable(round(result2,0))
+
+
+rel_mse <- function(bary, f, M, FP=0.0, FN=0.0) {
+  f0 = f/(bary*(M-1)+1)
+  f1 = M*f0
+  Delta = f1 - f0  
+  rho = Delta*sqrt(bary*(1-bary))/sqrt(f*(1-f))
+  num = FP*(1-bary) + FN*bary
+  den = f0*(1-bary) + f1*bary
+  D_m = 1 - Delta * bary/(1-bary) * num/den
+  term1 = D_m * rho * sqrt(bary*(1-bary)) * sqrt(1-f)/sqrt(f)
[…]

### 2020-05-11 write-up/covid-writeup.tex
covid paper updated

@@
 \usepackage[foot]{amsaddr}
-\usepackage{biblatex}
+% \usepackage{biblatex}
+\usepackage{natbib}
 %\bibliographystyle{apalike}
@@
 Through simple calculations, we demonstrate pitfalls and paradoxes that can arise when considering case-count data in the presence of selection bias and measurement-error.
-The discussion guides a series of suggestions to improve current case-count reporting.
+The discussion guides several suggestions on how to improve current case-count reporting.
 \end{abstract}
@@
 As of that afternoon, the New York Times reports at least 965,214 people across the United States have tested positive for the virus, and at least 49,465 patients with the virus have died.
-Aggressive policies had been put in place across the US with at least 50\% of the US population officially urged to stay home via state-wide executive actions.
+Aggressive policies had been put in place across the US with at least 50\% of the US population officially urged as of late April to stay home via state-wide executive actions.
 
 Despite these steps, the data landscape for understanding COVID-19 remains limited.
-Public databases maintained by \href{https://bit.ly/2UqFSuA}{Johns Hopkins University} (cite) and the \href{https://bit.ly/2vUHfrK}{New York Times} (cite) provide incoming county-level information of confirmed cases and deaths.
-Statisticians, epidemiologists, economists, and data scientists have all used this granular data to forecast COVID-19 case-counts, deaths, and hospitalizations (cite).
-Variations on the SIR models (cite) have been used to draw inferences about infection rates and intensity.
-Parameter inference most often uses observed case-count and/or observed COVID-19 related deaths to infer latent trajectories of the pandemic (cites).
+Public databases maintained by \href{https://bit.ly/2UqFSuA}{Johns Hopkins University}~\citep{JHU_Lancet} and the \href{https://bit.ly/2vUHfrK}{New York Times}~\cite{NYT} provide incoming county-level information of confirmed cases and deaths.
+Statisticians, epidemiologists, economists, and data scientists have used this granular data to forecast COVID-19 case-counts, deaths, and hospitalizations~\citep{Song2020,Ray2020,2020.IHME}.
+In many cases, variations on the SIR models have been used to draw inferences about infection rates and intensity.
+Parameter inference often uses observed case-count and/or observed COVID-19 related deaths to infer latent trajectories of the pandemic.
 
 The goal of this paper is to express reservations at the use of case-counts as a proxy for disease prevalence and intensity as well as its use as direct input into estimation of standard epidemiological models for inference and forecasting.
-The reason is simple: current models do not account simultaneously for selection bias and measurement-error.
+The reason is straightforward: current models do not account simultaneously for selection bias and measurement-error.
 Selection bias enters due to differences in testing strategies across countries and states.
 In the US, for example, limited testing capacity has caused local and state health departments to focus on testing only high-risk populations.
-Moreover, testing requires the individual to self-select into receiving the test.
+Moreover, testing requires the individual to self-select into asking for a test and then receiving approval by their doctor.
 Demands for increased testing capacity, while laudable, ignore the issues of self-selection and measurement-error.
 While increasing testing capacity increases data quantity, there is no guarantee for increased data quality.
-We will show that aggressive pushes for ramped up testing capacity that are tied to decreases in data quality may have a very mitigated impact  in our ability to estimate quantities of interest such as prevalence and effective reproduction number.
+We will show that aggressive pushes for ramped up testing capacity that are tied to decreases in data quality may have a mitigated impact  in our ability to estimate quantities of interest such as prevalence and effective reproduction number.
 
 In this paper, we will demonstrate the complex interaction between these two fundamental concepts.
-Measurement-error is often associated with parameter attenuation; we show, however, that the combination with selection effects can cause bias in parameter estimates to change sign and increase/decrease in magnitude.
+In classical statistics, measurement-error is often associated with parameter attenuation; we will show, however, that the combination with self-selection can cause bias in parameter estimates to change sign and increase/decrease in magnitude.
 Moreover, the bias may depend on the sampling fraction, prevalence, and population size.
@@
 
-\subsection{Related work}
+% \subsection{Related work}
 
-There has been an abundance..
+% There has been an abundance..
 
-However, most complain about testing \emph{capacity} (cite Nate Silver), there is clearly issues of data quality.
+% However, most complain about testing \emph{capacity} (cite Nate Silver), there is clearly issues of data quality.
 
-Measurement-error discussions.
-Selection bias . Proponents of this critique suggest well-designed studies.
+% Measurement-error discussions.
+% Selection bias . Proponents of this critique suggest well-designed studies.
 
-Stanford study was Ionides.  They however, forgot about Critique \#2: measurement
+% Stanford study was Ionides.  They however, forgot about Critique \#2: measurement
 
@@
 We show that the rate of change in COVID-19 observed case-counts overestimates the true rate of change prior to the peak time and underestimates it immediately after.
-This implies a data scientist analyzing observed rates will be overly pessimistic in the early stages of the pandemic, and overly optimistic subsequent to the peak times.
+This implies a data scientist analyzing observed rates will be (on average) overly pessimistic in the early stages of the pandemic, and overly optimistic subsequent to the peak times.
 We show that estimates of the effective reproduction number may be similarly impacted.
@@
 Fifth, we discuss a sensitivity analysis method for estimating data quality and its use in understanding the potential biases in COVID-19 case-count analyses.
-We then end the paper with a discussion of the benefits of randomized testing, where auxiliary information is most relevant, and the distinction between data quality and decision making in the current pandemic.
+We then end the paper with a discussion of the benefits of randomized testing, use of auxiliary information is most relevant, and the distinction between data quality and decision making in the current pandemic.
 
 \section{Analysis of case-count data}
+\label{section:casecount}
 
@@
 At a fixed moment in time, let $Y_j$ denote COVID-19 status for the $j$th individual in the population, $j=1,\ldots, N$.
-Here, like in survey methodology (cite), we treat COVID-19 status as a fixed but unknown quantity of interest.
+Here, like in survey methodology~\citep{Cochran77}, we treat COVID-19 status as a fixed but unknown quantity of interest.
 For simplicity, we start by ignoring the dynamic nature of the viral outbreak as well as the fact that individuals can recover from the disease and assume either individual $j$ is COVID-19 positive and $Y_j=1$ or is COVID-19 negative and $Y_j=0$.
@@
 
-When the sampling mechanism is non-random, selection effects may cause bias in the above estimates.
-To better understand this issue, Meng (2019) provided the following intuitive and powerful statistical decomposition of the error between $\bar y$ and the true proportion $\bar Y$
+The above selection mechanisms are random and independent of the outcome of interest.
+When this is not the case, selection effects may cause bias in the above estimates.
+To better understand this issue, \cite{Meng2018} provided the following intuitive and powerful statistical decomposition of the error between $\bar y$ and the true proportion $\bar Y$
 $$
@@
 
-Here we highlight the core insights relevant to the COVID-19 crisis.  First, comparing the mean-squared error under selection mechanism $\I$ and SRS, we see that
+Here we highlight the core insights from~\cite{Meng2018} that are relevant to the COVID-19 crisis.  First, comparing the mean-squared error under selection mechanism $\I$ and SRS, we see that
 $$
@@
 This points to a troubling and paradoxical situation: the error relative to SRS increases as a function of population size.
-Meng terms this the ``Law of Large Populations'' (LLP).
+\cite{Meng2018} termed this the ``Law of Large Populations'' (LLP).
 This points to a critical issue in the current media practices in communication of case-count data: two countries with the same testing strategy (i.e., $E_{\I} [\rho_{I,Y} ]$ equal) can yield wildly different estimates due to population size.
@@
 Under these parameters, the effective sample size is $14$.
-Recent proposals (cite) have suggested increasing testing capacity which may reduce the relative sampling rate.
+Recent proposals~\citep{Siddarth2020} have argued in favor of increasing testing capacity which may reduce the relative sampling rate.
 Supposing it drops to $1.2$ then the effective sample size will increase to $299$.
@@
 Let $P_j$ be an indicator of measurement error, equal to $1$ when we mismeasure the outcome and $0$ when we observe the true outcome.
-We suppose this is a stochastic variable that satisfies $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.
+We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.
 If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^\star = Y_j(1-P_j) + (1-Y_j) P_j$.
@@
 From here on, we refer to $\rho_{I,Y}$ as the \emph{true data quality}, and $\rho_{I,PZ}$ as the \emph{observed data quality} that accounts for both selection bias and measurement-error.
-In the appendix, we show the sign of $\rho_{I,PZ}$ is the opposite of the sign of $\rho_{I,Y}$, which implies that the observed data quality adds bias in the opposite direction from the true data quality.
+In the appendix, we show the sign of $\rho_{I,PZ}$ is the opposite of the sign of $\rho_{I,Y}$, which implies that the observed data quality adds error in the opposite direction from the true data quality.
 
@@
 The final term in brackets is the \emph{measurement-error adjustment to data quality} which is a complex function of sampling rate differential, the odds ratio, and the ratio of measurement-error interaction with prevalence and sampling rates interaction with prevalence.
-Note $\sgn(\Delta) = \sgn(\rho_{I,Y})$ by equation~\ref{eq:binaryrho} so the measurement-error adjustment either shrinks the data quality measure toward zero or reverses its sign.
+Note $\sgn(\Delta) = \sgn(\rho_{I,Y})$ by equation~\ref{eq:binaryrho} in the appendix.
+This implies the measurement-error adjustment either shrinks the data quality measure toward zero or reverses its sign.
 
-While prior investigations have noted the interaction between measurement-error and selection bias (cites), the interaction with the sample size relative to the population, i.e., $f$, has largely been ignored.
+While prior investigations have noted the interaction between measurement-error and selection bias~\citep{Beesley2020,Beesley2019,Smeden2019}, the interaction with the sample size relative to the population, i.e., $f$, has largely been ignored.
 The above statistical decomposition clarifies the importance of this quantity~$f$.
@@
 In the appendix, we show that the error now can be expressed as
-$$
-\tilde y_n - \bar Y = \rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
+\begin{equation}
+\label{eq:error}
+\rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
 \times \underbrace{\left[ 1 + FP + FN - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right]}_{D_M}.
-$$
+\end{equation}
 The first term is the same as before but increased by $(1 + FP + FN)$ to account for the additional uncertainty due to measurement-error.
@@
 These increases, however, may come at the cost of increases in false-positive and false-negative rates.
-In terms of effective sample size, if the data quality term remains constant, then testing increases can actually hurt effective sample size.
+In terms of effective sample size, if the data quality term remains constant, then testing increases may not significantly improve the effective sample size.
 Consider the simplest case where $f$ increases from $0.05$ to $0.1$; suppose this increase was associated with  false positive rate increasing from $0.5$\% to $5$\% and the false negative rate rising from $5\%$ to $20$\%.
 Then the standard analysis would suggest a reduction in the MSE by 47\% but the actual decrease is by 26\%.
-This implies that sampling $15\%$ of the population rather than $5\%$ will only increase the effective sample size by a factor $2.9$ rather than the expected $4$ fold increase that is expected.
+This implies that doubling the sampling rate will only increase the effective sample size by a factor $2.9$ rather than the expected $4$ fold increase that is expected.
 In a worst case scenario, if the false negative rate climbs to $0.30$ and the false positive rate to $0.10$ then the factor effective sample size increases by only a factor of $2.3$.
@@
 
-The data analyst, now least wary of estimating prevalence and total counts, pauses and thinks.
+The data analyst, now wary of estimating prevalence and total counts, pauses and thinks.
 They return shortly thereafter, with a follow-up consideration: \emph{Ok, perhaps estimating prevalence is difficult.
@@
 Consider the state of New York.
-Figure~\ref{ny-covid-tests} shows the number of tests and positive tests per day in New York state.
-Figure~\ref{ny-covid-frac} presents the fraction of tests that were positive per day in New York state.
+Figure~\ref{fig:ny-covid-test} shows the number of tests and positive tests per day in New York state.
+Figure~\ref{fig:ny-covid-frac} presents the fraction of tests that were positive per day in New York state.
 For cancellation, we would need the expected data qualities to be uncorrelated and equal.
-Testing appears relatively stable over a week which suggests assuming $\Delta_t$ (i.e., differential in the selection rate) varies slowly over small windows of time is plausible.
-Note, however, that constant $\Delta$ does not imply constant data quality (see Appendix \ref{section:effss} for details).
-Figure~\ref{ny-covid-tests} shows the rate of testing remains relatively constant in the last few weeks, specifically relative to the total NY population, so $f_t \approx f_{t-1}$ is reasonable.
+Testing appears relatively stable over a week which suggests assuming $\Delta_t$ (i.e., differential in the selection rate) varies slowly over time may be a plausible assumption.
+Note, however, that constant $\Delta_t$ does not imply constant data quality (see Appendix \ref{section:effss} for details).
+Figure~\ref{fig:ny-covid-test} shows the number of tests has remained relatively constant in the last few weeks, specifically relative to the total NY population, assuming $f_t$ also varies slowly in time is reasonable.
 Finally, the measurement-error adjustment is likely to be constant as no new testing procedures have been reported.
@@
 
-Figure \ref{fig:ratio-bias} shows a trajectory of the ratio and potential bias under an SIR model for the epidemic dynamis, with state evolutoin given by the following ODE
+Figure \ref{fig:ratio-bias} shows a trajectory of the true ratio and the potential biased estimators under an SIR model for the epidemic dynamics, with state evolution given by the following ODE
 $$
@@
 Figure~\ref{fig:fracpop} presents the fraction of population newly infected individuals at each time step for $\beta = 1.4$ and $\gamma = 0.2$.
-Figure \ref{fig:ratio-bias} shows the trajectory of $\bar Y_t / \bar Y_{t-1}$.
-We see that in each case, the rate of change in prevalence is likely to be  overestimated prior to the peak and then underestimated afterwards.
+In each case, the rate of change in prevalence is likely to be  overestimated prior to the peak and then underestimated afterwards.
 From a decision making perspective, such biases have potential impact.
@@
 Moreover, the question is ``Can we trust the observed data to let us know if we have reached the peak?''  It appears that the peak time is the easiest to estimate in terms of having minimal error using availability data.
-Note, however, that the standard errors on such estimates will be incredibly wide due to the effective sample size small based on our prior discussions.
+Note, however, that the standard errors on such estimates will be incredibly wide due to the low effective sample size as discussed in Section~\ref{section:casecount}.
 
-\subsubsection{Estimation of effective reproduction number}
-Many well-respected epidemiologists argue that tracking the effective reproduction number $R_t$ is the only way to manage through the crisis (cites).
+\subsection{Estimation of effective reproduction number}
+\label{section:r0-estimation}
+Many well-respected epidemiologists argue that tracking the effective reproduction number $R_t$ is the only way to manage through the crisis~\citep{Gabriel2020}.
 We show here how these estimates are also impacted by selection bias and measurement error.
 For simplicity, we again assume the estimates are adjusted for known false positive and false negative rates.
-Luis and Ribeiro (2008) show that under a Poisson likelihood, a simple relation between the trajectory of \emph{true new cases} and the effective reproduction number can be derived.
-In particular, under an SIR model, the number of case counts on day $t$ is Poisson distributed with rate $K_{t-1} \exp \left( \frac{1}{\gamma} (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval which is approximately $7$ days for COVID-19 (Sanche, 2020).
-Using this, a method of moments estimator of the effective reproduction number at time $t$ can be given by
+\cite{Bettencourt2008} show that under a Poisson likelihood, a simple relation between the trajectory of new cases and the effective reproduction number can be derived.
+In particular, under an SIR model, the number of case counts on day $t$ is Poisson distributed with rate $K_{t-1} \exp \left( \frac{1}{\gamma} (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval which is approximately $7$ days for COVID-19~\citep{Sanche2020}.
+Using this, a simple estimator of the effective reproduction number at time $t$ can be given by
 $$
@@
 So here we see the exact same trade-off as with the rates but on the logarithmic scale.
-For small values $\log(1+x) = x$ so the biases are similar.
+For small values of $x$, we have $\log(1+x) = x$ and so the biases are similar.
 However, the error is no longer scaled by the rate of change in prevalence $\bar Y_2/\bar Y_1$.
@@
 \begin{aligned}
-\frac{\bar y_1 - \bar y_2}{ V_{SRS} (\bar y_1 - \bar y_2) } &=
+\frac{\bar y_1 - \bar y_2}{ \sqrt{V_{SRS} (\bar y_1 - \bar y_2)} } &=
 \frac{\rho_{I_1, Y_1} \times D_{M_1} \times \sqrt{\frac{1-f_1}{f_1}}  - \rho_{I_2, Y_2} \times D_{M_2} \times \sqrt{\frac{1-f_2}{f_2}} }{ \sqrt{\frac{1}{N_1 -1} \times \frac{1-f_1}{f_1} + \frac{1}{N_2 - 1} \times \frac{1-f_2}{f_2}} }  \\
@@
 
-\textcolor{red}{Take, for example, a comparison of the US and Canada.
+Take, for example, a comparison of the US and Canada.
 The two population sizes are approximately 328 million and 38 million people.
 Then the population adjustment is $5.84$ million.
-As of May 5th, Canada has performed a total of 24.92 tests per 1,000 people while the US has performed a total of 22.01 tests per 1,000 people (cite: https://ourworldindata.org/grapher/full-list-cumulative-total-tests-per-thousand)} so the data quantities are approximately equal to $f = 23/1000 = 0.023$.
+As of May 5th, Canada has performed a total of 24.92 tests per 1,000 people while the US has performed a total of 22.01 tests per 1,000 people\footnote{According to \href{https://ourworldindata.org/grapher/full-list-cumulative-total-tests-per-thousand}{ourworldindata.org}} so the data quantities are approximately equal to $f = 23/1000 = 0.023$.
 Then, under a prevalence of 10\%, one would need the differential $\Delta_1-\Delta_2$ to be much smaller than $11.68 \times 10^{-6}$ in order for the selection bias to not impact.
 So only under the very strong assumptions of equal data quantity and quality will the comparisons be valid and the $Z$-score can be treated as in the classical setting.
-
-\textcolor{red}{What is adjustment to CIs}.   Inserting the effective sample sizes can give us a better sense for a appropriate $Z$-score.  For example, using $n_eff = 20$ for each. And then
+A simple alternative is to use the effective sample sizes to build more appropriate $Z$-score.
+That is, setting
 $$
-Z_n = \frac{\bar y_1 - \bar y_2}{ \frac{1-f}{f} \sigma_y \sqrt{ \frac{1}{n_1 - 1} + \frac{1}{n_2 - 1}}}.
+Z_n = \frac{\bar y_1 - \bar y_2}{ \frac{1-f}{f} \sigma_y \sqrt{ \frac{1}{n_{1,eff} - 1} + \frac{1}{n_{2,eff} - 1}}}
 $$
-So we see that we cannot reject the null. \textcolor{red}{We can do the same for different rates as well.}  This is not perfect but more appropriate.
+allows us to build Z-scores that account for the true sample size.  While still not perfect, this allows us to build Z-scores that are not overly confident due to selection bias and measurement error.
 
 \subsection{Case count comparisons}
+Two alternative comparisons are common.
+First, the data analyst may directly compare the observed number of new cases on a given day.
+If $Y_1$ and $Y_2$ are the number of cases in locations $j=1,2$ respectively, then we can express the error of $y_1 - y_2$ as
+$$
+\begin{aligned}
+y_1 - y_2 &= n_1 ( \bar y_1 - \bar Y_{1}) - n_2 ( \bar y_2 - \bar Y_{2}) + \left( \frac{n_1}{N_1} Y_{1} - \frac{n_2}{N_2} Y_2  \right) \\
+&= \left[ n_1 \sigma_{Y_1} \rho_{I_1, Y_1} \sqrt{\frac{1-f_1}{f_1}} D_{M_1} - n_2 \sigma_{Y_2} \rho_{I_2, Y_2} \sqrt{\frac{1-f_2}{f_2}} D_{M_2}  \right] + \left[ f_1 Y_1 - f_2 Y_2 \right].
+\end{aligned}
+$$
+Under simple random sampling, the first term is zero in expectation.
+However, the question is whether $f_1 \bar Y_1 - f_2 \bar Y_2$ is of scientific interest.
+We would argue that this is only of true interest when $f_1 = f_2$ and we are interested in comparing total number of cases.
+In this setting, however, the difference in error terms is unlikely to be zero.
+To see this, assume that we have similar expected data quality, data quantity, problem difficulty, and measurement error.
+Then the error is a function of $n_1 - n_2 = n_2 \left( \frac{N_1}{N_2} - 1 \right)$.
+So we see that the error now is a function of the relative difference in the population sizes.
 
-We could instead think about the number of new cases $K_1$ and $K_2$ within each population.  Then
+Alternatively, the data analyst may be wary of such concerns in population differences and scale counts by respective population size, i.e., $y_1/N_1 - y_2/ N_2$.  We can again express the error as
 $$
-y_1 - y_2 = \frac{N_1}{n_1} ( \bar y_1 - Y_{1}) + \left( \frac{N_1}{n_1} Y_{1} - \frac{N_2}{n_2} Y_2  \right).
+\begin{aligned}
+\frac{y_1}{N_1} - \frac{y_2}{N_2} &= \frac{n_1}{N_1} ( \bar y_1 - \bar Y_{1}) - \frac{n_2}{N_2} ( \bar y_2 - \bar Y_{2}) + \left( \frac{n_1}{N_1} \bar Y_{1} - \frac{n_2}{N_2} \bar Y_2  \right) \\
+&= \left[ \sigma_{Y_1} \rho_{I_1, Y_1} \sqrt{ f_1 (1-f_1) } D_{M_1} - \sigma_{Y_2} \rho_{I_2, Y_2} \sqrt{f_2 (1-f_2)} D_{M_2} \right] + \left[ f_1 \bar Y_1 - f_2 \bar Y_2 \right].
+\end{aligned}
 $$
-OK then we should compare relative counts?
-The question is then ``What are you comparing at a populatoin level?''
-$$
-y_1/N_1 - y_2/ N_2 = \left[ f_1 \sigma_{Y_1} \rho_{I_1, Y_1} \sqrt{\frac{1-f_1}{f_1}} D_{M_1} - f_2 \sigma_{Y_2} \rho_{I_1, Y_1} \sqrt{\frac{1-f_1}{f_1}}  \right] + \left[ f_1 \bar Y_1 - f_2 \bar Y_2 \right]
-$$
-Is this a quantity of genuine interest?
-Under SRS, this makes sense if $f_1 = f_2$.
-Otherwise, you are comparing apples and oranges.
-With selection bias, it's not clear that agreement in this quantity is even desirable.
+Under simple random sampling, the first term is again zero in expectation.
+However, the question is whether $f_1 \bar Y_1 - f_2 \bar Y_2$ is of scientific interest.
+We would argue that this is only of true interest when $f_1 = f_2$ and we wish to compare disease prevalence.
+Importantly, the error in this comparison is less problematic.
+To see this, assume that we have similar expected data quality, data quantity, problem difficulty, and measurement error.
+Then the error is equal to zero.
 
-A final question is whether an observed drop in prevalence can be interpreted as a ``real'' decrease in prevalence.  To this end, one would typically construct a Z-score and perform a
-
-$$
-\frac{n}{N} \times \rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y} \times D_M
-$$
+It is important to note that the standard law of large populations caveats apply to both comparisons.
+In both cases, the point is that we must ascertain what is the population-level contrast and whether this should be of genuine scientific interest.
+Otherwise, we are comparing apples and oranges.
+Not only this, but the two comparisons come with different levels of error induced by selection bias and measurement error.
 
@@
 Here, for simplicity, we focus on comparing the estimated effective reproductive rate.
-We assume the two time-series are aligned so that $t=0$ is the time of first .
-This negates the alignment issue and is common in the plots.
+We assume the two time-series are aligned so that $t=0$ is the time of first case.
+This negates alignment issues and is common in practice.
 
 While the above issues on $Z$-scores and effective sample size are important, here we highlight a separate issue.
-Namely, the interplay of bias when two trajectories when the peaks are slightly off.
-We show that the biases can make us think the gap between two similar trajectories is larger than we would like!!!!
+Namely, the interaction of biases in estimation of the two trajectories when the peak infection times differ slightly.
+We can write the difference in the rates as
 $$
@@
 $$
-where $e_{jt}$ is the error associated given in Section XX.
+where $e_{jt}$ is the error associated given in Section \ref{section:r0-estimation}.
 Recall that the error allowed for over-estimation prior to the peak and then under-estimation post-peak.
@@
 Consider two countries (A and B) in which the peak occurs 2 weeks prior for country A then country B but other.
-Figure XX presents 1 week prior to the peak of A up to one week after the peak for country B.
-Data for each country is simulated from an SIR model as described in Section XX.
-We see that the errors interact in complex ways.
-In week 1, the two countries appear the same; then in week 2, the gap between country A and B enlarges significantly as country A seems to see rapid improvement; then in week 3, country B improves and actually catches up to country A; then in month 4, country A sees rapid improvement while country B appears worse.
-We are not claiming this is always the case, and indeed for small enough selection bias and/or large gaps in case-counts such complex behavior will not occur; however, this is simpy to point out how observed information can tell a more complex story than the truth (i.e., country A recovers 1 month prior to country B).
+Figure~\ref{fig:comparison} presents such a comparison where each country's disease trajectory follows an SIR model as described in Section XX with different parameters.
+Figure \ref{fig:eff} shows how the errors interact in complex ways.
+First, the gap is correctly estimated; then the gap is over-estimated as country A sees a rapid rise in cases; then the gap is even more over-estimated as country A improves and the rate is under-estimated while country B sees a rapid increase in their case-count and their rate is over-estimated; then country A's rate is correctly estimated while country B's rate is under-estimated as it sees improvement in infection rates; finally, the gap disappears.
+We are not claiming this will always be the case, and indeed for small enough selection bias and/or large gaps in case-counts such complex behavior will not occur; however, this is simply to point out how observed information can tell a more complex story than the truth (i.e., country A recovers 1 month prior to country B).
 
@@
 Right: comparison of $\hat R_t$ across time with $FN = 0.2$, $FP = 0.01$, and $M = 4$.}
-\label{fig:test}
+\label{fig:comparison}
 \end{figure}
@@
 In Meng (2019), estimation of the data quality relied on observing the true outcome.
-In election data, for example, we observe the voting distribution and can use this observation as ground truth.
+In election data, for example, we observe the vote totals after the fact and can use this as ground truth.
 We are not so lucky in the current crisis.
@@
 It found 13.9\% of 3,000 people tested across the state had signs of the virus.
-The study did not report sensitivity and specificity; therefore, we take reported measures from the Santa Clara study (cite) \footnote{We recognize the debate regarding its conclusions.
+The study did not report sensitivity and specificity; therefore, we take reported measures from the Santa Clara study~\citep{Bendavid2020} \footnote{We recognize the debate regarding its conclusions.
 Here we are only using this study for its information regarding measurement-error rates.}.
@@
 
-Figure XXa plots the number of new total and positive tests per day.
-Figure XXb plots the fraction of tests that were positiver per day.
-In both cases, we see smoothed these plots as the spiky pattern is likely due to testing backlogs.
+Figure~\ref{fig:ny-covid-test} plots the number of new total and positive tests per day.
+Figure~\ref{fig:ny-covid} plots the fraction of tests that were positiver per day.
+In both cases, we applied an exponential smoothing to the reported values because the spiky pattern is likely due to testing backlogs.
 Given the report was on the 23rd, we suppose for simplicity that the study was run several days prior on the 20th of April.
@@
 The population of New York state is $19,378,102$.
-Subtracting off the number of individuals who had already been tested yields a sampling fraction of $f = 0.001$.
+Subtracting off the number of individuals who had already been tested yields a sampling fraction of $f = 0.001$ on the 20th.
 The smoothed prevalence estimate on the 20th is 28.3\%.
@@
 Here, we assume the New York survey is a simple random sample from the state's population.
-Comparison of states with similar $f$ and similar testing inclusion protocols.  Not perfect.
-
+We recognize that the survey is not such a sample, but techniques like poststratification and raking can be applied to calibrate estimates.
+Then the error is $16.6$\%.
+Using~\eqref{eq:error}, we can use the error to construct an estimate of the relative sampling rate:
 \[
@@
 While nothing above is definitive, it points to the interplay of selection bias and measurement-error.
-The outlined procedure gives the community tools to understand how sensitivity and specificity impact our estimators and therefore how much we should trust conclusions based on these assumptions.
+This calculation gives us simple tools to understand how sensitivity and specificity impact our estimators and therefore how much we should trust conclusions based on these assumptions.
 Under the Santa Clara study estimates of these quantities, while we expect impact, the bias is much smaller than under other estimates.
@@
 This article argues why random sampling is a powerful tool and the limitations of data obtained through a self-selection process.
-Few COVID-19 studies (cite) have been released that aim to build representative samples.
-One difficulty in the current crisis is the low prevalence relative to the population in many areas.
-Take the recent Santa Clara study, for example.
+As COVID-19 survey studies become more readily accessible, we encourage researchers to focus on building representative samples.
+One difficulty in the current crisis is the low prevalence relative to the population size in certain geographies.
+Take the recent Santa Clara study, for example, that estimated less than 2\% prevalence in the area.
 The key issue with low prevalence is that if the test's specificity is not lower than the prevalence, then the observed data is consistent with zero recovered individuals.
 
-In such settings, sampling at random is doomed; however, there are solutions to such settings.
-Stratify the population into risk categories (based on population density, occupation, age, and other important factors) and perform Neyman allocation (i.e., sample strata $h$ with $P_h \sqrt{\bar Y_h \times (1-\bar Y_h)}$ where $P_h$ is the strata frequency in the population and $\bar Y_h$ the prevalence in this.
-Of course, $P_h$ and $\bar Y_h$ are not known a priori; however, if we can get decent esitmates.
-\textcolor{red}{Huge benefits when the rates are very different.}
-
-One could go further.
-\textcolor{explain network stuff}.
-However, network information is difficult to obtain.
-Geographic location, occupation, age and basic demographic information are much easier to obtain and are a good proxy for the latent network connectivity.
+In such settings, simple random sampling is doomed to fail; however, there are potential solutions.
+In particular, one can stratify the population into risk categories (based on population density, occupation, age, and other important factors) and perform Neyman allocation~\citep{Cochran77}.
+That is, sample individuals at random in strata $h$ with size proportional to $P_h \sqrt{\bar Y_h \times (1-\bar Y_h)}$ where $P_h$ is the strata frequency in the population and $\bar Y_h$ the prevalence in the strata.
+Of course, $P_h$ and $\bar Y_h$ are not known a priori; however, if we can get decent estimates, we can see huge benefits in settings where the prevalence varies widely by strata.
+One could go attempt to go further and incorporate contact network information.
+However, such data is difficult to obtain in practice.
+Indeed, geographic location, occupation, age and basic demographic information are much easier to obtain and are a good proxy for the latent network connectivity.
 
@@
 
-Having read through the above technical discussions, one may come to the incorrect conclusion that because we are wary of trusting the observational data, we must wait for better data sources in order to act (Ionides cite).
-This conclusions mistakes data and information.
-Lack of high quality data means we should be skeptical regarding conclusions drawn from these data sources.
+Having read through the above technical discussions, one may come to the incorrect conclusion that because we are concerned about implicit biases in observational data, we must wait for better data sources in order to act~\citep{Ioannidis2020}.
+This conclusion confuses data and information.
+Lack of high quality data means we should be skeptical regarding conclusions drawn from these data sources alone.
 There are whole fields dedicated to careful analysis in such settings (e.g., causal inference from observational studies).
-Being a data skeptic, does not mean one must wait for hgih quality data to act.
+Being a data skeptic, does not mean one must wait for high quality data to act.
 One must wait for high quality information.
-Yet, in teh current crisis, we have just that.
+Yet, in the current crisis, we have just that.
 Indeed, epidemiologists and public health officials have a strong understanding of the basic facts of pandemics and disease spread.
-Social distancing and
-
-Moreover, the risks of interventions to por risk   the data and therefore
-Of course, there are tradeoffs and long-term stay-at-home orders may have negative economic consequences.
-This is beyond the scope of the current article.
-While not the main thrust of this article, we are well aware that discussions of data quality can sometimes be misinterpreted as claims that we do not have the requisite information to act immediately and decisively.
-It is that point that we are refuting.
-It is simply not the case.
+Social distancing, contact tracing and mass testing are all important.
+Individuals and populations with low quality data and only basic information will make decisions that respect inherent risks in unknown unknowns.
+Of course, there are trade-offs and long-term stay-at-home orders will have negative economic consequences.
+A full discussion of these topics is beyond the scope of the current article; we merely bring up the matter because we are well aware that discussions of data quality can sometimes be misinterpreted as claims that we do not have the requisite information to act immediately and decisively.
+This is simply not the case.
 
@@
 Governments are pushing increased testing capacity and contact tracing.
-Contact tracing is incredibly useful for identifying carriers early and preventing further spreading.
-I would argue that, after the current wave, understanding prevalence is also key.
+Contact tracing is incredibly useful for identifying carriers early and preventing the disease from spreading.
+We would argue that, after the current wave, understanding prevalence is also key.
 We know that the probability of an outbreak is a function of current prevalence and network connectivity.
-Within the SIR framework, there is a critical threshold for  prevalence below which the chance of a full-scale outbreak is near-zero and above the likelihood jumps rapidly (cite).
-Therefore, knowing prevalence will help governments determine long-term community-level risk and allow for more targeted interventions -- shutting down only certain locations when necessary.
+Epidemic critical thresholds have been derived for many infectious disease models~\cite{Pastor2001,Newman2002,Parshani2010}.
+Knowing prevalence will help governments determine long-term community-level risk and allow for more targeted interventions -- shutting down only certain locations when necessary.
 Once we ``return to normal''\footnote{or at least the new normal}, it is clear that testing strategies should include a focus on early detection to help head off future outbreaks.
@@
 Random testing removes these effect modifiers, giving governments more information to fight the disease.
-This is not to say random testing should replace current testing practices; only that the random testing should supplement and will provide valuable information even in small numbers.
+This is not to say random testing should replace current testing practices; only that the random testing should supplement and will provide valuable additional information even with moderate sample sizes.
 
@@
 
-There is nothing routine about COVID-19.  The resulting statistical questions are
-, ``routine statistical questions are less common than questionable statistical routines.''
+There is nothing routine about COVID-19, including the corresponding statistical questions.
+The goal of this paper was to point out questional statistical routines and how observed case-count data can be misleading.
+Indeed, case-count trajectories are being reported daily by governments and broadcast by the media.
+The precision in the reported numbers give the illusion of accuracy when what needs to be understood is the level of uncertainty.
+Selection bias not only means uncertainty, it also implies analysts feel certain about incorrect conclusions.
 
-In current crisis, data quality is often tied to data quantity due to rationing system.
+Not only that, in current crisis data quality is often tied to data quantity due to rationing system.
 Increasing testing under test rationing lowers correlation.
 There is a lot of hidden information in the selection indicator $I_j$.
-Up until recently, an individual had to first \emph{want} to be tested, then call their primary health care provider who decided whether they met thresholds for testing.
-Currently, we are unaware of reliable information on the number of cases where.
+Indeed, individuals must first seek testing and then call their primary health care provider must decide whether they met thresholds for testing.
+Our analysis shows the limitations and caveats that come from analysis of this data in terms of estimation of prevalence, rates of change, and cross-country comparisons.
+We show how one can build simple sensitivity analyses to understand the potential bias in case-count data using surveys that will become more prevalent in the next few months.
 
-We show how one can build sensitivity analyses to understand the potential bias in the observational data.
-These biases depend significantly on the interaction between self-selection and measurement-error.
+One potential approach is via model-based extensions of the mentioned SIR models to try and capture both selection bias and measurement-error issues (cite); however, without strong assumptions on the selection mechanism, the estimates are often not nonparametrically identifiable.
+As Cox and Hinckley (XXXX, p.
+96) write, when an issue ``cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically.'' In the absence of randomized experiments, the best route forward using any approach are detailed sensitivity analyses and humility in conclusions drawn from such data sources.
 
-Interventions in this area are designed experiments.  Unlike Fischer's null, here the objectives are settings, statisticians.
-
-The CDC reports 413,867 total tests performed in the US as of April 25th.
-
-Account for data quality in models by incorporating auxiliary information.  Perform advanced sensitivity analysis.  Forecasts should be called coutnerfactual forecasts.
-
-Poststratifiers and systematic measurement error.
-
-A critical question is whether there are alternative data streams that may be leveraged to understand the handling of the COVID-19 pandemic in the US.
-
-Extensions of the SIR models can try and capture these issues; however, without strong assumptions on the selection mechanism, the estimates are often not nonparametrically identifiable.
-As Cox and Hinckley (XXXX, p.
-96) identified, when an issue ``cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically.'' In the absence of randomized experiments, the best route forward are detailed sensitivity analyses and humility in conclusions drawn from such data sources.
-
-Social capital and models.
+\bibliographystyle{plainnat}
+\bibliography{covid-refs}
 
@@
 
-\subsection{Simple cases}
-
-We consider two simple cases here.  First, we assume no false positive results, i.e., set $FP=0$.  Then $\sigma_{PZ} = \bar Y \sqrt{2 FN}$.  Then the math simplifies:
-$$
-\bar y_n^\star - \bar Y = \bar Y \sqrt{\frac{1-f}{f}} \left[\rho_{I,Y} \sqrt{\frac{1-\bar Y}{\bar Y}} + \sqrt{FN} \left( \sqrt{2} \rho_{I, PZ} - \sqrt{\frac{f}{1-f}} \sqrt{FN} \right)  \right]
-$$
-The sign of the error therefore depends on true data quality ($\rho_{I, Y}$), odds ratio, observed data quality~$\rho_{I,PZ}$, false negative rates (FN), and sample fraction $f$.
-
-Second, we assume no false negative results, i.e., set $FN=0$.  Then $\sigma_{PZ} = \sqrt{2 FP \bar Y (1-\bar Y)}$.  Then the math simplifies:
-$$
-\bar y_n^\star - \bar Y = \sqrt{\bar Y (1-\bar Y)} \sqrt{\frac{1-f}{f}} \left[\rho_{I,Y} +  \sqrt{FP} \left( \sqrt{2} \rho_{I, PZ} + \sqrt{\frac{f}{1-f}} \sqrt{FP} \sqrt{1-\bar Y} \right)  \right]
-$$
-The sign of the error therefore depends on true data quality ($\rho_{I, Y}$),  observed data quality~$\rho_{I,PZ}$, false positive rates (FP), one minus prevalence ($1-\bar Y$), and sample fraction $f$.
-
-\subsection{Adjusted estimator}
-
-Define $\bar y_n^{\star \star} = \bar y_n^\star + FN \bar y_n^\star - FP(1-\bar y_n^\star)$; then the error is
-$$
-\bar y_n^{\star \star} - \bar Y = \sqrt{\frac{1-f}{f}} \left[ \rho_{I, Y} \sigma_{Y} (1+FP+FN)  + \rho_{I, PZ} \sqrt{ 2 \bar Y \left( FP (1-\bar Y) + FN \bar Y \right)} \right]
-$$
-In the case when false negative rate is zero, we have
-$$
-\bar y_n^{\star \star} - \bar Y = \sqrt{\frac{1-f}{f}} \sqrt{\bar Y(1-\bar Y)} \left[ \rho_{I, Y} (1+FP) + \rho_{I, PZ} \sqrt{ 2 FP } \right]
-$$
-In the case when false positive rate is zero, we have
-$$
-\bar y_n^{\star \star} - \bar Y = \sqrt{\frac{1-f}{f}} \sqrt{\bar Y(1-\bar Y)} \left[ \rho_{I, Y} (1+FN) + \rho_{I, PZ} \sqrt{ 2 FN } \times \sqrt{\frac{\bar Y}{1-\bar Y}} \right]
-$$
-
 \subsection{Estimate of observed data quality}
 
+By definition, we have
 $$
@@
 
-Let ${\bf u} = (u_1,u_2) \in \mathbb{R}^2$ and $g({\bf u}) = \frac{u_2}{u_1}$, i.e., a differentiable function $g:\mathbb{R}^2 \to \mathbb{R}$.
+Let ${\bf u} = (u_{t-1},u_t) \in \mathbb{R}^2$ and $g({\bf u}) = \frac{u_t}{u_{t-1}}$, i.e., a differentiable function $g:\mathbb{R}^2 \to \mathbb{R}$.
 Centering a Taylor series expansion of second-order around coordinates $(U_2, U_1) \in \mathbb{R}^2$ yields
@@
 \begin{aligned}
-g({\bf u}) =& g(U_1, U_2) - \frac{U_2}{U_1^2} (u_1 - U_1) + \frac{1}{U_1} (u_2 - U_2) \\
-&+ \frac{1}{2} \left[ \frac{2 U_2}{U_1^3} (u_1 - U_1)^2 + 0 \times (u_2 - U_2)^2 - 2 \times (u_1 - U_1) (u_2 - U_2) \frac{1}{U_1^2} \right]
+g({\bf u}) =& g(U_{t-1}, U_t) - \frac{U_t}{U_{t-1}^2} (u_{t-1} - U_{t-1}) + \frac{1}{U_{t-1}} (u_t - U_t) \\
+&+ \frac{1}{2} \left[ \frac{2 U_t}{U_{t-1}^3} (u_{t-1} - U_{t-1})^2 + 0 \times (u_t - U_t)^2 - 2 \times (u_{t-1} - U_{t-1}) (u_t - U_t) \frac{1}{U_t^2} \right]
 \end{aligned}
 $$
-Plugging in $(\bar y_1, \bar y_2)$ for $(u_1, u_2)$ and $(\bar Y_1, \bar Y_2)$ for $(U_1, U_2)$ yields
+Plugging in $(\bar y_{t-1}, \bar y_t)$ for $(u_{t-1}, u_t)$ and $(\bar Y_{t-1}, \bar Y_t)$ for $(U_{t-1}, U_t)$ yields the $\frac{\bar y_t}{\bar y_{t-1}} - \frac{\bar Y_t}{\bar Y_{t-1}} $ is equal to
 $$
 \begin{aligned}
-\frac{\bar y_2}{\bar y_1} - \frac{\bar Y_2}{\bar Y_1} =& - \frac{\bar Y_2}{\bar Y_1^2} (\bar y_1 - \bar Y_1) + \frac{1}{\bar Y_1} (\bar y_2 - \bar Y_2) \\
-&+ \frac{\bar Y_2}{\bar Y_1^3} (\bar y_1 - \bar Y_1)^2 -  (\bar y_1 - \bar Y_1) (\bar y_2 - \bar Y_2) \frac{1}{\bar Y_1^2} \\
-&= \frac{\bar Y_2}{\bar Y_1} \bigg[  \rho_{I_2,Y_2} \sqrt{\frac{1-f_2}{f_2}} CV (Y_2)  -\rho_{I_1,Y_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \\
-&+ \rho^2_{I_1,Y_1} \frac{1-f_1}{f_1} CV^2 (Y_1) -  \rho_{I_1,Y_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \times \rho_{I_2,Y_2} \sqrt{\frac{1-f_2}{f_2}} CV (Y_2)   \bigg] \\
-&= \frac{\bar Y_2}{\bar Y_1} \bigg[ \rho_{I_2,Y_2} \sqrt{\frac{1-f_2}{f_2}} CV (Y_2)  -\rho_{I_1,Y_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \bigg] \left[ 1 - \rho_{I_1,Y_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \right]
+=&
+- \frac{\bar Y_t}{\bar Y_{t-1}^2} (\bar y_{t-1} - \bar Y_{t-1}) + \frac{1}{\bar Y_{t-1}} (\bar y_t - \bar Y_t) \\
+&+ \frac{\bar Y_t}{\bar Y_{t-1}^3} (\bar y_{t-1} - \bar Y_{t-1})^2 -  (\bar y_{t-1} - \bar Y_{t-1}) (\bar y_t - \bar Y_t) \frac{1}{\bar Y_{t-1}^2} \\
+&= \frac{\bar Y_t}{\bar Y_{t-1}} \bigg[  \rho_{I_t,Y_t} \sqrt{\frac{1-f_t}{f_t}} CV (Y_t)  -\rho_{I_{t-1},Y_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \\
+&+ \rho^2_{I_{t-1},Y_{t-1}} \frac{1-f_{t-1}}{f_{t-1}} CV^2 (Y_{t-1}) -  \rho_{I_{t-1},Y_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \times
+\rho_{I_t,Y_t} \sqrt{\frac{1-f_t}{f_t}} CV (Y_t)   \bigg] \\
+&= \frac{\bar Y_t}{\bar Y_{t-1}} \bigg[ \rho_{I_t,Y_t} \sqrt{\frac{1-f_t}{f_t}} CV (Y_t)  -\rho_{I_{t-1},Y_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \bigg] \left[ 1 - \rho_{I_{t-1},Y_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \right]
 \end{aligned}
@@
 Under measurement-error, the extra terms $D_{2}$ and $D_1$ can be inserted in the correct locations.
+
+\subsection{Estimation of effective reproduction number}
+
+Let
+$$
+\delta_{t} = \rho_{I_t,Y_t} \sqrt{\frac{1-f_t}{f_t}} CV (Y_t)  -\rho_{I_{t-1},Y_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}).
+$$
 

### 2020-05-11 methods/sir_example.R

@@ -71,7 +71,27 @@ bias_fn <- function(M, bary, f, FP = 0.01, FN = 0.15) {
   
   bias = (lin_term+quad_term) * cur_bary/lag_bary
   
-  Rt = 1+log(cur_bary/lag_bary)/7
+  return(list("r"=cur_bary/lag_bary, "bias" = bias))
+}
+
+logbias_fn <- function(M, bary, f, FP = 0.01, FN = 0.15) {
+  const = (bary*M + (1-bary))
+  f0 = f/const
+  f1 = M*f0
+  Delta = f1 - f0
+  D_M = 1 + FP + FN - Delta * bary/(1-bary) * (FP*(1-bary) + FN * bary)/ (f0 * (1-bary) + f1 * bary)
+  rho = Delta*sqrt(bary*(1-bary))/sqrt(f*(1-f))
+  error = rho*sqrt((1-f)/f) * sqrt(1-bary)/sqrt(bary)*D_M
+  lag_error = error[1:(length(error)-1)]
+  cur_error = error[2:(length(error))]
+  
+  lin_term = cur_error - lag_error
+  quad_term = lag_error^2 - cur_error*lag_error
+  
+  cur_bary = bary[2:length(error)]
+  lag_bary = bary[1:(length(error)-1)]
+  
+  Rt = 1+log(cur_bary/lag_bary)/7 + log(survfrac)
   logbias = log(1+lin_term+quad_term)/7
   
   return(list("r"=cur_bary/lag_bary, "bias" = bias,

### 2020-05-12 write-up/covid-writeup.tex
Update covid-writeup.tex
final pre send out draft update

@@
 The hope is that $y_t/y_{t-1}$ is a good proxy for $K_t/K_{t-1}$.
-Let $\hat R_t = 1 + \frac{1}{\gamma} \log \left( y_t / y_{t-1} \right)$ be the estimated effective reproduction number on day $t$.
-Under SRS with the same sample size at each step, this is true for sufficiently large sample size.
-However, similar to Section~\ref{section:rates}, the proxy has error in both numerator and denominator that may cause issues.
+However, if we think about estimating number $K_t$ under SRS of new cases among those susceptible on day $t$, the natural estimator is $S_t \bar y_t$.
+Unfortunately the number of individuals susceptible on day $t$ is unknown.
+Here, we assume that $\hat R_t = 1 + \frac{1}{\gamma} \log \left( \bar y_t / \bar y_{t-1} \right)$ be the estimated effective reproduction number on day $t$.
+This estimator adjusts for sample size variation across days but not the varying population size.
+The estimator is close in spirit to the estimator derived under assuming the process is fully observed.
+Similar to Section~\ref{section:rates}, the proxy has error in both numerator and denominator that may cause issues.
 We again wish to express the statistical error of $\hat R_t - R_t$ in useful terms.
@@
 \frac{1}{\gamma}\log &\bigg( 1 + \bigg[ \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t)  -\rho_{I_{t-1},K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t) \bigg] \\
-&\times \left[ 1 - \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t) \right] \bigg)
+&\times \left[ 1 - \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t) \right] \bigg) - \frac{1}{\gamma} \log \left( \frac{S_t}{S_{t-1}} \right).
 \end{aligned}
@@
 Moreover, the error depends not on aggregate prevalence $\bar Y_t$ but fraction of new cases $\bar K_t$.
+The final term is the error due to varying population sizes.
+Since $S_t \leq S_{t-1}$, the final term $- \log (S_t/S_{t-1}) \geq 0$.
 Figure~\ref{fig:r0-bias} displays the bias for the effective reproductive number and how it is impacted by the relative sampling fraction.
-Results are similar to the ratio estimate, but the impact on the underestimation is more severe due to the logarithmic transformation.
+Here for simplicity, we assume the fraction of the population that is susceptible remains large and near the total population size.
+Under these assumption, the potential bias is equivalent to that for the ratio estimator adjusted to be on the logarithm scale.
+We note that changes to this assumption can drastically impact the potential bias.
 
@@
 One potential approach is via model-based extensions of the mentioned SIR models to try and capture both selection bias and measurement-error issues (cite); however, without strong assumptions on the selection mechanism, the estimates are often not nonparametrically identifiable.
-As Cox and Hinckley (XXXX, p.
-96) write, when an issue ``cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically.'' In the absence of randomized experiments, the best route forward using any approach are detailed sensitivity analyses and humility in conclusions drawn from such data sources.
+Indeed, when an issue ``cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically'' ~\cite[pp.
+96]{CoxHink74}.
+In the absence of randomized experiments, the best route forward using any approach are detailed sensitivity analyses and humility in conclusions drawn from such data sources.
 
@@
 \subsection{Ratio estimator}
+\label{app:ratio}
 
@@
 $$
-\delta_{t} = \rho_{I_t,Y_t} \sqrt{\frac{1-f_t}{f_t}} CV (Y_t)  -\rho_{I_{t-1},Y_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}).
+\delta_t := \bigg[ \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t)  -\rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_t}{f_t}} CV (K_{t-1}) \bigg].
+$$
+Then results from Section~\ref{app:ratio} show that the estimate of the number of new cases on day t is given by
+$$
+\frac{S_t \cdot \bar y_t}{S_{t-1} \cdot \bar y_{t-1}} =
+\frac{K_t}{K_{t-1}} \left( 1 + \delta_t \times \left[ 1 - \rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_t}{f_t}} CV (K_{t-1}) \right] \right)
+$$
+Then setting $e_t = \delta_t \times [1 - \rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_t}{f_t}} CV (K_{t-1}) ]$, we have
+$$
+\begin{aligned}
+\log \left( \frac{S_t \bar y_t}{S_{t-1} \bar y_{t-1}} \right) - \log \left( \frac{K_t}{K_{t-1}} \right) &= \log (1 + e_t) \\
+\log \left( \frac{\bar y_t}{\bar y_{t-1}} \right) - \log \left( \frac{K_t}{K_{t-1}} \right) &= 1 + e_t - \log \left( \frac{S_t}{S_{t-1}} \right) \\
+1 + \frac{1}{\gamma} \log \left( \frac{\bar y_t}{\bar y_{t-1}} \right) - \left[ 1 + \frac{1}{\gamma} \log \left( \frac{K_t}{K_{t-1}} \right) \right] &= \frac{1}{\gamma} \left[ \log \left( 1 + e_t \right) - \log \left( \frac{S_{t}}{S_{t-1}} \right) \right] \\
+\Rightarrow \hat R_t - R_t &= \frac{1}{\gamma} \left[ \log \left( 1 + e_t \right) - \log \left( \frac{S_{t}}{S_{t-1}} \right) \right]
+\end{aligned}
 $$

### 2020-05-12 methods/sir_example.R

@@ -48,9 +48,11 @@ dev.off()
 f = 0.02
 M = 4
 days = 1:365
-FP = 0.01; FN = 0.10
-bary1 = -diff(out1$S)
-bary2 = -diff(out2$S)
+FP = 0.01; FN = 0.1
+bary1_new = -diff(out1$S)
+bary1_all = out1$I
+bary2_new = -diff(out2$S)
+bary2_all = out2$I
 
 bias_fn <- function(M, bary, f, FP = 0.01, FN = 0.15) {
   const = (bary*M + (1-bary))
@@ -71,54 +73,38 @@ bias_fn <- function(M, bary, f, FP = 0.01, FN = 0.15) {
   
   bias = (lin_term+quad_term) * cur_bary/lag_bary
   
-  return(list("r"=cur_bary/lag_bary, "bias" = bias))
-}
-
-logbias_fn <- function(M, bary, f, FP = 0.01, FN = 0.15) {
-  const = (bary*M + (1-bary))
-  f0 = f/const
-  f1 = M*f0
-  Delta = f1 - f0
-  D_M = 1 + FP + FN - Delta * bary/(1-bary) * (FP*(1-bary) + FN * bary)/ (f0 * (1-bary) + f1 * bary)
-  rho = Delta*sqrt(bary*(1-bary))/sqrt(f*(1-f))
-  error = rho*sqrt((1-f)/f) * sqrt(1-bary)/sqrt(bary)*D_M
-  lag_error = error[1:(length(error)-1)]
-  cur_error = error[2:(length(error))]
-  
-  lin_term = cur_error - lag_error
-  quad_term = lag_error^2 - cur_error*lag_error
-  
-  cur_bary = bary[2:length(error)]
-  lag_bary = bary[1:(length(error)-1)]
-  
-  Rt = 1+log(cur_bary/lag_bary)/7 + log(survfrac)
+  Rt = 1+log(cur_bary/lag_bary)/7
   logbias = log(1+lin_term+quad_term)/7
   
   return(list("r"=cur_bary/lag_bary, "bias" = bias,
               "Rt"=Rt, "logbias"=logbias))
 }
 
-bias1 = bias_fn(M, bary1, f)
-bias2 = bias_fn(M, bary2, f)
+bias1_all = bias_fn(M, bary1_all, f)
+bias1_new = bias_fn(M, bary1_new, f)
+bias2_all = bias_fn(M, bary2_all, f)
[…]

### 2020-05-13 write-up/covid-writeup.tex
update covid paper

@@
 \DeclareGraphicsExtensions{.eps, .ps}
-\usepackage{soul,color,hyperref}
+\usepackage{hyperref}
 \usepackage{amsmath, amsthm, amsfonts}
@@
 
-This article discusses the relationship between three statistical concepts: selection bias, measurement-error, and the long-forgotten population size.
+This article discusses the relationship between three statistical concepts: selection bias, measurement-error, and the too often forgotten population size.
 We clarify mathematically why the situation is much more complex than it first appears.
@@
 Fifth, we discuss a sensitivity analysis method for estimating data quality and its use in understanding the potential biases in COVID-19 case-count analyses.
-We then end the paper with a discussion of the benefits of randomized testing, use of auxiliary information is most relevant, and the distinction between data quality and decision making in the current pandemic.
+We then end the paper with a discussion of the benefits of randomized testing, use of auxiliary information to improve over simple random sampling, and the distinction between data quality and decision making in the current pandemic.
 
@@
 
-Here we highlight the core insights from~\cite{Meng2018} that are relevant to the COVID-19 crisis.  First, comparing the mean-squared error under selection mechanism $\I$ and SRS, we see that
+Here we highlight two key insights from~\cite{Meng2018} that are relevant to the COVID-19 crisis.  First, comparing the mean-squared error under selection mechanism $\I$ and SRS, we see that
 $$
@@
 In discussions of inaccurate testing, the standard assumption is measurement-error leads to parameter attenuation.
-When paired with selection bias, however, the two sources of error become entangled, and resulting biases can become magnified, muted, or even switch signs.
+When paired with selection bias, however, the two sources of error become entangled, and resulting errors can become magnified, muted, or even switch signs.
 
@@
 We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.
-If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^\star = Y_j(1-P_j) + (1-Y_j) P_j$.
+If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^{\star} = Y_j(1-P_j) + (1-Y_j) P_j$.
 Suppose disease prevalence was estimated as the fraction who tested positive for COVID-19, i.e., $\bar y_n^\star = \frac{\sum_{i=1}^N I_j Y_j^\star}{\sum_{i=1}^N I_j}$.
@@
 $$
-where $Z = 1-2Y$.
+where $Z = 1-2Y$; see Appendix~\ref{app:imperfect} for the derivation.
 The first term represents the perfect testing regime, the second term represents the interaction between imperfect testing and selection bias, while the third time represents the bias due to imperfect testing.
@@
 The final term in brackets is the \emph{measurement-error adjustment to data quality} which is a complex function of sampling rate differential, the odds ratio, and the ratio of measurement-error interaction with prevalence and sampling rates interaction with prevalence.
-Note $\sgn(\Delta) = \sgn(\rho_{I,Y})$ by equation~\ref{eq:binaryrho} in the appendix.
-This implies the measurement-error adjustment either shrinks the data quality measure toward zero or reverses its sign.
+Note $\sgn(\Delta) = \sgn(\rho_{I,Y})$ by equation~\ref{eq:binaryrho} in the appendix, which implies the measurement-error adjustment either shrinks the data quality measure toward zero or reverses its sign.
 
@@
 Consider again the current COVID-19 pandemic.
-We continue to assume the ratio of conditional selection rates $f_1/f_0$ is equal to 1.5.
+For now, we continue to assume the ratio of conditional selection rates $f_1/f_0$ is equal to 1.5.
 In Section~\ref{section:est_dq}, we discuss recent research suggesting a false negative rate around 17.2\% and a false positive rate around 0.05\%.
@@
 This is merely to demonstrate that in some cases we see a huge increase in MSE and in other settings we have a huge decrease in MSE.
-What drives this is the false negative rate interaction with prevalence and sampling rates.
+What drives this is the false positive and negative rate interaction with prevalence and sampling rates.
 Therefore, whether we are better or worse off with respect to the MSE is a very difficult question to answer.
@@
 Figure~\ref{fig:heatmap} presents the measurement-error data quality adjustment as a function of the relative frequency and odds ratio.
-We see that the adjustment can be both positive and negative as well as greater than or less than zero.
+We see that the adjustment can be both positive and negative as well as a range of magnitudes.
 
@@
 \includegraphics[width = 0.4\textwidth]{../methods/figs/mem_heatmap_article.png}
-\caption{Measurement-error data quality adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.15$ and $FN=0.10$.
-Color scaled so blue = $-6$, white = 0, and red = $6$.}
+\caption{Measurement-error data quality adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.005$ and $FN=0.172$.
+Color scaled so blue = $-10$, white = $0$, and red = $10$.}
 \label{fig:heatmap}
@@
 Here we consider the ratio estimators for the relative change in the prevalence rate.
-We let $\bar Y_{t-1}$ and $\bar Y_{t}$ denotes the prevalence at time step $t-1$ and $t$ respectively (i.e., prevalence on two consecutive days).
+We let $\bar Y_{t-1}$ and $\bar Y_{t}$ denotes the prevalence (across the entire population) at time step $t-1$ and $t$ respectively (i.e., prevalence on two consecutive days).
 Then the ratio estimator is given by $r = \bar y_t / \bar y_{t-1}$.
@@
 where $S_t$ is the number of susceptible individuals at time $t$, $I_t$ is the number of infected individuals at time $t$, $N$ is the total poulation size, and $R_t$ is the number of removed (recovered or deceased) individuals at time $t$.
-Figure~\ref{fig:fracpop} presents the fraction of population newly infected individuals at each time step for $\beta = 1.4$ and $\gamma = 0.2$.
+Figure~\ref{fig:fracpop} presents the fraction of population newly infected individuals at each time step for $\beta = 1.4$ and $\gamma = 0.2$ (black line).
 In each case, the rate of change in prevalence is likely to be  overestimated prior to the peak and then underestimated afterwards.
@@
 Of course, we do not observe $K_t$ and $K_{t-1}$ but noisy proxies $y_t$ and $y_{t-1}$, i.e., the observed case-count on day $t$ and $t-1$ respectively.
-The hope is that $y_t/y_{t-1}$ is a good proxy for $K_t/K_{t-1}$.
+The hope is that $y_t/y_{t-1}$ is a good proxy for $K_t/K_{t-1}$ where here $y_t$ isc conceptualized as the number of new cases on day $t$.
 However, if we think about estimating number $K_t$ under SRS of new cases among those susceptible on day $t$, the natural estimator is $S_t \bar y_t$.
 Unfortunately the number of individuals susceptible on day $t$ is unknown.
-Here, we assume that $\hat R_t = 1 + \frac{1}{\gamma} \log \left( \bar y_t / \bar y_{t-1} \right)$ be the estimated effective reproduction number on day $t$.
+Here, we assume that $\hat R_t = 1 + \frac{1}{\gamma} \log \left( \bar y_t / \bar y_{t-1} \right)$ is the estimated effective reproduction number on day $t$.
 This estimator adjusts for sample size variation across days but not the varying population size.
@@
 \begin{aligned}
-\frac{1}{\gamma}\log &\bigg( 1 + \bigg[ \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t)  -\rho_{I_{t-1},K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t) \bigg] \\
-&\times \left[ 1 - \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t) \right] \bigg) - \frac{1}{\gamma} \log \left( \frac{S_t}{S_{t-1}} \right).
+\frac{1}{\gamma}\log &\bigg( 1 + \bigg[ \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t)  -\rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (K_{t-1}) \bigg] \\
+&\times \left[ 1 - \rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (K_{t-1}) \right] \bigg) - \frac{1}{\gamma} \log \left( \frac{S_t}{S_{t-1}} \right).
 \end{aligned}
@@
 Here for simplicity, we assume the fraction of the population that is susceptible remains large and near the total population size.
-Under these assumption, the potential bias is equivalent to that for the ratio estimator adjusted to be on the logarithm scale.
-We note that changes to this assumption can drastically impact the potential bias.
+Under these assumption, the potential bias is similar to that for the ratio estimator adjusted to be on the logarithm scale.
+Changes to this assumption can impact the potential bias.
+The key difference between Figure~\ref{fig:ratio-bias} and Figure~\ref{fig:r0-bias} is that prevalence depends on the fraction of the population infected at time $t$ while $\hat R_t$ depends on the fraction of new cases in the population.
+This leads to differences in when the bias is most pronounced.
 
@@
 \end{subfigure}
-\caption{Potential bias due to self-selection and measurement-error from a deterministic SIR model with $\beta = 1.4$ and $\gamma = 0.2$.
+\caption{Potential bias due to self-selection and measurement-error from a deterministic SIR model with $\beta = 1.4$ (black) and $\gamma = 0.2$ (red).
 Here we assume $f = 0.02$, $FP = 0.01$, $FN = 0.15$, and use a range of relative sampling fraction $M = f_1/f_0$.}
@@
 
-Here we show how selection bias can even creep into clinical trial analysis.
+Here we show how selection bias and measurement-error can creep into clinical trial analysis.
 The key concern is whether clinical trials on COVID-19 recruit from the pool of individuals who have tested positive for COVID-19 or whether they sample randomly from the population, test, and then recruit from this subset of tested individuals.
@@
 In such situations, the treatment effect estimates will tend to be overly optimistic as most individuals recruited are potentially higher on the severity index than the population average.
-Not only that, but for fixed data quality, the error compared to SRS in terms of effect estimation depends scales with the population size (i.e., the law of large populations).
+Not only that, but for fixed data quality, the error compared to SRS in terms of effect estimation scales with the population size (i.e., the law of large populations).
 
@@
 Therefore, for marginal treatment effects to be interpretable, there must be a well-defined population.
-As often our main interest is in causal effects on the population of COVID-19 positive.
+Most often, our main interest is in causal effects on the population of COVID-19 positive.
 Randomization within the clinical trial design yields internal validity, but we also need external validity to estimate the correct marginal effect of interest.
@@
 $$
-allows us to build Z-scores that account for the true sample size.  While still not perfect, this allows us to build Z-scores that are not overly confident due to selection bias and measurement error.
+allows us to build Z-scores that account for the true sample size.
+While still not perfect, this allows us to build Z-scores that are not overly confident due to selection bias and measurement error.
+See Section~\ref{section:est_dq} for discussion of how to estimate data quality.
 
@@
 Consider two countries (A and B) in which the peak occurs 2 weeks prior for country A then country B but other.
-Figure~\ref{fig:comparison} presents such a comparison where each country's disease trajectory follows an SIR model as described in Section XX with different parameters.
+Figure~\ref{fig:comparison} presents such a comparison where each country's disease trajectory follows an SIR model as described in Section~\ref{section:rates} with different parameters.
 Figure \ref{fig:eff} shows how the errors interact in complex ways.
@@
 That is, sample individuals at random in strata $h$ with size proportional to $P_h \sqrt{\bar Y_h \times (1-\bar Y_h)}$ where $P_h$ is the strata frequency in the population and $\bar Y_h$ the prevalence in the strata.
-Of course, $P_h$ and $\bar Y_h$ are not known a priori; however, if we can get decent estimates, we can see huge benefits in settings where the prevalence varies widely by strata.
+Of course, $P_h$ and $\bar Y_h$ are not known a priori; however, if decent estimates can be obtained, we can see huge benefits in settings where the prevalence varies widely by strata.
 One could go attempt to go further and incorporate contact network information.
@@
 There are whole fields dedicated to careful analysis in such settings (e.g., causal inference from observational studies).
-Being a data skeptic, does not mean one must wait for high quality data to act.
-One must wait for high quality information.
+Being a data skeptic, does not mean one must wait for high quality data to act, especially in high risk scenarios.
+One aims to have high quality information.
 Yet, in the current crisis, we have just that.
@@
 There is nothing routine about COVID-19, including the corresponding statistical questions.
-The goal of this paper was to point out questional statistical routines and how observed case-count data can be misleading.
+The goal of this paper was to point out questionable statistical routines.
 Indeed, case-count trajectories are being reported daily by governments and broadcast by the media.
-The precision in the reported numbers give the illusion of accuracy when what needs to be understood is the level of uncertainty.
+The precision in the reported numbers gives the illusion of accuracy when what needs to be understood is the level of uncertainty.
 Selection bias not only means uncertainty, it also implies analysts feel certain about incorrect conclusions.
@@
 
-One potential approach is via model-based extensions of the mentioned SIR models to try and capture both selection bias and measurement-error issues (cite); however, without strong assumptions on the selection mechanism, the estimates are often not nonparametrically identifiable.
+One alternative approach is model-based extensions of SIR models to try and capture both selection bias and measurement-error issues; however, without strong assumptions on the selection mechanism, the estimates are often not nonparametrically identifiable.
 Indeed, when an issue ``cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically'' ~\cite[pp.
 96]{CoxHink74}.
-In the absence of randomized experiments, the best route forward using any approach are detailed sensitivity analyses and humility in conclusions drawn from such data sources.
+In the absence of randomized experiments, the best route forward using any approach is the use of detailed sensitivity analyses and humility in conclusions drawn from such data sources.
 
@@
 \section{Imperfect testing: derivation}
-
+\label{app:imperfect}
 We considered the mean estimator

### 2020-05-13 methods/mem_heatmap.R

@@ -15,8 +15,8 @@ seq_oddsratio = exp(seq(log(0.01), log(50), length.out = 100))
 seq_M = exp(seq(log(0.5),log(20), length.out = 100))
 result = matrix(nrow = length(seq_oddsratio), ncol = length(seq_M))
 result_unadj = matrix(nrow = length(seq_oddsratio), ncol = length(seq_M))
-FP = 0.01
-FN = 0.15
+FP = 0.005
+FN = 0.172
 rho = 0.01
 f = 0.001
 

### 2020-05-16 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 In this paper, we attempt to clarify this interaction.
-Through simple calculations, we demonstrate pitfalls and paradoxes that can arise when considering case-count data in the presence of selection bias and measurement-error.
+Through simple calculations, we demonstrate pitfalls and paradoxes that can arise when considering case-count data in the presence of selection bias and measurement error.
 The discussion guides several suggestions on how to improve current case-count reporting.
@@
 The World Health Organization has declared the coronavirus disease 2019 (COVID-19) a public health emergency.
-As of April 27th, 2020, a total of 2,993,00 cases have been confirmed worldwide.
+As of April 27th, 2020, a total of 2,993,000 cases have been confirmed worldwide.
 As of that afternoon, the New York Times reports at least 965,214 people across the United States have tested positive for the virus, and at least 49,465 patients with the virus have died.
-Aggressive policies had been put in place across the US with at least 50\% of the US population officially urged as of late April to stay home via state-wide executive actions.
+Aggressive policies have been put in place across the US with at least 50\% of the US population officially urged as of late April to stay home via state-wide executive actions.
 
@@
 In many cases, variations on the SIR models have been used to draw inferences about infection rates and intensity.
-Parameter inference often uses observed case-count and/or observed COVID-19 related deaths to infer latent trajectories of the pandemic.
+Parametric inference often uses observed case-count and/or observed COVID-19 related deaths to infer latent trajectories of the pandemic.
 
 The goal of this paper is to express reservations at the use of case-counts as a proxy for disease prevalence and intensity as well as its use as direct input into estimation of standard epidemiological models for inference and forecasting.
-The reason is straightforward: current models do not account simultaneously for selection bias and measurement-error.
+The reason is straightforward: current models do not account simultaneously for selection bias and measurement error.
 Selection bias enters due to differences in testing strategies across countries and states.
@@
 Moreover, testing requires the individual to self-select into asking for a test and then receiving approval by their doctor.
-Demands for increased testing capacity, while laudable, ignore the issues of self-selection and measurement-error.
+Demands for increased testing capacity, while laudable, ignore the issues of self-selection and measurement error.
 While increasing testing capacity increases data quantity, there is no guarantee for increased data quality.
@@
 In this paper, we will demonstrate the complex interaction between these two fundamental concepts.
-In classical statistics, measurement-error is often associated with parameter attenuation; we will show, however, that the combination with self-selection can cause bias in parameter estimates to change sign and increase/decrease in magnitude.
+In classical statistics, measurement error is often associated with parameter attenuation; we will show, however, that the combination with self-selection can cause bias in parameter estimates to change sign and increase/decrease in magnitude.
 Moreover, the bias may depend on the sampling fraction, prevalence, and population size.
-Without a complete understanding of these interations, we are destined to misinterpret case-counts and come to erroneous conclusions.
+Without a complete understanding of these interactions, we are destined to misinterpret case-counts and come to erroneous conclusions.
 We will demonstrate why there is a strong need for more precision and care in presenting COVID-19 analyses to the broader research and non-academic communities.
@@
 
-% Measurement-error discussions.
+% measurement error discussions.
 % Selection bias . Proponents of this critique suggest well-designed studies.
@@
 
-This article discusses the relationship between three statistical concepts: selection bias, measurement-error, and the too often forgotten population size.
+This article discusses the relationship between three statistical concepts: selection bias, measurement error, and the too often forgotten population size.
 We clarify mathematically why the situation is much more complex than it first appears.
@@
 
-The remainder of this paper aims to build upon these fundamental insights by extending the decomposition in two directions: accounting for  measurement-error and the temporal nature of the pandemic.
+The remainder of this paper aims to build upon these fundamental insights by extending the decomposition in two directions: accounting for  measurement error and the temporal nature of the pandemic.
 We then discuss effective sampling methods and data versus decision-making.
@@
 Here we investigate the interplay between imperfect testing and selection bias.
-In discussions of inaccurate testing, the standard assumption is measurement-error leads to parameter attenuation.
+In discussions of inaccurate testing, the standard assumption is measurement error leads to parameter attenuation.
 When paired with selection bias, however, the two sources of error become entangled, and resulting errors can become magnified, muted, or even switch signs.
@@
 First we require some additional notation.
-Let $P_j$ be an indicator of measurement error, equal to $1$ when we mismeasure the outcome and $0$ when we observe the true outcome.
+Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the outcome and $0$ when we observe the true outcome.
 We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.
@@
 The first term represents the perfect testing regime, the second term represents the interaction between imperfect testing and selection bias, while the third time represents the bias due to imperfect testing.
-From here on, we refer to $\rho_{I,Y}$ as the \emph{true data quality}, and $\rho_{I,PZ}$ as the \emph{observed data quality} that accounts for both selection bias and measurement-error.
+From here on, we refer to $\rho_{I,Y}$ as the \emph{true data quality}, and $\rho_{I,PZ}$ as the \emph{observed data quality} that accounts for both selection bias and measurement error.
 In the appendix, we show the sign of $\rho_{I,PZ}$ is the opposite of the sign of $\rho_{I,Y}$, which implies that the observed data quality adds error in the opposite direction from the true data quality.
 
-We start by considering the first two terms and assess whether the sign of the bias can reverse due to the interaction of measurement-error and selection bias.
+We start by considering the first two terms and assess whether the sign of the bias can reverse due to the interaction of measurement error and selection bias.
 To do this, we define the sampling rates differential.
@@
 $$
-The final term in brackets is the \emph{measurement-error adjustment to data quality} which is a complex function of sampling rate differential, the odds ratio, and the ratio of measurement-error interaction with prevalence and sampling rates interaction with prevalence.
-Note $\sgn(\Delta) = \sgn(\rho_{I,Y})$ by equation~\ref{eq:binaryrho} in the appendix, which implies the measurement-error adjustment either shrinks the data quality measure toward zero or reverses its sign.
+The final term in brackets is the \emph{measurement error adjustment to data quality} which is a complex function of sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence.
+Note $\sgn(\Delta) = \sgn(\rho_{I,Y})$ by equation~\ref{eq:binaryrho} in the appendix, which implies the measurement error adjustment either shrinks the data quality measure toward zero or reverses its sign.
 
-While prior investigations have noted the interaction between measurement-error and selection bias~\citep{Beesley2020,Beesley2019,Smeden2019}, the interaction with the sample size relative to the population, i.e., $f$, has largely been ignored.
+While prior investigations have noted the interaction between measurement error and selection bias~\citep{Beesley2020,Beesley2019,Smeden2019}, the interaction with the sample size relative to the population, i.e., $f$, has largely been ignored.
 The above statistical decomposition clarifies the importance of this quantity~$f$.
-In particular, note that the statistical error also includes a bias term due to measurement-error and this term increases as the sampled fraction $f$ increases.
+In particular, note that the statistical error also includes a bias term due to measurement error and this term increases as the sampled fraction $f$ increases.
 Therefore, how the first two terms interact with the final term depends on the fraction of the population sampled.
@@
 \end{equation}
-The first term is the same as before but increased by $(1 + FP + FN)$ to account for the additional uncertainty due to measurement-error.
+The first term is the same as before but increased by $(1 + FP + FN)$ to account for the additional uncertainty due to measurement error.
 The second term is the interaction between selection bias and measurement error.
-Figure~\ref{fig:heatmap} presents the measurement-error data quality adjustment as a function of the relative frequency and odds ratio.
+Figure~\ref{fig:heatmap} presents the measurement error data quality adjustment as a function of the relative frequency and odds ratio.
 We see that the adjustment can be both positive and negative as well as a range of magnitudes.
 
-We next investigate the impact of measurement-error on the effective sample size calculation.  The effective sample size can now be bounded by
+We next investigate the impact of measurement error on the effective sample size calculation.  The effective sample size can now be bounded by
 $$
@@
 \includegraphics[width = 0.4\textwidth]{../methods/figs/mem_heatmap_article.png}
-\caption{Measurement-error data quality adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.005$ and $FN=0.172$.
+\caption{measurement error data quality adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.005$ and $FN=0.172$.
 Color scaled so blue = $-10$, white = $0$, and red = $10$.}
@@
 The above argument shows that the effective sample size is affected in a complex manner by the interaction between selection bias and measurement error.
-Here, we discuss the impact of the interaction on the effective sample size when testing capacity is tied to measurement-error.
+Here, we discuss the impact of the interaction on the effective sample size when testing capacity is tied to measurement error.
 In the current COVID-19 pandemic, there have been well-justified calls for increases in testing capacity.
@@
 Then the ratio estimator is given by $r = \bar y_t / \bar y_{t-1}$.
-We assume both numerator and denominator are the prevalence estimates adjusted for measurement-error.
+We assume both numerator and denominator are the prevalence estimates adjusted for measurement error.
 If the sample size at each time were equal (i.e., $n_1 = n_2 = n$) then this analysis would be equivalent to comparing the increase in observed case-counts (i.e., $\bar y_t/\bar y_{t-1} = y_t/y_{t-1}$).
@@
 $$
-where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement-error adjustment, and $CV(Y_j) = \sigma_{Y_j}/\bar Y_j$ is the coefficient of variation at time step $j$.
+where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement error adjustment, and $CV(Y_j) = \sigma_{Y_j}/\bar Y_j$ is the coefficient of variation at time step $j$.
 The magnitude of the error depends on the population-level ratio $\bar Y_{t} / \bar Y_{t-1}$ so a large decrease in the prevalence rate will have a relatively small error as compared to a large increase in the prevalence rate.
@@
 The analyst's intuition only holds when the difference is equal to zero.
-This occurs at times when data quality, sampling fraction, measurement-error, and prevalence are all constant across time.
+This occurs at times when data quality, sampling fraction, measurement error, and prevalence are all constant across time.
 
@@
 Figure~\ref{fig:ny-covid-test} shows the number of tests has remained relatively constant in the last few weeks, specifically relative to the total NY population, assuming $f_t$ also varies slowly in time is reasonable.
-Finally, the measurement-error adjustment is likely to be constant as no new testing procedures have been reported.
+Finally, the measurement error adjustment is likely to be constant as no new testing procedures have been reported.
 This would not be true in states like Michigan, where testing criteria have been expanding every few weeks.
@@
 $$
-where $S_t$ is the number of susceptible individuals at time $t$, $I_t$ is the number of infected individuals at time $t$, $N$ is the total poulation size, and $R_t$ is the number of removed (recovered or deceased) individuals at time $t$.
+where $S_t$ is the number of susceptible individuals at time $t$, $I_t$ is the number of infected individuals at time $t$, $N$ is the total population size, and $R_t$ is the number of removed (recovered or deceased) individuals at time $t$.
 Figure~\ref{fig:fracpop} presents the fraction of population newly infected individuals at each time step for $\beta = 1.4$ and $\gamma = 0.2$ (black line).
@@
 Of course, we do not observe $K_t$ and $K_{t-1}$ but noisy proxies $y_t$ and $y_{t-1}$, i.e., the observed case-count on day $t$ and $t-1$ respectively.
-The hope is that $y_t/y_{t-1}$ is a good proxy for $K_t/K_{t-1}$ where here $y_t$ isc conceptualized as the number of new cases on day $t$.
+The hope is that $y_t/y_{t-1}$ is a good proxy for $K_t/K_{t-1}$ where here $y_t$ is conceptualized as the number of new cases on day $t$.
 However, if we think about estimating number $K_t$ under SRS of new cases among those susceptible on day $t$, the natural estimator is $S_t \bar y_t$.
@@
 \end{subfigure}
-\caption{Potential bias due to self-selection and measurement-error from a deterministic SIR model with $\beta = 1.4$ (black) and $\gamma = 0.2$ (red).
+\caption{Potential bias due to self-selection and measurement error from a deterministic SIR model with $\beta = 1.4$ (black) and $\gamma = 0.2$ (red).
 Here we assume $f = 0.02$, $FP = 0.01$, $FN = 0.15$, and use a range of relative sampling fraction $M = f_1/f_0$.}
@@
 
-Here we show how selection bias and measurement-error can creep into clinical trial analysis.
+Here we show how selection bias and measurement error can creep into clinical trial analysis.
 The key concern is whether clinical trials on COVID-19 recruit from the pool of individuals who have tested positive for COVID-19 or whether they sample randomly from the population, test, and then recruit from this subset of tested individuals.
@@
 First, the gap is correctly estimated; then the gap is over-estimated as country A sees a rapid rise in cases; then the gap is even more over-estimated as country A improves and the rate is under-estimated while country B sees a rapid increase in their case-count and their rate is over-estimated; then country A's rate is correctly estimated while country B's rate is under-estimated as it sees improvement in infection rates; finally, the gap disappears.
-We are not claiming this will always be the case, and indeed for small enough selection bias and/or large gaps in case-counts such complex behavior will not occur; however, this is simply to point out how observed information can tell a more complex story than the truth (i.e., country A recovers 1 month prior to country B).
+We are not claiming this will always be the case, and indeed for small enough selection bias and/or large gaps in case-counts such complex behaviour will not occur; however, this is simply to point out how observed information can tell a more complex story than the truth (i.e., country A recovers 1 month prior to country B).
 
@@
 
-In prior sections, we have discussed the impact of selection bias and measurement-error on a range of outcomes of interest in COVID-19 research.
+In prior sections, we have discussed the impact of selection bias and measurement error on a range of outcomes of interest in COVID-19 research.
 Here, we try and address the empirical question of how to estimate the data quality $E_{\I} \left[ \rho_{I,Y} \right]$ in the current context.
-In Meng (2019), estimation of the data quality relied on observing the true outcome.
+In \cite{Meng2018}, estimation of the data quality relied on observing the true outcome.
 In election data, for example, we observe the vote totals after the fact and can use this as ground truth.
@@
 The study did not report sensitivity and specificity; therefore, we take reported measures from the Santa Clara study~\citep{Bendavid2020} \footnote{We recognize the debate regarding its conclusions.
-Here we are only using this study for its information regarding measurement-error rates.}.
+Here we are only using this study for its information regarding measurement error rates.}.
 The reported specificity is $82.8\%$ (95\% CI: 76.0-88.4\%) and sensitivity is $99.5\%$ (95\% CI: 99.2-99.7\%).
@@
 The smoothed prevalence estimate on the 20th is 28.3\%.
-Adjusting for measurement-error yields an estimate of $32.5\%$.
+Adjusting for measurement error yields an estimate of $32.5\%$.
 
@@
 We can also perform sensitivity as a function of the false negative and positive rates.
-Under these measurement-error ranges, we saw that the impact on the effective reproductive rate estimator was much smaller.
+Under these measurement error ranges, we saw that the impact on the effective reproductive rate estimator was much smaller.
 However, for $FP = 0.05$ and $FN = 0.005$ we have $M = 4.31$.
-While nothing above is definitive, it points to the interplay of selection bias and measurement-error.
+While nothing above is definitive, it points to the interplay of selection bias and measurement error.
 This calculation gives us simple tools to understand how sensitivity and specificity impact our estimators and therefore how much we should trust conclusions based on these assumptions.
@@
 
-One alternative approach is model-based extensions of SIR models to try and capture both selection bias and measurement-error issues; however, without strong assumptions on the selection mechanism, the estimates are often not nonparametrically identifiable.
+One alternative approach is model-based extensions of SIR models to try and capture both selection bias and measurement error issues; however, without strong assumptions on the selection mechanism, the estimates are often not nonparametrically identifiable.
 Indeed, when an issue ``cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically'' ~\cite[pp.
@@
 where the second equality is obtained by plugging in the statistical decomposition of the error for both time points and the coefficient of variation being defined as $CV(Y) := \sigma_Y/\mu_Y$.
-Under measurement-error, the extra terms $D_{2}$ and $D_1$ can be inserted in the correct locations.
+Under measurement error, the extra terms $D_{2}$ and $D_1$ can be inserted in the correct locations.
 
@@
 
-We also present the same plot under $FP = 0.005$ and $FN = 0.172$ to show the impact of measurement-error on effective sample size.
+We also present the same plot under $FP = 0.005$ and $FN = 0.172$ to show the impact of measurement error on effective sample size.
 

### 2020-05-17 write-up/covid-writeup.tex
figure updates

@@
 \usepackage{amsmath, amsthm, amsfonts}
-% \usepackage[margin=1.5in]{geometry}
+\usepackage[margin=1.5in]{geometry}
 
@@
 Many statisticians, epidemiologists, economists and data scientists have registered serious reservations regarding the reported coronavirus case-counts.
-Limited testing capacity across many states has been widely identified as a key driver of suppressed coronavirus case-counts.
-Calls for increased testing capacity are well-justified and increasingly frequent.
+Limited testing capacity across the country has been widely identified as a key driver of suppressed coronavirus case-counts.
+The calls to increase testing capacity are well-justified as they become a more frequent point of discussion in the public sphere.
 While expanded testing is a laudable goal, selection bias will impact estimates of disease prevalence and the effective reproduction number until the entire population is sampled.
-Moreover, tests are imperfect.
-False positive/negative rates interact in complex ways with selection bias.
+Moreover, tests are imperfect as false positive/negative rates interact in complex ways with selection bias.
 In this paper, we attempt to clarify this interaction.
@@
 As of April 27th, 2020, a total of 2,993,000 cases have been confirmed worldwide.
-As of that afternoon, the New York Times reports at least 965,214 people across the United States have tested positive for the virus, and at least 49,465 patients with the virus have died.
+The New York Times reports at least 965,214 people across the United States have tested positive for the virus, and at least 49,465 patients with the virus have died.
 Aggressive policies have been put in place across the US with at least 50\% of the US population officially urged as of late April to stay home via state-wide executive actions.
@@
 Under these parameters, the effective sample size is $14$.
-Recent proposals~\citep{Siddarth2020} have argued in favor of increasing testing capacity which may reduce the relative sampling rate.
+Recent proposals~\citep{Siddarth2020} have argued in favor of increasing overall testing capacity.
+With more tests available, the relative sampling rate may reduce.
 Supposing it drops to $1.2$ then the effective sample size will increase to $299$.
@@
 where $Z = 1-2Y$; see Appendix~\ref{app:imperfect} for the derivation.
-The first term represents the perfect testing regime, the second term represents the interaction between imperfect testing and selection bias, while the third time represents the bias due to imperfect testing.
-From here on, we refer to $\rho_{I,Y}$ as the \emph{true data quality}, and $\rho_{I,PZ}$ as the \emph{observed data quality} that accounts for both selection bias and measurement error.
+The first term in the large brackets represents the perfect testing regime; to this end, we refer to $\rho_{I,Y}$ as the \emph{true data quality}.
+The second term represents the interaction between imperfect testing and selection bias.
+The variable $PZ = 1[Y=0,P=1] - 1[Y=1, P=1]$ is non-zero only when $P=1$ (i.e., the outcome is incorrectly reported); to this end, we refer to $\rho_{I,PZ}$ as the \emph{observed data quality} that accounts for both selection bias and measurement error.
 In the appendix, we show the sign of $\rho_{I,PZ}$ is the opposite of the sign of $\rho_{I,Y}$, which implies that the observed data quality adds error in the opposite direction from the true data quality.
+Finally, the third term represents the bias due to imperfect testing.
 
@@
 To do this, we define the sampling rates differential.
-Let $f_1 := \pr (I_J = 1 \mid Y_J = 1)$ and $f_0 := \pr(I_J = 1 \mid Y_J = 1)$ be the sampling rates.
+Let $f_1 := \pr (I_J = 1 \mid Y_J = 1)$ and $f_0 := \pr(I_J = 1 \mid Y_J = 0)$ be the sampling rates.
 Then $\Delta = f_1 - f_0$ is the sampling rate differential.
@@
 $$
-n_{eff} = \frac{f}{1-f} \times \frac{1}{E_{I} \left[ \rho_{I,Y}^2 D_M^2 \right]}.
+n_{eff} \le \frac{f}{1-f} \times \frac{1}{E_{I} \left[ \rho_{I,Y}^2 D_M^2 \right]}.
 $$
@@
 \includegraphics[width = 0.4\textwidth]{../methods/figs/mem_heatmap_article.png}
-\caption{measurement error data quality adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.005$ and $FN=0.172$.
+\caption{Measurement error data quality adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.005$ and $FN=0.172$.
 Color scaled so blue = $-10$, white = $0$, and red = $10$.}
@@
 
-\subsection{Clinical trials}
+\subsection{Effect modifiers in COVID-19 clinical trials}
 

### 2020-05-17 methods/sir_example.R

@@ -36,12 +36,13 @@ out2$time <- NULL
 ## Show data
 png(filename = "C:/Users/Balthazar/Documents/GitHub/covid-umich/methods/figs/sir.png",
     width = 6.5, height = 4, units = "in", res = 1200, pointsize = 12)
-par(mar = c(1.5,2,1,1)+0.1)
+par(mar = c(2,2,1,1)+0.1)
 plot(times, out1$I, type = "l", axes = FALSE, bty = "n")
 lines(times, out2$I, col = "red")
 axis(side = 1, labels = FALSE)
 axis(side = 2, labels = FALSE)
 mtext("Fraction Infected", side = 2, line = 1, cex = 0.75)
+mtext("Time", side = 1, line = 1, cex = 0.75)
 dev.off()
 
 ## Setup for bias plots
@@ -99,7 +100,7 @@ ymax = max(bias1_new$Rt+bias1_new$logbias, bias1_new$Rt, bias2_new$Rt+bias2_new$
 ymin = min(bias1_new$Rt+bias1_new$logbias, bias1_new$Rt, bias2_new$Rt+bias2_new$logbias, bias2_new$Rt, na.rm = TRUE)
 png(filename = "C:/Users/Balthazar/Documents/GitHub/covid-umich/methods/figs/sir_rt_comparison.png",
     width = 6.5, height = 4, units = "in", res = 1200, pointsize = 12)
-par(mar = c(1.5,3,0,1)+0.1)
+par(mar = c(2,3,0,1)+0.1)
 plot(bias1_new$Rt, type = "l", ylim = c(0.7, 1.3), axes = FALSE, bty = "n", xlab = "", ylab = "")
 lines(bias1_new$Rt + bias1_new$logbias, lty = 2)
 abline(h = 1.0, col = "grey", lty = 2)
@@ -108,6 +109,7 @@ lines(bias2_new$Rt+bias2_new$logbias, lty = 2, col = "red")
 axis(side =1, labels = FALSE)
 axis(side =2, cex.axis = 0.75, at = seq(0.8,1.2,0.1))
 mtext(expression(R[t]), side =2, line = 2, cex = 0.75)
+mtext("Time", side = 1, line = 1, cex = 0.75)
 dev.off()
 
 ## Same curve, just different M value
@@ -121,14 +123,15 @@ ymax = max(bias3$r+bias3$bias)
 ymin = min(bias3$r+bias3$bias)
[…]

### 2020-05-17 write-up/covid-writeup.tex
Draft 2 done

@@
 Fifth, we discuss a sensitivity analysis method for estimating data quality and its use in understanding the potential biases in COVID-19 case-count analyses.
-We then end the paper with a discussion of the benefits of randomized testing, use of auxiliary information to improve over simple random sampling, and the distinction between data quality and decision making in the current pandemic.
+We then end the paper with a discussion of the benefits of randomized testing and use of auxiliary information to improve over simple random sampling.
 
@@
 While important, many conflate increased testing capacity with increased quality of testing.
-We will discuss this point further in Section \ref{section:testingcomments}.
+We will discuss this point further in Section \ref{section:discussion}.
 As of May 8th, the US has performed $8,412,095$ tests in total.
@@
 The remainder of this paper aims to build upon these fundamental insights by extending the decomposition in two directions: accounting for  measurement error and the temporal nature of the pandemic.
-We then discuss effective sampling methods and data versus decision-making.
+We then discuss effective sampling methods and their importance in the current pandemic.
 
@@
 
-\section{Decision-making: data versus information.}
+% \section{Decision-making: data versus information.}
 
-Having read through the above technical discussions, one may come to the incorrect conclusion that because we are concerned about implicit biases in observational data, we must wait for better data sources in order to act~\citep{Ioannidis2020}.
+% Having read through the above technical discussions, one may come to the incorrect conclusion that because we are concerned about implicit biases in observational data, we must wait for better data sources in order to act~\citep{Ioannidis2020}.
 This conclusion confuses data and information.
@@
 
-\section{That which does not break us, makes us stronger (but potentially not smarter)}
-\label{section:testingcomments}
-
-A slightly altered version of Nietzsche's famous quote has become a mantra for post-pandemic thinking: \emph{that which which does not break us, makes us stronger}.
-While a potentially true statement with respect to viral resistance, it is not clear that governments have yet to learn lessons on pandemic response.
-Governments are pushing increased testing capacity and contact tracing.
-Contact tracing is incredibly useful for identifying carriers early and preventing the disease from spreading.
-We would argue that, after the current wave, understanding prevalence is also key.
-We know that the probability of an outbreak is a function of current prevalence and network connectivity.
-Epidemic critical thresholds have been derived for many infectious disease models~\cite{Pastor2001,Newman2002,Parshani2010}.
-Knowing prevalence will help governments determine long-term community-level risk and allow for more targeted interventions -- shutting down only certain locations when necessary.
-Once we ``return to normal''\footnote{or at least the new normal}, it is clear that testing strategies should include a focus on early detection to help head off future outbreaks.
-
-Governments implicitly argue that increased testing capacity will perform this task.
-Without complete compliance, however, our understanding of future outbreaks may be plagued by self-selection bias, compounded by changing sensitivity and specificity rates.
-Random testing removes these effect modifiers, giving governments more information to fight the disease.
-This is not to say random testing should replace current testing practices; only that the random testing should supplement and will provide valuable additional information even with moderate sample sizes.
-
 \section{Discussion}
+\label{section:discussion}
 
@@
 Selection bias not only means uncertainty, it also implies analysts feel certain about incorrect conclusions.
+We end this paper with a brief discussion of several important topics related to selection bias and measurement error.
 
-Not only that, in current crisis data quality is often tied to data quantity due to rationing system.
+\subsection*{Decision-making: data versus information}
+
+Having read through the above technical discussions, one may come to the incorrect conclusion that because we are concerned about implicit biases in observational data, we must wait for better data sources in order to act~\citep{Ioannidis2020}.
+This conclusion confuses data and information.
+Lack of high quality data means we should be skeptical regarding conclusions drawn from these data sources alone.
+Being skeptical of observed data does not imply governments and communities should wait for high quality data to act; this is especially true in the current high risk scenario.
+Instead, we should aim to have high quality information.
+Yet in the current crisis we have just that.
+Indeed, epidemiologists and public health officials have a strong understanding of the basic facts of pandemics and disease spread.
+Social distancing, contact tracing and mass testing are all important.
+Individuals and communities with low quality data and only basic information will hopefully make decisions that respect inherent risks to their well being as well as the high degree of uncertainty in the observed data.
+Of course, there are trade-offs and long-term stay-at-home orders will have negative economic consequences.
+A full discussion of these topics is beyond the scope of the current article.
+
+\subsection*{High quality information on prevalence.}
+
+Governments are pushing for increased testing capacity and robust contact tracing.
+Contact tracing is incredibly useful for identifying carriers early and preventing spread of the disease.
+We would argue that, after the current wave, understanding prevalence is also key.
+We know that the probability of an outbreak is a function of current prevalence and network connectivity.
+Epidemic critical thresholds have been derived for many infectious disease models~\cite{Pastor2001,Newman2002,Parshani2010}.
+Knowing prevalence will help governments determine long-term community-level risk and allow for more targeted interventions -- shutting down only certain locations when necessary.
+
+\subsection*{Data quantity versus quality}
+
+Governments implicitly argue that increased testing capacity will perform the above task.
+Without complete compliance, however, our understanding of future outbreaks may be plagued by self-selection bias, compounded by changing sensitivity and specificity rates.
+Random testing removes these effect modifiers, giving governments more information to fight the disease.
+This is not to say random testing should replace current testing practices; only that the random testing should supplement and will provide valuable additional information even with moderate sample sizes.
+
+% Not only that, in current crisis data quality is often tied to data quantity due to rationing system.
 Increasing testing under test rationing lowers correlation.
@@
 
-One alternative approach is model-based extensions of SIR models to try and capture both selection bias and measurement error issues; however, without strong assumptions on the selection mechanism, the estimates are often not nonparametrically identifiable.
-Indeed, when an issue ``cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically'' ~\cite[pp.
+\subsection*{Model-based solutions.}
+
+One alternative approach to what is presented in this article is a model-based extension of SIR models to try and capture both selection bias and measurement error; however, without strong assumptions on the selection mechanism, the estimates are often not nonparametrically identifiable.
+When an issue ``cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically'' ~\cite[pp.
 96]{CoxHink74}.

### 2020-05-21 write-up/scifile.tex
science-paper

--- 2020-05-17 write-up/covid-writeup.tex
@@
-\documentclass[aoas]{amsart}
+% Use only LaTeX2e, calling the article.cls class and 12-point type.
 
-\RequirePackage[OT1]{fontenc}
+\documentclass[12pt]{article}
 
-\usepackage[foot]{amsaddr}
-% \usepackage{biblatex}
-\usepackage{natbib}
-%\bibliographystyle{apalike}
+% Users of the {thebibliography} environment or BibTeX should use the
+% scicite.sty package, downloadable from *Science* at
+% http://www.sciencemag.org/authors/preparing-manuscripts-using-latex
+% This package should properly format in-text
+% reference calls and reference-list numbers.
 
-\usepackage{tchdr}
-\boldshortcuts
+\usepackage{scicite}
 
-\usepackage{graphicx}
-\usepackage{wrapfig}
-\usepackage{setspace}
-\DeclareGraphicsExtensions{.eps, .ps}
-\usepackage{hyperref}
-\usepackage{amsmath, amsthm, amsfonts}
-% \usepackage[margin=1.5in]{geometry}
+\usepackage{times}
+\usepackage{amsmath,amsthm,amsfonts}
 
-% \numberwithin{equation}{section}
-% \theoremstyle{plain}
-% \newtheorem{theorem}{Theorem}[section]
-% \newtheorem{example}{Example}
-% \newtheorem{remark}{Remark}
-% \newtheorem{corollary}[theorem]{Corollary}
-% \newtheorem{criterion}[theorem]{Criterion}
-% \newtheorem{definition}[theorem]{Definition}
-% \newtheorem{exercise}[theorem]{Exercise}
-% \newtheorem{lemma}[theorem]{Lemma}
-% \newtheorem{proposition}[theorem]{Proposition}
+% The preamble here sets up a lot of new/revised commands and
+% environments.  It's annoying, but please do *not* try to strip these
+% out into a separate .sty file (which could lead to the loss of some
+% information when we convert the file to other formats).  Instead, keep
+% them in the preamble of your main LaTeX source file.
 
+% The following parameters seem to provide a reasonable page setup.
+
+\topmargin 0.0cm
+\oddsidemargin 0.2cm
+\textwidth 16cm
+\textheight 21cm
+\footskip 1.0cm
+
+%The next command sets up an environment for the abstract to your paper.
+
+\newenvironment{sciabstract}{%
+\begin{quote} \bf}
+{\end{quote}}
+
+\def\I{{\bf I}}
+\def\bary{{\bar y}}
 \def\pr{\text{pr}}
 \def\sgn{\text{sgn}}
-\def\I{\bf I}
+
+% Include your paper's title here
+
+\title{The Hypothesis of Testing: Paradoxes arising out of reported coronavirus case-counts}
+
+% Place the author information here.  Please hand-code the contact
+% information and notecalls; do *not* use \footnote commands.  Let the
+% author contact information appear immediately below the author names
+% as shown.  We would also prefer that you don't change the type-size
+% settings shown here.
+
+\author
+{Walter Dempsey$^{1\ast}$\\
+\\
+\normalsize{$^{1}$Department of Biostatistics, University of Michigan,}\\
+\normalsize{1415 Washington Heights, Ann Arbor, MI 48109, USA}\\
+\\
+\normalsize{$^\ast$To whom correspondence should be addressed; E-mail:  wdem@umich.edu.}
+}
+
+% Include the date command, but leave its argument blank.
+
+\date{}
+
+%%%%%%%%%%%%%%%%% END OF PREAMBLE %%%%%%%%%%%%%%%%
 
 \begin{document}
-\title[The Hypothesis of Testing]{The Hypothesis of Testing: Paradoxes arising out of reported coronavirus case-counts} %
 
-\author{Walter Dempsey}
-\address{Department of Biostatistics, University of Michigan, Ann Arbor, MI 48109}
-% \email{wdem@umich.edu}
+% Double-space the manuscript.
 
-\begin{abstract}
+\baselineskip24pt
+
+% Make the title.
+
+\maketitle
+
+% Place your abstract within the special {sciabstract} environment.
+
+\begin{sciabstract}
   Many statisticians, epidemiologists, economists and data scientists have registered serious reservations regarding the reported coronavirus case-counts.
@@
 The discussion guides several suggestions on how to improve current case-count reporting.
-\end{abstract}
+\end{sciabstract}
 
-\maketitle
+% In setting up this template for *Science* papers, we've used both
+% the \section* command and the \paragraph* command for topical
+% divisions.  Which you use will of course depend on the type of paper
+% you're writing.  Review Articles tend to have displayed headings, for
+% which \section* is more appropriate; Research Articles, when they have
+% formal topical divisions at all, tend to signal them with bold text
+% that runs into the paragraph, for which \paragraph* is the right
+% choice.  Either way, use the asterisk (*) modifier, as shown, to
+% suppress numbering.
 
-\section{Introduction}
+\section*{Introduction}
+
 The World Health Organization has declared the coronavirus disease 2019 (COVID-19) a public health emergency.
-As of April 27th, 2020, a total of 2,993,000 cases have been confirmed worldwide.
-The New York Times reports at least 965,214 people across the United States have tested positive for the virus, and at least 49,465 patients with the virus have died.
-Aggressive policies have been put in place across the US with at least 50\% of the US population officially urged as of late April to stay home via state-wide executive actions.
+As of May 21st, 2020, over 5 million cases have been confirmed worldwide with 1.5 million confirmed cases and over 90 thousand COVID-19 deaths across the United States.
+This pandemic
+%has ripped through our communities and
+has become the focal point of every day life; yet the data landscape for understanding COVID-19 remains limited.
+Public databases maintained by Johns Hopkins University~\cite{JHU_Lancet} and the New York Times~\cite{NYT} provide incoming county-level information of confirmed cases and deaths.
+Statisticians, epidemiologists, economists, and data scientists have attempted to use this granular data to forecast COVID-19 case-counts, deaths, and hospitalizations~\cite{Song2020,Ray2020,2020.IHME}.
 
-Despite these steps, the data landscape for understanding COVID-19 remains limited.
-Public databases maintained by \href{https://bit.ly/2UqFSuA}{Johns Hopkins University}~\citep{JHU_Lancet} and the \href{https://bit.ly/2vUHfrK}{New York Times}~\cite{NYT} provide incoming county-level information of confirmed cases and deaths.
-Statisticians, epidemiologists, economists, and data scientists have used this granular data to forecast COVID-19 case-counts, deaths, and hospitalizations~\citep{Song2020,Ray2020,2020.IHME}.
-In many cases, variations on the SIR models have been used to draw inferences about infection rates and intensity.
-Parametric inference often uses observed case-count and/or observed COVID-19 related deaths to infer latent trajectories of the pandemic.
-
-The goal of this paper is to express reservations at the use of case-counts as a proxy for disease prevalence and intensity as well as its use as direct input into estimation of standard epidemiological models for inference and forecasting.
+The goal of this paper is to express reservations at the use of case-counts as a proxy for disease prevalence and in estimation of standard epidemiological models for inference and forecasting.
 The reason is straightforward: current models do not account simultaneously for selection bias and measurement error.
-Selection bias enters due to differences in testing strategies across countries and states.
-In the US, for example, limited testing capacity has caused local and state health departments to focus on testing only high-risk populations.
-Moreover, testing requires the individual to self-select into asking for a test and then receiving approval by their doctor.
-Demands for increased testing capacity, while laudable, ignore the issues of self-selection and measurement error.
-While increasing testing capacity increases data quantity, there is no guarantee for increased data quality.
-We will show that aggressive pushes for ramped up testing capacity that are tied to decreases in data quality may have a mitigated impact  in our ability to estimate quantities of interest such as prevalence and effective reproduction number.
-
-In this paper, we will demonstrate the complex interaction between these two fundamental concepts.
-In classical statistics, measurement error is often associated with parameter attenuation; we will show, however, that the combination with self-selection can cause bias in parameter estimates to change sign and increase/decrease in magnitude.
-Moreover, the bias may depend on the sampling fraction, prevalence, and population size.
-Without a complete understanding of these interactions, we are destined to misinterpret case-counts and come to erroneous conclusions.
+In the US, for example, limited testing capacity has caused health departments to test only high-risk individuals.
+Moreover, testing requires individuals to request a test and receive approval from their doctor.
+Demands for increased testing capacity, while laudable, do not guarantee increased data quality.
+Aggressive measures to increase testing capacity that decrease data quality may have a mitigated impact in our ability to estimate quantities of interest such as prevalence and effective reproduction number.
 We will demonstrate why there is a strong need for more precision and care in presenting COVID-19 analyses to the broader research and non-academic communities.
 
-% \subsection{Related work}
-
-% There has been an abundance..
-
-% However, most complain about testing \emph{capacity} (cite Nate Silver), there is clearly issues of data quality.
-
-% measurement error discussions.
-% Selection bias . Proponents of this critique suggest well-designed studies.
-
-% Stanford study was Ionides.  They however, forgot about Critique \#2: measurement
-
-\subsection{Outline}
-
-This article discusses the relationship between three statistical concepts: selection bias, measurement error, and the too often forgotten population size.
-We clarify mathematically why the situation is much more complex than it first appears.
-Through simple mathematical arguments, we clarify five important issues related to COVID-19 case-count analysis.
-First, we show that unadjusted prevalence rates are (unsurprisingly) biased when tests are imperfect.
-What is surprising is that the direction and magnitude of the bias can vary substantially and interacts with the sampling fraction.
-Next we show that data quality for adjusted prevalence rates includes an additional term in the error decomposition that captures the interplay among measurement error, selection bias, and prevalence.
-Third, we show that daily trends in rate of positive tests is equally problematic even if testing rates and selection protocols are held constant.
-We show that the rate of change in COVID-19 observed case-counts overestimates the true rate of change prior to the peak time and underestimates it immediately after.
-This implies a data scientist analyzing observed rates will be (on average) overly pessimistic in the early stages of the pandemic, and overly optimistic subsequent to the peak times.
-We show that estimates of the effective reproduction number may be similarly impacted.
-Fourth, we show that cross-country comparisons are difficult at best with population size, sampling fraction, and data quality all interacting to impact null hypothesis testing.
-Fifth, we discuss a sensitivity analysis method for estimating data quality and its use in understanding the potential biases in COVID-19 case-count analyses.
-We then end the paper with a discussion of the benefits of randomized testing and use of auxiliary information to improve over simple random sampling.
-
-\section{Analysis of case-count data}
+\section*{Analysis of imperfect case-count data}
 \label{section:casecount}
 
-We start with some simple notation.
 Let $N$ denote the population size.
-For state-level analysis $N$ is the state's total population, while for country-level analysis $N$ is the country's total population.
-At a fixed moment in time, let $Y_j$ denote COVID-19 status for the $j$th individual in the population, $j=1,\ldots, N$.
-Here, like in survey methodology~\citep{Cochran77}, we treat COVID-19 status as a fixed but unknown quantity of interest.
-For simplicity, we start by ignoring the dynamic nature of the viral outbreak as well as the fact that individuals can recover from the disease and assume either individual $j$ is COVID-19 positive and $Y_j=1$ or is COVID-19 negative and $Y_j=0$.
+At a fixed time, let $Y_j$ denote COVID-19 status for the $j$th individual in the population, $j=1,\ldots, N$.
+Here, as in survey methodology~\cite{Cochran77}, we treat COVID-19 status as a fixed but unknown quantity of interest.
+For simplicity, we start by ignoring the dynamic nature of the outbreak and the fact that individuals can recover.
 We also let $I_j \in \{0,1\}$ be an indicator that the individual was selected for testing ($I = 1$) or not ($I=0$).
 
-To start, we assume the primary questions of interest are estimates of the overall number of COVID-19 cases and/or disease prevalence.
-That is, we are interested in either the population total $Y = \sum_{j=1}^N Y_j$ or the population average $\bar Y = Y/N$.
+Here we consider estimates of the number of COVID-19 cases and/or disease prevalence, i.e., the population total $Y = \sum_{j=1}^N Y_j$ or the population average $\bar Y = Y/N$.
 Suppose that $n$ tests are performed and we observe the values $y_1, \ldots, y_n \in \{0,1\}$.
-Then a natural candidate for prevalence is the proportion of positive tests $\bar y = \frac{1}{n} \sum_{i=1}^n y_i$, and a natural candidate for overall cases is $N \times \bar y$.
-Under simple random sampling (SRS) or any other epsem\footnote{equal probability of selection method} design, the above are unbiased estimators of the population-level quantities of interest.
-Under SRS, the variance of the estimator can be expressed as $\frac{1}{N-1} \times \frac{1-f}{f} \times \sigma_Y^2$ where $f = n/N$ is the sampling fraction and $\sigma_Y^2 = \frac{1}{N} \sum_{i=1}^N (Y_i - \bar Y)^2 = \bar Y (1- \bar Y)$.
+Then a natural candidate for prevalence is the proportion of positive tests $\bar y = n^{-1} \sum_{i=1}^n y_i$, and a natural candidate for overall cases is $N \times \bar y$.
+Under simple random sampling (SRS), both are unbiased estimators of the  quantities of interest and the variance of the prevalence estimator equals $(N-1)^{-1} \times (1-f)/f \times \sigma_Y^2$ where $f = n/N$ is the sampling fraction and $\sigma_Y^2 = \frac{1}{N} \sum_{i=1}^N (Y_i - \bar Y)^2 = \bar Y (1- \bar Y)$.
+
+In the current context, the true status is not observed.
+Instead, individuals are tested for COVID-19 status.
+Tests are imperfect.
+COVID-19 testing is no exception.
+The standard assumption is that measurement error leads to parameter attenuation.
+When paired with selection bias, however, the two sources of error become entangled, and resulting errors can become magnified, muted, or even switch signs~\cite{Beesley2020,Beesley2019,Smeden2019}.
+Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the outcome and $0$ when we observe the true outcome.
+We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.
+If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^{\star} = Y_j(1-P_j) + (1-Y_j) P_j$.
+Suppose disease prevalence is estimated by $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ which is unbiased under simple random sampling.
+We wish to express the error $\tilde  y_n - \bar Y$ in statistical terms.
+In the supplementary materials, we show that the error now can be expressed as
+\begin{equation}
+\label{eq:error}
+\rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
+\times \underbrace{\left[ 1 + FP + FN - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right]}_{D_M}.
+\end{equation}
 
@@
 
-\begin{wrapfigure}{r}{0.5\textwidth}
-\centering
-\includegraphics[width = 0.4\textwidth]{../methods/figs/US-casecount.png}
-\caption{US total tests and total positive tests per day}
-\label{fig:ustests}
-\vspace{-0.3cm}
-\end{wrapfigure}
-
 Second, there have been calls for increased testing.
@@
 As of May 8th, the US has performed $8,412,095$ tests in total.
-Figure~\ref{fig:ustests} shows the trajectory of testing per day and positive cases per day in the US.
+Figure~1 shows the trajectory of testing per day and positive cases per day in the US.
 The US population is roughly $328$ million, meaning the fraction sampled is $f = 0.026$.
@@
 Under these parameters, the effective sample size is $14$.
-Recent proposals~\citep{Siddarth2020} have argued in favor of increasing overall testing capacity.
+Recent proposals~\cite{Siddarth2020} have argued in favor of increasing overall testing capacity.
 With more tests available, the relative sampling rate may reduce.
@@
 
-\subsection{Imperfect testing}
+\subsection*{Imperfect testing}
 
-Tests are imperfect.
-COVID-19 testing is no exception.
-Here we investigate the interplay between imperfect testing and selection bias.
-In discussions of inaccurate testing, the standard assumption is measurement error leads to parameter attenuation.
-When paired with selection bias, however, the two sources of error become entangled, and resulting errors can become magnified, muted, or even switch signs.
-
-First we require some additional notation.
 Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the outcome and $0$ when we observe the true outcome.
@@
 
-While prior investigations have noted the interaction between measurement error and selection bias~\citep{Beesley2020,Beesley2019,Smeden2019}, the interaction with the sample size relative to the population, i.e., $f$, has largely been ignored.
+While prior investigations have noted the interaction between measurement error and selection bias~\cite{Beesley2020,Beesley2019,Smeden2019}, the interaction with the sample size relative to the population, i.e., $f$, has largely been ignored.
 The above statistical decomposition clarifies the importance of this quantity~$f$.
@@
 The second term is the interaction between selection bias and measurement error.
-Figure~\ref{fig:heatmap} presents the measurement error data quality adjustment as a function of the relative frequency and odds ratio.
+Figure~2 presents the measurement error data quality adjustment as a function of the relative frequency and odds ratio.
 We see that the adjustment can be both positive and negative as well as a range of magnitudes.
@@
 If the relative sampling fraction drops to $1.2$ then the effective sample size drops from $299$ to $216$.
-
-\begin{wrapfigure}{r}{0.5\textwidth}
-\centering
-\includegraphics[width = 0.4\textwidth]{../methods/figs/mem_heatmap_article.png}
-\caption{Measurement error data quality adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.005$ and $FN=0.172$.
-Color scaled so blue = $-10$, white = $0$, and red = $10$.}
-\label{fig:heatmap}
-\vspace{-0.3cm}
-\end{wrapfigure}
 
@@
 
-\subsection{Regrettable rates: complex biases resulting from self-selection}
+\subsection*{Regrettable rates: complex biases resulting from self-selection}
 \label{section:rates}
@@
 Consider the state of New York.
-Figure~\ref{fig:ny-covid-test} shows the number of tests and positive tests per day in New York state.
-Figure~\ref{fig:ny-covid-frac} presents the fraction of tests that were positive per day in New York state.
+Figure~5 shows the number of tests and positive tests per day in New York state.
+Figure~5b presents the fraction of tests that were positive per day in New York state.
 For cancellation, we would need the expected data qualities to be uncorrelated and equal.
@@
 Note, however, that constant $\Delta_t$ does not imply constant data quality (see Appendix \ref{section:effss} for details).
-Figure~\ref{fig:ny-covid-test} shows the number of tests has remained relatively constant in the last few weeks, specifically relative to the total NY population, assuming $f_t$ also varies slowly in time is reasonable.
+Figure~5a shows the number of tests has remained relatively constant in the last few weeks, specifically relative to the total NY population, assuming $f_t$ also varies slowly in time is reasonable.
 Finally, the measurement error adjustment is likely to be constant as no new testing procedures have been reported.
@@
 
-Figure \ref{fig:ratio-bias} shows a trajectory of the true ratio and the potential biased estimators under an SIR model for the epidemic dynamics, with state evolution given by the following ODE
+Figure~3a shows a trajectory of the true ratio and the potential biased estimators under an SIR model for the epidemic dynamics, with state evolution given by the following ODE
 $$
@@
 where $S_t$ is the number of susceptible individuals at time $t$, $I_t$ is the number of infected individuals at time $t$, $N$ is the total population size, and $R_t$ is the number of removed (recovered or deceased) individuals at time $t$.
-Figure~\ref{fig:fracpop} presents the fraction of population newly infected individuals at each time step for $\beta = 1.4$ and $\gamma = 0.2$ (black line).
+Figure~4a presents the fraction of population newly infected individuals at each time step for $\beta = 1.4$ and $\gamma = 0.2$ (black line).
 In each case, the rate of change in prevalence is likely to be  overestimated prior to the peak and then underestimated afterwards.
@@
 
-\subsection{Estimation of effective reproduction number}
+\subsection*{Estimation of effective reproduction number}
 \label{section:r0-estimation}
-Many well-respected epidemiologists argue that tracking the effective reproduction number $R_t$ is the only way to manage through the crisis~\citep{Gabriel2020}.
+Many well-respected epidemiologists argue that tracking the effective reproduction number $R_t$ is the only way to manage through the crisis~\cite{Gabriel2020}.
 We show here how these estimates are also impacted by selection bias and measurement error.
@@
 \cite{Bettencourt2008} show that under a Poisson likelihood, a simple relation between the trajectory of new cases and the effective reproduction number can be derived.
-In particular, under an SIR model, the number of case counts on day $t$ is Poisson distributed with rate $K_{t-1} \exp \left( \frac{1}{\gamma} (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval which is approximately $7$ days for COVID-19~\citep{Sanche2020}.
+In particular, under an SIR model, the number of case counts on day $t$ is Poisson distributed with rate $K_{t-1} \exp \left( \frac{1}{\gamma} (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval which is approximately $7$ days for COVID-19~\cite{Sanche2020}.
 Using this, a simple estimator of the effective reproduction number at time $t$ can be given by
@@
 Since $S_t \leq S_{t-1}$, the final term $- \log (S_t/S_{t-1}) \geq 0$.
-Figure~\ref{fig:r0-bias} displays the bias for the effective reproductive number and how it is impacted by the relative sampling fraction.
+Figure~4b displays the bias for the effective reproductive number and how it is impacted by the relative sampling fraction.
 Here for simplicity, we assume the fraction of the population that is susceptible remains large and near the total population size.
@@
 Changes to this assumption can impact the potential bias.
-The key difference between Figure~\ref{fig:ratio-bias} and Figure~\ref{fig:r0-bias} is that prevalence depends on the fraction of the population infected at time $t$ while $\hat R_t$ depends on the fraction of new cases in the population.
+The key difference between Figure~4a and Figure~4b is that prevalence depends on the fraction of the population infected at time $t$ while $\hat R_t$ depends on the fraction of new cases in the population.
 This leads to differences in when the bias is most pronounced.
 
-\begin{figure}
-\centering
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{../methods/figs/sir_ratio.png}
-  \caption{Ratio estimator}
-  \label{fig:ratio-bias}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{../methods/figs/sir_rt.png}
-  \caption{Effective reproductive rate estimator}
-  \label{fig:r0-bias}
-\end{subfigure}
-\caption{Potential bias due to self-selection and measurement error from a deterministic SIR model with $\beta = 1.4$ (black) and $\gamma = 0.2$ (red).
-Here we assume $f = 0.02$, $FP = 0.01$, $FN = 0.15$, and use a range of relative sampling fraction $M = f_1/f_0$.}
-\label{fig:rates}
-\end{figure}
-
-\subsection{Effect modifiers in COVID-19 clinical trials}
+\subsection*{Effect modifiers in COVID-19 clinical trials}
 
@@
 Most often, our main interest is in causal effects on the population of COVID-19 positive.
-Randomization within the clinical trial design yields internal validity, but we also need external validity to estimate the correct marginal effect of interest.
+Randomization within the clinical trial design yields internal validity, but we also need external validity to estimate the correct marginal effect of interest~\cite{Keiding2016}.
 
-\section{Cross-population comparisons}
+\section*{Cross-population comparisons}
 
@@
 
-\subsection{Prevalence comparisons}
+\subsection*{Prevalence comparisons}
 
@@
 Then the population adjustment is $5.84$ million.
-As of May 5th, Canada has performed a total of 24.92 tests per 1,000 people while the US has performed a total of 22.01 tests per 1,000 people\footnote{According to \href{https://ourworldindata.org/grapher/full-list-cumulative-total-tests-per-thousand}{ourworldindata.org}} so the data quantities are approximately equal to $f = 23/1000 = 0.023$.
+As of May 5th, Canada has performed a total of 24.92 tests per 1,000 people while the US has performed a total of 22.01 tests per 1,000 people~\cite{worlddata} so the data quantities are approximately equal to $f = 23/1000 = 0.023$.
 Then, under a prevalence of 10\%, one would need the differential $\Delta_1-\Delta_2$ to be much smaller than $11.68 \times 10^{-6}$ in order for the selection bias to not impact.
@@
 
-\subsection{Case count comparisons}
-Two alternative comparisons are common.
-First, the data analyst may directly compare the observed number of new cases on a given day.
-If $Y_1$ and $Y_2$ are the number of cases in locations $j=1,2$ respectively, then we can express the error of $y_1 - y_2$ as
-$$
-\begin{aligned}
-y_1 - y_2 &= n_1 ( \bar y_1 - \bar Y_{1}) - n_2 ( \bar y_2 - \bar Y_{2}) + \left( \frac{n_1}{N_1} Y_{1} - \frac{n_2}{N_2} Y_2  \right) \\
-&= \left[ n_1 \sigma_{Y_1} \rho_{I_1, Y_1} \sqrt{\frac{1-f_1}{f_1}} D_{M_1} - n_2 \sigma_{Y_2} \rho_{I_2, Y_2} \sqrt{\frac{1-f_2}{f_2}} D_{M_2}  \right] + \left[ f_1 Y_1 - f_2 Y_2 \right].
-\end{aligned}
-$$
-Under simple random sampling, the first term is zero in expectation.
-However, the question is whether $f_1 \bar Y_1 - f_2 \bar Y_2$ is of scientific interest.
-We would argue that this is only of true interest when $f_1 = f_2$ and we are interested in comparing total number of cases.
-In this setting, however, the difference in error terms is unlikely to be zero.
-To see this, assume that we have similar expected data quality, data quantity, problem difficulty, and measurement error.
-Then the error is a function of $n_1 - n_2 = n_2 \left( \frac{N_1}{N_2} - 1 \right)$.
-So we see that the error now is a function of the relative difference in the population sizes.
-
-Alternatively, the data analyst may be wary of such concerns in population differences and scale counts by respective population size, i.e., $y_1/N_1 - y_2/ N_2$.  We can again express the error as
-$$
-\begin{aligned}
-\frac{y_1}{N_1} - \frac{y_2}{N_2} &= \frac{n_1}{N_1} ( \bar y_1 - \bar Y_{1}) - \frac{n_2}{N_2} ( \bar y_2 - \bar Y_{2}) + \left( \frac{n_1}{N_1} \bar Y_{1} - \frac{n_2}{N_2} \bar Y_2  \right) \\
-&= \left[ \sigma_{Y_1} \rho_{I_1, Y_1} \sqrt{ f_1 (1-f_1) } D_{M_1} - \sigma_{Y_2} \rho_{I_2, Y_2} \sqrt{f_2 (1-f_2)} D_{M_2} \right] + \left[ f_1 \bar Y_1 - f_2 \bar Y_2 \right].
-\end{aligned}
-$$
-Under simple random sampling, the first term is again zero in expectation.
-However, the question is whether $f_1 \bar Y_1 - f_2 \bar Y_2$ is of scientific interest.
-We would argue that this is only of true interest when $f_1 = f_2$ and we wish to compare disease prevalence.
-Importantly, the error in this comparison is less problematic.
-To see this, assume that we have similar expected data quality, data quantity, problem difficulty, and measurement error.
-Then the error is equal to zero.
-
-It is important to note that the standard law of large populations caveats apply to both comparisons.
-In both cases, the point is that we must ascertain what is the population-level contrast and whether this should be of genuine scientific interest.
-Otherwise, we are comparing apples and oranges.
-Not only this, but the two comparisons come with different levels of error induced by selection bias and measurement error.
-
-\subsection{Rate comparisons}
+\subsection*{Rate comparisons}
 
@@
 Consider two countries (A and B) in which the peak occurs 2 weeks prior for country A then country B but other.
-Figure~\ref{fig:comparison} presents such a comparison where each country's disease trajectory follows an SIR model as described in Section~\ref{section:rates} with different parameters.
-Figure \ref{fig:eff} shows how the errors interact in complex ways.
+Figure~4b presents such a comparison where each country's disease trajectory follows an SIR model as described in Section~\ref{section:rates} with different parameters.
+Figure~4 shows how the errors interact in complex ways.
 First, the gap is correctly estimated; then the gap is over-estimated as country A sees a rapid rise in cases; then the gap is even more over-estimated as country A improves and the rate is under-estimated while country B sees a rapid increase in their case-count and their rate is over-estimated; then country A's rate is correctly estimated while country B's rate is under-estimated as it sees improvement in infection rates; finally, the gap disappears.
@@
 
-\begin{figure}
-\centering
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{../methods/figs/sir.png}
-  \caption{Fraction of new cases in population}
-  \label{fig:fracpop}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{../methods/figs/sir_rt_comparison.png}
-  \caption{Effective reproduction rate estimators}
-  \label{fig:eff}
-\end{subfigure}
-\caption{Left: fraction infected in two SIR models with $\beta = 1.4$ and $0.9$ respectively and $\gamma = 0.2$ with same initial conditions.
-Right: comparison of $\hat R_t$ across time with $FN = 0.2$, $FP = 0.01$, and $M = 4$.}
-\label{fig:comparison}
-\end{figure}
-
-\subsection{Estimation of Data Quality}
+\subsection*{Estimation of Data Quality}
 \label{section:est_dq}
@@
 It found 13.9\% of 3,000 people tested across the state had signs of the virus.
-The study did not report sensitivity and specificity; therefore, we take reported measures from the Santa Clara study~\citep{Bendavid2020} \footnote{We recognize the debate regarding its conclusions.
+The study did not report sensitivity and specificity; therefore, we take reported measures from the Santa Clara study~\cite{Bendavid2020} \footnote{We recognize the debate regarding its conclusions.
 Here we are only using this study for its information regarding measurement error rates.}.
@@
 
-Figure~\ref{fig:ny-covid-test} plots the number of new total and positive tests per day.
-Figure~\ref{fig:ny-covid} plots the fraction of tests that were positiver per day.
+Figure~5 plots the number of new total and positive tests per day.
+Figure~5a plots the fraction of tests that were positiver per day.
 In both cases, we applied an exponential smoothing to the reported values because the spiky pattern is likely due to testing backlogs.
@@
 
-\begin{figure}
-\centering
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{../methods/figs/NY_casecount.png}
-  \caption{NY new cases per day}
-  \label{fig:ny-covid-test}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{../methods/figs/NY_fracpos.png}
-  \caption{NY fraction of new tests positive per day}
-  \label{fig:ny-covid-frac}
-\end{subfigure}
-\caption{New York COVID-19 reported data }
-\label{fig:ny-covid}
-\end{figure}
-
-\section{Stratified sampling: improving precision in low-prevalence environments}
-
-This article argues why random sampling is a powerful tool and the limitations of data obtained through a self-selection process.
-As COVID-19 survey studies become more readily accessible, we encourage researchers to focus on building representative samples.
-One difficulty in the current crisis is the low prevalence relative to the population size in certain geographies.
-Take the recent Santa Clara study, for example, that estimated less than 2\% prevalence in the area.
-The key issue with low prevalence is that if the test's specificity is not lower than the prevalence, then the observed data is consistent with zero recovered individuals.
-
-In such settings, simple random sampling is doomed to fail; however, there are potential solutions.
-In particular, one can stratify the population into risk categories (based on population density, occupation, age, and other important factors) and perform Neyman allocation~\citep{Cochran77}.
-That is, sample individuals at random in strata $h$ with size proportional to $P_h \sqrt{\bar Y_h \times (1-\bar Y_h)}$ where $P_h$ is the strata frequency in the population and $\bar Y_h$ the prevalence in the strata.
-Of course, $P_h$ and $\bar Y_h$ are not known a priori; however, if decent estimates can be obtained, we can see huge benefits in settings where the prevalence varies widely by strata.
-One could go attempt to go further and incorporate contact network information.
-However, such data is difficult to obtain in practice.
-Indeed, geographic location, occupation, age and basic demographic information are much easier to obtain and are a good proxy for the latent network connectivity.
-
-% \section{Decision-making: data versus information.}
-
-% Having read through the above technical discussions, one may come to the incorrect conclusion that because we are concerned about implicit biases in observational data, we must wait for better data sources in order to act~\citep{Ioannidis2020}.
-This conclusion confuses data and information.
-Lack of high quality data means we should be skeptical regarding conclusions drawn from these data sources alone.
-There are whole fields dedicated to careful analysis in such settings (e.g., causal inference from observational studies).
-Being a data skeptic, does not mean one must wait for high quality data to act, especially in high risk scenarios.
-One aims to have high quality information.
-Yet, in the current crisis, we have just that.
-Indeed, epidemiologists and public health officials have a strong understanding of the basic facts of pandemics and disease spread.
-Social distancing, contact tracing and mass testing are all important.
-Individuals and populations with low quality data and only basic information will make decisions that respect inherent risks in unknown unknowns.
-Of course, there are trade-offs and long-term stay-at-home orders will have negative economic consequences.
-A full discussion of these topics is beyond the scope of the current article; we merely bring up the matter because we are well aware that discussions of data quality can sometimes be misinterpreted as claims that we do not have the requisite information to act immediately and decisively.
-This is simply not the case.
-
-\section{Discussion}
+\section*{Discussion}
 \label{section:discussion}
@@
 
+\subsection*{Stratified sampling: improving precision in low-prevalence environments}
+
+This article argues why random sampling is a powerful tool and the limitations of data obtained through a self-selection process.
+As COVID-19 survey studies become more readily accessible, we encourage researchers to focus on building representative samples.
+One difficulty in the current crisis is the low prevalence relative to the population size in certain geographies.
+Take the recent Santa Clara study, for example, that estimated less than 2\% prevalence in the area.
+The key issue with low prevalence is that if the test's specificity is not lower than the prevalence, then the observed data is consistent with zero recovered individuals.
+
+In such settings, simple random sampling is doomed to fail; however, there are potential solutions.
+In particular, one can stratify the population into risk categories (based on population density, occupation, age, and other important factors) and perform Neyman allocation~\cite{Cochran77}.
+That is, sample individuals at random in strata $h$ with size proportional to $P_h \sqrt{\bar Y_h \times (1-\bar Y_h)}$ where $P_h$ is the strata frequency in the population and $\bar Y_h$ the prevalence in the strata.
+Of course, $P_h$ and $\bar Y_h$ are not known a priori; however, if decent estimates can be obtained, we can see huge benefits in settings where the prevalence varies widely by strata.
+One could go attempt to go further and incorporate contact network information.
+However, such data is difficult to obtain in practice.
+Indeed, geographic location, occupation, age and basic demographic information are much easier to obtain and are a good proxy for the latent network connectivity.
+
 \subsection*{Decision-making: data versus information}
 
-Having read through the above technical discussions, one may come to the incorrect conclusion that because we are concerned about implicit biases in observational data, we must wait for better data sources in order to act~\citep{Ioannidis2020}.
+Having read through the above technical discussions, one may come to the incorrect conclusion that because we are concerned about implicit biases in observational data, we must wait for better data sources in order to act~\cite{Ioannidis2020}.
 This conclusion confuses data and information.
@@
 One alternative approach to what is presented in this article is a model-based extension of SIR models to try and capture both selection bias and measurement error; however, without strong assumptions on the selection mechanism, the estimates are often not nonparametrically identifiable.
-When an issue ``cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically'' ~\cite[pp.
-96]{CoxHink74}.
+When an issue ``cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically''\cite{CoxHink74}.
 In the absence of randomized experiments, the best route forward using any approach is the use of detailed sensitivity analyses and humility in conclusions drawn from such data sources.
 
-\bibliographystyle{plainnat}
+% \bibliography{scibib}
+\bibliographystyle{Science}
 \bibliography{covid-refs}
 
-\appendix
+\section*{Acknowledgments}
+Include acknowledgments of funding, any patents pending, where raw data for the paper are deposited, etc.
 
-\section{Imperfect testing: derivation}
+%Here you should list the contents of your Supplementary Materials -- below is an example.
+%You should include a list of Supplementary figures, Tables, and any references that appear only in the SM.
+%Note that the reference numbering continues from the main text to the SM.
+% In the example below, Refs. 4-10 were cited only in the SM.
+\section*{Supplementary materials}
+
+\subsection*{Imperfect testing: derivation}
 \label{app:imperfect}
@@
 
-\subsection{Estimate of observed data quality}
+\subsubsection*{Estimate of observed data quality}
 
@@
 
-\subsection{Ratio estimator}
+\subsection*{Ratio estimator}
 \label{app:ratio}
@@
 
-\subsection{Estimation of effective reproduction number}
+\subsubsection*{Estimation of effective reproduction number}
 
@@
 
-\section{Derivation for the effective sample size}
+\subsection*{Derivation for the effective sample size}
 \label{section:effss}
@@
 
-% \section{Quotes}
+Materials and Methods\\
+Supplementary Text\\
 
-% I prefer to think of a statistical sensibility rather than statistical thinking.
-It’s “less than an agenda but more than an attitude.”  It allows for methodological preference while avoiding dogma.
-Paired with data analytic humility and I think you have proper “data science”
+Figs. S1 to S3\\
+Tables S1 to S4\\
+References \textit{(4-10)}
 
-% A test of meaningfulness of a possible model for a data-generating process is whether it can be used directly to simulate data.” (p.104).
-In our current setting, this most certainly related to simulation while accounting for measruement error.
+% For your review copy (i.e., the file you initially send in for
+% evaluation), you can use the {figure} environment and the
+% \includegraphics command to stream your figures into the text, placing
+% all figures at the end.  For the final, revised manuscript for
+% acceptance and production, however, PostScript or other graphics
+% should not be streamed into your compliled file.  Instead, set
+% captions as simple paragraphs (with a \noindent tag), setting them
+% off from the rest of the text with a \clearpage as shown  below, and
+% submit figures as separate files according to the Art Department's
+% instructions.
+
+\clearpage
+
+\noindent {\bf Fig. 1.} US total tests and total positive tests per day\\
+\noindent {\bf Fig. 2.} US total tests and total positive tests per day\\
+\noindent {\bf Fig. 3.} US total tests and total positive tests per day\\
+\noindent {\bf Fig. 4.} US total tests and total positive tests per day\\
+\noindent {\bf Fig. 5.} US total tests and total positive tests per day
+
+% Please do not use figure environments to set
+% up your figures in the final (post-peer-review) draft, do not include graphics in your
+% source code, and do not cite figures in the text using \LaTeX\
+% \verb+\ref+ commands.  Instead, simply refer to the figure numbers in
+% the text per {\it Science\/} style, and include the list of captions at
+% the end of the document, coded as ordinary paragraphs as shown in the
+% \texttt{scifile.tex} template file.  Your actual figure files should
+% be submitted separately.
 

### 2020-05-21 write-up/scifile.tex
Update scifile.tex

@@
 
-\section*{Introduction}
+\paragraph*{Introduction}
 
@@
 This pandemic
-%has ripped through our communities and
 has become the focal point of every day life; yet the data landscape for understanding COVID-19 remains limited.
@@
 
-\section*{Analysis of imperfect case-count data}
-\label{section:casecount}
+\paragraph*{Analysis of imperfect case-count data}
 
@@
 COVID-19 testing is no exception.
-The standard assumption is that measurement error leads to parameter attenuation.
-When paired with selection bias, however, the two sources of error become entangled, and resulting errors can become magnified, muted, or even switch signs~\cite{Beesley2020,Beesley2019,Smeden2019}.
 Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the outcome and $0$ when we observe the true outcome.
@@
 If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^{\star} = Y_j(1-P_j) + (1-Y_j) P_j$.
-Suppose disease prevalence is estimated by $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ which is unbiased under simple random sampling.
-We wish to express the error $\tilde  y_n - \bar Y$ in statistical terms.
-In the supplementary materials, we show that the error now can be expressed as
+We suppose disease prevalence is estimated by $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ which is unbiased under simple random sampling.
+
+To better understand this issue, we provide the following intuitive and powerful statistical decomposition of the error between $\tilde y_n$ and the true proportion $\bar Y$
 \begin{equation}
@@
 \rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
-\times \underbrace{\left[ 1 + FP + FN - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right]}_{D_M}.
+\times \underbrace{\left[ 1 + FP + FN - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right]}_{D_M},
 \end{equation}
+where $f_1 := \pr (I_J = 1 \mid Y_J = 1)$ and $f_0 := \pr(I_J = 1 \mid Y_J = 0)$ are sampling rates for COVID-19 positive and negative individuals respectively, and $\Delta = f_1 - f_0$ is the sampling rate differential.
+This extends prior work by \cite{Meng2018} to account for imperfect testing.
+The first term represents \emph{data quality}, the second \emph{data quantity}, and the third \emph{problem difficulty}.
+The final term in brackets is the \emph{measurement error adjustment to data quality} which is a complex function of sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence.
+Figure~2 presents the measurement error data quality adjustment as a function of the relative frequency and odds ratio.
+We see that the adjustment can be both positive and negative as well as a range of magnitudes.
 
-The above selection mechanisms are random and independent of the outcome of interest.
-When this is not the case, selection effects may cause bias in the above estimates.
-To better understand this issue, \cite{Meng2018} provided the following intuitive and powerful statistical decomposition of the error between $\bar y$ and the true proportion $\bar Y$
-$$
-\bar y_n - \bar Y =  \rho_{I, Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_Y.
-$$
-The first term represents \emph{data quality}, the second \emph{data quantity}, and the third \emph{problem difficulty}.
 The term $\rho_{I,Y}$ is the empirical correlation between the population values~$\{ Y_j \}_{j=1}^N$ and the selection values $\{ I_j \}_{j=1}^N$.
@@
 The key issue with selective testing is that $E_{\I} [ \rho_{I,Y} ] \neq 0$.
-Meng identified this as the fundamental issue that can lead to paradoxes in the analysis of big data.
+We can see here  selection bias, however, the two sources of error become entangled, and resulting errors can become magnified, muted, or even switch signs~\cite{Beesley2020,Beesley2019,Smeden2019}.
 
@@
 
-The remainder of this paper aims to build upon these fundamental insights by extending the decomposition in two directions: accounting for  measurement error and the temporal nature of the pandemic.
-We then discuss effective sampling methods and their importance in the current pandemic.
-
 \subsection*{Imperfect testing}
-
-Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the outcome and $0$ when we observe the true outcome.
-We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.
-If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^{\star} = Y_j(1-P_j) + (1-Y_j) P_j$.
-Suppose disease prevalence was estimated as the fraction who tested positive for COVID-19, i.e., $\bar y_n^\star = \frac{\sum_{i=1}^N I_j Y_j^\star}{\sum_{i=1}^N I_j}$.
-We can again investigate the error compared to the true prevalence $\bar Y$ in statistical terms:
-$$
-\bar y_n^\star - \bar Y = \sqrt{\frac{1-f}{f}} \left[ \rho_{I,Y} \times \sigma_Y + \rho_{I,PZ} \times \sigma_{PZ} + \sqrt{\frac{f}{1-f}}  \left( FP - (FP+FN) \bar Y \right) \right] .
-$$
-where $Z = 1-2Y$; see Appendix~\ref{app:imperfect} for the derivation.
-The first term in the large brackets represents the perfect testing regime; to this end, we refer to $\rho_{I,Y}$ as the \emph{true data quality}.
-The second term represents the interaction between imperfect testing and selection bias.
-The variable $PZ = 1[Y=0,P=1] - 1[Y=1, P=1]$ is non-zero only when $P=1$ (i.e., the outcome is incorrectly reported); to this end, we refer to $\rho_{I,PZ}$ as the \emph{observed data quality} that accounts for both selection bias and measurement error.
-In the appendix, we show the sign of $\rho_{I,PZ}$ is the opposite of the sign of $\rho_{I,Y}$, which implies that the observed data quality adds error in the opposite direction from the true data quality.
-Finally, the third term represents the bias due to imperfect testing.
-
-We start by considering the first two terms and assess whether the sign of the bias can reverse due to the interaction of measurement error and selection bias.
-To do this, we define the sampling rates differential.
-Let $f_1 := \pr (I_J = 1 \mid Y_J = 1)$ and $f_0 := \pr(I_J = 1 \mid Y_J = 0)$ be the sampling rates.
-Then $\Delta = f_1 - f_0$ is the sampling rate differential.
-Using these terms, we can re-express the first two terms as
-$$
-\rho_{I,Y} \times \sigma_Y + \rho_{I,PZ} \times \sigma_{PZ} =
-\rho_{I,Y} \times \sigma_Y \left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right].
-$$
-The final term in brackets is the \emph{measurement error adjustment to data quality} which is a complex function of sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence.
-Note $\sgn(\Delta) = \sgn(\rho_{I,Y})$ by equation~\ref{eq:binaryrho} in the appendix, which implies the measurement error adjustment either shrinks the data quality measure toward zero or reverses its sign.
-
-While prior investigations have noted the interaction between measurement error and selection bias~\cite{Beesley2020,Beesley2019,Smeden2019}, the interaction with the sample size relative to the population, i.e., $f$, has largely been ignored.
-The above statistical decomposition clarifies the importance of this quantity~$f$.
-In particular, note that the statistical error also includes a bias term due to measurement error and this term increases as the sampled fraction $f$ increases.
-Therefore, how the first two terms interact with the final term depends on the fraction of the population sampled.
-This interaction is complex, but implies that whether the estimate $\bar y^\star_n$ is an overestimate or underestimate is a complicated question due to the relation amongst these three pieces.
-
-Consider again the current COVID-19 pandemic.
-For now, we continue to assume the ratio of conditional selection rates $f_1/f_0$ is equal to 1.5.
-In Section~\ref{section:est_dq}, we discuss recent research suggesting a false negative rate around 17.2\% and a false positive rate around 0.05\%.
-Under these rates and the current US prevalence rates, the ratio of the MSE to MSE under no measurement error is 0.436; if we switch false negative rates down to 5\% and increase the false positive rate to 5\% then the relative MSE is 2.89.
-This is merely to demonstrate that in some cases we see a huge increase in MSE and in other settings we have a huge decrease in MSE.
-What drives this is the false positive and negative rate interaction with prevalence and sampling rates.
-Therefore, whether we are better or worse off with respect to the MSE is a very difficult question to answer.
-
-The attentive data analyst will recognize the estimator $\bar y_n$ is biased even for simple random samples and, if sensitivity and specificity were known a priori, may suggest the alternative estimator $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ which is unbiased under simple random sampling.
-We again wish to express the error $\tilde  y_n - \bar Y$ in statistical terms.
-In the appendix, we show that the error now can be expressed as
-\begin{equation}
-\label{eq:error}
-\rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
-\times \underbrace{\left[ 1 + FP + FN - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right]}_{D_M}.
-\end{equation}
-The first term is the same as before but increased by $(1 + FP + FN)$ to account for the additional uncertainty due to measurement error.
-The second term is the interaction between selection bias and measurement error.
-Figure~2 presents the measurement error data quality adjustment as a function of the relative frequency and odds ratio.
-We see that the adjustment can be both positive and negative as well as a range of magnitudes.
 

### 2020-05-21 write-up/scifile.tex
Update scifile.tex

@@
 The reason is straightforward: current models do not account simultaneously for selection bias and measurement error.
-In the US, for example, limited testing capacity has caused health departments to test only high-risk individuals.
-Moreover, testing requires individuals to request a test and receive approval from their doctor.
+In the US, for example, individuals must to request a test and receive approval from their doctor.
 Demands for increased testing capacity, while laudable, do not guarantee increased data quality.
-Aggressive measures to increase testing capacity that decrease data quality may have a mitigated impact in our ability to estimate quantities of interest such as prevalence and effective reproduction number.
-We will demonstrate why there is a strong need for more precision and care in presenting COVID-19 analyses to the broader research and non-academic communities.
+We will demonstrate that there is limited information gain from increased testing when selection bias persists and testing inaccuracies potentially increase.
 

### 2020-05-21 write-up/scifile.tex
update science paper

@@
 
-% In setting up this template for *Science* papers, we've used both
-% the \section* command and the \paragraph* command for topical
-% divisions.  Which you use will of course depend on the type of paper
-% you're writing.  Review Articles tend to have displayed headings, for
-% which \section* is more appropriate; Research Articles, when they have
-% formal topical divisions at all, tend to signal them with bold text
-% that runs into the paragraph, for which \paragraph* is the right
-% choice.  Either way, use the asterisk (*) modifier, as shown, to
-% suppress numbering.
-
 \paragraph*{Introduction}
@@
 The World Health Organization has declared the coronavirus disease 2019 (COVID-19) a public health emergency.
-As of May 21st, 2020, over 5 million cases have been confirmed worldwide with 1.5 million confirmed cases and over 90 thousand COVID-19 deaths across the United States.
+As of May 21st, 2020, over 5 million cases have been confirmed worldwide with 1.5 million cases and over 90 thousand confirmed deaths across the United States.
 This pandemic
@@
 
-Let $N$ denote the population size.
-At a fixed time, let $Y_j$ denote COVID-19 status for the $j$th individual in the population, $j=1,\ldots, N$.
-Here, as in survey methodology~\cite{Cochran77}, we treat COVID-19 status as a fixed but unknown quantity of interest.
-For simplicity, we start by ignoring the dynamic nature of the outbreak and the fact that individuals can recover.
-We also let $I_j \in \{0,1\}$ be an indicator that the individual was selected for testing ($I = 1$) or not ($I=0$).
+Let $N$ denote the population size.  At a fixed time, let $Y_j$ denote COVID-19 status for the $j$th individual in the population, $j=1,\ldots, N$.
+Here, for simplicity, we ignore the dynamic nature of the outbreak and the fact that individuals can recover.
+As in survey methodology~\cite{Cochran77}, we treat COVID-19 status as a fixed but unknown quantity of interest.
+Let $I_j \in \{0,1\}$ be an indicator that the individual was selected for testing ($I = 1$) or not ($I=0$).
+Suppose we are interested in COVID-19 prevalence, i.e., the population average $\bar Y = Y/N$.
+Suppose that $n$ tests are performed and we observe the values $y_1, \ldots, y_n \in \{0,1\}$.
+Then a natural candidate for prevalence is the proportion of positive tests $\bar y = n^{-1} \sum_{i=1}^n y_i$.
 
-Here we consider estimates of the number of COVID-19 cases and/or disease prevalence, i.e., the population total $Y = \sum_{j=1}^N Y_j$ or the population average $\bar Y = Y/N$.
-Suppose that $n$ tests are performed and we observe the values $y_1, \ldots, y_n \in \{0,1\}$.
-Then a natural candidate for prevalence is the proportion of positive tests $\bar y = n^{-1} \sum_{i=1}^n y_i$, and a natural candidate for overall cases is $N \times \bar y$.
-Under simple random sampling (SRS), both are unbiased estimators of the  quantities of interest and the variance of the prevalence estimator equals $(N-1)^{-1} \times (1-f)/f \times \sigma_Y^2$ where $f = n/N$ is the sampling fraction and $\sigma_Y^2 = \frac{1}{N} \sum_{i=1}^N (Y_i - \bar Y)^2 = \bar Y (1- \bar Y)$.
-
-In the current context, the true status is not observed.
-Instead, individuals are tested for COVID-19 status.
 Tests are imperfect.
-COVID-19 testing is no exception.
-Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the outcome and $0$ when we observe the true outcome.
-We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.
-If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^{\star} = Y_j(1-P_j) + (1-Y_j) P_j$.
-We suppose disease prevalence is estimated by $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ which is unbiased under simple random sampling.
-
-To better understand this issue, we provide the following intuitive and powerful statistical decomposition of the error between $\tilde y_n$ and the true proportion $\bar Y$
+COVID-19 testing is no exception which implies a selected individual's true status is not observed.
+Here we assume the false positive rate (FP) and false negative rate (FN) are known.
+Disease prevalence is estimated by $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ which is unbiased under simple random sampling (SRS).
+To understand the impact of selection bias and imperfect testing, we provide a statistical decomposition of the error between $\tilde y_n$ and $\bar Y$
 \begin{equation}
@@
 where $f_1 := \pr (I_J = 1 \mid Y_J = 1)$ and $f_0 := \pr(I_J = 1 \mid Y_J = 0)$ are sampling rates for COVID-19 positive and negative individuals respectively, and $\Delta = f_1 - f_0$ is the sampling rate differential.
-This extends prior work by \cite{Meng2018} to account for imperfect testing.
+This extends work by \cite{Meng2018} to account for imperfect testing.
 The first term represents \emph{data quality}, the second \emph{data quantity}, and the third \emph{problem difficulty}.
-The final term in brackets is the \emph{measurement error adjustment to data quality} which is a complex function of sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence.
-Figure~2 presents the measurement error data quality adjustment as a function of the relative frequency and odds ratio.
-We see that the adjustment can be both positive and negative as well as a range of magnitudes.
+The final term represents the \emph{imperfect testing adjustment} which is a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence.
+Figure~2 presents $D_M$ as a function of the relative frequency and odds ratio can be both positive and negative as well as a range of magnitudes~\cite{Beesley2020,Beesley2019,Smeden2019}.
 
-The term $\rho_{I,Y}$ is the empirical correlation between the population values~$\{ Y_j \}_{j=1}^N$ and the selection values $\{ I_j \}_{j=1}^N$.
-Under simple random sampling, $E_{\I} [ \rho_{I,Y} ] = 0$, where the expectation is with respect to the selection mechanism~$\I$, so there is no bias.
-The SRS variance formula above shows that $E_{\I} [ \rho_{I,Y}^2 ]  = 1/(N-1)$.
-The key issue with selective testing is that $E_{\I} [ \rho_{I,Y} ] \neq 0$.
-We can see here  selection bias, however, the two sources of error become entangled, and resulting errors can become magnified, muted, or even switch signs~\cite{Beesley2020,Beesley2019,Smeden2019}.
-
-Here we highlight two key insights from~\cite{Meng2018} that are relevant to the COVID-19 crisis.  First, comparing the mean-squared error under selection mechanism $\I$ and SRS, we see that
+Comparing the mean-squared error (MSE) under a selection mechanism $\I$ with imperfect testing and SRS with perfect testing (i.e.,$FP=FN=0$), we see that
 $$
-\frac{E_{\I} \left[ (\bar y_n - \bar Y)^2 \right]}{\sqrt{V_{SRS} (\bar Y)}} = (N-1) E_{\I} \left[ \rho_{I,Y}^2 \right].
+\frac{E_{\I} \left[ (\bar y_n - \bar Y)^2 \right]}{\sqrt{V_{SRS} (\bar Y)}} = (N-1) E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right].
 $$
-This points to a troubling and paradoxical situation: the error relative to SRS increases as a function of population size.
-\cite{Meng2018} termed this the ``Law of Large Populations'' (LLP).
-This points to a critical issue in the current media practices in communication of case-count data: two countries with the same testing strategy (i.e., $E_{\I} [\rho_{I,Y} ]$ equal) can yield wildly different estimates due to population size.
-Large countries like the US may have similar true prevalence rates as smaller countries like the UK.
-Even under similar testing strategies, the mean-squared error in the prevalence rate for the US will have substantially more variation in comparison to SRS.
-Comparing the US to the UK, for example, the MSE will increase by a factor of almost five.
+This points to a paradoxical situation: the error relative to SRS increases as a function of population size.
+This was termed the ``Law of Large Populations'' (LLP)~\cite{Meng2018} and continues to hold in our imperfect testing context.
+This points to a critical issue in the current media practices in communication of case-count data: two countries with the same testing strategy can yield very different estimates due to population size.
+When comparing countries like the US to the UK, the MSE will increase by a factor of almost five.
 Thus conclusions drawn from observed case-count records may be not just wrong, but very wrong.
 
-Second, there have been calls for increased testing.
-While important, many conflate increased testing capacity with increased quality of testing.
-We will discuss this point further in Section \ref{section:discussion}.
-As of May 8th, the US has performed $8,412,095$ tests in total.
-Figure~1 shows the trajectory of testing per day and positive cases per day in the US.
-The US population is roughly $328$ million, meaning the fraction sampled is $f = 0.026$.
-After smoothing, the empirical prevalence is $9.1\%$ on May 8th.
-Supposing COVID-19 positive individuals are $2$ times more likely to get tested than those individuals who are COVID-19 negative, the question is ``What is the sample size from a SRS that would yield equivalent MSE in the estimated prevalence?'' Using the above MSE result, Meng showed the effective sample size is equal to $n_{eff} \leq \frac{1}{E_I [ \rho_{I,Y}^2 ]} \frac{f}{1-f}$.
-Under these parameters, the effective sample size is $14$.
-Recent proposals~\cite{Siddarth2020} have argued in favor of increasing overall testing capacity.
-With more tests available, the relative sampling rate may reduce.
-Supposing it drops to $1.2$ then the effective sample size will increase to $299$.
-See Appendix~\ref{section:effss} for the mathematical details and calculations in additional scenarios.
-The above calculations point to a sad state of affairs, saying that the effective sample size even in best case scenarios is not better than a small random sample from the population.
-
-\subsection*{Imperfect testing}
-
-We next investigate the impact of measurement error on the effective sample size calculation.  The effective sample size can now be bounded by
+An important follow-up question is ``What is the sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' We show this \emph{effective} sample size can be bounded by
 $$
@@
 $$
-In Section \ref{section:est_dq}, we discuss recent studies that estimate the COVID-19 false positive rate of $0.5$\% and false negative rate $17.2$\%.
-Supposing COVID-19 positive individuals are $2$ times more likely to get tested than those individuals who are COVID-19 negative, the effective sample size drops from $14$ to $10$.
-If the relative sampling fraction drops to $1.2$ then the effective sample size drops from $299$ to $216$.
+As of May 8th, the US has performed $8,412,095$ tests.
+The US population is roughly $328$ million, so the fraction sampled is $f = 0.026$.
+The empirical prevalence is $9.1\%$ on May 8th.
+A recent study estimates a false positive and negative rate for antibody testing using a lateral flow immunoassay are $0.5$\% and $17.2$\% respectively \cite{Bendavid2020}.
+Supposing COVID-19 positive individuals are $2$ times more likely to get tested, then using these parameters the effective sample size is $10$.
+Recent proposals~\cite{Siddarth2020} have argued for increased testing capacity, which will reduce the relative sampling rate.
+Supposing it drops to $1.2$ then the effective sample size will increase to $216$.
+This points to a sad state of affairs, saying that the effective sample size even in optimistic scenarios is not better than a small random sample from the population.
 
-The above argument shows that the effective sample size is affected in a complex manner by the interaction between selection bias and measurement error.
-Here, we discuss the impact of the interaction on the effective sample size when testing capacity is tied to measurement error.
-In the current COVID-19 pandemic, there have been well-justified calls for increases in testing capacity.
-These increases, however, may come at the cost of increases in false-positive and false-negative rates.
+Increased testing capacity may also come at the cost of increased in false positive and negative rates.
 In terms of effective sample size, if the data quality term remains constant, then testing increases may not significantly improve the effective sample size.
@@
 
-\subsection*{Regrettable rates: complex biases resulting from self-selection}
-\label{section:rates}
+\paragraph*{Regrettable rates: complex biases resulting from self-selection}
 
-The data analyst, now wary of estimating prevalence and total counts, pauses and thinks.
-They return shortly thereafter, with a follow-up consideration: \emph{Ok, perhaps estimating prevalence is difficult.
-Certainly, however, we can estimate the rate of growth.
-All I want to know is when we hit the point at which the curve flattens and number of deaths decrease.
-That can't be too hard, surely!}
-
-Unfortunately, ratio estimators do not cancel errors as now both the numerator and the denominator are uncertain.
-This means selection bias and measurement error can have paradoxical effects.
-Here we consider the ratio estimators for the relative change in the prevalence rate.
-We let $\bar Y_{t-1}$ and $\bar Y_{t}$ denotes the prevalence (across the entire population) at time step $t-1$ and $t$ respectively (i.e., prevalence on two consecutive days).
-Then the ratio estimator is given by $r = \bar y_t / \bar y_{t-1}$.
-We assume both numerator and denominator are the prevalence estimates adjusted for measurement error.
-If the sample size at each time were equal (i.e., $n_1 = n_2 = n$) then this analysis would be equivalent to comparing the increase in observed case-counts (i.e., $\bar y_t/\bar y_{t-1} = y_t/y_{t-1}$).
-We wish to express the error $\frac{\bar y_t}{\bar y_{t-1}} - \frac{\bar Y_{t}}{\bar Y_{t-1}}$ in statistical terms.
-Using a Taylor series approximation, we show in the appendix that the error can be expressed approximately as
+A separate question for communities facing COVID-19 is knowledge of rates of growth in case-count and the number of COVID-19 deaths decrease.
+Let $\bar Y_{t-1}$ and $\bar Y_{t}$ denotes the prevalence (across the entire population) at time step $t-1$ and $t$ respectively (i.e., prevalence on two consecutive days).
+Then the ratio estimator is $r = \bar y_t / \bar y_{t-1}$.
+Using a Taylor series approximation, the error ${\bar y_t}/{\bar y_{t-1}} - {\bar Y_{t}}/{\bar Y_{t-1}}$ can be expressed approximately as
 $$
@@
 
+\subsection*{Technical details}
+
+Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the outcome and $0$ when we observe the true outcome.
+We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.
+If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^{\star} = Y_j(1-P_j) + (1-Y_j) P_j$.
+
 \subsection*{Imperfect testing: derivation}

### 2020-05-21 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 \DeclareGraphicsExtensions{.eps, .ps}
-\usepackage{hyperref}
 \usepackage{amsmath, amsthm, amsfonts}
@@
 \begin{document}
+
 \title[The Hypothesis of Testing]{The Hypothesis of Testing: Paradoxes arising out of reported coronavirus case-counts} %
@@
 Despite these steps, the data landscape for understanding COVID-19 remains limited.
-Public databases maintained by \href{https://bit.ly/2UqFSuA}{Johns Hopkins University}~\citep{JHU_Lancet} and the \href{https://bit.ly/2vUHfrK}{New York Times}~\cite{NYT} provide incoming county-level information of confirmed cases and deaths.
+Public databases maintained by Johns Hopkins University~\citep{JHU_Lancet} and the New York Times~\cite{NYT} provide incoming county-level information of confirmed cases and deaths.
 Statisticians, epidemiologists, economists, and data scientists have used this granular data to forecast COVID-19 case-counts, deaths, and hospitalizations~\citep{Song2020,Ray2020,2020.IHME}.
@@
 Most often, our main interest is in causal effects on the population of COVID-19 positive.
-Randomization within the clinical trial design yields internal validity, but we also need external validity to estimate the correct marginal effect of interest.
+Randomization within the clinical trial design yields internal validity, but we also need external validity to estimate the correct marginal effect of interest~\cite{Keiding2016}.
 

### 2020-05-21 write-up/scifile.tex
Update scifile.tex

@@
 has become the focal point of every day life; yet the data landscape for understanding COVID-19 remains limited.
-Public databases maintained by Johns Hopkins University~\cite{JHU_Lancet} and the New York Times~\cite{NYT} provide incoming county-level information of confirmed cases and deaths.
-Statisticians, epidemiologists, economists, and data scientists have attempted to use this granular data to forecast COVID-19 case-counts, deaths, and hospitalizations~\cite{Song2020,Ray2020,2020.IHME}.
+Public databases~\cite{JHU_Lancet,NYT} provide incoming county-level information of confirmed cases and deaths.
+Statisticians, epidemiologists, economists, and data scientists have used this granular data to forecast COVID-19 case-counts, deaths, and hospitalizations~\cite{Song2020,Ray2020,2020.IHME}.
 
 The goal of this paper is to express reservations at the use of case-counts as a proxy for disease prevalence and in estimation of standard epidemiological models for inference and forecasting.
-The reason is straightforward: current models do not account simultaneously for selection bias and measurement error.
-In the US, for example, individuals must to request a test and receive approval from their doctor.
-Demands for increased testing capacity, while laudable, do not guarantee increased data quality.
-We will demonstrate that there is limited information gain from increased testing when selection bias persists and testing inaccuracies potentially increase.
+The reason is straightforward: observed case-count data is plagued by selection bias and measurement error.
+We will demonstrate that the information gained from increasing testing capacity is limited when selection bias and testing inaccuracies persist.
 
@@
 
-Let $N$ denote the population size.  At a fixed time, let $Y_j$ denote COVID-19 status for the $j$th individual in the population, $j=1,\ldots, N$.
-Here, for simplicity, we ignore the dynamic nature of the outbreak and the fact that individuals can recover.
+Let $N$ denote the population size.
+At a fixed time, let $Y_j$ denote COVID-19 status for the $j$th individual in the population, $j=1,\ldots, N$.
+To start, for simplicity, we ignore the dynamic nature of the outbreak and the fact that individuals can recover.
 As in survey methodology~\cite{Cochran77}, we treat COVID-19 status as a fixed but unknown quantity of interest.
 Let $I_j \in \{0,1\}$ be an indicator that the individual was selected for testing ($I = 1$) or not ($I=0$).
-Suppose we are interested in COVID-19 prevalence, i.e., the population average $\bar Y = Y/N$.
-Suppose that $n$ tests are performed and we observe the values $y_1, \ldots, y_n \in \{0,1\}$.
-Then a natural candidate for prevalence is the proportion of positive tests $\bar y = n^{-1} \sum_{i=1}^n y_i$.
+Suppose we are interested in COVID-19 prevalence, i.e., the population average $\bar Y = Y/N$, and that $n$ tests are performed and we observe the values $y_1, \ldots, y_n \in \{0,1\}$.
+The natural estimator is the proportion of positive tests $\bar y = n^{-1} \sum_{i=1}^n y_i$.
 
@@
 Here we assume the false positive rate (FP) and false negative rate (FN) are known.
-Disease prevalence is estimated by $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ which is unbiased under simple random sampling (SRS).
+Here we consider the estimator $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ which is unbiased under simple random sampling (SRS).
 To understand the impact of selection bias and imperfect testing, we provide a statistical decomposition of the error between $\tilde y_n$ and $\bar Y$
@@
 
-Comparing the mean-squared error (MSE) under a selection mechanism $\I$ with imperfect testing and SRS with perfect testing (i.e.,$FP=FN=0$), we see that
+Comparing the mean-squared error (MSE) under a selection mechanism $\I$ with imperfect testing and SRS with perfect testing, we see that
 $$
@@
 This was termed the ``Law of Large Populations'' (LLP)~\cite{Meng2018} and continues to hold in our imperfect testing context.
-This points to a critical issue in the current media practices in communication of case-count data: two countries with the same testing strategy can yield very different estimates due to population size.
-When comparing countries like the US to the UK, the MSE will increase by a factor of almost five.
+This points to a critical issue in current media practices in communication of case-count data: two countries with the same testing strategy can yield very different estimates due to population size.
+When comparing countries like the US to the UK, the MSE increasea by a factor of almost five.
 Thus conclusions drawn from observed case-count records may be not just wrong, but very wrong.
@@
 Consider the simplest case where $f$ increases from $0.05$ to $0.1$; suppose this increase was associated with  false positive rate increasing from $0.5$\% to $5$\% and the false negative rate rising from $5\%$ to $20$\%.
-Then the standard analysis would suggest a reduction in the MSE by 47\% but the actual decrease is by 26\%.
+Then the standard analysis would suggest a reduction in the MSE by 47\% but the actual reduction is 26\%.
 This implies that doubling the sampling rate will only increase the effective sample size by a factor $2.9$ rather than the expected $4$ fold increase that is expected.
-In a worst case scenario, if the false negative rate climbs to $0.30$ and the false positive rate to $0.10$ then the factor effective sample size increases by only a factor of $2.3$.
 
-\paragraph*{Regrettable rates: complex biases resulting from self-selection}
+\paragraph*{Regrettable rates.}
 
-A separate question for communities facing COVID-19 is knowledge of rates of growth in case-count and the number of COVID-19 deaths decrease.
-Let $\bar Y_{t-1}$ and $\bar Y_{t}$ denotes the prevalence (across the entire population) at time step $t-1$ and $t$ respectively (i.e., prevalence on two consecutive days).
-Then the ratio estimator is $r = \bar y_t / \bar y_{t-1}$.
-Using a Taylor series approximation, the error ${\bar y_t}/{\bar y_{t-1}} - {\bar Y_{t}}/{\bar Y_{t-1}}$ can be expressed approximately as
+Another important question is whether selection bias and imperfect testing impact the ability to estimate rates of growth in case-count.
+Let $\bar Y_{t-1}$ and $\bar Y_{t}$ denote the prevalence on two consecutive times $t-1$ and $t$.
+We consider the ratio estimator $r = \bar y_t / \bar y_{t-1}$.
+Using a Taylor series approximation, the error between ${\bar y_t}/{\bar y_{t-1}}$ and the true rate ${\bar Y_{t}}/{\bar Y_{t-1}}$ can be expressed approximately as
 $$
@@
 where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement error adjustment, and $CV(Y_j) = \sigma_{Y_j}/\bar Y_j$ is the coefficient of variation at time step $j$.
-The magnitude of the error depends on the population-level ratio $\bar Y_{t} / \bar Y_{t-1}$ so a large decrease in the prevalence rate will have a relatively small error as compared to a large increase in the prevalence rate.
-The second term represents the ``cancellation'' that the naive analyst is hoping will occur.
-The analyst's intuition only holds when the difference is equal to zero.
-This occurs at times when data quality, sampling fraction, measurement error, and prevalence are all constant across time.
+The error magnitude depends on the true rate $\bar Y_{t} / \bar Y_{t-1}$ so a large decrease will have a relatively small error compared to a large increase.
+The second term represents potential \emph{cancellation} which may occurs when data quality, sampling fraction, measurement error, and prevalence are all constant across time.
 
@@
 Figure~5b presents the fraction of tests that were positive per day in New York state.
-For cancellation, we would need the expected data qualities to be uncorrelated and equal.
-Testing appears relatively stable over a week which suggests assuming $\Delta_t$ (i.e., differential in the selection rate) varies slowly over time may be a plausible assumption.
-Note, however, that constant $\Delta_t$ does not imply constant data quality (see Appendix \ref{section:effss} for details).
-Figure~5a shows the number of tests has remained relatively constant in the last few weeks, specifically relative to the total NY population, assuming $f_t$ also varies slowly in time is reasonable.
-Finally, the measurement error adjustment is likely to be constant as no new testing procedures have been reported.
+For cancellation, the expected data qualities need to be uncorrelated and equal.
+Testing per day suggests it is plausible to assume $\Delta_t$ varies slowly over time.
+Note, however, that constant $\Delta_t$ does not imply constant data quality.
+Figure~5a shows the number of tests has remained approximately constant in the last few weeks relative to the total NY population.
+Therefore, the assumption that $f_t$ also varies slowly in time is reasonable.
+Finally, the assumption that the imperfect testing adjustment is constant is reasonable as no new testing procedures have been reported.
 This would not be true in states like Michigan, where testing criteria have been expanding every few weeks.
-The final component is an adjustment to this differential error based on the statistical error at the first time point.
 
-Figure~3a shows a trajectory of the true ratio and the potential biased estimators under an SIR model for the epidemic dynamics, with state evolution given by the following ODE
+Figure~3a shows a trajectory of the true ratio and the potential biased estimators under an SIR model~\cite{Pastor2001,Newman2002,Parshani2010} for the epidemic dynamics, with state evolution given by the following ODE
 $$
@@
 Figure~4a presents the fraction of population newly infected individuals at each time step for $\beta = 1.4$ and $\gamma = 0.2$ (black line).
-In each case, the rate of change in prevalence is likely to be  overestimated prior to the peak and then underestimated afterwards.
+In terms of bias, the rate is likely to be overestimated prior to the peak and then underestimated afterwards.
 From a decision making perspective, such biases have potential impact.
@@
 What both sides miss is that the direction of bias is non-constant over time.
-After we hit the peak, estimates will likely under-estimate prevalence.
+After the peak, the estimator likely under-estimates the rate of change.
 Such bias puts pressure on lawmakers and governors to prematurely relax social distancing measures.
-Moreover, the question is ``Can we trust the observed data to let us know if we have reached the peak?''  It appears that the peak time is the easiest to estimate in terms of having minimal error using availability data.
-Note, however, that the standard errors on such estimates will be incredibly wide due to the low effective sample size as discussed in Section~\ref{section:casecount}.
+Moreover, it appears that the peak time is the easiest to estimate in terms of having minimal error using availability data.
+Note, however, that the standard errors on such estimates will be wide due to relatively low effective sample size.
 
-\subsection*{Estimation of effective reproduction number}
-\label{section:r0-estimation}
-Many well-respected epidemiologists argue that tracking the effective reproduction number $R_t$ is the only way to manage through the crisis~\cite{Gabriel2020}.
-We show here how these estimates are also impacted by selection bias and measurement error.
-For simplicity, we again assume the estimates are adjusted for known false positive and false negative rates.
-\cite{Bettencourt2008} show that under a Poisson likelihood, a simple relation between the trajectory of new cases and the effective reproduction number can be derived.
-In particular, under an SIR model, the number of case counts on day $t$ is Poisson distributed with rate $K_{t-1} \exp \left( \frac{1}{\gamma} (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval which is approximately $7$ days for COVID-19~\cite{Sanche2020}.
-Using this, a simple estimator of the effective reproduction number at time $t$ can be given by
+\paragraph*{Estimation of effective reproduction number.}
+Many epidemiologists argue that tracking the effective reproduction number $R_t$ is the only way to manage through the crisis~\cite{Gabriel2020}.
+Under a Poisson likelihood, a simple relation between the trajectory of new cases and the effective reproduction number can be derived \cite{Bettencourt2008}.
+In particular, under an SIR model the number of case counts on day $t$ is Poisson distributed with rate $K_{t-1} \exp \left( \frac{1}{\gamma} (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval which is approximately $7$ days for COVID-19~\cite{Sanche2020}.
+Using this, a simple estimator can be given by
 $$
@@
 $$
-Of course, we do not observe $K_t$ and $K_{t-1}$ but noisy proxies $y_t$ and $y_{t-1}$, i.e., the observed case-count on day $t$ and $t-1$ respectively.
-The hope is that $y_t/y_{t-1}$ is a good proxy for $K_t/K_{t-1}$ where here $y_t$ is conceptualized as the number of new cases on day $t$.
-However, if we think about estimating number $K_t$ under SRS of new cases among those susceptible on day $t$, the natural estimator is $S_t \bar y_t$.
-Unfortunately the number of individuals susceptible on day $t$ is unknown.
-Here, we assume that $\hat R_t = 1 + \frac{1}{\gamma} \log \left( \bar y_t / \bar y_{t-1} \right)$ is the estimated effective reproduction number on day $t$.
-This estimator adjusts for sample size variation across days but not the varying population size.
-The estimator is close in spirit to the estimator derived under assuming the process is fully observed.
-Similar to Section~\ref{section:rates}, the proxy has error in both numerator and denominator that may cause issues.
-We again wish to express the statistical error of $\hat R_t - R_t$ in useful terms.
-We can re-arrange the error decomposition from Section~\ref{section:rates} to show that this error is given by
+Of course, we do not observe $K_t$ and $K_{t-1}$.
+Under SRS of new cases among those susceptible on day $t$, the natural estimator is $S_t \bar y_t$.
+Unfortunately the number of susceptible individuals on day $t$ is unknown.
+Here, we study the estimator $\hat R_t = 1 + \frac{1}{\gamma} \log \left( \bar y_t / \bar y_{t-1} \right)$.
+We can again express the statistical error of $\hat R_t - R_t$ in useful terms as follows
 $$
@@
 $$
-So here we see the exact same trade-off as with the rates but on the logarithmic scale.
+This demonstrates the exact same trade-off as with the rates but on the logarithmic scale.
 For small values of $x$, we have $\log(1+x) = x$ and so the biases are similar.
@@
 Here for simplicity, we assume the fraction of the population that is susceptible remains large and near the total population size.
-Under these assumption, the potential bias is similar to that for the ratio estimator adjusted to be on the logarithm scale.
 Changes to this assumption can impact the potential bias.
@@
 
-\subsection*{Effect modifiers in COVID-19 clinical trials}
+\paragraph*{Cross-population rate comparisons.}
+So far we have focused on understanding the limitations of using case-count data to understand population quantities of interest for a \emph{single} population.
+Many are interested in cross-population comparisons.
+Many pieces in the media as well as academic articles have plots of case-count over time aligned by time of first known case.
+Others have claimed such comparisons unfair due to unequal population sizes and instead plotted case-count per million.
+Here, for simplicity, we focus on comparing the estimated effective reproductive rate.
+We assume the two time-series are aligned so that $t=0$ is the time of first case.
+
+We focus on the interaction of biases in estimation of the two trajectories when the peak infection times differ slightly.
+Recall that the error allowed for over-estimation prior to the peak and then under-estimation post-peak.
+Here, these errors can mingle in interesting ways.
+Consider two countries (A and B) in which the peak occurs 2 weeks prior for country A then country B but other.
+Figure~4b presents such a comparison where each country's disease trajectory follows an SIR model as described above with different parameters.
+Figure~4 shows how the errors interact in complex ways.
+First, the gap is correctly estimated; then the gap is over-estimated as country A sees a rapid rise in cases; then the gap is even more over-estimated as country A improves and the rate is under-estimated while country B sees a rapid increase in their case-count and their rate is over-estimated; then country A's rate is correctly estimated while country B's rate is under-estimated as it sees improvement in infection rates; finally, the gap disappears.
+We are not claiming this will always be the case, and indeed for small enough selection bias and/or large gaps in case-counts such complex behaviour will not occur; however, this is simply to point out how observed information can tell a more complex story than the truth (i.e., country A recovers 1 month prior to country B).
+
+\paragraph*{Estimation of Data Quality.} Here, we try and address the empirical question of how to estimate data quality $E_{\I} \left[ \rho_{I,Y} \right]$ in the current context.
+In \cite{Meng2018}, estimation relied on observing the true outcome.
+In election data, for example, this is possible as election vote totals can be used as ground truth.
+This is not possible in the current crisis.
+Here, we propose a simple procedure that uses survey samples as noisy but unbiased estimates of the true prevalence.
+Our goal is less inferential and more to provide a sensitivity analysis that can aid in understanding the amount of information in observational COVID case-count data.
+
+On April 23rd, Andrew Cuomo announced results from a study in New York state.
+It found 13.9\% of 3,000 people tested across the state had signs of the virus.
+The study did not report sensitivity and specificity; therefore, we take reported measures from the Santa Clara study~\cite{Bendavid2020}.
+The reported specificity is $82.8\%$ (95\% CI: 76.0-88.4\%) and sensitivity is $99.5\%$ (95\% CI: 99.2-99.7\%).
+This corresponds to a false negative rate of $17.2\%$ (11.6-24.0\%) and false positive rate of $0.5\%$ (0.3 - 0.8\%).
+Correcting for these rates we have an estimated prevalence of $15.9\%$ with a range of 14.8\% to 17.0\% using the ranges of the false positive and negative rates.
+
+Given the report was on the 23rd, we suppose for simplicity that the study was run several days prior on the 20th of April.
+On the 20th, there were $19,654$ tests performed and $617,555$ tests had been performed up to to and including the 19th.
+The population of New York state is $19,378,102$.
+Subtracting off the number of individuals who had already been tested yields a sampling fraction of $f = 0.001$ on the 20th.
+The smoothed prevalence estimate on the 20th is 28.3\%.
+Adjusting for measurement error yields an estimate of $32.5\%$.
+
+Here, we assume the New York survey is a simple random sample from the state's population.
+We recognize that the survey is not such a sample, but techniques like poststratification and raking can be applied to calibrate estimates.
+Then the error is $16.6$\%.
+Using~\eqref{eq:error}, we can use the error to construct an estimate of the relative sampling rate:
+\[
+\rho D_M = \sqrt{\frac{f}{1-f}} \frac{\text{0.166}}{\sigma_Y} = 1.43 \times 10^{-2} \Rightarrow \Delta = 1.06 \times 10^{-3} \Rightarrow M = 2.29.
+\]
+The range for the CIs is (2.05, 2.54).
+We can also perform sensitivity as a function of the false negative and positive rates.
+Under these measurement error ranges, we saw that the impact on the effective reproductive rate estimator was much smaller.
+However, for $FP = 0.05$ and $FN = 0.005$ we have $M = 4.31$.
+While nothing above is definitive, it points to the interplay of selection bias and measurement error.
+This calculation gives us simple tools to understand how sensitivity and specificity impact our estimators and therefore how much we should trust conclusions based on these assumptions.
+Under the Santa Clara study estimates of these quantities, while we expect impact, the bias is much smaller than under other estimates.
+
+\paragraph*{Discussion}
+\label{section:discussion}
+
+There is nothing routine about COVID-19, including the corresponding statistical questions.
+The goal of this paper was to point out questionable statistical routines.
+Indeed, case-count trajectories are being reported daily by governments and broadcast by the media.
+The precision in the reported numbers gives the illusion of accuracy when what needs to be understood is the level of uncertainty.
+Selection bias not only means uncertainty, it also implies analysts feel certain about incorrect conclusions.
+We end this paper with a brief discussion of several important topics related to selection bias and measurement error.
+
+\paragraph*{Stratified sampling: improving precision in low-prevalence environments}
+
+This article argues why random sampling is a powerful tool and the limitations of data obtained through a self-selection process.
+As COVID-19 survey studies become more readily accessible, we encourage researchers to focus on building representative samples.
+One difficulty in the current crisis is the low prevalence relative to the population size in certain geographies.
+Take the recent Santa Clara study, for example, that estimated less than 2\% prevalence in the area.
+The key issue with low prevalence is that if the test's specificity is not lower than the prevalence, then the observed data is consistent with zero recovered individuals.
+
+In such settings, simple random sampling is likely to fail; however, there are potential solutions.
+For example, one can stratify the population into risk categories (based on population density, occupation, age, and other important factors) and perform Neyman allocation~\cite{Cochran77}.
+One could go attempt to go further and incorporate contact network information.
+However, such data is difficult to obtain in practice.
+Indeed, geographic location, occupation, age and basic demographic information are decent good proxies for latent network connectivity.
+
+\paragraph*{Effect modifiers in COVID-19 clinical trials}
 
@@
 The key concern is whether clinical trials on COVID-19 recruit from the pool of individuals who have tested positive for COVID-19 or whether they sample randomly from the population, test, and then recruit from this subset of tested individuals.
-To see this issue, suppose we have an outcome $Y$, a treatment $A$ and an \emph{unobserved} variable $U$.
-Suppose treatment is assigned at random, i.e., $A =1$ with probability $50\%$ and $A=0$ with probability $50\%$.
-Suppose the conditional mean of the outcome satisfies  $E(Y \mid U, A ) = \beta_0 + \beta_1 A + \beta_2 U + \beta_3 A U$.
-Typically, we are interested in the \emph{causal effect} of $A$ on $Y$.
-In counterfactual language,
-$$
-E( Y(A=1) - Y(A=0) ) = E_U ( E_A ( Y(A=1) - Y(A=0) \mid U=u ))
-= \beta_1 + \beta_3 E(U).
-$$
-Even in a randomized experiment, the question is what the correct value for $E(U)$ is.
-For COVID-19, the value of interest is the expected value of $U$ in the population of COVID-19 positive individuals.
-Due to selection bias, however, the bias in this marginal treatment effect compared to the true marginal effect of interest is $\beta_1 \cdot E_{\I} [ \rho_{I,U} ] \times \sqrt{ \frac{f}{1-f} } \times \sigma_U$.
-Selection bias then leads to the estimated effect being biased if $\beta_1 \neq 0$.
-Directionality of the bias will depend on the relationship between $U$ and $Y$.
-If $U$ is a measure of disease severity, then it could be expected to see $\beta_1 > 0$ and $E_{\I} [ \rho_{I, U} ] >0$.
-In such situations, the treatment effect estimates will tend to be overly optimistic as most individuals recruited are potentially higher on the severity index than the population average.
-Not only that, but for fixed data quality, the error compared to SRS in terms of effect estimation scales with the population size (i.e., the law of large populations).
-
 Randomization of treatment assignment in clinical trials negates unobserved confounders.
@@
 
-\section*{Cross-population comparisons}
+\paragraph*{Model-based solutions.}
 
-So far we have focused on understanding the limitations of using case-count data to understand population quantities of interest for a \emph{single} population (i.e., United States, Germany, France or Michigan, New York, London in isolation) at a single moment and over time.
-Many are interested in cross-population comparisons.
-Many pieces in the media as well as academic articles have plots of case-count over time aligned by time of first known case.
-Others have claimed such comparisons unfair due to unequal population sizes and instead plotted case-count per million.
-In this section, we focus on the statistical issues relevant to such comparisons.
-We do not make any comments on the multi-causal nature of success/failure in countries.
+One alternative approach to what is presented in this article is a model-based extension of SIR models to try and capture both selection bias and measurement error; however, without strong assumptions on the selection mechanism, the estimates are often not nonparametrically identifiable.
+When an issue ``cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically''\cite{CoxHink74}.
+In the absence of randomized experiments, the best route forward using any approach is the use of detailed sensitivity analyses and humility in conclusions drawn from such data sources.
+
+% \bibliography{scibib}
+\bibliographystyle{Science}
+\bibliography{covid-refs}
+
+\section*{Acknowledgments}
+Include acknowledgments of funding, any patents pending, where raw data for the paper are deposited, etc.
+
+%Here you should list the contents of your Supplementary Materials -- below is an example.
+%You should include a list of Supplementary figures, Tables, and any references that appear only in the SM.
+%Note that the reference numbering continues from the main text to the SM.
+% In the example below, Refs. 4-10 were cited only in the SM.
+\section*{Supplementary materials}
 
@@
 $$
-allows us to build Z-scores that account for the true sample size.
-While still not perfect, this allows us to build Z-scores that are not overly confident due to selection bias and measurement error.
-See Section~\ref{section:est_dq} for discussion of how to estimate data quality.
-
-\subsection*{Rate comparisons}
-
-An alternative is the comparison on the rates of case-count change over time.
-Here, for simplicity, we focus on comparing the estimated effective reproductive rate.
-We assume the two time-series are aligned so that $t=0$ is the time of first case.
-This negates alignment issues and is common in practice.
-
-While the above issues on $Z$-scores and effective sample size are important, here we highlight a separate issue.
-Namely, the interaction of biases in estimation of the two trajectories when the peak infection times differ slightly.
-We can write the difference in the rates as
-$$
-R_{t1} - R_{t2} = \frac{1}{\gamma} \log \left( \frac{1 + e_{1t}}{1+e_{2t}} \right)
-$$
-where $e_{jt}$ is the error associated given in Section \ref{section:r0-estimation}.
-Recall that the error allowed for over-estimation prior to the peak and then under-estimation post-peak.
-Here, these errors can mingle in interesting ways.
-Consider two countries (A and B) in which the peak occurs 2 weeks prior for country A then country B but other.
-Figure~4b presents such a comparison where each country's disease trajectory follows an SIR model as described in Section~\ref{section:rates} with different parameters.
-Figure~4 shows how the errors interact in complex ways.
-First, the gap is correctly estimated; then the gap is over-estimated as country A sees a rapid rise in cases; then the gap is even more over-estimated as country A improves and the rate is under-estimated while country B sees a rapid increase in their case-count and their rate is over-estimated; then country A's rate is correctly estimated while country B's rate is under-estimated as it sees improvement in infection rates; finally, the gap disappears.
-We are not claiming this will always be the case, and indeed for small enough selection bias and/or large gaps in case-counts such complex behaviour will not occur; however, this is simply to point out how observed information can tell a more complex story than the truth (i.e., country A recovers 1 month prior to country B).
-
-\subsection*{Estimation of Data Quality}
-\label{section:est_dq}
-
-In prior sections, we have discussed the impact of selection bias and measurement error on a range of outcomes of interest in COVID-19 research.
-Here, we try and address the empirical question of how to estimate the data quality $E_{\I} \left[ \rho_{I,Y} \right]$ in the current context.
-In \cite{Meng2018}, estimation of the data quality relied on observing the true outcome.
-In election data, for example, we observe the vote totals after the fact and can use this as ground truth.
-We are not so lucky in the current crisis.
-Here, we propose a simple procedure that uses survey samples as noisy but unbiased estimates of the true prevalence.
-Our goal is less inferential and more to provide a sensitivity analysis that can aid in understanding the amount of information in observational COVID case-count data.
-
-On April 23rd, Andrew Cuomo announced results from a study in New York state.
-It found 13.9\% of 3,000 people tested across the state had signs of the virus.
-The study did not report sensitivity and specificity; therefore, we take reported measures from the Santa Clara study~\cite{Bendavid2020} \footnote{We recognize the debate regarding its conclusions.
-Here we are only using this study for its information regarding measurement error rates.}.
-The reported specificity is $82.8\%$ (95\% CI: 76.0-88.4\%) and sensitivity is $99.5\%$ (95\% CI: 99.2-99.7\%).
-This corresponds to a false negative rate of $17.2\%$ (11.6-24.0\%) and false positive rate of $0.5\%$ (0.3 - 0.8\%).
-Correcting for these rates we have an estimated prevalence of $15.9\%$ with a range of 14.8\% to 17.0\% using the ranges of the false positive and negative rates.
-
-Figure~5 plots the number of new total and positive tests per day.
-Figure~5a plots the fraction of tests that were positiver per day.
-In both cases, we applied an exponential smoothing to the reported values because the spiky pattern is likely due to testing backlogs.
-Given the report was on the 23rd, we suppose for simplicity that the study was run several days prior on the 20th of April.
-On the 20th, there were $19,654$ tests performed and $617,555$ tests had been performed up to to and including the 19th.
-The population of New York state is $19,378,102$.
-Subtracting off the number of individuals who had already been tested yields a sampling fraction of $f = 0.001$ on the 20th.
-The smoothed prevalence estimate on the 20th is 28.3\%.
-Adjusting for measurement error yields an estimate of $32.5\%$.
-
-Here, we assume the New York survey is a simple random sample from the state's population.
-We recognize that the survey is not such a sample, but techniques like poststratification and raking can be applied to calibrate estimates.
-Then the error is $16.6$\%.
-Using~\eqref{eq:error}, we can use the error to construct an estimate of the relative sampling rate:
-\[
-\rho D_M = \sqrt{\frac{f}{1-f}} \frac{\text{0.166}}{\sigma_Y} = 1.43 \times 10^{-2} \Rightarrow \Delta = 1.06 \times 10^{-3} \Rightarrow M = 2.29.
-\]
-The range for the CIs is (2.05, 2.54).
-We can also perform sensitivity as a function of the false negative and positive rates.
-Under these measurement error ranges, we saw that the impact on the effective reproductive rate estimator was much smaller.
-However, for $FP = 0.05$ and $FN = 0.005$ we have $M = 4.31$.
-While nothing above is definitive, it points to the interplay of selection bias and measurement error.
-This calculation gives us simple tools to understand how sensitivity and specificity impact our estimators and therefore how much we should trust conclusions based on these assumptions.
-Under the Santa Clara study estimates of these quantities, while we expect impact, the bias is much smaller than under other estimates.
-
-\section*{Discussion}
-\label{section:discussion}
-
-There is nothing routine about COVID-19, including the corresponding statistical questions.
-The goal of this paper was to point out questionable statistical routines.
-Indeed, case-count trajectories are being reported daily by governments and broadcast by the media.
-The precision in the reported numbers gives the illusion of accuracy when what needs to be understood is the level of uncertainty.
-Selection bias not only means uncertainty, it also implies analysts feel certain about incorrect conclusions.
-We end this paper with a brief discussion of several important topics related to selection bias and measurement error.
-
-\subsection*{Stratified sampling: improving precision in low-prevalence environments}
-
-This article argues why random sampling is a powerful tool and the limitations of data obtained through a self-selection process.
-As COVID-19 survey studies become more readily accessible, we encourage researchers to focus on building representative samples.
-One difficulty in the current crisis is the low prevalence relative to the population size in certain geographies.
-Take the recent Santa Clara study, for example, that estimated less than 2\% prevalence in the area.
-The key issue with low prevalence is that if the test's specificity is not lower than the prevalence, then the observed data is consistent with zero recovered individuals.
-
-In such settings, simple random sampling is doomed to fail; however, there are potential solutions.
-In particular, one can stratify the population into risk categories (based on population density, occupation, age, and other important factors) and perform Neyman allocation~\cite{Cochran77}.
-That is, sample individuals at random in strata $h$ with size proportional to $P_h \sqrt{\bar Y_h \times (1-\bar Y_h)}$ where $P_h$ is the strata frequency in the population and $\bar Y_h$ the prevalence in the strata.
-Of course, $P_h$ and $\bar Y_h$ are not known a priori; however, if decent estimates can be obtained, we can see huge benefits in settings where the prevalence varies widely by strata.
-One could go attempt to go further and incorporate contact network information.
-However, such data is difficult to obtain in practice.
-Indeed, geographic location, occupation, age and basic demographic information are much easier to obtain and are a good proxy for the latent network connectivity.
-
-\subsection*{Decision-making: data versus information}
-
-Having read through the above technical discussions, one may come to the incorrect conclusion that because we are concerned about implicit biases in observational data, we must wait for better data sources in order to act~\cite{Ioannidis2020}.
-This conclusion confuses data and information.
-Lack of high quality data means we should be skeptical regarding conclusions drawn from these data sources alone.
-Being skeptical of observed data does not imply governments and communities should wait for high quality data to act; this is especially true in the current high risk scenario.
-Instead, we should aim to have high quality information.
-Yet in the current crisis we have just that.
-Indeed, epidemiologists and public health officials have a strong understanding of the basic facts of pandemics and disease spread.
-Social distancing, contact tracing and mass testing are all important.
-Individuals and communities with low quality data and only basic information will hopefully make decisions that respect inherent risks to their well being as well as the high degree of uncertainty in the observed data.
-Of course, there are trade-offs and long-term stay-at-home orders will have negative economic consequences.
-A full discussion of these topics is beyond the scope of the current article.
-
-\subsection*{High quality information on prevalence.}
-
-Governments are pushing for increased testing capacity and robust contact tracing.
-Contact tracing is incredibly useful for identifying carriers early and preventing spread of the disease.
-We would argue that, after the current wave, understanding prevalence is also key.
-We know that the probability of an outbreak is a function of current prevalence and network connectivity.
-Epidemic critical thresholds have been derived for many infectious disease models~\cite{Pastor2001,Newman2002,Parshani2010}.
-Knowing prevalence will help governments determine long-term community-level risk and allow for more targeted interventions -- shutting down only certain locations when necessary.
-
-\subsection*{Data quantity versus quality}
-
-Governments implicitly argue that increased testing capacity will perform the above task.
-Without complete compliance, however, our understanding of future outbreaks may be plagued by self-selection bias, compounded by changing sensitivity and specificity rates.
-Random testing removes these effect modifiers, giving governments more information to fight the disease.
-This is not to say random testing should replace current testing practices; only that the random testing should supplement and will provide valuable additional information even with moderate sample sizes.
-
-% Not only that, in current crisis data quality is often tied to data quantity due to rationing system.
-Increasing testing under test rationing lowers correlation.
-There is a lot of hidden information in the selection indicator $I_j$.
-Indeed, individuals must first seek testing and then call their primary health care provider must decide whether they met thresholds for testing.
-Our analysis shows the limitations and caveats that come from analysis of this data in terms of estimation of prevalence, rates of change, and cross-country comparisons.
-We show how one can build simple sensitivity analyses to understand the potential bias in case-count data using surveys that will become more prevalent in the next few months.
-
-\subsection*{Model-based solutions.}
-
-One alternative approach to what is presented in this article is a model-based extension of SIR models to try and capture both selection bias and measurement error; however, without strong assumptions on the selection mechanism, the estimates are often not nonparametrically identifiable.
-When an issue ``cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically''\cite{CoxHink74}.
-In the absence of randomized experiments, the best route forward using any approach is the use of detailed sensitivity analyses and humility in conclusions drawn from such data sources.
-
-% \bibliography{scibib}
-\bibliographystyle{Science}
-\bibliography{covid-refs}
-
-\section*{Acknowledgments}
-Include acknowledgments of funding, any patents pending, where raw data for the paper are deposited, etc.
-
-%Here you should list the contents of your Supplementary Materials -- below is an example.
-%You should include a list of Supplementary figures, Tables, and any references that appear only in the SM.
-%Note that the reference numbering continues from the main text to the SM.
-% In the example below, Refs. 4-10 were cited only in the SM.
-\section*{Supplementary materials}
+allows us to build Z-scores that account for the true sample size.  While still not perfect, this allows us to build Z-scores that are not overly confident due to selection bias and measurement error.
 

### 2020-05-22 write-up/scifile.tex
science paper updated

@@
 
+\usepackage{subcaption}
 \usepackage{scicite}
-
+\usepackage{graphicx}
 \usepackage{times}
@@
 
-\paragraph*{Introduction}
+\paragraph*{Introduction.}
 
@@
 Public databases~\cite{JHU_Lancet,NYT} provide incoming county-level information of confirmed cases and deaths.
-Statisticians, epidemiologists, economists, and data scientists have used this granular data to forecast COVID-19 case-counts, deaths, and hospitalizations~\cite{Song2020,Ray2020,2020.IHME}.
+Statisticians, epidemiologists, economists, and data scientists have used this granular data to forecast COVID-19 case-counts, deaths, and hospitalizations~\cite{Giordano2020,Song2020,Ray2020,2020.IHME,Wang2020.03,JTD36385}.
 
@@
 The reason is straightforward: observed case-count data is plagued by selection bias and measurement error.
-We will demonstrate that the information gained from increasing testing capacity is limited when selection bias and testing inaccuracies persist.
+We will demonstrate that the information gained from increasing testing capacity is limited in the presence of selection bias and when testing inaccuracies persist.
 
-\paragraph*{Analysis of imperfect case-count data}
+\paragraph*{Analysis of imperfect case-count data.}
 
@@
 At a fixed time, let $Y_j$ denote COVID-19 status for the $j$th individual in the population, $j=1,\ldots, N$.
-To start, for simplicity, we ignore the dynamic nature of the outbreak and the fact that individuals can recover.
+To start, the dynamic nature of the outbreak and recoverability of individuals are ignored.
 As in survey methodology~\cite{Cochran77}, we treat COVID-19 status as a fixed but unknown quantity of interest.
-Let $I_j \in \{0,1\}$ be an indicator that the individual was selected for testing ($I = 1$) or not ($I=0$).
-Suppose we are interested in COVID-19 prevalence, i.e., the population average $\bar Y = Y/N$, and that $n$ tests are performed and we observe the values $y_1, \ldots, y_n \in \{0,1\}$.
-The natural estimator is the proportion of positive tests $\bar y = n^{-1} \sum_{i=1}^n y_i$.
+Let $I_j \in \{0,1\}$ be an indicator that the individual was selected for testing.
+We are interested in COVID-19 prevalence, i.e., the population average $\bar Y = \sum_{j=1}^N Y_j/N$.
+Suppose $n$ tests are performed and the values $y_1, \ldots, y_n$ are observed.
 
@@
 Here we assume the false positive rate (FP) and false negative rate (FN) are known.
-Here we consider the estimator $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ which is unbiased under simple random sampling (SRS).
+Here we consider the estimator $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ where $\bar y = n^{-1} \sum_{i=1}^n y_i$ is the proportion of positive tests.
+This estimator is unbiased under simple random sampling (SRS).
 To understand the impact of selection bias and imperfect testing, we provide a statistical decomposition of the error between $\tilde y_n$ and $\bar Y$
@@
 The final term represents the \emph{imperfect testing adjustment} which is a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence.
-Figure~2 presents $D_M$ as a function of the relative frequency and odds ratio can be both positive and negative as well as a range of magnitudes~\cite{Beesley2020,Beesley2019,Smeden2019}.
+Figure~\ref{fig:heatmap} presents $D_M$ as a function of the relative frequency and odds ratio can be both positive and negative as well as a range of magnitudes~\cite{Beesley2020,Beesley2019,Smeden2019}.
 
@@
 $$
-\frac{E_{\I} \left[ (\bar y_n - \bar Y)^2 \right]}{\sqrt{V_{SRS} (\bar Y)}} = (N-1) E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right].
+\frac{E_{\I} \left[ (\bar y_n - \bar Y)^2 \right]}{\sqrt{V_{SRS} (\bar Y)}} = (N-1) E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right]
 $$
-This points to a paradoxical situation: the error relative to SRS increases as a function of population size.
-This was termed the ``Law of Large Populations'' (LLP)~\cite{Meng2018} and continues to hold in our imperfect testing context.
-This points to a critical issue in current media practices in communication of case-count data: two countries with the same testing strategy can yield very different estimates due to population size.
-When comparing countries like the US to the UK, the MSE increasea by a factor of almost five.
-Thus conclusions drawn from observed case-count records may be not just wrong, but very wrong.
+implying the error relative to SRS increases as a function of population size.
+This paradox is too often ignored in current communications of case-count data.
+When comparing the US and UK, for example, the MSE increases by a factor of five.
+Thus conclusions drawn from observed case-counts may be not just wrong, but very wrong.
 
-An important follow-up question is ``What is the sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' We show this \emph{effective} sample size can be bounded by
+A key question is ``What is the (effective) sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' We show this can be bounded by
 $$
-n_{eff} \le \frac{f}{1-f} \times \frac{1}{E_{I} \left[ \rho_{I,Y}^2 D_M^2 \right]}.
+\frac{f}{1-f} \times \frac{1}{E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right]}.
 $$
@@
 A recent study estimates a false positive and negative rate for antibody testing using a lateral flow immunoassay are $0.5$\% and $17.2$\% respectively \cite{Bendavid2020}.
-Supposing COVID-19 positive individuals are $2$ times more likely to get tested, then using these parameters the effective sample size is $10$.
+Supposing COVID-19 positive individuals are $2$ times more likely to get tested, then the effective sample size is $10$.
 Recent proposals~\cite{Siddarth2020} have argued for increased testing capacity, which will reduce the relative sampling rate.
 Supposing it drops to $1.2$ then the effective sample size will increase to $216$.
-This points to a sad state of affairs, saying that the effective sample size even in optimistic scenarios is not better than a small random sample from the population.
+Thus the effective sample size even in optimistic scenarios is equivalent to a small random sample from the population.
 
@@
 
-Another important question is whether selection bias and imperfect testing impact the ability to estimate rates of growth in case-count.
-Let $\bar Y_{t-1}$ and $\bar Y_{t}$ denote the prevalence on two consecutive times $t-1$ and $t$.
-We consider the ratio estimator $r = \bar y_t / \bar y_{t-1}$.
-Using a Taylor series approximation, the error between ${\bar y_t}/{\bar y_{t-1}}$ and the true rate ${\bar Y_{t}}/{\bar Y_{t-1}}$ can be expressed approximately as
+Another important question is whether selection bias and imperfect testing impact the ability to estimate rates of change in prevalence.
+Let $\bar Y_{t-1}$ and $\bar Y_{t}$ denote the prevalence on two consecutive days and consider the estimator $r = \bar y_t / \bar y_{t-1}$.
+Using a Taylor series approximation, the error between ${\bar y_t}/{\bar y_{t-1}}$ and ${\bar Y_{t}}/{\bar Y_{t-1}}$ can be expressed approximately as
 $$
@@
 $$
-where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement error adjustment, and $CV(Y_j) = \sigma_{Y_j}/\bar Y_j$ is the coefficient of variation at time step $j$.
+where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement error adjustment, and $CV(Y_j) = \sigma_{Y_j}/\bar Y_j$ is the coefficient of variation on day $j$.
 The error magnitude depends on the true rate $\bar Y_{t} / \bar Y_{t-1}$ so a large decrease will have a relatively small error compared to a large increase.
@@
 
-Consider the state of New York.
-Figure~5 shows the number of tests and positive tests per day in New York state.
-Figure~5b presents the fraction of tests that were positive per day in New York state.
+% Consider, for example, the state of New York.
+Figure~\ref{fig:ny-covid-test} shows the number of tests and positive tests per day in New York state.
+Figure~\ref{fig:ny-covid-frac} presents the fraction of tests that were positive per day in New York state.
 For cancellation, the expected data qualities need to be uncorrelated and equal.
 Testing per day suggests it is plausible to assume $\Delta_t$ varies slowly over time.
-Note, however, that constant $\Delta_t$ does not imply constant data quality.
-Figure~5a shows the number of tests has remained approximately constant in the last few weeks relative to the total NY population.
-Therefore, the assumption that $f_t$ also varies slowly in time is reasonable.
+Figure~\ref{fig:ny-covid-test} shows the number of tests has remained approximately constant recently relative to the NY population size.
+Therefore, the assumption that $f_t$ varies slowly in time is reasonable.
 Finally, the assumption that the imperfect testing adjustment is constant is reasonable as no new testing procedures have been reported.
-This would not be true in states like Michigan, where testing criteria have been expanding every few weeks.
 
-Figure~3a shows a trajectory of the true ratio and the potential biased estimators under an SIR model~\cite{Pastor2001,Newman2002,Parshani2010} for the epidemic dynamics, with state evolution given by the following ODE
+Figure~\ref{fig:ratio-bias} displays the trajectory of the true ratio and the potential biased estimators under an SIR model~\cite{Pastor2001,Newman2002,Parshani2010} for the epidemic dynamics, with state evolution given by
 $$
@@
 $$
-where $S_t$ is the number of susceptible individuals at time $t$, $I_t$ is the number of infected individuals at time $t$, $N$ is the total population size, and $R_t$ is the number of removed (recovered or deceased) individuals at time $t$.
-Figure~4a presents the fraction of population newly infected individuals at each time step for $\beta = 1.4$ and $\gamma = 0.2$ (black line).
-In terms of bias, the rate is likely to be overestimated prior to the peak and then underestimated afterwards.
+where $S_t$ is the number of susceptible individuals, $I_t$ is the number of infected individuals, and $R_t$ is the number of removed (recovered or deceased) individuals at time $t$.
+In terms of bias, the rate is overestimated prior to the peak and underestimated afterwards.
 From a decision making perspective, such biases have potential impact.
-First, the over-estimation may give lawmakers and governors more leverage in proposing aggressive actions that reduce prevalence.
-Of course, the analysis supports the argument that the estimates based on available data are overly pessimistic.
-What both sides miss is that the direction of bias is non-constant over time.
-After the peak, the estimator likely under-estimates the rate of change.
-Such bias puts pressure on lawmakers and governors to prematurely relax social distancing measures.
-Moreover, it appears that the peak time is the easiest to estimate in terms of having minimal error using availability data.
-Note, however, that the standard errors on such estimates will be wide due to relatively low effective sample size.
+First, the overestimation may give lawmakers and governors more leverage in proposing aggressive actions that reduce prevalence.
+Of course, the analysis supports the argument that the estimates based on available data may be pessimistic.
+What is missed is that the direction of bias is non-constant over time.
+Underestimation post-peak puts pressure on lawmakers and governors to prematurely relax social distancing measures.
+Moreover, it appears that the peak time is the easiest to estimate in terms of having minimal error using available data.
+Note, however, that the standard errors on such estimates will be wide due to effective sample sizes.
 
@@
 $$
-This demonstrates the exact same trade-off as with the rates but on the logarithmic scale.
-For small values of $x$, we have $\log(1+x) = x$ and so the biases are similar.
-However, the error is no longer scaled by the rate of change in prevalence $\bar Y_2/\bar Y_1$.
-Instead the error is scaled by the serial interval $\gamma^{-1} = 1/7$.
-Moreover, the error depends not on aggregate prevalence $\bar Y_t$ but fraction of new cases $\bar K_t$.
-The final term is the error due to varying population sizes.
-Since $S_t \leq S_{t-1}$, the final term $- \log (S_t/S_{t-1}) \geq 0$.
-Figure~4b displays the bias for the effective reproductive number and how it is impacted by the relative sampling fraction.
-Here for simplicity, we assume the fraction of the population that is susceptible remains large and near the total population size.
-Changes to this assumption can impact the potential bias.
-The key difference between Figure~4a and Figure~4b is that prevalence depends on the fraction of the population infected at time $t$ while $\hat R_t$ depends on the fraction of new cases in the population.
+This implies a similar trade-off as before but on the logarithmic scale.
+The error is no longer scaled by $\bar Y_t/\bar Y_{t-1}$ but by the serial interval and does not depend on prevalence $\bar Y_t$ but on the fraction of new cases out of those susceptible $\bar K_t$.
 This leads to differences in when the bias is most pronounced.
+The final term is the error due to varying numbers of susceptible individuals.
+Figure~\ref{fig:r0-bias} displays the bias as a function of the relative sampling fraction assuming the fraction of the population that is susceptible remains large.
 
@@
 Many are interested in cross-population comparisons.
-Many pieces in the media as well as academic articles have plots of case-count over time aligned by time of first known case.
-Others have claimed such comparisons unfair due to unequal population sizes and instead plotted case-count per million.
 Here, for simplicity, we focus on comparing the estimated effective reproductive rate.
-We assume the two time-series are aligned so that $t=0$ is the time of first case.
+See the supplementary materials for discussion of prevalence comparisons.
+We assume the time-series are aligned so that $t=0$ is the time of first case in each population.
 
-We focus on the interaction of biases in estimation of the two trajectories when the peak infection times differ slightly.
-Recall that the error allowed for over-estimation prior to the peak and then under-estimation post-peak.
-Here, these errors can mingle in interesting ways.
-Consider two countries (A and B) in which the peak occurs 2 weeks prior for country A then country B but other.
-Figure~4b presents such a comparison where each country's disease trajectory follows an SIR model as described above with different parameters.
-Figure~4 shows how the errors interact in complex ways.
-First, the gap is correctly estimated; then the gap is over-estimated as country A sees a rapid rise in cases; then the gap is even more over-estimated as country A improves and the rate is under-estimated while country B sees a rapid increase in their case-count and their rate is over-estimated; then country A's rate is correctly estimated while country B's rate is under-estimated as it sees improvement in infection rates; finally, the gap disappears.
-We are not claiming this will always be the case, and indeed for small enough selection bias and/or large gaps in case-counts such complex behaviour will not occur; however, this is simply to point out how observed information can tell a more complex story than the truth (i.e., country A recovers 1 month prior to country B).
+Consider two countries (A and B) in which the peak occurs 2 weeks prior for country A than country B.
+Figure~\ref{fig:comparison} presents such a comparison where each country's disease trajectory follows an SIR model (A=black and B=red).
+Of interest is the difference between the effective reproduction rate for country A and B, $R_{tA} - R_{tB}$.
+Figure~\ref{fig:eff} shows how biases in estimation of each rate interact in complex ways.
+At first, the difference  is correctly estimated; then the gap is over-estimated as country A sees a rapid rise in cases; then the magnitude of over-estimation increases as country A sees declining case-count while country B sees rapidly increasing case-count; then the gap country A's rate is correctly estimated while country B's rate is under-estimated as it sees declining case-count; finally, the gap disappears.
+We are not claiming this will always be the case, and indeed for small enough selection bias and/or large gaps in case-counts such complex behaviour will not occur; however, this is demonstrats how observed information can tell a more complex story than the truth (i.e., country A's peak is 2 weeks prior to country B's peak).
 
-\paragraph*{Estimation of Data Quality.} Here, we try and address the empirical question of how to estimate data quality $E_{\I} \left[ \rho_{I,Y} \right]$ in the current context.
-In \cite{Meng2018}, estimation relied on observing the true outcome.
-In election data, for example, this is possible as election vote totals can be used as ground truth.
+\paragraph*{Estimation of Data Quality.} We next consider estimation of data quality $E_{\I} \left[ \rho_{I,Y} \right]$ in the current context.
+In \cite{Meng2018}, estimation relied on observing the true outcome (i.e., election vote totals).
 This is not possible in the current crisis.
-Here, we propose a simple procedure that uses survey samples as noisy but unbiased estimates of the true prevalence.
-Our goal is less inferential and more to provide a sensitivity analysis that can aid in understanding the amount of information in observational COVID case-count data.
+Here, we propose use of survey samples as noisy but unbiased estimates of the true prevalence.
+Our goal is less inferential.
+Instead, the goal is to build sensitivity analyses that can aid in understanding the amount of information in observational COVID case-count data.
 
@@
 This corresponds to a false negative rate of $17.2\%$ (11.6-24.0\%) and false positive rate of $0.5\%$ (0.3 - 0.8\%).
-Correcting for these rates we have an estimated prevalence of $15.9\%$ with a range of 14.8\% to 17.0\% using the ranges of the false positive and negative rates.
+This leads to an estimated prevalence of $15.9\%$ with a range of 14.8\% to 17.0\%.
+For simplicity, we assume the study was conducted on April 20th.
+On the 20th,  $19,654$ tests were performed.
+The population of New York is $19,378,102$.
+Subtracting off the $617,555$ tests that had already performed yields a sampling fraction of $f = 0.001$.
+The smoothed prevalence estimate is 28.3\%.
+Adjusting for imperfect testing yields an estimate of $32.5\%$.
 
-Given the report was on the 23rd, we suppose for simplicity that the study was run several days prior on the 20th of April.
-On the 20th, there were $19,654$ tests performed and $617,555$ tests had been performed up to to and including the 19th.
-The population of New York state is $19,378,102$.
-Subtracting off the number of individuals who had already been tested yields a sampling fraction of $f = 0.001$ on the 20th.
-The smoothed prevalence estimate on the 20th is 28.3\%.
-Adjusting for measurement error yields an estimate of $32.5\%$.
-
-Here, we assume the New York survey is a simple random sample from the state's population.
-We recognize that the survey is not such a sample, but techniques like poststratification and raking can be applied to calibrate estimates.
-Then the error is $16.6$\%.
-Using~\eqref{eq:error}, we can use the error to construct an estimate of the relative sampling rate:
+Under the assumption that the survey is a SRS from New York, the error is $16.6$\%.  Using~\eqref{eq:error}, we construct an estimate of the relative sampling rate as follows
 \[
@@
 \]
-The range for the CIs is (2.05, 2.54).
-We can also perform sensitivity as a function of the false negative and positive rates.
-Under these measurement error ranges, we saw that the impact on the effective reproductive rate estimator was much smaller.
-However, for $FP = 0.05$ and $FN = 0.005$ we have $M = 4.31$.
-While nothing above is definitive, it points to the interplay of selection bias and measurement error.
-This calculation gives us simple tools to understand how sensitivity and specificity impact our estimators and therefore how much we should trust conclusions based on these assumptions.
-Under the Santa Clara study estimates of these quantities, while we expect impact, the bias is much smaller than under other estimates.
+We can perform a sensitivity analysis by considering the range of false negative and positive rates from~\cite{Bendavid2020}, which leads to a range for $M$ of $(2.05, 2.54)$.
+Therefore, the impact on bias of $\hat R_t$ appears moderate.
+However, for $FP = 0.05$ and $FN = 0.005$, the relative sampling rate is $M=4.31$.
+This simple tool helps us understand how sensitivity and specificity impact biases and  therefore how much we should trust conclusions based on these assumptions.
 
-\paragraph*{Discussion}
-\label{section:discussion}
-
+\paragraph*{Discussion.}
 There is nothing routine about COVID-19, including the corresponding statistical questions.
 The goal of this paper was to point out questionable statistical routines.
-Indeed, case-count trajectories are being reported daily by governments and broadcast by the media.
-The precision in the reported numbers gives the illusion of accuracy when what needs to be understood is the level of uncertainty.
-Selection bias not only means uncertainty, it also implies analysts feel certain about incorrect conclusions.
-We end this paper with a brief discussion of several important topics related to selection bias and measurement error.
+The precision in reported case-count data gives the illusion of information when what what is needed is quantification of uncertainty.
+Selection bias implies analysts feel certain about incorrect conclusions.
+We end with a brief discussion of important related topics.
 
-\paragraph*{Stratified sampling: improving precision in low-prevalence environments}
+\paragraph*{Stratified sampling: improving precision in low-prevalence environments.}
+Random sampling is a powerful tool.
+One difficulty with SRS is when prevalence is low.
+Consider the Santa Clara study~\cite{Bendavid2020} which estimated prevalence less than 2\% in the county.
+The key issue with low prevalence is that if the test's specificity is higher than the prevalence, then the observed data is consistent with zero recovered individuals.
+In these settings, SRS is likely to fail; however, there are potential solutions.
+One can, for example, stratify the population into risk categories (e.g., based on population density, occupation, age) and perform Neyman allocation~\cite{Cochran77}.
 
-This article argues why random sampling is a powerful tool and the limitations of data obtained through a self-selection process.
-As COVID-19 survey studies become more readily accessible, we encourage researchers to focus on building representative samples.
-One difficulty in the current crisis is the low prevalence relative to the population size in certain geographies.
-Take the recent Santa Clara study, for example, that estimated less than 2\% prevalence in the area.
-The key issue with low prevalence is that if the test's specificity is not lower than the prevalence, then the observed data is consistent with zero recovered individuals.
+\paragraph*{Effect modifiers in COVID-19 clinical trials.}
 
-In such settings, simple random sampling is likely to fail; however, there are potential solutions.
-For example, one can stratify the population into risk categories (based on population density, occupation, age, and other important factors) and perform Neyman allocation~\cite{Cochran77}.
-One could go attempt to go further and incorporate contact network information.
-However, such data is difficult to obtain in practice.
-Indeed, geographic location, occupation, age and basic demographic information are decent good proxies for latent network connectivity.
-
-\paragraph*{Effect modifiers in COVID-19 clinical trials}
-
-Here we show how selection bias and measurement error can creep into clinical trial analysis.
-The key concern is whether clinical trials on COVID-19 recruit from the pool of individuals who have tested positive for COVID-19 or whether they sample randomly from the population, test, and then recruit from this subset of tested individuals.
+Clinical trials that recruit individuals who have tested positive for COVID-19 will have analysis impacted by selection bias and imperfect testing.
 Randomization of treatment assignment in clinical trials negates unobserved confounders.
-This, however, does not negate effect modifiers.
-Therefore, for marginal treatment effects to be interpretable, there must be a well-defined population.
-Most often, our main interest is in causal effects on the population of COVID-19 positive.
-Randomization within the clinical trial design yields internal validity, but we also need external validity to estimate the correct marginal effect of interest~\cite{Keiding2016}.
+This does not negate effect modifiers.
+Randomization within the clinical trial design yields internal validity, but external validity is necessary to estimate the correct marginal effect of interest~\cite{Keiding2016}.
+See the supplementary materials for additional discussion.
 
@@
 
-One alternative approach to what is presented in this article is a model-based extension of SIR models to try and capture both selection bias and measurement error; however, without strong assumptions on the selection mechanism, the estimates are often not nonparametrically identifiable.
+One can extend the SIR model to attempt an account for selection bias and measurement error; however, without strong assumptions on the selection mechanism, the estimates are often not identifiable.
 When an issue ``cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically''\cite{CoxHink74}.
-In the absence of randomized experiments, the best route forward using any approach is the use of detailed sensitivity analyses and humility in conclusions drawn from such data sources.
+In the absence of random sampling, the best route forward regardless of methodology detailed is the use of sensitivity analyses and humility in data-driven conclusions.
 
@@
 \section*{Acknowledgments}
-Include acknowledgments of funding, any patents pending, where raw data for the paper are deposited, etc.
+We thank covidtracking.com for providing open-access to state case-count data.
 
@@
 allows us to build Z-scores that account for the true sample size.  While still not perfect, this allows us to build Z-scores that are not overly confident due to selection bias and measurement error.
+
+\subsection*{Effect modifiers in COVID-19 clinical trials}
+
+Here we show how selection bias and measurement error can creep into clinical trial analysis.
+The key concern is whether clinical trials on COVID-19 recruit from the pool of individuals who have tested positive for COVID-19 or whether they sample randomly from the population, test, and then recruit from this subset of tested individuals.
+To see this issue, suppose we have an outcome $Y$, a treatment $A$ and an \emph{unobserved} variable $U$.
+Suppose treatment is assigned at random, i.e., $A =1$ with probability $50\%$ and $A=0$ with probability $50\%$.
+Suppose the conditional mean of the outcome satisfies  $E(Y \mid U, A ) = \beta_0 + \beta_1 A + \beta_2 U + \beta_3 A U$.
+Typically, we are interested in the \emph{causal effect} of $A$ on $Y$.
+In counterfactual language,
+$$
+E( Y(A=1) - Y(A=0) ) = E_U ( E_A ( Y(A=1) - Y(A=0) \mid U=u ))
+= \beta_1 + \beta_3 E(U).
+$$
+Even in a randomized experiment, the question is what the correct value for $E(U)$ is.
+For COVID-19, the value of interest is the expected value of $U$ in the population of COVID-19 positive individuals.
+Due to selection bias, however, the bias in this marginal treatment effect compared to the true marginal effect of interest is $\beta_1 \cdot E_{\I} [ \rho_{I,U} ] \times \sqrt{ \frac{f}{1-f} } \times \sigma_U$.
+Selection bias then leads to the estimated effect being biased if $\beta_1 \neq 0$.
+Directionality of the bias will depend on the relationship between $U$ and $Y$.
+If $U$ is a measure of disease severity, then it could be expected to see $\beta_1 > 0$ and $E_{\I} [ \rho_{I, U} ] >0$.
+In such situations, the treatment effect estimates will tend to be overly optimistic as most individuals recruited are potentially higher on the severity index than the population average.
+Not only that, but for fixed data quality, the error compared to SRS in terms of effect estimation scales with the population size (i.e., the law of large populations).
+
+Randomization of treatment assignment in clinical trials negates unobserved confounders.
+This, however, does not negate effect modifiers.
+Therefore, for marginal treatment effects to be interpretable, there must be a well-defined population.
+Most often, our main interest is in causal effects on the population of COVID-19 positive.
+Randomization within the clinical trial design yields internal validity, but we also need external validity to estimate the correct marginal effect of interest~\cite{Keiding2016}.
 
@@
 
-Materials and Methods\\
-Supplementary Text\\
+% Materials and Methods\\
+% Supplementary Text\\
 
-Figs. S1 to S3\\
-Tables S1 to S4\\
-References \textit{(4-10)}
+% Figs. S1 to S3\\
+% Tables S1 to S4\\
+% References \textit{(4-10)}
 
@@
 
-\noindent {\bf Fig. 1.} US total tests and total positive tests per day\\
-\noindent {\bf Fig. 2.} US total tests and total positive tests per day\\
-\noindent {\bf Fig. 3.} US total tests and total positive tests per day\\
-\noindent {\bf Fig. 4.} US total tests and total positive tests per day\\
-\noindent {\bf Fig. 5.} US total tests and total positive tests per day
+\begin{figure}[!th]
+\centering
+\includegraphics[width = 0.6\textwidth]{../methods/figs/US-casecount.png}
+\caption{US total tests and total positive tests per day}
+\label{fig:ustests}
+\end{figure}
+
+\begin{figure}[!th]
+\centering
+\includegraphics[width = 0.6\textwidth]{../methods/figs/mem_heatmap_article.png}
+\caption{Measurement error data quality adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.005$ and $FN=0.172$.
+Color scaled so blue = $-10$, white = $0$, and red = $10$.}
+\label{fig:heatmap}
+\end{figure}
+
+\begin{figure}[!th]
+\centering
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{../methods/figs/sir_ratio.png}
+  \caption{Ratio estimator}
+  \label{fig:ratio-bias}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{../methods/figs/sir_rt.png}
+  \caption{Effective reproductive rate estimator}
+  \label{fig:r0-bias}
+\end{subfigure}
+\caption{Potential bias due to self-selection and measurement error from a deterministic SIR model with $\beta = 1.4$ (black) and $\gamma = 0.2$ (red).
+Here we assume $f = 0.02$, $FP = 0.01$, $FN = 0.15$, and use a range of relative sampling fraction $M = f_1/f_0$.}
+\label{fig:rates}
+\end{figure}
+
+\begin{figure}[!th]
+\centering
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{../methods/figs/sir.png}
+  \caption{Fraction of new cases in population}
+  \label{fig:fracpop}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{../methods/figs/sir_rt_comparison.png}
+  \caption{Effective reproduction rate estimators}
+  \label{fig:eff}
+\end{subfigure}
+\caption{Left: fraction infected in two SIR models with $\beta = 1.4$ and $0.9$ respectively and $\gamma = 0.2$ with same initial conditions.
+Right: comparison of $\hat R_t$ across time with $FN = 0.2$, $FP = 0.01$, and $M = 4$.}
+\label{fig:comparison}
+\end{figure}
+
+\begin{figure}[!th]
+\centering
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{../methods/figs/NY_casecount.png}
+  \caption{NY new cases per day}
+  \label{fig:ny-covid-test}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{../methods/figs/NY_fracpos.png}
+  \caption{NY fraction of new tests positive per day}
+  \label{fig:ny-covid-frac}
+\end{subfigure}
+\caption{New York COVID-19 reported data }
+\label{fig:ny-covid}
+\end{figure}
+
+% \noindent {\bf Fig. 1.} US total tests and total positive tests per day\\
+% \noindent {\bf Fig. 2.} US total tests and total positive tests per day\\
+% \noindent {\bf Fig. 3.} US total tests and total positive tests per day\\
+% \noindent {\bf Fig. 4.} US total tests and total positive tests per day\\
+% \noindent {\bf Fig. 5.} US total tests and total positive tests per day
 

### 2020-05-25 write-up/scifile.tex
deleted a good paragraph :(
tex updates

@@
 Through simple calculations, we demonstrate pitfalls and paradoxes that can arise when considering case-count data in the presence of selection bias and measurement error.
-The discussion guides several suggestions on how to improve current case-count reporting.
 \end{sciabstract}
@@
 COVID-19 testing is no exception which implies a selected individual's true status is not observed.
-Here we assume the false positive rate (FP) and false negative rate (FN) are known.
-Here we consider the estimator $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ where $\bar y = n^{-1} \sum_{i=1}^n y_i$ is the proportion of positive tests.
+Here we assume the false positive rate (FP) and false negative rate (FN) are known and consider the estimator $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ where $\bar y = n^{-1} \sum_{i=1}^n y_i$ is the proportion of positive tests.
 This estimator is unbiased under simple random sampling (SRS).
-To understand the impact of selection bias and imperfect testing, we provide a statistical decomposition of the error between $\tilde y_n$ and $\bar Y$
+To understand the impact of selection bias and imperfect testing, we provide a statistical decomposition of the error between $\tilde y_n$ and $\bar Y$ which is given by
 \begin{equation}
@@
 \end{equation}
-where $f_1 := \pr (I_J = 1 \mid Y_J = 1)$ and $f_0 := \pr(I_J = 1 \mid Y_J = 0)$ are sampling rates for COVID-19 positive and negative individuals respectively, and $\Delta = f_1 - f_0$ is the sampling rate differential.
+where $f=n/N$ is the sampling fraction, $f_1$ and $f_0$ are sampling fractions for COVID-19 positive and negative individuals respectively, and $\Delta = f_1 - f_0$ is the sampling rate differential.
 This extends work by \cite{Meng2018} to account for imperfect testing.
@@
 The final term represents the \emph{imperfect testing adjustment} which is a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence.
-Figure~\ref{fig:heatmap} presents $D_M$ as a function of the relative frequency and odds ratio can be both positive and negative as well as a range of magnitudes~\cite{Beesley2020,Beesley2019,Smeden2019}.
+Figure~\ref{fig:heatmap} shows that $D_M$, as a function of the relative frequency ($f_1/f_0$) and odds ratio, can be both positive and negative as well as a range of magnitudes~\cite{Beesley2020,Beesley2019,Smeden2019}.
 
@@
 implying the error relative to SRS increases as a function of population size.
-This paradox is too often ignored in current communications of case-count data.
+The paradox that two countries with the same testing strategy can yield wildly different estimates due to population size is too often ignored in current communications of case-count data.
 When comparing the US and UK, for example, the MSE increases by a factor of five.
@@
 
-A key question is ``What is the (effective) sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' We show this can be bounded by
+A key question is ``What is the (effective) sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' This quantity can be bounded by
 $$
@@
 In terms of effective sample size, if the data quality term remains constant, then testing increases may not significantly improve the effective sample size.
-Consider the simplest case where $f$ increases from $0.05$ to $0.1$; suppose this increase was associated with  false positive rate increasing from $0.5$\% to $5$\% and the false negative rate rising from $5\%$ to $20$\%.
+Consider the case where $f$ increases from $0.05$ to $0.1$ and the increase is associated with the false positive rate increasing from $0.5$\% to $5$\% and the false negative rate rising from $5\%$ to $20$\%.
 Then the standard analysis would suggest a reduction in the MSE by 47\% but the actual reduction is 26\%.
-This implies that doubling the sampling rate will only increase the effective sample size by a factor $2.9$ rather than the expected $4$ fold increase that is expected.
+This implies that doubling the sampling fraction will only increase the effective sample size by a factor $2.9$ rather than the expected $4$ fold increase that is expected.
 
@@
 $$
-where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement error adjustment, and $CV(Y_j) = \sigma_{Y_j}/\bar Y_j$ is the coefficient of variation on day $j$.
+where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement error adjustment, and $CV(Y_j) = \sigma_{Y_j}/Y_j$ is the coefficient of variation on day $j$.
 The error magnitude depends on the true rate $\bar Y_{t} / \bar Y_{t-1}$ so a large decrease will have a relatively small error compared to a large increase.
-The second term represents potential \emph{cancellation} which may occurs when data quality, sampling fraction, measurement error, and prevalence are all constant across time.
-
-% Consider, for example, the state of New York.
-Figure~\ref{fig:ny-covid-test} shows the number of tests and positive tests per day in New York state.
-Figure~\ref{fig:ny-covid-frac} presents the fraction of tests that were positive per day in New York state.
-For cancellation, the expected data qualities need to be uncorrelated and equal.
-Testing per day suggests it is plausible to assume $\Delta_t$ varies slowly over time.
-Figure~\ref{fig:ny-covid-test} shows the number of tests has remained approximately constant recently relative to the NY population size.
-Therefore, the assumption that $f_t$ varies slowly in time is reasonable.
-Finally, the assumption that the imperfect testing adjustment is constant is reasonable as no new testing procedures have been reported.
+The second term represents potential \emph{cancellation} which may occur when data quality, sampling fraction, measurement error, and prevalence are all constant across time.
 
@@
 Moreover, it appears that the peak time is the easiest to estimate in terms of having minimal error using available data.
-Note, however, that the standard errors on such estimates will be wide due to effective sample sizes.
+Note, however, that the standard errors on such estimates will be wide due to small   effective sample sizes.
 
@@
 \centering
-\includegraphics[width = 0.6\textwidth]{../methods/figs/US-casecount.png}
-\caption{US total tests and total positive tests per day}
-\label{fig:ustests}
-\end{figure}
-
-\begin{figure}[!th]
-\centering
 \includegraphics[width = 0.6\textwidth]{../methods/figs/mem_heatmap_article.png}
@@
 \end{figure}
+
+% \begin{figure}[!th]
+% \centering
+% \includegraphics[width = 0.6\textwidth]{../methods/figs/US-casecount.png}
+% \caption{US total tests and total positive tests per day}
+% \label{fig:ustests}
+% \end{figure}
 

### 2020-06-03 write-up/scifile.tex
Update scifile.tex

@@
 The World Health Organization has declared the coronavirus disease 2019 (COVID-19) a public health emergency.
-As of May 21st, 2020, over 5 million cases have been confirmed worldwide with 1.5 million cases and over 90 thousand confirmed deaths across the United States.
+As of June 2nd, 2020, over 6 million cases have been confirmed worldwide with 1.8 million cases and over 100 thousand confirmed deaths across the United States.
 This pandemic
@@
 The reason is straightforward: observed case-count data is plagued by selection bias and measurement error.
-We will demonstrate that the information gained from increasing testing capacity is limited in the presence of selection bias and when testing inaccuracies persist.
+Through simple calculations, we will demonstrate that the information gained from increasing testing capacity is limited in the presence of selection bias and when testing inaccuracies persist.
 
@@
 
+We start by considering estimation of COVID-19 prevalence using imperfect case-count data.
 Let $N$ denote the population size.
@@
 Let $I_j \in \{0,1\}$ be an indicator that the individual was selected for testing.
-We are interested in COVID-19 prevalence, i.e., the population average $\bar Y = \sum_{j=1}^N Y_j/N$.
-Suppose $n$ tests are performed and the values $y_1, \ldots, y_n$ are observed.
+COVID-19 prevalence is equal to the population average $\bar Y = \sum_{j=1}^N Y_j/N = N_1/N$ where $N_1$ is the number of COVID-19 positive individuals in the population.
+Suppose $n$ tests are performed and the values $y_1, \ldots, y_n$ are observed with $\sum_{i=1}^n y_i = n_1$.
 
 Tests are imperfect.
-COVID-19 testing is no exception which implies a selected individual's true status is not observed.
+COVID-19 testing is no exception.
+Imperfect testing means a selected individual's observed COVID-19 status may not be the same as their true status.
 Here we assume the false positive rate (FP) and false negative rate (FN) are known and consider the estimator $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ where $\bar y = n^{-1} \sum_{i=1}^n y_i$ is the proportion of positive tests.
 This estimator is unbiased under simple random sampling (SRS).
-To understand the impact of selection bias and imperfect testing, we provide a statistical decomposition of the error between $\tilde y_n$ and $\bar Y$ which is given by
+To understand the impact of selection bias and imperfect testing, we derive a statistical decomposition of the error between $\tilde y_n$ and $\bar Y$ which is given by
 \begin{equation}
@@
 \end{equation}
-where $f=n/N$ is the sampling fraction, $f_1$ and $f_0$ are sampling fractions for COVID-19 positive and negative individuals respectively, and $\Delta = f_1 - f_0$ is the sampling rate differential.
+where $f=n/N$ is the sampling fraction, $f_1=n_1/N_1$ and $f_0=n_0/N_0=(n-n_1)/(N-N_1)$ are sampling fractions for COVID-19 positive and negative individuals respectively, and $\Delta = f_1 - f_0$ is the sampling rate differential.
+See the supplementary materials for the derivation.
 This extends work by \cite{Meng2018} to account for imperfect testing.
@@
 Figure~\ref{fig:heatmap} shows that $D_M$, as a function of the relative frequency ($f_1/f_0$) and odds ratio, can be both positive and negative as well as a range of magnitudes~\cite{Beesley2020,Beesley2019,Smeden2019}.
+Under perfect testing, $D_M = 1$ so the relation between estimation and selection bias is simple, e.g., COVID-19 positive individuals more likely to receive test implies upward bias in prevalence estimates and vice versa.
+When tests are imperfect, this simple relationship no longer holds.
 
@@
 $$
-implying the error relative to SRS increases as a function of population size.
+implying the error relative to SRS increases as a function of population size~\cite{Meng2018}.
 The paradox that two countries with the same testing strategy can yield wildly different estimates due to population size is too often ignored in current communications of case-count data.
-When comparing the US and UK, for example, the MSE increases by a factor of five.
-Thus conclusions drawn from observed case-counts may be not just wrong, but very wrong.
+When comparing the US and UK under the assumption of similar testing strategies, for example, the MSE increases by a factor of five.
+Thus, conclusions drawn from observed case-counts may not just be wrong, but very wrong.
 
-A key question is ``What is the (effective) sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' This quantity can be bounded by
+A key question is ``What is the (effective) sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' In the supplement, we show the effective sample size $n_{eff}$ can be bounded by
 $$
@@
 Supposing COVID-19 positive individuals are $2$ times more likely to get tested, then the effective sample size is $10$.
-Recent proposals~\cite{Siddarth2020} have argued for increased testing capacity, which will reduce the relative sampling rate.
-Supposing it drops to $1.2$ then the effective sample size will increase to $216$.
+Recent proposals~\cite{Siddarth2020} have argued for increased testing capacity, which may likely reduce the relative sampling rate.
+If the relative sampling rate drops to $1.2$ then the effective sample size will increase to $216$.
 Thus the effective sample size even in optimistic scenarios is equivalent to a small random sample from the population.
 
-Increased testing capacity may also come at the cost of increased in false positive and negative rates.
+Increased testing capacity may also come at the cost of increases in false positive and negative rates.
 In terms of effective sample size, if the data quality term remains constant, then testing increases may not significantly improve the effective sample size.
@@
 Then the standard analysis would suggest a reduction in the MSE by 47\% but the actual reduction is 26\%.
-This implies that doubling the sampling fraction will only increase the effective sample size by a factor $2.9$ rather than the expected $4$ fold increase that is expected.
+This implies that doubling $f$ will increase the effective sample size by a factor $2.9$ rather than the expected $4$.
 
@@
 
-Another important question is whether selection bias and imperfect testing impact the ability to estimate rates of change in prevalence.
+The prior analysis demonstrates the limited information regarding COVID-19 prevalence in observational case-count data.
+Scientists, however, may claim that daily observed case-counts simply undercount daily total cases by a constant multiple over time.
+If true then the ratio of case-counts at consecutive times may be a good estimate of the true change in prevalence, helping scientists understand the disease trajectory.
+In this section, we demonstrate how selection bias and imperfect testing impact such estimates.
+
 Let $\bar Y_{t-1}$ and $\bar Y_{t}$ denote the prevalence on two consecutive days and consider the estimator $r = \bar y_t / \bar y_{t-1}$.
@@
 where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement error adjustment, and $CV(Y_j) = \sigma_{Y_j}/Y_j$ is the coefficient of variation on day $j$.
+See the supplementary materials for the derivation.
 The error magnitude depends on the true rate $\bar Y_{t} / \bar Y_{t-1}$ so a large decrease will have a relatively small error compared to a large increase.
@@
 In terms of bias, the rate is overestimated prior to the peak and underestimated afterwards.
-From a decision making perspective, such biases have potential impact.
+From a decision making perspective, such biases have a substantial impact.
 First, the overestimation may give lawmakers and governors more leverage in proposing aggressive actions that reduce prevalence.
 Of course, the analysis supports the argument that the estimates based on available data may be pessimistic.
-What is missed is that the direction of bias is non-constant over time.
+What is missing in the current public health discourse is that the bias direction of the prevalence estimate is non-constant over time.
 Underestimation post-peak puts pressure on lawmakers and governors to prematurely relax social distancing measures.
 Moreover, it appears that the peak time is the easiest to estimate in terms of having minimal error using available data.
-Note, however, that the standard errors on such estimates will be wide due to small   effective sample sizes.
+Note, however, that the standard errors will be large due to small effective sample sizes.
 
@@
 Under a Poisson likelihood, a simple relation between the trajectory of new cases and the effective reproduction number can be derived \cite{Bettencourt2008}.
-In particular, under an SIR model the number of case counts on day $t$ is Poisson distributed with rate $K_{t-1} \exp \left( \frac{1}{\gamma} (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval which is approximately $7$ days for COVID-19~\cite{Sanche2020}.
-Using this, a simple estimator can be given by
+In particular, under an SIR model the number of case counts on day $t$, denoted $K_t$, is Poisson distributed with rate $K_{t-1} \exp \left( \frac{1}{\gamma} (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval, which is approximately $7$ days for COVID-19~\cite{Sanche2020}.
+Using this relation, a moment-based estimator is given by
 $$
@@
 So far we have focused on understanding the limitations of using case-count data to understand population quantities of interest for a \emph{single} population.
-Many are interested in cross-population comparisons.
+Many are interested in cross-population comparisons to contrast the impact of countries' mitigation policies.
 Here, for simplicity, we focus on comparing the estimated effective reproductive rate.
@@
 Figure~\ref{fig:eff} shows how biases in estimation of each rate interact in complex ways.
-At first, the difference  is correctly estimated; then the gap is over-estimated as country A sees a rapid rise in cases; then the magnitude of over-estimation increases as country A sees declining case-count while country B sees rapidly increasing case-count; then the gap country A's rate is correctly estimated while country B's rate is under-estimated as it sees declining case-count; finally, the gap disappears.
-We are not claiming this will always be the case, and indeed for small enough selection bias and/or large gaps in case-counts such complex behaviour will not occur; however, this is demonstrats how observed information can tell a more complex story than the truth (i.e., country A's peak is 2 weeks prior to country B's peak).
+At first, the difference is correctly estimated; then the gap is over-estimated as country A sees a rapid rise in cases; then the magnitude of over-estimation increases as country A sees declining case-count while country B sees rapidly increasing case-count; then country A's rate is correctly estimated while country B's rate is under-estimated as it sees declining case-count; finally, the gap disappears.
+We do not claim this will always be the case; however, this demonstrates how observed information can tell a more complex story than the truth (i.e., country A's peak is 2 weeks prior to country B's peak).
 
-\paragraph*{Estimation of Data Quality.} We next consider estimation of data quality $E_{\I} \left[ \rho_{I,Y} \right]$ in the current context.
-In \cite{Meng2018}, estimation relied on observing the true outcome (i.e., election vote totals).
-This is not possible in the current crisis.
-Here, we propose use of survey samples as noisy but unbiased estimates of the true prevalence.
-Our goal is less inferential.
-Instead, the goal is to build sensitivity analyses that can aid in understanding the amount of information in observational COVID case-count data.
+\paragraph*{Estimation of Data Quality.} We next consider estimation of data quality $E_{\I} \left[ \rho_{I,Y} \right]$.
+In \cite{Meng2018}, estimation relied on observing the true outcome (i.e., election vote totals), which  is not possible in the current crisis.
+Here, we propose using surveys as noisy but unbiased estimates of the true prevalence.
+Our goal is not inferential.
+Instead, we aim to build sensitivity analyses that can aid in understanding the amount of information in observational COVID case-count data.
 
@@
 This corresponds to a false negative rate of $17.2\%$ (11.6-24.0\%) and false positive rate of $0.5\%$ (0.3 - 0.8\%).
-This leads to an estimated prevalence of $15.9\%$ with a range of 14.8\% to 17.0\%.
+The estimated prevalence is then $15.9\%$ with a range of 14.8\% to 17.0\%.
 For simplicity, we assume the study was conducted on April 20th.
@@
 The precision in reported case-count data gives the illusion of information when what what is needed is quantification of uncertainty.
-Selection bias implies analysts feel certain about incorrect conclusions.
+Selection bias leads data analysts to feel certain about incorrect conclusions.
 We end with a brief discussion of important related topics.
@@
 
-One can extend the SIR model to attempt an account for selection bias and measurement error; however, without strong assumptions on the selection mechanism, the estimates are often not identifiable.
+One can extend the SIR model to account for selection bias and measurement error.
+Without strong assumptions on the selection mechanism, however, the estimates are often not identifiable.
 When an issue ``cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically''\cite{CoxHink74}.
-In the absence of random sampling, the best route forward regardless of methodology detailed is the use of sensitivity analyses and humility in data-driven conclusions.
+Absent random sampling, the best route forward is careful sensitivity analyses and humility in data-driven conclusions.
 
@@
 As of May 5th, Canada has performed a total of 24.92 tests per 1,000 people while the US has performed a total of 22.01 tests per 1,000 people~\cite{worlddata} so the data quantities are approximately equal to $f = 23/1000 = 0.023$.
-Then, under a prevalence of 10\%, one would need the differential $\Delta_1-\Delta_2$ to be much smaller than $11.68 \times 10^{-6}$ in order for the selection bias to not impact.
+Then, under a prevalence of 10\%, one would need the differential $\Delta_1-\Delta_2$ to be much smaller than $11.68 \times 10^{-6}$ in order for the selection bias to not have substantial impact.
 So only under the very strong assumptions of equal data quantity and quality will the comparisons be valid and the $Z$-score can be treated as in the classical setting.
@@
 $$
-Even in a randomized experiment, the question is what the correct value for $E(U)$ is.
+Even in a randomized experiment, the question is what the correct value for $E(U)$.
 For COVID-19, the value of interest is the expected value of $U$ in the population of COVID-19 positive individuals.

### 2020-06-05 write-up/scifile.tex
Update scifile.tex

@@
 Let $\bar Y_{t-1}$ and $\bar Y_{t}$ denote the prevalence on two consecutive days and consider the estimator $r = \bar y_t / \bar y_{t-1}$.
-Using a Taylor series approximation, the error between ${\bar y_t}/{\bar y_{t-1}}$ and ${\bar Y_{t}}/{\bar Y_{t-1}}$ can be expressed approximately as
+Using a second-order Taylor series approximation, the error between ${\bar y_t}/{\bar y_{t-1}}$ and ${\bar Y_{t}}/{\bar Y_{t-1}}$ can be expressed approximately as
 $$

### 2020-06-06 write-up/scifile.tex
Fixing for pub

@@
   Many statisticians, epidemiologists, economists and data scientists have registered serious reservations regarding the reported coronavirus case-counts.
-Limited testing capacity across the country has been widely identified as a key driver of suppressed coronavirus case-counts.
+Limited testing capacity across the country has been  identified as a key driver of suppressed coronavirus case-counts.
 The calls to increase testing capacity are well-justified as they become a more frequent point of discussion in the public sphere.
@@
 This pandemic
-has become the focal point of every day life; yet the data landscape for understanding COVID-19 remains limited.
+has become the focal point of everyday life; yet the data landscape for understanding COVID-19 remains limited.
 Public databases~\cite{JHU_Lancet,NYT} provide incoming county-level information of confirmed cases and deaths.
@@
 Imperfect testing means a selected individual's observed COVID-19 status may not be the same as their true status.
-Here we assume the false positive rate (FP) and false negative rate (FN) are known and consider the estimator $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ where $\bar y = n^{-1} \sum_{i=1}^n y_i$ is the proportion of positive tests.
+Here we assume the false positive rate (FP) and false negative rate (FN) are known and consider the estimator $\tilde y_n = \bar y_n - (1-\bar y_n) FP + FN \bar y_n$ where $\bar y = n^{-1} \sum_{i=1}^n y_i$ is the proportion of positive tests.
 This estimator is unbiased under simple random sampling (SRS).
@@
 
-On April 23rd, Andrew Cuomo announced results from a study in New York state.
+On April 23rd, Governor Andrew Cuomo announced results from a study in New York state.
 It found 13.9\% of 3,000 people tested across the state had signs of the virus.
@@
 \[
-\rho D_M = \sqrt{\frac{f}{1-f}} \frac{\text{0.166}}{\sigma_Y} = 1.43 \times 10^{-2} \Rightarrow \Delta = 1.06 \times 10^{-3} \Rightarrow M = 2.29.
+\rho D_M = \sqrt{\frac{f}{1-f}} \frac{\text{0.166}}{\sigma_Y} = 1.48 \times 10^{-2} \Rightarrow \Delta = 1.09 \times 10^{-3} \Rightarrow M = 2.32.
 \]
-We can perform a sensitivity analysis by considering the range of false negative and positive rates from~\cite{Bendavid2020}, which leads to a range for $M$ of $(2.05, 2.54)$.
+We can perform a sensitivity analysis by considering the range of false negative and positive rates from~\cite{Bendavid2020}, which leads to a range for $M$ of $(2.26, 2.39)$.
 Therefore, the impact on bias of $\hat R_t$ appears moderate.
-However, for $FP = 0.05$ and $FN = 0.005$, the relative sampling rate is $M=4.31$.
+If $FP = 0.05$ and $FN = 0.005$, then the relative sampling rate increases to $M=3.00$.
 This simple tool helps us understand how sensitivity and specificity impact biases and  therefore how much we should trust conclusions based on these assumptions.

### 2020-06-06 methods/ny_calc.R

@@ -1,4 +1,4 @@
-FP = 0.00; FN = 0.000
+FP = 0.005; FN = 0.172
 bary0 = 0.139
 bary = bary0*(1+FN) - FP*(1-bary0)
 f= 0.001

### 2020-06-09 write-up/scifile.tex
Paper updated appendix

@@
 Imperfect testing means a selected individual's observed COVID-19 status may not be the same as their true status.
-Here we assume the false positive rate (FP) and false negative rate (FN) are known and consider the estimator $\tilde y_n = \bar y_n - (1-\bar y_n) FP + FN \bar y_n$ where $\bar y = n^{-1} \sum_{i=1}^n y_i$ is the proportion of positive tests.
-This estimator is unbiased under simple random sampling (SRS).
-To understand the impact of selection bias and imperfect testing, we derive a statistical decomposition of the error between $\tilde y_n$ and $\bar Y$ which is given by
+Here we assume the false positive rate (FP) and false negative rate (FN) are known.
+In this paper, we consider the estimator $\tilde y_n = (\bar y_n - FP)/(1-(FP+FN))$ where $\bar y = n^{-1} \sum_{i=1}^n y_i$ is the proportion of positive tests.
+This estimator is unbiased under simple random sampling (SRS); see the supplementary materials for a derivation and discussion of the proposed estimator.
+To understand the impact of selection bias and imperfect testing, we derive the following statistical decomposition of the error between $\tilde y_n$ and $\bar Y$:
 \begin{equation}
@@
 \rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
-\times \underbrace{\left[ 1 + FP + FN - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right]}_{D_M},
+\times \underbrace{\left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{D_M},
 \end{equation}
@@
 As of May 8th, the US has performed $8,412,095$ tests.
-The US population is roughly $328$ million, so the fraction sampled is $f = 0.026$.
-The empirical prevalence is $9.1\%$ on May 8th.
-A recent study estimates a false positive and negative rate for antibody testing using a lateral flow immunoassay are $0.5$\% and $17.2$\% respectively \cite{Bendavid2020}.
-Supposing COVID-19 positive individuals are $2$ times more likely to get tested, then the effective sample size is $10$.
+The US population is roughly $328$ million, so $f = 0.066$.
+The empirical prevalence is $9.5\%$ on May 8th.
+A recent study estimates a false positive and negative rate for antibody testing at $0.5$\% and $17.2$\% respectively \cite{Bendavid2020}.
+Supposing COVID-19 positive individuals are $2$ times more likely to get tested, then the effective sample size is $103$.
 Recent proposals~\cite{Siddarth2020} have argued for increased testing capacity, which may likely reduce the relative sampling rate.
-If the relative sampling rate drops to $1.2$ then the effective sample size will increase to $216$.
+If the relative sampling rate drops to $1.2$ then the effective sample size will increase to $2212$.
 Thus the effective sample size even in optimistic scenarios is equivalent to a small random sample from the population.
 
+{\bf Finish!}
 Increased testing capacity may also come at the cost of increases in false positive and negative rates.
@@
 This corresponds to a false negative rate of $17.2\%$ (11.6-24.0\%) and false positive rate of $0.5\%$ (0.3 - 0.8\%).
-The estimated prevalence is then $15.9\%$ with a range of 14.8\% to 17.0\%.
+The estimated prevalence is then $16.3\%$ with a range of 15.0\% to 18.0\%.
 For simplicity, we assume the study was conducted on April 20th.
@@
 Subtracting off the $617,555$ tests that had already performed yields a sampling fraction of $f = 0.001$.
-The smoothed prevalence estimate is 28.3\%.
-Adjusting for imperfect testing yields an estimate of $32.5\%$.
+The smoothed prevalence estimate is 28.4\%.
+Adjusting for imperfect testing yields an estimate of $33.9\%$.
 
-Under the assumption that the survey is a SRS from New York, the error is $16.6$\%.  Using~\eqref{eq:error}, we construct an estimate of the relative sampling rate as follows
+Under the assumption that the survey is a SRS from New York, the error is $17.6$\%.  Using~\eqref{eq:error}, we construct an estimate of the relative sampling rate as follows
 \[
-\rho D_M = \sqrt{\frac{f}{1-f}} \frac{\text{0.166}}{\sigma_Y} = 1.48 \times 10^{-2} \Rightarrow \Delta = 1.09 \times 10^{-3} \Rightarrow M = 2.32.
+\rho D_M = \sqrt{\frac{f}{1-f}} \frac{\text{0.176}}{\sigma_Y} = 1.07 \times 10^{-2} \Rightarrow \Delta = 1.07 \times 10^{-3} \Rightarrow M = 2.3.
 \]
-We can perform a sensitivity analysis by considering the range of false negative and positive rates from~\cite{Bendavid2020}, which leads to a range for $M$ of $(2.26, 2.39)$.
+We can perform a sensitivity analysis by considering the range of false negative and positive rates from~\cite{Bendavid2020}, which leads to a range for $M$ of $(2.21, 2.38)$.
 Therefore, the impact on bias of $\hat R_t$ appears moderate.
-If $FP = 0.05$ and $FN = 0.005$, then the relative sampling rate increases to $M=3.00$.
+If $FP = 0.05$ and $FN = 0.005$, then the relative sampling rate increases to $M=3.04$.
 This simple tool helps us understand how sensitivity and specificity impact biases and  therefore how much we should trust conclusions based on these assumptions.
@@
 \section*{Supplementary materials}
-
-\subsection*{Prevalence comparisons}
-
-At a fixed time, suppose prevalence estimates $\bar y_1$ and $\bar y_2$ are observed for two populations. We can express $\bar y_1 - \bar y_2$ in statistical terms as
-$$
-(\bar Y_1 - \bar Y_2) + \left[ \rho_{I_1, Y_1} \times D_{M_1} \times \sqrt{\frac{1-f_1}{f_1}} \times \sigma_{Y_1}  - \rho_{I_2, Y_2} \times D_{M_2} \times \sqrt{\frac{1-f_2}{f_2}} \times \sigma_{Y_2} \right].
-$$
-The first term is the difference in population prevalence (i.e., the quantity of interest).
-If random sampling were performed in each population, $E_{\I} [ \rho_{I_j, Y_j}] = 0$ and so the estimate would be unbiased.
-However, the second term represents the complex error that results from selection bias.
-The particular test of interest is $H_0: \bar Y_1 = \bar Y_2$.  Under the null, $\sigma_{Y_1} = \sigma_{Y_2}$.  In classical statistics, we would compute a Z-score:
-$$
-\begin{aligned}
-\frac{\bar y_1 - \bar y_2}{ \sqrt{V_{SRS} (\bar y_1 - \bar y_2)} } &=
-\frac{\rho_{I_1, Y_1} \times D_{M_1} \times \sqrt{\frac{1-f_1}{f_1}}  - \rho_{I_2, Y_2} \times D_{M_2} \times \sqrt{\frac{1-f_2}{f_2}} }{ \sqrt{\frac{1}{N_1 -1} \times \frac{1-f_1}{f_1} + \frac{1}{N_2 - 1} \times \frac{1-f_2}{f_2}} }  \\
-&\overset{\star}{=}
-\sqrt{\frac{(N_1-1)(N_2 -1)}{N_1 + N_2}} \left[ \rho_{I_1, Y_1} - \rho_{I_2, Y_2} \right]
-\end{aligned}
-$$
-Where the last equality holds under $f_1 = f_2$, i.e., same sampling rate relative to population size.
-This tells us that the exact error in the comparison of sample means, as an estimate of the difference in population means, scales as a function of the difference in data quality $\rho_{I_1, Y_1} - \rho_{I_2, Y_2}$ times a square root of a population adjustment.
-For highly unequal population sizes (e.g., $N_1 \gg N_2$), the adjustment is approximated well by the square root of the smaller population size minus one (e.g.
-$\sqrt{N_2 -1}$).
-
-Take, for example, a comparison of the US and Canada.
-The two population sizes are approximately 328 million and 38 million people.
-Then the population adjustment is $5.84$ million.
-As of May 5th, Canada has performed a total of 24.92 tests per 1,000 people while the US has performed a total of 22.01 tests per 1,000 people~\cite{worlddata} so the data quantities are approximately equal to $f = 23/1000 = 0.023$.
-Then, under a prevalence of 10\%, one would need the differential $\Delta_1-\Delta_2$ to be much smaller than $11.68 \times 10^{-6}$ in order for the selection bias to not have substantial impact.
-So only under the very strong assumptions of equal data quantity and quality will the comparisons be valid and the $Z$-score can be treated as in the classical setting.
-A simple alternative is to use the effective sample sizes to build more appropriate $Z$-score.
-That is, setting
-$$
-Z_n = \frac{\bar y_1 - \bar y_2}{ \frac{1-f}{f} \sigma_y \sqrt{ \frac{1}{n_{1,eff} - 1} + \frac{1}{n_{2,eff} - 1}}}
-$$
-allows us to build Z-scores that account for the true sample size.  While still not perfect, this allows us to build Z-scores that are not overly confident due to selection bias and measurement error.
-
-\subsection*{Effect modifiers in COVID-19 clinical trials}
-
-Here we show how selection bias and measurement error can creep into clinical trial analysis.
-The key concern is whether clinical trials on COVID-19 recruit from the pool of individuals who have tested positive for COVID-19 or whether they sample randomly from the population, test, and then recruit from this subset of tested individuals.
-To see this issue, suppose we have an outcome $Y$, a treatment $A$ and an \emph{unobserved} variable $U$.
-Suppose treatment is assigned at random, i.e., $A =1$ with probability $50\%$ and $A=0$ with probability $50\%$.
-Suppose the conditional mean of the outcome satisfies  $E(Y \mid U, A ) = \beta_0 + \beta_1 A + \beta_2 U + \beta_3 A U$.
-Typically, we are interested in the \emph{causal effect} of $A$ on $Y$.
-In counterfactual language,
-$$
-E( Y(A=1) - Y(A=0) ) = E_U ( E_A ( Y(A=1) - Y(A=0) \mid U=u ))
-= \beta_1 + \beta_3 E(U).
-$$
-Even in a randomized experiment, the question is what the correct value for $E(U)$.
-For COVID-19, the value of interest is the expected value of $U$ in the population of COVID-19 positive individuals.
-Due to selection bias, however, the bias in this marginal treatment effect compared to the true marginal effect of interest is $\beta_1 \cdot E_{\I} [ \rho_{I,U} ] \times \sqrt{ \frac{f}{1-f} } \times \sigma_U$.
-Selection bias then leads to the estimated effect being biased if $\beta_1 \neq 0$.
-Directionality of the bias will depend on the relationship between $U$ and $Y$.
-If $U$ is a measure of disease severity, then it could be expected to see $\beta_1 > 0$ and $E_{\I} [ \rho_{I, U} ] >0$.
-In such situations, the treatment effect estimates will tend to be overly optimistic as most individuals recruited are potentially higher on the severity index than the population average.
-Not only that, but for fixed data quality, the error compared to SRS in terms of effect estimation scales with the population size (i.e., the law of large populations).
-
-Randomization of treatment assignment in clinical trials negates unobserved confounders.
-This, however, does not negate effect modifiers.
-Therefore, for marginal treatment effects to be interpretable, there must be a well-defined population.
-Most often, our main interest is in causal effects on the population of COVID-19 positive.
-Randomization within the clinical trial design yields internal validity, but we also need external validity to estimate the correct marginal effect of interest~\cite{Keiding2016}.
 
@@
 $$
+We see the final term is given by $FP (1-\bar Y) - FN \bar Y$.  This motivates the adjusted estimate
+$$
+\tilde y_n^{(0)} = \bar y_n^\star - FP (1- \bar y_n^\star ) + FN \bar y_n^\star
+= \bar y_n^\star (1 + FN + FP) - FP.
+$$
+Now considering the error for the adjusted estimate, $\tilde y_n^{(0)} - \bar Y$, we have
+$$
+\begin{aligned}
+ &\bar y_n^\star (1 + FN + FP) - FP - \bar Y \\
+ =&(\bar y_n^\star - \bar Y) +  (FN + FP) \bar y_n^\star - FP  \\
+ =&\underbrace{\sqrt{\frac{1-f}{f}} \left[\rho_{I,Y} \sigma_Y + \rho_{I, PZ} \sigma_{PZ} \right]}_{\Psi} + (  FN + FP ) (\bar y_n^\star -  \bar Y) \\
+ =& \Psi + (FN + FP) \Psi + (FN + FP) (FP - (FP+FN) \bar Y).
+ \end{aligned}
+$$
+Now define $\tilde y_n^{(1)} = \tilde y_n^{(0)} + (FN+FP)  ( (FN+FP) \bar y_n - FP)$.  Deriving the error $\tilde y_n^{(1)} - \bar Y$, we have
+$$
+\begin{aligned}
+ &(\bar y_n^{(0)} - \bar Y) + (FN+FP)  ( (FN+FP) \bar y_n^\star - FP)\\
+ =&\Psi + (FN + FP) \Psi + (FN + FP) (FP - (FP+FN) \bar Y) + (FN + FP) ((FP+FN) \bar y_n^\star - FP)  \\
+  =& \Psi + (FN + FP) \Psi + (FN + FP)^2 \Psi + (FN+FP)^2 (FP - (FP+FN) \bar Y).
+ \end{aligned}
+$$
+Recursively defining  $\tilde y_n^{(t)} = \tilde y_n^{(t-1)} + (FN+FP)  ( (FN+FP) \bar y_n - FP)$ for $t=1,2,\ldots$ where $\tilde y_n^{(0)} = \bar y_n^\star$.  Then
+$$
+\tilde y_n^{(t)} = \bar y_n^\star \sum_{s=0}^{t+1} (FP+FN)^s - FP \sum_{s=0}^{t} (FP+FN)^s
+$$
+and the associated error at iteration $t$ given by
+$$
+\Psi \sum_{s=0}^t (FN+FP)^s = \Psi \frac{1 - (FN+FP)^t}{1 - (FN+FP)}.
+$$
+Define
+$$
+\tilde y = \lim_{t \to \infty} \tilde y_n^{(t)} = \frac{\bar y_n^\star - FP}{1 -(FN+FP)}.
+$$
+The associated error can be expressed as $\frac{\Psi}{1-(FN+FP)}$.
+
+\subsubsection*{Estimate of observed data quality}
+
 For the binary outcome $Y$, we have $\sigma_Y = \sqrt{\bar Y (1-\bar Y)}$. Moreover,
@@
 Then the formula for the error is given by:
-$$
-\sqrt{\frac{1-f}{f}} \left[\rho_{I,Y} \sqrt{\bar Y (1-\bar Y)} + \rho_{I, PZ} \sqrt{ 2 \bar Y \left( FP (1-\bar Y) + FN \cdot \bar Y \right )} + \sqrt{\frac{f}{1-f}} \left( FP - (FP+FN) \bar Y \right) \right]
-$$
-
-\subsubsection*{Estimate of observed data quality}
-
+\begin{equation}
+\label{eq:finalstep}
+\sqrt{\frac{1-f}{f}} \left[\rho_{I,Y} \sqrt{\bar Y (1-\bar Y)} + \rho_{I, PZ} \sqrt{ 2 \bar Y \left( FP (1-\bar Y) + FN \cdot \bar Y \right )} \right] \times \frac{1}{1- (FN+FP)}
+\end{equation}
 By definition, we have
@@
 $$
-\rho_{I,Y} \sigma_Y \left( 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right)
+\rho_{I,Y} \sigma_Y \left( 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right).
 $$
-Suppose $f_1 = M \cdot f_0$.  Then $\Delta = f_1 - f_0 = f_0 (M-1)$ and $f_0 (1-\bar Y) + f_1 \bar Y = f_0 ( (1-\bar Y) + M \bar Y )$ and we can re-write above as
-$$
-\rho_{I,Y} \sigma_Y \left( 1 - (M-1) \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{(1-\bar Y) + M \bar Y} \right).
-$$
+Inserting into equation~\eqref{eq:finalstep} yields the desired result.
 
@@
 \end{table}
+
+\subsection*{Prevalence comparisons}
+
+At a fixed time, suppose prevalence estimates $\bar y_1$ and $\bar y_2$ are observed for two populations. We can express $\bar y_1 - \bar y_2$ in statistical terms as
+$$
+(\bar Y_1 - \bar Y_2) + \left[ \rho_{I_1, Y_1} \times D_{M_1} \times \sqrt{\frac{1-f_1}{f_1}} \times \sigma_{Y_1}  - \rho_{I_2, Y_2} \times D_{M_2} \times \sqrt{\frac{1-f_2}{f_2}} \times \sigma_{Y_2} \right].
+$$
+The first term is the difference in population prevalence (i.e., the quantity of interest).
+If random sampling were performed in each population, $E_{\I} [ \rho_{I_j, Y_j}] = 0$ and so the estimate would be unbiased.
+However, the second term represents the complex error that results from selection bias.
+The particular test of interest is $H_0: \bar Y_1 = \bar Y_2$.  Under the null, $\sigma_{Y_1} = \sigma_{Y_2}$.  In classical statistics, we would compute a Z-score:
+$$
+\begin{aligned}
+\frac{\bar y_1 - \bar y_2}{ \sqrt{V_{SRS} (\bar y_1 - \bar y_2)} } &=
+\frac{\rho_{I_1, Y_1} \times D_{M_1} \times \sqrt{\frac{1-f_1}{f_1}}  - \rho_{I_2, Y_2} \times D_{M_2} \times \sqrt{\frac{1-f_2}{f_2}} }{ \sqrt{\frac{1}{N_1 -1} \times \frac{1-f_1}{f_1} + \frac{1}{N_2 - 1} \times \frac{1-f_2}{f_2}} }  \\
+&\overset{\star}{=}
+\sqrt{\frac{(N_1-1)(N_2 -1)}{N_1 + N_2}} \left[ \rho_{I_1, Y_1} - \rho_{I_2, Y_2} \right]
+\end{aligned}
+$$
+Where the last equality holds under $f_1 = f_2$, i.e., same sampling rate relative to population size.
+This tells us that the exact error in the comparison of sample means, as an estimate of the difference in population means, scales as a function of the difference in data quality $\rho_{I_1, Y_1} - \rho_{I_2, Y_2}$ times a square root of a population adjustment.
+For highly unequal population sizes (e.g., $N_1 \gg N_2$), the adjustment is approximated well by the square root of the smaller population size minus one (e.g.
+$\sqrt{N_2 -1}$).
+
+Take, for example, a comparison of the US and Canada.
+The two population sizes are approximately 328 million and 38 million people.
+Then the population adjustment is $5.84$ million.
+As of May 5th, Canada has performed a total of 24.92 tests per 1,000 people while the US has performed a total of 22.01 tests per 1,000 people~\cite{worlddata} so the data quantities are approximately equal to $f = 23/1000 = 0.023$.
+Then, under a prevalence of 10\%, one would need the differential $\Delta_1-\Delta_2$ to be much smaller than $11.68 \times 10^{-6}$ in order for the selection bias to not have substantial impact.
+So only under the very strong assumptions of equal data quantity and quality will the comparisons be valid and the $Z$-score can be treated as in the classical setting.
+A simple alternative is to use the effective sample sizes to build more appropriate $Z$-score.
+That is, setting
+$$
+Z_n = \frac{\bar y_1 - \bar y_2}{ \frac{1-f}{f} \sigma_y \sqrt{ \frac{1}{n_{1,eff} - 1} + \frac{1}{n_{2,eff} - 1}}}
+$$
+allows us to build Z-scores that account for the true sample size.  While still not perfect, this allows us to build Z-scores that are not overly confident due to selection bias and measurement error.
+
+\subsection*{Effect modifiers in COVID-19 clinical trials}
+
+Here we show how selection bias and measurement error can creep into clinical trial analysis.
+The key concern is whether clinical trials on COVID-19 recruit from the pool of individuals who have tested positive for COVID-19 or whether they sample randomly from the population, test, and then recruit from this subset of tested individuals.
+To see this issue, suppose we have an outcome $Y$, a treatment $A$ and an \emph{unobserved} variable $U$.
+Suppose treatment is assigned at random, i.e., $A =1$ with probability $50\%$ and $A=0$ with probability $50\%$.
+Suppose the conditional mean of the outcome satisfies  $E(Y \mid U, A ) = \beta_0 + \beta_1 A + \beta_2 U + \beta_3 A U$.
+Typically, we are interested in the \emph{causal effect} of $A$ on $Y$.
+In counterfactual language,
+$$
+E( Y(A=1) - Y(A=0) ) = E_U ( E_A ( Y(A=1) - Y(A=0) \mid U=u ))
+= \beta_1 + \beta_3 E(U).
+$$
+Even in a randomized experiment, the question is what the correct value for $E(U)$.
+For COVID-19, the value of interest is the expected value of $U$ in the population of COVID-19 positive individuals.
+Due to selection bias, however, the bias in this marginal treatment effect compared to the true marginal effect of interest is $\beta_1 \cdot E_{\I} [ \rho_{I,U} ] \times \sqrt{ \frac{f}{1-f} } \times \sigma_U$.
+Selection bias then leads to the estimated effect being biased if $\beta_1 \neq 0$.
+Directionality of the bias will depend on the relationship between $U$ and $Y$.
+If $U$ is a measure of disease severity, then it could be expected to see $\beta_1 > 0$ and $E_{\I} [ \rho_{I, U} ] >0$.
+In such situations, the treatment effect estimates will tend to be overly optimistic as most individuals recruited are potentially higher on the severity index than the population average.
+Not only that, but for fixed data quality, the error compared to SRS in terms of effect estimation scales with the population size (i.e., the law of large populations).
+
+Randomization of treatment assignment in clinical trials negates unobserved confounders.
+This, however, does not negate effect modifiers.
+Therefore, for marginal treatment effects to be interpretable, there must be a well-defined population.
+Most often, our main interest is in causal effects on the population of COVID-19 positive.
+Randomization within the clinical trial design yields internal validity, but we also need external validity to estimate the correct marginal effect of interest~\cite{Keiding2016}.
 

### 2020-06-09 write-up/scifile.tex
Update scifile.tex

@@
 
-\subsection*{Imperfect testing: derivation}
+\paragraph*{Derivations for imperfect testing framework.}
 \label{app:imperfect}
-We considered the mean estimator
+We start by considering the empirical mean estimator under imperfect testing,
 $$
@@
 $$
-\bar y_n^\star - \bar Y = \sqrt{\frac{1-f}{f}} \left[\rho_{I,Y} \sigma_Y + \rho_{I, PZ} \sigma_{PZ} + \sqrt{\frac{f}{1-f}} \left( FP - (FP+FN) \bar Y \right) \right]
+\bar y_n^\star - \bar Y = \sqrt{\frac{1-f}{f}} \left(\rho_{I,Y} \sigma_Y + \rho_{I, PZ} \sigma_{PZ} \right) + \left( FP - (FP+FN) \bar Y \right)
 $$
-We see the final term is given by $FP (1-\bar Y) - FN \bar Y$.  This motivates the adjusted estimate
+We see the final term is given by $FP (1-\bar Y) - FN \bar Y$ is the bias associated with using the unadjusted prevalence estimate $\bar y_n^\star$.  This motivates an adjusted estimate
 $$
@@
 $$
-Now define $\tilde y_n^{(1)} = \tilde y_n^{(0)} + (FN+FP)  ( (FN+FP) \bar y_n - FP)$.  Deriving the error $\tilde y_n^{(1)} - \bar Y$, we have
+The final term is a (smaller) bias term and so we propose another adjusted estimator $\tilde y_n^{(1)} = \tilde y_n^{(0)} + (FN+FP)  ( (FN+FP) \bar y_n^\star - FP)$, with associated error $\tilde y_n^{(1)} - \bar Y$ given by
 $$
@@
 $$
-Recursively defining  $\tilde y_n^{(t)} = \tilde y_n^{(t-1)} + (FN+FP)  ( (FN+FP) \bar y_n - FP)$ for $t=1,2,\ldots$ where $\tilde y_n^{(0)} = \bar y_n^\star$.  Then
+This motivates recursively defining estimators $\tilde y_n^{(t)} = \tilde y_n^{(t-1)} + (FN+FP)  ( (FN+FP) \bar y_n - FP)$ for $t=1,2,\ldots$ where $\tilde y_n^{(0)} = \bar y_n^\star$.  Then
 $$
@@
 $$
-Define
+We can then get an estimator with no residual bias term by taking the limit as $t$ goes to infinity; that is, define
 $$
-\tilde y = \lim_{t \to \infty} \tilde y_n^{(t)} = \frac{\bar y_n^\star - FP}{1 -(FN+FP)}.
+\tilde y_n = \lim_{t \to \infty} \tilde y_n^{(t)} = \frac{\bar y_n^\star - FP}{1 -(FN+FP)}.
 $$
-The associated error can be expressed as $\frac{\Psi}{1-(FN+FP)}$.
+Then the associated error $\tilde y_n - \bar Y$ can be expressed as $\frac{\Psi}{1-(FN+FP)}$.
 
-\subsubsection*{Estimate of observed data quality}
+\paragraph*{Model-based derivation of $\tilde y$.}
+The estimator $\tilde y$ was derived as a limit of a process that removes the residual bias term at each step.
+Here we consider a model-based explanation.
+Let $\theta = \pr( Y = 1)$ and $\phi = \pr( \text{test is positive})$.
+Given a known false negative (FN) and false positive (FP) rates, we have
+$$
+\begin{aligned}
+\phi &= \theta \cdot (1-FN) + (1-\theta) \cdot FP = \theta (1-FN-FP) + FP \\
+\Rightarrow \theta &= \frac{\phi - FP}{1-FN-FP}.
+\end{aligned}
+$$
+Thus, the estimator $\tilde y$ is also the appropriate estimator under a model-based approach.
+While the derivation here is more straightforward, the derivation in the prior section provides a simple formula for the associated error $\tilde y_n - \bar Y$ and gives a novel connection between the empirical estimator $\bar y_n^\star$ and the adjusted estimator $\tilde y_n$ without reference to the model-based approach.
 
+\paragraph*{Simplifying $\Psi$.}
 For the binary outcome $Y$, we have $\sigma_Y = \sqrt{\bar Y (1-\bar Y)}$. Moreover,
@@
 \begin{aligned}
-V_J(P_J Z_J) &= E_J[(P_J Z_J)^2] - E[P] E[Z] \\
-&= E[P] - E[P] (1 - 2 \bar Y) = 2 \bar Y E_J [ P_J ] \\
+V_J(P_J Z_J) &= E_J[(P_J Z_J)^2] - E_J[P_J] E_J[Z_J] \\
+&= E_J[P_J] - E_J[P_J] (1 - 2 \bar Y) = 2 \bar Y E_J [ P_J ] \\
 &= 2 \bar Y \left( FP (1-\bar Y) + FN \bar Y \right) \\

### 2020-06-09 write-up/scifile.tex
Finalized code

@@
 A recent study estimates a false positive and negative rate for antibody testing at $0.5$\% and $17.2$\% respectively \cite{Bendavid2020}.
-Supposing COVID-19 positive individuals are $2$ times more likely to get tested, then the effective sample size is $103$.
+Supposing COVID-19 positive individuals are $2$ times more likely to get tested, then the effective sample size is $89$.
 Recent proposals~\cite{Siddarth2020} have argued for increased testing capacity, which may likely reduce the relative sampling rate.
-If the relative sampling rate drops to $1.2$ then the effective sample size will increase to $2212$.
+If the relative sampling rate drops to $1.2$ then the effective sample size will increase to $1879$.
 Thus the effective sample size even in optimistic scenarios is equivalent to a small random sample from the population.
-
-{\bf Finish!}
-Increased testing capacity may also come at the cost of increases in false positive and negative rates.
-In terms of effective sample size, if the data quality term remains constant, then testing increases may not significantly improve the effective sample size.
-Consider the case where $f$ increases from $0.05$ to $0.1$ and the increase is associated with the false positive rate increasing from $0.5$\% to $5$\% and the false negative rate rising from $5\%$ to $20$\%.
-Then the standard analysis would suggest a reduction in the MSE by 47\% but the actual reduction is 26\%.
-This implies that doubling $f$ will increase the effective sample size by a factor $2.9$ rather than the expected $4$.
+Finally, increased testing capacity may also come at the cost of increases in false positive and negative rates.
+Consider the case where $f$ increases from $0.05$ to $0.1$ and the increase is associated with the false negative rate rising from $5\%$ to $20$\%.
+Then the effective sample size increases by a factor of $1.76$ rather than the expected $4$ fold increase.
 
@@
 
+\subsection*{Derivation of effective sample size}
+
+Let $S_Y^2 = (N-1)^{-1} \sum_{j=1}^N (Y_j - \bar Y)^2$ be the population variance as defined in survey sampling~\cite{Cochran77}.
+Then $\sigma_Y^2 = (N-1)/N \cdot S_Y^2$.
+Under SRS, the MSE is the variance as the estimate is unbiased and the variance is given by $(1-f)/n S_Y^2$.
+Then setting the MSE under general selection and SRS equal we have
+$$
+\begin{aligned}
+\underbrace{\frac{1-f}{f} \times E_{\I} \left[ \rho_{I, Y}^2 \times D_M^2 \right]}_{1/n_{eff}^\star} \times \sigma_Y^2 &= \frac{1-f}{n_{eff}} S_Y^2 \\
+\frac{1}{n_{eff}^\star} \times \frac{N-1}{N} S_Y^2 &= \frac{1-f}{n_{eff}} S_Y^2 \\
+\frac{1}{n_{eff}^\star} &=  \left( \frac{1}{n_{eff}} - \frac{1}{N} \right) \left( \frac{N}{N-1} \right) \\
+\frac{1}{n_{eff}^\star} \left[ 1 - \frac{1}{N} + \frac{n_{eff}^\star}{N} \right]  &=  \frac{1}{n_{eff}} \\
+n_{eff}^\star \left[ 1 - \frac{1}{N} + \frac{n_{eff}^\star}{N} \right]^{-1}  &=  n_{eff} \\
+\frac{n_{eff}^\star}{ 1 + (n_{eff}^\star -1) N^{-1}} &=  n_{eff} \\
+\end{aligned}
+$$
+Then if $n_{eff}^\star \geq 1$, we have that
+$$
+n_{eff} \leq n_{eff}^\star = \frac{f}{1-f} \times \frac{1}{E_{\I} \left[ \rho^2_{I, Y} \times D_M^2 \right]}
+$$
+
 \subsection*{Ratio estimator}
@@
 
-\subsection*{Derivation for the effective sample size}
+\subsection*{Computating the effective sample size}
 \label{section:effss}
@@
 \end{equation}
-where $\Delta = P_J (I_J = 1 \mid Y_J = 1) - P(I_J = 1 \mid Y_J = 0) = f_1 - f_0$.  Suppose that $M = f_1/f_0$; then
-$f_0 = f / (\bar Y \cdot (M-1) + 1)$.  For $\bar Y = 0.091$, $M = 2$ and $f = 0.026$ we have $f_0 = 0.024$ and $f_1 = 0.048$ which implies $\Delta = 0.024$.
-Then $\rho_{I,Y} = 0.043$.  Then $E_{\I} [ \rho_{I,Y}^2 ] \leq (E_{\I} [\rho_{I,Y} ]^2 = 0.0019$.  Then
-$$
-n_{eff} \leq \frac{f}{1-f} \frac{1}{0.0019} = 14.39.
-$$
-Using this computation, we compute effective sample size under a range of values $\bary$, $M$ with $f = 0.026$ (i.e., current sampling fraction) and $f =1/7$ (sample individuals roughly every 7 days).
+where $\Delta = P_J (I_J = 1 \mid Y_J = 1) - P(I_J = 1 \mid Y_J = 0) = f_1 - f_0$.
+Suppose that $M = f_1/f_0$; then $f_0 = f / (\bar Y \cdot (M-1) + 1)$.
+Using the upper bound $E_{\I} [ \rho_{I,Y}^2 ] \leq E_{\I} [\rho_{I,Y} ]^2$, we compute effective sample size under a range of values $\bary$, $M$ with $f = 0.066$ (i.e., current sampling fraction) and observed prevalence $0.095$.
 
@@
   \hline
-0.016 & 1598.00 & 402.00 & 180.00 & 102.00 & 66.00 \\
-  0.036 & 731.00 & 185.00 & 84.00 & 48.00 & 31.00 \\
-  0.056 & 484.00 & 124.00 & 56.00 & 32.00 & 21.00 \\
-  0.076 & 367.00 & 94.00 & 43.00 & 25.00 & 16.00 \\
-  0.096 & 299.00 & 78.00 & 36.00 & 21.00 & 14.00 \\
+0.05 & 11303 & 2882 & 1306 & 749 & 489 \\
+  0.07 & 6065 & 1558 & 712 & 411 & 270 \\
+  0.09 & 3862 & 1000 & 460 & 268 & 177 \\
+  0.11 & 2724 & 711 & 329 & 193 & 129 \\
+  0.13 & 2057 & 541 & 252 & 149 & 100 \\
    \hline
@@
   \hline
-0.016 & 1154.00 & 290.00 & 130.00 & 73.00 & 47.00 \\
-  0.036 & 528.00 & 134.00 & 60.00 & 34.00 & 22.00 \\
-  0.056 & 349.00 & 89.00 & 41.00 & 23.00 & 15.00 \\
-  0.076 & 265.00 & 68.00 & 31.00 & 18.00 & 12.00 \\
-  0.096 & 216.00 & 56.00 & 26.00 & 15.00 & 10.00 \\
+0.05 & 6480 & 1656 & 752 & 432 & 282 \\
+  0.07 & 3304 & 852 & 391 & 227 & 149 \\
+  0.09 & 2059 & 536 & 248 & 145 & 97 \\
+  0.11 & 1440 & 379 & 177 & 104 & 70 \\
+  0.13 & 1086 & 288 & 136 & 81 & 55 \\
    \hline
@@
 \includegraphics[width = 0.6\textwidth]{../methods/figs/mem_heatmap_article.png}
-\caption{Measurement error data quality adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.005$ and $FN=0.172$.
-Color scaled so blue = $-10$, white = $0$, and red = $10$.}
+\caption{Imperfect testing adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.005$ and $FN=0.172$. Color scaled so blue = $-10$, white = $0$, and red = $10$.}
 \label{fig:heatmap}
@@
 \end{subfigure}
-\caption{Potential bias due to self-selection and measurement error from a deterministic SIR model with $\beta = 1.4$ (black) and $\gamma = 0.2$ (red).
-Here we assume $f = 0.02$, $FP = 0.01$, $FN = 0.15$, and use a range of relative sampling fraction $M = f_1/f_0$.}
+\caption{Potential bias due to self-selection and measurement error from a deterministic SIR model with $\beta = 1.2$ (black) and $\gamma = 0.15$.
+Here we assume $f = 0.02$, $FP = 0.005$, $FN = 0.172$, and use a range of relative sampling fraction $M = f_1/f_0$.}
 \label{fig:rates}
@@
 \end{subfigure}
-\caption{Left: fraction infected in two SIR models with $\beta = 1.4$ and $0.9$ respectively and $\gamma = 0.2$ with same initial conditions.
-Right: comparison of $\hat R_t$ across time with $FN = 0.2$, $FP = 0.01$, and $M = 4$.}
+\caption{Left: fraction infected in two SIR models with $\beta = 1.2$ and $0.9$ respectively and $\gamma = 0.15$ with same initial conditions.
+Right: comparison of $\hat R_t$ across time with $FN = 0.172$, $FP = 0.005$, and $M = 4$.}
 \label{fig:comparison}

### 2020-06-09 methods/mem_heatmap.R

@@ -1,4 +1,11 @@
+# Construct a heatmap of the imperfect testing 
+# adjustment as a function of the odds ratio and relative fraction
+# for a fixed false positive and false negative rate
+
 memadj <- function(odds_ratio, M, FP, FN) {
+  ## Compute the Measurement error term D_M
+  ## INPUT: Odds ration, relative fraction (M), False Positive rate (FP), and false negative rate (FN)
+  ## OUPUT: Imperfect testing adjustment (D_M)
   prevalence = odds_ratio/(1+odds_ratio)
   firstterm = 1 - (M-1) * odds_ratio * (FP*(1-prevalence) + FN *(prevalence))/((1-prevalence) + M * prevalence)
   secondterm = 1/(1-FP-FN)
@@ -14,7 +21,6 @@ FN = 0.172
 rho = 0.01
 f = 0.001
 
-
 for(i in 1:length(seq_oddsratio)) {
   for (j in 1:length(seq_M)) {
     odds = seq_oddsratio[i]
@@ -28,6 +34,9 @@ row.names(result) = round(seq_oddsratio,3)
 colnames(result) = round(seq_M,2)
 
 library("gplots")
+png(filename = "./figs/mem_heatmap.png",
+    width = 960, height = 960, units = "px", pointsize = 16)
 heatmap.2(as.matrix(result), col = bluered(100), dendrogram='none',
           trace = "none", density.info = "none", Rowv=FALSE,
           Colv=FALSE, key.xlab = "Relative Sampling Frequency", key.ylab = "Odds Ratio")
+dev.off()

### 2020-06-09 methods/sir_example.R

@@ -49,7 +49,7 @@ dev.off()
 f = 0.02
 M = 4
 days = 1:365
-FP = 0.01; FN = 0.1
+FP = 0.005; FN = 0.172
 bary1_new = -diff(out1$S)
 bary1_all = out1$I
 bary2_new = -diff(out2$S)

### 2020-06-09 write-up/scifile.tex
updated a bit of code

@@
 \usepackage{amsmath,amsthm,amsfonts}
+\usepackage{hyperref}
 
@@
 Here we assume the false positive rate (FP) and false negative rate (FN) are known.
-In this paper, we consider the estimator $\tilde y_n = (\bar y_n - FP)/(1-(FP+FN))$ where $\bar y = n^{-1} \sum_{i=1}^n y_i$ is the proportion of positive tests.
+In this paper, we consider the estimator $\tilde y_n = (\bar y_n - FP)/(1-(FP+FN))$ where $\bar y_n = n^{-1} \sum_{i=1}^n y_i$ is the proportion of positive tests in the sample.
 This estimator is unbiased under simple random sampling (SRS); see the supplementary materials for a derivation and discussion of the proposed estimator.
@@
 Figure~\ref{fig:heatmap} shows that $D_M$, as a function of the relative frequency ($f_1/f_0$) and odds ratio, can be both positive and negative as well as a range of magnitudes~\cite{Beesley2020,Beesley2019,Smeden2019}.
-Under perfect testing, $D_M = 1$ so the relation between estimation and selection bias is simple, e.g., COVID-19 positive individuals more likely to receive test implies upward bias in prevalence estimates and vice versa.
+Under perfect testing, $D_M = 1$ so the relation between estimation and selection bias is simple, e.g., COVID-19 positive individuals more likely to receive test implies upward bias in prevalence estimates.
 When tests are imperfect, this simple relationship no longer holds.
@@
 Thus the effective sample size even in optimistic scenarios is equivalent to a small random sample from the population.
-Finally, increased testing capacity may also come at the cost of increases in false positive and negative rates.
+Moreover, increased testing capacity may also come at the cost of increases in false positive and negative rates.
 Consider the case where $f$ increases from $0.05$ to $0.1$ and the increase is associated with the false negative rate rising from $5\%$ to $20$\%.
@@
 where $S_t$ is the number of susceptible individuals, $I_t$ is the number of infected individuals, and $R_t$ is the number of removed (recovered or deceased) individuals at time $t$.
-In terms of bias, the rate is overestimated prior to the peak and underestimated afterwards.
+In terms of bias, the rate is overestimated prior to the peak and underestimated afterwards; the bias increases dramatically when the relative fraction exceeds $2$.
 From a decision making perspective, such biases have a substantial impact.
@@
 Under a Poisson likelihood, a simple relation between the trajectory of new cases and the effective reproduction number can be derived \cite{Bettencourt2008}.
-In particular, under an SIR model the number of case counts on day $t$, denoted $K_t$, is Poisson distributed with rate $K_{t-1} \exp \left( \frac{1}{\gamma} (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval, which is approximately $7$ days for COVID-19~\cite{Sanche2020}.
+In particular, under an SIR model the number of case counts on day $t$, denoted $K_t$, is Poisson distributed with rate $K_{t-1} \exp \left( \gamma (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval, which is approximately $7$ days for COVID-19~\cite{Sanche2020}.
 Using this relation, a moment-based estimator is given by
@@
 When an issue ``cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically''\cite{CoxHink74}.
-Absent random sampling, the best route forward is careful sensitivity analyses and humility in data-driven conclusions.
+Absent random sampling, the best route forward for all data analyses is careful associated sensitivity analyses and humility in data-driven conclusions.
 
@@
 \section*{Supplementary materials}
+
+All relevant code can be found at \url{https://github.com/wdempsey/covid-umich}.
 
@@
 where the second equality is obtained by plugging in the statistical decomposition of the error for both time points and the coefficient of variation being defined as $CV(Y) := \sigma_Y/\mu_Y$.
-Under measurement error, the extra terms $D_{2}$ and $D_1$ can be inserted in the correct locations.
+Under measurement error, the extra terms $D_{t}$ and $D_{t-1}$ can be inserted in the correct locations.
 
@@
 $$
-Then results from Section~\ref{app:ratio} show that the estimate of the number of new cases on day t is given by
+Then the previous sections derivation shows that the estimate of the number of new cases on day t is given by
 $$

### 2020-06-09 methods/sir_example.R

@@ -47,7 +47,7 @@ dev.off()
 
 ## Setup for bias plots
 f = 0.02
-M = 4
+M = 4 
 days = 1:365
 FP = 0.005; FN = 0.172
 bary1_new = -diff(out1$S)
@@ -61,7 +61,7 @@ bias_fn <- function(M, bary, f, FP = 0.01, FN = 0.15) {
   f1 = M*f0
   Delta = f1 - f0
   D_M1 = 1 - Delta * bary/(1-bary) * (FP*(1-bary) + FN * bary)/ (f0 * (1-bary) + f1 * bary)
-  D_M2 = 1 - Delta * bary/(1-bary) * (FP*(1-bary) + FN * bary)/ (f0 * (1-bary) + f1 * bary)
+  D_M2 = 1/(1-FP-FN)
   D_M = D_M1 * D_M2
   rho = Delta*sqrt(bary*(1-bary))/sqrt(f*(1-f))
   error = rho*sqrt((1-f)/f) * sqrt(1-bary)/sqrt(bary)*D_M
@@ -88,15 +88,6 @@ bias1_new = bias_fn(M, bary1_new, f)
 bias2_all = bias_fn(M, bary2_all, f)
 bias2_new = bias_fn(M, bary2_new, f)
 
-## Ratio Comparison: Use All for Prevalence
-
-ymax = max(bias1_all$r+bias1_all$bias, bias1_all$r, bias2_all$r+bias2_all$bias, bias2_all$r)
-ymin = min(bias1_all$r+bias1_all$bias, bias1_all$r, bias2_all$r+bias2_all$bias, bias2_all$r)
-plot(bias1_all$r, type = "l", ylim = c(ymin, ymax))
-lines(bias1_all$r+bias1_all$bias, lty = 2)
-lines(bias2_all$r, col = "red")
-lines(bias2_all$r+bias2_all$bias, col = "red", lty = 2)
-
 ## Ratio Comparison: Use New for Fraction of New Cases
 ymax = max(bias1_new$Rt+bias1_new$logbias, bias1_new$Rt, bias2_new$Rt+bias2_new$logbias, bias2_new$Rt, na.rm = TRUE)
 ymin = min(bias1_new$Rt+bias1_new$logbias, bias1_new$Rt, bias2_new$Rt+bias2_new$logbias, bias2_new$Rt, na.rm = TRUE)
[…]

### 2020-06-10 write-up/covid-writeup-science.tex
Create covid-writeup-science.tex

--- 2020-06-09 write-up/scifile.tex
@@
 
-\title{The Hypothesis of Testing: Paradoxes arising out of reported coronavirus case-counts}
+\title{Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic}
+
+% \title{The Hypothesis of Testing: Paradoxes arising out of reported coronavirus case-counts}
 
@@
 \begin{sciabstract}
-  Many statisticians, epidemiologists, economists and data scientists have registered serious reservations regarding the reported coronavirus case-counts.
-Limited testing capacity across the country has been  identified as a key driver of suppressed coronavirus case-counts.
+  Coronavirus case-counts have influenced governments on their policies and decision making, while researchers have used them to drive their forecast models and predictions.
+Throughout this, many have also cited limited testing as the key driver of weak information on the COVID-19 pandemic.
 The calls to increase testing capacity are well-justified as they become a more frequent point of discussion in the public sphere.
@@
 Moreover, tests are imperfect as false positive/negative rates interact in complex ways with selection bias.
-In this paper, we attempt to clarify this interaction.
+Indeed, measurement error and selection bias are the two greatest problems limiting our understanding of the current COVID-19 pandemic; neither can be fully addressed by increased testing capacity.
+In this paper, we attempt to clarify this point.
 Through simple calculations, we demonstrate pitfalls and paradoxes that can arise when considering case-count data in the presence of selection bias and measurement error.

### 2020-06-10 write-up/covid-writeup-science.tex
finalize submission

@@
 $$
-As of May 8th, the US has performed $8,412,095$ tests.
-The US population is roughly $328$ million, so $f = 0.066$.
-The empirical prevalence is $9.5\%$ on May 8th.
+As of June 10th, the US has performed $21,474,531$ tests.
+The US population is roughly $312$ million, so $f = 0.069$.
+The cumulative empirical prevalence is $9.3\%$ on May 8th.
 A recent study estimates a false positive and negative rate for antibody testing at $0.5$\% and $17.2$\% respectively \cite{Bendavid2020}.
-Supposing COVID-19 positive individuals are $2$ times more likely to get tested, then the effective sample size is $89$.
+Supposing COVID-19 positive individuals are $2$ times more likely to get tested, then the effective sample size is $92$.
 Recent proposals~\cite{Siddarth2020} have argued for increased testing capacity, which may likely reduce the relative sampling rate.
-If the relative sampling rate drops to $1.2$ then the effective sample size will increase to $1879$.
+If the relative sampling rate drops to $1.2$ then the effective sample size will increase to $1950$.
 Thus the effective sample size even in optimistic scenarios is equivalent to a small random sample from the population.
@@
 
-Let $\bar Y_{t-1}$ and $\bar Y_{t}$ denote the prevalence on two consecutive days and consider the estimator $r = \bar y_t / \bar y_{t-1}$.
-Using a second-order Taylor series approximation, the error between ${\bar y_t}/{\bar y_{t-1}}$ and ${\bar Y_{t}}/{\bar Y_{t-1}}$ can be expressed approximately as
+Let $\bar Y_{t-1}$ and $\bar Y_{t}$ denote the prevalence on two consecutive days and consider the estimator $r = \tilde y_t / \tilde y_{t-1}$.
+Using a second-order Taylor series approximation, the error between ${\tilde y_t}/{\tilde y_{t-1}}$ and ${\bar Y_{t}}/{\bar Y_{t-1}}$ can be expressed approximately as
 $$
@@
 Of course, we do not observe $K_t$ and $K_{t-1}$.
-Under SRS of new cases among those susceptible on day $t$, the natural estimator is $S_t \bar y_t$.
+Under SRS of new cases among those susceptible on day $t$, the natural estimator is $S_t \tilde y_t$.
 Unfortunately the number of susceptible individuals on day $t$ is unknown.
-Here, we study the estimator $\hat R_t = 1 + \frac{1}{\gamma} \log \left( \bar y_t / \bar y_{t-1} \right)$.
+Here, we study the estimator $\hat R_t = 1 + \frac{1}{\gamma} \log \left( \tilde y_t / \tilde y_{t-1} \right)$.
 We can again express the statistical error of $\hat R_t - R_t$ in useful terms as follows
@@
 
-%Here you should list the contents of your Supplementary Materials -- below is an example.
-%You should include a list of Supplementary figures, Tables, and any references that appear only in the SM.
-%Note that the reference numbering continues from the main text to the SM.
-% In the example below, Refs. 4-10 were cited only in the SM.
+% %Here you should list the contents of your Supplementary Materials -- below is an example.
+% %You should include a list of Supplementary figures, Tables, and any references that appear only in the SM.
+% %Note that the reference numbering continues from the main text to the SM.
+% % In the example below, Refs. 4-10 were cited only in the SM.
+% \newpage
+
 \section*{Supplementary materials}

### 2020-06-18 write-up/covid-writeup.tex
tex updates
Update covid-writeup.tex

@@
-\documentclass[aoas]{amsart}
+\documentclass[12pt]{amsart}
 
@@
 
+\setlength{\textwidth}{6.5in}
+\setlength{\textheight}{8.5in}
+\setlength{\topmargin}{-0.25in}
+\setlength{\evensidemargin}{0in}
+\setlength{\oddsidemargin}{0in}
+%\setlength{\evensidemargin}{.25in}
+%\setlength{\oddsidemargin}{.25in}
+
+\usepackage{setspace}
+\doublespacing
+
 \def\pr{\text{pr}}
@@
 
-\title[The Hypothesis of Testing]{The Hypothesis of Testing: Paradoxes arising out of reported coronavirus case-counts} %
+\title[Statistical paradoxes in coronavirus case-counts]{Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic} %
 
@@
 \begin{abstract}
-Many statisticians, epidemiologists, economists and data scientists have registered serious reservations regarding the reported coronavirus case-counts.
-Limited testing capacity across the country has been widely identified as a key driver of suppressed coronavirus case-counts.
+  Coronavirus case-counts have influenced governments on their policies and decision making, while researchers have used them to drive their forecast models and predictions.
+Throughout this, many have also cited limited testing as the key driver of weak information on the COVID-19 pandemic.
 The calls to increase testing capacity are well-justified as they become a more frequent point of discussion in the public sphere.
@@
 Moreover, tests are imperfect as false positive/negative rates interact in complex ways with selection bias.
-In this paper, we attempt to clarify this interaction.
+Indeed, measurement error and selection bias are the two greatest problems limiting our understanding of the current COVID-19 pandemic; neither can be fully addressed by increased testing capacity.
+In this paper, we attempt to clarify this point.
 Through simple calculations, we demonstrate pitfalls and paradoxes that can arise when considering case-count data in the presence of selection bias and measurement error.
-The discussion guides several suggestions on how to improve current case-count reporting.
 \end{abstract}
@@
 The World Health Organization has declared the coronavirus disease 2019 (COVID-19) a public health emergency.
-As of April 27th, 2020, a total of 2,993,000 cases have been confirmed worldwide.
-The New York Times reports at least 965,214 people across the United States have tested positive for the virus, and at least 49,465 patients with the virus have died.
-Aggressive policies have been put in place across the US with at least 50\% of the US population officially urged as of late April to stay home via state-wide executive actions.
+As of June 2nd, 2020, over 6 million cases have been confirmed worldwide with 1.8 million cases and over 100 thousand confirmed deaths across the United States.
+This pandemic
+has become the focal point of everyday life; yet the data landscape for understanding COVID-19 remains limited.
+Public databases~\cite{JHU_Lancet,NYT} provide incoming county-level information of confirmed cases and deaths.
+Statisticians, epidemiologists, economists, and data scientists have used this granular data to forecast COVID-19 case-counts, deaths, and hospitalizations~\cite{Giordano2020,Song2020,Ray2020,2020.IHME,Wang2020.03,JTD36385}.
 
-Despite these steps, the data landscape for understanding COVID-19 remains limited.
-Public databases maintained by Johns Hopkins University~\citep{JHU_Lancet} and the New York Times~\cite{NYT} provide incoming county-level information of confirmed cases and deaths.
-Statisticians, epidemiologists, economists, and data scientists have used this granular data to forecast COVID-19 case-counts, deaths, and hospitalizations~\citep{Song2020,Ray2020,2020.IHME}.
-In many cases, variations on the SIR models have been used to draw inferences about infection rates and intensity.
-Parametric inference often uses observed case-count and/or observed COVID-19 related deaths to infer latent trajectories of the pandemic.
+The goal of this paper is to express reservations at the use of case-counts as a proxy for disease prevalence and in estimation of standard epidemiological models for inference and forecasting.
+The reason is straightforward: observed case-count data is plagued by selection bias and measurement error.
+Through simple calculations, we will demonstrate that the information gained from increasing testing capacity is limited in the presence of selection bias and when testing inaccuracies persist.
 
-The goal of this paper is to express reservations at the use of case-counts as a proxy for disease prevalence and intensity as well as its use as direct input into estimation of standard epidemiological models for inference and forecasting.
-The reason is straightforward: current models do not account simultaneously for selection bias and measurement error.
-Selection bias enters due to differences in testing strategies across countries and states.
-In the US, for example, limited testing capacity has caused local and state health departments to focus on testing only high-risk populations.
-Moreover, testing requires the individual to self-select into asking for a test and then receiving approval by their doctor.
-Demands for increased testing capacity, while laudable, ignore the issues of self-selection and measurement error.
-While increasing testing capacity increases data quantity, there is no guarantee for increased data quality.
-We will show that aggressive pushes for ramped up testing capacity that are tied to decreases in data quality may have a mitigated impact  in our ability to estimate quantities of interest such as prevalence and effective reproduction number.
+% \subsection{Outline}
 
-In this paper, we will demonstrate the complex interaction between these two fundamental concepts.
-In classical statistics, measurement error is often associated with parameter attenuation; we will show, however, that the combination with self-selection can cause bias in parameter estimates to change sign and increase/decrease in magnitude.
-Moreover, the bias may depend on the sampling fraction, prevalence, and population size.
-Without a complete understanding of these interactions, we are destined to misinterpret case-counts and come to erroneous conclusions.
-We will demonstrate why there is a strong need for more precision and care in presenting COVID-19 analyses to the broader research and non-academic communities.
-
-% \subsection{Related work}
-
-% There has been an abundance..
-
-% However, most complain about testing \emph{capacity} (cite Nate Silver), there is clearly issues of data quality.
-
-% measurement error discussions.
-% Selection bias . Proponents of this critique suggest well-designed studies.
-
-% Stanford study was Ionides.  They however, forgot about Critique \#2: measurement
-
-\subsection{Outline}
-
-This article discusses the relationship between three statistical concepts: selection bias, measurement error, and the too often forgotten population size.
+% This article discusses the relationship between three statistical concepts: selection bias, measurement error, and the too often forgotten population size.
 We clarify mathematically why the situation is much more complex than it first appears.
@@
 Here, like in survey methodology~\citep{Cochran77}, we treat COVID-19 status as a fixed but unknown quantity of interest.
-For simplicity, we start by ignoring the dynamic nature of the viral outbreak as well as the fact that individuals can recover from the disease and assume either individual $j$ is COVID-19 positive and $Y_j=1$ or is COVID-19 negative and $Y_j=0$.
+For simplicity, the dynamic nature of the outbreak and recoverability of individuals are initially ignored.
+We assume either individual $j$ is COVID-19 positive and $Y_j=1$ or is COVID-19 negative and $Y_j=0$.
 We also let $I_j \in \{0,1\}$ be an indicator that the individual was selected for testing ($I = 1$) or not ($I=0$).
@@
 
-Here we highlight two key insights from~\cite{Meng2018} that are relevant to the COVID-19 crisis.  First, comparing the mean-squared error under selection mechanism $\I$ and SRS, we see that
-$$
-\frac{E_{\I} \left[ (\bar y_n - \bar Y)^2 \right]}{\sqrt{V_{SRS} (\bar Y)}} = (N-1) E_{\I} \left[ \rho_{I,Y}^2 \right].
-$$
-This points to a troubling and paradoxical situation: the error relative to SRS increases as a function of population size.
-\cite{Meng2018} termed this the ``Law of Large Populations'' (LLP).
-This points to a critical issue in the current media practices in communication of case-count data: two countries with the same testing strategy (i.e., $E_{\I} [\rho_{I,Y} ]$ equal) can yield wildly different estimates due to population size.
-Large countries like the US may have similar true prevalence rates as smaller countries like the UK.
-Even under similar testing strategies, the mean-squared error in the prevalence rate for the US will have substantially more variation in comparison to SRS.
-Comparing the US to the UK, for example, the MSE will increase by a factor of almost five.
-Thus conclusions drawn from observed case-count records may be not just wrong, but very wrong.
-
-\begin{wrapfigure}{r}{0.5\textwidth}
-\centering
-\includegraphics[width = 0.4\textwidth]{../methods/figs/US-casecount.png}
-\caption{US total tests and total positive tests per day}
-\label{fig:ustests}
-\vspace{-0.3cm}
-\end{wrapfigure}
-
-Second, there have been calls for increased testing.
-While important, many conflate increased testing capacity with increased quality of testing.
-We will discuss this point further in Section \ref{section:discussion}.
-As of May 8th, the US has performed $8,412,095$ tests in total.
-Figure~\ref{fig:ustests} shows the trajectory of testing per day and positive cases per day in the US.
-The US population is roughly $328$ million, meaning the fraction sampled is $f = 0.026$.
-After smoothing, the empirical prevalence is $9.1\%$ on May 8th.
-Supposing COVID-19 positive individuals are $2$ times more likely to get tested than those individuals who are COVID-19 negative, the question is ``What is the sample size from a SRS that would yield equivalent MSE in the estimated prevalence?'' Using the above MSE result, Meng showed the effective sample size is equal to $n_{eff} \leq \frac{1}{E_I [ \rho_{I,Y}^2 ]} \frac{f}{1-f}$.
-Under these parameters, the effective sample size is $14$.
-Recent proposals~\citep{Siddarth2020} have argued in favor of increasing overall testing capacity.
-With more tests available, the relative sampling rate may reduce.
-Supposing it drops to $1.2$ then the effective sample size will increase to $299$.
-See Appendix~\ref{section:effss} for the mathematical details and calculations in additional scenarios.
-The above calculations point to a sad state of affairs, saying that the effective sample size even in best case scenarios is not better than a small random sample from the population.
-
-The remainder of this paper aims to build upon these fundamental insights by extending the decomposition in two directions: accounting for  measurement error and the temporal nature of the pandemic.
+The remainder of this paper aims to build upon this fundamental work by extending the decomposition in two directions: accounting for measurement error and the temporal nature of the pandemic.
 We then discuss effective sampling methods and their importance in the current pandemic.
@@
 
+\subsubsection{Connecting true and observed data quality}
+
 We start by considering the first two terms and assess whether the sign of the bias can reverse due to the interaction of measurement error and selection bias.
@@
 
-While prior investigations have noted the interaction between measurement error and selection bias~\citep{Beesley2020,Beesley2019,Smeden2019}, the interaction with the sample size relative to the population, i.e., $f$, has largely been ignored.
+% While prior investigations have noted the interaction between measurement error and selection bias~\citep{Beesley2020,Beesley2019,Smeden2019}, the interaction with the sample size relative to the population, i.e., $f$, has largely been ignored.
 The above statistical decomposition clarifies the importance of this quantity~$f$.
@@
 
-Consider again the current COVID-19 pandemic.
+% Consider again the current COVID-19 pandemic.
 For now, we continue to assume the ratio of conditional selection rates $f_1/f_0$ is equal to 1.5.
@@
 
-The attentive data analyst will recognize the estimator $\bar y_n$ is biased even for simple random samples and, if sensitivity and specificity were known a priori, may suggest the alternative estimator $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ which is unbiased under simple random sampling.
-We again wish to express the error $\tilde  y_n - \bar Y$ in statistical terms.
-In the appendix, we show that the error now can be expressed as
+\subsection{Deriving an unbiased estimator}
+
+The attentive data analyst will recognize the estimator $\bar y_n$ is biased even for simple random samples and, if sensitivity and specificity were known a priori, may use the error decomposition to construct an unbiased estimator $\tilde y_n$.
+
+In this paper, we consider the estimator $\tilde y_n = (\bar y_n - FP)/(1-(FP+FN))$ where $\bar y_n = n^{-1} \sum_{i=1}^n y_i$ is the proportion of positive tests in the sample.
+This estimator is unbiased under simple random sampling (SRS); see the supplementary materials for a derivation and discussion of the proposed estimator.
+To understand the impact of selection bias and imperfect testing, we derive the following statistical decomposition of the error between $\tilde y_n$ and $\bar Y$:
 \begin{equation}
@@
 \rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
-\times \underbrace{\left[ 1 + FP + FN - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right]}_{D_M}.
+\times \underbrace{\left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{D_M},
 \end{equation}
-The first term is the same as before but increased by $(1 + FP + FN)$ to account for the additional uncertainty due to measurement error.
-The second term is the interaction between selection bias and measurement error.
-Figure~\ref{fig:heatmap} presents the measurement error data quality adjustment as a function of the relative frequency and odds ratio.
-We see that the adjustment can be both positive and negative as well as a range of magnitudes.
+where $f=n/N$ is the sampling fraction, $f_1=n_1/N_1$ and $f_0=n_0/N_0=(n-n_1)/(N-N_1)$ are sampling fractions for COVID-19 positive and negative individuals respectively, and $\Delta = f_1 - f_0$ is the sampling rate differential.
+See the supplementary materials for the derivation.
+This extends work by \cite{Meng2018} to account for imperfect testing.
+The first term represents \emph{data quality}, the second \emph{data quantity}, and the third \emph{problem difficulty}.
+The final term represents the \emph{imperfect testing adjustment} which is a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence.
+Figure~\ref{fig:heatmap} shows that $D_M$, as a function of the relative frequency ($f_1/f_0$) and odds ratio, can be both positive and negative as well as a range of magnitudes~\cite{Beesley2020,Beesley2019,Smeden2019}.
+Under perfect testing, $D_M = 1$ so the relation between estimation and selection bias is simple, e.g., COVID-19 positive individuals more likely to receive test implies upward bias in prevalence estimates.
+When tests are imperfect, this simple relationship no longer holds.
 
-We next investigate the impact of measurement error on the effective sample size calculation.  The effective sample size can now be bounded by
+Comparing the mean-squared error (MSE) under a selection mechanism $\I$ with imperfect testing and SRS with perfect testing, we see that
 $$
-n_{eff} \le \frac{f}{1-f} \times \frac{1}{E_{I} \left[ \rho_{I,Y}^2 D_M^2 \right]}.
+\frac{E_{\I} \left[ (\bar y_n - \bar Y)^2 \right]}{\sqrt{V_{SRS} (\bar Y)}} = (N-1) E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right]
 $$
-In Section \ref{section:est_dq}, we discuss recent studies that estimate the COVID-19 false positive rate of $0.5$\% and false negative rate $17.2$\%.
-Supposing COVID-19 positive individuals are $2$ times more likely to get tested than those individuals who are COVID-19 negative, the effective sample size drops from $14$ to $10$.
-If the relative sampling fraction drops to $1.2$ then the effective sample size drops from $299$ to $216$.
+implying the error relative to SRS increases as a function of population size~\cite{Meng2018}.
+The paradox that two countries with the same testing strategy can yield wildly different estimates due to population size is too often ignored in current communications of case-count data.
+When comparing the US and UK under the assumption of similar testing strategies, for example, the MSE increases by a factor of five.
+Thus, conclusions drawn from observed case-counts may not just be wrong, but very wrong.
 
@@
 
-The above argument shows that the effective sample size is affected in a complex manner by the interaction between selection bias and measurement error.
-Here, we discuss the impact of the interaction on the effective sample size when testing capacity is tied to measurement error.
-In the current COVID-19 pandemic, there have been well-justified calls for increases in testing capacity.
-These increases, however, may come at the cost of increases in false-positive and false-negative rates.
-In terms of effective sample size, if the data quality term remains constant, then testing increases may not significantly improve the effective sample size.
-Consider the simplest case where $f$ increases from $0.05$ to $0.1$; suppose this increase was associated with  false positive rate increasing from $0.5$\% to $5$\% and the false negative rate rising from $5\%$ to $20$\%.
-Then the standard analysis would suggest a reduction in the MSE by 47\% but the actual decrease is by 26\%.
-This implies that doubling the sampling rate will only increase the effective sample size by a factor $2.9$ rather than the expected $4$ fold increase that is expected.
-In a worst case scenario, if the false negative rate climbs to $0.30$ and the false positive rate to $0.10$ then the factor effective sample size increases by only a factor of $2.3$.
+A key question is ``What is the (effective) sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' In the supplement, we show the effective sample size $n_{eff}$ can be bounded by
+$$
+\frac{f}{1-f} \times \frac{1}{E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right]}.
+$$
+As of June 10th, the US has performed $21,474,531$ tests.
+The US population is roughly $312$ million, so $f = 0.069$.
+The cumulative empirical prevalence is $9.3\%$ on May 8th.
+A recent study estimates a false positive and negative rate for antibody testing at $0.5$\% and $17.2$\% respectively \cite{Bendavid2020}.
+Supposing COVID-19 positive individuals are $2$ times more likely to get tested, then the effective sample size is $92$.
+Recent proposals~\cite{Siddarth2020} have argued for increased testing capacity, which may likely reduce the relative sampling rate.
+If the relative sampling rate drops to $1.2$ then the effective sample size will increase to $1950$.
+Thus the effective sample size even in optimistic scenarios is equivalent to a small random sample from the population.
+Moreover, increased testing capacity may also come at the cost of increases in false positive and negative rates.
+Consider the case where $f$ increases from $0.05$ to $0.1$ and the increase is associated with the false negative rate rising from $5\%$ to $20$\%.
+Then the effective sample size increases by a factor of $1.76$ rather than the expected $4$ fold increase.
 
@@
 
-\subsection{Case count comparisons}
-Two alternative comparisons are common.
-First, the data analyst may directly compare the observed number of new cases on a given day.
-If $Y_1$ and $Y_2$ are the number of cases in locations $j=1,2$ respectively, then we can express the error of $y_1 - y_2$ as
-$$
-\begin{aligned}
-y_1 - y_2 &= n_1 ( \bar y_1 - \bar Y_{1}) - n_2 ( \bar y_2 - \bar Y_{2}) + \left( \frac{n_1}{N_1} Y_{1} - \frac{n_2}{N_2} Y_2  \right) \\
-&= \left[ n_1 \sigma_{Y_1} \rho_{I_1, Y_1} \sqrt{\frac{1-f_1}{f_1}} D_{M_1} - n_2 \sigma_{Y_2} \rho_{I_2, Y_2} \sqrt{\frac{1-f_2}{f_2}} D_{M_2}  \right] + \left[ f_1 Y_1 - f_2 Y_2 \right].
-\end{aligned}
-$$
-Under simple random sampling, the first term is zero in expectation.
-However, the question is whether $f_1 \bar Y_1 - f_2 \bar Y_2$ is of scientific interest.
-We would argue that this is only of true interest when $f_1 = f_2$ and we are interested in comparing total number of cases.
-In this setting, however, the difference in error terms is unlikely to be zero.
-To see this, assume that we have similar expected data quality, data quantity, problem difficulty, and measurement error.
-Then the error is a function of $n_1 - n_2 = n_2 \left( \frac{N_1}{N_2} - 1 \right)$.
-So we see that the error now is a function of the relative difference in the population sizes.
-
-Alternatively, the data analyst may be wary of such concerns in population differences and scale counts by respective population size, i.e., $y_1/N_1 - y_2/ N_2$.  We can again express the error as
-$$
-\begin{aligned}
-\frac{y_1}{N_1} - \frac{y_2}{N_2} &= \frac{n_1}{N_1} ( \bar y_1 - \bar Y_{1}) - \frac{n_2}{N_2} ( \bar y_2 - \bar Y_{2}) + \left( \frac{n_1}{N_1} \bar Y_{1} - \frac{n_2}{N_2} \bar Y_2  \right) \\
-&= \left[ \sigma_{Y_1} \rho_{I_1, Y_1} \sqrt{ f_1 (1-f_1) } D_{M_1} - \sigma_{Y_2} \rho_{I_2, Y_2} \sqrt{f_2 (1-f_2)} D_{M_2} \right] + \left[ f_1 \bar Y_1 - f_2 \bar Y_2 \right].
-\end{aligned}
-$$
-Under simple random sampling, the first term is again zero in expectation.
-However, the question is whether $f_1 \bar Y_1 - f_2 \bar Y_2$ is of scientific interest.
-We would argue that this is only of true interest when $f_1 = f_2$ and we wish to compare disease prevalence.
-Importantly, the error in this comparison is less problematic.
-To see this, assume that we have similar expected data quality, data quantity, problem difficulty, and measurement error.
-Then the error is equal to zero.
-
-It is important to note that the standard law of large populations caveats apply to both comparisons.
-In both cases, the point is that we must ascertain what is the population-level contrast and whether this should be of genuine scientific interest.
-Otherwise, we are comparing apples and oranges.
-Not only this, but the two comparisons come with different levels of error induced by selection bias and measurement error.
-
 \subsection{Rate comparisons}
@@
 Here, these errors can mingle in interesting ways.
-Consider two countries (A and B) in which the peak occurs 2 weeks prior for country A then country B but other.
+Consider two countries (A and B) in which the peak occurs 2 weeks prior for country A than country B.
 Figure~\ref{fig:comparison} presents such a comparison where each country's disease trajectory follows an SIR model as described in Section~\ref{section:rates} with different parameters.

### 2020-06-18 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 
-The data analyst, now wary of estimating prevalence and total counts, pauses and thinks.
-They return shortly thereafter, with a follow-up consideration: \emph{Ok, perhaps estimating prevalence is difficult.
-Certainly, however, we can estimate the rate of growth.
-All I want to know is when we hit the point at which the curve flattens and number of deaths decrease.
-That can't be too hard, surely!}
+The prior analysis demonstrates the limited information regarding COVID-19 prevalence in observational case-count data.
+Scientists, however, may claim that daily observed case-counts simply undercount daily total cases by a constant multiple over time.
+If true then the ratio of case-counts at consecutive times may be a good estimate of the true change in prevalence, helping scientists understand the disease trajectory.
+In this section, we demonstrate how selection bias and imperfect testing impact such estimates.
 
-Unfortunately, ratio estimators do not cancel errors as now both the numerator and the denominator are uncertain.
-This means selection bias and measurement error can have paradoxical effects.
-Here we consider the ratio estimators for the relative change in the prevalence rate.
-We let $\bar Y_{t-1}$ and $\bar Y_{t}$ denotes the prevalence (across the entire population) at time step $t-1$ and $t$ respectively (i.e., prevalence on two consecutive days).
-Then the ratio estimator is given by $r = \bar y_t / \bar y_{t-1}$.
-We assume both numerator and denominator are the prevalence estimates adjusted for measurement error.
-If the sample size at each time were equal (i.e., $n_1 = n_2 = n$) then this analysis would be equivalent to comparing the increase in observed case-counts (i.e., $\bar y_t/\bar y_{t-1} = y_t/y_{t-1}$).
-We wish to express the error $\frac{\bar y_t}{\bar y_{t-1}} - \frac{\bar Y_{t}}{\bar Y_{t-1}}$ in statistical terms.
-Using a Taylor series approximation, we show in the appendix that the error can be expressed approximately as
+Let $\bar Y_{t-1}$ and $\bar Y_{t}$ denote the prevalence on two consecutive days and consider the estimator $r = \tilde y_t / \tilde y_{t-1}$.
+Using a second-order Taylor series approximation, the error between ${\tilde y_t}/{\tilde y_{t-1}}$ and ${\bar Y_{t}}/{\bar Y_{t-1}}$ can be expressed approximately as
 $$
@@
 $$
-where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement error adjustment, and $CV(Y_j) = \sigma_{Y_j}/\bar Y_j$ is the coefficient of variation at time step $j$.
-The magnitude of the error depends on the population-level ratio $\bar Y_{t} / \bar Y_{t-1}$ so a large decrease in the prevalence rate will have a relatively small error as compared to a large increase in the prevalence rate.
-The second term represents the ``cancellation'' that the naive analyst is hoping will occur.
-The analyst's intuition only holds when the difference is equal to zero.
-This occurs at times when data quality, sampling fraction, measurement error, and prevalence are all constant across time.
+where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement error adjustment, and $CV(Y_j) = \sigma_{Y_j}/Y_j$ is the coefficient of variation on day $j$.
+See the supplementary materials for the derivation.
+The error magnitude depends on the true rate $\bar Y_{t} / \bar Y_{t-1}$ so a large decrease will have a relatively small error compared to a large increase.
+The second term represents potential \emph{cancellation} which may occur when data quality, sampling fraction, measurement error, and prevalence are all constant across time.
 
-Consider the state of New York.
-Figure~\ref{fig:ny-covid-test} shows the number of tests and positive tests per day in New York state.
-Figure~\ref{fig:ny-covid-frac} presents the fraction of tests that were positive per day in New York state.
-For cancellation, we would need the expected data qualities to be uncorrelated and equal.
-Testing appears relatively stable over a week which suggests assuming $\Delta_t$ (i.e., differential in the selection rate) varies slowly over time may be a plausible assumption.
-Note, however, that constant $\Delta_t$ does not imply constant data quality (see Appendix \ref{section:effss} for details).
-Figure~\ref{fig:ny-covid-test} shows the number of tests has remained relatively constant in the last few weeks, specifically relative to the total NY population, assuming $f_t$ also varies slowly in time is reasonable.
-Finally, the measurement error adjustment is likely to be constant as no new testing procedures have been reported.
-This would not be true in states like Michigan, where testing criteria have been expanding every few weeks.
-The final component is an adjustment to this differential error based on the statistical error at the first time point.
-
-Figure \ref{fig:ratio-bias} shows a trajectory of the true ratio and the potential biased estimators under an SIR model for the epidemic dynamics, with state evolution given by the following ODE
+Figure~\ref{fig:ratio-bias} displays the trajectory of the true ratio and the potential biased estimators under an SIR model~\cite{Pastor2001,Newman2002,Parshani2010} for the epidemic dynamics, with state evolution given by
 $$
@@
 $$
-where $S_t$ is the number of susceptible individuals at time $t$, $I_t$ is the number of infected individuals at time $t$, $N$ is the total population size, and $R_t$ is the number of removed (recovered or deceased) individuals at time $t$.
-Figure~\ref{fig:fracpop} presents the fraction of population newly infected individuals at each time step for $\beta = 1.4$ and $\gamma = 0.2$ (black line).
-In each case, the rate of change in prevalence is likely to be  overestimated prior to the peak and then underestimated afterwards.
-From a decision making perspective, such biases have potential impact.
-First, the over-estimation may give lawmakers and governors more leverage in proposing aggressive actions that reduce prevalence.
-Of course, the analysis supports the argument that the estimates based on available data are overly pessimistic.
-What both sides miss is that the direction of bias is non-constant over time.
-After we hit the peak, estimates will likely under-estimate prevalence.
-Such bias puts pressure on lawmakers and governors to prematurely relax social distancing measures.
-Moreover, the question is ``Can we trust the observed data to let us know if we have reached the peak?''  It appears that the peak time is the easiest to estimate in terms of having minimal error using availability data.
-Note, however, that the standard errors on such estimates will be incredibly wide due to the low effective sample size as discussed in Section~\ref{section:casecount}.
+where $S_t$ is the number of susceptible individuals, $I_t$ is the number of infected individuals, and $R_t$ is the number of removed (recovered or deceased) individuals at time $t$.
+In terms of bias, the rate is overestimated prior to the peak and underestimated afterwards; the bias increases dramatically when the relative fraction exceeds $2$.
+From a decision making perspective, such biases have a substantial impact.
+First, the overestimation may give lawmakers and governors more leverage in proposing aggressive actions that reduce prevalence.
+Of course, the analysis supports the argument that the estimates based on available data may be pessimistic.
+What is missing in the current public health discourse is that the bias direction of the prevalence estimate is non-constant over time.
+Underestimation post-peak puts pressure on lawmakers and governors to prematurely relax social distancing measures.
+Moreover, it appears that the peak time is the easiest to estimate in terms of having minimal error using available data.
+Note, however, that the standard errors will be large due to small effective sample sizes.
 
@@
 \label{section:r0-estimation}
-Many well-respected epidemiologists argue that tracking the effective reproduction number $R_t$ is the only way to manage through the crisis~\citep{Gabriel2020}.
-We show here how these estimates are also impacted by selection bias and measurement error.
-For simplicity, we again assume the estimates are adjusted for known false positive and false negative rates.
-\cite{Bettencourt2008} show that under a Poisson likelihood, a simple relation between the trajectory of new cases and the effective reproduction number can be derived.
-In particular, under an SIR model, the number of case counts on day $t$ is Poisson distributed with rate $K_{t-1} \exp \left( \frac{1}{\gamma} (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval which is approximately $7$ days for COVID-19~\citep{Sanche2020}.
-Using this, a simple estimator of the effective reproduction number at time $t$ can be given by
+Many epidemiologists argue that tracking the effective reproduction number $R_t$ is the only way to manage through the crisis~\cite{Gabriel2020}.
+Under a Poisson likelihood, a simple relation between the trajectory of new cases and the effective reproduction number can be derived \cite{Bettencourt2008}.
+In particular, under an SIR model the number of case counts on day $t$, denoted $K_t$, is Poisson distributed with rate $K_{t-1} \exp \left( \gamma (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval, which is approximately $7$ days for COVID-19~\cite{Sanche2020}.
+Using this relation, a moment-based estimator is given by
 $$
@@
 $$
-Of course, we do not observe $K_t$ and $K_{t-1}$ but noisy proxies $y_t$ and $y_{t-1}$, i.e., the observed case-count on day $t$ and $t-1$ respectively.
-The hope is that $y_t/y_{t-1}$ is a good proxy for $K_t/K_{t-1}$ where here $y_t$ is conceptualized as the number of new cases on day $t$.
-However, if we think about estimating number $K_t$ under SRS of new cases among those susceptible on day $t$, the natural estimator is $S_t \bar y_t$.
-Unfortunately the number of individuals susceptible on day $t$ is unknown.
-Here, we assume that $\hat R_t = 1 + \frac{1}{\gamma} \log \left( \bar y_t / \bar y_{t-1} \right)$ is the estimated effective reproduction number on day $t$.
-This estimator adjusts for sample size variation across days but not the varying population size.
-The estimator is close in spirit to the estimator derived under assuming the process is fully observed.
-Similar to Section~\ref{section:rates}, the proxy has error in both numerator and denominator that may cause issues.
-We again wish to express the statistical error of $\hat R_t - R_t$ in useful terms.
-We can re-arrange the error decomposition from Section~\ref{section:rates} to show that this error is given by
+Of course, we do not observe $K_t$ and $K_{t-1}$.
+Under SRS of new cases among those susceptible on day $t$, the natural estimator is $S_t \tilde y_t$.
+Unfortunately the number of susceptible individuals on day $t$ is unknown.
+Here, we study the estimator $\hat R_t = 1 + \frac{1}{\gamma} \log \left( \tilde y_t / \tilde y_{t-1} \right)$.
+We can again express the statistical error of $\hat R_t - R_t$ in useful terms as follows
 $$
@@
 $$
-So here we see the exact same trade-off as with the rates but on the logarithmic scale.
-For small values of $x$, we have $\log(1+x) = x$ and so the biases are similar.
-However, the error is no longer scaled by the rate of change in prevalence $\bar Y_2/\bar Y_1$.
-Instead the error is scaled by the serial interval $\gamma^{-1} = 1/7$.
-Moreover, the error depends not on aggregate prevalence $\bar Y_t$ but fraction of new cases $\bar K_t$.
-The final term is the error due to varying population sizes.
-Since $S_t \leq S_{t-1}$, the final term $- \log (S_t/S_{t-1}) \geq 0$.
-Figure~\ref{fig:r0-bias} displays the bias for the effective reproductive number and how it is impacted by the relative sampling fraction.
-Here for simplicity, we assume the fraction of the population that is susceptible remains large and near the total population size.
-Under these assumption, the potential bias is similar to that for the ratio estimator adjusted to be on the logarithm scale.
-Changes to this assumption can impact the potential bias.
-The key difference between Figure~\ref{fig:ratio-bias} and Figure~\ref{fig:r0-bias} is that prevalence depends on the fraction of the population infected at time $t$ while $\hat R_t$ depends on the fraction of new cases in the population.
+This implies a similar trade-off as before but on the logarithmic scale.
+The error is no longer scaled by $\bar Y_t/\bar Y_{t-1}$ but by the serial interval and does not depend on prevalence $\bar Y_t$ but on the fraction of new cases out of those susceptible $\bar K_t$.
 This leads to differences in when the bias is most pronounced.
+The final term is the error due to varying numbers of susceptible individuals.
+Figure~\ref{fig:r0-bias} displays the bias as a function of the relative sampling fraction assuming the fraction of the population that is susceptible remains large.
 
@@
 \section{Cross-population comparisons}
-
-So far we have focused on understanding the limitations of using case-count data to understand population quantities of interest for a \emph{single} population (i.e., United States, Germany, France or Michigan, New York, London in isolation) at a single moment and over time.
-Many are interested in cross-population comparisons.
+So far we have focused on understanding the limitations of using case-count data to understand population quantities of interest for a \emph{single} population.
+Many are interested in cross-population comparisons to contrast the impact of countries' mitigation policies.
 Many pieces in the media as well as academic articles have plots of case-count over time aligned by time of first known case.
@@
 In this section, we focus on the statistical issues relevant to such comparisons.
-We do not make any comments on the multi-causal nature of success/failure in countries.
 
@@
 Consider two countries (A and B) in which the peak occurs 2 weeks prior for country A than country B.
-Figure~\ref{fig:comparison} presents such a comparison where each country's disease trajectory follows an SIR model as described in Section~\ref{section:rates} with different parameters.
-Figure \ref{fig:eff} shows how the errors interact in complex ways.
-First, the gap is correctly estimated; then the gap is over-estimated as country A sees a rapid rise in cases; then the gap is even more over-estimated as country A improves and the rate is under-estimated while country B sees a rapid increase in their case-count and their rate is over-estimated; then country A's rate is correctly estimated while country B's rate is under-estimated as it sees improvement in infection rates; finally, the gap disappears.
-We are not claiming this will always be the case, and indeed for small enough selection bias and/or large gaps in case-counts such complex behaviour will not occur; however, this is simply to point out how observed information can tell a more complex story than the truth (i.e., country A recovers 1 month prior to country B).
+Figure~\ref{fig:comparison} presents such a comparison where each country's disease trajectory follows an SIR model (A=black and B=red).
+Of interest is the difference between the effective reproduction rate for country A and B, $R_{tA} - R_{tB}$.
+Figure~\ref{fig:eff} shows how biases in estimation of each rate interact in complex ways.
+At first, the difference is correctly estimated; then the gap is over-estimated as country A sees a rapid rise in cases; then the magnitude of over-estimation increases as country A sees declining case-count while country B sees rapidly increasing case-count; then country A's rate is correctly estimated while country B's rate is under-estimated as it sees declining case-count; finally, the gap disappears.
+We do not claim this will always be the case; however, this demonstrates how observed information can tell a more complex story than the truth (i.e., country A's peak is 2 weeks prior to country B's peak).
 
@@
 
-In prior sections, we have discussed the impact of selection bias and measurement error on a range of outcomes of interest in COVID-19 research.
-Here, we try and address the empirical question of how to estimate the data quality $E_{\I} \left[ \rho_{I,Y} \right]$ in the current context.
-In \cite{Meng2018}, estimation of the data quality relied on observing the true outcome.
-In election data, for example, we observe the vote totals after the fact and can use this as ground truth.
-We are not so lucky in the current crisis.
-Here, we propose a simple procedure that uses survey samples as noisy but unbiased estimates of the true prevalence.
-Our goal is less inferential and more to provide a sensitivity analysis that can aid in understanding the amount of information in observational COVID case-count data.
+We next consider estimation of data quality $E_{\I} \left[ \rho_{I,Y} \right]$.
+In \cite{Meng2018}, estimation relied on observing the true outcome (i.e., election vote totals), which  is not possible in the current crisis.
+Here, we propose using surveys as noisy but unbiased estimates of the true prevalence.
+Our goal is not inferential.
+Instead, we aim to build sensitivity analyses that can aid in understanding the amount of information in observational COVID case-count data.
 
-On April 23rd, Andrew Cuomo announced results from a study in New York state.
+On April 23rd, Governor Andrew Cuomo announced results from a study in New York state.
 It found 13.9\% of 3,000 people tested across the state had signs of the virus.
-The study did not report sensitivity and specificity; therefore, we take reported measures from the Santa Clara study~\citep{Bendavid2020} \footnote{We recognize the debate regarding its conclusions.
-Here we are only using this study for its information regarding measurement error rates.}.
+The study did not report sensitivity and specificity; therefore, we take reported measures from the Santa Clara study~\cite{Bendavid2020}.
 The reported specificity is $82.8\%$ (95\% CI: 76.0-88.4\%) and sensitivity is $99.5\%$ (95\% CI: 99.2-99.7\%).
 This corresponds to a false negative rate of $17.2\%$ (11.6-24.0\%) and false positive rate of $0.5\%$ (0.3 - 0.8\%).
-Correcting for these rates we have an estimated prevalence of $15.9\%$ with a range of 14.8\% to 17.0\% using the ranges of the false positive and negative rates.
+The estimated prevalence is then $16.3\%$ with a range of 15.0\% to 18.0\%.
+For simplicity, we assume the study was conducted on April 20th.
+On the 20th, $19,654$ tests were performed.
+The population of New York is $19,378,102$.
+Subtracting off the $617,555$ tests that had already performed yields a sampling fraction of $f = 0.001$.
+The smoothed prevalence estimate is 28.4\%.
+Adjusting for imperfect testing yields an estimate of $33.9\%$.
 
-Figure~\ref{fig:ny-covid-test} plots the number of new total and positive tests per day.
-Figure~\ref{fig:ny-covid} plots the fraction of tests that were positiver per day.
-In both cases, we applied an exponential smoothing to the reported values because the spiky pattern is likely due to testing backlogs.
-Given the report was on the 23rd, we suppose for simplicity that the study was run several days prior on the 20th of April.
-On the 20th, there were $19,654$ tests performed and $617,555$ tests had been performed up to to and including the 19th.
-The population of New York state is $19,378,102$.
-Subtracting off the number of individuals who had already been tested yields a sampling fraction of $f = 0.001$ on the 20th.
-The smoothed prevalence estimate on the 20th is 28.3\%.
-Adjusting for measurement error yields an estimate of $32.5\%$.
-
-Here, we assume the New York survey is a simple random sample from the state's population.
-We recognize that the survey is not such a sample, but techniques like poststratification and raking can be applied to calibrate estimates.
-Then the error is $16.6$\%.
-Using~\eqref{eq:error}, we can use the error to construct an estimate of the relative sampling rate:
+Under the assumption that the survey is a SRS from New York, the error is $17.6$\%.  Using~\eqref{eq:error}, we construct an estimate of the relative sampling rate as follows
 \[
-\rho D_M = \sqrt{\frac{f}{1-f}} \frac{\text{0.166}}{\sigma_Y} = 1.43 \times 10^{-2} \Rightarrow \Delta = 1.06 \times 10^{-3} \Rightarrow M = 2.29.
+\rho D_M = \sqrt{\frac{f}{1-f}} \frac{\text{0.176}}{\sigma_Y} = 1.07 \times 10^{-2} \Rightarrow \Delta = 1.07 \times 10^{-3} \Rightarrow M = 2.3.
 \]
-The range for the CIs is (2.05, 2.54).
-We can also perform sensitivity as a function of the false negative and positive rates.
-Under these measurement error ranges, we saw that the impact on the effective reproductive rate estimator was much smaller.
-However, for $FP = 0.05$ and $FN = 0.005$ we have $M = 4.31$.
-While nothing above is definitive, it points to the interplay of selection bias and measurement error.
-This calculation gives us simple tools to understand how sensitivity and specificity impact our estimators and therefore how much we should trust conclusions based on these assumptions.
-Under the Santa Clara study estimates of these quantities, while we expect impact, the bias is much smaller than under other estimates.
+We can perform a sensitivity analysis by considering the range of false negative and positive rates from~\cite{Bendavid2020}, which leads to a range for $M$ of $(2.21, 2.38)$.
+Therefore, the impact on bias of $\hat R_t$ appears moderate.
+If $FP = 0.05$ and $FN = 0.005$, then the relative sampling rate increases to $M=3.04$.
+This simple tool helps us understand how sensitivity and specificity impact biases and  therefore how much we should trust conclusions based on these assumptions.
 
@@
 
-\section{Stratified sampling: improving precision in low-prevalence environments}
-
-This article argues why random sampling is a powerful tool and the limitations of data obtained through a self-selection process.
-As COVID-19 survey studies become more readily accessible, we encourage researchers to focus on building representative samples.
-One difficulty in the current crisis is the low prevalence relative to the population size in certain geographies.
-Take the recent Santa Clara study, for example, that estimated less than 2\% prevalence in the area.
-The key issue with low prevalence is that if the test's specificity is not lower than the prevalence, then the observed data is consistent with zero recovered individuals.
-
-In such settings, simple random sampling is doomed to fail; however, there are potential solutions.
-In particular, one can stratify the population into risk categories (based on population density, occupation, age, and other important factors) and perform Neyman allocation~\citep{Cochran77}.
-That is, sample individuals at random in strata $h$ with size proportional to $P_h \sqrt{\bar Y_h \times (1-\bar Y_h)}$ where $P_h$ is the strata frequency in the population and $\bar Y_h$ the prevalence in the strata.
-Of course, $P_h$ and $\bar Y_h$ are not known a priori; however, if decent estimates can be obtained, we can see huge benefits in settings where the prevalence varies widely by strata.
-One could go attempt to go further and incorporate contact network information.
-However, such data is difficult to obtain in practice.
-Indeed, geographic location, occupation, age and basic demographic information are much easier to obtain and are a good proxy for the latent network connectivity.
-
-% \section{Decision-making: data versus information.}
-
-% Having read through the above technical discussions, one may come to the incorrect conclusion that because we are concerned about implicit biases in observational data, we must wait for better data sources in order to act~\citep{Ioannidis2020}.
-This conclusion confuses data and information.
-Lack of high quality data means we should be skeptical regarding conclusions drawn from these data sources alone.
-There are whole fields dedicated to careful analysis in such settings (e.g., causal inference from observational studies).
-Being a data skeptic, does not mean one must wait for high quality data to act, especially in high risk scenarios.
-One aims to have high quality information.
-Yet, in the current crisis, we have just that.
-Indeed, epidemiologists and public health officials have a strong understanding of the basic facts of pandemics and disease spread.
-Social distancing, contact tracing and mass testing are all important.
-Individuals and populations with low quality data and only basic information will make decisions that respect inherent risks in unknown unknowns.
-Of course, there are trade-offs and long-term stay-at-home orders will have negative economic consequences.
-A full discussion of these topics is beyond the scope of the current article; we merely bring up the matter because we are well aware that discussions of data quality can sometimes be misinterpreted as claims that we do not have the requisite information to act immediately and decisively.
-This is simply not the case.
-
-\section{Discussion}
-\label{section:discussion}
-
+\section{Discussion.}
 There is nothing routine about COVID-19, including the corresponding statistical questions.
 The goal of this paper was to point out questionable statistical routines.
-Indeed, case-count trajectories are being reported daily by governments and broadcast by the media.
-The precision in the reported numbers gives the illusion of accuracy when what needs to be understood is the level of uncertainty.
-Selection bias not only means uncertainty, it also implies analysts feel certain about incorrect conclusions.
-We end this paper with a brief discussion of several important topics related to selection bias and measurement error.
+The precision in reported case-count data gives the illusion of information when what what is needed is quantification of uncertainty.
+Selection bias leads data analysts to feel certain about incorrect conclusions.
+We end with a brief discussion of important related topics.
 
@@
 
+\subsection*{Data quantity versus quality}
+
+Governments implicitly argue that increased testing capacity will perform the above task.
+Without complete compliance, however, our understanding of future outbreaks may be plagued by self-selection bias, compounded by changing sensitivity and specificity rates.
+Random testing removes these effect modifiers, giving governments more information to fight the disease.
+This is not to say random testing should replace current testing practices; only that the random testing should supplement and will provide valuable additional information even with moderate sample sizes.
+
 \subsection*{High quality information on prevalence.}
@@
 
-\subsection*{Data quantity versus quality}
+\subsection*{Stratified sampling: improving precision in low-prevalence environments.}
+Random sampling is a powerful tool.
+One difficulty with SRS is when prevalence is low.
+Consider the Santa Clara study~\cite{Bendavid2020} which estimated prevalence less than 2\% in the county.
+The key issue with low prevalence is that if the test's specificity is higher than the prevalence, then the observed data is consistent with zero recovered individuals.
+In these settings, SRS is likely to fail; however, there are potential solutions.
+One can, for example, stratify the population into risk categories (e.g., based on population density, occupation, age) and perform Neyman allocation~\cite{Cochran77}.
 
-Governments implicitly argue that increased testing capacity will perform the above task.
-Without complete compliance, however, our understanding of future outbreaks may be plagued by self-selection bias, compounded by changing sensitivity and specificity rates.
-Random testing removes these effect modifiers, giving governments more information to fight the disease.
-This is not to say random testing should replace current testing practices; only that the random testing should supplement and will provide valuable additional information even with moderate sample sizes.
+\paragraph*{Model-based solutions.}
 
-% Not only that, in current crisis data quality is often tied to data quantity due to rationing system.
-Increasing testing under test rationing lowers correlation.
-There is a lot of hidden information in the selection indicator $I_j$.
-Indeed, individuals must first seek testing and then call their primary health care provider must decide whether they met thresholds for testing.
-Our analysis shows the limitations and caveats that come from analysis of this data in terms of estimation of prevalence, rates of change, and cross-country comparisons.
-We show how one can build simple sensitivity analyses to understand the potential bias in case-count data using surveys that will become more prevalent in the next few months.
-
-\subsection*{Model-based solutions.}
-
-One alternative approach to what is presented in this article is a model-based extension of SIR models to try and capture both selection bias and measurement error; however, without strong assumptions on the selection mechanism, the estimates are often not nonparametrically identifiable.
-When an issue ``cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically'' ~\cite[pp.
-96]{CoxHink74}.
-In the absence of randomized experiments, the best route forward using any approach is the use of detailed sensitivity analyses and humility in conclusions drawn from such data sources.
+One can extend the SIR model to account for selection bias and measurement error.
+Without strong assumptions on the selection mechanism, however, the estimates are often not identifiable.
+When an issue ``cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically''\cite{CoxHink74}.
+Absent random sampling, the best route forward for all data analyses is careful associated sensitivity analyses and humility in data-driven conclusions.
 

### 2020-08-24 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 has become the focal point of everyday life; yet the data landscape for understanding COVID-19 remains limited.
-Public databases~\cite{JHU_Lancet,NYT} provide incoming county-level information of confirmed cases and deaths.
-Statisticians, epidemiologists, economists, and data scientists have used this granular data to forecast COVID-19 case-counts, deaths, and hospitalizations~\cite{Giordano2020,Song2020,Ray2020,2020.IHME,Wang2020.03,JTD36385}.
+Public databases~\citep{JHU_Lancet,NYT} provide incoming county-level information of confirmed cases and deaths.
+Statisticians, epidemiologists, economists, and data scientists have used this granular data to forecast COVID-19 case-counts, deaths, and hospitalizations~\citep{Giordano2020,Song2020,Ray2020,2020.IHME,Wang2020.03,JTD36385}.
 
@@
 The SRS variance formula above shows that $E_{\I} [ \rho_{I,Y}^2 ]  = 1/(N-1)$.
+
 The key issue with selective testing is that $E_{\I} [ \rho_{I,Y} ] \neq 0$.
 Meng identified this as the fundamental issue that can lead to paradoxes in the analysis of big data.
-
 The remainder of this paper aims to build upon this fundamental work by extending the decomposition in two directions: accounting for measurement error and the temporal nature of the pandemic.
@@
 The second term represents the interaction between imperfect testing and selection bias.
-The variable $PZ = 1[Y=0,P=1] - 1[Y=1, P=1]$ is non-zero only when $P=1$ (i.e., the outcome is incorrectly reported); to this end, we refer to $\rho_{I,PZ}$ as the \emph{observed data quality} that accounts for both selection bias and measurement error.
+The variable $PZ = 1[Y=0,P=1] - 1[Y=1, P=1]$ is non-zero only when $P=1$ (i.e., the outcome is incorrectly reported) and the sign depends on the true outcome $Y$.
+Given this, we refer to $\rho_{I,PZ}$ as the \emph{observed data quality} that accounts for both selection bias and measurement error.
 In the appendix, we show the sign of $\rho_{I,PZ}$ is the opposite of the sign of $\rho_{I,Y}$, which implies that the observed data quality adds error in the opposite direction from the true data quality.
@@
 To do this, we define the sampling rates differential.
-Let $f_1 := \pr (I_J = 1 \mid Y_J = 1)$ and $f_0 := \pr(I_J = 1 \mid Y_J = 0)$ be the sampling rates.
+Let $f_1 := \pr (I_J = 1 \mid Y_J = 1)$ and $f_0 := \pr(I_J = 1 \mid Y_J = 0)$ be the sampling rates where $J$ is a random index defined on $\{1,\ldots, N\}$.
 Then $\Delta = f_1 - f_0$ is the sampling rate differential.
@@
 $$
-The final term in brackets is the \emph{measurement error adjustment to data quality} which is a complex function of sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence.
-Note $\sgn(\Delta) = \sgn(\rho_{I,Y})$ by equation~\ref{eq:binaryrho} in the appendix, which implies the measurement error adjustment either shrinks the data quality measure toward zero or reverses its sign.
+final term represents the \emph{imperfect testing adjustment} which is a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence.
+Note $\sgn(\Delta) = \sgn(\rho_{I,Y})$ by equation~\ref{eq:binaryrho} in the Appendix, which implies the measurement error adjustment either shrinks the data quality measure toward zero or reverses its sign.
 
@@
 
-\subsection{Deriving an unbiased estimator}
+\subsection{An improved estimator under imperfect testing}
 
-The attentive data analyst will recognize the estimator $\bar y_n$ is biased even for simple random samples and, if sensitivity and specificity were known a priori, may use the error decomposition to construct an unbiased estimator $\tilde y_n$.
-
-In this paper, we consider the estimator $\tilde y_n = (\bar y_n - FP)/(1-(FP+FN))$ where $\bar y_n = n^{-1} \sum_{i=1}^n y_i$ is the proportion of positive tests in the sample.
-This estimator is unbiased under simple random sampling (SRS); see the supplementary materials for a derivation and discussion of the proposed estimator.
+The attentive data analyst will recognize the estimator $\bar y_n$ is biased even for simple random samples.
+In Appendix~\ref{app:memestimator}, assuming sensitivity and specificity were known a priori, we present a novel iterative procedure that uses the error decomposition to construct the estimator $\tilde y_n = (\bar y_n - FP)/(1-(FP+FN))$, which is unbiased under simple random sampling (SRS); see Appendix~\ref{app:modelbased} for a discussion of the connection to model-based estimators.
 To understand the impact of selection bias and imperfect testing, we derive the following statistical decomposition of the error between $\tilde y_n$ and $\bar Y$:
@@
 \end{equation}
-where $f=n/N$ is the sampling fraction, $f_1=n_1/N_1$ and $f_0=n_0/N_0=(n-n_1)/(N-N_1)$ are sampling fractions for COVID-19 positive and negative individuals respectively, and $\Delta = f_1 - f_0$ is the sampling rate differential.
-See the supplementary materials for the derivation.
+where $f=n/N$ is the sampling fraction, $f_1$ and $f_0$ are sampling fractions for COVID-19 positive and negative individuals respectively, and $\Delta = f_1 - f_0$ is the sampling rate differential.
+See Appendix~\ref{app:memestimator} for the derivation.
 This extends work by \cite{Meng2018} to account for imperfect testing.
 The first term represents \emph{data quality}, the second \emph{data quantity}, and the third \emph{problem difficulty}.
-The final term represents the \emph{imperfect testing adjustment} which is a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence.
-Figure~\ref{fig:heatmap} shows that $D_M$, as a function of the relative frequency ($f_1/f_0$) and odds ratio, can be both positive and negative as well as a range of magnitudes~\cite{Beesley2020,Beesley2019,Smeden2019}.
+The final term represents the \emph{imperfect testing adjustment} which is again a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence but here is scaled by the factor $(1 - FP - FN)^{-1}$ compared to prior decomposition.
+
+Figure~\ref{fig:heatmap} shows that $D_M$, as a function of the relative frequency ($f_1/f_0$) and odds ratio, can be both positive and negative as well as a range of magnitudes~\citep{Beesley2020,Beesley2019,Smeden2019}.
 Under perfect testing, $D_M = 1$ so the relation between estimation and selection bias is simple, e.g., COVID-19 positive individuals more likely to receive test implies upward bias in prevalence estimates.
 When tests are imperfect, this simple relationship no longer holds.
+
+\begin{figure}[!th]
+\centering
+\includegraphics[width = 0.4\textwidth]{../methods/figs/mem_heatmap_article.png}
+\caption{Imperfect testing adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.005$ and $FN=0.172$. Color scaled so blue = $-10$, white = $0$, and red = $10$.}
+\label{fig:heatmap}
+\vspace{-0.3cm}
+\end{figure}
 
@@
 $$
-implying the error relative to SRS increases as a function of population size~\cite{Meng2018}.
+implying the error relative to SRS increases as a function of population size~\citep{Meng2018}.
 The paradox that two countries with the same testing strategy can yield wildly different estimates due to population size is too often ignored in current communications of case-count data.
@@
 
-\begin{wrapfigure}{r}{0.5\textwidth}
-\centering
-\includegraphics[width = 0.4\textwidth]{../methods/figs/mem_heatmap_article.png}
-\caption{Measurement error data quality adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.005$ and $FN=0.172$.
-Color scaled so blue = $-10$, white = $0$, and red = $10$.}
-\label{fig:heatmap}
-\vspace{-0.3cm}
-\end{wrapfigure}
-
-A key question is ``What is the (effective) sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' In the supplement, we show the effective sample size $n_{eff}$ can be bounded by
+A key question is ``What is the (effective) sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' In Appendix~\ref{app:effss}, we show the effective sample size $n_{eff}$ can be bounded by
 $$
@@
 The cumulative empirical prevalence is $9.3\%$ on May 8th.
-A recent study estimates a false positive and negative rate for antibody testing at $0.5$\% and $17.2$\% respectively \cite{Bendavid2020}.
+A recent study estimates a false positive and negative rate for antibody testing at $0.5$\% and $17.2$\% respectively \citep{Bendavid2020}.
 Supposing COVID-19 positive individuals are $2$ times more likely to get tested, then the effective sample size is $92$.
-Recent proposals~\cite{Siddarth2020} have argued for increased testing capacity, which may likely reduce the relative sampling rate.
+Recent proposals~\citep{Siddarth2020} have argued for increased testing capacity, which may likely reduce the relative sampling rate.
 If the relative sampling rate drops to $1.2$ then the effective sample size will increase to $1950$.
@@
 Then the effective sample size increases by a factor of $1.76$ rather than the expected $4$ fold increase.
+
+% \begin{wrapfigure}{r}{0.5\textwidth}
 
@@
 where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement error adjustment, and $CV(Y_j) = \sigma_{Y_j}/Y_j$ is the coefficient of variation on day $j$.
-See the supplementary materials for the derivation.
+See Appendix~\ref{app:ratio} for the derivation.
 The error magnitude depends on the true rate $\bar Y_{t} / \bar Y_{t-1}$ so a large decrease will have a relatively small error compared to a large increase.
@@
 
-Figure~\ref{fig:ratio-bias} displays the trajectory of the true ratio and the potential biased estimators under an SIR model~\cite{Pastor2001,Newman2002,Parshani2010} for the epidemic dynamics, with state evolution given by
+Figure~\ref{fig:ratio-bias} displays the trajectory of the true ratio and the potential biased estimators under an SIR model~\citep{Pastor2001,Newman2002,Parshani2010} for the epidemic dynamics, with state evolution given by
 $$
@@
 \label{section:r0-estimation}
-Many epidemiologists argue that tracking the effective reproduction number $R_t$ is the only way to manage through the crisis~\cite{Gabriel2020}.
-Under a Poisson likelihood, a simple relation between the trajectory of new cases and the effective reproduction number can be derived \cite{Bettencourt2008}.
-In particular, under an SIR model the number of case counts on day $t$, denoted $K_t$, is Poisson distributed with rate $K_{t-1} \exp \left( \gamma (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval, which is approximately $7$ days for COVID-19~\cite{Sanche2020}.
+Many epidemiologists argue that tracking the effective reproduction number $R_t$ is the only way to manage through the crisis~\citep{Gabriel2020}.
+Under a Poisson likelihood, a simple relation between the trajectory of new cases and the effective reproduction number can be derived \citep{Bettencourt2008}.
+In particular, under an SIR model the number of case counts on day $t$, denoted $K_t$, is Poisson distributed with rate $K_{t-1} \exp \left( \gamma (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval, which is approximately $7$ days for COVID-19~\citep{Sanche2020}.
 Using this relation, a moment-based estimator is given by
@@
 Even in a randomized experiment, the question is what the correct value for $E(U)$ is.
-For COVID-19, the value of interest is the expected value of $U$ in the population of COVID-19 positive individuals.
-Due to selection bias, however, the bias in this marginal treatment effect compared to the true marginal effect of interest is $\beta_1 \cdot E_{\I} [ \rho_{I,U} ] \times \sqrt{ \frac{f}{1-f} } \times \sigma_U$.
+For COVID-19, the value of interest is likely the expected value of $U$ in the population of COVID-19 positive individuals, i.e., $E_J [ U_J Y_J]/ E_J[Y_J] = \sum_{j=1}^N U_j Y_j / \sum_{j=1}^N Y_j$.
+Assuming no measurement error, the estimator is given by the ratio $\sum_{j=1}^n y_i u_i / \sum_{j=1} y_i$.
+Due to selection bias, the bias in the marginal treatment effect compared to the true marginal effect of interest is approximately equal to
+$$
+\beta_3 \times \sqrt{\frac{1-f}{f}} \times \frac{\bar U^\prime}{\bar Y} \bigg[ \rho_{I,U^\prime} \times CV (U^\prime)  - \rho_{I,Y} \times CV (Y) \bigg] \left[ 1 - \rho_{I,Y} \sqrt{\frac{1-f}{f}} CV (Y) \right]
+$$
+where $U^\prime = U Y$.
+The derivation again uses a second-order Taylor series approximation and follows exactly the same as in Appendix~\ref{app:ratio}.
 Selection bias then leads to the estimated effect being biased if $\beta_1 \neq 0$.
 Directionality of the bias will depend on the relationship between $U$ and $Y$.
-If $U$ is a measure of disease severity, then it could be expected to see $\beta_1 > 0$ and $E_{\I} [ \rho_{I, U} ] >0$.
+If $U$ is a measure of disease severity, then it could be expected to see $\beta_3 > 0$ and $\rho_{I,U^\prime} \times CV (U^\prime)  > \rho_{I,Y} CV (Y)$.
 In such situations, the treatment effect estimates will tend to be overly optimistic as most individuals recruited are potentially higher on the severity index than the population average.
@@
 Most often, our main interest is in causal effects on the population of COVID-19 positive.
-Randomization within the clinical trial design yields internal validity, but we also need external validity to estimate the correct marginal effect of interest~\cite{Keiding2016}.
+Randomization within the clinical trial design yields internal validity, but we also need external validity to estimate the correct marginal effect of interest~\citep{Keiding2016}.
 
@@
 It found 13.9\% of 3,000 people tested across the state had signs of the virus.
-The study did not report sensitivity and specificity; therefore, we take reported measures from the Santa Clara study~\cite{Bendavid2020}.
+The study did not report sensitivity and specificity; therefore, we take reported measures from the Santa Clara study~\citep{Bendavid2020}.
 The reported specificity is $82.8\%$ (95\% CI: 76.0-88.4\%) and sensitivity is $99.5\%$ (95\% CI: 99.2-99.7\%).
@@
 \end{subfigure}
-\caption{New York COVID-19 reported data }
+\caption{New York COVID-19 reported data}
 \label{fig:ny-covid}
@@
 The goal of this paper was to point out questionable statistical routines.
-The precision in reported case-count data gives the illusion of information when what what is needed is quantification of uncertainty.
+Precision in reported case-count data gives the illusion of information when what what is needed is quantification of uncertainty.
 Selection bias leads data analysts to feel certain about incorrect conclusions.
@@
 We know that the probability of an outbreak is a function of current prevalence and network connectivity.
-Epidemic critical thresholds have been derived for many infectious disease models~\cite{Pastor2001,Newman2002,Parshani2010}.
+Epidemic critical thresholds have been derived for many infectious disease models~\citep{Pastor2001,Newman2002,Parshani2010}.
 Knowing prevalence will help governments determine long-term community-level risk and allow for more targeted interventions -- shutting down only certain locations when necessary.
@@
 One difficulty with SRS is when prevalence is low.
-Consider the Santa Clara study~\cite{Bendavid2020} which estimated prevalence less than 2\% in the county.
+Consider the Santa Clara study~\citep{Bendavid2020} which estimated prevalence less than 2\% in the county.
 The key issue with low prevalence is that if the test's specificity is higher than the prevalence, then the observed data is consistent with zero recovered individuals.
 In these settings, SRS is likely to fail; however, there are potential solutions.
-One can, for example, stratify the population into risk categories (e.g., based on population density, occupation, age) and perform Neyman allocation~\cite{Cochran77}.
+One can, for example, stratify the population into risk categories (e.g., based on population density, occupation, age) and perform Neyman allocation~\citep{Cochran77}.
 
-\paragraph*{Model-based solutions.}
+\subsection*{Model-based solutions.}
 
@@
 Without strong assumptions on the selection mechanism, however, the estimates are often not identifiable.
-When an issue ``cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically''\cite{CoxHink74}.
+When an issue ``cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically''\citep{CoxHink74}.
 Absent random sampling, the best route forward for all data analyses is careful associated sensitivity analyses and humility in data-driven conclusions.
@@
 
-\section{Imperfect testing: derivation}
+All relevant code can be found at \url{https://github.com/wdempsey/covid-umich}.
+
+\section{Technical details}
+
+Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the outcome and $0$ when we observe the true outcome.
+We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.
+If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^{\star} = Y_j(1-P_j) + (1-Y_j) P_j$.
+
+\subsection{Derivations for imperfect testing framework.}
 \label{app:imperfect}
-We considered the mean estimator
+We start by considering the empirical mean estimator under imperfect testing,
 $$
@@
 $$
-\bar y_n^\star - \bar Y = \sqrt{\frac{1-f}{f}} \left[\rho_{I,Y} \sigma_Y + \rho_{I, PZ} \sigma_{PZ} + \sqrt{\frac{f}{1-f}} \left( FP - (FP+FN) \bar Y \right) \right]
+\bar y_n^\star - \bar Y = \sqrt{\frac{1-f}{f}} \left(\rho_{I,Y} \sigma_Y + \rho_{I, PZ} \sigma_{PZ} \right) + \left( FP - (FP+FN) \bar Y \right)
 $$
+
+\subsubsection{Derivation of an estimator unbiased under SRS}
+\label{app:memestimator}
+We see the final term is given by $FP (1-\bar Y) - FN \bar Y$ is the bias associated with using the unadjusted prevalence estimate $\bar y_n^\star$.  This motivates an adjusted estimate
+$$
+\tilde y_n^{(0)} = \bar y_n^\star - FP (1- \bar y_n^\star ) + FN \bar y_n^\star
+= \bar y_n^\star (1 + FN + FP) - FP.
+$$
+Now considering the error for the adjusted estimate, $\tilde y_n^{(0)} - \bar Y$, we have
+$$
+\begin{aligned}
+ &\bar y_n^\star (1 + FN + FP) - FP - \bar Y \\
+ =&(\bar y_n^\star - \bar Y) +  (FN + FP) \bar y_n^\star - FP  \\
+ =&\underbrace{\sqrt{\frac{1-f}{f}} \left[\rho_{I,Y} \sigma_Y + \rho_{I, PZ} \sigma_{PZ} \right]}_{\Psi} + (  FN + FP ) (\bar y_n^\star -  \bar Y) \\
+ =& \Psi + (FN + FP) \Psi + (FN + FP) (FP - (FP+FN) \bar Y).
+ \end{aligned}
+$$
+The final term is a (smaller) bias term and so we propose another adjusted estimator $\tilde y_n^{(1)} = \tilde y_n^{(0)} + (FN+FP)  ( (FN+FP) \bar y_n^\star - FP)$, with associated error $\tilde y_n^{(1)} - \bar Y$ given by
+$$
+\begin{aligned}
+ &(\bar y_n^{(0)} - \bar Y) + (FN+FP)  ( (FN+FP) \bar y_n^\star - FP)\\
+ =&\Psi + (FN + FP) \Psi + (FN + FP) (FP - (FP+FN) \bar Y) + (FN + FP) ((FP+FN) \bar y_n^\star - FP)  \\
+  =& \Psi + (FN + FP) \Psi + (FN + FP)^2 \Psi + (FN+FP)^2 (FP - (FP+FN) \bar Y).
+ \end{aligned}
+$$
+This motivates recursively defining estimators $\tilde y_n^{(t)} = \tilde y_n^{(t-1)} + (FN+FP)  ( (FN+FP) \bar y_n - FP)$ for $t=1,2,\ldots$ where $\tilde y_n^{(0)} = \bar y_n^\star$.  Then
+$$
+\tilde y_n^{(t)} = \bar y_n^\star \sum_{s=0}^{t+1} (FP+FN)^s - FP \sum_{s=0}^{t} (FP+FN)^s
+$$
+and the associated error at iteration $t$ given by
+$$
+\Psi \sum_{s=0}^t (FN+FP)^s = \Psi \frac{1 - (FN+FP)^t}{1 - (FN+FP)}.
+$$
+We can then get an estimator with no residual bias term by taking the limit as $t$ goes to infinity; that is, define
+$$
+\tilde y_n = \lim_{t \to \infty} \tilde y_n^{(t)} = \frac{\bar y_n^\star - FP}{1 -(FN+FP)}.
+$$
+Then the associated error $\tilde y_n - \bar Y$ can be expressed as $\frac{\Psi}{1-(FN+FP)}$.
+
+\subsection{Model-based derivation of the estimator.}
+\label{app:modelbased}
+The estimator $\tilde y$ was derived as a limit of a process that removes the residual bias term at each step.
+Here we consider a model-based explanation.
+Let $\theta = \pr( Y = 1)$ and $\phi = \pr( \text{test is positive})$.
+Given a known false negative (FN) and false positive (FP) rates, we have
+$$
+\begin{aligned}
+\phi &= \theta \cdot (1-FN) + (1-\theta) \cdot FP = \theta (1-FN-FP) + FP \\
+\Rightarrow \theta &= \frac{\phi - FP}{1-FN-FP}.
+\end{aligned}
+$$
+Thus, the estimator $\tilde y$ is also the appropriate estimator under a model-based approach.
+While the derivation here is more straightforward, the derivation in the prior section provides a simple formula for the associated error $\tilde y_n - \bar Y$ and gives a novel connection between the empirical estimator $\bar y_n^\star$ and the adjusted estimator $\tilde y_n$ without reference to the model-based approach.
+
+\subsection{Further simplification.}
 For the binary outcome $Y$, we have $\sigma_Y = \sqrt{\bar Y (1-\bar Y)}$. Moreover,
@@
 Then the formula for the error is given by:
-$$
-\sqrt{\frac{1-f}{f}} \left[\rho_{I,Y} \sqrt{\bar Y (1-\bar Y)} + \rho_{I, PZ} \sqrt{ 2 \bar Y \left( FP (1-\bar Y) + FN \cdot \bar Y \right )} + \sqrt{\frac{f}{1-f}} \left( FP - (FP+FN) \bar Y \right) \right]
-$$
-
-\subsection{Estimate of observed data quality}
-
+\begin{equation}
+\label{eq:finalstep}
+\sqrt{\frac{1-f}{f}} \left[\rho_{I,Y} \sqrt{\bar Y (1-\bar Y)} + \rho_{I, PZ} \sqrt{ 2 \bar Y \left( FP (1-\bar Y) + FN \cdot \bar Y \right )} \right] \times \frac{1}{1- (FN+FP)}
+\end{equation}
 By definition, we have
@@
 $$
-\rho_{I,Y} \sigma_Y \left( 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right)
+\rho_{I,Y} \sigma_Y \left( 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right).
 $$
-Suppose $f_1 = M \cdot f_0$.  Then $\Delta = f_1 - f_0 = f_0 (M-1)$ and $f_0 (1-\bar Y) + f_1 \bar Y = f_0 ( (1-\bar Y) + M \bar Y )$ and we can re-write above as
+Inserting into equation~\eqref{eq:finalstep} yields the desired result.
+
+\subsection{Derivation of effective sample size}
+\label{app:effss}
+
+Let $S_Y^2 = (N-1)^{-1} \sum_{j=1}^N (Y_j - \bar Y)^2$ be the population variance as defined in survey sampling~\citep{Cochran77}.
+Then $\sigma_Y^2 = (N-1)/N \cdot S_Y^2$.
+Under SRS, the MSE is the variance as the estimate is unbiased and the variance is given by $(1-f)/n S_Y^2$.
+Then setting the MSE under general selection and SRS equal we have
 $$
-\rho_{I,Y} \sigma_Y \left( 1 - (M-1) \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{(1-\bar Y) + M \bar Y} \right).
+\begin{aligned}
+\underbrace{\frac{1-f}{f} \times E_{\I} \left[ \rho_{I, Y}^2 \times D_M^2 \right]}_{1/n_{eff}^\star} \times \sigma_Y^2 &= \frac{1-f}{n_{eff}} S_Y^2 \\
+\frac{1}{n_{eff}^\star} \times \frac{N-1}{N} S_Y^2 &= \frac{1-f}{n_{eff}} S_Y^2 \\
+\frac{1}{n_{eff}^\star} &=  \left( \frac{1}{n_{eff}} - \frac{1}{N} \right) \left( \frac{N}{N-1} \right) \\
+\frac{1}{n_{eff}^\star} \left[ 1 - \frac{1}{N} + \frac{n_{eff}^\star}{N} \right]  &=  \frac{1}{n_{eff}} \\
+n_{eff}^\star \left[ 1 - \frac{1}{N} + \frac{n_{eff}^\star}{N} \right]^{-1}  &=  n_{eff} \\
+\frac{n_{eff}^\star}{ 1 + (n_{eff}^\star -1) N^{-1}} &=  n_{eff} \\
+\end{aligned}
+$$
+Then if $n_{eff}^\star \geq 1$, we have that
+$$
+n_{eff} \leq n_{eff}^\star = \frac{f}{1-f} \times \frac{1}{E_{\I} \left[ \rho^2_{I, Y} \times D_M^2 \right]}
 $$
@@
 
-\section{Derivation for the effective sample size}
+\subsection{Computating the effective sample size}
 \label{section:effss}
@@
 
-% \section{Quotes}
-
-% I prefer to think of a statistical sensibility rather than statistical thinking.
-It’s “less than an agenda but more than an attitude.”  It allows for methodological preference while avoiding dogma.
-Paired with data analytic humility and I think you have proper “data science”
-
-% A test of meaningfulness of a possible model for a data-generating process is whether it can be used directly to simulate data.” (p.104).
-In our current setting, this most certainly related to simulation while accounting for measruement error.
-
 \end{document}

### 2020-08-26 write-up/covid-writeup.tex
Update covid-writeup.tex
Revision underway: including doubly robust statistics

@@
-\documentclass[12pt]{amsart}
+\documentclass[11pt]{amsart}
 
@@
 Through simple calculations, we demonstrate pitfalls and paradoxes that can arise when considering case-count data in the presence of selection bias and measurement error.
+\textcolor{red}{Need to update}
 \end{abstract}
@@
 
-The goal of this paper is to express reservations at the use of case-counts as a proxy for disease prevalence and in estimation of standard epidemiological models for inference and forecasting.
+This paper has two goals.
+The first is to express reservations at the use of case-counts as a proxy for disease prevalence and in estimation of standard epidemiological models for inference and forecasting.
 The reason is straightforward: observed case-count data is plagued by selection bias and measurement error.
 Through simple calculations, we will demonstrate that the information gained from increasing testing capacity is limited in the presence of selection bias and when testing inaccuracies persist.
+In particular, the millions of tests in the US have a small effective sample size when compared to random sampling.
 
-% \subsection{Outline}
+The second is to show how probabilistic samples can also be used to derive a doubly robust approach that leverages epidemiological models along with pseudo survey weights built from combining nonprobability and probability samples.
+Using a random sample from Indiana, we show that  for several key covariates and, given several assumptions, produce corrected estimates of the prevalence and true case-counts.
+We end with several remarks on cautious use of these methods.
 
-% This article discusses the relationship between three statistical concepts: selection bias, measurement error, and the too often forgotten population size.
+\subsection{Related work.}
+
+Meng
+
+Combining nonprob and prob samples
+
+Epi forecasting
+
+\subsection{Objectives and outline}
+
+This article discusses the relationship between three statistical concepts: selection bias, measurement error, and the too often forgotten population size.
 We clarify mathematically why the situation is much more complex than it first appears.
@@
 We then end the paper with a discussion of the benefits of randomized testing and use of auxiliary information to improve over simple random sampling.
+
+\section{COVID-19 data}
+
+In this paper, we start with case-count data as reported by US state governments (cites).  Figure~\ref{fig:ny-covid}
+
+Case-count data
+
+Generally available public data (COVID trackers).
+
+COVID trackers tend to provide basic information.  From the JHU dashboard,
+
+Most states provide additional demographic data.
+
+We focus on strata built on age, race, and sex.  Only have marginal information!
+
+\subsection{US COVID-19 diagnostic and antibody testing}
+
+There are two types of tests.  The first is
+
+When an individual is infected with SARS-CoV-2 there is a
+
+During the first two days, an infected individual will not
+An infected individual will
+
+%https://www.npr.org/sections/goatsandsoda/2020/05/31/865932474/should-i-get-tested-for-coronavirus-just-for-the-heck-of-it
+ test positive three to five days after exposure, the individual enters the stage at which detection is possible
+
+We focus on US COVID-19 testing.
+
+The typical testing
+
+However, upon investi
+%https://www.coronavirus.in.gov/
+As of today, access to the history of testing requirements remains quite difficult.
+Early during the coronavirus pandemic, limited testing meant that only those at high risk were allowed to be tested (i.e., symptomatic or high risk populations); however, as testing capacity expanded these restrictions were gradually lifted.
+Such restrictions
+
+Given the difficult, we focus on reconstructing this history for the state of Indiana.
+As of May 12th, 2020, the Indiana State Department of Health laboratory was testing high-risk individuals for COVID-19.
+Any symptomatic individual could receive a tested.
+Moreover there were key factors for people who can be tested regardless of symptom status: over the age of 65, diabetic, high-blood pressure, obesity, pregnancy, and minority populations at greater risk of severe illness.
+% https://www.indystar.com/story/news/health/2020/05/12/coronavirus-testing-indiana-who-should-get-tested/3110592001/
+On April 28th, 2020, the criteria expanded to include any Indiana resident with virus symptoms, people in close contact with those who have tested positive and residents of congregant communities.
+% https://www.wishtv.com/news/medical/indiana-opens-up-covid-19-testing-to-more-hoosiers/
+On June 15, 2020, ISDH lifted all restrictions and testing expanded to include anyone in Indiana who wants to be tested for coronavirus.
+% https://www.indystar.com/story/news/health/2020/06/12/indiana-says-anyone-who-wants-coronavirus-test-can-get-one/3179151001/
+
+\subsection{Probabilistic samples}
+
+Indiana random sample description (see paper for more details)
+
+Among 15,495 randomly selected persons, 3,695 participated of whom 3,629 had at least one test result available.
 
@@
 If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^{\star} = Y_j(1-P_j) + (1-Y_j) P_j$.
-Suppose disease prevalence was estimated as the fraction who tested positive for COVID-19, i.e., $\bar y_n^\star = \frac{\sum_{i=1}^N I_j Y_j^\star}{\sum_{i=1}^N I_j}$.
-We can again investigate the error compared to the true prevalence $\bar Y$ in statistical terms:
-$$
-\bar y_n^\star - \bar Y = \sqrt{\frac{1-f}{f}} \left[ \rho_{I,Y} \times \sigma_Y + \rho_{I,PZ} \times \sigma_{PZ} + \sqrt{\frac{f}{1-f}}  \left( FP - (FP+FN) \bar Y \right) \right] .
-$$
-where $Z = 1-2Y$; see Appendix~\ref{app:imperfect} for the derivation.
-The first term in the large brackets represents the perfect testing regime; to this end, we refer to $\rho_{I,Y}$ as the \emph{true data quality}.
-The second term represents the interaction between imperfect testing and selection bias.
-The variable $PZ = 1[Y=0,P=1] - 1[Y=1, P=1]$ is non-zero only when $P=1$ (i.e., the outcome is incorrectly reported) and the sign depends on the true outcome $Y$.
-Given this, we refer to $\rho_{I,PZ}$ as the \emph{observed data quality} that accounts for both selection bias and measurement error.
-In the appendix, we show the sign of $\rho_{I,PZ}$ is the opposite of the sign of $\rho_{I,Y}$, which implies that the observed data quality adds error in the opposite direction from the true data quality.
-Finally, the third term represents the bias due to imperfect testing.
-
-\subsubsection{Connecting true and observed data quality}
-
-We start by considering the first two terms and assess whether the sign of the bias can reverse due to the interaction of measurement error and selection bias.
-To do this, we define the sampling rates differential.
-Let $f_1 := \pr (I_J = 1 \mid Y_J = 1)$ and $f_0 := \pr(I_J = 1 \mid Y_J = 0)$ be the sampling rates where $J$ is a random index defined on $\{1,\ldots, N\}$.
-Then $\Delta = f_1 - f_0$ is the sampling rate differential.
-Using these terms, we can re-express the first two terms as
-$$
-\rho_{I,Y} \times \sigma_Y + \rho_{I,PZ} \times \sigma_{PZ} =
-\rho_{I,Y} \times \sigma_Y \left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right].
-$$
-final term represents the \emph{imperfect testing adjustment} which is a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence.
-Note $\sgn(\Delta) = \sgn(\rho_{I,Y})$ by equation~\ref{eq:binaryrho} in the Appendix, which implies the measurement error adjustment either shrinks the data quality measure toward zero or reverses its sign.
-
-% While prior investigations have noted the interaction between measurement error and selection bias~\citep{Beesley2020,Beesley2019,Smeden2019}, the interaction with the sample size relative to the population, i.e., $f$, has largely been ignored.
-The above statistical decomposition clarifies the importance of this quantity~$f$.
-In particular, note that the statistical error also includes a bias term due to measurement error and this term increases as the sampled fraction $f$ increases.
-Therefore, how the first two terms interact with the final term depends on the fraction of the population sampled.
-This interaction is complex, but implies that whether the estimate $\bar y^\star_n$ is an overestimate or underestimate is a complicated question due to the relation amongst these three pieces.
-
-% Consider again the current COVID-19 pandemic.
-For now, we continue to assume the ratio of conditional selection rates $f_1/f_0$ is equal to 1.5.
-In Section~\ref{section:est_dq}, we discuss recent research suggesting a false negative rate around 17.2\% and a false positive rate around 0.05\%.
-Under these rates and the current US prevalence rates, the ratio of the MSE to MSE under no measurement error is 0.436; if we switch false negative rates down to 5\% and increase the false positive rate to 5\% then the relative MSE is 2.89.
-This is merely to demonstrate that in some cases we see a huge increase in MSE and in other settings we have a huge decrease in MSE.
-What drives this is the false positive and negative rate interaction with prevalence and sampling rates.
-Therefore, whether we are better or worse off with respect to the MSE is a very difficult question to answer.
-
-\subsection{An improved estimator under imperfect testing}
-
 The attentive data analyst will recognize the estimator $\bar y_n$ is biased even for simple random samples.
@@
 Then the effective sample size increases by a factor of $1.76$ rather than the expected $4$ fold increase.
-
 % \begin{wrapfigure}{r}{0.5\textwidth}
@@
 
-\subsection{Effect modifiers in COVID-19 clinical trials}
+\section{Doubly robust estimation}
 
-Here we show how selection bias and measurement error can creep into clinical trial analysis.
-The key concern is whether clinical trials on COVID-19 recruit from the pool of individuals who have tested positive for COVID-19 or whether they sample randomly from the population, test, and then recruit from this subset of tested individuals.
-To see this issue, suppose we have an outcome $Y$, a treatment $A$ and an \emph{unobserved} variable $U$.
-Suppose treatment is assigned at random, i.e., $A =1$ with probability $50\%$ and $A=0$ with probability $50\%$.
-Suppose the conditional mean of the outcome satisfies  $E(Y \mid U, A ) = \beta_0 + \beta_1 A + \beta_2 U + \beta_3 A U$.
-Typically, we are interested in the \emph{causal effect} of $A$ on $Y$.
-In counterfactual language,
+When presented with non-probability samples, one can often reduce bias in two ways.
+First, one may model the propensity for self-selection and use inverse probability weighting (IPW)~(cite: Elliott) techniques to adjust for selection bias.
+Unlike in causal inference~(cite: Robins) where the observational data permits direct estimation of propensities, one does not observe those who decide not to be  part of the .
+Instead, we need auxiliary information.
+Here, we leverage US census and the Indiana random testing demographic information to adjust the propensities.
+
+First generate the counts per cell using raking on the marginal summary statistics.  Suppose the selection indicator could be modeled parametrically and depends only on the strata $P($
+
+Start by using raking to get the ``number'' of individuals per strata.
+\begin{rmk}[Raking]
+All we get is testing demographic data.
+Indiana provides.
+Checked and it is often available upon digging.
+Demographic data allows us to build stratified estimators.
+We start by assuming that $d_k$ is the number of individuals in each strata is known.
+\end{rmk}
+
+Suppose we wanted to fit a logistic regression with the outcome being the selection indicator $Y_i$ given the strata information, coded as $X_i \in [K]$ for simplicity.
 $$
-E( Y(A=1) - Y(A=0) ) = E_U ( E_A ( Y(A=1) - Y(A=0) \mid U=u ))
-= \beta_1 + \beta_3 E(U).
+\sum_{i=1}^N R_i \log \left( \pi (X_i; \theta) \right) + (1-R_i) \log \left( 1 - \pi (X_i; \theta) \right)
+= \sum_{i=1}^N R_i \log \left( \frac{\pi (X_i; \theta)}{1-\pi(X_i; \theta)} \right) + \sum_{i=1}^N \log \left( 1 - \pi (X_i; \theta) \right)
 $$
-Even in a randomized experiment, the question is what the correct value for $E(U)$ is.
-For COVID-19, the value of interest is likely the expected value of $U$ in the population of COVID-19 positive individuals, i.e., $E_J [ U_J Y_J]/ E_J[Y_J] = \sum_{j=1}^N U_j Y_j / \sum_{j=1}^N Y_j$.
-Assuming no measurement error, the estimator is given by the ratio $\sum_{j=1}^n y_i u_i / \sum_{j=1} y_i$.
-Due to selection bias, the bias in the marginal treatment effect compared to the true marginal effect of interest is approximately equal to
+The first term involves quantities observed in the nonprobability sample and can be re-expressed as sums over the strata.
+The second term, however, depends on the entire population.
+We can re-write this as a sum over strata as well $\sum_{k=1}^K N_k \log \left( 1- \pi_k \right)$ where $N_k$ is the number of individuals in strata $k$.
+The strata totals may be available from Census information.
+Alternatively, one could use the probability sample to construct a design-unbiased estimator of this quantity $\sum_k N \cdot \tilde d_k/n \log ( 1- \pi_k )$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.
+Under the assumptoin that $\pi_k < 1$, this leads to two potential estimators
 $$
-\beta_3 \times \sqrt{\frac{1-f}{f}} \times \frac{\bar U^\prime}{\bar Y} \bigg[ \rho_{I,U^\prime} \times CV (U^\prime)  - \rho_{I,Y} \times CV (Y) \bigg] \left[ 1 - \rho_{I,Y} \sqrt{\frac{1-f}{f}} CV (Y) \right]
+\hat \pi_k^{(1)} = \frac{d_k}{N_k}, \quad \hat \pi_k^{(2)} = \frac{d_k}{N} \frac{n}{\tilde d_k}.
 $$
-where $U^\prime = U Y$.
-The derivation again uses a second-order Taylor series approximation and follows exactly the same as in Appendix~\ref{app:ratio}.
-Selection bias then leads to the estimated effect being biased if $\beta_1 \neq 0$.
-Directionality of the bias will depend on the relationship between $U$ and $Y$.
-If $U$ is a measure of disease severity, then it could be expected to see $\beta_3 > 0$ and $\rho_{I,U^\prime} \times CV (U^\prime)  > \rho_{I,Y} CV (Y)$.
-In such situations, the treatment effect estimates will tend to be overly optimistic as most individuals recruited are potentially higher on the severity index than the population average.
-Not only that, but for fixed data quality, the error compared to SRS in terms of effect estimation scales with the population size (i.e., the law of large populations).
+The assumptionWe start by assuming that $\pi_k < 1$ for all $k$.
+This is not true (need to deal with very bad symptoms cases).
+We unfortunately don't have $\tilde d_k$ nor $d_k$, but we can try and get these into using standard techniques (raking).
 
-Randomization of treatment assignment in clinical trials negates unobserved confounders.
-This, however, does not negate effect modifiers.
-Therefore, for marginal treatment effects to be interpretable, there must be a well-defined population.
-Most often, our main interest is in causal effects on the population of COVID-19 positive.
-Randomization within the clinical trial design yields internal validity, but we also need external validity to estimate the correct marginal effect of interest~\citep{Keiding2016}.
+Simplest case we have Age $<40$ in the random sample is $1,017$ and we have $13+17.3+15.2 = 45\% \times 1,002,038$ tests so $450917.1$ tests.
+Then $N = 6.85$ million while $n = 3658$ leading to $\pi_k = 0.24$.
+
+Then our new estimator is
+$$
+\frac{1}{1-FP-FN} \sum_{k} d_k (\tilde y - FP) / \pi_k.
+$$
+where $k$ indexes the strata and $d_k$ is the number of samples in strata $k$.
+
+\subsection{Statistical decomposition of error}
+
+For $i=1,\ldots,N$ let $\tilde Y_i = (Y_i - FP)/ (1- FP -FN)$ and $\tilde I_i = I_i / \pi_{K_i}$.  Then the error can now be expressed as:
+$$
+\rho_{\tilde I, Y} \times \sqrt{\frac{1-f+ CV^2_W}{f}} \sigma_{Y} \times D_M
+$$
+
+We can act like this is now $\rho_{I,\tilde Y}$ is $O(N^{-1})$ but this is most likely not the case; doing so, is an assumption that we "got something right".
+But the random sample has response bias and coverage issues as well.
+Therefore, we suspect it won't fix everything.
+In section XX, we will try and estimate the expecation of this quanitty after the fact; this will allow us to assess effective sample size and build corresponding CIs.
+
+\subsection{Model-based estimation}
+
+We circle back and add in a SEIR model with measurement-error.
+We can correct for selection bias two through simple rules (see Song and Mukherjee and others for more complex alternatives).
+With such limited data we can only predict total counts and use the demographics as is to be representative.
+This yields estimates at $\hat y_k$
+
+$$
+\sum_{k} P_k \hat y_k + \frac{1}{N} \sum_{i=1}^N I_i \frac{(Y_i - FP)/(1-FP-FN) - \hat y}{\pi_{K_i}}
+$$
+where $P_k$ is the population fraction in strata $k$.
+This estimator is ``doubly-robust'' in the sense that it is robust to either the SIR or the propensities being correctly specified.
+Setting $\pi_i = n/N$ and $\hat y_k = 0$ yields the estimator $(\bar y_n - FP)/(1-FP-FN)$.
+
+Trying to decompose this error we start by recognizing $\frac{1}{N} \sum_{i=1}^N Y_i = \sum_k P_k \bar Y_k$.  Then we can concern ourselves with the per strata errors:
+$$
+\left( P_k - \frac{n_k}{N \pi_k} \right) \hat y_k + \frac{n_k}{N \pi_k} \frac{\bar y_k - FP}{1-FP-FN}
+$$
 
@@
 
+\section{Alternative estimators}
+
 \subsection{Rate comparisons}
@@
 \end{figure}
+
+\section{Applications to binary outcome and the COVID-19 pandemic.}
 
@@
 
+\subsection*{Additional biases and measurement-error}
+
+Length bias and measurement-error?
+
 \subsection*{Decision-making: data versus information}
@@
 
+\subsection*{Random sampling for doubly robust estimation}
+
+Selection rates are likely time-varying; therefore, we need rapid survey samples to properly model the selection criteria for doubly robust inferential procedures.
+
+COLLECT COVARIATE INFORMATION (even aggregate) and provide it in accessible formats.
+
+- Easy access to history of testing requirements
+- Smaller random samples over time (at least as testing and prevalence changes drastically) to assess selection changes
+- Record symptom/asymptomatic status!
+
+\subsection*{Systematic data collection and reporting}
+
+I would say this is important but JHU is not enough; see above for benefits of random sampling to augment and improve decision making.
+
 \bibliographystyle{plainnat}
@@
 \section{Technical details}
+
+Suppose disease prevalence was estimated as the fraction who tested positive for COVID-19, i.e., $\bar y_n^\star = \frac{\sum_{i=1}^N I_j Y_j^\star}{\sum_{i=1}^N I_j}$.
+We can again investigate the error compared to the true prevalence $\bar Y$ in statistical terms:
+$$
+\bar y_n^\star - \bar Y = \sqrt{\frac{1-f}{f}} \left[ \rho_{I,Y} \times \sigma_Y + \rho_{I,PZ} \times \sigma_{PZ} + \sqrt{\frac{f}{1-f}}  \left( FP - (FP+FN) \bar Y \right) \right] .
+$$
+where $Z = 1-2Y$; see Appendix~\ref{app:imperfect} for the derivation.
+The first term in the large brackets represents the perfect testing regime; to this end, we refer to $\rho_{I,Y}$ as the \emph{true data quality}.
+The second term represents the interaction between imperfect testing and selection bias.
+The variable $PZ = 1[Y=0,P=1] - 1[Y=1, P=1]$ is non-zero only when $P=1$ (i.e., the outcome is incorrectly reported) and the sign depends on the true outcome $Y$.
+Given this, we refer to $\rho_{I,PZ}$ as the \emph{observed data quality} that accounts for both selection bias and measurement error.
+In the appendix, we show the sign of $\rho_{I,PZ}$ is the opposite of the sign of $\rho_{I,Y}$, which implies that the observed data quality adds error in the opposite direction from the true data quality.
+Finally, the third term represents the bias due to imperfect testing.
+
+\subsubsection{Connecting true and observed data quality}
+
+We start by considering the first two terms and assess whether the sign of the bias can reverse due to the interaction of measurement error and selection bias.
+To do this, we define the sampling rates differential.
+Let $f_1 := \pr (I_J = 1 \mid Y_J = 1)$ and $f_0 := \pr(I_J = 1 \mid Y_J = 0)$ be the sampling rates where $J$ is a random index defined on $\{1,\ldots, N\}$.
+Then $\Delta = f_1 - f_0$ is the sampling rate differential.
+Using these terms, we can re-express the first two terms as
+$$
+\rho_{I,Y} \times \sigma_Y + \rho_{I,PZ} \times \sigma_{PZ} =
+\rho_{I,Y} \times \sigma_Y \left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right].
+$$
+final term represents the \emph{imperfect testing adjustment} which is a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence.
+Note $\sgn(\Delta) = \sgn(\rho_{I,Y})$ by equation~\ref{eq:binaryrho} in the Appendix, which implies the measurement error adjustment either shrinks the data quality measure toward zero or reverses its sign.
+
+% While prior investigations have noted the interaction between measurement error and selection bias~\citep{Beesley2020,Beesley2019,Smeden2019}, the interaction with the sample size relative to the population, i.e., $f$, has largely been ignored.
+The above statistical decomposition clarifies the importance of this quantity~$f$.
+In particular, note that the statistical error also includes a bias term due to measurement error and this term increases as the sampled fraction $f$ increases.
+Therefore, how the first two terms interact with the final term depends on the fraction of the population sampled.
+This interaction is complex, but implies that whether the estimate $\bar y^\star_n$ is an overestimate or underestimate is a complicated question due to the relation amongst these three pieces.
+
+% Consider again the current COVID-19 pandemic.
+For now, we continue to assume the ratio of conditional selection rates $f_1/f_0$ is equal to 1.5.
+In Section~\ref{section:est_dq}, we discuss recent research suggesting a false negative rate around 17.2\% and a false positive rate around 0.05\%.
+Under these rates and the current US prevalence rates, the ratio of the MSE to MSE under no measurement error is 0.436; if we switch false negative rates down to 5\% and increase the false positive rate to 5\% then the relative MSE is 2.89.
+This is merely to demonstrate that in some cases we see a huge increase in MSE and in other settings we have a huge decrease in MSE.
+What drives this is the false positive and negative rate interaction with prevalence and sampling rates.
+Therefore, whether we are better or worse off with respect to the MSE is a very difficult question to answer.
+
+\subsection{An improved estimator under imperfect testing}
 
@@
 
+\section{Effect modifiers in COVID-19 clinical trials}
+
+Here we show how selection bias and measurement error can creep into clinical trial analysis.
+The key concern is whether clinical trials on COVID-19 recruit from the pool of individuals who have tested positive for COVID-19 or whether they sample randomly from the population, test, and then recruit from this subset of tested individuals.
+To see this issue, suppose we have an outcome $Y$, a treatment $A$ and an \emph{unobserved} variable $U$.
+Suppose treatment is assigned at random, i.e., $A =1$ with probability $50\%$ and $A=0$ with probability $50\%$.
+Suppose the conditional mean of the outcome satisfies  $E(Y \mid U, A ) = \beta_0 + \beta_1 A + \beta_2 U + \beta_3 A U$.
+Typically, we are interested in the \emph{causal effect} of $A$ on $Y$.
+In counterfactual language,
+$$
+E( Y(A=1) - Y(A=0) ) = E_U ( E_A ( Y(A=1) - Y(A=0) \mid U=u ))
+= \beta_1 + \beta_3 E(U).
+$$
+Even in a randomized experiment, the question is what the correct value for $E(U)$ is.
+For COVID-19, the value of interest is likely the expected value of $U$ in the population of COVID-19 positive individuals, i.e., $E_J [ U_J Y_J]/ E_J[Y_J] = \sum_{j=1}^N U_j Y_j / \sum_{j=1}^N Y_j$.
+Assuming no measurement error, the estimator is given by the ratio $\sum_{j=1}^n y_i u_i / \sum_{j=1} y_i$.
+Due to selection bias, the bias in the marginal treatment effect compared to the true marginal effect of interest is approximately equal to
+$$
+\beta_3 \times \sqrt{\frac{1-f}{f}} \times \frac{\bar U^\prime}{\bar Y} \bigg[ \rho_{I,U^\prime} \times CV (U^\prime)  - \rho_{I,Y} \times CV (Y) \bigg] \left[ 1 - \rho_{I,Y} \sqrt{\frac{1-f}{f}} CV (Y) \right]
+$$
+where $U^\prime = U Y$.
+The derivation again uses a second-order Taylor series approximation and follows exactly the same as in Appendix~\ref{app:ratio}.
+Selection bias then leads to the estimated effect being biased if $\beta_1 \neq 0$.
+Directionality of the bias will depend on the relationship between $U$ and $Y$.
+If $U$ is a measure of disease severity, then it could be expected to see $\beta_3 > 0$ and $\rho_{I,U^\prime} \times CV (U^\prime)  > \rho_{I,Y} CV (Y)$.
+In such situations, the treatment effect estimates will tend to be overly optimistic as most individuals recruited are potentially higher on the severity index than the population average.
+Not only that, but for fixed data quality, the error compared to SRS in terms of effect estimation scales with the population size (i.e., the law of large populations).
+
+Randomization of treatment assignment in clinical trials negates unobserved confounders.
+This, however, does not negate effect modifiers.
+Therefore, for marginal treatment effects to be interpretable, there must be a well-defined population.
+Most often, our main interest is in causal effects on the population of COVID-19 positive.
+Randomization within the clinical trial design yields internal validity, but we also need external validity to estimate the correct marginal effect of interest~\citep{Keiding2016}.
+
 \end{document}

### 2020-08-26 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 
+% https://www.regenstrief.org/covid-dashboard/
+%
+
 \subsection{US COVID-19 diagnostic and antibody testing}
@@
 Among 15,495 randomly selected persons, 3,695 participated of whom 3,629 had at least one test result available.
+
+Once a person showed up to our mobile testing sites, they were given both a PCR swab test that looks for current infections and an antibody blood test that looks for evidence of past infection.
 
@@
 
-When presented with non-probability samples, one can often reduce bias in two ways.
+The prior section focused on the negative consequences.
+
+The question is "What can be done".
+
+Here we present a technique for combining epidemilogical forecasts, IPW built using auxiliary information, and measurement-error corrections to generate 'doubly robust' estimates of the prevalence (true case count).
+
+We provide an appropriate extension of the statistical decomposition.
+Remark that the robustness is under certain assumptions which we know are likely false; therefore, we are interested in
+
+\subsection{Selection propensity estimation}
+
+With non-probability samples, one can often reduce bias in two ways.
 First, one may model the propensity for self-selection and use inverse probability weighting (IPW)~(cite: Elliott) techniques to adjust for selection bias.
-Unlike in causal inference~(cite: Robins) where the observational data permits direct estimation of propensities, one does not observe those who decide not to be  part of the .
-Instead, we need auxiliary information.
-Here, we leverage US census and the Indiana random testing demographic information to adjust the propensities.
-
-First generate the counts per cell using raking on the marginal summary statistics.  Suppose the selection indicator could be modeled parametrically and depends only on the strata $P($
-
-Start by using raking to get the ``number'' of individuals per strata.
-\begin{rmk}[Raking]
-All we get is testing demographic data.
-Indiana provides.
-Checked and it is often available upon digging.
-Demographic data allows us to build stratified estimators.
-We start by assuming that $d_k$ is the number of individuals in each strata is known.
-\end{rmk}
-
-Suppose we wanted to fit a logistic regression with the outcome being the selection indicator $Y_i$ given the strata information, coded as $X_i \in [K]$ for simplicity.
-$$
-\sum_{i=1}^N R_i \log \left( \pi (X_i; \theta) \right) + (1-R_i) \log \left( 1 - \pi (X_i; \theta) \right)
-= \sum_{i=1}^N R_i \log \left( \frac{\pi (X_i; \theta)}{1-\pi(X_i; \theta)} \right) + \sum_{i=1}^N \log \left( 1 - \pi (X_i; \theta) \right)
-$$
-The first term involves quantities observed in the nonprobability sample and can be re-expressed as sums over the strata.
-The second term, however, depends on the entire population.
-We can re-write this as a sum over strata as well $\sum_{k=1}^K N_k \log \left( 1- \pi_k \right)$ where $N_k$ is the number of individuals in strata $k$.
-The strata totals may be available from Census information.
-Alternatively, one could use the probability sample to construct a design-unbiased estimator of this quantity $\sum_k N \cdot \tilde d_k/n \log ( 1- \pi_k )$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.
-Under the assumptoin that $\pi_k < 1$, this leads to two potential estimators
-$$
-\hat \pi_k^{(1)} = \frac{d_k}{N_k}, \quad \hat \pi_k^{(2)} = \frac{d_k}{N} \frac{n}{\tilde d_k}.
-$$
-The assumptionWe start by assuming that $\pi_k < 1$ for all $k$.
+In causal inference~(cite: Robins), given certain ignorability assumptions, the observational data permits direct estimation of treatment propensity.
+In the COVID-19 context, one does not observe those who decide not to be tested.
+In order to estimate selection propensities, one instead needs auxiliary information.
+Here we present two methods for propensity estimation -- the first leveraging population-level summary statistics and the second leveraging the probability sample.
+
+Let $X_j$ denote a vector of covariates for the $j$th individual in the population, $j=1,\ldots,N$.
+Suppose the selection indicator $I_j$ is a Bernoulli random variable that depends only on these covariates, i.e., $P(I_j = 1 \mid X_j = x) = \pi (x; \theta)$.
+Maximum likelihood estimation follows by maximizing
+\begin{equation}
+\label{eq:propensity}
+\sum_{i=1}^N R_i \log \left( \frac{\pi (X_i; \theta)}{1-\pi(X_i; \theta)} \right) + \sum_{i=1}^N \log \left( 1 - \pi (X_i; \theta) \right).
+\end{equation}
+The first sum only involves individuals observed in the nonprobability sample.
+The second sum is over the entire population.
+Maximimum likelihood estimation therefore requires knowledge of covariate information for every individual in the population.
+Typically, this is not possible.
+We next present two options that are possible for analysis of selection propensities using only publically available COVID-19 case count data.
+
+\subsubsection{Auxiliary information through population-level summary statistics}
+
+For COVID-19 case-count data, the only covariate information provided by state and local governments are summary statistics for those factors discussed in Section XX -- age, sex, race, and (potentially) co-morbities.
+These are discrete factors and therefore define a stratification of the population; then we can write $X_i \in \{1,\ldots,K\}$ to denote the individual belongs to strata $k$.
+One strata, for example, could be defined as anyone who is greater than or equal to 60 years of age, male, and white.
+$$
+\sum_{k=1}^K d_k \log \left( \frac{\pi (k; \theta)}{1-\pi(k; \theta)} \right) + \sum_{k=1}^K N_k \log \left( 1 - \pi (k; \theta) \right).
+$$
+where $d_k$ and $N_k$ are the number of individuals in the nonprobabilistic sample and in the entire population respectively in the $k$th strata, $k=1,\ldots K$.
+
+Maximum likelihood estimation then requires knowledge of $N_k$, which is feasible when population-level summary statistics are available.
+The US Census, for example, can provide marginal summary statistics for the distribution of age, sex, and race in Indiana state.
+For co-morbitities such as hypertension, diabetes, and chronic obstructive pulmonary disease (COPD).
+Unfortunately the joint distribution is not provided.
+To circumvent this, we employ the standard technique of raking (cite: paper) to generate estimates $\hat N_k$ for the strata information.
+Estimation can then be performed by using standard algorithms such as the Newton-Rhapson methods.
+In the particular instance that the selection propensity is estimated non-parametrically then the resulting weight can be computed exactly and is given by $\hat P(I_j = 1 \mid X_j = k) = \hat \pi_k = d_k/\hat N_k$.
+
+\subsubsection{Auxiliary information through probability samples}
+
+(cite: Chen)
+Currently US testing does not collect additional covariate information; however, this need not be the case.
+Recommendation XX below states that additional covariate information should be collected from nonprobabilistic samples.
+If this were the case, we could not longer rely on population-level summary statistics so an alternative is required.
+
+Here, we assume access to a probability sample from the same population on which the outcome and exact same set of covariates are measured.
+Let $\tilde R_i$ be an indicator that the individual was included in the probability sample and $\tilde W_i$ be the probability of inclusion for the $i$th individual (i.e., equal to $n/N$ for a simple random sample of size $n$).
+The probability sample is used to construct a design-unbiased estimator of the second term
+$$
+\sum_{i=1}^N R_i \log \left( \frac{\pi (X_i; \theta)}{1-\pi(X_i; \theta)} \right)  + \sum_{i=1}^N \tilde R_i \tilde W_i \log ( 1 - \pi (X_i; \theta)).
+$$
+Taking the expecation of the second term with respect to the sampling design yields the second term in~\eqref{eq:propensity}.
+Recall Indiana's probabilistic sample was a stratified random sampling design; therefore, the weights will depend on strata (i.e., 10 strata for each public health preparadness district)
+
+In the simple stratified case with a non-parametric selection propensity, the weight is given by estimator of the
+quantity $\sum_k N \cdot \tilde d_k/n \log ( 1- \pi_k )$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.
+
+\subsubsection{An IPW estimator}
+
+Given the
+
+Then our new estimator is
+$$
+\bar y_n^\star
+= \frac{1}{1-FP-FN} \cdot \frac{1}{N} \sum_{i=1}^n \frac{y_i - FP}{\pi (x_i; \hat \theta)}
+= \frac{1}{1-FP-FN} \sum_{k=1}^K \frac{d_k}{N \pi_k} (\bar y_k - FP).
+$$
+where $k$ indexes the strata and $d_k$ is the number of samples in strata $k$.
+
 This is not true (need to deal with very bad symptoms cases).
-We unfortunately don't have $\tilde d_k$ nor $d_k$, but we can try and get these into using standard techniques (raking).
 
@@
 Then $N = 6.85$ million while $n = 3658$ leading to $\pi_k = 0.24$.
-
-Then our new estimator is
-$$
-\frac{1}{1-FP-FN} \sum_{k} d_k (\tilde y - FP) / \pi_k.
-$$
-where $k$ indexes the strata and $d_k$ is the number of samples in strata $k$.
 
@@
 
+In Section XX, we estimate $\rho_{I,Y}$ and $\rho_{\tilde I, Y}$ to see the impact of IPW.
+
 \subsection{Model-based estimation}
@@
 This yields estimates at $\hat y_k$
+
+\subsection{Doubly robust estimation}
 
@@
 See Section~\ref{section:est_dq} for discussion of how to estimate data quality.
-
-\section{Alternative estimators}
 

### 2020-08-27 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 Fifth, we discuss a sensitivity analysis method for estimating data quality and its use in understanding the potential biases in COVID-19 case-count analyses.
-We then end the paper with a discussion of the benefits of randomized testing and use of auxiliary information to improve over simple random sampling.
+{\bf We then end the paper with a discussion of the benefits of randomized testing and use of auxiliary information to improve over simple random sampling.
+NEED TO MAKE CLEAR THAT WE CAN MAKE USE OF RANDOM SAMPLING TO IMPROVE INFERENCE FROM NONPROB SAMPLES}
 
@@
 \subsection{Imperfect testing}
+\label{section:imperfecttesting}
 
@@
 
-\section{Doubly robust estimation}
-
-The prior section focused on the negative consequences.
-
-The question is "What can be done".
-
-Here we present a technique for combining epidemilogical forecasts, IPW built using auxiliary information, and measurement-error corrections to generate 'doubly robust' estimates of the prevalence (true case count).
-
-We provide an appropriate extension of the statistical decomposition.
-Remark that the robustness is under certain assumptions which we know are likely false; therefore, we are interested in
+\section{Potential improvements to prevalence estimation}
+
+The prior section presented negative consequences of analysing the measurement-error adjusted prevalence from observed case-count data.
+This raises the immediate question as to whether prevalence estimators can be improved using publicly available data.
+In this section, we consider two directions.
+First, we use auxiliary information to build selection propensities and consider an inverse probability weighting (IWP) estimator.
+We extend the statistical error decomposition to the weighted setting.
+Second, we consider an extension of compartmental models from infectious disease epidemiology that adjust for selection bias and measurement-error as model-based forecasts of prevalence.
+We then combine these epidemiological forecasts, selection propensities, and measurement-error corrections to generate \emph{doubly robust} prevalence estimators.
+We provide an appropriate extension of the statistical error decomposition and compare to the other prior decompositions.
+The proposed method and corresponding statistical error decompositions guide our recommendations in Section XX.
 
@@
 In order to estimate selection propensities, one instead needs auxiliary information.
-Here we present two methods for propensity estimation -- the first leveraging population-level summary statistics and the second leveraging the probability sample.
+Here we present two methods for propensity estimation -- use of population-level summary statistics and use of a probability sample.
 
@@
 Taking the expecation of the second term with respect to the sampling design yields the second term in~\eqref{eq:propensity}.
-Recall Indiana's probabilistic sample was a stratified random sampling design; therefore, the weights will depend on strata (i.e., 10 strata for each public health preparadness district)
-
-In the simple stratified case with a non-parametric selection propensity, the weight is given by estimator of the
-quantity $\sum_k N \cdot \tilde d_k/n \log ( 1- \pi_k )$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.
-
-\subsubsection{An IPW estimator}
-
-Given the
-
-Then our new estimator is
+Recall Indiana's probabilistic sample was a stratified random sampling design; therefore, the weights will depend on strata (i.e., 10 strata for each public health preparadness district).
+Under simple random sampling and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the $\hat \pi_k = (d_k / \tilde d_k) (n / N)$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.
+
+\subsubsection{An IPW estimator and statistical error decomposition.}
+Given a selection propensity, define a weight~$W(x) = \pi (x; \hat \theta)^{-1}$. Then the IPW estimator adjusted for measurement error is given by
 $$
 \bar y_n^\star
-= \frac{1}{1-FP-FN} \cdot \frac{1}{N} \sum_{i=1}^n \frac{y_i - FP}{\pi (x_i; \hat \theta)}
-= \frac{1}{1-FP-FN} \sum_{k=1}^K \frac{d_k}{N \pi_k} (\bar y_k - FP).
-$$
-where $k$ indexes the strata and $d_k$ is the number of samples in strata $k$.
-
-This is not true (need to deal with very bad symptoms cases).
-
-Simplest case we have Age $<40$ in the random sample is $1,017$ and we have $13+17.3+15.2 = 45\% \times 1,002,038$ tests so $450917.1$ tests.
-Then $N = 6.85$ million while $n = 3658$ leading to $\pi_k = 0.24$.
-
-\subsection{Statistical decomposition of error}
-
-For $i=1,\ldots,N$ let $\tilde Y_i = (Y_i - FP)/ (1- FP -FN)$ and $\tilde I_i = I_i / \pi_{K_i}$.  Then the error can now be expressed as:
-$$
-\rho_{\tilde I, Y} \times \sqrt{\frac{1-f+ CV^2_W}{f}} \sigma_{Y} \times D_M
-$$
+= \frac{1}{1-FP-FN} \cdot \frac{\sum_{i=1}^n w(x_i) (y_i - FP)}{\pi (x_i; \hat \theta)}
+\stackrel{(2)}{=} \frac{1}{1-FP-FN} \sum_{k=1}^K \frac{d_k}{N \pi_k} (\bar y_k - FP),
+$$
+where the second equality (2) is under the assumption that $x_i$ is a stratification variable with $k$ indexing the strata and $d_k$ is the number of samples in strata $k$.
+Under simple random sampling with one strata~$\hat \pi_k = n/N$ and we recover the estimator $\tilde y_n$ from Section~\ref{section:imperfecttesting}.
+
+Let $\tilde I_j = I_j / (N \cdot \pi (X_j; \theta)) := I_j W_j$ for $j=1,\ldots,N$.
+Then we can express $\bar y_n^\star$ as a weighted estimator $(1-FP-FN)^{-1} \sum_{i=1}^n $Then the error when comparing weighted estimator~$\bar y_n^\star$ to the true prevalence $\bar Y$ can be expressed as:
+$$
+\rho_{\tilde I, Y} \times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y} \times \underbrace{\left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{D_M} \times \frac{E_J[\tilde I_J]}{N}
+$$
+where $CV_W$ is the coefficient of variation (i.e., standard deviation/mean) of $W_J$ given $R_J = 1$.  See Appendix XX for the derivation.
+
+Interestingly, the impact of measurement-error on data quality is unchanged when considering a weighted estimand.  Similar to~\cite{Meng2018}, the data quality index....
 
@@
 
+Until now, this paper focused primarily on selection bias in coronavirus case-count data from a purely statistical perspective.
+Here, we extend the approach to whether it is possible bring in simple models from infectious diease epidemiology.
+Our goal is simply an estimate of $\bar y_k$, the prevalence in strata $k$.
+
+We rely heavily on recently available software~\cite{Song2020} for estimation of such models.
+
 We circle back and add in a SEIR model with measurement-error.
@@
 With such limited data we can only predict total counts and use the demographics as is to be representative.
-This yields estimates at $\hat y_k$
+This yields estimates at $\hat y_k$ for each strata.
 
@@
 Setting $\pi_i = n/N$ and $\hat y_k = 0$ yields the estimator $(\bar y_n - FP)/(1-FP-FN)$.
+
+Corresponding standard errors are
+$$
+\text{Var} (\bar y_{DR}^\star) =
+$$
+See Appendix XX for a derivation.
+
+{\bf where to describe difference from Chen}.
+
+Remark that the robustness is under certain assumptions which we know are likely false; therefore, we are interested in
 
@@
 
+\subsection{COVID-19 active infection prevalence in Indiana}
+
+We have a small non-random sample to supplement and get these; then use raking on the larger study.  Appendix XX presents the marginal distributions for the non-random sample per
+
+Heuristically, one may suppose that asymptomatic individuals are less likely to select.  This is born out.
+
+This is not true (need to deal with very bad symptoms cases).
+
+Simplest case we have Age $<40$ in the random sample is $1,017$ and we have $13+17.3+15.2 = 45\% \times 1,002,038$ tests so $450917.1$ tests.
+Then $N = 6.85$ million while $n = 3658$ leading to $\pi_k = 0.24$.
+
+\begin{rmk}[Use of second nonrandom sample]
+The key issue with selection bias is whether the individual is displaying symptoms and had contact with someone who was COVID+.  These are not captured by the
+
+In random sample, $44.2$\% of individuals who tested positive for SARS-CoV-2 positive individuals reported no symptoms during the 2 weeks before testing.
+
+In nonrandom sample, $20.2$\% of individuals who tested positive for SARS-CoV-2 reported no symptoms during the 2 weeks before testing.
+
+\end{rmk}
+
 \subsection{Estimation of Data Quality}
 \label{section:est_dq}
+
+SEROPREVALENCE (i.e., used to be infected).  We can attempt to estimate using our methods and assume non-overlapping samples.  Then back out the correlation.
 
@@
 - Smaller random samples over time (at least as testing and prevalence changes drastically) to assess selection changes
-- Record symptom/asymptomatic status!
+- Record symptom/asymptomatic status and COVID contact.  If not for all, then for at least some we need these covariates!!!!!
 

### 2020-08-27 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 
+(Watch MICHAEL MINA VIDEO again)
 There are two types of tests.  The first is
@@
 When an individual is infected with SARS-CoV-2 there is a
+
+When someone becomes infected, there will be an incubation period which may last between two to five days (cite: NPR) before the indivudal turn son the molecular/viral test.
+Very quickly the viral load will be high.
+
+Seven days post-symptom, the individual starts producing anti-bodies and so the seroprevalance test may be positive. A week later lasting a
 
@@
 $$
-Taking the expecation of the second term with respect to the sampling design yields the second term in~\eqref{eq:propensity}.
-Recall Indiana's probabilistic sample was a stratified random sampling design; therefore, the weights will depend on strata (i.e., 10 strata for each public health preparadness district).
+Taking the expectation of the second term with respect to the sampling design yields the second term in~\eqref{eq:propensity}.
+Recall Indiana's probabilistic sample was a stratified random sampling design; therefore, the weights will depend on strata (i.e., 10 strata for each public health preparedness district).
 Under simple random sampling and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the $\hat \pi_k = (d_k / \tilde d_k) (n / N)$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.
@@
 \subsubsection{An IPW estimator and statistical error decomposition.}
-Given a selection propensity, define a weight~$W(x) = \pi (x; \hat \theta)^{-1}$. Then the IPW estimator adjusted for measurement error is given by
+Given a selection propensity, define the inverse probability weight~$w(x) = \pi (x; \hat \theta)^{-1}$. Then the IPW estimator adjusted for measurement error is given by
 $$
 \bar y_n^\star
-= \frac{1}{1-FP-FN} \cdot \frac{\sum_{i=1}^n w(x_i) (y_i - FP)}{\pi (x_i; \hat \theta)}
-\stackrel{(2)}{=} \frac{1}{1-FP-FN} \sum_{k=1}^K \frac{d_k}{N \pi_k} (\bar y_k - FP),
-$$
-where the second equality (2) is under the assumption that $x_i$ is a stratification variable with $k$ indexing the strata and $d_k$ is the number of samples in strata $k$.
+= \frac{1}{1-FP-FN} \cdot \frac{\sum_{i=1}^n w(x_i) (y_i - FP)}{\sum_{i=1}^n w (x_i)}
+\stackrel{(2)}{=} \frac{1}{1-FP-FN} \sum_{k=1}^K \frac{d_k w_k}{w} (\bar y_k - FP),
+$$
+where the second equality (2) is under the assumption that $x_i$ is a stratification variable with $k$ indexing the strata, $w_k$ is the weight and $d_k$ is the number of samples in strata $k$, and $w = \sum_{k=1}^K d_k w_k$.
 Under simple random sampling with one strata~$\hat \pi_k = n/N$ and we recover the estimator $\tilde y_n$ from Section~\ref{section:imperfecttesting}.
 
-Let $\tilde I_j = I_j / (N \cdot \pi (X_j; \theta)) := I_j W_j$ for $j=1,\ldots,N$.
-Then we can express $\bar y_n^\star$ as a weighted estimator $(1-FP-FN)^{-1} \sum_{i=1}^n $Then the error when comparing weighted estimator~$\bar y_n^\star$ to the true prevalence $\bar Y$ can be expressed as:
-$$
-\rho_{\tilde I, Y} \times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y} \times \underbrace{\left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{D_M} \times \frac{E_J[\tilde I_J]}{N}
+Let $\tilde I_j = I_j  \cdot W(X_j)$ for $j=1,\ldots,N$.  Then the error when comparing weighted estimator~$\bar y_n^\star$ to the true prevalence $\bar Y$ can be expressed as:
+$$
+\rho_{\tilde I, Y} \times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y} \times \underbrace{\left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{D_M}
 $$
@@
 
-Until now, this paper focused primarily on selection bias in coronavirus case-count data from a purely statistical perspective.
-Here, we extend the approach to whether it is possible bring in simple models from infectious diease epidemiology.
-Our goal is simply an estimate of $\bar y_k$, the prevalence in strata $k$.
-
+So far the focus has primarily been on selection bias in coronavirus case-count data from a purely survey sampling perspective.
+Here, we consider  compartmental model approaches from infectious disease epidemiology.
+Our primary objective is a model-based forecast of strata-level prevalence $\bar y_k$ using data up to but not including the current time.
+
+Here, we present briefly a basic epidemiological state-space model -- the susceptible, infected, and removed (recovered and death) model, or SIR model.
+The SIR model was originally proposed by (cite: Osthus, Forecasting seasonal influenza with a state-space sir model) with only one-dimensional time series of infected proportions.
+We refer the reader to Chapters 9-12 of [30] for a substantive introduction to state-space models.
+
+More advanced versions of the SIR model can be used to make epidemiological forecasts.
+Here, we focus on the SIER model due to availability of existing software.
 We rely heavily on recently available software~\cite{Song2020} for estimation of such models.
+
+The current prediction does not (A)
 
@@
 
-\subsection{Doubly robust estimation}
-
+\subsection{Doubly robust estimation.}
+Two methods have been presented.  The first uses IPW to adjust for differential selection into the
+
+Rather than relying on either independently, here we show how to combine these.
+$$
+\sum_{i=1}^N \mu (x_i; \hat \beta) + \frac{1}{\sum_{i=1}^n w(x_i)}
+$$
+
+There are again two options.  In the first,
 $$
@@
 
+In the second, we rely on
+
 Corresponding standard errors are
@@
 
-{\bf where to describe difference from Chen}.
-
 Remark that the robustness is under certain assumptions which we know are likely false; therefore, we are interested in
@@
 $$
+
+{\bf where to describe difference from Chen. (1) weight normalized, (2) use aux info from multiple sources and use raking, (3) use a epi model rather than covariate pred model}.
 

### 2020-08-28 write-up/covid-writeup.tex
Update covid-writeup.tex
Can't use song software?

@@
 Figure~\ref{fig:ratio-bias} displays the trajectory of the true ratio and the potential biased estimators under an SIR model~\citep{Pastor2001,Newman2002,Parshani2010} for the epidemic dynamics, with state evolution given by
-$$
-\begin{aligned}
+\begin{equation}
+\label{eq:sir}
 \frac{\partial S_t}{\partial t} &= - \beta S_t I_t N^{-1}; \quad
@@
 \frac{\partial R_t}{\partial t} = - \gamma I_t \\
-\end{aligned}
 $$
@@
 
+Here we follow closely the description from~\cite{Song2020}.
+Let $Y_t^I$ and $Y_t^R$ denote the proportion of currently infected (i.e., would test positive given a RT-PCR test with 100\% sensitivity) and removed cases (i.e., including both recovered cases and deaths) at time $t$.
+Assume the two-dimensional process follows
+$$
+\theta_t \mid \theta_{t-1}; \tau \sim \text{Dirichlet} \left(\kappa f(\theta_{t-1}, \beta, \gamma) \right)
+$$
+where $\kappa$ scales the variance of the Dirichlet distribution and $f(\cdot)$ is a function that determines infection dynamics.
+Here, we assume the classical infectious disease SIR model as given by~\eqref{eq:sir} with $(S_t,I_t, R_t)$ replaced by $(\theta_t^S, \theta_t^I, \theta_t^R)$.
+\cite{Song2020}
+
+A key concern with the proposed model is lack of .
+We could address this by using IPW methods but this would mimic Section XX and therefore lead to potentially similar results.
+Instead, we opt to follow (cite: HDSR) and model death data.
+This avoids the selection bias problem by instead trying to back)
+
+$$
+D_t \mid p, \theta, \nu \sim \text{Poisson} \left( p \sum_{t=1}^r \nu_t \theta_{(r-t)} \right)
+$$
+
+Unlike in (cite: HDRS), we are interested in predictions of case-counts.  We split deaths into $<60$ and $>60$, using two different IFR's to back out the counts for each subset.  Then we recombine
+
 More advanced versions of the SIR model can be used to make epidemiological forecasts.
@@
 
-The current prediction does not (A)
-
-We circle back and add in a SEIR model with measurement-error.
 We can correct for selection bias two through simple rules (see Song and Mukherjee and others for more complex alternatives).
@@
 
+Alternative forecasts are
+
 \subsection{Doubly robust estimation.}
-Two methods have been presented.  The first uses IPW to adjust for differential selection into the
-
-Rather than relying on either independently, here we show how to combine these.
-$$
-\sum_{i=1}^N \mu (x_i; \hat \beta) + \frac{1}{\sum_{i=1}^n w(x_i)}
-$$
+Two methods have been presented.
+The first uses IPW to adjust for differential selection.
+The second uses epidemiological forecasts based on SIR models.
+Rather than relying solely on either approach, here we show how to combine forecasting and inverse-probability weighting by building a \emph{doubly robust} approach (cite: Robins, etc), which is a common approach for causal inference from observational data.
+In the survey literature, if the weights are known, the approach is referred to as model-assisted estimation.
+Recent work by (cite: Chen) extends the classical survey approach to non-probability samples.
+Here, we do the same.
+We reserve comparison of our approach with (citE: Chen) after introducing the method.
+
+We define the doubly-robust estimator as
+$$
+\bar y_{n}^{(DR)} = \frac{1}{N} \sum_{i=1}^N \mu (x_i; \hat \beta) + \frac{1}{\sum_{i=1}^n w(x_i)} \sum_{j=1}^N I_j w(x_i) (Y_i - \mu(x_i; \hat \beta)).
+$$
+If the model is correct, then $\bar y_{n}^DR - \bar Y_N$ will be zero.  If the inverse-probability weights
 

### 2020-08-28 write-up/covid-writeup.tex
Update covid-writeup.tex
realworld attempt

@@
 The World Health Organization has declared the coronavirus disease 2019 (COVID-19) a public health emergency.
-As of June 2nd, 2020, over 6 million cases have been confirmed worldwide with 1.8 million cases and over 100 thousand confirmed deaths across the United States.
+As of August 28th, 2020, over 24.5 million cases have been confirmed worldwide with 5.95 million cases and over 180 thousand confirmed deaths across the United States.
 This pandemic
@@
 
-This paper has two goals.
-The first is to express reservations at the use of case-counts as a proxy for disease prevalence and in estimation of standard epidemiological models for inference and forecasting.
+This paper has two main objectives.
+The first objective is to express reservations at the use of case-counts as a proxy for disease prevalence and in estimation of standard epidemiological models for inference and forecasting.
 The reason is straightforward: observed case-count data is plagued by selection bias and measurement error.
@@
 
-The second is to show how probabilistic samples can also be used to derive a doubly robust approach that leverages epidemiological models along with pseudo survey weights built from combining nonprobability and probability samples.
-Using a random sample from Indiana, we show that  for several key covariates and, given several assumptions, produce corrected estimates of the prevalence and true case-counts.
-We end with several remarks on cautious use of these methods.
+A real-world constraint for countries, states, and municipalities is the monetary and time cost of random sampling.
+The second objective of this paper is to demonstrate how random samples provide necessary auxiliary information to address selection bias in coronavirus case-count data.
+
+Using a random sample from Indiana, we construct improved  corrected estimates of the prevalence and true case-counts.  We end with several remarks on cautious use of these methods.
 
@@
 
+Model-assisted surveys.
+
 Combining nonprob and prob samples
 
-Epi forecasting
-
-\subsection{Objectives and outline}
-
-This article discusses the relationship between three statistical concepts: selection bias, measurement error, and the too often forgotten population size.
-We clarify mathematically why the situation is much more complex than it first appears.
-Through simple mathematical arguments, we clarify five important issues related to COVID-19 case-count analysis.
-First, we show that unadjusted prevalence rates are (unsurprisingly) biased when tests are imperfect.
+The coronavirus research literature has exploded. One large component of the volume of research is epidemiological forecasting.  Here we focus on
+These forcasts are often used to estimate effective reproduction number.  They can also be used to estimate the   estimating prevalence and true case count,
+
+\subsection{Outline}
+
+This article discusses the relationship between three statistical concepts: selection bias, measurement error, and population size.
+
+In Section XX, we present simple mathematical arguments to show that unadjusted prevalence rates are (unsurprisingly) biased when tests are imperfect.
 What is surprising is that the direction and magnitude of the bias can vary substantially and interacts with the sampling fraction.
-Next we show that data quality for adjusted prevalence rates includes an additional term in the error decomposition that captures the interplay among measurement error, selection bias, and prevalence.
-Third, we show that daily trends in rate of positive tests is equally problematic even if testing rates and selection protocols are held constant.
 We show that the rate of change in COVID-19 observed case-counts overestimates the true rate of change prior to the peak time and underestimates it immediately after.
@@
 We show that estimates of the effective reproduction number may be similarly impacted.
+
+In Section XX, we present the doubly robust estimation method for combining epidemiological forecasts with nonprobability and probability samples to estimate prevalence
+
   Fourth, we show that cross-country comparisons are difficult at best with population size, sampling fraction, and data quality all interacting to impact null hypothesis testing.
 Fifth, we discuss a sensitivity analysis method for estimating data quality and its use in understanding the potential biases in COVID-19 case-count analyses.
-{\bf We then end the paper with a discussion of the benefits of randomized testing and use of auxiliary information to improve over simple random sampling.
-NEED TO MAKE CLEAR THAT WE CAN MAKE USE OF RANDOM SAMPLING TO IMPROVE INFERENCE FROM NONPROB SAMPLES}
+
+In Section XX, we apply the derived IPW and doubly-robust estimation methods to the case study of Indiana statewide prevalence from March 1st until August 27th.
+We estimate the data quality for IPW estimates using the random sample on April 28th.
+
+In Section XX, we provide a series of recommendations to help improve statistical inference in settings with large nonprobability samples from a stochastic process where small random sampling over time is possible.
 
@@
 \label{eq:sir}
-\frac{\partial S_t}{\partial t} &= - \beta S_t I_t N^{-1}; \quad
+\frac{\partial S_t}{\partial t} = - \beta S_t I_t N^{-1}; \quad
 \frac{\partial I_t}{\partial t} = \beta S_t I_t N^{-1} - \gamma I_t; \quad
-\frac{\partial R_t}{\partial t} = - \gamma I_t \\
-$$
+\frac{\partial R_t}{\partial t} = - \gamma I_t
+\end{equation}
 where $S_t$ is the number of susceptible individuals, $I_t$ is the number of infected individuals, and $R_t$ is the number of removed (recovered or deceased) individuals at time $t$.
@@
 
-The prior section presented negative consequences of analysing the measurement-error adjusted prevalence from observed case-count data.
-This raises the immediate question as to whether prevalence estimators can be improved using publicly available data.
+The prior section presented negative consequences of selection-bias and measurement error when estimating prevalence from observed case-count data.
+This raises the immediate question whether prevalence estimators can be improved using publicly available data.
 In this section, we consider two directions.
@@
 We extend the statistical error decomposition to the weighted setting.
-Second, we consider an extension of compartmental models from infectious disease epidemiology that adjust for selection bias and measurement-error as model-based forecasts of prevalence.
+Second, we consider an extension of compartmental models from infectious disease epidemiology using death data to try and avoid selection bias.
+After adjusting for measurement-error, this leads to a model-based forecasts of prevalence.
 We then combine these epidemiological forecasts, selection propensities, and measurement-error corrections to generate \emph{doubly robust} prevalence estimators.
-We provide an appropriate extension of the statistical error decomposition and compare to the other prior decompositions.
-The proposed method and corresponding statistical error decompositions guide our recommendations in Section XX.
+We provide an appropriate extension of the statistical error decomposition.
+The method and statistical error decompositions guide our recommendations in Section XX.
 
@@
 
-With non-probability samples, one can often reduce bias in two ways.
-First, one may model the propensity for self-selection and use inverse probability weighting (IPW)~(cite: Elliott) techniques to adjust for selection bias.
-In causal inference~(cite: Robins), given certain ignorability assumptions, the observational data permits direct estimation of treatment propensity.
+With non-probability samples, one can reduce bias by modelling the self-selection propensity and use inverse probability weighting (IPW)~(cite: Elliott) techniques to adjust for selection bias.
+In causal inference~(cite: Robins), under certain assumptions, the observational data permits direct estimation of treatment propensity.
 In the COVID-19 context, one does not observe those who decide not to be tested.
-In order to estimate selection propensities, one instead needs auxiliary information.
-Here we present two methods for propensity estimation -- use of population-level summary statistics and use of a probability sample.
-
-Let $X_j$ denote a vector of covariates for the $j$th individual in the population, $j=1,\ldots,N$.
+In order to estimate selection propensities, one needs auxiliary information.
+
+To see this mathematically, let $X_j$ denote a vector of covariates for the $j$th individual in the population, $j=1,\ldots,N$.
 Suppose the selection indicator $I_j$ is a Bernoulli random variable that depends only on these covariates, i.e., $P(I_j = 1 \mid X_j = x) = \pi (x; \theta)$.
@@
 The second sum is over the entire population.
-Maximimum likelihood estimation therefore requires knowledge of covariate information for every individual in the population.
+Maximum likelihood estimation therefore requires knowledge of covariate information for every individual in the population.
 Typically, this is not possible.
-We next present two options that are possible for analysis of selection propensities using only publically available COVID-19 case count data.
+Here we present two methods for propensity estimation -- use of population-level summary statistics and use of a probability sample.
 
@@
 
-For COVID-19 case-count data, the only covariate information provided by state and local governments are summary statistics for those factors discussed in Section XX -- age, sex, race, and (potentially) co-morbities.
-These are discrete factors and therefore define a stratification of the population; then we can write $X_i \in \{1,\ldots,K\}$ to denote the individual belongs to strata $k$.
-One strata, for example, could be defined as anyone who is greater than or equal to 60 years of age, male, and white.
+For COVID-19 case-count data, the only covariate information provided by state and local governments are summary statistics for those factors discussed in Section XX -- age, sex, race, and (potentially) comorbidities.
+These are discrete factors and therefore define a stratification of the population.
+Then $X_j = k \in \{1,\ldots,K\}$ denotes the $j$th individual belongs to strata $k$ and the likelihood can be re-written as
 $$
@@
 The US Census, for example, can provide marginal summary statistics for the distribution of age, sex, and race in Indiana state.
-For co-morbitities such as hypertension, diabetes, and chronic obstructive pulmonary disease (COPD).
+For comorbidities such as hypertension, diabetes, and chronic obstructive pulmonary disease (COPD) the Indiana State Department of Health releases summary statistics.
 Unfortunately the joint distribution is not provided.
-To circumvent this, we employ the standard technique of raking (cite: paper) to generate estimates $\hat N_k$ for the strata information.
-Estimation can then be performed by using standard algorithms such as the Newton-Rhapson methods.
-In the particular instance that the selection propensity is estimated non-parametrically then the resulting weight can be computed exactly and is given by $\hat P(I_j = 1 \mid X_j = k) = \hat \pi_k = d_k/\hat N_k$.
+To circumvent this, we employ the standard technique of raking (cite: paper) to generate estimates $\hat N_k$ for each strata.
+Estimation can then be performed by using standard algorithms such as Newton-Rhapson.
+If the selection propensity is estimated non-parametrically then the resulting weight can be computed exactly and is give nby $\hat P(I_j = 1 \mid X_j = k) = \hat \pi_k = d_k/\hat N_k$.
 
@@
 Instead, we opt to follow (cite: HDSR) and model death data.
-This avoids the selection bias problem by instead trying to back)
-
 $$
@@
 $$
-
-Unlike in (cite: HDRS), we are interested in predictions of case-counts.  We split deaths into $<60$ and $>60$, using two different IFR's to back out the counts for each subset.  Then we recombine
-
-More advanced versions of the SIR model can be used to make epidemiological forecasts.
-Here, we focus on the SIER model due to availability of existing software.
-We rely heavily on recently available software~\cite{Song2020} for estimation of such models.
-
-We can correct for selection bias two through simple rules (see Song and Mukherjee and others for more complex alternatives).
-With such limited data we can only predict total counts and use the demographics as is to be representative.
+Fully Bayesian algorithms that combine (cite: Song) and (cite: Johndrow) can be derived.
+Here, we focus on the deterministic case-count and resulting SIR model due to availability of existing software~\cite{Song2020} for estimation of such models.
+Unlike in (cite: HDRS), we are interested in predictions of case-counts.
+We split deaths into $<60$ and $>60$, using two different IFR's to back out the counts for each subset and then split these among the sub-populations using the raking weights.
 This yields estimates at $\hat y_k$ for each strata.
-
-Alternative forecasts are
 
@@
 $$
-\bar y_{n}^{(DR)} = \frac{1}{N} \sum_{i=1}^N \mu (x_i; \hat \beta) + \frac{1}{\sum_{i=1}^n w(x_i)} \sum_{j=1}^N I_j w(x_i) (Y_i - \mu(x_i; \hat \beta)).
-$$
-If the model is correct, then $\bar y_{n}^DR - \bar Y_N$ will be zero.  If the inverse-probability weights
-
-There are again two options.  In the first,
-$$
-\sum_{k} P_k \hat y_k + \frac{1}{N} \sum_{i=1}^N I_i \frac{(Y_i - FP)/(1-FP-FN) - \hat y}{\pi_{K_i}}
-$$
-where $P_k$ is the population fraction in strata $k$.
+\bar y_{n}^{(DR)} = \frac{1}{1 - FP - FN} \left[ \frac{1}{N} \sum_{i=1}^N \mu (x_i; \hat \beta) + \frac{1}{\sum_{i=1}^n w(x_i)} \sum_{j=1}^N I_j w(x_i) (Y_i - \mu(x_i; \hat \beta)) \right] - FP.
+$$
 This estimator is ``doubly-robust'' in the sense that it is robust to either the SIR or the propensities being correctly specified.
-Setting $\pi_i = n/N$ and $\hat y_k = 0$ yields the estimator $(\bar y_n - FP)/(1-FP-FN)$.
-
-In the second, we rely on
-
-Corresponding standard errors are
+Setting $\pi_i = n/N$ and $\mu(x_i; \hat \beta) = 0$ we again recover the estimator $\tilde y_n$ from Section~\ref{section:imperfecttesting}.
+
+The key issue with a doubly-robust estimator (cite: guy from Chen who assumes) is we again don't observe $X_j$ for each $j=1,\ldots,N$.
+There are again two options.
+If we rely on auxiliary information through population-level summary statistics, we have
+$$
+\frac{1}{1 - FP - FN} \left[ \sum_{k} P_k \hat y_k + \sum_{k=1}^K \frac{d_k w_k}{w} (\bar y_k - \hat y_k) \right] - FP
+$$
+where $P_k$ is the population fraction in strata $k$ and $\hat y_k = \mu(k; \hat \beta)$. Alternatively, we can use auxiliary information through probability samples
+$$
+\frac{1}{1-FP-FN} \left[ \frac{1}{N} \sum_{i=1}^N \tilde R_i \tilde W_i \mu (X_i; \hat \theta) + \sum_{i=1}^n \frac{w_i}{w} \left(y_i - \mu(x_i; \hat \beta)\right) \right] - FP.
+$$
+
+A plug-in estimator for the variance
 $$
@@
 $$
-\left( P_k - \frac{n_k}{N \pi_k} \right) \hat y_k + \frac{n_k}{N \pi_k} \frac{\bar y_k - FP}{1-FP-FN}
+\left( P_k - \frac{d_k w_k}{w} \right) \hat y_k + \frac{d_k w_k}{w} \frac{\bar y_k - FP}{1-FP-FN}
 $$

### 2020-08-28 write-up/covid-writeup.tex
Update covid-writeup.tex
better intro

@@
 
-A real-world constraint for countries, states, and municipalities is the monetary and time cost of random sampling.
-The second objective of this paper is to demonstrate how random samples provide necessary auxiliary information to address selection bias in coronavirus case-count data.
-
-Using a random sample from Indiana, we construct improved  corrected estimates of the prevalence and true case-counts.  We end with several remarks on cautious use of these methods.
+Due to monetary and time constraints, random testing may be performed infrequently.
+As of August 28, 2020, Indiana is the only state to conduct statewide random sample testing, which was collected from April 25--29, 2020.
+Such infrequent random testing may be insufficient to understand the trajectory of infectious disease prevalence which may change rapidly over time.
+Therefore, while random testing is preferred, governments, researchers, and policy makers continue to look toward the coronavirus case-counts to understand the impact of COVID-19 on the population and make data-informed decisions.
+
+The second objective is to demonstrate how random samples provide necessary auxiliary information to address selection bias in coronavirus case-count data.
+Random samples are used in estimation of selection propensities which are then employed in an inverse-probability weighting scheme to construct adjusted estimators of disease prevalence.
+A doubly robust extension allows researchers to combine these estimates with epidemiological forecasts based on compartmental models common in the study .
+While we demonstrate improvements over simple estimates of disease prevalence, we highlight how selection bias may persist.
 

### 2020-08-28 write-up/covid-writeup.tex
Update covid-writeup.tex
adding covid description

--- 2020-08-28 write-up/covid-writeup.tex
@@
 
-A real-world constraint for countries, states, and municipalities is the monetary and time cost of random sampling.
-The second objective of this paper is to demonstrate how random samples provide necessary auxiliary information to address selection bias in coronavirus case-count data.
-
-Using a random sample from Indiana, we construct improved  corrected estimates of the prevalence and true case-counts.  We end with several remarks on cautious use of these methods.
+Due to monetary and time constraints, random testing may be performed infrequently.
+As of August 28, 2020, Indiana is the only state to conduct a statewide random sample testing, which was collected from April 25--29, 2020.
+Infrequent random testing may be insufficient to understand the trajectory of infectious diseases which change rapidly over time.
+Therefore, while random testing may be preferable in theory, in practice governments, researchers, and policy makers continue to use coronavirus case-counts to understand the impact of COVID-19 on the population and make data-informed decisions.
+
+The second objective is to demonstrate how random samples provide necessary auxiliary information to address selection bias in coronavirus case-count data.
+Random samples provide the necessary covariate information from a representative sample from the population to estimate selection propensities.
+These propensities are then used in an inverse-probability weighting scheme to construct estimators of disease prevalence that attempt to control for selection bias.
+A doubly robust extension allows researchers to combine these estimates with epidemiological forecasts based on compartmental models that are common in the study of infectious diseases (cite: book).
+While we demonstrate improvements over simple estimates of disease prevalence, we highlight how selection bias may persist.
 
@@
 
-In Section XX, we present simple mathematical arguments to show that unadjusted prevalence rates are (unsurprisingly) biased when tests are imperfect.
+In Section~\ref{section:data}, we summarize the publicly available data on COVID-19.
+We highlight how testing information is collected as well as demographic and comorbidity information.
+In Section~\ref{section:testinginfo}, we briefly summarize COVID-19 testing.
+We clarify the different between viral and antibody testing.
+For viral testing, we briefly discuss the reasons behind measurement-error in RT-PCR tests including test timing and sample collection techniques.
+
+In Section~\ref{section:casecount}, we present simple mathematical arguments to show that unadjusted prevalence rates are (unsurprisingly) biased when tests are imperfect.
 What is surprising is that the direction and magnitude of the bias can vary substantially and interacts with the sampling fraction.
@@
 
-In Section XX, we present the doubly robust estimation method for combining epidemiological forecasts with nonprobability and probability samples to estimate prevalence
-
-  Fourth, we show that cross-country comparisons are difficult at best with population size, sampling fraction, and data quality all interacting to impact null hypothesis testing.
-Fifth, we discuss a sensitivity analysis method for estimating data quality and its use in understanding the potential biases in COVID-19 case-count analyses.
-
-In Section XX, we apply the derived IPW and doubly-robust estimation methods to the case study of Indiana statewide prevalence from March 1st until August 27th.
-We estimate the data quality for IPW estimates using the random sample on April 28th.
-
-In Section XX, we provide a series of recommendations to help improve statistical inference in settings with large nonprobability samples from a stochastic process where small random sampling over time is possible.
+In Section~\ref{section:improvedcasecount}, we present a doubly robust estimation method for combining epidemiological forecasts with inverse-probability weighting methods to estimate prevalence.
+Weight estimation requires auxiliary information provided by random samples.
+A statistical error decomposition illustrates a trade-off in use of weights: a potential increase in data quality but a reduction in data quantity.
+
+In Section~\ref{section:applications}, the IPW and doubly-robust estimation methods are applied to the case study of Indiana statewide prevalence from March 1st until August 27th.
+We estimate data quality for IPW estimates using the random sample on April 28th.
+We then discuss a sensitivity analysis method for estimating data quality and its use in understanding the potential biases in COVID-19 case-count analyses by analysing New York statewide prevalence and comparing to the results from their statewide random testing on April 23rd.
+Both demonstrate improvements over simple prevalence estimation but that data quality is not of the same order as random sampling.
+In Section~\ref{section:crosspop}, the case study results are used to suggest cross-country comparisons are difficult at best with population size, sampling fraction, and data quality all interacting to impact hypothesis testing.
+
+In Section~\ref{section:discussion}, a series of recommendations are provided that aim to help improve statistical inference in settings with large nonprobability samples from a stochastic process where small random sampling over time is possible.
 
 \section{COVID-19 data}
-
-In this paper, we start with case-count data as reported by US state governments (cites).  Figure~\ref{fig:ny-covid}
-
-Case-count data
-
-Generally available public data (COVID trackers).
-
-COVID trackers tend to provide basic information.  From the JHU dashboard,
-
-Most states provide additional demographic data.
-
-We focus on strata built on age, race, and sex.  Only have marginal information!
-
+\label{section:data}
+
+Coronavirus case-count data refers to the number of positive tests performed on each day.
+Figure~\ref{fig:ny-covid} plots (dotted orange line) the number of reported confirmed COVID-19 cases per day in the state of Indiana.
+Departments of Health in each state also typically report the total number of reported tests per day.
+On August 5th, 2020, for example, there were 1075 total daily positive cases and 11,663 total daily tests.
+Public databases maintained by \href{https://bit.ly/2UqFSuA}{Johns Hopkins University} and the \href{https://bit.ly/2vUHfrK}{New York Times} provide incoming county-level information of confirmed cases and deaths.
+
+Many states provide aggregate demographic information on those who requested a test as well as those who tested positive for SARS-CoV-2.
+Indiana's \href{https://www.coronavirus.in.gov/2393.htm}{COVID-19 dashboard}, for instance, provides statewide demographics for age, gender, ethnicity, and race.
+The demographic information is based on cumulative testing.
+Daily versions cannot be easily reconstructed from available public data.
+In addition, the Regenstrief Institute provides additional comorbidity information among those who tested positive for COVID-19 in the state of Indiana, including hypertension, diabetes, and chronic obstructive pulmonary disease (COPD).
+These comprise the main set of covariate information available from public sources for Indiana.
+Similar demographic information is likely available for other states.
 % https://www.regenstrief.org/covid-dashboard/
@@
 
-\subsection{US COVID-19 diagnostic and antibody testing}
+\subsection{Viral and antibody testing}
+\label{section:testinginfo}
 
@@
 When an individual is infected with SARS-CoV-2 there is a
+
+A viral test such as the RT-PCR
 
@@
 
-The key issue with selective testing is that $E_{\I} [ \rho_{I,Y} ] \neq 0$.
+% The key issue with selective testing is that $E_{\I} [ \rho_{I,Y} ] \neq 0$.
 Meng identified this as the fundamental issue that can lead to paradoxes in the analysis of big data.
@@
 \section{Potential improvements to prevalence estimation}
+\label{section:improvedcasecount}
 
@@
 
+\section{Applications to binary outcome and the COVID-19 pandemic.}
+\label{section:applications}
+
+\subsection{COVID-19 active infection prevalence in Indiana}
+
+We have a small non-random sample to supplement and get these; then use raking on the larger study.  Appendix XX presents the marginal distributions for the non-random sample per
+
+Heuristically, one may suppose that asymptomatic individuals are less likely to select.  This is born out.
+
+This is not true (need to deal with very bad symptoms cases).
+
+Simplest case we have Age $<40$ in the random sample is $1,017$ and we have $13+17.3+15.2 = 45\% \times 1,002,038$ tests so $450917.1$ tests.
+Then $N = 6.85$ million while $n = 3658$ leading to $\pi_k = 0.24$.
+
+\begin{rmk}[Use of second nonrandom sample]
+The key issue with selection bias is whether the individual is displaying symptoms and had contact with someone who was COVID+.  These are not captured by the
+
+In random sample, $44.2$\% of individuals who tested positive for SARS-CoV-2 positive individuals reported no symptoms during the 2 weeks before testing.
+
+In nonrandom sample, $20.2$\% of individuals who tested positive for SARS-CoV-2 reported no symptoms during the 2 weeks before testing.
+
+\end{rmk}
+
 \section{Cross-population comparisons}
+\label{section:crosspop}
 So far we have focused on understanding the limitations of using case-count data to understand population quantities of interest for a \emph{single} population.
@@
 
-\section{Applications to binary outcome and the COVID-19 pandemic.}
-
-\subsection{COVID-19 active infection prevalence in Indiana}
-
-We have a small non-random sample to supplement and get these; then use raking on the larger study.  Appendix XX presents the marginal distributions for the non-random sample per
-
-Heuristically, one may suppose that asymptomatic individuals are less likely to select.  This is born out.
-
-This is not true (need to deal with very bad symptoms cases).
-
-Simplest case we have Age $<40$ in the random sample is $1,017$ and we have $13+17.3+15.2 = 45\% \times 1,002,038$ tests so $450917.1$ tests.
-Then $N = 6.85$ million while $n = 3658$ leading to $\pi_k = 0.24$.
-
-\begin{rmk}[Use of second nonrandom sample]
-The key issue with selection bias is whether the individual is displaying symptoms and had contact with someone who was COVID+.  These are not captured by the
-
-In random sample, $44.2$\% of individuals who tested positive for SARS-CoV-2 positive individuals reported no symptoms during the 2 weeks before testing.
-
-In nonrandom sample, $20.2$\% of individuals who tested positive for SARS-CoV-2 reported no symptoms during the 2 weeks before testing.
-
-\end{rmk}
-
 \subsection{Estimation of Data Quality}
@@
 
-\section{Discussion.}
+\section{Discussion}
+\label{section:discussion}
 There is nothing routine about COVID-19, including the corresponding statistical questions.
@@
 I would say this is important but JHU is not enough; see above for benefits of random sampling to augment and improve decision making.
+
+\subsection*{Frequency and turnaround time}
+
+This supposes no selection bias.
+
+What is your goal?
+"get people who are currently sick off the street." Then test frequently and often.
+But we dont live in a mandatory testing environment (nor is it currently capable).
+Random testing is one way to improve upone current practice whiel we wait for that
 

### 2020-08-28 write-up/covid-writeup.tex
Update covid-writeup.tex
almost done with covid testing section!

@@
 These comprise the main set of covariate information available from public sources for Indiana.
-Similar demographic information is likely available for other states.
+Similar demographic information is likely available for other states.co
 % https://www.regenstrief.org/covid-dashboard/
@@
 
-(Watch MICHAEL MINA VIDEO again)
+Here we briefly describe the difference between COVID-19 molecular and serological tests as well as the relationship between COVID-19 testing and the infection and immune response time frame.
+These distinctions will be useful later in the paper when discussing different random sample tests.
+
+When an individual is infected with SARS-CoV-2, there will typically be an incubation period which may last between two to five days (cite: NPR) before the viral load is high enough to be detected.
+A molecular test refers to tests that look increased viral load in the patient (such as RT-PCR or antigen tests).
+These are tests such that if someone has an active infection, a positive result is expected after the incubation period.
+A test with perfect sensitivity will come back positive for the next few weeks (cite: wlos).
+At that point, the viral load will decrease and a molecular test with perfect specificity will come back negative.
+
+Serological (or antibody) tests look for an immunological response to the virus.
+A week or so after an individual is infected with SARS-CoV-2, the individual will start producing antibodies.
+At this point, a serological test with perfect sensitivity will come back positive.
+This test provides evidence of past infection while the molecular test provides evidence of an \emph{active} infection.
+
+While an individual may test positive on a molecular test for several weeks, it is important to note that the transmissibility o fth
+This is particularly important for epidemiological forecasts build from compartmental models.
+
+Recent systematic reviews have indicated, however, that  (cite:  NEJM)
 There are two types of tests.  The first is
@@
 
-When someone becomes infected, there will be an incubation period which may last between two to five days (cite: NPR) before the indivudal turn son the molecular/viral test.
-Very quickly the viral load will be high.
-
 Seven days post-symptom, the individual starts producing anti-bodies and so the seroprevalance test may be positive. A week later lasting a
 
-During the first two days, an infected individual will not
-An infected individual will
-
 %https://www.npr.org/sections/goatsandsoda/2020/05/31/865932474/should-i-get-tested-for-coronavirus-just-for-the-heck-of-it
- test positive three to five days after exposure, the individual enters the stage at which detection is possible
 

### 2020-08-30 write-up/covid-writeup.tex
Update covid-writeup.tex
working on covid data section

@@
 The calls to increase testing capacity are well-justified as they become a more frequent point of discussion in the public sphere.
-While expanded testing is a laudable goal, selection bias will impact estimates of disease prevalence and the effective reproduction number until the entire population is sampled.
-Moreover, tests are imperfect as false positive/negative rates interact in complex ways with selection bias.
+While expanded testing is a laudable goal, selection bias will impact estimates of active infection rate and the effective reproduction number until the entire population is sampled.
+Moreover, tests are imperfect.
+Test sensitivity and specificity interact in complex ways with selection bias.
 Indeed, measurement error and selection bias are the two greatest problems limiting our understanding of the current COVID-19 pandemic; neither can be fully addressed by increased testing capacity.
@@
 
-\section{COVID-19 data}
+\section{COVID-19 testing and data}
 \label{section:data}
 
-Coronavirus case-count data refers to the number of positive tests performed on each day.
+In this section we provide the necessary background to understand COVID-19 diagnostic and serologic testing, their respective scientific uses in managing the pandemic, and the corresponding data streams.
+
+\subsection{Diagnostic versus serologic testing}
+\label{section:testinginfo}
+
+% We start with a brief description of the difference between COVID-19 molecular and serological tests as well as the relationship between COVID-19 testing, the infection, and immune response time frame.
+These distinctions will be useful later in the paper when discussing different random sample tests.
+
+When an individual is infected with SARS-CoV-2, there will typically be an incubation period which may last between two to five days (cite: NPR) before the viral load is high enough to be detected.
+A \emph{molecular test} refers to tests that aim to detect increased viral load (e.g., RT-PCR or antigen tests).
+If someone has an active infection, then after the incubation period a molecular test with perfect sensitivity will yield a positive result for the next few weeks (cite: wlos).
+After that, the viral load will decrease and a molecular test with perfect specificity will come back negative.
+
+While an infected individual may yield a positive molecular test for several weeks, they will likely stop transmitting the disease after the first few days after infection, meaning a positive molecular test does not imply transmissibility.
+This is of particular importance for epidemiological forecasts built from population-level data and impacts our model choice in Section~\ref{section:modelbased}.
+
+Molecular tests are referred to as diagnostic tests.
+
+Serological (or antibody) tests look for an immunological response to the virus.
+A week or so after an individual is infected with SARS-CoV-2, the individual will start producing antibodies.
+At this point, a serological test with perfect sensitivity will come back positive.
+This test provides evidence of past infection while the molecular test provides evidence of an \emph{active} infection.
+
+Seroprevalence studies are useful for disease surveillance.
+
+To date, however, the number of seroprevalance surveys are limited.
+
+See XX for further discussion.
+
+Rapid and frequent molecular testing is cited as a key component (citEe: OECD and others) in effective strategies to identify active infections and prevent further spread.  ,
+
+Significant logistic and capacity constraints have
+
+Our primary goal is to estimate accurately estimate the fraction of active infections in the population.
+
+This is the most relevant
+Secondary goals include estimation of rates of change and the effective reproduction number which are also useful in policy making.
+
+Alternatively,
+Such testing would result in higher detection rates.
+Will these results be reported to local authorities.
+If not, selection bias in who uses these tests and then reports the results is likely to matter.
+
+When an individual is infected with SARS-CoV-2 there is a
+
+A viral test such as the RT-PCR
+
+Seven days post-symptom, the individual starts producing anti-bodies and so the seroprevalance test may be positive. A week later lasting a
+
+\subsection{COVID-19 case counts, viral testing counts, and deaths}
+\label{section:testinginfo}
+
+Coronavirus case-count data refers to the number of positive diagnostic tests performed on each day.
 Figure~\ref{fig:ny-covid} plots (dotted orange line) the number of reported confirmed COVID-19 cases per day in the state of Indiana.
@@
 Public databases maintained by \href{https://bit.ly/2UqFSuA}{Johns Hopkins University} and the \href{https://bit.ly/2vUHfrK}{New York Times} provide incoming county-level information of confirmed cases and deaths.
+
+{\bf key point: WHAT IS THE GOAL OF CASE COUNT DATA?
+WHAT IS THE GOAL OF SEROPREVALENCE STUDIES?
+are we using it for that reason.
+cse count data is colelcted as part of surveillance programs wehre teh goal is to identify and isolate active infections.
+However, many people are trying to use these to udnerstand the disease.
+This is our issue.}
 
@@
 
-\subsection{Viral and antibody testing}
-\label{section:testinginfo}
-
-Here we briefly describe the difference between COVID-19 molecular and serological tests as well as the relationship between COVID-19 testing and the infection and immune response time frame.
-These distinctions will be useful later in the paper when discussing different random sample tests.
-
-When an individual is infected with SARS-CoV-2, there will typically be an incubation period which may last between two to five days (cite: NPR) before the viral load is high enough to be detected.
-A molecular test refers to tests that look increased viral load in the patient (such as RT-PCR or antigen tests).
-These are tests such that if someone has an active infection, a positive result is expected after the incubation period.
-A test with perfect sensitivity will come back positive for the next few weeks (cite: wlos).
-At that point, the viral load will decrease and a molecular test with perfect specificity will come back negative.
-
-Serological (or antibody) tests look for an immunological response to the virus.
-A week or so after an individual is infected with SARS-CoV-2, the individual will start producing antibodies.
-At this point, a serological test with perfect sensitivity will come back positive.
-This test provides evidence of past infection while the molecular test provides evidence of an \emph{active} infection.
-
-While an individual may test positive on a molecular test for several weeks, it is important to note that the transmissibility o fth
-This is particularly important for epidemiological forecasts build from compartmental models.
-
-Recent systematic reviews have indicated, however, that  (cite:  NEJM)
-There are two types of tests.  The first is
-
-When an individual is infected with SARS-CoV-2 there is a
-
-A viral test such as the RT-PCR
-
-Seven days post-symptom, the individual starts producing anti-bodies and so the seroprevalance test may be positive. A week later lasting a
-
 %https://www.npr.org/sections/goatsandsoda/2020/05/31/865932474/should-i-get-tested-for-coronavirus-just-for-the-heck-of-it
+
+\subsubsection{Molecular testing restrictions in Indiana}
 
@@
 \subsection{Model-based estimation}
+\label{section:modelbased}
 

### 2020-08-30 write-up/covid-writeup.tex
Update covid-writeup.tex
closer- just getting question right

@@
 Weight estimation requires auxiliary information provided by random samples.
-A statistical error decomposition illustrates a trade-off in use of weights: a potential increase in data quality but a reduction in data quantity.
+A statistical error decomposition illustrates a trade-off in use of weights: a potential increase in data quality but a reduction in effective data quantity.
 
@@
 
-When an individual is infected with SARS-CoV-2, there will typically be an incubation period which may last between two to five days (cite: NPR) before the viral load is high enough to be detected.
-A \emph{molecular test} refers to tests that aim to detect increased viral load (e.g., RT-PCR or antigen tests).
-If someone has an active infection, then after the incubation period a molecular test with perfect sensitivity will yield a positive result for the next few weeks (cite: wlos).
+After an individual is infected with SARS-CoV-2, there is an incubation period which lasts between two to five days (cite: NPR) before the viral load is high enough to be detected.
+A \emph{molecular test} refers to diagnostic tests that aim to detect increased viral load (e.g., RT-PCR or antigen tests).
+If someone has an active infection then after the incubation period a molecular test with perfect sensitivity will yield a positive result for the next few weeks (cite: wlos).
 After that, the viral load will decrease and a molecular test with perfect specificity will come back negative.
-
-While an infected individual may yield a positive molecular test for several weeks, they will likely stop transmitting the disease after the first few days after infection, meaning a positive molecular test does not imply transmissibility.
-This is of particular importance for epidemiological forecasts built from population-level data and impacts our model choice in Section~\ref{section:modelbased}.
-
-Molecular tests are referred to as diagnostic tests.
-
-Serological (or antibody) tests look for an immunological response to the virus.
+While an infected individual may yield a positive molecular test for several weeks, they will likely stop transmitting the disease within a few days of infection, meaning a positive molecular test does not imply transmissibility.
+
+A molecular test conducted on an infected individual may return a negative result.
+Such false negatives are very strongly associated with when the test is conducted.
+In the incubation phase, most molecular tests will return a false negative result.
+Molecular tests are most sensitive when the viral loads are highest which occurs during the first few days of transmissibility (cite: ).
+Moreover, most molecular tests are performed via nasopharyngeal swab (cite: NEJM).
+Specimen collection by swab is known to impact false negative/positive rates regardless of test timing.
+Systematic reviews suggest that 70\% sensitivity and 95\% specificity are reasonable estimates for current RT-PCR tests (cite: NEJM).
+
+The primary goal is of molecular tests is in diagnosis of active infections in the population.
+Such diagnostic tests generate important information about the presence of SARS-Cov-2 in the population, and help scientists and policy-makers understand patterns of transmission and propagation.
+Rapid and frequent molecular testing is cited as a key component (citEe: OECD and others) in effective strategies to identify active infections and prevent systemic outbreaks.
+There has been work on developing faster, cheaper testing options (cite: Mina) that can be used at-home as well as group testing (citE: Mina); however, these developments are beyond the scope of this paper.
+Here we consider the current system of testing which involves both self-selection into testing and historical limits on who was able to receive diagnostic tests due to limited testing capacity.
+
+\emph{Serological tests} look for an immunological response to the virus.
 A week or so after an individual is infected with SARS-CoV-2, the individual will start producing antibodies.
@@
 
-Seroprevalence studies are useful for disease surveillance.
-
-To date, however, the number of seroprevalance surveys are limited.
-
-See XX for further discussion.
-
-Rapid and frequent molecular testing is cited as a key component (citEe: OECD and others) in effective strategies to identify active infections and prevent further spread.  ,
-
-Significant logistic and capacity constraints have
-
-Our primary goal is to estimate accurately estimate the fraction of active infections in the population.
-
-This is the most relevant
-Secondary goals include estimation of rates of change and the effective reproduction number which are also useful in policy making.
-
-Alternatively,
-Such testing would result in higher detection rates.
-Will these results be reported to local authorities.
-If not, selection bias in who uses these tests and then reports the results is likely to matter.
-
-When an individual is infected with SARS-CoV-2 there is a
-
-A viral test such as the RT-PCR
-
-Seven days post-symptom, the individual starts producing anti-bodies and so the seroprevalance test may be positive. A week later lasting a
-
-\subsection{COVID-19 case counts, viral testing counts, and deaths}
+To better estimate SARS-CoV-2 immunity in a population, seroprevalence studies that generate a probabilistic population sample and perform serological tests on the sample can be collected.
+These studies are useful for disease surveillance.
+When the number of active infections in a population are low, serological surveys are cheaper and more accurate than molecular tests and are more useful in detecting outbreaks (cite: Mina).
+To date, the CDC has conducted ten large-scale geographic serological surveys with three rounds.
+While population-based sampling strategies provide a more representative nationwide sample, they are very time intensive and expensive.
+Here we consider whether non-probability diagnostic testing can be used in estimation of prevalence of immunity.
+
+\subsection{Publicly available data on COVID-19}
 \label{section:testinginfo}
 
-Coronavirus case-count data refers to the number of positive diagnostic tests performed on each day.
-Figure~\ref{fig:ny-covid} plots (dotted orange line) the number of reported confirmed COVID-19 cases per day in the state of Indiana.
-Departments of Health in each state also typically report the total number of reported tests per day.
-On August 5th, 2020, for example, there were 1075 total daily positive cases and 11,663 total daily tests.
-Public databases maintained by \href{https://bit.ly/2UqFSuA}{Johns Hopkins University} and the \href{https://bit.ly/2vUHfrK}{New York Times} provide incoming county-level information of confirmed cases and deaths.
-
-{\bf key point: WHAT IS THE GOAL OF CASE COUNT DATA?
+In this paper, our primary goal is to accurately estimate the active infection rate in the population over time using \emph{publicly available} test results.
+Secondary goals include estimation of rates of change in these quantities and the effective reproduction number (cite:) which characterize the infectious disease spread.
+A final goal is using these estimates to characterize prevalence of immunity in the general population.
+As previously stated, these quantities are fundamental to public health policy and general critical information on presence and transmission of SARS-CoV-2 as well as the impact of public health  interventions.
+
+% {\bf key point: WHAT IS THE GOAL OF CASE COUNT DATA?
 WHAT IS THE GOAL OF SEROPREVALENCE STUDIES?
@@
 This is our issue.}
+
+Coronavirus case-count data refers to the number of positive molecular tests performed on each day.
+Figure~\ref{fig:ny-covid} plots (dotted orange line) the number of reported confirmed COVID-19 cases per day in the state of Indiana.
+Total number of COVID-19 diagnostic tests performed per day is also reported (Figure~\ref{fig:ny-covid}, dotted blue line).
+On August 5th, 2020, for example, there were 1075 total daily positive cases and 11,663 total daily tests.
+Public databases maintained by \href{https://bit.ly/2UqFSuA}{Johns Hopkins University} and the \href{https://bit.ly/2vUHfrK}{New York Times} provide accessible incoming county-level information of confirmed cases and deaths.
 
@@
 
+This is of particular importance for epidemiological forecasts built from population-level data and impacts our model choice in Section~\ref{section:modelbased}.
+{\bf Check if SIR models account for this?}
+
 \subsection{Doubly robust estimation.}

### 2020-08-30 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 There has been work on developing faster, cheaper testing options (cite: Mina) that can be used at-home as well as group testing (citE: Mina); however, these developments are beyond the scope of this paper.
-Here we consider the current system of testing which involves both self-selection into testing and historical limits on who was able to receive diagnostic tests due to limited testing capacity.
+% Here we consider the current system of testing which involves both self-selection into testing and historical limits on who was able to receive diagnostic tests due to limited testing capacity.
 
@@
 While population-based sampling strategies provide a more representative nationwide sample, they are very time intensive and expensive.
-Here we consider whether non-probability diagnostic testing can be used in estimation of prevalence of immunity.
 
@@
 
-In this paper, our primary goal is to accurately estimate the active infection rate in the population over time using \emph{publicly available} test results.
+In this paper, our primary goal is accurate estimates of the population-level active infection rates over time using \emph{publicly available} testing data.
 Secondary goals include estimation of rates of change in these quantities and the effective reproduction number (cite:) which characterize the infectious disease spread.
@@
 
-Many states provide aggregate demographic information on those who requested a test as well as those who tested positive for SARS-CoV-2.
+Many states also provide aggregate demographic information on those who requested a test as well as those who tested positive for SARS-CoV-2.
 Indiana's \href{https://www.coronavirus.in.gov/2393.htm}{COVID-19 dashboard}, for instance, provides statewide demographics for age, gender, ethnicity, and race.
 The demographic information is based on cumulative testing.
-Daily versions cannot be easily reconstructed from available public data.
-In addition, the Regenstrief Institute provides additional comorbidity information among those who tested positive for COVID-19 in the state of Indiana, including hypertension, diabetes, and chronic obstructive pulmonary disease (COPD).
+In addition, the Regenstrief Institute provides additional comorbidity information among those who tested positive for COVID-19 in the state of Indiana, including hypertension, diabetes, and chronic obstructive pulmonary disease (COPD) summary statistics.
 These comprise the main set of covariate information available from public sources for Indiana.
-Similar demographic information is likely available for other states.co
+Similar demographic information is likely available for other states.
 % https://www.regenstrief.org/covid-dashboard/
@@
 
-We focus on US COVID-19 testing.
-
-The typical testing
-
-However, upon investi
+Due to limited testing capacity, many US states have instituted restrictions on who can request a molecular test at a given time.
 %https://www.coronavirus.in.gov/
-As of today, access to the history of testing requirements remains quite difficult.
-Early during the coronavirus pandemic, limited testing meant that only those at high risk were allowed to be tested (i.e., symptomatic or high risk populations); however, as testing capacity expanded these restrictions were gradually lifted.
-Such restrictions
-
-Given the difficult, we focus on reconstructing this history for the state of Indiana.
-As of May 12th, 2020, the Indiana State Department of Health laboratory was testing high-risk individuals for COVID-19.
+As of August 30th, 2020, a repository of the history of testing requirements does not exist.
+Early on, limited testing meant that only those at high risk were allowed to be tested (i.e., symptomatic or high risk populations); as testing capacity expanded, these restrictions were gradually lifted.
+
+Here, we focus on reconstructing the testing restriction history for the state of Indiana.
+As of May 12th, 2020, the Indiana State Department of Health (ISDH) laboratory was testing high-risk individuals for COVID-19.
 Any symptomatic individual could receive a tested.

### 2020-08-30 write-up/covid-writeup.tex
Update covid-writeup.tex
closer but i need to get my samples better

@@
 
-\subsection{Probabilistic samples}
-
-Indiana random sample description (see paper for more details)
-
-Among 15,495 randomly selected persons, 3,695 participated of whom 3,629 had at least one test result available.
-
-Once a person showed up to our mobile testing sites, they were given both a PCR swab test that looks for current infections and an antibody blood test that looks for evidence of past infection.
+\subsection{Probabilistic sample in Indiana}
+
+Between April 25--29, 2020, Indiana conducted statewide random sample testing (both molecular and serological) of persons ages $\geq 12$ years to assess prevalence of active infection and presence of antibodies to SARS-CoV-2 (cite: Indiana report).
+A stratified random sampling design was conducted using Indiana’s 10 public health preparedness districts as sampling strata.
+15,495 participants were contacted resulting in a final sample size of 3,658.
+The same demographic data as provided by ISDH for Indian molecular testing data was included (e.g., summary statistics on age, sex, and race).
+Participants in the probability sample were also asked if they experienced any COVID-19 compatible symptoms during the past 2 weeks or had shared a household with someone who had a positive test result for SARS-CoV-2.
+
+During May 2--3, 2020, an additional nonrandom sample of 898 individuals was also collected.
+In this paper, the nonrandom sample is used to assess whether the cumulative summary statistics are appropriate in constructing the inverse-probability weights as well as in constructing appropriate weights using symptom information.
 
@@
 
-The prior section presented negative consequences of selection-bias and measurement error when estimating prevalence from observed case-count data.
-This raises the immediate question whether prevalence estimators can be improved using publicly available data.
+The prior section presented negative consequences of selection-bias and measurement error when estimating infection prevalence from observed case-count data.
+This raises the immediate question whether prevalence estimators can be improved.
 In this section, we consider two directions.
-First, we use auxiliary information to build selection propensities and consider an inverse probability weighting (IWP) estimator.
+First, we use publicly available auxiliary information to build selection propensities and consider an inverse probability weighting (IWP) estimator.
 We extend the statistical error decomposition to the weighted setting.

### 2020-09-01 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 %\bibliographystyle{apalike}
+\usepackage{multirow}
 
@@
 \begin{abstract}
-  Coronavirus case-counts have influenced governments on their policies and decision making, while researchers have used them to drive their forecast models and predictions.
-Throughout this, many have also cited limited testing as the key driver of weak information on the COVID-19 pandemic.
-The calls to increase testing capacity are well-justified as they become a more frequent point of discussion in the public sphere.
-While expanded testing is a laudable goal, selection bias will impact estimates of active infection rate and the effective reproduction number until the entire population is sampled.
-Moreover, tests are imperfect.
-Test sensitivity and specificity interact in complex ways with selection bias.
-Indeed, measurement error and selection bias are the two greatest problems limiting our understanding of the current COVID-19 pandemic; neither can be fully addressed by increased testing capacity.
-In this paper, we attempt to clarify this point.
-Through simple calculations, we demonstrate pitfalls and paradoxes that can arise when considering case-count data in the presence of selection bias and measurement error.
-\textcolor{red}{Need to update}
+  Coronavirus case-count data has influenced government policies and drives many epidemiological forecasts.
+Limited testing is cited as the key driver behind minimal information on the COVID-19 pandemic.
+While expanded testing is laudable, measurement error and selection bias are the two greatest problems limiting our understanding of the current COVID-19 pandemic; neither can be fully addressed by increased testing capacity.
+In this paper, we demonstrate their impact on estimation of point prevalence and the effective reproduction number.
+We show that estimates based on the millions of molecular tests in the US has the same mean square error as a genuine small simple random sample.
+To address this, a procedure is presented that combines case-count data with small random samples over time to estimate selection propensities based on key covariate information.
+We then combine these selection propensities with epidemiological forecast models using death-only data to construct a \emph{doubly-robust} estimation method that accounts for both measurement-error and selection bias.
+This method is then applied this to the United States's only statewide random molecular sample collected from April 25--29th in Indiana.
+We end with a series of recommendations based on the proposed methodology.
 \end{abstract}
@@
 In particular, the millions of tests in the US have a small effective sample size when compared to random sampling.
+These calculations demonstrate the importance of COVID-19 government-led data collection including random samples over time from the population.
 
 Due to monetary and time constraints, random testing may be performed infrequently.
-As of August 28, 2020, Indiana is the only state to conduct a statewide random sample testing, which was collected from April 25--29, 2020.
+As of August 28, 2020, Indiana is the only state to conduct a statewide random sample testing\footnote{This is the only random sample to collect both seroprevalence and diagnostic testing results.
+The CDC has conducted several seroprevalence studies.
+See Section~\ref{section:testinginfo} for further discussion.}, which was collected from April 25--29, 2020.
 Infrequent random testing may be insufficient to understand the trajectory of infectious diseases which change rapidly over time.
@@
 A doubly robust extension allows researchers to combine these estimates with epidemiological forecasts based on compartmental models that are common in the study of infectious diseases (cite: book).
-While we demonstrate improvements over simple estimates of disease prevalence, we highlight how selection bias may persist.
+While we demonstrate improvements over simple estimates of disease prevalence, we highlight how selection bias may persist and impact uncertainty quantification.
 
@@
 
-Meng
-
-Model-assisted surveys.
-
-Combining nonprob and prob samples
-
-The coronavirus research literature has exploded. One large component of the volume of research is epidemiological forecasting.  Here we focus on
-These forcasts are often used to estimate effective reproduction number.  They can also be used to estimate the   estimating prevalence and true case count,
+This article discusses the relationship between three statistical concepts: selection bias, measurement error, and population size.
+We build on the work of \cite{Meng2018} who studied an error decomposition to understand the relationship between selection bias and population size.
+In particular, we quantify the interaction between measurement-error and selection bias.
+The sign and magnitude of the statistical decomposition can change drastically; the impact on the effective sample size can therefore be quite large.
+
+After demonstrating the limitations of case-count analysis when compared to random sampling, we then assess whether their is potential for combining the \emph{nonprobability samples} (cite: nonprob, Elliott) with random samples from the population.
+(cite: Chen) considered combining nonprobability and probability samples.
+Here, we also rely on  samples
+
+A key issue is also that the random sampling will occur over time, potentially infrequently compared to the nonprobability sample.   Here
+We then provide an extension of the statistical error decomposition
+
+The coronavirus research literature has exploded. One large component of the volume of research is epidemiological forecasting.
+These forecast models are often also used to estimate the effective reproduction number.  We extend (cite: Johndrow) to handle the more complex SIR dynamics and covariate information.
 
 \subsection{Outline}
-
-This article discusses the relationship between three statistical concepts: selection bias, measurement error, and population size.
 
@@
 
-\subsubsection{Molecular testing restrictions in Indiana}
+\subsubsection{Testing restrictions and public health policy in Indiana}
 
@@
 As of May 12th, 2020, the Indiana State Department of Health (ISDH) laboratory was testing high-risk individuals for COVID-19.
-Any symptomatic individual could receive a tested.
+Any symptomatic individual could receive a test.
 Moreover there were key factors for people who can be tested regardless of symptom status: over the age of 65, diabetic, high-blood pressure, obesity, pregnancy, and minority populations at greater risk of severe illness.
@@
 % https://www.indystar.com/story/news/health/2020/06/12/indiana-says-anyone-who-wants-coronavirus-test-can-get-one/3179151001/
+
+On March 23, 2020, Indiana's governor issues a \emph{stay at home} order effective March 26th through April 5th.
+The order was extended until April 30th.
+On May 1st, 2020, a five-stage plan for gradual reopening was announced by Governon Holcomb -- See XX for exact details (cite: plan).
+Stage two went from May 4th to May 21st; stage three went from May 22nd to June 13th; stage 4 went from June 12th to July 3rd; stage 4.5 began on July 4th and will continue until September 25th; stage 5's start time has yet to be determined.
+Such  policies target reduction in active infection rates.
+In this paper, we incorporate Indiana's public health policy in our construction of epidemiological forecasts in Section~\ref{section:modelbased}.
 
@@
 Taking the expectation of the second term with respect to the sampling design yields the second term in~\eqref{eq:propensity}.
-Recall Indiana's probabilistic sample was a stratified random sampling design; therefore, the weights will depend on strata (i.e., 10 strata for each public health preparedness district).
 Under simple random sampling and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the $\hat \pi_k = (d_k / \tilde d_k) (n / N)$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.
@@
 In Section XX, we estimate $\rho_{I,Y}$ and $\rho_{\tilde I, Y}$ to see the impact of IPW.
+
+\subsubsection{Time-varying weights and random sampling}
+
+Suppose we have two random samples at times $t$ and $t+k$ (i.e., $k$ days apart).
 
@@
 This yields estimates at $\hat y_k$ for each strata.
+We use $p$ as 0.6 from the
 
@@
 \subsection{COVID-19 active infection prevalence in Indiana}
+
+Here we consider estimation of the active infection rate in Indiana.  We start with IPW estimation.  Table .
+
+The report does not include
+
+\begin{table}[!th]
+\begin{tabular}{c | c | l l | l | l l }
+& & \multicolumn{2}{c}{Total Tests} &  & \multicolumn{2}{c}{Positive Tests} \\
+\cline{3-4} \cline{6-7}
+& & NonRandom & Random & Census & NonRandom & Random \\ \hline
+\multirow{2}{*}{Sex$^\star$} & Female & 57\% & 55\% & 49\% & 53\% & - \\
+& Male & 43 & 45 & 51 & 47 & -  \\ \hline
+\multirow{3}{*}{Age$^\star$} & $<40$ & 47 & 28 & 53 & 47\\
+& $40-59$ & 28 & 36 & 25 & 30\\
+& $\geq 60$ & 25 & 36 & 22 & 13\\ \hline
+\multirow{2}{*}{Race$^\star$} & White & 60 & 98 & 85 & 50\\
+& Nonwhite & 21 & 8 & 15 & 50 \\ \hline
+\multirow{2}{*}{Symptoms$^{\star \star}$} & Yes & - & - & - \\
+& No & - & - & - \\ \hline
+\multirow{2}{*}{Contact$^{\star \star}$} & Yes & 11 & 1 & - \\
+& No & 89 & 99 & - \\ \hline
+\end{tabular}
+\caption{Temp \\
+$\star$: ISDH case-count nonrandom sample \\
+$\star\star$: May 2--3 nonrandom sample
+}
+\end{table}
+
+1. Unknown assumed MAR for Race, Gender, and Age
+2. 20\% are asymptomatic (so 78\% + 22\% (20\%) showed no symptoms)
 

### 2020-09-03 write-up/covid-writeup.tex
wrtieup taking form

@@
 In this paper, we demonstrate their impact on estimation of point prevalence and the effective reproduction number.
-We show that estimates based on the millions of molecular tests in the US has the same mean square error as a genuine small simple random sample.
+We show that estimates based on the millions of molecular tests in the US has the same mean square error as a small simple random sample.
 To address this, a procedure is presented that combines case-count data with small random samples over time to estimate selection propensities based on key covariate information.
 We then combine these selection propensities with epidemiological forecast models using death-only data to construct a \emph{doubly-robust} estimation method that accounts for both measurement-error and selection bias.
-This method is then applied this to the United States's only statewide random molecular sample collected from April 25--29th in Indiana.
+This method is then applied to estimate Indiana's prevalence using case-count and death data as well as the United States's only statewide random molecular sample collected from April 25--29th.
 We end with a series of recommendations based on the proposed methodology.
@@
 In particular, the millions of tests in the US have a small effective sample size when compared to random sampling.
-These calculations demonstrate the importance of COVID-19 government-led data collection including random samples over time from the population.
+These calculations demonstrate the importance of probabilistic sampling designs  over time for estimation of point prevalence and effective reproduction number.
 
@@
 These propensities are then used in an inverse-probability weighting scheme to construct estimators of disease prevalence that attempt to control for selection bias.
-A doubly robust extension allows researchers to combine these estimates with epidemiological forecasts based on compartmental models that are common in the study of infectious diseases (cite: book).
+A doubly robust extension allows researchers to combine these estimates with epidemiological forecasts based on compartmental models that are common in the study of infectious diseases~\cite{Hao2020,Song2020,Ray2020,Johndrow2020}.
 While we demonstrate improvements over simple estimates of disease prevalence, we highlight how selection bias may persist and impact uncertainty quantification.
@@
 
-After demonstrating the limitations of case-count analysis when compared to random sampling, we then assess whether their is potential for combining the \emph{nonprobability samples} (cite: nonprob, Elliott) with random samples from the population.
-(cite: Chen) considered combining nonprobability and probability samples.
-Here, we also rely on  samples
-
-A key issue is also that the random sampling will occur over time, potentially infrequently compared to the nonprobability sample.   Here
-We then provide an extension of the statistical error decomposition
-
-The coronavirus research literature has exploded. One large component of the volume of research is epidemiological forecasting.
-These forecast models are often also used to estimate the effective reproduction number.  We extend (cite: Johndrow) to handle the more complex SIR dynamics and covariate information.
+After demonstrating the limitations of case-count analysis when compared to random sampling, we then assess whether their is potential for combining the \emph{nonprobability samples} with \emph{probability samples} to improve point prevalence estimation.
+For any probability sampling design, the Horvitz-Thompson estimator~\citep{HT1952} incorporates design information via inverse-probability weights (IPW).
+For nonprobability samples, the IPW estimator requires modelling the propensity scores.
+Its use in the survey context is also referred to as the quasi-randomization approach~\citep{Elliott2017}.
+\cite{Valliant2011} consider a weighted logistic regression procedure using the pooled probability and nonprobability samples.
+\cite{Chen2019} consider a pseudo-log-likelihood approach that uses the random samples as a proxy for a term in the log-likelihood.
+Here, we extend this approach to account for measurement-error as well as random sampling at multiple times.
+We then provide an extension of the statistical error decomposition and discuss the trade-offs inherent in such a weighting approach.
+
+One core component of coronavirus research is epidemiological state-space modelling of case-count and death data.
+These models can be used to answer a variety of research questions including case-count forecasting, estimation of the effective reproduction number, and estimation of quarantine and other health policies on infectious disease dynamics.
+The basic approach is a deterministic compartmental model called the susceptible-infectious-recovered (SIR) model.
+A probabilistic extension was proposed by~\cite{Osthus2017} to model one-dimensional time series of infected proportions.
+\cite{Song2020} extended this approach to incorporate interventions and assess interventions on COVID-19 epidemic in China.
+\cite{Hao2020} builds an extended SEIR model (SAPHIRE) to account for various presymptomatic infectiousness, time-varying ascertainment rates, transmission rates and population movements.
+Given a probability sampling design, individual predictions can be leveraged to improve estimation via model-assisted approaches~\citep{Breidt2017}.
+For nonprobability samples, \cite{Chen2019} derive a doubly robust approach that uses outcome predictions given covariates on the nonprobability and probability samples.
+Here, we combine the state-space SIR approach of \cite{Song2020} but instead, as in \cite{Johndrow2020}, focus on COVID-19 confirmed death count data.
+We generate epidemiological forecasts for point prevalence of each population strata.
+We then demonstrate how to combine these forecasts with the IPW approach to construct doubly-robust estimates of point prevalence.
+A derived statistical error decomposition guides this discussion.
 
@@
 In Section~\ref{section:testinginfo}, we briefly summarize COVID-19 testing.
-We clarify the different between viral and antibody testing.
+We clarify the difference between viral and antibody testing.
 For viral testing, we briefly discuss the reasons behind measurement-error in RT-PCR tests including test timing and sample collection techniques.
@@
 
-\subsection{Probabilistic sample in Indiana}
+\subsection{Probabilistic samples in Indiana}
 
@@
 In this paper, the nonrandom sample is used to assess whether the cumulative summary statistics are appropriate in constructing the inverse-probability weights as well as in constructing appropriate weights using symptom information.
+
+{\bf Facebook symptoms. }
+Finally, since XX, Facebook has conducted random sample testing ().
+In particular, the COVID Symptom tracker provides an estimate of the number of individuals who have experienced any COVID-19 compatible symptoms.
 
@@
 We provide an appropriate extension of the statistical error decomposition.
-The method and statistical error decompositions guide our recommendations in Section XX.
+The method and statistical error decompositions guide our recommendations in Section~\ref{section:discussion}.
 
@@
 
-With non-probability samples, one can reduce bias by modelling the self-selection propensity and use inverse probability weighting (IPW)~(cite: Elliott) techniques to adjust for selection bias.
-In causal inference~(cite: Robins), under certain assumptions, the observational data permits direct estimation of treatment propensity.
+With non-probability samples, one can reduce bias by modelling the self-selection propensity and use inverse probability weighting (IPW)~\citep{Elliott2017} techniques to adjust for selection bias.
+In causal inference~\citep{Hernan2020}, under certain assumptions, the observational data permits direct estimation of treatment propensity.
 In the COVID-19 context, one does not observe those who decide not to be tested.
@@
 To see this mathematically, let $X_j$ denote a vector of covariates for the $j$th individual in the population, $j=1,\ldots,N$.
+For simplicity, the dynamic nature of the outbreak and recoverability of individuals is ignored for now.
 Suppose the selection indicator $I_j$ is a Bernoulli random variable that depends only on these covariates, i.e., $P(I_j = 1 \mid X_j = x) = \pi (x; \theta)$.
@@
 \label{eq:propensity}
-\sum_{i=1}^N R_i \log \left( \frac{\pi (X_i; \theta)}{1-\pi(X_i; \theta)} \right) + \sum_{i=1}^N \log \left( 1 - \pi (X_i; \theta) \right).
+\sum_{j=1}^N I_j \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi (X_j; \theta) \right).
 \end{equation}
@@
 
-(cite: Chen)
 Currently US testing does not collect additional covariate information; however, this need not be the case.
@@
 Taking the expectation of the second term with respect to the sampling design yields the second term in~\eqref{eq:propensity}.
+This was proposed by~\cite{Chen2019} when covariates were observed on a random sample from the population.
+Here, we only observe strata information.
 Under simple random sampling and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the $\hat \pi_k = (d_k / \tilde d_k) (n / N)$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.
@@
 
-Let $\tilde I_j = I_j  \cdot W(X_j)$ for $j=1,\ldots,N$.  Then the error when comparing weighted estimator~$\bar y_n^\star$ to the true prevalence $\bar Y$ can be expressed as:
-$$
-\rho_{\tilde I, Y} \times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y} \times \underbrace{\left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{D_M}
+Let $\tilde I_j (X_j) = I_j  \cdot W(X_j)$ for $j=1,\ldots,N$.  Then the error when comparing weighted estimator~$\bar y_n^\star$ to the true prevalence $\bar Y$ can be expressed as:
+\begin{equation}
+\label{eq:statdecomp2}
+\rho_{\tilde I (X), Y} \times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y} \times \underbrace{\left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{D_M}
 $$
 where $CV_W$ is the coefficient of variation (i.e., standard deviation/mean) of $W_J$ given $R_J = 1$.  See Appendix XX for the derivation.
+
+Comparing~\eqref{eq:statdecomp} to~\eqref{eq:error} shows that weighting impacts the estimation error in two ways.
+First, there is a negative impact on the data quantity component; taking the ratio of these quantities yields
+$\sqrt{1 + \frac{CV_W^2}{1-f}} \geq 1$.
+Hence, if the data quality does not increase (i.e., $| \rho_{\tilde I (X), Y} | = | \rho_{I,Y}|$ ) then weighting increases the error magnitude.
+Note this relative error increase depends on the fraction of population sampled $f$, implying that for large samples there is a larger potential increase in the error if the weights do not improve data quality.
+
+As demonstrated in the derivation of IPW approach, if the weights are correctly specified then the $E_{\bf R} [ \rho_{\tilde I (X), Y} ] = 0$ and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then...
+
+Thi
+The weighting does not interact with the
 
@@
 
-\subsubsection{Time-varying weights and random sampling}
-
-Suppose we have two random samples at times $t$ and $t+k$ (i.e., $k$ days apart).
+\subsubsection{Time-varying weights and infrequent random sampling}
+
+Suppose we have two random samples at times $t$ and $t+k$ (i.e., $k$ days apart). Here, we extend the IPW approach by~\cite{Chen2019} to the temporal setting.
+
+\begin{equation}
+\label{eq:tvpropensity}
+\sum_{t=1}^T \left[ \sum_{j=1}^N I_j \log \left( \frac{\pi (X_{j,t}; \theta)}{1-\pi(X_{j,t}; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi (X_{j,t}; \theta) \right) \right].
+\end{equation}
+
+We could estimate separately if we had 1 random sample at each.  But perhaps it is better to smooth across; especially
 

### 2020-09-10 write-up/covid-writeup.tex
updated bib and writeup

@@
 We show that estimates based on the millions of molecular tests in the US has the same mean square error as a small simple random sample.
-To address this, a procedure is presented that combines case-count data with small random samples over time to estimate selection propensities based on key covariate information.
+To address this, a procedure is presented that combines case-count data with random samples over time to estimate selection propensities based on key covariate information.
 We then combine these selection propensities with epidemiological forecast models using death-only data to construct a \emph{doubly-robust} estimation method that accounts for both measurement-error and selection bias.
-This method is then applied to estimate Indiana's prevalence using case-count and death data as well as the United States's only statewide random molecular sample collected from April 25--29th.
+This method is then applied to estimate Indiana's prevalence using case-count, hospitalization, and death data with cumulative demographic information, the United States's only statewide random molecular sample collected from April 25--29th, and Facebook's COVID-19 symptom survey.
 We end with a series of recommendations based on the proposed methodology.
@@
 
-{\bf Facebook symptoms. }
+\subsubsection{Facebook/CMU symptom survey.}
 Finally, since XX, Facebook has conducted random sample testing ().
 In particular, the COVID Symptom tracker provides an estimate of the number of individuals who have experienced any COVID-19 compatible symptoms.
+
+Symptom report is a key covariate to understand.
+
+So its poststratification but are we sure the FB people reflect..
 
@@
 \begin{equation}
-\label{eq:error}
+\label{eq:statdecomp}
 \rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
@@
 
+% {\bf Figure of individual going through the states? Can discuss exposure length as selection more likely.}
+
 \subsection{Regrettable rates: complex biases resulting from self-selection}
@@
 \label{eq:sir}
-\frac{\partial S_t}{\partial t} = - \beta S_t I_t N^{-1}; \quad
-\frac{\partial I_t}{\partial t} = \beta S_t I_t N^{-1} - \gamma I_t; \quad
-\frac{\partial R_t}{\partial t} = - \gamma I_t
+\frac{\partial s_t}{\partial t} = - \beta s_t i_t; \quad
+\frac{\partial i_t}{\partial t} = \beta s_t i_t - \gamma i_t; \quad
+\frac{\partial r_t}{\partial t} = - \gamma i_t
 \end{equation}
-where $S_t$ is the number of susceptible individuals, $I_t$ is the number of infected individuals, and $R_t$ is the number of removed (recovered or deceased) individuals at time $t$.
+where $s_t$ is the fraction of susceptible individuals, $i_t$ is the fraction of infected individuals, and $r_t$ is the fraction of removed (recovered or deceased) individuals in the population at time $t$.
 In terms of bias, the rate is overestimated prior to the peak and underestimated afterwards; the bias increases dramatically when the relative fraction exceeds $2$.
@@
 \subsubsection{Auxiliary information through population-level summary statistics}
-
-For COVID-19 case-count data, the only covariate information provided by state and local governments are summary statistics for those factors discussed in Section XX -- age, sex, race, and (potentially) comorbidities.
+\label{subsec:auxsum}
+
+For COVID-19 case-count data, the only covariate information provided by state and local governments are summary statistics for those factors discussed in Section~\ref{section:testinginfo} -- age, sex, race, and (potentially) comorbidities.
 These are discrete factors and therefore define a stratification of the population.
@@
 For comorbidities such as hypertension, diabetes, and chronic obstructive pulmonary disease (COPD) the Indiana State Department of Health releases summary statistics.
-Unfortunately the joint distribution is not provided.
-To circumvent this, we employ the standard technique of raking (cite: paper) to generate estimates $\hat N_k$ for each strata.
+Typically  joint distribution is not publicly available.
+Here, we employ the standard survey technique of raking~\citep{deville1993} to generate estimates $\hat N_k$ for each strata.
 Estimation can then be performed by using standard algorithms such as Newton-Rhapson.
-If the selection propensity is estimated non-parametrically then the resulting weight can be computed exactly and is give nby $\hat P(I_j = 1 \mid X_j = k) = \hat \pi_k = d_k/\hat N_k$.
+If the selection propensity is estimated non-parametrically then the resulting weight can be computed exactly and is given by $\hat P(I_j = 1 \mid X_j = k) = \hat \pi_k = d_k/\hat N_k$.
 
 \subsubsection{Auxiliary information through probability samples}
-
-Currently US testing does not collect additional covariate information; however, this need not be the case.
+\label{subsec:auxprob}
+
+Currently, US testing does not collect additional covariate information such as symptom information; however, this need not be the case moving forward.
 Recommendation XX below states that additional covariate information should be collected from nonprobabilistic samples.
-If this were the case, we could not longer rely on population-level summary statistics so an alternative is required.
+If this were the case, population-level summary statistics are no longer sufficient for selection propensity estimation so an alternative source of auxiliary information is required.
 
@@
 The probability sample is used to construct a design-unbiased estimator of the second term
-$$
+\begin{equation}
+\label{eq:auxinfoprob}
 \sum_{i=1}^N R_i \log \left( \frac{\pi (X_i; \theta)}{1-\pi(X_i; \theta)} \right)  + \sum_{i=1}^N \tilde R_i \tilde W_i \log ( 1 - \pi (X_i; \theta)).
-$$
+\end{equation}
 Taking the expectation of the second term with respect to the sampling design yields the second term in~\eqref{eq:propensity}.
@@
 \subsubsection{An IPW estimator and statistical error decomposition.}
-Given a selection propensity, define the inverse probability weight~$w(x) = \pi (x; \hat \theta)^{-1}$. Then the IPW estimator adjusted for measurement error is given by
+
+Selection propensities can be constructed to use both types of auxiliary information (i.e., combine Sections~\ref{subsec:auxsum} and~\ref{subsec:auxprob}).
+Given a selection propensity, define the inverse probability weight~$w(x) = \pi (x; \hat \theta)^{-1}$.
+Then the IPW estimator adjusted for measurement error is given by
 $$
@@
 \rho_{\tilde I (X), Y} \times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y} \times \underbrace{\left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{D_M}
-$$
-where $CV_W$ is the coefficient of variation (i.e., standard deviation/mean) of $W_J$ given $R_J = 1$.  See Appendix XX for the derivation.
-
-Comparing~\eqref{eq:statdecomp} to~\eqref{eq:error} shows that weighting impacts the estimation error in two ways.
+\end{equation}
+where $CV_W$ is the coefficient of variation (i.e., standard deviation/mean) of $W_J$ given $R_J = 1$, and $\rho_{\tilde I(X), Y}$ is the empirical correlation which here depends on covariate distribution.
+See Appendix XX for the derivation.
+
+Comparing~\eqref{eq:statdecomp2} to~\eqref{eq:statdecomp} shows that weighting impacts the estimation error in two ways.
 First, there is a negative impact on the data quantity component; taking the ratio of these quantities yields
@@
 
-As demonstrated in the derivation of IPW approach, if the weights are correctly specified then the $E_{\bf R} [ \rho_{\tilde I (X), Y} ] = 0$ and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then...
-
-Thi
-The weighting does not interact with the
-
-Interestingly, the impact of measurement-error on data quality is unchanged when considering a weighted estimand.  Similar to~\cite{Meng2018}, the data quality index....
-
-We can act like this is now $\rho_{I,\tilde Y}$ is $O(N^{-1})$ but this is most likely not the case; doing so, is an assumption that we "got something right".
-But the random sample has response bias and coverage issues as well.
-Therefore, we suspect it won't fix everything.
-In section XX, we will try and estimate the expecation of this quanitty after the fact; this will allow us to assess effective sample size and build corresponding CIs.
-
-In Section XX, we estimate $\rho_{I,Y}$ and $\rho_{\tilde I, Y}$ to see the impact of IPW.
+As demonstrated in the derivation of IPW approach, if the weights are correctly specified then the $E_{\bf R} [ \rho_{\tilde I (X), Y} ] = 0$ and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then the data quality index is unlikely to inversely scale with population size.
+Indeed, Horvitz-Thompson estimators are sensitity to errors in the estimated weights.
+Low selection propensity imply large weights which can have deleterious impacts by certain observations dominating estimation.
+Interestingly, the impact of measurement-error on data quality is unchanged when considering a weighted estimand.
+Similar to~\cite{Meng2018}, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from an IPW estimator is likely to put too much confidence in the sheer size of the data.
+In section XX, random sampled outcomes are used to assess effective sample sizes and adjust corresponding CIs.
 
@@
 
-Suppose we have two random samples at times $t$ and $t+k$ (i.e., $k$ days apart). Here, we extend the IPW approach by~\cite{Chen2019} to the temporal setting.
-
+Here we extend the IPW approach by~\cite{Chen2019} to this temporal setting by considering the joint likelihood
 \begin{equation}
 \label{eq:tvpropensity}
-\sum_{t=1}^T \left[ \sum_{j=1}^N I_j \log \left( \frac{\pi (X_{j,t}; \theta)}{1-\pi(X_{j,t}; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi (X_{j,t}; \theta) \right) \right].
+\sum_{t=1}^T \left[ \sum_{j=1}^N I_{j,t} \log \left( \frac{\pi_t (X_{j,t}; \theta)}{1-\pi(X_{j,t}; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi_t (X_{j,t}; \theta) \right) \right].
 \end{equation}
+{\bf $I_{j,t}$ will strongly depend on the selection history. So $X_{j,t}$ can contain features of past such as testing and symptom onset}
+
+Suppose random samples are collected at each time $t =1,\ldots,T$.
+Then the pseudo-likelihood can be re-written as in~\eqref{eq:auxinfoprob}, but with one term for each time point.
+In practice, covariate information is available on the nonprobability sample at each time $t=1,\ldots,T$; however, random samples and therefore the necessary covariate information is typically only available at a subset of times.
+Here, we briefly discuss a non-parametric kernel approach.
+
+$$
+\sum_{t=1}^T \sum_{i=1}^N K_h(|t^\prime - t|) \tilde W_{j,t} \tilde I_j
+$$
+
+We can think of this quantity as a $Z_t$ and each random sample as an estimate $\tilde Z_t$
+$$
+\| \hat Z_t - Z_t \| = \int \log \left( 1 - \pi_t (x; \theta) \right) dF_t (x)
+$$
+Where $dF_t$ is the associated measure at time~$t$
 
@@
 
-So far the focus has primarily been on selection bias in coronavirus case-count data from a purely survey sampling perspective.
+To this point, the primary focus of this paper has been on selection bias in coronavirus case-count data from a survey sampling perspective.
 Here, we consider compartmental model approaches from infectious disease epidemiology.
-Our primary objective is a model-based forecast of strata-level prevalence $\bar y_k$ using data up to but not including the current time.
+Our primary objective is a model-based forecast of strata-level prevalence $\bar y_k$.
 
 Here, we present briefly a basic epidemiological state-space model -- the susceptible, infected, and removed (recovered and death) model, or SIR model.
-The SIR model was originally proposed by (cite: Osthus, Forecasting seasonal influenza with a state-space sir model) with only one-dimensional time series of infected proportions.
-We refer the reader to Chapters 9-12 of [30] for a substantive introduction to state-space models.
-
-Here we follow closely the description from~\cite{Song2020}.
-Let $Y_t^I$ and $Y_t^R$ denote the proportion of currently infected (i.e., would test positive given a RT-PCR test with 100\% sensitivity) and removed cases (i.e., including both recovered cases and deaths) at time $t$.
+The probabilistic SIR model was originally proposed by~\cite{Osthus2017} with only one-dimensional time series of infected proportions; this formulation was extended by~\cite{Song2020} to model coronavirus case-counts.
+Given the weights built in
+
+Let $i_t$ and $Y_t^R$ denote the proportion of currently infected (i.e., would test positive given a RT-PCR test with 100\% sensitivity) and removed cases (i.e., including both recovered cases and deaths) at time $t$.
+Let $\theta_t = (s_t, e_t, i_t^{(T)}, i_t^{(NT)}, r_t)$ where
+
 Assume the two-dimensional process follows
@@
 \cite{Song2020}
+\begin{align*}
+\frac{\partial s_t}{\partial t} &= - \beta s_t i^{(T)}_t; \quad \frac{\partial e_t}{\partial t} = \beta s_t i^{(T)}_t - \sigma e_t; \quad \\
+\frac{\partial i^{(T)}_t}{\partial t} &= \sigma e_t - \beta_2 i^{(T)}; \quad
+\frac{\partial i^{(NT)}_t}{\partial t} = \beta_2 i^{(NT)}  - \gamma i^{(NT)}; \quad \\
+\frac{\partial r_t}{\partial t} &= - \gamma i^{(NT)}_t
+\end{align*}
+where $S_t$ is the number of susceptible individuals, $I_t$ is the number of infected individuals, and $R_t$ is the number of removed (recovered or deceased) individuals at time $t$.
 
@@
 \label{section:applications}
+
+What are the propensity measures of importance:
+\begin{itemize}
+	\item Symptom
+	\item Demographics
+	\item Prior testing (have you been tested before)
+	\item Prevalence (overall worry)
+\end{itemize}
+There is no good data on individual prior testing so we will ignore.
+
+{\bf Combines death, case count with covariate info, symptoms from Fb,  }
 
@@
 
+\begin{rmk}[Use of symptom-to-hospitalization ratio to estimate symptom subgroup size]
+
+\end{rmk}
+
 \section{Cross-population comparisons}
@@
 
+{\bf COORDINATION ON DATA SOURCES}
+
 \bibliographystyle{plainnat}

### 2020-09-11 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 Here, we assume access to a probability sample from the same population on which the outcome and exact same set of covariates are measured.
-Let $\tilde R_i$ be an indicator that the individual was included in the probability sample and $\tilde W_i$ be the probability of inclusion for the $i$th individual (i.e., equal to $n/N$ for a simple random sample of size $n$).
+Let $\tilde I_j$ be an indicator that the individual was included in the probability sample and $\tilde W_j$ be the probability of inclusion for the $j$th individual (i.e., equal to $n/N$ for a simple random sample of size $n$).
 The probability sample is used to construct a design-unbiased estimator of the second term
@@
 \label{eq:auxinfoprob}
-\sum_{i=1}^N R_i \log \left( \frac{\pi (X_i; \theta)}{1-\pi(X_i; \theta)} \right)  + \sum_{i=1}^N \tilde R_i \tilde W_i \log ( 1 - \pi (X_i; \theta)).
+\sum_{j=1}^N I_j \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
 \end{equation}
@@
 
-\subsubsection{Time-varying weights and infrequent random sampling}
-
-Here we extend the IPW approach by~\cite{Chen2019} to this temporal setting by considering the joint likelihood
+\subsubsection{Time-varying propensities and adjustments for application to COVID-19}
+
+In prior sections on selection propensity estimation, we ignored the dynamic nature of the outbreak.
+Here we extend the IPW approach by~\cite{Chen2019} to the temporal setting by considering the joint likelihood
 \begin{equation}
 \label{eq:tvpropensity}
-\sum_{t=1}^T \left[ \sum_{j=1}^N I_{j,t} \log \left( \frac{\pi_t (X_{j,t}; \theta)}{1-\pi(X_{j,t}; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi_t (X_{j,t}; \theta) \right) \right].
+\sum_{t=1}^T \left[ \sum_{j=1}^N I_{j,t} \log \left( \frac{\pi_t (X_{j,t}; \theta)}{1-\pi_t(X_{j,t}; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi_t (X_{j,t}; \theta) \right) \right]
 \end{equation}
-{\bf $I_{j,t}$ will strongly depend on the selection history. So $X_{j,t}$ can contain features of past such as testing and symptom onset}
-
-Suppose random samples are collected at each time $t =1,\ldots,T$.
+where $t=1,\ldots,T$ are the sequence of days on which case-count data is reported.
+Here, $I_{j,t}$ denotes self-selection into testing on day $t$, which is highly correlated with prior testing and results, i.e., $I_{j,t^\prime}$ and $Y_{j,t^\prime}$ for $t^\prime < t$.
+For example, an individual who tests positive on day $t$ is unlikely to seek testing on subsequent days.
+Moreover, an individual in a high prevalence area may be more likely to seek out testing.
+Here, we assume that the covariate vector $X_{j,t}$ contains all features of past relevant for selection.
+
+Suppose sufficiently large random samples are collected at each time $t =1,\ldots,T$.
 Then the pseudo-likelihood can be re-written as in~\eqref{eq:auxinfoprob}, but with one term for each time point.
-In practice, covariate information is available on the nonprobability sample at each time $t=1,\ldots,T$; however, random samples and therefore the necessary covariate information is typically only available at a subset of times.
+For COVID-19, there are three issues with this approach.
+First, large random samples are not available at every time.
+Second, only cumulative demographics are available for  covariate information is  the necessary covariate information is typically only available at a subset of times.
+Third, random samples from a potentially biased Facebook survey are available upon request.
+These ask if individuals have been tested
+
+To address
 Here, we briefly discuss a non-parametric kernel approach.
 
+large random samples
+
+covariate information is available on the nonprobability sample at each time $t=1,\ldots,T$;
+
 $$
@@
 We could estimate separately if we had 1 random sample at each.  But perhaps it is better to smooth across; especially
+
+\begin{rmk}[Cumulative demographics]
+Most states only report cumulative demographic information.
+\end{rmk}
 

### 2020-09-12 write-up/covid-writeup.tex
update from laptop

@@
 
-In prior sections on selection propensity estimation, we ignored the dynamic nature of the outbreak.
-Here we extend the IPW approach by~\cite{Chen2019} to the temporal setting by considering the joint likelihood
+In prior sections on selection propensity estimation, we ignored the dynamic nature of the outbreak.  Here we extend the IPW approach to the temporal setting by considering the joint likelihood
 \begin{equation}
@@
 
-Suppose sufficiently large random samples are collected at each time $t =1,\ldots,T$.
-Then the pseudo-likelihood can be re-written as in~\eqref{eq:auxinfoprob}, but with one term for each time point.
-For COVID-19, there are three issues with this approach.
-First, large random samples are not available at every time.
-Second, only cumulative demographics are available for  covariate information is  the necessary covariate information is typically only available at a subset of times.
-Third, random samples from a potentially biased Facebook survey are available upon request.
-These ask if individuals have been tested
-
-To address
+As the necessary covariates (e.g., symptom status, prior testing, COVID-19 exposure) go beyond demographic information, random sampling approaches are required.
+If sufficiently large random samples are collected at each time $t =1,\ldots,T$, then the pseudo-likelihood can be re-written as in~\eqref{eq:auxinfoprob}, but with one term for each time point.
+For COVID-19, there are two key issues with this approach.
+First, large probabilistic samples are not available at every time.
+Second, only cumulative demographics on case counts are available from states' department of health\footnote{Historical daily demographic data can be requested via FOIA, but such requests take months to complete -- as the author has learned through sweat and tears.}.
+Moreover, key covariate information such as symptom status is not reported (and potentially not recorded) by states' department of health.
+
+To address these issues, we rely on two additional sources of data.
+First, the Facebook symptom survey collects symptom data on a random subset of the Facebook population.
+Use of these random samples lead to questions of target population coverage (i.e., not all Hoosiers have Facebook); however, such data fills in the gaps for symptom information.
+While the Facebook survey also asks if individuals have been tested, exact date of testing is not recorded; therefore, the survey's use as a resource for understanding COVID-19 testing is limited.
+To address this, we rely on COVID-19 hospitalization data to estimate a relative fraction of cases that display symptoms.
+
+Specific weight calculations and adjustments are deferred to Section~\ref{section:applications}.
 Here, we briefly discuss a non-parametric kernel approach.
+We start by using considering a pseudo-likelihood
+\begin{equation}
+\label{eq:tvpropensity}
+\sum_{t=1}^T \left[ \sum_{j=1}^N I_{j,t} I^{(2)}_{j,t} W_{j,t} \log \left( \frac{\pi_t (X_{j,t}; \theta)}{1-\pi_t(X_{j,t}; \theta)} \right) + \sum_{j=1}^N \tilde W_{j,t} \tilde I_{j,t} \log \left( 1 - \pi_t (X_{j,t}; \theta) \right) \right]
+\end{equation}
 

### 2020-09-15 write-up/covid-writeup.tex
Update covid-writeup.tex

--- 2020-09-11 write-up/covid-writeup.tex
@@
 \multirow{3}{*}{Age$^\star$} & $<40$ & 47 & 28 & 53 & 47\\
-& $40-59$ & 28 & 36 & 25 & 30\\
+& $40-59$ & 28 & 36 & 25 & 3 0\\
 & $\geq 60$ & 25 & 36 & 22 & 13\\ \hline
@@
 & Nonwhite & 21 & 8 & 15 & 50 \\ \hline
-\multirow{2}{*}{Symptoms$^{\star \star}$} & Yes & - & - & - \\
-& No & - & - & - \\ \hline
+\multirow{2}{*}{Symptoms$^{\star \star}$} & Yes & 62 & 43 & - & 83 & 62 \\
+& No & 38 & 57 & - & 17 & 38  \\ \hline
 \multirow{2}{*}{Contact$^{\star \star}$} & Yes & 11 & 1 & - \\
@@
 \end{table}
+
+% Random sample
+% Symptom 0, Test 0 2040
+% Symptom 0, Test 1 18
+% Symptom 1, Test 0 1518
+% Symptom 1, Test 1 29
+% Total symptom 1; (1518+29)/3605 = 43
+% Pos test symptom 1; (29)/(29+18) = 62
+
+% Avg 100 COVID-19 hospitalizations in state in that time
+% 650 cases
+% percent of cases in nonrandom sample with symptoms = 184/(184+37) = 83%
+% Then 83% * 650 = 540 symptomatic cases
+% 100/540 = 19% of symptomatic cases were hospitalized
+
+% Example: prior to April 28th, only symptoms and a few others
+% 04/11: around 500 cases
+% 04/11: 117 hospitalizations -> 615 cases
 
@@
 \label{section:discussion}
+
+{\bf Our goal with surveillance is to identify areas of weakness; but selection bias is an issue!}
+
 There is nothing routine about COVID-19, including the corresponding statistical questions.

### 2020-09-16 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 \begin{tabular}{c | c | l l | l | l l }
-& & \multicolumn{2}{c}{Total Tests} &  & \multicolumn{2}{c}{Positive Tests} \\
+& & \multicolumn{2}{c}{Total Tests} & US & \multicolumn{2}{c}{Positive Test Rate}\\
 \cline{3-4} \cline{6-7}
 & & NonRandom & Random & Census & NonRandom & Random \\ \hline
-\multirow{2}{*}{Sex$^\star$} & Female & 57\% & 55\% & 49\% & 53\% & - \\
-& Male & 43 & 45 & 51 & 47 & -  \\ \hline
-\multirow{3}{*}{Age$^\star$} & $<40$ & 47 & 28 & 53 & 47\\
-& $40-59$ & 28 & 36 & 25 & 3 0\\
-& $\geq 60$ & 25 & 36 & 22 & 13\\ \hline
-\multirow{2}{*}{Race$^\star$} & White & 60 & 98 & 85 & 50\\
-& Nonwhite & 21 & 8 & 15 & 50 \\ \hline
-\multirow{2}{*}{Symptoms$^{\star \star}$} & Yes & 62 & 43 & - & 83 & 62 \\
-& No & 38 & 57 & - & 17 & 38  \\ \hline
-\multirow{2}{*}{Contact$^{\star \star}$} & Yes & 11 & 1 & - \\
+\multirow{2}{*}{Sex} & Female & 58\% & 55\% & 51\% & 21.7\% & 1.42\% \\
+& Male & 42 & 45 & 49 & 24.2 & 2.13 \\ \hline
+\multirow{3}{*}{Age} & $<40$ & 40 & 28 & 53 & 29.7 & 1.71 \\
+& $40-59$ & 41 & 36 & 25 & 24.9 & 2.09 \\
+& $\geq 60$ & 19 & 36 & 22 & 6.7 & 0.92 \\ \hline
+\multirow{2}{*}{Race} & White & 23 & 92 & 87 \\
+& Nonwhite & 77 & 8 & 13 \\ \hline
+\multirow{2}{*}{Symptom} & Yes & 62 & 43 & - &  \\
+& No & 38 & 57 & - & \\ \hline
+\multirow{2}{*}{Contact} & Yes & 11 & 1 & - \\
 & No & 89 & 99 & - \\ \hline
+\multirow{2}{*}{Prior test} & Yes & 6 & 1 & - \\
+& No & 94 & 99 & - \\ \hline
 \end{tabular}
-\caption{Temp \\
-$\star$: ISDH case-count nonrandom sample \\
-$\star\star$: May 2--3 nonrandom sample
-}
+\caption{Temp}
 \end{table}
+
+Average between May 2nd and 3rd is 11.7\% statewide.  So we need to adjust all these estimates down and then re-weight.
 

### 2020-09-17 write-up/covid-writeup.tex
Update covid-writeup.tex
switchin to fever inst of symp

--- 2020-09-15 write-up/covid-writeup.tex
@@
 In this section, we consider two directions.
-First, we use publicly available auxiliary information to build selection propensities and consider an inverse probability weighting (IWP) estimator.
+First, we use publicly available auxiliary information to build selection propensities and consider an inverse probability weighting (IPW) estimator.
 We extend the statistical error decomposition to the weighted setting.
@@
 This was proposed by~\cite{Chen2019} when covariates were observed on a random sample from the population.
-Here, we only observe strata information.
 Under simple random sampling and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the $\hat \pi_k = (d_k / \tilde d_k) (n / N)$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.
+
+Equation~\eqref{eq:auxinfoprob} presupposes covariate information is collected for all individuals in the nonprobability sample; however, this is often not feasible for state and local governments where rapid sample collection is prioritized over additional collection of covariate information.
+We can extend~\eqref{eq:auxinfoprob} by randomly sampling a subset of the nonprobability sample on which to collect the covariate information, yielding the design-unbiased estimator
+\begin{equation}
+\label{eq:auxinfoprob2}
+\sum_{j=1}^N I_j I_j^{(2)} W_j^{(2)} \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
+\end{equation}
+Here, $I_j^{(2)}$ is an indicator that the individual is in the probability subsample and $W_j^{(2)}$ is the probability of inclusion.
+The estimate no relies on random sampling in both the first and second term.
 
@@
 Here, we briefly discuss a non-parametric kernel approach.
-We start by using considering a pseudo-likelihood
-\begin{equation}
-\label{eq:tvpropensity}
-\sum_{t=1}^T \left[ \sum_{j=1}^N I_{j,t} I^{(2)}_{j,t} W_{j,t} \log \left( \frac{\pi_t (X_{j,t}; \theta)}{1-\pi_t(X_{j,t}; \theta)} \right) + \sum_{j=1}^N \tilde W_{j,t} \tilde I_{j,t} \log \left( 1 - \pi_t (X_{j,t}; \theta) \right) \right]
-\end{equation}
-
-large random samples
-
-covariate information is available on the nonprobability sample at each time $t=1,\ldots,T$;
-
-$$
-\sum_{t=1}^T \sum_{i=1}^N K_h(|t^\prime - t|) \tilde W_{j,t} \tilde I_j
-$$
-
-We can think of this quantity as a $Z_t$ and each random sample as an estimate $\tilde Z_t$
-$$
-\| \hat Z_t - Z_t \| = \int \log \left( 1 - \pi_t (x; \theta) \right) dF_t (x)
-$$
-Where $dF_t$ is the associated measure at time~$t$
-
-We could estimate separately if we had 1 random sample at each.  But perhaps it is better to smooth across; especially
-
-\begin{rmk}[Cumulative demographics]
-Most states only report cumulative demographic information.
-\end{rmk}
+We start by considering a time-varying version of equation~\eqref{eq:auxinfoprob2}
+% \begin{equation}
+% \label{eq:tvpropensity}
+% \sum_{t=1}^T \left[ \sum_{j=1}^N I_{j,t} I^{(2)}_{j,t} W_{j,t} \log \left( \frac{\pi_t (X_{j,t}; \theta)}{1-\pi_t(X_{j,t}; \theta)} \right) + \sum_{j=1}^N \tilde W_{j,t} \tilde I_{j,t} \log \left( 1 - \pi_t (X_{j,t}; \theta) \right) \right]
+% \end{equation}
+that accounts for random sampling from within the non-random sample.
+Note random samples may be quite small at each time while we expect $\pi_t(x;\theta)$ to be a smooth function of time; therefore, we introduce a non-parametric approach to estimation of the selection propensity at time~$t$,
+$$
+\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \left[ \sum_{j=1}^N I_{j,t^\prime} I^{(2)}_{j,t^\prime} W_{j,t^\prime} \log \left( \frac{\pi_t (X_{j,t^\prime}; \theta)}{1-\pi_t(X_{j,t^\prime}; \theta)} \right) + \sum_{j=1}^N \tilde W_{j,t^\prime} \tilde I_{j,t^\prime} \log \left( 1 - \pi_t (X_{j,t^\prime}; \theta) \right) \right]
+$$
+where $K_h$ is a kernel function with tunable parameter $h$.
+Under simple random sampling of equal size~$\tilde n$ from a fixed population size~$N$ at each time, and s, and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the
+$$
+\hat \pi_{k,t} = \frac{n}{N} \times \frac{\sum_{t^\prime=1}^T K(|t^\prime - t|) \frac{n_t}{m} \hat d_{k,t^\prime}}{\sum_{t^\prime=1}^T \tilde d_{k,t^\prime}}
+$$
+where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.
 
@@
 \begin{table}[!th]
-\begin{tabular}{c | c | l l | l | l l }
-& & \multicolumn{2}{c}{Total Tests} &  & \multicolumn{2}{c}{Positive Tests} \\
+\begin{tabular}{c | c | r r | r | r r }
+& & \multicolumn{2}{c}{Total Tests (\%)} & US & \multicolumn{2}{c}{Positive Test Rate (\%)}\\
 \cline{3-4} \cline{6-7}
 & & NonRandom & Random & Census & NonRandom & Random \\ \hline
-\multirow{2}{*}{Sex$^\star$} & Female & 57\% & 55\% & 49\% & 53\% & - \\
-& Male & 43 & 45 & 51 & 47 & -  \\ \hline
-\multirow{3}{*}{Age$^\star$} & $<40$ & 47 & 28 & 53 & 47\\
-& $40-59$ & 28 & 36 & 25 & 3 0\\
-& $\geq 60$ & 25 & 36 & 22 & 13\\ \hline
-\multirow{2}{*}{Race$^\star$} & White & 60 & 98 & 85 & 50\\
-& Nonwhite & 21 & 8 & 15 & 50 \\ \hline
-\multirow{2}{*}{Symptoms$^{\star \star}$} & Yes & 62 & 43 & - & 83 & 62 \\
-& No & 38 & 57 & - & 17 & 38  \\ \hline
-\multirow{2}{*}{Contact$^{\star \star}$} & Yes & 11 & 1 & - \\
-& No & 89 & 99 & - \\ \hline
+\multirow{2}{*}{Sex} & Female & 58 & 55 & 51 & 21.7 & 1.4 \\
+& Male & 42 & 45 & 49 & 24.2 & 2.1 \\ \hline
+\multirow{3}{*}{Age} & $<40$ & 40 & 28 & 53 & 29.7 & 1.7 \\
+& $40-59$ & 41 & 36 & 25 & 24.9 & 2.1 \\
+& $\geq 60$ & 19 & 36 & 22 & 6.7 & 0.9 \\ \hline
+\multirow{2}{*}{Race} & White & 23 & 92 & 87 & 19.5 & 1.5 \\
+& Nonwhite & 77 & 8 & 13 & 25.0 & 3.4 \\ \hline
+Hispanic & Hispanic & - & - & - & 37.6 & 6.9 \\
+Origin & Non-Hispanic & - & - & - & 13.0 & 1.3 \\ \hline
+\multirow{2}{*}{Fever} & Yes & 17.0 & 43 & - & 31.6 & 1.9 \\
+& No & 38 & 57 & -& 10.4 & 0.9 \\ \hline
+\multirow{2}{*}{Household} & Yes & 11 & 1 & - & 46.1 & 24.4 \\
+& No & 89 & 99 & - & 21.6 & 1.3 \\ \hline
+Prior $+$ & Yes & 6 & 1 & - & 39.2 & 24.4 \\
+Test & No & 94 & 99 & - & 21.6 & 1.3 \\ \hline
 \end{tabular}
-\caption{Temp \\
-$\star$: ISDH case-count nonrandom sample \\
-$\star\star$: May 2--3 nonrandom sample
-}
+\caption{Estimated point prevalence of active infection with SARS-CoV-2 by demographics in Indiana.}
 \end{table}
 
-% Random sample
-% Symptom 0, Test 0 2040
-% Symptom 0, Test 1 18
-% Symptom 1, Test 0 1518
-% Symptom 1, Test 1 29
-% Total symptom 1; (1518+29)/3605 = 43
-% Pos test symptom 1; (29)/(29+18) = 62
-
-% Avg 100 COVID-19 hospitalizations in state in that time
-% 650 cases
-% percent of cases in nonrandom sample with symptoms = 184/(184+37) = 83%
-% Then 83% * 650 = 540 symptomatic cases
-% 100/540 = 19% of symptomatic cases were hospitalized
-
-% Example: prior to April 28th, only symptoms and a few others
-% 04/11: around 500 cases
-% 04/11: 117 hospitalizations -> 615 cases
-
-1. Unknown assumed MAR for Race, Gender, and Age
-2. 20\% are asymptomatic (so 78\% + 22\% (20\%) showed no symptoms)
-
-We have a small non-random sample to supplement and get these; then use raking on the larger study.  Appendix XX presents the marginal distributions for the non-random sample per
-
-Heuristically, one may suppose that asymptomatic individuals are less likely to select.  This is born out.
-
-This is not true (need to deal with very bad symptoms cases).
-
-Simplest case we have Age $<40$ in the random sample is $1,017$ and we have $13+17.3+15.2 = 45\% \times 1,002,038$ tests so $450917.1$ tests.
-Then $N = 6.85$ million while $n = 3658$ leading to $\pi_k = 0.24$.
-
-\begin{rmk}[Use of second nonrandom sample]
-The key issue with selection bias is whether the individual is displaying symptoms and had contact with someone who was COVID+.  These are not captured by the
-
-In random sample, $44.2$\% of individuals who tested positive for SARS-CoV-2 positive individuals reported no symptoms during the 2 weeks before testing.
-
-In nonrandom sample, $20.2$\% of individuals who tested positive for SARS-CoV-2 reported no symptoms during the 2 weeks before testing.
-
-\end{rmk}
+We correct the nonrandom sample to match the Indiana state-wide prevalence
+
+\begin{subsubsection}{Time-varying doubly-robust prevalence estimation}
 

### 2020-09-17 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 & & NonRandom & Random & Census & NonRandom & Random \\ \hline
-\multirow{2}{*}{Sex} & Female & 58 & 55 & 51 & 21.7 & 1.4 \\
-& Male & 42 & 45 & 49 & 24.2 & 2.1 \\ \hline
-\multirow{3}{*}{Age} & $<40$ & 40 & 28 & 53 & 29.7 & 1.7 \\
-& $40-59$ & 41 & 36 & 25 & 24.9 & 2.1 \\
-& $\geq 60$ & 19 & 36 & 22 & 6.7 & 0.9 \\ \hline
-\multirow{2}{*}{Race} & White & 23 & 92 & 87 & 19.5 & 1.5 \\
-& Nonwhite & 77 & 8 & 13 & 25.0 & 3.4 \\ \hline
-Hispanic & Hispanic & - & - & - & 37.6 & 6.9 \\
-Origin & Non-Hispanic & - & - & - & 13.0 & 1.3 \\ \hline
-\multirow{2}{*}{Fever} & Yes & 17.0 & 43 & - & 31.6 & 1.9 \\
-& No & 38 & 57 & -& 10.4 & 0.9 \\ \hline
-\multirow{2}{*}{Household} & Yes & 11 & 1 & - & 46.1 & 24.4 \\
-& No & 89 & 99 & - & 21.6 & 1.3 \\ \hline
-Prior $+$ & Yes & 6 & 1 & - & 39.2 & 24.4 \\
-Test & No & 94 & 99 & - & 21.6 & 1.3 \\ \hline
+\multirow{2}{*}{Sex} & Female & 58.2 & 55.0 & 51 & 21.7 & 1.4 \\
+& Male & 42 & 45.0 & 49 & 24.2 & 2.1 \\ \hline
+\multirow{3}{*}{Age} & $<40$ & 40 & 28.0 & 53 & 29.7 & 1.7 \\
+& $40-59$ & 41 & 36.0 & 25 & 24.9 & 2.1 \\
+& $\geq 60$ & 19 & 36.0 & 22 & 6.7 & 0.9 \\ \hline
+\multirow{2}{*}{Race} & White & 23 & 92.0 & 87 & 19.5 & 1.5 \\
+& Nonwhite & 77 & 8.0 & 13 & 25.0 & 3.4 \\ \hline
+Hispanic & Hispanic & - & 2.0 & - & 37.6 & 6.9 \\
+Origin & Non-Hispanic & - & 98.0 & - & 13.0 & 1.3 \\ \hline
+\multirow{2}{*}{Fever} & Yes & 17.0 & 1.8 & - & 66.4 & 4.5 \\
+& No & 83.0 & 98.2 & - & 15.6 & 1.3 \\ \hline
+\multirow{2}{*}{Household} & Yes & 10.8 & 1.4 & - & 46.1 & 24.4 \\
+& No & 89.2 & 98.6 & - & 21.6 & 1.3 \\ \hline
+Prior $+$ & Yes & 6.1 & 1.6 & - & 39.2 & 24.4 \\
+Test & No & 93.9 & 98.6 & - & 21.6 & 1.3 \\ \hline
 \end{tabular}
@@
 
-We correct the nonrandom sample to match the Indiana state-wide prevalence
+We correct the nonrandom sample to match the Indiana state-wide prevalence.
+
+$66.4 * 0.17 + 15.6 * 0.83 = XX = 11.7 * c $
 

### 2020-09-18 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 
-Here we consider estimation of the active infection rate in Indiana.  We start with IPW estimation.  Table .
-
-The report does not include
-
-\begin{table}[!th]
+Here we consider estimation of the active infection rate in Indiana.
+Table~\ref{tab:indiana} summarizes data from a random sample from April 25--29th as well as a nonrandom sample obtained between May 2--3 in racial/ethnic minority communities.
+The nonrandom sample is therefore not a random subsample of the overall case-counts; however, this data provides important supplementary covariate information.
+Namely, the samples collected included covariate information on (1) prior symptoms, (2) COVID-19 positive individual in household, and (3) prior COVID-19 positive test.
+To account for the nonrandom sample being from high risk areas, we adjust estimates to match the statewide positivity rate of 11.7\% between May 1--2.
+These adjusted rates are presented in parentheses in Table~\ref{tab:indiana}.
+
+\begin{table}[th]
 \begin{tabular}{c | c | r r | r | r r }
@@
 & & NonRandom & Random & Census & NonRandom & Random \\ \hline
-\multirow{2}{*}{Sex} & Female & 58.2 & 55.0 & 51 & 21.7 & 1.4 \\
-& Male & 42 & 45.0 & 49 & 24.2 & 2.1 \\ \hline
-\multirow{3}{*}{Age} & $<40$ & 40 & 28.0 & 53 & 29.7 & 1.7 \\
-& $40-59$ & 41 & 36.0 & 25 & 24.9 & 2.1 \\
-& $\geq 60$ & 19 & 36.0 & 22 & 6.7 & 0.9 \\ \hline
-\multirow{2}{*}{Race} & White & 23 & 92.0 & 87 & 19.5 & 1.5 \\
-& Nonwhite & 77 & 8.0 & 13 & 25.0 & 3.4 \\ \hline
-Hispanic & Hispanic & - & 2.0 & - & 37.6 & 6.9 \\
-Origin & Non-Hispanic & - & 98.0 & - & 13.0 & 1.3 \\ \hline
+\multirow{2}{*}{Sex} & Female & 58.2 & 55.0 & 50.7 & 21.7 (11.2) & 1.4 \\
+& Male & 41.8 & 45.0 & 49.3 & 24.2 (12.4) & 2.1 \\ \hline
+\multirow{3}{*}{Age} & $<40$ & 39.4 & 28.0 & 52.7 & 29.7 & 1.7 \\
+& $40-59$ & 41.1 & 36.0 & 25.2 & 24.9 & 2.1 \\
+& $\geq 60$ & 19.5 & 36.0 & 22 & 6.7 & 0.9 \\ \hline
+\multirow{2}{*}{Race} & White & 23.1 & 92.0 & 86.9 & 19.5 & 1.5 \\
+& Nonwhite & 76.9 & 8.0 & 13.1 & 25.0 & 3.4 \\ \hline
+Hispanic & Hispanic & 44.1 & 2.0 & 7.1 & 37.6 & 6.9 \\
+Origin & Non-Hispanic & 55.9 & 98.0 & 92.9 & 13.0 & 1.3 \\ \hline
 \multirow{2}{*}{Fever} & Yes & 17.0 & 1.8 & - & 66.4 & 4.5 \\
 & No & 83.0 & 98.2 & - & 15.6 & 1.3 \\ \hline
-\multirow{2}{*}{Household} & Yes & 10.8 & 1.4 & - & 46.1 & 24.4 \\
-& No & 89.2 & 98.6 & - & 21.6 & 1.3 \\ \hline
+Household & Yes & 10.8 & 1.4 & - & 46.1 & 24.4 \\
+$+$ Case & No & 89.2 & 98.6 & - & 21.6 & 1.3 \\ \hline
 Prior $+$ & Yes & 6.1 & 1.6 & - & 39.2 & 24.4 \\
@@
 \caption{Estimated point prevalence of active infection with SARS-CoV-2 by demographics in Indiana.}
+\label{tab:indiana}
 \end{table}

### 2020-10-09 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 Participants in the probability sample were also asked if they experienced any COVID-19 compatible symptoms during the past 2 weeks or had shared a household with someone who had a positive test result for SARS-CoV-2.
-
 During May 2--3, 2020, an additional nonrandom sample of 898 individuals was also collected.
@@
 \subsubsection{Facebook/CMU symptom survey.}
-Finally, since XX, Facebook has conducted random sample testing ().
-In particular, the COVID Symptom tracker provides an estimate of the number of individuals who have experienced any COVID-19 compatible symptoms.
-
-Symptom report is a key covariate to understand.
-
-So its poststratification but are we sure the FB people reflect..
+Since April 2020, Facebook has started randomly showing an opt-in symptom survey to its users.
+The survey is run by health researchers at Carnegie Mellon University.
+Data collected includes basic demographic information and if the respondent has symptoms such as fever, coughing, shortness of breath, or loss of smell which are associated with COVID-19.
+The survey defines an individual as displaying \emph{COVID-like symptoms} if they exhibit a fever along with a cough, or shortness of breath, or difficulty breathing.
+Figure~\ref{fig:fbsymptoms} displays nonparametric regression estimates of the fraction of individuals within age and gender strata.
+
+\begin{figure}[!th]
+\centering
+\includegraphics[width = 0.9\textwidth]{../methods/figs/fbcovid19symptoms.png}
+\caption{Reported rate of COVID-like symptoms in strata.
+Daily rates were estimated using weighted method on each day separately and then smoothed over time using local-linear nonparametric regression.}
+\label{fig:fbsymptoms}
+\vspace{-0.3cm}
+\end{figure}
 
@@
 
-We correct the nonrandom sample to match the Indiana state-wide prevalence.
-
-$66.4 * 0.17 + 15.6 * 0.83 = XX = 11.7 * c $
+Comparison to April 20th-25th using Fb data
+\begin{itemize}
+\item Fever: {\bf FB} = [0.017, 0.020, 0.024], {\bf Random}: 0.018
+\item Cough: {\bf FB} = [0.142, 0.148, 0.154], {\bf Random}: 0.149
+\item Shortness: {\bf FB} = [0.058, 0.060, 0.062], {\bf Random}: 0.062
+\end{itemize}
 

### 2020-10-15 write-up/covid-writeup.tex
updating for fb data

@@
-\documentclass[11pt]{amsart}
+\documentclass[12pt]{amsart}
 
@@
 
-For COVID-19 case-count data, the only covariate information provided by state and local governments are summary statistics for those factors discussed in Section~\ref{section:testinginfo} -- age, sex, race, and (potentially) comorbidities.
+For COVID-19 case-count data, the only covariate information provided by state and local governments are summary statistics for those factors discussed in Section~\ref{section:testinginfo} -- age, sex,  and race.
 These are discrete factors and therefore define a stratification of the population.
@@
 The US Census, for example, can provide marginal summary statistics for the distribution of age, sex, and race in Indiana state.
-For comorbidities such as hypertension, diabetes, and chronic obstructive pulmonary disease (COPD) the Indiana State Department of Health releases summary statistics.
 Typically  joint distribution is not publicly available.
@@
 
-To address these issues, we rely on two additional sources of data.
-First, the Facebook symptom survey collects symptom data on a random subset of the Facebook population.
-Use of these random samples lead to questions of target population coverage (i.e., not all Hoosiers have Facebook); however, such data fills in the gaps for symptom information.
-While the Facebook survey also asks if individuals have been tested, exact date of testing is not recorded; therefore, the survey's use as a resource for understanding COVID-19 testing is limited.
-To address this, we rely on COVID-19 hospitalization data to estimate a relative fraction of cases that display symptoms.
+To address these issues, we rely on the Facebook symptom survey.
+Daily random surveys provide adequate covariate information; however, the daily estimates vary substantially while we expect  while we expect to
 

### 2020-10-15 methods/fb-analysis/fb-cleaning-code.R

@@ -1,0 +1,40 @@
+data_files = list.files(path = "./all-fb-data/")
+
+file_name = "indiana_data.csv"
+
+library("lubridate")
+
+for(i in 1:length(data_files)) {
+
+  temp = read.csv(paste("./all-fb-data/",data_files[i], sep=""))
+  
+  indiana_temp = subset(temp,A3b == 15)
+  
+  time <- ymd_hms(indiana_temp$StartDatetime)
+  day_df = data.frame("date" = date(time), "weight" = indiana_temp$weight)
+  
+  fever <- function(x) {any(as.numeric(unlist(strsplit(x, split = ",")))==1)}
+  cough <- function(x) {any(as.numeric(unlist(strsplit(x, split = ",")))==2)}
+  shortness <- function(x) {any(as.numeric(unlist(strsplit(x, split = ",")))==3)}
+  diffbreath <- function(x) {any(as.numeric(unlist(strsplit(x, split = ",")))==4)}
+  
+  day_df$fever = unlist(lapply(indiana_temp$B2,fever))
+  day_df$cough = unlist(lapply(indiana_temp$B2,cough))
+  day_df$shortness = unlist(lapply(indiana_temp$B2,shortness))
+  day_df$diffbreath = unlist(lapply(indiana_temp$B2,diffbreath))
+  day_df$symptoms = (day_df$fever & day_df$cough) | day_df$shortness | day_df$diffbreath
+  
+  day_df$gottested = as.numeric(is.element(indiana_temp$B5, c(1,2,3)))
+  day_df$postest =  as.numeric(is.element(indiana_temp$B5, c(1)))
+  day_df$negtest =  as.numeric(is.element(indiana_temp$B5, c(2)))
+  
+  day_df$gender = indiana_temp$D1
+  day_df$age = indiana_temp$D2
+  
+  if(file.exists(file_name)){
+    write.table(file = file_name, x = day_df, append = TRUE, col.names = FALSE, row.names = FALSE, sep = ",")
+  } else{
+    write.table(file = file_name, x = day_df, row.names = FALSE, sep = ",")
+  }
+  
+}

### 2020-10-15 methods/fb-analysis/fb-indiana-plotsv2.R

@@ -1,0 +1,171 @@
+library(lubridate)
+library(ggplot2)
+library(np)
+
+indiana_data = read.csv("indiana_data.csv")
+
+indiana_data$symptoms = (indiana_data$fever & (indiana_data$cough | indiana_data$shortness | indiana_data$diffbreath))
+
+indiana_data$weightedfever = (indiana_data$symptoms) * indiana_data$weight
+# png(filename = "fbcovid19symptoms.png",
+#     width = 960, height = 480, units = "px", pointsize = 12)
+par(mfrow = c(1,2), mar = c(4,4,0.5,1)+0.1)
+
+denom = aggregate(weight ~ date + gender, data = indiana_data, sum)
+
+numer = aggregate(weightedfever ~ date + gender, data = indiana_data, sum)
+
+tempdf = data.frame('x' = difftime(date(numer$date), "2020-01-01"), 'y' = numer$weightedfever/denom$weight, 'gender' = denom$gender)
+tempdf$x = as.vector(tempdf$x)
+
+model.np <- npreg(y~x, regtype = 'll', bwmethod = 'cv.aic', gradients = TRUE, data = subset(tempdf, gender == 2))
+
+plot(date(numer$date[numer$gender == 2]), fitted(model.np), type = "l", 
+     bty= 'n', axes = FALSE, ylab = "Reported COVID-19 Symptoms", xlab = "Date",
+     ylim = c(0.0,0.015), lwd = 3)
+axis.Date(1, at=seq(min(date(numer$date[numer$gender == 2])), max(date(numer$date[numer$gender == 2])), by="months"), format="%b")
+axis(side = 2)
+
+lines(date(numer$date[numer$gender == 2]), fitted(model.np)+1.96*se(model.np), lty = 2)
+lines(date(numer$date[numer$gender == 2]), fitted(model.np)-1.96*se(model.np), lty = 2)
+
+model.np <- npreg(y~x, regtype = 'll', bwmethod = 'cv.aic', gradients = TRUE, data = subset(tempdf, gender == 1))
+
+lines(date(numer$date[numer$gender == 1]), fitted(model.np), col = "red", lwd = 3)
+
+lines(date(numer$date[numer$gender == 1]), fitted(model.np)+1.96*se(model.np), lty = 2, col = "red")
+lines(date(numer$date[numer$gender == 1]), fitted(model.np)-1.96*se(model.np), lty = 2, col = "red")
+
[…]

### 2020-10-19 write-up/covid-writeup.tex
Update covid-writeup.tex
Need to shorten all sections!

@@
 First, we use publicly available auxiliary information to build selection propensities and consider an inverse probability weighting (IPW) estimator.
-We extend the statistical error decomposition to the weighted setting.
-Second, we consider an extension of compartmental models from infectious disease epidemiology using death data to try and avoid selection bias.
-After adjusting for measurement-error, this leads to a model-based forecasts of prevalence.
+We then extend the statistical error decomposition to the weighted setting.
+Second, we consider an extension of compartmental models from infectious disease epidemiology using death data.
+After adjusting for measurement-error, this leads to an alternative, model-based forecast of prevalence.
 We then combine these epidemiological forecasts, selection propensities, and measurement-error corrections to generate \emph{doubly robust} prevalence estimators.
@@
 Here, $I_j^{(2)}$ is an indicator that the individual is in the probability subsample and $W_j^{(2)}$ is the probability of inclusion.
-The estimate no relies on random sampling in both the first and second term.
+The estimate now relies on random sampling in both the first and second term.
 
@@
 Then the IPW estimator adjusted for measurement error is given by
-$$
+\begin{equation}
+\label{eq:ipwest}
 \bar y_n^\star
@@
 \stackrel{(2)}{=} \frac{1}{1-FP-FN} \sum_{k=1}^K \frac{d_k w_k}{w} (\bar y_k - FP),
-$$
-where the second equality (2) is under the assumption that $x_i$ is a stratification variable with $k$ indexing the strata, $w_k$ is the weight and $d_k$ is the number of samples in strata $k$, and $w = \sum_{k=1}^K d_k w_k$.
+\end{equation}
+where equality (2) is under the assumption that $x_i$ is a stratification variable with $k$ indexing the strata, $w_k$ is the weight and $d_k$ is the number of samples in strata $k$, and $w = \sum_{k=1}^K d_k w_k$.
 Under simple random sampling with one strata~$\hat \pi_k = n/N$ and we recover the estimator $\tilde y_n$ from Section~\ref{section:imperfecttesting}.
@@
 \end{equation}
-where $CV_W$ is the coefficient of variation (i.e., standard deviation/mean) of $W_J$ given $R_J = 1$, and $\rho_{\tilde I(X), Y}$ is the empirical correlation which here depends on covariate distribution.
+where $CV_W$ is the coefficient of variation (i.e., standard deviation/mean) of $W_J$ given $I_J = 1$, and $\rho_{\tilde I(X), Y}$ is the empirical correlation which here depends on covariate distribution.
 See Appendix XX for the derivation.
@@
 As demonstrated in the derivation of IPW approach, if the weights are correctly specified then the $E_{\bf R} [ \rho_{\tilde I (X), Y} ] = 0$ and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then the data quality index is unlikely to inversely scale with population size.
-Indeed, Horvitz-Thompson estimators are sensitity to errors in the estimated weights.
+Indeed, Horvitz-Thompson estimators are sensitive to errors in the estimated weights.
 Low selection propensity imply large weights which can have deleterious impacts by certain observations dominating estimation.
 Interestingly, the impact of measurement-error on data quality is unchanged when considering a weighted estimand.
-Similar to~\cite{Meng2018}, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from an IPW estimator is likely to put too much confidence in the sheer size of the data.
-In section XX, random sampled outcomes are used to assess effective sample sizes and adjust corresponding CIs.
-
-\subsubsection{Time-varying propensities and adjustments for application to COVID-19}
+Similar to~\cite{Meng2018}, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from an IPW estimator are likely to put too much confidence in the sheer size of the data.
+In section XX, independent random samples are used to assess effective sample sizes and adjust corresponding CIs.
+
+\subsubsection{Time-varying propensities}
 
@@
 \end{equation}
-where $t=1,\ldots,T$ are the sequence of days on which case-count data is reported.
-Here, $I_{j,t}$ denotes self-selection into testing on day $t$, which is highly correlated with prior testing and results, i.e., $I_{j,t^\prime}$ and $Y_{j,t^\prime}$ for $t^\prime < t$.
-For example, an individual who tests positive on day $t$ is unlikely to seek testing on subsequent days.
+where $t=1,\ldots,T$ are the days when case-count data is reported.
+Here, $I_{j,t}$ denotes self-selection into testing on day $t$, which is highly correlated with prior testing and results.
+For example, an individual who tests positive is unlikely to seek testing on subsequent days.
 Moreover, an individual in a high prevalence area may be more likely to seek out testing.
@@
 
-As the necessary covariates (e.g., symptom status, prior testing, COVID-19 exposure) go beyond demographic information, random sampling approaches are required.
-If sufficiently large random samples are collected at each time $t =1,\ldots,T$, then the pseudo-likelihood can be re-written as in~\eqref{eq:auxinfoprob}, but with one term for each time point.
-For COVID-19, there are two key issues with this approach.
-First, large probabilistic samples are not available at every time.
-Second, only cumulative demographics on case counts are available from states' department of health\footnote{Historical daily demographic data can be requested via FOIA, but such requests take months to complete -- as the author has learned through sweat and tears.}.
-Moreover, key covariate information such as symptom status is not reported (and potentially not recorded) by states' department of health.
-
-To address these issues, we rely on the Facebook symptom survey.
-Daily random surveys provide adequate covariate information; however, the daily estimates vary substantially while we expect  while we expect to
-
-Specific weight calculations and adjustments are deferred to Section~\ref{section:applications}.
-Here, we briefly discuss a non-parametric kernel approach.
-We start by considering a time-varying version of equation~\eqref{eq:auxinfoprob2}
-% \begin{equation}
-% \label{eq:tvpropensity}
-% \sum_{t=1}^T \left[ \sum_{j=1}^N I_{j,t} I^{(2)}_{j,t} W_{j,t} \log \left( \frac{\pi_t (X_{j,t}; \theta)}{1-\pi_t(X_{j,t}; \theta)} \right) + \sum_{j=1}^N \tilde W_{j,t} \tilde I_{j,t} \log \left( 1 - \pi_t (X_{j,t}; \theta) \right) \right]
-% \end{equation}
-that accounts for random sampling from within the non-random sample.
-Note random samples may be quite small at each time while we expect $\pi_t(x;\theta)$ to be a smooth function of time; therefore, we introduce a non-parametric approach to estimation of the selection propensity at time~$t$,
-$$
-\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \left[ \sum_{j=1}^N I_{j,t^\prime} I^{(2)}_{j,t^\prime} W_{j,t^\prime} \log \left( \frac{\pi_t (X_{j,t^\prime}; \theta)}{1-\pi_t(X_{j,t^\prime}; \theta)} \right) + \sum_{j=1}^N \tilde W_{j,t^\prime} \tilde I_{j,t^\prime} \log \left( 1 - \pi_t (X_{j,t^\prime}; \theta) \right) \right]
+As   necessary covariates go beyond demographic information, random sampling approaches as discussed in Section~\ref{subsec:auxprob} are required.
+If sufficiently large random samples are collected at each time $t =1,\ldots,T$, then the pseudo-likelihood can be re-written as in~\eqref{eq:auxinfoprob} and propensities estimated separately per time point.
+Unfortunately for COVID-19, large probabilistic samples are not available at every time within a given region.
+For simplicity, here we discuss a non-parametric kernel-based approach where the selection propensity at time~$t$, denoted $\hat \theta_t$, maximizes the smoothed pseudo-likelihood
+$$
+\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \left[ \sum_{j=1}^N I_{j,t^\prime} \log \left( \frac{\pi_t (X_{j,t^\prime}; \theta)}{1-\pi_t(X_{j,t^\prime}; \theta)} \right) + \sum_{j=1}^N \tilde W_{j,t^\prime} \tilde I_{j,t^\prime} \log \left( 1 - \pi_t (X_{j,t^\prime}; \theta) \right) \right]
 $$
 where $K_h$ is a kernel function with tunable parameter $h$.
-Under simple random sampling of equal size~$\tilde n$ from a fixed population size~$N$ at each time, and s, and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the
-$$
-\hat \pi_{k,t} = \frac{n}{N} \times \frac{\sum_{t^\prime=1}^T K(|t^\prime - t|) \frac{n_t}{m} \hat d_{k,t^\prime}}{\sum_{t^\prime=1}^T \tilde d_{k,t^\prime}}
-$$
-where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.
+Under simple random sampling of equal size~$\tilde n$ from a fixed population size~$N$ at each time and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the
+$$
+\hat \pi_{k,t} = \frac{n}{N} \times \frac{\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \hat d_{k,t^\prime}}{\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \tilde d_{k,t^\prime}}
+$$
+where $\tilde d_k$ is the number of observations in strata $k$ in the random sample. Given $\pi (x;\hat \theta_t)$, the prevalence estimator $\bar y_{n,t}^\star$ is given by~\ref{eq:ipwest}.
 
@@
 
-To this point, the primary focus of this paper has been on selection bias in coronavirus case-count data from a survey sampling perspective.
+Up to this point, the primary focus of this paper has been on selection bias in coronavirus case-count data from a survey sampling perspective.
 Here, we consider compartmental model approaches from infectious disease epidemiology.
 Our primary objective is a model-based forecast of strata-level prevalence $\bar y_k$.
-
-Here, we present briefly a basic epidemiological state-space model -- the susceptible, infected, and removed (recovered and death) model, or SIR model.
-The probabilistic SIR model was originally proposed by~\cite{Osthus2017} with only one-dimensional time series of infected proportions; this formulation was extended by~\cite{Song2020} to model coronavirus case-counts.
-Given the weights built in
-
-Let $i_t$ and $Y_t^R$ denote the proportion of currently infected (i.e., would test positive given a RT-PCR test with 100\% sensitivity) and removed cases (i.e., including both recovered cases and deaths) at time $t$.
-Let $\theta_t = (s_t, e_t, i_t^{(T)}, i_t^{(NT)}, r_t)$ where
+To do this, an extension of a standard epidemiological state-space model -- the susceptible, infected, and removed (recovered and death) model, or SIR model -- is presented.
+
+Let $s_t$, $e_t$, $i_t^{(T)}$, $i^{(NT)}$, and $r_t$ denote the proportion of survivors, exposed, currently transmitting and not trasmitting infected, and removed cases (i.e., including both recovered cases and deaths) at time $t$.
+Suppose $K$ strata such that the proportions are defined per strata, i.e., $s_t = (s_{1,t},\ldots, s_{K,t})$ is the proportion of survivors in each strata.
+Let $\theta_t = (s_t, e_t, i_t^{(T)}, i_t^{(NT)}, r_t)$ denote the vector
 

### 2020-10-20 write-up/covid-writeup.tex
Bib and tex update
bib and tex

@@
 
+\begin{lemma}[IPW Variance]
+\label{lemma:ipw}
+Under assumptions, variance is given by:
+$$
+V_{IPW} =
+$$
+An estimator is given by
+$$
+\hat V_{IPW} =
+$$
+\end{lemma}
+
+Lemma~\ref{}
 Let $\tilde I_j (X_j) = I_j  \cdot W(X_j)$ for $j=1,\ldots,N$.  Then the error when comparing weighted estimator~$\bar y_n^\star$ to the true prevalence $\bar Y$ can be expressed as:
@@
 To do this, an extension of a standard epidemiological state-space model -- the susceptible, infected, and removed (recovered and death) model, or SIR model -- is presented.
+Here we consider the probabilistic SIR model was originally proposed by~\cite{Osthus2017} with only one-dimensional time series of infected proportions; this formulation was extended by~\cite{Song2020} to model coronavirus case-counts.
 
@@
 Suppose $K$ strata such that the proportions are defined per strata, i.e., $s_t = (s_{1,t},\ldots, s_{K,t})$ is the proportion of survivors in each strata.
-Let $\theta_t = (s_t, e_t, i_t^{(T)}, i_t^{(NT)}, r_t)$ denote the vector
-
-Assume the two-dimensional process follows
+Let $\theta_t = (s_t, e_t, i_t^{(T)}, i_t^{(NT)}, r_t)$, then we assume the process is governed by the following Markov model:
 $$
@@
 $$
-where $\kappa$ scales the variance of the Dirichlet distribution and $f(\cdot)$ is a function that determines infection dynamics.
-Here, we assume the classical infectious disease SIR model as given by~\eqref{eq:sir} with $(S_t,I_t, R_t)$ replaced by $(\theta_t^S, \theta_t^I, \theta_t^R)$.
-\cite{Song2020}
+where $\kappa$ scales the variance of the Dirichlet distribution and $f(\cdot)$ is a function that determines infection dynamics.  Here, we extend the SIR dynamics to account for strata-level dynamics
 \begin{align*}
-\frac{\partial s_t}{\partial t} &= - \beta s_t i^{(T)}_t; \quad \frac{\partial e_t}{\partial t} = \beta s_t i^{(T)}_t - \sigma e_t; \quad \\
+\frac{\partial s_t}{\partial t} &= - s_t \circ (B \, i^{(T)}_t); \quad \frac{\partial e_t}{\partial t} = s_t \circ (B \, i^{(T)}_t) - \sigma e_t; \quad \\
 \frac{\partial i^{(T)}_t}{\partial t} &= \sigma e_t - \beta_2 i^{(T)}; \quad
@@
 \end{align*}
-where $S_t$ is the number of susceptible individuals, $I_t$ is the number of infected individuals, and $R_t$ is the number of removed (recovered or deceased) individuals at time $t$.
-
-A key concern with the proposed model is lack of .
-We could address this by using IPW methods but this would mimic Section XX and therefore lead to potentially similar results.
-Instead, we opt to follow (cite: HDSR) and model death data.
-$$
-D_t \mid p, \theta, \nu \sim \text{Poisson} \left( p \sum_{t=1}^r \nu_t \theta_{(r-t)} \right)
-$$
-Fully Bayesian algorithms that combine (cite: Song) and (cite: Johndrow) can be derived.
-Here, we focus on the deterministic case-count and resulting SIR model due to availability of existing software~\cite{Song2020} for estimation of such models.
-Unlike in (cite: HDRS), we are interested in predictions of case-counts.
-We split deaths into $<60$ and $>60$, using two different IFR's to back out the counts for each subset and then split these among the sub-populations using the raking weights.
-This yields estimates at $\hat y_k$ for each strata.
-We use $p$ as 0.6 from the
-
-This is of particular importance for epidemiological forecasts built from population-level data and impacts our model choice in Section~\ref{section:modelbased}.
-{\bf Check if SIR models account for this?}
+where $B \in \mathbb{R}^{K \times K}$ and $\circ$ denotes elementwise product of vectors.
+The matrix $B$ allows for strata-specific infection dynamics with the SIR model being recovered if $B_{k,\cdot} = \beta$ for each $k \in [K]$.
+Following~\cite{Song2020}, we employ Runge-Kutta (RK4) approximations for discretization; see Appendix XX for details.
+
+Selection bias in case count data could be direcly addressed by extending the above compartmental model.
+Here, we instead opt to bring in a different source of information by considering COVID-19 death data.
+Let $D_{k,r}$ denote the number of individuals who pass away on day $r$ from strata $k$ and $\nu_{k,t}$ to denote the number of new infections on day $t$ in the $k$th strata.
+Then
+$$
+D_{k,r} \mid p, \theta, \nu \sim \text{Poisson} \left( \sum_{t=1}^r p_{k} \nu_{k,t} \theta_{(r-t)} \right)
+$$
+where $p_{k}$ is the infection fatality rate for the $k$th strata.
+This extends prior analysis of death data~\cite{Johndrow2020} by allowing the infection fatality rate to depend on age.
+Indeed,~\cite{Levin2020} demonstrate that ``the estimated IFR is close to zero for children and younger adults but rises exponentially with age, reaching 0.4\% at age 55, 1.4\% at age 65, 4.6\% at age 75, and 15\% at age 85.'' For conciseness, parameter choices and inference are discussed in depth in Appendix XX.
 
 \subsection{Doubly robust estimation.}
-Two methods have been presented.
-The first uses IPW to adjust for differential selection.
-The second uses epidemiological forecasts based on SIR models.
-Rather than relying solely on either approach, here we show how to combine forecasting and inverse-probability weighting by building a \emph{doubly robust} approach (cite: Robins, etc), which is a common approach for causal inference from observational data.
-In the survey literature, if the weights are known, the approach is referred to as model-assisted estimation.
-Recent work by (cite: Chen) extends the classical survey approach to non-probability samples.
-Here, we do the same.
-We reserve comparison of our approach with (citE: Chen) after introducing the method.
-
-We define the doubly-robust estimator as
-$$
-\bar y_{n}^{(DR)} = \frac{1}{1 - FP - FN} \left[ \frac{1}{N} \sum_{i=1}^N \mu (x_i; \hat \beta) + \frac{1}{\sum_{i=1}^n w(x_i)} \sum_{j=1}^N I_j w(x_i) (Y_i - \mu(x_i; \hat \beta)) \right] - FP.
-$$
-This estimator is ``doubly-robust'' in the sense that it is robust to either the SIR or the propensities being correctly specified.
+Rather than relying solely on IPW or epidemiological forecasts, here we combine forecasting and inverse-probability weighting by extending recent work by~\cite{Chen2019} to account time-varying propensities, measurement-error and random subsampling of nonprobability samples.
+In this context, the doubly-robust estimator is given by
+$$
+\bar y_{n}^{(DR)} = \frac{1}{1 - FP - FN} \left[ \frac{1}{N} \sum_{j=1}^N \hat P_k \hat \mu (x_i) + \frac{1}{\sum_{i=1}^n w(x_i)} \sum_{j=1}^N I_j w(x_i) (Y_i - \hat \mu_k ) \right] - FP.
+$$
+where $x_i$ is the strata information for individual $i$.
+This estimator is ``doubly-robust'' because it is robust to either the model-based forecasts or the propensities being correctly specified.
 Setting $\pi_i = n/N$ and $\mu(x_i; \hat \beta) = 0$ we again recover the estimator $\tilde y_n$ from Section~\ref{section:imperfecttesting}.
 
-The key issue with a doubly-robust estimator (cite: guy from Chen who assumes) is we again don't observe $X_j$ for each $j=1,\ldots,N$.
-There are again two options.
-If we rely on auxiliary information through population-level summary statistics, we have
-$$
-\frac{1}{1 - FP - FN} \left[ \sum_{k} P_k \hat y_k + \sum_{k=1}^K \frac{d_k w_k}{w} (\bar y_k - \hat y_k) \right] - FP
-$$
-where $P_k$ is the population fraction in strata $k$ and $\hat y_k = \mu(k; \hat \beta)$. Alternatively, we can use auxiliary information through probability samples
-$$
-\frac{1}{1-FP-FN} \left[ \frac{1}{N} \sum_{i=1}^N \tilde R_i \tilde W_i \mu (X_i; \hat \theta) + \sum_{i=1}^n \frac{w_i}{w} \left(y_i - \mu(x_i; \hat \beta)\right) \right] - FP.
-$$
-
+The key issue with a doubly-robust estimator is the strata information is not observed for every individual in the population.
+Here, we again use auxiliary information through population-level summary statistics and probability samples and consider the estimator
+$$
+\frac{1}{1-FP-FN} \left[ \sum_{i=1}^N \hat P_k \hat \mu_k + \sum_{i=1}^n \frac{w_i}{w} \left(y_i - \hat \mu (x_i; \theta) \right) \right] - FP.
+$$
+where $\hat P_k$ is the estimated fraction in strata $k$.
+
+\subsection{Variance estimator and error decomposition}
 A plug-in estimator for the variance

### 2020-10-20 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 
-% \numberwithin{equation}{section}
-% \theoremstyle{plain}
-% \newtheorem{theorem}{Theorem}[section]
-% \newtheorem{example}{Example}
-% \newtheorem{remark}{Remark}
-% \newtheorem{corollary}[theorem]{Corollary}
-% \newtheorem{criterion}[theorem]{Criterion}
-% \newtheorem{definition}[theorem]{Definition}
-% \newtheorem{exercise}[theorem]{Exercise}
-% \newtheorem{lemma}[theorem]{Lemma}
-% \newtheorem{proposition}[theorem]{Proposition}
+\numberwithin{equation}{section}
+\theoremstyle{plain}
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{example}{Example}
+\newtheorem{remark}{Remark}
+\newtheorem{corollary}[theorem]{Corollary}
+\newtheorem{criterion}[theorem]{Criterion}
+\newtheorem{definition}[theorem]{Definition}
+\newtheorem{exercise}[theorem]{Exercise}
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{proposition}[theorem]{Proposition}
 
@@
 
-\subsubsection{An IPW estimator and statistical error decomposition.}
+\subsubsection{An IPW estimator}
 
@@
 
-\begin{lemma}[IPW Variance]
-\label{lemma:ipw}
-Under assumptions, variance is given by:
-$$
-V_{IPW} =
-$$
-An estimator is given by
-$$
-\hat V_{IPW} =
-$$
-\end{lemma}
-
-Lemma~\ref{}
-Let $\tilde I_j (X_j) = I_j  \cdot W(X_j)$ for $j=1,\ldots,N$.  Then the error when comparing weighted estimator~$\bar y_n^\star$ to the true prevalence $\bar Y$ can be expressed as:
-\begin{equation}
-\label{eq:statdecomp2}
-\rho_{\tilde I (X), Y} \times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y} \times \underbrace{\left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{D_M}
-\end{equation}
-where $CV_W$ is the coefficient of variation (i.e., standard deviation/mean) of $W_J$ given $I_J = 1$, and $\rho_{\tilde I(X), Y}$ is the empirical correlation which here depends on covariate distribution.
-See Appendix XX for the derivation.
-
-Comparing~\eqref{eq:statdecomp2} to~\eqref{eq:statdecomp} shows that weighting impacts the estimation error in two ways.
-First, there is a negative impact on the data quantity component; taking the ratio of these quantities yields
-$\sqrt{1 + \frac{CV_W^2}{1-f}} \geq 1$.
-Hence, if the data quality does not increase (i.e., $| \rho_{\tilde I (X), Y} | = | \rho_{I,Y}|$ ) then weighting increases the error magnitude.
-Note this relative error increase depends on the fraction of population sampled $f$, implying that for large samples there is a larger potential increase in the error if the weights do not improve data quality.
-
-As demonstrated in the derivation of IPW approach, if the weights are correctly specified then the $E_{\bf R} [ \rho_{\tilde I (X), Y} ] = 0$ and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then the data quality index is unlikely to inversely scale with population size.
-Indeed, Horvitz-Thompson estimators are sensitive to errors in the estimated weights.
-Low selection propensity imply large weights which can have deleterious impacts by certain observations dominating estimation.
-Interestingly, the impact of measurement-error on data quality is unchanged when considering a weighted estimand.
-Similar to~\cite{Meng2018}, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from an IPW estimator are likely to put too much confidence in the sheer size of the data.
-In section XX, independent random samples are used to assess effective sample sizes and adjust corresponding CIs.
-
 \subsubsection{Time-varying propensities}
@@
 where $\tilde d_k$ is the number of observations in strata $k$ in the random sample. Given $\pi (x;\hat \theta_t)$, the prevalence estimator $\bar y_{n,t}^\star$ is given by~\ref{eq:ipwest}.
+
+\subsubsection{Asymptotics and a statistical error decomposition}
+
+Assuming the selection propensity model is correctly specificed, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:ipw} shows that the IPW estimator is  consistent and calculates the estimator's variance:
+
+\begin{lemma}[IPW Variance] \normalfont
+\label{lemma:ipw}
+Under regularity assumptions specified in the Appendix, we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$ (check)  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ (check) where:
+$$
+V_t^{(IPW)} = \frac{1}{N^2} \sum_{i=1}^N (1 - \pi_{i,t}) \pi_{i,t} \left( \right) + b_1^\top D b_1 + \tilde b_2^\top D_2 b_2.
+$$
+where $b_1 = $, $b_2 = $, $D_1 = N^{-2} V_p ()$, $D_2 = N^{-2} V_p$, and $V_{p,1}$ denotes the design-based variance under the probability sampling designs.
+
+A plug-in estimator is given by
+$$
+V_t^{(IPW)} = \frac{1}{N^2} \sum_{i=1}^N I_{j,t} (1 - \pi_{i,t}) \left( \right) + \hat b_{1,t}^\top \hat D_t  \hat b_{1,t} + \tilde \hat b_{2,t}^\top \hat D_{2,t}  \hat b_{2,t}.
+$$
+
+\end{lemma}
+
+Let $\tilde I_j (X_j) = I_j  \cdot W(X_j)$ for $j=1,\ldots,N$.  Then the error when comparing weighted estimator~$\bar y_n^\star$ to the true prevalence $\bar Y$ can be expressed as:
+\begin{equation}
+\label{eq:statdecomp2}
+\rho_{\tilde I (X), Y} \times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y} \times \underbrace{\left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{D_M}
+\end{equation}
+where $CV_W$ is the coefficient of variation (i.e., standard deviation/mean) of $W_J$ given $I_J = 1$, and $\rho_{\tilde I(X), Y}$ is the empirical correlation which here depends on covariate distribution.
+See Appendix XX for the derivation.
+
+Comparing~\eqref{eq:statdecomp2} to~\eqref{eq:statdecomp} shows that weighting impacts the estimation error in two ways.
+First, there is a negative impact on the data quantity component; taking the ratio of these quantities yields
+$\sqrt{1 + \frac{CV_W^2}{1-f}} \geq 1$.
+Hence, if the data quality does not increase (i.e., $| \rho_{\tilde I (X), Y} | = | \rho_{I,Y}|$ ) then weighting increases the error magnitude.
+Note this relative error increase depends on the fraction of population sampled $f$, implying that for large samples there is a larger potential increase in the error if the weights do not improve data quality.
+
+As demonstrated in Lemma~\ref{lemma:ipw}, if the propensity model is correctly specified then the $E_{\bf R} [ \rho_{\tilde I (X), Y} ] = 0$ and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then the data quality index is unlikely to inversely scale with population size.
+Indeed, Horvitz-Thompson estimators are sensitive to errors in the estimated weights.
+Low selection propensity imply large weights which can have deleterious impacts by certain observations dominating estimation.
+Interestingly, the impact of measurement-error on data quality is unchanged when considering a weighted estimand.
+Similar to~\cite{Meng2018}, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from an IPW estimator are likely to put too much confidence in the sheer size of the data.
+In section XX, independent random samples are used to assess effective sample sizes and adjust corresponding CIs.
 

### 2020-10-20 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 Suppose the selection indicator $I_j$ is a Bernoulli random variable that depends only on these covariates, i.e., $P(I_j = 1 \mid X_j = x) = \pi (x; \theta)$.
+For COVID-19 case-count data, the only covariate information provided by state and local governments are summary statistics for those factors discussed in Section~\ref{section:testinginfo} -- age, sex,  and race.
+These are discrete factors and therefore define a stratification of the population.
+Currently, US testing does not collect additional covariate information such as symptom information; however, this need not be the case moving forward.
+Recommendation XX below states that additional covariate information should be collected from nonprobabilistic samples.
+
 Maximum likelihood estimation follows by maximizing
@@
 Typically, this is not possible.
-Here we present two methods for propensity estimation -- use of population-level summary statistics and use of a probability sample.
-
-\subsubsection{Auxiliary information through population-level summary statistics}
-\label{subsec:auxsum}
-
-For COVID-19 case-count data, the only covariate information provided by state and local governments are summary statistics for those factors discussed in Section~\ref{section:testinginfo} -- age, sex,  and race.
-These are discrete factors and therefore define a stratification of the population.
-Then $X_j = k \in \{1,\ldots,K\}$ denotes the $j$th individual belongs to strata $k$ and the likelihood can be re-written as
-$$
-\sum_{k=1}^K d_k \log \left( \frac{\pi (k; \theta)}{1-\pi(k; \theta)} \right) + \sum_{k=1}^K N_k \log \left( 1 - \pi (k; \theta) \right).
-$$
-where $d_k$ and $N_k$ are the number of individuals in the nonprobabilistic sample and in the entire population respectively in the $k$th strata, $k=1,\ldots K$.
-
-Maximum likelihood estimation then requires knowledge of $N_k$, which is feasible when population-level summary statistics are available.
-The US Census, for example, can provide marginal summary statistics for the distribution of age, sex, and race in Indiana state.
-Typically  joint distribution is not publicly available.
-Here, we employ the standard survey technique of raking~\citep{deville1993} to generate estimates $\hat N_k$ for each strata.
-Estimation can then be performed by using standard algorithms such as Newton-Rhapson.
-If the selection propensity is estimated non-parametrically then the resulting weight can be computed exactly and is given by $\hat P(I_j = 1 \mid X_j = k) = \hat \pi_k = d_k/\hat N_k$.
+Here we present a method that uses auxiliary information obtained from probability samples as well as probability subsamples from the self-selected population.
 
@@
 \label{subsec:auxprob}
-
-Currently, US testing does not collect additional covariate information such as symptom information; however, this need not be the case moving forward.
-Recommendation XX below states that additional covariate information should be collected from nonprobabilistic samples.
-If this were the case, population-level summary statistics are no longer sufficient for selection propensity estimation so an alternative source of auxiliary information is required.
 
@@
 $$
-V_t^{(IPW)} = \frac{1}{N^2} \sum_{i=1}^N (1 - \pi_{i,t}) \pi_{i,t} \left( \right) + b_1^\top D b_1 + \tilde b_2^\top D_2 b_2.
-$$
-where $b_1 = $, $b_2 = $, $D_1 = N^{-2} V_p ()$, $D_2 = N^{-2} V_p$, and $V_{p,1}$ denotes the design-based variance under the probability sampling designs.
+V_t^{(IPW)} = \frac{1}{N^2} \sum_{t^\prime = 1}^T \sum_{i=1}^N (1 - \pi_{i,t}) \pi_{i,t} \left( \frac{y_{i,t} - \mu_t}{\pi_{i,t}} -  \right) + b_{1,t}^\top D b_1 + \tilde b_2^\top D_2 b_2.
+$$
+where
+$$
+b_{1,t} = \left[ \sum_{t^\prime=1}^T K_h (|t-t^\prime|)\sum_{j=1}^N (1-\pi_{i,t}) (y_{i,t^\prime} - \mu_{t^\prime}) x_{i,t^\prime}) \right] \left[ N^{-1} \sum_{t^\prime=1^T} K_h (|t-t^\prime|)\sum_{j=1}^N (1-\pi_{i,t} (y_{i,t^\prime} - \mu_{t^\prime}) x_{i,t^\prime})\right] $$,
+$b_2 = $, $D_1 = N^{-2} V_p ()$, $D_2 = N^{-2} V_p$, and $V_{p,1}$ denotes the design-based variance under the probability sampling designs.
 

### 2020-10-28 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 \newtheorem{theorem}{Theorem}[section]
-\newtheorem{example}{Example}
-\newtheorem{remark}{Remark}
+% \newtheorem{example}{Example}
+% \newtheorem{remark}{Remark}
 \newtheorem{corollary}[theorem]{Corollary}
@@
 \usepackage{setspace}
+% \onehalfspacing
 \doublespacing
@@
 In the incubation phase, most molecular tests will return a false negative result.
-Molecular tests are most sensitive when the viral loads are highest which occurs during the first few days of transmissibility (cite: ).
-Moreover, most molecular tests are performed via nasopharyngeal swab (cite: NEJM).
+Molecular tests are most sensitive when the viral loads are highest which occurs during the first few days of transmissibility (cite: that guy ).
+Moreover, most molecular tests are performed via nasopharyngeal swab.
 Specimen collection by swab is known to impact false negative/positive rates regardless of test timing.
-Systematic reviews suggest that 70\% sensitivity and 95\% specificity are reasonable estimates for current RT-PCR tests (cite: NEJM).
+Systematic reviews suggest that 70\% sensitivity and 95\% specificity are reasonable estimates for current RT-PCR tests~\cite{Woloshin2020}.
 
@@
 In this section, we consider two directions.
-First, we use publicly available auxiliary information to build selection propensities and consider an inverse probability weighting (IPW) estimator.
+% First, we use publicly available auxiliary information to build selection propensities and consider an inverse probability weighting (IPW) estimator.
 We then extend the statistical error decomposition to the weighted setting.
@@
 We provide an appropriate extension of the statistical error decomposition.
-The method and statistical error decompositions guide our recommendations in Section~\ref{section:discussion}.
+The proposed methods and statistical error decompositions guide our recommendations in Section~\ref{section:discussion}.
 
@@
 Suppose the selection indicator $I_j$ is a Bernoulli random variable that depends only on these covariates, i.e., $P(I_j = 1 \mid X_j = x) = \pi (x; \theta)$.
-For COVID-19 case-count data, the only covariate information provided by state and local governments are summary statistics for those factors discussed in Section~\ref{section:testinginfo} -- age, sex,  and race.
-These are discrete factors and therefore define a stratification of the population.
-Currently, US testing does not collect additional covariate information such as symptom information; however, this need not be the case moving forward.
-Recommendation XX below states that additional covariate information should be collected from nonprobabilistic samples.
-
 Maximum likelihood estimation follows by maximizing
@@
 This was proposed by~\cite{Chen2019} when covariates were observed on a random sample from the population.
-Under simple random sampling and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the $\hat \pi_k = (d_k / \tilde d_k) (n / N)$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.
-
-Equation~\eqref{eq:auxinfoprob} presupposes covariate information is collected for all individuals in the nonprobability sample; however, this is often not feasible for state and local governments where rapid sample collection is prioritized over additional collection of covariate information.
-We can extend~\eqref{eq:auxinfoprob} by randomly sampling a subset of the nonprobability sample on which to collect the covariate information, yielding the design-unbiased estimator
-\begin{equation}
-\label{eq:auxinfoprob2}
-\sum_{j=1}^N I_j I_j^{(2)} W_j^{(2)} \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
-\end{equation}
-Here, $I_j^{(2)}$ is an indicator that the individual is in the probability subsample and $W_j^{(2)}$ is the probability of inclusion.
-The estimate now relies on random sampling in both the first and second term.
-
-\subsubsection{An IPW estimator}
-
-Selection propensities can be constructed to use both types of auxiliary information (i.e., combine Sections~\ref{subsec:auxsum} and~\ref{subsec:auxprob}).
-Given a selection propensity, define the inverse probability weight~$w(x) = \pi (x; \hat \theta)^{-1}$.
-Then the IPW estimator adjusted for measurement error is given by
+% Under simple random sampling and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the $\hat \pi_k = (d_k / \tilde d_k) (n / N)$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.
+
+% Equation~\eqref{eq:auxinfoprob} presupposes covariate information is collected for all individuals in the nonprobability sample; however, this is often not feasible for state and local governments where rapid sample collection is prioritized over additional collection of covariate information.
+We can extend~\eqref{eq:auxinfoprob} by randomly sampling a subset of the nonprobability sample on which to collect the covariate information, yielding the pseudo-likelihood
+% \begin{equation}
+% \label{eq:auxinfoprob2}
+% \sum_{j=1}^N I_j I_j^{(2)} W_j^{(2)} \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
+% \end{equation}
+% Here, $I_j^{(2)}$ is an indicator that the individual is in the probability subsample and $W_j^{(2)}$ is the probability of inclusion.
+Estimation proceeds by maximizing the pseudo-likelihood which now depends on random sampling in both the first and second term.
+
+\subsubsection{An IPW estimator and statistical error decomposition}
+\label{section:IPWerrordecomp}
+
+Given a selection propensity, define the inverse probability weight~$w(x) = \pi (x; \hat \theta)^{-1}$. Then the IPW estimator adjusted for measurement error is given by
 \begin{equation}
@@
 Under simple random sampling with one strata~$\hat \pi_k = n/N$ and we recover the estimator $\tilde y_n$ from Section~\ref{section:imperfecttesting}.
+
+Let $\tilde I_j (X_j) = I_j  \cdot W(X_j)$ for $j=1,\ldots,N$.  Then the error when comparing weighted estimator~$\bar y_n^\star$ to the true prevalence $\bar Y$ can be expressed as:
+\begin{equation}
+\label{eq:statdecomp2}
+\rho_{\tilde I (X), Y} \times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y} \times \underbrace{\left[ 1 - \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{\tilde D_M}
+\end{equation}
+where $CV_W$ is the coefficient of variation (i.e., standard deviation/mean) of $W_J$ given $I_J = 1$, $\rho_{\tilde I(X), Y}$ is the empirical correlation which here depends on covariate distribution, $\tilde f_k = E[ W_J I_J \mid Y_j = k]$ for $k=0,1$, and $\tilde \Delta = \tilde f_1 - \tilde f_0$.
+See Appendix~\ref{app:ipwderivation} for the derivation.
+
+Comparing~\eqref{eq:statdecomp2} to~\eqref{eq:statdecomp} shows that weighting impacts the estimation error in three ways.
+First, there is a negative impact on the data quantity component; taking the ratio of these quantities yields
+$\sqrt{1 + \frac{CV_W^2}{1-f}} \geq 1$.
+Hence, if the data quality does not increase (i.e., $| \rho_{\tilde I (X), Y} | = | \rho_{I,Y}|$ ) then weighting increases the error magnitude.
+Note this relative error increase depends on the fraction of population sampled $f$, implying that for large samples there is a larger potential increase in the error if the weights do not improve data quality.
+Third, the impact of measurement-error on data quality is changed when considering a weighted estimand.
+{\red state that it may switch sign!als}
+
+As demonstrated in Lemma~\ref{lemma:ipw}, if the propensity model is correctly specified then the $E_{\bf R} [ \rho_{\tilde I (X), Y} ] = 0$ and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then the data quality index is unlikely to inversely scale with population size.
+Indeed, Horvitz-Thompson estimators are sensitive to errors in the estimated weights.
+Low selection propensity imply large weights which can have deleterious impacts by certain observations dominating estimation.
+Similar to~\cite{Meng2018}, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from an IPW estimator are likely to put too much confidence in the sheer size of the data.
 
@@
 
-As   necessary covariates go beyond demographic information, random sampling approaches as discussed in Section~\ref{subsec:auxprob} are required.
 If sufficiently large random samples are collected at each time $t =1,\ldots,T$, then the pseudo-likelihood can be re-written as in~\eqref{eq:auxinfoprob} and propensities estimated separately per time point.
@@
 $$
-where $K_h$ is a kernel function with tunable parameter $h$.
-Under simple random sampling of equal size~$\tilde n$ from a fixed population size~$N$ at each time and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the
-$$
-\hat \pi_{k,t} = \frac{n}{N} \times \frac{\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \hat d_{k,t^\prime}}{\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \tilde d_{k,t^\prime}}
-$$
-where $\tilde d_k$ is the number of observations in strata $k$ in the random sample. Given $\pi (x;\hat \theta_t)$, the prevalence estimator $\bar y_{n,t}^\star$ is given by~\ref{eq:ipwest}.
-
-\subsubsection{Asymptotics and a statistical error decomposition}
-
-Assuming the selection propensity model is correctly specificed, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:ipw} shows that the IPW estimator is  consistent and calculates the estimator's variance:
+where $K_h$ is a kernel function with tunable parameter $h$. Given $\pi (x;\hat \theta_t)$, the prevalence estimator $\bar y_{n,t}^\star$ is given by~\ref{eq:ipwest}.
+
+% Under simple random sampling of equal size~$\tilde n$ from a fixed population size~$N$ at each time and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the
+% $$
+% \hat \pi_{k,t} = \frac{n}{N} \times \frac{\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \hat d_{k,t^\prime}}{\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \tilde d_{k,t^\prime}}
+% $$
+% where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.
+
+\subsubsection{Asymptotics}
+
+Assuming the selection propensity model is correctly specified, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:ipw} shows that the IPW estimator is  consistent and calculates the estimator's variance.
+In Lemma~\ref{lemma:ipw}, sensitivity and specificity are not assumed known.
+Instead, we suppose that these are estimated using a pseudo-likelihood on a random sample from the population whose size is chosen so the resulting estimated sensitivity and specificity have desired levels of uncertainty.
+See Appendix~\ref{app:asympderivations} for additional details.
 
@@
 \label{lemma:ipw}
-Under regularity assumptions specified in the Appendix, we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$ (check)  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ (check) where:
-$$
-V_t^{(IPW)} = \frac{1}{N^2} \sum_{t^\prime = 1}^T \sum_{i=1}^N (1 - \pi_{i,t}) \pi_{i,t} \left( \frac{y_{i,t} - \mu_t}{\pi_{i,t}} -  \right) + b_{1,t}^\top D b_1 + \tilde b_2^\top D_2 b_2.
-$$
-where
-$$
-b_{1,t} = \left[ \sum_{t^\prime=1}^T K_h (|t-t^\prime|)\sum_{j=1}^N (1-\pi_{i,t}) (y_{i,t^\prime} - \mu_{t^\prime}) x_{i,t^\prime}) \right] \left[ N^{-1} \sum_{t^\prime=1^T} K_h (|t-t^\prime|)\sum_{j=1}^N (1-\pi_{i,t} (y_{i,t^\prime} - \mu_{t^\prime}) x_{i,t^\prime})\right] $$,
-$b_2 = $, $D_1 = N^{-2} V_p ()$, $D_2 = N^{-2} V_p$, and $V_{p,1}$ denotes the design-based variance under the probability sampling designs.
-
-A plug-in estimator is given by
-$$
-V_t^{(IPW)} = \frac{1}{N^2} \sum_{i=1}^N I_{j,t} (1 - \pi_{i,t}) \left( \right) + \hat b_{1,t}^\top \hat D_t  \hat b_{1,t} + \tilde \hat b_{2,t}^\top \hat D_{2,t}  \hat b_{2,t}.
-$$
-
+The estimates~$\mu_t := \bar y_{n,t}^\star$ and $\hat \pi_{j,t}$
+are solutions to the following set of estimating equations
+$$
+\Phi_n (\eta_t) =
+\left (
+\begin{array}{c}
+\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - (1-FP-FN) \cdot (\mu_t+FP)}{\pi_{j,t}} \\
+\frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
+\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
+\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
+\end{array}
+\right ) = {\bf 0}
+$$
+where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
+Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$ (check)  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ where $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$.
 \end{lemma}
-
-Let $\tilde I_j (X_j) = I_j  \cdot W(X_j)$ for $j=1,\ldots,N$.  Then the error when comparing weighted estimator~$\bar y_n^\star$ to the true prevalence $\bar Y$ can be expressed as:
-\begin{equation}
-\label{eq:statdecomp2}
-\rho_{\tilde I (X), Y} \times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y} \times \underbrace{\left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{D_M}
-\end{equation}
-where $CV_W$ is the coefficient of variation (i.e., standard deviation/mean) of $W_J$ given $I_J = 1$, and $\rho_{\tilde I(X), Y}$ is the empirical correlation which here depends on covariate distribution.
-See Appendix XX for the derivation.
-
-Comparing~\eqref{eq:statdecomp2} to~\eqref{eq:statdecomp} shows that weighting impacts the estimation error in two ways.
-First, there is a negative impact on the data quantity component; taking the ratio of these quantities yields
-$\sqrt{1 + \frac{CV_W^2}{1-f}} \geq 1$.
-Hence, if the data quality does not increase (i.e., $| \rho_{\tilde I (X), Y} | = | \rho_{I,Y}|$ ) then weighting increases the error magnitude.
-Note this relative error increase depends on the fraction of population sampled $f$, implying that for large samples there is a larger potential increase in the error if the weights do not improve data quality.
-
-As demonstrated in Lemma~\ref{lemma:ipw}, if the propensity model is correctly specified then the $E_{\bf R} [ \rho_{\tilde I (X), Y} ] = 0$ and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then the data quality index is unlikely to inversely scale with population size.
-Indeed, Horvitz-Thompson estimators are sensitive to errors in the estimated weights.
-Low selection propensity imply large weights which can have deleterious impacts by certain observations dominating estimation.
-Interestingly, the impact of measurement-error on data quality is unchanged when considering a weighted estimand.
-Similar to~\cite{Meng2018}, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from an IPW estimator are likely to put too much confidence in the sheer size of the data.
-In section XX, independent random samples are used to assess effective sample sizes and adjust corresponding CIs.
 
@@
 This extends prior analysis of death data~\cite{Johndrow2020} by allowing the infection fatality rate to depend on age.
-Indeed,~\cite{Levin2020} demonstrate that ``the estimated IFR is close to zero for children and younger adults but rises exponentially with age, reaching 0.4\% at age 55, 1.4\% at age 65, 4.6\% at age 75, and 15\% at age 85.'' For conciseness, parameter choices and inference are discussed in depth in Appendix XX.
+Indeed,~\cite{Levin2020} demonstrate that ``the estimated IFR is close to zero for children and younger adults but rises exponentially with age, reaching 0.4\% at age 55, 1.4\% at age 65, 4.6\% at age 75, and 15\% at age 85.'' For conciseness, parameter choices and inference are discussed in depth in Appendix~\ref{app:modelbased}.
 
 \subsection{Doubly robust estimation.}
-Rather than relying solely on IPW or epidemiological forecasts, here we combine forecasting and inverse-probability weighting by extending recent work by~\cite{Chen2019} to account time-varying propensities, measurement-error and random subsampling of nonprobability samples.
+Rather than relying solely on IPW or epidemiological forecasts, here we combine forecasting and inverse-probability weighting by extending recent work by~\cite{Chen2019} to account time-varying propensities and measurement-error.
 In this context, the doubly-robust estimator is given by
 $$
-\bar y_{n}^{(DR)} = \frac{1}{1 - FP - FN} \left[ \frac{1}{N} \sum_{j=1}^N \hat P_k \hat \mu (x_i) + \frac{1}{\sum_{i=1}^n w(x_i)} \sum_{j=1}^N I_j w(x_i) (Y_i - \hat \mu_k ) \right] - FP.
-$$
-where $x_i$ is the strata information for individual $i$.
-This estimator is ``doubly-robust'' because it is robust to either the model-based forecasts or the propensities being correctly specified.
-Setting $\pi_i = n/N$ and $\mu(x_i; \hat \beta) = 0$ we again recover the estimator $\tilde y_n$ from Section~\ref{section:imperfecttesting}.
-
-The key issue with a doubly-robust estimator is the strata information is not observed for every individual in the population.
-Here, we again use auxiliary information through population-level summary statistics and probability samples and consider the estimator
-$$
-\frac{1}{1-FP-FN} \left[ \sum_{i=1}^N \hat P_k \hat \mu_k + \sum_{i=1}^n \frac{w_i}{w} \left(y_i - \hat \mu (x_i; \theta) \right) \right] - FP.
-$$
-where $\hat P_k$ is the estimated fraction in strata $k$.
-
-\subsection{Variance estimator and error decomposition}
-A plug-in estimator for the variance
-$$
-\text{Var} (\bar y_{DR}^\star) =
-$$
-See Appendix XX for a derivation.
-
-Remark that the robustness is under certain assumptions which we know are likely false; therefore, we are interested in
-
-Trying to decompose this error we start by recognizing $\frac{1}{N} \sum_{i=1}^N Y_i = \sum_k P_k \bar Y_k$.  Then we can concern ourselves with the per strata errors:
-$$
-\left( P_k - \frac{d_k w_k}{w} \right) \hat y_k + \frac{d_k w_k}{w} \frac{\bar y_k - FP}{1-FP-FN}
-$$
-
-{\bf where to describe difference from Chen. (1) weight normalized, (2) use aux info from multiple sources and use raking, (3) use a epi model rather than covariate pred model}.
+\bar y_{n}^{(DR)} = \frac{1}{N} \sum_{j=1}^N \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \mu(x_j) \right).
+$$
+where the estimates $\hat \mu(x_j)$ are not corrected for measurement-error as they are built to estimate true case count from death-only data.
+This estimator is called ``doubly-robust'' because it is consistent if either the model-based forecasts or the time-varying propensities are correctly specified.
+A statistical error decomposition can be derived
+\begin{align}
+\label{eq:statdecomp3}
+\begin{split}
+\rho_{\tilde I (X), Y-\mu(X)} &\times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y-\mu(X)}  \\
+&\times \underbrace{\left[ 1 - \frac{\rho_{\tilde I(X),Y} \sigma_{Y}}{\rho_{\tilde I(X),Y-\mu(X)} \sigma_{Y-\mu(X)}} \times \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right]}_{\tilde D_M}.
+\end{split}
+\end{align}
+See Appendix~\ref{app:DRderivation} for the derivation.
+Comparing~\eqref{eq:statdecomp3} to~\eqref{eq:statdecomp2} shows that the addition of the model-based focast impacts the estimation error in two ways.
+First, if the model is adequate, then one may expect a reduction in the problem difficulty and (potentially) in the data quality components.
+Note that the model-based forecast does not impact the data quantity component.
+
+If the model-based forecast is correct then $E_{\bf R} [ \rho_{\tilde I (X), Y-\mu(X)} ] = 0$  and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the forecasts are not correctly specified then the data quality index is unlikely to inversely scale with population size.
+Interestingly, the impact of measurement-error on data quality now depends on a relative comparison of the data quality and problem difficulty of the weighted estimator and the doubly-robust estimator.
+For decent models, one would expect this ratio to be greater than one.
+Therefore, $\tilde \Delta < 0$ implies an expected increase in impact of measurement-error and $\tilde \Delta > 0$ implies the magnitude could shrink toward zero or change sign and have an even large magnitude in the opposite direction.
+Again, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from a DR estimator are likely to put too much confidence in the sheer size of the data.
+
+Unfortunately, the first term cannot be computed as covariate information is not collected on the entire population.
+Here, we again use the probability samples and consider the doubly-robust estimator
+$$
+\frac{1}{\sum_{j=1}^N \tilde I_j \tilde W_j} \sum_{j=1}^N  \tilde I_j \tilde W_j \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \mu(x_j) \right).
+$$
+where $\tilde I_j$ indicates an individual is included in the random sample and $\tilde W_j$ is the associated weight for that unit.
+Assuming the selection propensity model is correctly specified, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:dr} shows that the IPW estimator is  consistent and calculates the estimator's variance.
+See Appendix~\ref{app:asympderivations} for additional details.
+
+\begin{lemma}[DR Variance] \normalfont
+\label{lemma:dr}
+The estimates~$\mu_t := \bar y_{n,t}^\star$ and $\hat \pi_{j,t}$
+are solutions to the following set of estimating equations
+$$
+\Phi_n (\eta_t) =
+\left (
+\begin{array}{c}
+\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - (1-FP-FN) \cdot (\mu_t+FP)}{\pi_{j,t}} \\
+\frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
+\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
+\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
+\end{array}
+\right ) = {\bf 0}
+$$
+where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
+Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$ (check)  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ where $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$.
+\end{lemma}
 
@@
 \subsection{COVID-19 active infection prevalence in Indiana}
+
+% For COVID-19 case-count data, the only covariate information provided by state and local governments are summary statistics for those factors discussed in Section~\ref{section:testinginfo} -- age, sex,  and race.
+These are discrete factors and therefore define a stratification of the population.
+Currently, US testing does not collect additional covariate information such as symptom information; however, this need not be the case moving forward.
+Recommendation XX below states that additional covariate information should be collected from nonprobabilistic samples.
 
@@
 
-\begin{subsubsection}{Time-varying doubly-robust prevalence estimation}
+\subsubsection{Time-varying doubly-robust prevalence estimation}
 
@@
 
+\newpage
 \appendix
@@
 
+\section{IPW statistical error decomposition derivation}
+\label{app:ipwderivation}
+We start by considering the empirical weighted mean estimator under imperfect testing,
+$$
+\bar y_n^\star = \frac{\sum_{j=1}^N W_j Y_j^\star I_j}{\sum_{j=1}^N W_j I_j} = \frac{\sum_{i=1}^N  I_j W_j Y_j^\star }{\sum_{j=1}^N  I_j } = \frac{\sum_{i=1}^N  I_j \left[ Y_j (1-P_j) + (1-Y_j) P_j \right]}{\sum_{j=1}^N  I_j }
+$$
+Then
+$$
+\begin{aligned}
+\bar y_n^\star  - \bar Y_N &= \frac{E_J \left[ I_J W_J \left[ Y_J (1-P_J) + (1-Y_J) P_J \right] \right]}{E_J [ I_J W_J ] } - E_J[Y_J] \\
+&= \frac{E_J \left[ I_J W_J P_J (1-2Y_J) \right]}{E_J [ I_J W_J ] } + \left( \frac{E_J [I_J W_J Y_J]}{E_J [ I_J W_J ] } - \frac{E_J[Y_J] E_J[I_J W_J]}{E_J[I_J W_J]} \right) \\
+\end{aligned}
+$$
+The term in parentheses can be re-written as
+$$
+\begin{aligned}
+\frac{E_J [I_J W_J Y_J]- E_J[Y_J] E_J[I_J W_J]}{E_J[I_J W_J]} &=  \frac{E_J [I_J W_J Y_J]- E_J[Y_J] E_J[I_J W_J]}{\sqrt{V_J(I_J W_J) V_J(Y_J)}} \frac{\sqrt{V_J(I_J W_J)}}{E_J[I_J]} \times \sqrt{V_J(Y_J)} \\
+&= \rho_{\tilde I (X),Y} \times \frac{\sqrt{V_J(I_J W_J)}}{E_J[I_J]} \times \sigma_Y
+\end{aligned}
+$$
+Then
+\begin{align*}
+E_J (I_J W_J) &= P(I_J = 1) E_J [W_J \mid I_J = 1] = f \times E_J [W_J \mid I_J = 1] \\
+V_J(I_J W_J) &= E[ I_J W_J^2] - E[I_J W_J]^2 \\
+&= f \cdot \left( E [W_J^2 \mid I_J = 1] - f E[W_J \mid I_J = 1 ]^2 \right) \\
+&= f \cdot \left( E [W_J^2 \mid I_J = 1] \pm E[W_J \mid I_j = 1]^2 - f E[W_J \mid I_J = 1 ]^2 \right) \\
+&= f \cdot \left( V (W_J \mid I_J = 1) + E[W_J \mid I_J = 1]^2 (1-f) \right)
+\end{align*}
+Taking the ratio:
+$$
+\sqrt{\frac{f \cdot \left( V (W_J \mid I_J = 1) + E[W_J \mid I_J = 1]^2 (1-f) \right)}{f^2 E_J [W_J \mid I_J = 1]^2}} = \sqrt{\frac{1 - f + CV(W)^2}{f}}
+$$
+which agrees with Meng's (2019) decomposition of weighted outcome.
+For the other term, first we define $Z_j := 1 - 2 Y_j $.
+Then $Z_j = 1$ if $Y_j = 0$ and $Z_j = -1$ if $Y_j = 1$.
+Then the term can be re-written as
+$$
+\begin{aligned}
+\frac{E_J \left[ I_J W_J P_J (1-2Y_J) \right]}{E_J [ I_J W_J ] } &= \left( \frac{E_J \left[ I_J W_J P_J Z_J \right]}{E_J [ I_J W_J ] } -  \frac{E_J \left[ P_J Z_J \right] E_J[ I_J W_J]}{E_J [ I_J W_J ] } \right) +  \frac{E_J \left[ P_J Z_J \right] E_J[ I_J W_J]}{E_J [ I_J W_J ] } \\
+\end{aligned}
+$$
+The term in parentheses can be re-expressed using the previous technique as:
+$$
+\rho_{\tilde I, PZ} \times \sqrt{\frac{1-f+CV(W)^2}{f}} \times \sigma_{PZ}
+$$
+where now the ``data defect'' and ``problem difficulty'' are with respect to $PZ$ rather than $Y$. The final term is equal to
+$$
+\begin{aligned}
+E_J [P_J Z_J ] &= E_J [ E_J [ P_J Z_J \mid Y_J ] ] \\
+&= \pr (P = 1 \mid Y = 0) (1-\bar Y) - \pr(P=1 \mid Y = 1) \bar Y \\
+&= FP - (FP + FN) \cdot \bar Y
+\end{aligned}
+$$
+Combining these yields:
+$$
+\bar y_n^\star - \bar Y = \sqrt{\frac{1-f+CV(W)^2}{f}} \left(\rho_{\tilde I(X),Y} \sigma_Y + \rho_{\tilde I(X), PZ} \sigma_{PZ} \right) + \left( FP - (FP+FN) \bar Y \right)
+$$
+By previous arguments, we have
+$$
+\rho_{\tilde I (X), PZ} = \rho_{\tilde I(X), Y} \times \frac{C(\tilde I(X), PZ)}{C(\tilde I(X), Y)} \sqrt{ \frac{1-\bar Y}{ 2 (FP(1-\bar Y) + FN \bar Y)} }
+$$
+Then
+\begin{align*}
+C(IW, PZ) &= E[IWPZ] - E[IW] E[PZ] \\
+&=  [FP \tilde f_0  - (FP \tilde f_0 + FN \tilde f_1) \bar Y]  - \tilde f \cdot ( FP - (FP+FN) \bar Y )\\
+&= -FP \tilde \Delta \bar Y + FP \bar Y^2 \Delta - FN \bar Y^2 \Delta \\
+&= -\tilde \Delta \bar Y ( FP \cdot (1-\bar Y) + FN \cdot \bar Y)
+\end{align*}
+where $\tilde f_i = E[ I_J W_J \mid Y_J = i]$ for $i \in \{0,1 \}$ and $f = \tilde f_1 \bar Y + \tilde f_0 (1-\bar Y)$. Moreover,
+\begin{align*}
+C(\tilde I,Y) &= E[ I W Y ] - \tilde f \bar Y \\
+&= \tilde f_1 \bar Y + \tilde f_0 (1-\bar Y) - \tilde f \bar Y \\
+&= \tilde f_0 (1-\bar Y) + \tilde \Delta (1-\bar Y) \bar Y \\
+&= (1- \bar Y) (\tilde f_0 + \tilde \Delta \bar Y)
+\end{align*}
+Combining yields
+$$
+\begin{aligned}
+\rho_{\tilde I (X),PZ} &= \rho_{\tilde I(X),Y} \times \frac{- \tilde \Delta \bar Y (FP \cdot (1-\bar Y) + FN \cdot \bar Y) }{(1-\bar Y) (\tilde f_0 + \tilde \Delta \bar Y)} \times \sqrt{ \frac{(1-\bar Y)}{2 ( FP (1-\bar Y) + FN \cdot \bar Y)} } \\
+&= - \rho_{\tilde I(X), Y} \times \Delta \times \sqrt{\frac{\bar Y}{1-\bar Y}} \frac{\sqrt{FP(1-\bar Y) + FN \cdot \bar Y}}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \times \sqrt{\frac{\bar Y}{2}}
+\end{aligned}
+$$
+We can then re-write $\rho_{\tilde I (X),Y} \sigma_Y + \rho_{\tilde I(X),PZ} \sigma_{PZ}$ as
+$$
+\rho_{\tilde I(X),Y} \sigma_Y \left( 1 - \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right).
+$$
+Thus yielding the desired result.
+
+\section{Doubly robust statistical error decomposition derivation}
+\label{app:DRderivation}
+We start by considering the empirical weighted mean estimator under imperfect testing,
+\begin{align*}
+&\frac{1}{N} \sum_{j=1}^N \mu(X_j) - \frac{\sum_{j=1}^N I_j W_j \mu(X_j)}{\sum_{j=1}^N I_j W_j} + \frac{\sum_{j=1}^N I_j W_j Y_j^\star}{\sum_{j=1}^N I_j W_j} \\
+=& \frac{\sum_{i=1}^N  I_j W_j \left( \left[ Y_j (1-P_j) + (1-Y_j) P_j \right] - \mu (X_j) \right)}{\sum_{j=1}^N  I_j W_j } + \frac{1}{N} \sum_{j=1}^N \mu(X_j
+)
+\end{align*}
+Then
+$$
+\begin{aligned}
+&\frac{E_J \left[ I_J W_J \left( \left[ Y_J (1-P_J) + (1-Y_J) P_J \right] - \mu(X_J) \right) \right]}{E_J [ I_J W_J ] } - E_J[Y_J -\mu (X_J)] \\
+= &\frac{E_J \left[ I_J W_J P_J (1-2Y_J) \right]}{E_J [ I_J W_J ] } + \left( \frac{E_J [I_J W_J (Y_J -\mu (X_J))]}{E_J [ I_J W_J ] } - \frac{E_J[Y_J-\mu(X_J)] E_J[I_J W_J]}{E_J[I_J W_J]} \right) \\
+\end{aligned}
+$$
+The term in parentheses can be re-written as
+$$
+\begin{aligned}
+&\frac{E_J [I_J W_J (Y_J-\mu(X_J))]- E_J[(Y_J-\mu(X_J))] E_J[I_J W_J]}{E_J[I_J W_J]} \\
+=  &\frac{E_J [I_J W_J (Y_J-\mu(X_J))]- E_J[(Y_J-\mu(X_J))] E_J[I_J W_J]}{\sqrt{V_J(I_J W_J) V_J(Y_J-\mu(X_J))}} \frac{\sqrt{V_J(I_J W_J)}}{E_J[I_J]} \times \sqrt{V_J(Y_J-\mu(X_J))} \\
+= &\rho_{\tilde I (X),Y-\mu(X)} \times \frac{\sqrt{V_J(I_J W_J)}}{E_J[I_J]} \times \sigma_{Y-\mu(X)}
+\end{aligned}
+$$
+From prior derivations, we then have the desired result.
+
+\section{Asymptotic derivations}
+\label{app:asympderivations}
+
+% https://www.thelancet.com/journals/lanres/article/PIIS2213-2600%2820%2930453-7/fulltext
+% https://www.medrxiv.org/content/10.1101/2020.04.26.20080911v1.full.pdf
+
+Based on~\cite{Woloshin2020}, we set the false positive rate to 2.4\% and false negative rate to 17.5\%.
+To allow for potential uncertainty, we assume two \emph{synthetic} simple random samples.
+The first is a simple random sample of true negative cases denoted $S_{FP}$; the second is simple random sample of true positive cases denoted $S_{FN}$.
+As $\mu_t$ and $\theta_{t}$ are estimated independently at each time $t$, we can focus on the estimating equations per time point separately.
+The estimates are given by:
+$$
+\left (
+\begin{array}{c}
+\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - (1-FP-FN) \cdot (\mu_t+FP)}{\pi_{j,t}} \\
+\frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
+\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
+\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
+\end{array}
+\right ) = {\bf 0}
+$$
+where $\pi_{j,t^\prime} = \pi (X_{j,t^\prime}; \theta_t)$, $Z_j$ is an indicator of a false positive in the sample $S_{FP}$  and $\tilde Z_j$ is an indicator of a false negative in the sample $S_{FN}$.
+We assume the synthetic $|S_{FP}|^{-1} \sum_{j \in S_{FP}} Z_j = 0.024$ and $|S_{FP}|^{-1} \sum_{j \in S_{FP}} Z_j = 0.175$.
+Sample sizes, denoted $n_{FP}$ and $n_{FN}$ respectively, are chosen to achieve desired estimator variance: $1.96/0.155^2 \cdot \sqrt{0.175 \cdot (1-0.175)} = 31$.
+
+Under joint randomization of propensity score model and sampling designs, we have $E [ \Phi_n (\eta) ] = {\bf 0}$.
+Consistency then follows by arguments in Section 3.2 of Tsiatis (2006).
+Under conditions C1-C6 from~\cite{Chen2019} and the additional assumption that $h_n \to 0$ as the sample size per timepoint of the sample size grows, we have $\Phi_n (\hat \eta) = 0$ and $\Phi_n (\eta_0) = O_p ( \tilde n^{-1/2} )$.
+By a first-order Taylor expansion we have $\hat \eta - \eta_0 = \left[ \right]^{-1} \Phi_n(\eta_0) + o_p (\tilde n^{-1/2})$ where $\phi_n(\eta) = \partial \Phi_n (\eta)/\partial \eta$.
+To calculate
+$$
+\frac{\partial \Phi (\eta)}{\partial \mu_t} =
+\left (
+\begin{array}{c}
+-\frac{(1-FP-FN)}{N} \sum_{j=1}^N \frac{I_{j,t}}{\pi_{j,t}} \\
+0 \\
+0 \\
+0
+\end{array}
+\right )
+$$
+
+$$
+\frac{\partial \Phi (\eta)}{\partial \theta_t} =
+\left (
+\begin{array}{c}
+-\frac{1}{N} \sum_{j=1}^N I_{j,t} (Y_{j,t} - (1-FP-FN) \cdot (\mu_t + FP)) \frac{1-\pi_{j,t}}{\pi_{j,t}} X_{j,t}^\top \\
+- \frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} (1-\pi_{j,t^\prime}) X_{j,t^\prime} X_{j,t^\prime}^\top\\
+0 \\
+0
+\end{array}
+\right )
+$$
+
+$$
+\frac{\partial \Phi (\eta)}{\partial FP} =
+\left (
+\begin{array}{c}
+-\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{(1-2 \cdot FP-FN -\mu_t)}{\pi_{j,t}} \\
+0\\
+-\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP^2} + \frac{1-Z_j}{(1-FP)^2}  \\
+0
+\end{array}
+\right )
+$$
+
+$$
+\frac{\partial \Phi (\eta)}{\partial FN} =
+\left (
+\begin{array}{c}
+\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{\mu_t+FP}{\pi_{j,t}} \\
+0\\
+0 \\
+-\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN^2} + \frac{1-\tilde Z_j}{(1-FN)^2}
+\end{array}
+\right )
+$$
+
+We are interested in computing $- E \left[ \frac{\partial \Phi (\eta)}{\partial \mu_t} \right]^{-1}$ where the expectation is with respect to the random indicators.
+First, $- E \left[ \frac{\partial \Phi (\eta)}{\partial \mu_t} \right]$ is equal to
+$$
+\left (
+\begin{array}{c c c c}
+1-FP-FN & \frac{1}{N} \sum_{j=1}^N \zeta_{t,j} (1-\pi_{j,t}) X_{j,t}^\top & 1 - 2FP-FN - \mu_t & \mu_t + FP \\
+0 & \frac{1}{N} \sum_{t^\prime=1}^T K_{t,t^\prime} \sum_{j=1}^N \pi_{j,t} (1-\pi_{j,t}) X_{j,t} X_{j,t}^\top & 0 & 0\\
+0 & 0 & FP (1-FP)  & 0\\
+0 & 0 & 0 & FN (1-FN)
+\end{array}
+\right )
+$$
+where $\zeta_{t,j} = (Y_{j,t} - (1-FP-FN) \cdot (\mu_t + FP))$.
+This matrix can be written as a diagonal matrix equal to the diagonal of $\partial \Phi_n (\eta)/\partial \eta$ and then a rank-one matrix that is zero except for the first row on the off-diagonal.
+Let $A + B$ denote the sum then
+
+$$
+(A+B)^{-1} = A^{-1} - \frac{1}{1+g} A^{-1} B A^{-1}
+$$
+where $g = \text{trace}(BA^{-1}) = 0$ by design.  In particular,
+$$
+\left (
+\begin{array}{c c c c}
+\frac{1}{1-FP-FN} & b_1 & b_2  & b_3a  \\
+0 & \left[\frac{1}{N} \sum_{t^\prime=1}^T K_{t,t^\prime} \sum_{j=1}^N \pi_{j,t} (1-\pi_{j,t}) X_{j,t} X_{j,t}^\top\right]^{-1} & 0 & 0\\
+0 & 0 & \left[ FP (1-FP) \right]^{-1}  & 0\\
+0 & 0 & 0 & \left[ FN (1-FN) \right]^{-1}
+\end{array}
+\right)
+$$
+where
+\begin{align*}
+b_1 =& (1-FP-FN)^{-1} \left[ N^{-1} \sum_{t^\prime=1}^T K_{t,t^\prime} \sum_{j=1}^N (1-\pi_{j,t^\prime}) \zeta_{t,j} X_{i,t^\prime}^\top \right] \\
+&\times \left[N^{-1} \sum_{t^\prime=1}^T K_{t,t^\prime} \sum_{j=1}^N \pi_{j,t^\prime} (1-\pi_{j,t^\prime}) X_{i,t^\prime} X_{i,t^\prime}^\top \right]^{-1},
+\end{align*}
+$b_2 = (1-FP-FN)^{-1} (1-2FP-FN-\mu_t) / (FP(1-FP))$, and $b_3 = (1-FP-FN)^{-1} (\mu_t+FP) / (FN(1-FN))$.
+
+Note that $\text{Var}(\Phi_n(\eta))$ can be decomposed into $A_{1,t} + A_{2,t} + A_{3,t} + A_{4,t}$
+
+$$
+A_{1,t} = \frac{1}{N} \left( \begin{array}{c}
+\sum_{j=1}^N \frac{I_{j,t} (Y_{j,t} - (1-FP-FN) (\mu_t + FP)}{\pi_{j,t}} \\
+\sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \pi_{j,t^\prime} X_{j,t^\prime} \\
+0 \\
+0
+\end{array}
+\right)
+$$
+and
+$$
+A_{2,t} = \frac{1}{N} \left( \begin{array}{c}
+0 \\
+\sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N \pi_{j,t^\prime} X_{j,t^\prime} - \sum_{j=1}^N \tilde I_{j,t^\prime} \pi_{j,t^\prime} X_{j,t} \\
+0 \\
+0
+\end{array}
+\right),
+$$
+and
+$$
+A_{3,t} = \left( \begin{array}{c}
+0 \\
+0 \\
+\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{(Z_j-FP)^2}{FP(1-FP)} \\
+0
+\end{array}
+\right),
+$$
+and
+$$
+A_{4,t} = \left( \begin{array}{c}
+0 \\
+0 \\
+0 \\
+\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{(\tilde Z_j - FN)^2}{FN(1-FN)}
+\end{array}
+\right),
+$$
+By independence of these components, the variance can be calculated by summation over variance of each term individually yielding. First, $V_1$
+{
+\tiny
+$$
+ N^{-2} \sum_{i=1}^N \left(
+\begin{array}{c c c c}
+\frac{(1-\pi_{j,t})}{\pi_{j,t}} (Y_{j,t} - (1-FP-FN) (\mu_{j,t} +FP))^2 & \sum_{t^\prime} K_{t,t^\prime} (1-\pi_{j,t^\prime}) \zeta_{t^\prime, j} X_{j,t^\prime}^\top & 0 & 0 \\
+\sum_{t^\prime} K_{t,t^\prime} (1-\pi_{j,t^\prime}) \zeta_{t^\prime, j} X_{j,t}&
+\sum_{t^\prime} K_{t,t^\prime} \pi_{j,t^\prime} (1-\pi_{j,t^\prime}) X_{j,t} X_{j,t}^\top
+& 0 & 0 \\
+0 & 0 & 0 & 0 \\
+0 & 0 & 0 & 0 \\
+\end{array}
+\right)
+$$
+}
+and $V_{j,t}$ are block-diagonal matrices with only one non-zero block that is equal to $D_{j,t}$ which is the design-based variance-covariance matrix under each probability sampling design.
+The asymptotic variance for the IPW estimator is then the first diagonal element of the matrix $E \left[ \phi_n (\eta_0) \right]^{-1} \times \left[ V_{1,t} + V_{2,t} + V_{3,t} + V_{4,t} \right] E \left[ \phi_n (\eta_0) \right]^{-1}$.
+
+\begin{align*}
+N^{-2} \bigg[ &(1-FP-FN)^{-2} \sum_{j=1}^N \zeta_{t,j}^2 \frac{(1-\pi_{j,t})}{\pi_{t,j}} + (1-FP-FN)^{-1} b_{1} \sum_{j=1}^N \sum_{t^\prime} K_{t,t^\prime} (1-\pi_{j,t^\prime}) \zeta_{t^\prime,j} X_{j,t} \\
+&+ (1-FP-FN)^{-1} \sum_{j=1}^N \sum_{t^\prime} K_{t,t^\prime} (1-\pi_{j,t^\prime} \zeta_{t^\prime, j} ) X_{j,t^\prime}^\top + b_1 \sum_{j=1}^N \sum_{t^\prime} K_{t,t^\prime} K_{t,t^\prime} \pi_{j,t^\prime} X_{j,t} X_{j,t}^\top \bigg]
+\end{align*}
+
+\section{Model-based estimation: parameter choices and inference}
+\label{app:modelbased}
+
 \end{document}

### 2020-11-09 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 \usepackage{multirow}
+\usepackage{enumitem}
 
@@
 Coronavirus case-count data refers to the number of positive molecular tests performed on each day.
-Figure~\ref{fig:ny-covid} plots (dotted orange line) the number of reported confirmed COVID-19 cases per day in the state of Indiana.
-Total number of COVID-19 diagnostic tests performed per day is also reported (Figure~\ref{fig:ny-covid}, dotted blue line).
+Figure~XX plots (dotted orange line) the number of reported confirmed COVID-19 cases per day in the state of Indiana.
+Total number of COVID-19 diagnostic tests performed per day is also reported (Figure~XX, dotted blue line).
 On August 5th, 2020, for example, there were 1075 total daily positive cases and 11,663 total daily tests.
@@
 Third, the impact of measurement-error on data quality is changed when considering a weighted estimand.
-{\red state that it may switch sign!als}
+In particular, weighting may result in $\text{sgn}(\Delta) \neq \text{sgn} (\tilde \Delta)$ which implies the impact of measurement-error may be in a different direction.
 
@@
 
-\begin{lemma}[IPW Variance] \normalfont
+\begin{lemma}[Variance of IPW estimator] \normalfont
 \label{lemma:ipw}
+\textcolor{red}{Must check.}
 The estimates~$\mu_t := \bar y_{n,t}^\star$ and $\hat \pi_{j,t}$
@@
 
-\begin{lemma}[DR Variance] \normalfont
+\begin{lemma}[Variance of doubly-robust estimator] \normalfont
 \label{lemma:dr}
+\textcolor{red}{Must fix.}
 The estimates~$\mu_t := \bar y_{n,t}^\star$ and $\hat \pi_{j,t}$
@@
 
-\section{Applications to binary outcome and the COVID-19 pandemic.}
+\section{COVID-19 active infection prevalence in Indiana}
 \label{section:applications}
-
-What are the propensity measures of importance:
-\begin{itemize}
-	\item Symptom
-	\item Demographics
-	\item Prior testing (have you been tested before)
-	\item Prevalence (overall worry)
-\end{itemize}
-There is no good data on individual prior testing so we will ignore.
-
-{\bf Combines death, case count with covariate info, symptoms from Fb,  }
-
-\subsection{COVID-19 active infection prevalence in Indiana}
 
@@
 
-Here we consider estimation of the active infection rate in Indiana.
-Table~\ref{tab:indiana} summarizes data from a random sample from April 25--29th as well as a nonrandom sample obtained between May 2--3 in racial/ethnic minority communities.
-The nonrandom sample is therefore not a random subsample of the overall case-counts; however, this data provides important supplementary covariate information.
+Here we consider estimation of the active infection rate in Indiana. We consider five sources of data:
+\begin{itemize}[leftmargin=*]
+\item {\bf Testing data}:  Each day, the number of tests performed and number of positive tests are reported, with counts broken out jointly by age, gender, and racial demographic information.
+Figure~\ref{fig:in-cases} plots the number of COVID-19 positive cases based on testing date by age.
+
+\begin{figure}
+\centering
+\begin{subfigure}{.45\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{../methods/figs/indianacasecounts.png}
+  \caption{Indiana case-counts per day}
+  \label{fig:in-cases}
+\end{subfigure}
+\begin{subfigure}{.45\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{../methods/figs/indianadeaths.png}
+  \caption{Indiana death-counts per day}
+  \label{fig:in-deaths}
+\end{subfigure}
+\end{figure}
+
+\item {\bf Random/Nonrandom statewide sample}: Table~\ref{tab:indiana} summarizes data from a random sample from April 25--29th as well as a nonrandom sample obtained between May 2--3 in racial/ethnic minority communities (cite).
+The nonrandom sample is not a random subsample of the overall case-counts; however, this data provides important supplementary covariate information.
 Namely, the samples collected included covariate information on (1) prior symptoms, (2) COVID-19 positive individual in household, and (3) prior COVID-19 positive test.
@@
 & Male & 41.8 & 45.0 & 49.3 & 24.2 (12.4) & 2.1 \\ \hline
-\multirow{3}{*}{Age} & $<40$ & 39.4 & 28.0 & 52.7 & 29.7 & 1.7 \\
-& $40-59$ & 41.1 & 36.0 & 25.2 & 24.9 & 2.1 \\
-& $\geq 60$ & 19.5 & 36.0 & 22 & 6.7 & 0.9 \\ \hline
-\multirow{2}{*}{Race} & White & 23.1 & 92.0 & 86.9 & 19.5 & 1.5 \\
-& Nonwhite & 76.9 & 8.0 & 13.1 & 25.0 & 3.4 \\ \hline
-Hispanic & Hispanic & 44.1 & 2.0 & 7.1 & 37.6 & 6.9 \\
-Origin & Non-Hispanic & 55.9 & 98.0 & 92.9 & 13.0 & 1.3 \\ \hline
-\multirow{2}{*}{Fever} & Yes & 17.0 & 1.8 & - & 66.4 & 4.5 \\
-& No & 83.0 & 98.2 & - & 15.6 & 1.3 \\ \hline
-Household & Yes & 10.8 & 1.4 & - & 46.1 & 24.4 \\
+\multirow{3}{*}{Age} & $<40$ & 39.4 & 28.0 & 52.7 & 29.7 (15.0) & 1.7 \\
+& $40-59$ & 41.1 & 36.0 & 25.2 & 24.9 (12.5) & 2.1 \\
+& $\geq 60$ & 19.5 & 36.0 & 22 & 6.7 (3.4) & 0.9 \\ \hline
+\multirow{2}{*}{Race} & White & 23.1 & 92.0 & 86.9 & 19.5 (9.6) & 1.5 \\
+& Nonwhite & 76.9 & 8.0 & 13.1 & 25.0 (12.3) & 3.4 \\ \hline
+% Hispanic & Hispanic & 44.1 & 2.0 & 7.1 & 37.6 & 6.9 \\
+% Origin & Non-Hispanic & 55.9 & 98.0 & 92.9 & 13.0 & 1.3 \\ \hline
+\multirow{2}{*}{Fever} & Yes & 17.0 & 1.8 & - & 66.4 (32.1) & 4.5 \\
+& No & 83.0 & 98.2 & - & 15.6 (7.5) & 1.3 \\ \hline
+Household & Yes & 10.8 & 1.4 & - & 46.1 & 29.4 \\
 $+$ Case & No & 89.2 & 98.6 & - & 21.6 & 1.3 \\ \hline
-Prior $+$ & Yes & 6.1 & 1.6 & - & 39.2 & 24.4 \\
+Prior $+$ & Yes & 6.1 & 1.4 & - & 39.2 & 24.4 \\
 Test & No & 93.9 & 98.6 & - & 21.6 & 1.3 \\ \hline
@@
 
-Comparison to April 20th-25th using Fb data
-\begin{itemize}
-\item Fever: {\bf FB} = [0.017, 0.020, 0.024], {\bf Random}: 0.018
-\item Cough: {\bf FB} = [0.142, 0.148, 0.154], {\bf Random}: 0.149
-\item Shortness: {\bf FB} = [0.058, 0.060, 0.062], {\bf Random}: 0.062
+\item {\bf Facebook symptom survey}: We extract daily symptom surveys completed by individuals on Facebook who identify as living in Indiana.
+We collect age, sex, and demographic information as well as COVID-19 related symptoms, which Facebook defines as having (1) a fever and cough, (2) shortness of breath, \emph{or} (3) difficulty breathing.
+To assess potential bias, we compare the random sample from Indiana to the Facebook symptom survey data collected from April 20th to April 25th.
+Table~\ref{tab:comparison} demonstrates minimal bias in the Facebook symptom survey.
+\begin{table}
+\centering
+\begin{tabular}{c | c c c | c}
+& \multicolumn{3}{c}{Facebook} & Random \\ \cline{2-5}
+& Lower CI & Estimate & Upper CI & Estimate \\ \hline
+Fever &  0.017 & 0.020 & 0.024 & 0.018 \\
+Cough &  0.142 & 0.148 & 0.154 & 0.149 \\
+Shortness & 0.058 & 0.060 & 0.062 & 0.062 \\ \hline
+\end{tabular}
+\caption{Comparison of Facebook symptom survey and random sample demonstrates minimal bias.}
+\label{tab:comparison}
+\end{table}
+
+% \item {\bf Hospitalization data}: On each day, we observe the number of individuals who are COVID-19 positive that visited a hospital on that day.
+Unfortunately, hospitalization data is not broken out by age, gender, and racial demographic information.
+Instead, we rely on combining Facebook data to calculate the fraction of those who sought the hospital per week in each strata.
+We then use this to figure out how many individuals per strata were hospitalized and then back out how many
+
+\item {\bf Death data}: On each day, we observe the number of deaths due to COVID-19 reported on that day, with deaths broken out jointly by age, gender, and racial demographic information.
+Figure~\ref{fig:in-deaths} plots the cumulative number of COVID-19 related deaths over time by age group.
 \end{itemize}
 
-\subsubsection{Time-varying doubly-robust prevalence estimation}
+\subsection{Inverse-probability weighting approach}
+
+We start with using Table~\ref{tab:indiana} to compute IPW weights.
+The random sample is $n=3,658$.
+Here, we demonstrate how to compute the weight for a particular strata.
+For example, the strata defined as Male, 40--59, White, has a Fever, no household cases, and no prior tests has a weight proportional to
+$$
+\frac{\left( 0.493 \times 0.252 \times 0.869 \times 0.018 \right)}{\left(0.418 \times 0.411 \times 0.231 \times 0.17 \right)} = 0.628
+$$
+Using these weights and the adjusted strata-level prevalence estimates, the IPW estimate is $7.7$.
+
+\subsubsection{Time-varying IPW estimator}
+
+\subsection{Model-based estimation}
+
+\subsection{Doubly robust estimation}
 
@@
 \end{rmk}
+
+SEROPREVALENCE (i.e., used to be infected).  We can attempt to estimate using our methods and assume non-overlapping samples.  Then back out the correlation.
+
+We next consider estimation of data quality $E_{\I} \left[ \rho_{I,Y} \right]$.
+In \cite{Meng2018}, estimation relied on observing the true outcome (i.e., election vote totals), which  is not possible in the current crisis.
+Here, we propose using surveys as noisy but unbiased estimates of the true prevalence.
+Our goal is not inferential.
+Instead, we aim to build sensitivity analyses that can aid in understanding the amount of information in observational COVID case-count data.
+
+On April 23rd, Governor Andrew Cuomo announced results from a study in New York state.
+It found 13.9\% of 3,000 people tested across the state had signs of the virus.
+The study did not report sensitivity and specificity; therefore, we take reported measures from the Santa Clara study~\citep{Bendavid2020}.
+The reported specificity is $82.8\%$ (95\% CI: 76.0-88.4\%) and sensitivity is $99.5\%$ (95\% CI: 99.2-99.7\%).
+This corresponds to a false negative rate of $17.2\%$ (11.6-24.0\%) and false positive rate of $0.5\%$ (0.3 - 0.8\%).
+The estimated prevalence is then $16.3\%$ with a range of 15.0\% to 18.0\%.
+For simplicity, we assume the study was conducted on April 20th.
+On the 20th, $19,654$ tests were performed.
+The population of New York is $19,378,102$.
+Subtracting off the $617,555$ tests that had already performed yields a sampling fraction of $f = 0.001$.
+The smoothed prevalence estimate is 28.4\%.
+Adjusting for imperfect testing yields an estimate of $33.9\%$.
+
+Under the assumption that the survey is a SRS from New York, the error is $17.6$\%.  Using~\eqref{eq:error}, we construct an estimate of the relative sampling rate as follows
+\[
+\rho D_M = \sqrt{\frac{f}{1-f}} \frac{\text{0.176}}{\sigma_Y} = 1.07 \times 10^{-2} \Rightarrow \Delta = 1.07 \times 10^{-3} \Rightarrow M = 2.3.
+\]
+We can perform a sensitivity analysis by considering the range of false negative and positive rates from~\cite{Bendavid2020}, which leads to a range for $M$ of $(2.21, 2.38)$.
+Therefore, the impact on bias of $\hat R_t$ appears moderate.
+If $FP = 0.05$ and $FN = 0.005$, then the relative sampling rate increases to $M=3.04$.
+This simple tool helps us understand how sensitivity and specificity impact biases and  therefore how much we should trust conclusions based on these assumptions.
+
+% \begin{figure}
+% \centering
+% \begin{subfigure}{.5\textwidth}
+%   \centering
+%   \includegraphics[width=.9\linewidth]{../methods/figs/NY_casecount.png}
+%   \caption{NY new cases per day}
+%   \label{fig:ny-covid-test}
+% \end{subfigure}%
+% \begin{subfigure}{.5\textwidth}
+%   \centering
+%   \includegraphics[width=.9\linewidth]{../methods/figs/NY_fracpos.png}
+%   \caption{NY fraction of new tests positive per day}
+%   \label{fig:ny-covid-frac}
+% \end{subfigure}
+% \caption{New York COVID-19 reported data}
+% \label{fig:ny-covid}
+% \end{figure}
 
@@
 
-\subsection{Estimation of Data Quality}
-\label{section:est_dq}
-
-SEROPREVALENCE (i.e., used to be infected).  We can attempt to estimate using our methods and assume non-overlapping samples.  Then back out the correlation.
-
-We next consider estimation of data quality $E_{\I} \left[ \rho_{I,Y} \right]$.
-In \cite{Meng2018}, estimation relied on observing the true outcome (i.e., election vote totals), which  is not possible in the current crisis.
-Here, we propose using surveys as noisy but unbiased estimates of the true prevalence.
-Our goal is not inferential.
-Instead, we aim to build sensitivity analyses that can aid in understanding the amount of information in observational COVID case-count data.
-
-On April 23rd, Governor Andrew Cuomo announced results from a study in New York state.
-It found 13.9\% of 3,000 people tested across the state had signs of the virus.
-The study did not report sensitivity and specificity; therefore, we take reported measures from the Santa Clara study~\citep{Bendavid2020}.
-The reported specificity is $82.8\%$ (95\% CI: 76.0-88.4\%) and sensitivity is $99.5\%$ (95\% CI: 99.2-99.7\%).
-This corresponds to a false negative rate of $17.2\%$ (11.6-24.0\%) and false positive rate of $0.5\%$ (0.3 - 0.8\%).
-The estimated prevalence is then $16.3\%$ with a range of 15.0\% to 18.0\%.
-For simplicity, we assume the study was conducted on April 20th.
-On the 20th, $19,654$ tests were performed.
-The population of New York is $19,378,102$.
-Subtracting off the $617,555$ tests that had already performed yields a sampling fraction of $f = 0.001$.
-The smoothed prevalence estimate is 28.4\%.
-Adjusting for imperfect testing yields an estimate of $33.9\%$.
-
-Under the assumption that the survey is a SRS from New York, the error is $17.6$\%.  Using~\eqref{eq:error}, we construct an estimate of the relative sampling rate as follows
-\[
-\rho D_M = \sqrt{\frac{f}{1-f}} \frac{\text{0.176}}{\sigma_Y} = 1.07 \times 10^{-2} \Rightarrow \Delta = 1.07 \times 10^{-3} \Rightarrow M = 2.3.
-\]
-We can perform a sensitivity analysis by considering the range of false negative and positive rates from~\cite{Bendavid2020}, which leads to a range for $M$ of $(2.21, 2.38)$.
-Therefore, the impact on bias of $\hat R_t$ appears moderate.
-If $FP = 0.05$ and $FN = 0.005$, then the relative sampling rate increases to $M=3.04$.
-This simple tool helps us understand how sensitivity and specificity impact biases and  therefore how much we should trust conclusions based on these assumptions.
-
-\begin{figure}
-\centering
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{../methods/figs/NY_casecount.png}
-  \caption{NY new cases per day}
-  \label{fig:ny-covid-test}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{../methods/figs/NY_fracpos.png}
-  \caption{NY fraction of new tests positive per day}
-  \label{fig:ny-covid-frac}
-\end{subfigure}
-\caption{New York COVID-19 reported data}
-\label{fig:ny-covid}
-\end{figure}
-
 \section{Discussion}

### 2021-02-14 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 \usepackage{setspace}
-% \onehalfspacing
-\doublespacing
+\onehalfspacing
+% \doublespacing
 
@@
 We show that estimates based on the millions of molecular tests in the US has the same mean square error as a small simple random sample.
-To address this, a procedure is presented that combines case-count data with random samples over time to estimate selection propensities based on key covariate information.
+To address this, a procedure is presented that combines case-count data and random samples over time to estimate selection propensities based on key covariate information.
 We then combine these selection propensities with epidemiological forecast models using death-only data to construct a \emph{doubly-robust} estimation method that accounts for both measurement-error and selection bias.
@@
 The World Health Organization has declared the coronavirus disease 2019 (COVID-19) a public health emergency.
-As of August 28th, 2020, over 24.5 million cases have been confirmed worldwide with 5.95 million cases and over 180 thousand confirmed deaths across the United States.
-This pandemic
+\textcolor{red}{As of August 28th, 2020, over 24.5 million cases have been confirmed worldwide with 5.95 million cases and over 180 thousand confirmed deaths across the United States.} This pandemic
 has become the focal point of everyday life; yet the data landscape for understanding COVID-19 remains limited.
@@
 The reason is straightforward: observed case-count data is plagued by selection bias and measurement error.
-Through simple calculations, we will demonstrate that the information gained from increasing testing capacity is limited in the presence of selection bias and when testing inaccuracies persist.
+Through a series of calculations, we will demonstrate that the information gained from increasing testing capacity is limited in the presence of selection bias and when testing inaccuracies persist.
 In particular, the millions of tests in the US have a small effective sample size when compared to random sampling.
@@
 Due to monetary and time constraints, random testing may be performed infrequently.
-As of August 28, 2020, Indiana is the only state to conduct a statewide random sample testing\footnote{This is the only random sample to collect both seroprevalence and diagnostic testing results.
+\textcolor{red}{As of August 28, 2020}, Indiana is the only state to conduct a statewide random sample testing\footnote{This is the only random sample to collect both seroprevalence and diagnostic testing results.
 The CDC has conducted several seroprevalence studies.
@@
 A doubly robust extension allows researchers to combine these estimates with epidemiological forecasts based on compartmental models that are common in the study of infectious diseases~\cite{Hao2020,Song2020,Ray2020,Johndrow2020}.
+% Hospitalization Gelman
 While we demonstrate improvements over simple estimates of disease prevalence, we highlight how selection bias may persist and impact uncertainty quantification.
@@
 After an individual is infected with SARS-CoV-2, there is an incubation period which lasts between two to five days (cite: NPR) before the viral load is high enough to be detected.
-A \emph{molecular test} refers to diagnostic tests that aim to detect increased viral load (e.g., RT-PCR or antigen tests).
+A \emph{molecular test} refers to any diagnostic test that aims to detect increased viral load (e.g., RT-PCR or antigen tests).
 If someone has an active infection then after the incubation period a molecular test with perfect sensitivity will yield a positive result for the next few weeks (cite: wlos).
@@
 
-The primary goal is of molecular tests is in diagnosis of active infections in the population.
+The primary goal of molecular tests is diagnosis of active infections in the population.
 Such diagnostic tests generate important information about the presence of SARS-Cov-2 in the population, and help scientists and policy-makers understand patterns of transmission and propagation.
@@
 % where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.
+
+\subsubsection{Uncertain sensitivity and specificity}
+
+For prevalence there is some
+
+For
+
+For seroprevalence, \cite{Bendavid2020} report
+
+, which leads to a range for $M$ of $(2.21, 2.38)$.
+Therefore, the impact on bias of $\hat R_t$ appears moderate.
+If $FP = 0.05$ and $FN = 0.005$, then the relative sampling rate increases to $M=3.04$.
+This simple tool helps us understand how sensitivity and specificity impact biases and  therefore how much we should trust conclusions based on these assumptions.
 
@@
 $$
-Using these weights and the adjusted strata-level prevalence estimates, the IPW estimate is $7.7$.
-
-\subsubsection{Time-varying IPW estimator}
-
-\subsection{Model-based estimation}
-
-\subsection{Doubly robust estimation}
-
-\begin{rmk}[Use of symptom-to-hospitalization ratio to estimate symptom subgroup size]
-
-\end{rmk}
-
-SEROPREVALENCE (i.e., used to be infected).  We can attempt to estimate using our methods and assume non-overlapping samples.  Then back out the correlation.
-
-We next consider estimation of data quality $E_{\I} \left[ \rho_{I,Y} \right]$.
-In \cite{Meng2018}, estimation relied on observing the true outcome (i.e., election vote totals), which  is not possible in the current crisis.
-Here, we propose using surveys as noisy but unbiased estimates of the true prevalence.
-Our goal is not inferential.
-Instead, we aim to build sensitivity analyses that can aid in understanding the amount of information in observational COVID case-count data.
-
-On April 23rd, Governor Andrew Cuomo announced results from a study in New York state.
-It found 13.9\% of 3,000 people tested across the state had signs of the virus.
-The study did not report sensitivity and specificity; therefore, we take reported measures from the Santa Clara study~\citep{Bendavid2020}.
-The reported specificity is $82.8\%$ (95\% CI: 76.0-88.4\%) and sensitivity is $99.5\%$ (95\% CI: 99.2-99.7\%).
-This corresponds to a false negative rate of $17.2\%$ (11.6-24.0\%) and false positive rate of $0.5\%$ (0.3 - 0.8\%).
-The estimated prevalence is then $16.3\%$ with a range of 15.0\% to 18.0\%.
-For simplicity, we assume the study was conducted on April 20th.
-On the 20th, $19,654$ tests were performed.
-The population of New York is $19,378,102$.
+Using the constructed weights and the adjusted strata-level prevalence estimates, the IPW estimate under no measurement-error is $7.7$, a decrease of four percentage points from the observed prevalence of $11.7$.
+Using Table~\ref{tab:indiana} to compute weights requires the assumption of independence among factors.
+To alleviate this, we compute weights using Facebook's survey data collected in Indiana over that window of time; the resulting IPW estimate under no measurement-error is
+
+The study did not report sensitivity and specificity; therefore, we take the suggested measures from~\cite{Woloshin2020} which report 70\% sensitivity is a reasonable estimates and~\cite{Cohen2020} which report 97.6\% specificity.
+This corresponds to a false negative rate of $30\%$ and false positive rate of $2.4\%$.
+Therefore, the IPW estimate accounting for measurement-error is $7.7\%$.
+To account for uncertainty in
+
+Indiana's population as of 2019 was $6.732$ million.
 Subtracting off the $617,555$ tests that had already performed yields a sampling fraction of $f = 0.001$.
@@
 Adjusting for imperfect testing yields an estimate of $33.9\%$.
+
+For this estimate, we consider estimation of data quality $E_{\I} \left[ \rho_{\tilde I(X),Y} \right]$.
+In \cite{Meng2018}, estimation relied on observing the true outcome (i.e., election vote totals), which is not possible in the current crisis.
+Here, we use the random sample as a noisy but unbiased estimate of the true prevalence.
+Our goal is not inferential.
+Instead, we aim to build sensitivity analyses that can aid in understanding the amount of information in observational COVID case-count data.
 
@@
 This simple tool helps us understand how sensitivity and specificity impact biases and  therefore how much we should trust conclusions based on these assumptions.
+
+\subsubsection{Time-varying IPW estimator}
+
+\subsection{Model-based estimation}
+
+\subsection{Doubly robust estimation}
 
@@
 
+\section{Reproducible code}
+
 All relevant code can be found at \url{https://github.com/wdempsey/covid-umich}.
@@
 
+Following~\cite{Song2020}, the fourth-order Runge-Kutta (RK4) method gives an approximation to the proposed SEIR model in equation XX as follows:
+
+$$
+f(\theta_{t-1}, \beta, \gamma) = \left(
+\begin{array}{c}
+\theta_{t-1}^{S} + \frac{1}{6} \left[ k_{t-1}^{S_1} + 2 k_{t-1}^{S_2} + 2 k_{t-1}^{S_3} + k_{t-1}^{S_4} \right] \\
+\theta_{t-1}^{E} + \frac{1}{6} \left[ k_{t-1}^{E_1} + 2 k_{t-1}^{E_2} + 2 k_{t-1}^{E_3} + k_{t-1}^{E_4} \right] \\
+\theta_{t-1}^{I^{(T)}} + \frac{1}{6} \left[ k_{t-1}^{I_1^{(T)}} + 2 k_{t-1}^{I_2^{(T)}} + 2 k_{t-1}^{I_3^{(T)}} + k_{t-1}^{I_4^{(T)}} \right] \\
+\theta_{t-1}^{I^{(NT)}} + \frac{1}{6} \left[ k_{t-1}^{I_1^{(NT)}} + 2 k_{t-1}^{I_2^{(NT)}} + 2 k_{t-1}^{I_3^{(NT)}} + k_{t-1}^{I_4^{(NT)}} \right] \\
+\theta_{t-1}^{R} + \frac{1}{6} \left[ k_{t-1}^{R_1} + 2 k_{t-1}^{R_2} + 2 k_{t-1}^{R_3} + k_{t-1}^{R_4} \right] \\
+\end{array}
+\right)
+$$
+where
+\begin{align*}
+k_t^{S_1} &= - s_t \circ (B i_t^{(T)}) \\
+k_t^{S_2} &= - (s_t + 0.5 k_t^{S_1}) \circ (B (i_t^{(T)} + 0.5 k_t^{I_1}) \\
+k_t^{S_3} &= - (s_t + 0.5 k_t^{S_2}) \circ (B i_t^{(T)} + 0.5 k_t^{I_2}) \\
+k_t^{S_4} &= - (s_t + k_t^{S_3}) \circ (B i_t^{(T)} + k_t^{I_3}),
+\end{align*}
+and
+\begin{align*}
+k_t^{E_1} &= s_t \circ (B i_t^{(T)}) - \sigma e_t \\
+k_t^{E_2} &= (s_t + 0.5 k_t^{S_1}) \circ (B (i_t^{(T)} + 0.5 k_t^{I_1}) - \sigma (e_t + 0.5 k_t^{e_1}) \\
+k_t^{E_3} &= (s_t + 0.5 k_t^{S_2}) \circ (B (i_t^{(T)} + 0.5 k_t^{I_2}) - \sigma (e_t + 0.5 k_t^{e_2}) \\
+k_t^{E_4} &= (s_t + k_t^{S_3}) \circ (B (i_t^{(T)} + k_t^{I_4}) - \sigma (e_t + k_t^{E_3}) \\
+\end{align*}
+and
+\begin{align*}
+k_t^{I_1^{(T)}} &= \sigma e_t - \beta_2 i^{(T)} \\
+k_t^{I_2^{(T)}} &= \sigma (e_t + 0.5 k_t^{E_1}) - \beta_2 (i^{(T)} + 0.5 k_t^{I_1^{(T)}}) \\
+k_t^{I_3^{(T)}} &= \sigma (e_t + 0.5 k_t^{E_2}) - \beta_2 (i^{(T)} + 0.5 k_t^{I_2^{(T)}}) \\
+k_t^{I_4^{(T)}} &= \sigma (e_t + k_t^{E_3}) - \beta_2 (i^{(T)} + k_t^{I_3^{(T)}}),
+\end{align*}
+and
+\begin{align*}
+k_t^{I_1^{(NT)}} &= \beta_2 i^{(T)} - \gamma i^{(NT)} \\
+k_t^{I_2^{(NT)}} &= \beta_2 (i^{(T)} + 0.5 k_t^{I_1^{(T)}}) - \gamma (i^{(NT)} + 0.5 k_t^{I_1^{(NT)}})\\
+k_t^{I_3^{(NT)}} &= \beta_2 (i^{(T)} + 0.5 k_t^{I_2^{(T)}}) - \gamma (i^{(NT)} + 0.5 k_t^{I_2^{(NT)}})\\
+k_t^{I_4^{(NT)}} &= \beta_2 (i^{(T)} + k_t^{I_3^{(T)}}) - \gamma (i^{(NT)} + k_t^{I_3^{(NT)}}) ,
+\end{align*}
+and
+\begin{align*}
+k_t^{R_1} &= - \gamma i_t^{(NT)} \\
+k_t^{R_2} &= - \gamma (i^{(NT)} + 0.5 k_t^{I_1^{(NT)}}) \\
+k_t^{R_3} &= - \gamma (i^{(NT)} + 0.5 k_t^{I_2^{(NT)}}) \\
+k_t^{R_4} &= - \gamma (i^{(NT)} + k_t^{I_3^{(NT)}}) \\
+\end{align*}
+
 \end{document}

### 2021-02-22 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 
-\subsection{Related work.}
+\subsection{Related work}
 
@@
 
-In this section we provide the necessary background to understand COVID-19 diagnostic and serologic testing, their respective scientific uses in managing the pandemic, and the corresponding data streams.
+In this section we provide the necessary background to understand COVID-19 diagnostic and serologic testing, their respective scientific uses in managing the pandemic, and the data streams we consider in this manuscript.
 

### 2021-05-14 write-up/cover letter-covid.tex
Changed to covid given symp rather than reverse due to math
This lets marginal models cancel in the back calc of number of symptomatic people with covid negative test results

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Plain Cover Letter
% LaTeX Template
%
% This template has been downloaded from:
% http://www.latextemplates.com
%
% Original author:
% Rensselaer Polytechnic Institute (http://www.rpi.edu/dept/arc/training/latex/resumes/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[10pt]{letter} % Default font size of the document, change to 10pt to fit more text

\usepackage{newcent} % Default font is the New Century Schoolbook PostScript font
%\usepackage{helvet} % Uncomment this (while commenting the above line) to use the Helvetica font

% Margins
\topmargin=-0.5in % Moves the top of the document 1 inch above the default
\textheight=8.75in % Total height of the text on the page before text goes on to the next page, this can be increased in a longer letter
\oddsidemargin=0in % Position of the left margin, can be negative or positive if you want more or less room
\textwidth=6.5in % Total width of the text, increase this if the left margin was decreased and vice-versa

\let\raggedleft\raggedright % Pushes the date (at the top) to the left, comment this line to have the date on the right

%\address{Rutgers University and University of Michigan}

\begin{document}

%----------------------------------------------------------------------------------------
%	ADDRESSEE SECTION
%----------------------------------------------------------------------------------------

\begin{letter}{Professor Karen Kafadar \\
    Editor-In-Chief, {\em Annals of Applied Statistics}
         }
%----------------------------------------------------------------------------------------
%	YOUR NAME & ADDRESS SECTION
%----------------------------------------------------------------------------------------

%\begin{center}
%\large
%\end{center}
%\vfill

\signature{Walter Dempsey\\
University of Michigan \\
Department of Biostatistics\\
M4057 SPH II \\
1415 Washington Heights \\
wdem@umich.edu} % Your name for the signature at the bottom

%----------------------------------------------------------------------------------------
%	LETTER CONTENT SECTION
%----------------------------------------------------------------------------------------

\date\today

\opening{Dear Dr. Kafadar:}

Kindly find enclosed the Report titled ``Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic,'' by Walter Dempsey.

The report contains novel empirical findings and an investigation of statistical paradoxes arising from the analysis of reported coronavirus case-counts that are being used by governments to guide policies and decision making. We have discovered an estimator and statistical decomposition of the associated error that shows the 21.5 million tests conducted in the US provides information on COVID-19 prevalence equivalent to a simple random sample of size 92. Analogous decompositions for the rate of change in prevalence and the effective reproduction number are used to demonstrate how the magnitude and sign of estimation bias varies over time, a bias which has strong policy implications. Overall, our findings demonstrate the statistical limitations of increased testing in the presence of selection bias and measurement error as well as how easy it is for researchers analysing reported coronavirus case-counts to not only fool themselves but be confident in their incorrect conclusions. We believe the work presents an important advance in the statistical understanding of the COVID-19 pandemic.

Thank you for your consideration. We look forward to hearing back from you.

\closing{Sincerely,}

\end{letter}

\end{document}

### 2021-05-14 write-up/covid-writeup.tex
Changed to covid given symp rather than reverse due to math
This lets marginal models cancel in the back calc of number of symptomatic people with covid negative test results

@@
 \usepackage{setspace}
-\onehalfspacing
-% \doublespacing
+% \onehalfspacing
+\doublespacing
 
@@
 
+\section{IRLS}
+
+First derivative with respect to $\theta$
+$$
+\nabla E(\theta) = \sum_{t^\prime=1}^T K_h(|t^\prime - t|) \left[ \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} + \sum_{j=1}^N \tilde W_{j,t^\prime} \tilde I_{j,t^\prime} \pi_t (X_{j,t^\prime}; \theta) X_{j,t^\prime} \right]
+$$
+Second derivative with respect to $\theta$
+$$
+H(\theta) = \sum_{t^\prime=1}^T K_h(|t^\prime - t|) \sum_{j=1}^N \tilde W_{j,t^\prime} \tilde I_{j,t^\prime} \pi_t (X_{j,t^\prime}; \theta) \left( 1- \pi_t (X_{j,t^\prime}; \theta) \right) X_{j,t^\prime} X_{j,t^\prime}^\top
+$$
+Then the Fisher scoring method sets
+$$
+\theta^{(t+1)} = \theta^{(t)} + H \left(\theta^{(t)} \right)^{-1} \nabla E \left(\theta^{(t)} \right)
+$$
+
 \end{document}

### 2021-05-14 preprocessing/5_hospfb_merge.R

@@ -13,8 +13,8 @@ weekweight = aggregate(day ~ week, fb_data, function(x){length(unique(x))})
 
 ## Build weekly version of FB data first
 weeksin2020 = 14:53
-all_data_neg_hospital = all_data_neg_symptom = 
-  all_data_pos_hospital = all_data_pos_symptom = 
+all_data_neg_hospital = all_data_pos_hospital = 
+  all_data_covid = 
   data.frame(gender = rep(0,0), age = rep(0,0), 
              weight = rep(0,0), week = rep(0,0),
              year = rep(0,0))
@@ -43,19 +43,6 @@ for (week in weeksin2020) {
   agg_temp$year = rep(2020, nrow(agg_temp))
   all_data_neg_hospital = rbind(all_data_neg_hospital, agg_temp)
   
-  ## Adding symptoms given contact
-  temp$weightsymptoms =  temp$weight * temp$fever
-  total_weight = sum(temp$weight[temp$gottested ==1 & temp$postest == 0], na.rm = TRUE)
-  total_weightsymptoms = sum(temp$weightsymptoms[temp$gottested ==1 & temp$postest == 0], na.rm = TRUE)
-  agg_temp = aggregate(weightsymptoms ~ gender+ age, subset(temp, gottested == 1 & postest == 0), sum)
-  agg_temp2 = aggregate(weight ~ gender+ age, subset(temp, gottested == 1 & postest == 0), sum)
-  agg_temp$weight = agg_temp2$weight
-  agg_temp$weightsymptoms = agg_temp$weightsymptoms/sum(agg_temp$weightsymptoms) * total_weightsymptoms * addweight
-  agg_temp$weight = agg_temp$weight/sum(agg_temp$weight) * total_weight * addweight
-  agg_temp$week = rep(week, nrow(agg_temp))
-  agg_temp$year = rep(2020, nrow(agg_temp))
-  all_data_neg_symptom = rbind(all_data_neg_symptom, agg_temp)
-  
   ## Positive 
   total_weight = sum(temp$weight[temp$postest ==1], na.rm = TRUE)
   total_weighthospital = sum(temp$weighthospital[temp$postest == 1], na.rm = TRUE)
@@ -68,18 +55,19 @@ for (week in weeksin2020) {
   agg_temp$year = rep(2020, nrow(agg_temp))
   all_data_pos_hospital = rbind(all_data_pos_hospital, agg_temp)
   
[…]

### 2021-05-14 visualizations/tv_propensities.R

@@ -9,7 +9,7 @@ propensities$date = MMWRweek::MMWRweek2Date(MMWRyear = propensities$year,
                                      MMWRday = 1)
 
 ## Example propensity: Fever + Contact + Not H + White + Male + 35to44
-x = as.matrix(c(0,1,1,0,1,0,1,0,0,1,0,0,0,0), ncol = 1)
+x = as.matrix(c(0,1,1,0,1,0,0,1,0,1,0,0,0,0), ncol = 1)
 propensities$probs = as.vector(1/(1+exp(-as.matrix(propensities[,3:16])%*%x)))
 
 ggplot(data = propensities, aes(x = date, y = probs)) +

### 2021-05-18 methods/4_covid_propensity_estimation_alt.R
Doing hospitalization to correct symptom only for 2nd pass

@@ -24,7 +24,7 @@ fb_distance <-  function(current_week, current_year) {
 kernelweight <- function(distance) {
   kw2 <- function(row) {
     dis = distance(row)
-    h = sqrt(-1/(2*log(0.9))) # Set a deterministic bandwidth for now
+    h = sqrt(-1/(2*log(0.75))) # Set a deterministic bandwidth for now
     kernel = exp(-dis^2/(2*h^2))
     return(kernel)
   }

### 2021-05-18 methods/4_symptom_propensity_estimation_alt.R

@@ -40,7 +40,7 @@ irls <- function(current_week, current_year, fb_X, fb_data,
   
   fb_dist = fb_distance(current_week, current_year)
   weights = apply(fb_data, 1, kernelweight(fb_dist))
-  weighted_symptoms = weights*fb_data$weighthospital
+  weighted_symptoms = weights*fb_data$weightsymptom
   first_term = t(fb_X)%*%weighted_symptoms
 
   fbweighted = weights*fb_data$weight
@@ -71,15 +71,14 @@ irls <- function(current_week, current_year, fb_X, fb_data,
 
 
 ## Brining in complete FB and Indiana data
-fb_data_pos_hospital = readRDS("../data/fb_weeklycomplete_pos_hospital_alt.RDS")
-names(fb_data_pos_hospital)[3] = "symptoms"
+fb_data_pos_symptom = readRDS("../data/fb_weeklycomplete_pos_symptom_alt.RDS")
 
 ## Build \pi(x; \theta)
 ## Logistic regression with no interactions
 
 ## First term can be precomputed given design matrix
-model = ~ -1+as.factor(gender)+as.factor(age) + symptoms
-fb_X = model.matrix(model, fb_data_pos_hospital)
+model = ~ -1+as.factor(gender)+as.factor(age) + hospital
+fb_X = model.matrix(model, fb_data_pos_symptom)
 
 weeks = c(14:53,1:5)
 years = c(rep(2020, length = length(c(14:53))),rep(2021, length = length(1:5)))
@@ -89,13 +88,13 @@ for(i in 1:length(weeks)) {
   print(paste("On week", weeks[i], "in year", years[i]))
   current_week = weeks[i]
   current_year = years[i]
-  current_estimated_theta = irls(current_week, current_year, fb_X, fb_data_pos_hospital)  
+  current_estimated_theta = irls(current_week, current_year, fb_X, fb_data_pos_symptom)  
   print(current_estimated_theta)
   results[i,] = c(current_week, current_year, as.vector(current_estimated_theta))
 }
 
 results = data.frame(results)
 names(results) = c("week", "year", "gender1", "gender2", "25to34", "35to44", "45to54",
-                   "55to64", "65to74", "75plus", "symptom")
+                   "55to64", "65to74", "75plus", "hospital")
 
-saveRDS(results, "../data/smoothedpropensities_pos_hospital_alt.RDS")
+saveRDS(results, "../data/smoothedpropensities_pos_symptom_alt.RDS")

### 2021-05-18 preprocessing/5_hospfb--preproc.R

@@ -85,10 +85,10 @@ for(all_row in 1:nrow(all_data_pos_symptom)) {
   for (row in 1:nrow(temp)) {
     if (temp$ethnicity[row] != "Hispanic or Latino") {
       temp$weight[row] = temp$weight[row] * ethnicity_census[ethnicity_levels == temp$ethnicity[row]] * race_census[race_levels == temp$race[row]]  
-      temp$weighthospital[row] = temp$weighthospital[row] * ethnicity_census[ethnicity_levels == temp$ethnicity[row]] * race_census[race_levels == temp$race[row]]
+      temp$weightsymptom[row] = temp$weightsymptom[row] * ethnicity_census[ethnicity_levels == temp$ethnicity[row]] * race_census[race_levels == temp$race[row]]
     } else {
       temp$weight[row] = temp$weight[row] * ethnicity_census[ethnicity_levels == temp$ethnicity[row]] * hispanic_match$census[hispanic_match$race == temp$race[row]]  
-      temp$weighthospital[row] = temp$weighthospital[row] * ethnicity_census[ethnicity_levels == temp$ethnicity[row]] * hispanic_match$census[hispanic_match$race == temp$race[row]]  
+      temp$weightsymptom[row] = temp$weightsymptom[row] * ethnicity_census[ethnicity_levels == temp$ethnicity[row]] * hispanic_match$census[hispanic_match$race == temp$race[row]]  
     }
     
   }

### 2021-05-18 preprocessing/5_hospfb_merge.R

@@ -10,6 +10,10 @@ fb_data = read.csv("../data/fb_test_data.csv")
 fb_data$day = day(fb_data$date)
 fb_data$week = week(fb_data$date) 
 weekweight = aggregate(day ~ week, fb_data, function(x){length(unique(x))})
+tested_fbdata = subset(fb_data, gottested == 1)
+rm(fb_data)
+
+tested_fbdata$hospital = is.element(tested_fbdata$hospital, c("TRUE","1"))
 
 ## Build weekly version of FB data first
 weeksin2020 = 14:53
@@ -18,10 +22,6 @@ all_data_pos_hospital = all_data_pos_symptom = all_data_covid =
              weight = rep(0,0), week = rep(0,0),
              year = rep(0,0))
 
-tested_fbdata = subset(fb_data, gottested == 1)
-rm(fb_data)
-
-tested_fbdata$hospital = is.element(tested_fbdata$hospital, c("TRUE","1"))
 
 for (week in weeksin2020) {
   print(paste("At week", week, "in year 2020"))
@@ -57,10 +57,10 @@ for (week in weeksin2020) {
   
   ## Adding covid positive given symptoms
   temp$weightcovid =  temp$weight * temp$postest
-  total_weight = sum(temp$weight[temp$symptoms == FALSE], na.rm = TRUE)
-  total_weightcovid = sum(temp$weightcovid[temp$symptoms == FALSE], na.rm = TRUE)
-  agg_temp = aggregate(weightcovid ~ gender+ age, subset(temp, symptoms == FALSE), sum)
-  agg_temp2 = aggregate(weight ~ gender+ age, subset(temp, symptoms == FALSE), sum)
+  total_weight = sum(temp$weight[temp$symptoms == FALSE & temp$state == 15], na.rm = TRUE)
+  total_weightcovid = sum(temp$weightcovid[temp$symptoms == FALSE & temp$state == 15], na.rm = TRUE)
+  agg_temp = aggregate(weightcovid ~ gender+ age, subset(temp, symptoms == FALSE & temp$state == 15), sum)
+  agg_temp2 = aggregate(weight ~ gender+ age, subset(temp, symptoms == FALSE & state == 15), sum)
   agg_temp$weight = agg_temp2$weight
   agg_temp$weightcovid = agg_temp$weightcovid/sum(agg_temp$weightcovid) * total_weightcovid * addweight
[…]

### 2021-05-18 preprocessing/5_in_hops_cleaning--preproc.R

@@ -2,10 +2,8 @@ weeklycoviddata_hospital = readRDS(file = "../data/weeklycoviddata_hospital.RDS"
 weeklycoviddata_testing = readRDS(file = "../data/weeklycoviddata.RDS")
 
 propensity_pos_hospital = readRDS("../data/smoothedpropensities_pos_hospital_alt.RDS")
+propensity_pos_symptom = readRDS("../data/smoothedpropensities_pos_symptom_alt.RDS")
 propensity_covid = readRDS("../data/smoothedpropensities_covid_alt.RDS")
-
-
-propensity_pos_hospital[1,]
 
 library(lubridate)
 weeklycoviddata_hospital$week = week(mdy(weeklycoviddata_hospital$startdate))
@@ -45,30 +43,27 @@ for(current_row in 1:nrow(weeklycoviddata_hospital)) {
     weeklycoviddata_hospital$pos_symptom_count[current_row] = NA
     weeklycoviddata_hospital$pos_nosymptom_count[current_row] = NA
   } else {
-    prop_temp_pos_hospital = propensity_pos_hospital[propensity_pos_hospital$week == temp$week & propensity_pos_hospital$year == temp$year,]
-    pos_hospital_term = sum(prop_temp_pos_hospital[3:11]*feature_vector)
-    prob_hosp_temp = 1/(1+exp(-pos_hospital_term))
-    
-    prop_temp_pos_covid = propensity_covid[propensity_covid$week == temp$week & propensity_covid$year == temp$year,]
-    pos_covid_term = sum(prop_temp_pos_covid[3:10]*feature_vector[1:8])
-    prob_pos_covid = 1/(1+exp(-pos_covid_term))
-    covid_odds_temp = prob_pos_covid/(1-prob_pos_covid)
-    
     covidtesting_match = which(weeklycoviddata_testing$week == temp$week & weeklycoviddata_testing$year == temp$year &
                                  weeklycoviddata_testing$age == temp$age & weeklycoviddata_testing$gender == temp$gender &
                                  weeklycoviddata_testing$ethnicity == temp$ethnicity & weeklycoviddata_testing$race == temp$race)
-    # covid_posrate_temp = sum(weeklycoviddata_testing$covid_counts[covidtesting_match])/sum(weeklycoviddata_testing$covid_tests[covidtesting_match])
-    # 
-    # covid_odds_temp = covid_posrate_temp/(1-covid_posrate_temp)
[…]

### 2021-05-18 preprocessing/5_in_hops_cleaning.R

@@ -56,7 +56,7 @@ construct_combos <- function(week, year) {
   addweight = 7/weekweight$day[weekweight$week == week]
 
   subset_hospital = indiana_hospital_data[indiana_hospital_data$week == week & indiana_hospital_data$year == year,]
-  weekly_hospitalization_count = sum(subset_hospital$Hospitalization.Count)
+  weekly_hospitalization_count = sum(subset_hospital$Hospitalization.Count, subset_hospital$ED.Visit.Count)
 
   allcombinations = expand.grid(startdate = min(subset_hospital$Date), age = unique(indiana_test_data$age), 
                                 gender = c("M", "F"), ethnicity = unique(indiana_test_data$ethnicity), 

### 2021-06-16 write-up/covid-writeup.tex
Doing hospitalization to correct symptom only for 2nd pass
Revising paper (story needs to emphasize "lessons learned")

@@
 \begin{abstract}
-  Coronavirus case-count data has influenced government policies and drives many epidemiological forecasts.
+  Coronavirus case-count data has influenced government policies and drives most epidemiological forecasts.
 Limited testing is cited as the key driver behind minimal information on the COVID-19 pandemic.
-While expanded testing is laudable, measurement error and selection bias are the two greatest problems limiting our understanding of the current COVID-19 pandemic; neither can be fully addressed by increased testing capacity.
+While expanded testing is laudable, measurement error and selection bias are the two greatest problems limiting our understanding of the COVID-19 pandemic; neither can be fully addressed by increased testing capacity.
 In this paper, we demonstrate their impact on estimation of point prevalence and the effective reproduction number.
@@
 The World Health Organization has declared the coronavirus disease 2019 (COVID-19) a public health emergency.
-\textcolor{red}{As of August 28th, 2020, over 24.5 million cases have been confirmed worldwide with 5.95 million cases and over 180 thousand confirmed deaths across the United States.} This pandemic
+\textcolor{red}{As of June 15th, 2021, over 176 million cases have been confirmed worldwide with 33.4 million cases and over 590 thousand confirmed deaths across the United States.} This pandemic
 has become the focal point of everyday life; yet the data landscape for understanding COVID-19 remains limited.
@@
 
-Due to monetary and time constraints, random testing may be performed infrequently.
-\textcolor{red}{As of August 28, 2020}, Indiana is the only state to conduct a statewide random sample testing\footnote{This is the only random sample to collect both seroprevalence and diagnostic testing results.
+Due to monetary and time constraints, however, random testing is either not performed or performed very infrequently.
+As of June 2021, Indiana is the only state to conduct statewide random sample testing\footnote{This is the only random sample to collect both seroprevalence and diagnostic testing results.
 The CDC has conducted several seroprevalence studies.
-See Section~\ref{section:testinginfo} for further discussion.}, which was collected from April 25--29, 2020.
-Infrequent random testing may be insufficient to understand the trajectory of infectious diseases which change rapidly over time.
+See Section~\ref{section:testinginfo} for further discussion.}, which was collected from April 25--29, 2020~\citep{Yiannoutsos2021}.
+Such infrequent random testing is likely to provide insufficient information to help researchers and policy makers better understand the disease trajectory which can change rapidly over time.
 Therefore, while random testing may be preferable in theory, in practice governments, researchers, and policy makers continue to use coronavirus case-counts to understand the impact of COVID-19 on the population and make data-informed decisions.
 
-The second objective is to demonstrate how random samples provide necessary auxiliary information to address selection bias in coronavirus case-count data.
+The second objective is to demonstrate how random samples provide necessary \emph{auxiliary information} to address selection bias in coronavirus case-count data.
 Random samples provide the necessary covariate information from a representative sample from the population to estimate selection propensities.
-These propensities are then used in an inverse-probability weighting scheme to construct estimators of disease prevalence that attempt to control for selection bias.
+These propensities can then used in an inverse-probability weighting scheme to construct estimators of disease prevalence that attempt to control for selection bias.
 A doubly robust extension allows researchers to combine these estimates with epidemiological forecasts based on compartmental models that are common in the study of infectious diseases~\cite{Hao2020,Song2020,Ray2020,Johndrow2020}.
+
+Note that the proposed approach requires covariate information to be collected on tested individuals.
+Unfortunately, many states do not require or report auxiliary covariate information beyond demographics such as gender, age, race, and ethnicity.
+To address this, we discuss how the proposal suggests minor changes to current practice that may improve performance over time.
+While we demonstrate empirical improvements over simple disease prevalence estimates, we also highlight how selection bias may persist and impact uncertainty quantification.
+
 % Hospitalization Gelman
-While we demonstrate improvements over simple estimates of disease prevalence, we highlight how selection bias may persist and impact uncertainty quantification.
 
@@
 
-In Section~\ref{section:applications}, the IPW and doubly-robust estimation methods are applied to the case study of Indiana statewide prevalence from March 1st until August 27th.
+In Section~\ref{section:applications}, the IPW and doubly-robust estimation methods are applied to the case study of Indiana statewide prevalence from March 2020 until June 2021.
 We estimate data quality for IPW estimates using the random sample on April 28th.
-We then discuss a sensitivity analysis method for estimating data quality and its use in understanding the potential biases in COVID-19 case-count analyses by analysing New York statewide prevalence and comparing to the results from their statewide random testing on April 23rd.
-Both demonstrate improvements over simple prevalence estimation but that data quality is not of the same order as random sampling.
-In Section~\ref{section:crosspop}, the case study results are used to suggest cross-country comparisons are difficult at best with population size, sampling fraction, and data quality all interacting to impact hypothesis testing.
+This demonstrates the proposed approach is an improvement over simple prevalence estimation but that data quality is not of the same order as random sampling.
 
@@
 
-After an individual is infected with SARS-CoV-2, there is an incubation period which lasts between two to five days (cite: NPR) before the viral load is high enough to be detected.
-A \emph{molecular test} refers to any diagnostic test that aims to detect increased viral load (e.g., RT-PCR or antigen tests).
-If someone has an active infection then after the incubation period a molecular test with perfect sensitivity will yield a positive result for the next few weeks (cite: wlos).
+Upon infection with SARS-CoV-2, there is an incubation period which lasts approximately five days~\citep{Lauer2020} before the viral load is high enough to be detected.
+A \emph{molecular test} refers to diagnostic tests that aim to detect increased viral load (e.g., RT-PCR or antigen tests).
+If someone has an active infection then after the incubation period, a molecular test with perfect sensitivity will yield a positive result for the next few weeks.
 After that, the viral load will decrease and a molecular test with perfect specificity will come back negative.
@@
 In the incubation phase, most molecular tests will return a false negative result.
-Molecular tests are most sensitive when the viral loads are highest which occurs during the first few days of transmissibility (cite: that guy ).
+Molecular tests are most sensitive when the viral loads are highest which occurs during the first few days of transmissibility~\citep{Mina2020}.
 Moreover, most molecular tests are performed via nasopharyngeal swab.
@@
 Such diagnostic tests generate important information about the presence of SARS-Cov-2 in the population, and help scientists and policy-makers understand patterns of transmission and propagation.
-Rapid and frequent molecular testing is cited as a key component (citEe: OECD and others) in effective strategies to identify active infections and prevent systemic outbreaks.
-There has been work on developing faster, cheaper testing options (cite: Mina) that can be used at-home as well as group testing (citE: Mina); however, these developments are beyond the scope of this paper.
+Rapid and frequent molecular testing is cited as a key component~\cite{OECD2021} in effective strategies to identify active infections and prevent systemic outbreaks.
 % Here we consider the current system of testing which involves both self-selection into testing and historical limits on who was able to receive diagnostic tests due to limited testing capacity.
@@
 These studies are useful for disease surveillance.
-When the number of active infections in a population are low, serological surveys are cheaper and more accurate than molecular tests and are more useful in detecting outbreaks (cite: Mina).
 To date, the CDC has conducted ten large-scale geographic serological surveys with three rounds.
@@
 \subsection{Publicly available data on COVID-19}
-\label{section:testinginfo}
-
-In this paper, our primary goal is accurate estimates of the population-level active infection rates over time using \emph{publicly available} testing data.
-Secondary goals include estimation of rates of change in these quantities and the effective reproduction number (cite:) which characterize the infectious disease spread.
-A final goal is using these estimates to characterize prevalence of immunity in the general population.
-As previously stated, these quantities are fundamental to public health policy and general critical information on presence and transmission of SARS-CoV-2 as well as the impact of public health  interventions.
+\label{subsection:testinginfo}
+
+Our primary goal is accurate estimates of the population-level active infection rates over time using \emph{publicly available} viral testing data.
+Secondary goals include estimation of rates of change and the effective reproduction number which characterize disease trajectory.
+A final goal is characterizing prevalence of population-level immunity.
+These quantities are fundamental to public health policy and provide critical information on the presence and transmission of SARS-CoV-2.
+ % as well as the impact of public health  interventions.
 
@@
 \centering
-\includegraphics[width = 0.9\textwidth]{../methods/figs/fbcovid19symptoms.png}
+\includegraphics[width = 0.9\textwidth]{../figs/fbcovid19symptoms.png}
 \caption{Reported rate of COVID-like symptoms in strata.
@@
 \centering
-\includegraphics[width = 0.4\textwidth]{../methods/figs/mem_heatmap_article.png}
+\includegraphics[width = 0.4\textwidth]{../figs/mem_heatmap_article.png}
 \caption{Imperfect testing adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.005$ and $FN=0.172$. Color scaled so blue = $-10$, white = $0$, and red = $10$.}
@@
   \centering
-  \includegraphics[width=.9\linewidth]{../methods/figs/sir_ratio.png}
+  \includegraphics[width=.9\linewidth]{../figs/sir_ratio.png}
   \caption{Ratio estimator}
@@
   \centering
-  \includegraphics[width=.9\linewidth]{../methods/figs/sir_rt.png}
+  \includegraphics[width=.9\linewidth]{../figs/sir_rt.png}
   \caption{Effective reproductive rate estimator}
@@
 \label{fig:rates}
+\end{figure}
+
+\subsection{Rate comparisons}
+
+So far we have focused on understanding the limitations of using case-count data to understand population quantities of interest for a \emph{single} population.
+Many are interested in cross-population comparisons to contrast the impact of countries' mitigation policies.
+Here, for simplicity, we focus on comparing the estimated effective reproductive rate.
+We assume the two time-series are aligned so that $t=0$ is the time of first case in each population respectively.
+% This negates alignment issues and is common in practice.
+
+% While the above issues on $Z$-scores and effective sample size are important, here we highlight a separate issue.
+Here we demonstrate the interaction of biases in estimation of the two trajectories when the peak infection times differ slightly.  First, we can write the difference in the rates as
+$$
+R_{t1} - R_{t2} = \frac{1}{\gamma} \log \left( \frac{1 + e_{1t}}{1+e_{2t}} \right)
+$$
+where $e_{jt}$ is the error associated given in Section \ref{section:r0-estimation}.
+Recall that the error allowed for over-estimation prior to the peak and then under-estimation post-peak.
+Here, these errors can mingle in interesting ways.
+Consider two countries (A and B) in which the peak occurs 2 weeks prior for country A than country B.
+Figure~\ref{fig:comparison} presents such a comparison where each country's disease trajectory follows an SIR model (A=black and B=red).
+Of interest is the difference between the effective reproduction rate for country A and B, $R_{tA} - R_{tB}$.
+Figure~\ref{fig:eff} shows how biases in estimation of each rate interact in complex ways.
+At first, the difference is correctly estimated; then the gap is over-estimated as country A sees a rapid rise in cases; then the magnitude of over-estimation increases as country A sees declining case-count while country B sees rapidly increasing case-count; then country A's rate is correctly estimated while country B's rate is under-estimated as it sees declining case-count; finally, the gap disappears.
+We do not claim this will always be the case; however, this demonstrates how observed information can tell a more complex story than the truth (i.e., country A's peak is 2 weeks prior to country B's peak).
+
+\begin{figure}
+\centering
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{../figs/sir.png}
+  \caption{Fraction of new cases in population}
+  \label{fig:fracpop}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{../figs/sir_rt_comparison.png}
+  \caption{Effective reproduction rate estimators}
+  \label{fig:eff}
+\end{subfigure}
+\caption{Left: fraction infected in two SIR models with $\beta = 1.2$ and $0.9$ respectively and $\gamma = 0.15$ with same initial conditions.
+Right: comparison of $\hat R_t$ across time with $FN = 0.172$, $FP = 0.005$, and $M = 4$.}
+\label{fig:comparison}
 \end{figure}
@@
   \centering
-  \includegraphics[width=.9\linewidth]{../methods/figs/indianacasecounts.png}
+  \includegraphics[width=.9\linewidth]{../figs/indianacasecounts.png}
   \caption{Indiana case-counts per day}
@@
   \centering
-  \includegraphics[width=.9\linewidth]{../methods/figs/indianadeaths.png}
+  \includegraphics[width=.9\linewidth]{../figs/indianadeaths.png}
   \caption{Indiana death-counts per day}
@@
 
+\emph{Need to show Hospitalization C+ versus Covid count to show that most of the tests are in that subset for early parts.}
+
 \subsection{Model-based estimation}
@@
 
-\section{Cross-population comparisons}
-\label{section:crosspop}
-So far we have focused on understanding the limitations of using case-count data to understand population quantities of interest for a \emph{single} population.
+% \section{Cross-population comparisons}
+% \label{section:crosspop}
+% So far we have focused on understanding the limitations of using case-count data to understand population quantities of interest for a \emph{single} population.
 Many are interested in cross-population comparisons to contrast the impact of countries' mitigation policies.
@@
 
-\subsection{Prevalence comparisons}
-
-At a fixed time, suppose prevalence estimates $\bar y_1$ and $\bar y_2$ are observed for two populations. We can express $\bar y_1 - \bar y_2$ in statistical terms as
-$$
-(\bar Y_1 - \bar Y_2) + \left[ \rho_{I_1, Y_1} \times D_{M_1} \times \sqrt{\frac{1-f_1}{f_1}} \times \sigma_{Y_1}  - \rho_{I_2, Y_2} \times D_{M_2} \times \sqrt{\frac{1-f_2}{f_2}} \times \sigma_{Y_2} \right].
-$$
-The first term is the difference in population prevalence (i.e., the quantity of interest).
+% \subsection{Prevalence comparisons}
+
+% At a fixed time, suppose prevalence estimates $\bar y_1$ and $\bar y_2$ are observed for two populations. We can express $\bar y_1 - \bar y_2$ in statistical terms as
+% $$
+% (\bar Y_1 - \bar Y_2) + \left[ \rho_{I_1, Y_1} \times D_{M_1} \times \sqrt{\frac{1-f_1}{f_1}} \times \sigma_{Y_1}  - \rho_{I_2, Y_2} \times D_{M_2} \times \sqrt{\frac{1-f_2}{f_2}} \times \sigma_{Y_2} \right].
+% $$
+% The first term is the difference in population prevalence (i.e., the quantity of interest).
 If random sampling were performed in each population, $E_{\I} [ \rho_{I_j, Y_j}] = 0$ and so the estimate would be unbiased.
 However, the second term represents the complex error that results from selection bias.
-The particular test of interest is $H_0: \bar Y_1 = \bar Y_2$.  Under the null, $\sigma_{Y_1} = \sigma_{Y_2}$.  In classical statistics, we would compute a Z-score:
-$$
-\begin{aligned}
-\frac{\bar y_1 - \bar y_2}{ \sqrt{V_{SRS} (\bar y_1 - \bar y_2)} } &=
-\frac{\rho_{I_1, Y_1} \times D_{M_1} \times \sqrt{\frac{1-f_1}{f_1}}  - \rho_{I_2, Y_2} \times D_{M_2} \times \sqrt{\frac{1-f_2}{f_2}} }{ \sqrt{\frac{1}{N_1 -1} \times \frac{1-f_1}{f_1} + \frac{1}{N_2 - 1} \times \frac{1-f_2}{f_2}} }  \\
-&\overset{\star}{=}
-\sqrt{\frac{(N_1-1)(N_2 -1)}{N_1 + N_2}} \left[ \rho_{I_1, Y_1} - \rho_{I_2, Y_2} \right]
-\end{aligned}
-$$
-Where the last equality holds under $f_1 = f_2$, i.e., same sampling rate relative to population size.
+% The particular test of interest is $H_0: \bar Y_1 = \bar Y_2$.  Under the null, $\sigma_{Y_1} = \sigma_{Y_2}$.  In classical statistics, we would compute a Z-score:
+% $$
+% \begin{aligned}
+% \frac{\bar y_1 - \bar y_2}{ \sqrt{V_{SRS} (\bar y_1 - \bar y_2)} } &=
+% \frac{\rho_{I_1, Y_1} \times D_{M_1} \times \sqrt{\frac{1-f_1}{f_1}}  - \rho_{I_2, Y_2} \times D_{M_2} \times \sqrt{\frac{1-f_2}{f_2}} }{ \sqrt{\frac{1}{N_1 -1} \times \frac{1-f_1}{f_1} + \frac{1}{N_2 - 1} \times \frac{1-f_2}{f_2}} }  \\
+% &\overset{\star}{=}
+% \sqrt{\frac{(N_1-1)(N_2 -1)}{N_1 + N_2}} \left[ \rho_{I_1, Y_1} - \rho_{I_2, Y_2} \right]
+% \end{aligned}
+% $$
+% Where the last equality holds under $f_1 = f_2$, i.e., same sampling rate relative to population size.
 This tells us that the exact error in the comparison of sample means, as an estimate of the difference in population means, scales as a function of the difference in data quality $\rho_{I_1, Y_1} - \rho_{I_2, Y_2}$ times a square root of a population adjustment.
@@
 
-Take, for example, a comparison of the US and Canada.
+% Take, for example, a comparison of the US and Canada.
 The two population sizes are approximately 328 million and 38 million people.
@@
 That is, setting
-$$
-Z_n = \frac{\bar y_1 - \bar y_2}{ \frac{1-f}{f} \sigma_y \sqrt{ \frac{1}{n_{1,eff} - 1} + \frac{1}{n_{2,eff} - 1}}}
-$$
-allows us to build Z-scores that account for the true sample size.
+% $$
+% Z_n = \frac{\bar y_1 - \bar y_2}{ \frac{1-f}{f} \sigma_y \sqrt{ \frac{1}{n_{1,eff} - 1} + \frac{1}{n_{2,eff} - 1}}}
+% $$
+% allows us to build Z-scores that account for the true sample size.
 While still not perfect, this allows us to build Z-scores that are not overly confident due to selection bias and measurement error.
 See Section~\ref{section:est_dq} for discussion of how to estimate data quality.
-
-\subsection{Rate comparisons}
-
-An alternative is the comparison on the rates of case-count change over time.
-Here, for simplicity, we focus on comparing the estimated effective reproductive rate.
-We assume the two time-series are aligned so that $t=0$ is the time of first case.
-This negates alignment issues and is common in practice.
-
-While the above issues on $Z$-scores and effective sample size are important, here we highlight a separate issue.
-Namely, the interaction of biases in estimation of the two trajectories when the peak infection times differ slightly.
-We can write the difference in the rates as
-$$
-R_{t1} - R_{t2} = \frac{1}{\gamma} \log \left( \frac{1 + e_{1t}}{1+e_{2t}} \right)
-$$
-where $e_{jt}$ is the error associated given in Section \ref{section:r0-estimation}.
-Recall that the error allowed for over-estimation prior to the peak and then under-estimation post-peak.
-Here, these errors can mingle in interesting ways.
-Consider two countries (A and B) in which the peak occurs 2 weeks prior for country A than country B.
-Figure~\ref{fig:comparison} presents such a comparison where each country's disease trajectory follows an SIR model (A=black and B=red).
-Of interest is the difference between the effective reproduction rate for country A and B, $R_{tA} - R_{tB}$.
-Figure~\ref{fig:eff} shows how biases in estimation of each rate interact in complex ways.
-At first, the difference is correctly estimated; then the gap is over-estimated as country A sees a rapid rise in cases; then the magnitude of over-estimation increases as country A sees declining case-count while country B sees rapidly increasing case-count; then country A's rate is correctly estimated while country B's rate is under-estimated as it sees declining case-count; finally, the gap disappears.
-We do not claim this will always be the case; however, this demonstrates how observed information can tell a more complex story than the truth (i.e., country A's peak is 2 weeks prior to country B's peak).
-
-\begin{figure}
-\centering
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{../methods/figs/sir.png}
-  \caption{Fraction of new cases in population}
-  \label{fig:fracpop}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{../methods/figs/sir_rt_comparison.png}
-  \caption{Effective reproduction rate estimators}
-  \label{fig:eff}
-\end{subfigure}
-\caption{Left: fraction infected in two SIR models with $\beta = 1.2$ and $0.9$ respectively and $\gamma = 0.15$ with same initial conditions.
-Right: comparison of $\hat R_t$ across time with $FN = 0.172$, $FP = 0.005$, and $M = 4$.}
-\label{fig:comparison}
-\end{figure}
 
@@
 \section{Model-based estimation: parameter choices and inference}
-\label{app:modelbased}
+\label{app:modelbased_approx}
 

### 2021-06-16 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 
-Many states also provide aggregate demographic information on those who requested a test as well as those who tested positive for SARS-CoV-2.
-Indiana's \href{https://www.coronavirus.in.gov/2393.htm}{COVID-19 dashboard}, for instance, provides statewide demographics for age, gender, ethnicity, and race.
-The demographic information is based on cumulative testing.
-In addition, the Regenstrief Institute provides additional comorbidity information among those who tested positive for COVID-19 in the state of Indiana, including hypertension, diabetes, and chronic obstructive pulmonary disease (COPD) summary statistics.
-These comprise the main set of covariate information available from public sources for Indiana.
-Similar demographic information is likely available for other states.
+Many states do not provide any demographic information on who requested a test nor on who tested positive for SARS-CoV-2.
+Working closely with the State of Indiana, the author's were able to request public availability of COVID-19 total tests, positive tests, and related deaths per day broken out marginally by age, gender, ethnicity, and race.
+Similar demographic breakdowns may be available for other states, but to date the data has not been made publicly available at the daily level.
 % https://www.regenstrief.org/covid-dashboard/

### 2021-06-17 write-up/covid-writeup.tex
edits to first few sections (minor)

@@
 \boldshortcuts
+
+\usepackage{url}
 
@@
 \maketitle
+
+\newpage
 
@@
 
-Many states do not provide any demographic information on who requested a test nor on who tested positive for SARS-CoV-2.
-Working closely with the State of Indiana, the author's were able to request public availability of COVID-19 total tests, positive tests, and related deaths per day broken out marginally by age, gender, ethnicity, and race.
-Similar demographic breakdowns may be available for other states, but to date the data has not been made publicly available at the daily level.
+Public databases most often contain aggregate information.
+The Johns Hopkins dashboard, for example, provides demographic breakdown of case counts as well as the total confirmed cases and deaths by county.
+Aggregate time series of case count and deaths can also be extracted.
+Unfortunately, most dashboards do not provide any demographic information on who requested a test nor on who tested positive for SARS-CoV-2 over time.
+Working closely with the State of Indiana, we were able to generate publicly available datasets for COVID-19 total tests, positive tests, and related deaths per day broken out marginally by age, gender, ethnicity, and race~\citep{IndianaData2021}.
+This granular dataset motivates the proposed approach.
 % https://www.regenstrief.org/covid-dashboard/
@@
 
-Due to limited testing capacity, many US states have instituted restrictions on who can request a molecular test at a given time.
+Due to limited testing capacity, many US states instituted testing restrictions early on in the pandemic.
+% on who can request a molecular test at a given time.
 %https://www.coronavirus.in.gov/
-As of August 30th, 2020, a repository of the history of testing requirements does not exist.
-Early on, limited testing meant that only those at high risk were allowed to be tested (i.e., symptomatic or high risk populations); as testing capacity expanded, these restrictions were gradually lifted.
-
-Here, we focus on reconstructing the testing restriction history for the state of Indiana.
+% Early on, limited testing meant that only those at high risk were allowed to be tested (i.e., symptomatic or high risk populations); as testing capacity expanded, these restrictions were gradually lifted.
+Here, we reconstruct the testing restriction history for the state of Indiana.
 As of May 12th, 2020, the Indiana State Department of Health (ISDH) laboratory was testing high-risk individuals for COVID-19.
 Any symptomatic individual could receive a test.
-Moreover there were key factors for people who can be tested regardless of symptom status: over the age of 65, diabetic, high-blood pressure, obesity, pregnancy, and minority populations at greater risk of severe illness.
+Moreover there were key factors for people who can be tested regardless of symptom status: over the age of 65, diabetic, high-blood pressure, obesity, pregnancy, and minority populations at greater risk of severe illness~\citep{indystar2020}.
 % https://www.indystar.com/story/news/health/2020/05/12/coronavirus-testing-indiana-who-should-get-tested/3110592001/
-On April 28th, 2020, the criteria expanded to include any Indiana resident with virus symptoms, people in close contact with those who have tested positive and residents of congregant communities.
+On April 28th, 2020, the criteria expanded to include any Indiana resident with virus symptoms, people in close contact with those who have tested positive and residents of congregant communities~\citep{wishtv2020}.
 % https://www.wishtv.com/news/medical/indiana-opens-up-covid-19-testing-to-more-hoosiers/
-On June 15, 2020, ISDH lifted all restrictions and testing expanded to include anyone in Indiana who wants to be tested for coronavirus.
+On June 15, 2020, ISDH lifted all restrictions and testing expanded to include anyone in Indiana who wants to be tested for coronavirus~\citep{indystar2020v2}.
 % https://www.indystar.com/story/news/health/2020/06/12/indiana-says-anyone-who-wants-coronavirus-test-can-get-one/3179151001/
@@
 The order was extended until April 30th.
-On May 1st, 2020, a five-stage plan for gradual reopening was announced by Governon Holcomb -- See XX for exact details (cite: plan).
-Stage two went from May 4th to May 21st; stage three went from May 22nd to June 13th; stage 4 went from June 12th to July 3rd; stage 4.5 began on July 4th and will continue until September 25th; stage 5's start time has yet to be determined.
+On May 1st, 2020, a five-stage plan for gradual reopening was announced by Governon Holcomb~\citep{fivestageplan}.
 Such policies target reduction in active infection rates.
-In this paper, we incorporate Indiana's public health policy in our construction of epidemiological forecasts in Section~\ref{section:modelbased}.
+In this paper, Indiana's public health policy is incorporated in our construction of epidemiological forecasts in Section~\ref{section:modelbased}.
 
@@
 
-\section{Analysis of case-count data}
-\label{section:casecount}
-
-We start with some simple notation.
+%\section{Analysis of case-count data}
+%\label{section:casecount}
+%
+%We start with some simple notation.
 Let $N$ denote the population size.
@@
 We also let $I_j \in \{0,1\}$ be an indicator that the individual was selected for testing ($I = 1$) or not ($I=0$).
-
-To start, we assume the primary questions of interest are estimates of the overall number of COVID-19 cases and/or disease prevalence.
+%
+%To start, we assume the primary questions of interest are estimates of the overall number of COVID-19 cases and/or disease prevalence.
 That is, we are interested in either the population total $Y = \sum_{j=1}^N Y_j$ or the population average $\bar Y = Y/N$.
@@
 Then a natural candidate for prevalence is the proportion of positive tests $\bar y = \frac{1}{n} \sum_{i=1}^n y_i$, and a natural candidate for overall cases is $N \times \bar y$.
-Under simple random sampling (SRS) or any other epsem\footnote{equal probability of selection method} design, the above are unbiased estimators of the population-level quantities of interest.
+%Under simple random sampling (SRS) or any other epsem\footnote{equal probability of selection method} design, the above are unbiased estimators of the population-level quantities of interest.
 Under SRS, the variance of the estimator can be expressed as $\frac{1}{N-1} \times \frac{1-f}{f} \times \sigma_Y^2$ where $f = n/N$ is the sampling fraction and $\sigma_Y^2 = \frac{1}{N} \sum_{i=1}^N (Y_i - \bar Y)^2 = \bar Y (1- \bar Y)$.
-
-The above selection mechanisms are random and independent of the outcome of interest.
+%
+%The above selection mechanisms are random and independent of the outcome of interest.
 When this is not the case, selection effects may cause bias in the above estimates.
 To better understand this issue, \cite{Meng2018} provided the following intuitive and powerful statistical decomposition of the error between $\bar y$ and the true proportion $\bar Y$
-$$
-\bar y_n - \bar Y =  \rho_{I, Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_Y.
-$$
-The first term represents \emph{data quality}, the second \emph{data quantity}, and the third \emph{problem difficulty}.
+%$$
+%\bar y_n - \bar Y =  \rho_{I, Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_Y.
+%$$
+%The first term represents \emph{data quality}, the second \emph{data quantity}, and the third \emph{problem difficulty}.
 The term $\rho_{I,Y}$ is the empirical correlation between the population values~$\{ Y_j \}_{j=1}^N$ and the selection values $\{ I_j \}_{j=1}^N$.
@@
 The SRS variance formula above shows that $E_{\I} [ \rho_{I,Y}^2 ]  = 1/(N-1)$.
-
-% The key issue with selective testing is that $E_{\I} [ \rho_{I,Y} ] \neq 0$.
+%
+%% The key issue with selective testing is that $E_{\I} [ \rho_{I,Y} ] \neq 0$.
 Meng identified this as the fundamental issue that can lead to paradoxes in the analysis of big data.
@@
 We then discuss effective sampling methods and their importance in the current pandemic.
-
-\subsection{Imperfect testing}
-\label{section:imperfecttesting}
-
-Tests are imperfect.
+%
+%\subsection{Imperfect testing}
+%\label{section:imperfecttesting}
+%
+%Tests are imperfect.
 COVID-19 testing is no exception.
@@
 When paired with selection bias, however, the two sources of error become entangled, and resulting errors can become magnified, muted, or even switch signs.
-
-First we require some additional notation.
+%
+%First we require some additional notation.
 Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the outcome and $0$ when we observe the true outcome.
@@
 To understand the impact of selection bias and imperfect testing, we derive the following statistical decomposition of the error between $\tilde y_n$ and $\bar Y$:
-\begin{equation}
-\label{eq:statdecomp}
-\rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
-\times \underbrace{\left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{D_M},
-\end{equation}
-where $f=n/N$ is the sampling fraction, $f_1$ and $f_0$ are sampling fractions for COVID-19 positive and negative individuals respectively, and $\Delta = f_1 - f_0$ is the sampling rate differential.
+%\begin{equation}
+%\label{eq:statdecomp}
+%\rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
+%\times \underbrace{\left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{D_M},
+%\end{equation}
+%where $f=n/N$ is the sampling fraction, $f_1$ and $f_0$ are sampling fractions for COVID-19 positive and negative individuals respectively, and $\Delta = f_1 - f_0$ is the sampling rate differential.
 See Appendix~\ref{app:memestimator} for the derivation.
@@
 The final term represents the \emph{imperfect testing adjustment} which is again a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence but here is scaled by the factor $(1 - FP - FN)^{-1}$ compared to prior decomposition.
-
-Figure~\ref{fig:heatmap} shows that $D_M$, as a function of the relative frequency ($f_1/f_0$) and odds ratio, can be both positive and negative as well as a range of magnitudes~\citep{Beesley2020,Beesley2019,Smeden2019}.
+%
+%Figure~\ref{fig:heatmap} shows that $D_M$, as a function of the relative frequency ($f_1/f_0$) and odds ratio, can be both positive and negative as well as a range of magnitudes~\citep{Beesley2020,Beesley2019,Smeden2019}.
 Under perfect testing, $D_M = 1$ so the relation between estimation and selection bias is simple, e.g., COVID-19 positive individuals more likely to receive test implies upward bias in prevalence estimates.
 When tests are imperfect, this simple relationship no longer holds.
-
-\begin{figure}[!th]
-\centering
-\includegraphics[width = 0.4\textwidth]{../figs/mem_heatmap_article.png}
-\caption{Imperfect testing adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.005$ and $FN=0.172$. Color scaled so blue = $-10$, white = $0$, and red = $10$.}
-\label{fig:heatmap}
-\vspace{-0.3cm}
-\end{figure}
-
-Comparing the mean-squared error (MSE) under a selection mechanism $\I$ with imperfect testing and SRS with perfect testing, we see that
-$$
-\frac{E_{\I} \left[ (\bar y_n - \bar Y)^2 \right]}{\sqrt{V_{SRS} (\bar Y)}} = (N-1) E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right]
-$$
-implying the error relative to SRS increases as a function of population size~\citep{Meng2018}.
+%
+%\begin{figure}[!th]
+%\centering
+%\includegraphics[width = 0.4\textwidth]{../figs/mem_heatmap_article.png}
+%\caption{Imperfect testing adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.005$ and $FN=0.172$. Color scaled so blue = $-10$, white = $0$, and red = $10$.}
+%\label{fig:heatmap}
+%\vspace{-0.3cm}
+%\end{figure}
+%
+%Comparing the mean-squared error (MSE) under a selection mechanism $\I$ with imperfect testing and SRS with perfect testing, we see that
+%$$
+%\frac{E_{\I} \left[ (\bar y_n - \bar Y)^2 \right]}{\sqrt{V_{SRS} (\bar Y)}} = (N-1) E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right]
+%$$
+%implying the error relative to SRS increases as a function of population size~\citep{Meng2018}.
 The paradox that two countries with the same testing strategy can yield wildly different estimates due to population size is too often ignored in current communications of case-count data.
@@
 Thus, conclusions drawn from observed case-counts may not just be wrong, but very wrong.
-
-A key question is ``What is the (effective) sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' In Appendix~\ref{app:effss}, we show the effective sample size $n_{eff}$ can be bounded by
-$$
-\frac{f}{1-f} \times \frac{1}{E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right]}.
-$$
-As of June 10th, the US has performed $21,474,531$ tests.
+%
+%A key question is ``What is the (effective) sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' In Appendix~\ref{app:effss}, we show the effective sample size $n_{eff}$ can be bounded by
+%$$
+%\frac{f}{1-f} \times \frac{1}{E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right]}.
+%$$
+%As of June 10th, the US has performed $21,474,531$ tests.
 The US population is roughly $312$ million, so $f = 0.069$.
@@
 Then the effective sample size increases by a factor of $1.76$ rather than the expected $4$ fold increase.
-% \begin{wrapfigure}{r}{0.5\textwidth}
-
-% {\bf Figure of individual going through the states? Can discuss exposure length as selection more likely.}
-
-\subsection{Regrettable rates: complex biases resulting from self-selection}
-\label{section:rates}
-
-The prior analysis demonstrates the limited information regarding COVID-19 prevalence in observational case-count data.
+%% \begin{wrapfigure}{r}{0.5\textwidth}
+%
+%% {\bf Figure of individual going through the states? Can discuss exposure length as selection more likely.}
+%
+%\subsection{Regrettable rates: complex biases resulting from self-selection}
+%\label{section:rates}
+%
+%The prior analysis demonstrates the limited information regarding COVID-19 prevalence in observational case-count data.
 Scientists, however, may claim that daily observed case-counts simply undercount daily total cases by a constant multiple over time.
@@
 In this section, we demonstrate how selection bias and imperfect testing impact such estimates.
-
-Let $\bar Y_{t-1}$ and $\bar Y_{t}$ denote the prevalence on two consecutive days and consider the estimator $r = \tilde y_t / \tilde y_{t-1}$.
+%
+%Let $\bar Y_{t-1}$ and $\bar Y_{t}$ denote the prevalence on two consecutive days and consider the estimator $r = \tilde y_t / \tilde y_{t-1}$.
 Using a second-order Taylor series approximation, the error between ${\tilde y_t}/{\tilde y_{t-1}}$ and ${\bar Y_{t}}/{\bar Y_{t-1}}$ can be expressed approximately as
-$$
-\begin{aligned}
-\frac{\bar Y_t}{\bar Y_{t-1}} &\times \bigg[ \rho_{I_t,Y_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (Y_t)  -\rho_{I_{t-1},Y_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \bigg] \\
-&\times \left[ 1 - \rho_{I_{t-1},Y_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \right]
-\end{aligned}
-$$
-where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement error adjustment, and $CV(Y_j) = \sigma_{Y_j}/Y_j$ is the coefficient of variation on day $j$.
+%$$
+%\begin{aligned}
+%\frac{\bar Y_t}{\bar Y_{t-1}} &\times \bigg[ \rho_{I_t,Y_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (Y_t)  -\rho_{I_{t-1},Y_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \bigg] \\
+%&\times \left[ 1 - \rho_{I_{t-1},Y_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \right]
+%\end{aligned}
+%$$
+%where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement error adjustment, and $CV(Y_j) = \sigma_{Y_j}/Y_j$ is the coefficient of variation on day $j$.
 See Appendix~\ref{app:ratio} for the derivation.
@@
 The second term represents potential \emph{cancellation} which may occur when data quality, sampling fraction, measurement error, and prevalence are all constant across time.
-
-Figure~\ref{fig:ratio-bias} displays the trajectory of the true ratio and the potential biased estimators under an SIR model~\citep{Pastor2001,Newman2002,Parshani2010} for the epidemic dynamics, with state evolution given by
-\begin{equation}
-\label{eq:sir}
-\frac{\partial s_t}{\partial t} = - \beta s_t i_t; \quad
-\frac{\partial i_t}{\partial t} = \beta s_t i_t - \gamma i_t; \quad
-\frac{\partial r_t}{\partial t} = - \gamma i_t
-\end{equation}
-where $s_t$ is the fraction of susceptible individuals, $i_t$ is the fraction of infected individuals, and $r_t$ is the fraction of removed (recovered or deceased) individuals in the population at time $t$.
+%
+%Figure~\ref{fig:ratio-bias} displays the trajectory of the true ratio and the potential biased estimators under an SIR model~\citep{Pastor2001,Newman2002,Parshani2010} for the epidemic dynamics, with state evolution given by
+%\begin{equation}
+%\label{eq:sir}
+%\frac{\partial s_t}{\partial t} = - \beta s_t i_t; \quad
+%\frac{\partial i_t}{\partial t} = \beta s_t i_t - \gamma i_t; \quad
+%\frac{\partial r_t}{\partial t} = - \gamma i_t
+%\end{equation}
+%where $s_t$ is the fraction of susceptible individuals, $i_t$ is the fraction of infected individuals, and $r_t$ is the fraction of removed (recovered or deceased) individuals in the population at time $t$.
 In terms of bias, the rate is overestimated prior to the peak and underestimated afterwards; the bias increases dramatically when the relative fraction exceeds $2$.
@@
 Note, however, that the standard errors will be large due to small effective sample sizes.
-
-\subsection{Estimation of effective reproduction number}
-\label{section:r0-estimation}
-Many epidemiologists argue that tracking the effective reproduction number $R_t$ is the only way to manage through the crisis~\citep{Gabriel2020}.
+%
+%\subsection{Estimation of effective reproduction number}
+%\label{section:r0-estimation}
+%Many epidemiologists argue that tracking the effective reproduction number $R_t$ is the only way to manage through the crisis~\citep{Gabriel2020}.
 Under a Poisson likelihood, a simple relation between the trajectory of new cases and the effective reproduction number can be derived \citep{Bettencourt2008}.
@@
 Using this relation, a moment-based estimator is given by
-$$
-R_t = 1 + \frac{1}{\gamma} \log \left( \frac{K_t}{K_{t-1}} \right).
-$$
-Of course, we do not observe $K_t$ and $K_{t-1}$.
+%$$
+%R_t = 1 + \frac{1}{\gamma} \log \left( \frac{K_t}{K_{t-1}} \right).
+%$$
+%Of course, we do not observe $K_t$ and $K_{t-1}$.
 Under SRS of new cases among those susceptible on day $t$, the natural estimator is $S_t \tilde y_t$.
@@
 We can again express the statistical error of $\hat R_t - R_t$ in useful terms as follows
-$$
-\begin{aligned}
-\frac{1}{\gamma}\log &\bigg( 1 + \bigg[ \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t)  -\rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (K_{t-1}) \bigg] \\
-&\times \left[ 1 - \rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (K_{t-1}) \right] \bigg) - \frac{1}{\gamma} \log \left( \frac{S_t}{S_{t-1}} \right).
-\end{aligned}
-$$
-This implies a similar trade-off as before but on the logarithmic scale.
+%$$
+%\begin{aligned}
+%\frac{1}{\gamma}\log &\bigg( 1 + \bigg[ \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t)  -\rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (K_{t-1}) \bigg] \\
+%&\times \left[ 1 - \rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (K_{t-1}) \right] \bigg) - \frac{1}{\gamma} \log \left( \frac{S_t}{S_{t-1}} \right).
+%\end{aligned}
+%$$
+%This implies a similar trade-off as before but on the logarithmic scale.
 The error is no longer scaled by $\bar Y_t/\bar Y_{t-1}$ but by the serial interval and does not depend on prevalence $\bar Y_t$ but on the fraction of new cases out of those susceptible $\bar K_t$.
@@
 Figure~\ref{fig:r0-bias} displays the bias as a function of the relative sampling fraction assuming the fraction of the population that is susceptible remains large.
-
-\begin{figure}
-\centering
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{../figs/sir_ratio.png}
-  \caption{Ratio estimator}
-  \label{fig:ratio-bias}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{../figs/sir_rt.png}
-  \caption{Effective reproductive rate estimator}
-  \label{fig:r0-bias}
-\end{subfigure}
-\caption{Potential bias due to self-selection and measurement error from a deterministic SIR model with $\beta = 1.2$ (black) and $\gamma = 0.15$.
+%
+%\begin{figure}
+%\centering
+%\begin{subfigure}{.5\textwidth}
+%  \centering
+%  \includegraphics[width=.9\linewidth]{../figs/sir_ratio.png}
+%  \caption{Ratio estimator}
+%  \label{fig:ratio-bias}
+%\end{subfigure}%
+%\begin{subfigure}{.5\textwidth}
+%  \centering
+%  \includegraphics[width=.9\linewidth]{../figs/sir_rt.png}
+%  \caption{Effective reproductive rate estimator}
+%  \label{fig:r0-bias}
+%\end{subfigure}
+%\caption{Potential bias due to self-selection and measurement error from a deterministic SIR model with $\beta = 1.2$ (black) and $\gamma = 0.15$.
 Here we assume $f = 0.02$, $FP = 0.005$, $FN = 0.172$, and use a range of relative sampling fraction $M = f_1/f_0$.}
-\label{fig:rates}
-\end{figure}
-
-\subsection{Rate comparisons}
-
-So far we have focused on understanding the limitations of using case-count data to understand population quantities of interest for a \emph{single} population.
+%\label{fig:rates}
+%\end{figure}
+%
+%\subsection{Rate comparisons}
+%
+%So far we have focused on understanding the limitations of using case-count data to understand population quantities of interest for a \emph{single} population.
 Many are interested in cross-population comparisons to contrast the impact of countries' mitigation policies.
@@
 We assume the two time-series are aligned so that $t=0$ is the time of first case in each population respectively.
-% This negates alignment issues and is common in practice.
-
-% While the above issues on $Z$-scores and effective sample size are important, here we highlight a separate issue.
-Here we demonstrate the interaction of biases in estimation of the two trajectories when the peak infection times differ slightly.  First, we can write the difference in the rates as
-$$
-R_{t1} - R_{t2} = \frac{1}{\gamma} \log \left( \frac{1 + e_{1t}}{1+e_{2t}} \right)
-$$
-where $e_{jt}$ is the error associated given in Section \ref{section:r0-estimation}.
+%% This negates alignment issues and is common in practice.
+%
+%% While the above issues on $Z$-scores and effective sample size are important, here we highlight a separate issue.
+%Here we demonstrate the interaction of biases in estimation of the two trajectories when the peak infection times differ slightly.  First, we can write the difference in the rates as
+%$$
+%R_{t1} - R_{t2} = \frac{1}{\gamma} \log \left( \frac{1 + e_{1t}}{1+e_{2t}} \right)
+%$$
+%where $e_{jt}$ is the error associated given in Section \ref{section:r0-estimation}.
 Recall that the error allowed for over-estimation prior to the peak and then under-estimation post-peak.
@@
 We do not claim this will always be the case; however, this demonstrates how observed information can tell a more complex story than the truth (i.e., country A's peak is 2 weeks prior to country B's peak).
-
-\begin{figure}
-\centering
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{../figs/sir.png}
-  \caption{Fraction of new cases in population}
-  \label{fig:fracpop}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{../figs/sir_rt_comparison.png}
-  \caption{Effective reproduction rate estimators}
-  \label{fig:eff}
-\end{subfigure}
-\caption{Left: fraction infected in two SIR models with $\beta = 1.2$ and $0.9$ respectively and $\gamma = 0.15$ with same initial conditions.
+%
+%\begin{figure}
+%\centering
+%\begin{subfigure}{.5\textwidth}
+%  \centering
+%  \includegraphics[width=.9\linewidth]{../figs/sir.png}
+%  \caption{Fraction of new cases in population}
+%  \label{fig:fracpop}
+%\end{subfigure}%
+%\begin{subfigure}{.5\textwidth}
+%  \centering
+%  \includegraphics[width=.9\linewidth]{../figs/sir_rt_comparison.png}
+%  \caption{Effective reproduction rate estimators}
+%  \label{fig:eff}
+%\end{subfigure}
+%\caption{Left: fraction infected in two SIR models with $\beta = 1.2$ and $0.9$ respectively and $\gamma = 0.15$ with same initial conditions.
 Right: comparison of $\hat R_t$ across time with $FN = 0.172$, $FP = 0.005$, and $M = 4$.}
-\label{fig:comparison}
-\end{figure}
-
-\section{Potential improvements to prevalence estimation}
-\label{section:improvedcasecount}
-
-The prior section presented negative consequences of selection-bias and measurement error when estimating infection prevalence from observed case-count data.
+%\label{fig:comparison}
+%\end{figure}
+%
+%\section{Potential improvements to prevalence estimation}
+%\label{section:improvedcasecount}
+%
+%The prior section presented negative consequences of selection-bias and measurement error when estimating infection prevalence from observed case-count data.
 This raises the immediate question whether prevalence estimators can be improved.
 In this section, we consider two directions.
-% First, we use publicly available auxiliary information to build selection propensities and consider an inverse probability weighting (IPW) estimator.
+%% First, we use publicly available auxiliary information to build selection propensities and consider an inverse probability weighting (IPW) estimator.
 We then extend the statistical error decomposition to the weighted setting.
@@
 We provide an appropriate extension of the statistical error decomposition.
-The proposed methods and statistical error decompositions guide our recommendations in Section~\ref{section:discussion}.
-
-\subsection{Selection propensity estimation}
-
-With non-probability samples, one can reduce bias by modelling the self-selection propensity and use inverse probability weighting (IPW)~\citep{Elliott2017} techniques to adjust for selection bias.
+%The proposed methods and statistical error decompositions guide our recommendations in Section~\ref{section:discussion}.
+%
+%\subsection{Selection propensity estimation}
+%
+%With non-probability samples, one can reduce bias by modelling the self-selection propensity and use inverse probability weighting (IPW)~\citep{Elliott2017} techniques to adjust for selection bias.
 In causal inference~\citep{Hernan2020}, under certain assumptions, the observational data permits direct estimation of treatment propensity.
@@
 In order to estimate selection propensities, one needs auxiliary information.
-
-To see this mathematically, let $X_j$ denote a vector of covariates for the $j$th individual in the population, $j=1,\ldots,N$.
+%
+%To see this mathematically, let $X_j$ denote a vector of covariates for the $j$th individual in the population, $j=1,\ldots,N$.
 For simplicity, the dynamic nature of the outbreak and recoverability of individuals is ignored for now.
@@
 Maximum likelihood estimation follows by maximizing
-\begin{equation}
-\label{eq:propensity}
-\sum_{j=1}^N I_j \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi (X_j; \theta) \right).
-\end{equation}
-The first sum only involves individuals observed in the nonprobability sample.
-The second sum is over the entire population.
+%\begin{equation}
+%\label{eq:propensity}
+%\sum_{j=1}^N I_j \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi (X_j; \theta) \right).
+%\end{equation}
+%The first sum only involves individuals observed in the nonprobability sample.
+%The second sum is over the entire population.
 Maximum likelihood estimation therefore requires knowledge of covariate information for every individual in the population.
@@
 Here we present a method that uses auxiliary information obtained from probability samples as well as probability subsamples from the self-selected population.
-
-\subsubsection{Auxiliary information through probability samples}
-\label{subsec:auxprob}
-
-Here, we assume access to a probability sample from the same population on which the outcome and exact same set of covariates are measured.
+%
+%\subsubsection{Auxiliary information through probability samples}
+%\label{subsec:auxprob}
+%
+%Here, we assume access to a probability sample from the same population on which the outcome and exact same set of covariates are measured.
 Let $\tilde I_j$ be an indicator that the individual was included in the probability sample and $\tilde W_j$ be the probability of inclusion for the $j$th individual (i.e., equal to $n/N$ for a simple random sample of size $n$).
 The probability sample is used to construct a design-unbiased estimator of the second term
-\begin{equation}
-\label{eq:auxinfoprob}
-\sum_{j=1}^N I_j \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
-\end{equation}
-Taking the expectation of the second term with respect to the sampling design yields the second term in~\eqref{eq:propensity}.
+%\begin{equation}
+%\label{eq:auxinfoprob}
+%\sum_{j=1}^N I_j \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
+%\end{equation}
+%Taking the expectation of the second term with respect to the sampling design yields the second term in~\eqref{eq:propensity}.
 This was proposed by~\cite{Chen2019} when covariates were observed on a random sample from the population.
-% Under simple random sampling and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the $\hat \pi_k = (d_k / \tilde d_k) (n / N)$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.
-
-% Equation~\eqref{eq:auxinfoprob} presupposes covariate information is collected for all individuals in the nonprobability sample; however, this is often not feasible for state and local governments where rapid sample collection is prioritized over additional collection of covariate information.
+%% Under simple random sampling and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the $\hat \pi_k = (d_k / \tilde d_k) (n / N)$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.
+%
+%% Equation~\eqref{eq:auxinfoprob} presupposes covariate information is collected for all individuals in the nonprobability sample; however, this is often not feasible for state and local governments where rapid sample collection is prioritized over additional collection of covariate information.
 We can extend~\eqref{eq:auxinfoprob} by randomly sampling a subset of the nonprobability sample on which to collect the covariate information, yielding the pseudo-likelihood
-% \begin{equation}
-% \label{eq:auxinfoprob2}
-% \sum_{j=1}^N I_j I_j^{(2)} W_j^{(2)} \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
-% \end{equation}
-% Here, $I_j^{(2)}$ is an indicator that the individual is in the probability subsample and $W_j^{(2)}$ is the probability of inclusion.
+%% \begin{equation}
+%% \label{eq:auxinfoprob2}
+%% \sum_{j=1}^N I_j I_j^{(2)} W_j^{(2)} \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
+%% \end{equation}
+%% Here, $I_j^{(2)}$ is an indicator that the individual is in the probability subsample and $W_j^{(2)}$ is the probability of inclusion.
 Estimation proceeds by maximizing the pseudo-likelihood which now depends on random sampling in both the first and second term.
-
-\subsubsection{An IPW estimator and statistical error decomposition}
-\label{section:IPWerrordecomp}
-
-Given a selection propensity, define the inverse probability weight~$w(x) = \pi (x; \hat \theta)^{-1}$. Then the IPW estimator adjusted for measurement error is given by
-\begin{equation}
-\label{eq:ipwest}
-\bar y_n^\star
-= \frac{1}{1-FP-FN} \cdot \frac{\sum_{i=1}^n w(x_i) (y_i - FP)}{\sum_{i=1}^n w (x_i)}
-\stackrel{(2)}{=} \frac{1}{1-FP-FN} \sum_{k=1}^K \frac{d_k w_k}{w} (\bar y_k - FP),
-\end{equation}
-where equality (2) is under the assumption that $x_i$ is a stratification variable with $k$ indexing the strata, $w_k$ is the weight and $d_k$ is the number of samples in strata $k$, and $w = \sum_{k=1}^K d_k w_k$.
+%
+%\subsubsection{An IPW estimator and statistical error decomposition}
+%\label{section:IPWerrordecomp}
+%
+%Given a selection propensity, define the inverse probability weight~$w(x) = \pi (x; \hat \theta)^{-1}$. Then the IPW estimator adjusted for measurement error is given by
+%\begin{equation}
+%\label{eq:ipwest}
+%\bar y_n^\star
+%= \frac{1}{1-FP-FN} \cdot \frac{\sum_{i=1}^n w(x_i) (y_i - FP)}{\sum_{i=1}^n w (x_i)}
+%\stackrel{(2)}{=} \frac{1}{1-FP-FN} \sum_{k=1}^K \frac{d_k w_k}{w} (\bar y_k - FP),
+%\end{equation}
+%where equality (2) is under the assumption that $x_i$ is a stratification variable with $k$ indexing the strata, $w_k$ is the weight and $d_k$ is the number of samples in strata $k$, and $w = \sum_{k=1}^K d_k w_k$.
 Under simple random sampling with one strata~$\hat \pi_k = n/N$ and we recover the estimator $\tilde y_n$ from Section~\ref{section:imperfecttesting}.
-
-Let $\tilde I_j (X_j) = I_j  \cdot W(X_j)$ for $j=1,\ldots,N$.  Then the error when comparing weighted estimator~$\bar y_n^\star$ to the true prevalence $\bar Y$ can be expressed as:
-\begin{equation}
-\label{eq:statdecomp2}
-\rho_{\tilde I (X), Y} \times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y} \times \underbrace{\left[ 1 - \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{\tilde D_M}
-\end{equation}
-where $CV_W$ is the coefficient of variation (i.e., standard deviation/mean) of $W_J$ given $I_J = 1$, $\rho_{\tilde I(X), Y}$ is the empirical correlation which here depends on covariate distribution, $\tilde f_k = E[ W_J I_J \mid Y_j = k]$ for $k=0,1$, and $\tilde \Delta = \tilde f_1 - \tilde f_0$.
+%
+%Let $\tilde I_j (X_j) = I_j  \cdot W(X_j)$ for $j=1,\ldots,N$.  Then the error when comparing weighted estimator~$\bar y_n^\star$ to the true prevalence $\bar Y$ can be expressed as:
+%\begin{equation}
+%\label{eq:statdecomp2}
+%\rho_{\tilde I (X), Y} \times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y} \times \underbrace{\left[ 1 - \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{\tilde D_M}
+%\end{equation}
+%where $CV_W$ is the coefficient of variation (i.e., standard deviation/mean) of $W_J$ given $I_J = 1$, $\rho_{\tilde I(X), Y}$ is the empirical correlation which here depends on covariate distribution, $\tilde f_k = E[ W_J I_J \mid Y_j = k]$ for $k=0,1$, and $\tilde \Delta = \tilde f_1 - \tilde f_0$.
 See Appendix~\ref{app:ipwderivation} for the derivation.
-
-Comparing~\eqref{eq:statdecomp2} to~\eqref{eq:statdecomp} shows that weighting impacts the estimation error in three ways.
+%
+%Comparing~\eqref{eq:statdecomp2} to~\eqref{eq:statdecomp} shows that weighting impacts the estimation error in three ways.
 First, there is a negative impact on the data quantity component; taking the ratio of these quantities yields
-$\sqrt{1 + \frac{CV_W^2}{1-f}} \geq 1$.
+%$\sqrt{1 + \frac{CV_W^2}{1-f}} \geq 1$.
 Hence, if the data quality does not increase (i.e., $| \rho_{\tilde I (X), Y} | = | \rho_{I,Y}|$ ) then weighting increases the error magnitude.
@@
 In particular, weighting may result in $\text{sgn}(\Delta) \neq \text{sgn} (\tilde \Delta)$ which implies the impact of measurement-error may be in a different direction.
-
-As demonstrated in Lemma~\ref{lemma:ipw}, if the propensity model is correctly specified then the $E_{\bf R} [ \rho_{\tilde I (X), Y} ] = 0$ and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then the data quality index is unlikely to inversely scale with population size.
+%
+%As demonstrated in Lemma~\ref{lemma:ipw}, if the propensity model is correctly specified then the $E_{\bf R} [ \rho_{\tilde I (X), Y} ] = 0$ and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then the data quality index is unlikely to inversely scale with population size.
 Indeed, Horvitz-Thompson estimators are sensitive to errors in the estimated weights.
@@
 Similar to~\cite{Meng2018}, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from an IPW estimator are likely to put too much confidence in the sheer size of the data.
-
-\subsubsection{Time-varying propensities}
-
-In prior sections on selection propensity estimation, we ignored the dynamic nature of the outbreak.  Here we extend the IPW approach to the temporal setting by considering the joint likelihood
-\begin{equation}
-\label{eq:tvpropensity}
-\sum_{t=1}^T \left[ \sum_{j=1}^N I_{j,t} \log \left( \frac{\pi_t (X_{j,t}; \theta)}{1-\pi_t(X_{j,t}; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi_t (X_{j,t}; \theta) \right) \right]
-\end{equation}
-where $t=1,\ldots,T$ are the days when case-count data is reported.
+%
+%\subsubsection{Time-varying propensities}
+%
+%In prior sections on selection propensity estimation, we ignored the dynamic nature of the outbreak.  Here we extend the IPW approach to the temporal setting by considering the joint likelihood
+%\begin{equation}
+%\label{eq:tvpropensity}
+%\sum_{t=1}^T \left[ \sum_{j=1}^N I_{j,t} \log \left( \frac{\pi_t (X_{j,t}; \theta)}{1-\pi_t(X_{j,t}; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi_t (X_{j,t}; \theta) \right) \right]
+%\end{equation}
+%where $t=1,\ldots,T$ are the days when case-count data is reported.
 Here, $I_{j,t}$ denotes self-selection into testing on day $t$, which is highly correlated with prior testing and results.
@@
 Here, we assume that the covariate vector $X_{j,t}$ contains all features of past relevant for selection.
-
-If sufficiently large random samples are collected at each time $t =1,\ldots,T$, then the pseudo-likelihood can be re-written as in~\eqref{eq:auxinfoprob} and propensities estimated separately per time point.
+%
+%If sufficiently large random samples are collected at each time $t =1,\ldots,T$, then the pseudo-likelihood can be re-written as in~\eqref{eq:auxinfoprob} and propensities estimated separately per time point.
 Unfortunately for COVID-19, large probabilistic samples are not available at every time within a given region.
 For simplicity, here we discuss a non-parametric kernel-based approach where the selection propensity at time~$t$, denoted $\hat \theta_t$, maximizes the smoothed pseudo-likelihood
-$$
-\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \left[ \sum_{j=1}^N I_{j,t^\prime} \log \left( \frac{\pi_t (X_{j,t^\prime}; \theta)}{1-\pi_t(X_{j,t^\prime}; \theta)} \right) + \sum_{j=1}^N \tilde W_{j,t^\prime} \tilde I_{j,t^\prime} \log \left( 1 - \pi_t (X_{j,t^\prime}; \theta) \right) \right]
-$$
-where $K_h$ is a kernel function with tunable parameter $h$. Given $\pi (x;\hat \theta_t)$, the prevalence estimator $\bar y_{n,t}^\star$ is given by~\ref{eq:ipwest}.
-
-% Under simple random sampling of equal size~$\tilde n$ from a fixed population size~$N$ at each time and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the
-% $$
-% \hat \pi_{k,t} = \frac{n}{N} \times \frac{\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \hat d_{k,t^\prime}}{\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \tilde d_{k,t^\prime}}
-% $$
-% where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.
-
-\subsubsection{Uncertain sensitivity and specificity}
-
-For prevalence there is some
-
-For
-
-For seroprevalence, \cite{Bendavid2020} report
-
-, which leads to a range for $M$ of $(2.21, 2.38)$.
+%$$
+%\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \left[ \sum_{j=1}^N I_{j,t^\prime} \log \left( \frac{\pi_t (X_{j,t^\prime}; \theta)}{1-\pi_t(X_{j,t^\prime}; \theta)} \right) + \sum_{j=1}^N \tilde W_{j,t^\prime} \tilde I_{j,t^\prime} \log \left( 1 - \pi_t (X_{j,t^\prime}; \theta) \right) \right]
+%$$
+%where $K_h$ is a kernel function with tunable parameter $h$. Given $\pi (x;\hat \theta_t)$, the prevalence estimator $\bar y_{n,t}^\star$ is given by~\ref{eq:ipwest}.
+%
+%% Under simple random sampling of equal size~$\tilde n$ from a fixed population size~$N$ at each time and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the
+%% $$
+%% \hat \pi_{k,t} = \frac{n}{N} \times \frac{\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \hat d_{k,t^\prime}}{\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \tilde d_{k,t^\prime}}
+%% $$
+%% where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.
+%
+%\subsubsection{Uncertain sensitivity and specificity}
+%
+%For prevalence there is some
+%
+%For
+%
+%For seroprevalence, \cite{Bendavid2020} report
+%
+%
+%
+%, which leads to a range for $M$ of $(2.21, 2.38)$.
 Therefore, the impact on bias of $\hat R_t$ appears moderate.
@@
 This simple tool helps us understand how sensitivity and specificity impact biases and  therefore how much we should trust conclusions based on these assumptions.
-
-\subsubsection{Asymptotics}
-
-Assuming the selection propensity model is correctly specified, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:ipw} shows that the IPW estimator is  consistent and calculates the estimator's variance.
+%
+%
+%\subsubsection{Asymptotics}
+%
+%Assuming the selection propensity model is correctly specified, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:ipw} shows that the IPW estimator is  consistent and calculates the estimator's variance.
 In Lemma~\ref{lemma:ipw}, sensitivity and specificity are not assumed known.
@@
 See Appendix~\ref{app:asympderivations} for additional details.
-
-\begin{lemma}[Variance of IPW estimator] \normalfont
-\label{lemma:ipw}
-\textcolor{red}{Must check.}
-The estimates~$\mu_t := \bar y_{n,t}^\star$ and $\hat \pi_{j,t}$
-are solutions to the following set of estimating equations
-$$
-\Phi_n (\eta_t) =
-\left (
-\begin{array}{c}
-\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - (1-FP-FN) \cdot (\mu_t+FP)}{\pi_{j,t}} \\
-\frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
-\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
-\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
-\end{array}
-\right ) = {\bf 0}
-$$
-where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
+%
+%\begin{lemma}[Variance of IPW estimator] \normalfont
+%\label{lemma:ipw}
+%\textcolor{red}{Must check.}
+%The estimates~$\mu_t := \bar y_{n,t}^\star$ and $\hat \pi_{j,t}$
+%are solutions to the following set of estimating equations
+%$$
+%\Phi_n (\eta_t) =
+%\left (
+%\begin{array}{c}
+%\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - (1-FP-FN) \cdot (\mu_t+FP)}{\pi_{j,t}} \\
+%\frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
+%\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
+%\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
+%\end{array}
+%\right ) = {\bf 0}
+%$$
+%where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
 Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$ (check)  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ where $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$.
-\end{lemma}
-
-\subsection{Model-based estimation}
-\label{section:modelbased}
-
-Up to this point, the primary focus of this paper has been on selection bias in coronavirus case-count data from a survey sampling perspective.
+%\end{lemma}
+%
+%\subsection{Model-based estimation}
+%\label{section:modelbased}
+%
+%Up to this point, the primary focus of this paper has been on selection bias in coronavirus case-count data from a survey sampling perspective.
 Here, we consider compartmental model approaches from infectious disease epidemiology.
@@
 Here we consider the probabilistic SIR model was originally proposed by~\cite{Osthus2017} with only one-dimensional time series of infected proportions; this formulation was extended by~\cite{Song2020} to model coronavirus case-counts.
-
-Let $s_t$, $e_t$, $i_t^{(T)}$, $i^{(NT)}$, and $r_t$ denote the proportion of survivors, exposed, currently transmitting and not trasmitting infected, and removed cases (i.e., including both recovered cases and deaths) at time $t$.
+%
+%Let $s_t$, $e_t$, $i_t^{(T)}$, $i^{(NT)}$, and $r_t$ denote the proportion of survivors, exposed, currently transmitting and not trasmitting infected, and removed cases (i.e., including both recovered cases and deaths) at time $t$.
 Suppose $K$ strata such that the proportions are defined per strata, i.e., $s_t = (s_{1,t},\ldots, s_{K,t})$ is the proportion of survivors in each strata.
 Let $\theta_t = (s_t, e_t, i_t^{(T)}, i_t^{(NT)}, r_t)$, then we assume the process is governed by the following Markov model:
-$$
-\theta_t \mid \theta_{t-1}; \tau \sim \text{Dirichlet} \left(\kappa f(\theta_{t-1}, \beta, \gamma) \right)
-$$
-where $\kappa$ scales the variance of the Dirichlet distribution and $f(\cdot)$ is a function that determines infection dynamics.  Here, we extend the SIR dynamics to account for strata-level dynamics
-\begin{align*}
-\frac{\partial s_t}{\partial t} &= - s_t \circ (B \, i^{(T)}_t); \quad \frac{\partial e_t}{\partial t} = s_t \circ (B \, i^{(T)}_t) - \sigma e_t; \quad \\
-\frac{\partial i^{(T)}_t}{\partial t} &= \sigma e_t - \beta_2 i^{(T)}; \quad
-\frac{\partial i^{(NT)}_t}{\partial t} = \beta_2 i^{(NT)}  - \gamma i^{(NT)}; \quad \\
-\frac{\partial r_t}{\partial t} &= - \gamma i^{(NT)}_t
-\end{align*}
-where $B \in \mathbb{R}^{K \times K}$ and $\circ$ denotes elementwise product of vectors.
+%$$
+%\theta_t \mid \theta_{t-1}; \tau \sim \text{Dirichlet} \left(\kappa f(\theta_{t-1}, \beta, \gamma) \right)
+%$$
+%where $\kappa$ scales the variance of the Dirichlet distribution and $f(\cdot)$ is a function that determines infection dynamics.
+Here, we extend the SIR dynamics to account for strata-level dynamics
+%\begin{align*}
+%\frac{\partial s_t}{\partial t} &= - s_t \circ (B \, i^{(T)}_t); \quad \frac{\partial e_t}{\partial t} = s_t \circ (B \, i^{(T)}_t) - \sigma e_t; \quad \\
+%\frac{\partial i^{(T)}_t}{\partial t} &= \sigma e_t - \beta_2 i^{(T)}; \quad
+%\frac{\partial i^{(NT)}_t}{\partial t} = \beta_2 i^{(NT)}  - \gamma i^{(NT)}; \quad \\
+%\frac{\partial r_t}{\partial t} &= - \gamma i^{(NT)}_t
+%\end{align*}
+%where $B \in \mathbb{R}^{K \times K}$ and $\circ$ denotes elementwise product of vectors.
 The matrix $B$ allows for strata-specific infection dynamics with the SIR model being recovered if $B_{k,\cdot} = \beta$ for each $k \in [K]$.
 Following~\cite{Song2020}, we employ Runge-Kutta (RK4) approximations for discretization; see Appendix XX for details.
-
-Selection bias in case count data could be direcly addressed by extending the above compartmental model.
+%
+%Selection bias in case count data could be direcly addressed by extending the above compartmental model.
 Here, we instead opt to bring in a different source of information by considering COVID-19 death data.
@@
 Then
-$$
-D_{k,r} \mid p, \theta, \nu \sim \text{Poisson} \left( \sum_{t=1}^r p_{k} \nu_{k,t} \theta_{(r-t)} \right)
-$$
-where $p_{k}$ is the infection fatality rate for the $k$th strata.
+%$$
+%D_{k,r} \mid p, \theta, \nu \sim \text{Poisson} \left( \sum_{t=1}^r p_{k} \nu_{k,t} \theta_{(r-t)} \right)
+%$$
+%where $p_{k}$ is the infection fatality rate for the $k$th strata.
 This extends prior analysis of death data~\cite{Johndrow2020} by allowing the infection fatality rate to depend on age.
 Indeed,~\cite{Levin2020} demonstrate that ``the estimated IFR is close to zero for children and younger adults but rises exponentially with age, reaching 0.4\% at age 55, 1.4\% at age 65, 4.6\% at age 75, and 15\% at age 85.'' For conciseness, parameter choices and inference are discussed in depth in Appendix~\ref{app:modelbased}.
-
-\subsection{Doubly robust estimation.}
-Rather than relying solely on IPW or epidemiological forecasts, here we combine forecasting and inverse-probability weighting by extending recent work by~\cite{Chen2019} to account time-varying propensities and measurement-error.
+%
+%\subsection{Doubly robust estimation.}
+%Rather than relying solely on IPW or epidemiological forecasts, here we combine forecasting and inverse-probability weighting by extending recent work by~\cite{Chen2019} to account time-varying propensities and measurement-error.
 In this context, the doubly-robust estimator is given by
-$$
-\bar y_{n}^{(DR)} = \frac{1}{N} \sum_{j=1}^N \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \mu(x_j) \right).
-$$
-where the estimates $\hat \mu(x_j)$ are not corrected for measurement-error as they are built to estimate true case count from death-only data.
+%$$
+%\bar y_{n}^{(DR)} = \frac{1}{N} \sum_{j=1}^N \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \mu(x_j) \right).
+%$$
+%where the estimates $\hat \mu(x_j)$ are not corrected for measurement-error as they are built to estimate true case count from death-only data.
 This estimator is called ``doubly-robust'' because it is consistent if either the model-based forecasts or the time-varying propensities are correctly specified.
 A statistical error decomposition can be derived
-\begin{align}
-\label{eq:statdecomp3}
-\begin{split}
-\rho_{\tilde I (X), Y-\mu(X)} &\times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y-\mu(X)}  \\
-&\times \underbrace{\left[ 1 - \frac{\rho_{\tilde I(X),Y} \sigma_{Y}}{\rho_{\tilde I(X),Y-\mu(X)} \sigma_{Y-\mu(X)}} \times \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right]}_{\tilde D_M}.
-\end{split}
-\end{align}
-See Appendix~\ref{app:DRderivation} for the derivation.
+%\begin{align}
+%\label{eq:statdecomp3}
+%\begin{split}
+%\rho_{\tilde I (X), Y-\mu(X)} &\times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y-\mu(X)}  \\
+%&\times \underbrace{\left[ 1 - \frac{\rho_{\tilde I(X),Y} \sigma_{Y}}{\rho_{\tilde I(X),Y-\mu(X)} \sigma_{Y-\mu(X)}} \times \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right]}_{\tilde D_M}.
+%\end{split}
+%\end{align}
+%See Appendix~\ref{app:DRderivation} for the derivation.
 Comparing~\eqref{eq:statdecomp3} to~\eqref{eq:statdecomp2} shows that the addition of the model-based focast impacts the estimation error in two ways.
@@
 Note that the model-based forecast does not impact the data quantity component.
-
-If the model-based forecast is correct then $E_{\bf R} [ \rho_{\tilde I (X), Y-\mu(X)} ] = 0$  and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the forecasts are not correctly specified then the data quality index is unlikely to inversely scale with population size.
+%
+%If the model-based forecast is correct then $E_{\bf R} [ \rho_{\tilde I (X), Y-\mu(X)} ] = 0$  and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the forecasts are not correctly specified then the data quality index is unlikely to inversely scale with population size.
 Interestingly, the impact of measurement-error on data quality now depends on a relative comparison of the data quality and problem difficulty of the weighted estimator and the doubly-robust estimator.
@@
 Again, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from a DR estimator are likely to put too much confidence in the sheer size of the data.
-
-Unfortunately, the first term cannot be computed as covariate information is not collected on the entire population.
+%
+%Unfortunately, the first term cannot be computed as covariate information is not collected on the entire population.
 Here, we again use the probability samples and consider the doubly-robust estimator
-$$
-\frac{1}{\sum_{j=1}^N \tilde I_j \tilde W_j} \sum_{j=1}^N  \tilde I_j \tilde W_j \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \mu(x_j) \right).
-$$
-where $\tilde I_j$ indicates an individual is included in the random sample and $\tilde W_j$ is the associated weight for that unit.
+%$$
+%\frac{1}{\sum_{j=1}^N \tilde I_j \tilde W_j} \sum_{j=1}^N  \tilde I_j \tilde W_j \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \mu(x_j) \right).
+%$$
+%where $\tilde I_j$ indicates an individual is included in the random sample and $\tilde W_j$ is the associated weight for that unit.
 Assuming the selection propensity model is correctly specified, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:dr} shows that the IPW estimator is  consistent and calculates the estimator's variance.
 See Appendix~\ref{app:asympderivations} for additional details.
-
-\begin{lemma}[Variance of doubly-robust estimator] \normalfont
-\label{lemma:dr}
-\textcolor{red}{Must fix.}
-The estimates~$\mu_t := \bar y_{n,t}^\star$ and $\hat \pi_{j,t}$
-are solutions to the following set of estimating equations
-$$
-\Phi_n (\eta_t) =
-\left (
-\begin{array}{c}
-\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - (1-FP-FN) \cdot (\mu_t+FP)}{\pi_{j,t}} \\
-\frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
-\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
-\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
-\end{array}
-\right ) = {\bf 0}
-$$
-where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
+%
+%\begin{lemma}[Variance of doubly-robust estimator] \normalfont
+%\label{lemma:dr}
+%\textcolor{red}{Must fix.}
+%The estimates~$\mu_t := \bar y_{n,t}^\star$ and $\hat \pi_{j,t}$
+%are solutions to the following set of estimating equations
+%$$
+%\Phi_n (\eta_t) =
+%\left (
+%\begin{array}{c}
+%\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - (1-FP-FN) \cdot (\mu_t+FP)}{\pi_{j,t}} \\
+%\frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
+%\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
+%\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
+%\end{array}
+%\right ) = {\bf 0}
+%$$
+%where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
 Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$ (check)  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ where $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$.
-\end{lemma}
-
-\section{COVID-19 active infection prevalence in Indiana}
-\label{section:applications}
-
-% For COVID-19 case-count data, the only covariate information provided by state and local governments are summary statistics for those factors discussed in Section~\ref{section:testinginfo} -- age, sex,  and race.
+%\end{lemma}
+%
+%
+%\section{COVID-19 active infection prevalence in Indiana}
+%\label{section:applications}
+%
+%% For COVID-19 case-count data, the only covariate information provided by state and local governments are summary statistics for those factors discussed in Section~\ref{section:testinginfo} -- age, sex,  and race.
 These are discrete factors and therefore define a stratification of the population.
@@
 Recommendation XX below states that additional covariate information should be collected from nonprobabilistic samples.
-
-Here we consider estimation of the active infection rate in Indiana. We consider five sources of data:
-\begin{itemize}[leftmargin=*]
-\item {\bf Testing data}:  Each day, the number of tests performed and number of positive tests are reported, with counts broken out jointly by age, gender, and racial demographic information.
+%
+%Here we consider estimation of the active infection rate in Indiana. We consider five sources of data:
+%\begin{itemize}[leftmargin=*]
+%\item {\bf Testing data}:  Each day, the number of tests performed and number of positive tests are reported, with counts broken out jointly by age, gender, and racial demographic information.
 Figure~\ref{fig:in-cases} plots the number of COVID-19 positive cases based on testing date by age.
-
-\begin{figure}
-\centering
-\begin{subfigure}{.45\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{../figs/indianacasecounts.png}
-  \caption{Indiana case-counts per day}
-  \label{fig:in-cases}
-\end{subfigure}
-\begin{subfigure}{.45\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{../figs/indianadeaths.png}
-  \caption{Indiana death-counts per day}
-  \label{fig:in-deaths}
-\end{subfigure}
-\end{figure}
-
-\item {\bf Random/Nonrandom statewide sample}: Table~\ref{tab:indiana} summarizes data from a random sample from April 25--29th as well as a nonrandom sample obtained between May 2--3 in racial/ethnic minority communities (cite).
+%
+%\begin{figure}
+%\centering
+%\begin{subfigure}{.45\textwidth}
+%  \centering
+%  \includegraphics[width=.9\linewidth]{../figs/indianacasecounts.png}
+%  \caption{Indiana case-counts per day}
+%  \label{fig:in-cases}
+%\end{subfigure}
+%\begin{subfigure}{.45\textwidth}
+%  \centering
+%  \includegraphics[width=.9\linewidth]{../figs/indianadeaths.png}
+%  \caption{Indiana death-counts per day}
+%  \label{fig:in-deaths}
+%\end{subfigure}
+%\end{figure}
+%
+%\item {\bf Random/Nonrandom statewide sample}: Table~\ref{tab:indiana} summarizes data from a random sample from April 25--29th as well as a nonrandom sample obtained between May 2--3 in racial/ethnic minority communities (cite).
 The nonrandom sample is not a random subsample of the overall case-counts; however, this data provides important supplementary covariate information.
@@
 These adjusted rates are presented in parentheses in Table~\ref{tab:indiana}.
-
-\begin{table}[th]
-\begin{tabular}{c | c | r r | r | r r }
-& & \multicolumn{2}{c}{Total Tests (\%)} & US & \multicolumn{2}{c}{Positive Test Rate (\%)}\\
-\cline{3-4} \cline{6-7}
-& & NonRandom & Random & Census & NonRandom & Random \\ \hline
-\multirow{2}{*}{Sex} & Female & 58.2 & 55.0 & 50.7 & 21.7 (11.2) & 1.4 \\
-& Male & 41.8 & 45.0 & 49.3 & 24.2 (12.4) & 2.1 \\ \hline
-\multirow{3}{*}{Age} & $<40$ & 39.4 & 28.0 & 52.7 & 29.7 (15.0) & 1.7 \\
-& $40-59$ & 41.1 & 36.0 & 25.2 & 24.9 (12.5) & 2.1 \\
-& $\geq 60$ & 19.5 & 36.0 & 22 & 6.7 (3.4) & 0.9 \\ \hline
-\multirow{2}{*}{Race} & White & 23.1 & 92.0 & 86.9 & 19.5 (9.6) & 1.5 \\
-& Nonwhite & 76.9 & 8.0 & 13.1 & 25.0 (12.3) & 3.4 \\ \hline
-% Hispanic & Hispanic & 44.1 & 2.0 & 7.1 & 37.6 & 6.9 \\
-% Origin & Non-Hispanic & 55.9 & 98.0 & 92.9 & 13.0 & 1.3 \\ \hline
-\multirow{2}{*}{Fever} & Yes & 17.0 & 1.8 & - & 66.4 (32.1) & 4.5 \\
-& No & 83.0 & 98.2 & - & 15.6 (7.5) & 1.3 \\ \hline
-Household & Yes & 10.8 & 1.4 & - & 46.1 & 29.4 \\
-$+$ Case & No & 89.2 & 98.6 & - & 21.6 & 1.3 \\ \hline
-Prior $+$ & Yes & 6.1 & 1.4 & - & 39.2 & 24.4 \\
-Test & No & 93.9 & 98.6 & - & 21.6 & 1.3 \\ \hline
-\end{tabular}
-\caption{Estimated point prevalence of active infection with SARS-CoV-2 by demographics in Indiana.}
-\label{tab:indiana}
-\end{table}
-
-\item {\bf Facebook symptom survey}: We extract daily symptom surveys completed by individuals on Facebook who identify as living in Indiana.
+%
+%
+%\begin{table}[th]
+%\begin{tabular}{c | c | r r | r | r r }
+%& & \multicolumn{2}{c}{Total Tests (\%)} & US & \multicolumn{2}{c}{Positive Test Rate (\%)}\\
+%\cline{3-4} \cline{6-7}
+%& & NonRandom & Random & Census & NonRandom & Random \\ \hline
+%\multirow{2}{*}{Sex} & Female & 58.2 & 55.0 & 50.7 & 21.7 (11.2) & 1.4 \\
+%& Male & 41.8 & 45.0 & 49.3 & 24.2 (12.4) & 2.1 \\ \hline
+%\multirow{3}{*}{Age} & $<40$ & 39.4 & 28.0 & 52.7 & 29.7 (15.0) & 1.7 \\
+%& $40-59$ & 41.1 & 36.0 & 25.2 & 24.9 (12.5) & 2.1 \\
+%& $\geq 60$ & 19.5 & 36.0 & 22 & 6.7 (3.4) & 0.9 \\ \hline
+%\multirow{2}{*}{Race} & White & 23.1 & 92.0 & 86.9 & 19.5 (9.6) & 1.5 \\
+%& Nonwhite & 76.9 & 8.0 & 13.1 & 25.0 (12.3) & 3.4 \\ \hline
+%% Hispanic & Hispanic & 44.1 & 2.0 & 7.1 & 37.6 & 6.9 \\
+%% Origin & Non-Hispanic & 55.9 & 98.0 & 92.9 & 13.0 & 1.3 \\ \hline
+%\multirow{2}{*}{Fever} & Yes & 17.0 & 1.8 & - & 66.4 (32.1) & 4.5 \\
+%& No & 83.0 & 98.2 & - & 15.6 (7.5) & 1.3 \\ \hline
+%Household & Yes & 10.8 & 1.4 & - & 46.1 & 29.4 \\
+%$+$ Case & No & 89.2 & 98.6 & - & 21.6 & 1.3 \\ \hline
+%Prior $+$ & Yes & 6.1 & 1.4 & - & 39.2 & 24.4 \\
+%Test & No & 93.9 & 98.6 & - & 21.6 & 1.3 \\ \hline
+%\end{tabular}
+%\caption{Estimated point prevalence of active infection with SARS-CoV-2 by demographics in Indiana.}
+%\label{tab:indiana}
+%\end{table}
+%
+%\item {\bf Facebook symptom survey}: We extract daily symptom surveys completed by individuals on Facebook who identify as living in Indiana.
 We collect age, sex, and demographic information as well as COVID-19 related symptoms, which Facebook defines as having (1) a fever and cough, (2) shortness of breath, \emph{or} (3) difficulty breathing.
@@
 Table~\ref{tab:comparison} demonstrates minimal bias in the Facebook symptom survey.
-\begin{table}
-\centering
-\begin{tabular}{c | c c c | c}
-& \multicolumn{3}{c}{Facebook} & Random \\ \cline{2-5}
-& Lower CI & Estimate & Upper CI & Estimate \\ \hline
-Fever &  0.017 & 0.020 & 0.024 & 0.018 \\
-Cough &  0.142 & 0.148 & 0.154 & 0.149 \\
-Shortness & 0.058 & 0.060 & 0.062 & 0.062 \\ \hline
-\end{tabular}
-\caption{Comparison of Facebook symptom survey and random sample demonstrates minimal bias.}
-\label{tab:comparison}
-\end{table}
-
-% \item {\bf Hospitalization data}: On each day, we observe the number of individuals who are COVID-19 positive that visited a hospital on that day.
+%\begin{table}
+%\centering
+%\begin{tabular}{c | c c c | c}
+%& \multicolumn{3}{c}{Facebook} & Random \\ \cline{2-5}
+%& Lower CI & Estimate & Upper CI & Estimate \\ \hline
+%Fever &  0.017 & 0.020 & 0.024 & 0.018 \\
+%Cough &  0.142 & 0.148 & 0.154 & 0.149 \\
+%Shortness & 0.058 & 0.060 & 0.062 & 0.062 \\ \hline
+%\end{tabular}
+%\caption{Comparison of Facebook symptom survey and random sample demonstrates minimal bias.}
+%\label{tab:comparison}
+%\end{table}
+%
+%% \item {\bf Hospitalization data}: On each day, we observe the number of individuals who are COVID-19 positive that visited a hospital on that day.
 Unfortunately, hospitalization data is not broken out by age, gender, and racial demographic information.
@@
 We then use this to figure out how many individuals per strata were hospitalized and then back out how many
-
-\item {\bf Death data}: On each day, we observe the number of deaths due to COVID-19 reported on that day, with deaths broken out jointly by age, gender, and racial demographic information.
+%
+%\item {\bf Death data}: On each day, we observe the number of deaths due to COVID-19 reported on that day, with deaths broken out jointly by age, gender, and racial demographic information.
 Figure~\ref{fig:in-deaths} plots the cumulative number of COVID-19 related deaths over time by age group.
-\end{itemize}
-
-\subsection{Inverse-probability weighting approach}
-
-We start with using Table~\ref{tab:indiana} to compute IPW weights.
+%\end{itemize}
+%
+%
+%\subsection{Inverse-probability weighting approach}
+%
+%We start with using Table~\ref{tab:indiana} to compute IPW weights.
 The random sample is $n=3,658$.
@@
 For example, the strata defined as Male, 40--59, White, has a Fever, no household cases, and no prior tests has a weight proportional to
-$$
-\frac{\left( 0.493 \times 0.252 \times 0.869 \times 0.018 \right)}{\left(0.418 \times 0.411 \times 0.231 \times 0.17 \right)} = 0.628
-$$
-Using the constructed weights and the adjusted strata-level prevalence estimates, the IPW estimate under no measurement-error is $7.7$, a decrease of four percentage points from the observed prevalence of $11.7$.
+%$$
+%\frac{\left( 0.493 \times 0.252 \times 0.869 \times 0.018 \right)}{\left(0.418 \times 0.411 \times 0.231 \times 0.17 \right)} = 0.628
+%$$
+%Using the constructed weights and the adjusted strata-level prevalence estimates, the IPW estimate under no measurement-error is $7.7$, a decrease of four percentage points from the observed prevalence of $11.7$.
 Using Table~\ref{tab:indiana} to compute weights requires the assumption of independence among factors.
 To alleviate this, we compute weights using Facebook's survey data collected in Indiana over that window of time; the resulting IPW estimate under no measurement-error is
-
-The study did not report sensitivity and specificity; therefore, we take the suggested measures from~\cite{Woloshin2020} which report 70\% sensitivity is a reasonable estimates and~\cite{Cohen2020} which report 97.6\% specificity.
+%
+%The study did not report sensitivity and specificity; therefore, we take the suggested measures from~\cite{Woloshin2020} which report 70\% sensitivity is a reasonable estimates and~\cite{Cohen2020} which report 97.6\% specificity.
 This corresponds to a false negative rate of $30\%$ and false positive rate of $2.4\%$.
@@
 To account for uncertainty in
-
-Indiana's population as of 2019 was $6.732$ million.
+%
+%Indiana's population as of 2019 was $6.732$ million.
 Subtracting off the $617,555$ tests that had already performed yields a sampling fraction of $f = 0.001$.
@@
 Adjusting for imperfect testing yields an estimate of $33.9\%$.
-
-For this estimate, we consider estimation of data quality $E_{\I} \left[ \rho_{\tilde I(X),Y} \right]$.
+%
+%
+%
+%For this estimate, we consider estimation of data quality $E_{\I} \left[ \rho_{\tilde I(X),Y} \right]$.
 In \cite{Meng2018}, estimation relied on observing the true outcome (i.e., election vote totals), which is not possible in the current crisis.
@@
 Instead, we aim to build sensitivity analyses that can aid in understanding the amount of information in observational COVID case-count data.
-
-Under the assumption that the survey is a SRS from New York, the error is $17.6$\%.  Using~\eqref{eq:error}, we construct an estimate of the relative sampling rate as follows
-\[
-\rho D_M = \sqrt{\frac{f}{1-f}} \frac{\text{0.176}}{\sigma_Y} = 1.07 \times 10^{-2} \Rightarrow \Delta = 1.07 \times 10^{-3} \Rightarrow M = 2.3.
-\]
-We can perform a sensitivity analysis by considering the range of false negative and positive rates from~\cite{Bendavid2020}, which leads to a range for $M$ of $(2.21, 2.38)$.
+%
+%
+%Under the assumption that the survey is a SRS from New York, the error is $17.6$\%.  Using~\eqref{eq:error}, we construct an estimate of the relative sampling rate as follows
+%\[
+%\rho D_M = \sqrt{\frac{f}{1-f}} \frac{\text{0.176}}{\sigma_Y} = 1.07 \times 10^{-2} \Rightarrow \Delta = 1.07 \times 10^{-3} \Rightarrow M = 2.3.
+%\]
+%We can perform a sensitivity analysis by considering the range of false negative and positive rates from~\cite{Bendavid2020}, which leads to a range for $M$ of $(2.21, 2.38)$.
 Therefore, the impact on bias of $\hat R_t$ appears moderate.
@@
 This simple tool helps us understand how sensitivity and specificity impact biases and  therefore how much we should trust conclusions based on these assumptions.
-
-\subsubsection{Time-varying IPW estimator}
-
-\emph{Need to show Hospitalization C+ versus Covid count to show that most of the tests are in that subset for early parts.}
-
-\subsection{Model-based estimation}
-
-\subsection{Doubly robust estimation}
-
-% \begin{figure}
-% \centering
-% \begin{subfigure}{.5\textwidth}
-%   \centering
-%   \includegraphics[width=.9\linewidth]{../methods/figs/NY_casecount.png}
-%   \caption{NY new cases per day}
-%   \label{fig:ny-covid-test}
-% \end{subfigure}%
-% \begin{subfigure}{.5\textwidth}
-%   \centering
-%   \includegraphics[width=.9\linewidth]{../methods/figs/NY_fracpos.png}
-%   \caption{NY fraction of new tests positive per day}
-%   \label{fig:ny-covid-frac}
-% \end{subfigure}
-% \caption{New York COVID-19 reported data}
-% \label{fig:ny-covid}
-% \end{figure}
-
-% \section{Cross-population comparisons}
-% \label{section:crosspop}
-% So far we have focused on understanding the limitations of using case-count data to understand population quantities of interest for a \emph{single} population.
+%
+%
+%\subsubsection{Time-varying IPW estimator}
+%
+%\emph{Need to show Hospitalization C+ versus Covid count to show that most of the tests are in that subset for early parts.}
+%
+%
+%
+%
+%
+%
+%\subsection{Model-based estimation}
+%
+%\subsection{Doubly robust estimation}
+%
+%% \begin{figure}
+%% \centering
+%% \begin{subfigure}{.5\textwidth}
+%%   \centering
+%%   \includegraphics[width=.9\linewidth]{../methods/figs/NY_casecount.png}
+%%   \caption{NY new cases per day}
+%%   \label{fig:ny-covid-test}
+%% \end{subfigure}%
+%% \begin{subfigure}{.5\textwidth}
+%%   \centering
+%%   \includegraphics[width=.9\linewidth]{../methods/figs/NY_fracpos.png}
+%%   \caption{NY fraction of new tests positive per day}
+%%   \label{fig:ny-covid-frac}
+%% \end{subfigure}
+%% \caption{New York COVID-19 reported data}
+%% \label{fig:ny-covid}
+%% \end{figure}
+%
+%
+%% \section{Cross-population comparisons}
+%% \label{section:crosspop}
+%% So far we have focused on understanding the limitations of using case-count data to understand population quantities of interest for a \emph{single} population.
 Many are interested in cross-population comparisons to contrast the impact of countries' mitigation policies.
@@
 In this section, we focus on the statistical issues relevant to such comparisons.
-
-% \subsection{Prevalence comparisons}
-
-% At a fixed time, suppose prevalence estimates $\bar y_1$ and $\bar y_2$ are observed for two populations. We can express $\bar y_1 - \bar y_2$ in statistical terms as
-% $$
-% (\bar Y_1 - \bar Y_2) + \left[ \rho_{I_1, Y_1} \times D_{M_1} \times \sqrt{\frac{1-f_1}{f_1}} \times \sigma_{Y_1}  - \rho_{I_2, Y_2} \times D_{M_2} \times \sqrt{\frac{1-f_2}{f_2}} \times \sigma_{Y_2} \right].
-% $$
-% The first term is the difference in population prevalence (i.e., the quantity of interest).
+%
+%% \subsection{Prevalence comparisons}
+%
+%% At a fixed time, suppose prevalence estimates $\bar y_1$ and $\bar y_2$ are observed for two populations. We can express $\bar y_1 - \bar y_2$ in statistical terms as
+%% $$
+%% (\bar Y_1 - \bar Y_2) + \left[ \rho_{I_1, Y_1} \times D_{M_1} \times \sqrt{\frac{1-f_1}{f_1}} \times \sigma_{Y_1}  - \rho_{I_2, Y_2} \times D_{M_2} \times \sqrt{\frac{1-f_2}{f_2}} \times \sigma_{Y_2} \right].
+%% $$
+%% The first term is the difference in population prevalence (i.e., the quantity of interest).
 If random sampling were performed in each population, $E_{\I} [ \rho_{I_j, Y_j}] = 0$ and so the estimate would be unbiased.
 However, the second term represents the complex error that results from selection bias.
-% The particular test of interest is $H_0: \bar Y_1 = \bar Y_2$.  Under the null, $\sigma_{Y_1} = \sigma_{Y_2}$.  In classical statistics, we would compute a Z-score:
-% $$
-% \begin{aligned}
-% \frac{\bar y_1 - \bar y_2}{ \sqrt{V_{SRS} (\bar y_1 - \bar y_2)} } &=
-% \frac{\rho_{I_1, Y_1} \times D_{M_1} \times \sqrt{\frac{1-f_1}{f_1}}  - \rho_{I_2, Y_2} \times D_{M_2} \times \sqrt{\frac{1-f_2}{f_2}} }{ \sqrt{\frac{1}{N_1 -1} \times \frac{1-f_1}{f_1} + \frac{1}{N_2 - 1} \times \frac{1-f_2}{f_2}} }  \\
-% &\overset{\star}{=}
-% \sqrt{\frac{(N_1-1)(N_2 -1)}{N_1 + N_2}} \left[ \rho_{I_1, Y_1} - \rho_{I_2, Y_2} \right]
-% \end{aligned}
-% $$
-% Where the last equality holds under $f_1 = f_2$, i.e., same sampling rate relative to population size.
+%% The particular test of interest is $H_0: \bar Y_1 = \bar Y_2$.  Under the null, $\sigma_{Y_1} = \sigma_{Y_2}$.  In classical statistics, we would compute a Z-score:
+%% $$
+%% \begin{aligned}
+%% \frac{\bar y_1 - \bar y_2}{ \sqrt{V_{SRS} (\bar y_1 - \bar y_2)} } &=
+%% \frac{\rho_{I_1, Y_1} \times D_{M_1} \times \sqrt{\frac{1-f_1}{f_1}}  - \rho_{I_2, Y_2} \times D_{M_2} \times \sqrt{\frac{1-f_2}{f_2}} }{ \sqrt{\frac{1}{N_1 -1} \times \frac{1-f_1}{f_1} + \frac{1}{N_2 - 1} \times \frac{1-f_2}{f_2}} }  \\
+%% &\overset{\star}{=}
+%% \sqrt{\frac{(N_1-1)(N_2 -1)}{N_1 + N_2}} \left[ \rho_{I_1, Y_1} - \rho_{I_2, Y_2} \right]
+%% \end{aligned}
+%% $$
+%% Where the last equality holds under $f_1 = f_2$, i.e., same sampling rate relative to population size.
 This tells us that the exact error in the comparison of sample means, as an estimate of the difference in population means, scales as a function of the difference in data quality $\rho_{I_1, Y_1} - \rho_{I_2, Y_2}$ times a square root of a population adjustment.
@@
 $\sqrt{N_2 -1}$).
-
-% Take, for example, a comparison of the US and Canada.
+%
+%% Take, for example, a comparison of the US and Canada.
 The two population sizes are approximately 328 million and 38 million people.
@@
 That is, setting
-% $$
-% Z_n = \frac{\bar y_1 - \bar y_2}{ \frac{1-f}{f} \sigma_y \sqrt{ \frac{1}{n_{1,eff} - 1} + \frac{1}{n_{2,eff} - 1}}}
-% $$
-% allows us to build Z-scores that account for the true sample size.
+%% $$
+%% Z_n = \frac{\bar y_1 - \bar y_2}{ \frac{1-f}{f} \sigma_y \sqrt{ \frac{1}{n_{1,eff} - 1} + \frac{1}{n_{2,eff} - 1}}}
+%% $$
+%% allows us to build Z-scores that account for the true sample size.
 While still not perfect, this allows us to build Z-scores that are not overly confident due to selection bias and measurement error.
 See Section~\ref{section:est_dq} for discussion of how to estimate data quality.
-
-\section{Discussion}
-\label{section:discussion}
-
-{\bf Our goal with surveillance is to identify areas of weakness; but selection bias is an issue!}
-
-There is nothing routine about COVID-19, including the corresponding statistical questions.
+%
+%\section{Discussion}
+%\label{section:discussion}
+%
+%{\bf Our goal with surveillance is to identify areas of weakness; but selection bias is an issue!}
+%
+%There is nothing routine about COVID-19, including the corresponding statistical questions.
 The goal of this paper was to point out questionable statistical routines.
@@
 We end with a brief discussion of important related topics.
-
-\subsection*{Additional biases and measurement-error}
-
-Length bias and measurement-error?
-
-\subsection*{Decision-making: data versus information}
-
-Having read through the above technical discussions, one may come to the incorrect conclusion that because we are concerned about implicit biases in observational data, we must wait for better data sources in order to act~\citep{Ioannidis2020}.
+%
+%\subsection*{Additional biases and measurement-error}
+%
+%Length bias and measurement-error?
+%
+%\subsection*{Decision-making: data versus information}
+%
+%Having read through the above technical discussions, one may come to the incorrect conclusion that because we are concerned about implicit biases in observational data, we must wait for better data sources in order to act~\citep{Ioannidis2020}.
 This conclusion confuses data and information.
@@
 A full discussion of these topics is beyond the scope of the current article.
-
-\subsection*{Data quantity versus quality}
-
-Governments implicitly argue that increased testing capacity will perform the above task.
+%
+%\subsection*{Data quantity versus quality}
+%
+%Governments implicitly argue that increased testing capacity will perform the above task.
 Without complete compliance, however, our understanding of future outbreaks may be plagued by self-selection bias, compounded by changing sensitivity and specificity rates.
@@
 This is not to say random testing should replace current testing practices; only that the random testing should supplement and will provide valuable additional information even with moderate sample sizes.
-
-\subsection*{High quality information on prevalence.}
-
-Governments are pushing for increased testing capacity and robust contact tracing.
+%
+%\subsection*{High quality information on prevalence.}
+%
+%Governments are pushing for increased testing capacity and robust contact tracing.
 Contact tracing is incredibly useful for identifying carriers early and preventing spread of the disease.
@@
 Knowing prevalence will help governments determine long-term community-level risk and allow for more targeted interventions -- shutting down only certain locations when necessary.
-
-\subsection*{Stratified sampling: improving precision in low-prevalence environments.}
-Random sampling is a powerful tool.
+%
+%\subsection*{Stratified sampling: improving precision in low-prevalence environments.}
+%Random sampling is a powerful tool.
 One difficulty with SRS is when prevalence is low.
@@
 One can, for example, stratify the population into risk categories (e.g., based on population density, occupation, age) and perform Neyman allocation~\citep{Cochran77}.
-
-\subsection*{Model-based solutions.}
-
-One can extend the SIR model to account for selection bias and measurement error.
+%
+%\subsection*{Model-based solutions.}
+%
+%One can extend the SIR model to account for selection bias and measurement error.
 Without strong assumptions on the selection mechanism, however, the estimates are often not identifiable.
@@
 Absent random sampling, the best route forward for all data analyses is careful associated sensitivity analyses and humility in data-driven conclusions.
-
-\subsection*{Random sampling for doubly robust estimation}
-
-Selection rates are likely time-varying; therefore, we need rapid survey samples to properly model the selection criteria for doubly robust inferential procedures.
-
-COLLECT COVARIATE INFORMATION (even aggregate) and provide it in accessible formats.
-
-- Easy access to history of testing requirements
-- Smaller random samples over time (at least as testing and prevalence changes drastically) to assess selection changes
-- Record symptom/asymptomatic status and COVID contact.  If not for all, then for at least some we need these covariates!!!!!
-
-\subsection*{Systematic data collection and reporting}
-
-I would say this is important but JHU is not enough; see above for benefits of random sampling to augment and improve decision making.
-
-\subsection*{Frequency and turnaround time}
-
-This supposes no selection bias.
-
-What is your goal?
+%
+%\subsection*{Random sampling for doubly robust estimation}
+%
+%Selection rates are likely time-varying; therefore, we need rapid survey samples to properly model the selection criteria for doubly robust inferential procedures.
+%
+%COLLECT COVARIATE INFORMATION (even aggregate) and provide it in accessible formats.
+%
+%- Easy access to history of testing requirements
+%- Smaller random samples over time (at least as testing and prevalence changes drastically) to assess selection changes
+%- Record symptom/asymptomatic status and COVID contact.  If not for all, then for at least some we need these covariates!!!!!
+%
+%\subsection*{Systematic data collection and reporting}
+%
+%I would say this is important but JHU is not enough; see above for benefits of random sampling to augment and improve decision making.
+%
+%\subsection*{Frequency and turnaround time}
+%
+%This supposes no selection bias.
+%
+%
+%
+%
+%What is your goal?
 "get people who are currently sick off the street." Then test frequently and often.
@@
 Random testing is one way to improve upone current practice whiel we wait for that
-
-{\bf COORDINATION ON DATA SOURCES}
+%
+%{\bf COORDINATION ON DATA SOURCES}
 

### 2021-06-22 methods/8_death_modeling.R
Death version prior to August that WORKS!

@@ -15,7 +15,7 @@ c_prior = "aquamarine2"
 df_swiss$date = ymd(df_swiss$startdate)
 df_swiss$death_dt = df_swiss$covid_deaths
 
-df_swiss = df_swiss[df_swiss$date < "2020-07-01",] # single bump to start
+df_swiss = df_swiss[df_swiss$date < "2020-07-15",] # single bump to start
 
 df_swiss %>% 
   ggplot() + 
@@ -46,7 +46,7 @@ t <- seq(1, n_days+max_death_day, by = 1)
 t0 = 0
 t <- t
 
-date_switch <- "2020-03-19" # date of introduction of control measures (empirical)
+date_switch <- "2020-04-01" # date of introduction of control measures (empirical)
 tswitch <- df_swiss %>% filter(date < date_switch) %>% nrow() + 1 # convert time to number
 # date_switch <- "2020-09-25" # date of ending of control measures
 # tswitch <- df_swiss %>% filter(date < date_switch) %>% nrow() + 1 # convert time to number
@@ -60,12 +60,12 @@ fit_forcing <- sampling(model_forcing,
                         data_forcing, 
                         iter=1000,
                         control = list(max_treedepth = 13, adapt_delta=0.9),
-                        seed=1,
+                        seed=2,
                         chains = 1)
 
 check_hmc_diagnostics(fit_forcing)
 
-pairs(fit_forcing, pars = c("beta", "gamma", "a", "exp_eta", "nu", "xi", "phi"))
+pairs(fit_forcing, pars = c("beta", "gamma", "a", "eta", "nu", "xi", "phi"))
  
 smr_pred <- cbind(as.data.frame(summary(fit_forcing, pars = "pred_cases", probs = c(0.025, 0.05, 0.1, 0.5, 0.9, 0.95, 0.975))$summary[(max_death_day+1):(max_death_day+length(deaths)-1),]),
                                 t=1:(n_days-1))

### 2021-06-22 write-up/covid-writeup.tex
Fixing theory

--- 2021-06-16 write-up/covid-writeup.tex
@@
 
-\subsection{Diagnostic versus serologic testing}
+\subsection{Diagnostic testing}
 \label{section:testinginfo}
@@
 
-\emph{Serological tests} look for an immunological response to the virus.
+% \emph{Serological tests} look for an immunological response to the virus.
 A week or so after an individual is infected with SARS-CoV-2, the individual will start producing antibodies.
@@
 
-To better estimate SARS-CoV-2 immunity in a population, seroprevalence studies that generate a probabilistic population sample and perform serological tests on the sample can be collected.
+% To better estimate SARS-CoV-2 immunity in a population, seroprevalence studies that generate a probabilistic population sample and perform serological tests on the sample can be collected.
 These studies are useful for disease surveillance.
@@
 
-Between April 25--29, 2020, Indiana conducted statewide random sample testing (both molecular and serological) of persons ages $\geq 12$ years to assess prevalence of active infection and presence of antibodies to SARS-CoV-2 (cite: Indiana report).
+Due to testing restrictions and other potential selection biases, publicly reported covid testing data may not be sufficient to understand active infection rates or the disease trajectory.
+Here, we discuss two random samples that can be leveraged to get either unbiased estimates of the statewide active infection rate or as auxiliary information to address selection bias in the testing data.
+
+\subsubsection{Random statewide testing}
+
+Between April 25--29, 2020, Indiana conducted statewide random molecular testing of persons ages $\geq 12$ years to assess prevalence of active infection to SARS-CoV-2~\citep{Yiannoutsos2021}.
 A stratified random sampling design was conducted using Indiana’s 10 public health preparedness districts as sampling strata.
@@
 During May 2--3, 2020, an additional nonrandom sample of 898 individuals was also collected.
-In this paper, the nonrandom sample is used to assess whether the cumulative summary statistics are appropriate in constructing the inverse-probability weights as well as in constructing appropriate weights using symptom information.
+% In this paper, the nonrandom sample is used to assess whether the cumulative summary statistics are appropriate in constructing the inverse-probability weights as well as in constructing appropriate weights using symptom information.
 
 \subsubsection{Facebook/CMU symptom survey.}
-Since April 2020, Facebook has started randomly showing an opt-in symptom survey to its users.
-The survey is run by health researchers at Carnegie Mellon University.
+Since April 2020, Facebook has started randomly showing an opt-in symptom survey to its users~\citep{delphisurvey}.
 Data collected includes basic demographic information and if the respondent has symptoms such as fever, coughing, shortness of breath, or loss of smell which are associated with COVID-19.
@@
 Figure~\ref{fig:fbsymptoms} displays nonparametric regression estimates of the fraction of individuals within age and gender strata.
+In this paper, we use this as the main source of auxiliary information on time-varying characteristics of the population of Indiana, e.g., fraction of popopulation displaying symptoms or in contact with COVID-19 positive individuals).
 
@@
 Let $N$ denote the population size.
-For state-level analysis $N$ is the state's total population, while for country-level analysis $N$ is the country's total population.
-At a fixed moment in time, let $Y_j$ denote COVID-19 status for the $j$th individual in the population, $j=1,\ldots, N$.
+At a given time, let $Y_j$ denote COVID-19 status for the $j$th individual in the population, $j=1,\ldots, N$.
 Here, like in survey methodology~\citep{Cochran77}, we treat COVID-19 status as a fixed but unknown quantity of interest.
@@
 We assume either individual $j$ is COVID-19 positive and $Y_j=1$ or is COVID-19 negative and $Y_j=0$.
-We also let $I_j \in \{0,1\}$ be an indicator that the individual was selected for testing ($I = 1$) or not ($I=0$).
-
-To start, we assume the primary questions of interest are estimates of the overall number of COVID-19 cases and/or disease prevalence.
+We also let $I_j \in \{0,1\}$ be an indicator that the individual was tested ($I = 1$) or not ($I=0$).
+
+To start, we assume the overall number of COVID-19 cases and/or disease prevalence are of primary interest.
 That is, we are interested in either the population total $Y = \sum_{j=1}^N Y_j$ or the population average $\bar Y = Y/N$.
@@
 Tests are imperfect.
-COVID-19 testing is no exception.
+As discussed in Section~\ref{section:testinginfo}, COVID-19 testing is no exception.
 Here we investigate the interplay between imperfect testing and selection bias.
@@
 \includegraphics[width = 0.4\textwidth]{../figs/mem_heatmap_article.png}
-\caption{Imperfect testing adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.005$ and $FN=0.172$. Color scaled so blue = $-10$, white = $0$, and red = $10$.}
+\caption{Imperfect testing adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.005$ and $FN=0.172$.
+Color scaled so blue = $-10$, white = $0$, and red = $10$.
+\textcolor{red}{NEED TO DO FOR FP = 0.05 and FN = 0.3}}
 \label{fig:heatmap}
@@
 $$
-As of June 10th, the US has performed $21,474,531$ tests.
+As of June 10th 2020, the US has performed $21,474,531$ tests.
 The US population is roughly $312$ million, so $f = 0.069$.
-The cumulative empirical prevalence is $9.3\%$ on May 8th.
+\textcolor{red}{FIX: The cumulative empirical prevalence is $9.3\%$ on May 8th.
 A recent study estimates a false positive and negative rate for antibody testing at $0.5$\% and $17.2$\% respectively \citep{Bendavid2020}.
-Supposing COVID-19 positive individuals are $2$ times more likely to get tested, then the effective sample size is $92$.
-Recent proposals~\citep{Siddarth2020} have argued for increased testing capacity, which may likely reduce the relative sampling rate.
+Supposing COVID-19 positive individuals are $2$ times more likely to get tested, then the effective sample size is $92$.Recent proposals~\citep{Siddarth2020} have argued for increased testing capacity, which may likely reduce the relative sampling rate.
 If the relative sampling rate drops to $1.2$ then the effective sample size will increase to $1950$.
@@
 Consider the case where $f$ increases from $0.05$ to $0.1$ and the increase is associated with the false negative rate rising from $5\%$ to $20$\%.
-Then the effective sample size increases by a factor of $1.76$ rather than the expected $4$ fold increase.
+Then the effective sample size increases by a factor of $1.76$ rather than the expected $4$ fold increase.}
 % \begin{wrapfigure}{r}{0.5\textwidth}
@@
 In this section, we consider two directions.
-% First, we use publicly available auxiliary information to build selection propensities and consider an inverse probability weighting (IPW) estimator.
-We then extend the statistical error decomposition to the weighted setting.
-Second, we consider an extension of compartmental models from infectious disease epidemiology using death data.
-After adjusting for measurement-error, this leads to an alternative, model-based forecast of prevalence.
-We then combine these epidemiological forecasts, selection propensities, and measurement-error corrections to generate \emph{doubly robust} prevalence estimators.
-We provide an appropriate extension of the statistical error decomposition.
 The proposed methods and statistical error decompositions guide our recommendations in Section~\ref{section:discussion}.
@@
 
-Here, we assume access to a probability sample from the same population on which the outcome and exact same set of covariates are measured.
+Here we assume access to a probability sample measuring both the outcome and exact same set of covariates.
 Let $\tilde I_j$ be an indicator that the individual was included in the probability sample and $\tilde W_j$ be the probability of inclusion for the $j$th individual (i.e., equal to $n/N$ for a simple random sample of size $n$).
@@
 Taking the expectation of the second term with respect to the sampling design yields the second term in~\eqref{eq:propensity}.
-This was proposed by~\cite{Chen2019} when covariates were observed on a random sample from the population.
+This extends the work by~\cite{Chen2019} which considered auxiliary information coming only from a simple random sample.
 % Under simple random sampling and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the $\hat \pi_k = (d_k / \tilde d_k) (n / N)$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.
@@
 where equality (2) is under the assumption that $x_i$ is a stratification variable with $k$ indexing the strata, $w_k$ is the weight and $d_k$ is the number of samples in strata $k$, and $w = \sum_{k=1}^K d_k w_k$.
-Under simple random sampling with one strata~$\hat \pi_k = n/N$ and we recover the estimator $\tilde y_n$ from Section~\ref{section:imperfecttesting}.
+% Under simple random sampling with one strata~$\hat \pi_k = n/N$ and we recover the estimator $\tilde y_n$ from Section~\ref{section:imperfecttesting}.
 
@@
 Hence, if the data quality does not increase (i.e., $| \rho_{\tilde I (X), Y} | = | \rho_{I,Y}|$ ) then weighting increases the error magnitude.
-Note this relative error increase depends on the fraction of population sampled $f$, implying that for large samples there is a larger potential increase in the error if the weights do not improve data quality.
+Second, the relative error increase depends on the fraction of population sampled $f$, implying that for large samples there is a larger potential increase in the error if the weights do not improve data quality.
 Third, the impact of measurement-error on data quality is changed when considering a weighted estimand.
@@
 
-As demonstrated in Lemma~\ref{lemma:ipw}, if the propensity model is correctly specified then the $E_{\bf R} [ \rho_{\tilde I (X), Y} ] = 0$ and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then the data quality index is unlikely to inversely scale with population size.
+As demonstrated below in Lemma~\ref{lemma:ipw}, if the propensity model is correctly specified then the $E_{\bf R} [ \rho_{\tilde I (X), Y} ] = 0$ and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then the data quality index is unlikely to inversely scale with population size.
 Indeed, Horvitz-Thompson estimators are sensitive to errors in the estimated weights.
@@
 Moreover, an individual in a high prevalence area may be more likely to seek out testing.
-Here, we assume that the covariate vector $X_{j,t}$ contains all features of past relevant for selection.
+Here, we assume that the covariate vector $X_{j,t}$ contains all features of the past relevant for selection.
 
@@
 
-\subsubsection{Uncertain sensitivity and specificity}
-
-For prevalence there is some
-
-For
-
-For seroprevalence, \cite{Bendavid2020} report
-
-, which leads to a range for $M$ of $(2.21, 2.38)$.
+%\subsubsection{Uncertain sensitivity and specificity}
+%
+%For prevalence there is some
+%
+%For
+%
+%For seroprevalence, \cite{Bendavid2020} report
+%
+%
+%
+%, which leads to a range for $M$ of $(2.21, 2.38)$.
 Therefore, the impact on bias of $\hat R_t$ appears moderate.
@@
 This simple tool helps us understand how sensitivity and specificity impact biases and  therefore how much we should trust conclusions based on these assumptions.
-
+%
+%
 \subsubsection{Asymptotics}
@@
 \textcolor{red}{Must check.}
-The estimates~$\mu_t := \bar y_{n,t}^\star$ and $\hat \pi_{j,t}$
-are solutions to the following set of estimating equations
+The estimates~$\hat \mu_t := \bar y_{n,t}^\star$, $\hat \pi_{j,t}$, $\hat FP$, and $\hat FN$ are solutions to the following set of estimating equations
 $$
@@
 \begin{array}{c}
-\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - (1-FP-FN) \cdot (\mu_t+FP)}{\pi_{j,t}} \\
+\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t}{\pi_{j,t}} \\
 \frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
@@
 
-\subsection{Doubly robust estimation.}
-Rather than relying solely on IPW or epidemiological forecasts, here we combine forecasting and inverse-probability weighting by extending recent work by~\cite{Chen2019} to account time-varying propensities and measurement-error.
+%\subsection{Doubly robust estimation.}
+%Rather than relying solely on IPW or epidemiological forecasts, here we combine forecasting and inverse-probability weighting by extending recent work by~\cite{Chen2019} to account time-varying propensities and measurement-error.
 In this context, the doubly-robust estimator is given by
-$$
-\bar y_{n}^{(DR)} = \frac{1}{N} \sum_{j=1}^N \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \mu(x_j) \right).
-$$
-where the estimates $\hat \mu(x_j)$ are not corrected for measurement-error as they are built to estimate true case count from death-only data.
+%$$
+%\bar y_{n}^{(DR)} = \frac{1}{N} \sum_{j=1}^N \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \mu(x_j) \right).
+%$$
+%where the estimates $\hat \mu(x_j)$ are not corrected for measurement-error as they are built to estimate true case count from death-only data.
 This estimator is called ``doubly-robust'' because it is consistent if either the model-based forecasts or the time-varying propensities are correctly specified.
 A statistical error decomposition can be derived
-\begin{align}
-\label{eq:statdecomp3}
-\begin{split}
-\rho_{\tilde I (X), Y-\mu(X)} &\times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y-\mu(X)}  \\
-&\times \underbrace{\left[ 1 - \frac{\rho_{\tilde I(X),Y} \sigma_{Y}}{\rho_{\tilde I(X),Y-\mu(X)} \sigma_{Y-\mu(X)}} \times \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right]}_{\tilde D_M}.
-\end{split}
-\end{align}
-See Appendix~\ref{app:DRderivation} for the derivation.
+%\begin{align}
+%\label{eq:statdecomp3}
+%\begin{split}
+%\rho_{\tilde I (X), Y-\mu(X)} &\times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y-\mu(X)}  \\
+%&\times \underbrace{\left[ 1 - \frac{\rho_{\tilde I(X),Y} \sigma_{Y}}{\rho_{\tilde I(X),Y-\mu(X)} \sigma_{Y-\mu(X)}} \times \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right]}_{\tilde D_M}.
+%\end{split}
+%\end{align}
+%See Appendix~\ref{app:DRderivation} for the derivation.
 Comparing~\eqref{eq:statdecomp3} to~\eqref{eq:statdecomp2} shows that the addition of the model-based focast impacts the estimation error in two ways.
@@
 Random testing is one way to improve upone current practice whiel we wait for that
-
-{\bf COORDINATION ON DATA SOURCES}
-
-\bibliographystyle{plainnat}
+%
+%{\bf COORDINATION ON DATA SOURCES}
+
+\bibliographystyle{plain}
 \bibliography{covid-refs}
@@
 
-Based on~\cite{Woloshin2020}, we set the false positive rate to 2.4\% and false negative rate to 17.5\%.
+Here, we take the suggested measures from~\cite{Woloshin2020} which report 70\% sensitivity is a reasonable estimates and~\cite{Cohen2020} which report 97.6\% specificity.
+This corresponds to a false negative rate of $30\%$ and false positive rate of $2.4\%$.
 To allow for potential uncertainty, we assume two \emph{synthetic} simple random samples.
@@
 As $\mu_t$ and $\theta_{t}$ are estimated independently at each time $t$, we can focus on the estimating equations per time point separately.
-The estimates are given by:
-$$
-\left (
+
+Here, we consider asymptotics under the following sequence (indexed by $L=1,2,\ldots$) of sampling regimes.
+For $L >0$, consider $L$ random samples of size $n$ drawn at equally spaced times $\{ t^\prime_l \}_{l=1}^L$ over the study window~$[0,T]$.
+Let~$\Delta_L$ denote the gap times between consecutive sample times.
+
+For each~$L=1,2,\ldots$, the estimates of $\eta = (\mu_t, \theta_t, FP, FN)$ is given by:
+$$
+\Phi_L (\eta) = \left (
 \begin{array}{c}
-\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - (1-FP-FN) \cdot (\mu_t+FP)}{\pi_{j,t}} \\
-\frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
+\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t}{\pi_{j,t}} \\
+\frac{1}{N \sum_{l=1}^L K_{t,t_l^\prime}} \sum_{l=1}^L K_{t,t_l^\prime} \left[ \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime_l} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime}  \pi_{j,t_l^\prime} X_{j,t_l^\prime} \right] \\
 \frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
@@
 where $\pi_{j,t^\prime} = \pi (X_{j,t^\prime}; \theta_t)$, $Z_j$ is an indicator of a false positive in the sample $S_{FP}$  and $\tilde Z_j$ is an indicator of a false negative in the sample $S_{FN}$.
-We assume the synthetic $|S_{FP}|^{-1} \sum_{j \in S_{FP}} Z_j = 0.024$ and $|S_{FP}|^{-1} \sum_{j \in S_{FP}} Z_j = 0.175$.
-Sample sizes, denoted $n_{FP}$ and $n_{FN}$ respectively, are chosen to achieve desired estimator variance: $1.96/0.155^2 \cdot \sqrt{0.175 \cdot (1-0.175)} = 31$.
-
-Under joint randomization of propensity score model and sampling designs, we have $E [ \Phi_n (\eta) ] = {\bf 0}$.
+We assume the synthetic $|S_{FP}|^{-1} \sum_{j \in S_{FP}} Z_j = 0.024$ and $|S_{FP}|^{-1} \sum_{j \in S_{FP}} Z_j = 0.30$.
+\textcolor{red}{Sample sizes, denoted $n_{FP}$ and $n_{FN}$ respectively, are chosen to achieve desired estimator variance, i.e., $\frac{0.024 \cdot (0.976)}{n_{FP}} = \sigma^2_{FP}$ and $\frac{0.30 \cdot 0.70}{n_{FN}} = \sigma^2_{FN}$.}
+
+To prove consistency, first consider the setting where a random sample of size $n$ is observed at time $t$.
+Let $\Phi_n (\eta)$ denote the corresponding estimating equations, which are equal to $\Phi_L(\eta)$ except for the second component which is given by
+$$
+\frac{1}{N} \left[ \sum_{j=1}^N I_{j,t} X_{j,t} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t} \tilde W_{j,t}  \pi_{j,t} X_{j,t} \right].
+$$
+Then under joint randomization of propensity score model and sampling designs, we have $E [ \Phi_n (\eta_0) ] = {\bf 0}$.
 Consistency then follows by arguments in Section 3.2 of Tsiatis (2006).
-Under conditions C1-C6 from~\cite{Chen2019} and the additional assumption that $h_n \to 0$ as the sample size per timepoint of the sample size grows, we have $\Phi_n (\hat \eta) = 0$ and $\Phi_n (\eta_0) = O_p ( \tilde n^{-1/2} )$.
-By a first-order Taylor expansion we have $\hat \eta - \eta_0 = \left[ \right]^{-1} \Phi_n(\eta_0) + o_p (\tilde n^{-1/2})$ where $\phi_n(\eta) = \partial \Phi_n (\eta)/\partial \eta$.
+Under conditions C1-C6 from~\cite{Chen2019} and as $n$ increases, we have $\Phi_n (\hat \eta) = 0$ and $\Phi_n (\eta_0) = O_p ( n^{-1/2} )$.
+
+To prove consistency under the sequence of sampling regimes, we simply need to show that $E[ \Phi_L (\eta_0) ] \to 0$ as $L$ increases.  First, we see that
+$$
+E \left[ \sum_{j = 1}^N \tilde I_{j,t} \tilde W_{j,t}  \pi_{j,t} X_{j,t} - \sum_{l=1}^L  \sum_{j = 1}^N \tilde K_{t,t^\prime_l} \tilde I_{j,t} \tilde W_{j,t}  \pi_{j,t} X_{j,t} \right]
+$$
+
+the additional assumption that $h_n \to 0$
+
+By a first-order Taylor expansion we have $\hat \eta - \eta_0 = \left[ \phi_n (\eta) \right]^{-1} \Phi_n(\eta_0) + o_p (\tilde n^{-1/2})$ where $\phi_n(\eta) = \partial \Phi_n (\eta)/\partial \eta$.
 To calculate
@@
 \begin{array}{c}
--\frac{1}{N} \sum_{j=1}^N I_{j,t} (Y_{j,t} - (1-FP-FN) \cdot (\mu_t + FP)) \frac{1-\pi_{j,t}}{\pi_{j,t}} X_{j,t}^\top \\
+-\frac{1}{N} \sum_{j=1}^N I_{j,t} (Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t) \frac{1-\pi_{j,t}}{\pi_{j,t}} X_{j,t}^\top \\
 - \frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} (1-\pi_{j,t^\prime}) X_{j,t^\prime} X_{j,t^\prime}^\top\\
@@
 \begin{array}{c}
--\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{(1-2 \cdot FP-FN -\mu_t)}{\pi_{j,t}} \\
+-\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{(1-\mu_t)}{\pi_{j,t}} \\
 0\\
--\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP^2} + \frac{1-Z_j}{(1-FP)^2}  \\
+-\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \left( \frac{Z_j}{FP^2} + \frac{1-Z_j}{(1-FP)^2} \right)  \\
 0
@@
 \begin{array}{c}
-\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{\mu_t+FP}{\pi_{j,t}} \\
+- \frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{\mu_t}{\pi_{j,t}} \\
 0\\
 0 \\
--\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN^2} + \frac{1-\tilde Z_j}{(1-FN)^2}
+-\frac{1}{n_{FN}}  \sum_{j \in S_{FN}} \left( \frac{\tilde Z_j}{FN^2} + \frac{1-\tilde Z_j}{(1-FN)^2} \right)
 \end{array}
@@
 \begin{array}{c c c c}
-1-FP-FN & \frac{1}{N} \sum_{j=1}^N \zeta_{t,j} (1-\pi_{j,t}) X_{j,t}^\top & 1 - 2FP-FN - \mu_t & \mu_t + FP \\
+1-FP-FN & \frac{1}{N} \sum_{j=1}^N \zeta_{t,j} (1-\pi_{j,t}) X_{j,t}^\top & 1 - \mu_t & \mu_t \\
 0 & \frac{1}{N} \sum_{t^\prime=1}^T K_{t,t^\prime} \sum_{j=1}^N \pi_{j,t} (1-\pi_{j,t}) X_{j,t} X_{j,t}^\top & 0 & 0\\
+0 & 0 & \left( FP (1-FP) \right)^{-1}  & 0\\
+0 & 0 & 0 & \left( FN (1-FN) \right)^{-1}
+\end{array}
+\right )
+$$
+where $\zeta_{t,j} = (Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t)$.
+This matrix can be written as a diagonal matrix equal to the diagonal of $\partial \Phi_n (\eta)/\partial \eta$ and then a rank-three matrix that is zero except for the first row on the off-diagonal.
+Let $A + B$ denote the sum broken into these 2 components.
+Then Woodbury matrix identity gives us
+$$
+(A+B)^{-1} = A^{-1} - A^{-1} B A^{-1}
+$$
+In particular, the resulting inverse is of the form
+$$
+\left (
+\begin{array}{c c c c}
+\frac{1}{1-FP-FN} & b_1 & b_2  & b_3 \\
+0 & \left[\frac{1}{N} \sum_{t^\prime=1}^T K_{t,t^\prime} \sum_{j=1}^N \pi_{j,t} (1-\pi_{j,t}) X_{j,t} X_{j,t}^\top\right]^{-1} & 0 & 0\\
 0 & 0 & FP (1-FP)   & 0\\
@@
 \end{array}
-\right )
-$$
-where $\zeta_{t,j} = (Y_{j,t} - (1-FP-FN) \cdot (\mu_t + FP))$.
-This matrix can be written as a diagonal matrix equal to the diagonal of $\partial \Phi_n (\eta)/\partial \eta$ and then a rank-one matrix that is zero except for the first row on the off-diagonal.
-Let $A + B$ denote the sum then
-
-$$
-(A+B)^{-1} = A^{-1} - \frac{1}{1+g} A^{-1} B A^{-1}
-$$
-where $g = \text{trace}(BA^{-1}) = 0$ by design.  In particular,
-$$
-\left (
-\begin{array}{c c c c}
-\frac{1}{1-FP-FN} & b_1 & b_2  & b_3a  \\
-0 & \left[\frac{1}{N} \sum_{t^\prime=1}^T K_{t,t^\prime} \sum_{j=1}^N \pi_{j,t} (1-\pi_{j,t}) X_{j,t} X_{j,t}^\top\right]^{-1} & 0 & 0\\
-0 & 0 & \left[ FP (1-FP) \right]^{-1}  & 0\\
-0 & 0 & 0 & \left[ FN (1-FN) \right]^{-1}
-\end{array}
 \right)
 $$
-where
-\begin{align*}
-b_1 =& (1-FP-FN)^{-1} \left[ N^{-1} \sum_{t^\prime=1}^T K_{t,t^\prime} \sum_{j=1}^N (1-\pi_{j,t^\prime}) \zeta_{t,j} X_{i,t^\prime}^\top \right] \\
-&\times \left[N^{-1} \sum_{t^\prime=1}^T K_{t,t^\prime} \sum_{j=1}^N \pi_{j,t^\prime} (1-\pi_{j,t^\prime}) X_{i,t^\prime} X_{i,t^\prime}^\top \right]^{-1},
-\end{align*}
-$b_2 = (1-FP-FN)^{-1} (1-2FP-FN-\mu_t) / (FP(1-FP))$, and $b_3 = (1-FP-FN)^{-1} (\mu_t+FP) / (FN(1-FN))$.
-
-Note that $\text{Var}(\Phi_n(\eta))$ can be decomposed into $A_{1,t} + A_{2,t} + A_{3,t} + A_{4,t}$
+where $b_1, b_2,$ and $b_3$ can be calculated using the Woodbury identity.
+
+Next, note that $\text{Var}(\Phi_n(\eta))$ can be decomposed into $A_{1,t} + A_{2,t} + A_{3,t} + A_{4,t}$
 
@@
 0 \\
-\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{(Z_j-FP)^2}{FP(1-FP)} \\
+\frac{1}{n_{FP}} \frac{1}{FP(1-FP)} \\
 0
@@
 0 \\
-\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{(\tilde Z_j - FN)^2}{FN(1-FN)}
+\frac{1}{n_{FN}} \frac{1}{FN(1-FN)}
 \end{array}

### 2021-06-25 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 The first is a simple random sample of true negative cases denoted $S_{FP}$; the second is simple random sample of true positive cases denoted $S_{FN}$.
-As $\mu_t$ and $\theta_{t}$ are estimated independently at each time $t$, we can focus on the estimating equations per time point separately.
-
-Here, we consider asymptotics under the following sequence (indexed by $L=1,2,\ldots$) of sampling regimes.
-For $L >0$, consider $L$ random samples of size $n$ drawn at equally spaced times $\{ t^\prime_l \}_{l=1}^L$ over the study window~$[0,T]$.
-Let~$\Delta_L$ denote the gap times between consecutive sample times.
-
-For each~$L=1,2,\ldots$, the estimates of $\eta = (\mu_t, \theta_t, FP, FN)$ is given by:
-$$
-\Phi_L (\eta) = \left (
+As $\mu_t$ and $\theta_{t}$ are estimated independently at each time $t$, we can focus on the estimating equations per time point~$t$ separately.
+
+Suppose there is a sequence of finite populations of size $N_{\nu}$ indexed by $\nu$.
+Each finite population has an associated non-probability at time $t$ of  size~$\tilde n$ and $L_\nu$ probability samples of fixed size $n$ drawn at equally spaced times $\{ t^\prime_l \}_{l=1}^L$ over the study window~$[0,T]$.
+Let~$\Delta_\nu$ denote the gap times between consecutive sample times.
+Finally, let~$n_{FP}$ and $n_{FN}$ denote the sample sizes of $S_{FP}$ and $S_{FN}$ respectively.
+For simplicity, dependency on $\nu$ is suppressed and the limiting process is represented by $N \to \infty$.
+
+For each~$\nu$ and $t$, the estimates of $\eta = (\mu_t, \theta_t, FP, FN)$ are given by:
+$$
+\Phi (\eta) = \left (
 \begin{array}{c}
@@
 To prove consistency, first consider the setting where a random sample of size $n$ is observed at time $t$.
-Let $\Phi_n (\eta)$ denote the corresponding estimating equations, which are equal to $\Phi_L(\eta)$ except for the second component which is given by
+Let $\tilde \Phi (\eta)$ denote the corresponding estimating equations, which is the same as $\Phi(\eta)$ except for the second component which is given by
 $$
@@
 $$
-Then under joint randomization of propensity score model and sampling designs, we have $E [ \Phi_n (\eta_0) ] = {\bf 0}$.
+Then under joint randomization of propensity score model and sampling designs, we have $E [ \tilde \Phi (\eta_0) ] = {\bf 0}$.
 Consistency then follows by arguments in Section 3.2 of Tsiatis (2006).
-Under conditions C1-C6 from~\cite{Chen2019} and as $n$ increases, we have $\Phi_n (\hat \eta) = 0$ and $\Phi_n (\eta_0) = O_p ( n^{-1/2} )$.
-
-To prove consistency under the sequence of sampling regimes, we simply need to show that $E[ \Phi_L (\eta_0) ] \to 0$ as $L$ increases.  First, we see that
-$$
-E \left[ \sum_{j = 1}^N \tilde I_{j,t} \tilde W_{j,t}  \pi_{j,t} X_{j,t} - \sum_{l=1}^L  \sum_{j = 1}^N \tilde K_{t,t^\prime_l} \tilde I_{j,t} \tilde W_{j,t}  \pi_{j,t} X_{j,t} \right]
-$$
-
-the additional assumption that $h_n \to 0$
-
-By a first-order Taylor expansion we have $\hat \eta - \eta_0 = \left[ \phi_n (\eta) \right]^{-1} \Phi_n(\eta_0) + o_p (\tilde n^{-1/2})$ where $\phi_n(\eta) = \partial \Phi_n (\eta)/\partial \eta$.
+Under conditions C1-C6 from~\cite{Chen2019} and as $n$ increases, we have $\tilde \Phi (\hat \eta) = 0$ and $\tilde \Phi(\eta_0) = O_p ( n^{-1/2} )$.
+
+To prove consistency under the sequence of sampling regimes, we need to show that $E[ \Phi (\eta_0) ] \to 0$ as $N \to \infty$.  First, we see that
+\begin{align*}
+&\frac{1}{N}  E \left[ \sum_{j = 1}^N \tilde I_{j,t} \tilde W_{j,t}  \pi_{j,t} X_{j,t} - \sum_{l=1}^L  \sum_{j = 1}^N \tilde K_{t,t^\prime_l} \tilde I_{j,t} \tilde W_{j,t}  \pi_{j,t} X_{j,t^\prime_l} \right] \\
+= &\frac{1}{N} \sum_{j = 1}^N \left[ \pi_{j,t} X_{j,t} - \sum_{l=1}^L \tilde K_{t,t^\prime_l} \pi_{j,t^\prime_l} X_{j,t^\prime_l} \right], \\
+= &\frac{1}{N} \sum_{j = 1}^N \left[ \left( \sum_{l=1}^L \tilde K_{t,t^\prime_l}  \left( \pi (X_{j,t}; \theta) - \pi (X_{j,t^\prime_l}; \theta) \right) \right) X_{j,t} + \sum_{l=1}^L \tilde K_{t,t^\prime_l} \pi_{j,t^\prime_l} (X_{j,t} - X_{j,t^\prime_l} ) \right],
+\end{align*}
+where $\tilde K_{t, t^\prime_l}$ is the normalized kernel such that $\sum_{l=1}^L \tilde K_{t, t^\prime_l} = 1$.  Moreover, we have that
+
+Let~$X_{j,t,l}$ denote the $l$th component of the vector $X_{j,t}$.  Then consider the population-level distribution $\{ X_{j,t,l} \}_{j=1}^N$.  We suppose
+$$
+\lim_{\epsilon \to 0} \lim_{N \to \infty} \frac{1}{N} \sum_{j=1}^N X_{j,t+\epsilon, l}  = \lim_{N \to \infty} \frac{1}{N} \sum_{j=1}^N X_{j,t, l}.
+$$
+That is, the limiting distributions are continuous as a function of $t$.
+If $\pi(x; \theta)$ is a continuous function in~$x$ and under suitable conditions on the kernel density~$\tilde K_{t, \cdot}$ with the bandwidth $h_\nu \to \infty$ such that $\sum_{l=1}^L \tilde K_{t,t_l} \to \infty$ as $\nu \to \infty$, we have
+$$
+\left | \sum_{l=1}^L \tilde K_{t, t_l^\prime} \pi (X_{j, t_l^\prime}; \theta) - \pi (X_{j, t}; \theta) \right|  \to 0
+$$
+
+Then by a first-order Taylor expansion we have $\hat \eta - \eta_0 = \left[ \phi (\eta) \right]^{-1} \Phi(\eta_0) + o_p (\tilde n^{-1/2})$ where $\phi(\eta) = \partial \Phi (\eta)/\partial \eta$ and $\tilde n = n \sum_{l=1}^L \tilde K_{t,t_l}$.
 To calculate
@@
 
-Next, note that $\text{Var}(\Phi_n(\eta))$ can be decomposed into $A_{1,t} + A_{2,t} + A_{3,t} + A_{4,t}$
+Next, note that $\text{Var}(\Phi(\eta))$ can be decomposed into $A_{1,t} + A_{2,t} + A_{3,t} + A_{4,t}$
 
@@
 A_{1,t} = \frac{1}{N} \left( \begin{array}{c}
-\sum_{j=1}^N \frac{I_{j,t} (Y_{j,t} - (1-FP-FN) (\mu_t + FP)}{\pi_{j,t}} \\
+\sum_{j=1}^N \frac{I_{j,t} (Y_{j,t} - FP - (1-FP-FN) \mu_t)}{\pi_{j,t}} \\
 \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \pi_{j,t^\prime} X_{j,t^\prime} \\
@@
 The asymptotic variance for the IPW estimator is then the first diagonal element of the matrix $E \left[ \phi_n (\eta_0) \right]^{-1} \times \left[ V_{1,t} + V_{2,t} + V_{3,t} + V_{4,t} \right] E \left[ \phi_n (\eta_0) \right]^{-1}$.
-
-\begin{align*}
-N^{-2} \bigg[ &(1-FP-FN)^{-2} \sum_{j=1}^N \zeta_{t,j}^2 \frac{(1-\pi_{j,t})}{\pi_{t,j}} + (1-FP-FN)^{-1} b_{1} \sum_{j=1}^N \sum_{t^\prime} K_{t,t^\prime} (1-\pi_{j,t^\prime}) \zeta_{t^\prime,j} X_{j,t} \\
-&+ (1-FP-FN)^{-1} \sum_{j=1}^N \sum_{t^\prime} K_{t,t^\prime} (1-\pi_{j,t^\prime} \zeta_{t^\prime, j} ) X_{j,t^\prime}^\top + b_1 \sum_{j=1}^N \sum_{t^\prime} K_{t,t^\prime} K_{t,t^\prime} \pi_{j,t^\prime} X_{j,t} X_{j,t}^\top \bigg]
-\end{align*}
 

### 2021-06-28 write-up/covid-writeup.tex
Theory almost good (need case study and plots next time)

@@
 Our primary objective is a model-based forecast of strata-level prevalence $\bar y_k$.
-To do this, an extension of a standard epidemiological state-space model -- the susceptible, infected, and removed (recovered and death) model, or SIR model -- is presented.
-Here we consider the probabilistic SIR model was originally proposed by~\cite{Osthus2017} with only one-dimensional time series of infected proportions; this formulation was extended by~\cite{Song2020} to model coronavirus case-counts.
-
-Let $s_t$, $e_t$, $i_t^{(T)}$, $i^{(NT)}$, and $r_t$ denote the proportion of survivors, exposed, currently transmitting and not trasmitting infected, and removed cases (i.e., including both recovered cases and deaths) at time $t$.
-Suppose $K$ strata such that the proportions are defined per strata, i.e., $s_t = (s_{1,t},\ldots, s_{K,t})$ is the proportion of survivors in each strata.
-Let $\theta_t = (s_t, e_t, i_t^{(T)}, i_t^{(NT)}, r_t)$, then we assume the process is governed by the following Markov model:
-$$
-\theta_t \mid \theta_{t-1}; \tau \sim \text{Dirichlet} \left(\kappa f(\theta_{t-1}, \beta, \gamma) \right)
-$$
-where $\kappa$ scales the variance of the Dirichlet distribution and $f(\cdot)$ is a function that determines infection dynamics.  Here, we extend the SIR dynamics to account for strata-level dynamics
+To do this, a probabilistic extension of a standard epidemiological state-space model -- the susceptible, exposed, infected, and removed (recovered and death) model, or SEIR model -- is presented.
+A probabilistic SIR model was originally proposed by~\cite{Osthus2017} with only one-dimensional time series of infected proportions; this formulation was extended by~\cite{Song2020} to model coronavirus case-counts.
+
+Let $s_t$, $e_t$, $i_t^{(T)}$, and $r_t$ denote the proportion of survivors, exposed, infected, and removed cases (i.e., including both recovered cases and deaths) at time $t$.
+The population-level SEIR dynamics are then given by the following set of differential equations:
 \begin{align*}
-\frac{\partial s_t}{\partial t} &= - s_t \circ (B \, i^{(T)}_t); \quad \frac{\partial e_t}{\partial t} = s_t \circ (B \, i^{(T)}_t) - \sigma e_t; \quad \\
-\frac{\partial i^{(T)}_t}{\partial t} &= \sigma e_t - \beta_2 i^{(T)}; \quad
-\frac{\partial i^{(NT)}_t}{\partial t} = \beta_2 i^{(NT)}  - \gamma i^{(NT)}; \quad \\
-\frac{\partial r_t}{\partial t} &= - \gamma i^{(NT)}_t
+\frac{\partial s_t}{\partial t} &= - \beta s_t i_t; \quad \frac{\partial e_t}{\partial t} = \beta s_t i_t - \sigma e_t; \quad \\
+\frac{\partial i^{(T)}_t}{\partial t} &= \sigma e_t - \gamma i_t; \quad
+\frac{\partial r_t}{\partial t} = - \gamma i_t.
 \end{align*}
-where $B \in \mathbb{R}^{K \times K}$ and $\circ$ denotes elementwise product of vectors.
-The matrix $B$ allows for strata-specific infection dynamics with the SIR model being recovered if $B_{k,\cdot} = \beta$ for each $k \in [K]$.
-Following~\cite{Song2020}, we employ Runge-Kutta (RK4) approximations for discretization; see Appendix XX for details.
-
-Selection bias in case count data could be direcly addressed by extending the above compartmental model.
-Here, we instead opt to bring in a different source of information by considering COVID-19 death data.
-Let $D_{k,r}$ denote the number of individuals who pass away on day $r$ from strata $k$ and $\nu_{k,t}$ to denote the number of new infections on day $t$ in the $k$th strata.
+Following~\cite{Song2020}, we employ Runge-Kutta (RK4) approximations for discretization.
+Here, we consider covariate information to take the form of a stratificatoin variable $k$ with $K$ strata.
+Let $\nu_{k,t}$ to denote the number of new infections on day $t$ in the $k$th strata.
+AS there are a total of $N \cdot i_t := I_t$ new infections on day $t$, we assume the joint distribution of strata-specific new infection totals takes a multinomial distribution
+$$
+( \nu_{1,t}, \ldots, \nu_{K,t} ) \sim \text{Multinomial} \left( I_t, (p_{i1,t}, \ldots, p_{iK, t}) \right).
+$$
+
+Selection bias in case count data is addressed by bringing in a different source of information, COVID-19 death data.
+Let $D_{k,r}$ denote the number of individuals who pass away on day $r$ from strata $k$.
 Then
@@
 
-%\subsection{Doubly robust estimation.}
-%Rather than relying solely on IPW or epidemiological forecasts, here we combine forecasting and inverse-probability weighting by extending recent work by~\cite{Chen2019} to account time-varying propensities and measurement-error.
+\subsection{Doubly robust estimation.}
+Rather than relying solely on IPW or epidemiological forecasts, here we combine forecasting and inverse-probability weighting by extending recent work by~\cite{Chen2019} to account time-varying propensities and measurement-error.
 In this context, the doubly-robust estimator is given by
-%$$
-%\bar y_{n}^{(DR)} = \frac{1}{N} \sum_{j=1}^N \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \mu(x_j) \right).
-%$$
-%where the estimates $\hat \mu(x_j)$ are not corrected for measurement-error as they are built to estimate true case count from death-only data.
+$$
+\bar y_{n}^{(DR)} = \frac{1}{N} \sum_{j=1}^N \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \mu(x_j) \right).
+$$
+where the estimates $\hat \mu(x_j)$ are not corrected for measurement-error as they are built to estimate true case count from death-only data.
 This estimator is called ``doubly-robust'' because it is consistent if either the model-based forecasts or the time-varying propensities are correctly specified.
 A statistical error decomposition can be derived
-%\begin{align}
-%\label{eq:statdecomp3}
-%\begin{split}
-%\rho_{\tilde I (X), Y-\mu(X)} &\times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y-\mu(X)}  \\
-%&\times \underbrace{\left[ 1 - \frac{\rho_{\tilde I(X),Y} \sigma_{Y}}{\rho_{\tilde I(X),Y-\mu(X)} \sigma_{Y-\mu(X)}} \times \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right]}_{\tilde D_M}.
-%\end{split}
-%\end{align}
-%See Appendix~\ref{app:DRderivation} for the derivation.
+\begin{align}
+\label{eq:statdecomp3}
+\begin{split}
+\rho_{\tilde I (X), Y-\mu(X)} &\times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y-\mu(X)}  \\
+&\times \underbrace{\left[ 1 - \frac{\rho_{\tilde I(X),Y} \sigma_{Y}}{\rho_{\tilde I(X),Y-\mu(X)} \sigma_{Y-\mu(X)}} \times \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right]}_{\tilde D_M}.
+\end{split}
+\end{align}
+See Appendix~\ref{app:DRderivation} for the derivation.
 Comparing~\eqref{eq:statdecomp3} to~\eqref{eq:statdecomp2} shows that the addition of the model-based focast impacts the estimation error in two ways.
@@
 Note that the model-based forecast does not impact the data quantity component.
-%
-%If the model-based forecast is correct then $E_{\bf R} [ \rho_{\tilde I (X), Y-\mu(X)} ] = 0$  and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the forecasts are not correctly specified then the data quality index is unlikely to inversely scale with population size.
+
+If the model-based forecast is correct then $E_{\bf R} [ \rho_{\tilde I (X), Y-\mu(X)} ] = 0$  and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the forecasts are not correctly specified then the data quality index is unlikely to inversely scale with population size.
 Interestingly, the impact of measurement-error on data quality now depends on a relative comparison of the data quality and problem difficulty of the weighted estimator and the doubly-robust estimator.
@@
 Again, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from a DR estimator are likely to put too much confidence in the sheer size of the data.
-%
-%Unfortunately, the first term cannot be computed as covariate information is not collected on the entire population.
+
+Unfortunately, the first term cannot be computed as covariate information is not collected on the entire population.
 Here, we again use the probability samples and consider the doubly-robust estimator
-%$$
-%\frac{1}{\sum_{j=1}^N \tilde I_j \tilde W_j} \sum_{j=1}^N  \tilde I_j \tilde W_j \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \mu(x_j) \right).
-%$$
-%where $\tilde I_j$ indicates an individual is included in the random sample and $\tilde W_j$ is the associated weight for that unit.
+$$
+\frac{1}{\sum_{j=1}^N \tilde I_j \tilde W_j} \sum_{j=1}^N  \tilde I_j \tilde W_j \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \mu(x_j) \right).
+$$
+where $\tilde I_j$ indicates an individual is included in the random sample and $\tilde W_j$ is the associated weight for that unit.
 Assuming the selection propensity model is correctly specified, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:dr} shows that the IPW estimator is  consistent and calculates the estimator's variance.
 See Appendix~\ref{app:asympderivations} for additional details.
-%
-%\begin{lemma}[Variance of doubly-robust estimator] \normalfont
-%\label{lemma:dr}
-%\textcolor{red}{Must fix.}
-%The estimates~$\mu_t := \bar y_{n,t}^\star$ and $\hat \pi_{j,t}$
-%are solutions to the following set of estimating equations
-%$$
-%\Phi_n (\eta_t) =
-%\left (
-%\begin{array}{c}
-%\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - (1-FP-FN) \cdot (\mu_t+FP)}{\pi_{j,t}} \\
-%\frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
-%\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
-%\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
-%\end{array}
-%\right ) = {\bf 0}
-%$$
-%where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
+
+\begin{lemma}[Variance of doubly-robust estimator] \normalfont
+\label{lemma:dr}
+\textcolor{red}{Must fix.}
+The estimates~$\mu_t := \bar y_{n,t}^\star$ and $\hat \pi_{j,t}$
+are solutions to the following set of estimating equations
+$$
+\Phi_n (\eta_t) =
+\left (
+\begin{array}{c}
+\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - (1-FP-FN) \cdot (\mu_t+FP)}{\pi_{j,t}} \\
+\frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
+\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
+\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
+\end{array}
+\right ) = {\bf 0}
+$$
+where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
 Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$ (check)  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ where $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$.
-%\end{lemma}
-%
-%
+\end{lemma}
+
 %\section{COVID-19 active infection prevalence in Indiana}
@@
 0 \\
-\frac{1}{n_{FP}} \frac{1}{FP(1-FP)} \\
+\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
 0
@@
 0 \\
-\frac{1}{n_{FN}} \frac{1}{FN(1-FN)}
+\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
 \end{array}
@@
 and $V_{j,t}$ are block-diagonal matrices with only one non-zero block that is equal to $D_{j,t}$ which is the design-based variance-covariance matrix under each probability sampling design.
+$$
+V_{3,t} = \left( \begin{array}{c}
+0 \\
+0 \\
+\frac{1}{n_{FP}} \frac{1}{FP(1-FP)} \\
+0
+\end{array}
+\right),
+$$
+and
+$$
+V_{4,t} = \left( \begin{array}{c}
+0 \\
+0 \\
+0 \\
+\frac{1}{n_{FN}} \frac{1}{FN(1-FN)}
+\end{array}
+\right),
+$$
+
 The asymptotic variance for the IPW estimator is then the first diagonal element of the matrix $E \left[ \phi_n (\eta_0) \right]^{-1} \times \left[ V_{1,t} + V_{2,t} + V_{3,t} + V_{4,t} \right] E \left[ \phi_n (\eta_0) \right]^{-1}$.
 
-\section{Model-based estimation: parameter choices and inference}
-\label{app:modelbased_approx}
-
-Following~\cite{Song2020}, the fourth-order Runge-Kutta (RK4) method gives an approximation to the proposed SEIR model in equation XX as follows:
-
-$$
-f(\theta_{t-1}, \beta, \gamma) = \left(
-\begin{array}{c}
-\theta_{t-1}^{S} + \frac{1}{6} \left[ k_{t-1}^{S_1} + 2 k_{t-1}^{S_2} + 2 k_{t-1}^{S_3} + k_{t-1}^{S_4} \right] \\
-\theta_{t-1}^{E} + \frac{1}{6} \left[ k_{t-1}^{E_1} + 2 k_{t-1}^{E_2} + 2 k_{t-1}^{E_3} + k_{t-1}^{E_4} \right] \\
-\theta_{t-1}^{I^{(T)}} + \frac{1}{6} \left[ k_{t-1}^{I_1^{(T)}} + 2 k_{t-1}^{I_2^{(T)}} + 2 k_{t-1}^{I_3^{(T)}} + k_{t-1}^{I_4^{(T)}} \right] \\
-\theta_{t-1}^{I^{(NT)}} + \frac{1}{6} \left[ k_{t-1}^{I_1^{(NT)}} + 2 k_{t-1}^{I_2^{(NT)}} + 2 k_{t-1}^{I_3^{(NT)}} + k_{t-1}^{I_4^{(NT)}} \right] \\
-\theta_{t-1}^{R} + \frac{1}{6} \left[ k_{t-1}^{R_1} + 2 k_{t-1}^{R_2} + 2 k_{t-1}^{R_3} + k_{t-1}^{R_4} \right] \\
-\end{array}
-\right)
-$$
-where
-\begin{align*}
-k_t^{S_1} &= - s_t \circ (B i_t^{(T)}) \\
-k_t^{S_2} &= - (s_t + 0.5 k_t^{S_1}) \circ (B (i_t^{(T)} + 0.5 k_t^{I_1}) \\
-k_t^{S_3} &= - (s_t + 0.5 k_t^{S_2}) \circ (B i_t^{(T)} + 0.5 k_t^{I_2}) \\
-k_t^{S_4} &= - (s_t + k_t^{S_3}) \circ (B i_t^{(T)} + k_t^{I_3}),
-\end{align*}
-and
-\begin{align*}
-k_t^{E_1} &= s_t \circ (B i_t^{(T)}) - \sigma e_t \\
-k_t^{E_2} &= (s_t + 0.5 k_t^{S_1}) \circ (B (i_t^{(T)} + 0.5 k_t^{I_1}) - \sigma (e_t + 0.5 k_t^{e_1}) \\
-k_t^{E_3} &= (s_t + 0.5 k_t^{S_2}) \circ (B (i_t^{(T)} + 0.5 k_t^{I_2}) - \sigma (e_t + 0.5 k_t^{e_2}) \\
-k_t^{E_4} &= (s_t + k_t^{S_3}) \circ (B (i_t^{(T)} + k_t^{I_4}) - \sigma (e_t + k_t^{E_3}) \\
-\end{align*}
-and
-\begin{align*}
-k_t^{I_1^{(T)}} &= \sigma e_t - \beta_2 i^{(T)} \\
-k_t^{I_2^{(T)}} &= \sigma (e_t + 0.5 k_t^{E_1}) - \beta_2 (i^{(T)} + 0.5 k_t^{I_1^{(T)}}) \\
-k_t^{I_3^{(T)}} &= \sigma (e_t + 0.5 k_t^{E_2}) - \beta_2 (i^{(T)} + 0.5 k_t^{I_2^{(T)}}) \\
-k_t^{I_4^{(T)}} &= \sigma (e_t + k_t^{E_3}) - \beta_2 (i^{(T)} + k_t^{I_3^{(T)}}),
-\end{align*}
-and
-\begin{align*}
-k_t^{I_1^{(NT)}} &= \beta_2 i^{(T)} - \gamma i^{(NT)} \\
-k_t^{I_2^{(NT)}} &= \beta_2 (i^{(T)} + 0.5 k_t^{I_1^{(T)}}) - \gamma (i^{(NT)} + 0.5 k_t^{I_1^{(NT)}})\\
-k_t^{I_3^{(NT)}} &= \beta_2 (i^{(T)} + 0.5 k_t^{I_2^{(T)}}) - \gamma (i^{(NT)} + 0.5 k_t^{I_2^{(NT)}})\\
-k_t^{I_4^{(NT)}} &= \beta_2 (i^{(T)} + k_t^{I_3^{(T)}}) - \gamma (i^{(NT)} + k_t^{I_3^{(NT)}}) ,
-\end{align*}
-and
-\begin{align*}
-k_t^{R_1} &= - \gamma i_t^{(NT)} \\
-k_t^{R_2} &= - \gamma (i^{(NT)} + 0.5 k_t^{I_1^{(NT)}}) \\
-k_t^{R_3} &= - \gamma (i^{(NT)} + 0.5 k_t^{I_2^{(NT)}}) \\
-k_t^{R_4} &= - \gamma (i^{(NT)} + k_t^{I_3^{(NT)}}) \\
-\end{align*}
+% \section{Model-based estimation: parameter choices and inference}
+% \label{app:modelbased_approx}
+
+% Following~\cite{Song2020}, the fourth-order Runge-Kutta (RK4) method gives an approximation to the proposed SEIR model in equation XX as follows:
+
+% $$
+% f(\theta_{t-1}, \beta, \gamma) = \left(
+% \begin{array}{c}
+% \theta_{t-1}^{S} + \frac{1}{6} \left[ k_{t-1}^{S_1} + 2 k_{t-1}^{S_2} + 2 k_{t-1}^{S_3} + k_{t-1}^{S_4} \right] \\
+% \theta_{t-1}^{E} + \frac{1}{6} \left[ k_{t-1}^{E_1} + 2 k_{t-1}^{E_2} + 2 k_{t-1}^{E_3} + k_{t-1}^{E_4} \right] \\
+% \theta_{t-1}^{I^{(T)}} + \frac{1}{6} \left[ k_{t-1}^{I_1^{(T)}} + 2 k_{t-1}^{I_2^{(T)}} + 2 k_{t-1}^{I_3^{(T)}} + k_{t-1}^{I_4^{(T)}} \right] \\
+% \theta_{t-1}^{I^{(NT)}} + \frac{1}{6} \left[ k_{t-1}^{I_1^{(NT)}} + 2 k_{t-1}^{I_2^{(NT)}} + 2 k_{t-1}^{I_3^{(NT)}} + k_{t-1}^{I_4^{(NT)}} \right] \\
+% \theta_{t-1}^{R} + \frac{1}{6} \left[ k_{t-1}^{R_1} + 2 k_{t-1}^{R_2} + 2 k_{t-1}^{R_3} + k_{t-1}^{R_4} \right] \\
+% \end{array}
+% \right)
+% $$
+% where
+% \begin{align*}
+% k_t^{S_1} &= - s_t \circ (B i_t^{(T)}) \\
+% k_t^{S_2} &= - (s_t + 0.5 k_t^{S_1}) \circ (B (i_t^{(T)} + 0.5 k_t^{I_1}) \\
+% k_t^{S_3} &= - (s_t + 0.5 k_t^{S_2}) \circ (B i_t^{(T)} + 0.5 k_t^{I_2}) \\
+% k_t^{S_4} &= - (s_t + k_t^{S_3}) \circ (B i_t^{(T)} + k_t^{I_3}),
+% \end{align*}
+% and
+% \begin{align*}
+% k_t^{E_1} &= s_t \circ (B i_t^{(T)}) - \sigma e_t \\
+% k_t^{E_2} &= (s_t + 0.5 k_t^{S_1}) \circ (B (i_t^{(T)} + 0.5 k_t^{I_1}) - \sigma (e_t + 0.5 k_t^{e_1}) \\
+% k_t^{E_3} &= (s_t + 0.5 k_t^{S_2}) \circ (B (i_t^{(T)} + 0.5 k_t^{I_2}) - \sigma (e_t + 0.5 k_t^{e_2}) \\
+% k_t^{E_4} &= (s_t + k_t^{S_3}) \circ (B (i_t^{(T)} + k_t^{I_4}) - \sigma (e_t + k_t^{E_3}) \\
+% \end{align*}
+% and
+% \begin{align*}
+% k_t^{I_1^{(T)}} &= \sigma e_t - \beta_2 i^{(T)} \\
+% k_t^{I_2^{(T)}} &= \sigma (e_t + 0.5 k_t^{E_1}) - \beta_2 (i^{(T)} + 0.5 k_t^{I_1^{(T)}}) \\
+% k_t^{I_3^{(T)}} &= \sigma (e_t + 0.5 k_t^{E_2}) - \beta_2 (i^{(T)} + 0.5 k_t^{I_2^{(T)}}) \\
+% k_t^{I_4^{(T)}} &= \sigma (e_t + k_t^{E_3}) - \beta_2 (i^{(T)} + k_t^{I_3^{(T)}}),
+% \end{align*}
+% and
+% \begin{align*}
+% k_t^{I_1^{(NT)}} &= \beta_2 i^{(T)} - \gamma i^{(NT)} \\
+% k_t^{I_2^{(NT)}} &= \beta_2 (i^{(T)} + 0.5 k_t^{I_1^{(T)}}) - \gamma (i^{(NT)} + 0.5 k_t^{I_1^{(NT)}})\\
+% k_t^{I_3^{(NT)}} &= \beta_2 (i^{(T)} + 0.5 k_t^{I_2^{(T)}}) - \gamma (i^{(NT)} + 0.5 k_t^{I_2^{(NT)}})\\
+% k_t^{I_4^{(NT)}} &= \beta_2 (i^{(T)} + k_t^{I_3^{(T)}}) - \gamma (i^{(NT)} + k_t^{I_3^{(NT)}}) ,
+% \end{align*}
+% and
+% \begin{align*}
+% k_t^{R_1} &= - \gamma i_t^{(NT)} \\
+% k_t^{R_2} &= - \gamma (i^{(NT)} + 0.5 k_t^{I_1^{(NT)}}) \\
+% k_t^{R_3} &= - \gamma (i^{(NT)} + 0.5 k_t^{I_2^{(NT)}}) \\
+% k_t^{R_4} &= - \gamma (i^{(NT)} + k_t^{I_3^{(NT)}}) \\
+% \end{align*}
 

### 2021-06-29 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 To address this, a procedure is presented that combines case-count data and random samples over time to estimate selection propensities based on key covariate information.
-We then combine these selection propensities with epidemiological forecast models using death-only data to construct a \emph{doubly-robust} estimation method that accounts for both measurement-error and selection bias.
+We then combine these selection propensities with epidemiological forecast models using death-only data to construct a \emph{doubly robust} estimation method that accounts for both measurement-error and selection bias.
 This method is then applied to estimate Indiana's prevalence using case-count, hospitalization, and death data with cumulative demographic information, the United States's only statewide random molecular sample collected from April 25--29th, and Facebook's COVID-19 symptom survey.
@@
 This extends prior analysis of death data~\cite{Johndrow2020} by allowing the infection fatality rate to depend on age.
-Indeed,~\cite{Levin2020} demonstrate that ``the estimated IFR is close to zero for children and younger adults but rises exponentially with age, reaching 0.4\% at age 55, 1.4\% at age 65, 4.6\% at age 75, and 15\% at age 85.'' For conciseness, parameter choices and inference are discussed in depth in Appendix~\ref{app:modelbased}.
+Indeed,~\cite{Levin2020} demonstrate that ``the estimated IFR is close to zero for children and younger adults but rises exponentially with age, reaching 0.4\% at age 55, 1.4\% at age 65, 4.6\% at age 75, and 15\% at age 85.''  See Appendix~\ref{app:ifr} for details on choice of $p_k$.
+
+% 0-17: 20/1000000 * 100 -> 0.002 (at age 9)
+% 18-49: 500/1000000 * 100 -> 0.05 (at age 33)
+% 50–64 years old: 6,000/1000000 * 100 -> 0.6
+% 65+ years old: 90,000/1000000 * 100 -> 9.0
+
+%
+
+% For conciseness, parameter choices and inference are discussed in depth in Appendix~\ref{app:modelbased}.
 
@@
 \label{lemma:dr}
-\textcolor{red}{Must fix.}
 The estimates~$\mu_t := \bar y_{n,t}^\star$ and $\hat \pi_{j,t}$
@@
 \begin{array}{c}
-\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - (1-FP-FN) \cdot (\mu_t+FP)}{\pi_{j,t}} \\
+\frac{1}{N} \left[ \sum_{j=1}^N \frac{I_{j,t}}{\pi_{j,t}} \left( \left( \frac{Y_{j,t} - FP}{1-FP-FN} - \mu(x_j)  \right)  - \mu_t \right) + \frac{1}{\sum_{j=1}^N \tilde I_j \tilde W_j} \sum_{j=1}^N  \tilde I_j \tilde W_j \hat \mu (x_j) \right] \\
 \frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
@@
 
-%\section{COVID-19 active infection prevalence in Indiana}
-%\label{section:applications}
-%
-%% For COVID-19 case-count data, the only covariate information provided by state and local governments are summary statistics for those factors discussed in Section~\ref{section:testinginfo} -- age, sex,  and race.
+\section{COVID-19 active infection prevalence in Indiana}
+\label{section:applications}
+
+% For COVID-19 case-count data, the only covariate information provided by state and local governments are summary statistics for those factors discussed in Section~\ref{section:testinginfo} -- age, sex,  and race.
 These are discrete factors and therefore define a stratification of the population.
@@
 Recommendation XX below states that additional covariate information should be collected from nonprobabilistic samples.
-%
-%Here we consider estimation of the active infection rate in Indiana. We consider five sources of data:
-%\begin{itemize}[leftmargin=*]
-%\item {\bf Testing data}:  Each day, the number of tests performed and number of positive tests are reported, with counts broken out jointly by age, gender, and racial demographic information.
+
+Here we consider estimation of the active infection rate in Indiana. We consider five sources of data:
+\begin{itemize}[leftmargin=*]
+\item {\bf Testing data}:  Each day, the number of tests performed and number of positive tests are reported, with counts broken out jointly by age, gender, and racial demographic information.
 Figure~\ref{fig:in-cases} plots the number of COVID-19 positive cases based on testing date by age.
-%
-%\begin{figure}
-%\centering
-%\begin{subfigure}{.45\textwidth}
-%  \centering
-%  \includegraphics[width=.9\linewidth]{../figs/indianacasecounts.png}
-%  \caption{Indiana case-counts per day}
-%  \label{fig:in-cases}
-%\end{subfigure}
-%\begin{subfigure}{.45\textwidth}
-%  \centering
-%  \includegraphics[width=.9\linewidth]{../figs/indianadeaths.png}
-%  \caption{Indiana death-counts per day}
-%  \label{fig:in-deaths}
-%\end{subfigure}
-%\end{figure}
-%
-%\item {\bf Random/Nonrandom statewide sample}: Table~\ref{tab:indiana} summarizes data from a random sample from April 25--29th as well as a nonrandom sample obtained between May 2--3 in racial/ethnic minority communities (cite).
+
+\begin{figure}
+\centering
+\begin{subfigure}{.45\textwidth}
+ \centering
+ \includegraphics[width=.9\linewidth]{../figs/indianacasecounts.png}
+ \caption{Indiana case-counts per day}
+ \label{fig:in-cases}
+\end{subfigure}
+\begin{subfigure}{.45\textwidth}
+ \centering
+ \includegraphics[width=.9\linewidth]{../figs/indianadeaths.png}
+ \caption{Indiana death-counts per day}
+ \label{fig:in-deaths}
+\end{subfigure}
+\end{figure}
+
+\item {\bf Random/Nonrandom statewide sample}: Table~\ref{tab:indiana} summarizes data from a random sample from April 25--29th as well as a nonrandom sample obtained between May 2--3 in racial/ethnic minority communities~\cite{Yiannoutsos2021}.
 The nonrandom sample is not a random subsample of the overall case-counts; however, this data provides important supplementary covariate information.
@@
 These adjusted rates are presented in parentheses in Table~\ref{tab:indiana}.
-%
-%
-%\begin{table}[th]
-%\begin{tabular}{c | c | r r | r | r r }
-%& & \multicolumn{2}{c}{Total Tests (\%)} & US & \multicolumn{2}{c}{Positive Test Rate (\%)}\\
-%\cline{3-4} \cline{6-7}
-%& & NonRandom & Random & Census & NonRandom & Random \\ \hline
-%\multirow{2}{*}{Sex} & Female & 58.2 & 55.0 & 50.7 & 21.7 (11.2) & 1.4 \\
-%& Male & 41.8 & 45.0 & 49.3 & 24.2 (12.4) & 2.1 \\ \hline
-%\multirow{3}{*}{Age} & $<40$ & 39.4 & 28.0 & 52.7 & 29.7 (15.0) & 1.7 \\
-%& $40-59$ & 41.1 & 36.0 & 25.2 & 24.9 (12.5) & 2.1 \\
-%& $\geq 60$ & 19.5 & 36.0 & 22 & 6.7 (3.4) & 0.9 \\ \hline
-%\multirow{2}{*}{Race} & White & 23.1 & 92.0 & 86.9 & 19.5 (9.6) & 1.5 \\
-%& Nonwhite & 76.9 & 8.0 & 13.1 & 25.0 (12.3) & 3.4 \\ \hline
-%% Hispanic & Hispanic & 44.1 & 2.0 & 7.1 & 37.6 & 6.9 \\
-%% Origin & Non-Hispanic & 55.9 & 98.0 & 92.9 & 13.0 & 1.3 \\ \hline
-%\multirow{2}{*}{Fever} & Yes & 17.0 & 1.8 & - & 66.4 (32.1) & 4.5 \\
-%& No & 83.0 & 98.2 & - & 15.6 (7.5) & 1.3 \\ \hline
-%Household & Yes & 10.8 & 1.4 & - & 46.1 & 29.4 \\
-%$+$ Case & No & 89.2 & 98.6 & - & 21.6 & 1.3 \\ \hline
-%Prior $+$ & Yes & 6.1 & 1.4 & - & 39.2 & 24.4 \\
-%Test & No & 93.9 & 98.6 & - & 21.6 & 1.3 \\ \hline
-%\end{tabular}
-%\caption{Estimated point prevalence of active infection with SARS-CoV-2 by demographics in Indiana.}
-%\label{tab:indiana}
-%\end{table}
-%
-%\item {\bf Facebook symptom survey}: We extract daily symptom surveys completed by individuals on Facebook who identify as living in Indiana.
+
+\begin{table}[th]
+\begin{tabular}{c | c | r r | r | r r }
+& & \multicolumn{2}{c}{Total Tests (\%)} & US & \multicolumn{2}{c}{Positive Test Rate (\%)}\\
+\cline{3-4} \cline{6-7}
+& & NonRandom & Random & Census & NonRandom & Random \\ \hline
+\multirow{2}{*}{Sex} & Female & 58.2 & 55.0 & 50.7 & 21.7 (11.2) & 1.4 \\
+& Male & 41.8 & 45.0 & 49.3 & 24.2 (12.4) & 2.1 \\ \hline
+\multirow{3}{*}{Age} & $<40$ & 39.4 & 28.0 & 52.7 & 29.7 (15.0) & 1.7 \\
+& $40-59$ & 41.1 & 36.0 & 25.2 & 24.9 (12.5) & 2.1 \\
+& $\geq 60$ & 19.5 & 36.0 & 22 & 6.7 (3.4) & 0.9 \\ \hline
+\multirow{2}{*}{Race} & White & 23.1 & 92.0 & 86.9 & 19.5 (9.6) & 1.5 \\
+& Nonwhite & 76.9 & 8.0 & 13.1 & 25.0 (12.3) & 3.4 \\ \hline
+% Hispanic & Hispanic & 44.1 & 2.0 & 7.1 & 37.6 & 6.9 \\
+% Origin & Non-Hispanic & 55.9 & 98.0 & 92.9 & 13.0 & 1.3 \\ \hline
+\multirow{2}{*}{Fever} & Yes & 17.0 & 1.8 & - & 66.4 (32.1) & 4.5 \\
+& No & 83.0 & 98.2 & - & 15.6 (7.5) & 1.3 \\ \hline
+Household & Yes & 10.8 & 1.4 & - & 46.1 & 29.4 \\
+$+$ Case & No & 89.2 & 98.6 & - & 21.6 & 1.3 \\ \hline
+Prior $+$ & Yes & 6.1 & 1.4 & - & 39.2 & 24.4 \\
+Test & No & 93.9 & 98.6 & - & 21.6 & 1.3 \\ \hline
+\end{tabular}
+\caption{Estimated point prevalence of active infection with SARS-CoV-2 by demographics in Indiana.}
+\label{tab:indiana}
+\end{table}
+
+\item {\bf Facebook symptom survey}: We extract daily symptom surveys completed by individuals on Facebook who identify as living in Indiana.
 We collect age, sex, and demographic information as well as COVID-19 related symptoms, which Facebook defines as having (1) a fever and cough, (2) shortness of breath, \emph{or} (3) difficulty breathing.
@@
 Table~\ref{tab:comparison} demonstrates minimal bias in the Facebook symptom survey.
-%\begin{table}
-%\centering
-%\begin{tabular}{c | c c c | c}
-%& \multicolumn{3}{c}{Facebook} & Random \\ \cline{2-5}
-%& Lower CI & Estimate & Upper CI & Estimate \\ \hline
-%Fever &  0.017 & 0.020 & 0.024 & 0.018 \\
-%Cough &  0.142 & 0.148 & 0.154 & 0.149 \\
-%Shortness & 0.058 & 0.060 & 0.062 & 0.062 \\ \hline
-%\end{tabular}
-%\caption{Comparison of Facebook symptom survey and random sample demonstrates minimal bias.}
-%\label{tab:comparison}
-%\end{table}
-%
-%% \item {\bf Hospitalization data}: On each day, we observe the number of individuals who are COVID-19 positive that visited a hospital on that day.
+\begin{table}
+\centering
+\begin{tabular}{c | c c c | c}
+& \multicolumn{3}{c}{Facebook} & Random \\ \cline{2-5}
+& Lower CI & Estimate & Upper CI & Estimate \\ \hline
+Fever &  0.017 & 0.020 & 0.024 & 0.018 \\
+Cough &  0.142 & 0.148 & 0.154 & 0.149 \\
+Shortness & 0.058 & 0.060 & 0.062 & 0.062 \\ \hline
+\end{tabular}
+\caption{Comparison of Facebook symptom survey and random sample demonstrates minimal bias.}
+\label{tab:comparison}
+\end{table}
+
+% \item {\bf Hospitalization data}: On each day, we observe the number of individuals who are COVID-19 positive that visited a hospital on that day.
 Unfortunately, hospitalization data is not broken out by age, gender, and racial demographic information.
@@
 We then use this to figure out how many individuals per strata were hospitalized and then back out how many
-%
-%\item {\bf Death data}: On each day, we observe the number of deaths due to COVID-19 reported on that day, with deaths broken out jointly by age, gender, and racial demographic information.
+
+\item {\bf Death data}: On each day, we observe the number of deaths due to COVID-19 reported on that day, with deaths broken out jointly by age, gender, and racial demographic information.
 Figure~\ref{fig:in-deaths} plots the cumulative number of COVID-19 related deaths over time by age group.
-%\end{itemize}
-%
-%
-%\subsection{Inverse-probability weighting approach}
-%
-%We start with using Table~\ref{tab:indiana} to compute IPW weights.
+\end{itemize}
+
+\subsection{Inverse-probability weighting approach}
+
+We start with using Table~\ref{tab:indiana} to compute IPW weights.
 The random sample is $n=3,658$.
@@
 For example, the strata defined as Male, 40--59, White, has a Fever, no household cases, and no prior tests has a weight proportional to
-%$$
-%\frac{\left( 0.493 \times 0.252 \times 0.869 \times 0.018 \right)}{\left(0.418 \times 0.411 \times 0.231 \times 0.17 \right)} = 0.628
-%$$
-%Using the constructed weights and the adjusted strata-level prevalence estimates, the IPW estimate under no measurement-error is $7.7$, a decrease of four percentage points from the observed prevalence of $11.7$.
+$$
+\frac{\left( 0.493 \times 0.252 \times 0.869 \times 0.018 \right)}{\left(0.418 \times 0.411 \times 0.231 \times 0.17 \right)} = 0.628
+$$
+Using the constructed weights and the adjusted strata-level prevalence estimates, the IPW estimate under no measurement-error is $7.7$, a decrease of four percentage points from the observed prevalence of $11.7$.
 Using Table~\ref{tab:indiana} to compute weights requires the assumption of independence among factors.
 To alleviate this, we compute weights using Facebook's survey data collected in Indiana over that window of time; the resulting IPW estimate under no measurement-error is
-%
-%The study did not report sensitivity and specificity; therefore, we take the suggested measures from~\cite{Woloshin2020} which report 70\% sensitivity is a reasonable estimates and~\cite{Cohen2020} which report 97.6\% specificity.
+
+The study did not report sensitivity and specificity; therefore, we take the suggested measures from~\cite{Woloshin2020} which report 70\% sensitivity is a reasonable estimates and~\cite{Cohen2020} which report 97.6\% specificity.
 This corresponds to a false negative rate of $30\%$ and false positive rate of $2.4\%$.
@@
 To account for uncertainty in
-%
-%Indiana's population as of 2019 was $6.732$ million.
+
+Indiana's population as of 2019 was $6.732$ million.
 Subtracting off the $617,555$ tests that had already performed yields a sampling fraction of $f = 0.001$.
@@
 Adjusting for imperfect testing yields an estimate of $33.9\%$.
-%
-%
-%
-%For this estimate, we consider estimation of data quality $E_{\I} \left[ \rho_{\tilde I(X),Y} \right]$.
+
+For this estimate, we consider estimation of data quality $E_{\I} \left[ \rho_{\tilde I(X),Y} \right]$.
 In \cite{Meng2018}, estimation relied on observing the true outcome (i.e., election vote totals), which is not possible in the current crisis.
@@
 Instead, we aim to build sensitivity analyses that can aid in understanding the amount of information in observational COVID case-count data.
-%
-%
-%Under the assumption that the survey is a SRS from New York, the error is $17.6$\%.  Using~\eqref{eq:error}, we construct an estimate of the relative sampling rate as follows
-%\[
-%\rho D_M = \sqrt{\frac{f}{1-f}} \frac{\text{0.176}}{\sigma_Y} = 1.07 \times 10^{-2} \Rightarrow \Delta = 1.07 \times 10^{-3} \Rightarrow M = 2.3.
-%\]
-%We can perform a sensitivity analysis by considering the range of false negative and positive rates from~\cite{Bendavid2020}, which leads to a range for $M$ of $(2.21, 2.38)$.
+
+Under the assumption that the survey is a SRS from New York, the error is $17.6$\%.  Using~\eqref{eq:error}, we construct an estimate of the relative sampling rate as follows
+\[
+\rho D_M = \sqrt{\frac{f}{1-f}} \frac{\text{0.176}}{\sigma_Y} = 1.07 \times 10^{-2} \Rightarrow \Delta = 1.07 \times 10^{-3} \Rightarrow M = 2.3.
+\]
+We can perform a sensitivity analysis by considering the range of false negative and positive rates from~\cite{Bendavid2020}, which leads to a range for $M$ of $(2.21, 2.38)$.
 Therefore, the impact on bias of $\hat R_t$ appears moderate.
@@
 This simple tool helps us understand how sensitivity and specificity impact biases and  therefore how much we should trust conclusions based on these assumptions.
-%
-%
+
 %\subsubsection{Time-varying IPW estimator}
@@
 
+\section{Infection fatality rate (IFR) by age group}
+\label{app:ifr}
+
 \end{document}

### 2021-06-29 write-up/covid-writeup.tex
Editing the first analysis of sample / checking'

@@
 
-We start with using Table~\ref{tab:indiana} to compute IPW weights.
-The random sample is $n=3,658$.
-Here, we demonstrate how to compute the weight for a particular strata.
-For example, the strata defined as Male, 40--59, White, has a Fever, no household cases, and no prior tests has a weight proportional to
+% FP: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7307014/
+% FN: https://www.medrxiv.org/content/10.1101/2020.04.16.20066787v2.full.pdf
+
+We start by using Table~\ref{tab:indiana} to compute IPW weights for end of April, early May.
+The random sample is $n=3,658$ and IPW weights are computed per strata which depend on Gender, Age, Race, Fever, Positive Case in Household, and Prior Positive Test.
+The weight for the strata defined as Male, 40--59, White, has a Fever, no household cases, and no prior tests is proportional to:
 $$

### 2021-06-29 methods/7_indiana_randomsample_analysis.R

@@ -96,8 +96,8 @@ for(n in 1:2) {
 
 haty_nomem = sum(NR_array_test * weights)
 
-FP = 0.05
-FN = 0.30
+FP = 0.03 # https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7307014/
+FN = 0.13 # https://www.medrxiv.org/content/10.1101/2020.04.16.20066787v2.full.pdf
 
 haty = (haty_nomem - FP)/(1-FP-FN)
 
@@ -239,7 +239,7 @@ rbind(april, july, october)
 ## Death data
 
 deaths = read.csv("data/covid_report_death_date_agegrp.csv", header = T)
-deaths$date = mdy_hm(deaths$ï..date)
+deaths$date = mdy_hm(deaths$?..date)
 age = unique(deaths$agegrp)[1]
 subcase = deaths[deaths$agegrp == age,]
 temp = aggregate(covid_deaths ~ date, subcase, FUN = sum)

### 2021-06-29 visualizations/mem_heatmap.R

@@ -34,7 +34,7 @@ row.names(result) = round(seq_oddsratio,3)
 colnames(result) = round(seq_M,2)
 
 library("gplots")
-png(filename = "./figs/mem_heatmap.png",
+png(filename = "../figs/mem_heatmap.png",
     width = 960, height = 960, units = "px", pointsize = 16)
 heatmap.2(as.matrix(result), col = bluered(100), dendrogram='none',
           trace = "none", density.info = "none", Rowv=FALSE,

### 2021-07-02 write-up/covid-writeup.tex
Update covid-writeup.tex
Fixed the M calc

@@
 & $40-59$ & 41.1 & 36.0 & 25.2 & 24.9 (12.5) & 2.1 \\
-& $\geq 60$ & 19.5 & 36.0 & 22 & 6.7 (3.4) & 0.9 \\ \hline
+& $\geq 60$ & 19.5 & 36.0 & 22.1 & 6.7 (3.4) & 0.9 \\ \hline
 \multirow{2}{*}{Race} & White & 23.1 & 92.0 & 86.9 & 19.5 (9.6) & 1.5 \\
@@
 & No & 83.0 & 98.2 & - & 15.6 (7.5) & 1.3 \\ \hline
-Household & Yes & 10.8 & 1.4 & - & 46.1 & 29.4 \\
-$+$ Case & No & 89.2 & 98.6 & - & 21.6 & 1.3 \\ \hline
-Prior $+$ & Yes & 6.1 & 1.4 & - & 39.2 & 24.4 \\
-Test & No & 93.9 & 98.6 & - & 21.6 & 1.3 \\ \hline
+Household & Yes & 10.8 & 1.4 & - & 46.1 (22.4) & 29.4 \\
+$+$ Case & No & 89.2 & 98.6 & - & 21.6 (10.4) & 1.3 \\ \hline
+Prior $+$ & Yes & 6.1 & 1.4 & - & 39.2 (20.2) & 24.4 \\
+Test & No & 93.9 & 98.6 & - & 21.6 (11.1) & 1.3 \\ \hline
 \end{tabular}
@@
 $$
-\frac{\left( 0.493 \times 0.252 \times 0.869 \times 0.018 \right)}{\left(0.418 \times 0.411 \times 0.231 \times 0.17 \right)} = 0.628
+\frac{\left( 0.493 \times 0.252 \times 0.869 \times 0.018 \times 0.986 \times 0.986 \right)}{\left(0.418 \times 0.411 \times 0.231 \times 0.17 \times 0.892 \times 0.939 \right)} \approx 0.334
 $$
 Using the constructed weights and the adjusted strata-level prevalence estimates, the IPW estimate under no measurement-error is $7.7$, a decrease of four percentage points from the observed prevalence of $11.7$.
-Using Table~\ref{tab:indiana} to compute weights requires the assumption of independence among factors.
-To alleviate this, we compute weights using Facebook's survey data collected in Indiana over that window of time; the resulting IPW estimate under no measurement-error is
+We also compute weights using Facebook's survey data collected in Indiana over that window of time but ignoring the household and prior testing information.
+The resulting IPW estimate under no measurement-error is $6.$
 
@@
 This corresponds to a false negative rate of $30\%$ and false positive rate of $2.4\%$.
-Therefore, the IPW estimate accounting for measurement-error is $7.7\%$.
+Therefore, the IPW estimates accounting for measurement-error are $5.6\%$ and $4.2\%$ respectively.
 To account for uncertainty in
 
-Indiana's population as of 2019 was $6.732$ million.
-Subtracting off the $617,555$ tests that had already performed yields a sampling fraction of $f = 0.001$.
-The smoothed prevalence estimate is 28.4\%.
-Adjusting for imperfect testing yields an estimate of $33.9\%$.
-
-For this estimate, we consider estimation of data quality $E_{\I} \left[ \rho_{\tilde I(X),Y} \right]$.
-In \cite{Meng2018}, estimation relied on observing the true outcome (i.e., election vote totals), which is not possible in the current crisis.
-Here, we use the random sample as a noisy but unbiased estimate of the true prevalence.
-Our goal is not inferential.
-Instead, we aim to build sensitivity analyses that can aid in understanding the amount of information in observational COVID case-count data.
-
-Under the assumption that the survey is a SRS from New York, the error is $17.6$\%.  Using~\eqref{eq:error}, we construct an estimate of the relative sampling rate as follows
-\[
-\rho D_M = \sqrt{\frac{f}{1-f}} \frac{\text{0.176}}{\sigma_Y} = 1.07 \times 10^{-2} \Rightarrow \Delta = 1.07 \times 10^{-3} \Rightarrow M = 2.3.
-\]
-We can perform a sensitivity analysis by considering the range of false negative and positive rates from~\cite{Bendavid2020}, which leads to a range for $M$ of $(2.21, 2.38)$.
-Therefore, the impact on bias of $\hat R_t$ appears moderate.
-If $FP = 0.05$ and $FN = 0.005$, then the relative sampling rate increases to $M=3.04$.
-This simple tool helps us understand how sensitivity and specificity impact biases and  therefore how much we should trust conclusions based on these assumptions.
-
-%\subsubsection{Time-varying IPW estimator}
-%
-%\emph{Need to show Hospitalization C+ versus Covid count to show that most of the tests are in that subset for early parts.}
-%
-%
-%
-%
-%
-%
 %\subsection{Model-based estimation}
@@
 %
-%
-%% \section{Cross-population comparisons}
-%% \label{section:crosspop}
-%% So far we have focused on understanding the limitations of using case-count data to understand population quantities of interest for a \emph{single} population.
-Many are interested in cross-population comparisons to contrast the impact of countries' mitigation policies.
-Many pieces in the media as well as academic articles have plots of case-count over time aligned by time of first known case.
-Others have claimed such comparisons unfair due to unequal population sizes and instead plotted case-count per million.
-In this section, we focus on the statistical issues relevant to such comparisons.
-%
-%% \subsection{Prevalence comparisons}
-%
-%% At a fixed time, suppose prevalence estimates $\bar y_1$ and $\bar y_2$ are observed for two populations. We can express $\bar y_1 - \bar y_2$ in statistical terms as
-%% $$
-%% (\bar Y_1 - \bar Y_2) + \left[ \rho_{I_1, Y_1} \times D_{M_1} \times \sqrt{\frac{1-f_1}{f_1}} \times \sigma_{Y_1}  - \rho_{I_2, Y_2} \times D_{M_2} \times \sqrt{\frac{1-f_2}{f_2}} \times \sigma_{Y_2} \right].
-%% $$
-%% The first term is the difference in population prevalence (i.e., the quantity of interest).
-If random sampling were performed in each population, $E_{\I} [ \rho_{I_j, Y_j}] = 0$ and so the estimate would be unbiased.
-However, the second term represents the complex error that results from selection bias.
-%% The particular test of interest is $H_0: \bar Y_1 = \bar Y_2$.  Under the null, $\sigma_{Y_1} = \sigma_{Y_2}$.  In classical statistics, we would compute a Z-score:
-%% $$
-%% \begin{aligned}
-%% \frac{\bar y_1 - \bar y_2}{ \sqrt{V_{SRS} (\bar y_1 - \bar y_2)} } &=
-%% \frac{\rho_{I_1, Y_1} \times D_{M_1} \times \sqrt{\frac{1-f_1}{f_1}}  - \rho_{I_2, Y_2} \times D_{M_2} \times \sqrt{\frac{1-f_2}{f_2}} }{ \sqrt{\frac{1}{N_1 -1} \times \frac{1-f_1}{f_1} + \frac{1}{N_2 - 1} \times \frac{1-f_2}{f_2}} }  \\
-%% &\overset{\star}{=}
-%% \sqrt{\frac{(N_1-1)(N_2 -1)}{N_1 + N_2}} \left[ \rho_{I_1, Y_1} - \rho_{I_2, Y_2} \right]
-%% \end{aligned}
-%% $$
-%% Where the last equality holds under $f_1 = f_2$, i.e., same sampling rate relative to population size.
-This tells us that the exact error in the comparison of sample means, as an estimate of the difference in population means, scales as a function of the difference in data quality $\rho_{I_1, Y_1} - \rho_{I_2, Y_2}$ times a square root of a population adjustment.
-For highly unequal population sizes (e.g., $N_1 \gg N_2$), the adjustment is approximated well by the square root of the smaller population size minus one (e.g.
-$\sqrt{N_2 -1}$).
-%
-%% Take, for example, a comparison of the US and Canada.
-The two population sizes are approximately 328 million and 38 million people.
-Then the population adjustment is $5.84$ million.
-As of May 5th, Canada has performed a total of 24.92 tests per 1,000 people while the US has performed a total of 22.01 tests per 1,000 people\footnote{According to \href{https://ourworldindata.org/grapher/full-list-cumulative-total-tests-per-thousand}{ourworldindata.org}} so the data quantities are approximately equal to $f = 23/1000 = 0.023$.
-Then, under a prevalence of 10\%, one would need the differential $\Delta_1-\Delta_2$ to be much smaller than $11.68 \times 10^{-6}$ in order for the selection bias to not impact.
-So only under the very strong assumptions of equal data quantity and quality will the comparisons be valid and the $Z$-score can be treated as in the classical setting.
-A simple alternative is to use the effective sample sizes to build more appropriate $Z$-score.
-That is, setting
-%% $$
-%% Z_n = \frac{\bar y_1 - \bar y_2}{ \frac{1-f}{f} \sigma_y \sqrt{ \frac{1}{n_{1,eff} - 1} + \frac{1}{n_{2,eff} - 1}}}
-%% $$
-%% allows us to build Z-scores that account for the true sample size.
-While still not perfect, this allows us to build Z-scores that are not overly confident due to selection bias and measurement error.
-See Section~\ref{section:est_dq} for discussion of how to estimate data quality.
-%
+
+\subsection{Disease prevalence by April 2020}
+
+Indiana's population as of 2019 was $6.732$ million.
+Subtracting off the $617,555$ tests that had already performed yields a sampling fraction of $f = 0.001$.
+The smoothed prevalence estimate is 28.4\%.
+Adjusting for imperfect testing yields an estimate of $33.9\%$.
+
+For this estimate, we consider estimation of data quality $E_{\I} \left[ \rho_{\tilde I(X),Y} \right]$.
+In \cite{Meng2018}, estimation relied on observing the true outcome (i.e., election vote totals), which is not possible in the current crisis.
+Here, we use the random sample which estimates the true prevalence of active COVID-19 disease at 1.81\%~\cite{Yiannoutsos2021}.
+Our goal is not inferential.
+Instead, we aim to build sensitivity analyses that can aid in understanding the amount of information in observational COVID case-count data.
+
+The non-weighted estimate error is $8.6$\%.  Using~\eqref{eq:statdecomp}, we construct an estimate of the relative sampling rate as follows
+\[
+\rho D_M = \sqrt{\frac{f}{1-f}} \frac{\text{0.085}}{\sigma_Y} = 3.84 \times 10^{-2} \Rightarrow \Delta = 1.39 \times 10^{-2} \Rightarrow M = 6.1.
+\]
+For the IPW estimates, $3.8$\% and $2.4$\% respectively
+\[
+\tilde \rho D_M = \sqrt{\frac{f}{1-f+CV_W^2}} \frac{\text{0.038}}{\sigma_Y} = 6.93 \times 10^{-3} \Rightarrow \tilde \Delta = 8.24 \times 10^{-3}  \Rightarrow \tilde M = 3.9.
+\]
+For the IPW estimate using Facebook data, we have $\tilde M = 3.3$.
+
+We can perform a sensitivity analysis by considering the range of false negative and false positive rates.
+Here, we assume sensitivity between $81$\% and $91$\% and specificity between $95$\% and $98.8$\%~\cite{Katz2020}, which leads to a range for $M$ of $(5.60, 6.39)$ in the unweighted analysis.
+Therefore, the impact on bias of $\hat R_t$ appears moderate.
+
+This simple tool helps us understand how sensitivity and specificity impact biases and  therefore how much we should trust conclusions based on these assumptions.
+
+%\subsubsection{Time-varying IPW estimator}
+%
+%\emph{Need to show Hospitalization C+ versus Covid count to show that most of the tests are in that subset for early parts.}
+
 %\section{Discussion}
@@
 \end{table}
+
+For binary outcomes, we have
+\begin{equation} \label{eq:binaryrho}
+\rho_{\tilde I (X),Y} = \tilde \Delta \sqrt{\frac{\bar Y (1 - \bar Y)}{f (1-f) \text{E}(W_J \mid I_J = 1)^2 + f \text{Var}(W_J \mid I_J = 1)} }
+\end{equation}
 

### 2021-07-12 write-up/covid-writeup.tex
Update covid-writeup.tex
Edits through section 3.4

@@
 \usepackage{enumitem}
+\usepackage{caption,subcaption}
 
@@
 
-Due to monetary and time constraints, however, random testing is either not performed or performed very infrequently.
+Due to monetary and time constraints, however, random testing is either not performed or performed infrequently.
 As of June 2021, Indiana is the only state to conduct statewide random sample testing\footnote{This is the only random sample to collect both seroprevalence and diagnostic testing results.
@@
 
-Note that the proposed approach requires covariate information to be collected on tested individuals.
-Unfortunately, many states do not require or report auxiliary covariate information beyond demographics such as gender, age, race, and ethnicity.
-To address this, we discuss how the proposal suggests minor changes to current practice that may improve performance over time.
+The proposed approach requires covariate information to be collected on individuals who receive a COVID-19 test.
+Unfortunately, many states do not require or report auxiliary covariate information beyond basic demographic information (e.g., gender, age, race, and ethnicity).
+We end with a brief list of suggestions of minor changes to current practice based on the proposed methodology.
 While we demonstrate empirical improvements over simple disease prevalence estimates, we also highlight how selection bias may persist and impact uncertainty quantification.
@@
 
-\subsection{Outline}
-
-In Section~\ref{section:data}, we summarize the publicly available data on COVID-19.
+% \subsection{Outline}
+
+% In Section~\ref{section:data}, we summarize the publicly available data on COVID-19.
 We highlight how testing information is collected as well as demographic and comorbidity information.
@@
 
-In Section~\ref{section:casecount}, we present simple mathematical arguments to show that unadjusted prevalence rates are (unsurprisingly) biased when tests are imperfect.
+% In Section~\ref{section:casecount}, we present simple mathematical arguments to show that unadjusted prevalence rates are (unsurprisingly) biased when tests are imperfect.
 What is surprising is that the direction and magnitude of the bias can vary substantially and interacts with the sampling fraction.
@@
 
-In Section~\ref{section:improvedcasecount}, we present a doubly robust estimation method for combining epidemiological forecasts with inverse-probability weighting methods to estimate prevalence.
+% In Section~\ref{section:improvedcasecount}, we present a doubly robust estimation method for combining epidemiological forecasts with inverse-probability weighting methods to estimate prevalence.
 Weight estimation requires auxiliary information provided by random samples.
@@
 
-In Section~\ref{section:applications}, the IPW and doubly-robust estimation methods are applied to the case study of Indiana statewide prevalence from March 2020 until June 2021.
+% In Section~\ref{section:applications}, the IPW and doubly-robust estimation methods are applied to the case study of Indiana statewide prevalence from March 2020 until June 2021.
 We estimate data quality for IPW estimates using the random sample on April 28th.
@@
 
-In Section~\ref{section:discussion}, a series of recommendations are provided that aim to help improve statistical inference in settings with large nonprobability samples from a stochastic process where small random sampling over time is possible.
+% In Section~\ref{section:discussion}, a series of recommendations are provided that aim to help improve statistical inference in settings with large nonprobability samples from a stochastic process where small random sampling over time is possible.
 
@@
 
-In this section we provide the necessary background to understand COVID-19 diagnostic and serologic testing, their respective scientific uses in managing the pandemic, and the data streams we consider in this manuscript.
+In this section we provide the necessary background to understand COVID-19 diagnostic and serologic testing, discuss their respective scientific uses in managing the pandemic, and introduce the data streams considered in this manuscript.
 
@@
 Specimen collection by swab is known to impact false negative/positive rates regardless of test timing.
-Systematic reviews suggest that 70\% sensitivity and 95\% specificity are reasonable estimates for current RT-PCR tests~\cite{Woloshin2020}.
+Systematic reviews suggest that 70\% sensitivity and 97.6\% specificity are reasonable estimates for current RT-PCR tests~\cite{Woloshin2020,Cohen2020}.
 
@@
 
-Our primary goal is accurate estimates of the population-level active infection rates over time using \emph{publicly available} viral testing data.
+The primary goal of this manuscript is to produce accurate estimates of the population-level active infection rates over time using \emph{publicly available} viral testing data.
 Secondary goals include estimation of rates of change and the effective reproduction number which characterize disease trajectory.
@@
 Coronavirus case-count data refers to the number of positive molecular tests performed on each day.
-Figure~XX plots (dotted orange line) the number of reported confirmed COVID-19 cases per day in the state of Indiana.
-Total number of COVID-19 diagnostic tests performed per day is also reported (Figure~XX, dotted blue line).
-On August 5th, 2020, for example, there were 1075 total daily positive cases and 11,663 total daily tests.
+Figure~\ref{fig:in-cases} plots the number of reported confirmed COVID-19 cases per day in the state of Indiana.
+Figure~\ref{fig:in-tests} plots the total number of COVID-19 molecularw tests performed per day.
+Figure~\ref{fig:in-deaths} plots the total number of COVID-19 related reported deaths per day.
 Public databases maintained by \href{https://bit.ly/2UqFSuA}{Johns Hopkins University} and the \href{https://bit.ly/2vUHfrK}{New York Times} provide accessible incoming county-level information of confirmed cases and deaths.
@@
 Unfortunately, most dashboards do not provide any demographic information on who requested a test nor on who tested positive for SARS-CoV-2 over time.
-Working closely with the State of Indiana, we were able to generate publicly available datasets for COVID-19 total tests, positive tests, and related deaths per day broken out marginally by age, gender, ethnicity, and race~\citep{IndianaData2021}.
+Working closely with the State of Indiana, we were able to produce publicly available datasets for COVID-19 total tests, positive tests, and related deaths per day broken out marginally by age, gender, ethnicity, and race~\citep{IndianaData2021}.
 This granular dataset motivates the proposed approach.
@@
 
-Due to testing restrictions and other potential selection biases, publicly reported covid testing data may not be sufficient to understand active infection rates or the disease trajectory.
-Here, we discuss two random samples that can be leveraged to get either unbiased estimates of the statewide active infection rate or as auxiliary information to address selection bias in the testing data.
+Due to testing restrictions and other potential selection biases, publicly reported Covid testing data may not be sufficient to understand active infection rates or the disease trajectory.
+Here, we discuss two random samples that can be leveraged to as auxiliary information to address selection bias.
 
@@
 The survey defines an individual as displaying \emph{COVID-like symptoms} if they exhibit a fever along with a cough, or shortness of breath, or difficulty breathing.
-Figure~\ref{fig:fbsymptoms} displays nonparametric regression estimates of the fraction of individuals within age and gender strata.
-In this paper, we use this as the main source of auxiliary information on time-varying characteristics of the population of Indiana, e.g., fraction of popopulation displaying symptoms or in contact with COVID-19 positive individuals).
+Figure~\ref{fig:fbsymptoms} displays smoothed estimates of the fraction of individuals within age and gender strata who report a fever within the past 24-hours.
+This paper uses the Facebook symptom survey as the main source of auxiliary information on time-varying characteristics of the population of Indiana, e.g., fraction of population displaying symptoms or in contact with COVID-19 positive individuals).
 
@@
 \includegraphics[width = 0.9\textwidth]{../figs/fbcovid19symptoms.png}
-\caption{Reported rate of COVID-like symptoms in strata.
-Daily rates were estimated using weighted method on each day separately and then smoothed over time using local-linear nonparametric regression.}
+\caption{Rate of reported fever per strata.  Daily rates were estimated using weighted method on each day separately and then smoothed over time using local-linear nonparametric regression.}
 \label{fig:fbsymptoms}
@@
 \end{figure}
+
+%% GOOD THROUGH HERE
 
@@
 
-The above selection mechanisms are random and independent of the outcome of interest.
-When this is not the case, selection effects may cause bias in the above estimates.
-To better understand this issue, \cite{Meng2018} provided the following intuitive and powerful statistical decomposition of the error between $\bar y$ and the true proportion $\bar Y$
+These random selection mechanisms are independent of the outcome of interest.
+When this is not the case, selection effects may cause bias.
+To better understand this issue, \cite{Meng2018} recently provided the following intuitive and powerful statistical decomposition of the error between $\bar y$ and the true proportion $\bar Y$
 $$
@@
 Here we investigate the interplay between imperfect testing and selection bias.
-In discussions of inaccurate testing, the standard assumption is measurement error leads to parameter attenuation.
+Researchers most often assume measurement error will lead to parameter attenuation.
 When paired with selection bias, however, the two sources of error become entangled, and resulting errors can become magnified, muted, or even switch signs.
@@
 The first term represents \emph{data quality}, the second \emph{data quantity}, and the third \emph{problem difficulty}.
-The final term represents the \emph{imperfect testing adjustment} which is again a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence but here is scaled by the factor $(1 - FP - FN)^{-1}$ compared to prior decomposition.
+The final term represents the \emph{imperfect testing adjustment} which is a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence but here is scaled by the factor $(1 - FP - FN)^{-1}$.
 
 Figure~\ref{fig:heatmap} shows that $D_M$, as a function of the relative frequency ($f_1/f_0$) and odds ratio, can be both positive and negative as well as a range of magnitudes~\citep{Beesley2020,Beesley2019,Smeden2019}.
-Under perfect testing, $D_M = 1$ so the relation between estimation and selection bias is simple, e.g., COVID-19 positive individuals more likely to receive test implies upward bias in prevalence estimates.
-When tests are imperfect, this simple relationship no longer holds.
+Under either perfect testing (i.e., $FP=FN=0$), $D_M = 1$ so the relation between estimation and selection bias is simple, e.g., COVID-19 positive individuals more likely to receive test implies upward bias in prevalence estimates.
+Under random testing (i.e., $f_0 = f_1$), $D_M = (1-FP-FN)^{-1}$ so measurement error simply scales the error.
+When tests are imperfect and selection bias exists, this simple relationship no longer holds.
 
@@
 \includegraphics[width = 0.4\textwidth]{../figs/mem_heatmap_article.png}
-\caption{Imperfect testing adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.005$ and $FN=0.172$.
-Color scaled so blue = $-10$, white = $0$, and red = $10$.
-\textcolor{red}{NEED TO DO FOR FP = 0.05 and FN = 0.3}}
+\caption{Imperfect testing adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.024$ and $FN=0.30$. Color scaled so blue = $-10$, white = $0$, and red = $10$.}
+%\textcolor{red}{NEED TO DO FOR FP = 0.05 and FN = 0.3}}
 \label{fig:heatmap}
@@
 If true then the ratio of case-counts at consecutive times may be a good estimate of the true change in prevalence, helping scientists understand the disease trajectory.
-In this section, we demonstrate how selection bias and imperfect testing impact such estimates.
+We next demonstrate how selection bias and imperfect testing impact such estimates.
 
@@
 $$
-where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement error adjustment, and $CV(Y_j) = \sigma_{Y_j}/Y_j$ is the coefficient of variation on day $j$.
+where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement error adjustment, and $CV(Y_j) = \sigma_{Y_j}/ \bar Y_j$ is the coefficient of variation on day $j$.
 See Appendix~\ref{app:ratio} for the derivation.
-The error magnitude depends on the true rate $\bar Y_{t} / \bar Y_{t-1}$ so a large decrease will have a relatively small error compared to a large increase.
-The second term represents potential \emph{cancellation} which may occur when data quality, sampling fraction, measurement error, and prevalence are all constant across time.
+The error magnitude depends on the true rate $\bar Y_{t} / \bar Y_{t-1}$ so a large decrease will have a small error relative to a large increase.
+The second term represents potential \emph{cancellation} which can  occur when data quality, sampling fraction, measurement error, and prevalence are constant across time.
 
@@
 In terms of bias, the rate is overestimated prior to the peak and underestimated afterwards; the bias increases dramatically when the relative fraction exceeds $2$.
-From a decision making perspective, such biases have a substantial impact.
-First, the overestimation may give lawmakers and governors more leverage in proposing aggressive actions that reduce prevalence.
-Of course, the analysis supports the argument that the estimates based on available data may be pessimistic.
-What is missing in the current public health discourse is that the bias direction of the prevalence estimate is non-constant over time.
-Underestimation post-peak puts pressure on lawmakers and governors to prematurely relax social distancing measures.
+Such biases may have a substantial impact on policy making.
+First, overestimation may give policy makers more leverage in proposing aggressive actions to reduce prevalence.
+Of course, the analysis provides evidence that estimates based on available data may be pessimistic.
+What is missing in the current discourse is that the direction of bias is non-constant over time.
+Underestimation post-peak puts pressure on policy makers to prematurely relax social distancing measures.
 Moreover, it appears that the peak time is the easiest to estimate in terms of having minimal error using available data.
 Note, however, that the standard errors will be large due to small effective sample sizes.
+
+%%  MADE IT TO HERE
 
@@
 \label{lemma:ipw}
-\textcolor{red}{Must check.}
+% \textcolor{red}{Must check.}
 The estimates~$\hat \mu_t := \bar y_{n,t}^\star$, $\hat \pi_{j,t}$, $\hat FP$, and $\hat FN$ are solutions to the following set of estimating equations
@@
 where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
-Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$ (check)  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ where $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$.
+Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ where $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$.
 \end{lemma}
@@
 Following~\cite{Song2020}, we employ Runge-Kutta (RK4) approximations for discretization.
-Here, we consider covariate information to take the form of a stratificatoin variable $k$ with $K$ strata.
+Here, we consider covariate information that takes the form of a stratification variable with $K$ strata.
 Let $\nu_{k,t}$ to denote the number of new infections on day $t$ in the $k$th strata.
-AS there are a total of $N \cdot i_t := I_t$ new infections on day $t$, we assume the joint distribution of strata-specific new infection totals takes a multinomial distribution
-$$
-( \nu_{1,t}, \ldots, \nu_{K,t} ) \sim \text{Multinomial} \left( I_t, (p_{i1,t}, \ldots, p_{iK, t}) \right).
-$$
-
-Selection bias in case count data is addressed by bringing in a different source of information, COVID-19 death data.
-Let $D_{k,r}$ denote the number of individuals who pass away on day $r$ from strata $k$.
-Then
+As there are a total of $N \cdot i_t := I_t$ new infections on day $t$, we assume the joint distribution of strata-specific new infection totals takes a multinomial distribution
+$$
+( \nu_{1,t}, \ldots, \nu_{K,t} ) \sim \text{Multinomial} \left( I_t, (p_{1,t}, \ldots, p_{K, t}) \right).
+$$
+
+Selection bias in case count data is addressed by bringing in COVID-19 death data.  Let $D_{k,r}$ denote the number of individuals who pass away on day $r$ from strata $k$.  Then
 $$
@@
 where $p_{k}$ is the infection fatality rate for the $k$th strata.
-This extends prior analysis of death data~\cite{Johndrow2020} by allowing the infection fatality rate to depend on age.
-Indeed,~\cite{Levin2020} demonstrate that ``the estimated IFR is close to zero for children and younger adults but rises exponentially with age, reaching 0.4\% at age 55, 1.4\% at age 65, 4.6\% at age 75, and 15\% at age 85.''  See Appendix~\ref{app:ifr} for details on choice of $p_k$.
+This extends prior analysis of death data~\cite{Johndrow2020} by allowing the infection fatality rate to depend on stratification variable.
+Specifically, ``the estimated IFR is close to zero for children and younger adults but rises exponentially with age, reaching 0.4\% at age 55, 1.4\% at age 65, 4.6\% at age 75, and 15\% at age 85''~\cite{Levin2020}.
 
@@
 % 65+ years old: 90,000/1000000 * 100 -> 9.0
-
 %
@@
 where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
-Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$ (check)  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ where $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$.
+Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$ and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ where $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$.
 \end{lemma}
@@
 
-Here we consider estimation of the active infection rate in Indiana. We consider five sources of data:
+Here we consider estimation of the active infection rate in Indiana using unweighted, IPW, model-based, and doubly robust estimates. To do so, we consider five data sources:
 \begin{itemize}[leftmargin=*]
 \item {\bf Testing data}:  Each day, the number of tests performed and number of positive tests are reported, with counts broken out jointly by age, gender, and racial demographic information.
-Figure~\ref{fig:in-cases} plots the number of COVID-19 positive cases based on testing date by age.
+Figure~\ref{fig:in-cases} and~\ref{fig:in-tests} plots daily COVID-19 positive cases and tests by age strata.
 
@@
  \centering
- \includegraphics[width=.9\linewidth]{../figs/indianacasecounts.png}
- \caption{Indiana case-counts per day}
+ \includegraphics[width=.9\linewidth]{../figs/indianacasecounts_byage.png}
+ \caption{Case-counts per day}
  \label{fig:in-cases}
@@
  \centering
- \includegraphics[width=.9\linewidth]{../figs/indianadeaths.png}
- \caption{Indiana death-counts per day}
+ \includegraphics[width=.9\linewidth]{../figs/indianacovidtests_byage.png}
+\caption{Tests per day}
+\label{fig:in-tests}
+\end{subfigure} \\ [1ex]
+\begin{subfigure}{\linewidth}
+\centering
+\includegraphics[width=.45\linewidth]{../figs/indianadeaths_byage.png}
+ \caption{Deaths per per day}
  \label{fig:in-deaths}
 \end{subfigure}
+\caption{Indiana daily case count, testing, and death data by age strata}
 \end{figure}
@@
 The nonrandom sample is not a random subsample of the overall case-counts; however, this data provides important supplementary covariate information.
-Namely, the samples collected included covariate information on (1) prior symptoms, (2) COVID-19 positive individual in household, and (3) prior COVID-19 positive test.
+% Namely, the samples collected included covariate information on (1) prior symptoms, (2) COVID-19 positive individual in household, and (3) prior COVID-19 positive test.
 To account for the nonrandom sample being from high risk areas, we adjust estimates to match the statewide positivity rate of 11.7\% between May 1--2.
-These adjusted rates are presented in parentheses in Table~\ref{tab:indiana}.
+These adjusted rates are presented in parentheses in Table~\ref{tab:indiana} along with the relevant covariate information.
+
+\item {\bf Facebook symptom survey}: We extract daily symptom surveys completed by individuals on Facebook who identify as living in Indiana.
+We collect age, sex, and demographic information as well as COVID-19 related symptoms, which Facebook defines as having (1) a fever and cough, (2) shortness of breath, \emph{or} (3) difficulty breathing.
+To assess potential bias, we compare the random sample from Indiana to the Facebook symptom survey data collected from April 20th to April 25th.
+Table~\ref{tab:comparison} demonstrates minimal bias in the Facebook symptom survey.
+
+\begin{table}[!th]
+\centering
+\begin{tabular}{c | c c c | c}
+& \multicolumn{3}{c}{Facebook} & Random \\ \cline{2-5}
+& Lower CI & Estimate & Upper CI & Estimate \\ \hline
+Fever &  0.017 & 0.020 & 0.024 & 0.018 \\
+Cough &  0.142 & 0.148 & 0.154 & 0.149 \\
+Shortness & 0.058 & 0.060 & 0.062 & 0.062 \\ \hline
+\end{tabular}
+\caption{Comparison of Facebook symptom survey and random sample demonstrates minimal bias.}
+\label{tab:comparison}
+\end{table}
+
+% \item {\bf Hospitalization data}: On each day, we observe the number of individuals who are COVID-19 positive that visited a hospital on that day.
+Unfortunately, hospitalization data is not broken out by age, gender, and racial demographic information.
+Instead, we rely on combining Facebook data to calculate the fraction of those who sought the hospital per week in each strata.
+We then use this to figure out how many individuals per strata were hospitalized and then back out how many
+
+\item {\bf Death data}: Daily COVID-19 related deaths are observed by age, gender, and racial demographics.
+Figure~\ref{fig:in-deaths} plots the cumulative number of COVID-19 related deaths over time by age group.
+While most COVID-19 tests and positive cases are within the younger age strata, most deaths are within the 70+ age strata.
+\end{itemize}
 
@@
 
-\item {\bf Facebook symptom survey}: We extract daily symptom surveys completed by individuals on Facebook who identify as living in Indiana.
-We collect age, sex, and demographic information as well as COVID-19 related symptoms, which Facebook defines as having (1) a fever and cough, (2) shortness of breath, \emph{or} (3) difficulty breathing.
-To assess potential bias, we compare the random sample from Indiana to the Facebook symptom survey data collected from April 20th to April 25th.
-Table~\ref{tab:comparison} demonstrates minimal bias in the Facebook symptom survey.
-\begin{table}
-\centering
-\begin{tabular}{c | c c c | c}
-& \multicolumn{3}{c}{Facebook} & Random \\ \cline{2-5}
-& Lower CI & Estimate & Upper CI & Estimate \\ \hline
-Fever &  0.017 & 0.020 & 0.024 & 0.018 \\
-Cough &  0.142 & 0.148 & 0.154 & 0.149 \\
-Shortness & 0.058 & 0.060 & 0.062 & 0.062 \\ \hline
-\end{tabular}
-\caption{Comparison of Facebook symptom survey and random sample demonstrates minimal bias.}
-\label{tab:comparison}
-\end{table}
-
-% \item {\bf Hospitalization data}: On each day, we observe the number of individuals who are COVID-19 positive that visited a hospital on that day.
-Unfortunately, hospitalization data is not broken out by age, gender, and racial demographic information.
-Instead, we rely on combining Facebook data to calculate the fraction of those who sought the hospital per week in each strata.
-We then use this to figure out how many individuals per strata were hospitalized and then back out how many
-
-\item {\bf Death data}: On each day, we observe the number of deaths due to COVID-19 reported on that day, with deaths broken out jointly by age, gender, and racial demographic information.
-Figure~\ref{fig:in-deaths} plots the cumulative number of COVID-19 related deaths over time by age group.
-\end{itemize}
-
 \subsection{Inverse-probability weighting approach}
@@
 The random sample is $n=3,658$ and IPW weights are computed per strata which depend on Gender, Age, Race, Fever, Positive Case in Household, and Prior Positive Test.
-The weight for the strata defined as Male, 40--59, White, has a Fever, no household cases, and no prior tests is proportional to:
+The weight for the strata defined as Male, 40--59, White, has a Fever, no household cases, and no prior tests, for example, is proportional to:
 $$
@@
 Using the constructed weights and the adjusted strata-level prevalence estimates, the IPW estimate under no measurement-error is $7.7$, a decrease of four percentage points from the observed prevalence of $11.7$.
-We also compute weights using Facebook's survey data collected in Indiana over that window of time but ignoring the household and prior testing information.
-The resulting IPW estimate under no measurement-error is $6.$
-
-The study did not report sensitivity and specificity; therefore, we take the suggested measures from~\cite{Woloshin2020} which report 70\% sensitivity is a reasonable estimates and~\cite{Cohen2020} which report 97.6\% specificity.
+We also make use of Facebook's symptom survey over that window of time to construct a second IPW estimate conditoinal on demographics and fever but ignoring the household and prior testing information, which under no measurement-error is $6.5$\%.
+
+The Indiana study did not report sensitivity and specificity; therefore, we take the suggested measures from~\cite{Woloshin2020} which report 70\% sensitivity is a reasonable estimates and~\cite{Cohen2020} which report 97.6\% specificity.
 This corresponds to a false negative rate of $30\%$ and false positive rate of $2.4\%$.
@@
 
-%\subsection{Model-based estimation}
-%
-%\subsection{Doubly robust estimation}
+\subsubsection{Disease prevalence by April 2020}
+
+Indiana's population as of 2019 was $6.732$ million.
+A total of $898$ were administered between April 25th to 29th.
+Subtracting off the $95,879$ tests that had already performed yields a sampling fraction of $f = 2.96 \times 10^{-3}$.
+Next we consider estimation of data quality $E_{\I} \left[ \rho_{\tilde I(X),Y} \right]$.
+In \cite{Meng2018}, estimation relied on observing the true outcome (i.e., election vote totals), which is not possible in the current crisis.
+Here, we use the random sample which estimates the true prevalence of active COVID-19 disease at 1.81\%~\cite{Yiannoutsos2021}.
+Our goal is not inferential.
+Instead, we aim to build sensitivity analyses that can aid in understanding the amount of information in observational COVID case-count data.
+
+The unweighted estimate for the true prevalence of active COVID-19 disease between April 25th and April 29th is 11.7\%.
+Assuming a false negative rate of $3$\% and false positive rate of $3$\%, the unweighted estimate is $10.4$\%, leading to an error of $8.5$\%.
+Using~\eqref{eq:statdecomp}, we construct an estimate of the relative sampling rate as follows
+\begin{equation*}
+\rho D_M = \sqrt{\frac{f}{1-f}} \frac{\text{0.085}}{\sigma_Y} = 3.84 \times 10^{-2} \Rightarrow \Delta = 1.39 \times 10^{-2} \Rightarrow M = 6.1.
+\end{equation*}
+Using inverse-probability weights, the IPW estimates for the true prevalence of active COVID-19 disease between April 25th and April 29th are $7.7$\% and $6.5\%$ respectively.
+Under the same sensitivity and specificity, the weighted estimates adjusted for measurement-error are $5.6$\% and $4.2$\%, leading to errors $3.8$\% and $2.4$\% respectively.
+Using~\eqref{eq:statdecomp2}, we construct an estimate of the relative sampling rate as follows
+\begin{equation*}
+\tilde \rho D_M = \sqrt{\frac{f}{1-f+CV_W^2}} \frac{\text{0.038}}{\sigma_Y} = 6.93 \times 10^{-3} \Rightarrow \tilde \Delta = 8.24 \times 10^{-3}  \Rightarrow \tilde M = 3.9,
+\end{equation*}
+using the Indiana non-random sample.
+For the IPW estimate using Facebook data, we have $\tilde M = 3.3$.
+Therefore, the impact of self selection on bias of $\hat R_t$ appears strong using unweighted data, remaining moderate even using the weighted estimates.
+
+A sensitivity analysis is performed by considering the range of false negative and false positive rates.
+Here, sensitivity is assumed to fall between $81$\% and $91$\% and specificity between $95$\% and $98.8$\%~\cite{Katz2020}, which leads to a range for $M$ of $(5.60, 6.39)$ for the unweighted analysis, and $(3.63,4.07)$ and $(3.10, 3.45)$ for the weighted analyses respectively.
+
+\subsubsection{Time-varying IPW estimator}
+
+Here we extend the IPW analysis to the time-varying setting.
+To do so, we make use of the Facebook symptom survey, hospitalization, and testing data.
+Recall the COVID-19 testing and positive case counts are known by age, gender, and racial demographics.
+Symptom status (e.g., fever, coughing, shortness of breath) and related covariate information (COVID-19 positive contact), however, are currently unavailable\footnote{Whether this is due to data privacy considerations or lack of data collection efforts is unclear.}.
+As COVID-19 symptom status is likely to alter the testing propensity and likelihood of active infection, Facebook symptom survey and hospitalization data are both used to estimate symptom status per strata.
+We do this in two ways which are each detailed below.
+
+In the first approach, we estimate two logistic regressions using weighted pseudo-likelihoods with the Facebook survey symptom data.
+The first estimates the probability of contact with a COVID-19 positive individual given demographic and COVID-19 test status (positive or negative COVID-19 test).
+The second estimates the probability of fever given demographic, COVID-19 test status, \emph{and} whether the individual has had contact with a COVID-19 positive individual.
+Using these two models, mean imputation is used to calculate the number of positive and negative tests given demographic strata, fever status, and COVID-19 contact status.
+With the resulting dataset, inverse probability weights are computed based.
+
+In the second model, for negative cases,  people fever
+
+For positive cases, we impute the number of positive cases
+
+Figure~\ref{fig:testinglik_mainpaper} presents the testing propensity for non-hispanic, white males across age strata and fever status using the second approach.
+We note here that the ratio of testing propensity within strata across fever status is time-varying.
+The ratio starting nine times more likely in early April and
+
+Fever
+
+\begin{figure}[!th]
+\centering
+\begin{subfigure}{.5\textwidth}
+ \centering
+ \includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig1_mainpaper.png}
+ \caption{Testing Likelihood Given Fever}
+ \label{fig:testinglik1_mainpaper}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+ \centering
+\includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig2_mainpaper.png}
+ \caption{Testing Propensity Given No Fever}
+ \label{fig:testinglik2_mainpaper}
+\end{subfigure}
+\caption{Testing Propensity across age strata and fever status}
+\label{fig:testinglik_mainpaper}
+\end{figure}
+
+%\emph{Need to show Hospitalization C+ versus Covid count to show that most of the tests are in that subset for early parts.}
+
+\subsection{Model-based estimation}
+
+\begin{figure}[!th]
+\centering
+\includegraphics[width=.6\linewidth]{../figs/deaths_ppc.png}
+\caption{Posterior predictive check on aggregate daily death counts}
+\label{fig:deathsppc}
+\end{figure}
+
+\subsection{Doubly robust estimation}
 %
@@
 
-\subsection{Disease prevalence by April 2020}
-
-Indiana's population as of 2019 was $6.732$ million.
-Subtracting off the $617,555$ tests that had already performed yields a sampling fraction of $f = 0.001$.
-The smoothed prevalence estimate is 28.4\%.
-Adjusting for imperfect testing yields an estimate of $33.9\%$.
-
-For this estimate, we consider estimation of data quality $E_{\I} \left[ \rho_{\tilde I(X),Y} \right]$.
-In \cite{Meng2018}, estimation relied on observing the true outcome (i.e., election vote totals), which is not possible in the current crisis.
-Here, we use the random sample which estimates the true prevalence of active COVID-19 disease at 1.81\%~\cite{Yiannoutsos2021}.
-Our goal is not inferential.
-Instead, we aim to build sensitivity analyses that can aid in understanding the amount of information in observational COVID case-count data.
-
-The non-weighted estimate error is $8.6$\%.  Using~\eqref{eq:statdecomp}, we construct an estimate of the relative sampling rate as follows
-\[
-\rho D_M = \sqrt{\frac{f}{1-f}} \frac{\text{0.085}}{\sigma_Y} = 3.84 \times 10^{-2} \Rightarrow \Delta = 1.39 \times 10^{-2} \Rightarrow M = 6.1.
-\]
-For the IPW estimates, $3.8$\% and $2.4$\% respectively
-\[
-\tilde \rho D_M = \sqrt{\frac{f}{1-f+CV_W^2}} \frac{\text{0.038}}{\sigma_Y} = 6.93 \times 10^{-3} \Rightarrow \tilde \Delta = 8.24 \times 10^{-3}  \Rightarrow \tilde M = 3.9.
-\]
-For the IPW estimate using Facebook data, we have $\tilde M = 3.3$.
-
-We can perform a sensitivity analysis by considering the range of false negative and false positive rates.
-Here, we assume sensitivity between $81$\% and $91$\% and specificity between $95$\% and $98.8$\%~\cite{Katz2020}, which leads to a range for $M$ of $(5.60, 6.39)$ in the unweighted analysis.
-Therefore, the impact on bias of $\hat R_t$ appears moderate.
-
-This simple tool helps us understand how sensitivity and specificity impact biases and  therefore how much we should trust conclusions based on these assumptions.
-
-%\subsubsection{Time-varying IPW estimator}
-%
-%\emph{Need to show Hospitalization C+ versus Covid count to show that most of the tests are in that subset for early parts.}
-
-%\section{Discussion}
-%\label{section:discussion}
+\begin{figure}[!th]
+\centering
+\includegraphics[width=.6\linewidth]{../figs/tv_air.png}
+\caption{Active Infection Rate Estimates}
+\label{fig:testinglik_mainpaper}
+\end{figure}
+
+\section{Discussion}
+\label{section:discussion}
 %
@@
 
+\section{Image combos (temporary)}
+
+\subsection{Model 1}
+
+In this section, we describe.
+
+\begin{figure}[!th]
+\centering
+\begin{subfigure}{.5\textwidth}
+ \centering
+ \includegraphics[width=.9\linewidth]{../figs/tvprop_contact_fig1.png}
+ \caption{COVID-19 Contact Likelihood Given Negative Test}
+ \label{fig:contactlik1}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+ \centering
+\includegraphics[width=.9\linewidth]{../figs/tvprop_contact_fig2.png}
+ \caption{COVID-19 Contact Likelihood Given Positive Test}
+ \label{fig:contactlik2}
+\end{subfigure}
+\caption{Likelihood of COVID-19 contact}
+\label{fig:contactlik}
+\end{figure}
+
+We then fit a model for likelihood of fever given
+
+\begin{figure}[!th]
+\centering
+\begin{subfigure}{.5\textwidth}
+ \centering
+ \includegraphics[width=.9\linewidth]{../figs/tvprop_symptom_fig1.png}
+ \caption{Symptom Likelihood Given Negative Test}
+ \label{fig:symptomlik1}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+ \centering
+\includegraphics[width=.9\linewidth]{../figs/tvprop_symptom_fig2.png}
+ \caption{Symptom Likelihood Given Positive Test}
+ \label{fig:symptomlik2}
+\end{subfigure}
+\caption{Likelihood of COVID-19 contact}
+\label{fig:symptomlik}
+\end{figure}
+
+The testing propensity likelihood
+
+\begin{figure}[!th]
+\centering
+\begin{subfigure}{.5\textwidth}
+ \centering
+ \includegraphics[width=.9\linewidth]{../figs/tvprop_fig1.png}
+ \caption{COVID-10 Testing Likelihood Given Fever and COVID-19 Contact}
+ \label{fig:testinglik1}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+ \centering
+\includegraphics[width=.9\linewidth]{../figs/tvprop_fig2.png}
+ \caption{Testing Propensity}
+ \label{fig:testinglik2}
+\end{subfigure}
+\caption{COVID-10 Testing Likelihood Given no Fever nor COVID-19 Contact}
+\label{fig:testinglik}
+\end{figure}
+
+\newpage
+
+\subsection{Model 2 (Hospitalization)}
+
+We then fit a model for likelihood of fever given
+
+\begin{figure}[!th]
+\centering
+\begin{subfigure}{.5\textwidth}
+ \centering
+ \includegraphics[width=.9\linewidth]{../figs/tvprop_symptom_alt_fig1.png}
+ \caption{Symptom Likelihood Given Negative Test}
+ \label{fig:symptomlik1}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+ \centering
+\includegraphics[width=.9\linewidth]{../figs/tvprop_symptom_alt_fig2.png}
+ \caption{Symptom Likelihood Given Positive Test}
+ \label{fig:symptomlik2}
+\end{subfigure}
+\caption{Likelihood of COVID-19 contact}
+\label{fig:symptomlik}
+\end{figure}
+
+The testing propensity likelihood
+
+\begin{figure}[!th]
+\centering
+\begin{subfigure}{.5\textwidth}
+ \centering
+ \includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig1.png}
+ \caption{COVID-10 Testing Likelihood Given Fever and COVID-19 Contact}
+ \label{fig:testinglik1}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+ \centering
+\includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig2.png}
+ \caption{Testing Propensity}
+ \label{fig:testinglik2}
+\end{subfigure}
+\caption{COVID-10 Testing Likelihood Given no Fever nor COVID-19 Contact}
+\label{fig:testinglik}
+\end{figure}
+
 \end{document}

### 2021-07-12 visualizations/sir_example_plot.R
Update sir_example_plot.R

@@ -37,7 +37,7 @@ out2 <- as.data.frame(out2)
 out1$time <- NULL
 out2$time <- NULL
 ## Show data
-png(filename = "C:/Users/Balthazar/Documents/GitHub/covid-umich/methods/figs/sir.png",
+png(filename = "../figs/sir.png",
     width = 6.5, height = 4, units = "in", res = 1200, pointsize = 12)
 par(mar = c(2,2,1,1)+0.1)
 plot(times, out1$I, type = "l", axes = FALSE, bty = "n")

### 2021-07-12 write-up/covid-writeup.tex
Update covid-writeup.tex
Edits through 4.1

@@
 
-%%  MADE IT TO HERE
-
 \subsection{Estimation of effective reproduction number}
@@
 This implies a similar trade-off as before but on the logarithmic scale.
-The error is no longer scaled by $\bar Y_t/\bar Y_{t-1}$ but by the serial interval and does not depend on prevalence $\bar Y_t$ but on the fraction of new cases out of those susceptible $\bar K_t$.
+The error is no longer scaled by $\bar Y_t/\bar Y_{t-1}$ but by the serial interval and does not depend on prevalence but on the fraction of new cases out of those susceptible.
 This leads to differences in when the bias is most pronounced.
-The final term is the error due to varying numbers of susceptible individuals.
-Figure~\ref{fig:r0-bias} displays the bias as a function of the relative sampling fraction assuming the fraction of the population that is susceptible remains large.
+Figure~\ref{fig:r0-bias} displays the bias as a function of the relative sampling fraction assuming the population fraction that is susceptible remains large.
 
@@
 \caption{Potential bias due to self-selection and measurement error from a deterministic SIR model with $\beta = 1.2$ (black) and $\gamma = 0.15$.
-Here we assume $f = 0.02$, $FP = 0.005$, $FN = 0.172$, and use a range of relative sampling fraction $M = f_1/f_0$.}
+Here we assume $f = 0.02$, $FP = 0.024$, $FN = 0.30$, and use a range of relative sampling fraction $M = f_1/f_0$.}
 \label{fig:rates}
@@
 \caption{Left: fraction infected in two SIR models with $\beta = 1.2$ and $0.9$ respectively and $\gamma = 0.15$ with same initial conditions.
-Right: comparison of $\hat R_t$ across time with $FN = 0.172$, $FP = 0.005$, and $M = 4$.}
+Right: comparison of $\hat R_t$ across time with $FN = 0.30$, $FP = 0.024$, and $M = 4$.}
 \label{fig:comparison}
@@
 Typically, this is not possible.
-Here we present a method that uses auxiliary information obtained from probability samples as well as probability subsamples from the self-selected population.
+Here we present a method that uses auxiliary information obtained from probability samples.
 

### 2021-07-13 methods/7_indiana_randomsample_analysis.R
Update 7_indiana_randomsample_analysis.R

@@ -96,8 +96,8 @@ for(n in 1:2) {
 
 haty_nomem = sum(NR_array_test * weights)
 
-FP = 0.03 # https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7307014/
-FN = 0.13 # https://www.medrxiv.org/content/10.1101/2020.04.16.20066787v2.full.pdf
+FP = 0.024 # https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7307014/
+FN = 0.30 # https://www.medrxiv.org/content/10.1101/2020.04.16.20066787v2.full.pdf
 
 haty = (haty_nomem - FP)/(1-FP-FN)
 
@@ -153,8 +153,8 @@ for(l in 1:2) {
 
 fb_haty_nomem = sum(fb_NR_array_test * fb_weights)
  
-FP = 0.03 # https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7307014/
-FN = 0.13 # https://www.medrxiv.org/content/10.1101/2020.04.16.20066787v2.full.pdf
+FP = 0.024 # https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7307014/
+FN = 0.30 # https://www.medrxiv.org/content/10.1101/2020.04.16.20066787v2.full.pdf
 
 fb_haty = (fb_haty_nomem - FP)/(1-FP-FN)
 

### 2021-07-13 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 \subsubsection{Facebook/CMU symptom survey.}
+\label{subsection:fbsymptom}
 Since April 2020, Facebook has started randomly showing an opt-in symptom survey to its users~\citep{delphisurvey}.
@@
 Let $\tilde I_j$ be an indicator that the individual was included in the probability sample and $\tilde W_j$ be the probability of inclusion for the $j$th individual (i.e., equal to $n/N$ for a simple random sample of size $n$).
-The probability sample is used to construct a design-unbiased estimator of the second term
+In recent work,~\cite{Chen2019} use a probability sample to construct a design-unbiased estimator of the second term
 \begin{equation}
@@
 \end{equation}
-Taking the expectation of the second term with respect to the sampling design yields the second term in~\eqref{eq:propensity}.
-This extends the work by~\cite{Chen2019} which considered auxiliary information coming only from a simple random sample.
+Taking an expectation with respect to the sampling design yields~\eqref{eq:propensity}.
 % Under simple random sampling and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the $\hat \pi_k = (d_k / \tilde d_k) (n / N)$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.
-
-% Equation~\eqref{eq:auxinfoprob} presupposes covariate information is collected for all individuals in the nonprobability sample; however, this is often not feasible for state and local governments where rapid sample collection is prioritized over additional collection of covariate information.
-We can extend~\eqref{eq:auxinfoprob} by randomly sampling a subset of the nonprobability sample on which to collect the covariate information, yielding the pseudo-likelihood
-% \begin{equation}
-% \label{eq:auxinfoprob2}
-% \sum_{j=1}^N I_j I_j^{(2)} W_j^{(2)} \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
-% \end{equation}
-% Here, $I_j^{(2)}$ is an indicator that the individual is in the probability subsample and $W_j^{(2)}$ is the probability of inclusion.
-Estimation proceeds by maximizing the pseudo-likelihood which now depends on random sampling in both the first and second term.
 
@@
 
-In prior sections on selection propensity estimation, we ignored the dynamic nature of the outbreak.  Here we extend the IPW approach to the temporal setting by considering the joint likelihood
+Here we extend the IPW approach to the temporal setting to account for the dynamic nature of the outbreak by considering the joint likelihood
 \begin{equation}
@@
 Unfortunately for COVID-19, large probabilistic samples are not available at every time within a given region.
-For simplicity, here we discuss a non-parametric kernel-based approach where the selection propensity at time~$t$, denoted $\hat \theta_t$, maximizes the smoothed pseudo-likelihood
+Here we consider a non-parametric kernel-based approach where the selection propensity at time~$t$, denoted $\hat \theta_t$, maximizes the smoothed pseudo-likelihood
 $$
@@
 
+Suppose there is a sequence of finite populations of size $N_{\nu}$ indexed by $\nu=1,2,\ldots$.
+Each finite population has an associated non-probability at time $t$ and probability samples of size~$n$ at each time~$t=1,\ldots,T$.
 Assuming the selection propensity model is correctly specified, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:ipw} shows that the IPW estimator is  consistent and calculates the estimator's variance.
@@
 Here, we consider covariate information that takes the form of a stratification variable with $K$ strata.
-Let $\nu_{k,t}$ to denote the number of new infections on day $t$ in the $k$th strata.
+Let $I_{k,t}$ to denote the number of new infections on day $t$ in the $k$th strata.
 As there are a total of $N \cdot i_t := I_t$ new infections on day $t$, we assume the joint distribution of strata-specific new infection totals takes a multinomial distribution
 $$
-( \nu_{1,t}, \ldots, \nu_{K,t} ) \sim \text{Multinomial} \left( I_t, (p_{1,t}, \ldots, p_{K, t}) \right).
+( I_{1,t}, \ldots, I_{K,t} ) \sim \text{Multinomial} \left( I_t, (p_{1,t}, \ldots, p_{K, t}) \right).
 $$
@@
 $$
-D_{k,r} \mid p, \theta, \nu \sim \text{Poisson} \left( \sum_{t=1}^r p_{k} \nu_{k,t} \theta_{(r-t)} \right)
-$$
-where $p_{k}$ is the infection fatality rate for the $k$th strata.
+D_{k,r} \mid p, \theta, \nu \sim \text{Poisson} \left( \sum_{t=1}^r p_{k} I_{k,t} \theta_{(r-t)} \right)
+$$
+where $p_{k}$ is the infection fatality rate for the $k$th strata and $\theta_{j}$ is a discrete-time distribution for time from infection to death.
 This extends prior analysis of death data~\cite{Johndrow2020} by allowing the infection fatality rate to depend on stratification variable.
 Specifically, ``the estimated IFR is close to zero for children and younger adults but rises exponentially with age, reaching 0.4\% at age 55, 1.4\% at age 65, 4.6\% at age 75, and 15\% at age 85''~\cite{Levin2020}.
+For conciseness, parameter choices and additional details on inference are discussed in Appendix~\ref{app:modelbased}.
 
@@
 %
-
-% For conciseness, parameter choices and inference are discussed in depth in Appendix~\ref{app:modelbased}.
 
@@
 
-Unfortunately, the first term cannot be computed as covariate information is not collected on the entire population.
+Unfortunately, the first term of the doubly robust estimator cannot be computed as covariate information is not collected on the entire population.
 Here, we again use the probability samples and consider the doubly-robust estimator
@@
 where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
-Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$ and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ where $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$.
+Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$ and $\text{var} (\hat y_{n,t}) = V_{t}^{(DR)} + o (n^{-1})$ where $V_t^{(DR)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$.
 \end{lemma}
@@
 
-Here we consider estimation of the active infection rate in Indiana using unweighted, IPW, model-based, and doubly robust estimates. To do so, we consider five data sources:
+We next consider estimation of the active infection rate in Indiana using unweighted, IPW, model-based, and doubly robust estimates. To start, we recap the data sources used in estimation:
 \begin{itemize}[leftmargin=*]
-\item {\bf Testing data}:  Each day, the number of tests performed and number of positive tests are reported, with counts broken out jointly by age, gender, and racial demographic information.
+\item {\bf Testing data}:  The number of daily tests performed and number of daily positive tests are reported.
+These counts are broken out jointly by age, gender, and racial demographic information.
 Figure~\ref{fig:in-cases} and~\ref{fig:in-tests} plots daily COVID-19 positive cases and tests by age strata.
@@
 
-\item {\bf Facebook symptom survey}: We extract daily symptom surveys completed by individuals on Facebook who identify as living in Indiana.
+\item {\bf Facebook symptom survey}: From the daily symptom surveys completed on Facebook (see Section~\ref{subsection:fbsymptom} for details), we extract survey responses for individuals who identify as living in Indiana.
 We collect age, sex, and demographic information as well as COVID-19 related symptoms, which Facebook defines as having (1) a fever and cough, (2) shortness of breath, \emph{or} (3) difficulty breathing.
 To assess potential bias, we compare the random sample from Indiana to the Facebook symptom survey data collected from April 20th to April 25th.
-Table~\ref{tab:comparison} demonstrates minimal bias in the Facebook symptom survey.
+Table~\ref{tab:comparison} demonstrates minimal bias in the Facebook symptom survey with respect to symptom distributions.
 
@@
 \end{tabular}
-\caption{Comparison of Facebook symptom survey and random sample demonstrates minimal bias.}
+\caption{Comparison of Facebook survey and random sample}
 \label{tab:comparison}
@@
 \end{tabular}
-\caption{Estimated point prevalence of active infection with SARS-CoV-2 by demographics in Indiana.}
+\caption{Estimated point prevalence of active infection with SARS-CoV-2 by demographics in Indiana~\cite{Yiannoutsos2021}.}
 \label{tab:indiana}
@@
 Using the constructed weights and the adjusted strata-level prevalence estimates, the IPW estimate under no measurement-error is $7.7$, a decrease of four percentage points from the observed prevalence of $11.7$.
-We also make use of Facebook's symptom survey over that window of time to construct a second IPW estimate conditoinal on demographics and fever but ignoring the household and prior testing information, which under no measurement-error is $6.5$\%.
+We also make use of Facebook's symptom survey over that window of time to construct a second IPW estimate conditional on demographics and fever but ignoring the household and prior testing information, which under no measurement-error is $6.5$\%.
 
@@
 Therefore, the IPW estimates accounting for measurement-error are $5.6\%$ and $4.2\%$ respectively.
-To account for uncertainty in
 
@@
 Indiana's population as of 2019 was $6.732$ million.
-A total of $898$ were administered between April 25th to 29th.
+A total of $898$ tests were administered in Indiana between April 25th to 29th.
 Subtracting off the $95,879$ tests that had already performed yields a sampling fraction of $f = 2.96 \times 10^{-3}$.
-Next we consider estimation of data quality $E_{\I} \left[ \rho_{\tilde I(X),Y} \right]$.
+Here we consider estimation of data quality $E_{\I} \left[ \rho_{\tilde I(X),Y} \right]$.
 In \cite{Meng2018}, estimation relied on observing the true outcome (i.e., election vote totals), which is not possible in the current crisis.
@@
 The unweighted estimate for the true prevalence of active COVID-19 disease between April 25th and April 29th is 11.7\%.
-Assuming a false negative rate of $3$\% and false positive rate of $3$\%, the unweighted estimate is $10.4$\%, leading to an error of $8.5$\%.
+Assuming a false negative rate of $30$\% and false positive rate of $2.4$\%, the unweighted estimate is $10.4$\%, leading to an error of $8.5$\%.
 Using~\eqref{eq:statdecomp}, we construct an estimate of the relative sampling rate as follows
@@
 %{\bf COORDINATION ON DATA SOURCES}
+
+% Equation~\eqref{eq:auxinfoprob} presupposes covariate information is collected for all individuals in the nonprobability sample; however, this is often not feasible for state and local governments where rapid sample collection is prioritized over additional collection of covariate information.
+We can extend~\eqref{eq:auxinfoprob} by randomly sampling a subset of the nonprobability sample on which to collect the covariate information, yielding the pseudo-likelihood
+% \begin{equation}
+% \label{eq:auxinfoprob2}
+% \sum_{j=1}^N I_j I_j^{(2)} W_j^{(2)} \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
+% \end{equation}
+% Here, $I_j^{(2)}$ is an indicator that the individual is in the probability subsample and $W_j^{(2)}$ is the probability of inclusion.
+Estimation proceeds by maximizing the pseudo-likelihood which now depends on random sampling in both the first and second term.
 

### 2021-07-13 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 Specimen collection by swab is known to impact false negative/positive rates regardless of test timing.
-Systematic reviews suggest that 70\% sensitivity and 97.6\% specificity are reasonable estimates for current RT-PCR tests~\cite{Woloshin2020,Cohen2020}.
+Systematic reviews suggest that 87\% sensitivity and 97.6\% specificity are reasonable estimates for current RT-PCR tests~\cite{Arevalo2020, Woloshin2020,Cohen2020}.
 
@@
 \includegraphics[width = 0.4\textwidth]{../figs/mem_heatmap_article.png}
-\caption{Imperfect testing adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.024$ and $FN=0.30$. Color scaled so blue = $-10$, white = $0$, and red = $10$.}
+\caption{Imperfect testing adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.024$ and $FN=0.13$. Color scaled so blue = $-10$, white = $0$, and red = $10$.}
 %\textcolor{red}{NEED TO DO FOR FP = 0.05 and FN = 0.3}}
@@
 
-The Indiana study did not report sensitivity and specificity; therefore, we take the suggested measures from~\cite{Woloshin2020} which report 70\% sensitivity is a reasonable estimates and~\cite{Cohen2020} which report 97.6\% specificity.
-This corresponds to a false negative rate of $30\%$ and false positive rate of $2.4\%$.
-Therefore, the IPW estimates accounting for measurement-error are $5.6\%$ and $4.2\%$ respectively.
+The Indiana study did not report sensitivity and specificity; therefore, we take the suggested measures from~\cite{Arevalo2020} which report 87\% sensitivity is a reasonable estimates and~\cite{Cohen2020} which report 97.6\% specificity.
+This corresponds to a false negative rate of $13$\% and false positive rate of $2.4\%$.
+Therefore, the IPW estimates accounting for measurement-error are $6.2\%$ and $6.1\%$ respectively.
 
@@
 The unweighted estimate for the true prevalence of active COVID-19 disease between April 25th and April 29th is 11.7\%.
-Assuming a false negative rate of $30$\% and false positive rate of $2.4$\%, the unweighted estimate is $10.4$\%, leading to an error of $8.5$\%.
+Assuming a false negative rate of $13$\% and false positive rate of $2.4$\%, the unweighted estimate is $11.0$\%, leading to an error of $9.2$\%.
 Using~\eqref{eq:statdecomp}, we construct an estimate of the relative sampling rate as follows
 \begin{equation*}
-\rho D_M = \sqrt{\frac{f}{1-f}} \frac{\text{0.085}}{\sigma_Y} = 3.84 \times 10^{-2} \Rightarrow \Delta = 1.39 \times 10^{-2} \Rightarrow M = 6.1.
+\rho D_M = \sqrt{\frac{f}{1-f}} \frac{\text{0.092}}{\sigma_Y} = 3.75 \times 10^{-2} \Rightarrow \Delta = 1.39 \times 10^{-2} \Rightarrow M = 6.1.
 \end{equation*}
 Using inverse-probability weights, the IPW estimates for the true prevalence of active COVID-19 disease between April 25th and April 29th are $7.7$\% and $6.5\%$ respectively.
-Under the same sensitivity and specificity, the weighted estimates adjusted for measurement-error are $5.6$\% and $4.2$\%, leading to errors $3.8$\% and $2.4$\% respectively.
+Under the same sensitivity and specificity, the weighted estimates adjusted for measurement-error are $6.2$\% and $4.9$\%, leading to errors $4.4$\% and $3.1$\% respectively.
 Using~\eqref{eq:statdecomp2}, we construct an estimate of the relative sampling rate as follows
 \begin{equation*}
-\tilde \rho D_M = \sqrt{\frac{f}{1-f+CV_W^2}} \frac{\text{0.038}}{\sigma_Y} = 6.93 \times 10^{-3} \Rightarrow \tilde \Delta = 8.24 \times 10^{-3}  \Rightarrow \tilde M = 3.9,
+\tilde \rho D_M = \sqrt{\frac{f}{1-f+CV_W^2}} \frac{\text{0.044}}{\sigma_Y} = 8.14 \times 10^{-3} \Rightarrow \tilde \Delta = 8.22 \times 10^{-3}  \Rightarrow \tilde M = 3.9,
 \end{equation*}
@@
 A sensitivity analysis is performed by considering the range of false negative and false positive rates.
-Here, sensitivity is assumed to fall between $81$\% and $91$\% and specificity between $95$\% and $98.8$\%~\cite{Katz2020}, which leads to a range for $M$ of $(5.60, 6.39)$ for the unweighted analysis, and $(3.63,4.07)$ and $(3.10, 3.45)$ for the weighted analyses respectively.
+Here, sensitivity is assumed to fall between $81$\% and $91$\% and specificity between $95$\% and $98.8$\%~\cite{Katz2020}, which leads to a range for $M$ of $(5.60, 6.44)$ for the unweighted analysis, and $(3.63,4.07)$ and $(3.10, 3.45)$ for the weighted analyses respectively.
 
@@
 
-Here we extend the IPW analysis to the time-varying setting.
+The IPW analysis is next extended to the time-varying setting.
 To do so, we make use of the Facebook symptom survey, hospitalization, and testing data.
@@
 Symptom status (e.g., fever, coughing, shortness of breath) and related covariate information (COVID-19 positive contact), however, are currently unavailable\footnote{Whether this is due to data privacy considerations or lack of data collection efforts is unclear.}.
-As COVID-19 symptom status is likely to alter the testing propensity and likelihood of active infection, Facebook symptom survey and hospitalization data are both used to estimate symptom status per strata.
-We do this in two ways which are each detailed below.
+As COVID-19 symptom status is likely to alter the testing propensity and likelihood of active infection, Facebook symptom survey and hospitalization data are used to estimate symptom status per strata in two ways which are each detailed below.
 
@@
 
-Here, we take the suggested measures from~\cite{Woloshin2020} which report 70\% sensitivity is a reasonable estimates and~\cite{Cohen2020} which report 97.6\% specificity.
-This corresponds to a false negative rate of $30\%$ and false positive rate of $2.4\%$.
+Here, we take the suggested measures from~\cite{Arevalo2020} which report 87\% sensitivity is a reasonable estimates and~\cite{Cohen2020} which report 97.6\% specificity.
+This corresponds to a false negative rate of $13\%$ and false positive rate of $2.4\%$.
 To allow for potential uncertainty, we assume two \emph{synthetic} simple random samples.

### 2021-07-15 write-up/covid-writeup.tex
IPW done except for Variance calculation

@@
 As COVID-19 symptom status is likely to alter the testing propensity and likelihood of active infection, Facebook symptom survey and hospitalization data are used to estimate symptom status per strata in two ways which are each detailed below.
+Here, two simple imputation methods are considered for illustrative purposes; see Section~\ref{section:discussion} for additional discussion.
 
@@
 The first estimates the probability of contact with a COVID-19 positive individual given demographic and COVID-19 test status (positive or negative COVID-19 test).
+Figure~\ref{fig:contactlik1} and~\ref{fig:contactlik2} in Appendix~\ref{app:in_add_details} presents the estimated likelihood of contact with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.
 The second estimates the probability of fever given demographic, COVID-19 test status, \emph{and} whether the individual has had contact with a COVID-19 positive individual.
+Figure~\ref{fig:symptomlik1} and~\ref{fig:symptomlik2} in Appendix~\ref{app:in_add_details} presents the estimated likelihood of reporting a fever with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.
+We see that likelihood of fever varies greatly based on whether the individual also reported contact with a COVID-19 positive individual.
 Using these two models, mean imputation is used to calculate the number of positive and negative tests given demographic strata, fever status, and COVID-19 contact status.
-With the resulting dataset, inverse probability weights are computed based.
-
-In the second model, for negative cases,  people fever
-
-For positive cases, we impute the number of positive cases
-
-Figure~\ref{fig:testinglik_mainpaper} presents the testing propensity for non-hispanic, white males across age strata and fever status using the second approach.
+With the resulting dataset, inverse probability weights are computed based (termed IPW1).
+
+In the second approach, we leverage hospitalization information.
+The first estimates the probability of fever given demographic and COVID-19 test status.
+When the COVID-19 test status is positive, the probability of fever depends on whether the individual was hospitalized.
+Figure~\ref{fig:symptomlik1_model2} and~\ref{fig:symptomlik2_model2} in Appendix~\ref{app:in_add_details} presents the estimated likelihood of reporting a fever with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.
+We see that hospitalization significantly increases the likelihood of fever as expected.
+We use these models and the hospitalization rates by demographic to perform mean impuation of the number of positive and negative tests given demographic strata, fever status, and COVID-19 contact status.
+With the resulting dataset, inverse probability weights are computed based (termed IPW2).
+
+To make the approach concrete, Figure~\ref{fig:testinglik_mainpaper} presents the testing propensity for non-hispanic, white males across age strata and fever status using the second approach.
 We note here that the ratio of testing propensity within strata across fever status is time-varying.
-The ratio starting nine times more likely in early April and
-
-Fever
+The ratio starting nine times more likely in early April and dropping to twice as likely by end of 2020.
 
@@
 
-%\emph{Need to show Hospitalization C+ versus Covid count to show that most of the tests are in that subset for early parts.}
-
 \subsection{Model-based estimation}
+
+Ages we use
+
+Stan.
+
+Figure~\ref{fig:deathsppc} performs a posterior predictive check on aggregate
 
@@
 
-\subsection{Doubly robust estimation}
-%
-%% \begin{figure}
-%% \centering
-%% \begin{subfigure}{.5\textwidth}
-%%   \centering
-%%   \includegraphics[width=.9\linewidth]{../methods/figs/NY_casecount.png}
-%%   \caption{NY new cases per day}
-%%   \label{fig:ny-covid-test}
-%% \end{subfigure}%
-%% \begin{subfigure}{.5\textwidth}
-%%   \centering
-%%   \includegraphics[width=.9\linewidth]{../methods/figs/NY_fracpos.png}
-%%   \caption{NY fraction of new tests positive per day}
-%%   \label{fig:ny-covid-frac}
-%% \end{subfigure}
-%% \caption{New York COVID-19 reported data}
-%% \label{fig:ny-covid}
-%% \end{figure}
-%
+\subsection{Time-varying prevalence estimates}
+
+Present IPW1, IPW2, Model-based, DR1, and DR2.
+
+We notice that IPW1/2 don't improve greatly but this may be due to the
+
+We notice model-based rates are much lower, but
 
@@
 \label{section:discussion}
-%
-%{\bf Our goal with surveillance is to identify areas of weakness; but selection bias is an issue!}
-%
-%There is nothing routine about COVID-19, including the corresponding statistical questions.
+
+{\bf Our goal with surveillance is to identify areas of weakness; but selection bias is an issue!}
+
+There is nothing routine about COVID-19, including the corresponding statistical questions.
 The goal of this paper was to point out questionable statistical routines.
@@
 We end with a brief discussion of important related topics.
-%
-%\subsection*{Additional biases and measurement-error}
-%
-%Length bias and measurement-error?
-%
-%\subsection*{Decision-making: data versus information}
-%
-%Having read through the above technical discussions, one may come to the incorrect conclusion that because we are concerned about implicit biases in observational data, we must wait for better data sources in order to act~\citep{Ioannidis2020}.
+
+\subsection*{Additional biases and measurement-error}
+
+Length bias and measurement-error?
+
+\subsection*{Decision-making: data versus information}
+
+Having read through the above technical discussions, one may come to the incorrect conclusion that because we are concerned about implicit biases in observational data, we must wait for better data sources in order to act~\citep{Ioannidis2020}.
 This conclusion confuses data and information.
@@
 A full discussion of these topics is beyond the scope of the current article.
-%
-%\subsection*{Data quantity versus quality}
-%
-%Governments implicitly argue that increased testing capacity will perform the above task.
+
+\subsection*{Data quantity versus quality}
+
+Governments implicitly argue that increased testing capacity will perform the above task.
 Without complete compliance, however, our understanding of future outbreaks may be plagued by self-selection bias, compounded by changing sensitivity and specificity rates.
@@
 This is not to say random testing should replace current testing practices; only that the random testing should supplement and will provide valuable additional information even with moderate sample sizes.
-%
-%\subsection*{High quality information on prevalence.}
-%
-%Governments are pushing for increased testing capacity and robust contact tracing.
+
+\subsection*{High quality information on prevalence.}
+
+Governments are pushing for increased testing capacity and robust contact tracing.
 Contact tracing is incredibly useful for identifying carriers early and preventing spread of the disease.
@@
 Knowing prevalence will help governments determine long-term community-level risk and allow for more targeted interventions -- shutting down only certain locations when necessary.
-%
-%\subsection*{Stratified sampling: improving precision in low-prevalence environments.}
-%Random sampling is a powerful tool.
+
+\subsection*{Stratified sampling: improving precision in low-prevalence environments.}
+Random sampling is a powerful tool.
 One difficulty with SRS is when prevalence is low.
@@
 One can, for example, stratify the population into risk categories (e.g., based on population density, occupation, age) and perform Neyman allocation~\citep{Cochran77}.
-%
-%\subsection*{Model-based solutions.}
-%
-%One can extend the SIR model to account for selection bias and measurement error.
+
+\subsection*{Model-based solutions.}
+
+One can extend the SIR model to account for selection bias and measurement error.
 Without strong assumptions on the selection mechanism, however, the estimates are often not identifiable.
@@
 Absent random sampling, the best route forward for all data analyses is careful associated sensitivity analyses and humility in data-driven conclusions.
-%
-%\subsection*{Random sampling for doubly robust estimation}
-%
-%Selection rates are likely time-varying; therefore, we need rapid survey samples to properly model the selection criteria for doubly robust inferential procedures.
-%
-%COLLECT COVARIATE INFORMATION (even aggregate) and provide it in accessible formats.
-%
-%- Easy access to history of testing requirements
-%- Smaller random samples over time (at least as testing and prevalence changes drastically) to assess selection changes
-%- Record symptom/asymptomatic status and COVID contact.  If not for all, then for at least some we need these covariates!!!!!
-%
-%\subsection*{Systematic data collection and reporting}
-%
-%I would say this is important but JHU is not enough; see above for benefits of random sampling to augment and improve decision making.
-%
-%\subsection*{Frequency and turnaround time}
-%
-%This supposes no selection bias.
-%
-%
-%
-%
-%What is your goal?
+
+\subsection*{Random sampling for doubly robust estimation}
+
+Selection rates are likely time-varying; therefore, we need rapid survey samples to properly model the selection criteria for doubly robust inferential procedures.
+
+COLLECT COVARIATE INFORMATION (even aggregate) and provide it in accessible formats.
+
+- Easy access to history of testing requirements
+- Smaller random samples over time (at least as testing and prevalence changes drastically) to assess selection changes
+- Record symptom/asymptomatic status and COVID contact.  If not for all, then for at least some we need these covariates!!!!!
+
+\subsection*{Systematic data collection and reporting}
+
+I would say this is important but JHU is not enough; see above for benefits of random sampling to augment and improve decision making.
+
+\subsection*{Frequency and turnaround time}
+
+This supposes no selection bias.
+
+What is your goal?
 "get people who are currently sick off the street." Then test frequently and often.
@@
 Random testing is one way to improve upone current practice whiel we wait for that
-%
-%{\bf COORDINATION ON DATA SOURCES}
-
-% Equation~\eqref{eq:auxinfoprob} presupposes covariate information is collected for all individuals in the nonprobability sample; however, this is often not feasible for state and local governments where rapid sample collection is prioritized over additional collection of covariate information.
+
+{\bf COORDINATION ON DATA SOURCES}
+
+Equation~\eqref{eq:auxinfoprob} presupposes covariate information is collected for all individuals in the nonprobability sample; however, this is often not feasible for state and local governments where rapid sample collection is prioritized over additional collection of covariate information.
 We can extend~\eqref{eq:auxinfoprob} by randomly sampling a subset of the nonprobability sample on which to collect the covariate information, yielding the pseudo-likelihood
-% \begin{equation}
-% \label{eq:auxinfoprob2}
-% \sum_{j=1}^N I_j I_j^{(2)} W_j^{(2)} \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
-% \end{equation}
-% Here, $I_j^{(2)}$ is an indicator that the individual is in the probability subsample and $W_j^{(2)}$ is the probability of inclusion.
+\begin{equation}
+\label{eq:auxinfoprob2}
+\sum_{j=1}^N I_j I_j^{(2)} W_j^{(2)} \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
+\end{equation}
+Here, $I_j^{(2)}$ is an indicator that the individual is in the probability subsample and $W_j^{(2)}$ is the probability of inclusion.
 Estimation proceeds by maximizing the pseudo-likelihood which now depends on random sampling in both the first and second term.
@@
 
-\section{Image combos (temporary)}
-
+\section{Indiana COVID-19 Analysis: Additional details}
+\label{app:in_add_details}
 \subsection{Model 1}
@@
  \includegraphics[width=.9\linewidth]{../figs/tvprop_symptom_fig1.png}
- \caption{Symptom Likelihood Given Negative Test}
+ \caption{Symptom Likelihood Given Negative Test \textcolor{red}{FIX!!}}
  \label{fig:symptomlik1}
@@
  \caption{Symptom Likelihood Given Negative Test}
- \label{fig:symptomlik1}
+ \label{fig:symptomlik1_model2}
 \end{subfigure}%
@@
  \caption{Symptom Likelihood Given Positive Test}
- \label{fig:symptomlik2}
+ \label{fig:symptomlik2_model2}
 \end{subfigure}

### 2021-07-15 methods/5_propensity_estimation_alt.R

@@ -92,7 +92,7 @@ irls <- function(current_week, current_year, fb_X, indiana_X,
 
 
 
-## Brining in complete FB and Indiana data
+## Bringing in complete FB and Indiana data
 
 indiana_data = readRDS("../data/weeklycoviddata_withsymptoms_alt.RDS")
 indiana_data$startdate = mdy(indiana_data$startdate)
@@ -137,4 +137,4 @@ names(results) = c("week", "year", "feverFALSE", "feverTRUE",
                    "genderF", "NotHoL", "raceAA", "raceOther", "raceWhite", 
                    "25to34", "35to44", "45to54", "55to64", "65to74", "75plus")
 
-saveRDS(results, "../data/smoothedpropensities_alt.RDS")
+saveRDS(results, "../data/smoothedpropensities_alt_071521.RDS")

### 2021-07-15 methods/6_invweight_estimation_alt.R

@@ -10,6 +10,7 @@ model = ~ -1+as.factor(symptoms) + as.factor(gender) + ethnicity + race + as.fac
 indiana_X = model.matrix(model, indiana_data)
 
 library(lubridate)
+indiana_data$startdate = ymd(indiana_data$startdate)
 indiana_data$week = week(indiana_data$startdate)
 indiana_data$year = year(indiana_data$startdate)
 
@@ -49,7 +50,5 @@ for(i in 1:length(weeks)) {
   results[i,4] = sum(current_counts*weights)/sum(current_tests*weights)
 }
 
-saveRDS(results, "../data/invweights_alt.RDS")
-
-results
+saveRDS(results, "../data/invweights_alt_071521.RDS")
 

### 2021-07-15 preprocessing/5_in_hops_cleaning.R

@@ -36,7 +36,7 @@ race_hosp_frac = dashboard_race/sum(dashboard_race)
 
 
 ## No hispanic information so use the census data here
-## Census: 95.5% non-hispanic 4.5% hispanic
+## Census: 92.7% non-hispanic 7.3% hispanic
 ethnicity_levels = levels(indiana_test_data$ethnicity)
 ethnicity_census = c(0.073, 0.927)
 

### 2021-07-20 visualizations/model_based.R
Figs updated

@@ -1,0 +1,123 @@
+## Libraries and Colors
+library(wesanderson)
+library(tidyverse)
+library(tidybayes)
+library(gridExtra)
+library(ggplot2)
+library(rstan)
+library(lubridate)
+
+c_mid <- c("#fc9272")
+c_dark <- c("#de2d26")
+c_posterior = "orange"
+c_prior = "aquamarine2"
+
+## Death model
+temp = read.csv("../data/covid_indiana_age.csv")
+
+temp$date = ymd(temp$DATE)
+
+aggregate(temp$COVID_COUNT ~ date, data = temp, sum)
+
+df_coviddeath <- readRDS("../data/dailycoviddata.RDS")
+df_coviddeath <- aggregate(covid_deaths ~ startdate, data = df_coviddeath, FUN = sum)
+df_coviddeath$date = ymd(df_coviddeath$startdate)
+df_coviddeath$death_dt = df_coviddeath$covid_deaths
+
+# Swiss population
+N <- 6.732E6;
+
+# Deaths
+deaths = df_coviddeath$death_dt[-length(df_coviddeath$death_dt)]
+dates = df_coviddeath$date[-length(df_coviddeath$death_dt)]
+
+# times
+n_days <- length(deaths)
+
+fit_forcing = readRDS("../data/fit_forcing_threejumps.RDS")
+max_death_day <- 41 # Hard coded
+
+check_hmc_diagnostics(fit_forcing)
+
+pairs(fit_forcing, pars = c("beta", "eta", "eta_two", "eta_three"))
+
+pairs(fit_forcing, pars = c("nu", "nu_two", "nu_three"))
+
+smr_pred <- cbind(as.data.frame(summary(fit_forcing, pars = "pred_cases", probs = c(0.025, 0.05, 0.1, 0.5, 0.9, 0.95, 0.975))$summary[(max_death_day+1):(max_death_day+length(deaths)-1),]),
+                  t=1:(n_days-1))
+colnames(smr_pred) <- make.names(colnames(smr_pred)) # to remove % in the col names
+
+smr_pred$date = date(dates[1:(length(dates)-1)])
+
+ggplot(smr_pred, mapping = aes(x = date)) +
+  #geom_ribbon(aes(ymin = X2.5., ymax = X97.5.), fill = c_dark, ) +
+  geom_ribbon(aes(ymin = X5., ymax = X95.), fill = c_posterior, alpha=0.35) +
[…]

### 2021-07-20 visualizations/tv_prevalence.R

@@ -58,19 +58,31 @@ deaths = df_coviddeath$death_dt[-length(df_coviddeath$death_dt)]
 n_days <- length(deaths)
 max_death_day <- 41 # Hard coded
 
-fit_forcing = readRDS("../data/fit_forcing_threejumps.RDS")
+fit_forcing = readRDS("../data/fit_forcing_byage.RDS")
 
 smr_pred <- cbind(as.data.frame(summary(fit_forcing, pars = "pred_cases", probs = c(0.025, 0.05, 0.1, 0.5, 0.9, 0.95, 0.975))$summary[(max_death_day+1):(max_death_day+length(deaths)-1),]),
                   t=1:(n_days-1))
 colnames(smr_pred) <- make.names(colnames(smr_pred)) # to remove % in the col names
 
 smr_pred$current_date = df_coviddeath$date[-c(469:470)]
-smr_pred$air = smr_pred$mean/100000 ## TEMP
+smr_pred$air = smr_pred$mean/N ## TEMP
 
 prevalence_temp = data.frame(date = smr_pred$current_date, Method = rep("Model-based", nrow(smr_pred)), 
                              estimate = smr_pred$air)
 
 prevalence_temp = prevalence_temp[prevalence_temp$date <= max(prevalence_long$date),]
+
+prevalence_temp$week = week(prevalence_temp$date)
+prevalence_temp$year = year(prevalence_temp$date)
+
+prevalence_temp = aggregate(estimate ~ week + year, data = prevalence_temp, FUN = sum)
+
+prevalence_temp$date = MMWRweek::MMWRweek2Date(MMWRyear = prevalence_temp$year,
+                                            MMWRweek = prevalence_temp$week,
+                                            MMWRday = 1)
+
+prevalence_temp = data.frame(date = prevalence_temp$date, Method = rep("Model-based", nrow(prevalence_temp)), 
+                             estimate = prevalence_temp$estimate)
 
 prevalence_long = rbind(prevalence_long, prevalence_temp)
 

### 2021-07-20 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 \subsection{Model-based estimation}
-
-Ages we use
-
-Stan.
+\label{section:modelbased}
+
+Here we consider the multinomial SEIR model using observed death data as presented in Figure~\ref{fig:in-deaths}.
+The model, presented in Section~\ref{section:modelbased}, requires infection fatality rates to be specified per-strata.
+In this paper, similar to~\cite{Johndrow2020}, we specify fixed per-strata infection fatality rates.
+Based on published age-specific IFRs~\cite{Levin2020}, IFR by age seems to closely follow a log-linear relationship.
+Combining across published age-specific IFRs, Table~\ref{tab:ifrperage} presents estimates of the average IFR within each strata using the age-distribution.
+\begin{table}[!th]
+\begin{tabular}{c | c c c c c c c c}
+Age Strata & 0-19 & 20-29 & 30-39 & 40-49 & 50-59 & 60-69 & 70-79 & 80+ \\ \hline
+IFR & 0.003\% & 0.014\% & 0.046\% & 0.144\% & 0.456\% & 1.443\% & 4.567\% & 15.468\%
+\end{tabular}
+\caption{IFR by age-strata}
+\label{tab:ifrperage}
+\end{table}
+
+The discrete-time distribution~$\{ \theta_{j} \}$ is a discretized version of a truncated normal distribution with mean $25$, standard deviation~$5$, minimum value $0$, and maximum value $44$.
+This was chosen to closely mimic the distribution from~\cite{Johndrow2020}.
 

### 2021-07-29 write-up/covid-writeup.tex
Updated DR model and latex

@@
 The World Health Organization has declared the coronavirus disease 2019 (COVID-19) a public health emergency.
-\textcolor{red}{As of June 15th, 2021, over 176 million cases have been confirmed worldwide with 33.4 million cases and over 590 thousand confirmed deaths across the United States.} This pandemic
+As of July 29th, 2021, over 196 million cases have been confirmed worldwide with 34.8 million cases and over 612 thousand confirmed deaths across the United States.
+This pandemic
 has become the focal point of everyday life; yet the data landscape for understanding COVID-19 remains limited.
@@
 \subsection{Inverse-probability weighting approach}
+\label{section:ipwapproach}
 
@@
 
-\subsection{Model-based estimation}
+\subsection{Application of SEIR model}
 \label{section:modelbased}
@@
 \end{table}
-
 The discrete-time distribution~$\{ \theta_{j} \}$ is a discretized version of a truncated normal distribution with mean $25$, standard deviation~$5$, minimum value $0$, and maximum value $44$.
 This was chosen to closely mimic the distribution from~\cite{Johndrow2020}.
-
-Figure~\ref{fig:deathsppc} performs a posterior predictive check on aggregate
+See~\ref{app:modelbased} for additional details on parametrization and choice of priors.
+To check how well our model fits the observed data, a posterior predictive check on aggregate death counts was performed.
+Figure~\ref{fig:deathsppc} shows the results which suggests chosen SEIR model fits the observed data well.
+The model produces age-specific case counts which can then be used for model-based estimation of the active infection rate as well as in constructing the doubly robust estimator.
 
@@
 \includegraphics[width=.6\linewidth]{../figs/deaths_ppc.png}
-\caption{Posterior predictive check on aggregate daily death counts}
+\caption{Posterior predictive check on aggregate daily death counts.}
 \label{fig:deathsppc}
@@
 
-Present IPW1, IPW2, Model-based, DR1, and DR2.
-
-We notice that IPW1/2 don't improve greatly but this may be due to the
-
-We notice model-based rates are much lower, but
+Here, we construct active infection rate estimates using the unweighted, inverse-probability weighted, model-based, and doubly robust estimators.
+Figure~\ref{fig:air} presents the results.
+The inverse-probability weighting methods do not significantly reduce bias compared to the unweighted estimates.
+We conjecture this is due to the limited availability of symptom and other important covariates that may impact the likelihood an individual will get  tested.
+Our imputation strategies use available data to try and alleviate this issue; however, this suggests better data collection in the future may strongly improve the IPW estimates as seen in Section~\ref{section:ipwapproach} when analyzing the statewide sample from April 25th--29th.
+
+While the model-based estimates appear more reasonable, note that the estimate on the last week of April is $0.1$\% which is a severe underestimate.
+Caution is therefore warranted when interpreting these estimates.
+We know that the age-specific infection fatality rate is likely time-varying and may vary by other factors beyond age.
+Therefore, improvements to these model-based predictions could be made by improving the IFR estimates; however, there is currently minimal publicly available data to do so.
+Moreover, the (thankfully) low number of deaths per strata make uncertainty in these estimates quite high.
+The 95\% credible range for active infections on April 25th, for example, is $[XX,XX]$ with a posterior mean of $XX$.
+Finally, the doubly robust estimates appear similar to the IPW estimate.
 
@@
 \caption{Active Infection Rate Estimates}
-\label{fig:testinglik_mainpaper}
+\label{fig:air}
 \end{figure}
@@
 \label{section:discussion}
-
-{\bf Our goal with surveillance is to identify areas of weakness; but selection bias is an issue!}
 
@@
 Precision in reported case-count data gives the illusion of information when what what is needed is quantification of uncertainty.
-Selection bias leads data analysts to feel certain about incorrect conclusions.
+This paper demonstrates how selection bias leads data analysts to feel certain about incorrect conclusions.
+As case-count data is routinely used for public health policy making, we presented an inverse-probability weighting method and a doubly-robust estimation method that leverage auxiliary information collected through random samples to overcome these issues.
 We end with a brief discussion of important related topics.
-
-\subsection*{Additional biases and measurement-error}
-
-Length bias and measurement-error?
 
@@
 Contact tracing is incredibly useful for identifying carriers early and preventing spread of the disease.
-We would argue that, after the current wave, understanding prevalence is also key.
+We would argue that, after the current wave, understanding active infections is also key.
 We know that the probability of an outbreak is a function of current prevalence and network connectivity.
@@
 
-\subsection*{Stratified sampling: improving precision in low-prevalence environments.}
-Random sampling is a powerful tool.
+% \subsection*{Stratified sampling: improving precision in low-prevalence environments.}
+% Random sampling is a powerful tool.
 One difficulty with SRS is when prevalence is low.
@@
 
-\subsection*{Random sampling for doubly robust estimation}
+\subsection*{Real-world implementation of the proposed method}
 
@@
 Estimation proceeds by maximizing the pseudo-likelihood which now depends on random sampling in both the first and second term.
+
+\subsection*{Additional biases and measurement-error}
+
+Other biases exist and are incredible important when trying to estimate quantities such as the incubation period and mean serial interval.
+
+These estimates could be severely biased due to selection bias and measurement error, but also suffer from
+
+Nevertheless,
+Length bias and measurement-error?
 
@@
 
+\section{Model-based: Additional details}
+\label{app:modelbased}
+
 \end{document}

### 2021-07-29 methods/10_dr_estimation.R

@@ -1,0 +1,136 @@
+## Libraries
+library(tidyverse)
+library(tidybayes)
+library(gridExtra)
+library(ggplot2)
+library(rstan)
+library(lubridate)
+library(MMWRweek)
+
+## INDIANA AND PROPENSITIES
+indiana_data = readRDS("../data/weeklycoviddata_withsymptoms_alt.RDS")
+indiana_data$startdate = mdy(indiana_data$startdate)
+propensities = readRDS("../data/smoothedpropensities_alt.RDS")
+
+### MODEL BASED ESTIMATES
+aggregate_air = readRDS("../data/aggregate_air.RDS")
+strata_air = readRDS("../data/modelbased_air_bystrata.RDS")
+air_flags = readRDS("../data/modelbased_flags.RDS")
+
+### FB Data
+fb_data = readRDS("../data/fb_weeklycomplete.RDS")
+
+##  Death DATA
+df_coviddeath <- readRDS("../data/dailycoviddata.RDS")
+df_coviddeath$week = week(df_coviddeath$startdate)
+df_coviddeath$year = year(df_coviddeath$startdate)
+df_coviddeath <- aggregate(covid_deaths ~ age + gender + ethnicity + race + week + year, data = df_coviddeath, FUN = sum)
+
+propensities$date = MMWRweek::MMWRweek2Date(MMWRyear = propensities$year,
+                                            MMWRweek = propensities$week,
+                                            MMWRday = 1)
+## First term can be precomputed given design matrix
+model = ~ -1+as.factor(symptoms) + as.factor(gender) + ethnicity + race + as.factor(age)
+indiana_X = model.matrix(model, indiana_data)
+indiana_data$week = week(indiana_data$startdate)
+indiana_data$year = year(indiana_data$startdate)
+indiana_data$age[indiana_data$age > 1] = (indiana_data$age[indiana_data$age > 1] - 1)
+
+invweight = complete_air = vector(length = nrow(indiana_data))
+for(i in 1:nrow(indiana_data)) {
+  modelmatrix_row = indiana_X[i,]
+  week_row = as.numeric(week(indiana_data$startdate[i]))
+  year_row = as.numeric(year(indiana_data$startdate[i]))
+  if(week_row > 5 & year_row == 2021) {
+    invweight[i] = NA
+  } else {
+    isflagged = subset(air_flags, race == indiana_data$race[i] &
+                         ethnicity == indiana_data$ethnicity[i] &
[…]

### 2021-07-29 methods/8_death_modeling.R

@@ -8,12 +8,10 @@ library(rstan)
 library(lubridate)
 
 df_coviddeath <- readRDS("../data/dailycoviddata.RDS")
+levels(df_coviddeath$age)[1:3] = "0-39" # Collapsing due to limited data
 df_coviddeath_age <- aggregate(covid_deaths ~ startdate + age, data = df_coviddeath, FUN = sum)
-
 df_coviddeath_age$date = ymd(df_coviddeath_age$startdate)
 df_coviddeath_age$death_dt = df_coviddeath_age$covid_deaths
-
-# df_swiss = df_swiss[df_swiss$date < "2020-07-15",] # single bump to start
 
 ## Death Distribution
 death_distribution = pnorm(seq(0.5,40.5, by =1), mean = 25, sd = 5)-pnorm(c(0,seq(0.5,39.5, by =1)), mean = 25, sd = 5)
@@ -50,7 +48,7 @@ tswitch_three <- subset(df_coviddeath_age, age == "80+") %>% filter(date < date_
 
 ## PLOTTING DEATH DATA BY AGE CATEGORY
 df_coviddeath_age_plot = df_coviddeath_age
-levels(df_coviddeath_age_plot$age) = c(1,1,1,2,2,2,3,3,3)
+levels(df_coviddeath_age_plot$age) = c(1,2,2,2,3,3)
 levels(df_coviddeath_age_plot$age) = c("0-39", "40-69", "70+")
 
 df_coviddeath_age_plot %>% 
@@ -64,8 +62,8 @@ df_coviddeath_age_plot %>%
   scale_fill_manual(values=wes_palette(n = 3, "IsleofDogs1"))
 
 
-death_rates = c(2.975607e-05, 1.440472e-04, 4.557553e-04, 1.441978e-03, 
-                4.562317e-03, 1.443485e-02, 4.567086e-02, 1.546816e-01)
+death_rates = c(0.01631191, 0.14419778, 0.45623168,
+                1.44348510, 4.56708582, 15.46815604)/100
 
 data_forcing <- list(n_days = n_days, t0 = t0, ts = t, N = N, deaths = deaths, 
                      tswitch = tswitch+max_death_day, tswitch_two = tswitch_two + max_death_day,
@@ -83,5 +81,4 @@ fit_forcing <- sampling(model_forcing,
                         seed=2,
                         chains = 1)
 
[…]

### 2021-07-29 preprocessing/6_in_death_cleaning.R

@@ -1,10 +1,10 @@
 ### Input: 
 ### Daily total tests, positive tests, deaths
-### Aggregate by age, gender, ethnicitiy, and race separately
+### Aggregate by age, gender, ethnicity, and race separately
 
 ### Output: 
 ### Weekly total tests, positive tests, deaths
-### Aggregate by age, gender, ethnicitiy, and race separately
+### Aggregate by age, gender, ethnicity, and race separately
 
 age_data = read.csv("../data/covid_indiana_age.csv")
 gender_data = read.csv("../data/covid_indiana_gender.csv")

### 2021-07-29 preprocessing/7_ifr.R

@@ -11,8 +11,8 @@ fit <- lm(log(ifr) ~ age)
 
 
 ## Age Ranges
-lower_age = c(0,20,30,40,50,60,70, 80)
-upper_age = c(19,29,39,49,59,69,79,90)
+lower_age = c(0,40,50,60,70, 80)
+upper_age = c(39,49,59,69,79,90)
 mean_ifr = vector(length = length(lower_age))
 
 for (i in 1:length(lower_age)) {

### 2021-07-29 preprocessing/8_modelbased_preproc.R

@@ -1,0 +1,267 @@
+## Libraries
+library(tidyverse)
+library(tidybayes)
+library(gridExtra)
+library(ggplot2)
+library(rstan)
+library(lubridate)
+
+## FUNCTIONS
+distance_function <-  function(current_week, current_year) {
+  fb_dist <- function(row) {
+    rowyear = as.numeric(row[3])
+    rowweek = as.numeric(row[2])
+    if (rowyear == current_year){
+      dis = abs(rowweek - current_week)
+    } else {
+      if (rowyear == 2020) {
+        dis = (53 - rowweek) + current_week
+      } else {
+        dis = (rowweek) + (53-current_week)
+      }
+    }
+    return(dis)
+  }
+  return(fb_dist)
+}
+
+kernelweight <- function(distance) {
+  kw2 <- function(row) {
+    dis = distance(row)
+    h = sqrt(-1/(2*log(0.98))) # Set a deterministic bandwidth for now
+    kernel = exp(-dis^2/(2*h^2))
+    return(kernel)
+  }
+  return(kw2)
+}
+
+## Model-based estimation
+## How do we match ages?
+## FB <--> INDIANA
+## 18-24 (1) <--> 0-19 and 20-29.
+## 25--34 (2) <--> 30--39.
+## 35--44 (3) <--> 40--49
+## 45-54 years (4) <--> 50--59
+## 55-64 years (5) <--> 60--69
+## 65-74 years (6) <--> 70--79
+## 75 years or older (7) <--> 80+
+
+## FIRST THREE INDIANA WERE COLLAPSE DUE TO LIMITED
+## DEATH DATA ON INDIVIDUALS <40 YEARS OF AGE.
+
+df_coviddeath <- readRDS("../data/dailycoviddata.RDS")
+df_coviddeath$date = ymd(df_coviddeath$startdate)
+df_coviddeath_agg <- aggregate(covid_deaths ~ date, data = df_coviddeath, FUN = sum)
+dates = df_coviddeath_agg$date[-length(df_coviddeath_agg$date)]
+
+fit_forcing = readRDS("../data/fit_forcing_byage_072621.RDS")
+summary_fit = summary(fit_forcing, pars = "pred_cases_per_agegroup", probs = c(0.025, 0.05, 0.1, 0.5, 0.9, 0.95, 0.975))$summary
+n_days = nrow(summary_fit)/6
[…]

### 2021-07-29 visualizations/model_based.R

@@ -18,7 +18,7 @@ df_coviddeath <- aggregate(covid_deaths ~ startdate, data = df_coviddeath, FUN =
 df_coviddeath$date = ymd(df_coviddeath$startdate)
 df_coviddeath$death_dt = df_coviddeath$covid_deaths
 
-# Swiss population
+# Indiana population
 N <- 6.732E6;
 
 # Deaths
@@ -28,11 +28,11 @@ dates = df_coviddeath$date[-length(df_coviddeath$death_dt)]
 # times
 n_days <- length(deaths)
 
-fit_forcing = readRDS("../data/fit_forcing_byage.RDS")
+fit_forcing = readRDS("../data/fit_forcing_byage_072621.RDS")
 
 test = summary(fit_forcing, pars = "pred_deaths", probs = c(0.025, 0.05, 0.1, 0.5, 0.9, 0.95, 0.975))$summary
-n_days = nrow(test)/8
-timepoints = rep(1:n_days, 8)
+n_days = nrow(test)/6
+timepoints = rep(1:n_days, 6)
 
 test = data.frame(test)
 test$timepoints = timepoints

### 2021-07-29 visualizations/tv_prevalence.R

@@ -22,6 +22,8 @@ prevalence_alt$date = MMWRweek::MMWRweek2Date(MMWRyear = prevalence$year,
                                           MMWRweek = prevalence$week,
                                           MMWRday = 1)
 
+dates = prevalence_alt$date
+
 prevalence_long = rbind(prevalence, prevalence, prevalence_alt)
 
 prevalence_long$Method = c(rep("Unweighted", nrow(prevalence)),
@@ -36,55 +38,18 @@ prevalence_long$estimate = prevalence_long$prev * (prevalence_long$Method == "Un
 prevalence_long = data.frame(date = prevalence_long$date, Method = prevalence_long$Method, estimate = prevalence_long$estimate)
 
 
-## Death model
-temp = read.csv("../data/covid_indiana_age.csv")
-
-temp$date = ymd(temp$DATE)
-
-aggregate(temp$COVID_COUNT ~ date, data = temp, sum)
-
-df_coviddeath <- readRDS("../data/dailycoviddata.RDS")
-df_coviddeath <- aggregate(covid_deaths ~ startdate, data = df_coviddeath, FUN = sum)
-df_coviddeath$date = ymd(df_coviddeath$startdate)
-df_coviddeath$death_dt = df_coviddeath$covid_deaths
-
-# Swiss population
-N <- 6.732E6;
-
-# Deaths
-deaths = df_coviddeath$death_dt[-length(df_coviddeath$death_dt)]
-
-# times
-n_days <- length(deaths)
-max_death_day <- 41 # Hard coded
-
-fit_forcing = readRDS("../data/fit_forcing_byage.RDS")
-
-smr_pred <- cbind(as.data.frame(summary(fit_forcing, pars = "pred_cases", probs = c(0.025, 0.05, 0.1, 0.5, 0.9, 0.95, 0.975))$summary[(max_death_day+1):(max_death_day+length(deaths)-1),]),
-                  t=1:(n_days-1))
-colnames(smr_pred) <- make.names(colnames(smr_pred)) # to remove % in the col names
-
-smr_pred$current_date = df_coviddeath$date[-c(469:470)]
-smr_pred$air = smr_pred$mean/N ## TEMP
-
-prevalence_temp = data.frame(date = smr_pred$current_date, Method = rep("Model-based", nrow(smr_pred)), 
-                             estimate = smr_pred$air)
-
-prevalence_temp = prevalence_temp[prevalence_temp$date <= max(prevalence_long$date),]
-
[…]

### 2021-07-29 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 
-\subsection*{Decision-making: data versus information}
-
-Having read through the above technical discussions, one may come to the incorrect conclusion that because we are concerned about implicit biases in observational data, we must wait for better data sources in order to act~\citep{Ioannidis2020}.
-This conclusion confuses data and information.
-Lack of high quality data means we should be skeptical regarding conclusions drawn from these data sources alone.
-Being skeptical of observed data does not imply governments and communities should wait for high quality data to act; this is especially true in the current high risk scenario.
-Instead, we should aim to have high quality information.
-Yet in the current crisis we have just that.
-Indeed, epidemiologists and public health officials have a strong understanding of the basic facts of pandemics and disease spread.
-Social distancing, contact tracing and mass testing are all important.
-Individuals and communities with low quality data and only basic information will hopefully make decisions that respect inherent risks to their well being as well as the high degree of uncertainty in the observed data.
-Of course, there are trade-offs and long-term stay-at-home orders will have negative economic consequences.
-A full discussion of these topics is beyond the scope of the current article.
-
 \subsection*{Data quantity versus quality}
 
-Governments implicitly argue that increased testing capacity will perform the above task.
+Governments and policy makers often implicitly argue that increased testing capacity will alleviate selection bias.
 Without complete compliance, however, our understanding of future outbreaks may be plagued by self-selection bias, compounded by changing sensitivity and specificity rates.
 Random testing removes these effect modifiers, giving governments more information to fight the disease.
-This is not to say random testing should replace current testing practices; only that the random testing should supplement and will provide valuable additional information even with moderate sample sizes.
-
-\subsection*{High quality information on prevalence.}
-
-Governments are pushing for increased testing capacity and robust contact tracing.
+This paper emphasizes that data quanitity is secondary to data quality and methodological considerations to account for selection bias.
+
+% \subsection*{High quality information on prevalence.}
+
+% Governments are pushing for increased testing capacity and robust contact tracing.
 Contact tracing is incredibly useful for identifying carriers early and preventing spread of the disease.
@@
 
-\subsection*{Model-based solutions.}
-
-One can extend the SIR model to account for selection bias and measurement error.
+\subsection*{Purely model-based solutions.}
+
+A common argument is that the SIR model to account for selection bias and measurement error; therefore, there is no need for auxiliary random samples.
 Without strong assumptions on the selection mechanism, however, the estimates are often not identifiable.
 When an issue ``cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically''\citep{CoxHink74}.
-Absent random sampling, the best route forward for all data analyses is careful associated sensitivity analyses and humility in data-driven conclusions.
+Absent some type of random sampling, the best route forward for all data analyses is careful associated sensitivity analyses and humility in data-driven conclusions.
 

### 2021-07-29 write-up/covid-writeup.tex
Updated writeup- discussion almost done

@@
 A derived statistical error decomposition guides this discussion.
+
+\textcolor{red}{Discuss the additional biases and Qing paper.}
+
+% Other biases exist and are incredible important when trying to estimate quantities such as the incubation period and mean serial interval.
+
+% These estimates could be severely biased due to selection bias and measurement error, but also suffer from
+
+% Nevertheless,
+% Length bias and measurement-error?
 
@@
 Precision in reported case-count data gives the illusion of information when what what is needed is quantification of uncertainty.
-This paper demonstrates how selection bias leads data analysts to feel certain about incorrect conclusions.
+Extensions of recent statistical error decompositions~\cite{Meng2018} demonstrate how selection bias leads data analysts to feel certain about incorrect conclusions.
 As case-count data is routinely used for public health policy making, we presented an inverse-probability weighting method and a doubly-robust estimation method that leverage auxiliary information collected through random samples to overcome these issues.
@@
 Governments and policy makers often implicitly argue that increased testing capacity will alleviate selection bias.
-Without complete compliance, however, our understanding of future outbreaks may be plagued by self-selection bias, compounded by changing sensitivity and specificity rates.
+Without complete compliance, however, our understanding of future outbreaks may be plagued by self-selection bias, compounded by changing sensitivity and specificity rates of RT-PCR testing.
 Random testing removes these effect modifiers, giving governments more information to fight the disease.
-This paper emphasizes that data quanitity is secondary to data quality and methodological considerations to account for selection bias.
-
-% \subsection*{High quality information on prevalence.}
-
-% Governments are pushing for increased testing capacity and robust contact tracing.
+This paper emphasizes that data quantity is secondary to both data quality and methodological considerations to account for selection bias.
+
+\subsection*{High quality information on prevalence.}
+
+Governments are pushing for increased testing capacity and robust contact tracing.
 Contact tracing is incredibly useful for identifying carriers early and preventing spread of the disease.
@@
 
-% \subsection*{Stratified sampling: improving precision in low-prevalence environments.}
-% Random sampling is a powerful tool.
-One difficulty with SRS is when prevalence is low.
-Consider the Santa Clara study~\citep{Bendavid2020} which estimated prevalence less than 2\% in the county.
-The key issue with low prevalence is that if the test's specificity is higher than the prevalence, then the observed data is consistent with zero recovered individuals.
-In these settings, SRS is likely to fail; however, there are potential solutions.
-One can, for example, stratify the population into risk categories (e.g., based on population density, occupation, age) and perform Neyman allocation~\citep{Cochran77}.
-
-\subsection*{Purely model-based solutions.}
-
-A common argument is that the SIR model to account for selection bias and measurement error; therefore, there is no need for auxiliary random samples.
+\subsection*{Model-based solutions.}
+
+A common argument is that the SEIR model can be extended to account for selection bias and measurement error directly; therefore, there is no need for auxiliary random sampling.
 Without strong assumptions on the selection mechanism, however, the estimates are often not identifiable.
@@
 
-Selection rates are likely time-varying; therefore, we need rapid survey samples to properly model the selection criteria for doubly robust inferential procedures.
-
-COLLECT COVARIATE INFORMATION (even aggregate) and provide it in accessible formats.
-
-- Easy access to history of testing requirements
-- Smaller random samples over time (at least as testing and prevalence changes drastically) to assess selection changes
-- Record symptom/asymptomatic status and COVID contact.  If not for all, then for at least some we need these covariates!!!!!
-
-\subsection*{Systematic data collection and reporting}
-
-I would say this is important but JHU is not enough; see above for benefits of random sampling to augment and improve decision making.
-
-\subsection*{Frequency and turnaround time}
-
-This supposes no selection bias.
-
-What is your goal?
-"get people who are currently sick off the street." Then test frequently and often.
-But we dont live in a mandatory testing environment (nor is it currently capable).
-Random testing is one way to improve upone current practice whiel we wait for that
-
-{\bf COORDINATION ON DATA SOURCES}
-
-Equation~\eqref{eq:auxinfoprob} presupposes covariate information is collected for all individuals in the nonprobability sample; however, this is often not feasible for state and local governments where rapid sample collection is prioritized over additional collection of covariate information.
+Application of the proposed methods relied on access to two critical datasets: (1) COVID-19 testing, case, hospitalization, and death data by demographic strata and (2) Facebook symptom survey.
+Public access to (1) required the author to submit a data request directly to the state of Indiana.
+Through conversations with XXXX, the author's request was one of only five to be approved by Indiana for public release.
+Most states do not release such granular data, making selection bias adjustments difficult.
+More systematic data collection and reporting is critical in future efforts.
+We acknowledge that this proposal will lead to important data privacy issues that will need to be addressed (cite).
+
+While we are eternally grateful to Indiana for the release, symptom information was not made available.
+This severely limits our ability to estimate correct selection propensity models.
+What is required in the future is a more detailed effort to collect relevant measures of an individual's past history that may impact their likelihood of seeking testing and being a positive COVID-19 case.
+Such data collection and dissemination is critical to address selection bias.
+
+A valid criticism of the proposed approach is that recording covariates on every tested individual is too time-consuming and costly, i.e., the data collection does not scale well.
+We argue that the list of relevant factors is not too long.
+For example, symptom status and COVID-19 contact are clearly relevant factors in testing selection.
+Future work may consider how to include/exclude various factors over time that are not significant to limit citizen burden in reporting.
+Moreover, equation~\eqref{eq:auxinfoprob} presupposes covariate information is collected for all individuals in the nonprobability sample; however, this is often not feasible for state and local governments where rapid sample collection is prioritized over additional collection of covariate information.
 We can extend~\eqref{eq:auxinfoprob} by randomly sampling a subset of the nonprobability sample on which to collect the covariate information, yielding the pseudo-likelihood
-\begin{equation}
-\label{eq:auxinfoprob2}
+\begin{equation*}
+% \label{eq:auxinfoprob2}
 \sum_{j=1}^N I_j I_j^{(2)} W_j^{(2)} \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
-\end{equation}
+\end{equation*}
 Here, $I_j^{(2)}$ is an indicator that the individual is in the probability subsample and $W_j^{(2)}$ is the probability of inclusion.
 Estimation proceeds by maximizing the pseudo-likelihood which now depends on random sampling in both the first and second term.
-
-\subsection*{Additional biases and measurement-error}
-
-Other biases exist and are incredible important when trying to estimate quantities such as the incubation period and mean serial interval.
-
-These estimates could be severely biased due to selection bias and measurement error, but also suffer from
-
-Nevertheless,
-Length bias and measurement-error?
+This would allow for rapid testing and minimizes cost of covariate collection.
 

### 2021-08-12 methods/11_effss_calcs.R
Create 11_effss_calcs.R

@@ -1,0 +1,78 @@
+## Bring in data and libraries
+age_data = read.csv("../Documents/GitHub/covid-umich/data/covid_indiana_age.csv")
+
+library(lubridate)
+library(xtable)
+
+## Clean Data and Define inputs
+
+age_data$DATE = date(age_data$DATE)
+temp = age_data[age_data$DATE >= "2020-04-25" & age_data$DATE <= "2020-04-29",]
+prevalence = 0.117
+N = 6.732e6
+
+f = sum(temp$COVID_TEST)/(N-sum(age_data$COVID_TEST[age_data$DATE < "2020-04-25"]))
+FP = 0.024
+FN = 0.13
+M = 1.3
+
+effss_calc <- function(prevalence, f, M, FN, FP) {
+  ## Effective SS calculation
+  f0 = f/((M-1) * prevalence + 1 )
+  f1 = M*f0
+  Delta = f1 - f0
+  ratio = sqrt(prevalence*(1-prevalence) / (f*(1-f)))
+  rho = Delta * ratio
+  odds = prevalence/(1-prevalence)
+  ratio2 = (FP*(1-prevalence) + FN * prevalence)/ (f0 *(1-prevalence) + f1 *prevalence)
+  DM = (1- Delta * ratio2)/(1-FP-FN)
+  effss = 1/(rho^2 * DM^2) * (f/(1-f))
+  return(effss) 
+}
+
+effss1 = effss_calc(prevalence, f, M, FN, FP)
+
+## Results in Section 3.1:
+
+print(paste("Number of tests between 25th and 29th of April 2020:", sum(temp$COVID_TEST)))
+print(paste("Indiana population:", N))
+print(paste("Sampling rate:", round(f,4)))
+print(paste("False positive rate:", FP))
+print(paste("False negative rate:", FN))
+print(paste("Relative sampling rate:", M))
+print(paste("Effective sample size:", round(effss,0)))
+
+
+## Even if 
+f2 = 0.01 # Increased sampling rate
+M2 = 1.1 # Decrease relative sampling rate
+effss2 = effss_calc(prevalence, f2, M2, FN, FP) 
+
+print(paste("Sampling rate:", round(f2,4)))
+print(paste("Relative sampling rate:", M2))
+print(paste("Effective sample size:", round(effss_2,0)))
+
+FN2 = 0.20 # Increase in the FN rate
[…]

### 2021-08-12 write-up/covid-writeup.tex
Paper edits

@@
 To address this, a procedure is presented that combines case-count data and random samples over time to estimate selection propensities based on key covariate information.
-We then combine these selection propensities with epidemiological forecast models using death-only data to construct a \emph{doubly robust} estimation method that accounts for both measurement-error and selection bias.
+We then combine these selection propensities with epidemiological forecast models to construct a \emph{doubly robust} estimation method that accounts for both measurement-error and selection bias.
 This method is then applied to estimate Indiana's prevalence using case-count, hospitalization, and death data with cumulative demographic information, the United States's only statewide random molecular sample collected from April 25--29th, and Facebook's COVID-19 symptom survey.
@@
 \cite{Hao2020} builds an extended SEIR model (SAPHIRE) to account for various presymptomatic infectiousness, time-varying ascertainment rates, transmission rates and population movements.
+
+Recent work by \cite{Zhao2021} pointed out that estimation of key epidemiological parameters such as the incubation time using standard epidemiological models can suffer from severe bias due to issues beyond selection bias and measurement error.
+Namely, right truncation and epidemic growth leading to patients ``being more likely to be infected towards the end of their exposure period''~\cite[pp.
+3]{Zhao2021}.
+Their approach carefully constructs the study sample and statistical model to account for these issues.
+In this paper, rather than buildinga  study sample, we ask whether we can collect auxiliary information to address selection bias in the observed case count data.
+
 Given a probability sampling design, individual predictions can be leveraged to improve estimation via model-assisted approaches~\citep{Breidt2017}.
 For nonprobability samples, \cite{Chen2019} derive a doubly robust approach that uses outcome predictions given covariates on the nonprobability and probability samples.
-Here, we combine the state-space SIR approach of \cite{Song2020} but instead, as in \cite{Johndrow2020}, focus on COVID-19 confirmed death count data.
+Here, we combine the state-space SEIR approach of \cite{Song2020} but instead, as in \cite{Johndrow2020}, focus on COVID-19 confirmed death count data.
 We generate epidemiological forecasts for point prevalence of each population strata.
@@
 A derived statistical error decomposition guides this discussion.
-
-\textcolor{red}{Discuss the additional biases and Qing paper.}
 
@@
 Secondary goals include estimation of rates of change and the effective reproduction number which characterize disease trajectory.
-A final goal is characterizing prevalence of population-level immunity.
+% A final goal is characterizing prevalence of population-level immunity.
 These quantities are fundamental to public health policy and provide critical information on the presence and transmission of SARS-CoV-2.
@@
 Figure~\ref{fig:in-cases} plots the number of reported confirmed COVID-19 cases per day in the state of Indiana.
-Figure~\ref{fig:in-tests} plots the total number of COVID-19 molecularw tests performed per day.
+Figure~\ref{fig:in-tests} plots the total number of COVID-19 molecular tests performed per day.
 Figure~\ref{fig:in-deaths} plots the total number of COVID-19 related reported deaths per day.
@@
 Unfortunately, most dashboards do not provide any demographic information on who requested a test nor on who tested positive for SARS-CoV-2 over time.
-Working closely with the State of Indiana, we were able to produce publicly available datasets for COVID-19 total tests, positive tests, and related deaths per day broken out marginally by age, gender, ethnicity, and race~\citep{IndianaData2021}.
-This granular dataset motivates the proposed approach.
+Working closely with the State of Indiana, we were able to obtain access to COVID-19 total tests, positive tests, and related deaths per day broken out marginally by age, gender, ethnicity, and race~\citep{IndianaData2021}.
+These granular datasets are now publicly available and motivate the proposed approach.
 % https://www.regenstrief.org/covid-dashboard/
@@
 
-Due to testing restrictions and other potential selection biases, publicly reported Covid testing data may not be sufficient to understand active infection rates or the disease trajectory.
-Here, we discuss two random samples that can be leveraged to as auxiliary information to address selection bias.
+Due to testing restrictions and other potential selection biases, publicly reported COVID-19 testing data may not be sufficient to understand active infection rates or the disease trajectory.
+Here, we discuss two random samples that provide auxiliary information and may help address selection bias.
 
@@
 Figure~\ref{fig:fbsymptoms} displays smoothed estimates of the fraction of individuals within age and gender strata who report a fever within the past 24-hours.
-This paper uses the Facebook symptom survey as the main source of auxiliary information on time-varying characteristics of the population of Indiana, e.g., fraction of population displaying symptoms or in contact with COVID-19 positive individuals).
+This paper uses the Facebook symptom survey as the main source of auxiliary information on time-varying characteristics of the population of Indiana, e.g., fraction of population displaying symptoms or in contact with COVID-19 positive individuals.
 
@@
 
-To start, we assume the overall number of COVID-19 cases and/or disease prevalence are of primary interest.
+To start, we assume the overall number of active COVID-19 cases and/or active infection rate (AIR) are of primary interest.
 That is, we are interested in either the population total $Y = \sum_{j=1}^N Y_j$ or the population average $\bar Y = Y/N$.
 Suppose that $n$ tests are performed and we observe the values $y_1, \ldots, y_n \in \{0,1\}$.
-Then a natural candidate for prevalence is the proportion of positive tests $\bar y = \frac{1}{n} \sum_{i=1}^n y_i$, and a natural candidate for overall cases is $N \times \bar y$.
+Then a natural candidate for AIR is the proportion of positive tests $\bar y = \frac{1}{n} \sum_{i=1}^n y_i$, and a natural candidate for overall active cases is $N \times \bar y$.
 Under simple random sampling (SRS) or any other epsem\footnote{equal probability of selection method} design, the above are unbiased estimators of the population-level quantities of interest.
@@
 The attentive data analyst will recognize the estimator $\bar y_n$ is biased even for simple random samples.
-In Appendix~\ref{app:memestimator}, assuming sensitivity and specificity were known a priori, we present a novel iterative procedure that uses the error decomposition to construct the estimator $\tilde y_n = (\bar y_n - FP)/(1-(FP+FN))$, which is unbiased under simple random sampling (SRS); see Appendix~\ref{app:modelbased} for a discussion of the connection to model-based estimators.
+In Appendix~\ref{app:memestimator}, assuming sensitivity and specificity are known a priori, a novel iterative procedure is used to construct the estimator $\tilde y_n = (\bar y_n - FP)/(1-(FP+FN))$, which is unbiased under simple random sampling (SRS); see Appendix~\ref{app:modelbased} for a discussion of the connection to model-based estimators.
 To understand the impact of selection bias and imperfect testing, we derive the following statistical decomposition of the error between $\tilde y_n$ and $\bar Y$:
@@
 This extends work by \cite{Meng2018} to account for imperfect testing.
-The first term represents \emph{data quality}, the second \emph{data quantity}, and the third \emph{problem difficulty}.
-The final term represents the \emph{imperfect testing adjustment} which is a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence but here is scaled by the factor $(1 - FP - FN)^{-1}$.
+The first three terms continue to represent \emph{data quality}, \emph{data quantity}, and \emph{problem difficulty} respectively.
+The new term represents the \emph{imperfect testing adjustment} which is a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence, which is scaled by the factor $(1 - FP - FN)^{-1}$.
 
 Figure~\ref{fig:heatmap} shows that $D_M$, as a function of the relative frequency ($f_1/f_0$) and odds ratio, can be both positive and negative as well as a range of magnitudes~\citep{Beesley2020,Beesley2019,Smeden2019}.
-Under either perfect testing (i.e., $FP=FN=0$), $D_M = 1$ so the relation between estimation and selection bias is simple, e.g., COVID-19 positive individuals more likely to receive test implies upward bias in prevalence estimates.
+Under perfect testing (i.e., $FP=FN=0$), $D_M = 1$ so the relation between estimation and selection bias is simple, e.g., COVID-19 positive individuals more likely to receive test implies upward bias in prevalence estimates.
 Under random testing (i.e., $f_0 = f_1$), $D_M = (1-FP-FN)^{-1}$ so measurement error simply scales the error.
@@
 The paradox that two countries with the same testing strategy can yield wildly different estimates due to population size is too often ignored in current communications of case-count data.
-When comparing the US and UK under the assumption of similar testing strategies, for example, the MSE increases by a factor of five.
 Thus, conclusions drawn from observed case-counts may not just be wrong, but very wrong.
@@
 $$
-As of June 10th 2020, the US has performed $21,474,531$ tests.
-The US population is roughly $312$ million, so $f = 0.069$.
-\textcolor{red}{FIX: The cumulative empirical prevalence is $9.3\%$ on May 8th.
-A recent study estimates a false positive and negative rate for antibody testing at $0.5$\% and $17.2$\% respectively \citep{Bendavid2020}.
-Supposing COVID-19 positive individuals are $2$ times more likely to get tested, then the effective sample size is $92$.Recent proposals~\citep{Siddarth2020} have argued for increased testing capacity, which may likely reduce the relative sampling rate.
-If the relative sampling rate drops to $1.2$ then the effective sample size will increase to $1950$.
+Between April 25th to 29th 2020, Indiana performed $95,879$ tests.
+Indiana's population is roughly $6.732$ million, so $f = 0.003$.
+The active infection rate is $11.7\%$ in this time interval.
+Recent studies have suggested 87\% sensitivity~\cite{Arevalo2020} and 97.6\% specificity~\cite{Cohen2020} are reasonable measurement error estimates for RT-PCR molecular tests.
+Supposing COVID-19 positive individuals are $1.3$ times more likely to get tested, then the effective sample size is $84$.
+Recent proposals~\citep{Siddarth2020} have argued for increased testing capacity, which may likely reduce the relative sampling rate.
+Even if the relative sampling rate drops to $1.1$ and $f$ increases to $0.01$ then the effective sample size will increase to $714$.
 Thus the effective sample size even in optimistic scenarios is equivalent to a small random sample from the population.
 Moreover, increased testing capacity may also come at the cost of increases in false positive and negative rates.
-Consider the case where $f$ increases from $0.05$ to $0.1$ and the increase is associated with the false negative rate rising from $5\%$ to $20$\%.
-Then the effective sample size increases by a factor of $1.76$ rather than the expected $4$ fold increase.}
+Consider the case where $f$ increasing from $0.003$ to $0.01$ and is associated with the false negative rate rising from $13\%$ to $20$\%.
+Then the effective sample size is $602$, representing a $7.2$ factor increase rather than the expected $8.5$ factor increase.
 % \begin{wrapfigure}{r}{0.5\textwidth}
@@
 
-\subsection{Computating the effective sample size}
+\subsection{Computing the effective sample size}
 \label{section:effss}
@@
 Suppose that $M = f_1/f_0$; then $f_0 = f / (\bar Y \cdot (M-1) + 1)$.
-Using the upper bound $E_{\I} [ \rho_{I,Y}^2 ] \leq E_{\I} [\rho_{I,Y} ]^2$, we compute effective sample size under a range of values $\bary$, $M$ with $f = 0.066$ (i.e., current sampling fraction) and observed prevalence $0.095$.
+Using the upper bound $E_{\I} [ \rho_{I,Y}^2 ] \leq E_{\I} [\rho_{I,Y} ]^2$, we compute effective sample size under a range of prevalences $\bary$, and relative sample rates $M$ given $f = 0.003$ (i.e., current sampling fraction).
 

### 2021-08-15 write-up/covid-writeup.tex
updaed paper through case study (all look good)

@@
 
-Due to monetary and time constraints, however, random testing is either not performed or performed infrequently.
+Due to monetary and time constraints, however, random testing is performed infrequently.
 As of June 2021, Indiana is the only state to conduct statewide random sample testing\footnote{This is the only random sample to collect both seroprevalence and diagnostic testing results.
@@
 While we demonstrate empirical improvements over simple disease prevalence estimates, we also highlight how selection bias may persist and impact uncertainty quantification.
-
-% Hospitalization Gelman
 
@@
 Their approach carefully constructs the study sample and statistical model to account for these issues.
-In this paper, rather than buildinga  study sample, we ask whether we can collect auxiliary information to address selection bias in the observed case count data.
+In this paper, rather than focusing on construction of a study sample, we ask whether one can collect auxiliary information to address selection bias in the observed case count data directly.
 
@@
 
-% Other biases exist and are incredible important when trying to estimate quantities such as the incubation period and mean serial interval.
-
-% These estimates could be severely biased due to selection bias and measurement error, but also suffer from
-
-% Nevertheless,
-% Length bias and measurement-error?
-
 % \subsection{Outline}
@@
 \label{section:testinginfo}
-
-% We start with a brief description of the difference between COVID-19 molecular and serological tests as well as the relationship between COVID-19 testing, the infection, and immune response time frame.
-These distinctions will be useful later in the paper when discussing different random sample tests.
 
@@
 The primary goal of molecular tests is diagnosis of active infections in the population.
-Such diagnostic tests generate important information about the presence of SARS-Cov-2 in the population, and help scientists and policy-makers understand patterns of transmission and propagation.
+Such diagnostic tests generate important information about the presence of SARS-CoV-2 in the population, and help scientists and policy-makers understand patterns of transmission and propagation.
 Rapid and frequent molecular testing is cited as a key component~\cite{OECD2021} in effective strategies to identify active infections and prevent systemic outbreaks.
-% Here we consider the current system of testing which involves both self-selection into testing and historical limits on who was able to receive diagnostic tests due to limited testing capacity.
-
-% \emph{Serological tests} look for an immunological response to the virus.
-A week or so after an individual is infected with SARS-CoV-2, the individual will start producing antibodies.
-At this point, a serological test with perfect sensitivity will come back positive.
-This test provides evidence of past infection while the molecular test provides evidence of an \emph{active} infection.
-
-% To better estimate SARS-CoV-2 immunity in a population, seroprevalence studies that generate a probabilistic population sample and perform serological tests on the sample can be collected.
-These studies are useful for disease surveillance.
-To date, the CDC has conducted ten large-scale geographic serological surveys with three rounds.
-While population-based sampling strategies provide a more representative nationwide sample, they are very time intensive and expensive.
 
@@
 Secondary goals include estimation of rates of change and the effective reproduction number which characterize disease trajectory.
-% A final goal is characterizing prevalence of population-level immunity.
 These quantities are fundamental to public health policy and provide critical information on the presence and transmission of SARS-CoV-2.
- % as well as the impact of public health  interventions.
 
@@
 
-Public databases most often contain aggregate information.
+Public databases most often, however, only contain aggregate information.
 The Johns Hopkins dashboard, for example, provides demographic breakdown of case counts as well as the total confirmed cases and deaths by county.
 Aggregate time series of case count and deaths can also be extracted.
-Unfortunately, most dashboards do not provide any demographic information on who requested a test nor on who tested positive for SARS-CoV-2 over time.
+Unfortunately, most dashboards do not provide demographic information on who requested a test nor on who tested positive for SARS-CoV-2 \emph{over time}.
 Working closely with the State of Indiana, we were able to obtain access to COVID-19 total tests, positive tests, and related deaths per day broken out marginally by age, gender, ethnicity, and race~\citep{IndianaData2021}.
@@
 % https://www.indystar.com/story/news/health/2020/06/12/indiana-says-anyone-who-wants-coronavirus-test-can-get-one/3179151001/
+Testing restrictions impact the propensity of an individual to receive a test and therefore, if COVID-19 positive, contribute to the case count.
+In this paper, testing restrictions are addressed by fitting time-varying testing propensities that depend on relevant covariate information.
 
@@
 
-Due to testing restrictions and other potential selection biases, publicly reported COVID-19 testing data may not be sufficient to understand active infection rates or the disease trajectory.
+Due to testing restrictions and other potential selection biases, publicly reported COVID-19 case cound data may not be sufficient to understand active infection rates or the disease trajectory.
 Here, we discuss two random samples that provide auxiliary information and may help address selection bias.
@@
 \includegraphics[width = 0.9\textwidth]{../figs/fbcovid19symptoms.png}
-\caption{Rate of reported fever per strata.  Daily rates were estimated using weighted method on each day separately and then smoothed over time using local-linear nonparametric regression.}
+\caption{Rate of reported fever per strata.
+Daily rates were estimated using weighted method suggested by~\citep{delphisurvey} on each day separately and then smoothed over time using local-linear nonparametric regression.}
 \label{fig:fbsymptoms}
@@
 If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^{\star} = Y_j(1-P_j) + (1-Y_j) P_j$.
-The attentive data analyst will recognize the estimator $\bar y_n$ is biased even for simple random samples.
+The attentive data analyst will recognize the estimator $\bar y_n$ is now biased for simple random samples.
 In Appendix~\ref{app:memestimator}, assuming sensitivity and specificity are known a priori, a novel iterative procedure is used to construct the estimator $\tilde y_n = (\bar y_n - FP)/(1-(FP+FN))$, which is unbiased under simple random sampling (SRS); see Appendix~\ref{app:modelbased} for a discussion of the connection to model-based estimators.
@@
 Figure~\ref{fig:heatmap} shows that $D_M$, as a function of the relative frequency ($f_1/f_0$) and odds ratio, can be both positive and negative as well as a range of magnitudes~\citep{Beesley2020,Beesley2019,Smeden2019}.
-Under perfect testing (i.e., $FP=FN=0$), $D_M = 1$ so the relation between estimation and selection bias is simple, e.g., COVID-19 positive individuals more likely to receive test implies upward bias in prevalence estimates.
-Under random testing (i.e., $f_0 = f_1$), $D_M = (1-FP-FN)^{-1}$ so measurement error simply scales the error.
+Assuming no measurement error, $D_M = 1$ so the relation between estimation and selection bias is simple, e.g., COVID-19 positive individuals more likely to receive test implies upward bias in prevalence estimates.
+Under random testing (i.e., $f_0 = f_1$), $D_M = (1-FP-FN)^{-1}$ so measurement error simply magnifies the error.
 When tests are imperfect and selection bias exists, this simple relationship no longer holds.
@@
 Indiana's population is roughly $6.732$ million, so $f = 0.003$.
-The active infection rate is $11.7\%$ in this time interval.
-Recent studies have suggested 87\% sensitivity~\cite{Arevalo2020} and 97.6\% specificity~\cite{Cohen2020} are reasonable measurement error estimates for RT-PCR molecular tests.
-Supposing COVID-19 positive individuals are $1.3$ times more likely to get tested, then the effective sample size is $84$.
+The active infection rate was estimated to be~$1.81\%$~\cite{Yiannoutsos2021} in this time interval.
+Recent studies have suggested 87\% sensitivity~\cite{Arevalo2020} and 97.6\% specificity~\cite{Cohen2020} are reasonable measurement error rates for RT-PCR tests.
+Supposing COVID-19 positive individuals are $1.5$ times more likely to get tested, then the effective sample size is $168$.
 Recent proposals~\citep{Siddarth2020} have argued for increased testing capacity, which may likely reduce the relative sampling rate.
-Even if the relative sampling rate drops to $1.1$ and $f$ increases to $0.01$ then the effective sample size will increase to $714$.
-Thus the effective sample size even in optimistic scenarios is equivalent to a small random sample from the population.
+Even if the relative sampling rate drops to $1.2$ and $f$ increases to $0.01$ then the effective sample size will increase to $1025$.
+Thus the effective sample size even in optimistic scenarios is equivalent to a moderate random sample from the population.
 Moreover, increased testing capacity may also come at the cost of increases in false positive and negative rates.
 Consider the case where $f$ increasing from $0.003$ to $0.01$ and is associated with the false negative rate rising from $13\%$ to $20$\%.
-Then the effective sample size is $602$, representing a $7.2$ factor increase rather than the expected $8.5$ factor increase.
+Then the effective sample size is $863$, representing a $5.1$ factor increase rather than the expected $6.1$ factor increase.
+See Section~\ref{section:effss} in the Supplementary Materials for additional effective sample size calculations.
 % \begin{wrapfigure}{r}{0.5\textwidth}
@@
 This leads to differences in when the bias is most pronounced.
-Figure~\ref{fig:r0-bias} displays the bias as a function of the relative sampling fraction assuming the population fraction that is susceptible remains large.
+Figure~\ref{fig:r0-bias} displays the bias as a function of the relative sampling fraction assuming the population fraction that is susceptible remains large (i.e., $S_t / S_{t-1} \approx 1$).
 
@@
 
-The prior section presented negative consequences of selection-bias and measurement error when estimating infection prevalence from observed case-count data.
-This raises the immediate question whether prevalence estimators can be improved.
+The prior section presented negative consequences of selection-bias and measurement error when estimating infection prevalence, rates of change, and the effective reproduction number from observed case-count data.
+This raises the immediate question whether we can can improve upon these estimators.
 In this section, we consider two directions.
@@
 In Lemma~\ref{lemma:ipw}, sensitivity and specificity are not assumed known.
-Instead, we suppose that these are estimated using a pseudo-likelihood on a random sample from the population whose size is chosen so the resulting estimated sensitivity and specificity have desired levels of uncertainty.
+Instead, we suppose that these are estimated using a pseudo-likelihood on two random samples from the population whose size (denoted~$n_{FP}$ and $n_{FN}$ respectively) is chosen so the resulting estimated sensitivity and specificity have desired levels of uncertainty.
 See Appendix~\ref{app:asympderivations} for additional details.
@@
 Let $I_{k,t}$ to denote the number of new infections on day $t$ in the $k$th strata.
-As there are a total of $N \cdot i_t := I_t$ new infections on day $t$, we assume the joint distribution of strata-specific new infection totals takes a multinomial distribution
+As there are a total of $N \cdot i_t := I_t$ new infections on day $t$, we assume the joint distribution of strata-specific new infection totals takes a multinomial distribution with
 $$
@@
 $$
-
-Selection bias in case count data is addressed by bringing in COVID-19 death data.  Let $D_{k,r}$ denote the number of individuals who pass away on day $r$ from strata $k$.  Then
-$$
-D_{k,r} \mid p, \theta, \nu \sim \text{Poisson} \left( \sum_{t=1}^r p_{k} I_{k,t} \theta_{(r-t)} \right)
-$$
-where $p_{k}$ is the infection fatality rate for the $k$th strata and $\theta_{j}$ is a discrete-time distribution for time from infection to death.
-This extends prior analysis of death data~\cite{Johndrow2020} by allowing the infection fatality rate to depend on stratification variable.
+where $( {\bf p}_t )_{t=1}^T = ( (p_{1,t}, \ldots, p_{K, t}) )_{t=1}^T$ are a sequence of parameters on the simplex, i.e., $\sum_{k=1}^K p_{k,t} = 1$.
+
+Selection bias is addressed by analyzing COVID-19 death data rather than case counts.  Let $D_{k,r}$ denote the number of individuals who pass away on day $r$ from strata $k$.  Then
+$$
+D_{k,r} \mid p, \theta, \nu \sim \text{Poisson} \left( \sum_{t=1}^r \text{IFR}_{k} \cdot I_{k,t} \theta_{(r-t)} \right)
+$$
+where $\text{IFR}_{k}$ is the infection fatality rate for the $k$th strata and $\theta_{j}$ is a discrete-time distribution for time from infection to death.
+This extends prior analysis of death data~\cite{Johndrow2020} by allowing the infection fatality rate to depend on a stratification variable, which is important as IFR depends heavily on age.
 Specifically, ``the estimated IFR is close to zero for children and younger adults but rises exponentially with age, reaching 0.4\% at age 55, 1.4\% at age 65, 4.6\% at age 75, and 15\% at age 85''~\cite{Levin2020}.
@@
 See Appendix~\ref{app:DRderivation} for the derivation.
-Comparing~\eqref{eq:statdecomp3} to~\eqref{eq:statdecomp2} shows that the addition of the model-based focast impacts the estimation error in two ways.
+Comparing~\eqref{eq:statdecomp3} to~\eqref{eq:statdecomp2} shows that the addition of the model-based forecast impacts the estimation error in two ways.
 First, if the model is adequate, then one may expect a reduction in the problem difficulty and (potentially) in the data quality components.
@@
 
-If the model-based forecast is correct then $E_{\bf R} [ \rho_{\tilde I (X), Y-\mu(X)} ] = 0$  and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the forecasts are not correctly specified then the data quality index is unlikely to inversely scale with population size.
+If the model-based forecast is correct then $E_{\bf R} [ \rho_{\tilde I (X), Y-\mu(X)} ] = 0$  and therefore $\rho_{\tilde I(X), Y-\mu(X)} = O(N^{-1})$; however, if the forecasts are not correctly specified then the data quality index is unlikely to inversely scale with population size.
 Interestingly, the impact of measurement-error on data quality now depends on a relative comparison of the data quality and problem difficulty of the weighted estimator and the doubly-robust estimator.
@@
 \includegraphics[width=.45\linewidth]{../figs/indianadeaths_byage.png}
- \caption{Deaths per per day}
+ \caption{Deaths per day}
  \label{fig:in-deaths}
 \end{subfigure}
-\caption{Indiana daily case count, testing, and death data by age strata}
+\caption{Indiana daily COVID-19 case count, testing, and death data by age strata}
 \end{figure}
@@
 We collect age, sex, and demographic information as well as COVID-19 related symptoms, which Facebook defines as having (1) a fever and cough, (2) shortness of breath, \emph{or} (3) difficulty breathing.
-To assess potential bias, we compare the random sample from Indiana to the Facebook symptom survey data collected from April 20th to April 25th.
+To assess potential bias, we compare the random sample from Indiana to the Facebook symptom survey data collected from April 25th to April 29th.
 Table~\ref{tab:comparison} demonstrates minimal bias in the Facebook symptom survey with respect to symptom distributions.
@@
 $$
-Using the constructed weights and the adjusted strata-level prevalence estimates, the IPW estimate under no measurement-error is $7.7$, a decrease of four percentage points from the observed prevalence of $11.7$.
+Using the constructed weights and the adjusted strata-level prevalence estimates, the IPW estimate under no measurement-error is $7.7\%$, a decrease of four percentage points from the observed prevalence of $11.7\%$.
 We also make use of Facebook's symptom survey over that window of time to construct a second IPW estimate conditional on demographics and fever but ignoring the household and prior testing information, which under no measurement-error is $6.5$\%.
@@
 This corresponds to a false negative rate of $13$\% and false positive rate of $2.4\%$.
-Therefore, the IPW estimates accounting for measurement-error are $6.2\%$ and $6.1\%$ respectively.
+Therefore, the IPW estimates accounting for measurement-error are $6.2\%$ and $4.9\%$ respectively.
 
@@
 Indiana's population as of 2019 was $6.732$ million.
-A total of $898$ tests were administered in Indiana between April 25th to 29th.
+A total of $19,649$ tests were administered in Indiana between April 25th to 29th.
 Subtracting off the $95,879$ tests that had already performed yields a sampling fraction of $f = 2.96 \times 10^{-3}$.
@@
 Symptom status (e.g., fever, coughing, shortness of breath) and related covariate information (COVID-19 positive contact), however, are currently unavailable\footnote{Whether this is due to data privacy considerations or lack of data collection efforts is unclear.}.
-As COVID-19 symptom status is likely to alter the testing propensity and likelihood of active infection, Facebook symptom survey and hospitalization data are used to estimate symptom status per strata in two ways which are each detailed below.
-Here, two simple imputation methods are considered for illustrative purposes; see Section~\ref{section:discussion} for additional discussion.
+As COVID-19 symptom status is likely to alter the testing propensity and likelihood of active infection, for illustrative purposes Facebook symptom survey and hospitalization data are used in two ways to imputate symptom status per strata.
+We outline the imputation methods below; see Section~\ref{section:discussion} for additional discussion.
 
@@
 
+\textcolor{red}{Underascertainmnet as pre-processed or modelled but often is an assumption that must be resolved parametrically. }
+
 \subsection*{Real-world implementation of the proposed method}
@@
 \centering
-\begin{tabular}{rrrrrr}
-& \multicolumn{5}{c}{$M$} \\ \cline{2-6}
-$\bar y$ & 1.2 & 1.4 & 1.6 & 1.8 & 2 \\
+\begin{tabular}{rrrrrrrr}
+& \multicolumn{7}{c}{$M$} \\ \cline{2-8}
+$\bar y$ & 1.05 & 1.15 & 1.25 & 1.35 & 1.45 & 1.55 & 1.65 \\
   \hline
-0.05 & 11303 & 2882 & 1306 & 749 & 489 \\
-  0.07 & 6065 & 1558 & 712 & 411 & 270 \\
-  0.09 & 3862 & 1000 & 460 & 268 & 177 \\
-  0.11 & 2724 & 711 & 329 & 193 & 129 \\
-  0.13 & 2057 & 541 & 252 & 149 & 100 \\
+0.01 & 40444 & 4503 & 1624 & 830 & 503 & 338 & 242 \\
+0.03 & 13787 & 1541 & 558 & 286 & 174 & 117 & 85 \\
+0.05 & 8463 & 950 & 345 & 178 & 109 & 73 & 53 \\
+0.07 & 6187 & 697 & 254 & 132 & 81 & 55 & 40 \\
+0.09 & 4928 & 557 & 204 & 106 & 65 & 44 & 32 \\
+0.11 & 4131 & 469 & 173 & 90 & 56 & 38 & 28 \\
    \hline
@@
 
-We also present the same plot under $FP = 0.005$ and $FN = 0.172$ to show the impact of measurement error on effective sample size.
+We also present the same plot under $FP = 0.024$ and $FN = 0.13$ to show the impact of measurement error on effective sample size.
 
@@
 \centering
-\begin{tabular}{rrrrrr}
-& \multicolumn{5}{c}{$M$} \\ \cline{2-6}
-$\bar y$ & 1.2 & 1.4 & 1.6 & 1.8 & 2 \\
+\begin{tabular}{rrrrrrrr}
+& \multicolumn{7}{c}{$M$} \\ \cline{2-8}
+ & 1.05 & 1.15 & 1.25 & 1.35 & 1.45 & 1.55 & 1.65 \\
   \hline
-0.05 & 6480 & 1656 & 752 & 432 & 282 \\
-  0.07 & 3304 & 852 & 391 & 227 & 149 \\
-  0.09 & 2059 & 536 & 248 & 145 & 97 \\
-  0.11 & 1440 & 379 & 177 & 104 & 70 \\
-  0.13 & 1086 & 288 & 136 & 81 & 55 \\
+0.01 & 29019 & 3247 & 1177 & 605 & 368 & 248 & 179 \\
+0.03 & 9894 & 1112 & 405 & 209 & 128 & 87 & 63 \\
+0.05 & 6075 & 686 & 251 & 130 & 80 & 54 & 39 \\
+0.07 & 4442 & 504 & 185 & 96 & 59 & 41 & 30 \\
+0.09 & 3539 & 403 & 149 & 78 & 48 & 33 & 24 \\
+0.11 & 2967 & 339 & 126 & 66 & 41 & 28 & 21 \\
    \hline
@@
 
-\section{Effect modifiers in COVID-19 clinical trials}
-
-Here we show how selection bias and measurement error can creep into clinical trial analysis.
+% \section{Effect modifiers in COVID-19 clinical trials}
+
+% Here we show how selection bias and measurement error can creep into clinical trial analysis.
 The key concern is whether clinical trials on COVID-19 recruit from the pool of individuals who have tested positive for COVID-19 or whether they sample randomly from the population, test, and then recruit from this subset of tested individuals.
@@
 In counterfactual language,
-$$
-E( Y(A=1) - Y(A=0) ) = E_U ( E_A ( Y(A=1) - Y(A=0) \mid U=u ))
-= \beta_1 + \beta_3 E(U).
-$$
-Even in a randomized experiment, the question is what the correct value for $E(U)$ is.
+% $$
+% E( Y(A=1) - Y(A=0) ) = E_U ( E_A ( Y(A=1) - Y(A=0) \mid U=u ))
+% = \beta_1 + \beta_3 E(U).
+% $$
+% Even in a randomized experiment, the question is what the correct value for $E(U)$ is.
 For COVID-19, the value of interest is likely the expected value of $U$ in the population of COVID-19 positive individuals, i.e., $E_J [ U_J Y_J]/ E_J[Y_J] = \sum_{j=1}^N U_j Y_j / \sum_{j=1}^N Y_j$.
@@
 Due to selection bias, the bias in the marginal treatment effect compared to the true marginal effect of interest is approximately equal to
-$$
-\beta_3 \times \sqrt{\frac{1-f}{f}} \times \frac{\bar U^\prime}{\bar Y} \bigg[ \rho_{I,U^\prime} \times CV (U^\prime)  - \rho_{I,Y} \times CV (Y) \bigg] \left[ 1 - \rho_{I,Y} \sqrt{\frac{1-f}{f}} CV (Y) \right]
-$$
-where $U^\prime = U Y$.
+% $$
+% \beta_3 \times \sqrt{\frac{1-f}{f}} \times \frac{\bar U^\prime}{\bar Y} \bigg[ \rho_{I,U^\prime} \times CV (U^\prime)  - \rho_{I,Y} \times CV (Y) \bigg] \left[ 1 - \rho_{I,Y} \sqrt{\frac{1-f}{f}} CV (Y) \right]
+% $$
+% where $U^\prime = U Y$.
 The derivation again uses a second-order Taylor series approximation and follows exactly the same as in Appendix~\ref{app:ratio}.
@@
 
-Randomization of treatment assignment in clinical trials negates unobserved confounders.
+% Randomization of treatment assignment in clinical trials negates unobserved confounders.
 This, however, does not negate effect modifiers.

### 2021-08-15 methods/11_effss_calcs.R

@@ -14,7 +14,7 @@ N = 6.732e6
 f = sum(temp$COVID_TEST)/(N-sum(age_data$COVID_TEST[age_data$DATE < "2020-04-25"]))
 FP = 0.024
 FN = 0.13
-M = 1.3
+M = 1.5
 
 effss_calc <- function(prevalence, f, M, FN, FP) {
   ## Effective SS calculation
@@ -40,23 +40,25 @@ print(paste("Sampling rate:", round(f,4)))
 print(paste("False positive rate:", FP))
 print(paste("False negative rate:", FN))
 print(paste("Relative sampling rate:", M))
-print(paste("Effective sample size:", round(effss,0)))
+print(paste("Effective sample size:", round(effss1,0)))
 
 
 ## Even if 
 f2 = 0.01 # Increased sampling rate
-M2 = 1.1 # Decrease relative sampling rate
+M2 = 1.2 # Decrease relative sampling rate
 effss2 = effss_calc(prevalence, f2, M2, FN, FP) 
 
 print(paste("Sampling rate:", round(f2,4)))
 print(paste("Relative sampling rate:", M2))
-print(paste("Effective sample size:", round(effss_2,0)))
+print(paste("Effective sample size:", round(effss2,0)))
+print(paste("Factor increase:", round(effss2/effss1,1)))
 
 FN2 = 0.20 # Increase in the FN rate
 effss_newFN = effss_calc(prevalence, f2, M2, FN2, FP) 
 
 print(paste("Effective sample size:", round(effss_newFN,0)))
 print(paste("% Reduction in effective sample size:", round((effss2-effss_newFN)/effss2,3)))
+print(paste("Factor increase:", round(effss_newFN/effss1,1)))
 
 
 ## APPENDIX B.8 

### 2021-08-17 write-up/covid-writeup.tex
ifr in writeup - need more model details

@@
 \def\I{\bf I}
+\def\code#1{\texttt{#1}}
 
@@
 
+We fit the
+
+\begin{align*}
+\frac{\partial s_t}{\partial t} &= - \beta s_t i_t; \quad \frac{\partial e_t}{\partial t} = \beta_t s_t i_t - \sigma e_t; \quad \\
+\frac{\partial i^{(T)}_t}{\partial t} &= \sigma e_t - \gamma i_t; \quad
+\frac{\partial r_t}{\partial t} = - \gamma i_t.
+\end{align*}
+
+We allow $\beta_t$ to be time-varying to account for the
+
+"2020-03-23", "2020-06-15", "2020-10-01".
+
+$w_t = 1/ \left( 1 + \exp \left( - \xi (t - t_1 - \nu)  \right) \right)$
+
+We defined a weight $w^{(1)}_t = 1$ prior to time $t_1$ and goes to $1$ as $t \to \infty$
+
+to accoun tfor
+
+$\beta (1-w_t) + \eta weight$
+
+For simplicity, we list the priors used in STAN below:
+\begin{itemize}
+  \item \code{beta $\sim$ Normal(1.5, 1) T[0,];}
+  \item \code{gamma $\sim$ normal(0.3, 0.5) T[0,];}
+  \item \code{sigma $\sim$ normal(0.4, 0.5) T[0,];}
+  \item \code{phi\_inv $\sim$ exponential(5);}
+  \item \code{i0 $\sim$ normal(0, 10);}
+  \item \code{e0 $\sim$ normal(0, 10);}
+  \item \code{eta $\sim$ normal(1, 1) T[0,];}
+  \item \code{eta\_two $\sim$ normal(1, 1) T[0,];}
+  \item \code{eta\_three $\sim$ normal(1, 1) T[0,];}
+  \item \code{nu $\sim$ exponential(1./5);}
+  \item \code{nu\_two $\sim$ exponential(1./5);}
+  \item \code{nu\_three $\sim$ exponential(1./5);}
+  \item \code{xi\_raw $\sim$ beta(1, 1);}
+  \item \code{phi = 1/phi\_inv;}
+  \item \code{xi = xi\_raw + 0.5;}
+\end{itemize}
+
+\begin{figure}[!th]
+\centering
+ \includegraphics[width=.9\linewidth]{../figs/logifr.png}
+ \caption{Logarithm of Infection Fatality Rate (IFR) by Age.  Linear model on the log-scale fits well and allows us to extrapolate publicly-reported IFR numbers to entire age range.}
+ \label{fig:logifr}
+\end{figure}
+
 \end{document}

### 2021-08-17 preprocessing/7_ifr.R

@@ -3,8 +3,14 @@
 
 age = c(9, 33, 55,65,75,85)
 ifr = c(0.002, 0.05, 0.4,1.4,4.6,15)
-plot(age,log(ifr))
-abline(lm(log(ifr) ~ age))
+par(mar = c(4,4,0.1,1)+0.1)
+png(filename = "../figs/logifr.png",
+    width = 960, height = 480, units = "px", pointsize = 25)
+plot(age,log(ifr), bty = "n", xlab = "Age", ylab = "log(IFR)", xlim = c(0,80), ylim = c(-8,2))
+seq_age = seq(2,80,1)
+seq_logifr = predict(lm(log(ifr) ~ age), newdata=data.frame(age = seq_age))
+lines(x = seq_age, y = seq_logifr)
+dev.off()
 
 ## Suggests a linear fit on log-scale 
 fit <- lm(log(ifr) ~ age)

### 2021-08-17 write-up/covid-writeup.tex
updated paper. now in appendix

@@
 \subsubsection{Testing restrictions and public health policy in Indiana}
+\label{section:publicpolicyindiana}
 
@@
 The population-level SEIR dynamics are then given by the following set of differential equations:
-\begin{align*}
-\frac{\partial s_t}{\partial t} &= - \beta s_t i_t; \quad \frac{\partial e_t}{\partial t} = \beta s_t i_t - \sigma e_t; \quad \\
+\begin{align}
+\frac{\partial s_t}{\partial t} &= - \beta s_t i_t; \quad \frac{\partial e_t}{\partial t} = \beta s_t i_t - \sigma e_t; \quad \label{eq:seir} \\
 \frac{\partial i^{(T)}_t}{\partial t} &= \sigma e_t - \gamma i_t; \quad
-\frac{\partial r_t}{\partial t} = - \gamma i_t.
-\end{align*}
+\frac{\partial r_t}{\partial t} = - \gamma i_t. \nonumber
+\end{align}
 Following~\cite{Song2020}, we employ Runge-Kutta (RK4) approximations for discretization.
@@
 Using these two models, mean imputation is used to calculate the number of positive and negative tests given demographic strata, fever status, and COVID-19 contact status.
-With the resulting dataset, inverse probability weights are computed based (termed IPW1).
+Inverse probability weights are then computed based on the resulting dataset (termed IPW1).
 
 In the second approach, we leverage hospitalization information.
-The first estimates the probability of fever given demographic and COVID-19 test status.
-When the COVID-19 test status is positive, the probability of fever depends on whether the individual was hospitalized.
+We first estimate the probability of fever given demographic and COVID-19 test status.
+When the COVID-19 test status is positive, the probability of fever is allowed to depend on whether the individual was hospitalized.
 Figure~\ref{fig:symptomlik1_model2} and~\ref{fig:symptomlik2_model2} in Appendix~\ref{app:in_add_details} presents the estimated likelihood of reporting a fever with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.
@@
 We use these models and the hospitalization rates by demographic to perform mean impuation of the number of positive and negative tests given demographic strata, fever status, and COVID-19 contact status.
-With the resulting dataset, inverse probability weights are computed based (termed IPW2).
+Inverse probability weights are then computed based on the resulting dataset (termed IPW2).
 
@@
 In this paper, similar to~\cite{Johndrow2020}, we specify fixed per-strata infection fatality rates.
-Based on published age-specific IFRs~\cite{Levin2020}, IFR by age seems to closely follow a log-linear relationship.
+Based on published age-specific IFRs~\cite{Levin2020}, Figure~\ref{fig:logifr} shows that IFR seems to closely follow a log-linear relationship with age.
 Combining across published age-specific IFRs, Table~\ref{tab:ifrperage} presents estimates of the average IFR within each strata using the age-distribution.
@@
 This was chosen to closely mimic the distribution from~\cite{Johndrow2020}.
-See~\ref{app:modelbased} for additional details on parametrization and choice of priors.
+
+The SEIR model in equation~\eqref{eq:seir} is extended to allow for time-varying transmission rates~$\beta_t$ to account for public policy changes discussed in Section~\ref{section:publicpolicyindiana} over the time period considered.
+Here, we insert three change points $(t_1,t_2,t_3)$ on March 23rd, June 15th, and October 1st of 2020 respectively.
+Given sensitivity to these choices, instead of considering a multiplicative structure, we allow for delayed implementation and slow change in the transmission rate by using weights:
+$$
+\beta_t = \sum_{k=1}^{3} \left(\beta_{k-1} (1-w^{(k)}_t ) + \beta_{k} w_t\right),\quad w^{(k)}_t = (1 + \exp \left( -\xi (t - t_k - \nu) \right))^{-1}
+$$
+where $\nu \geq 0$ is the potential delay and $\xi$ is the rate of change.
+Each weight~$w_t^{(k)} = 0$ for $t < t_k$ and $w_t^{(k)} \to 1$ as $t \to \infty$.
+See Section~\ref{app:modelbased} in the supplementary materials for details on prior specification.
+
 To check how well our model fits the observed data, a posterior predictive check on aggregate death counts was performed.
@@
 
-While the model-based estimates appear more reasonable, note that the estimate on the last week of April is $0.1$\% which is a severe underestimate.
+\textcolor{red}{
+While the model-based estimates appear more reasonable, note that the estimate on the last week of April is XX\% which is a severe underestimate.
 Caution is therefore warranted when interpreting these estimates.
 We know that the age-specific infection fatality rate is likely time-varying and may vary by other factors beyond age.
-Therefore, improvements to these model-based predictions could be made by improving the IFR estimates; however, there is currently minimal publicly available data to do so.
+Therefore, improvements could be made by more finely grained, and accurate IFR estimates; however, there is currently minimal publicly available data to do so.
 Moreover, the (thankfully) low number of deaths per strata make uncertainty in these estimates quite high.
@@
 Finally, the doubly robust estimates appear similar to the IPW estimate.
+}
 
@@
 \end{figure}
+
+\textcolor{red}{
+We also construct ratio and effective reproduction number estimates as discussed in Section XX.
+Figure XX presents the results.
+Unweighted, IPW, and doubly robust estimates appear similar.
+Recall the severe bias during the growth phase due to selection bias dicussed in SEction XX may imply the observed rates are severely overestimated between October 2020 and January 2021.
+The model-based methods may be conservative in this time window.
+A similar argument holds for the effective reproduction rate in Figure XXb.
+}
 
@@
 Contact tracing is incredibly useful for identifying carriers early and preventing spread of the disease.
-We would argue that, after the current wave, understanding active infections is also key.
+We argue that understanding active infections is also key.
 We know that the probability of an outbreak is a function of current prevalence and network connectivity.
@@
 
-\textcolor{red}{Underascertainmnet as pre-processed or modelled but often is an assumption that must be resolved parametrically. }
+%\textcolor{red}{Underascertainmnet as pre-processed or modelled but often is an assumption that must be resolved parametrically. }
 
@@
 Public access to (1) required the author to submit a data request directly to the state of Indiana.
-Through conversations with XXXX, the author's request was one of only five to be approved by Indiana for public release.
+Through conversations with the state of Indiana, the author's request was one of only five data requests to be approved by Indiana for public release.
 Most states do not release such granular data, making selection bias adjustments difficult.
 More systematic data collection and reporting is critical in future efforts.
-We acknowledge that this proposal will lead to important data privacy issues that will need to be addressed (cite).
+We acknowledge that this proposal will lead to data privacy issues that will need to be addressed.
 
@@
 Moreover, equation~\eqref{eq:auxinfoprob} presupposes covariate information is collected for all individuals in the nonprobability sample; however, this is often not feasible for state and local governments where rapid sample collection is prioritized over additional collection of covariate information.
-We can extend~\eqref{eq:auxinfoprob} by randomly sampling a subset of the nonprobability sample on which to collect the covariate information, yielding the pseudo-likelihood
+We can easily extend~\eqref{eq:auxinfoprob} by randomly sampling a subset of the nonprobability sample on which to collect the covariate information, yielding a new pseudo-likelihood
 \begin{equation*}
@@
 
-Suppose disease prevalence was estimated as the fraction who tested positive for COVID-19, i.e., $\bar y_n^\star = \frac{\sum_{i=1}^N I_j Y_j^\star}{\sum_{i=1}^N I_j}$.
-We can again investigate the error compared to the true prevalence $\bar Y$ in statistical terms:
-$$
-\bar y_n^\star - \bar Y = \sqrt{\frac{1-f}{f}} \left[ \rho_{I,Y} \times \sigma_Y + \rho_{I,PZ} \times \sigma_{PZ} + \sqrt{\frac{f}{1-f}}  \left( FP - (FP+FN) \bar Y \right) \right] .
-$$
-where $Z = 1-2Y$; see Appendix~\ref{app:imperfect} for the derivation.
-The first term in the large brackets represents the perfect testing regime; to this end, we refer to $\rho_{I,Y}$ as the \emph{true data quality}.
-The second term represents the interaction between imperfect testing and selection bias.
-The variable $PZ = 1[Y=0,P=1] - 1[Y=1, P=1]$ is non-zero only when $P=1$ (i.e., the outcome is incorrectly reported) and the sign depends on the true outcome $Y$.
-Given this, we refer to $\rho_{I,PZ}$ as the \emph{observed data quality} that accounts for both selection bias and measurement error.
-In the appendix, we show the sign of $\rho_{I,PZ}$ is the opposite of the sign of $\rho_{I,Y}$, which implies that the observed data quality adds error in the opposite direction from the true data quality.
-Finally, the third term represents the bias due to imperfect testing.
-
-\subsubsection{Connecting true and observed data quality}
-
-We start by considering the first two terms and assess whether the sign of the bias can reverse due to the interaction of measurement error and selection bias.
-To do this, we define the sampling rates differential.
-Let $f_1 := \pr (I_J = 1 \mid Y_J = 1)$ and $f_0 := \pr(I_J = 1 \mid Y_J = 0)$ be the sampling rates where $J$ is a random index defined on $\{1,\ldots, N\}$.
-Then $\Delta = f_1 - f_0$ is the sampling rate differential.
-Using these terms, we can re-express the first two terms as
-$$
-\rho_{I,Y} \times \sigma_Y + \rho_{I,PZ} \times \sigma_{PZ} =
-\rho_{I,Y} \times \sigma_Y \left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right].
-$$
-final term represents the \emph{imperfect testing adjustment} which is a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence.
-Note $\sgn(\Delta) = \sgn(\rho_{I,Y})$ by equation~\ref{eq:binaryrho} in the Appendix, which implies the measurement error adjustment either shrinks the data quality measure toward zero or reverses its sign.
-
-% While prior investigations have noted the interaction between measurement error and selection bias~\citep{Beesley2020,Beesley2019,Smeden2019}, the interaction with the sample size relative to the population, i.e., $f$, has largely been ignored.
-The above statistical decomposition clarifies the importance of this quantity~$f$.
-In particular, note that the statistical error also includes a bias term due to measurement error and this term increases as the sampled fraction $f$ increases.
-Therefore, how the first two terms interact with the final term depends on the fraction of the population sampled.
-This interaction is complex, but implies that whether the estimate $\bar y^\star_n$ is an overestimate or underestimate is a complicated question due to the relation amongst these three pieces.
-
-% Consider again the current COVID-19 pandemic.
-For now, we continue to assume the ratio of conditional selection rates $f_1/f_0$ is equal to 1.5.
-In Section~\ref{section:est_dq}, we discuss recent research suggesting a false negative rate around 17.2\% and a false positive rate around 0.05\%.
-Under these rates and the current US prevalence rates, the ratio of the MSE to MSE under no measurement error is 0.436; if we switch false negative rates down to 5\% and increase the false positive rate to 5\% then the relative MSE is 2.89.
-This is merely to demonstrate that in some cases we see a huge increase in MSE and in other settings we have a huge decrease in MSE.
-What drives this is the false positive and negative rate interaction with prevalence and sampling rates.
-Therefore, whether we are better or worse off with respect to the MSE is a very difficult question to answer.
-
-\subsection{An improved estimator under imperfect testing}
-
-Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the outcome and $0$ when we observe the true outcome.
+Recall that $P_j$ is  an indicator of measurement error, equal to $1$ when we incorrectly measure the outcome and $0$ when we observe the true outcome.
 We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.
@@
 Suppose there is a sequence of finite populations of size $N_{\nu}$ indexed by $\nu$.
-Each finite population has an associated non-probability at time $t$ of  size~$\tilde n$ and $L_\nu$ probability samples of fixed size $n$ drawn at equally spaced times $\{ t^\prime_l \}_{l=1}^L$ over the study window~$[0,T]$.
-Let~$\Delta_\nu$ denote the gap times between consecutive sample times.
+Each finite population has an associated non-probability at time $t$ and probability samples of size $n$ at each time~$t=1,\ldots, T$.
 Finally, let~$n_{FP}$ and $n_{FN}$ denote the sample sizes of $S_{FP}$ and $S_{FN}$ respectively.
@@
 
-\section{Model-based: Additional details}
+\section{Model-based: Prior specification}
 \label{app:modelbased}
-
-We fit the
-
-\begin{align*}
-\frac{\partial s_t}{\partial t} &= - \beta s_t i_t; \quad \frac{\partial e_t}{\partial t} = \beta_t s_t i_t - \sigma e_t; \quad \\
-\frac{\partial i^{(T)}_t}{\partial t} &= \sigma e_t - \gamma i_t; \quad
-\frac{\partial r_t}{\partial t} = - \gamma i_t.
-\end{align*}
-
-We allow $\beta_t$ to be time-varying to account for the
-
-"2020-03-23", "2020-06-15", "2020-10-01".
-
-$w_t = 1/ \left( 1 + \exp \left( - \xi (t - t_1 - \nu)  \right) \right)$
-
-We defined a weight $w^{(1)}_t = 1$ prior to time $t_1$ and goes to $1$ as $t \to \infty$
-
-to accoun tfor
-
-$\beta (1-w_t) + \eta weight$
 
@@
 
-\begin{figure}[!th]
-\centering
- \includegraphics[width=.9\linewidth]{../figs/logifr.png}
- \caption{Logarithm of Infection Fatality Rate (IFR) by Age.  Linear model on the log-scale fits well and allows us to extrapolate publicly-reported IFR numbers to entire age range.}
- \label{fig:logifr}
-\end{figure}
-
 \end{document}

### 2021-08-17 write-up/covid-writeup.tex
keeping general theory but going to clarify.  Then finalize paper!

@@
 
-Suppose there is a sequence of finite populations of size $N_{\nu}$ indexed by $\nu$.
-Each finite population has an associated non-probability at time $t$ and probability samples of size $n$ at each time~$t=1,\ldots, T$.
+uppose there is a sequence of finite populations of size $N_{\nu}$ indexed by $\nu$.
+Each finite population has an associated non-probability at time $t$ of  size~$\tilde n$ and $L_\nu$ probability samples of fixed size $n$ drawn at equally spaced times $\{ t^\prime_l \}_{l=1}^L$ over the study window~$[0,T]$.
+Let~$\Delta_\nu$ denote the gap times between consecutive sample times.
 Finally, let~$n_{FP}$ and $n_{FN}$ denote the sample sizes of $S_{FP}$ and $S_{FN}$ respectively.
@@
 where $\pi_{j,t^\prime} = \pi (X_{j,t^\prime}; \theta_t)$, $Z_j$ is an indicator of a false positive in the sample $S_{FP}$  and $\tilde Z_j$ is an indicator of a false negative in the sample $S_{FN}$.
-We assume the synthetic $|S_{FP}|^{-1} \sum_{j \in S_{FP}} Z_j = 0.024$ and $|S_{FP}|^{-1} \sum_{j \in S_{FP}} Z_j = 0.30$.
-\textcolor{red}{Sample sizes, denoted $n_{FP}$ and $n_{FN}$ respectively, are chosen to achieve desired estimator variance, i.e., $\frac{0.024 \cdot (0.976)}{n_{FP}} = \sigma^2_{FP}$ and $\frac{0.30 \cdot 0.70}{n_{FN}} = \sigma^2_{FN}$.}
+We assume the $|S_{FP}|^{-1} \sum_{j \in S_{FP}} Z_j = 0.024$ and $|S_{FP}|^{-1} \sum_{j \in S_{FP}} Z_j = 0.30$.
+Sample sizes, denoted $n_{FP}$ and $n_{FN}$ respectively, are chosen to achieve desired estimator variance, i.e., $\frac{0.024 \cdot (0.976)}{n_{FP}} = \sigma^2_{FP}$ and $\frac{0.13 \cdot 0.87}{n_{FN}} = \sigma^2_{FN}$.
+Here we set $\sigma^2_{FP} = 0.02^2$ and $\sigma_{FN}^2 = 0.05^2$ which yield $n_{FP} \approx 59$ and $n_{FN} \approx 45$ respectively.
 

### 2021-08-18 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 
-Suppose there is a sequence of finite populations of size $N_{\nu}$ indexed by $\nu=1,2,\ldots$.
-Each finite population has an associated non-probability at time $t$ and probability samples of size~$n$ at each time~$t=1,\ldots,T$.
+Suppose there is a sequence of finite populations of size $N_{\nu}$ indexed by $\nu$.
+Each finite population has an associated non-probability at time $t$ of  size~$\tilde n$ and $L_\nu$ probability samples of fixed size $n$ drawn at equally spaced times $\{ t^\prime_l \}_{l=1}^L$ over the study window~$[0,T]$.
 Assuming the selection propensity model is correctly specified, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:ipw} shows that the IPW estimator is  consistent and calculates the estimator's variance.
@@
 where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
-Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ where $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$.
+Under certain regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ where $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$.
 \end{lemma}
@@
 \subsubsection{Time-varying IPW estimator}
+\label{section:tvipw}
 
@@
 
-uppose there is a sequence of finite populations of size $N_{\nu}$ indexed by $\nu$.
+Suppose there is a sequence of finite populations of size $N_{\nu}$ indexed by $\nu$.
 Each finite population has an associated non-probability at time $t$ of  size~$\tilde n$ and $L_\nu$ probability samples of fixed size $n$ drawn at equally spaced times $\{ t^\prime_l \}_{l=1}^L$ over the study window~$[0,T]$.
@@
 For simplicity, dependency on $\nu$ is suppressed and the limiting process is represented by $N \to \infty$.
+The asymptotic argument below relies on regularity conditions C1-C6 from~\cite{Chen2019}, smoothness of the population-level distribution of the time-varying covariates~$\{ X_{j,t} \}$, and smoothness of the propensity function~$\pi(x; \theta)$ as a function of $x \in \mathbb{R}^p$.
 
@@
 \frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t}{\pi_{j,t}} \\
-\frac{1}{N \sum_{l=1}^L K_{t,t_l^\prime}} \sum_{l=1}^L K_{t,t_l^\prime} \left[ \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime_l} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime}  \pi_{j,t_l^\prime} X_{j,t_l^\prime} \right] \\
+\frac{1}{N \sum_{l=1}^L K_{t,t_l^\prime}} \sum_{l=1}^L K_{t,t_l^\prime} \left[ \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime_l} - \sum_{j = 1}^N \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime}  \pi_{j,t_l^\prime} X_{j,t_l^\prime} \right] \\
 \frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
@@
 Sample sizes, denoted $n_{FP}$ and $n_{FN}$ respectively, are chosen to achieve desired estimator variance, i.e., $\frac{0.024 \cdot (0.976)}{n_{FP}} = \sigma^2_{FP}$ and $\frac{0.13 \cdot 0.87}{n_{FN}} = \sigma^2_{FN}$.
-Here we set $\sigma^2_{FP} = 0.02^2$ and $\sigma_{FN}^2 = 0.05^2$ which yield $n_{FP} \approx 59$ and $n_{FN} \approx 45$ respectively.
-
-To prove consistency, first consider the setting where a random sample of size $n$ is observed at time $t$.
-Let $\tilde \Phi (\eta)$ denote the corresponding estimating equations, which is the same as $\Phi(\eta)$ except for the second component which is given by
+In this paper, we set $\sigma^2_{FP} = 0.02^2$ and $\sigma_{FN}^2 = 0.05^2$ which means $n_{FP} \approx 59$ and $n_{FN} \approx 45$ respectively.
+
+To prove consistency, first consider the setting where a single random sample of size $n$ is observed at time $t$.
+Let $\tilde \Phi (\eta)$ denote the corresponding estimating equations, which is the same as $\Phi(\eta)$ except for the second component which is now given by
 $$
@@
 
-To prove consistency under the sequence of sampling regimes, we need to show that $E[ \Phi (\eta_0) ] \to 0$ as $N \to \infty$.  First, we see that
+To prove consistency under the sequence of sampling regimes, we need to show that $E[ \Phi (\eta_0) ] \to 0$ as $N \to \infty$.  First, we study the difference in the 2nd term:
 \begin{align*}
-&\frac{1}{N}  E \left[ \sum_{j = 1}^N \tilde I_{j,t} \tilde W_{j,t}  \pi_{j,t} X_{j,t} - \sum_{l=1}^L  \sum_{j = 1}^N \tilde K_{t,t^\prime_l} \tilde I_{j,t} \tilde W_{j,t}  \pi_{j,t} X_{j,t^\prime_l} \right] \\
+&\frac{1}{N}  E \left[ \sum_{j = 1}^N \tilde I_{j,t} \tilde W_{j,t}  \pi_{j,t} X_{j,t} - \sum_{l=1}^L  \sum_{j = 1}^N \tilde K_{t,t^\prime_l} \tilde I_{j,t^\prime_l} \tilde W_{j,t^\prime_l}  \pi_{j,t^\prime_l} X_{j,t^\prime_l} \right] \\
 = &\frac{1}{N} \sum_{j = 1}^N \left[ \pi_{j,t} X_{j,t} - \sum_{l=1}^L \tilde K_{t,t^\prime_l} \pi_{j,t^\prime_l} X_{j,t^\prime_l} \right], \\
+= &\frac{1}{N} \sum_{j = 1}^N \left[ \sum_{l=1}^L \tilde K_{t,t^\prime_l} \pi_{j,t} X_{j,t} \pm \sum_{l=1}^L \tilde K_{t,t^\prime_l} \pi_{j,t_l^\prime} X_{j,t}  - \sum_{l=1}^L \tilde K_{t,t^\prime_l} \pi_{j,t^\prime_l} X_{j,t^\prime_l} \right], \\
 = &\frac{1}{N} \sum_{j = 1}^N \left[ \left( \sum_{l=1}^L \tilde K_{t,t^\prime_l}  \left( \pi (X_{j,t}; \theta) - \pi (X_{j,t^\prime_l}; \theta) \right) \right) X_{j,t} + \sum_{l=1}^L \tilde K_{t,t^\prime_l} \pi_{j,t^\prime_l} (X_{j,t} - X_{j,t^\prime_l} ) \right],
 \end{align*}
-where $\tilde K_{t, t^\prime_l}$ is the normalized kernel such that $\sum_{l=1}^L \tilde K_{t, t^\prime_l} = 1$.  Moreover, we have that
-
-Let~$X_{j,t,l}$ denote the $l$th component of the vector $X_{j,t}$.  Then consider the population-level distribution $\{ X_{j,t,l} \}_{j=1}^N$.  We suppose
+where $\tilde K_{t, t^\prime_l}$ is the normalized kernel such that $\sum_{l=1}^L \tilde K_{t, t^\prime_l} = 1$. The difference in the 1st term is similarly defined.
+
+Let~$X_{j,t,l}$ denote the $l$th component of the vector $X_{j,t}$.  Then consider the population-level distribution $\{ X_{j,t,l} \}_{j=1}^N$.  We make the additional assumption that
 $$
@@
 $$
+Combining this, $E \left[ \Phi(\eta_0)  - \tilde \Phi (\eta_0) \right] \to 0$ which implies consistency as desired.
 
@@
 The asymptotic variance for the IPW estimator is then the first diagonal element of the matrix $E \left[ \phi_n (\eta_0) \right]^{-1} \times \left[ V_{1,t} + V_{2,t} + V_{3,t} + V_{4,t} \right] E \left[ \phi_n (\eta_0) \right]^{-1}$.
-
-% \section{Model-based estimation: parameter choices and inference}
-% \label{app:modelbased_approx}
-
-% Following~\cite{Song2020}, the fourth-order Runge-Kutta (RK4) method gives an approximation to the proposed SEIR model in equation XX as follows:
-
-% $$
-% f(\theta_{t-1}, \beta, \gamma) = \left(
-% \begin{array}{c}
-% \theta_{t-1}^{S} + \frac{1}{6} \left[ k_{t-1}^{S_1} + 2 k_{t-1}^{S_2} + 2 k_{t-1}^{S_3} + k_{t-1}^{S_4} \right] \\
-% \theta_{t-1}^{E} + \frac{1}{6} \left[ k_{t-1}^{E_1} + 2 k_{t-1}^{E_2} + 2 k_{t-1}^{E_3} + k_{t-1}^{E_4} \right] \\
-% \theta_{t-1}^{I^{(T)}} + \frac{1}{6} \left[ k_{t-1}^{I_1^{(T)}} + 2 k_{t-1}^{I_2^{(T)}} + 2 k_{t-1}^{I_3^{(T)}} + k_{t-1}^{I_4^{(T)}} \right] \\
-% \theta_{t-1}^{I^{(NT)}} + \frac{1}{6} \left[ k_{t-1}^{I_1^{(NT)}} + 2 k_{t-1}^{I_2^{(NT)}} + 2 k_{t-1}^{I_3^{(NT)}} + k_{t-1}^{I_4^{(NT)}} \right] \\
-% \theta_{t-1}^{R} + \frac{1}{6} \left[ k_{t-1}^{R_1} + 2 k_{t-1}^{R_2} + 2 k_{t-1}^{R_3} + k_{t-1}^{R_4} \right] \\
-% \end{array}
-% \right)
-% $$
-% where
-% \begin{align*}
-% k_t^{S_1} &= - s_t \circ (B i_t^{(T)}) \\
-% k_t^{S_2} &= - (s_t + 0.5 k_t^{S_1}) \circ (B (i_t^{(T)} + 0.5 k_t^{I_1}) \\
-% k_t^{S_3} &= - (s_t + 0.5 k_t^{S_2}) \circ (B i_t^{(T)} + 0.5 k_t^{I_2}) \\
-% k_t^{S_4} &= - (s_t + k_t^{S_3}) \circ (B i_t^{(T)} + k_t^{I_3}),
-% \end{align*}
-% and
-% \begin{align*}
-% k_t^{E_1} &= s_t \circ (B i_t^{(T)}) - \sigma e_t \\
-% k_t^{E_2} &= (s_t + 0.5 k_t^{S_1}) \circ (B (i_t^{(T)} + 0.5 k_t^{I_1}) - \sigma (e_t + 0.5 k_t^{e_1}) \\
-% k_t^{E_3} &= (s_t + 0.5 k_t^{S_2}) \circ (B (i_t^{(T)} + 0.5 k_t^{I_2}) - \sigma (e_t + 0.5 k_t^{e_2}) \\
-% k_t^{E_4} &= (s_t + k_t^{S_3}) \circ (B (i_t^{(T)} + k_t^{I_4}) - \sigma (e_t + k_t^{E_3}) \\
-% \end{align*}
-% and
-% \begin{align*}
-% k_t^{I_1^{(T)}} &= \sigma e_t - \beta_2 i^{(T)} \\
-% k_t^{I_2^{(T)}} &= \sigma (e_t + 0.5 k_t^{E_1}) - \beta_2 (i^{(T)} + 0.5 k_t^{I_1^{(T)}}) \\
-% k_t^{I_3^{(T)}} &= \sigma (e_t + 0.5 k_t^{E_2}) - \beta_2 (i^{(T)} + 0.5 k_t^{I_2^{(T)}}) \\
-% k_t^{I_4^{(T)}} &= \sigma (e_t + k_t^{E_3}) - \beta_2 (i^{(T)} + k_t^{I_3^{(T)}}),
-% \end{align*}
-% and
-% \begin{align*}
-% k_t^{I_1^{(NT)}} &= \beta_2 i^{(T)} - \gamma i^{(NT)} \\
-% k_t^{I_2^{(NT)}} &= \beta_2 (i^{(T)} + 0.5 k_t^{I_1^{(T)}}) - \gamma (i^{(NT)} + 0.5 k_t^{I_1^{(NT)}})\\
-% k_t^{I_3^{(NT)}} &= \beta_2 (i^{(T)} + 0.5 k_t^{I_2^{(T)}}) - \gamma (i^{(NT)} + 0.5 k_t^{I_2^{(NT)}})\\
-% k_t^{I_4^{(NT)}} &= \beta_2 (i^{(T)} + k_t^{I_3^{(T)}}) - \gamma (i^{(NT)} + k_t^{I_3^{(NT)}}) ,
-% \end{align*}
-% and
-% \begin{align*}
-% k_t^{R_1} &= - \gamma i_t^{(NT)} \\
-% k_t^{R_2} &= - \gamma (i^{(NT)} + 0.5 k_t^{I_1^{(NT)}}) \\
-% k_t^{R_3} &= - \gamma (i^{(NT)} + 0.5 k_t^{I_2^{(NT)}}) \\
-% k_t^{R_4} &= - \gamma (i^{(NT)} + k_t^{I_3^{(NT)}}) \\
-% \end{align*}
-
 \section{IRLS}
@@
 $$
-
-\section{Infection fatality rate (IFR) by age group}
-\label{app:ifr}
 
@@
 \label{app:in_add_details}
+
+In this section, we describe the two mean-imputation methods.
 \subsection{Model 1}
 
-In this section, we describe.
+First, we use the pseudo-likelihood to compute the likelihood of contact given age, gender, and test result.
+Figure~\ref{fig:contactlik1} and~\ref{fig:contactlik2} plots the likelihood for 25-34 year olds given a negative test and positive test respectfully.
+We see that the likelihood of COVID-19 is much higher for those with a positive test and the likelihood is time-varying.
 
@@
  \includegraphics[width=.9\linewidth]{../figs/tvprop_contact_fig1.png}
- \caption{COVID-19 Contact Likelihood Given Negative Test}
+ \caption{Contact Likelihood Given Negative Test}
  \label{fig:contactlik1}
@@
 \includegraphics[width=.9\linewidth]{../figs/tvprop_contact_fig2.png}
- \caption{COVID-19 Contact Likelihood Given Positive Test}
+ \caption{Contact Likelihood Given Positive Test}
  \label{fig:contactlik2}
@@
 
-We then fit a model for likelihood of fever given
+We then use the pseudo-likelihood to compute the likelihood of fever given age, gender, COVID-19 contact status and test result.
+Figure~\ref{fig:symptomlik1} and~\ref{fig:symptomlik2} plots the likelihood for 35-44 year olds given a negative test and positive test respectfully.
+We see that the likelihood of fever depends heavily on whether they had a COVID-19 contact and their test results.
+Again, the likelihood is time-varying for most configurations.
 
@@
  \includegraphics[width=.9\linewidth]{../figs/tvprop_symptom_fig1.png}
- \caption{Symptom Likelihood Given Negative Test \textcolor{red}{FIX!!}}
+ \caption{Fever Likelihood Given Negative Test \textcolor{red}{FIX!!}}
  \label{fig:symptomlik1}
@@
 \includegraphics[width=.9\linewidth]{../figs/tvprop_symptom_fig2.png}
- \caption{Symptom Likelihood Given Positive Test}
+ \caption{Fever Likelihood Given Positive Test}
  \label{fig:symptomlik2}
 \end{subfigure}
-\caption{Likelihood of COVID-19 contact}
+\caption{Likelihood of reported fever}
 \label{fig:symptomlik}
@@
 
-The testing propensity likelihood
+Based on mean-imputation of COVID-19 contact and fever indicators, we can use the proposed pseudo-likelihood approach to compute the likelihood of getting tested for COVID-19 given age, ethnicity, race, gender, fever status, and COVID-19 contact indicator.
+Figure~\ref{fig:testinglik1} and~\ref{fig:testinglik2} plots the likelihood of testing given fever and COVID-19 contact and no fever nor COVID-19 contact respectfully.
+We see that the likelihood is time-varying and depends heavily on both fever and COVID-19 contact indicators.
 
@@
  \includegraphics[width=.9\linewidth]{../figs/tvprop_fig1.png}
- \caption{COVID-10 Testing Likelihood Given Fever and COVID-19 Contact}
+ \caption{Testing Likelihood Given Fever and COVID-19 Contact}
  \label{fig:testinglik1}
@@
 \end{subfigure}
-\caption{COVID-10 Testing Likelihood Given no Fever nor COVID-19 Contact}
+\caption{Testing Likelihood Given no Fever nor COVID-19 Contact}
 \label{fig:testinglik}
@@
 
-We then fit a model for likelihood of fever given
+This model uses hospitalization records to try and improve our mean-imputation strategy.
+First, we use the pseudo-likelihood to compute the likelihood of fever given age, gender, and test result.
+The model only depends on hospitalization status for positive tests.
+Figure~\ref{fig:symptomlik1_model2} and~\ref{fig:symptomlik2_model2} plots the likelihood for 35-44 year olds given a negative test and positive test respectfully.
+We see that the likelihood of fever is time-varying.
+For a negative test, the likelihood of ever fever is very high in early April, which accounts for testing restrictions.
+We see that hospitalization significantly increases the risk of fever given a positive test.
 
@@
 
-The testing propensity likelihood
-
-\begin{figure}[!th]
-\centering
-\begin{subfigure}{.5\textwidth}
- \centering
- \includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig1.png}
- \caption{COVID-10 Testing Likelihood Given Fever and COVID-19 Contact}
- \label{fig:testinglik1}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
- \centering
-\includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig2.png}
- \caption{Testing Propensity}
- \label{fig:testinglik2}
-\end{subfigure}
-\caption{COVID-10 Testing Likelihood Given no Fever nor COVID-19 Contact}
-\label{fig:testinglik}
-\end{figure}
+First, we computed the probability of fever given COVID-19 positive test and hospitalization as well as COVID-19 positive test and no hospitalization.
+We then computed the lieklihood of symptom given a COVID-19 positive test by weighting these two likelihoods by the fraction of COVID-19 positive hospitalizations to COVID-19 total positive cases per week.
+Mean imputation of fever status for COVID-19 negative tests was performed based on the above model without use of hospitalization data.
+
+Based on mean-imputation of COVID-19 fever using hospitalization records indicators, we can use the proposed pseudo-likelihood approach to compute the likelihood of getting tested for COVID-19 given age, ethnicity, race, gender, and fever status.
+Figure~\ref{fig:testinglik1_mainpaper} and~\ref{fig:testinglik1_mainpaper} in Section~\ref{section:tvipw} of the manuscript plots the likelihood of testing given fever and no fever respectfully for all age ranges.
+We see that the likelihood is time-varying and depends heavily on both fever status.
+
+% \begin{figure}[!th]
+% \centering
+% \begin{subfigure}{.5\textwidth}
+%  \centering
+%  \includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig1.png}
+%  \caption{Testing Likelihood Given Fever}
+%  \label{fig:tvprop_alt_fig1.png}
+% \end{subfigure}%
+% \begin{subfigure}{.5\textwidth}
+%  \centering
+% \includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig2.png}
+%  \caption{Testing Propensity Given No Fever}
+%  \label{fig:tvprop_alt_fig2.png}
+% \end{subfigure}
+% \caption{COVID-10 Testing Likelihood Given no Fever nor COVID-19 Contact}
+% \label{fig:testinglik}
+% \end{figure}
 
@@
 \end{itemize}
+where \code{beta, eta, eta\_two, eta\_three} refer to the four values that form the time-varying parameter $\beta_t$ when combined with \code{xi, eta, eta\_two, eta\_three} in the SEIR model as described in Section~\ref{section:modelbased}.
 

### 2021-08-18 write-up/covid-writeup.tex
updating prev figures

@@
 \textcolor{red}{
-While the model-based estimates appear more reasonable, note that the estimate on the last week of April is XX\% which is a severe underestimate.
+While the model-based estimates appear more reasonable, note that the estimate on the last week of April is 0.25\% which is a severe underestimate.
 Caution is therefore warranted when interpreting these estimates.
@@
 Moreover, the (thankfully) low number of deaths per strata make uncertainty in these estimates quite high.
-The 95\% credible range for active infections on April 25th, for example, is $[XX,XX]$ with a posterior mean of $XX$.
 Finally, the doubly robust estimates appear similar to the IPW estimate.

### 2021-08-18 visualizations/tv_prevalence.R

@@ -37,6 +37,19 @@ prevalence_long$estimate = prevalence_long$prev * (prevalence_long$Method == "Un
 
 prevalence_long = data.frame(date = prevalence_long$date, Method = prevalence_long$Method, estimate = prevalence_long$estimate)
 
+for(method in c("Unweighted", "IPW1", "IPW2")) {
+  prevalence_temp = subset(prevalence_long, Method == method)
+  ratio = prevalence_temp$estimate[2:length(prevalence_temp$estimate)]/prevalence_temp$estimate[1:(length(prevalence_temp$estimate)-1)]
+  date = prevalence_temp$date[2:length(prevalence_temp$estimate)]
+  x = as.numeric(prevalence_temp$date - min(prevalence_temp$date))
+  smoothed_estimate = (predict(loess(prevalence_temp$estimate ~ x)))
+  ratio = smoothed_estimate[2:length(smoothed_estimate)]/smoothed_estimate[1:(length(smoothed_estimate)-1)]
+  ratio_temp = data.frame(date = date, Method = rep(method, length(date)), 
+                               estimate = ratio)
+}
+
+plot(prevalence_temp$date, prevalence_temp$estimate)
+lines(prevalence_temp$date, smoothed_estimate)
 
 prevalence_temp = readRDS("../data/aggregate_air.RDS")
 prevalence_temp = data.frame(date = prevalence_temp$date, Method = rep("Model-based", nrow(prevalence_temp)), 
@@ -64,3 +77,7 @@ ggplot(data = prevalence_long, aes(x = date, y = estimate, col = Method)) +
   scale_color_manual(values=wes_palette(n = 5, "IsleofDogs1"))
 
 dev.off()
+
+
+
+

### 2021-08-18 write-up/covid-writeup.tex
adding variance and then done

@@
 
-\textcolor{red}{
-While the model-based estimates appear more reasonable, note that the estimate on the last week of April is 0.25\% which is a severe underestimate.
+While the model-based estimates appear more reasonable, note that the estimate on the last week of April is 0.25\% which is a severe underestimate compared to the estimated prevalence using the stratified random sample.
 Caution is therefore warranted when interpreting these estimates.
@@
 Finally, the doubly robust estimates appear similar to the IPW estimate.
-}
-
-\begin{figure}[!th]
-\centering
-\includegraphics[width=.6\linewidth]{../figs/tv_air.png}
+
+\begin{figure}
+\centering
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{../figs/tv_air.png}
   \caption{Active Infection Rate Estimates}
   \label{fig:air}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+ \centering
+ \includegraphics[width=.9\linewidth]{../figs/tv_air_ratio.png}
+ \caption{Ratio estimates}
+ \label{fig:air_ratio}
+\end{subfigure}
+\caption{Time-varying estimates based on unweighted, model-based, IPW and doubly robust methods.}
+\label{fig:tvestimates}
 \end{figure}
 
-\textcolor{red}{
-We also construct ratio and effective reproduction number estimates as discussed in Section XX.
-Figure XX presents the results.
+We also construct ratio estimates as discussed in Section~\ref{section:rates}.
+Figure~\ref{fig:air_ratio} presents the results.
 Unweighted, IPW, and doubly robust estimates appear similar.
-Recall the severe bias during the growth phase due to selection bias dicussed in SEction XX may imply the observed rates are severely overestimated between October 2020 and January 2021.
-The model-based methods may be conservative in this time window.
-A similar argument holds for the effective reproduction rate in Figure XXb.
-}
+Recall the severe bias during the growth phase due to selection bias dicussed in Section~\ref{section:rates} may imply the observed rates are  overestimated between October 2020 and January 2021.
+The model-based methods may be as well.
 

### 2021-08-18 methods/6_invweight_var_estimation.R

@@ -1,0 +1,54 @@
+indiana_data = readRDS("../data/weeklycoviddata_withsymptoms_alt.RDS")
+indiana_data$startdate = mdy(indiana_data$startdate)
+propensities = readRDS("../data/smoothedpropensities_alt.RDS")
+
+propensities$date = MMWRweek::MMWRweek2Date(MMWRyear = propensities$year,
+                                            MMWRweek = propensities$week,
+                                            MMWRday = 1)
+## First term can be precomputed given design matrix
+model = ~ -1+as.factor(symptoms) + as.factor(gender) + ethnicity + race + as.factor(age)
+indiana_X = model.matrix(model, indiana_data)
+
+library(lubridate)
+indiana_data$startdate = ymd(indiana_data$startdate)
+indiana_data$week = week(indiana_data$startdate)
+indiana_data$year = year(indiana_data$startdate)
+
+invweight = vector(length = nrow(indiana_data))
+for(i in 1:nrow(indiana_data)) {
+  modelmatrix_row = indiana_X[i,]
+  week_row = as.numeric(week(indiana_data$startdate[i]))
+  year_row = as.numeric(year(indiana_data$startdate[i]))
+  if(week_row > 5 & year_row == 2021) {
+    invweight[i] = NA
+  } else {
+    prop_row = which(propensities$week == week_row & propensities$year == year_row)
+    term = as.matrix(propensities[prop_row,3:15])%*%modelmatrix_row
+    invweight[i] = (1+exp(-term))
+  }
+}
+
+indiana_data$invweight = invweight
+
+current_week = 48
+current_year = 2020
+
+weeks = c(14:53,1:5)
+years = c(rep(2020, length = length(c(14:53))),rep(2021, length = length(1:5)))
+results = matrix(nrow = length(weeks), ncol = 4) 
+
+for(i in 1:length(weeks)) {
+  current_week = weeks[i]
+  current_year = years[i]
+  print(paste("At week", current_week, "in year", current_year))
+  current_counts = indiana_data$covid_counts[which(indiana_data$week == current_week & indiana_data$year == current_year)]
+  current_tests = indiana_data$covid_tests[which(indiana_data$week == current_week & indiana_data$year == current_year)]
+  weights = indiana_data$invweight[which(indiana_data$week == current_week & indiana_data$year == current_year)]
+  results[i,1] = current_week
+  results[i,2] = current_year
[…]

### 2021-08-18 visualizations/tv_prevalence.R

@@ -37,6 +37,8 @@ prevalence_long$estimate = prevalence_long$prev * (prevalence_long$Method == "Un
 
 prevalence_long = data.frame(date = prevalence_long$date, Method = prevalence_long$Method, estimate = prevalence_long$estimate)
 
+ratio_long = data.frame(date = rep(0,0), Method = rep(0,0), estimate = rep(0,0))
+
 for(method in c("Unweighted", "IPW1", "IPW2")) {
   prevalence_temp = subset(prevalence_long, Method == method)
   ratio = prevalence_temp$estimate[2:length(prevalence_temp$estimate)]/prevalence_temp$estimate[1:(length(prevalence_temp$estimate)-1)]
@@ -46,26 +48,40 @@ for(method in c("Unweighted", "IPW1", "IPW2")) {
   ratio = smoothed_estimate[2:length(smoothed_estimate)]/smoothed_estimate[1:(length(smoothed_estimate)-1)]
   ratio_temp = data.frame(date = date, Method = rep(method, length(date)), 
                                estimate = ratio)
+  ratio_long = rbind(ratio_long, ratio_temp)
 }
 
-plot(prevalence_temp$date, prevalence_temp$estimate)
-lines(prevalence_temp$date, smoothed_estimate)
 
+## ADD IN MODEL-BASED
 prevalence_temp = readRDS("../data/aggregate_air.RDS")
 prevalence_temp = data.frame(date = prevalence_temp$date, Method = rep("Model-based", nrow(prevalence_temp)), 
                              estimate = prevalence_temp$estimate)
-
+ratio = prevalence_temp$estimate[2:length(prevalence_temp$estimate)]/prevalence_temp$estimate[1:(length(prevalence_temp$estimate)-1)]
+date = prevalence_temp$date[2:length(prevalence_temp$estimate)]
+x = as.numeric(prevalence_temp$date - min(prevalence_temp$date))
+smoothed_estimate = (predict(loess(prevalence_temp$estimate ~ x)))
+ratio = smoothed_estimate[2:length(smoothed_estimate)]/smoothed_estimate[1:(length(smoothed_estimate)-1)]
+ratio_temp = data.frame(date = date, Method = rep("Model-based", length(date)), 
+                        estimate = ratio)
+ratio_long = rbind(ratio_long, ratio_temp)
 prevalence_long = rbind(prevalence_long, prevalence_temp)
 
[…]

### 2021-08-20 write-up/covid-writeup.tex
Fixing the figures in paper

@@
 \centering
-\includegraphics[width = 0.4\textwidth]{../figs/mem_heatmap_article.png}
+\includegraphics[width = 0.7\textwidth]{../figs/mem_heatmap.png}
 \caption{Imperfect testing adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.024$ and $FN=0.13$. Color scaled so blue = $-10$, white = $0$, and red = $10$.}
@@
 \caption{Potential bias due to self-selection and measurement error from a deterministic SIR model with $\beta = 1.2$ (black) and $\gamma = 0.15$.
-Here we assume $f = 0.02$, $FP = 0.024$, $FN = 0.30$, and use a range of relative sampling fraction $M = f_1/f_0$.}
+Here we assume $f = 0.02$, $FP = 0.024$, $FN = 0.30$, and use a range of relative sampling fraction $M = f_1/f_0$.
+\textcolor{red}{FIX!}}
 \label{fig:rates}

### 2021-08-20 visualizations/mem_heatmap.R

@@ -30,13 +30,17 @@ for(i in 1:length(seq_oddsratio)) {
 }
 
 result = data.frame(result)
-row.names(result) = round(seq_oddsratio,3)
-colnames(result) = round(seq_M,2)
+row.names(result) = round(seq_oddsratio,4)
+colnames(result) = round(seq_M,4)
 
-library("gplots")
 png(filename = "../figs/mem_heatmap.png",
-    width = 960, height = 960, units = "px", pointsize = 16)
-heatmap.2(as.matrix(result), col = bluered(100), dendrogram='none',
-          trace = "none", density.info = "none", Rowv=FALSE,
-          Colv=FALSE, key.xlab = "Relative Sampling Frequency", key.ylab = "Odds Ratio")
+    width = 960, height = 640, units = "px", pointsize = 16)
+par(mar = c(4,4,1,1)+0.1)
+contour( z = as.matrix(result),
+         y = log(seq_oddsratio),
+         x = seq_M, nlevels = 30,
+         xlab = "Relative Sampling Frequency (M)",
+         ylab = "Log Odds Ratio", bty = "n",
+         labcex = 0.9)
+axis(side =1, cex = 1.5); axis(side = 2, cex = 1.5)
 dev.off()

### 2021-08-20 visualizations/sir_example_plot.R

@@ -39,13 +39,13 @@ out2$time <- NULL
 ## Show data
 png(filename = "../figs/sir.png",
     width = 6.5, height = 4, units = "in", res = 1200, pointsize = 12)
-par(mar = c(2,2,1,1)+0.1)
+par(mar = c(3,3,1,1)+0.1)
 plot(times, out1$I, type = "l", axes = FALSE, bty = "n")
-lines(times, out2$I, col = "red")
+lines(times, out2$I, col = "grey50")
 axis(side = 1, labels = FALSE)
 axis(side = 2, labels = FALSE)
-mtext("Fraction Infected", side = 2, line = 1, cex = 0.75)
-mtext("Time", side = 1, line = 1, cex = 0.75)
+mtext("Fraction Infected", side = 2, line = 1, cex = 1.5)
+mtext("Time", side = 1, line = 1, cex = 1.5)
 dev.off()
 
 ## Setup for bias plots
@@ -96,16 +96,16 @@ ymax = max(bias1_new$Rt+bias1_new$logbias, bias1_new$Rt, bias2_new$Rt+bias2_new$
 ymin = min(bias1_new$Rt+bias1_new$logbias, bias1_new$Rt, bias2_new$Rt+bias2_new$logbias, bias2_new$Rt, na.rm = TRUE)
 png(filename = "../figs/sir_rt_comparison.png",
     width = 6.5, height = 4, units = "in", res = 1200, pointsize = 12)
-par(mar = c(2,3,1,1)+0.1)
+par(mar = c(3,4,1,1)+0.1)
 plot(bias1_new$Rt, type = "l", ylim = c(0.7, 1.3), axes = FALSE, bty = "n", xlab = "", ylab = "")
 lines(bias1_new$Rt + bias1_new$logbias, lty = 2)
 abline(h = 1.0, col = "grey", lty = 2)
-lines(bias2_new$Rt, lty = 1, col = "red")
-lines(bias2_new$Rt+bias2_new$logbias, lty = 2, col = "red")
+lines(bias2_new$Rt, lty = 1, col = "grey50")
+lines(bias2_new$Rt+bias2_new$logbias, lty = 2, col = "grey50")
 axis(side =1, labels = FALSE)
 axis(side =2, cex.axis = 0.75, at = seq(0.8,1.2,0.1))
-mtext(expression(R[t]), side =2, line = 2, cex = 0.75)
[…]

### 2021-08-22 write-up/covid-writeup.tex
Update covid-writeup.tex
Fixed all visualizations to grayscale

@@
 The sign and magnitude of the statistical decomposition can change drastically; the impact on the effective sample size can therefore be quite large.
+\textcolor{red}{Acknowledge selection bias and measurement error.}
 
@@
 %
+
+\begin{rmk}[Reporting Delays]
+
+\end{rmk}
 
@@
 \subsection{Application of SEIR model}
-\label{section:modelbased}
+\label{section:application_of_modelbased}
 
@@
 In this paper, similar to~\cite{Johndrow2020}, we specify fixed per-strata infection fatality rates.
-Based on published age-specific IFRs~\cite{Levin2020}, Figure~\ref{fig:logifr} shows that IFR seems to closely follow a log-linear relationship with age.
+Based on published age-specific IFRs~\cite{Levin2020}, IFR closely follows a log-linear relationship with age.
 Combining across published age-specific IFRs, Table~\ref{tab:ifrperage} presents estimates of the average IFR within each strata using the age-distribution.
@@
 Each weight~$w_t^{(k)} = 0$ for $t < t_k$ and $w_t^{(k)} \to 1$ as $t \to \infty$.
-See Section~\ref{app:modelbased} in the supplementary materials for details on prior specification.
+See Section~\ref{app:prior_modelbased} in the supplementary materials for details on prior specification.
 
@@
 
+\newpage
+\textcolor{red}{Fix references!}
 \bibliographystyle{plain}
@@
 For binary outcomes, we have
-\begin{equation} \label{eq:binaryrho}
+\begin{equation} \label{eq:weightedbinaryrho}
 \rho_{\tilde I (X),Y} = \tilde \Delta \sqrt{\frac{\bar Y (1 - \bar Y)}{f (1-f) \text{E}(W_J \mid I_J = 1)^2 + f \text{Var}(W_J \mid I_J = 1)} }
@@
 \caption{Likelihood of COVID-19 contact}
-\label{fig:symptomlik}
+\label{fig:symptomlik_model2}
 \end{figure}
@@
 \section{Model-based: Prior specification}
-\label{app:modelbased}
+\label{app:prior_modelbased}
 

### 2021-08-22 visualizations/indiana_plots.R

@@ -33,7 +33,7 @@ df_case_counts %>%
   ggplot() + 
   geom_bar(mapping = aes(x = date, y = covid_count, fill = Age), stat = "identity") +
   labs(y="COVID-19 Reported Cases", x = "Date") + 
-  scale_x_date(date_breaks = "months" , date_labels = "%b-%y") +
+  scale_x_date(date_breaks = "3 month" , date_labels = "%b-%y") +
   # geom_vline(aes(xintercept = date(date_switch)), size = 1.5, linetype="dotted") + 
   # geom_vline(aes(xintercept = date(date_switch_two)), size = 1.5, linetype="dotted") + 
   # geom_vline(aes(xintercept = date(date_switch_three)), size = 1.5, linetype="dotted") +
@@ -49,7 +49,7 @@ df_case_counts %>%
   ggplot() + 
   geom_bar(mapping = aes(x = date, y = covid_test, fill = Age), stat = "identity") +
   labs(y="COVID-19 Reported Tests", x = "Date") + 
-  scale_x_date(date_breaks = "months" , date_labels = "%b-%y") +
+  scale_x_date(date_breaks = "3 months" , date_labels = "%b-%y") +
   # geom_vline(aes(xintercept = date(date_switch)), size = 1.5, linetype="dotted") + 
   # geom_vline(aes(xintercept = date(date_switch_two)), size = 1.5, linetype="dotted") + 
   # geom_vline(aes(xintercept = date(date_switch_three)), size = 1.5, linetype="dotted") +
@@ -65,7 +65,7 @@ df_case_counts %>%
   ggplot() + 
   geom_bar(mapping = aes(x = date, y = covid_deaths, fill = Age), stat = "identity") +
   labs(y="COVID-19 Reported Tests", x = "Date") + 
-  scale_x_date(date_breaks = "months" , date_labels = "%b-%y") +
+  scale_x_date(date_breaks = "3 months" , date_labels = "%b-%y") +
   # geom_vline(aes(xintercept = date(date_switch)), size = 1.5, linetype="dotted") + 
   # geom_vline(aes(xintercept = date(date_switch_two)), size = 1.5, linetype="dotted") + 
   # geom_vline(aes(xintercept = date(date_switch_three)), size = 1.5, linetype="dotted") +

### 2021-08-22 visualizations/model_based.R

@@ -1,5 +1,5 @@
 ## Libraries and Colors
-library(wesanderson)
+library("RColorBrewer")
 library(tidyverse)
 library(tidybayes)
 library(gridExtra)
@@ -11,6 +11,7 @@ c_mid <- c("#fc9272")
 c_dark <- c("#de2d26")
 c_posterior = "orange"
 c_prior = "aquamarine2"
+my_palette <- brewer.pal(name="Greys",n=9)[7]
 
 ## Death model
 df_coviddeath <- readRDS("../data/dailycoviddata.RDS")
@@ -51,7 +52,7 @@ smr_pred <- cbind(agg_deaths,
 colnames(smr_pred) <- make.names(colnames(smr_pred)) # to remove % in the col names
 smr_pred$date = date(dates)
 smr_pred = smr_pred[smr_pred$date < "2021-02-01",]
-c_posterior = "orange"
+c_posterior = my_palette # "orange"
 
 
 ggplot(smr_pred, mapping = aes(x = date)) +

### 2021-08-22 visualizations/tv_contact_propensities.R

@@ -1,3 +1,4 @@
+library("RColorBrewer")
 library(lubridate)
 library(MMWRweek)
 library(ggplot2)
@@ -8,6 +9,8 @@ propensities_neg$date = MMWRweek::MMWRweek2Date(MMWRyear = propensities_neg$year
                                                 MMWRday = 1)
 names(propensities_neg) = c("week", "year", "gender1", "gender2", "25to34", "35to44", "45to54",
                             "55to64", "65to74", "75plus", "date")
+
+my_palette <- brewer.pal(name="Greys",n=9)[c(5,9)]
 
 ## Example propensity: Gender + 25-34
 xfemale = as.matrix(c(0,1,1,0,0,0,0,0), ncol = 1)
@@ -29,6 +32,7 @@ ggplot(data = propensities_neg_long, aes(x = date, y = probs, col = Gender)) +
   labs(x = "Date",
        y = "Likelihood of Contact",
        title = "Individuals 25-34 tested negative in the past 24 hours") +
+  scale_color_manual(values = my_palette) +
   theme(text = element_text(size=25))
 
 dev.off()
@@ -61,6 +65,7 @@ ggplot(data = propensities_pos_long, aes(x = date, y = probs, col = Gender)) +
   labs(x = "Date",
        y = "Likelihood of Contact",
        title = "Individuals 25-34 tested positive in the past 24 hours") +
-  theme(text = element_text(size=25))
+  theme(text = element_text(size=25)) + 
+  scale_color_manual(values = my_palette)
 
 dev.off()

### 2021-08-22 visualizations/tv_propensities_alt.R

@@ -1,13 +1,15 @@
 library(lubridate)
 library(MMWRweek)
 library(ggplot2)
-library(wesanderson)
+library("RColorBrewer")
 
 propensities = readRDS("../data/smoothedpropensities_alt.RDS")
 
 propensities$date = MMWRweek::MMWRweek2Date(MMWRyear = propensities$year,
                                             MMWRweek = propensities$week,
                                             MMWRday = 1)
+
+my_palette <- brewer.pal(name="Greys",n=9)[3:(3+5)]
 
 ## Example propensity: Fever + Not H + White + Male + 35to44
 x = as.matrix(c(0,1,0,1,0,0,1,0,1,0,0,0,0), ncol = 1)
@@ -68,7 +70,7 @@ ggplot(data = propensities_long, aes(x = date, y = propensity, col = age)) +
        y = "COVID-19 Testing Propensity",
        title = "Non-Hispanic, White Male with Fever") +
   theme(text = element_text(size=25)) +
-  scale_color_manual(values=wes_palette(n = 6, "IsleofDogs1"))
+  scale_color_manual(values=my_palette)
 
 dev.off()
 
@@ -133,7 +135,7 @@ ggplot(data = propensities_long, aes(x = date, y = propensity, col = age)) +
        y = "COVID-19 Testing Propensity",
        title = "Non-Hispanic, White Male without Fever") +
   theme(text = element_text(size=25)) +
-  scale_color_manual(values=wes_palette(n = 6, "IsleofDogs1"))
+  scale_color_manual(values=my_palette)
 
 dev.off()
 

### 2021-08-22 visualizations/tv_symptom_propensities.R

@@ -1,3 +1,4 @@
+library("RColorBrewer")
 library(lubridate)
 library(MMWRweek)
 library(ggplot2)
@@ -8,6 +9,8 @@ propensities_neg$date = MMWRweek::MMWRweek2Date(MMWRyear = propensities_neg$year
                                                 MMWRday = 1)
 names(propensities_neg) = c("week", "year", "gender1", "gender2", "25to34", "35to44", "45to54",
                             "55to64", "65to74", "75plus", "contact", "date")
+
+my_palette <- brewer.pal(name="Greys",n=9)[c(4,6)]
 
 ## Example propensity: Fever + Not H + White + Male
 xmalenocontact =   as.matrix(c(1,0,0,1,0,0,0,0,1), ncol = 1)
@@ -29,6 +32,7 @@ ggplot(data = propensities_neg_long, aes(x = date, y = probs, col = Contact)) +
   labs(x = "Date",
        y = "Likelihood of Fever",
        title = "Male Individuals 35-44 tested negative in the past 24 hours") + 
+  scale_color_manual(values = my_palette) +
   theme(text = element_text(size=25))
 
 dev.off()
@@ -61,6 +65,7 @@ ggplot(data = propensities_pos_long, aes(x = date, y = probs, col = Contact)) +
   labs(x = "Date",
        y = "Likelihood of Fever",
        title = "Female individuals 35-44 tested positive in the past 24 hours") +
+  scale_color_manual(values = my_palette) +
   theme(text = element_text(size=25))
 
 dev.off()

### 2021-08-22 visualizations/tv_symptom_propensities_alt.R

@@ -1,3 +1,4 @@
+library("RColorBrewer")
 library(lubridate)
 library(MMWRweek)
 library(ggplot2)
@@ -29,6 +30,7 @@ ggplot(data = propensities_neg_long, aes(x = date, y = probs, col = Gender)) +
   labs(x = "Date",
        y = "Likelihood of Fever",
        title = "Individuals 35-44 tested negative in the past 24 hours")+ 
+  scale_color_manual(values = my_palette) +
   theme(text = element_text(size=25))
 
 dev.off()
@@ -61,6 +63,7 @@ ggplot(data = propensities_pos_long, aes(x = date, y = probs, col = Hospitalized
   labs(x = "Date",
        y = "Likelihood of Fever",
        title = "Individuals 35-44 tested positive in the past 24 hours") + 
+  scale_color_manual(values = my_palette) +
   theme(text = element_text(size=25))
 
 dev.off()

### 2021-08-23 write-up/covid-supplement-writeup.tex
Switched to 2 documents

--- 2021-08-18 write-up/covid-writeup.tex
@@
 \usepackage{caption,subcaption}
+
+\usepackage{xr}
+\makeatletter
+\newcommand*{\addFileDependency}[1]{% argument=file name and extension
+  \typeout{(#1)}% latexmk will find this if $recorder=0 (however, in that case, it will ignore #1 if it is a .aux or .pdf file etc and it exists!
+if it doesn't exist, it will appear in the list of dependents regardless)
+  \@addtofilelist{#1}% if you want it to appear in \listfiles, not really necessary and latexmk doesn't use this
+  \IfFileExists{#1}{}{\typeout{No file #1.}}% latexmk will find this message if #1 doesn't exist (yet)
+}
+\makeatother
+
+\newcommand*{\myexternaldocument}[1]{%
+    \externaldocument{#1}%
+    \addFileDependency{#1.tex}%
+    \addFileDependency{#1.aux}%
+}
+%%% END HELPER CODE
+
+% put all the external documents here!
+\myexternaldocument{covid-writeup}
 
@@
 
-\title[Statistical paradoxes in coronavirus case-counts]{Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic} %
+\title[Statistical paradoxes in coronavirus case-counts]{Supplementary Materials to ``Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic''} %
 
@@
 \begin{abstract}
-  Coronavirus case-count data has influenced government policies and drives most epidemiological forecasts.
-Limited testing is cited as the key driver behind minimal information on the COVID-19 pandemic.
-While expanded testing is laudable, measurement error and selection bias are the two greatest problems limiting our understanding of the COVID-19 pandemic; neither can be fully addressed by increased testing capacity.
-In this paper, we demonstrate their impact on estimation of point prevalence and the effective reproduction number.
-We show that estimates based on the millions of molecular tests in the US has the same mean square error as a small simple random sample.
-To address this, a procedure is presented that combines case-count data and random samples over time to estimate selection propensities based on key covariate information.
-We then combine these selection propensities with epidemiological forecast models to construct a \emph{doubly robust} estimation method that accounts for both measurement-error and selection bias.
-This method is then applied to estimate Indiana's prevalence using case-count, hospitalization, and death data with cumulative demographic information, the United States's only statewide random molecular sample collected from April 25--29th, and Facebook's COVID-19 symptom survey.
-We end with a series of recommendations based on the proposed methodology.
+This is the supplementary materials to the manuscript titled ``Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic.''  Section~\ref{app:imperfect} derives the statistical decompositions for the measurement-error corrected estimator and provides the connection to model-based estimators.
+Sections~\ref{app:ipwderivation} and~\ref{app:DRderivation} derives the statistical decomposition for the inverse-probability weighted and doubly robust estimators respectively.
+Section~\ref{app:asympderivations} gives the proof for the asymptotics of the inverse-probability weighted estimator.  
+Section~\ref{app:irls} outlines the iteratively re-weighted least squares algorithm.
+Section~\ref{app:in_add_details} provides additional detals on the COVID-19 analysis of Indiana's case-count data.
+Section~\ref{app:prior_modelbased} provides prior specifications for the SEIR model.
 \end{abstract}
@@
 \maketitle
-
-\newpage
-
-\section{Introduction}
-The World Health Organization has declared the coronavirus disease 2019 (COVID-19) a public health emergency.
-As of July 29th, 2021, over 196 million cases have been confirmed worldwide with 34.8 million cases and over 612 thousand confirmed deaths across the United States.
-This pandemic
-has become the focal point of everyday life; yet the data landscape for understanding COVID-19 remains limited.
-Public databases~\citep{JHU_Lancet,NYT} provide incoming county-level information of confirmed cases and deaths.
-Statisticians, epidemiologists, economists, and data scientists have used this granular data to forecast COVID-19 case-counts, deaths, and hospitalizations~\citep{Giordano2020,Song2020,Ray2020,2020.IHME,Wang2020.03,JTD36385}.
-
-This paper has two main objectives.
-The first objective is to express reservations at the use of case-counts as a proxy for disease prevalence and in estimation of standard epidemiological models for inference and forecasting.
-The reason is straightforward: observed case-count data is plagued by selection bias and measurement error.
-Through a series of calculations, we will demonstrate that the information gained from increasing testing capacity is limited in the presence of selection bias and when testing inaccuracies persist.
-In particular, the millions of tests in the US have a small effective sample size when compared to random sampling.
-These calculations demonstrate the importance of probabilistic sampling designs over time for estimation of point prevalence and effective reproduction number.
-
-Due to monetary and time constraints, however, random testing is performed infrequently.
-As of June 2021, Indiana is the only state to conduct statewide random sample testing\footnote{This is the only random sample to collect both seroprevalence and diagnostic testing results.
-The CDC has conducted several seroprevalence studies.
-See Section~\ref{section:testinginfo} for further discussion.}, which was collected from April 25--29, 2020~\citep{Yiannoutsos2021}.
-Such infrequent random testing is likely to provide insufficient information to help researchers and policy makers better understand the disease trajectory which can change rapidly over time.
-Therefore, while random testing may be preferable in theory, in practice governments, researchers, and policy makers continue to use coronavirus case-counts to understand the impact of COVID-19 on the population and make data-informed decisions.
-
-The second objective is to demonstrate how random samples provide necessary \emph{auxiliary information} to address selection bias in coronavirus case-count data.
-Random samples provide the necessary covariate information from a representative sample from the population to estimate selection propensities.
-These propensities can then used in an inverse-probability weighting scheme to construct estimators of disease prevalence that attempt to control for selection bias.
-A doubly robust extension allows researchers to combine these estimates with epidemiological forecasts based on compartmental models that are common in the study of infectious diseases~\cite{Hao2020,Song2020,Ray2020,Johndrow2020}.
-
-The proposed approach requires covariate information to be collected on individuals who receive a COVID-19 test.
-Unfortunately, many states do not require or report auxiliary covariate information beyond basic demographic information (e.g., gender, age, race, and ethnicity).
-We end with a brief list of suggestions of minor changes to current practice based on the proposed methodology.
-While we demonstrate empirical improvements over simple disease prevalence estimates, we also highlight how selection bias may persist and impact uncertainty quantification.
-
-\subsection{Related work}
-
-This article discusses the relationship between three statistical concepts: selection bias, measurement error, and population size.
-We build on the work of \cite{Meng2018} who studied an error decomposition to understand the relationship between selection bias and population size.
-In particular, we quantify the interaction between measurement-error and selection bias.
-The sign and magnitude of the statistical decomposition can change drastically; the impact on the effective sample size can therefore be quite large.
-
-After demonstrating the limitations of case-count analysis when compared to random sampling, we then assess whether their is potential for combining the \emph{nonprobability samples} with \emph{probability samples} to improve point prevalence estimation.
-For any probability sampling design, the Horvitz-Thompson estimator~\citep{HT1952} incorporates design information via inverse-probability weights (IPW).
-For nonprobability samples, the IPW estimator requires modelling the propensity scores.
-Its use in the survey context is also referred to as the quasi-randomization approach~\citep{Elliott2017}.
-\cite{Valliant2011} consider a weighted logistic regression procedure using the pooled probability and nonprobability samples.
-\cite{Chen2019} consider a pseudo-log-likelihood approach that uses the random samples as a proxy for a term in the log-likelihood.
-Here, we extend this approach to account for measurement-error as well as random sampling at multiple times.
-We then provide an extension of the statistical error decomposition and discuss the trade-offs inherent in such a weighting approach.
-
-One core component of coronavirus research is epidemiological state-space modelling of case-count and death data.
-These models can be used to answer a variety of research questions including case-count forecasting, estimation of the effective reproduction number, and estimation of quarantine and other health policies on infectious disease dynamics.
-The basic approach is a deterministic compartmental model called the susceptible-infectious-recovered (SIR) model.
-A probabilistic extension was proposed by~\cite{Osthus2017} to model one-dimensional time series of infected proportions.
-\cite{Song2020} extended this approach to incorporate interventions and assess interventions on COVID-19 epidemic in China.
-\cite{Hao2020} builds an extended SEIR model (SAPHIRE) to account for various presymptomatic infectiousness, time-varying ascertainment rates, transmission rates and population movements.
-
-Recent work by \cite{Zhao2021} pointed out that estimation of key epidemiological parameters such as the incubation time using standard epidemiological models can suffer from severe bias due to issues beyond selection bias and measurement error.
-Namely, right truncation and epidemic growth leading to patients ``being more likely to be infected towards the end of their exposure period''~\cite[pp.
-3]{Zhao2021}.
-Their approach carefully constructs the study sample and statistical model to account for these issues.
-In this paper, rather than focusing on construction of a study sample, we ask whether one can collect auxiliary information to address selection bias in the observed case count data directly.
-
-Given a probability sampling design, individual predictions can be leveraged to improve estimation via model-assisted approaches~\citep{Breidt2017}.
-For nonprobability samples, \cite{Chen2019} derive a doubly robust approach that uses outcome predictions given covariates on the nonprobability and probability samples.
-Here, we combine the state-space SEIR approach of \cite{Song2020} but instead, as in \cite{Johndrow2020}, focus on COVID-19 confirmed death count data.
-We generate epidemiological forecasts for point prevalence of each population strata.
-We then demonstrate how to combine these forecasts with the IPW approach to construct doubly-robust estimates of point prevalence.
-A derived statistical error decomposition guides this discussion.
-
-% \subsection{Outline}
-
-% In Section~\ref{section:data}, we summarize the publicly available data on COVID-19.
-We highlight how testing information is collected as well as demographic and comorbidity information.
-In Section~\ref{section:testinginfo}, we briefly summarize COVID-19 testing.
-We clarify the difference between viral and antibody testing.
-For viral testing, we briefly discuss the reasons behind measurement-error in RT-PCR tests including test timing and sample collection techniques.
-
-% In Section~\ref{section:casecount}, we present simple mathematical arguments to show that unadjusted prevalence rates are (unsurprisingly) biased when tests are imperfect.
-What is surprising is that the direction and magnitude of the bias can vary substantially and interacts with the sampling fraction.
-We show that the rate of change in COVID-19 observed case-counts overestimates the true rate of change prior to the peak time and underestimates it immediately after.
-This implies a data scientist analyzing observed rates will be (on average) overly pessimistic in the early stages of the pandemic, and overly optimistic subsequent to the peak times.
-We show that estimates of the effective reproduction number may be similarly impacted.
-
-% In Section~\ref{section:improvedcasecount}, we present a doubly robust estimation method for combining epidemiological forecasts with inverse-probability weighting methods to estimate prevalence.
-Weight estimation requires auxiliary information provided by random samples.
-A statistical error decomposition illustrates a trade-off in use of weights: a potential increase in data quality but a reduction in effective data quantity.
-
-% In Section~\ref{section:applications}, the IPW and doubly-robust estimation methods are applied to the case study of Indiana statewide prevalence from March 2020 until June 2021.
-We estimate data quality for IPW estimates using the random sample on April 28th.
-This demonstrates the proposed approach is an improvement over simple prevalence estimation but that data quality is not of the same order as random sampling.
-
-% In Section~\ref{section:discussion}, a series of recommendations are provided that aim to help improve statistical inference in settings with large nonprobability samples from a stochastic process where small random sampling over time is possible.
-
-\section{COVID-19 testing and data}
-\label{section:data}
-
-In this section we provide the necessary background to understand COVID-19 diagnostic and serologic testing, discuss their respective scientific uses in managing the pandemic, and introduce the data streams considered in this manuscript.
-
-\subsection{Diagnostic testing}
-\label{section:testinginfo}
-
-Upon infection with SARS-CoV-2, there is an incubation period which lasts approximately five days~\citep{Lauer2020} before the viral load is high enough to be detected.
-A \emph{molecular test} refers to diagnostic tests that aim to detect increased viral load (e.g., RT-PCR or antigen tests).
-If someone has an active infection then after the incubation period, a molecular test with perfect sensitivity will yield a positive result for the next few weeks.
-After that, the viral load will decrease and a molecular test with perfect specificity will come back negative.
-While an infected individual may yield a positive molecular test for several weeks, they will likely stop transmitting the disease within a few days of infection, meaning a positive molecular test does not imply transmissibility.
-
-A molecular test conducted on an infected individual may return a negative result.
-Such false negatives are very strongly associated with when the test is conducted.
-In the incubation phase, most molecular tests will return a false negative result.
-Molecular tests are most sensitive when the viral loads are highest which occurs during the first few days of transmissibility~\citep{Mina2020}.
-Moreover, most molecular tests are performed via nasopharyngeal swab.
-Specimen collection by swab is known to impact false negative/positive rates regardless of test timing.
-Systematic reviews suggest that 87\% sensitivity and 97.6\% specificity are reasonable estimates for current RT-PCR tests~\cite{Arevalo2020, Woloshin2020,Cohen2020}.
-
-The primary goal of molecular tests is diagnosis of active infections in the population.
-Such diagnostic tests generate important information about the presence of SARS-CoV-2 in the population, and help scientists and policy-makers understand patterns of transmission and propagation.
-Rapid and frequent molecular testing is cited as a key component~\cite{OECD2021} in effective strategies to identify active infections and prevent systemic outbreaks.
-
-\subsection{Publicly available data on COVID-19}
-\label{subsection:testinginfo}
-
-The primary goal of this manuscript is to produce accurate estimates of the population-level active infection rates over time using \emph{publicly available} viral testing data.
-Secondary goals include estimation of rates of change and the effective reproduction number which characterize disease trajectory.
-These quantities are fundamental to public health policy and provide critical information on the presence and transmission of SARS-CoV-2.
-
-% {\bf key point: WHAT IS THE GOAL OF CASE COUNT DATA?
-WHAT IS THE GOAL OF SEROPREVALENCE STUDIES?
-are we using it for that reason.
-cse count data is colelcted as part of surveillance programs wehre teh goal is to identify and isolate active infections.
-However, many people are trying to use these to udnerstand the disease.
-This is our issue.}
-
-Coronavirus case-count data refers to the number of positive molecular tests performed on each day.
-Figure~\ref{fig:in-cases} plots the number of reported confirmed COVID-19 cases per day in the state of Indiana.
-Figure~\ref{fig:in-tests} plots the total number of COVID-19 molecular tests performed per day.
-Figure~\ref{fig:in-deaths} plots the total number of COVID-19 related reported deaths per day.
-Public databases maintained by \href{https://bit.ly/2UqFSuA}{Johns Hopkins University} and the \href{https://bit.ly/2vUHfrK}{New York Times} provide accessible incoming county-level information of confirmed cases and deaths.
-
-Public databases most often, however, only contain aggregate information.
-The Johns Hopkins dashboard, for example, provides demographic breakdown of case counts as well as the total confirmed cases and deaths by county.
-Aggregate time series of case count and deaths can also be extracted.
-Unfortunately, most dashboards do not provide demographic information on who requested a test nor on who tested positive for SARS-CoV-2 \emph{over time}.
-Working closely with the State of Indiana, we were able to obtain access to COVID-19 total tests, positive tests, and related deaths per day broken out marginally by age, gender, ethnicity, and race~\citep{IndianaData2021}.
-These granular datasets are now publicly available and motivate the proposed approach.
-% https://www.regenstrief.org/covid-dashboard/
-%
-
-%https://www.npr.org/sections/goatsandsoda/2020/05/31/865932474/should-i-get-tested-for-coronavirus-just-for-the-heck-of-it
-
-\subsubsection{Testing restrictions and public health policy in Indiana}
-\label{section:publicpolicyindiana}
-
-Due to limited testing capacity, many US states instituted testing restrictions early on in the pandemic.
-% on who can request a molecular test at a given time.
-%https://www.coronavirus.in.gov/
-% Early on, limited testing meant that only those at high risk were allowed to be tested (i.e., symptomatic or high risk populations); as testing capacity expanded, these restrictions were gradually lifted.
-Here, we reconstruct the testing restriction history for the state of Indiana.
-As of May 12th, 2020, the Indiana State Department of Health (ISDH) laboratory was testing high-risk individuals for COVID-19.
-Any symptomatic individual could receive a test.
-Moreover there were key factors for people who can be tested regardless of symptom status: over the age of 65, diabetic, high-blood pressure, obesity, pregnancy, and minority populations at greater risk of severe illness~\citep{indystar2020}.
-% https://www.indystar.com/story/news/health/2020/05/12/coronavirus-testing-indiana-who-should-get-tested/3110592001/
-On April 28th, 2020, the criteria expanded to include any Indiana resident with virus symptoms, people in close contact with those who have tested positive and residents of congregant communities~\citep{wishtv2020}.
-% https://www.wishtv.com/news/medical/indiana-opens-up-covid-19-testing-to-more-hoosiers/
-On June 15, 2020, ISDH lifted all restrictions and testing expanded to include anyone in Indiana who wants to be tested for coronavirus~\citep{indystar2020v2}.
-% https://www.indystar.com/story/news/health/2020/06/12/indiana-says-anyone-who-wants-coronavirus-test-can-get-one/3179151001/
-Testing restrictions impact the propensity of an individual to receive a test and therefore, if COVID-19 positive, contribute to the case count.
-In this paper, testing restrictions are addressed by fitting time-varying testing propensities that depend on relevant covariate information.
-
-On March 23, 2020, Indiana's governor issues a \emph{stay at home} order effective March 26th through April 5th.
-The order was extended until April 30th.
-On May 1st, 2020, a five-stage plan for gradual reopening was announced by Governon Holcomb~\citep{fivestageplan}.
-Such policies target reduction in active infection rates.
-In this paper, Indiana's public health policy is incorporated in our construction of epidemiological forecasts in Section~\ref{section:modelbased}.
-
-\subsection{Probabilistic samples in Indiana}
-
-Due to testing restrictions and other potential selection biases, publicly reported COVID-19 case cound data may not be sufficient to understand active infection rates or the disease trajectory.
-Here, we discuss two random samples that provide auxiliary information and may help address selection bias.
-
-\subsubsection{Random statewide testing}
-
-Between April 25--29, 2020, Indiana conducted statewide random molecular testing of persons ages $\geq 12$ years to assess prevalence of active infection to SARS-CoV-2~\citep{Yiannoutsos2021}.
-A stratified random sampling design was conducted using Indiana’s 10 public health preparedness districts as sampling strata.
-15,495 participants were contacted resulting in a final sample size of 3,658.
-The same demographic data as provided by ISDH for Indian molecular testing data was included (e.g., summary statistics on age, sex, and race).
-Participants in the probability sample were also asked if they experienced any COVID-19 compatible symptoms during the past 2 weeks or had shared a household with someone who had a positive test result for SARS-CoV-2.
-During May 2--3, 2020, an additional nonrandom sample of 898 individuals was also collected.
-% In this paper, the nonrandom sample is used to assess whether the cumulative summary statistics are appropriate in constructing the inverse-probability weights as well as in constructing appropriate weights using symptom information.
-
-\subsubsection{Facebook/CMU symptom survey.}
-\label{subsection:fbsymptom}
-Since April 2020, Facebook has started randomly showing an opt-in symptom survey to its users~\citep{delphisurvey}.
-Data collected includes basic demographic information and if the respondent has symptoms such as fever, coughing, shortness of breath, or loss of smell which are associated with COVID-19.
-The survey defines an individual as displaying \emph{COVID-like symptoms} if they exhibit a fever along with a cough, or shortness of breath, or difficulty breathing.
-Figure~\ref{fig:fbsymptoms} displays smoothed estimates of the fraction of individuals within age and gender strata who report a fever within the past 24-hours.
-This paper uses the Facebook symptom survey as the main source of auxiliary information on time-varying characteristics of the population of Indiana, e.g., fraction of population displaying symptoms or in contact with COVID-19 positive individuals.
-
-\begin{figure}[!th]
-\centering
-\includegraphics[width = 0.9\textwidth]{../figs/fbcovid19symptoms.png}
-\caption{Rate of reported fever per strata.
-Daily rates were estimated using weighted method suggested by~\citep{delphisurvey} on each day separately and then smoothed over time using local-linear nonparametric regression.}
-\label{fig:fbsymptoms}
-\vspace{-0.3cm}
-\end{figure}
-
-%% GOOD THROUGH HERE
-
-\section{Analysis of case-count data}
-\label{section:casecount}
-
-We start with some simple notation.
-Let $N$ denote the population size.
-At a given time, let $Y_j$ denote COVID-19 status for the $j$th individual in the population, $j=1,\ldots, N$.
-Here, like in survey methodology~\citep{Cochran77}, we treat COVID-19 status as a fixed but unknown quantity of interest.
-For simplicity, the dynamic nature of the outbreak and recoverability of individuals are initially ignored.
-We assume either individual $j$ is COVID-19 positive and $Y_j=1$ or is COVID-19 negative and $Y_j=0$.
-We also let $I_j \in \{0,1\}$ be an indicator that the individual was tested ($I = 1$) or not ($I=0$).
-
-To start, we assume the overall number of active COVID-19 cases and/or active infection rate (AIR) are of primary interest.
-That is, we are interested in either the population total $Y = \sum_{j=1}^N Y_j$ or the population average $\bar Y = Y/N$.
-Suppose that $n$ tests are performed and we observe the values $y_1, \ldots, y_n \in \{0,1\}$.
-Then a natural candidate for AIR is the proportion of positive tests $\bar y = \frac{1}{n} \sum_{i=1}^n y_i$, and a natural candidate for overall active cases is $N \times \bar y$.
-Under simple random sampling (SRS) or any other epsem\footnote{equal probability of selection method} design, the above are unbiased estimators of the population-level quantities of interest.
-Under SRS, the variance of the estimator can be expressed as $\frac{1}{N-1} \times \frac{1-f}{f} \times \sigma_Y^2$ where $f = n/N$ is the sampling fraction and $\sigma_Y^2 = \frac{1}{N} \sum_{i=1}^N (Y_i - \bar Y)^2 = \bar Y (1- \bar Y)$.
-
-These random selection mechanisms are independent of the outcome of interest.
-When this is not the case, selection effects may cause bias.
-To better understand this issue, \cite{Meng2018} recently provided the following intuitive and powerful statistical decomposition of the error between $\bar y$ and the true proportion $\bar Y$
-$$
-\bar y_n - \bar Y =  \rho_{I, Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_Y.
-$$
-The first term represents \emph{data quality}, the second \emph{data quantity}, and the third \emph{problem difficulty}.
-The term $\rho_{I,Y}$ is the empirical correlation between the population values~$\{ Y_j \}_{j=1}^N$ and the selection values $\{ I_j \}_{j=1}^N$.
-Under simple random sampling, $E_{\I} [ \rho_{I,Y} ] = 0$, where the expectation is with respect to the selection mechanism~$\I$, so there is no bias.
-The SRS variance formula above shows that $E_{\I} [ \rho_{I,Y}^2 ]  = 1/(N-1)$.
-
-% The key issue with selective testing is that $E_{\I} [ \rho_{I,Y} ] \neq 0$.
-Meng identified this as the fundamental issue that can lead to paradoxes in the analysis of big data.
-The remainder of this paper aims to build upon this fundamental work by extending the decomposition in two directions: accounting for measurement error and the temporal nature of the pandemic.
-We then discuss effective sampling methods and their importance in the current pandemic.
-
-\subsection{Imperfect testing}
-\label{section:imperfecttesting}
-
-Tests are imperfect.
-As discussed in Section~\ref{section:testinginfo}, COVID-19 testing is no exception.
-Here we investigate the interplay between imperfect testing and selection bias.
-Researchers most often assume measurement error will lead to parameter attenuation.
-When paired with selection bias, however, the two sources of error become entangled, and resulting errors can become magnified, muted, or even switch signs.
-
-First we require some additional notation.
-Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the outcome and $0$ when we observe the true outcome.
-We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.
-If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^{\star} = Y_j(1-P_j) + (1-Y_j) P_j$.
-The attentive data analyst will recognize the estimator $\bar y_n$ is now biased for simple random samples.
-In Appendix~\ref{app:memestimator}, assuming sensitivity and specificity are known a priori, a novel iterative procedure is used to construct the estimator $\tilde y_n = (\bar y_n - FP)/(1-(FP+FN))$, which is unbiased under simple random sampling (SRS); see Appendix~\ref{app:modelbased} for a discussion of the connection to model-based estimators.
-To understand the impact of selection bias and imperfect testing, we derive the following statistical decomposition of the error between $\tilde y_n$ and $\bar Y$:
-\begin{equation}
-\label{eq:statdecomp}
-\rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
-\times \underbrace{\left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{D_M},
-\end{equation}
-where $f=n/N$ is the sampling fraction, $f_1$ and $f_0$ are sampling fractions for COVID-19 positive and negative individuals respectively, and $\Delta = f_1 - f_0$ is the sampling rate differential.
-See Appendix~\ref{app:memestimator} for the derivation.
-This extends work by \cite{Meng2018} to account for imperfect testing.
-The first three terms continue to represent \emph{data quality}, \emph{data quantity}, and \emph{problem difficulty} respectively.
-The new term represents the \emph{imperfect testing adjustment} which is a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence, which is scaled by the factor $(1 - FP - FN)^{-1}$.
-
-Figure~\ref{fig:heatmap} shows that $D_M$, as a function of the relative frequency ($f_1/f_0$) and odds ratio, can be both positive and negative as well as a range of magnitudes~\citep{Beesley2020,Beesley2019,Smeden2019}.
-Assuming no measurement error, $D_M = 1$ so the relation between estimation and selection bias is simple, e.g., COVID-19 positive individuals more likely to receive test implies upward bias in prevalence estimates.
-Under random testing (i.e., $f_0 = f_1$), $D_M = (1-FP-FN)^{-1}$ so measurement error simply magnifies the error.
-When tests are imperfect and selection bias exists, this simple relationship no longer holds.
-
-\begin{figure}[!th]
-\centering
-\includegraphics[width = 0.4\textwidth]{../figs/mem_heatmap_article.png}
-\caption{Imperfect testing adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.024$ and $FN=0.13$. Color scaled so blue = $-10$, white = $0$, and red = $10$.}
-%\textcolor{red}{NEED TO DO FOR FP = 0.05 and FN = 0.3}}
-\label{fig:heatmap}
-\vspace{-0.3cm}
-\end{figure}
-
-Comparing the mean-squared error (MSE) under a selection mechanism $\I$ with imperfect testing and SRS with perfect testing, we see that
-$$
-\frac{E_{\I} \left[ (\bar y_n - \bar Y)^2 \right]}{\sqrt{V_{SRS} (\bar Y)}} = (N-1) E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right]
-$$
-implying the error relative to SRS increases as a function of population size~\citep{Meng2018}.
-The paradox that two countries with the same testing strategy can yield wildly different estimates due to population size is too often ignored in current communications of case-count data.
-Thus, conclusions drawn from observed case-counts may not just be wrong, but very wrong.
-
-A key question is ``What is the (effective) sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' In Appendix~\ref{app:effss}, we show the effective sample size $n_{eff}$ can be bounded by
-$$
-\frac{f}{1-f} \times \frac{1}{E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right]}.
-$$
-Between April 25th to 29th 2020, Indiana performed $95,879$ tests.
-Indiana's population is roughly $6.732$ million, so $f = 0.003$.
-The active infection rate was estimated to be~$1.81\%$~\cite{Yiannoutsos2021} in this time interval.
-Recent studies have suggested 87\% sensitivity~\cite{Arevalo2020} and 97.6\% specificity~\cite{Cohen2020} are reasonable measurement error rates for RT-PCR tests.
-Supposing COVID-19 positive individuals are $1.5$ times more likely to get tested, then the effective sample size is $168$.
-Recent proposals~\citep{Siddarth2020} have argued for increased testing capacity, which may likely reduce the relative sampling rate.
-Even if the relative sampling rate drops to $1.2$ and $f$ increases to $0.01$ then the effective sample size will increase to $1025$.
-Thus the effective sample size even in optimistic scenarios is equivalent to a moderate random sample from the population.
-Moreover, increased testing capacity may also come at the cost of increases in false positive and negative rates.
-Consider the case where $f$ increasing from $0.003$ to $0.01$ and is associated with the false negative rate rising from $13\%$ to $20$\%.
-Then the effective sample size is $863$, representing a $5.1$ factor increase rather than the expected $6.1$ factor increase.
-See Section~\ref{section:effss} in the Supplementary Materials for additional effective sample size calculations.
-% \begin{wrapfigure}{r}{0.5\textwidth}
-
-% {\bf Figure of individual going through the states? Can discuss exposure length as selection more likely.}
-
-\subsection{Regrettable rates: complex biases resulting from self-selection}
-\label{section:rates}
-
-The prior analysis demonstrates the limited information regarding COVID-19 prevalence in observational case-count data.
-Scientists, however, may claim that daily observed case-counts simply undercount daily total cases by a constant multiple over time.
-If true then the ratio of case-counts at consecutive times may be a good estimate of the true change in prevalence, helping scientists understand the disease trajectory.
-We next demonstrate how selection bias and imperfect testing impact such estimates.
-
-Let $\bar Y_{t-1}$ and $\bar Y_{t}$ denote the prevalence on two consecutive days and consider the estimator $r = \tilde y_t / \tilde y_{t-1}$.
-Using a second-order Taylor series approximation, the error between ${\tilde y_t}/{\tilde y_{t-1}}$ and ${\bar Y_{t}}/{\bar Y_{t-1}}$ can be expressed approximately as
-$$
-\begin{aligned}
-\frac{\bar Y_t}{\bar Y_{t-1}} &\times \bigg[ \rho_{I_t,Y_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (Y_t)  -\rho_{I_{t-1},Y_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \bigg] \\
-&\times \left[ 1 - \rho_{I_{t-1},Y_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \right]
-\end{aligned}
-$$
-where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement error adjustment, and $CV(Y_j) = \sigma_{Y_j}/ \bar Y_j$ is the coefficient of variation on day $j$.
-See Appendix~\ref{app:ratio} for the derivation.
-The error magnitude depends on the true rate $\bar Y_{t} / \bar Y_{t-1}$ so a large decrease will have a small error relative to a large increase.
-The second term represents potential \emph{cancellation} which can  occur when data quality, sampling fraction, measurement error, and prevalence are constant across time.
-
-Figure~\ref{fig:ratio-bias} displays the trajectory of the true ratio and the potential biased estimators under an SIR model~\citep{Pastor2001,Newman2002,Parshani2010} for the epidemic dynamics, with state evolution given by
-\begin{equation}
-\label{eq:sir}
-\frac{\partial s_t}{\partial t} = - \beta s_t i_t; \quad
-\frac{\partial i_t}{\partial t} = \beta s_t i_t - \gamma i_t; \quad
-\frac{\partial r_t}{\partial t} = - \gamma i_t
-\end{equation}
-where $s_t$ is the fraction of susceptible individuals, $i_t$ is the fraction of infected individuals, and $r_t$ is the fraction of removed (recovered or deceased) individuals in the population at time $t$.
-In terms of bias, the rate is overestimated prior to the peak and underestimated afterwards; the bias increases dramatically when the relative fraction exceeds $2$.
-Such biases may have a substantial impact on policy making.
-First, overestimation may give policy makers more leverage in proposing aggressive actions to reduce prevalence.
-Of course, the analysis provides evidence that estimates based on available data may be pessimistic.
-What is missing in the current discourse is that the direction of bias is non-constant over time.
-Underestimation post-peak puts pressure on policy makers to prematurely relax social distancing measures.
-Moreover, it appears that the peak time is the easiest to estimate in terms of having minimal error using available data.
-Note, however, that the standard errors will be large due to small effective sample sizes.
-
-\subsection{Estimation of effective reproduction number}
-\label{section:r0-estimation}
-Many epidemiologists argue that tracking the effective reproduction number $R_t$ is the only way to manage through the crisis~\citep{Gabriel2020}.
-Under a Poisson likelihood, a simple relation between the trajectory of new cases and the effective reproduction number can be derived \citep{Bettencourt2008}.
-In particular, under an SIR model the number of case counts on day $t$, denoted $K_t$, is Poisson distributed with rate $K_{t-1} \exp \left( \gamma (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval, which is approximately $7$ days for COVID-19~\citep{Sanche2020}.
-Using this relation, a moment-based estimator is given by
-$$
-R_t = 1 + \frac{1}{\gamma} \log \left( \frac{K_t}{K_{t-1}} \right).
-$$
-Of course, we do not observe $K_t$ and $K_{t-1}$.
-Under SRS of new cases among those susceptible on day $t$, the natural estimator is $S_t \tilde y_t$.
-Unfortunately the number of susceptible individuals on day $t$ is unknown.
-Here, we study the estimator $\hat R_t = 1 + \frac{1}{\gamma} \log \left( \tilde y_t / \tilde y_{t-1} \right)$.
-We can again express the statistical error of $\hat R_t - R_t$ in useful terms as follows
-$$
-\begin{aligned}
-\frac{1}{\gamma}\log &\bigg( 1 + \bigg[ \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t)  -\rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (K_{t-1}) \bigg] \\
-&\times \left[ 1 - \rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (K_{t-1}) \right] \bigg) - \frac{1}{\gamma} \log \left( \frac{S_t}{S_{t-1}} \right).
-\end{aligned}
-$$
-This implies a similar trade-off as before but on the logarithmic scale.
-The error is no longer scaled by $\bar Y_t/\bar Y_{t-1}$ but by the serial interval and does not depend on prevalence but on the fraction of new cases out of those susceptible.
-This leads to differences in when the bias is most pronounced.
-Figure~\ref{fig:r0-bias} displays the bias as a function of the relative sampling fraction assuming the population fraction that is susceptible remains large (i.e., $S_t / S_{t-1} \approx 1$).
-
-\begin{figure}
-\centering
-\begin{subfigure}{.5\textwidth}
- \centering
- \includegraphics[width=.9\linewidth]{../figs/sir_ratio.png}
- \caption{Ratio estimator}
- \label{fig:ratio-bias}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
- \centering
- \includegraphics[width=.9\linewidth]{../figs/sir_rt.png}
- \caption{Effective reproductive rate estimator}
- \label{fig:r0-bias}
-\end{subfigure}
-\caption{Potential bias due to self-selection and measurement error from a deterministic SIR model with $\beta = 1.2$ (black) and $\gamma = 0.15$.
-Here we assume $f = 0.02$, $FP = 0.024$, $FN = 0.30$, and use a range of relative sampling fraction $M = f_1/f_0$.}
-\label{fig:rates}
-\end{figure}
-
-\subsection{Rate comparisons}
-
-So far we have focused on understanding the limitations of using case-count data to understand population quantities of interest for a \emph{single} population.
-Many are interested in cross-population comparisons to contrast the impact of countries' mitigation policies.
-Here, for simplicity, we focus on comparing the estimated effective reproductive rate.
-We assume the two time-series are aligned so that $t=0$ is the time of first case in each population respectively.
-% This negates alignment issues and is common in practice.
-
-% While the above issues on $Z$-scores and effective sample size are important, here we highlight a separate issue.
-Here we demonstrate the interaction of biases in estimation of the two trajectories when the peak infection times differ slightly.  First, we can write the difference in the rates as
-$$
-R_{t1} - R_{t2} = \frac{1}{\gamma} \log \left( \frac{1 + e_{1t}}{1+e_{2t}} \right)
-$$
-where $e_{jt}$ is the error associated given in Section \ref{section:r0-estimation}.
-Recall that the error allowed for over-estimation prior to the peak and then under-estimation post-peak.
-Here, these errors can mingle in interesting ways.
-Consider two countries (A and B) in which the peak occurs 2 weeks prior for country A than country B.
-Figure~\ref{fig:comparison} presents such a comparison where each country's disease trajectory follows an SIR model (A=black and B=red).
-Of interest is the difference between the effective reproduction rate for country A and B, $R_{tA} - R_{tB}$.
-Figure~\ref{fig:eff} shows how biases in estimation of each rate interact in complex ways.
-At first, the difference is correctly estimated; then the gap is over-estimated as country A sees a rapid rise in cases; then the magnitude of over-estimation increases as country A sees declining case-count while country B sees rapidly increasing case-count; then country A's rate is correctly estimated while country B's rate is under-estimated as it sees declining case-count; finally, the gap disappears.
-We do not claim this will always be the case; however, this demonstrates how observed information can tell a more complex story than the truth (i.e., country A's peak is 2 weeks prior to country B's peak).
-
-\begin{figure}
-\centering
-\begin{subfigure}{.5\textwidth}
- \centering
- \includegraphics[width=.9\linewidth]{../figs/sir.png}
- \caption{Fraction of new cases in population}
- \label{fig:fracpop}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
- \centering
- \includegraphics[width=.9\linewidth]{../figs/sir_rt_comparison.png}
- \caption{Effective reproduction rate estimators}
- \label{fig:eff}
-\end{subfigure}
-\caption{Left: fraction infected in two SIR models with $\beta = 1.2$ and $0.9$ respectively and $\gamma = 0.15$ with same initial conditions.
-Right: comparison of $\hat R_t$ across time with $FN = 0.30$, $FP = 0.024$, and $M = 4$.}
-\label{fig:comparison}
-\end{figure}
-
-\section{Potential improvements to prevalence estimation}
-\label{section:improvedcasecount}
-
-The prior section presented negative consequences of selection-bias and measurement error when estimating infection prevalence, rates of change, and the effective reproduction number from observed case-count data.
-This raises the immediate question whether we can can improve upon these estimators.
-In this section, we consider two directions.
-The proposed methods and statistical error decompositions guide our recommendations in Section~\ref{section:discussion}.
-
-\subsection{Selection propensity estimation}
-
-With non-probability samples, one can reduce bias by modelling the self-selection propensity and use inverse probability weighting (IPW)~\citep{Elliott2017} techniques to adjust for selection bias.
-In causal inference~\citep{Hernan2020}, under certain assumptions, the observational data permits direct estimation of treatment propensity.
-In the COVID-19 context, one does not observe those who decide not to be tested.
-In order to estimate selection propensities, one needs auxiliary information.
-
-To see this mathematically, let $X_j$ denote a vector of covariates for the $j$th individual in the population, $j=1,\ldots,N$.
-For simplicity, the dynamic nature of the outbreak and recoverability of individuals is ignored for now.
-Suppose the selection indicator $I_j$ is a Bernoulli random variable that depends only on these covariates, i.e., $P(I_j = 1 \mid X_j = x) = \pi (x; \theta)$.
-Maximum likelihood estimation follows by maximizing
-\begin{equation}
-\label{eq:propensity}
-\sum_{j=1}^N I_j \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi (X_j; \theta) \right).
-\end{equation}
-The first sum only involves individuals observed in the nonprobability sample.
-The second sum is over the entire population.
-Maximum likelihood estimation therefore requires knowledge of covariate information for every individual in the population.
-Typically, this is not possible.
-Here we present a method that uses auxiliary information obtained from probability samples.
-
-\subsubsection{Auxiliary information through probability samples}
-\label{subsec:auxprob}
-
-Here we assume access to a probability sample measuring both the outcome and exact same set of covariates.
-Let $\tilde I_j$ be an indicator that the individual was included in the probability sample and $\tilde W_j$ be the probability of inclusion for the $j$th individual (i.e., equal to $n/N$ for a simple random sample of size $n$).
-In recent work,~\cite{Chen2019} use a probability sample to construct a design-unbiased estimator of the second term
-\begin{equation}
-\label{eq:auxinfoprob}
-\sum_{j=1}^N I_j \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
-\end{equation}
-Taking an expectation with respect to the sampling design yields~\eqref{eq:propensity}.
-% Under simple random sampling and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the $\hat \pi_k = (d_k / \tilde d_k) (n / N)$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.
-
-\subsubsection{An IPW estimator and statistical error decomposition}
-\label{section:IPWerrordecomp}
-
-Given a selection propensity, define the inverse probability weight~$w(x) = \pi (x; \hat \theta)^{-1}$. Then the IPW estimator adjusted for measurement error is given by
-\begin{equation}
-\label{eq:ipwest}
-\bar y_n^\star
-= \frac{1}{1-FP-FN} \cdot \frac{\sum_{i=1}^n w(x_i) (y_i - FP)}{\sum_{i=1}^n w (x_i)}
-\stackrel{(2)}{=} \frac{1}{1-FP-FN} \sum_{k=1}^K \frac{d_k w_k}{w} (\bar y_k - FP),
-\end{equation}
-where equality (2) is under the assumption that $x_i$ is a stratification variable with $k$ indexing the strata, $w_k$ is the weight and $d_k$ is the number of samples in strata $k$, and $w = \sum_{k=1}^K d_k w_k$.
-% Under simple random sampling with one strata~$\hat \pi_k = n/N$ and we recover the estimator $\tilde y_n$ from Section~\ref{section:imperfecttesting}.
-
-Let $\tilde I_j (X_j) = I_j  \cdot W(X_j)$ for $j=1,\ldots,N$.  Then the error when comparing weighted estimator~$\bar y_n^\star$ to the true prevalence $\bar Y$ can be expressed as:
-\begin{equation}
-\label{eq:statdecomp2}
-\rho_{\tilde I (X), Y} \times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y} \times \underbrace{\left[ 1 - \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{\tilde D_M}
-\end{equation}
-where $CV_W$ is the coefficient of variation (i.e., standard deviation/mean) of $W_J$ given $I_J = 1$, $\rho_{\tilde I(X), Y}$ is the empirical correlation which here depends on covariate distribution, $\tilde f_k = E[ W_J I_J \mid Y_j = k]$ for $k=0,1$, and $\tilde \Delta = \tilde f_1 - \tilde f_0$.
-See Appendix~\ref{app:ipwderivation} for the derivation.
-
-Comparing~\eqref{eq:statdecomp2} to~\eqref{eq:statdecomp} shows that weighting impacts the estimation error in three ways.
-First, there is a negative impact on the data quantity component; taking the ratio of these quantities yields
-$\sqrt{1 + \frac{CV_W^2}{1-f}} \geq 1$.
-Hence, if the data quality does not increase (i.e., $| \rho_{\tilde I (X), Y} | = | \rho_{I,Y}|$ ) then weighting increases the error magnitude.
-Second, the relative error increase depends on the fraction of population sampled $f$, implying that for large samples there is a larger potential increase in the error if the weights do not improve data quality.
-Third, the impact of measurement-error on data quality is changed when considering a weighted estimand.
-In particular, weighting may result in $\text{sgn}(\Delta) \neq \text{sgn} (\tilde \Delta)$ which implies the impact of measurement-error may be in a different direction.
-
-As demonstrated below in Lemma~\ref{lemma:ipw}, if the propensity model is correctly specified then the $E_{\bf R} [ \rho_{\tilde I (X), Y} ] = 0$ and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then the data quality index is unlikely to inversely scale with population size.
-Indeed, Horvitz-Thompson estimators are sensitive to errors in the estimated weights.
-Low selection propensity imply large weights which can have deleterious impacts by certain observations dominating estimation.
-Similar to~\cite{Meng2018}, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from an IPW estimator are likely to put too much confidence in the sheer size of the data.
-
-\subsubsection{Time-varying propensities}
-
-Here we extend the IPW approach to the temporal setting to account for the dynamic nature of the outbreak by considering the joint likelihood
-\begin{equation}
-\label{eq:tvpropensity}
-\sum_{t=1}^T \left[ \sum_{j=1}^N I_{j,t} \log \left( \frac{\pi_t (X_{j,t}; \theta)}{1-\pi_t(X_{j,t}; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi_t (X_{j,t}; \theta) \right) \right]
-\end{equation}
-where $t=1,\ldots,T$ are the days when case-count data is reported.
-Here, $I_{j,t}$ denotes self-selection into testing on day $t$, which is highly correlated with prior testing and results.
-For example, an individual who tests positive is unlikely to seek testing on subsequent days.
-Moreover, an individual in a high prevalence area may be more likely to seek out testing.
-Here, we assume that the covariate vector $X_{j,t}$ contains all features of the past relevant for selection.
-
-If sufficiently large random samples are collected at each time $t =1,\ldots,T$, then the pseudo-likelihood can be re-written as in~\eqref{eq:auxinfoprob} and propensities estimated separately per time point.
-Unfortunately for COVID-19, large probabilistic samples are not available at every time within a given region.
-Here we consider a non-parametric kernel-based approach where the selection propensity at time~$t$, denoted $\hat \theta_t$, maximizes the smoothed pseudo-likelihood
-$$
-\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \left[ \sum_{j=1}^N I_{j,t^\prime} \log \left( \frac{\pi_t (X_{j,t^\prime}; \theta)}{1-\pi_t(X_{j,t^\prime}; \theta)} \right) + \sum_{j=1}^N \tilde W_{j,t^\prime} \tilde I_{j,t^\prime} \log \left( 1 - \pi_t (X_{j,t^\prime}; \theta) \right) \right]
-$$
-where $K_h$ is a kernel function with tunable parameter $h$. Given $\pi (x;\hat \theta_t)$, the prevalence estimator $\bar y_{n,t}^\star$ is given by~\ref{eq:ipwest}.
-
-% Under simple random sampling of equal size~$\tilde n$ from a fixed population size~$N$ at each time and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the
-% $$
-% \hat \pi_{k,t} = \frac{n}{N} \times \frac{\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \hat d_{k,t^\prime}}{\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \tilde d_{k,t^\prime}}
-% $$
-% where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.
-
-%\subsubsection{Uncertain sensitivity and specificity}
-%
-%For prevalence there is some
-%
-%For
-%
-%For seroprevalence, \cite{Bendavid2020} report
-%
-%
-%
-%, which leads to a range for $M$ of $(2.21, 2.38)$.
-Therefore, the impact on bias of $\hat R_t$ appears moderate.
-If $FP = 0.05$ and $FN = 0.005$, then the relative sampling rate increases to $M=3.04$.
-This simple tool helps us understand how sensitivity and specificity impact biases and  therefore how much we should trust conclusions based on these assumptions.
-%
-%
-\subsubsection{Asymptotics}
-
-Suppose there is a sequence of finite populations of size $N_{\nu}$ indexed by $\nu$.
-Each finite population has an associated non-probability at time $t$ of  size~$\tilde n$ and $L_\nu$ probability samples of fixed size $n$ drawn at equally spaced times $\{ t^\prime_l \}_{l=1}^L$ over the study window~$[0,T]$.
-Assuming the selection propensity model is correctly specified, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:ipw} shows that the IPW estimator is  consistent and calculates the estimator's variance.
-In Lemma~\ref{lemma:ipw}, sensitivity and specificity are not assumed known.
-Instead, we suppose that these are estimated using a pseudo-likelihood on two random samples from the population whose size (denoted~$n_{FP}$ and $n_{FN}$ respectively) is chosen so the resulting estimated sensitivity and specificity have desired levels of uncertainty.
-See Appendix~\ref{app:asympderivations} for additional details.
-
-\begin{lemma}[Variance of IPW estimator] \normalfont
-\label{lemma:ipw}
-% \textcolor{red}{Must check.}
-The estimates~$\hat \mu_t := \bar y_{n,t}^\star$, $\hat \pi_{j,t}$, $\hat FP$, and $\hat FN$ are solutions to the following set of estimating equations
-$$
-\Phi_n (\eta_t) =
-\left (
-\begin{array}{c}
-\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t}{\pi_{j,t}} \\
-\frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
-\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
-\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
-\end{array}
-\right ) = {\bf 0}
-$$
-where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
-Under certain regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ where $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$.
-\end{lemma}
-
-\subsection{Model-based estimation}
-\label{section:modelbased}
-
-Up to this point, the primary focus of this paper has been on selection bias in coronavirus case-count data from a survey sampling perspective.
-Here, we consider compartmental model approaches from infectious disease epidemiology.
-Our primary objective is a model-based forecast of strata-level prevalence $\bar y_k$.
-To do this, a probabilistic extension of a standard epidemiological state-space model -- the susceptible, exposed, infected, and removed (recovered and death) model, or SEIR model -- is presented.
-A probabilistic SIR model was originally proposed by~\cite{Osthus2017} with only one-dimensional time series of infected proportions; this formulation was extended by~\cite{Song2020} to model coronavirus case-counts.
-
-Let $s_t$, $e_t$, $i_t^{(T)}$, and $r_t$ denote the proportion of survivors, exposed, infected, and removed cases (i.e., including both recovered cases and deaths) at time $t$.
-The population-level SEIR dynamics are then given by the following set of differential equations:
-\begin{align}
-\frac{\partial s_t}{\partial t} &= - \beta s_t i_t; \quad \frac{\partial e_t}{\partial t} = \beta s_t i_t - \sigma e_t; \quad \label{eq:seir} \\
-\frac{\partial i^{(T)}_t}{\partial t} &= \sigma e_t - \gamma i_t; \quad
-\frac{\partial r_t}{\partial t} = - \gamma i_t. \nonumber
-\end{align}
-Following~\cite{Song2020}, we employ Runge-Kutta (RK4) approximations for discretization.
-Here, we consider covariate information that takes the form of a stratification variable with $K$ strata.
-Let $I_{k,t}$ to denote the number of new infections on day $t$ in the $k$th strata.
-As there are a total of $N \cdot i_t := I_t$ new infections on day $t$, we assume the joint distribution of strata-specific new infection totals takes a multinomial distribution with
-$$
-( I_{1,t}, \ldots, I_{K,t} ) \sim \text{Multinomial} \left( I_t, (p_{1,t}, \ldots, p_{K, t}) \right).
-$$
-where $( {\bf p}_t )_{t=1}^T = ( (p_{1,t}, \ldots, p_{K, t}) )_{t=1}^T$ are a sequence of parameters on the simplex, i.e., $\sum_{k=1}^K p_{k,t} = 1$.
-
-Selection bias is addressed by analyzing COVID-19 death data rather than case counts.  Let $D_{k,r}$ denote the number of individuals who pass away on day $r$ from strata $k$.  Then
-$$
-D_{k,r} \mid p, \theta, \nu \sim \text{Poisson} \left( \sum_{t=1}^r \text{IFR}_{k} \cdot I_{k,t} \theta_{(r-t)} \right)
-$$
-where $\text{IFR}_{k}$ is the infection fatality rate for the $k$th strata and $\theta_{j}$ is a discrete-time distribution for time from infection to death.
-This extends prior analysis of death data~\cite{Johndrow2020} by allowing the infection fatality rate to depend on a stratification variable, which is important as IFR depends heavily on age.
-Specifically, ``the estimated IFR is close to zero for children and younger adults but rises exponentially with age, reaching 0.4\% at age 55, 1.4\% at age 65, 4.6\% at age 75, and 15\% at age 85''~\cite{Levin2020}.
-For conciseness, parameter choices and additional details on inference are discussed in Appendix~\ref{app:modelbased}.
-
-% 0-17: 20/1000000 * 100 -> 0.002 (at age 9)
-% 18-49: 500/1000000 * 100 -> 0.05 (at age 33)
-% 50–64 years old: 6,000/1000000 * 100 -> 0.6
-% 65+ years old: 90,000/1000000 * 100 -> 9.0
-%
-
-\subsection{Doubly robust estimation.}
-Rather than relying solely on IPW or epidemiological forecasts, here we combine forecasting and inverse-probability weighting by extending recent work by~\cite{Chen2019} to account time-varying propensities and measurement-error.
-In this context, the doubly-robust estimator is given by
-$$
-\bar y_{n}^{(DR)} = \frac{1}{N} \sum_{j=1}^N \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \mu(x_j) \right).
-$$
-where the estimates $\hat \mu(x_j)$ are not corrected for measurement-error as they are built to estimate true case count from death-only data.
-This estimator is called ``doubly-robust'' because it is consistent if either the model-based forecasts or the time-varying propensities are correctly specified.
-A statistical error decomposition can be derived
-\begin{align}
-\label{eq:statdecomp3}
-\begin{split}
-\rho_{\tilde I (X), Y-\mu(X)} &\times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y-\mu(X)}  \\
-&\times \underbrace{\left[ 1 - \frac{\rho_{\tilde I(X),Y} \sigma_{Y}}{\rho_{\tilde I(X),Y-\mu(X)} \sigma_{Y-\mu(X)}} \times \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right]}_{\tilde D_M}.
-\end{split}
-\end{align}
-See Appendix~\ref{app:DRderivation} for the derivation.
-Comparing~\eqref{eq:statdecomp3} to~\eqref{eq:statdecomp2} shows that the addition of the model-based forecast impacts the estimation error in two ways.
-First, if the model is adequate, then one may expect a reduction in the problem difficulty and (potentially) in the data quality components.
-Note that the model-based forecast does not impact the data quantity component.
-
-If the model-based forecast is correct then $E_{\bf R} [ \rho_{\tilde I (X), Y-\mu(X)} ] = 0$  and therefore $\rho_{\tilde I(X), Y-\mu(X)} = O(N^{-1})$; however, if the forecasts are not correctly specified then the data quality index is unlikely to inversely scale with population size.
-Interestingly, the impact of measurement-error on data quality now depends on a relative comparison of the data quality and problem difficulty of the weighted estimator and the doubly-robust estimator.
-For decent models, one would expect this ratio to be greater than one.
-Therefore, $\tilde \Delta < 0$ implies an expected increase in impact of measurement-error and $\tilde \Delta > 0$ implies the magnitude could shrink toward zero or change sign and have an even large magnitude in the opposite direction.
-Again, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from a DR estimator are likely to put too much confidence in the sheer size of the data.
-
-Unfortunately, the first term of the doubly robust estimator cannot be computed as covariate information is not collected on the entire population.
-Here, we again use the probability samples and consider the doubly-robust estimator
-$$
-\frac{1}{\sum_{j=1}^N \tilde I_j \tilde W_j} \sum_{j=1}^N  \tilde I_j \tilde W_j \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \mu(x_j) \right).
-$$
-where $\tilde I_j$ indicates an individual is included in the random sample and $\tilde W_j$ is the associated weight for that unit.
-Assuming the selection propensity model is correctly specified, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:dr} shows that the IPW estimator is  consistent and calculates the estimator's variance.
-See Appendix~\ref{app:asympderivations} for additional details.
-
-\begin{lemma}[Variance of doubly-robust estimator] \normalfont
-\label{lemma:dr}
-The estimates~$\mu_t := \bar y_{n,t}^\star$ and $\hat \pi_{j,t}$
-are solutions to the following set of estimating equations
-$$
-\Phi_n (\eta_t) =
-\left (
-\begin{array}{c}
-\frac{1}{N} \left[ \sum_{j=1}^N \frac{I_{j,t}}{\pi_{j,t}} \left( \left( \frac{Y_{j,t} - FP}{1-FP-FN} - \mu(x_j)  \right)  - \mu_t \right) + \frac{1}{\sum_{j=1}^N \tilde I_j \tilde W_j} \sum_{j=1}^N  \tilde I_j \tilde W_j \hat \mu (x_j) \right] \\
-\frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
-\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
-\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
-\end{array}
-\right ) = {\bf 0}
-$$
-where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
-Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$ and $\text{var} (\hat y_{n,t}) = V_{t}^{(DR)} + o (n^{-1})$ where $V_t^{(DR)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$.
-\end{lemma}
-
-\section{COVID-19 active infection prevalence in Indiana}
-\label{section:applications}
-
-% For COVID-19 case-count data, the only covariate information provided by state and local governments are summary statistics for those factors discussed in Section~\ref{section:testinginfo} -- age, sex,  and race.
-These are discrete factors and therefore define a stratification of the population.
-Currently, US testing does not collect additional covariate information such as symptom information; however, this need not be the case moving forward.
-Recommendation XX below states that additional covariate information should be collected from nonprobabilistic samples.
-
-We next consider estimation of the active infection rate in Indiana using unweighted, IPW, model-based, and doubly robust estimates. To start, we recap the data sources used in estimation:
-\begin{itemize}[leftmargin=*]
-\item {\bf Testing data}:  The number of daily tests performed and number of daily positive tests are reported.
-These counts are broken out jointly by age, gender, and racial demographic information.
-Figure~\ref{fig:in-cases} and~\ref{fig:in-tests} plots daily COVID-19 positive cases and tests by age strata.
-
-\begin{figure}
-\centering
-\begin{subfigure}{.45\textwidth}
- \centering
- \includegraphics[width=.9\linewidth]{../figs/indianacasecounts_byage.png}
- \caption{Case-counts per day}
- \label{fig:in-cases}
-\end{subfigure}
-\begin{subfigure}{.45\textwidth}
- \centering
- \includegraphics[width=.9\linewidth]{../figs/indianacovidtests_byage.png}
-\caption{Tests per day}
-\label{fig:in-tests}
-\end{subfigure} \\ [1ex]
-\begin{subfigure}{\linewidth}
-\centering
-\includegraphics[width=.45\linewidth]{../figs/indianadeaths_byage.png}
- \caption{Deaths per day}
- \label{fig:in-deaths}
-\end{subfigure}
-\caption{Indiana daily COVID-19 case count, testing, and death data by age strata}
-\end{figure}
-
-\item {\bf Random/Nonrandom statewide sample}: Table~\ref{tab:indiana} summarizes data from a random sample from April 25--29th as well as a nonrandom sample obtained between May 2--3 in racial/ethnic minority communities~\cite{Yiannoutsos2021}.
-The nonrandom sample is not a random subsample of the overall case-counts; however, this data provides important supplementary covariate information.
-% Namely, the samples collected included covariate information on (1) prior symptoms, (2) COVID-19 positive individual in household, and (3) prior COVID-19 positive test.
-To account for the nonrandom sample being from high risk areas, we adjust estimates to match the statewide positivity rate of 11.7\% between May 1--2.
-These adjusted rates are presented in parentheses in Table~\ref{tab:indiana} along with the relevant covariate information.
-
-\item {\bf Facebook symptom survey}: From the daily symptom surveys completed on Facebook (see Section~\ref{subsection:fbsymptom} for details), we extract survey responses for individuals who identify as living in Indiana.
-We collect age, sex, and demographic information as well as COVID-19 related symptoms, which Facebook defines as having (1) a fever and cough, (2) shortness of breath, \emph{or} (3) difficulty breathing.
-To assess potential bias, we compare the random sample from Indiana to the Facebook symptom survey data collected from April 25th to April 29th.
-Table~\ref{tab:comparison} demonstrates minimal bias in the Facebook symptom survey with respect to symptom distributions.
-
-\begin{table}[!th]
-\centering
-\begin{tabular}{c | c c c | c}
-& \multicolumn{3}{c}{Facebook} & Random \\ \cline{2-5}
-& Lower CI & Estimate & Upper CI & Estimate \\ \hline
-Fever &  0.017 & 0.020 & 0.024 & 0.018 \\
-Cough &  0.142 & 0.148 & 0.154 & 0.149 \\
-Shortness & 0.058 & 0.060 & 0.062 & 0.062 \\ \hline
-\end{tabular}
-\caption{Comparison of Facebook survey and random sample}
-\label{tab:comparison}
-\end{table}
-
-% \item {\bf Hospitalization data}: On each day, we observe the number of individuals who are COVID-19 positive that visited a hospital on that day.
-Unfortunately, hospitalization data is not broken out by age, gender, and racial demographic information.
-Instead, we rely on combining Facebook data to calculate the fraction of those who sought the hospital per week in each strata.
-We then use this to figure out how many individuals per strata were hospitalized and then back out how many
-
-\item {\bf Death data}: Daily COVID-19 related deaths are observed by age, gender, and racial demographics.
-Figure~\ref{fig:in-deaths} plots the cumulative number of COVID-19 related deaths over time by age group.
-While most COVID-19 tests and positive cases are within the younger age strata, most deaths are within the 70+ age strata.
-\end{itemize}
-
-\begin{table}[th]
-\begin{tabular}{c | c | r r | r | r r }
-& & \multicolumn{2}{c}{Total Tests (\%)} & US & \multicolumn{2}{c}{Positive Test Rate (\%)}\\
-\cline{3-4} \cline{6-7}
-& & NonRandom & Random & Census & NonRandom & Random \\ \hline
-\multirow{2}{*}{Sex} & Female & 58.2 & 55.0 & 50.7 & 21.7 (11.2) & 1.4 \\
-& Male & 41.8 & 45.0 & 49.3 & 24.2 (12.4) & 2.1 \\ \hline
-\multirow{3}{*}{Age} & $<40$ & 39.4 & 28.0 & 52.7 & 29.7 (15.0) & 1.7 \\
-& $40-59$ & 41.1 & 36.0 & 25.2 & 24.9 (12.5) & 2.1 \\
-& $\geq 60$ & 19.5 & 36.0 & 22.1 & 6.7 (3.4) & 0.9 \\ \hline
-\multirow{2}{*}{Race} & White & 23.1 & 92.0 & 86.9 & 19.5 (9.6) & 1.5 \\
-& Nonwhite & 76.9 & 8.0 & 13.1 & 25.0 (12.3) & 3.4 \\ \hline
-% Hispanic & Hispanic & 44.1 & 2.0 & 7.1 & 37.6 & 6.9 \\
-% Origin & Non-Hispanic & 55.9 & 98.0 & 92.9 & 13.0 & 1.3 \\ \hline
-\multirow{2}{*}{Fever} & Yes & 17.0 & 1.8 & - & 66.4 (32.1) & 4.5 \\
-& No & 83.0 & 98.2 & - & 15.6 (7.5) & 1.3 \\ \hline
-Household & Yes & 10.8 & 1.4 & - & 46.1 (22.4) & 29.4 \\
-$+$ Case & No & 89.2 & 98.6 & - & 21.6 (10.4) & 1.3 \\ \hline
-Prior $+$ & Yes & 6.1 & 1.4 & - & 39.2 (20.2) & 24.4 \\
-Test & No & 93.9 & 98.6 & - & 21.6 (11.1) & 1.3 \\ \hline
-\end{tabular}
-\caption{Estimated point prevalence of active infection with SARS-CoV-2 by demographics in Indiana~\cite{Yiannoutsos2021}.}
-\label{tab:indiana}
-\end{table}
-
-\subsection{Inverse-probability weighting approach}
-\label{section:ipwapproach}
-
-% FP: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7307014/
-% FN: https://www.medrxiv.org/content/10.1101/2020.04.16.20066787v2.full.pdf
-
-We start by using Table~\ref{tab:indiana} to compute IPW weights for end of April, early May.
-The random sample is $n=3,658$ and IPW weights are computed per strata which depend on Gender, Age, Race, Fever, Positive Case in Household, and Prior Positive Test.
-The weight for the strata defined as Male, 40--59, White, has a Fever, no household cases, and no prior tests, for example, is proportional to:
-$$
-\frac{\left( 0.493 \times 0.252 \times 0.869 \times 0.018 \times 0.986 \times 0.986 \right)}{\left(0.418 \times 0.411 \times 0.231 \times 0.17 \times 0.892 \times 0.939 \right)} \approx 0.334
-$$
-Using the constructed weights and the adjusted strata-level prevalence estimates, the IPW estimate under no measurement-error is $7.7\%$, a decrease of four percentage points from the observed prevalence of $11.7\%$.
-We also make use of Facebook's symptom survey over that window of time to construct a second IPW estimate conditional on demographics and fever but ignoring the household and prior testing information, which under no measurement-error is $6.5$\%.
-
-The Indiana study did not report sensitivity and specificity; therefore, we take the suggested measures from~\cite{Arevalo2020} which report 87\% sensitivity is a reasonable estimates and~\cite{Cohen2020} which report 97.6\% specificity.
-This corresponds to a false negative rate of $13$\% and false positive rate of $2.4\%$.
-Therefore, the IPW estimates accounting for measurement-error are $6.2\%$ and $4.9\%$ respectively.
-
-\subsubsection{Disease prevalence by April 2020}
-
-Indiana's population as of 2019 was $6.732$ million.
-A total of $19,649$ tests were administered in Indiana between April 25th to 29th.
-Subtracting off the $95,879$ tests that had already performed yields a sampling fraction of $f = 2.96 \times 10^{-3}$.
-Here we consider estimation of data quality $E_{\I} \left[ \rho_{\tilde I(X),Y} \right]$.
-In \cite{Meng2018}, estimation relied on observing the true outcome (i.e., election vote totals), which is not possible in the current crisis.
-Here, we use the random sample which estimates the true prevalence of active COVID-19 disease at 1.81\%~\cite{Yiannoutsos2021}.
-Our goal is not inferential.
-Instead, we aim to build sensitivity analyses that can aid in understanding the amount of information in observational COVID case-count data.
-
-The unweighted estimate for the true prevalence of active COVID-19 disease between April 25th and April 29th is 11.7\%.
-Assuming a false negative rate of $13$\% and false positive rate of $2.4$\%, the unweighted estimate is $11.0$\%, leading to an error of $9.2$\%.
-Using~\eqref{eq:statdecomp}, we construct an estimate of the relative sampling rate as follows
-\begin{equation*}
-\rho D_M = \sqrt{\frac{f}{1-f}} \frac{\text{0.092}}{\sigma_Y} = 3.75 \times 10^{-2} \Rightarrow \Delta = 1.39 \times 10^{-2} \Rightarrow M = 6.1.
-\end{equation*}
-Using inverse-probability weights, the IPW estimates for the true prevalence of active COVID-19 disease between April 25th and April 29th are $7.7$\% and $6.5\%$ respectively.
-Under the same sensitivity and specificity, the weighted estimates adjusted for measurement-error are $6.2$\% and $4.9$\%, leading to errors $4.4$\% and $3.1$\% respectively.
-Using~\eqref{eq:statdecomp2}, we construct an estimate of the relative sampling rate as follows
-\begin{equation*}
-\tilde \rho D_M = \sqrt{\frac{f}{1-f+CV_W^2}} \frac{\text{0.044}}{\sigma_Y} = 8.14 \times 10^{-3} \Rightarrow \tilde \Delta = 8.22 \times 10^{-3}  \Rightarrow \tilde M = 3.9,
-\end{equation*}
-using the Indiana non-random sample.
-For the IPW estimate using Facebook data, we have $\tilde M = 3.3$.
-Therefore, the impact of self selection on bias of $\hat R_t$ appears strong using unweighted data, remaining moderate even using the weighted estimates.
-
-A sensitivity analysis is performed by considering the range of false negative and false positive rates.
-Here, sensitivity is assumed to fall between $81$\% and $91$\% and specificity between $95$\% and $98.8$\%~\cite{Katz2020}, which leads to a range for $M$ of $(5.60, 6.44)$ for the unweighted analysis, and $(3.63,4.07)$ and $(3.10, 3.45)$ for the weighted analyses respectively.
-
-\subsubsection{Time-varying IPW estimator}
-\label{section:tvipw}
-
-The IPW analysis is next extended to the time-varying setting.
-To do so, we make use of the Facebook symptom survey, hospitalization, and testing data.
-Recall the COVID-19 testing and positive case counts are known by age, gender, and racial demographics.
-Symptom status (e.g., fever, coughing, shortness of breath) and related covariate information (COVID-19 positive contact), however, are currently unavailable\footnote{Whether this is due to data privacy considerations or lack of data collection efforts is unclear.}.
-As COVID-19 symptom status is likely to alter the testing propensity and likelihood of active infection, for illustrative purposes Facebook symptom survey and hospitalization data are used in two ways to imputate symptom status per strata.
-We outline the imputation methods below; see Section~\ref{section:discussion} for additional discussion.
-
-In the first approach, we estimate two logistic regressions using weighted pseudo-likelihoods with the Facebook survey symptom data.
-The first estimates the probability of contact with a COVID-19 positive individual given demographic and COVID-19 test status (positive or negative COVID-19 test).
-Figure~\ref{fig:contactlik1} and~\ref{fig:contactlik2} in Appendix~\ref{app:in_add_details} presents the estimated likelihood of contact with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.
-The second estimates the probability of fever given demographic, COVID-19 test status, \emph{and} whether the individual has had contact with a COVID-19 positive individual.
-Figure~\ref{fig:symptomlik1} and~\ref{fig:symptomlik2} in Appendix~\ref{app:in_add_details} presents the estimated likelihood of reporting a fever with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.
-We see that likelihood of fever varies greatly based on whether the individual also reported contact with a COVID-19 positive individual.
-Using these two models, mean imputation is used to calculate the number of positive and negative tests given demographic strata, fever status, and COVID-19 contact status.
-Inverse probability weights are then computed based on the resulting dataset (termed IPW1).
-
-In the second approach, we leverage hospitalization information.
-We first estimate the probability of fever given demographic and COVID-19 test status.
-When the COVID-19 test status is positive, the probability of fever is allowed to depend on whether the individual was hospitalized.
-Figure~\ref{fig:symptomlik1_model2} and~\ref{fig:symptomlik2_model2} in Appendix~\ref{app:in_add_details} presents the estimated likelihood of reporting a fever with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.
-We see that hospitalization significantly increases the likelihood of fever as expected.
-We use these models and the hospitalization rates by demographic to perform mean impuation of the number of positive and negative tests given demographic strata, fever status, and COVID-19 contact status.
-Inverse probability weights are then computed based on the resulting dataset (termed IPW2).
-
-To make the approach concrete, Figure~\ref{fig:testinglik_mainpaper} presents the testing propensity for non-hispanic, white males across age strata and fever status using the second approach.
-We note here that the ratio of testing propensity within strata across fever status is time-varying.
-The ratio starting nine times more likely in early April and dropping to twice as likely by end of 2020.
-
-\begin{figure}[!th]
-\centering
-\begin{subfigure}{.5\textwidth}
- \centering
- \includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig1_mainpaper.png}
- \caption{Testing Likelihood Given Fever}
- \label{fig:testinglik1_mainpaper}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
- \centering
-\includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig2_mainpaper.png}
- \caption{Testing Propensity Given No Fever}
- \label{fig:testinglik2_mainpaper}
-\end{subfigure}
-\caption{Testing Propensity across age strata and fever status}
-\label{fig:testinglik_mainpaper}
-\end{figure}
-
-\subsection{Application of SEIR model}
-\label{section:modelbased}
-
-Here we consider the multinomial SEIR model using observed death data as presented in Figure~\ref{fig:in-deaths}.
-The model, presented in Section~\ref{section:modelbased}, requires infection fatality rates to be specified per-strata.
-In this paper, similar to~\cite{Johndrow2020}, we specify fixed per-strata infection fatality rates.
-Based on published age-specific IFRs~\cite{Levin2020}, Figure~\ref{fig:logifr} shows that IFR seems to closely follow a log-linear relationship with age.
-Combining across published age-specific IFRs, Table~\ref{tab:ifrperage} presents estimates of the average IFR within each strata using the age-distribution.
-\begin{table}[!th]
-\begin{tabular}{c | c c c c c c c c}
-Age Strata & 0-19 & 20-29 & 30-39 & 40-49 & 50-59 & 60-69 & 70-79 & 80+ \\ \hline
-IFR & 0.003\% & 0.014\% & 0.046\% & 0.144\% & 0.456\% & 1.443\% & 4.567\% & 15.468\%
-\end{tabular}
-\caption{IFR by age-strata}
-\label{tab:ifrperage}
-\end{table}
-The discrete-time distribution~$\{ \theta_{j} \}$ is a discretized version of a truncated normal distribution with mean $25$, standard deviation~$5$, minimum value $0$, and maximum value $44$.
-This was chosen to closely mimic the distribution from~\cite{Johndrow2020}.
-
-The SEIR model in equation~\eqref{eq:seir} is extended to allow for time-varying transmission rates~$\beta_t$ to account for public policy changes discussed in Section~\ref{section:publicpolicyindiana} over the time period considered.
-Here, we insert three change points $(t_1,t_2,t_3)$ on March 23rd, June 15th, and October 1st of 2020 respectively.
-Given sensitivity to these choices, instead of considering a multiplicative structure, we allow for delayed implementation and slow change in the transmission rate by using weights:
-$$
-\beta_t = \sum_{k=1}^{3} \left(\beta_{k-1} (1-w^{(k)}_t ) + \beta_{k} w_t\right),\quad w^{(k)}_t = (1 + \exp \left( -\xi (t - t_k - \nu) \right))^{-1}
-$$
-where $\nu \geq 0$ is the potential delay and $\xi$ is the rate of change.
-Each weight~$w_t^{(k)} = 0$ for $t < t_k$ and $w_t^{(k)} \to 1$ as $t \to \infty$.
-See Section~\ref{app:modelbased} in the supplementary materials for details on prior specification.
-
-To check how well our model fits the observed data, a posterior predictive check on aggregate death counts was performed.
-Figure~\ref{fig:deathsppc} shows the results which suggests chosen SEIR model fits the observed data well.
-The model produces age-specific case counts which can then be used for model-based estimation of the active infection rate as well as in constructing the doubly robust estimator.
-
-\begin{figure}[!th]
-\centering
-\includegraphics[width=.6\linewidth]{../figs/deaths_ppc.png}
-\caption{Posterior predictive check on aggregate daily death counts.}
-\label{fig:deathsppc}
-\end{figure}
-
-\subsection{Time-varying prevalence estimates}
-
-Here, we construct active infection rate estimates using the unweighted, inverse-probability weighted, model-based, and doubly robust estimators.
-Figure~\ref{fig:air} presents the results.
-The inverse-probability weighting methods do not significantly reduce bias compared to the unweighted estimates.
-We conjecture this is due to the limited availability of symptom and other important covariates that may impact the likelihood an individual will get  tested.
-Our imputation strategies use available data to try and alleviate this issue; however, this suggests better data collection in the future may strongly improve the IPW estimates as seen in Section~\ref{section:ipwapproach} when analyzing the statewide sample from April 25th--29th.
-
-While the model-based estimates appear more reasonable, note that the estimate on the last week of April is 0.25\% which is a severe underestimate compared to the estimated prevalence using the stratified random sample.
-Caution is therefore warranted when interpreting these estimates.
-We know that the age-specific infection fatality rate is likely time-varying and may vary by other factors beyond age.
-Therefore, improvements could be made by more finely grained, and accurate IFR estimates; however, there is currently minimal publicly available data to do so.
-Moreover, the (thankfully) low number of deaths per strata make uncertainty in these estimates quite high.
-Finally, the doubly robust estimates appear similar to the IPW estimate.
-
-\begin{figure}
-\centering
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{../figs/tv_air.png}
-  \caption{Active Infection Rate Estimates}
-  \label{fig:air}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
- \centering
- \includegraphics[width=.9\linewidth]{../figs/tv_air_ratio.png}
- \caption{Ratio estimates}
- \label{fig:air_ratio}
-\end{subfigure}
-\caption{Time-varying estimates based on unweighted, model-based, IPW and doubly robust methods.}
-\label{fig:tvestimates}
-\end{figure}
-
-We also construct ratio estimates as discussed in Section~\ref{section:rates}.
-Figure~\ref{fig:air_ratio} presents the results.
-Unweighted, IPW, and doubly robust estimates appear similar.
-Recall the severe bias during the growth phase due to selection bias dicussed in Section~\ref{section:rates} may imply the observed rates are  overestimated between October 2020 and January 2021.
-The model-based methods may be as well.
-
-\section{Discussion}
-\label{section:discussion}
-
-There is nothing routine about COVID-19, including the corresponding statistical questions.
-The goal of this paper was to point out questionable statistical routines.
-Precision in reported case-count data gives the illusion of information when what what is needed is quantification of uncertainty.
-Extensions of recent statistical error decompositions~\cite{Meng2018} demonstrate how selection bias leads data analysts to feel certain about incorrect conclusions.
-As case-count data is routinely used for public health policy making, we presented an inverse-probability weighting method and a doubly-robust estimation method that leverage auxiliary information collected through random samples to overcome these issues.
-We end with a brief discussion of important related topics.
-
-\subsection*{Data quantity versus quality}
-
-Governments and policy makers often implicitly argue that increased testing capacity will alleviate selection bias.
-Without complete compliance, however, our understanding of future outbreaks may be plagued by self-selection bias, compounded by changing sensitivity and specificity rates of RT-PCR testing.
-Random testing removes these effect modifiers, giving governments more information to fight the disease.
-This paper emphasizes that data quantity is secondary to both data quality and methodological considerations to account for selection bias.
-
-\subsection*{High quality information on prevalence.}
-
-Governments are pushing for increased testing capacity and robust contact tracing.
-Contact tracing is incredibly useful for identifying carriers early and preventing spread of the disease.
-We argue that understanding active infections is also key.
-We know that the probability of an outbreak is a function of current prevalence and network connectivity.
-Epidemic critical thresholds have been derived for many infectious disease models~\citep{Pastor2001,Newman2002,Parshani2010}.
-Knowing prevalence will help governments determine long-term community-level risk and allow for more targeted interventions -- shutting down only certain locations when necessary.
-
-\subsection*{Model-based solutions.}
-
-A common argument is that the SEIR model can be extended to account for selection bias and measurement error directly; therefore, there is no need for auxiliary random sampling.
-Without strong assumptions on the selection mechanism, however, the estimates are often not identifiable.
-When an issue ``cannot be resolved nonparametrically then it is usually dangerous to resolve it parametrically''\citep{CoxHink74}.
-Absent some type of random sampling, the best route forward for all data analyses is careful associated sensitivity analyses and humility in data-driven conclusions.
-
-%\textcolor{red}{Underascertainmnet as pre-processed or modelled but often is an assumption that must be resolved parametrically. }
-
-\subsection*{Real-world implementation of the proposed method}
-
-Application of the proposed methods relied on access to two critical datasets: (1) COVID-19 testing, case, hospitalization, and death data by demographic strata and (2) Facebook symptom survey.
-Public access to (1) required the author to submit a data request directly to the state of Indiana.
-Through conversations with the state of Indiana, the author's request was one of only five data requests to be approved by Indiana for public release.
-Most states do not release such granular data, making selection bias adjustments difficult.
-More systematic data collection and reporting is critical in future efforts.
-We acknowledge that this proposal will lead to data privacy issues that will need to be addressed.
-
-While we are eternally grateful to Indiana for the release, symptom information was not made available.
-This severely limits our ability to estimate correct selection propensity models.
-What is required in the future is a more detailed effort to collect relevant measures of an individual's past history that may impact their likelihood of seeking testing and being a positive COVID-19 case.
-Such data collection and dissemination is critical to address selection bias.
-
-A valid criticism of the proposed approach is that recording covariates on every tested individual is too time-consuming and costly, i.e., the data collection does not scale well.
-We argue that the list of relevant factors is not too long.
-For example, symptom status and COVID-19 contact are clearly relevant factors in testing selection.
-Future work may consider how to include/exclude various factors over time that are not significant to limit citizen burden in reporting.
-Moreover, equation~\eqref{eq:auxinfoprob} presupposes covariate information is collected for all individuals in the nonprobability sample; however, this is often not feasible for state and local governments where rapid sample collection is prioritized over additional collection of covariate information.
-We can easily extend~\eqref{eq:auxinfoprob} by randomly sampling a subset of the nonprobability sample on which to collect the covariate information, yielding a new pseudo-likelihood
-\begin{equation*}
-% \label{eq:auxinfoprob2}
-\sum_{j=1}^N I_j I_j^{(2)} W_j^{(2)} \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
-\end{equation*}
-Here, $I_j^{(2)}$ is an indicator that the individual is in the probability subsample and $W_j^{(2)}$ is the probability of inclusion.
-Estimation proceeds by maximizing the pseudo-likelihood which now depends on random sampling in both the first and second term.
-This would allow for rapid testing and minimizes cost of covariate collection.
-
-\bibliographystyle{plain}
-\bibliography{covid-refs}
 
@@
 \section{IRLS}
-
+\label{app:irls}
 First derivative with respect to $\theta$
@@
 
+\bibliographystyle{plainnat}
+\bibliography{covid-refs}
+
 \end{document}

### 2021-08-23 write-up/covid-writeup.tex
Switched to 2 documents

@@
 \usepackage{natbib}
+%\setcitestyle{author-year,open={((},close={))}} %Citation-related commands
 %\bibliographystyle{apalike}
@@
 \usepackage{caption,subcaption}
+
+\usepackage{xr}
+\makeatletter
+\newcommand*{\addFileDependency}[1]{% argument=file name and extension
+  \typeout{(#1)}% latexmk will find this if $recorder=0 (however, in that case, it will ignore #1 if it is a .aux or .pdf file etc and it exists!
+if it doesn't exist, it will appear in the list of dependents regardless)
+  \@addtofilelist{#1}% if you want it to appear in \listfiles, not really necessary and latexmk doesn't use this
+  \IfFileExists{#1}{}{\typeout{No file #1.}}% latexmk will find this message if #1 doesn't exist (yet)
+}
+\makeatother
+
+\newcommand*{\myexternaldocument}[1]{%
+    \externaldocument{#1}%
+    \addFileDependency{#1.tex}%
+    \addFileDependency{#1.aux}%
+}
+%%% END HELPER CODE
+
+% put all the external documents here!
+\myexternaldocument{covid-supplement-writeup}
 
@@
 These propensities can then used in an inverse-probability weighting scheme to construct estimators of disease prevalence that attempt to control for selection bias.
-A doubly robust extension allows researchers to combine these estimates with epidemiological forecasts based on compartmental models that are common in the study of infectious diseases~\cite{Hao2020,Song2020,Ray2020,Johndrow2020}.
+A doubly robust extension allows researchers to combine these estimates with epidemiological forecasts based on compartmental models that are common in the study of infectious diseases~\citep{Hao2020,Song2020,Ray2020,Johndrow2020}.
 
@@
  Recent work by \cite{Zhao2021} pointed out that estimation of key epidemiological parameters such as the incubation time using standard epidemiological models can suffer from severe bias due to issues beyond selection bias and measurement error.
-Namely, right truncation and epidemic growth leading to patients ``being more likely to be infected towards the end of their exposure period''~\cite[pp.
+Namely, right truncation and epidemic growth leading to patients ``being more likely to be infected towards the end of their exposure period''~\citep[pp.
 3]{Zhao2021}.
@@
 Specimen collection by swab is known to impact false negative/positive rates regardless of test timing.
-Systematic reviews suggest that 87\% sensitivity and 97.6\% specificity are reasonable estimates for current RT-PCR tests~\cite{Arevalo2020, Woloshin2020,Cohen2020}.
+Systematic reviews suggest that 87\% sensitivity and 97.6\% specificity are reasonable estimates for current RT-PCR tests~\citep{Arevalo2020, Woloshin2020,Cohen2020}.
 
@@
 Such diagnostic tests generate important information about the presence of SARS-CoV-2 in the population, and help scientists and policy-makers understand patterns of transmission and propagation.
-Rapid and frequent molecular testing is cited as a key component~\cite{OECD2021} in effective strategies to identify active infections and prevent systemic outbreaks.
+Rapid and frequent molecular testing is cited as a key component~\citep{OECD2021} in effective strategies to identify active infections and prevent systemic outbreaks.
 
@@
 Indiana's population is roughly $6.732$ million, so $f = 0.003$.
-The active infection rate was estimated to be~$1.81\%$~\cite{Yiannoutsos2021} in this time interval.
-Recent studies have suggested 87\% sensitivity~\cite{Arevalo2020} and 97.6\% specificity~\cite{Cohen2020} are reasonable measurement error rates for RT-PCR tests.
+The active infection rate was estimated to be~$1.81\%$~\citep{Yiannoutsos2021} in this time interval.
+Recent studies have suggested 87\% sensitivity~\citep{Arevalo2020} and 97.6\% specificity~\citep{Cohen2020} are reasonable measurement error rates for RT-PCR tests.
 Supposing COVID-19 positive individuals are $1.5$ times more likely to get tested, then the effective sample size is $168$.
@@
  % While the above issues on $Z$-scores and effective sample size are important, here we highlight a separate issue.
-Here we demonstrate the interaction of biases in estimation of the two trajectories when the peak infection times differ slightly.  First, we can write the difference in the rates as
-$$
-R_{t1} - R_{t2} = \frac{1}{\gamma} \log \left( \frac{1 + e_{1t}}{1+e_{2t}} \right)
-$$
-where $e_{jt}$ is the error associated given in Section \ref{section:r0-estimation}.
+ Here we demonstrate the interaction of biases in estimation of the two trajectories when the peak infection times differ slightly.
+First, we can write the difference in the rates as $R_{t1} - R_{t2} = \frac{1}{\gamma} \log \left( \frac{1 + e_{1t}}{1+e_{2t}} \right)$ where $e_{jt}$ is the error associated given in Section \ref{section:r0-estimation}.
 Recall that the error allowed for over-estimation prior to the peak and then under-estimation post-peak.
@@
 Precision in reported case-count data gives the illusion of information when what what is needed is quantification of uncertainty.
-Extensions of recent statistical error decompositions~\cite{Meng2018} demonstrate how selection bias leads data analysts to feel certain about incorrect conclusions.
+Extensions of recent statistical error decompositions~\citep{Meng2018} demonstrate how selection bias leads data analysts to feel certain about incorrect conclusions.
 As case-count data is routinely used for public health policy making, we presented an inverse-probability weighting method and a doubly-robust estimation method that leverage auxiliary information collected through random samples to overcome these issues.
@@
 \newpage
-\textcolor{red}{Fix references!}
-\bibliographystyle{plain}
+% \textcolor{red}{Fix references!}
+\bibliographystyle{plainnat}
 \bibliography{covid-refs}
 
-\newpage
-\appendix
-
-\section{Reproducible code}
-
-All relevant code can be found at \url{https://github.com/wdempsey/covid-umich}.
-
-\section{Technical details}
-
-Recall that $P_j$ is  an indicator of measurement error, equal to $1$ when we incorrectly measure the outcome and $0$ when we observe the true outcome.
+% \newpage
+% % \appendix
+
+% % \section{Reproducible code}
+
+% % All relevant code can be found at \url{https://github.com/wdempsey/covid-umich}.
+
+% % \section{Technical details}
+
+% % Recall that $P_j$ is  an indicator of measurement error, equal to $1$ when we incorrectly measure the outcome and $0$ when we observe the true outcome.
 We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.
@@
 
-\subsection{Derivations for imperfect testing framework.}
-\label{app:imperfect}
-We start by considering the empirical mean estimator under imperfect testing,
-$$
-\bar y_n^\star = \frac{\sum_{j=1}^N Y_j^\star I_j}{\sum_{j=1}^N I_j} = \frac{\sum_{i=1}^N  I_j Y_j^\star }{\sum_{j=1}^N  I_j } = \frac{\sum_{i=1}^N  I_j \left[ Y_j (1-P_j) + (1-Y_j) P_j \right]}{\sum_{j=1}^N  I_j }
-$$
-For any set of numbers $\{ A_1, \ldots, A_N \}$ we can view it as the support of a random variable $A_J$ induced by the random index $J$ defined on $\{1,\ldots, N\}$.
+% % \subsection{Derivations for imperfect testing framework.}
+% % \label{app:imperfect}
+% % We start by considering the empirical mean estimator under imperfect testing,
+% % $$
+% % \bar y_n^\star = \frac{\sum_{j=1}^N Y_j^\star I_j}{\sum_{j=1}^N I_j} = \frac{\sum_{i=1}^N  I_j Y_j^\star }{\sum_{j=1}^N  I_j } = \frac{\sum_{i=1}^N  I_j \left[ Y_j (1-P_j) + (1-Y_j) P_j \right]}{\sum_{j=1}^N  I_j }
+% % $$
+% % For any set of numbers $\{ A_1, \ldots, A_N \}$ we can view it as the support of a random variable $A_J$ induced by the random index $J$ defined on $\{1,\ldots, N\}$.
 When $J$ is uniformly distributed $E_J (A_J) = \sum_{j=1}^N A_j / N \equiv \bar A_N$.
 Then
-$$
-\begin{aligned}
-\bar y_n^\star  - \bar Y_N &= \frac{E_J \left[ I_J \left[ Y_J (1-P_J) + (1-Y_J) P_J \right] \right]}{E_J [ I_J ] } - E_J[Y_J] \\
-&= \frac{E_J \left[ I_J P_J (1-2Y_J) \right]}{E_J [ I_J ] } + \left( \frac{E_J [I_J Y_J]}{E_J [ I_J ] } - \frac{E_J[Y_J] E_J[I_J]}{E_J[I_J]} \right) \\
-\end{aligned}
-$$
-The term in parentheses can be re-written as
-$$
-\begin{aligned}
-\frac{E_J [I_J Y_J]- E_J[Y_J] E_J[I_J]}{E_J[I_J]} &=  \frac{E_J [I_J Y_J]- E_J[Y_J] E_J[I_J]}{\sqrt{V_J(I_J) V_J(Y_J)}} \frac{\sqrt{V_J(I_J)}}{E_J[I_J]} \times \sqrt{V_J(Y_J)} \\
-&= \rho_{I,Y} \times \sqrt{\frac{(1-f)}{f}} \times \sigma_Y
-\end{aligned}
-$$
-which agrees with Meng's (2019) decomposition. For the other term, first we define $Z_j := 1 - 2 Y_j $. Then $Z_j = 1$ if $Y_j = 0$ and $Z_j = -1$ if $Y_j = 1$. Then the term can be re-written as
-$$
-\begin{aligned}
-\frac{E_J \left[ I_J P_J (1-2Y_J) \right]}{E_J [ I_J ] } &= \left( \frac{E_J \left[ I_J P_J Z_J \right]}{E_J [ I_J ] } -  \frac{E_J \left[ P_J Z_J \right] E_J[ I_J]}{E_J [ I_J ] } \right) +  \frac{E_J \left[ P_J Z_J \right] E_J[ I_J]}{E_J [ I_J ] } \\
-\end{aligned}
-$$
-The term in parentheses can be re-expressed using the previous technique as:
-$$
-\rho_{I, PZ} \times \sqrt{\frac{1-f}{f}} \times \sigma_{PZ}
-$$
-where now the ``data defect'' and ``problem difficulty'' are with respect to $PZ$ rather than $Y$. The final term is equal to
-$$
-\begin{aligned}
-E_J [P_J Z_J ] &= E_J [ E_J [ P_J Z_J \mid Y_J ] ] \\
-&= \pr (P = 1 \mid Y = 0) (1-\bar Y) - \pr(P=1 \mid Y = 1) \bar Y \\
-&= FP - (FP + FN) \cdot \bar Y
-\end{aligned}
-$$
-Combining these yields:
-$$
-\bar y_n^\star - \bar Y = \sqrt{\frac{1-f}{f}} \left(\rho_{I,Y} \sigma_Y + \rho_{I, PZ} \sigma_{PZ} \right) + \left( FP - (FP+FN) \bar Y \right)
-$$
-
-\subsubsection{Derivation of an estimator unbiased under SRS}
-\label{app:memestimator}
-We see the final term is given by $FP (1-\bar Y) - FN \bar Y$ is the bias associated with using the unadjusted prevalence estimate $\bar y_n^\star$.  This motivates an adjusted estimate
-$$
-\tilde y_n^{(0)} = \bar y_n^\star - FP (1- \bar y_n^\star ) + FN \bar y_n^\star
-= \bar y_n^\star (1 + FN + FP) - FP.
-$$
-Now considering the error for the adjusted estimate, $\tilde y_n^{(0)} - \bar Y$, we have
-$$
-\begin{aligned}
- &\bar y_n^\star (1 + FN + FP) - FP - \bar Y \\
- =&(\bar y_n^\star - \bar Y) +  (FN + FP) \bar y_n^\star - FP  \\
- =&\underbrace{\sqrt{\frac{1-f}{f}} \left[\rho_{I,Y} \sigma_Y + \rho_{I, PZ} \sigma_{PZ} \right]}_{\Psi} + (  FN + FP ) (\bar y_n^\star -  \bar Y) \\
- =& \Psi + (FN + FP) \Psi + (FN + FP) (FP - (FP+FN) \bar Y).
- \end{aligned}
-$$
-The final term is a (smaller) bias term and so we propose another adjusted estimator $\tilde y_n^{(1)} = \tilde y_n^{(0)} + (FN+FP)  ( (FN+FP) \bar y_n^\star - FP)$, with associated error $\tilde y_n^{(1)} - \bar Y$ given by
-$$
-\begin{aligned}
- &(\bar y_n^{(0)} - \bar Y) + (FN+FP)  ( (FN+FP) \bar y_n^\star - FP)\\
- =&\Psi + (FN + FP) \Psi + (FN + FP) (FP - (FP+FN) \bar Y) + (FN + FP) ((FP+FN) \bar y_n^\star - FP)  \\
-  =& \Psi + (FN + FP) \Psi + (FN + FP)^2 \Psi + (FN+FP)^2 (FP - (FP+FN) \bar Y).
- \end{aligned}
-$$
-This motivates recursively defining estimators $\tilde y_n^{(t)} = \tilde y_n^{(t-1)} + (FN+FP)  ( (FN+FP) \bar y_n - FP)$ for $t=1,2,\ldots$ where $\tilde y_n^{(0)} = \bar y_n^\star$.  Then
-$$
-\tilde y_n^{(t)} = \bar y_n^\star \sum_{s=0}^{t+1} (FP+FN)^s - FP \sum_{s=0}^{t} (FP+FN)^s
-$$
-and the associated error at iteration $t$ given by
-$$
-\Psi \sum_{s=0}^t (FN+FP)^s = \Psi \frac{1 - (FN+FP)^t}{1 - (FN+FP)}.
-$$
-We can then get an estimator with no residual bias term by taking the limit as $t$ goes to infinity; that is, define
-$$
-\tilde y_n = \lim_{t \to \infty} \tilde y_n^{(t)} = \frac{\bar y_n^\star - FP}{1 -(FN+FP)}.
-$$
-Then the associated error $\tilde y_n - \bar Y$ can be expressed as $\frac{\Psi}{1-(FN+FP)}$.
-
-\subsection{Model-based derivation of the estimator.}
-\label{app:modelbased}
-The estimator $\tilde y$ was derived as a limit of a process that removes the residual bias term at each step.
+% % $$
+% % \begin{aligned}
+% % \bar y_n^\star  - \bar Y_N &= \frac{E_J \left[ I_J \left[ Y_J (1-P_J) + (1-Y_J) P_J \right] \right]}{E_J [ I_J ] } - E_J[Y_J] \\
+% % &= \frac{E_J \left[ I_J P_J (1-2Y_J) \right]}{E_J [ I_J ] } + \left( \frac{E_J [I_J Y_J]}{E_J [ I_J ] } - \frac{E_J[Y_J] E_J[I_J]}{E_J[I_J]} \right) \\
+% % \end{aligned}
+% % $$
+% % The term in parentheses can be re-written as
+% % $$
+% % \begin{aligned}
+% % \frac{E_J [I_J Y_J]- E_J[Y_J] E_J[I_J]}{E_J[I_J]} &=  \frac{E_J [I_J Y_J]- E_J[Y_J] E_J[I_J]}{\sqrt{V_J(I_J) V_J(Y_J)}} \frac{\sqrt{V_J(I_J)}}{E_J[I_J]} \times \sqrt{V_J(Y_J)} \\
+% % &= \rho_{I,Y} \times \sqrt{\frac{(1-f)}{f}} \times \sigma_Y
+% % \end{aligned}
+% % $$
+% % which agrees with Meng's (2019) decomposition. For the other term, first we define $Z_j := 1 - 2 Y_j $. Then $Z_j = 1$ if $Y_j = 0$ and $Z_j = -1$ if $Y_j = 1$. Then the term can be re-written as
+% % $$
+% % \begin{aligned}
+% % \frac{E_J \left[ I_J P_J (1-2Y_J) \right]}{E_J [ I_J ] } &= \left( \frac{E_J \left[ I_J P_J Z_J \right]}{E_J [ I_J ] } -  \frac{E_J \left[ P_J Z_J \right] E_J[ I_J]}{E_J [ I_J ] } \right) +  \frac{E_J \left[ P_J Z_J \right] E_J[ I_J]}{E_J [ I_J ] } \\
+% % \end{aligned}
+% % $$
+% % The term in parentheses can be re-expressed using the previous technique as:
+% % $$
+% % \rho_{I, PZ} \times \sqrt{\frac{1-f}{f}} \times \sigma_{PZ}
+% % $$
+% % where now the ``data defect'' and ``problem difficulty'' are with respect to $PZ$ rather than $Y$. The final term is equal to
+% % $$
+% % \begin{aligned}
+% % E_J [P_J Z_J ] &= E_J [ E_J [ P_J Z_J \mid Y_J ] ] \\
+% % &= \pr (P = 1 \mid Y = 0) (1-\bar Y) - \pr(P=1 \mid Y = 1) \bar Y \\
+% % &= FP - (FP + FN) \cdot \bar Y
+% % \end{aligned}
+% % $$
+% % Combining these yields:
+% % $$
+% % \bar y_n^\star - \bar Y = \sqrt{\frac{1-f}{f}} \left(\rho_{I,Y} \sigma_Y + \rho_{I, PZ} \sigma_{PZ} \right) + \left( FP - (FP+FN) \bar Y \right)
+% % $$
+
+% % \subsubsection{Derivation of an estimator unbiased under SRS}
+% % \label{app:memestimator}
+% % We see the final term is given by $FP (1-\bar Y) - FN \bar Y$ is the bias associated with using the unadjusted prevalence estimate $\bar y_n^\star$.  This motivates an adjusted estimate
+% % $$
+% % \tilde y_n^{(0)} = \bar y_n^\star - FP (1- \bar y_n^\star ) + FN \bar y_n^\star
+% % = \bar y_n^\star (1 + FN + FP) - FP.
+% % $$
+% % Now considering the error for the adjusted estimate, $\tilde y_n^{(0)} - \bar Y$, we have
+% % $$
+% % \begin{aligned}
+% %  &\bar y_n^\star (1 + FN + FP) - FP - \bar Y \\
+% %  =&(\bar y_n^\star - \bar Y) +  (FN + FP) \bar y_n^\star - FP  \\
+% %  =&\underbrace{\sqrt{\frac{1-f}{f}} \left[\rho_{I,Y} \sigma_Y + \rho_{I, PZ} \sigma_{PZ} \right]}_{\Psi} + (  FN + FP ) (\bar y_n^\star -  \bar Y) \\
+% %  =& \Psi + (FN + FP) \Psi + (FN + FP) (FP - (FP+FN) \bar Y).
+% %  \end{aligned}
+% % $$
+% % The final term is a (smaller) bias term and so we propose another adjusted estimator $\tilde y_n^{(1)} = \tilde y_n^{(0)} + (FN+FP)  ( (FN+FP) \bar y_n^\star - FP)$, with associated error $\tilde y_n^{(1)} - \bar Y$ given by
+% % $$
+% % \begin{aligned}
+% %  &(\bar y_n^{(0)} - \bar Y) + (FN+FP)  ( (FN+FP) \bar y_n^\star - FP)\\
+% %  =&\Psi + (FN + FP) \Psi + (FN + FP) (FP - (FP+FN) \bar Y) + (FN + FP) ((FP+FN) \bar y_n^\star - FP)  \\
+% %   =& \Psi + (FN + FP) \Psi + (FN + FP)^2 \Psi + (FN+FP)^2 (FP - (FP+FN) \bar Y).
+% %  \end{aligned}
+% % $$
+% % This motivates recursively defining estimators $\tilde y_n^{(t)} = \tilde y_n^{(t-1)} + (FN+FP)  ( (FN+FP) \bar y_n - FP)$ for $t=1,2,\ldots$ where $\tilde y_n^{(0)} = \bar y_n^\star$.  Then
+% % $$
+% % \tilde y_n^{(t)} = \bar y_n^\star \sum_{s=0}^{t+1} (FP+FN)^s - FP \sum_{s=0}^{t} (FP+FN)^s
+% % $$
+% % and the associated error at iteration $t$ given by
+% % $$
+% % \Psi \sum_{s=0}^t (FN+FP)^s = \Psi \frac{1 - (FN+FP)^t}{1 - (FN+FP)}.
+% % $$
+% % We can then get an estimator with no residual bias term by taking the limit as $t$ goes to infinity; that is, define
+% % $$
+% % \tilde y_n = \lim_{t \to \infty} \tilde y_n^{(t)} = \frac{\bar y_n^\star - FP}{1 -(FN+FP)}.
+% % $$
+% % Then the associated error $\tilde y_n - \bar Y$ can be expressed as $\frac{\Psi}{1-(FN+FP)}$.
+
+% % \subsection{Model-based derivation of the estimator.}
+% % \label{app:modelbased}
+% % The estimator $\tilde y$ was derived as a limit of a process that removes the residual bias term at each step.
 Here we consider a model-based explanation.
@@
 Given a known false negative (FN) and false positive (FP) rates, we have
-$$
-\begin{aligned}
-\phi &= \theta \cdot (1-FN) + (1-\theta) \cdot FP = \theta (1-FN-FP) + FP \\
-\Rightarrow \theta &= \frac{\phi - FP}{1-FN-FP}.
-\end{aligned}
-$$
-Thus, the estimator $\tilde y$ is also the appropriate estimator under a model-based approach.
+% % $$
+% % \begin{aligned}
+% % \phi &= \theta \cdot (1-FN) + (1-\theta) \cdot FP = \theta (1-FN-FP) + FP \\
+% % \Rightarrow \theta &= \frac{\phi - FP}{1-FN-FP}.
+% % \end{aligned}
+% % $$
+% % Thus, the estimator $\tilde y$ is also the appropriate estimator under a model-based approach.
 While the derivation here is more straightforward, the derivation in the prior section provides a simple formula for the associated error $\tilde y_n - \bar Y$ and gives a novel connection between the empirical estimator $\bar y_n^\star$ and the adjusted estimator $\tilde y_n$ without reference to the model-based approach.
 
-\subsection{Further simplification.}
-For the binary outcome $Y$, we have $\sigma_Y = \sqrt{\bar Y (1-\bar Y)}$. Moreover,
-$$
-\begin{aligned}
-V_J(P_J Z_J) &= E_J[(P_J Z_J)^2] - E_J[P_J] E_J[Z_J] \\
-&= E_J[P_J] - E_J[P_J] (1 - 2 \bar Y) = 2 \bar Y E_J [ P_J ] \\
-&= 2 \bar Y \left( FP (1-\bar Y) + FN \bar Y \right) \\
-\Rightarrow \sigma_{PZ} &= \sqrt{ 2 \bar Y \left( FP (1-\bar Y) + FN \cdot  \bar Y \right) }
-\end{aligned}
-$$
-Then the formula for the error is given by:
-\begin{equation}
-\label{eq:finalstep}
-\sqrt{\frac{1-f}{f}} \left[\rho_{I,Y} \sqrt{\bar Y (1-\bar Y)} + \rho_{I, PZ} \sqrt{ 2 \bar Y \left( FP (1-\bar Y) + FN \cdot \bar Y \right )} \right] \times \frac{1}{1- (FN+FP)}
-\end{equation}
-By definition, we have
-$$
-\begin{aligned}
-\rho_{I,PZ} &= \frac{C(I, PZ)}{\sqrt{V(PZ) V(I)}} \\
-&= \frac{C(I, PZ)}{\sqrt{V(Y) V(I)}} \sqrt{\frac{V(Y)}{V(PZ)}} \\
-&= \rho_{I,Y} \frac{C(I,PZ)}{C(I,Y)} \sqrt{ \frac{(1-\bar Y)}{2 ( FP (1-\bar Y) + FN \cdot \bar Y)} }
-\end{aligned}
-$$
-
-$$
-\begin{aligned}
-C(I, PZ) &= E[ I P Z ] - E[I] E[PZ] \\
-&=  [FP f_0 - (FP f_0 + FN f_1) \bar Y] - f [ FP - (FP+FN) \bar Y ] \\
-&=  - FP \Delta \bar Y + FP \bar Y^2 \Delta - FN \bar Y^2 \Delta \\
-&=  - \Delta \bar Y (FP \cdot (1-\bar Y) + FN \cdot \bar Y) \\
-\end{aligned}
-$$
-where $f = f_1 \bar Y + f_0 (1-\bar Y)$ so $f_0 - f = -\Delta \bar Y$ and $f_1 - f = \Delta (1-\bar Y)$.
-$$
-\begin{aligned}
-C(I, Y) &= E[ I Y ] - f \bar Y \\
-&=  f_1 \bar Y + f_0 (1-\bar Y) - f \bar Y \\
-&=  f_0 (1-\bar Y) + \Delta (1-\bar Y) \bar Y \\
-&= (1-\bar Y) (f_0 + \Delta \bar Y)
-\end{aligned}
-$$
-Combining yields
-$$
-\begin{aligned}
-\rho_{I,PZ} &= \rho_{I,Y} \times \frac{- \Delta \bar Y (FP \cdot (1-\bar Y) + FN \cdot \bar Y) }{(1-\bar Y) (f_0 + \Delta \bar Y)} \times \sqrt{ \frac{(1-\bar Y)}{2 ( FP (1-\bar Y) + FN \cdot \bar Y)} } \\
-&= - \rho_{I, Y} \times \Delta \times \sqrt{\frac{\bar Y}{1-\bar Y}} \frac{\sqrt{FP(1-\bar Y) + FN \cdot \bar Y}}{f_0 (1-\bar Y) + f_1 \bar Y} \times \sqrt{\frac{\bar Y}{2}}
-\end{aligned}
-$$
-We can then re-write $\rho_{I,Y} \sigma_Y + \rho_{I,PZ} \sigma_{PZ}$ as
-$$
-\rho_{I,Y} \sigma_Y \left( 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right).
-$$
-Inserting into equation~\eqref{eq:finalstep} yields the desired result.
-
-\subsection{Derivation of effective sample size}
-\label{app:effss}
-
-Let $S_Y^2 = (N-1)^{-1} \sum_{j=1}^N (Y_j - \bar Y)^2$ be the population variance as defined in survey sampling~\citep{Cochran77}.
+% % \subsection{Further simplification.}
+% % For the binary outcome $Y$, we have $\sigma_Y = \sqrt{\bar Y (1-\bar Y)}$. Moreover,
+% % $$
+% % \begin{aligned}
+% % V_J(P_J Z_J) &= E_J[(P_J Z_J)^2] - E_J[P_J] E_J[Z_J] \\
+% % &= E_J[P_J] - E_J[P_J] (1 - 2 \bar Y) = 2 \bar Y E_J [ P_J ] \\
+% % &= 2 \bar Y \left( FP (1-\bar Y) + FN \bar Y \right) \\
+% % \Rightarrow \sigma_{PZ} &= \sqrt{ 2 \bar Y \left( FP (1-\bar Y) + FN \cdot  \bar Y \right) }
+% % \end{aligned}
+% % $$
+% % Then the formula for the error is given by:
+% % \begin{equation}
+% % \label{eq:finalstep}
+% % \sqrt{\frac{1-f}{f}} \left[\rho_{I,Y} \sqrt{\bar Y (1-\bar Y)} + \rho_{I, PZ} \sqrt{ 2 \bar Y \left( FP (1-\bar Y) + FN \cdot \bar Y \right )} \right] \times \frac{1}{1- (FN+FP)}
+% % \end{equation}
+% % By definition, we have
+% % $$
+% % \begin{aligned}
+% % \rho_{I,PZ} &= \frac{C(I, PZ)}{\sqrt{V(PZ) V(I)}} \\
+% % &= \frac{C(I, PZ)}{\sqrt{V(Y) V(I)}} \sqrt{\frac{V(Y)}{V(PZ)}} \\
+% % &= \rho_{I,Y} \frac{C(I,PZ)}{C(I,Y)} \sqrt{ \frac{(1-\bar Y)}{2 ( FP (1-\bar Y) + FN \cdot \bar Y)} }
+% % \end{aligned}
+% % $$
+
+% % $$
+% % \begin{aligned}
+% % C(I, PZ) &= E[ I P Z ] - E[I] E[PZ] \\
+% % &=  [FP f_0 - (FP f_0 + FN f_1) \bar Y] - f [ FP - (FP+FN) \bar Y ] \\
+% % &=  - FP \Delta \bar Y + FP \bar Y^2 \Delta - FN \bar Y^2 \Delta \\
+% % &=  - \Delta \bar Y (FP \cdot (1-\bar Y) + FN \cdot \bar Y) \\
+% % \end{aligned}
+% % $$
+% % where $f = f_1 \bar Y + f_0 (1-\bar Y)$ so $f_0 - f = -\Delta \bar Y$ and $f_1 - f = \Delta (1-\bar Y)$.
+% % $$
+% % \begin{aligned}
+% % C(I, Y) &= E[ I Y ] - f \bar Y \\
+% % &=  f_1 \bar Y + f_0 (1-\bar Y) - f \bar Y \\
+% % &=  f_0 (1-\bar Y) + \Delta (1-\bar Y) \bar Y \\
+% % &= (1-\bar Y) (f_0 + \Delta \bar Y)
+% % \end{aligned}
+% % $$
+% % Combining yields
+% % $$
+% % \begin{aligned}
+% % \rho_{I,PZ} &= \rho_{I,Y} \times \frac{- \Delta \bar Y (FP \cdot (1-\bar Y) + FN \cdot \bar Y) }{(1-\bar Y) (f_0 + \Delta \bar Y)} \times \sqrt{ \frac{(1-\bar Y)}{2 ( FP (1-\bar Y) + FN \cdot \bar Y)} } \\
+% % &= - \rho_{I, Y} \times \Delta \times \sqrt{\frac{\bar Y}{1-\bar Y}} \frac{\sqrt{FP(1-\bar Y) + FN \cdot \bar Y}}{f_0 (1-\bar Y) + f_1 \bar Y} \times \sqrt{\frac{\bar Y}{2}}
+% % \end{aligned}
+% % $$
+% % We can then re-write $\rho_{I,Y} \sigma_Y + \rho_{I,PZ} \sigma_{PZ}$ as
+% % $$
+% % \rho_{I,Y} \sigma_Y \left( 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right).
+% % $$
+% % Inserting into equation~\eqref{eq:finalstep} yields the desired result.
+
+% % \subsection{Derivation of effective sample size}
+% % \label{app:effss}
+
+% % Let $S_Y^2 = (N-1)^{-1} \sum_{j=1}^N (Y_j - \bar Y)^2$ be the population variance as defined in survey sampling~\citep{Cochran77}.
 Then $\sigma_Y^2 = (N-1)/N \cdot S_Y^2$.
@@
 Then setting the MSE under general selection and SRS equal we have
-$$
-\begin{aligned}
-\underbrace{\frac{1-f}{f} \times E_{\I} \left[ \rho_{I, Y}^2 \times D_M^2 \right]}_{1/n_{eff}^\star} \times \sigma_Y^2 &= \frac{1-f}{n_{eff}} S_Y^2 \\
-\frac{1}{n_{eff}^\star} \times \frac{N-1}{N} S_Y^2 &= \frac{1-f}{n_{eff}} S_Y^2 \\
-\frac{1}{n_{eff}^\star} &=  \left( \frac{1}{n_{eff}} - \frac{1}{N} \right) \left( \frac{N}{N-1} \right) \\
-\frac{1}{n_{eff}^\star} \left[ 1 - \frac{1}{N} + \frac{n_{eff}^\star}{N} \right]  &=  \frac{1}{n_{eff}} \\
-n_{eff}^\star \left[ 1 - \frac{1}{N} + \frac{n_{eff}^\star}{N} \right]^{-1}  &=  n_{eff} \\
-\frac{n_{eff}^\star}{ 1 + (n_{eff}^\star -1) N^{-1}} &=  n_{eff} \\
-\end{aligned}
-$$
-Then if $n_{eff}^\star \geq 1$, we have that
-$$
-n_{eff} \leq n_{eff}^\star = \frac{f}{1-f} \times \frac{1}{E_{\I} \left[ \rho^2_{I, Y} \times D_M^2 \right]}
-$$
-
-\subsection{Ratio estimator}
-\label{app:ratio}
-
-Let ${\bf u} = (u_{t-1},u_t) \in \mathbb{R}^2$ and $g({\bf u}) = \frac{u_t}{u_{t-1}}$, i.e., a differentiable function $g:\mathbb{R}^2 \to \mathbb{R}$.
+% % $$
+% % \begin{aligned}
+% % \underbrace{\frac{1-f}{f} \times E_{\I} \left[ \rho_{I, Y}^2 \times D_M^2 \right]}_{1/n_{eff}^\star} \times \sigma_Y^2 &= \frac{1-f}{n_{eff}} S_Y^2 \\
+% % \frac{1}{n_{eff}^\star} \times \frac{N-1}{N} S_Y^2 &= \frac{1-f}{n_{eff}} S_Y^2 \\
+% % \frac{1}{n_{eff}^\star} &=  \left( \frac{1}{n_{eff}} - \frac{1}{N} \right) \left( \frac{N}{N-1} \right) \\
+% % \frac{1}{n_{eff}^\star} \left[ 1 - \frac{1}{N} + \frac{n_{eff}^\star}{N} \right]  &=  \frac{1}{n_{eff}} \\
+% % n_{eff}^\star \left[ 1 - \frac{1}{N} + \frac{n_{eff}^\star}{N} \right]^{-1}  &=  n_{eff} \\
+% % \frac{n_{eff}^\star}{ 1 + (n_{eff}^\star -1) N^{-1}} &=  n_{eff} \\
+% % \end{aligned}
+% % $$
+% % Then if $n_{eff}^\star \geq 1$, we have that
+% % $$
+% % n_{eff} \leq n_{eff}^\star = \frac{f}{1-f} \times \frac{1}{E_{\I} \left[ \rho^2_{I, Y} \times D_M^2 \right]}
+% % $$
+
+% % \subsection{Ratio estimator}
+% % \label{app:ratio}
+
+% % Let ${\bf u} = (u_{t-1},u_t) \in \mathbb{R}^2$ and $g({\bf u}) = \frac{u_t}{u_{t-1}}$, i.e., a differentiable function $g:\mathbb{R}^2 \to \mathbb{R}$.
 Centering a Taylor series expansion of second-order around coordinates $(U_2, U_1) \in \mathbb{R}^2$ yields
-$$
-\begin{aligned}
-g({\bf u}) =& g(U_{t-1}, U_t) - \frac{U_t}{U_{t-1}^2} (u_{t-1} - U_{t-1}) + \frac{1}{U_{t-1}} (u_t - U_t) \\
-&+ \frac{1}{2} \left[ \frac{2 U_t}{U_{t-1}^3} (u_{t-1} - U_{t-1})^2 + 0 \times (u_t - U_t)^2 - 2 \times (u_{t-1} - U_{t-1}) (u_t - U_t) \frac{1}{U_t^2} \right]
-\end{aligned}
-$$
-Plugging in $(\bar y_{t-1}, \bar y_t)$ for $(u_{t-1}, u_t)$ and $(\bar Y_{t-1}, \bar Y_t)$ for $(U_{t-1}, U_t)$ yields the $\frac{\bar y_t}{\bar y_{t-1}} - \frac{\bar Y_t}{\bar Y_{t-1}} $ is equal to
-$$
-\begin{aligned}
-=&
-- \frac{\bar Y_t}{\bar Y_{t-1}^2} (\bar y_{t-1} - \bar Y_{t-1}) + \frac{1}{\bar Y_{t-1}} (\bar y_t - \bar Y_t) \\
-&+ \frac{\bar Y_t}{\bar Y_{t-1}^3} (\bar y_{t-1} - \bar Y_{t-1})^2 -  (\bar y_{t-1} - \bar Y_{t-1}) (\bar y_t - \bar Y_t) \frac{1}{\bar Y_{t-1}^2} \\
-&= \frac{\bar Y_t}{\bar Y_{t-1}} \bigg[  \rho_{I_t,Y_t} \sqrt{\frac{1-f_t}{f_t}} CV (Y_t)  -\rho_{I_{t-1},Y_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \\
-&+ \rho^2_{I_{t-1},Y_{t-1}} \frac{1-f_{t-1}}{f_{t-1}} CV^2 (Y_{t-1}) -  \rho_{I_{t-1},Y_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \times
-\rho_{I_t,Y_t} \sqrt{\frac{1-f_t}{f_t}} CV (Y_t)   \bigg] \\
-&= \frac{\bar Y_t}{\bar Y_{t-1}} \bigg[ \rho_{I_t,Y_t} \sqrt{\frac{1-f_t}{f_t}} CV (Y_t)  -\rho_{I_{t-1},Y_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \bigg] \left[ 1 - \rho_{I_{t-1},Y_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \right]
-\end{aligned}
-$$
-where the second equality is obtained by plugging in the statistical decomposition of the error for both time points and the coefficient of variation being defined as $CV(Y) := \sigma_Y/\mu_Y$.
+% % $$
+% % \begin{aligned}
+% % g({\bf u}) =& g(U_{t-1}, U_t) - \frac{U_t}{U_{t-1}^2} (u_{t-1} - U_{t-1}) + \frac{1}{U_{t-1}} (u_t - U_t) \\
+% % &+ \frac{1}{2} \left[ \frac{2 U_t}{U_{t-1}^3} (u_{t-1} - U_{t-1})^2 + 0 \times (u_t - U_t)^2 - 2 \times (u_{t-1} - U_{t-1}) (u_t - U_t) \frac{1}{U_t^2} \right]
+% % \end{aligned}
+% % $$
+% % Plugging in $(\bar y_{t-1}, \bar y_t)$ for $(u_{t-1}, u_t)$ and $(\bar Y_{t-1}, \bar Y_t)$ for $(U_{t-1}, U_t)$ yields the $\frac{\bar y_t}{\bar y_{t-1}} - \frac{\bar Y_t}{\bar Y_{t-1}} $ is equal to
+% % $$
+% % \begin{aligned}
+% % =&
+% % - \frac{\bar Y_t}{\bar Y_{t-1}^2} (\bar y_{t-1} - \bar Y_{t-1}) + \frac{1}{\bar Y_{t-1}} (\bar y_t - \bar Y_t) \\
+% % &+ \frac{\bar Y_t}{\bar Y_{t-1}^3} (\bar y_{t-1} - \bar Y_{t-1})^2 -  (\bar y_{t-1} - \bar Y_{t-1}) (\bar y_t - \bar Y_t) \frac{1}{\bar Y_{t-1}^2} \\
+% % &= \frac{\bar Y_t}{\bar Y_{t-1}} \bigg[  \rho_{I_t,Y_t} \sqrt{\frac{1-f_t}{f_t}} CV (Y_t)  -\rho_{I_{t-1},Y_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \\
+% % &+ \rho^2_{I_{t-1},Y_{t-1}} \frac{1-f_{t-1}}{f_{t-1}} CV^2 (Y_{t-1}) -  \rho_{I_{t-1},Y_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \times
+% % \rho_{I_t,Y_t} \sqrt{\frac{1-f_t}{f_t}} CV (Y_t)   \bigg] \\
+% % &= \frac{\bar Y_t}{\bar Y_{t-1}} \bigg[ \rho_{I_t,Y_t} \sqrt{\frac{1-f_t}{f_t}} CV (Y_t)  -\rho_{I_{t-1},Y_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \bigg] \left[ 1 - \rho_{I_{t-1},Y_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \right]
+% % \end{aligned}
+% % $$
+% % where the second equality is obtained by plugging in the statistical decomposition of the error for both time points and the coefficient of variation being defined as $CV(Y) := \sigma_Y/\mu_Y$.
 Under measurement error, the extra terms $D_{t}$ and $D_{t-1}$ can be inserted in the correct locations.
 
-\subsection{Estimation of effective reproduction number}
-
-Let
-$$
-\delta_t := \bigg[ \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t)  -\rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_t}{f_t}} CV (K_{t-1}) \bigg].
-$$
-Then the previous sections derivation shows that the estimate of the number of new cases on day t is given by
-$$
-\frac{S_t \cdot \bar y_t}{S_{t-1} \cdot \bar y_{t-1}} =
-\frac{K_t}{K_{t-1}} \left( 1 + \delta_t \times \left[ 1 - \rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_t}{f_t}} CV (K_{t-1}) \right] \right)
-$$
-Then setting $e_t = \delta_t \times [1 - \rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_t}{f_t}} CV (K_{t-1}) ]$, we have
-$$
-\begin{aligned}
-\log \left( \frac{S_t \bar y_t}{S_{t-1} \bar y_{t-1}} \right) - \log \left( \frac{K_t}{K_{t-1}} \right) &= \log (1 + e_t) \\
-\log \left( \frac{\bar y_t}{\bar y_{t-1}} \right) - \log \left( \frac{K_t}{K_{t-1}} \right) &= 1 + e_t - \log \left( \frac{S_t}{S_{t-1}} \right) \\
-1 + \frac{1}{\gamma} \log \left( \frac{\bar y_t}{\bar y_{t-1}} \right) - \left[ 1 + \frac{1}{\gamma} \log \left( \frac{K_t}{K_{t-1}} \right) \right] &= \frac{1}{\gamma} \left[ \log \left( 1 + e_t \right) - \log \left( \frac{S_{t}}{S_{t-1}} \right) \right] \\
-\Rightarrow \hat R_t - R_t &= \frac{1}{\gamma} \left[ \log \left( 1 + e_t \right) - \log \left( \frac{S_{t}}{S_{t-1}} \right) \right]
-\end{aligned}
-$$
-
-\subsection{Computing the effective sample size}
-\label{section:effss}
-
-For binary outcomes, we have
-\begin{equation} \label{eq:binaryrho}
-\rho_{I,Y} = \Delta \sqrt{\frac{\bar Y (1 - \bar Y)}{f (1-f)} }
-\end{equation}
-where $\Delta = P_J (I_J = 1 \mid Y_J = 1) - P(I_J = 1 \mid Y_J = 0) = f_1 - f_0$.
+% % \subsection{Estimation of effective reproduction number}
+
+% % Let
+% % $$
+% % \delta_t := \bigg[ \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t)  -\rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_t}{f_t}} CV (K_{t-1}) \bigg].
+% % $$
+% % Then the previous sections derivation shows that the estimate of the number of new cases on day t is given by
+% % $$
+% % \frac{S_t \cdot \bar y_t}{S_{t-1} \cdot \bar y_{t-1}} =
+% % \frac{K_t}{K_{t-1}} \left( 1 + \delta_t \times \left[ 1 - \rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_t}{f_t}} CV (K_{t-1}) \right] \right)
+% % $$
+% % Then setting $e_t = \delta_t \times [1 - \rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_t}{f_t}} CV (K_{t-1}) ]$, we have
+% % $$
+% % \begin{aligned}
+% % \log \left( \frac{S_t \bar y_t}{S_{t-1} \bar y_{t-1}} \right) - \log \left( \frac{K_t}{K_{t-1}} \right) &= \log (1 + e_t) \\
+% % \log \left( \frac{\bar y_t}{\bar y_{t-1}} \right) - \log \left( \frac{K_t}{K_{t-1}} \right) &= 1 + e_t - \log \left( \frac{S_t}{S_{t-1}} \right) \\
+% % 1 + \frac{1}{\gamma} \log \left( \frac{\bar y_t}{\bar y_{t-1}} \right) - \left[ 1 + \frac{1}{\gamma} \log \left( \frac{K_t}{K_{t-1}} \right) \right] &= \frac{1}{\gamma} \left[ \log \left( 1 + e_t \right) - \log \left( \frac{S_{t}}{S_{t-1}} \right) \right] \\
+% % \Rightarrow \hat R_t - R_t &= \frac{1}{\gamma} \left[ \log \left( 1 + e_t \right) - \log \left( \frac{S_{t}}{S_{t-1}} \right) \right]
+% % \end{aligned}
+% % $$
+
+% % \subsection{Computing the effective sample size}
+% % \label{section:effss}
+
+% % For binary outcomes, we have
+% % \begin{equation} \label{eq:binaryrho}
+% % \rho_{I,Y} = \Delta \sqrt{\frac{\bar Y (1 - \bar Y)}{f (1-f)} }
+% % \end{equation}
+% % where $\Delta = P_J (I_J = 1 \mid Y_J = 1) - P(I_J = 1 \mid Y_J = 0) = f_1 - f_0$.
 Suppose that $M = f_1/f_0$; then $f_0 = f / (\bar Y \cdot (M-1) + 1)$.
@@
 
-\begin{table}[ht]
-\centering
-\begin{tabular}{rrrrrrrr}
-& \multicolumn{7}{c}{$M$} \\ \cline{2-8}
-$\bar y$ & 1.05 & 1.15 & 1.25 & 1.35 & 1.45 & 1.55 & 1.65 \\
-  \hline
-0.01 & 40444 & 4503 & 1624 & 830 & 503 & 338 & 242 \\
-0.03 & 13787 & 1541 & 558 & 286 & 174 & 117 & 85 \\
-0.05 & 8463 & 950 & 345 & 178 & 109 & 73 & 53 \\
-0.07 & 6187 & 697 & 254 & 132 & 81 & 55 & 40 \\
-0.09 & 4928 & 557 & 204 & 106 & 65 & 44 & 32 \\
-0.11 & 4131 & 469 & 173 & 90 & 56 & 38 & 28 \\
-   \hline
-\end{tabular}
-\end{table}
-
-We also present the same plot under $FP = 0.024$ and $FN = 0.13$ to show the impact of measurement error on effective sample size.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{rrrrrrrr}
-& \multicolumn{7}{c}{$M$} \\ \cline{2-8}
- & 1.05 & 1.15 & 1.25 & 1.35 & 1.45 & 1.55 & 1.65 \\
-  \hline
-0.01 & 29019 & 3247 & 1177 & 605 & 368 & 248 & 179 \\
-0.03 & 9894 & 1112 & 405 & 209 & 128 & 87 & 63 \\
-0.05 & 6075 & 686 & 251 & 130 & 80 & 54 & 39 \\
-0.07 & 4442 & 504 & 185 & 96 & 59 & 41 & 30 \\
-0.09 & 3539 & 403 & 149 & 78 & 48 & 33 & 24 \\
-0.11 & 2967 & 339 & 126 & 66 & 41 & 28 & 21 \\
-   \hline
-\end{tabular}
-\end{table}
-
-For binary outcomes, we have
-\begin{equation} \label{eq:weightedbinaryrho}
-\rho_{\tilde I (X),Y} = \tilde \Delta \sqrt{\frac{\bar Y (1 - \bar Y)}{f (1-f) \text{E}(W_J \mid I_J = 1)^2 + f \text{Var}(W_J \mid I_J = 1)} }
-\end{equation}
-
-% \section{Effect modifiers in COVID-19 clinical trials}
-
-% Here we show how selection bias and measurement error can creep into clinical trial analysis.
+% % \begin{table}[ht]
+% % \centering
+% % \begin{tabular}{rrrrrrrr}
+% % & \multicolumn{7}{c}{$M$} \\ \cline{2-8}
+% % $\bar y$ & 1.05 & 1.15 & 1.25 & 1.35 & 1.45 & 1.55 & 1.65 \\
+% %   \hline
+% % 0.01 & 40444 & 4503 & 1624 & 830 & 503 & 338 & 242 \\
+% % 0.03 & 13787 & 1541 & 558 & 286 & 174 & 117 & 85 \\
+% % 0.05 & 8463 & 950 & 345 & 178 & 109 & 73 & 53 \\
+% % 0.07 & 6187 & 697 & 254 & 132 & 81 & 55 & 40 \\
+% % 0.09 & 4928 & 557 & 204 & 106 & 65 & 44 & 32 \\
+% % 0.11 & 4131 & 469 & 173 & 90 & 56 & 38 & 28 \\
+% %    \hline
+% % \end{tabular}
+% % \end{table}
+
+% % We also present the same plot under $FP = 0.024$ and $FN = 0.13$ to show the impact of measurement error on effective sample size.
+
+% % \begin{table}[ht]
+% % \centering
+% % \begin{tabular}{rrrrrrrr}
+% % & \multicolumn{7}{c}{$M$} \\ \cline{2-8}
+% %  & 1.05 & 1.15 & 1.25 & 1.35 & 1.45 & 1.55 & 1.65 \\
+% %   \hline
+% % 0.01 & 29019 & 3247 & 1177 & 605 & 368 & 248 & 179 \\
+% % 0.03 & 9894 & 1112 & 405 & 209 & 128 & 87 & 63 \\
+% % 0.05 & 6075 & 686 & 251 & 130 & 80 & 54 & 39 \\
+% % 0.07 & 4442 & 504 & 185 & 96 & 59 & 41 & 30 \\
+% % 0.09 & 3539 & 403 & 149 & 78 & 48 & 33 & 24 \\
+% % 0.11 & 2967 & 339 & 126 & 66 & 41 & 28 & 21 \\
+% %    \hline
+% % \end{tabular}
+% % \end{table}
+
+% % For binary outcomes, we have
+% % \begin{equation} \label{eq:weightedbinaryrho}
+% % \rho_{\tilde I (X),Y} = \tilde \Delta \sqrt{\frac{\bar Y (1 - \bar Y)}{f (1-f) \text{E}(W_J \mid I_J = 1)^2 + f \text{Var}(W_J \mid I_J = 1)} }
+% % \end{equation}
+
+% % % \section{Effect modifiers in COVID-19 clinical trials}
+
+% % % Here we show how selection bias and measurement error can creep into clinical trial analysis.
 The key concern is whether clinical trials on COVID-19 recruit from the pool of individuals who have tested positive for COVID-19 or whether they sample randomly from the population, test, and then recruit from this subset of tested individuals.
@@
 In counterfactual language,
-% $$
-% E( Y(A=1) - Y(A=0) ) = E_U ( E_A ( Y(A=1) - Y(A=0) \mid U=u ))
-% = \beta_1 + \beta_3 E(U).
-% $$
-% Even in a randomized experiment, the question is what the correct value for $E(U)$ is.
+% % % $$
+% % % E( Y(A=1) - Y(A=0) ) = E_U ( E_A ( Y(A=1) - Y(A=0) \mid U=u ))
+% % % = \beta_1 + \beta_3 E(U).
+% % % $$
+% % % Even in a randomized experiment, the question is what the correct value for $E(U)$ is.
 For COVID-19, the value of interest is likely the expected value of $U$ in the population of COVID-19 positive individuals, i.e., $E_J [ U_J Y_J]/ E_J[Y_J] = \sum_{j=1}^N U_j Y_j / \sum_{j=1}^N Y_j$.
@@
 Due to selection bias, the bias in the marginal treatment effect compared to the true marginal effect of interest is approximately equal to
-% $$
-% \beta_3 \times \sqrt{\frac{1-f}{f}} \times \frac{\bar U^\prime}{\bar Y} \bigg[ \rho_{I,U^\prime} \times CV (U^\prime)  - \rho_{I,Y} \times CV (Y) \bigg] \left[ 1 - \rho_{I,Y} \sqrt{\frac{1-f}{f}} CV (Y) \right]
-% $$
-% where $U^\prime = U Y$.
+% % % $$
+% % % \beta_3 \times \sqrt{\frac{1-f}{f}} \times \frac{\bar U^\prime}{\bar Y} \bigg[ \rho_{I,U^\prime} \times CV (U^\prime)  - \rho_{I,Y} \times CV (Y) \bigg] \left[ 1 - \rho_{I,Y} \sqrt{\frac{1-f}{f}} CV (Y) \right]
+% % % $$
+% % % where $U^\prime = U Y$.
 The derivation again uses a second-order Taylor series approximation and follows exactly the same as in Appendix~\ref{app:ratio}.
@@
 
-% Randomization of treatment assignment in clinical trials negates unobserved confounders.
+% % % Randomization of treatment assignment in clinical trials negates unobserved confounders.
 This, however, does not negate effect modifiers.
@@
 
-\section{IPW statistical error decomposition derivation}
-\label{app:ipwderivation}
-We start by considering the empirical weighted mean estimator under imperfect testing,
-$$
-\bar y_n^\star = \frac{\sum_{j=1}^N W_j Y_j^\star I_j}{\sum_{j=1}^N W_j I_j} = \frac{\sum_{i=1}^N  I_j W_j Y_j^\star }{\sum_{j=1}^N  I_j } = \frac{\sum_{i=1}^N  I_j \left[ Y_j (1-P_j) + (1-Y_j) P_j \right]}{\sum_{j=1}^N  I_j }
-$$
-Then
-$$
-\begin{aligned}
-\bar y_n^\star  - \bar Y_N &= \frac{E_J \left[ I_J W_J \left[ Y_J (1-P_J) + (1-Y_J) P_J \right] \right]}{E_J [ I_J W_J ] } - E_J[Y_J] \\
-&= \frac{E_J \left[ I_J W_J P_J (1-2Y_J) \right]}{E_J [ I_J W_J ] } + \left( \frac{E_J [I_J W_J Y_J]}{E_J [ I_J W_J ] } - \frac{E_J[Y_J] E_J[I_J W_J]}{E_J[I_J W_J]} \right) \\
-\end{aligned}
-$$
-The term in parentheses can be re-written as
-$$
-\begin{aligned}
-\frac{E_J [I_J W_J Y_J]- E_J[Y_J] E_J[I_J W_J]}{E_J[I_J W_J]} &=  \frac{E_J [I_J W_J Y_J]- E_J[Y_J] E_J[I_J W_J]}{\sqrt{V_J(I_J W_J) V_J(Y_J)}} \frac{\sqrt{V_J(I_J W_J)}}{E_J[I_J]} \times \sqrt{V_J(Y_J)} \\
-&= \rho_{\tilde I (X),Y} \times \frac{\sqrt{V_J(I_J W_J)}}{E_J[I_J]} \times \sigma_Y
-\end{aligned}
-$$
-Then
-\begin{align*}
-E_J (I_J W_J) &= P(I_J = 1) E_J [W_J \mid I_J = 1] = f \times E_J [W_J \mid I_J = 1] \\
-V_J(I_J W_J) &= E[ I_J W_J^2] - E[I_J W_J]^2 \\
-&= f \cdot \left( E [W_J^2 \mid I_J = 1] - f E[W_J \mid I_J = 1 ]^2 \right) \\
-&= f \cdot \left( E [W_J^2 \mid I_J = 1] \pm E[W_J \mid I_j = 1]^2 - f E[W_J \mid I_J = 1 ]^2 \right) \\
-&= f \cdot \left( V (W_J \mid I_J = 1) + E[W_J \mid I_J = 1]^2 (1-f) \right)
-\end{align*}
-Taking the ratio:
-$$
-\sqrt{\frac{f \cdot \left( V (W_J \mid I_J = 1) + E[W_J \mid I_J = 1]^2 (1-f) \right)}{f^2 E_J [W_J \mid I_J = 1]^2}} = \sqrt{\frac{1 - f + CV(W)^2}{f}}
-$$
-which agrees with Meng's (2019) decomposition of weighted outcome.
+% % \section{IPW statistical error decomposition derivation}
+% % \label{app:ipwderivation}
+% % We start by considering the empirical weighted mean estimator under imperfect testing,
+% % $$
+% % \bar y_n^\star = \frac{\sum_{j=1}^N W_j Y_j^\star I_j}{\sum_{j=1}^N W_j I_j} = \frac{\sum_{i=1}^N  I_j W_j Y_j^\star }{\sum_{j=1}^N  I_j } = \frac{\sum_{i=1}^N  I_j \left[ Y_j (1-P_j) + (1-Y_j) P_j \right]}{\sum_{j=1}^N  I_j }
+% % $$
+% % Then
+% % $$
+% % \begin{aligned}
+% % \bar y_n^\star  - \bar Y_N &= \frac{E_J \left[ I_J W_J \left[ Y_J (1-P_J) + (1-Y_J) P_J \right] \right]}{E_J [ I_J W_J ] } - E_J[Y_J] \\
+% % &= \frac{E_J \left[ I_J W_J P_J (1-2Y_J) \right]}{E_J [ I_J W_J ] } + \left( \frac{E_J [I_J W_J Y_J]}{E_J [ I_J W_J ] } - \frac{E_J[Y_J] E_J[I_J W_J]}{E_J[I_J W_J]} \right) \\
+% % \end{aligned}
+% % $$
+% % The term in parentheses can be re-written as
+% % $$
+% % \begin{aligned}
+% % \frac{E_J [I_J W_J Y_J]- E_J[Y_J] E_J[I_J W_J]}{E_J[I_J W_J]} &=  \frac{E_J [I_J W_J Y_J]- E_J[Y_J] E_J[I_J W_J]}{\sqrt{V_J(I_J W_J) V_J(Y_J)}} \frac{\sqrt{V_J(I_J W_J)}}{E_J[I_J]} \times \sqrt{V_J(Y_J)} \\
+% % &= \rho_{\tilde I (X),Y} \times \frac{\sqrt{V_J(I_J W_J)}}{E_J[I_J]} \times \sigma_Y
+% % \end{aligned}
+% % $$
+% % Then
+% % \begin{align*}
+% % E_J (I_J W_J) &= P(I_J = 1) E_J [W_J \mid I_J = 1] = f \times E_J [W_J \mid I_J = 1] \\
+% % V_J(I_J W_J) &= E[ I_J W_J^2] - E[I_J W_J]^2 \\
+% % &= f \cdot \left( E [W_J^2 \mid I_J = 1] - f E[W_J \mid I_J = 1 ]^2 \right) \\
+% % &= f \cdot \left( E [W_J^2 \mid I_J = 1] \pm E[W_J \mid I_j = 1]^2 - f E[W_J \mid I_J = 1 ]^2 \right) \\
+% % &= f \cdot \left( V (W_J \mid I_J = 1) + E[W_J \mid I_J = 1]^2 (1-f) \right)
+% % \end{align*}
+% % Taking the ratio:
+% % $$
+% % \sqrt{\frac{f \cdot \left( V (W_J \mid I_J = 1) + E[W_J \mid I_J = 1]^2 (1-f) \right)}{f^2 E_J [W_J \mid I_J = 1]^2}} = \sqrt{\frac{1 - f + CV(W)^2}{f}}
+% % $$
+% % which agrees with Meng's (2019) decomposition of weighted outcome.
 For the other term, first we define $Z_j := 1 - 2 Y_j $.
@@
 Then the term can be re-written as
-$$
-\begin{aligned}
-\frac{E_J \left[ I_J W_J P_J (1-2Y_J) \right]}{E_J [ I_J W_J ] } &= \left( \frac{E_J \left[ I_J W_J P_J Z_J \right]}{E_J [ I_J W_J ] } -  \frac{E_J \left[ P_J Z_J \right] E_J[ I_J W_J]}{E_J [ I_J W_J ] } \right) +  \frac{E_J \left[ P_J Z_J \right] E_J[ I_J W_J]}{E_J [ I_J W_J ] } \\
-\end{aligned}
-$$
-The term in parentheses can be re-expressed using the previous technique as:
-$$
-\rho_{\tilde I, PZ} \times \sqrt{\frac{1-f+CV(W)^2}{f}} \times \sigma_{PZ}
-$$
-where now the ``data defect'' and ``problem difficulty'' are with respect to $PZ$ rather than $Y$. The final term is equal to
-$$
-\begin{aligned}
-E_J [P_J Z_J ] &= E_J [ E_J [ P_J Z_J \mid Y_J ] ] \\
-&= \pr (P = 1 \mid Y = 0) (1-\bar Y) - \pr(P=1 \mid Y = 1) \bar Y \\
-&= FP - (FP + FN) \cdot \bar Y
-\end{aligned}
-$$
-Combining these yields:
-$$
-\bar y_n^\star - \bar Y = \sqrt{\frac{1-f+CV(W)^2}{f}} \left(\rho_{\tilde I(X),Y} \sigma_Y + \rho_{\tilde I(X), PZ} \sigma_{PZ} \right) + \left( FP - (FP+FN) \bar Y \right)
-$$
-By previous arguments, we have
-$$
-\rho_{\tilde I (X), PZ} = \rho_{\tilde I(X), Y} \times \frac{C(\tilde I(X), PZ)}{C(\tilde I(X), Y)} \sqrt{ \frac{1-\bar Y}{ 2 (FP(1-\bar Y) + FN \bar Y)} }
-$$
-Then
-\begin{align*}
-C(IW, PZ) &= E[IWPZ] - E[IW] E[PZ] \\
-&=  [FP \tilde f_0  - (FP \tilde f_0 + FN \tilde f_1) \bar Y]  - \tilde f \cdot ( FP - (FP+FN) \bar Y )\\
-&= -FP \tilde \Delta \bar Y + FP \bar Y^2 \Delta - FN \bar Y^2 \Delta \\
-&= -\tilde \Delta \bar Y ( FP \cdot (1-\bar Y) + FN \cdot \bar Y)
-\end{align*}
-where $\tilde f_i = E[ I_J W_J \mid Y_J = i]$ for $i \in \{0,1 \}$ and $f = \tilde f_1 \bar Y + \tilde f_0 (1-\bar Y)$. Moreover,
-\begin{align*}
-C(\tilde I,Y) &= E[ I W Y ] - \tilde f \bar Y \\
-&= \tilde f_1 \bar Y + \tilde f_0 (1-\bar Y) - \tilde f \bar Y \\
-&= \tilde f_0 (1-\bar Y) + \tilde \Delta (1-\bar Y) \bar Y \\
-&= (1- \bar Y) (\tilde f_0 + \tilde \Delta \bar Y)
-\end{align*}
-Combining yields
-$$
-\begin{aligned}
-\rho_{\tilde I (X),PZ} &= \rho_{\tilde I(X),Y} \times \frac{- \tilde \Delta \bar Y (FP \cdot (1-\bar Y) + FN \cdot \bar Y) }{(1-\bar Y) (\tilde f_0 + \tilde \Delta \bar Y)} \times \sqrt{ \frac{(1-\bar Y)}{2 ( FP (1-\bar Y) + FN \cdot \bar Y)} } \\
-&= - \rho_{\tilde I(X), Y} \times \Delta \times \sqrt{\frac{\bar Y}{1-\bar Y}} \frac{\sqrt{FP(1-\bar Y) + FN \cdot \bar Y}}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \times \sqrt{\frac{\bar Y}{2}}
-\end{aligned}
-$$
-We can then re-write $\rho_{\tilde I (X),Y} \sigma_Y + \rho_{\tilde I(X),PZ} \sigma_{PZ}$ as
-$$
-\rho_{\tilde I(X),Y} \sigma_Y \left( 1 - \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right).
-$$
-Thus yielding the desired result.
-
-\section{Doubly robust statistical error decomposition derivation}
-\label{app:DRderivation}
-We start by considering the empirical weighted mean estimator under imperfect testing,
-\begin{align*}
-&\frac{1}{N} \sum_{j=1}^N \mu(X_j) - \frac{\sum_{j=1}^N I_j W_j \mu(X_j)}{\sum_{j=1}^N I_j W_j} + \frac{\sum_{j=1}^N I_j W_j Y_j^\star}{\sum_{j=1}^N I_j W_j} \\
-=& \frac{\sum_{i=1}^N  I_j W_j \left( \left[ Y_j (1-P_j) + (1-Y_j) P_j \right] - \mu (X_j) \right)}{\sum_{j=1}^N  I_j W_j } + \frac{1}{N} \sum_{j=1}^N \mu(X_j
-)
-\end{align*}
-Then
-$$
-\begin{aligned}
-&\frac{E_J \left[ I_J W_J \left( \left[ Y_J (1-P_J) + (1-Y_J) P_J \right] - \mu(X_J) \right) \right]}{E_J [ I_J W_J ] } - E_J[Y_J -\mu (X_J)] \\
-= &\frac{E_J \left[ I_J W_J P_J (1-2Y_J) \right]}{E_J [ I_J W_J ] } + \left( \frac{E_J [I_J W_J (Y_J -\mu (X_J))]}{E_J [ I_J W_J ] } - \frac{E_J[Y_J-\mu(X_J)] E_J[I_J W_J]}{E_J[I_J W_J]} \right) \\
-\end{aligned}
-$$
-The term in parentheses can be re-written as
-$$
-\begin{aligned}
-&\frac{E_J [I_J W_J (Y_J-\mu(X_J))]- E_J[(Y_J-\mu(X_J))] E_J[I_J W_J]}{E_J[I_J W_J]} \\
-=  &\frac{E_J [I_J W_J (Y_J-\mu(X_J))]- E_J[(Y_J-\mu(X_J))] E_J[I_J W_J]}{\sqrt{V_J(I_J W_J) V_J(Y_J-\mu(X_J))}} \frac{\sqrt{V_J(I_J W_J)}}{E_J[I_J]} \times \sqrt{V_J(Y_J-\mu(X_J))} \\
-= &\rho_{\tilde I (X),Y-\mu(X)} \times \frac{\sqrt{V_J(I_J W_J)}}{E_J[I_J]} \times \sigma_{Y-\mu(X)}
-\end{aligned}
-$$
-From prior derivations, we then have the desired result.
-
-\section{Asymptotic derivations}
-\label{app:asympderivations}
-
-% https://www.thelancet.com/journals/lanres/article/PIIS2213-2600%2820%2930453-7/fulltext
-% https://www.medrxiv.org/content/10.1101/2020.04.26.20080911v1.full.pdf
-
-Here, we take the suggested measures from~\cite{Arevalo2020} which report 87\% sensitivity is a reasonable estimates and~\cite{Cohen2020} which report 97.6\% specificity.
+% % $$
+% % \begin{aligned}
+% % \frac{E_J \left[ I_J W_J P_J (1-2Y_J) \right]}{E_J [ I_J W_J ] } &= \left( \frac{E_J \left[ I_J W_J P_J Z_J \right]}{E_J [ I_J W_J ] } -  \frac{E_J \left[ P_J Z_J \right] E_J[ I_J W_J]}{E_J [ I_J W_J ] } \right) +  \frac{E_J \left[ P_J Z_J \right] E_J[ I_J W_J]}{E_J [ I_J W_J ] } \\
+% % \end{aligned}
+% % $$
+% % The term in parentheses can be re-expressed using the previous technique as:
+% % $$
+% % \rho_{\tilde I, PZ} \times \sqrt{\frac{1-f+CV(W)^2}{f}} \times \sigma_{PZ}
+% % $$
+% % where now the ``data defect'' and ``problem difficulty'' are with respect to $PZ$ rather than $Y$. The final term is equal to
+% % $$
+% % \begin{aligned}
+% % E_J [P_J Z_J ] &= E_J [ E_J [ P_J Z_J \mid Y_J ] ] \\
+% % &= \pr (P = 1 \mid Y = 0) (1-\bar Y) - \pr(P=1 \mid Y = 1) \bar Y \\
+% % &= FP - (FP + FN) \cdot \bar Y
+% % \end{aligned}
+% % $$
+% % Combining these yields:
+% % $$
+% % \bar y_n^\star - \bar Y = \sqrt{\frac{1-f+CV(W)^2}{f}} \left(\rho_{\tilde I(X),Y} \sigma_Y + \rho_{\tilde I(X), PZ} \sigma_{PZ} \right) + \left( FP - (FP+FN) \bar Y \right)
+% % $$
+% % By previous arguments, we have
+% % $$
+% % \rho_{\tilde I (X), PZ} = \rho_{\tilde I(X), Y} \times \frac{C(\tilde I(X), PZ)}{C(\tilde I(X), Y)} \sqrt{ \frac{1-\bar Y}{ 2 (FP(1-\bar Y) + FN \bar Y)} }
+% % $$
+% % Then
+% % \begin{align*}
+% % C(IW, PZ) &= E[IWPZ] - E[IW] E[PZ] \\
+% % &=  [FP \tilde f_0  - (FP \tilde f_0 + FN \tilde f_1) \bar Y]  - \tilde f \cdot ( FP - (FP+FN) \bar Y )\\
+% % &= -FP \tilde \Delta \bar Y + FP \bar Y^2 \Delta - FN \bar Y^2 \Delta \\
+% % &= -\tilde \Delta \bar Y ( FP \cdot (1-\bar Y) + FN \cdot \bar Y)
+% % \end{align*}
+% % where $\tilde f_i = E[ I_J W_J \mid Y_J = i]$ for $i \in \{0,1 \}$ and $f = \tilde f_1 \bar Y + \tilde f_0 (1-\bar Y)$. Moreover,
+% % \begin{align*}
+% % C(\tilde I,Y) &= E[ I W Y ] - \tilde f \bar Y \\
+% % &= \tilde f_1 \bar Y + \tilde f_0 (1-\bar Y) - \tilde f \bar Y \\
+% % &= \tilde f_0 (1-\bar Y) + \tilde \Delta (1-\bar Y) \bar Y \\
+% % &= (1- \bar Y) (\tilde f_0 + \tilde \Delta \bar Y)
+% % \end{align*}
+% % Combining yields
+% % $$
+% % \begin{aligned}
+% % \rho_{\tilde I (X),PZ} &= \rho_{\tilde I(X),Y} \times \frac{- \tilde \Delta \bar Y (FP \cdot (1-\bar Y) + FN \cdot \bar Y) }{(1-\bar Y) (\tilde f_0 + \tilde \Delta \bar Y)} \times \sqrt{ \frac{(1-\bar Y)}{2 ( FP (1-\bar Y) + FN \cdot \bar Y)} } \\
+% % &= - \rho_{\tilde I(X), Y} \times \Delta \times \sqrt{\frac{\bar Y}{1-\bar Y}} \frac{\sqrt{FP(1-\bar Y) + FN \cdot \bar Y}}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \times \sqrt{\frac{\bar Y}{2}}
+% % \end{aligned}
+% % $$
+% % We can then re-write $\rho_{\tilde I (X),Y} \sigma_Y + \rho_{\tilde I(X),PZ} \sigma_{PZ}$ as
+% % $$
+% % \rho_{\tilde I(X),Y} \sigma_Y \left( 1 - \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right).
+% % $$
+% % Thus yielding the desired result.
+
+% % \section{Doubly robust statistical error decomposition derivation}
+% % \label{app:DRderivation}
+% % We start by considering the empirical weighted mean estimator under imperfect testing,
+% % \begin{align*}
+% % &\frac{1}{N} \sum_{j=1}^N \mu(X_j) - \frac{\sum_{j=1}^N I_j W_j \mu(X_j)}{\sum_{j=1}^N I_j W_j} + \frac{\sum_{j=1}^N I_j W_j Y_j^\star}{\sum_{j=1}^N I_j W_j} \\
+% % =& \frac{\sum_{i=1}^N  I_j W_j \left( \left[ Y_j (1-P_j) + (1-Y_j) P_j \right] - \mu (X_j) \right)}{\sum_{j=1}^N  I_j W_j } + \frac{1}{N} \sum_{j=1}^N \mu(X_j
+% % )
+% % \end{align*}
+% % Then
+% % $$
+% % \begin{aligned}
+% % &\frac{E_J \left[ I_J W_J \left( \left[ Y_J (1-P_J) + (1-Y_J) P_J \right] - \mu(X_J) \right) \right]}{E_J [ I_J W_J ] } - E_J[Y_J -\mu (X_J)] \\
+% % = &\frac{E_J \left[ I_J W_J P_J (1-2Y_J) \right]}{E_J [ I_J W_J ] } + \left( \frac{E_J [I_J W_J (Y_J -\mu (X_J))]}{E_J [ I_J W_J ] } - \frac{E_J[Y_J-\mu(X_J)] E_J[I_J W_J]}{E_J[I_J W_J]} \right) \\
+% % \end{aligned}
+% % $$
+% % The term in parentheses can be re-written as
+% % $$
+% % \begin{aligned}
+% % &\frac{E_J [I_J W_J (Y_J-\mu(X_J))]- E_J[(Y_J-\mu(X_J))] E_J[I_J W_J]}{E_J[I_J W_J]} \\
+% % =  &\frac{E_J [I_J W_J (Y_J-\mu(X_J))]- E_J[(Y_J-\mu(X_J))] E_J[I_J W_J]}{\sqrt{V_J(I_J W_J) V_J(Y_J-\mu(X_J))}} \frac{\sqrt{V_J(I_J W_J)}}{E_J[I_J]} \times \sqrt{V_J(Y_J-\mu(X_J))} \\
+% % = &\rho_{\tilde I (X),Y-\mu(X)} \times \frac{\sqrt{V_J(I_J W_J)}}{E_J[I_J]} \times \sigma_{Y-\mu(X)}
+% % \end{aligned}
+% % $$
+% % From prior derivations, we then have the desired result.
+
+% % \section{Asymptotic derivations}
+% % \label{app:asympderivations}
+
+% % % https://www.thelancet.com/journals/lanres/article/PIIS2213-2600%2820%2930453-7/fulltext
+% % % https://www.medrxiv.org/content/10.1101/2020.04.26.20080911v1.full.pdf
+
+% % Here, we take the suggested measures from~\cite{Arevalo2020} which report 87\% sensitivity is a reasonable estimates and~\cite{Cohen2020} which report 97.6\% specificity.
 This corresponds to a false negative rate of $13\%$ and false positive rate of $2.4\%$.
@@
 
-Suppose there is a sequence of finite populations of size $N_{\nu}$ indexed by $\nu$.
+% % Suppose there is a sequence of finite populations of size $N_{\nu}$ indexed by $\nu$.
 Each finite population has an associated non-probability at time $t$ of  size~$\tilde n$ and $L_\nu$ probability samples of fixed size $n$ drawn at equally spaced times $\{ t^\prime_l \}_{l=1}^L$ over the study window~$[0,T]$.
@@
 
-For each~$\nu$ and $t$, the estimates of $\eta = (\mu_t, \theta_t, FP, FN)$ are given by:
-$$
-\Phi (\eta) = \left (
-\begin{array}{c}
-\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t}{\pi_{j,t}} \\
-\frac{1}{N \sum_{l=1}^L K_{t,t_l^\prime}} \sum_{l=1}^L K_{t,t_l^\prime} \left[ \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime_l} - \sum_{j = 1}^N \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime}  \pi_{j,t_l^\prime} X_{j,t_l^\prime} \right] \\
-\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
-\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
-\end{array}
-\right ) = {\bf 0}
-$$
-where $\pi_{j,t^\prime} = \pi (X_{j,t^\prime}; \theta_t)$, $Z_j$ is an indicator of a false positive in the sample $S_{FP}$  and $\tilde Z_j$ is an indicator of a false negative in the sample $S_{FN}$.
+% % For each~$\nu$ and $t$, the estimates of $\eta = (\mu_t, \theta_t, FP, FN)$ are given by:
+% % $$
+% % \Phi (\eta) = \left (
+% % \begin{array}{c}
+% % \frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t}{\pi_{j,t}} \\
+% % \frac{1}{N \sum_{l=1}^L K_{t,t_l^\prime}} \sum_{l=1}^L K_{t,t_l^\prime} \left[ \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime_l} - \sum_{j = 1}^N \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime}  \pi_{j,t_l^\prime} X_{j,t_l^\prime} \right] \\
+% % \frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
+% % \frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
+% % \end{array}
+% % \right ) = {\bf 0}
+% % $$
+% % where $\pi_{j,t^\prime} = \pi (X_{j,t^\prime}; \theta_t)$, $Z_j$ is an indicator of a false positive in the sample $S_{FP}$  and $\tilde Z_j$ is an indicator of a false negative in the sample $S_{FN}$.
 We assume the $|S_{FP}|^{-1} \sum_{j \in S_{FP}} Z_j = 0.024$ and $|S_{FP}|^{-1} \sum_{j \in S_{FP}} Z_j = 0.30$.
@@
 
-To prove consistency, first consider the setting where a single random sample of size $n$ is observed at time $t$.
+% % To prove consistency, first consider the setting where a single random sample of size $n$ is observed at time $t$.
 Let $\tilde \Phi (\eta)$ denote the corresponding estimating equations, which is the same as $\Phi(\eta)$ except for the second component which is now given by
-$$
-\frac{1}{N} \left[ \sum_{j=1}^N I_{j,t} X_{j,t} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t} \tilde W_{j,t}  \pi_{j,t} X_{j,t} \right].
-$$
-Then under joint randomization of propensity score model and sampling designs, we have $E [ \tilde \Phi (\eta_0) ] = {\bf 0}$.
+% % $$
+% % \frac{1}{N} \left[ \sum_{j=1}^N I_{j,t} X_{j,t} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t} \tilde W_{j,t}  \pi_{j,t} X_{j,t} \right].
+% % $$
+% % Then under joint randomization of propensity score model and sampling designs, we have $E [ \tilde \Phi (\eta_0) ] = {\bf 0}$.
 Consistency then follows by arguments in Section 3.2 of Tsiatis (2006).
@@
 
-To prove consistency under the sequence of sampling regimes, we need to show that $E[ \Phi (\eta_0) ] \to 0$ as $N \to \infty$.  First, we study the difference in the 2nd term:
-\begin{align*}
-&\frac{1}{N}  E \left[ \sum_{j = 1}^N \tilde I_{j,t} \tilde W_{j,t}  \pi_{j,t} X_{j,t} - \sum_{l=1}^L  \sum_{j = 1}^N \tilde K_{t,t^\prime_l} \tilde I_{j,t^\prime_l} \tilde W_{j,t^\prime_l}  \pi_{j,t^\prime_l} X_{j,t^\prime_l} \right] \\
-= &\frac{1}{N} \sum_{j = 1}^N \left[ \pi_{j,t} X_{j,t} - \sum_{l=1}^L \tilde K_{t,t^\prime_l} \pi_{j,t^\prime_l} X_{j,t^\prime_l} \right], \\
-= &\frac{1}{N} \sum_{j = 1}^N \left[ \sum_{l=1}^L \tilde K_{t,t^\prime_l} \pi_{j,t} X_{j,t} \pm \sum_{l=1}^L \tilde K_{t,t^\prime_l} \pi_{j,t_l^\prime} X_{j,t}  - \sum_{l=1}^L \tilde K_{t,t^\prime_l} \pi_{j,t^\prime_l} X_{j,t^\prime_l} \right], \\
-= &\frac{1}{N} \sum_{j = 1}^N \left[ \left( \sum_{l=1}^L \tilde K_{t,t^\prime_l}  \left( \pi (X_{j,t}; \theta) - \pi (X_{j,t^\prime_l}; \theta) \right) \right) X_{j,t} + \sum_{l=1}^L \tilde K_{t,t^\prime_l} \pi_{j,t^\prime_l} (X_{j,t} - X_{j,t^\prime_l} ) \right],
-\end{align*}
-where $\tilde K_{t, t^\prime_l}$ is the normalized kernel such that $\sum_{l=1}^L \tilde K_{t, t^\prime_l} = 1$. The difference in the 1st term is similarly defined.
-
-Let~$X_{j,t,l}$ denote the $l$th component of the vector $X_{j,t}$.  Then consider the population-level distribution $\{ X_{j,t,l} \}_{j=1}^N$.  We make the additional assumption that
-$$
-\lim_{\epsilon \to 0} \lim_{N \to \infty} \frac{1}{N} \sum_{j=1}^N X_{j,t+\epsilon, l}  = \lim_{N \to \infty} \frac{1}{N} \sum_{j=1}^N X_{j,t, l}.
-$$
-That is, the limiting distributions are continuous as a function of $t$.
+% % To prove consistency under the sequence of sampling regimes, we need to show that $E[ \Phi (\eta_0) ] \to 0$ as $N \to \infty$.  First, we study the difference in the 2nd term:
+% % \begin{align*}
+% % &\frac{1}{N}  E \left[ \sum_{j = 1}^N \tilde I_{j,t} \tilde W_{j,t}  \pi_{j,t} X_{j,t} - \sum_{l=1}^L  \sum_{j = 1}^N \tilde K_{t,t^\prime_l} \tilde I_{j,t^\prime_l} \tilde W_{j,t^\prime_l}  \pi_{j,t^\prime_l} X_{j,t^\prime_l} \right] \\
+% % = &\frac{1}{N} \sum_{j = 1}^N \left[ \pi_{j,t} X_{j,t} - \sum_{l=1}^L \tilde K_{t,t^\prime_l} \pi_{j,t^\prime_l} X_{j,t^\prime_l} \right], \\
+% % = &\frac{1}{N} \sum_{j = 1}^N \left[ \sum_{l=1}^L \tilde K_{t,t^\prime_l} \pi_{j,t} X_{j,t} \pm \sum_{l=1}^L \tilde K_{t,t^\prime_l} \pi_{j,t_l^\prime} X_{j,t}  - \sum_{l=1}^L \tilde K_{t,t^\prime_l} \pi_{j,t^\prime_l} X_{j,t^\prime_l} \right], \\
+% % = &\frac{1}{N} \sum_{j = 1}^N \left[ \left( \sum_{l=1}^L \tilde K_{t,t^\prime_l}  \left( \pi (X_{j,t}; \theta) - \pi (X_{j,t^\prime_l}; \theta) \right) \right) X_{j,t} + \sum_{l=1}^L \tilde K_{t,t^\prime_l} \pi_{j,t^\prime_l} (X_{j,t} - X_{j,t^\prime_l} ) \right],
+% % \end{align*}
+% % where $\tilde K_{t, t^\prime_l}$ is the normalized kernel such that $\sum_{l=1}^L \tilde K_{t, t^\prime_l} = 1$. The difference in the 1st term is similarly defined.
+
+% % Let~$X_{j,t,l}$ denote the $l$th component of the vector $X_{j,t}$.  Then consider the population-level distribution $\{ X_{j,t,l} \}_{j=1}^N$.  We make the additional assumption that
+% % $$
+% % \lim_{\epsilon \to 0} \lim_{N \to \infty} \frac{1}{N} \sum_{j=1}^N X_{j,t+\epsilon, l}  = \lim_{N \to \infty} \frac{1}{N} \sum_{j=1}^N X_{j,t, l}.
+% % $$
+% % That is, the limiting distributions are continuous as a function of $t$.
 If $\pi(x; \theta)$ is a continuous function in~$x$ and under suitable conditions on the kernel density~$\tilde K_{t, \cdot}$ with the bandwidth $h_\nu \to \infty$ such that $\sum_{l=1}^L \tilde K_{t,t_l} \to \infty$ as $\nu \to \infty$, we have
-$$
-\left | \sum_{l=1}^L \tilde K_{t, t_l^\prime} \pi (X_{j, t_l^\prime}; \theta) - \pi (X_{j, t}; \theta) \right|  \to 0
-$$
-Combining this, $E \left[ \Phi(\eta_0)  - \tilde \Phi (\eta_0) \right] \to 0$ which implies consistency as desired.
-
-Then by a first-order Taylor expansion we have $\hat \eta - \eta_0 = \left[ \phi (\eta) \right]^{-1} \Phi(\eta_0) + o_p (\tilde n^{-1/2})$ where $\phi(\eta) = \partial \Phi (\eta)/\partial \eta$ and $\tilde n = n \sum_{l=1}^L \tilde K_{t,t_l}$.
+% % $$
+% % \left | \sum_{l=1}^L \tilde K_{t, t_l^\prime} \pi (X_{j, t_l^\prime}; \theta) - \pi (X_{j, t}; \theta) \right|  \to 0
+% % $$
+% % Combining this, $E \left[ \Phi(\eta_0)  - \tilde \Phi (\eta_0) \right] \to 0$ which implies consistency as desired.
+
+% % Then by a first-order Taylor expansion we have $\hat \eta - \eta_0 = \left[ \phi (\eta) \right]^{-1} \Phi(\eta_0) + o_p (\tilde n^{-1/2})$ where $\phi(\eta) = \partial \Phi (\eta)/\partial \eta$ and $\tilde n = n \sum_{l=1}^L \tilde K_{t,t_l}$.
 To calculate
-$$
-\frac{\partial \Phi (\eta)}{\partial \mu_t} =
-\left (
-\begin{array}{c}
--\frac{(1-FP-FN)}{N} \sum_{j=1}^N \frac{I_{j,t}}{\pi_{j,t}} \\
-0 \\
-0 \\
-0
-\end{array}
-\right )
-$$
-
-$$
-\frac{\partial \Phi (\eta)}{\partial \theta_t} =
-\left (
-\begin{array}{c}
--\frac{1}{N} \sum_{j=1}^N I_{j,t} (Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t) \frac{1-\pi_{j,t}}{\pi_{j,t}} X_{j,t}^\top \\
-- \frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} (1-\pi_{j,t^\prime}) X_{j,t^\prime} X_{j,t^\prime}^\top\\
-0 \\
-0
-\end{array}
-\right )
-$$
-
-$$
-\frac{\partial \Phi (\eta)}{\partial FP} =
-\left (
-\begin{array}{c}
--\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{(1-\mu_t)}{\pi_{j,t}} \\
-0\\
--\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \left( \frac{Z_j}{FP^2} + \frac{1-Z_j}{(1-FP)^2} \right)  \\
-0
-\end{array}
-\right )
-$$
-
-$$
-\frac{\partial \Phi (\eta)}{\partial FN} =
-\left (
-\begin{array}{c}
-- \frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{\mu_t}{\pi_{j,t}} \\
-0\\
-0 \\
--\frac{1}{n_{FN}}  \sum_{j \in S_{FN}} \left( \frac{\tilde Z_j}{FN^2} + \frac{1-\tilde Z_j}{(1-FN)^2} \right)
-\end{array}
-\right )
-$$
-
-We are interested in computing $- E \left[ \frac{\partial \Phi (\eta)}{\partial \mu_t} \right]^{-1}$ where the expectation is with respect to the random indicators.
+% % $$
+% % \frac{\partial \Phi (\eta)}{\partial \mu_t} =
+% % \left (
+% % \begin{array}{c}
+% % -\frac{(1-FP-FN)}{N} \sum_{j=1}^N \frac{I_{j,t}}{\pi_{j,t}} \\
+% % 0 \\
+% % 0 \\
+% % 0
+% % \end{array}
+% % \right )
+% % $$
+
+% % $$
+% % \frac{\partial \Phi (\eta)}{\partial \theta_t} =
+% % \left (
+% % \begin{array}{c}
+% % -\frac{1}{N} \sum_{j=1}^N I_{j,t} (Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t) \frac{1-\pi_{j,t}}{\pi_{j,t}} X_{j,t}^\top \\
+% % - \frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} (1-\pi_{j,t^\prime}) X_{j,t^\prime} X_{j,t^\prime}^\top\\
+% % 0 \\
+% % 0
+% % \end{array}
+% % \right )
+% % $$
+
+% % $$
+% % \frac{\partial \Phi (\eta)}{\partial FP} =
+% % \left (
+% % \begin{array}{c}
+% % -\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{(1-\mu_t)}{\pi_{j,t}} \\
+% % 0\\
+% % -\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \left( \frac{Z_j}{FP^2} + \frac{1-Z_j}{(1-FP)^2} \right)  \\
+% % 0
+% % \end{array}
+% % \right )
+% % $$
+
+% % $$
+% % \frac{\partial \Phi (\eta)}{\partial FN} =
+% % \left (
+% % \begin{array}{c}
+% % - \frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{\mu_t}{\pi_{j,t}} \\
+% % 0\\
+% % 0 \\
+% % -\frac{1}{n_{FN}}  \sum_{j \in S_{FN}} \left( \frac{\tilde Z_j}{FN^2} + \frac{1-\tilde Z_j}{(1-FN)^2} \right)
+% % \end{array}
+% % \right )
+% % $$
+
+% % We are interested in computing $- E \left[ \frac{\partial \Phi (\eta)}{\partial \mu_t} \right]^{-1}$ where the expectation is with respect to the random indicators.
 First, $- E \left[ \frac{\partial \Phi (\eta)}{\partial \mu_t} \right]$ is equal to
-$$
-\left (
-\begin{array}{c c c c}
-1-FP-FN & \frac{1}{N} \sum_{j=1}^N \zeta_{t,j} (1-\pi_{j,t}) X_{j,t}^\top & 1 - \mu_t & \mu_t \\
-0 & \frac{1}{N} \sum_{t^\prime=1}^T K_{t,t^\prime} \sum_{j=1}^N \pi_{j,t} (1-\pi_{j,t}) X_{j,t} X_{j,t}^\top & 0 & 0\\
-0 & 0 & \left( FP (1-FP) \right)^{-1}  & 0\\
-0 & 0 & 0 & \left( FN (1-FN) \right)^{-1}
-\end{array}
-\right )
-$$
-where $\zeta_{t,j} = (Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t)$.
+% % $$
+% % \left (
+% % \begin{array}{c c c c}
+% % 1-FP-FN & \frac{1}{N} \sum_{j=1}^N \zeta_{t,j} (1-\pi_{j,t}) X_{j,t}^\top & 1 - \mu_t & \mu_t \\
+% % 0 & \frac{1}{N} \sum_{t^\prime=1}^T K_{t,t^\prime} \sum_{j=1}^N \pi_{j,t} (1-\pi_{j,t}) X_{j,t} X_{j,t}^\top & 0 & 0\\
+% % 0 & 0 & \left( FP (1-FP) \right)^{-1}  & 0\\
+% % 0 & 0 & 0 & \left( FN (1-FN) \right)^{-1}
+% % \end{array}
+% % \right )
+% % $$
+% % where $\zeta_{t,j} = (Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t)$.
 This matrix can be written as a diagonal matrix equal to the diagonal of $\partial \Phi_n (\eta)/\partial \eta$ and then a rank-three matrix that is zero except for the first row on the off-diagonal.
@@
 Then Woodbury matrix identity gives us
-$$
-(A+B)^{-1} = A^{-1} - A^{-1} B A^{-1}
-$$
-In particular, the resulting inverse is of the form
-$$
-\left (
-\begin{array}{c c c c}
-\frac{1}{1-FP-FN} & b_1 & b_2  & b_3 \\
-0 & \left[\frac{1}{N} \sum_{t^\prime=1}^T K_{t,t^\prime} \sum_{j=1}^N \pi_{j,t} (1-\pi_{j,t}) X_{j,t} X_{j,t}^\top\right]^{-1} & 0 & 0\\
-0 & 0 & FP (1-FP)   & 0\\
-0 & 0 & 0 & FN (1-FN)
-\end{array}
-\right)
-$$
-where $b_1, b_2,$ and $b_3$ can be calculated using the Woodbury identity.
-
-Next, note that $\text{Var}(\Phi(\eta))$ can be decomposed into $A_{1,t} + A_{2,t} + A_{3,t} + A_{4,t}$
-
-$$
-A_{1,t} = \frac{1}{N} \left( \begin{array}{c}
-\sum_{j=1}^N \frac{I_{j,t} (Y_{j,t} - FP - (1-FP-FN) \mu_t)}{\pi_{j,t}} \\
-\sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \pi_{j,t^\prime} X_{j,t^\prime} \\
-0 \\
-0
-\end{array}
-\right)
-$$
-and
-$$
-A_{2,t} = \frac{1}{N} \left( \begin{array}{c}
-0 \\
-\sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N \pi_{j,t^\prime} X_{j,t^\prime} - \sum_{j=1}^N \tilde I_{j,t^\prime} \pi_{j,t^\prime} X_{j,t} \\
-0 \\
-0
-\end{array}
-\right),
-$$
-and
-$$
-A_{3,t} = \left( \begin{array}{c}
-0 \\
-0 \\
-\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
-0
-\end{array}
-\right),
-$$
-and
-$$
-A_{4,t} = \left( \begin{array}{c}
-0 \\
-0 \\
-0 \\
-\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
-\end{array}
-\right),
-$$
-By independence of these components, the variance can be calculated by summation over variance of each term individually yielding. First, $V_1$
-{
-\tiny
-$$
- N^{-2} \sum_{i=1}^N \left(
-\begin{array}{c c c c}
-\frac{(1-\pi_{j,t})}{\pi_{j,t}} (Y_{j,t} - (1-FP-FN) (\mu_{j,t} +FP))^2 & \sum_{t^\prime} K_{t,t^\prime} (1-\pi_{j,t^\prime}) \zeta_{t^\prime, j} X_{j,t^\prime}^\top & 0 & 0 \\
-\sum_{t^\prime} K_{t,t^\prime} (1-\pi_{j,t^\prime}) \zeta_{t^\prime, j} X_{j,t}&
-\sum_{t^\prime} K_{t,t^\prime} \pi_{j,t^\prime} (1-\pi_{j,t^\prime}) X_{j,t} X_{j,t}^\top
-& 0 & 0 \\
-0 & 0 & 0 & 0 \\
-0 & 0 & 0 & 0 \\
-\end{array}
-\right)
-$$
-}
-and $V_{j,t}$ are block-diagonal matrices with only one non-zero block that is equal to $D_{j,t}$ which is the design-based variance-covariance matrix under each probability sampling design.
-$$
-V_{3,t} = \left( \begin{array}{c}
-0 \\
-0 \\
-\frac{1}{n_{FP}} \frac{1}{FP(1-FP)} \\
-0
-\end{array}
-\right),
-$$
-and
-$$
-V_{4,t} = \left( \begin{array}{c}
-0 \\
-0 \\
-0 \\
-\frac{1}{n_{FN}} \frac{1}{FN(1-FN)}
-\end{array}
-\right),
-$$
-
-The asymptotic variance for the IPW estimator is then the first diagonal element of the matrix $E \left[ \phi_n (\eta_0) \right]^{-1} \times \left[ V_{1,t} + V_{2,t} + V_{3,t} + V_{4,t} \right] E \left[ \phi_n (\eta_0) \right]^{-1}$.
-\section{IRLS}
-
-First derivative with respect to $\theta$
-$$
-\nabla E(\theta) = \sum_{t^\prime=1}^T K_h(|t^\prime - t|) \left[ \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} + \sum_{j=1}^N \tilde W_{j,t^\prime} \tilde I_{j,t^\prime} \pi_t (X_{j,t^\prime}; \theta) X_{j,t^\prime} \right]
-$$
-Second derivative with respect to $\theta$
-$$
-H(\theta) = \sum_{t^\prime=1}^T K_h(|t^\prime - t|) \sum_{j=1}^N \tilde W_{j,t^\prime} \tilde I_{j,t^\prime} \pi_t (X_{j,t^\prime}; \theta) \left( 1- \pi_t (X_{j,t^\prime}; \theta) \right) X_{j,t^\prime} X_{j,t^\prime}^\top
-$$
-Then the Fisher scoring method sets
-$$
-\theta^{(t+1)} = \theta^{(t)} + H \left(\theta^{(t)} \right)^{-1} \nabla E \left(\theta^{(t)} \right)
-$$
-
-\section{Indiana COVID-19 Analysis: Additional details}
-\label{app:in_add_details}
-
-In this section, we describe the two mean-imputation methods.
-\subsection{Model 1}
-
-First, we use the pseudo-likelihood to compute the likelihood of contact given age, gender, and test result.
+% % $$
+% % (A+B)^{-1} = A^{-1} - A^{-1} B A^{-1}
+% % $$
+% % In particular, the resulting inverse is of the form
+% % $$
+% % \left (
+% % \begin{array}{c c c c}
+% % \frac{1}{1-FP-FN} & b_1 & b_2  & b_3 \\
+% % 0 & \left[\frac{1}{N} \sum_{t^\prime=1}^T K_{t,t^\prime} \sum_{j=1}^N \pi_{j,t} (1-\pi_{j,t}) X_{j,t} X_{j,t}^\top\right]^{-1} & 0 & 0\\
+% % 0 & 0 & FP (1-FP)   & 0\\
+% % 0 & 0 & 0 & FN (1-FN)
+% % \end{array}
+% % \right)
+% % $$
+% % where $b_1, b_2,$ and $b_3$ can be calculated using the Woodbury identity.
+
+% % Next, note that $\text{Var}(\Phi(\eta))$ can be decomposed into $A_{1,t} + A_{2,t} + A_{3,t} + A_{4,t}$
+
+% % $$
+% % A_{1,t} = \frac{1}{N} \left( \begin{array}{c}
+% % \sum_{j=1}^N \frac{I_{j,t} (Y_{j,t} - FP - (1-FP-FN) \mu_t)}{\pi_{j,t}} \\
+% % \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \pi_{j,t^\prime} X_{j,t^\prime} \\
+% % 0 \\
+% % 0
+% % \end{array}
+% % \right)
+% % $$
+% % and
+% % $$
+% % A_{2,t} = \frac{1}{N} \left( \begin{array}{c}
+% % 0 \\
+% % \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N \pi_{j,t^\prime} X_{j,t^\prime} - \sum_{j=1}^N \tilde I_{j,t^\prime} \pi_{j,t^\prime} X_{j,t} \\
+% % 0 \\
+% % 0
+% % \end{array}
+% % \right),
+% % $$
+% % and
+% % $$
+% % A_{3,t} = \left( \begin{array}{c}
+% % 0 \\
+% % 0 \\
+% % \frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
+% % 0
+% % \end{array}
+% % \right),
+% % $$
+% % and
+% % $$
+% % A_{4,t} = \left( \begin{array}{c}
+% % 0 \\
+% % 0 \\
+% % 0 \\
+% % \frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
+% % \end{array}
+% % \right),
+% % $$
+% % By independence of these components, the variance can be calculated by summation over variance of each term individually yielding. First, $V_1$
+% % {
+% % \tiny
+% % $$
+% %  N^{-2} \sum_{i=1}^N \left(
+% % \begin{array}{c c c c}
+% % \frac{(1-\pi_{j,t})}{\pi_{j,t}} (Y_{j,t} - (1-FP-FN) (\mu_{j,t} +FP))^2 & \sum_{t^\prime} K_{t,t^\prime} (1-\pi_{j,t^\prime}) \zeta_{t^\prime, j} X_{j,t^\prime}^\top & 0 & 0 \\
+% % \sum_{t^\prime} K_{t,t^\prime} (1-\pi_{j,t^\prime}) \zeta_{t^\prime, j} X_{j,t}&
+% % \sum_{t^\prime} K_{t,t^\prime} \pi_{j,t^\prime} (1-\pi_{j,t^\prime}) X_{j,t} X_{j,t}^\top
+% % & 0 & 0 \\
+% % 0 & 0 & 0 & 0 \\
+% % 0 & 0 & 0 & 0 \\
+% % \end{array}
+% % \right)
+% % $$
+% % }
+% % and $V_{j,t}$ are block-diagonal matrices with only one non-zero block that is equal to $D_{j,t}$ which is the design-based variance-covariance matrix under each probability sampling design.
+% % $$
+% % V_{3,t} = \left( \begin{array}{c}
+% % 0 \\
+% % 0 \\
+% % \frac{1}{n_{FP}} \frac{1}{FP(1-FP)} \\
+% % 0
+% % \end{array}
+% % \right),
+% % $$
+% % and
+% % $$
+% % V_{4,t} = \left( \begin{array}{c}
+% % 0 \\
+% % 0 \\
+% % 0 \\
+% % \frac{1}{n_{FN}} \frac{1}{FN(1-FN)}
+% % \end{array}
+% % \right),
+% % $$
+
+% % The asymptotic variance for the IPW estimator is then the first diagonal element of the matrix $E \left[ \phi_n (\eta_0) \right]^{-1} \times \left[ V_{1,t} + V_{2,t} + V_{3,t} + V_{4,t} \right] E \left[ \phi_n (\eta_0) \right]^{-1}$.
+% % \section{IRLS}
+
+% % First derivative with respect to $\theta$
+% % $$
+% % \nabla E(\theta) = \sum_{t^\prime=1}^T K_h(|t^\prime - t|) \left[ \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} + \sum_{j=1}^N \tilde W_{j,t^\prime} \tilde I_{j,t^\prime} \pi_t (X_{j,t^\prime}; \theta) X_{j,t^\prime} \right]
+% % $$
+% % Second derivative with respect to $\theta$
+% % $$
+% % H(\theta) = \sum_{t^\prime=1}^T K_h(|t^\prime - t|) \sum_{j=1}^N \tilde W_{j,t^\prime} \tilde I_{j,t^\prime} \pi_t (X_{j,t^\prime}; \theta) \left( 1- \pi_t (X_{j,t^\prime}; \theta) \right) X_{j,t^\prime} X_{j,t^\prime}^\top
+% % $$
+% % Then the Fisher scoring method sets
+% % $$
+% % \theta^{(t+1)} = \theta^{(t)} + H \left(\theta^{(t)} \right)^{-1} \nabla E \left(\theta^{(t)} \right)
+% % $$
+
+% % \section{Indiana COVID-19 Analysis: Additional details}
+% % \label{app:in_add_details}
+
+% % In this section, we describe the two mean-imputation methods.
+% % \subsection{Model 1}
+
+% % First, we use the pseudo-likelihood to compute the likelihood of contact given age, gender, and test result.
 Figure~\ref{fig:contactlik1} and~\ref{fig:contactlik2} plots the likelihood for 25-34 year olds given a negative test and positive test respectfully.
@@
 
-\begin{figure}[!th]
-\centering
-\begin{subfigure}{.5\textwidth}
- \centering
- \includegraphics[width=.9\linewidth]{../figs/tvprop_contact_fig1.png}
- \caption{Contact Likelihood Given Negative Test}
- \label{fig:contactlik1}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
- \centering
-\includegraphics[width=.9\linewidth]{../figs/tvprop_contact_fig2.png}
- \caption{Contact Likelihood Given Positive Test}
- \label{fig:contactlik2}
-\end{subfigure}
-\caption{Likelihood of COVID-19 contact}
-\label{fig:contactlik}
-\end{figure}
-
-We then use the pseudo-likelihood to compute the likelihood of fever given age, gender, COVID-19 contact status and test result.
+% % \begin{figure}[!th]
+% % \centering
+% % \begin{subfigure}{.5\textwidth}
+% %  \centering
+% %  \includegraphics[width=.9\linewidth]{../figs/tvprop_contact_fig1.png}
+% %  \caption{Contact Likelihood Given Negative Test}
+% %  \label{fig:contactlik1}
+% % \end{subfigure}%
+% % \begin{subfigure}{.5\textwidth}
+% %  \centering
+% % \includegraphics[width=.9\linewidth]{../figs/tvprop_contact_fig2.png}
+% %  \caption{Contact Likelihood Given Positive Test}
+% %  \label{fig:contactlik2}
+% % \end{subfigure}
+% % \caption{Likelihood of COVID-19 contact}
+% % \label{fig:contactlik}
+% % \end{figure}
+
+% % We then use the pseudo-likelihood to compute the likelihood of fever given age, gender, COVID-19 contact status and test result.
 Figure~\ref{fig:symptomlik1} and~\ref{fig:symptomlik2} plots the likelihood for 35-44 year olds given a negative test and positive test respectfully.
@@
 
-\begin{figure}[!th]
-\centering
-\begin{subfigure}{.5\textwidth}
- \centering
- \includegraphics[width=.9\linewidth]{../figs/tvprop_symptom_fig1.png}
- \caption{Fever Likelihood Given Negative Test \textcolor{red}{FIX!!}}
- \label{fig:symptomlik1}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
- \centering
-\includegraphics[width=.9\linewidth]{../figs/tvprop_symptom_fig2.png}
- \caption{Fever Likelihood Given Positive Test}
- \label{fig:symptomlik2}
-\end{subfigure}
-\caption{Likelihood of reported fever}
-\label{fig:symptomlik}
-\end{figure}
-
-Based on mean-imputation of COVID-19 contact and fever indicators, we can use the proposed pseudo-likelihood approach to compute the likelihood of getting tested for COVID-19 given age, ethnicity, race, gender, fever status, and COVID-19 contact indicator.
+% % \begin{figure}[!th]
+% % \centering
+% % \begin{subfigure}{.5\textwidth}
+% %  \centering
+% %  \includegraphics[width=.9\linewidth]{../figs/tvprop_symptom_fig1.png}
+% %  \caption{Fever Likelihood Given Negative Test \textcolor{red}{FIX!!}}
+% %  \label{fig:symptomlik1}
+% % \end{subfigure}%
+% % \begin{subfigure}{.5\textwidth}
+% %  \centering
+% % \includegraphics[width=.9\linewidth]{../figs/tvprop_symptom_fig2.png}
+% %  \caption{Fever Likelihood Given Positive Test}
+% %  \label{fig:symptomlik2}
+% % \end{subfigure}
+% % \caption{Likelihood of reported fever}
+% % \label{fig:symptomlik}
+% % \end{figure}
+
+% % Based on mean-imputation of COVID-19 contact and fever indicators, we can use the proposed pseudo-likelihood approach to compute the likelihood of getting tested for COVID-19 given age, ethnicity, race, gender, fever status, and COVID-19 contact indicator.
 Figure~\ref{fig:testinglik1} and~\ref{fig:testinglik2} plots the likelihood of testing given fever and COVID-19 contact and no fever nor COVID-19 contact respectfully.
@@
 
-\begin{figure}[!th]
-\centering
-\begin{subfigure}{.5\textwidth}
- \centering
- \includegraphics[width=.9\linewidth]{../figs/tvprop_fig1.png}
- \caption{Testing Likelihood Given Fever and COVID-19 Contact}
- \label{fig:testinglik1}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
- \centering
-\includegraphics[width=.9\linewidth]{../figs/tvprop_fig2.png}
- \caption{Testing Propensity}
- \label{fig:testinglik2}
-\end{subfigure}
-\caption{Testing Likelihood Given no Fever nor COVID-19 Contact}
-\label{fig:testinglik}
-\end{figure}
-
-\newpage
-
-\subsection{Model 2 (Hospitalization)}
-
-This model uses hospitalization records to try and improve our mean-imputation strategy.
+% % \begin{figure}[!th]
+% % \centering
+% % \begin{subfigure}{.5\textwidth}
+% %  \centering
+% %  \includegraphics[width=.9\linewidth]{../figs/tvprop_fig1.png}
+% %  \caption{Testing Likelihood Given Fever and COVID-19 Contact}
+% %  \label{fig:testinglik1}
+% % \end{subfigure}%
+% % \begin{subfigure}{.5\textwidth}
+% %  \centering
+% % \includegraphics[width=.9\linewidth]{../figs/tvprop_fig2.png}
+% %  \caption{Testing Propensity}
+% %  \label{fig:testinglik2}
+% % \end{subfigure}
+% % \caption{Testing Likelihood Given no Fever nor COVID-19 Contact}
+% % \label{fig:testinglik}
+% % \end{figure}
+
+% % \newpage
+
+% % \subsection{Model 2 (Hospitalization)}
+
+% % This model uses hospitalization records to try and improve our mean-imputation strategy.
 First, we use the pseudo-likelihood to compute the likelihood of fever given age, gender, and test result.
@@
 
-\begin{figure}[!th]
-\centering
-\begin{subfigure}{.5\textwidth}
- \centering
- \includegraphics[width=.9\linewidth]{../figs/tvprop_symptom_alt_fig1.png}
- \caption{Symptom Likelihood Given Negative Test}
- \label{fig:symptomlik1_model2}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
- \centering
-\includegraphics[width=.9\linewidth]{../figs/tvprop_symptom_alt_fig2.png}
- \caption{Symptom Likelihood Given Positive Test}
- \label{fig:symptomlik2_model2}
-\end{subfigure}
-\caption{Likelihood of COVID-19 contact}
-\label{fig:symptomlik_model2}
-\end{figure}
-
-First, we computed the probability of fever given COVID-19 positive test and hospitalization as well as COVID-19 positive test and no hospitalization.
+% % \begin{figure}[!th]
+% % \centering
+% % \begin{subfigure}{.5\textwidth}
+% %  \centering
+% %  \includegraphics[width=.9\linewidth]{../figs/tvprop_symptom_alt_fig1.png}
+% %  \caption{Symptom Likelihood Given Negative Test}
+% %  \label{fig:symptomlik1_model2}
+% % \end{subfigure}%
+% % \begin{subfigure}{.5\textwidth}
+% %  \centering
+% % \includegraphics[width=.9\linewidth]{../figs/tvprop_symptom_alt_fig2.png}
+% %  \caption{Symptom Likelihood Given Positive Test}
+% %  \label{fig:symptomlik2_model2}
+% % \end{subfigure}
+% % \caption{Likelihood of COVID-19 contact}
+% % \label{fig:symptomlik_model2}
+% % \end{figure}
+
+% % First, we computed the probability of fever given COVID-19 positive test and hospitalization as well as COVID-19 positive test and no hospitalization.
 We then computed the lieklihood of symptom given a COVID-19 positive test by weighting these two likelihoods by the fraction of COVID-19 positive hospitalizations to COVID-19 total positive cases per week.
@@
 
-Based on mean-imputation of COVID-19 fever using hospitalization records indicators, we can use the proposed pseudo-likelihood approach to compute the likelihood of getting tested for COVID-19 given age, ethnicity, race, gender, and fever status.
+% % Based on mean-imputation of COVID-19 fever using hospitalization records indicators, we can use the proposed pseudo-likelihood approach to compute the likelihood of getting tested for COVID-19 given age, ethnicity, race, gender, and fever status.
 Figure~\ref{fig:testinglik1_mainpaper} and~\ref{fig:testinglik1_mainpaper} in Section~\ref{section:tvipw} of the manuscript plots the likelihood of testing given fever and no fever respectfully for all age ranges.
@@
 
-% \begin{figure}[!th]
-% \centering
-% \begin{subfigure}{.5\textwidth}
-%  \centering
-%  \includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig1.png}
-%  \caption{Testing Likelihood Given Fever}
-%  \label{fig:tvprop_alt_fig1.png}
-% \end{subfigure}%
-% \begin{subfigure}{.5\textwidth}
-%  \centering
-% \includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig2.png}
-%  \caption{Testing Propensity Given No Fever}
-%  \label{fig:tvprop_alt_fig2.png}
-% \end{subfigure}
-% \caption{COVID-10 Testing Likelihood Given no Fever nor COVID-19 Contact}
-% \label{fig:testinglik}
-% \end{figure}
-
-\section{Model-based: Prior specification}
-\label{app:prior_modelbased}
-
-For simplicity, we list the priors used in STAN below:
-\begin{itemize}
-  \item \code{beta $\sim$ Normal(1.5, 1) T[0,];}
-  \item \code{gamma $\sim$ normal(0.3, 0.5) T[0,];}
-  \item \code{sigma $\sim$ normal(0.4, 0.5) T[0,];}
-  \item \code{phi\_inv $\sim$ exponential(5);}
-  \item \code{i0 $\sim$ normal(0, 10);}
-  \item \code{e0 $\sim$ normal(0, 10);}
-  \item \code{eta $\sim$ normal(1, 1) T[0,];}
-  \item \code{eta\_two $\sim$ normal(1, 1) T[0,];}
-  \item \code{eta\_three $\sim$ normal(1, 1) T[0,];}
-  \item \code{nu $\sim$ exponential(1./5);}
-  \item \code{nu\_two $\sim$ exponential(1./5);}
-  \item \code{nu\_three $\sim$ exponential(1./5);}
-  \item \code{xi\_raw $\sim$ beta(1, 1);}
-  \item \code{phi = 1/phi\_inv;}
-  \item \code{xi = xi\_raw + 0.5;}
-\end{itemize}
-where \code{beta, eta, eta\_two, eta\_three} refer to the four values that form the time-varying parameter $\beta_t$ when combined with \code{xi, eta, eta\_two, eta\_three} in the SEIR model as described in Section~\ref{section:modelbased}.
+% % % \begin{figure}[!th]
+% % % \centering
+% % % \begin{subfigure}{.5\textwidth}
+% % %  \centering
+% % %  \includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig1.png}
+% % %  \caption{Testing Likelihood Given Fever}
+% % %  \label{fig:tvprop_alt_fig1.png}
+% % % \end{subfigure}%
+% % % \begin{subfigure}{.5\textwidth}
+% % %  \centering
+% % % \includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig2.png}
+% % %  \caption{Testing Propensity Given No Fever}
+% % %  \label{fig:tvprop_alt_fig2.png}
+% % % \end{subfigure}
+% % % \caption{COVID-10 Testing Likelihood Given no Fever nor COVID-19 Contact}
+% % % \label{fig:testinglik}
+% % % \end{figure}
+
+% % \section{Model-based: Prior specification}
+% % \label{app:prior_modelbased}
+
+% % For simplicity, we list the priors used in STAN below:
+% % \begin{itemize}
+% %   \item \code{beta $\sim$ Normal(1.5, 1) T[0,];}
+% %   \item \code{gamma $\sim$ normal(0.3, 0.5) T[0,];}
+% %   \item \code{sigma $\sim$ normal(0.4, 0.5) T[0,];}
+% %   \item \code{phi\_inv $\sim$ exponential(5);}
+% %   \item \code{i0 $\sim$ normal(0, 10);}
+% %   \item \code{e0 $\sim$ normal(0, 10);}
+% %   \item \code{eta $\sim$ normal(1, 1) T[0,];}
+% %   \item \code{eta\_two $\sim$ normal(1, 1) T[0,];}
+% %   \item \code{eta\_three $\sim$ normal(1, 1) T[0,];}
+% %   \item \code{nu $\sim$ exponential(1./5);}
+% %   \item \code{nu\_two $\sim$ exponential(1./5);}
+% %   \item \code{nu\_three $\sim$ exponential(1./5);}
+% %   \item \code{xi\_raw $\sim$ beta(1, 1);}
+% %   \item \code{phi = 1/phi\_inv;}
+% %   \item \code{xi = xi\_raw + 0.5;}
+% % \end{itemize}
+% % where \code{beta, eta, eta\_two, eta\_three} refer to the four values that form the time-varying parameter $\beta_t$ when combined with \code{xi, eta, eta\_two, eta\_three} in the SEIR model as described in Section~\ref{section:modelbased}.
 

### 2021-08-23 preprocessing/3_feverfb_merge.R
Fixing the code to use COVID 19 symptoms

@@ -2,7 +2,7 @@
 ## (1) Cleaned Facebook Data
 ## (2) Cleaned Weekly Indiana Data
 ## Outputs:
-## (1) Collaping down to FEVER only
+## (1) Collapsing down to Symptom and Contact only
 
 library("lubridate")
 
@@ -39,8 +39,8 @@ for (week in weeksin2020) {
   agg_temp$year = rep(2020, nrow(agg_temp))
   all_data_neg_contact = rbind(all_data_neg_contact, agg_temp)
   
-  ## Adding fever given contact
-  temp$weightsymptoms =  temp$weight * temp$fever
+  ## Adding symptoms given contact
+  temp$weightsymptoms =  temp$weight * temp$symptoms
   total_weight = sum(temp$weight[temp$gottested ==1 & temp$postest == 0], na.rm = TRUE)
   total_weightsymptoms = sum(temp$weightsymptoms[temp$gottested ==1 & temp$postest == 0], na.rm = TRUE)
   agg_temp = aggregate(weightsymptoms ~ gender+ age + contact, subset(temp, gottested == 1 & postest == 0), sum)
@@ -64,8 +64,8 @@ for (week in weeksin2020) {
   agg_temp$year = rep(2020, nrow(agg_temp))
   all_data_pos_contact = rbind(all_data_pos_contact, agg_temp)
   
-  ## Adding fever given contact
-  temp$weightsymptoms =  temp$weight * temp$fever
+  ## Adding symptoms given contact
+  temp$weightsymptoms =  temp$weight * temp$symptoms
   total_weight = sum(temp$weight[temp$postest == 1], na.rm = TRUE)
   total_weightsymptoms = sum(temp$weightsymptoms[temp$postest == 1], na.rm = TRUE)
   agg_temp = aggregate(weightsymptoms ~ gender+ age + contact, subset(temp, postest == 1), sum)
@@ -98,8 +98,8 @@ for (week in weeksin2021) {
   agg_temp$year = rep(2021, nrow(agg_temp))
   all_data_neg_contact = rbind(all_data_neg_contact, agg_temp)
   
-  ## Adding fever given contact
-  temp$weightsymptoms =  temp$weight * temp$fever
+  ## Adding symptoms given contact
+  temp$weightsymptoms =  temp$weight * temp$symptoms
   total_weight = sum(temp$weight[temp$gottested ==1 & temp$postest == 0], na.rm = TRUE)
[…]

### 2021-08-23 visualizations/fb_symptoms_plot.R

@@ -9,22 +9,22 @@ indiana_data = read.csv("../data/fb_indiana_data.csv")
 
 indiana_data$symptoms = (indiana_data$fever & (indiana_data$cough | indiana_data$shortness | indiana_data$diffbreath))
 
-indiana_data$weightedfever = (indiana_data$fever) * indiana_data$weight
+indiana_data$weightedsymptoms = (indiana_data$symptoms) * indiana_data$weight
 png(filename = "../figs/fbcovid19symptoms.png",
     width = 960, height = 480, units = "px", pointsize = 12)
 par(mfrow = c(1,2), mar = c(4,4,0.5,1)+0.1)
 
 denom = aggregate(weight ~ date + gender, data = indiana_data, sum)
 
-numer = aggregate(weightedfever ~ date + gender, data = indiana_data, sum)
+numer = aggregate(weightedsymptoms ~ date + gender, data = indiana_data, sum)
 
-tempdf = data.frame('x' = difftime(date(numer$date), "2020-01-01"), 'y' = numer$weightedfever/denom$weight, 'gender' = denom$gender)
+tempdf = data.frame('x' = difftime(date(numer$date), "2020-01-01"), 'y' = numer$weightedsymptoms/denom$weight, 'gender' = denom$gender)
 tempdf$x = as.vector(tempdf$x)
 
 model.np <- npreg(y~x, regtype = 'll', bwmethod = 'cv.aic', gradients = TRUE, data = subset(tempdf, gender == 2))
 
 plot(date(numer$date[numer$gender == 2]), fitted(model.np), type = "l", 
-     bty= 'n', axes = FALSE, ylab = "Reported Fever", xlab = "Date",
+     bty= 'n', axes = FALSE, ylab = "Reported COVID-19 Symptoms", xlab = "Date",
      ylim = c(0.0,0.030), lwd = 3)
 axis.Date(1, at=seq(min(date(numer$date[numer$gender == 2])), max(date(numer$date[numer$gender == 2])), by="months"), format="%b")
 axis(side = 2)
@@ -47,9 +47,9 @@ indiana_data$blockage = is.element(indiana_data$age, c(1,2,3)) + 2* is.element(i
 
 denomage = aggregate(weight ~ date + blockage, data = subset(indiana_data, blockage != 0), sum)
 
-numerage = aggregate(weightedfever ~ date + blockage, data = subset(indiana_data, blockage != 0), sum)
[…]

### 2021-08-24 visualizations/tv_contact_propensities.R
updating figures and numbers for symptoms

@@ -31,7 +31,7 @@ ggplot(data = propensities_neg_long, aes(x = date, y = probs, col = Gender)) +
   geom_point(size = 5) +
   labs(x = "Date",
        y = "Likelihood of Contact",
-       title = "Individuals 25-34 tested negative in the past 24 hours") +
+       title = "Individuals 25-34, Negative Test") +
   scale_color_manual(values = my_palette) +
   theme(text = element_text(size=25))
 
@@ -64,7 +64,7 @@ ggplot(data = propensities_pos_long, aes(x = date, y = probs, col = Gender)) +
   geom_point(size = 5) +
   labs(x = "Date",
        y = "Likelihood of Contact",
-       title = "Individuals 25-34 tested positive in the past 24 hours") +
+       title = "Individuals 25-34, Positive Test") +
   theme(text = element_text(size=25)) + 
   scale_color_manual(values = my_palette)
 

### 2021-08-24 visualizations/tv_symptom_propensities.R

@@ -30,8 +30,8 @@ png(filename = "../figs/tvprop_symptom_fig1.png",
 ggplot(data = propensities_neg_long, aes(x = date, y = probs, col = Contact)) +
   geom_point(size=5) +
   labs(x = "Date",
-       y = "Likelihood of Fever",
-       title = "Male Individuals 35-44 tested negative in the past 24 hours") + 
+       y = "Likelihood of Symptoms",
+       title = "Male, 35-44, Negative Test") + 
   scale_color_manual(values = my_palette) +
   theme(text = element_text(size=25))
 
@@ -63,8 +63,8 @@ png(filename = "../figs/tvprop_symptom_fig2.png",
 ggplot(data = propensities_pos_long, aes(x = date, y = probs, col = Contact)) +
   geom_point(size=5) +
   labs(x = "Date",
-       y = "Likelihood of Fever",
-       title = "Female individuals 35-44 tested positive in the past 24 hours") +
+       y = "Likelihood of Symptoms",
+       title = "Female, 35-44, Positive Test") +
   scale_color_manual(values = my_palette) +
   theme(text = element_text(size=25))
 

### 2021-08-25 write-up/covid-supplement-writeup.tex
updating figures and numbers for symptoms
writeups checked (close)

@@
 Suppose there is a sequence of finite populations of size $N_{\nu}$ indexed by $\nu$.
-Each finite population has an associated non-probability at time $t$ of  size~$\tilde n$ and $L_\nu$ probability samples of fixed size $n$ drawn at equally spaced times $\{ t^\prime_l \}_{l=1}^L$ over the study window~$[0,T]$.
-Let~$\Delta_\nu$ denote the gap times between consecutive sample times.
+Each finite population has an associated non-probability sample at time $t$ of  size~$\tilde n$ and $L_\nu$ probability samples of fixed size $n$ drawn at equally spaced times $\{ t^\prime_l \}_{l=1}^L$ over the study window~$[0,T]$.
+Let~$\Delta_\nu$ denote the gap times between consecutive sample times such that as $\nu \to \infty$ we have that $\Delta_\nu \to 0$.
 Finally, let~$n_{FP}$ and $n_{FN}$ denote the sample sizes of $S_{FP}$ and $S_{FN}$ respectively.
 For simplicity, dependency on $\nu$ is suppressed and the limiting process is represented by $N \to \infty$.
-The asymptotic argument below relies on regularity conditions C1-C6 from~\cite{Chen2019}, smoothness of the population-level distribution of the time-varying covariates~$\{ X_{j,t} \}$, and smoothness of the propensity function~$\pi(x; \theta)$ as a function of $x \in \mathbb{R}^p$.
+The asymptotic argument below relies on regularity conditions C1-C6 from~\cite{Chen2019}, smoothness of the population-level distribution of the time-varying covariates~$\{ X_{j,t} \}$ as a function of $t > 0$, and smoothness of the propensity function~$\pi(x; \theta)$ as a function of $x \in \mathbb{R}^p$.
 
@@
 Consistency then follows by arguments in Section 3.2 of Tsiatis (2006).
-Under conditions C1-C6 from~\cite{Chen2019} and as $n$ increases, we have $\tilde \Phi (\hat \eta) = 0$ and $\tilde \Phi(\eta_0) = O_p ( n^{-1/2} )$.
+Under conditions C1-C6 from~\cite{Chen2019} and as $n$ increases, we have $\tilde \Phi (\hat \eta^\prime) = 0$ and $\tilde \Phi(\eta_0) = O_p ( n^{-1/2} )$ where $\hat \eta^\prime$ here refers to the solution to $\tilde \Phi (\eta)$.
 
@@
 \sum_{j=1}^N \frac{I_{j,t} (Y_{j,t} - FP - (1-FP-FN) \mu_t)}{\pi_{j,t}} \\
-\sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \pi_{j,t^\prime} X_{j,t^\prime} \\
+\sum_{t^\prime = 1}^T K_{t,t^\prime} \left[ \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \pi_{j,t^\prime} X_{j,t^\prime} \right] \\
 0 \\
@@
 0 \\
-\sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N \pi_{j,t^\prime} X_{j,t^\prime} - \sum_{j=1}^N \tilde I_{j,t^\prime} \pi_{j,t^\prime} X_{j,t} \\
+\sum_{t^\prime = 1}^T K_{t,t^\prime} \left[ \sum_{j=1}^N \pi_{j,t^\prime} X_{j,t^\prime} - \sum_{j=1}^N \tilde I_{j,t^\prime} \pi_{j,t^\prime} X_{j,t} \right] \\
 0 \\
@@
 $$
-By independence of these components, the variance can be calculated by summation over variance of each term individually yielding. First, $V_1$
+By independence of these components, the variance can be calculated by summation over variance of each term individually yielding. First, $V_{1,t}$
 {
@@
 }
-and $V_{j,t}$ are block-diagonal matrices with only one non-zero block that is equal to $D_{j,t}$ which is the design-based variance-covariance matrix under each probability sampling design.
+and $V_{2,t}$ is a block-diagonal matrix with only one non-zero block that is equal to $D_{j,t} = N^{-2} \sum_{t^\prime} K_{t,t^\prime}^2 V_p\left( \sum_{i=1}^n \tilde W_{i,t^\prime} \tilde I_{i,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime} \right)$ which is the design-based variance-covariance matrix under the probability sampling design; and
 $$
@@
  \includegraphics[width=.9\linewidth]{../figs/tvprop_symptom_fig1.png}
- \caption{Fever Likelihood Given Negative Test \textcolor{red}{FIX!!}}
+ \caption{Symptom Likelihood Given Negative Test}
  \label{fig:symptomlik1}
@@
 \includegraphics[width=.9\linewidth]{../figs/tvprop_symptom_fig2.png}
- \caption{Fever Likelihood Given Positive Test}
+ \caption{Symptom Likelihood Given Positive Test}
  \label{fig:symptomlik2}
 \end{subfigure}
-\caption{Likelihood of reported fever}
+\caption{Likelihood of reported COVID-19 symptoms}
 \label{fig:symptomlik}

### 2021-08-25 write-up/covid-writeup.tex
Fixing the code to use COVID 19 symptoms
writeups checked (close)

@@
 % \newtheorem{example}{Example}
-% \newtheorem{remark}{Remark}
+ \newtheorem{remark}{Remark}
 \newtheorem{corollary}[theorem]{Corollary}
@@
 This paper has two main objectives.
-The first objective is to express reservations at the use of case-counts as a proxy for disease prevalence and in estimation of standard epidemiological models for inference and forecasting.
+The first objective is to express reservations at the use of observed case-counts as a proxy for disease prevalence and in estimation of standard epidemiological models for inference and forecasting.
 The reason is straightforward: observed case-count data is plagued by selection bias and measurement error.
@@
 Due to monetary and time constraints, however, random testing is performed infrequently.
-As of June 2021, Indiana is the only state to conduct statewide random sample testing\footnote{This is the only random sample to collect both seroprevalence and diagnostic testing results.
+As of June 2021, to the best of our knowledge, Indiana is the only state to conduct statewide random sample testing\footnote{This is the only random sample to collect both seroprevalence and diagnostic testing results.
 The CDC has conducted several seroprevalence studies.
@@
  This article discusses the relationship between three statistical concepts: selection bias, measurement error, and population size.
-We build on the work of \cite{Meng2018} who studied an error decomposition to understand the relationship between selection bias and population size.
-In particular, we quantify the interaction between measurement-error and selection bias.
-The sign and magnitude of the statistical decomposition can change drastically; the impact on the effective sample size can therefore be quite large.
- \textcolor{red}{Acknowledge selection bias and measurement error.}
+While the impact of measurement error~\citep{Smeden2019} and selection bias~\citep{Keiding2016} on estimation are both well-studied topics, here we provide a new perspective by building on the work of \cite{Meng2018} who studied an error decomposition to understand the relationship between selection bias and population size.
+Speficially, we quantify the interaction between measurement-error and selection bias on statistical error, showing how the sign and magnitude can change drastically.
+ % the impact on the effective sample size can therefore be quite large.
 
@@
 For nonprobability samples, the IPW estimator requires modelling the propensity scores.
-Its use in the survey context is also referred to as the quasi-randomization approach~\citep{Elliott2017}.
+Its use in the survey context is also referred to as quasi-randomization~\citep{Elliott2017}.
 \cite{Valliant2011} consider a weighted logistic regression procedure using the pooled probability and nonprobability samples.
 \cite{Chen2019} consider a pseudo-log-likelihood approach that uses the random samples as a proxy for a term in the log-likelihood.
-Here, we extend this approach to account for measurement-error as well as random sampling at multiple times.
+Here, we extend this approach to account for measurement-error as well as observing random samples at multiple times.
 We then provide an extension of the statistical error decomposition and discuss the trade-offs inherent in such a weighting approach.
@@
 
- Recent work by \cite{Zhao2021} pointed out that estimation of key epidemiological parameters such as the incubation time using standard epidemiological models can suffer from severe bias due to issues beyond selection bias and measurement error.
-Namely, right truncation and epidemic growth leading to patients ``being more likely to be infected towards the end of their exposure period''~\citep[pp.
-3]{Zhao2021}.
-Their approach carefully constructs the study sample and statistical model to account for these issues.
-In this paper, rather than focusing on construction of a study sample, we ask whether one can collect auxiliary information to address selection bias in the observed case count data directly.
-
 Given a probability sampling design, individual predictions can be leveraged to improve estimation via model-assisted approaches~\citep{Breidt2017}.
@@
 
+Recent work by \cite{Zhao2021} pointed out that estimation of key epidemiological parameters such as the incubation time using standard epidemiological models can suffer from severe bias due to issues beyond selection bias and measurement error.
+Namely, right truncation and epidemic growth leading to patients ``being more likely to be infected towards the end of their exposure period''~\citep[pp.
+3]{Zhao2021}.
+Their approach carefully constructs the study sample and statistical model to account for these issues.
+In this paper, rather than focusing on construction of a study sample, we ask whether one can collect auxiliary information to address selection bias in the observed case count data directly.
+
  % \subsection{Outline}
@@
 
- In this section we provide the necessary background to understand COVID-19 diagnostic and serologic testing, discuss their respective scientific uses in managing the pandemic, and introduce the data streams considered in this manuscript.
+ Here we provide the necessary background to understand COVID-19 diagnostic testing, its scientific use in managing the pandemic, and the data streams considered in this manuscript.
 
@@
 
- Upon infection with SARS-CoV-2, there is an incubation period which lasts approximately five days~\citep{Lauer2020} before the viral load is high enough to be detected.
+ Upon infection with SARS-CoV-2, an incubation period starts and lasts approximately five days~\citep{Lauer2020} before the viral load is high enough to be detected.
 A \emph{molecular test} refers to diagnostic tests that aim to detect increased viral load (e.g., RT-PCR or antigen tests).
@@
 
- \begin{rmk}[Reporting Delays]
-
- \end{rmk}
+\begin{remark}[Reporting Delays]
+Figure~\ref{fig:in-data} shows clear reduced testing and case count reporting on weekends compared to weekdays.
+Moreover, COVID-19 tests are reported on the day they were administered, while case counts are reported based on the date the positive test was reported to and confirmed by the Indiana Department of Health system.
+To try and address both issues, analyses are conducted on the weekly level in this manuscipt.
+ \end{remark}
 
@@
  % https://www.wishtv.com/news/medical/indiana-opens-up-covid-19-testing-to-more-hoosiers/
- On June 15, 2020, ISDH lifted all restrictions and testing expanded to include anyone in Indiana who wants to be tested for coronavirus~\citep{indystar2020v2}.
+ On June 15th, 2020, ISDH lifted all restrictions and testing expanded to include anyone in Indiana who wants to be tested for coronavirus~\citep{indystar2020v2}.
  % https://www.indystar.com/story/news/health/2020/06/12/indiana-says-anyone-who-wants-coronavirus-test-can-get-one/3179151001/
@@
 
- On March 23, 2020, Indiana's governor issues a \emph{stay at home} order effective March 26th through April 5th.
+ On March 23rd, 2020, Indiana's governor issued a \emph{stay at home} order effective March 26th through April 5th.
 The order was extended until April 30th.
@@
 
- Due to testing restrictions and other potential selection biases, publicly reported COVID-19 case cound data may not be sufficient to understand active infection rates or the disease trajectory.
+ Due to testing restrictions and other potential selection biases, publicly reported COVID-19 case count data may not be sufficient to understand active infection rates or the disease trajectory.
 Here, we discuss two random samples that provide auxiliary information and may help address selection bias.
@@
 15,495 participants were contacted resulting in a final sample size of 3,658.
-The same demographic data as provided by ISDH for Indian molecular testing data was included (e.g., summary statistics on age, sex, and race).
-Participants in the probability sample were also asked if they experienced any COVID-19 compatible symptoms during the past 2 weeks or had shared a household with someone who had a positive test result for SARS-CoV-2.
+Demographic data was collected (e.g., summary statistics on age, sex, and race) as well as data on whether they experienced any COVID-19 compatible symptoms during the past 2 weeks or had shared a household with someone who had a positive test result for SARS-CoV-2.
 During May 2--3, 2020, an additional nonrandom sample of 898 individuals was also collected.
@@
  \includegraphics[width = 0.9\textwidth]{../figs/fbcovid19symptoms.png}
- \caption{Rate of reported fever per strata.
+ \caption{Rate of reported COVID-19 symptoms per strata.
 Daily rates were estimated using weighted method suggested by~\citep{delphisurvey} on each day separately and then smoothed over time using local-linear nonparametric regression.}
@@
 Here, like in survey methodology~\citep{Cochran77}, we treat COVID-19 status as a fixed but unknown quantity of interest.
-For simplicity, the dynamic nature of the outbreak and recoverability of individuals are initially ignored.
+For simplicity, we start by ignoring the dynamic nature of the outbreak and recoverability of individuals.
 We assume either individual $j$ is COVID-19 positive and $Y_j=1$ or is COVID-19 negative and $Y_j=0$.
@@
  First we require some additional notation.
-Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the outcome and $0$ when we observe the true outcome.
+Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the binary outcome and $0$ when we observe the true binary outcome.
 We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.
@@
 The first three terms continue to represent \emph{data quality}, \emph{data quantity}, and \emph{problem difficulty} respectively.
-The new term represents the \emph{imperfect testing adjustment} which is a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence, which is scaled by the factor $(1 - FP - FN)^{-1}$.
-
- Figure~\ref{fig:heatmap} shows that $D_M$, as a function of the relative frequency ($f_1/f_0$) and odds ratio, can be both positive and negative as well as a range of magnitudes~\citep{Beesley2020,Beesley2019,Smeden2019}.
-Assuming no measurement error, $D_M = 1$ so the relation between estimation and selection bias is simple, e.g., COVID-19 positive individuals more likely to receive test implies upward bias in prevalence estimates.
+The new term~$D_M$ represents the \emph{imperfect testing adjustment} which is a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence, which is scaled by the factor $(1 - FP - FN)^{-1}$.
+
+ Figure~\ref{fig:heatmap} shows that $D_M$, as a function of the relative frequency ($f_1/f_0$) and log odds ratio, can be both positive and negative as well as a range of magnitudes~\citep{Beesley2020,Beesley2019,Smeden2019}.
+Assuming no measurement error, $D_M = 1$ so the relation between estimation and selection bias is simple, e.g., if COVID-19 positive individuals were more likely to receive test then this implies upward bias in prevalence estimates.
 Under random testing (i.e., $f_0 = f_1$), $D_M = (1-FP-FN)^{-1}$ so measurement error simply magnifies the error.
@@
 
- Figure~\ref{fig:ratio-bias} displays the trajectory of the true ratio and the potential biased estimators under an SIR model~\citep{Pastor2001,Newman2002,Parshani2010} for the epidemic dynamics, with state evolution given by
+ Figure~\ref{fig:ratio-bias} displays the trajectory of the true ratio and the potential biased estimators under a susceptible-infected-recovered (SIR) model~\citep{Pastor2001,Newman2002,Parshani2010} for the epidemic dynamics, with state evolution given by
  \begin{equation}
@@
  \caption{Potential bias due to self-selection and measurement error from a deterministic SIR model with $\beta = 1.2$ (black) and $\gamma = 0.15$.
-Here we assume $f = 0.02$, $FP = 0.024$, $FN = 0.30$, and use a range of relative sampling fraction $M = f_1/f_0$.
-\textcolor{red}{FIX!}}
+Here we assume $f = 0.02$, $FP = 0.024$, $FN = 0.13$, and use a range of relative sampling fractions $M = f_1/f_0$.}
  \label{fig:rates}
@@
  Taking an expectation with respect to the sampling design yields~\eqref{eq:propensity}.
+Solving~\eqref{eq:auxinfoprob} can be done by iteratively re-weighted least squares (IRLS); see Section~\ref{app:irls} in the suppementary materials for algorithmic details.
+
  % Under simple random sampling and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the $\hat \pi_k = (d_k / \tilde d_k) (n / N)$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.
@@
 Low selection propensity imply large weights which can have deleterious impacts by certain observations dominating estimation.
-Similar to~\cite{Meng2018}, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from an IPW estimator are likely to put too much confidence in the sheer size of the data.
+Similar to~\cite{Meng2018}, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from an IPW estimator are likely to put too much confidence in the sheer data size.
 
@@
 Here, $I_{j,t}$ denotes self-selection into testing on day $t$, which is highly correlated with prior testing and results.
-For example, an individual who tests positive is unlikely to seek testing on subsequent days.
+For example, an individual who tests positive may be unlikely to seek testing in the subsequent few days/weeks.
 Moreover, an individual in a high prevalence area may be more likely to seek out testing.
@@
  Suppose there is a sequence of finite populations of size $N_{\nu}$ indexed by $\nu$.
-Each finite population has an associated non-probability at time $t$ of  size~$\tilde n$ and $L_\nu$ probability samples of fixed size $n$ drawn at equally spaced times $\{ t^\prime_l \}_{l=1}^L$ over the study window~$[0,T]$.
-Assuming the selection propensity model is correctly specified, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:ipw} shows that the IPW estimator is  consistent and calculates the estimator's variance.
+Each finite population has an associated non-probability sample at time $t$ of size~$\tilde n$ and $L_\nu$ probability samples of size $n$ drawn at equally spaced times $\{ t^\prime_l \}_{l=1}^L$ over the study window~$[0,T]$.
+Assuming the selection propensity model is correctly specified, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:ipw} shows that the IPW estimator is  consistent as the number of probability samples grows and calculates the estimator's variance.
 In Lemma~\ref{lemma:ipw}, sensitivity and specificity are not assumed known.
@@
  where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
-Under certain regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ where $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$.
+Under certain regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ where $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$ with $E[\cdot]$ and $\text{Var} (\cdot)$ are under the joint randomization of propensity score and sampling designs.
  \end{lemma}
@@
 
- Up to this point, the primary focus of this paper has been on selection bias in coronavirus case-count data from a survey sampling perspective.
+ Up to this point, the primary focus has been selection bias in coronavirus case-count data from a survey sampling perspective.
 Here, we consider compartmental model approaches from infectious disease epidemiology.
@@
 
- Let $s_t$, $e_t$, $i_t^{(T)}$, and $r_t$ denote the proportion of survivors, exposed, infected, and removed cases (i.e., including both recovered cases and deaths) at time $t$.
+ Let $s_t$, $e_t$, $i_t$, and $r_t$ denote the proportion of survivors, exposed, infected, and removed cases (i.e., including both recovered cases and deaths) at time $t$.
 The population-level SEIR dynamics are then given by the following set of differential equations:
@@
  \frac{\partial s_t}{\partial t} &= - \beta s_t i_t; \quad \frac{\partial e_t}{\partial t} = \beta s_t i_t - \sigma e_t; \quad \label{eq:seir} \\
- \frac{\partial i^{(T)}_t}{\partial t} &= \sigma e_t - \gamma i_t; \quad
+ \frac{\partial i_t}{\partial t} &= \sigma e_t - \gamma i_t; \quad
  \frac{\partial r_t}{\partial t} = - \gamma i_t. \nonumber
@@
  \caption{Indiana daily COVID-19 case count, testing, and death data by age strata}
+ \label{fig:in-data}
  \end{figure}
@@
  \end{tabular}
- \caption{Estimated point prevalence of active infection with SARS-CoV-2 by demographics in Indiana~\cite{Yiannoutsos2021}.}
+ \caption{Estimated point prevalence of active infection with SARS-CoV-2 by demographics in Indiana~\citep{Yiannoutsos2021}.}
  \label{tab:indiana}

### 2021-08-25 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
 See Appendix~\ref{app:asympderivations} for additional details.
+ \textcolor{red}{clarify what you are predicting as we go from no time to longitudinal setting. WE can't assume the survey at the time so we need the kernel again here.}
 
@@
  \begin{array}{c}
- \frac{1}{N} \left[ \sum_{j=1}^N \frac{I_{j,t}}{\pi_{j,t}} \left( \left( \frac{Y_{j,t} - FP}{1-FP-FN} - \mu(x_j)  \right)  - \mu_t \right) + \frac{1}{\sum_{j=1}^N \tilde I_j \tilde W_j} \sum_{j=1}^N  \tilde I_j \tilde W_j \hat \mu (x_j) \right] \\
+ \frac{1}{N} \left[ \sum_{j=1}^N \frac{I_{j,t}}{\pi_{j,t}} \left( \left( \frac{Y_{j,t} - FP}{1-FP-FN} - \hat \mu_{j,t}  \right)  - \mu_t \right) + \frac{1}{\sum_{j=1}^N \tilde I_j \tilde W_j} \sum_{j=1}^N  \tilde I_j \tilde W_j \hat \mu_t (X_{j,t}) \right] \\
  \frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
@@
  where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
-Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$ and $\text{var} (\hat y_{n,t}) = V_{t}^{(DR)} + o (n^{-1})$ where $V_t^{(DR)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$.
+Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$ and $\text{var} (\hat y_{n,t}) = V_{t}^{(DR)} + o (n^{-1})$ where $V_t^{(DR)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$ with $E[\cdot]$ and $\text{Var} (\cdot)$ are under the joint randomization of propensity score and sampling designs.
  \end{lemma}
@@
  \begin{tabular}{c | c c c | c}
- & \multicolumn{3}{c}{Facebook} & Random \\ \cline{2-5}
+ & \multicolumn{3}{c}{Facebook} & Random \\ \cline{2-4}
  & Lower CI & Estimate & Upper CI & Estimate \\ \hline

### 2021-08-25 methods/6_invweight_estimation_alt.R
paper almost done

@@ -1,6 +1,6 @@
 indiana_data = readRDS("../data/weeklycoviddata_withsymptoms_alt.RDS")
 indiana_data$startdate = mdy(indiana_data$startdate)
-propensities = readRDS("../data/smoothedpropensities_alt.RDS")
+propensities = readRDS("../data/smoothedpropensities_alt_08252021.RDS")
 
 propensities$date = MMWRweek::MMWRweek2Date(MMWRyear = propensities$year,
                                             MMWRweek = propensities$week,

### 2021-08-25 visualizations/tv_propensities_alt.R

@@ -3,7 +3,7 @@ library(MMWRweek)
 library(ggplot2)
 library("RColorBrewer")
 
-propensities = readRDS("../data/smoothedpropensities_alt.RDS")
+propensities = readRDS("../data/smoothedpropensities_alt_08252021.RDS")
 
 propensities$date = MMWRweek::MMWRweek2Date(MMWRyear = propensities$year,
                                             MMWRweek = propensities$week,
@@ -22,8 +22,8 @@ ggplot(data = propensities, aes(x = date, y = probs)) +
   geom_point(size = 5) +
   labs(x = "Date",
        y = "COVID-19 Testing Propensity",
-       title = "35-44 year old non-Hispanic, White Male with Fever") +
-  theme(text = element_text(size=25))
+       title = "35-44, non-Hispanic, White Male with Symptoms") +
+  theme(text = element_text(size=20))
 
 dev.off()
 
@@ -61,6 +61,8 @@ propensities_long$propensity = propensities_long$probs1 * (propensities_long$age
 
 propensities_long_fever = propensities_long
 
+my_palette <- brewer.pal(name="Greys",n=9)[3:(3+5)]
+
 png(filename = "../figs/tvprop_alt_fig1_mainpaper.png",
     width = 960, height = 480, units = "px", pointsize = 25)
 
@@ -68,7 +70,7 @@ ggplot(data = propensities_long, aes(x = date, y = propensity, col = age)) +
   geom_point(size = 5) +
   labs(x = "Date",
        y = "COVID-19 Testing Propensity",
-       title = "Non-Hispanic, White Male with Fever") +
+       title = "Non-Hispanic, White Male with Symptoms") +
   theme(text = element_text(size=25)) +
   scale_color_manual(values=my_palette)
 
@@ -87,8 +89,8 @@ ggplot(data = propensities, aes(x = date, y = probs)) +
   geom_point(size = 5) +
   labs(x = "Date",
        y = "Testing Propensity",
-       title = "65-74 year old non-Hispanic, White Female without fever") +
-  theme(text = element_text(size=25))
+       title = "65-74, non-Hispanic, White Female without Symptoms") +
+  theme(text = element_text(size=20))
 
 dev.off()
 
@@ -133,7 +135,7 @@ ggplot(data = propensities_long, aes(x = date, y = propensity, col = age)) +
   geom_point(size = 5) +
   labs(x = "Date",
        y = "COVID-19 Testing Propensity",
[…]

### 2021-08-25 visualizations/tv_symptom_propensities_alt.R

@@ -33,7 +33,7 @@ ggplot(data = propensities_neg_long, aes(x = date, y = probs, col = Gender)) +
        y = "Likelihood of Symptoms",
        title = "Individuals 35-44, Negative Test")+ 
   scale_color_manual(values = my_palette) +
-  theme(text = element_text(size=25))
+  theme(text = element_text(size=15))
 
 dev.off()
 
@@ -66,6 +66,6 @@ ggplot(data = propensities_pos_long, aes(x = date, y = probs, col = Hospitalized
        y = "Likelihood of Symptoms",
        title = "Individuals 35-44, Positive test") + 
   scale_color_manual(values = my_palette) +
-  theme(text = element_text(size=25))
+  theme(text = element_text(size=15))
 
 dev.off()

### 2021-08-25 write-up/covid-writeup.tex
paper almost done
Paper done outside final checks (need to carefully explain why analysis isnt great)

@@
  \subsubsection{Asymptotics}
+ \label{section:asymptotics}
 
@@
  Unfortunately, the first term of the doubly robust estimator cannot be computed as covariate information is not collected on the entire population.
-Here, we again use the probability samples and consider the doubly-robust estimator
- $$
- \frac{1}{\sum_{j=1}^N \tilde I_j \tilde W_j} \sum_{j=1}^N  \tilde I_j \tilde W_j \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \mu(x_j) \right).
- $$
- where $\tilde I_j$ indicates an individual is included in the random sample and $\tilde W_j$ is the associated weight for that unit.
-Assuming the selection propensity model is correctly specified, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:dr} shows that the IPW estimator is  consistent and calculates the estimator's variance.
+Here, we suppose the same asymptotic regime as in Section~\ref{section:asymptotics} and therefore use the probability samples and consider the doubly-robust estimator at time~$t$
+ \begin{align*}
+ &\frac{1}{\sum_{l=1}^L K_{t,t_l^\prime} \sum_{j=1}^N \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime}} \sum_{l=1}^L K_{t,t_l^\prime} \sum_{j=1}^N  \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime} \hat \mu (X_{j,t^\prime}) \\
+ + &\quad \frac{1}{\sum_{j=1}^N I_j w (X_{j,t})} \sum_{j=1}^N I_j w(X_{j,t}) \left( \frac{Y_{j,t} - FP}{1 - FP - FN} - \mu_t(X_{j,t}) \right).
+ \end{align*}
+ where $\tilde I_{j,t_l^\prime}$ indicates an individual is included in the random sample at time~$t_l^\prime$ and $\tilde W_{j,t_l^\prime}$ is the associated weight for that unit.
+Assuming the selection propensity model is correctly specified, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:dr} shows that the doubly robust estimator is consistent and calculates the estimator's variance.
 See Appendix~\ref{app:asympderivations} for additional details.
- \textcolor{red}{clarify what you are predicting as we go from no time to longitudinal setting. WE can't assume the survey at the time so we need the kernel again here.}
 
@@
 We note here that the ratio of testing propensity within strata across fever status is time-varying.
-The ratio starting nine times more likely in early April and dropping to twice as likely by end of 2020.
+The ratio starting 50 times more likely in early April and dropping to ten times as likely by end of 2020.
 
@@
 
- The SEIR model in equation~\eqref{eq:seir} is extended to allow for time-varying transmission rates~$\beta_t$ to account for public policy changes discussed in Section~\ref{section:publicpolicyindiana} over the time period considered.
-Here, we insert three change points $(t_1,t_2,t_3)$ on March 23rd, June 15th, and October 1st of 2020 respectively.
-Given sensitivity to these choices, instead of considering a multiplicative structure, we allow for delayed implementation and slow change in the transmission rate by using weights:
- $$
- \beta_t = \sum_{k=1}^{3} \left(\beta_{k-1} (1-w^{(k)}_t ) + \beta_{k} w_t\right),\quad w^{(k)}_t = (1 + \exp \left( -\xi (t - t_k - \nu) \right))^{-1}
- $$
- where $\nu \geq 0$ is the potential delay and $\xi$ is the rate of change.
-Each weight~$w_t^{(k)} = 0$ for $t < t_k$ and $w_t^{(k)} \to 1$ as $t \to \infty$.
-See Section~\ref{app:prior_modelbased} in the supplementary materials for details on prior specification.
-
- To check how well our model fits the observed data, a posterior predictive check on aggregate death counts was performed.
-Figure~\ref{fig:deathsppc} shows the results which suggests chosen SEIR model fits the observed data well.
-The model produces age-specific case counts which can then be used for model-based estimation of the active infection rate as well as in constructing the doubly robust estimator.
-
  \begin{figure}[!th]
@@
 
+ The SEIR model in equation~\eqref{eq:seir} is extended to allow for time-varying transmission rates~$\beta_t$ to account for public policy changes discussed in Section~\ref{section:publicpolicyindiana} over the time period considered.
+Here, we insert three change points $(t_1,t_2,t_3)$ on March 23rd, June 15th, and October 1st of 2020 respectively.
+Given sensitivity to these choices, instead of considering a multiplicative structure, we allow for delayed implementation and slow change in the transmission rate by using weights:
+ $$
+ \beta_t = \sum_{k=1}^{3} \left(\beta_{k-1} (1-w^{(k)}_t ) + \beta_{k} w_t\right),\quad w^{(k)}_t = (1 + \exp \left( -\xi (t - t_k - \nu) \right))^{-1}
+ $$
+ where $\nu \geq 0$ is the potential delay and $\xi$ is the rate of change.
+Each weight~$w_t^{(k)} = 0$ for $t < t_k$ and $w_t^{(k)} \to 1$ as $t \to \infty$.
+See Section~\ref{app:prior_modelbased} in the supplementary materials for details on prior specification.
+
+ To check how well our model fits the observed data, a posterior predictive check on aggregate death counts was performed.
+Figure~\ref{fig:deathsppc} shows the results which suggests chosen SEIR model fits the observed data well.
+The model produces age-specific case counts which can then be used for model-based estimation of the active infection rate as well as in constructing the doubly robust estimator.
+
  \subsection{Time-varying prevalence estimates}
@@
  Here, we construct active infection rate estimates using the unweighted, inverse-probability weighted, model-based, and doubly robust estimators.
-Figure~\ref{fig:air} presents the results.
+Figure~\ref{fig:tvestimates} presents the results.
 The inverse-probability weighting methods do not significantly reduce bias compared to the unweighted estimates.
-We conjecture this is due to the limited availability of symptom and other important covariates that may impact the likelihood an individual will get  tested.
-Our imputation strategies use available data to try and alleviate this issue; however, this suggests better data collection in the future may strongly improve the IPW estimates as seen in Section~\ref{section:ipwapproach} when analyzing the statewide sample from April 25th--29th.
+We conjecture the poor performance is due to the limited availability of symptom and other important covariate information to estimate strata-specific infection rates.
+This suggests better data collection in the future may strongly improve the IPW estimates as seen in Section~\ref{section:ipwapproach} when analyzing the statewide sample from April 25th--29th.
 
@@
 
- \begin{figure}
- \centering
- \begin{subfigure}{.5\textwidth}
-   \centering
-   \includegraphics[width=.9\linewidth]{../figs/tv_air.png}
-   \caption{Active Infection Rate Estimates}
-   \label{fig:air}
- \end{subfigure}%
- \begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{../figs/tv_air_ratio.png}
-  \caption{Ratio estimates}
-  \label{fig:air_ratio}
- \end{subfigure}
- \caption{Time-varying estimates based on unweighted, model-based, IPW and doubly robust methods.}
- \label{fig:tvestimates}
- \end{figure}
-
- We also construct ratio estimates as discussed in Section~\ref{section:rates}.
+  % We also construct ratio estimates as discussed in Section~\ref{section:rates}.
 Figure~\ref{fig:air_ratio} presents the results.
@@
 The model-based methods may be as well.
+
+  \begin{figure}[!th]
+ % \centering
+ % \begin{subfigure}{.5\textwidth}
+   \centering
+   \includegraphics[width=.6\linewidth]{../figs/tv_air.png}
+   % \caption{Active Infection Rate Estimates}
+   % \label{fig:air}
+ % \end{subfigure}%
+ % \begin{subfigure}{.5\textwidth}
+ %  \centering
+ %  \includegraphics[width=.9\linewidth]{../figs/tv_air_ratio.png}
+ %  \caption{Ratio estimates}
+ %  \label{fig:air_ratio}
+ % \end{subfigure}
+ \caption{Time-varying active infectino rate estimates based on unweighted, model-based, IPW and doubly robust methods.}
+ \label{fig:tvestimates}
+ \end{figure}
 
@@
 
- \subsection*{High quality information on prevalence.}
-
- Governments are pushing for increased testing capacity and robust contact tracing.
+ % \subsection*{High quality information on prevalence.}
+
+ % Governments are pushing for increased testing capacity and robust contact tracing.
 Contact tracing is incredibly useful for identifying carriers early and preventing spread of the disease.
@@
  While we are eternally grateful to Indiana for the release, symptom information was not made available.
-This severely limits our ability to estimate correct selection propensity models.
-What is required in the future is a more detailed effort to collect relevant measures of an individual's past history that may impact their likelihood of seeking testing and being a positive COVID-19 case.
+This severely limits our ability to estimate correct selection propensity models and strata-specific infection rates.
+A more sustained effort to collect relevant measures of an individual's past history that may impact their likelihood of seeking testing and being COVID-19 positive is required.
 Such data collection and dissemination is critical to address selection bias.
@@
 Moreover, equation~\eqref{eq:auxinfoprob} presupposes covariate information is collected for all individuals in the nonprobability sample; however, this is often not feasible for state and local governments where rapid sample collection is prioritized over additional collection of covariate information.
-We can easily extend~\eqref{eq:auxinfoprob} by randomly sampling a subset of the nonprobability sample on which to collect the covariate information, yielding a new pseudo-likelihood
- \begin{equation*}
- % \label{eq:auxinfoprob2}
- \sum_{j=1}^N I_j I_j^{(2)} W_j^{(2)} \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
- \end{equation*}
- Here, $I_j^{(2)}$ is an indicator that the individual is in the probability subsample and $W_j^{(2)}$ is the probability of inclusion.
+We can easily extend~\eqref{eq:auxinfoprob} by randomly sampling a subset of the nonprobability sample on which to collect the covariate information.
+ %yielding a new pseudo-likelihood
+ % \begin{equation*}
+ % % \label{eq:auxinfoprob2}
+ % \sum_{j=1}^N I_j I_j^{(2)} W_j^{(2)} \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
+ % \end{equation*}
+ % Here, $I_j^{(2)}$ is an indicator that the individual is in the probability subsample and $W_j^{(2)}$ is the probability of inclusion.
 Estimation proceeds by maximizing the pseudo-likelihood which now depends on random sampling in both the first and second term.

### 2021-08-25 methods/6_invweight_var_estimation.R

@@ -1,54 +1,0 @@
-indiana_data = readRDS("../data/weeklycoviddata_withsymptoms_alt.RDS")
-indiana_data$startdate = mdy(indiana_data$startdate)
-propensities = readRDS("../data/smoothedpropensities_alt.RDS")
-
-propensities$date = MMWRweek::MMWRweek2Date(MMWRyear = propensities$year,
-                                            MMWRweek = propensities$week,
-                                            MMWRday = 1)
-## First term can be precomputed given design matrix
-model = ~ -1+as.factor(symptoms) + as.factor(gender) + ethnicity + race + as.factor(age)
-indiana_X = model.matrix(model, indiana_data)
-
-library(lubridate)
-indiana_data$startdate = ymd(indiana_data$startdate)
-indiana_data$week = week(indiana_data$startdate)
-indiana_data$year = year(indiana_data$startdate)
-
-invweight = vector(length = nrow(indiana_data))
-for(i in 1:nrow(indiana_data)) {
-  modelmatrix_row = indiana_X[i,]
-  week_row = as.numeric(week(indiana_data$startdate[i]))
-  year_row = as.numeric(year(indiana_data$startdate[i]))
-  if(week_row > 5 & year_row == 2021) {
-    invweight[i] = NA
-  } else {
-    prop_row = which(propensities$week == week_row & propensities$year == year_row)
-    term = as.matrix(propensities[prop_row,3:15])%*%modelmatrix_row
-    invweight[i] = (1+exp(-term))
-  }
-}
-
-indiana_data$invweight = invweight
-
-current_week = 48
-current_year = 2020
-
-weeks = c(14:53,1:5)
-years = c(rep(2020, length = length(c(14:53))),rep(2021, length = length(1:5)))
-results = matrix(nrow = length(weeks), ncol = 4) 
-
-for(i in 1:length(weeks)) {
-  current_week = weeks[i]
-  current_year = years[i]
-  print(paste("At week", current_week, "in year", current_year))
-  current_counts = indiana_data$covid_counts[which(indiana_data$week == current_week & indiana_data$year == current_year)]
-  current_tests = indiana_data$covid_tests[which(indiana_data$week == current_week & indiana_data$year == current_year)]
-  weights = indiana_data$invweight[which(indiana_data$week == current_week & indiana_data$year == current_year)]
-  results[i,1] = current_week
-  results[i,2] = current_year
[…]

### 2021-08-25 preprocessing/8_modelbased_preproc.R

@@ -68,6 +68,7 @@ summary_fit$year = year(summary_fit$timepoints)
 
 summary_aggregate = aggregate(mean ~ age_group + week + year, data = summary_fit, FUN = sum)
 
+
 ## WE NOW NEED DEATH FRACTIONS PER SUBGROUP
 ## We DO THIS AS A KERNEL ESTIMATOR 
 levels(df_coviddeath$age) = c(1,1,1:6)
@@ -227,7 +228,7 @@ df_coviddeath <- aggregate(covid_deaths ~ startdate, data = df_coviddeath, FUN =
 df_coviddeath$date = ymd(df_coviddeath$startdate)
 df_coviddeath$death_dt = df_coviddeath$covid_deaths
 
-# Swiss population
+# Indiana population
 N <- 6.732E6;
 
 # Deaths

### 2021-08-25 visualizations/tv_prevalence.R

@@ -105,7 +105,7 @@ ggplot(data = ratio_long, aes(x = date, y = estimate, col = Method)) +
   labs(x = "Date",
        y = "Active Infection Rate Estimate") + 
   theme(text = element_text(size=25)) +
-  scale_color_manual(values=wes_palette(n = 5, "IsleofDogs1"))
+  scale_color_manual(values=my_palette)
 
 dev.off()
 

### 2021-08-27 methods/6_invweight_variance_alt.R
Updated prevs!

@@ -1,0 +1,129 @@
+## Distance Functions
+
+indiana_distance <- function(current_week, current_year) {
+  ind_dist <- function(row) {
+    rowyear = year(row$startdate)
+    rowweek = week(row$startdate)
+    if (rowyear == current_year){
+      dis = abs(rowweek - current_week)
+    } else {
+      if (rowyear == 2020) {
+        dis = (53 - rowweek) + current_week
+      } else {
+        dis = (rowweek) + (53-current_week)
+      }
+    }
+    return(dis)
+  }
+  return(ind_dist)
+}
+
+fb_distance <-  function(current_week, current_year) {
+  fb_dist <- function(row) {
+    rowyear = as.numeric(row[7])
+    rowweek = as.numeric(row[6])
+    if (rowyear == current_year){
+      dis = abs(rowweek - current_week)
+    } else {
+      if (rowyear == 2020) {
+        dis = (53 - rowweek) + current_week
+      } else {
+        dis = (rowweek) + (53-current_week)
+      }
+    }
+    return(dis)
+  }
+  return(fb_dist)
+}
+
+kernelweight <- function(distance) {
+  kw2 <- function(row) {
+    dis = distance(row)
+    h = sqrt(-1/(2*log(0.7))) # Set a deterministic bandwidth for now
+    kernel = exp(-dis^2/(2*h^2))
+    return(kernel)
+  }
+  return(kw2)
+}
+
+
+## INV WEIGHT COMPUTATION
+indiana_data = readRDS("../data/weeklycoviddata_withsymptoms_alt.RDS")
+propensities = readRDS("../data/smoothedpropensities_alt_08262021.RDS")
+estimates = readRDS("../data/invweights_alt_08262021.RDS")
+estimates = data.frame(estimates)
+names(estimates) = c("week", "year", "unweighted", "ipw")
+
+library(MMWRweek)
+library(lubridate)
+
+propensities$date = MMWRweek::MMWRweek2Date(MMWRyear = propensities$year,
+                                            MMWRweek = propensities$week,
+                                            MMWRday = 1)
+## First term can be precomputed given design matrix
+model = ~ -1+as.factor(symptoms) + as.factor(gender) + ethnicity + race + as.factor(age)
+indiana_X = model.matrix(model, indiana_data)
+
+library(lubridate)
[…]

### 2021-08-27 methods/9_dr_estimation.R

@@ -10,7 +10,7 @@ library(MMWRweek)
 ## INDIANA AND PROPENSITIES
 indiana_data = readRDS("../data/weeklycoviddata_withsymptoms_alt.RDS")
 indiana_data$startdate = mdy(indiana_data$startdate)
-propensities = readRDS("../data/smoothedpropensities_alt.RDS")
+propensities = readRDS("../data/smoothedpropensities_alt_08262021.RDS")
 
 ### MODEL BASED ESTIMATES
 aggregate_air = readRDS("../data/aggregate_air.RDS")
@@ -114,6 +114,8 @@ current_year = 2020
 weeks = c(14:53,1:5)
 years = c(rep(2020, length = length(c(14:53))),rep(2021, length = length(1:5)))
 results = matrix(nrow = length(weeks), ncol = 5) 
+FP= 0.024
+FN = 0.13
 
 for(i in 1:length(weeks)) {
   current_week = weeks[i]
@@ -127,10 +129,10 @@ for(i in 1:length(weeks)) {
   fb_air = sum(fb_data$air * fb_data$weight, na.rm = TRUE)/ sum(fb_data$weight, na.rm = TRUE)
   results[i,1] = current_week
   results[i,2] = current_year
-  results[i,3] = sum(current_counts)/sum(current_tests)
-  results[i,4] = sum(current_counts*weights)/sum(current_tests*weights)
-  results[i,5] =  fb_air + (results[i,4] - invweighted_air)
+  results[i,3] = (sum(current_counts)/sum(current_tests) - FP)/(1-FP-FN)
+  results[i,4] = (sum(current_counts*weights)/sum(current_tests*weights) - FP)/(1-FP-FN)
+  results[i,5] = fb_air + (results[i,4] - invweighted_air)
 }
 
-saveRDS(results,"../data/drestimates_alt_071521.RDS")
+saveRDS(results,"../data/drestimates_alt_082621.RDS")
 

### 2021-08-27 visualizations/tv_prevalence.R

@@ -69,7 +69,7 @@ ratio_long = rbind(ratio_long, ratio_temp)
 prevalence_long = rbind(prevalence_long, prevalence_temp)
 
 ## ADD IN DOUBLY ROBUST
-prevalence_temp = readRDS("../data/drestimates_alt_071521.RDS")
+prevalence_temp = readRDS("../data/drestimates_alt_082621.RDS")
 prevalence_temp = data.frame(date = dates, Method = rep("Doubly Robust", nrow(prevalence_temp)), 
                              estimate = prevalence_temp[,5])
 ratio = prevalence_temp$estimate[2:length(prevalence_temp$estimate)]/prevalence_temp$estimate[1:(length(prevalence_temp$estimate)-1)]
@@ -81,9 +81,6 @@ ratio_temp = data.frame(date = date, Method = rep("Doubly Robust", length(date))
                         estimate = ratio)
 ratio_long = rbind(ratio_long, ratio_temp)
 prevalence_long = rbind(prevalence_long, prevalence_temp)
-
-
-prevalence_long = prevalence_long[prevalence_long$Method != "Doubly Robust",] # JUST FOR NOW
 
 ## FINAL FIGURES
 png(filename = "../figs/tv_air.png",

### 2021-08-30 write-up/covid-writeup.tex
Updated prevs!
final major paper update

@@
 We then combine these selection propensities with epidemiological forecast models to construct a \emph{doubly robust} estimation method that accounts for both measurement-error and selection bias.
-This method is then applied to estimate Indiana's prevalence using case-count, hospitalization, and death data with cumulative demographic information, the United States's only statewide random molecular sample collected from April 25--29th, and Facebook's COVID-19 symptom survey.
+This method is then applied to estimate Indiana's prevalence using case-count, hospitalization, and death data with cumulative demographic information, one of United States's only two statewide random molecular sample collected from April 25--29th, and Facebook's COVID-19 symptom survey.
 We end with a series of recommendations based on the proposed methodology.
@@
 Due to monetary and time constraints, however, random testing is performed infrequently.
-As of June 2021, to the best of our knowledge, Indiana is the only state to conduct statewide random sample testing\footnote{This is the only random sample to collect both seroprevalence and diagnostic testing results.
+As of June 2021, to the best of our knowledge, Indiana and Ohio are the only states to conduct statewide random sample testing\footnote{This is the only random sample to collect both seroprevalence and diagnostic testing results.
 The CDC has conducted several seroprevalence studies.
@@
 
- \begin{figure}[!th]
+  \begin{figure}
  \centering
- \includegraphics[width=.6\linewidth]{../figs/deaths_ppc.png}
- \caption{Posterior predictive check on aggregate daily death counts.}
+ \begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{../figs/deaths_ppc.png}
+ \caption{PPC on aggregate daily death counts.}
  \label{fig:deathsppc}
+ \end{subfigure}%
+ \begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{../figs/cases_ppc.png}
+ \caption{Posterior on daily case counts.}
+ \label{fig:casesppc}
+ \end{subfigure}
+ \caption{Posterior predictive check (PPC) on death counts and posterior distribution of case counts.}
+ \label{fig:ppcs}
  \end{figure}

### 2021-08-30 methods/6_invweight_variance_alt.R

@@ -49,6 +49,9 @@ kernelweight <- function(distance) {
 
 ## INV WEIGHT COMPUTATION
 indiana_data = readRDS("../data/weeklycoviddata_withsymptoms_alt.RDS")
+fb_data = readRDS("../data/fb_weeklycomplete.RDS")
+names(fb_data)[3] = "symptoms"
+
 propensities = readRDS("../data/smoothedpropensities_alt_08262021.RDS")
 estimates = readRDS("../data/invweights_alt_08262021.RDS")
 estimates = data.frame(estimates)
@@ -69,61 +72,178 @@ indiana_data$startdate = mdy(indiana_data$startdate)
 indiana_data$week = week(indiana_data$startdate)
 indiana_data$year = year(indiana_data$startdate)
 
-current_week = 42
-current_year = 2020
-ind_dist <- indiana_distance(current_week, current_year)
-## COMPUTE INDIANA DATA TERMS
+set_of_weeks = c(14:52, 1:5)
+set_of_years = c(rep(2020, length(14:52)), rep(2021, length(1:5)))
+results = rep(0,0)
 
-invweight = vector(length = nrow(indiana_data))
-top_row = rep(0, ncol(indiana_X)+1)
-for(i in 1:nrow(indiana_data)) {
-  modelmatrix_row = indiana_X[i,]
-  ind_dist(indiana_data[i,])
-  week_row = as.numeric(week(indiana_data$startdate[i]))
-  year_row = as.numeric(year(indiana_data$startdate[i]))
-  total_tests = indiana_data$covid_tests[i]
-  pos_tests = indiana_data$covid_counts[i]
-  neg_tests = total_tests - pos_tests
-  prev_estimate = estimates$ipw[estimates$week == week_row & estimates$year == current_year]
-  if(week_row > 5 & year_row == 2021) {
-    invweight[i] = NA
-  } else {
-    prop_row = which(propensities$week == week_row & propensities$year == year_row)
-    term = as.matrix(propensities[prop_row,3:15])%*%modelmatrix_row
-    invweight[i] = (1+exp(-term))
-  }
-  temp_pi = 1/invweight[i]
-  temp_odds = (1-temp_pi)/temp_pi
-  constant = modelmatrix_row*temp_odds
-  top_row[1] = top_row[1] + -(1-FP-FN) * invweight[i]
[…]

### 2021-08-31 write-up/covid-supplement-writeup.tex
final major paper update
Updated IFR to match Raftery

@@
 
-Then by a first-order Taylor expansion we have $\hat \eta - \eta_0 = \left[ \phi (\eta) \right]^{-1} \Phi(\eta_0) + o_p (\tilde n^{-1/2})$ where $\phi(\eta) = \partial \Phi (\eta)/\partial \eta$ and $\tilde n = n \sum_{l=1}^L \tilde K_{t,t_l}$.
+Then by a first-order Taylor expansion we have $\hat \eta - \eta_0 = \left[ \phi (\eta) \right]^{-1} \Phi(\eta_0) + o_p (\bar n^{-1/2})$ where $\phi(\eta) = \partial \Phi (\eta)/\partial \eta$ and $\bar n = n \sum_{l=1}^L K_{t,t_l}$.
 To calculate
@@
 \begin{array}{c c c c}
-\frac{(1-\pi_{j,t})}{\pi_{j,t}} (Y_{j,t} - (1-FP-FN) (\mu_{j,t} +FP))^2 & \sum_{t^\prime} K_{t,t^\prime} (1-\pi_{j,t^\prime}) \zeta_{t^\prime, j} X_{j,t^\prime}^\top & 0 & 0 \\
+\frac{(1-\pi_{j,t})}{\pi_{j,t}} (Y_{j,t} - FP - (1-FP-FN) \mu_{j,t})^2 & \sum_{t^\prime} K_{t,t^\prime} (1-\pi_{j,t^\prime}) \zeta_{t^\prime, j} X_{j,t^\prime}^\top & 0 & 0 \\
 \sum_{t^\prime} K_{t,t^\prime} (1-\pi_{j,t^\prime}) \zeta_{t^\prime, j} X_{j,t}&
@@
 
+Figure~\ref{fig:air_cis} in the supplementary materials presents the confidence intervals per time point for the IPW2 estimator.
+The confidence interval length decreases substantially over time, reflecting the increased testing capacity.
+Due to the number of surveys per week, under correct model specification there is minimal uncertainty in the parameter estimates.
+As the number of tests per week increases to over one hundred thousand, there is minimal uncertainty in the active infection rate estimates.
+This points to the importance of the statistical decomposition~\eqref{eq:statdecomp2} and the discussion in Section~\ref{section:IPWerrordecomp}.
+
+\begin{figure}[!th]
+ \centering
+ \includegraphics[width=.9\linewidth]{../figs/tv_air_cis.png}
+ \caption{IPW2 estimate with confidence intervals}
+ \label{fig:air_cis}
+\end{figure}
+
 % \begin{figure}[!th]
@@
 
+Figure~\ref{fig:undercounting} below is a plot of undercount based on the SEIR model.  We see the undercount goes away by January; however, this points to potential misspecification of the IFR.
+
+\begin{figure}[!th]
+ \centering
+ \includegraphics[width=.7\linewidth]{../figs/undercounting.png}
+ \caption{Undercount based on SEIR model}
+ \label{fig:undercounting}
+\end{figure}
+
 \bibliographystyle{plainnat}

### 2021-08-31 write-up/covid-writeup.tex
Updated IFR to match Raftery

@@
 We then combine these selection propensities with epidemiological forecast models to construct a \emph{doubly robust} estimation method that accounts for both measurement-error and selection bias.
-This method is then applied to estimate Indiana's prevalence using case-count, hospitalization, and death data with cumulative demographic information, one of United States's only two statewide random molecular sample collected from April 25--29th, and Facebook's COVID-19 symptom survey.
+This method is then applied to estimate Indiana's prevalence using case-count, hospitalization, and death data with cumulative demographic information, a statewide random molecular sample collected from April 25--29th, and Facebook's COVID-19 symptom survey.
 We end with a series of recommendations based on the proposed methodology.
@@
 Due to monetary and time constraints, however, random testing is performed infrequently.
-As of June 2021, to the best of our knowledge, Indiana and Ohio are the only states to conduct statewide random sample testing\footnote{This is the only random sample to collect both seroprevalence and diagnostic testing results.
-The CDC has conducted several seroprevalence studies.
-See Section~\ref{section:testinginfo} for further discussion.}, which was collected from April 25--29, 2020~\citep{Yiannoutsos2021}.
+As of June 2021, Indiana and Ohio are the only states to conduct statewide random sample testing\footnote{These are is the only random samples to collect both seroprevalence and diagnostic testing results.
+The CDC and other states have conducted seroprevalence-only studies.}.
+Indiana's sample was collected from April 25--29, 2020~\citep{Yiannoutsos2021}.
 Such infrequent random testing is likely to provide insufficient information to help researchers and policy makers better understand the disease trajectory which can change rapidly over time.
@@
 While the impact of measurement error~\citep{Smeden2019} and selection bias~\citep{Keiding2016} on estimation are both well-studied topics, here we provide a new perspective by building on the work of \cite{Meng2018} who studied an error decomposition to understand the relationship between selection bias and population size.
-Speficially, we quantify the interaction between measurement-error and selection bias on statistical error, showing how the sign and magnitude can change drastically.
+Specifically, we quantify the interaction between measurement-error and selection bias on statistical error, showing how the sign and magnitude can change drastically.
  % the impact on the effective sample size can therefore be quite large.
@@
 
- One core component of coronavirus research is epidemiological state-space modelling of case-count and death data.
+ One core component of coronavirus research is epidemiological compartmental modelling of case-count and death data.
 These models can be used to answer a variety of research questions including case-count forecasting, estimation of the effective reproduction number, and estimation of quarantine and other health policies on infectious disease dynamics.
@@
 Demographic data was collected (e.g., summary statistics on age, sex, and race) as well as data on whether they experienced any COVID-19 compatible symptoms during the past 2 weeks or had shared a household with someone who had a positive test result for SARS-CoV-2.
-During May 2--3, 2020, an additional nonrandom sample of 898 individuals was also collected.
+During May 2--3, 2020, an additional non-random sample of 898 individuals was also collected.
  % In this paper, the nonrandom sample is used to assess whether the cumulative summary statistics are appropriate in constructing the inverse-probability weights as well as in constructing appropriate weights using symptom information.
@@
 
- Here we assume access to a probability sample measuring both the outcome and exact same set of covariates.
+ Here we assume access to a probability sample measuring the same set of covariates.
 Let $\tilde I_j$ be an indicator that the individual was included in the probability sample and $\tilde W_j$ be the probability of inclusion for the $j$th individual (i.e., equal to $n/N$ for a simple random sample of size $n$).
@@
  Taking an expectation with respect to the sampling design yields~\eqref{eq:propensity}.
-Solving~\eqref{eq:auxinfoprob} can be done by iteratively re-weighted least squares (IRLS); see Section~\ref{app:irls} in the suppementary materials for algorithmic details.
+Solving~\eqref{eq:auxinfoprob} can be done by iteratively re-weighted least squares (IRLS); see Section~\ref{app:irls} in the supplementary materials for algorithmic details.
 
@@
  $$
- where $K_h$ is a kernel function with tunable parameter $h$. Given $\pi (x;\hat \theta_t)$, the prevalence estimator $\bar y_{n,t}^\star$ is given by~\ref{eq:ipwest}.
+ where $K_h$ is a kernel function with tunable parameter $h$.
+Given $\pi (x;\hat \theta_t)$, the prevalence estimator $\bar y_{n,t}^\star$ is given by~\ref{eq:ipwest} using case-count data observed on day~$t$.
 
@@
  where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
-Under certain regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ where $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$ with $E[\cdot]$ and $\text{Var} (\cdot)$ are under the joint randomization of propensity score and sampling designs.
+Under certain regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (\bar n^{-1/2})$  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (\bar n^{-1})$ where $\bar n = n \sum_{l=1}^L K_{t,t_l}$ and $V_t^{(IPW)}$ is the first diagonal element of $E [\Phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\Phi_n(\eta_0)]^{-1}$ with $E[\cdot]$ and $\text{Var} (\cdot)$ are under the joint randomization of propensity score and sampling designs.
  \end{lemma}
@@
  See Appendix~\ref{app:DRderivation} for the derivation.
-Comparing~\eqref{eq:statdecomp3} to~\eqref{eq:statdecomp2} shows that the addition of the model-based forecast impacts the estimation error in two ways.
-First, if the model is adequate, then one may expect a reduction in the problem difficulty and (potentially) in the data quality components.
-Note that the model-based forecast does not impact the data quantity component.
-
- If the model-based forecast is correct then $E_{\bf R} [ \rho_{\tilde I (X), Y-\mu(X)} ] = 0$  and therefore $\rho_{\tilde I(X), Y-\mu(X)} = O(N^{-1})$; however, if the forecasts are not correctly specified then the data quality index is unlikely to inversely scale with population size.
+ % Comparing~\eqref{eq:statdecomp3} to~\eqref{eq:statdecomp2} shows that the addition of the model-based forecast impacts the estimation error in two ways.
+ If the model is adequate, then one may expect a reduction in the problem difficulty and (potentially) in the data quality components.
+ % Note that the model-based forecast does not impact the data quantity component.
+% Second, if the model-based forecast is correct then $E_{\bf R} [ \rho_{\tilde I (X), Y-\mu(X)} ] = 0$  and therefore $\rho_{\tilde I(X), Y-\mu(X)} = O(N^{-1})$; however, if the forecasts are not correctly specified then the data quality index is unlikely to inversely scale with population size.
  Interestingly, the impact of measurement-error on data quality now depends on a relative comparison of the data quality and problem difficulty of the weighted estimator and the doubly-robust estimator.
-For decent models, one would expect this ratio to be greater than one.
+ % For decent models, one would expect this ratio to be greater than one.
 Therefore, $\tilde \Delta < 0$ implies an expected increase in impact of measurement-error and $\tilde \Delta > 0$ implies the magnitude could shrink toward zero or change sign and have an even large magnitude in the opposite direction.
@@
  where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
-Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$ and $\text{var} (\hat y_{n,t}) = V_{t}^{(DR)} + o (n^{-1})$ where $V_t^{(DR)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$ with $E[\cdot]$ and $\text{Var} (\cdot)$ are under the joint randomization of propensity score and sampling designs.
+Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (\bar n^{-1/2})$ and $\text{var} (\hat y_{n,t}) = V_{t}^{(DR)} + o (\bar n^{-1})$ where $\bar n = n \sum_{l=1}^L K_{t,t_l}$ and $V_t^{(DR)}$ is the first diagonal element of $E [\Phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\Phi_n(\eta_0)]^{-1}$ with $E[\cdot]$ and $\text{Var} (\cdot)$ are under the joint randomization of propensity score and sampling designs.
  \end{lemma}
@@
  % Namely, the samples collected included covariate information on (1) prior symptoms, (2) COVID-19 positive individual in household, and (3) prior COVID-19 positive test.
- To account for the nonrandom sample being from high risk areas, we adjust estimates to match the statewide positivity rate of 11.7\% between May 1--2.
+ To account for the nonrandom sample being from high risk areas, we adjust estimates to match the statewide positivity rate of 11.7\% on August 30th.
 These adjusted rates are presented in parentheses in Table~\ref{tab:indiana} along with the relevant covariate information.
@@
 Symptom status (e.g., fever, coughing, shortness of breath) and related covariate information (COVID-19 positive contact), however, are currently unavailable\footnote{Whether this is due to data privacy considerations or lack of data collection efforts is unclear.}.
-As COVID-19 symptom status is likely to alter the testing propensity and likelihood of active infection, for illustrative purposes Facebook symptom survey and hospitalization data are used in two ways to imputate symptom status per strata.
+As COVID-19 symptom status is likely to alter the testing propensity and likelihood of active infection, for illustrative purposes Facebook symptom survey and hospitalization data are used in two ways to impute symptom status per strata.
 We outline the imputation methods below; see Section~\ref{section:discussion} for additional discussion.
@@
 We see that hospitalization significantly increases the likelihood of fever as expected.
-We use these models and the hospitalization rates by demographic to perform mean impuation of the number of positive and negative tests given demographic strata, fever status, and COVID-19 contact status.
+We use these models and the hospitalization rates by demographic to perform mean imputation of the number of positive and negative tests given demographic strata, fever status, and COVID-19 contact status.
 Inverse probability weights are then computed based on the resulting dataset (termed IPW2).
 
- To make the approach concrete, Figure~\ref{fig:testinglik_mainpaper} presents the testing propensity for non-hispanic, white males across age strata and fever status using the second approach.
-We note here that the ratio of testing propensity within strata across fever status is time-varying.
-The ratio starting 50 times more likely in early April and dropping to ten times as likely by end of 2020.
+ Figure~\ref{fig:testinglik_mainpaper} presents the testing propensity for non-Hispanic, white males across age strata and fever status using the second approach.
+Note that the ratio of testing propensity within strata across fever status is time-varying, starting 50 times more likely in early April and dropping to ten times more likely by end of 2020.
 
@@
 Based on published age-specific IFRs~\cite{Levin2020}, IFR closely follows a log-linear relationship with age.
-Combining across published age-specific IFRs, Table~\ref{tab:ifrperage} presents estimates of the average IFR within each strata using the age-distribution.
+\cite{Ironse2103272118} uses Indiana's random survey and death data to estimate a marginal IFR of $0.84\%$.
+Combining across published age-specific IFRs and anchoring our analysis to the estimated marginal IFR of $0.84\%$ for Indiana, Table~\ref{tab:ifrperage} presents the age-specific IFRs used in the analysis.
+Note we collapse ages less than $40$ given the limited number of COVID-19 related deaths in this age range.
  \begin{table}[!th]
  \begin{tabular}{c | c c c c c c c c}
- Age Strata & 0-19 & 20-29 & 30-39 & 40-49 & 50-59 & 60-69 & 70-79 & 80+ \\ \hline
- IFR & 0.003\% & 0.014\% & 0.046\% & 0.144\% & 0.456\% & 1.443\% & 4.567\% & 15.468\%
+ Age Strata & 0-40 & 50-59 & 60-69 & 70-79 & 80+ \\ \hline
+ IFR & $0.014\%$ & $0.120\%$ & $1.206\%$ & $3.815\%$ & $12.920\%$
  \end{tabular}
@@
  \label{tab:ifrperage}
+ \vspace{-1cm}
  \end{table}
@@
  $$
- \beta_t = \sum_{k=1}^{3} \left(\beta_{k-1} (1-w^{(k)}_t ) + \beta_{k} w_t\right),\quad w^{(k)}_t = (1 + \exp \left( -\xi (t - t_k - \nu) \right))^{-1}
- $$
- where $\nu \geq 0$ is the potential delay and $\xi$ is the rate of change.
+ \beta_t = \sum_{k=1}^{3} \left(\beta_{k-1} (1-w^{(k)}_t ) + \beta_{k} w_t\right),\quad w^{(k)}_t = (1 + \exp \left( -\xi (t - t_k - \nu) \right))^{-1} {\bf 1}[t \geq t_k]
+ $$
+ where $\nu \geq 0$ is the potential delay, $\xi$ is the rate of change, and ${\bf 1}[\cdot]$ is an indicator function.
 Each weight~$w_t^{(k)} = 0$ for $t < t_k$ and $w_t^{(k)} \to 1$ as $t \to \infty$.
@@
 Figure~\ref{fig:deathsppc} shows the results which suggests chosen SEIR model fits the observed data well.
+Figure~\ref{fig:casesppc} presents the posterior distribution of total cases.
 The model produces age-specific case counts which can then be used for model-based estimation of the active infection rate as well as in constructing the doubly robust estimator.
@@
 Figure~\ref{fig:tvestimates} presents the results.
-The inverse-probability weighting methods do not significantly reduce bias compared to the unweighted estimates.
-We conjecture the poor performance is due to the limited availability of symptom and other important covariate information to estimate strata-specific infection rates.
-This suggests better data collection in the future may strongly improve the IPW estimates as seen in Section~\ref{section:ipwapproach} when analyzing the statewide sample from April 25th--29th.
+The inverse-probability weighting methods does reduce bias compared to the unweighted estimates.
+Bias, however, does remain which we conjecture is due to the limited availability of symptom and other important covariate information to estimate strata-specific infection rates.
+This suggests better data collection in the future may strongly improve performance of the IPW methodology.
 
@@
 Caution is therefore warranted when interpreting these estimates.
-We know that the age-specific infection fatality rate is likely time-varying and may vary by other factors beyond age.
-Therefore, improvements could be made by more finely grained, and accurate IFR estimates; however, there is currently minimal publicly available data to do so.
+Age-specific infection fatality rates are likely time-varying and may vary by other factors such as quality of health care.
+Therefore, improvements could be made by using more accurate IFR estimates; however, there is currently minimal publicly available data to do so.
 Moreover, the (thankfully) low number of deaths per strata make uncertainty in these estimates quite high.
-Finally, the doubly robust estimates appear similar to the IPW estimate.
+Finally, the doubly robust estimate appears similar to the IPW estimate.
+
+Figure~\ref{fig:air_cis} in the supplementary materials presents the confidence intervals per time point for the IPW2 estimator.
+The confidence interval length decreases substantially over time, reflecting the increased testing capacity.
+Due to the number of surveys per week, there is minimal uncertainty in the parameter estimates.
+As the number of tests per week increases to over one hundred thousand, there is therefore minimal uncertainty in the active infection rate estimates.
+This points to the importance of the statistical decomposition~\eqref{eq:statdecomp2} and the discussion in Section~\ref{section:IPWerrordecomp}.
 
@@
  % \centering
- % \begin{subfigure}{.5\textwidth}
+ %\begin{subfigure}{.5\textwidth}
  \centering
-   \includegraphics[width=.6\linewidth]{../figs/tv_air.png}
+ \includegraphics[width=.7\linewidth]{../figs/tv_air.png}
  % \caption{Active Infection Rate Estimates}
  % \label{fig:air}
- % \end{subfigure}%
- % \begin{subfigure}{.5\textwidth}
+ %\end{subfigure}%
+ %\begin{subfigure}{.5\textwidth}
  % \centering
- %  \includegraphics[width=.9\linewidth]{../figs/tv_air_ratio.png}
- %  \caption{Ratio estimates}
- %  \label{fig:air_ratio}
+ % \includegraphics[width=.9\linewidth]{../figs/tv_air_cis.png}
+ % \caption{IPW2 estimate with confidence intervals}
+ % \label{fig:air_cis}
  % \end{subfigure}
- \caption{Time-varying active infectino rate estimates based on unweighted, model-based, IPW and doubly robust methods.}
+ \caption{Time-varying active infectionrate estimates based on unweighted, model-based, IPW and doubly robust methods.}
  \label{fig:tvestimates}
@@
 This paper emphasizes that data quantity is secondary to both data quality and methodological considerations to account for selection bias.
-
- % \subsection*{High quality information on prevalence.}
+To that end, we emphasize the importance of harmony in data collection and the need for coordinated, systematic data collection.
+Lack of covariate information including symptom and COVID-19 contacts on individuals seeking COVID-19 tests is unacceptable.
+Infrequent random sampling to supplement this data with auxiliary information is desperately needed.
+
+% \subsection*{Higher quality information on prevalence.}
 
@@
 Absent some type of random sampling, the best route forward for all data analyses is careful associated sensitivity analyses and humility in data-driven conclusions.
+\cite{Ironse2103272118}, for example, use the Indiana seroprevalence random survey and to anchor their analysis; however, infrequent random surveys imply time-varying aspects such as the IFR may not be resolved by use of a single anchor.
 

### 2021-08-31 methods/8_death_modeling.R

@@ -61,9 +61,9 @@ df_coviddeath_age_plot %>%
   geom_vline(aes(xintercept = date(date_switch_three)), linetype="dotted") +
   scale_fill_manual(values=wes_palette(n = 3, "IsleofDogs1"))
 
+death_rates = readRDS("../data/mean_ifr.RDS")
 
-death_rates = c(0.01631191, 0.14419778, 0.45623168,
-                1.44348510, 4.56708582, 15.46815604)/100
+death_rates = death_rates/100
 
 data_forcing <- list(n_days = n_days, t0 = t0, ts = t, N = N, deaths = deaths, 
                      tswitch = tswitch+max_death_day, tswitch_two = tswitch_two + max_death_day,
@@ -81,4 +81,4 @@ fit_forcing <- sampling(model_forcing,
                         seed=2,
                         chains = 1)
 
-saveRDS(fit_forcing, "../data/fit_forcing_byage_072621.RDS")
+saveRDS(fit_forcing, "../data/fit_forcing_byage_083121.RDS")

### 2021-08-31 preprocessing/7_ifr.R

@@ -20,11 +20,27 @@ fit <- lm(log(ifr) ~ age)
 lower_age = c(0,40,50,60,70, 80)
 upper_age = c(39,49,59,69,79,90)
 mean_ifr = vector(length = length(lower_age))
+offset = 0.18 # To match the IFR estimate from https://www.pnas.org/content/pnas/118/31/e2103272118.full.pdf
 
 for (i in 1:length(lower_age)) {
-  log_ifr = predict(fit, newdata = data.frame(age = seq(lower_age[i], upper_age[i], length.out = 100)) )
+  log_ifr = predict(fit, newdata = data.frame(age = seq(lower_age[i], upper_age[i], length.out = 100)) ) - offset
   pred_ifr = exp(log_ifr)
   mean_ifr[i] = mean(pred_ifr)
 }
 
 print(mean_ifr)
+
+saveRDS(mean_ifr, "../data/mean_ifr.RDS")
+
+## https://www.infoplease.com/us/census/indiana/demographic-statistics
+## Pop breakdown
+pop = c(7+7.3+7.3+7.5+7+13.7+15.8/2, 
+        15.8/2+13.4/2, 13.4/2 + 4.8, 
+        3.9+6.5/2, 6.5/2 + 4.4/2, 4.4/2+1.5 ) 
+pop = pop/sum(pop)
+
+sum(mean_ifr * pop)  # should match marginal IFR of Raftery.
+
+
+
+

### 2021-08-31 visualizations/fb_symptoms_plot.R

@@ -55,7 +55,7 @@ tempdf$x = as.vector(tempdf$x)
 model.np <- npreg(y~x, regtype = 'll', bwmethod = 'cv.aic', gradients = TRUE, data = subset(tempdf, age == 1))
 
 plot(date(numerage$date[numerage$blockage == 1]), fitted(model.np), type = "l", 
-     bty= 'n', axes = FALSE, ylab = "Reported Fever", xlab = "Date",
+     bty= 'n', axes = FALSE, ylab = "Reported Symptoms", xlab = "Date",
      ylim = c(0.0,0.030), lwd = 3)
 axis.Date(1, at=seq(min(date(numerage$date[numerage$blockage == 1])), max(date(numerage$date[numerage$blockage == 1])), by="months"), format="%b")
 axis(side = 2)

### 2021-08-31 visualizations/model_based.R

@@ -88,7 +88,7 @@ png(filename = "../figs/cases_ppc.png",
     width = 6.5, height = 4, units = "in", res = 1200, pointsize = 12)
 par(mar = c(2,2,1,1)+0.1)
 
-smr_pred <- cbind(agg_cases)
+smr_pred <- agg_cases
 smr_pred$date = date(dates)
 smr_pred = smr_pred[smr_pred$date < "2021-02-01",]
 c_posterior = my_palette # "orange"
@@ -100,6 +100,58 @@ ggplot(smr_pred, mapping = aes(x = date)) +
   #geom_ribbon(aes(ymin = X10., ymax = X90.), fill = c_light) +
   geom_line(mapping = aes(x = date, y = X50.), color = c_posterior) +
   labs(x = "Day", y = "Cases") +
+  scale_x_date(date_labels = "%b %Y")
+
+dev.off()
+
+## Aggregate Undercount calculation
+library(MMWRweek)
+df_case_counts = read.csv("../data/covid_indiana_age.csv", header = T)
+names(df_case_counts) = c("date", "Age", "covid_test", "covid_count", "covid_deaths")
+df_case_counts$date = ymd(df_case_counts$date)
+df_case_counts_total = aggregate(covid_count ~ date, df_case_counts, sum)
+
+
+weeks_for_undercount = unique(week(smr_pred$date[smr_pred$date >= "2020-04-01"]))
+years_for_undercount = c(rep(2020, 40), rep(2021, 5))
+number_of_dates = length(weeks_for_undercount)
+undercount = data.frame(week = weeks_for_undercount, year = years_for_undercount, 
+                        undercountfactor = rep(0, number_of_dates),
+                        undercountfactor.X5 = rep(0, number_of_dates),
+                        undercountfactor.X95 = rep(0, number_of_dates))
+
+for(i in 1:nrow(undercount)) {
+  observed_count = sum(df_case_counts_total$covid_count[week(df_case_counts_total$date) == weeks_for_undercount[i] &
+                                                      year(df_case_counts_total$date) == years_for_undercount[i]
+                                                    ], na.rm = TRUE)
+  mean_latent_count = sum(smr_pred$mean[week(smr_pred$date) == weeks_for_undercount[i] &
+                                      year(smr_pred$date) == years_for_undercount[i]], na.rm = TRUE)
[…]

### 2021-08-31 visualizations/tv_prevalence.R

@@ -107,3 +107,25 @@ dev.off()
 #   scale_color_manual(values=my_palette)
 # 
 # dev.off()
+
+my_palette <- brewer.pal(name="Greys",n=9)[7]
+
+prevalence_alt_cis = readRDS("../data/ipw_cis.RDS")
+prevalence_alt_cis = data.frame(prevalence_alt_cis)
+names(prevalence_alt_cis)  = c("week", "year", "estimate", "stderr", "lowci", "upperci")
+prevalence_alt_cis$date = MMWRweek::MMWRweek2Date(MMWRyear = prevalence_alt_cis$year,
+                                          MMWRweek = prevalence_alt_cis$week,
+                                          MMWRday = 1)
+
+pd <- position_dodge(0.1) # move them .05 to the left and right
+
+png(filename = "../figs/tv_air_cis.png",
+    width = 960, height = 480, units = "px", pointsize = 25)
+
+ggplot(data = prevalence_alt_cis, aes(x = date, y = estimate)) +
+  geom_point(size = 2, color = my_palette) +
+  geom_errorbar(aes(ymin=lowci, ymax=upperci), width=.2, position = pd) +  labs(x = "Date",
+       y = "Active Infection Rate Estimate") + 
+  theme(text = element_text(size=25))
+
+dev.off()

### 2021-09-01 write-up/covid-supplement-writeup.tex
Updated SEIR one last time

@@
 Suppose there is a sequence of finite populations of size $N_{\nu}$ indexed by $\nu$.
-Each finite population has an associated non-probability sample at time $t$ of  size~$\tilde n$ and $L_\nu$ probability samples of fixed size $n$ drawn at equally spaced times $\{ t^\prime_l \}_{l=1}^L$ over the study window~$[0,T]$.
+Each finite population has $L_\nu$ non-probability samples and  probability samples of fixed size $n$ drawn at equally spaced times $\{ t^\prime_l \}_{l=1}^L$ over the study window~$[0,T]$.
 Let~$\Delta_\nu$ denote the gap times between consecutive sample times such that as $\nu \to \infty$ we have that $\Delta_\nu \to 0$.
@@
 \frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t}{\pi_{j,t}} \\
-\frac{1}{N \sum_{l=1}^L K_{t,t_l^\prime}} \sum_{l=1}^L K_{t,t_l^\prime} \left[ \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime_l} - \sum_{j = 1}^N \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime}  \pi_{j,t_l^\prime} X_{j,t_l^\prime} \right] \\
-\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
-\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
+\frac{1}{N \times \sum_{l=1}^L K_{h} \left(|t - t_l^\prime| \right)} \sum_{l=1}^L K_{h} \left(|t - t_l^\prime| \right) \left[ \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime_l} - \sum_{j = 1}^N \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime}  \pi_{j,t_l^\prime} X_{j,t_l^\prime} \right] \\
+\frac{1}{n_{FP}} \sum_{i=1}^{n_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
+\frac{1}{n_{FN}} \sum_{i=1}^{n_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
 \end{array}
@@
 $$
-\frac{1}{N} \left[ \sum_{j=1}^N I_{j,t} X_{j,t} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t} \tilde W_{j,t}  \pi_{j,t} X_{j,t} \right].
+\frac{1}{N} \left[ \sum_{j=1}^N I_{j,t} X_{j,t} - \sum_{j = 1}^N \tilde I_{j,t} \tilde W_{j,t}  \pi_{j,t} X_{j,t} \right].
 $$
@@
 \begin{align*}
-&\frac{1}{N}  E \left[ \sum_{j = 1}^N \tilde I_{j,t} \tilde W_{j,t}  \pi_{j,t} X_{j,t} - \sum_{l=1}^L  \sum_{j = 1}^N \tilde K_{t,t^\prime_l} \tilde I_{j,t^\prime_l} \tilde W_{j,t^\prime_l}  \pi_{j,t^\prime_l} X_{j,t^\prime_l} \right] \\
-= &\frac{1}{N} \sum_{j = 1}^N \left[ \pi_{j,t} X_{j,t} - \sum_{l=1}^L \tilde K_{t,t^\prime_l} \pi_{j,t^\prime_l} X_{j,t^\prime_l} \right], \\
-= &\frac{1}{N} \sum_{j = 1}^N \left[ \sum_{l=1}^L \tilde K_{t,t^\prime_l} \pi_{j,t} X_{j,t} \pm \sum_{l=1}^L \tilde K_{t,t^\prime_l} \pi_{j,t_l^\prime} X_{j,t}  - \sum_{l=1}^L \tilde K_{t,t^\prime_l} \pi_{j,t^\prime_l} X_{j,t^\prime_l} \right], \\
-= &\frac{1}{N} \sum_{j = 1}^N \left[ \left( \sum_{l=1}^L \tilde K_{t,t^\prime_l}  \left( \pi (X_{j,t}; \theta) - \pi (X_{j,t^\prime_l}; \theta) \right) \right) X_{j,t} + \sum_{l=1}^L \tilde K_{t,t^\prime_l} \pi_{j,t^\prime_l} (X_{j,t} - X_{j,t^\prime_l} ) \right],
+&\frac{1}{N}  E \left[ \sum_{j = 1}^N \tilde I_{j,t} \tilde W_{j,t}  \pi_{j,t} X_{j,t} - \sum_{l=1}^L  \sum_{j = 1}^N K_{h,l} \tilde I_{j,t^\prime_l} \tilde W_{j,t^\prime_l}  \pi_{j,t^\prime_l} X_{j,t^\prime_l} \right] \\
+= &\frac{1}{N} \sum_{j = 1}^N \left[ \pi_{j,t} X_{j,t} - \sum_{l=1}^L K_{h,l} \pi_{j,t^\prime_l} X_{j,t^\prime_l} \right], \\
+= &\frac{1}{N} \sum_{j = 1}^N \left[ \sum_{l=1}^L K_{h,l} \pi_{j,t} X_{j,t} \pm \sum_{l=1}^L K_{h,l} \pi_{j,t_l^\prime} X_{j,t}  - \sum_{l=1}^L K_{h, l} \pi_{j,t^\prime_l} X_{j,t^\prime_l} \right], \\
+= &\frac{1}{N} \sum_{j = 1}^N \left[ \left( \sum_{l=1}^L K_{h,l}  \left( \pi (X_{j,t}; \theta) - \pi (X_{j,t^\prime_l}; \theta) \right) \right) X_{j,t} + \sum_{l=1}^L K_{h,l} \pi_{j,t^\prime_l} (X_{j,t} - X_{j,t^\prime_l} ) \right],
 \end{align*}
-where $\tilde K_{t, t^\prime_l}$ is the normalized kernel such that $\sum_{l=1}^L \tilde K_{t, t^\prime_l} = 1$. The difference in the 1st term is similarly defined.
+where~$K_{h,l} := K_h(|t - t_l^\prime|) / \sum_{l=1}^L K_{h} \left(|t - t_l^\prime| \right)$ is the normalized kernel for shorthand. The difference in the 1st term is similarly defined.
 
@@
 That is, the limiting distributions are continuous as a function of $t$.
-If $\pi(x; \theta)$ is a continuous function in~$x$ and under suitable conditions on the kernel density~$\tilde K_{t, \cdot}$ with the bandwidth $h_\nu \to \infty$ such that $\sum_{l=1}^L \tilde K_{t,t_l} \to \infty$ as $\nu \to \infty$, we have
+If $\pi(x; \theta)$ is a continuous function in~$x$ and under suitable conditions on the kernel density~$K_{h_\nu} (\cdot)$ with the bandwidth $h_\nu \to \infty$ such that $\sum_{l=1}^L K_{h_{\nu}} \left( | t - t_l^\prime | \right) \to \infty$ as $\nu \to \infty$, we have
 $$
-\left | \sum_{l=1}^L \tilde K_{t, t_l^\prime} \pi (X_{j, t_l^\prime}; \theta) - \pi (X_{j, t}; \theta) \right|  \to 0
+\left | \sum_{l=1}^L K_{h, l} \pi (X_{j, t_l^\prime}; \theta) - \pi (X_{j, t}; \theta) \right|  \to 0
 $$
@@
 
-Then by a first-order Taylor expansion we have $\hat \eta - \eta_0 = \left[ \phi (\eta) \right]^{-1} \Phi(\eta_0) + o_p (\bar n^{-1/2})$ where $\phi(\eta) = \partial \Phi (\eta)/\partial \eta$ and $\bar n = n \sum_{l=1}^L K_{t,t_l}$.
+Then by a first-order Taylor expansion we have $\hat \eta - \eta_0 = \left[ \phi (\eta) \right]^{-1} \Phi(\eta_0) + o_p (\bar n^{-1/2})$ where $\phi(\eta) = \partial \Phi (\eta)/\partial \eta$ and $\bar n = n \sum_{l=1}^L K_{h,l}$.
 To calculate
@@
 0\\
--\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \left( \frac{Z_j}{FP^2} + \frac{1-Z_j}{(1-FP)^2} \right)  \\
+-\frac{1}{n_{FP}} \sum_{j=1}^{n_{FP}} \left( \frac{Z_j}{FP^2} + \frac{1-Z_j}{(1-FP)^2} \right)  \\
 0
@@
 0 \\
--\frac{1}{n_{FN}}  \sum_{j \in S_{FN}} \left( \frac{\tilde Z_j}{FN^2} + \frac{1-\tilde Z_j}{(1-FN)^2} \right)
+-\frac{1}{n_{FN}}  \sum_{j =1}^{n_{FN}} \left( \frac{\tilde Z_j}{FN^2} + \frac{1-\tilde Z_j}{(1-FN)^2} \right)
 \end{array}
@@
 0 \\
-\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
+\frac{1}{n_{FP}} \sum_{j=1}^{n_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
 0
@@
 0 \\
-\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
+\frac{1}{n_{FN}} \sum_{j=1}^{n_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
 \end{array}
@@
 First, we use the pseudo-likelihood to compute the likelihood of contact given age, gender, and test result.
-Figure~\ref{fig:contactlik1} and~\ref{fig:contactlik2} plots the likelihood for 25-34 year olds given a negative test and positive test respectfully.
+Figure~\ref{fig:contactlik1} and~\ref{fig:contactlik2} plots the likelihood for 25-34 years old given a negative test and positive test respectfully.
 We see that the likelihood of COVID-19 is much higher for those with a positive test and the likelihood is time-varying.
@@
 The model only depends on hospitalization status for positive tests.
-Figure~\ref{fig:symptomlik1_model2} and~\ref{fig:symptomlik2_model2} plots the likelihood for 35-44 year olds given a negative test and positive test respectfully.
+Figure~\ref{fig:symptomlik1_model2} and~\ref{fig:symptomlik2_model2} plots the likelihood for 35-44 years old given a negative test and positive test respectfully.
 We see that the likelihood of fever is time-varying.
@@
 First, we computed the probability of fever given COVID-19 positive test and hospitalization as well as COVID-19 positive test and no hospitalization.
-We then computed the lieklihood of symptom given a COVID-19 positive test by weighting these two likelihoods by the fraction of COVID-19 positive hospitalizations to COVID-19 total positive cases per week.
+We then computed the likelihood of symptom given a COVID-19 positive test by weighting these two likelihoods by the fraction of COVID-19 positive hospitalizations to COVID-19 total positive cases per week.
 Mean imputation of fever status for COVID-19 negative tests was performed based on the above model without use of hospitalization data.

### 2021-09-01 write-up/covid-writeup.tex
Updated SEIR one last time

@@
 We then combine these selection propensities with epidemiological forecast models to construct a \emph{doubly robust} estimation method that accounts for both measurement-error and selection bias.
-This method is then applied to estimate Indiana's prevalence using case-count, hospitalization, and death data with cumulative demographic information, a statewide random molecular sample collected from April 25--29th, and Facebook's COVID-19 symptom survey.
+This method is then applied to estimate Indiana's prevalence using case-count, hospitalization, and death data with demographic information, a statewide random molecular sample collected from April 25--29th, and Facebook's COVID-19 symptom survey.
 We end with a series of recommendations based on the proposed methodology.
@@
 
- After demonstrating the limitations of case-count analysis when compared to random sampling, we then assess whether their is potential for combining the \emph{nonprobability samples} with \emph{probability samples} to improve point prevalence estimation.
+ After demonstrating the limitations of case-count analysis when compared to random sampling, we then assess whether there is potential for combining the \emph{nonprobability samples} with \emph{probability samples} to improve point prevalence estimation.
 For any probability sampling design, the Horvitz-Thompson estimator~\citep{HT1952} incorporates design information via inverse-probability weights (IPW).
@@
 \cite{Song2020} extended this approach to incorporate interventions and assess interventions on COVID-19 epidemic in China.
-\cite{Hao2020} builds an extended SEIR model (SAPHIRE) to account for various presymptomatic infectiousness, time-varying ascertainment rates, transmission rates and population movements.
+\cite{Hao2020} extends this work further to account for various presymptomatic infectiousness, time-varying ascertainment rates, transmission rates and population movements.
 
@@
 For nonprobability samples, \cite{Chen2019} derive a doubly robust approach that uses outcome predictions given covariates on the nonprobability and probability samples.
-Here, we combine the state-space SEIR approach of \cite{Song2020} but instead, as in \cite{Johndrow2020}, focus on COVID-19 confirmed death count data.
+Here, we combine the compartmental model of \cite{Song2020} but instead, as in \cite{Johndrow2020}, focus on COVID-19 confirmed death count data.
 We generate epidemiological forecasts for point prevalence of each population strata.
@@
 Moreover, COVID-19 tests are reported on the day they were administered, while case counts are reported based on the date the positive test was reported to and confirmed by the Indiana Department of Health system.
-To try and address both issues, analyses are conducted on the weekly level in this manuscipt.
+To minimize the impact of reporting on analysis, testing and case-count data are aggregated at the weekly level for analysis.
  \end{remark}
@@
 During May 2--3, 2020, an additional non-random sample of 898 individuals was also collected.
+ Table~\ref{tab:indiana} summarizes the data.
  % In this paper, the nonrandom sample is used to assess whether the cumulative summary statistics are appropriate in constructing the inverse-probability weights as well as in constructing appropriate weights using symptom information.
@@
 We assume either individual $j$ is COVID-19 positive and $Y_j=1$ or is COVID-19 negative and $Y_j=0$.
-We also let $I_j \in \{0,1\}$ be an indicator that the individual was tested ($I = 1$) or not ($I=0$).
+We also let $I_j \in \{0,1\}$ be an indicator that the individual was tested ($I_j = 1$) or not ($I_j=0$).
 
@@
  Under simple random sampling (SRS) or any other epsem\footnote{equal probability of selection method} design, the above are unbiased estimators of the population-level quantities of interest.
-Under SRS, the variance of the estimator can be expressed as $\frac{1}{N-1} \times \frac{1-f}{f} \times \sigma_Y^2$ where $f = n/N$ is the sampling fraction and $\sigma_Y^2 = \frac{1}{N} \sum_{i=1}^N (Y_i - \bar Y)^2 = \bar Y (1- \bar Y)$.
+Under SRS, the variance of the estimator can be expressed as $\frac{1}{N-1} \times \frac{1-f}{f} \times \sigma_Y^2$ where $f = n/N$ is the sampling fraction and $\sigma_Y^2 = \frac{1}{N} \sum_{j=1}^N (Y_j - \bar Y)^2 = \bar Y (1- \bar Y)$.
 
@@
 Assuming no measurement error, $D_M = 1$ so the relation between estimation and selection bias is simple, e.g., if COVID-19 positive individuals were more likely to receive test then this implies upward bias in prevalence estimates.
-Under random testing (i.e., $f_0 = f_1$), $D_M = (1-FP-FN)^{-1}$ so measurement error simply magnifies the error.
+Under random testing (i.e., $f_0 = f_1$), $D_M = (1-FP-FN)^{-1}$ so measurement error simply magnifies this error.
 When tests are imperfect and selection bias exists, this simple relationship no longer holds.
@@
  \end{subfigure}
- \caption{Potential bias due to self-selection and measurement error from a deterministic SIR model with $\beta = 1.2$ (black) and $\gamma = 0.15$.
-Here we assume $f = 0.02$, $FP = 0.024$, $FN = 0.13$, and use a range of relative sampling fractions $M = f_1/f_0$.}
+ \caption{Potential bias in ratio and effective reproductive rate estimators under an SIR model with $\beta = 1.2$ (black) and $\gamma = 0.15$.
+Here, $f = 0.02$, $FP = 0.024$, $FN = 0.13$, and a range of relative sampling fractions $M = f_1/f_0$ are considered.}
  \label{fig:rates}
@@
  % While the above issues on $Z$-scores and effective sample size are important, here we highlight a separate issue.
- Here we demonstrate the interaction of biases in estimation of the two trajectories when the peak infection times differ slightly.
-First, we can write the difference in the rates as $R_{t1} - R_{t2} = \frac{1}{\gamma} \log \left( \frac{1 + e_{1t}}{1+e_{2t}} \right)$ where $e_{jt}$ is the error associated given in Section \ref{section:r0-estimation}.
-Recall that the error allowed for over-estimation prior to the peak and then under-estimation post-peak.
-Here, these errors can mingle in interesting ways.
+ %Here we demonstrate the interaction of biases in estimation of the two trajectories when the peak infection times differ slightly.
+ %First, we can write the difference in the rates as $R_{t1} - R_{t2} = \frac{1}{\gamma} \log \left( \frac{1 + e_{1t}}{1+e_{2t}} \right)$ where $e_{jt}$ is the error associated given in Section \ref{section:r0-estimation}.
+ %Recall that the error allowed for over-estimation prior to the peak and then under-estimation post-peak.  Here, these errors can mingle in interesting ways.
  Consider two countries (A and B) in which the peak occurs 2 weeks prior for country A than country B.
-Figure~\ref{fig:comparison} presents such a comparison where each country's disease trajectory follows an SIR model (A=black and B=red).
-Of interest is the difference between the effective reproduction rate for country A and B, $R_{tA} - R_{tB}$.
-Figure~\ref{fig:eff} shows how biases in estimation of each rate interact in complex ways.
+Figure~\ref{fig:comparison} presents such a comparison where each country's disease trajectory follows an SIR model (A=black and B=grey).
+Figure~\ref{fig:eff} shows how biases interact in complex ways.
 At first, the difference is correctly estimated; then the gap is over-estimated as country A sees a rapid rise in cases; then the magnitude of over-estimation increases as country A sees declining case-count while country B sees rapidly increasing case-count; then country A's rate is correctly estimated while country B's rate is under-estimated as it sees declining case-count; finally, the gap disappears.
-We do not claim this will always be the case; however, this demonstrates how observed information can tell a more complex story than the truth (i.e., country A's peak is 2 weeks prior to country B's peak).
+While this may not always be the case, the analysis demonstrates how estimates can tell a more complex story than the truth (i.e., country A's peak is 2 weeks prior to country B's peak).
 
@@
 Suppose the selection indicator $I_j$ is a Bernoulli random variable that depends only on these covariates, i.e., $P(I_j = 1 \mid X_j = x) = \pi (x; \theta)$.
+Here, we focus on logistic regression models, i.e., $\pi(x; \theta) = \text{expit} \left( x^\top \theta \right)$.
 Maximum likelihood estimation follows by maximizing
@@
  As demonstrated below in Lemma~\ref{lemma:ipw}, if the propensity model is correctly specified then the $E_{\bf R} [ \rho_{\tilde I (X), Y} ] = 0$ and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then the data quality index is unlikely to inversely scale with population size.
-Indeed, Horvitz-Thompson estimators are sensitive to errors in the estimated weights.
+ % Indeed, Horvitz-Thompson estimators are sensitive to errors in the estimated weights.
 Low selection propensity imply large weights which can have deleterious impacts by certain observations dominating estimation.
@@
  If sufficiently large random samples are collected at each time $t =1,\ldots,T$, then the pseudo-likelihood can be re-written as in~\eqref{eq:auxinfoprob} and propensities estimated separately per time point.
-Unfortunately for COVID-19, large probabilistic samples are not available at every time within a given region.
-Here we consider a non-parametric kernel-based approach where the selection propensity at time~$t$, denoted $\hat \theta_t$, maximizes the smoothed pseudo-likelihood
+Unfortunately, large probabilistic samples are not available at every time within a given region.
+To address this, here we consider a non-parametric kernel-based approach where the selection propensity at time~$t$, denoted $\hat \theta_t$, maximizes the smoothed pseudo-likelihood
  $$
@@
 
- % Under simple random sampling of equal size~$\tilde n$ from a fixed population size~$N$ at each time and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the
- % $$
- % \hat \pi_{k,t} = \frac{n}{N} \times \frac{\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \hat d_{k,t^\prime}}{\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \tilde d_{k,t^\prime}}
- % $$
- % where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.
-
- %\subsubsection{Uncertain sensitivity and specificity}
- %
- %For prevalence there is some
- %
- %For
- %
- %For seroprevalence, \cite{Bendavid2020} report
- %
- %
- %
- %, which leads to a range for $M$ of $(2.21, 2.38)$.
-Therefore, the impact on bias of $\hat R_t$ appears moderate.
-If $FP = 0.05$ and $FN = 0.005$, then the relative sampling rate increases to $M=3.04$.
-This simple tool helps us understand how sensitivity and specificity impact biases and  therefore how much we should trust conclusions based on these assumptions.
- %
- %
  \subsubsection{Asymptotics}
@@
  Suppose there is a sequence of finite populations of size $N_{\nu}$ indexed by $\nu$.
-Each finite population has an associated non-probability sample at time $t$ of size~$\tilde n$ and $L_\nu$ probability samples of size $n$ drawn at equally spaced times $\{ t^\prime_l \}_{l=1}^L$ over the study window~$[0,T]$.
-Assuming the selection propensity model is correctly specified, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:ipw} shows that the IPW estimator is  consistent as the number of probability samples grows and calculates the estimator's variance.
+For each finite population, we assume there are $L_\nu$ non-probability samples and probability samples of size $n$ drawn at equally spaced times $\{ t^\prime_l \}_{l=1}^{L_{\nu}}$ over the study window~$[0,T]$.
+Assuming the selection propensity model is correctly specified, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:ipw} shows that the IPW estimator is consistent as $\nu$ goes to infinity (i.e., $N_{\nu}$ and $L_{\nu} \to \infty$) and calculates the estimator's variance.
 In Lemma~\ref{lemma:ipw}, sensitivity and specificity are not assumed known.
@@
  \frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t}{\pi_{j,t}} \\
- \frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
- \frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
- \frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
+ \frac{1}{N \times \sum_{l = 1}^L K_h \left(|t - t_l^\prime| \right)} \sum_{l = 1}^L K_h \left(|t - t_l^\prime| \right) \left[ \sum_{j=1}^N I_{j,t_l^\prime} X_{j,t_l^\prime} - \sum_{j = 1}^N \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime}  \pi_{j,t_l^\prime} X_{j,t_l^\prime}  \right] \\
+ \frac{1}{n_{FP}} \sum_{i=1}^{n_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
+ \frac{1}{n_{FN}} \sum_{i=1}^{n_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
  \end{array}
@@
  where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
-Under certain regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (\bar n^{-1/2})$  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (\bar n^{-1})$ where $\bar n = n \sum_{l=1}^L K_{t,t_l}$ and $V_t^{(IPW)}$ is the first diagonal element of $E [\Phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\Phi_n(\eta_0)]^{-1}$ with $E[\cdot]$ and $\text{Var} (\cdot)$ are under the joint randomization of propensity score and sampling designs.
+Under certain regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (\bar n^{-1/2})$  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (\bar n^{-1})$ where $\bar n = n \sum_{l=1}^L K_{h} (|t - t_l^\prime|)$ and $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$ where $\phi_n = \frac{\partial \Phi (\eta)}{\partial \eta}$ with $E[\cdot]$ and $\text{Var} (\cdot)$ are under the joint randomization of propensity score and sampling designs.
  \end{lemma}
@@
  where $\text{IFR}_{k}$ is the infection fatality rate for the $k$th strata and $\theta_{j}$ is a discrete-time distribution for time from infection to death.
-This extends prior analysis of death data~\cite{Johndrow2020} by allowing the infection fatality rate to depend on a stratification variable, which is important as IFR depends heavily on age.
-Specifically, ``the estimated IFR is close to zero for children and younger adults but rises exponentially with age, reaching 0.4\% at age 55, 1.4\% at age 65, 4.6\% at age 75, and 15\% at age 85''~\cite{Levin2020}.
-For conciseness, parameter choices and additional details on inference are discussed in Appendix~\ref{app:modelbased}.
+This extends prior analysis of death data~\cite{Johndrow2020} by allowing the infection fatality rate to depend on a stratification variable, which is important as IFR depends heavily on age~\citep{Levin2020}.
+See Section~\ref{section:application_of_modelbased} for discussion of parameter choices.
+%Additional details on inference are discussed in Appendix~\ref{app:modelbased}.
 
@@
  $$
- \bar y_{n}^{(DR)} = \frac{1}{N} \sum_{j=1}^N \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \mu(x_j) \right).
+ \bar y_{n}^{(DR)} = \frac{1}{N} \sum_{j=1}^N \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \hat \mu(x_j) \right).
  $$
@@
  Unfortunately, the first term of the doubly robust estimator cannot be computed as covariate information is not collected on the entire population.
-Here, we suppose the same asymptotic regime as in Section~\ref{section:asymptotics} and therefore use the probability samples and consider the doubly-robust estimator at time~$t$
+Here, we suppose the same asymptotic regime as in Section~\ref{section:asymptotics} and use the probability samples and consider the following estimator at time~$t$
  \begin{align*}
- &\frac{1}{\sum_{l=1}^L K_{t,t_l^\prime} \sum_{j=1}^N \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime}} \sum_{l=1}^L K_{t,t_l^\prime} \sum_{j=1}^N  \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime} \hat \mu (X_{j,t^\prime}) \\
- + &\quad \frac{1}{\sum_{j=1}^N I_j w (X_{j,t})} \sum_{j=1}^N I_j w(X_{j,t}) \left( \frac{Y_{j,t} - FP}{1 - FP - FN} - \mu_t(X_{j,t}) \right).
+ &\frac{1}{\sum_{l=1}^L K_{h} (|t - t_l^\prime|) \sum_{j=1}^N \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime}} \sum_{l=1}^L K_{h} (|t - t_l^\prime|) \sum_{j=1}^N  \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime} \hat \mu_{t^\prime_l} (X_{j,t^\prime}) \\
+ + &\quad \frac{1}{\sum_{j=1}^N I_{j,t} w (X_{j,t})} \sum_{j=1}^N I_{j,t} w(X_{j,t}) \left( \frac{Y_{j,t} - FP}{1 - FP - FN} - \mu_t(X_{j,t}) \right).
  \end{align*}
@@
  \frac{1}{N} \left[ \sum_{j=1}^N \frac{I_{j,t}}{\pi_{j,t}} \left( \left( \frac{Y_{j,t} - FP}{1-FP-FN} - \hat \mu_{j,t}  \right)  - \mu_t \right) + \frac{1}{\sum_{j=1}^N \tilde I_j \tilde W_j} \sum_{j=1}^N  \tilde I_j \tilde W_j \hat \mu_t (X_{j,t}) \right] \\
- \frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
- \frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
- \frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
+ \frac{1}{N \times \sum_{l=1}^L K_{h} (|t - t_l^\prime|)} \sum_{l=1}^L K_{h} (|t - t_l^\prime|) \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
+ \frac{1}{n_{FP}} \sum_{j=1}^{n_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
+ \frac{1}{n_{FN}} \sum_{j=1}^{n_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
  \end{array}
@@
  where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
-Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (\bar n^{-1/2})$ and $\text{var} (\hat y_{n,t}) = V_{t}^{(DR)} + o (\bar n^{-1})$ where $\bar n = n \sum_{l=1}^L K_{t,t_l}$ and $V_t^{(DR)}$ is the first diagonal element of $E [\Phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\Phi_n(\eta_0)]^{-1}$ with $E[\cdot]$ and $\text{Var} (\cdot)$ are under the joint randomization of propensity score and sampling designs.
+Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (\bar n^{-1/2})$ and $\text{var} (\hat y_{n,t}) = V_{t}^{(DR)} + o (\bar n^{-1})$ where $\bar n = n \sum_{l=1}^L K_h(|t-t_l|)$ and $V_t^{(DR)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$ where $\phi_n = \frac{\partial \Phi (\eta)}{\partial \eta}$ with $E[\cdot]$ and $\text{Var} (\cdot)$ are under the joint randomization of propensity score and sampling designs.
  \end{lemma}
@@
 The first estimates the probability of contact with a COVID-19 positive individual given demographic and COVID-19 test status (positive or negative COVID-19 test).
-Figure~\ref{fig:contactlik1} and~\ref{fig:contactlik2} in Appendix~\ref{app:in_add_details} presents the estimated likelihood of contact with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.
+Figure~\ref{fig:contactlik1} and~\ref{fig:contactlik2} in Section~\ref{app:in_add_details} in the supplementary materials presents the estimated likelihood of contact with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.
 The second estimates the probability of fever given demographic, COVID-19 test status, \emph{and} whether the individual has had contact with a COVID-19 positive individual.
-Figure~\ref{fig:symptomlik1} and~\ref{fig:symptomlik2} in Appendix~\ref{app:in_add_details} presents the estimated likelihood of reporting a fever with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.
+Figure~\ref{fig:symptomlik1} and~\ref{fig:symptomlik2} in Section~\ref{app:in_add_details} in the supplementary materials presents the estimated likelihood of reporting a fever with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.
 We see that likelihood of fever varies greatly based on whether the individual also reported contact with a COVID-19 positive individual.
 Using these two models, mean imputation is used to calculate the number of positive and negative tests given demographic strata, fever status, and COVID-19 contact status.
-Inverse probability weights are then computed based on the resulting dataset (termed IPW1).
-
- In the second approach, we leverage hospitalization information.
-We first estimate the probability of fever given demographic and COVID-19 test status.
-When the COVID-19 test status is positive, the probability of fever is allowed to depend on whether the individual was hospitalized.
-Figure~\ref{fig:symptomlik1_model2} and~\ref{fig:symptomlik2_model2} in Appendix~\ref{app:in_add_details} presents the estimated likelihood of reporting a fever with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.
-We see that hospitalization significantly increases the likelihood of fever as expected.
-We use these models and the hospitalization rates by demographic to perform mean imputation of the number of positive and negative tests given demographic strata, fever status, and COVID-19 contact status.
-Inverse probability weights are then computed based on the resulting dataset (termed IPW2).
-
- Figure~\ref{fig:testinglik_mainpaper} presents the testing propensity for non-Hispanic, white males across age strata and fever status using the second approach.
-Note that the ratio of testing propensity within strata across fever status is time-varying, starting 50 times more likely in early April and dropping to ten times more likely by end of 2020.
+Inverse probability weights are computed based on the resulting dataset (termed IPW1).
 
@@
   \includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig1_mainpaper.png}
-  \caption{Testing Likelihood Given Fever}
+  \caption{Testing Likelihood Given Symptoms}
   \label{fig:testinglik1_mainpaper}
@@
  \includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig2_mainpaper.png}
-  \caption{Testing Propensity Given No Fever}
+  \caption{Testing Propensity Given No Symptoms}
   \label{fig:testinglik2_mainpaper}
  \end{subfigure}
- \caption{Testing Propensity across age strata and fever status}
+ \caption{Testing Propensity across age strata and COVID-19 symptom status}
  \label{fig:testinglik_mainpaper}
  \end{figure}
+
+ In the second approach, we leverage hospitalization information.
+We first estimate the probability of fever given demographic and COVID-19 test status.
+When the COVID-19 test status is positive, the probability of fever is allowed to depend on whether the individual was hospitalized.
+Figure~\ref{fig:symptomlik1_model2} and~\ref{fig:symptomlik2_model2} in Section~\ref{app:in_add_details} in the supplementary materials presents the estimated likelihood of reporting a fever with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.
+We see that hospitalization significantly increases the likelihood of fever as expected.
+We use these models and the case-to-hospitalization rates by demographic strata to perform mean imputation of the number of positive and negative tests given demographic strata, fever status, and COVID-19 contact status.
+Inverse probability weights are then computed based on the resulting dataset (termed IPW2).
+
+ Figure~\ref{fig:testinglik_mainpaper} presents the testing propensity for non-Hispanic, white males across age strata and fever status using the second approach.
+Note that the ratio of testing propensity within strata across fever status is time-varying, starting 50 times more likely in early April and dropping to ten times more likely by end of 2020.
 
@@
 In this paper, similar to~\cite{Johndrow2020}, we specify fixed per-strata infection fatality rates.
-Based on published age-specific IFRs~\cite{Levin2020}, IFR closely follows a log-linear relationship with age.
+Based on published age-specific IFRs~\citep{Levin2020}, IFR closely follows a log-linear relationship with age.
 \cite{Ironse2103272118} uses Indiana's random survey and death data to estimate a marginal IFR of $0.84\%$.
 Combining across published age-specific IFRs and anchoring our analysis to the estimated marginal IFR of $0.84\%$ for Indiana, Table~\ref{tab:ifrperage} presents the age-specific IFRs used in the analysis.
-Note we collapse ages less than $40$ given the limited number of COVID-19 related deaths in this age range.
+Note the model-based approach collapses strata for ages less than $40$ given the limited number of COVID-19 related deaths in this age range.
  \begin{table}[!th]
@@
  The discrete-time distribution~$\{ \theta_{j} \}$ is a discretized version of a truncated normal distribution with mean $25$, standard deviation~$5$, minimum value $0$, and maximum value $44$.
-This was chosen to closely mimic the distribution from~\cite{Johndrow2020}.
+This was chosen to closely mirror the distribution from~\cite{Johndrow2020}.
 
@@
 Here, we insert three change points $(t_1,t_2,t_3)$ on March 23rd, June 15th, and October 1st of 2020 respectively.
-Given sensitivity to these choices, instead of considering a multiplicative structure, we allow for delayed implementation and slow change in the transmission rate by using weights:
+Given sensitivity to these choices, instead of considering a simple multiplicative structure, we allow for delayed implementation and slow change in the transmission rate by using weights:
  $$
@@
  $$
- where $\nu \geq 0$ is the potential delay, $\xi$ is the rate of change, and ${\bf 1}[\cdot]$ is an indicator function.
+ where~$(\beta_0,\beta_1, \beta_2, \beta_3)$ is the vector of transmission rate parameters, $\nu \geq 0$ is the potential delay, $\xi$ is the rate of change, and ${\bf 1}[\cdot]$ is an indicator function.
 Each weight~$w_t^{(k)} = 0$ for $t < t_k$ and $w_t^{(k)} \to 1$ as $t \to \infty$.
@@
 Figure~\ref{fig:deathsppc} shows the results which suggests chosen SEIR model fits the observed data well.
-Figure~\ref{fig:casesppc} presents the posterior distribution of total cases.
-The model produces age-specific case counts which can then be used for model-based estimation of the active infection rate as well as in constructing the doubly robust estimator.
+Figure~\ref{fig:casesppc} presents the posterior distribution of total cases which can be used to estimate the active infection rate.
+Figure~\ref{fig:undercounting} in the supplementary materials computes the weekly undercount factor of observed case cases.
+Posterior distributions for the age-specific case counts will be used as part of the doubly-robust estimator of the active infection rate.
 
@@
 Caution is therefore warranted when interpreting these estimates.
-Age-specific infection fatality rates are likely time-varying and may vary by other factors such as quality of health care.
-Therefore, improvements could be made by using more accurate IFR estimates; however, there is currently minimal publicly available data to do so.
+Infection fatality rates are likely time-varying and may vary by other factors such as quality of healthcare.
+Therefore, improvements could be made by using more accurate IFR estimates, but there is currently minimal publicly available data to do so.
 Moreover, the (thankfully) low number of deaths per strata make uncertainty in these estimates quite high.
@@
 This paper emphasizes that data quantity is secondary to both data quality and methodological considerations to account for selection bias.
-To that end, we emphasize the importance of harmony in data collection and the need for coordinated, systematic data collection.
-Lack of covariate information including symptom and COVID-19 contacts on individuals seeking COVID-19 tests is unacceptable.
-Infrequent random sampling to supplement this data with auxiliary information is desperately needed.
 
@@
 Most states do not release such granular data, making selection bias adjustments difficult.
-More systematic data collection and reporting is critical in future efforts.
-We acknowledge that this proposal will lead to data privacy issues that will need to be addressed.
-
- While we are eternally grateful to Indiana for the release, symptom information was not made available.
+Coordinated, systematic data collection and reporting is critical.
+ Lack of covariate information including symptom and COVID-19 contacts on individuals seeking COVID-19 tests is unacceptable.
+Random samples over time to supplement this data with auxiliary information is desperately needed.
+We acknowledge that this proposal will lead to data privacy concenrs that will need to be addressed.
+
+ % While we are eternally grateful to Indiana for the release, symptom information was not made available.
 This severely limits our ability to estimate correct selection propensity models and strata-specific infection rates.
@@
 
- A valid criticism of the proposed approach is that recording covariates on every tested individual is too time-consuming and costly, i.e., the data collection does not scale well.
+ A valid criticism of the proposed approach is that recording covariates on every tested individual is too time-consuming and costly, i.e., the proposed data collection protocol does not scale well.
 We argue that the list of relevant factors is not too long.
@@
 Estimation proceeds by maximizing the pseudo-likelihood which now depends on random sampling in both the first and second term.
- This would allow for rapid testing and minimizes cost of covariate collection.
+ This would allow for rapid testing and minimize data collection costs.
 

### 2021-09-01 visualizations/model_based.R

@@ -29,7 +29,7 @@ dates = df_coviddeath$date[-length(df_coviddeath$death_dt)]
 # times
 n_days <- length(deaths)
 
-fit_forcing = readRDS("../data/fit_forcing_byage_072621.RDS")
+fit_forcing = readRDS("../data/fit_forcing_byage_083121.RDS")
 
 test = summary(fit_forcing, pars = "pred_deaths", probs = c(0.025, 0.05, 0.1, 0.5, 0.9, 0.95, 0.975))$summary
 n_days = nrow(test)/6

### 2021-09-01 write-up/covid-writeup.tex
Update MB AIR

@@
 
- While the model-based estimates appear more reasonable, note that the estimate on the last week of April is 0.25\% which is a severe underestimate compared to the estimated prevalence using the stratified random sample.
+ While the model-based estimates appear more reasonable outside the summer months, note that the estimate on the last week of April is $0.30\%$ which is a severe underestimate compared to the estimated prevalence using the stratified random sample of $1.81\%$.
 Caution is therefore warranted when interpreting these estimates.

### 2021-09-01 preprocessing/8_modelbased_preproc.R

@@ -54,7 +54,7 @@ df_coviddeath$date = ymd(df_coviddeath$startdate)
 df_coviddeath_agg <- aggregate(covid_deaths ~ date, data = df_coviddeath, FUN = sum)
 dates = df_coviddeath_agg$date[-length(df_coviddeath_agg$date)]
 
-fit_forcing = readRDS("../data/fit_forcing_byage_072621.RDS")
+fit_forcing = readRDS("../data/fit_forcing_byage_083121.RDS")
 summary_fit = summary(fit_forcing, pars = "pred_cases_per_agegroup", probs = c(0.025, 0.05, 0.1, 0.5, 0.9, 0.95, 0.975))$summary
 n_days = nrow(summary_fit)/6
 dates = c(min(dates)-1:(n_days - length(dates)), dates) ## EXTEND BACKWARDS FOR CASES
@@ -265,3 +265,6 @@ saveRDS(prevalence_temp, "../data/aggregate_air.RDS")
 
 
 plot(prevalence_temp$date, prevalence_temp$estimate)
+
+
+cumsum(prevalence_temp$mean)[10]/N

### 2021-09-01 write-up/covid-writeup.tex
Now just wait for model to run and then done!

@@
 
- Suppose there is a sequence of finite populations of size $N_{\nu}$ indexed by $\nu$.
-For each finite population, we assume there are $L_\nu$ non-probability samples and probability samples of size $n$ drawn at equally spaced times $\{ t^\prime_l \}_{l=1}^{L_{\nu}}$ over the study window~$[0,T]$.
-Assuming the selection propensity model is correctly specified, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:ipw} shows that the IPW estimator is consistent as $\nu$ goes to infinity (i.e., $N_{\nu}$ and $L_{\nu} \to \infty$) and calculates the estimator's variance.
-In Lemma~\ref{lemma:ipw}, sensitivity and specificity are not assumed known.
-Instead, we suppose that these are estimated using a pseudo-likelihood on two random samples from the population whose size (denoted~$n_{FP}$ and $n_{FN}$ respectively) is chosen so the resulting estimated sensitivity and specificity have desired levels of uncertainty.
+ Here we suppose there is a sequence of finite populations of size $N_{\nu}$ indexed by $\nu$ and that there are $L_\nu$ non-probability samples and probability samples of size $n$ drawn at equally spaced times $\{ t^\prime_l \}_{l=1}^{L_{\nu}}$ over the study window~$[0,T]$.
+Assuming the selection model is correctly specified, then under the design of the probability sample and propensities of the nonprobability sample, Lemma~\ref{lemma:ipw} shows that the IPW estimator is consistent as $\nu$ goes to infinity (i.e., $N_{\nu}, L_{\nu} \to \infty$) and calculates the estimator's variance.
+In Lemma~\ref{lemma:ipw}, sensitivity/specificity are not assumed known but estimated using a pseudo-likelihood on two random samples whose size (denoted~$n_{FP}$ and $n_{FN}$) is chosen so the estimated sensitivity/specificity have desired levels of uncertainty.
+On each sample, a binary outcome denoted $Z_j$ and $\tilde Z_j$ is observed representing a false positive/negative.
 See Appendix~\ref{app:asympderivations} for additional details.
@@
  &\frac{1}{\sum_{l=1}^L K_{h} (|t - t_l^\prime|) \sum_{j=1}^N \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime}} \sum_{l=1}^L K_{h} (|t - t_l^\prime|) \sum_{j=1}^N  \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime} \hat \mu_{t^\prime_l} (X_{j,t^\prime}) \\
- + &\quad \frac{1}{\sum_{j=1}^N I_{j,t} w (X_{j,t})} \sum_{j=1}^N I_{j,t} w(X_{j,t}) \left( \frac{Y_{j,t} - FP}{1 - FP - FN} - \mu_t(X_{j,t}) \right).
+ + &\quad \frac{1}{\sum_{j=1}^N I_{j,t} w (X_{j,t})} \sum_{j=1}^N I_{j,t} w(X_{j,t}) \left( \frac{Y_{j,t} - FP}{1 - FP - FN} - \hat \mu_t(X_{j,t}) \right).
  \end{align*}
@@
  \begin{array}{c}
- \frac{1}{N} \left[ \sum_{j=1}^N \frac{I_{j,t}}{\pi_{j,t}} \left( \left( \frac{Y_{j,t} - FP}{1-FP-FN} - \hat \mu_{j,t}  \right)  - \mu_t \right) + \frac{1}{\sum_{j=1}^N \tilde I_j \tilde W_j} \sum_{j=1}^N  \tilde I_j \tilde W_j \hat \mu_t (X_{j,t}) \right] \\
+ \frac{1}{N} \left[ \sum_{j=1}^N \frac{I_{j,t}}{\pi_{j,t}} \left( \left( \frac{Y_{j,t} - FP}{1-FP-FN} - \hat \mu_{j,t}  \right)  - \mu_t \right) + \frac{\sum_{l=1}^L K_{h} (|t - t_l^\prime|) \sum_{j=1}^N \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime} \hat \mu_{j,t_l^\prime}}{\sum_{l=1}^L K_{h} (|t - t_l^\prime|) \sum_{j=1}^N \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime}} \right] \\
  \frac{1}{N \times \sum_{l=1}^L K_{h} (|t - t_l^\prime|)} \sum_{l=1}^L K_{h} (|t - t_l^\prime|) \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
@@
 Figure~\ref{fig:deathsppc} shows the results which suggests chosen SEIR model fits the observed data well.
-Figure~\ref{fig:casesppc} presents the posterior distribution of total cases which can be used to estimate the active infection rate.
-Figure~\ref{fig:undercounting} in the supplementary materials computes the weekly undercount factor of observed case cases.
-Posterior distributions for the age-specific case counts will be used as part of the doubly-robust estimator of the active infection rate.
+Figure~\ref{fig:casesppc} presents the posterior distribution of total new cases.
+Figure~\ref{fig:undercounting} in the supplementary materials computes the weekly undercount factor of observed case counts.
+Posterior distributions for the age-specific number of infected individuals will be used as part of the doubly-robust estimator of the active infection rate.
 

### 2021-09-01 methods/8_death_modeling.R

@@ -39,27 +39,13 @@ t <- seq(1, n_days+max_death_day, by = 1)
 t0 = 0
 t <- t
 
+## Switch Dates
 date_switch <- "2020-03-23" # date of introduction of control measures (empirical)
 tswitch <- subset(df_coviddeath_age, age == "80+") %>% filter(date < date_switch) %>% nrow() + 1 # convert time to number
 date_switch_two <- "2020-06-15" # date of ending of control measures
 tswitch_two <- subset(df_coviddeath_age, age == "80+") %>% filter(date < date_switch_two) %>% nrow() + 1 # convert time to number
 date_switch_three <- "2020-10-01" # date of ending of control measures
 tswitch_three <- subset(df_coviddeath_age, age == "80+") %>% filter(date < date_switch_three) %>% nrow() + 1 # convert time to number
-
-# ## PLOTTING DEATH DATA BY AGE CATEGORY
-# df_coviddeath_age_plot = df_coviddeath_age
-# levels(df_coviddeath_age_plot$age) = c(1,2,2,2,3,3)
-# levels(df_coviddeath_age_plot$age) = c("0-39", "40-69", "70+")
-# 
-# df_coviddeath_age_plot %>% 
-#   ggplot() + 
-#   geom_bar(mapping = aes(x = date, y = death_dt, fill = age), stat = "identity") +
-#   labs(y="Number of COVID-19 Reported Deaths") + 
-#   scale_x_date(date_breaks = "months" , date_labels = "%b-%y")+ 
-#   geom_vline(aes(xintercept = date(date_switch)), linetype="dotted") + 
-#   geom_vline(aes(xintercept = date(date_switch_two)), linetype="dotted") + 
-#   geom_vline(aes(xintercept = date(date_switch_three)), linetype="dotted") +
-#   scale_fill_manual(values=wes_palette(n = 3, "IsleofDogs1"))
 
 death_rates = readRDS("../data/mean_ifr.RDS")
 
@@ -74,6 +60,7 @@ data_forcing <- list(n_days = n_days, t0 = t0, ts = t, N = N, deaths = deaths,
                      alpha = rep(1, length(death_rates)))
 model_forcing <- stan_model("./8_sir_model.stan")
 
+set.seed(17194871)
 fit_forcing <- sampling(model_forcing, 
                         data_forcing, 
                         iter=2000,

### 2021-09-02 write-up/covid-writeup.tex
Updated paper to shorten a bit

@@
 Due to monetary and time constraints, however, random testing is performed infrequently.
-As of June 2021, Indiana and Ohio are the only states to conduct statewide random sample testing\footnote{These are is the only random samples to collect both seroprevalence and diagnostic testing results.
+As of June 2021, Indiana and Ohio are the only states to conduct statewide random sample testing\footnote{These are the only random samples to collect both seroprevalence and diagnostic testing results.
 The CDC and other states have conducted seroprevalence-only studies.}.
@@
 Unfortunately, many states do not require or report auxiliary covariate information beyond basic demographic information (e.g., gender, age, race, and ethnicity).
-We end with a brief list of suggestions of minor changes to current practice based on the proposed methodology.
+We end with a brief list of suggestions of changes to current practice based on the proposed methodology.
 While we demonstrate empirical improvements over simple disease prevalence estimates, we also highlight how selection bias may persist and impact uncertainty quantification.
@@
 
- A molecular test conducted on an infected individual may return a negative result.
+ A molecular test conducted on an actively infected individual may return a negative result.
 Such false negatives are very strongly associated with when the test is conducted.
@@
 Moreover, COVID-19 tests are reported on the day they were administered, while case counts are reported based on the date the positive test was reported to and confirmed by the Indiana Department of Health system.
-To minimize the impact of reporting on analysis, testing and case-count data are aggregated at the weekly level for analysis.
+To minimize the impact of reporting, testing and case-count data are aggregated at the weekly level for analysis.
  \end{remark}
@@
 The survey defines an individual as displaying \emph{COVID-like symptoms} if they exhibit a fever along with a cough, or shortness of breath, or difficulty breathing.
-Figure~\ref{fig:fbsymptoms} displays smoothed estimates of the fraction of individuals within age and gender strata who report a fever within the past 24-hours.
-This paper uses the Facebook symptom survey as the main source of auxiliary information on time-varying characteristics of the population of Indiana, e.g., fraction of population displaying symptoms or in contact with COVID-19 positive individuals.
+Figure~\ref{fig:fbsymptoms} displays smoothed estimates of the fraction of individuals who report COVID-like symptoms within the past 24-hours by age and gender.
+The Facebook symptom survey is used as the main source of auxiliary information on time-varying characteristics of the population of Indiana, e.g., displaying COVID-like symptoms or in contact with COVID-19 positive individuals.
 
@@
  Tests are imperfect.
-As discussed in Section~\ref{section:testinginfo}, COVID-19 testing is no exception.
+COVID-19 testing is no exception.
 Here we investigate the interplay between imperfect testing and selection bias.
-Researchers most often assume measurement error will lead to parameter attenuation.
-When paired with selection bias, however, the two sources of error become entangled, and resulting errors can become magnified, muted, or even switch signs.
+Researchers often assume measurement error leads to parameter attenuation.
+When paired with selection bias, however, the two sources become entangled, and resulting errors in magnified, muted, or even switch signs.
 
@@
 The first three terms continue to represent \emph{data quality}, \emph{data quantity}, and \emph{problem difficulty} respectively.
-The new term~$D_M$ represents the \emph{imperfect testing adjustment} which is a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence, which is scaled by the factor $(1 - FP - FN)^{-1}$.
+The new term~$D_M$ represents the \emph{imperfect testing adjustment} which is a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence.
+%, which is scaled by the factor $(1 - FP - FN)^{-1}$.
 
@@
  implying the error relative to SRS increases as a function of population size~\citep{Meng2018}.
-The paradox that two countries with the same testing strategy can yield wildly different estimates due to population size is too often ignored in current communications of case-count data.
-Thus, conclusions drawn from observed case-counts may not just be wrong, but very wrong.
+That two countries with the same testing strategy can yield wildly different estimates due to population size is too often ignored in current communications of case-count data.
+ % Thus, conclusions drawn from observed case-counts may not just be wrong, but very wrong.
 
@@
  \end{equation}
- where $s_t$ is the fraction of susceptible individuals, $i_t$ is the fraction of infected individuals, and $r_t$ is the fraction of removed (recovered or deceased) individuals in the population at time $t$.
+ where $s_t, i_t$ and $r_t$ are the fraction of susceptible, infected, and removed (recovered or deceased) individuals in the population at time $t$ respectively.
 In terms of bias, the rate is overestimated prior to the peak and underestimated afterwards; the bias increases dramatically when the relative fraction exceeds $2$.
 Such biases may have a substantial impact on policy making.
-First, overestimation may give policy makers more leverage in proposing aggressive actions to reduce prevalence.
-Of course, the analysis provides evidence that estimates based on available data may be pessimistic.
-What is missing in the current discourse is that the direction of bias is non-constant over time.
+Overestimation pre-peak may give policy makers more leverage in proposing aggressive actions to reduce prevalence.
 Underestimation post-peak puts pressure on policy makers to prematurely relax social distancing measures.
-Moreover, it appears that the peak time is the easiest to estimate in terms of having minimal error using available data.
-Note, however, that the standard errors will be large due to small effective sample sizes.
+Estimates at the peak time appear to have minimal bias.
+ %Note, however, that the standard errors will be large due to small effective sample sizes.
 
@@
  The prior section presented negative consequences of selection-bias and measurement error when estimating infection prevalence, rates of change, and the effective reproduction number from observed case-count data.
-This raises the immediate question whether we can can improve upon these estimators.
-In this section, we consider two directions.
+In this section, we consider two directions to improve upon these estimators.
  The proposed methods and statistical error decompositions guide our recommendations in Section~\ref{section:discussion}.
@@
 
- With non-probability samples, one can reduce bias by modelling the self-selection propensity and use inverse probability weighting (IPW)~\citep{Elliott2017} techniques to adjust for selection bias.
-In causal inference~\citep{Hernan2020}, under certain assumptions, the observational data permits direct estimation of treatment propensity.
-In the COVID-19 context, one does not observe those who decide not to be tested.
-In order to estimate selection propensities, one needs auxiliary information.
+ With non-probability samples, bias can be reduced by modelling the self-selection propensity and using inverse probability weighting (IPW)~\citep{Elliott2017} to adjust for selection bias.
+ % In causal inference~\citep{Hernan2020}, under certain assumptions, the observational data permits direct estimation of treatment propensity.
+ In the current context, however, one does not observe those who are not tested.  To estimate selection propensities, auxiliary information is needed.
 
@@
  Here we assume access to a probability sample measuring the same set of covariates.
-Let $\tilde I_j$ be an indicator that the individual was included in the probability sample and $\tilde W_j$ be the probability of inclusion for the $j$th individual (i.e., equal to $n/N$ for a simple random sample of size $n$).
-In recent work,~\cite{Chen2019} use a probability sample to construct a design-unbiased estimator of the second term
+Let $\tilde I_j$ be an indicator that the individual was included in the probability sample and $\tilde W_j$ be the probability of inclusion for the $j$th individual.
+\cite{Chen2019} use a probability sample to construct a design-unbiased estimator of the second term
  \begin{equation}

### 2021-09-02 write-up/covid-supplement-writeup.tex
So close - just final edits

@@
 
-In this section, we describe the two mean-imputation methods.
+Table~\ref{tab:comparison} demonstrates minimal bias in the Facebook symptom survey with respect to symptom distributions.
+
+ \begin{table}[!th]
+ \centering
+ \begin{tabular}{c | c c c | c}
+ & \multicolumn{3}{c}{Facebook} & Random \\ \cline{2-4}
+ & Lower CI & Estimate & Upper CI & Estimate \\ \hline
+ Fever &  0.017 & 0.020 & 0.024 & 0.018 \\
+ Cough &  0.142 & 0.148 & 0.154 & 0.149 \\
+ Shortness & 0.058 & 0.060 & 0.062 & 0.062 \\ \hline
+ \end{tabular}
+ \caption{Comparison of Facebook survey and random sample}
+ \label{tab:comparison}
+ \end{table}
+
+Next, we describe the two mean-imputation methods.
 \subsection{Model 1}
@@
 where \code{beta, eta, eta\_two, eta\_three} refer to the four values that form the time-varying parameter $\beta_t$ when combined with \code{xi, eta, eta\_two, eta\_three} in the SEIR model as described in Section~\ref{section:modelbased}.
+Following~\cite{Song2020}, we employ Runge-Kutta (RK4) approximations for discretization.
 

### 2021-09-02 write-up/covid-writeup.tex
So close - just final edits

@@
  \end{equation}
- Taking an expectation with respect to the sampling design yields~\eqref{eq:propensity}.
-Solving~\eqref{eq:auxinfoprob} can be done by iteratively re-weighted least squares (IRLS); see Section~\ref{app:irls} in the supplementary materials for algorithmic details.
+ Expectation of~\ref{eq:auxinfoprob} with respect to the sampling design yields~\eqref{eq:propensity}.
+Solving~\eqref{eq:auxinfoprob} is done by iteratively re-weighted least squares (IRLS); see Section~\ref{app:irls} in the supplementary materials for details.
 
@@
  Here we suppose there is a sequence of finite populations of size $N_{\nu}$ indexed by $\nu$ and that there are $L_\nu$ non-probability samples and probability samples of size $n$ drawn at equally spaced times $\{ t^\prime_l \}_{l=1}^{L_{\nu}}$ over the study window~$[0,T]$.
-Assuming the selection model is correctly specified, then under the design of the probability sample and propensities of the nonprobability sample, Lemma~\ref{lemma:ipw} shows that the IPW estimator is consistent as $\nu$ goes to infinity (i.e., $N_{\nu}, L_{\nu} \to \infty$) and calculates the estimator's variance.
-In Lemma~\ref{lemma:ipw}, sensitivity/specificity are not assumed known but estimated using a pseudo-likelihood on two random samples whose size (denoted~$n_{FP}$ and $n_{FN}$) is chosen so the estimated sensitivity/specificity have desired levels of uncertainty.
-On each sample, a binary outcome denoted $Z_j$ and $\tilde Z_j$ is observed representing a false positive/negative.
+Assuming correct selection model specification, then under the probability sample design and nonprobability sample propensities, Lemma~\ref{lemma:ipw} shows that the IPW estimator at a time~$t \in \{ t^\prime_l \}$ is consistent as $\nu$ goes to infinity (i.e., $N_{\nu}, L_{\nu} \to \infty$) and calculates the estimator's variance.
+In Lemma~\ref{lemma:ipw}, sensitivity/specificity are unknown and estimated using a pseudo-likelihood on two random samples whose sizes (denoted~$n_{FP}$ and $n_{FN}$) reflect uncertainty in these quantities.
 See Appendix~\ref{app:asympderivations} for additional details.
@@
  % \textcolor{red}{Must check.}
- The estimates~$\hat \mu_t := \bar y_{n,t}^\star$, $\hat \pi_{j,t}$, $\hat FP$, and $\hat FN$ are solutions to the following set of estimating equations
+The estimates~$\hat \mu_t := \bar y_{n,t}^\star$, $\hat \pi_{j,t}$, $\hat FP$, and $\hat FN$ at a time~$t \in \{t_l^\prime\}$ are solutions to the following set of estimating equations
  $$
@@
 \end{align}
- Following~\cite{Song2020}, we employ Runge-Kutta (RK4) approximations for discretization.
+%Following~\cite{Song2020}, we employ Runge-Kutta (RK4) approximations for discretization.
 Here, we consider covariate information that takes the form of a stratification variable with $K$ strata.
 Let $I_{k,t}$ to denote the number of new infections on day $t$ in the $k$th strata.
-As there are a total of $N \cdot i_t := I_t$ new infections on day $t$, we assume the joint distribution of strata-specific new infection totals takes a multinomial distribution with
+The joint distribution of strata-specific new infection totals follows a multinomial distribution with
  $$
@@
  $$
- where the estimates $\hat \mu(x_j)$ are not corrected for measurement-error as they are built to estimate true case count from death-only data.
+ where $\hat \mu(x_j)$ is not corrected for measurement-error as it estimates true case counts.
 This estimator is called ``doubly-robust'' because it is consistent if either the model-based forecasts or the time-varying propensities are correctly specified.
@@
  Unfortunately, the first term of the doubly robust estimator cannot be computed as covariate information is not collected on the entire population.
-Here, we suppose the same asymptotic regime as in Section~\ref{section:asymptotics} and use the probability samples and consider the following estimator at time~$t$
- \begin{align*}
- &\frac{1}{\sum_{l=1}^L K_{h} (|t - t_l^\prime|) \sum_{j=1}^N \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime}} \sum_{l=1}^L K_{h} (|t - t_l^\prime|) \sum_{j=1}^N  \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime} \hat \mu_{t^\prime_l} (X_{j,t^\prime}) \\
- + &\quad \frac{1}{\sum_{j=1}^N I_{j,t} w (X_{j,t})} \sum_{j=1}^N I_{j,t} w(X_{j,t}) \left( \frac{Y_{j,t} - FP}{1 - FP - FN} - \hat \mu_t(X_{j,t}) \right).
- \end{align*}
- where $\tilde I_{j,t_l^\prime}$ indicates an individual is included in the random sample at time~$t_l^\prime$ and $\tilde W_{j,t_l^\prime}$ is the associated weight for that unit.
+Here, we suppose the same asymptotic regime as in Section~\ref{section:asymptotics} and use the probability sample at the time~$t \in \{t_l^\prime\}$ to estimate this term:
+ $$
+ \frac{1}{\sum_{j=1}^N \tilde I_{j,t} \tilde W_{j,t}} \sum_{j=1}^N  \tilde I_{j,t} \tilde W_{j,t} \hat \mu_{t} (X_{j,t}) + \frac{1}{\sum_{j=1}^N I_{j,t} w (X_{j,t})} \sum_{j=1}^N I_{j,t} w(X_{j,t}) \left( \frac{Y_{j,t} - FP}{1 - FP - FN} - \hat \mu_t(X_{j,t}) \right).
+ $$
  Assuming the selection propensity model is correctly specified, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:dr} shows that the doubly robust estimator is consistent and calculates the estimator's variance.
@@
 
- \item {\bf Random/Nonrandom statewide sample}: Table~\ref{tab:indiana} summarizes data from a random sample from April 25--29th as well as a nonrandom sample obtained between May 2--3 in racial/ethnic minority communities~\cite{Yiannoutsos2021}.
+ \item {\bf Random/Nonrandom statewide sample}: Table~\ref{tab:indiana} summarizes data from a random sample from April 25--29th as well as a nonrandom sample obtained between May 2--3 in racial/ethnic minority communities~\citep{Yiannoutsos2021}.
 The nonrandom sample is not a random subsample of the overall case-counts; however, this data provides important supplementary covariate information.
@@
 We collect age, sex, and demographic information as well as COVID-19 related symptoms, which Facebook defines as having (1) a fever and cough, (2) shortness of breath, \emph{or} (3) difficulty breathing.
-To assess potential bias, we compare the random sample from Indiana to the Facebook symptom survey data collected from April 25th to April 29th.
-Table~\ref{tab:comparison} demonstrates minimal bias in the Facebook symptom survey with respect to symptom distributions.
-
- \begin{table}[!th]
- \centering
- \begin{tabular}{c | c c c | c}
- & \multicolumn{3}{c}{Facebook} & Random \\ \cline{2-4}
- & Lower CI & Estimate & Upper CI & Estimate \\ \hline
- Fever &  0.017 & 0.020 & 0.024 & 0.018 \\
- Cough &  0.142 & 0.148 & 0.154 & 0.149 \\
- Shortness & 0.058 & 0.060 & 0.062 & 0.062 \\ \hline
- \end{tabular}
- \caption{Comparison of Facebook survey and random sample}
- \label{tab:comparison}
- \end{table}
+Table~\ref{tab:comparison} in Section~\ref{app:in_add_details} of the supplementary materials shows minimal bias with respect to symptom distributions when comparing to Indiana's random sample to the Facebook symptom survey data collected from April 25th to April 29th.
+
+ % \begin{table}[!th]
+ % \centering
+ % \begin{tabular}{c | c c c | c}
+ % & \multicolumn{3}{c}{Facebook} & Random \\ \cline{2-4}
+ % & Lower CI & Estimate & Upper CI & Estimate \\ \hline
+ % Fever &  0.017 & 0.020 & 0.024 & 0.018 \\
+ % Cough &  0.142 & 0.148 & 0.154 & 0.149 \\
+ % Shortness & 0.058 & 0.060 & 0.062 & 0.062 \\ \hline
+ % \end{tabular}
+ % \caption{Comparison of Facebook survey and random sample}
+ % \label{tab:comparison}
+ % \end{table}
 
@@
  \item {\bf Death data}: Daily COVID-19 related deaths are observed by age, gender, and racial demographics.
-Figure~\ref{fig:in-deaths} plots the cumulative number of COVID-19 related deaths over time by age group.
-While most COVID-19 tests and positive cases are within the younger age strata, most deaths are within the 70+ age strata.
+Figure~\ref{fig:in-deaths} plots COVID-19 related deaths over time by age group.
+While most tests and positive cases are within the younger age strata, most deaths are within the 70+ age strata.
  \end{itemize}
@@
  The Indiana study did not report sensitivity and specificity; therefore, we take the suggested measures from~\cite{Arevalo2020} which report 87\% sensitivity is a reasonable estimates and~\cite{Cohen2020} which report 97.6\% specificity.
-This corresponds to a false negative rate of $13$\% and false positive rate of $2.4\%$.
-Therefore, the IPW estimates accounting for measurement-error are $6.2\%$ and $4.9\%$ respectively.
+This corresponds to a false negative rate of $13$\% and false positive rate of $2.4\%$, resulting in estimates of $6.2\%$ and $4.9\%$ respectively.
 
@@
 Here we consider estimation of data quality $E_{\I} \left[ \rho_{\tilde I(X),Y} \right]$.
-In \cite{Meng2018}, estimation relied on observing the true outcome (i.e., election vote totals), which is not possible in the current crisis.
+In \cite{Meng2018}, estimation relied on observing the true outcome (i.e., election vote totals).
 Here, we use the random sample which estimates the true prevalence of active COVID-19 disease at 1.81\%~\cite{Yiannoutsos2021}.
-Our goal is not inferential.
-Instead, we aim to build sensitivity analyses that can aid in understanding the amount of information in observational COVID case-count data.
+Our goal is to build sensitivity analyses to understand the amount of information in observational case-count data.
 
@@
 Assuming a false negative rate of $13$\% and false positive rate of $2.4$\%, the unweighted estimate is $11.0$\%, leading to an error of $9.2$\%.
-Using~\eqref{eq:statdecomp}, we construct an estimate of the relative sampling rate as follows
+Using~\eqref{eq:statdecomp}, an estimate of the relative sampling rate is
  \begin{equation*}
@@
  \end{equation*}
- Using inverse-probability weights, the IPW estimates for the true prevalence of active COVID-19 disease between April 25th and April 29th are $7.7$\% and $6.5\%$ respectively.
-Under the same sensitivity and specificity, the weighted estimates adjusted for measurement-error are $6.2$\% and $4.9$\%, leading to errors $4.4$\% and $3.1$\% respectively.
-Using~\eqref{eq:statdecomp2}, we construct an estimate of the relative sampling rate as follows
+ The IPW estimates for the prevalence of active COVID-19 disease between April 25th and April 29th adjusted for measurement-error are $6.2$\% and $4.9$\%, leading to errors $4.4$\% and $3.1$\% respectively.
+Using~\eqref{eq:statdecomp2}, an estimate of the relative sampling rate is
  \begin{equation*}
@@
  \end{equation*}
- using the Indiana non-random sample.
-For the IPW estimate using Facebook data, we have $\tilde M = 3.3$.
-Therefore, the impact of self selection on bias of $\hat R_t$ appears strong using unweighted data, remaining moderate even using the weighted estimates.
-
- A sensitivity analysis is performed by considering the range of false negative and false positive rates.
-Here, sensitivity is assumed to fall between $81$\% and $91$\% and specificity between $95$\% and $98.8$\%~\cite{Katz2020}, which leads to a range for $M$ of $(5.60, 6.44)$ for the unweighted analysis, and $(3.63,4.07)$ and $(3.10, 3.45)$ for the weighted analyses respectively.
+ using the Indiana non-random sample.  For the IPW estimate using Facebook data, $\tilde M = 3.3$. Therefore, bias is high using unweighted data and remains moderate using the weighted estimates.
+
+ A sensitivity analysis can be performed by considering the range of false negative/positive rates.
+Here, sensitivity ranges between $81$\% and $91$\% and specificity between $95$\% and $98.8$\%~\cite{Katz2020}.
+This leads to a range for $M$ of $(5.60, 6.44)$ for the unweighted analysis, and $(3.63,4.07)$ and $(3.10, 3.45)$ for the weighted analyses.
 
@@
 Combining across published age-specific IFRs and anchoring our analysis to the estimated marginal IFR of $0.84\%$ for Indiana, Table~\ref{tab:ifrperage} presents the age-specific IFRs used in the analysis.
-Note the model-based approach collapses strata for ages less than $40$ given the limited number of COVID-19 related deaths in this age range.
 \begin{table}[!th]
@@
  The discrete-time distribution~$\{ \theta_{j} \}$ is a discretized version of a truncated normal distribution with mean $25$, standard deviation~$5$, minimum value $0$, and maximum value $44$.
-This was chosen to closely mirror the distribution from~\cite{Johndrow2020}.
+This closely mirrors the distribution from~\cite{Johndrow2020}.
+Strata for ages less than $40$ are collapsed due to the limited number of COVID-19 related deaths in this age range.
+For similar reasons, the multinomial parameter~${\bf p}_t$ is assumed to be constant in time with a Dirichlet prior.
 
@@
 
- To check how well our model fits the observed data, a posterior predictive check on aggregate death counts was performed.
+ To check model fit, a posterior predictive check on aggregate death counts was performed.
 Figure~\ref{fig:deathsppc} shows the results which suggests chosen SEIR model fits the observed data well.
@@
 Figure~\ref{fig:undercounting} in the supplementary materials computes the weekly undercount factor of observed case counts.
-Posterior distributions for the age-specific number of infected individuals will be used as part of the doubly-robust estimator of the active infection rate.
+Posterior distributions for the age-specific case counts are used as part of the doubly-robust estimator of the active infection rate.
 
@@
 Bias, however, does remain which we conjecture is due to the limited availability of symptom and other important covariate information to estimate strata-specific infection rates.
-This suggests better data collection in the future may strongly improve performance of the IPW methodology.
+This suggests better data collection in the future may strongly improve performance.
 
@@
  % \end{subfigure}
- \caption{Time-varying active infectionrate estimates based on unweighted, model-based, IPW and doubly robust methods.}
+ \caption{Time-varying active infection rate estimates based on unweighted, model-based, IPW and doubly robust methods.}
  \label{fig:tvestimates}
@@
 
- Application of the proposed methods relied on access to two critical datasets: (1) COVID-19 testing, case, hospitalization, and death data by demographic strata and (2) Facebook symptom survey.
-Public access to (1) required the author to submit a data request directly to the state of Indiana.
-Through conversations with the state of Indiana, the author's request was one of only five data requests to be approved by Indiana for public release.
+ The proposed methodology relied on access to two critical datasets: (1) COVID-19 testing, case, hospitalization, and death data by demographic strata and (2) Facebook symptom survey.
+Public access to (1) required submission of data requests to the state of Indiana.
+The author's request was one of only five data requests to be approved by Indiana for public release.
 Most states do not release such granular data, making selection bias adjustments difficult.
 Coordinated, systematic data collection and reporting is critical.
- Lack of covariate information including symptom and COVID-19 contacts on individuals seeking COVID-19 tests is unacceptable.
+ Lack of covariate information on individuals seeking COVID-19 tests is unacceptable.
 Random samples over time to supplement this data with auxiliary information is desperately needed.
@@
 
- A valid criticism of the proposed approach is that recording covariates on every tested individual is too time-consuming and costly, i.e., the proposed data collection protocol does not scale well.
-We argue that the list of relevant factors is not too long.
+ A valid criticism of the proposed approach is that recording covariates on every tested individual is time-consuming and costly.
+We argue that the list of relevant factors is of reasonable length.
 For example, symptom status and COVID-19 contact are clearly relevant factors in testing selection.
-Future work may consider how to include/exclude various factors over time that are not significant to limit citizen burden in reporting.
-Moreover, equation~\eqref{eq:auxinfoprob} presupposes covariate information is collected for all individuals in the nonprobability sample; however, this is often not feasible for state and local governments where rapid sample collection is prioritized over additional collection of covariate information.
-We can easily extend~\eqref{eq:auxinfoprob} by randomly sampling a subset of the nonprobability sample on which to collect the covariate information.
- %yielding a new pseudo-likelihood
- % \begin{equation*}
- % % \label{eq:auxinfoprob2}
- % \sum_{j=1}^N I_j I_j^{(2)} W_j^{(2)} \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
- % \end{equation*}
- % Here, $I_j^{(2)}$ is an indicator that the individual is in the probability subsample and $W_j^{(2)}$ is the probability of inclusion.
-Estimation proceeds by maximizing the pseudo-likelihood which now depends on random sampling in both the first and second term.
- This would allow for rapid testing and minimize data collection costs.
+Future work may consider how to include/exclude factors over time that are not significant to limit citizen reporting burden.
+Moreover, equation~\eqref{eq:auxinfoprob} presupposes covariate information is collected for all individuals in the nonprobability sample, which is often not feasible for state/local governments where rapid sample collection is prioritized.
+One can easily extend~\eqref{eq:auxinfoprob} by randomly sampling a subset of the nonprobability sample on which to collect the covariate information.
+ This would balance goals of rapid testing and auxiliary data collection.
 

### 2021-09-02 write-up/covid-writeup.tex
Update covid-writeup.tex

@@
  $$
- Assuming the selection propensity model is correctly specified, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:dr} shows that the doubly robust estimator is consistent and calculates the estimator's variance.
+ Assuming correct selection propensity model specification, then under the probability sample design and nonprobability sample design, Lemma~\ref{lemma:dr} shows that the doubly robust estimator is consistent and calculates the estimator's variance.
 See Appendix~\ref{app:asympderivations} for additional details.

### 2022-02-09 write-up/letters/response-ref-1-v1.tex
amanda is the best

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Plain Cover Letter
% LaTeX Template
%
% This template has been downloaded from:
% http://www.latextemplates.com
%
% Original author:
% Rensselaer Polytechnic Institute (http://www.rpi.edu/dept/arc/training/latex/resumes/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[11pt]{letter} % Default font size of the document, change to 10pt to fit more text

\usepackage{newcent} % Default font is the New Century Schoolbook PostScript font
%\usepackage{helvet} % Uncomment this (while commenting the above line) to use the Helvetica font

% Margins
\topmargin=-1in % Moves the top of the document 1 inch above the default
\textheight=8.5in % Total height of the text on the page before text goes on to the next page, this can be increased in a longer letter
\oddsidemargin=-10pt % Position of the left margin, can be negative or positive if you want more or less room
\textwidth=6.5in % Total width of the text, increase this if the left margin was decreased and vice-versa

\let\raggedleft\raggedright % Pushes the date (at the top) to the left, comment this line to have the date on the right
\usepackage{amsthm, amsfonts}
\usepackage{hyperref}

\def\Y{{\bf Y}}

\begin{document}

%----------------------------------------------------------------------------------------
%	ADDRESSEE SECTION
%----------------------------------------------------------------------------------------

\begin{letter}{Professor
	Peter Glynn\\
	Editor, {\em Journal of Applied Probability}}

%----------------------------------------------------------------------------------------
%	YOUR NAME & ADDRESS SECTION
%----------------------------------------------------------------------------------------

%\begin{center}
%\large
%\end{center}
%\vfill

\signature{Walter Dempsey\\
University of Michigan\\
Department of Biostatistics\\
1415 Washington Heights\\
Ann Arbor, MI 48103} % Your name for the signature at the bottom

%----------------------------------------------------------------------------------------
%	LETTER CONTENT SECTION
%----------------------------------------------------------------------------------------

\vspace{5mm}

\newpage

{\bf Response to Referees}

We appreciate all of the very helpful comments from the editor, associate editor, and the two referees which has helped improve the paper substantially. We have addressed the concerns as best as possible in this revision. We provide a further point-by-point explanation below. Any comments in {\it italics} indicate a statement from the referee report.

{\bf Response to Editor and Associate Editor}
\begin{enumerate}
\item {\it The referees have many specific concerns they are asking to address, and their expertise and viewpoints are important to revise this manuscript to improve its clarity and impact.}

\vspace{5mm}
We thank the editor-in-chief as this is an excellent point.  We now acknowledge that both topics are well known and cite relevant literature in Section 1.1; however, while infectious disease researchers and survey statisticians have been aware of the perils of self-selection and measurement error, the interplay in this context has been less well studied.   By extending the statistical decomposition of Meng (2018), we help shed light on how measurement error does not act to simply alter the magnitude of error but may in fact change the sign as well.  This decomposition motivates the new inverse-probability and doubly robust methodologies in the revised manuscript.
\vspace{5mm}

\item {\it Some of these deal with the fast-evolving nature of the pandemic with new information emerging every week and changing variants with different characteristics also changing our understanding of the disease course and spread and timing, that affect the content in the paper. I understand that it is not feasible for a peer-reviewed paper like this to include all the latest information in a fast-changing environment like this in which understanding can change quickly, but it is important in this revision to update the KEY points as much as possible based on updated knowledge, but otherwise revise the language to make it “forward compatible” in terms of relevance as things continue to change, e.g. by acknowledging some details of spread dynamics may change with future variants, but presenting the current work based on a particular SARS-CoV-2 variant as illustrative of the issues that arise with each new variant and other viruses, as well. That is, recognizing it may not be feasible to keep updating based on emerging knowledge from each new strain, but to lock down the discussion context but add enough explanation so the reader can see its relevance for new strains.}

\vspace{5mm}
x
\vspace{5mm}

\item {\it The AE also mentions a few comments as well:}
\begin{itemize}
	\item {\it Title: As noted by one of the reviewers, the “paradox” is never discussed/defined in the manuscript. Consider changing to “Statistical estimation of Coronavirus Case Counts: Measurement…” or some alternative.}
	\item {\it Organization: The figures and tables are not located anywhere near much of their introduction/discussion. For example, Table 1 was discussed on page 7 and is itself located on page 22. Consider the reorganization such that Figures and Tables are located near their discussion.}
	\item {\it Discussion on the Indiana Random Sample: On page 22, table 1, it appears that the random sample is not similar to the census estimates, potentially indicating non-response bias.}
	\item {\it Additionally, there may be significant potentially non-ignorable variables (e.g., political opinions and rural status) that drive non-response and are not presented. Is there any comment/discussion on the success of the randomization to be used as the “gold-standard” in this example? Much of the results appear to rely on the SRS and it ignores the potential for nonresponse bias.}
\end{itemize}

\vspace{5mm}
Thank you for raising this issue.  We discuss reporting delays and the relevant literature in Remark 1 in Section 2.2.  Given currently available public data, one cannot correct at the daily level for reporting delays.  Instead, we analyze COVID-19 data at the weekly level to account for potential reporting delays.  While not a perfect solution, it avoids clear reporting variation across day-of-week which can be seen in Figure 5.
\vspace{5mm}
\end{enumerate}
\newpage
{\bf Response to Reviewer 1}
I am incredibly grateful for this review. (PRAISE OUT THE WAZZOOO).

\begin{enumerate}
\item {\bf Section 1: Introduction.} {\it It may be useful in the introduction to briefly discuss the various purposes of testing for diseases – for COVID we have largely focused on diagnostic testing of people who are symptomatic or have a known or suspect exposure. Diagnostic testing will never provide valid prevalence or incidence estimates, especially for a disease where a significant proportion of cases are asymptomatic or pauci-symptomatic and thus unlikely to present for diagnostic testing. Screening tests are more appropriate as a potential tool for reconstructing prevalence or incidence tests, although these are seldom conducted at random. The random and non-random testing samples appear to represent screening and diagnostic testing goals, respectively, and this is part of why they can be expected not to deliver the same prevalence estimates.}

\vspace{5mm}
Section XX now discusses the purposes of testing.
\vspace{5mm}

\item {\bf Section 1.1}
\begin{itemize}
	\item {\it For the first goal of the paper, it might be helpful to link the biases with those outlined in a set of papers from the Harvard Center for Disease Dynamics: \url{https://doi.org/10.1007/s10654-021-00727-7} and DOI:\url{10.1093/aje/kwaa188}}
	\vspace{5mm}
	We now connect with XX
	\vspace{5mm}
	\item {\it For the second goal of the paper, it would be useful to discuss how the proposed method differs from the more traditional epidemiologic method of obtaining a validation sample. As I see it, the chief distinction is that here the “gold standard” is not a different test (although one could argue that the gold standard being used here is a screening test), but rather it is a less biased selection mechanism. Clarifying the links with validation studies could be helpful for epidemiologists \& statisticians alike, as the use of a validation study to correct for selection bias is a fairly novel suggestion.}
	\vspace{5mm}
	Remark wihtin validation subsection to contrast with validation sample.
	\vspace{5mm}
	\item {\it Finally, consider linking the arguments you are making with recent work on ‘target validity’ (see: \url{https://doi.org/10.1093/aje/kwy228}). Literature on target validity has largely been focused on specifying the problem of tackling internal and external validity together, but the method you propose here is a concrete attempt to address both types of validity and extends this conversation nicely.}
	\vspace{5mm}
	Validation subsection
	\vspace{5mm}
\end{itemize}
\item {\bf Section 2.1}
\begin{itemize}
	\item {\it Please clarify in the first paragraph the distinction between incubation period (time to symptom onset) and latent period (time to infectiousness). The viral load will be detectable by at least the end of the latent period which for SARS-CoV-2 occurs before the end of the incubation period. However, individuals are unlikely to present for diagnostic testing until after the end of the incubation period.}
	\vspace{5mm}
	We now connect with XX
	\vspace{5mm}
	\item {\it Please also clarify the terminology ‘active infection’ which in some instances appears to refer to individuals in the symptomatic period of infection, and in other instances appears to refer to individuals in the infectious period of infection. We know that these are not identical in SARS-CoV-2 infection and that difference is a key reason why screening and diagnostic tests are not easily comparable.}
	\vspace{5mm}
	Remark wihtin validation subsection to contrast with validation sample.
	\vspace{5mm}
	\item {\it Consider adding some additional information on screening tests, particular anything relevant to the random sample conducted in Indiana which is used as the validation set. Were the screening tests conducted using nasopharyngeal swabs (in many places screening tests use nasal swabs instead and often self-collected samples)? Were the molecular tests used to assay the samples the same for screening and diagnostic tests? Differences in sample collection and testing processes could undermine the utility of this random sample as a validation set.}
	\vspace{5mm}
	Validation subsection
	\vspace{5mm}
\end{itemize}
\item {\bf Section 2.1}
\begin{itemize}
	\item {\it This section could benefit from a few remarks on why it is important / useful to focus on publicly available testing data. There are certainly distinct statistical issues, if public data are aggregated, but are the selection bias and measurement error issues different between public and government datasets? If so, how do the solutions presented here apply to government datasets? Can similar or the same approaches be used by health departments to assess the pandemic and guide responses?}
	\vspace{5mm}
	Clarify that Indiana released a lot of aggregate information.  Two distinct goals: (1) what can we say with additional surveys but not additional testing, (2) guide health departments who have access to
	\vspace{5mm}
	\item {\it A timeline of restrictions and testing eligibility would be a useful addition to Section 2.2.1. Also, what there no time period where testing in Indiana was restricted on the basis of travel history or suspected exposure?}
	\vspace{5mm}
	Remark wihtin validation subsection to contrast with validation sample.
	\vspace{5mm}
\end{itemize}

\end{enumerate}
\newpage

{\bf Response to AE3}
\begin{itemize}
\item {\it XX}

\vspace{5mm}
WXX
\vspace{5mm}

\end{itemize}

\end{letter}

\end{document}

### 2022-02-09 write-up/letters/round1/response-ref-1-v1.tex
amanda is the best

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Plain Cover Letter
% LaTeX Template
%
% This template has been downloaded from:
% http://www.latextemplates.com
%
% Original author:
% Rensselaer Polytechnic Institute (http://www.rpi.edu/dept/arc/training/latex/resumes/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[11pt]{letter} % Default font size of the document, change to 10pt to fit more text

\usepackage{newcent} % Default font is the New Century Schoolbook PostScript font
%\usepackage{helvet} % Uncomment this (while commenting the above line) to use the Helvetica font

% Margins
\topmargin=-1in % Moves the top of the document 1 inch above the default
\textheight=8.5in % Total height of the text on the page before text goes on to the next page, this can be increased in a longer letter
\oddsidemargin=-10pt % Position of the left margin, can be negative or positive if you want more or less room
\textwidth=6.5in % Total width of the text, increase this if the left margin was decreased and vice-versa

\let\raggedleft\raggedright % Pushes the date (at the top) to the left, comment this line to have the date on the right
\usepackage{amsthm, amsfonts}

\def\Y{{\bf Y}}

\begin{document}

%----------------------------------------------------------------------------------------
%	ADDRESSEE SECTION
%----------------------------------------------------------------------------------------

\begin{letter}{Professor
	Peter Glynn\\
	Editor, {\em Journal of Applied Probability}}

%----------------------------------------------------------------------------------------
%	YOUR NAME & ADDRESS SECTION
%----------------------------------------------------------------------------------------

%\begin{center}
%\large
%\end{center}
%\vfill

\signature{Walter Dempsey\\
University of Michigan\\
Department of Biostatistics\\
1415 Washington Heights\\
Ann Arbor, MI 48103} % Your name for the signature at the bottom

%----------------------------------------------------------------------------------------
%	LETTER CONTENT SECTION
%----------------------------------------------------------------------------------------

\vspace{5mm}

\newpage

{\bf Response to Referees}

We appreciate all of the very helpful comments from the editor-in-chief, four associate editors, and the referee which has helped improve the paper substantially. We have addressed the concerns as best as possible in this revision. We provide a further point-by-point explanation below. Any comments in {\it italics} indicate a quote from the referee report.

{\bf Response to Editor-in-Chief}
\begin{enumerate}
\item {\it Selection bias and measurement error are certainly well known.
Certainly worth pointing out as both apply here.  But both should be acknowledged as being very well known.}

\vspace{5mm}
We thank the editor-in-chief as this is an excellent point.  We now acknowledge that both topics are well known and cite relevant literature in Section 1.1; however, while infectious disease researchers and survey statisticians have been aware of the perils of self-selection and measurement error, the interplay in this context has been less well studied.   By extending the statistical decomposition of Meng (2018), we help shed light on how measurement error does not act to simply alter the magnitude of error but may in fact change the sign as well.  This decomposition motivates the new inverse-probability and doubly robust methodologies in the revised manuscript.
\vspace{5mm}

\item {\it My reaction after reading the introduction was: "OK, we know that.
But, what do you suggest we do?  Throw up our hands and go home?
Just tell people, `We can't help you.  The data won't help us to
model the epidemic or to forecast what will happen. Forget the data,
we'll just have to wait and see what happens.'" I don't think that's
the message you want to convey.  And we certainly do not want non-
statisticians (interested in Covid-19 data) to read your paper and
advise public health officials against collecting data altogether
because it has too many biases!  Rather, I think you want to alert
the reader that adjustments need to be made before modeling the
data, and, more importantly, we may need to develop new models
for this situation where most of the outcomes refer to suspected
individuals, not asymptomatic cases.}

\vspace{5mm}
We agree that the original manuscript was too focused on the scientific communication perspective and may have given the wrong impression to non-statisticians.  The revised manuscript now fits well within the mission of AOAS by focusing on novel methods that provide a route forward.  We then apply these methods to a current and important scientific problem. In particular, we use the decomposition to motivate new methodology that leverages auxiliary information to address selection bias. In Section 2, we now introduce the data streams that illustrate this methodology. First, we worked with the state of Indiana to release daily COVID-19 testing, case-count, and deaths by demographic strata. Second, we were granted access to the Facebook symptom survey which provides daily measures of important time-varying covariates such as the fraction of individuals currently experiencing COVID-19 related symptoms and/or in recent contact with a COVID-19 positive individual.  The Facebook survey provides auxiliary information that is used to construct selection propensities, which are then incorporated into an inverse-probability weighting method (introduced in Section 4.1) to estimate time-varying active infection rates. Of course, many infectious disease experts prefer compartmental model-based approaches built from differential equations such as the susceptible-exposed-infected-recovered (SEIR) model.  In Section 4.2, we build a novel compartmental model to take advantage of strata-level COVID-19 death count data.  In Section 4.3, we show how to combine these models with inverse weighting to construct doubly-robust estimates of active infection rates.  We extend the decompositions to the weighting and doubly robust setting which then guide our discussion of practical implications.
\vspace{5mm}

\item {\it BTW, this surely is not a new problem - likely HIV testing is
generally done on only those individuals suspected to be in
high-risk groups.  Nor is the problem of reporting delays;
CDC routinely adjusts monthly counts of AIDS for reporting delays.
You may need to reference that literature.}

\vspace{5mm}
Thank you for raising this issue.  We discuss reporting delays and the relevant literature in Remark 1 in Section 2.2.  Given currently available public data, one cannot correct at the daily level for reporting delays.  Instead, we analyze COVID-19 data at the weekly level to account for potential reporting delays.  While not a perfect solution, it avoids clear reporting variation across day-of-week which can be seen in Figure 5.
\vspace{5mm}

\item {\it Another point that may be worth mentioning is the potential
effect of length-biased sampling: the longer an individual
has been exposed, the more likely (s)he is to seek symptomatic
testing).  This effect hopefully will be absent in context of
random selection of individuals for asymptomatic testing.}

\vspace{5mm}
This is an excellent point and was the motivation for considering how to address length-biased sampling and other potential reasons that make individuals more likely to seek out testing.  Here, we address this by building selection propensities that may depend both on symptom status and the length of exposure as reported by the individual at the time of testing.   The lack of data to address length-biased sampling is pointed out in Section 6 and guides the suggested real-world improvements to data collection that will enable our proposed approach to combine survey statistical methods with infectious disease modeling to improve our understanding of active infection rates over time.
\vspace{5mm}

\item {\it Incidentally, I see the most frustrating part of the case count
reports as lack of two critical covariates on each case: (i) age;
(ii) presence or absence of relevant co-morbidity (auto-immune
condition, suppressed immune system, compromised respiratory or
lung condition).  I think the case count data, adjusted for
reporting delays and incorporating undertainty due to measurement
error as best we can, would be far more useful with these two
covariates.  And if we can get a handle on how to adjust for
selection bias, even better.}

\vspace{5mm}
We agree wholeheartedly with this critique.  To try and improve covariate reporting, we worked with the state of Indiana to release daily COVID-19 testing, case-count, and deaths by demographic strata. We acknowledge the limitations of available demographic information and use the Facebook symptom survey to try and impute the remaining important covariates.  For a separate stratified random sample from Indiana that we were able to access, these critical covariates are available and we show how this data can be used to improve over unweighted approaches.  Overall, the proposed methodology uses these critical covariates to account for selection bias.  Model-based approaches are extended to handle this information and combined to build doubly robust estimation strategies as well.  These new methods, aimed directly at this frustrating issue, motivate our practical recommendations to improve policy making and increase accuracy in estimation using case count data.
\vspace{5mm}

\end{enumerate}

\noindent {\bf Editor-in-Chief: Specific comments}

\begin{itemize}
\item {\it Abstract, l.3: "Throughout this" this what? "this pandemic"?}

\vspace{5mm}
Yes, we meant pandemic.  This has now been made explicit.
\vspace{5mm}

\item {\it p2, p.5-6: "For simplicity, the dynamic nature of the outbreak and recoverability of individuals are initially ignored." Drastic oversimplification, don't you think?}

\vspace{5mm}
We agree that this is a simplification.  However, we continue to start in this setting as a means to help readers understand the statistical decomposition in Section 3.1 as well as the pseudo-likelihood approach detailed in Section 4.  Both are extended to account for the dynamic nature of the outbreak in subsequent sections.
\vspace{5mm}

\item {\it p3, l.-3: Generally, $X_j = \mu_j + \epsilon_j$.Here, it seems as those you are saying: $P_j = 1$ if $\epsilon_j \neq 0$ and $P_j = 0$ if $\epsilon_j = 0$, is that right?}

\vspace{5mm}
We are focused exclusively on the binary outcome setting where $Y_j \in \{0,1\}$ as one is either COVID-19 positive~($Y_j = 1$) or negative~($Y_j = 0$).  Under measurement error, we observe $Y_j^\star \in \{0,1\}$ which is equal to $Y_j (1-P_j) + (1-Y_j) P_j$ where $P_j$ is the binary indicator of observing the incorrect value.
\vspace{5mm}

\item {\it p5, l.-2:  I must have missed the def'n of $D_M$ -- I looked back
three times and still cannot find it.}

\vspace{5mm}
The term $D_M$ is defined in equation (3.1) and represents the \emph{imperfect testing adjustment} which is a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence.  Figure 2 visualizes the adjustment as a contour plot to help the reader understand how the adjustment
\vspace{5mm}

\item {\it You'll also need to define SIR.  See AE2's note.}

\vspace{5mm}
We apologize for the oversight.  We thank you and AE2 for bringing this to our attention.  We now mathematically define the susceptible-infected-recovered (SIR) model in Section 3.2 as well as the susceptible-exposed-infected-recovered (SEIR) in Section 4.2.
\vspace{5mm}

\item {\it p6, Fig 1 has no legend.  It also looks like most of the action is in the region (Sampling Frequency 0.5 to 5), (Odds Ratio 1-50). I also think that a plot with contour levels would both avoid the need for color altogether and would be much easier to read.  Did you try to read the labels on the axes?  It's impossible unless you magnify the pdf 8-fold.}

\vspace{5mm}
We have fixed Figure 1 (now Figure 2) to be a contour plot with appropriately sized labels on the axes. We apologize for needing a magnifying glass to read it before.
\vspace{5mm}

\item {\it p10, Fig 2: you definitely do not need color here; different
line types suffice.  Likewise for Fig 3 (p14) and Fig 4 (p15).}

\vspace{5mm}
We have fixed all figures to be in grayscale and use different line types.
\vspace{5mm}

\item {\it p18: Please follow AOAS Guidelines for references; see the
web page from where you submitted your MS.}

\vspace{5mm}
The references now follow AOAS Guidelines.
\vspace{5mm}

\item {\it p20: Appendices can go into a Supplementary File (unless
you have a strong reason for including them here after the References).}

\vspace{5mm}
The appendices have been moved to a supplementary file as suggested.
\vspace{5mm}

\end{itemize}
\newpage

{\bf Response to AE1}
\begin{itemize}
\item {\it I was confused about what prevalence and case numbers the author is addressing. My understanding is that there are primary two assays, one for serology (evidence of ever having the virus) measured in the blood, and the other is presence of the active virus measured by swabbing the throat.  The characteristics of the assays, in terms of measurement error, are different.  In addition, there exist several different laboratory procedures for doing these two types of assays with varying accuracy.  The paper does not seem to address these basic issues.}

\vspace{5mm}
We thank the reviewer for points this out. Section 2.1 now discusses diagnostic testing.  Given the revision is focused mostly on molecular tests, we limit our discussion to these types of assays and their scientific use in managing the pandemic.  We discuss reasons behind false positives/negatives and cite the most recent systematic reviews of sensitivity and specificity for RT-PCR tests.
\vspace{5mm}

\item {\it My primary concern relates to the fundamental mission of AOAS. The manuscript has very limited data analysis, limited primarily to pp.14-15, in part because the data are very limited: it is difficult to obtain data about the various aspects of the testing for this virus. The data analysis that is presented on pp.14-15 is mostly about accounting for sensitivity and specificity of the serology assays, not the testing for presence of the Covid-19 virus at a particular time.  There is an assumption with the serology that anyone who has had Covid-19 at some time would have antibodies. These assumptions need to be explicitly made.  I am not sure if anyone knows whether this assumption is even correct or to what degree it might be correct. How does this fit into the measurement error that is discussed and illustrated in the data analysis?}

\vspace{5mm}
We agree that the original manuscript was too focused on the scientific communication perspective although it did contain original derivations related to extending the statistical error decomposition of Meng (2018) to account for measurement error.  The revised manuscript now fits well within the mission of AOAS by focusing on novel methods and clear application to a current and important scientific problem. In particular, we use the decomposition to motivate new methodology that leverages auxiliary information to address selection bias. In Section 2, we now introduce the data streams that illustrate this methodology. First, we worked with the state of Indiana to release daily COVID-19 testing, case-count, and deaths by demographic strata. Second, we were granted access to the Facebook symptom survey which provides daily measures of important time-varying covariates such as the fraction of individuals currently experiencing COVID-19 related symptoms and/or in recent contact with a COVID-19 positive individual.  The Facebook survey provides auxiliary information that is used to construct selection propensities, which are then incorporated into an inverse-probability weighting method (introduced in Section 4.1) to estimate time-varying active infection rates. Of course, many infectious disease experts prefer compartmental model-based approaches built from differential equations such as the susceptible-exposed-infected-recovered (SEIR) model.  In Section 4.2, we build a novel compartmental model to take advantage of strata-level COVID-19 death count data.  In Section 4.3, we show how to combine these models with inverse weighting to construct doubly-robust estimates of active infection rates.  We extend the decompositions to the weighting and doubly robust setting which then guide our discussion of practical implications.
\vspace{5mm}

\item {\it The beginning of the manuscript describes the very important issue of selection bias.  But the discussion is regrettably limited.  It is unfortunate that the manuscript only refers to the Covid-19 data (not the analysis) and offers insufficient attention to what I view as the most important issue.  The manuscript could have offered both greater depth and wider coverage of selection bias and other issues raised by the author.}

\vspace{5mm}
We now expand our coverage of selection bias issue by considering what it will take to overcome it with respect to COVID-19 case count data.  The auxiliary information helps address selection bias by allowing us to build selection propensity models while the doubly robust methodology helps us incorporate well known infectious disease models.  Overall, the paper has moved from just acknowledging the issue of selection bias and measurement error -- albeit from a new lens using the decomposition of Meng (2018)) -- to providing solutions and practical recommendations based on the proposed methodology.
\vspace{5mm}

\item {\it All in all, I don't see enough innovation in the data analysis and methods described here that would make AOAS a good home for it.  This manuscript appears to be a methodological review of data issues related to the measuring of the public health impact of the Covid-19 pandemic. I think this manuscript would be better suited for quantitative epidemiology journal or a quantitative journal in infectious diseases.}

\vspace{5mm}
As stated above, we agree that the original manuscript was too focused on the scientific communication perspective.  The revised manuscript, with its methodological innovation and application to a current and important relevant scientific application, now fits well within the mission of AOAS.
\vspace{5mm}

\end{itemize}

\newpage

{\bf Response to AE2}
\begin{enumerate}
\item {\it I read the paper (AOAS2006-048) but did not study any derivations. It seems straightforward, but I'm not well acquainted with the statistical area and do not have much intuition for the author's development. I'm not suggesting that it is incorrect. To get a reasonable understanding of it, I would have to work carefully through the logic and details.  In the interest of time, I've not done so.}

\vspace{5mm}
The original derivations were straightforward and followed from some arithmetic calculations, but the main point was the resulting statistical error decomposition.  While infectious disease researchers and survey statisticians have been aware of the perils of self-selection and measurement error, the interplay in this context has been less well studied.  By extending the statistical decomposition of Meng (2018), we help shed light on how measurement error does not act to simply alter the magnitude of error but may in fact change the sign as well.  This decomposition motivates the new inverse-probability and doubly robust methodologies in the revised manuscript.
\vspace{5mm}

\item {\it I noticed quite a few rough edges.}

\begin{itemize}
\item {\it Page 4, line 4: The summation variable should be j.}

\vspace{5mm}
We thank the reviewer for pointing this out.  This has been fixed.
\vspace{5mm}

\item {\it Page 8, middle: It should not be difficult to guess what "SIR" means (if one doesn't know), but the author should explain it.}

\vspace{5mm}
We thank the reviewer for pointing this out.  We now mathematically define the susceptible-infected-recovered (SIR) model in Section 3.2 as well as the susceptible-exposed-infected-recovered (SEIR) in Section 4.2.  The SEIR model motivates a new compartmental model that uses daily demographic strata-level COVID-19 death data to estimate daily COVID-19 case counts.  These models are then used in a novel doubly robust method to estimate weekly active infection rates over time.
\vspace{5mm}

\item {\it Figure 3: The caption should explain the dashed curves in part B.}

\vspace{5mm}
We thank the reviewer for pointing this out.  These are the ratio estimator and effective reproductive rate estimator under different relative sampling fractions $M = f_1/f_0$. We see the bias increases as a function of the relative sampling fraction.  We now explain the dashed curves in Figure 3a and 3b.
\vspace{5mm}

\item {\it Page 14, bottom: I wonder about the validity of the estimates of specificity and sensitivity from the Santa Clara study. I recall seeing criticism when the results appeared, but I may be mistaken.}

\vspace{5mm}
We agree that the study has come under significant criticism.  We have updated our choices of specificity and sensitivity to mirror the recent literature on specificity and sensitivity of RT-PCR testing which has grown substantially over the past few months.
\vspace{5mm}
\end{itemize}

\end{enumerate}
\newpage

{\bf Response to AE3}
\begin{itemize}
\item {\it I took some time to read the paper this evening.  The problem considered in this paper is very important and I appreciated that someone is looking into it.  However, I am not sure AOAS is the best home for the paper.  In my humble opinion, epidemiology or general science journals, or maybe medical journals, might be a better fit.  It might receive even more attention if published in those journals than a statistical journal.}

\vspace{5mm}
We agree that the original manuscript was too focused on the scientific communication perspective although it did contain original derivations related to extending the statistical error decomposition of Meng (2018) to account for measurement error.  The revised manuscript now fits well within the mission of AOAS by focusing on novel methods and clear application to a current and important scientific problem. In particular, we use the decomposition to motivate new methodology that leverages auxiliary information to address selection bias. In Section 2, we now introduce the data streams that illustrate this methodology. First, we worked with the state of Indiana to release daily COVID-19 testing, case-count, and deaths by demographic strata. Second, we were granted access to the Facebook symptom survey which provides daily measures of important time-varying covariates such as the fraction of individuals currently experiencing COVID-19 related symptoms and/or in recent contact with a COVID-19 positive individual.  The Facebook survey provides auxiliary information that is used to construct selection propensities, which are then incorporated into an inverse-probability weighting method (introduced in Section 4.1) to estimate time-varying active infection rates. Of course, many infectious disease experts prefer compartmental model-based approaches built from differential equations such as the susceptible-exposed-infected-recovered (SEIR) model.  In Section 4.2 we build a novel compartmental model to take advantage of strata-level COVID-19 death count data.  In Section 4.3, we show how to combine these models with inverse weighting to construct doubly-robust estimates of active infection rates.  We extend the decompositions to the weighting and doubly robust setting which then guide our discussion of practical implications.
\vspace{5mm}

\end{itemize}
\newpage

{\bf Response to AE4}
\begin{itemize}
\item {\it I’ve read through it now, and think there is value to what they are doing.  It is not rocket science, but that is OK to me if there is applied impact. I think that the author is trying to substantively think deeply about the measurement error and selection bias in case counts, which I constantly encounter and see is a major issue in managing this pandemic.  But I would like to see more practical guidelines in the discussion.  I see some promise here but no guarantees.}

\vspace{5mm}

\vspace{5mm}
We agree that the original work was not rocket science. While infectious disease researchers and survey statisticians have been aware of the perils of self-selection and measurement error, the interplay in this context has been less well studied.  By extending the statistical decomposition of Meng (2018), we help shed light on how measurement error does not act to simply alter the magnitude of error but may in fact change the sign as well.
\vspace{5mm}

To build practical guidelines, we first worked on accessing high quality data.
To this end, we worked with the state of Indiana to release daily COVID-19 testing, case-count, and deaths by demographic strata. Second, we were granted access to the Facebook symptom survey which provides daily measures of important time-varying covariates such as the fraction of individuals currently experiencing COVID-19 related symptoms and/or in recent contact with a COVID-19 positive individual.
\vspace{5mm}

This data motivated our practical proposal: use time-varying surveys such as the Facebook symptom survey to provide auxiliary information that allows us to construct our inverse-probability weights (introduced in Section 4.1) to estimate time-varying active infection rates. Of course, many infectious disease experts prefer compartmental model-based approaches built from differential equations such as the susceptible-exposed-infected-recovered (SEIR) model.  In Section 4.2, we build a novel compartmental model to take advantage of strata-level COVID-19 death count data.  In Section 4.3, we show how to combine these models with inverse weighting to construct doubly-robust estimates of active infection rates.
\vspace{5mm}

The main limitation of our proposed approach is the limited covariate information being currently reported with case counts.  Specifically, symptom status, contact with a COVID-19 positive individual, and other covariate information is either not collected or not reported.  This guides the suggested real-world improvements to data collection that will enable our proposed approach to combine survey statistical methods with infectious disease modeling to improve our understanding of active infection rates over time.
\vspace{5mm}

\end{itemize}
\newpage

{\bf Response to Referee}
\begin{itemize}
\item {\it I skimmed through the MS very quickly. Although it doesn’t seem like rocket science, and may also be making points that are very well known to infectious disease researchers, it may be that the author is making some good points and, more importantly, may be translating a message well known in one field to another field. If si, then the MS might be worthwhile from a science communication perspective.}

\vspace{5mm}
We agree that the original work was not rocket science. While infectious disease researchers and survey statisticians have been aware of the perils of self-selection and measurement error, the interplay in this context has been less well studied.  By extending the statistical decomposition of Meng (2018), we help shed light on how measurement error does not act to simply alter the magnitude of error but may in fact change the sign as well.
\vspace{5mm}

We have now moved well beyond the science communication perspective.  In particular, we use the decomposition to motivate new methodology that leverages auxiliary information to address selection bias. In Section 2, we now introduce the data streams that illustrate this methodology. First, we worked with the state of Indiana to release daily COVID-19 testing, case-count, and deaths by demographic strata. Second, we were granted access to the Facebook symptom survey which provides daily measures of important time-varying covariates such as the fraction of individuals currently experiencing COVID-19 related symptoms and/or in recent contact with a COVID-19 positive individual.  The Facebook survey provides auxiliary information that is used to construct selection propensities, which are then incorporated into an inverse-probability weighting method (introduced in Section 4.1) to estimate time-varying active infection rates.  Of course, many infectious disease experts prefer compartmental model-based approaches built from differential equations such as the susceptible-exposed-infected-recovered (SEIR) model.  In Section 4.2, we build a novel compartmental model to take advantage of strata-level COVID-19 death count data.  In Section 4.3, we show how to combine these models with inverse weighting to construct doubly-robust estimates of active infection rates.  We extend the decompositions to the weighting and doubly robust setting which then guide our discussion of practical implications.
\vspace{5mm}

The main limitation of our proposed approach is the limited covariate information being currently reported with case counts.  Specifically, symptom status, contact with a COVID-19 positive individual, and other covariate information is either not collected or not reported.  This guides the suggested real-world improvements to data collection that will enable our proposed approach to combine survey statistical methods with infectious disease modeling to improve our understanding of active infection rates over time.
\vspace{5mm}

\end{itemize}

\end{letter}

\end{document}

### 2022-02-09 write-up/letters/cover letter-covid.tex
amanda is the best

--- 2021-05-14 write-up/cover letter-covid.tex
@@
 
-Kindly find enclosed the Report titled ``Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic,'' by Walter Dempsey.
+Kindly find enclosed the revised version of the manuscript number AOAS2006-048 titled ``Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic.''  We have revised the article to address the referees’ requests about content, exposition, and other points of clarification.
+The reports called for a major revision to re-structure the manuscript around a timely data analysis and related methodology, a more careful discussion of COVID-19 molecular versus serological testing, and to more carefully distinguish the current work from past work on the issues of selection bias and measurement error.
+The new version complies with this request.
 
-The report contains novel empirical findings and an investigation of statistical paradoxes arising from the analysis of reported coronavirus case-counts that are being used by governments to guide policies and decision making.
-We have discovered an estimator and statistical decomposition of the associated error that shows the 21.5 million tests conducted in the US provides information on COVID-19 prevalence equivalent to a simple random sample of size 92.
-Analogous decompositions for the rate of change in prevalence and the effective reproduction number are used to demonstrate how the magnitude and sign of estimation bias varies over time, a bias which has strong policy implications.
-Overall, our findings demonstrate the statistical limitations of increased testing in the presence of selection bias and measurement error as well as how easy it is for researchers analysing reported coronavirus case-counts to not only fool themselves but be confident in their incorrect conclusions.
-We believe the work presents an important advance in the statistical understanding of the COVID-19 pandemic.
+The revision focuses on taking the statistical decomposition previously derived and answering the question of what is to be done given potential selection bias and known measurement-error.
+To start, we worked directly with the state of Indiana and are one of only five data requests to be approved by the Governor's office.
+The newly public dataset reports daily COVID-19 testing, case counts, and deaths broken out by demographic strata.
+To address selection bias and estimate active infection rates, we introduce a novel inverse-probability weighting method that uses auxiliary information to form selection propensities.
+The necessary auxiliary information comes from Facebook's large-scale symptom survey which provides relevant information on time-varying covariates such as the fraction of individuals in the population reporting COVID-19 symptoms and/or recent contact with a COVID-19 positive individual.
+We extend to a doubly robust method that allows researchers to incorporate compartmental models into their estimation.
+The statistical error decomposition is extended to each setting and guides our discussion.
+The revision now goes well beyond a reframing of the classical problems of selection bias and measurement error by providing health policy experts with new tools to leverage auxiliary information in addressing the ongoing COVID-19 pandemic and future infectious disease outbreaks.
+
+We believe the work presents an important advance in how researchers should think about addressing time-varying selection bias to better understanding the COVID-19 pandemic trajectory.
+Attached please find a detailed response to the referees outlining how the changes to the manuscript address the concerns raised in their reports.
 

### 2022-02-09 write-up/letters/response-ref-1-v1.tex
amanda thx

@@
 	\vspace{5mm}
-	Remark wihtin validation subsection to contrast with validation sample.
+	Remark within validation subsection to contrast with validation sample.
 	\vspace{5mm}
@@
 \end{itemize}
-\item {\bf Section 2.1}
+\item {\bf Section 2.1: Diagnostic testing}
 \begin{itemize}
@@
 	\vspace{5mm}
+
 	We now connect with XX
@@
 	\vspace{5mm}
-	Remark wihtin validation subsection to contrast with validation sample.
+
+	We clarify that here we are interested in all individuals who .
+
 	\vspace{5mm}
@@
 	\vspace{5mm}
-	Validation subsection
+
+	We clarify the means. These were individuals who report to the exact same set of locations as in the non-random sample.  So they are representative.
+	Note, we do not use the outcome data in the method for validation, only for adjustment for self-selection in the diagnostic tests.
 	\vspace{5mm}
@@
 	\vspace{5mm}
+
 	Clarify that Indiana released a lot of aggregate information.
@@
 	\vspace{5mm}
-	Remark wihtin validation subsection to contrast with validation sample.
+
+	We have a timeline
 	\vspace{5mm}
+\end{itemize}
+\item {\bf Section 2.3: Probabilistic Samples}
+\begin{itemize}
+	\item {\it The $\approx 30\%$ response rate in the supposedly random sample requires some remarks.
+Even if the original list of eligible individuals was random, it seems unlikely that the final sample remains random with such low response.
+Is it possible to describe any characteristics that may differ between the intended sample and the final sample?
+Certainly the final random sample differs in many ways from the Census data presented in Table 1.
+Also, please clarify if the Census data is US total or Indiana specific.
+}
+	\vspace{5mm}
+
+	We can discuss the non-response weights used in their paper to adjust.  And that this is a different and solvable issue (we can just weight by the )
+	\vspace{5mm}
+	\item {\it For the Facebook / CMU survey sample, please present the age breakdown of sampled individuals.
+It is well-known that the age-distribution of Facebook users differs from the general population, so this is unlikely to be a representative sample of the Indiana population.
+Other demographic information on this sample would be valuable too.
+(Why is it not included in Table 1?)}
+	\vspace{5mm}
+
+	We include an age breakdown of the data.
+Note that weights are used to adjust and we are using these in the calculations.
+We now include a table in the appendix a breakdown of the demographic information for each sample.
+	\vspace{5mm}
+\end{itemize}
+\item {\bf Section 3: Analysis of case count data}
+\begin{itemize}
+	\item {\it Please clarify the definition of ‘active infection’ (any infection, vs infectious, vs symptomatic, vs detectable by molecular testing?).
+This has important implications for the construction of the compartmental model and for the definition of your period prevalence (AIR).}
+	\vspace{5mm}
+
+	XXXXX
+	\vspace{5mm}
+	\item {\it Throughout, there is a general issue of too few distinct variable identifiers – for example, using Y to refer to infection status and y to refer to test status will likely lead to confusion.}
+	\vspace{5mm}
+
+	We clarify our choice of notation by adding an appendix with notation.
+	\vspace{5mm}
+
+	\item {\it The “natural candidate for AIR” presented is the quantity that has more commonly been referred to as the “test positivity rate” (assuming that yi does in fact indicate test status rather than infection status which is obviously unknown).
+It would be helpful to mention this.}
+	\vspace{5mm}
+
+	This is now stated explicitly
+	\vspace{5mm}
+	\item {\it }
 \end{itemize}

### 2022-02-10 write-up/letters/response-ref-1-v1.tex
Amanda is mean

@@
 
+Let us start by saying that these are the most thorough and helpful reviews we have ever received on a submitted manuscript.
 We appreciate all of the very helpful comments from the editor, associate editor, and the two referees which has helped improve the paper substantially.
@@
 	This is now stated explicitly
+\end{itemize}
+\item {\bf Section 3.1: Imperfect testing}
+\begin{itemize}
+	\item {\it Here, the value of distinguishing between test status and infection status becomes clearer.
+By using $P_j$ to mean ‘error status of testing’, the derivations in this section lead to a focus on false positives and false negatives.
+However, the real quantity of interest both for individuals and for determining prevalence is surely the positive predictive value and the negative predictive value – that is, Pr($Y_j$=1|positive test result) and Pr($Y_j$=0|negative test result).
+Unlike the FP and FN which (because they are strictly functions of sensitivity and specificity) are only affected by test characteristics, the PPV and NPV are affected also by prevalence.
+Acknowledging this leads more directly to the benefit of an iterative method for validating the test results (since FP \& FN should not depend on Y or y \& thus not need an iterative method).}
 	\vspace{5mm}
-	\item {\it }
+
+	The adjustment is exactly related to the Bayes rule PPV/NPV relationship with FP, FN, and $\bar Y$.
+	\vspace{5mm}
+	\item {\it Also worth noting that test positivity is not strictly binary (since it’s generally based on an underlying continuous cycle threshold), and so measurement error could potentially lead towards bias away from, or across, the null even in the absence of selection bias.}
+	\vspace{5mm}
+
+	We now include a discussion of the underlying continuous cycle (with proper credit).
+The binary variable of interest can depend on infectious is what is often most important.
+So we highlight that most but acknowledge the binary variable can be defined....
+	\vspace{5mm}
+
+	\item {\it The estimator $y = (yn-FP) /(1-(FP+FN))$ is a standard estimator for correcting for measurement error using FP \& FN estimated from a validation sample, where yn is the mis-measured average \& y is the corrected average.
+The use of this estimator to iteratively estimate the point prevalence is a good contribution of this paper, but the failure to distinguish between infection \& test status in the variable names somewhat obscures that this is what the author is doing.}
+	\vspace{5mm}
+
+	We now more carefully distinguish.
+And yes, this is the standard estimator.
+We have simply derived it from a new perspective (the decomposition perspective which didn't require any Bayesian machinery).
+	\vspace{5mm}
+
+	\item {\it Finally, in the section on effective sample size, please clarify why the FP and FN would be dependent on testing capacity.
+False positive and false negative rates, as well as sensitivity and specificity, are strictly characteristics of a given test – if increased test capacity or burden would alter the sample collection quality then FP \& FN could change.
+But if increased test capacity simply changes the underlying prevalence in the group tested (because of opening up testing to uninfected individuals), then the FP \& FN should not change, but the PPV and the NPV will change since these are dependent on prevalence.}
+	\vspace{5mm}
+
+	I now give an explicit example.
+The test characteristics may change based on the nurse (cite) and/or what exact test is used.
+For example, a center that moves from nasal swab to spit to increase testing capacity will see a ...(cite).
 \end{itemize}
+\item {\bf Section 3.2: Regrettable Rates}
+\begin{itemize}
+	\item {\it Here, and throughout the theoretical example, the model is an SIR model.
+However, we know that SARS-CoV-2 is better represented with an SEIR model, and indeed the applied example later in the manuscript uses an SEIR approach.
+Since this paper is very particularly addressing an issue with COVID data it would be beneficial to use an SEIR model throughout rather than the simpler SIR model.}
+	\vspace{5mm}
 
+	We now match the SEIR and still see the same thing for the ratio.
+\end{itemize}
+\item {\bf Section 3.3: Effective Reproduction Number}
+\begin{itemize}
+	\item {\it Here again, an SEIR model would be valuable.
+It is also important to clarify that the effective reproductive number can be defined in a number of subtly different ways, and that these impact both the method of estimation and the interpretation of the effect.
+See for example Cori et al; Am J Epi, 178(9):1505-12.
+2013 and Wallinga \& Tuenis Am J Epi; 160(6):509-16.
+2004 for methods that, while similar, estimate subtly different quantities.
+It would be helpful to clarify exactly what effective reproductive number is being estimated here – from the formulae, I suspect it is the type described by Cori et al rather than Wallinga \& Tuenis.}
+	\vspace{5mm}
+
+	We clarify the subtle differences and that we are of the Cori type.
+	\vspace{5mm}
+	\item {\it The Rt estimator will differ if an SEIR model is used, and it is not clear (to me) that the same intuition about the error will hold.}
+	\vspace{5mm}
+
+	This estimator is motivated by the SIR but is used more generally in COVID plots (see \href{here}{https://academic.oup.com/aje/article/189/11/1438/5864525?searchresult=1} for use in plots).
+This is closely related to Cori paper \href{as reviewer identified}{https://academic.oup.com/aje/article/178/9/1505/89262?searchresult=1}.
+See \href{here}{https://www.nature.com/articles/s41598-020-76563-8.pdf?origin=ppub} where we have the basic reproduction number for SEIR and SIR.
+We used this to alter the proposed moment-based estimator to be more appropriate for SEIR in two ways.
+First, the constant is now $(1/\gamma + 1/\sigma)$.
+Second, we acknowledge the estimator $K_t/K_{t-1}$ may have issues of smoothness.
+We can smooth these by considering aggregate differences.
+We provide estimates of ratio and eff repro number in the appendix for 3- and 7- day window lengths.
+We see a similar story.
+\end{itemize}
+\item {\bf Section 3.4: Rate comparisons}
+\begin{itemize}
+	\item {\it This section would also be more relevant with an SEIR model}
+	\vspace{5mm}
+
+	You got it dude.
+	\vspace{5mm}
+\end{itemize}
+\item {\bf Section 4.1: Selection propensity estimation}
+\begin{itemize}
+	\item {\it The use of weights to correct for sampling bias is an established practice in epidemiology.
+It would be helpful to discuss how the proposed method differs from survey sampling weights or transportability weights (see Am J Epidemiol 2017 Oct 15;186(8):1010-1014.
+doi: 10.1093/aje/kwx164.).}
+	\vspace{5mm}
+
+	Inverse probability of selection weights describe by Cole and Stuart are exactly what we are building.  However, how to build them is the question.
+	As far as we are aware, the \href{approaches}{https://academic.oup.com/aje/article/172/1/107/82724} require knowledge of the selection mechanism which is not available in our setting.
+Instead, we require auxiliary information.
+Here, (cite elliott stuff).
+A key different is the time-varying nature which we highlight.
+	\vspace{5mm}
+	\item {\it This section, and the sections below, also become somewhat confusing to follow because of the multiple different variables which are all denoted Ij.
+Since a chief goal is to describe and correct for errors in the sampling process, it would be much more useful for the reader if there were different notation for each of the different sampling mechanisms being discussed / used (eg, inclusion in the random sample, inclusion in the non-random validation sample, inclusion in the original diagnostic testing data, inclusion in ??
+on day x).}
+	\vspace{5mm}
+
+	We have no built a notation table and made the notation different.  Apologies for the confusion. We use $I_j^{R}$ and $I_j^{NR}$.
+	\vspace{5mm}
+	\item {\it In the final formula (eq 4.3), I am not clear on why the weights are applied to both yk and FP (ie why sum(w*(yk-FP)) and not sum(w*yk) – FP)?
+This is very likely just me not understanding the derivation in Appendix C, but some explanation of the intuition would be helpful.}
+	\vspace{5mm}
+
+	Since we normalizing the weights, this is just to write it in simple form but basically $\sum w y$ is the estimate and we subtract FP.  So you are right!
+	\vspace{5mm}
+	\item {\it The time-varying section seems somewhat unnecessary since it is simply taking the expectation of the data over some time period – the daily data also does this, but where the time-period is 1 day.
+}
+	\vspace{5mm}
+
+	Necessary for smoothing.
+	\vspace{5mm}
+\end{itemize}
+\item {\bf Section 4.2: model-based estimation}
+\begin{itemize}
+	\item {\it The SEIR model described in the text and in the supplement differ in a way that I find confusing – the code and equations describe a simple SEIR model but the written text describes an SEmInR model with stratum-specific covariates which appear to apply to at least the I compartments and possibly the E compartments.
+From the code and equations, presumably the assumption is that all strata mix with identical probability and all strata experience the same probabilities of infection, exposure, and transmission.
+However, these assumptions are (1) unlikely to hold and (2) make it hard to understand why the model includes strata at all.
+}
+	\vspace{5mm}
+
+	Prior is Dirichlet (uniform) but not time-varying.  This allows us to learn the links.
+	\vspace{5mm}
+\end{itemize}
+\item {\bf Section 4.3: Doubly Robust Estimation}
+\begin{itemize}
+	\item {\it The term $\mu(x_j)$ in the unnumbered equation does not seem to be defined anywhere – this is presumably coming from the SEIR model given that the rest of the equation is essentially the IPW-based estimator.
+However, it is unclear to me what output from the SEIR this term includes.}
+	\vspace{5mm}
+
+	The SEIR model is used to generate the number of new weekly cases which allows us to calculate the active infection rate.  $\mu(x_j)$ is the model-based estimate.
+	\vspace{5mm}
+\end{itemize}
+\item {\bf Section 5: Applied analysis}
+\begin{itemize}
+	\item {\it The applied example relies heavily on the idea that the random sample is a true random sample of the population of Indiana eligible to be infected with COVID, and that the non-random sample and the Facebook / CMU sample similarly provide representative measures of covariate data.
+This is not well-established by the short data descriptions given and deserves further emphasis.}
+	\vspace{5mm}
+
+	We provide further emphasis and evidence.  A key point is that this data is used to illustrate the 'what to do' point that was raised in the first round.
+	\vspace{5mm}
+\end{itemize}
+\item {\bf Section 5.1.1. Disease Prevalence}
+\begin{itemize}
+	\item {\it This section seems to assume that individuals can only be tested once but that assumption is unlikely to be true.
+In particular, people with higher exposures will get tested more frequently than people with lower exposures, and similarly people at higher risk of severe outcomes may be tested more frequently than people with lower risk.
+Calculating the sampling fraction as the (number of tests)/(total pop – previously tested) thus seems likely to provide a biased sampling fraction estimate.}
+	\vspace{5mm}
+
+	We acknowledge this limitation.  As an alternative we assume the risk set is always total pop.  This presents a range.
+	\vspace{5mm}
+	\item {\it Also, it would be useful to clarify whether the random sample, which is being used as the gold standard, was corrected for measurement error, as well as how the issue of 30\% response rate was dealt with in obtaining the prevalence estimate.
+Given these issues, the random sample estimate could easily also be biased.}
+	\vspace{5mm}
+
+	Cite PNAS for measurement error and response rate cleaning.
+	\vspace{5mm}
+\end{itemize}
+\item {\bf Section 5.1.2 Time varying estimate}
+\begin{itemize}
+	\item {\it 5.1.2.
+In the second approach, using hospitalization data to attempt to back-predict fever status seems highly likely to be subject to survivor bias as well as biases driven by differential access to care.
+Some discussion of the issues with this approach is warranted.}
+	\vspace{5mm}
+
+	We have added Remark XX to discuss the bias.
+In Remark XX, we state that availability of symptom status (which we know the state collects but has told us is unavailable due to data privacy / confidentiality issues) is important to use.
+	\vspace{5mm}
+	\item {\it Also, focusing on the non-Hispanic, white male groups seems potentially to limit understanding of the scope of error.
+Although Indiana has a very high percent of white individuals, the individuals who present for testing (as well as those who test positive, are exposed in the workplace and home, and hospitalized or die) are disproportionately non-white.
+It is important to show the performance of the method in both white and non-white individuals because non-white individuals are the ones who are experiencing higher rates of illness and death and therefore will likely be more negatively impacted by biased estimates of disease prevalence.}
+	\vspace{5mm}
+
+	We acknowledge this shortcoming.  We now include versions of Figure 6 for non-white fr
+
+\end{itemize}
 \end{enumerate}
@@
 \vspace{5mm}
-
 \end{itemize}
@@
 
+5.2 SEIR model
+As stated above, some description of the assumptions related to contact patterns and transmission between and among age strata is needed.
+In addition, please clarify which compartments of the model the infection-fatality rate applies to (E \& I or I only), as well as how the model handles death from non-COVID causes from these and other model compartments.
+Also, the discussion of IFR would benefit greatly from being contextualized to the rest of the paper – the goal of the manuscript is specifically to provide a better estimate of who is infected.
+Such an estimate would therefore be expected to provide a more valid infection fatality rate, and this in turn implies the IFR used here must be biased.
+This should be discussed.
+
 \end{document}

### 2022-02-11 write-up/letters/response-ref-1-v1.tex
Updated response to referees

@@
 	We acknowledge this shortcoming.  We now include versions of Figure 6 for non-white fr
+	\vspace{5mm}
+\end{itemize}
+\item {\bf 5.2 SEIR model}
+\begin{itemize}
+	\item {\it As stated above, some description of the assumptions related to contact patterns and transmission between and among age strata is needed.
+In addition, please clarify which compartments of the model the infection-fatality rate applies to (E \& I or I only), as well as how the model handles death from non-COVID causes from these and other model compartments.}
+	\vspace{5mm}
 
+	We now describe the model in more detail.
+	\vspace{5mm}
+	\item {\it Also, the discussion of IFR would benefit greatly from being contextualized to the rest of the paper – the goal of the manuscript is specifically to provide a better estimate of who is infected.
+Such an estimate would therefore be expected to provide a more valid infection fatality rate, and this in turn implies the IFR used here must be biased.
+This should be discussed.}
+	\vspace{5mm}
+
+    We have now acknowledged this along with the fact that IFR is time-varying.
+So this shows the IFR is likely biased (or was only accurate early).
+We performed sensitivity analysis with time-varying IFR based on estimates to improve estimation.
+	\vspace{5mm}
 \end{itemize}
@@
 
-{\bf Response to AE3}
+{\bf Response to AE2}
 \begin{itemize}
@@
 
-5.2 SEIR model
-As stated above, some description of the assumptions related to contact patterns and transmission between and among age strata is needed.
-In addition, please clarify which compartments of the model the infection-fatality rate applies to (E \& I or I only), as well as how the model handles death from non-COVID causes from these and other model compartments.
-Also, the discussion of IFR would benefit greatly from being contextualized to the rest of the paper – the goal of the manuscript is specifically to provide a better estimate of who is infected.
-Such an estimate would therefore be expected to provide a more valid infection fatality rate, and this in turn implies the IFR used here must be biased.
-This should be discussed.
-
 \end{document}

### 2022-02-11 write-up/letters/response-ref-1-v1.tex
Updated response to referees

@@
 
-{\bf Response to AE2}
+{\bf Response to Reviewer \#2}
+
+I am incredibly grateful for this review.
+I agree that the paper has evolved beyond the current title.
+This was not updated in the prior submission simply to ensure continuity of presentation (i.e., didn't want to change from first submission).
+Here we replace with the title ``Good Title''.
+We next address the reviewers specific comments.
+
 \begin{itemize}
+\item {\it The paper states some facts about the SARS-CoV-2 virus that are outdated, for example “Upon infection with SARS-CoV-2, an incubation period starts and lasts approximately five days [Lauer et al., 2020] before the viral load is high enough to be detected.” (Page 5).
+Lauer was based on the original
+strain, we see different disease dynamics with the delta strain.
+Perhaps to be more precise, the author could clarify that this is an example of the original SARS-CoV-2 incubation rather than definitative as there will likely be future variants that will change this further.
+Similarly, on page 13, the serial
+interval of 7 is stated as fact – this is again based on the original strain rather than the current disease dynamics.
+Stating these numbers as an example rather than definitive would be preferable, I believe.}
+\vspace{5mm}
+
+We have added remark XX and more carefully put this paper within proper context.
+\vspace{5mm}
+\item {\it The author states: “If someone has an active infection then after the incubation period, a molecular test with perfect sensitivity will yield a positive result for the next few weeks.” (Page 5) The author states “A molecular test refers to diagnostic tests that aim to detect increased viral load” (page 5) - I think a more precise deinition is needed, for example aim to detect viral load above a certain threshold.}
+\vspace{5mm}
+
+We use Figure XX to make precise the importance of viral threshold.
+\vspace{5mm}
+\item {\it The author states “If someone has an active infection then after the incubation period, a molecular test with perfect sensitivity will yield a positive result for the next few weeks.” (page 5) I’m not sure it is accurate that the patient will have a positive result for the next few weeks, perhaps “If someone has an active infection then after the incubation period, a molecular test with perfect sensitivity will yield a positive result while the patient has a viral load above the threshold of detection”.}
+\vspace{5mm}
+
+We thank the reviewer and now match this more careful language to be more exact in our description.
+\vspace{5mm}
+\item {\it The author lumps RT-PCR and antigen tests in together and discusses that a patients will likely have positive tests for several weeks while not infectious (able to transmit) – for the most part this is the case for RT-PCR tests, however antigen tests give a clearer picture of infectiousness as the viral threshold is higher and often correlates with the likelihood to transmit.}
+
+\vspace{5mm}
+We now make this distinction much clearer.
+\vspace{5mm}
 \item {\it XX}
@@
 \vspace{5mm}
+\item {\it I’m not sure this statement “The primary goal of molecular tests is diagnosis of active infections in the population.” (page 5) is entirely true.
+I definitely think this is a worthwhile goal, but many in the medical community use molecular tests to determine the diagnosis of an individual patient, regardless if the infection is “active” or not (i.e.
+whether they are currently infectious rather than infected which is what I assume the author means by “active”).}
+
+\vspace{5mm}
+WXX
+\vspace{5mm}
+\item {\it On page 13, the author discusses the bias relative to the “peak” – is this the peak in cases? I believe so, but it would be nice to define a bit more clearly.}
+
+\vspace{5mm}
+WXX
+\vspace{5mm}
+\item {\it We would expect the IFR to dramatically change after the introductions of vaccines in December 2020, making this method less applicable without introducing a time-varying component.}
+
+\vspace{5mm}
+WXX
+\vspace{5mm}
+\item {\it I found Figure 3 in the supplement conusing.
+Figure 3A shows the profile for a 35-44 white male with symptoms and COVID-19 contact and Figure 3B shows the profile for a 65-74 white female without symptoms nor COVID-19 contact – I would have expected the covariates between these (age, ethnicity, race, gender) to be the same rather than differ with the only thing differing being symptoms 1/ COVID-19 contact?
+As a minor note, the y-axis labels do not match for this plot.}
+
+\vspace{5mm}
+WXX
+\vspace{5mm}
+\item {\it The model based estimates in Figure 8 seem so different from the other estimates that it is potentially implausible and suggests misspecification of the IFR.
+Is this potentially due to the age distribution of cases and the IFR collapsing choices?
+Perhaps as a sensitivity analysis different IFR values should be examined}
+
+\vspace{5mm}
+WXX
+\vspace{5mm}
+\item {\bf Minor comments:}
+	\begin{enumerate}
+		\item {\it Figure 3 mention “β = 1.2 (black)” – I’m not sure what “black” refers to?}
+		\item {\it In the supplement, page 21, the author refers to “Figure 6a and 6a” – I believe they mean “Figure 6a and 6b”.
+This also claims to show the likelihood for all age ranges - I believe it is only for males.}
+	\end{enumerate}
 \end{itemize}

### 2022-02-15 visualizations/seir_example_plot.R
SEIR now in place with approximate solution

@@ -26,7 +26,8 @@ init       <- c(S = 1-1e-6, E = 1e-6, I = 0.0, R = 0.0)
 parameters1 <- c(beta = 1.2, sigma = 0.3, gamma = 0.15)
 parameters2 <- c(beta = 0.9, sigma = 0.3, gamma = 0.15)
 ## Time frame
-times      <- seq(0, 100, by = 1)
+gap = 1
+times      <- seq(0, 100, by = gap)
 
 ## Solve using ode (General Solver for Ordinary Differential Equations)
 out1 <- ode(y = init, times = times, func = sir, parms = parameters1)
@@ -80,8 +81,8 @@ bias_fn <- function(M, bary, f, FP = 0.01, FN = 0.15) {
   
   bias = (lin_term+quad_term) * cur_bary/lag_bary
   
-  Rt = 1+log(cur_bary/lag_bary)* (1/7 + 1/3)
-  logbias = log(1+lin_term+quad_term) * (1/7 + 1/3)
+  Rt = 1+log(cur_bary/lag_bary)* (1/7 + 1/3) / gap
+  logbias = log(1+lin_term+quad_term) * (1/7 + 1/3) / gap
   
   return(list("r"=cur_bary/lag_bary, "bias" = bias,
               "Rt"=Rt, "logbias"=logbias))

### 2022-02-15 write-up/letters/response-ref-1-v1.tex
SEIR now in place with approximate solution
SEIR example done. Onto the early parts ot he paper

@@
 
-	We now match the SEIR and still see the same thing for the ratio.
+	The SIR model in Section 3.2 is replaced by the SEIR model.
+We highlight that the SEIR model has been shown to better represent the SARS-CoV-2 dynamics.
+Figure XX which presents the ratio estimate and potential bias has been updated accordingly.
+The conclusions remain largely the same.
+The only difference from the prior manuscript is that the peak in bias prior to the ratio going less than 1 is less pronounced (was up to 5, now up to 2).
 \end{itemize}
@@
 
-	We clarify the subtle differences and that we are of the Cori type.
+	We clarify the subtle differences between the definitions of the effective reproductive number and establish the approach is in line with the type described by Cor et. al (2013).
 	\vspace{5mm}
@@
 
-	This estimator is motivated by the SIR but is used more generally in COVID plots (see \href{here}{https://academic.oup.com/aje/article/189/11/1438/5864525?searchresult=1} for use in plots).
-This is closely related to Cori paper \href{as reviewer identified}{https://academic.oup.com/aje/article/178/9/1505/89262?searchresult=1}.
-See \href{here}{https://www.nature.com/articles/s41598-020-76563-8.pdf?origin=ppub} where we have the basic reproduction number for SEIR and SIR.
-We used this to alter the proposed moment-based estimator to be more appropriate for SEIR in two ways.
-First, the constant is now $(1/\gamma + 1/\sigma)$.
-Second, we acknowledge the estimator $K_t/K_{t-1}$ may have issues of smoothness.
-We can smooth these by considering aggregate differences.
-We provide estimates of ratio and eff repro number in the appendix for 3- and 7- day window lengths.
-We see a similar story.
+	%https://academic.oup.com/aje/article/189/11/1438/5864525?searchresult=1, plot SIR
+	%https://academic.oup.com/aje/article/178/9/1505/89262?searchresult=1, CORI
+	%https://www.nature.com/articles/s41598-020-76563-8.pdf?origin=ppub, Althaus approximate
+
+	This estimator is originally motivated by the SIR model and is exact under the poisson likelihood; however, the moment-based estimator has been proposed as a general tool for assessing epidemic trends in real time (\href{https://academic.oup.com/aje/article/189/11/1438/5864525?searchresult=1}{cite}).
+The estimator is closely related to the instantaneous reproduction numbers proposed by Cori et.
+al.
+(20) \href{https://academic.oup.com/aje/article/178/9/1505/89262?searchresult=1}{cite}.
+We now rely on recent work \href{https://www.nature.com/articles/s41598-020-76563-8.pdf?origin=ppub}{(cite)} to propose an approximate moment-based estimator in the SEIR setting.
+	\vspace{5mm}
+
+	\noindent Heng and Althaus (2020) derive the following approximate formula for the basic reproduction number under SEIR dynamics:
+	$$
+	R_0 = \frac{1 + \Lambda \left(\frac{1}{\sigma} + \frac{1}{\gamma} \right)}{S_0 + I_0 (1 + \gamma/\sigma)}.
+	$$
+	where~$\Lambda$ is the epidemic growth rate, and $S_0$/$I_0$ are the initial fractions of the population that are susceptible and infectious respectively.
+For $S_0 \approx 1$ and $I_0 \ll 1$, we have the approximate realtionship $R_0 \approx 1 + (\frac{1}{\gamma} + \frac{1}{\sigma})\Lambda$.
+	, this is an approximate solution which lacks the term $\Lambda^2 \frac{1}{\gamma \sigma}$.
+``It corresponds to the case of an exponentially distributed generation time with mean duration $1/\gamma + 1/\sigma$, which is the same as the solution for the SIR model assuming an infectious period of $1/\gamma + 1/\sigma$.
+For realistic growth rates and choices of $\sigma$ and $\gamma$ for the COVID-19 pandemic, this term is negligible and therefore we work with the approximate formula.
+	\vspace{5mm}
+
+	\noindent Using the connection to the SIR model with infectious period $1/\gamma + 1/\sigma$, and replacing the true epidemic growth rate~$\Lambda$ by the estimated growth rate~$\hat \Lambda = \log \left( K_t / K_{t-1} \right)$ yields our new moment-based estimator.
+The updated $R_t$ estimator shows that the same intuition about the error holds when we are considering an SEIR rather than an SIR model.
+Figure XX shows the reproduction number estimates and potential biases under SEIR dynamics which shows the same general conclusions hold.
 \end{itemize}
@@
 
-	You got it dude.
+	This section has been also updated to consider an SEIR model.  All conclusions remain the same.
 	\vspace{5mm}
@@
 	\begin{enumerate}
-		\item {\it Figure 3 mention “β = 1.2 (black)” – I’m not sure what “black” refers to?}
+		\item {\it Figure 3 mention “$\beta = 1.2$ (black)” – I’m not sure what “black” refers to?}
+		\vspace{5mm}
+
+		We have removed this and simply state the SEIR parameter values ($\beta = 1.2, \gamma = 0.14, \sigma = 0.3$) underlying the Figure.
+
 		\item {\it In the supplement, page 21, the author refers to “Figure 6a and 6a” – I believe they mean “Figure 6a and 6b”.

### 2022-02-15 write-up/covid-writeup.tex
SEIR example done. Onto the early parts ot he paper

@@
 
- The prior analysis demonstrates the limited information regarding COVID-19 prevalence in observational case-count data.
-Scientists, however, may claim that daily observed case-counts simply undercount daily total cases by a constant multiple over time.
+ The prior analysis demonstrates the potentially limited information regarding COVID-19 prevalence in observational case-count data.
+Scientists, however, may claim that daily observed case-counts simply undercount daily total cases by a constant multiple over time (i.e., undercounting).
 If true then the ratio of case-counts at consecutive times may be a good estimate of the true change in prevalence, helping scientists understand the disease trajectory.
@@
 
- Figure~\ref{fig:ratio-bias} displays the trajectory of the true ratio and the potential biased estimators under a susceptible-infected-recovered (SIR) model~\citep{Pastor2001,Newman2002,Parshani2010} for the epidemic dynamics, with state evolution given by
- \begin{equation}
- \label{eq:sir}
- \frac{\partial s_t}{\partial t} = - \beta s_t i_t; \quad
- \frac{\partial i_t}{\partial t} = \beta s_t i_t - \gamma i_t; \quad
- \frac{\partial r_t}{\partial t} = - \gamma i_t
- \end{equation}
- where $s_t, i_t$ and $r_t$ are the fraction of susceptible, infected, and removed (recovered or deceased) individuals in the population at time $t$ respectively.
+ Figure~\ref{fig:ratio-bias} displays the trajectory of the true ratio and the potential biased estimators under a susceptible-exposed-infected-recovered (SEIR) model~\citep{Pastor2001,Newman2002,Parshani2010} for the epidemic dynamics, with state evolution given by
+ \begin{align}
+ \frac{\partial s_t}{\partial t} &= - \beta s_t i_t; \quad \frac{\partial e_t}{\partial t} = \beta s_t i_t - \sigma e_t; \quad \label{eq:seir} \\
+ \frac{\partial i_t}{\partial t} &= \sigma e_t - \gamma i_t; \quad
+ \frac{\partial r_t}{\partial t} = - \gamma i_t. \nonumber
+\end{align}
+ % \begin{equation}
+ % \label{eq:sir}
+ % \frac{\partial s_t}{\partial t} = - \beta s_t i_t; \quad
+ % \frac{\partial i_t}{\partial t} = \beta s_t i_t - \gamma i_t; \quad
+ % \frac{\partial r_t}{\partial t} = - \gamma i_t
+ % \end{equation}
+ where $s_t, e_t, i_t$ and $r_t$ are the fraction of susceptible, exposed infected, and removed (recovered or deceased) individuals in the population at time $t$ respectively.
 In terms of bias, the rate is overestimated prior to the peak and underestimated afterwards; the bias increases dramatically when the relative fraction exceeds $2$.
@@
 In particular, under an SIR model the number of case counts on day $t$, denoted $K_t$, is Poisson distributed with rate $K_{t-1} \exp \left( \gamma (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval, which is approximately $7$ days for COVID-19~\citep{Sanche2020}.
-Using this relation, a moment-based estimator is given by
- $$
- R_t = 1 + \frac{1}{\gamma} \log \left( \frac{K_t}{K_{t-1}} \right).
- $$
+
+\cite{Heng2020} derive an approximate formula for the basic reproduction number under SEIR dynamics such that for $S_0 \approx 1$ and $I_0 \ll 1$, we have the approximate realtionship, which ``corresponds to [...] the solution for the SIR model assuming an infectious period of $1/\gamma + 1/\sigma$.
+Using this connection, a moment-based estimator is given by
+$$
+R_t \approx 1 + \left( \frac{1}{\gamma} + \frac{1}{\sigma} \right) \log \left( \frac{K_t}{K_{t-1}} \right).
+$$
+This approximation ignores a term that is quadratic in the epidemic growth rate and inversely linear in $\sigma \times \gamma$; for realistic values of these quantities, however, the term is relatively negligible and therefore ignored.
 Of course, we do not observe $K_t$ and $K_{t-1}$.
@@
  \begin{aligned}
- \frac{1}{\gamma}\log &\bigg( 1 + \bigg[ \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t)  -\rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (K_{t-1}) \bigg] \\
- &\times \left[ 1 - \rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (K_{t-1}) \right] \bigg) - \frac{1}{\gamma} \log \left( \frac{S_t}{S_{t-1}} \right).
+ \left( \frac{1}{\gamma} + \frac{1}{\sigma} \right)\log &\bigg( 1 + \bigg[ \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t)  -\rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (K_{t-1}) \bigg] \\
+ &\times \left[ 1 - \rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (K_{t-1}) \right] \bigg) - \left( \frac{1}{\gamma} + \frac{1}{\sigma} \right) \log \left( \frac{S_t}{S_{t-1}} \right).
  \end{aligned}
@@
 This leads to differences in when the bias is most pronounced.
-Figure~\ref{fig:r0-bias} displays the bias as a function of the relative sampling fraction assuming the population fraction that is susceptible remains large (i.e., $S_t / S_{t-1} \approx 1$).
+Figure~\ref{fig:r0-bias} displays the bias as a function of the relative sampling fraction ignoring the final term (i.e., $S_t / S_{t-1} \approx 1$).
 
@@
   \centering
-  \includegraphics[width=.9\linewidth]{../figs/sir_ratio.png}
+  \includegraphics[width=.9\linewidth]{../figs/seir_ratio.png}
   \caption{Ratio estimator}
@@
   \centering
-  \includegraphics[width=.9\linewidth]{../figs/sir_rt.png}
+  \includegraphics[width=.9\linewidth]{../figs/seir_rt.png}
   \caption{Effective reproductive rate estimator}
@@
  \end{subfigure}
- \caption{Potential bias in ratio and effective reproductive rate estimators under an SIR model with $\beta = 1.2$ (black) and $\gamma = 0.15$.
+ \caption{Potential bias in ratio and effective reproductive rate estimators under an SEIR model with $\beta = 1.2$, $\gamma = 0.15$, and $\sigma = 0.3$.
 Here, $f = 0.02$, $FP = 0.024$, $FN = 0.13$, and a range of relative sampling fractions $M = f_1/f_0$ are considered.}
@@
  Consider two countries (A and B) in which the peak occurs 2 weeks prior for country A than country B.
-Figure~\ref{fig:comparison} presents such a comparison where each country's disease trajectory follows an SIR model (A=black and B=grey).
+Figure~\ref{fig:comparison} presents such a comparison where each country's disease trajectory follows an SEIR model (A=black and B=grey).
 Figure~\ref{fig:eff} shows how biases interact in complex ways.
@@
   \centering
-  \includegraphics[width=.9\linewidth]{../figs/sir.png}
+  \includegraphics[width=.9\linewidth]{../figs/seir.png}
   \caption{Fraction of new cases in population}
@@
   \centering
-  \includegraphics[width=.9\linewidth]{../figs/sir_rt_comparison.png}
+  \includegraphics[width=.9\linewidth]{../figs/seir_rt_comparison.png}
   \caption{Effective reproduction rate estimators}
@@
  \end{subfigure}
- \caption{Left: fraction infected in two SIR models with $\beta = 1.2$ and $0.9$ respectively and $\gamma = 0.15$ with same initial conditions.
+ \caption{Left: fraction infected in two SEIR models with $\beta = 1.2$ and $0.9$ respectively, $\sigma = 0.3$, and $\gamma = 0.15$ with same initial conditions.
 Right: comparison of $\hat R_t$ across time with $FN = 0.30$, $FP = 0.024$, and $M = 4$.}
@@
  Let $s_t$, $e_t$, $i_t$, and $r_t$ denote the proportion of survivors, exposed, infected, and removed cases (i.e., including both recovered cases and deaths) at time $t$.
-The population-level SEIR dynamics are then given by the following set of differential equations:
-\begin{align}
- \frac{\partial s_t}{\partial t} &= - \beta s_t i_t; \quad \frac{\partial e_t}{\partial t} = \beta s_t i_t - \sigma e_t; \quad \label{eq:seir} \\
- \frac{\partial i_t}{\partial t} &= \sigma e_t - \gamma i_t; \quad
- \frac{\partial r_t}{\partial t} = - \gamma i_t. \nonumber
-\end{align}
+The population-level SEIR dynamics are given by the set of differential equations in~\eqref{eq:seir}.
+% \begin{align}
+%  \frac{\partial s_t}{\partial t} &= - \beta s_t i_t; \quad \frac{\partial e_t}{\partial t} = \beta s_t i_t - \sigma e_t; \quad \label{eq:seir} \\
+%  \frac{\partial i_t}{\partial t} &= \sigma e_t - \gamma i_t; \quad
+%  \frac{\partial r_t}{\partial t} = - \gamma i_t. \nonumber
+% \end{align}
 %Following~\cite{Song2020}, we employ Runge-Kutta (RK4) approximations for discretization.

### 2022-02-16 write-up/letters/response-ref-1-v1.tex
finalizing section 3.  then back to the early parts of paper for setup fixes

@@
 We highlight that the SEIR model has been shown to better represent the SARS-CoV-2 dynamics.
-Figure XX which presents the ratio estimate and potential bias has been updated accordingly.
-The conclusions remain largely the same.
+Figure 3a presents updated versions of the potnetial bias in ratio estimators under an SEIR model.
+The conclusion remains largely the same.
 The only difference from the prior manuscript is that the peak in bias prior to the ratio going less than 1 is less pronounced (was up to 5, now up to 2).
+	\vspace{5mm}
 \end{itemize}
@@
 
-	We clarify the subtle differences between the definitions of the effective reproductive number and establish the approach is in line with the type described by Cor et. al (2013).
+	We clarify the subtle differences between the definitions of the instantaneous and case reproductive numbers.
+We establish that our approach aims to estimate the instantaneous reproduction number, i.e., is closely related to the proposal by Cori et al.
+(2013)).
+
 	\vspace{5mm}
@@
 
-	This estimator is originally motivated by the SIR model and is exact under the poisson likelihood; however, the moment-based estimator has been proposed as a general tool for assessing epidemic trends in real time (\href{https://academic.oup.com/aje/article/189/11/1438/5864525?searchresult=1}{cite}).
+	This estimator is originally motivated by the SIR model and is exact under the poisson likelihood; however, the moment-based estimator has been proposed as a general tool for assessing epidemic trends in real time (Xinhua Yu, Assessing Epidemic Trends in Real Time With a Simple Ratio Plot, Am J of Epid, 189(11), 2020).
 The estimator is closely related to the instantaneous reproduction numbers proposed by Cori et.
 al.
-(20) \href{https://academic.oup.com/aje/article/178/9/1505/89262?searchresult=1}{cite}.
-We now rely on recent work \href{https://www.nature.com/articles/s41598-020-76563-8.pdf?origin=ppub}{(cite)} to propose an approximate moment-based estimator in the SEIR setting.
-	\vspace{5mm}
-
-	\noindent Heng and Althaus (2020) derive the following approximate formula for the basic reproduction number under SEIR dynamics:
+(2013), but we now rely on recent work (Heng and Althaus, The approximately universal shapes of epidemic curves in the SEIR model, Sci Reports, 2020) to propose an approximate moment-based estimator in the SEIR setting.
+Specifically, Heng and Althaus (2020) derive the following approximate formula for the basic reproduction number under SEIR dynamics:
 	$$
@@
 	where~$\Lambda$ is the epidemic growth rate, and $S_0$/$I_0$ are the initial fractions of the population that are susceptible and infectious respectively.
-For $S_0 \approx 1$ and $I_0 \ll 1$, we have the approximate realtionship $R_0 \approx 1 + (\frac{1}{\gamma} + \frac{1}{\sigma})\Lambda$.
+For $S_0 \approx 1$ and $I_0 \ll 1$, we have the approximate relationship $R_0 \approx 1 + (\frac{1}{\gamma} + \frac{1}{\sigma})\Lambda$.
 	, this is an approximate solution which lacks the term $\Lambda^2 \frac{1}{\gamma \sigma}$.
@@
 
-	\noindent Using the connection to the SIR model with infectious period $1/\gamma + 1/\sigma$, and replacing the true epidemic growth rate~$\Lambda$ by the estimated growth rate~$\hat \Lambda = \log \left( K_t / K_{t-1} \right)$ yields our new moment-based estimator.
+	\noindent Note that for the SIR model, the basic reproduction number is~$R_0 = 1 + \frac{1}{\gamma} \Lambda$.
+So estimate of the instantaneous reproduction number is obatined by replacing $\Lambda$ by an estimate of the  instantaneous growth rate $\hat \Lambda_t = \log (K_t / K_{t-1})$.
+Then using the approximate relationship between the SEIR model and an SIR model with infectious period $1/\gamma + 1/\sigma$, we arrive at the new moment-based estimator
+	$$
+    R_t \approx 1 + \left( \frac{1}{\gamma} + \frac{1}{\sigma} \right) \log \left( \frac{K_t}{K_{t-1}} \right).
+	$$
 	The updated $R_t$ estimator shows that the same intuition about the error holds when we are considering an SEIR rather than an SIR model.
-Figure XX shows the reproduction number estimates and potential biases under SEIR dynamics which shows the same general conclusions hold.
+Figure 3b shows the reproduction number estimates and potential biases under SEIR dynamics which shows the same general conclusions hold.
 \end{itemize}

### 2022-02-16 write-up/covid-writeup.tex
finalizing section 3.  then back to the early parts of paper for setup fixes

@@
  \label{section:r0-estimation}
- Many epidemiologists argue that tracking the effective reproduction number $R_t$ is the only way to manage through the crisis~\citep{Gabriel2020}.
-Under a Poisson likelihood, a simple relation between the trajectory of new cases and the effective reproduction number can be derived \citep{Bettencourt2008}.
+ Many epidemiologists argue that tracking the effective reproduction number is the only way to manage through the crisis~\citep{Gabriel2020}.
+Here, we study the instantaneous reproduction number~\citep{Cori20113,Fraser2007}, denoted $R_t$, which is the average number of secondary cases that each infected individual would infect if the conditions remained as they were at time~$t$.
+This is distinct from the case reproduction number,~$R_t^c$, which is the average number of secondary cases that a case infected at time step $t$ will eventually infect~\citep{Wallinga2004}.
+The case reproduction number accounts for potential changes to contact rates and transmissibility, which include impact of control measures.
+The instantaneous reproduction number~$R_t$ is the only reproduction number easily estimated in real time, and therefore has been a key focal point in the COVID-19 pandemic.
+
+Under a Poisson likelihood, a simple relation between the trajectory of new cases and the instantaneous reproduction number can be derived \citep{Bettencourt2008}.
 In particular, under an SIR model the number of case counts on day $t$, denoted $K_t$, is Poisson distributed with rate $K_{t-1} \exp \left( \gamma (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval, which is approximately $7$ days for COVID-19~\citep{Sanche2020}.

### 2022-02-18 visualizations/seir_example_plot.R
Update seir_example_plot.R

@@ -79,17 +79,6 @@ bias_fn <- function(M, bary, f, FP = 0.01, FN = 0.15) {
   cur_bary = bary[2:length(error)]
   lag_bary = bary[1:(length(error)-1)]
   
-  cori = vector(length = length(bary)-1)
-  for(i in 2:length(bary)) {
-    temp_cur_bary = bary[i]
-    prior_bary = bary[1:(i-1)]
-    prior_times = 1:(i-1)
-    weights = dgamma((i-1):1, shape = 4)
-    weights = weights/sum(weights)
-    denominator = sum(prior_bary*weights)
-    cori[i-1] = temp_cur_bary/denominator
-  }
-  plot(cori)
   bias = (lin_term+quad_term) * cur_bary/lag_bary
   
   Rt = 1+log(cur_bary/lag_bary)* (1/7 + 1/3) / gap
@@ -168,4 +157,75 @@ axis(side =2, cex.axis = 1)
 mtext(expression(R[t]), side =2, line = 2, cex = 1.5)
 mtext("Time", side = 1, line = 1, cex = 1.5)
 legend(55, 1.3, legend = c("M=1", "M=2", "M=3", "M=4"), lty = c(1,2,2,3), col = c("black", "black", "grey50", "grey50"), cex = 1.5, bty = "n", lwd = c(1,1,2,2))
+dev.off()
+
+
+#### CORI BIAS
+library(lubridate)
+library(EpiEstim)
+## Time frame
+gap = 1/7
+times <- seq(0, 100, by = gap)
+
+## Solve using ode (General Solver for Ordinary Differential Equations)
+cori_out1 <- ode(y = init, times = times, func = sir, parms = parameters1)
+cori_out1 <- as.data.frame(cori_out1)
+cori_bary <- cori_out1$I[-1]
+dates = seq(ymd('2009-04-07'),ymd('2009-04-07')+length(cori_bary)-1, by = '1 day')
+
+cori_bias_fn <- function(M, bary, f, FP = 0.01, FN = 0.15) {
+  const = (bary*M + (1-bary))
+  f0 = f/const
+  f1 = M*f0
+  Delta = f1 - f0
+  D_M1 = 1 - Delta * bary/(1-bary) * (FP*(1-bary) + FN * bary)/ (f0 * (1-bary) + f1 * bary)
+  D_M2 = 1/(1-FP-FN)
+  D_M = D_M1 * D_M2
[…]

### 2022-02-18 write-up/covid-supplement-writeup.tex
Update covid-supplement-writeup.tex

@@
 
-\newpage
-
 \subsection{Model 2 (Hospitalization)}
@@
  \centering
- \includegraphics[width=.9\linewidth]{../figs/tv_air_cis.png}
+ \includegraphics[width=.6\linewidth]{../figs/tv_air_cis.png}
  \caption{IPW2 estimate with confidence intervals}
@@
  \centering
- \includegraphics[width=.7\linewidth]{../figs/undercounting.png}
+ \includegraphics[width=.6\linewidth]{../figs/undercounting.png}
  \caption{Undercount based on SEIR model}
  \label{fig:undercounting}
+\end{figure}
+
+\section{Alternative estimator }
+\label{app:cori_rt}
+
+\begin{figure}
+ \centering
+  \includegraphics[width=.9\linewidth]{../figs/seir_cori_rt.png}
+  \caption{Effective reproductive rate estimator}
+  \label{fig:r-bias}
+ \caption{Potential bias in instantaneous reproductive rate estimator based on~\cite{Cori20113} under an SEIR model with $\beta = 1.2$, $\gamma = 0.15$, and $\sigma = 0.3$.
+Here, $f = 0.02$, $FP = 0.024$, $FN = 0.13$, and a range of relative sampling fractions $M = f_1/f_0$ are considered.}
+ \label{fig:rates}
  \end{figure}

### 2022-02-18 write-up/covid-supplement-writeup.tex
Eff R_t now all set

@@
 
-\section{Alternative estimator }
+\section{Alternative estimator of the instantaneous reproductive number}
 \label{app:cori_rt}
+
+Here we present the estimate of~\cite{Cori20113}.
+Let~$I_t$ detnoe the total infectiousness of infected individuals at time~$t$.
+Then~$E[I_t] = R_t \sum_{s=1}^{t} I_{t-s} w_s$ where $w_{u}$ is the infectivity function.
+Here, we follow~\cite{Cori20113} and choose a discretized shifted Gamma distribution with mean $7$ days and standard deviation $2$ days.
+Then a moment-based estimator can be obtained by $\hat R_t = I_t / \sum_{s=1}^t I_{t-s} w_s$, which takes the form of the ratio estimators in Section~\ref{section:rates}.
+Therefore, the bias can be readily obtained from the associated Taylor series decomposition, with the terms related to time~$t-1$ replaced by a weighted version.
 
@@
   \caption{Effective reproductive rate estimator}
-  \label{fig:r-bias}
  \caption{Potential bias in instantaneous reproductive rate estimator based on~\cite{Cori20113} under an SEIR model with $\beta = 1.2$, $\gamma = 0.15$, and $\sigma = 0.3$.
 Here, $f = 0.02$, $FP = 0.024$, $FN = 0.13$, and a range of relative sampling fractions $M = f_1/f_0$ are considered.}
- \label{fig:rates}
+ \label{fig:cori-bias}
  \end{figure}
+
+Figure~\ref{fig:cori-bias} presents the potential bias in instantaneous reproductive rate estimator.
+The general conclusions follow similarly as in  Figure~\ref{fig:ratio-bias}.
+The rate is overestimated prior to the peak and underestimated afterwards.
+Estimates at the peak time appear to have minimal bias.
 

### 2022-02-18 write-up/covid-writeup.tex
Update covid-writeup.tex
Eff R_t now all set

@@
  where $s_t, e_t, i_t$ and $r_t$ are the fraction of susceptible, exposed infected, and removed (recovered or deceased) individuals in the population at time $t$ respectively.
+The SEIR model has been used extensively as a model for SARS-CoV-2 dynamics~\cite{Song2020}.
 In terms of bias, the rate is overestimated prior to the peak and underestimated afterwards; the bias increases dramatically when the relative fraction exceeds $2$.
-Such biases may have a substantial impact on policy making.
+Such biases may impact policy making.
 Overestimation pre-peak may give policy makers more leverage in proposing aggressive actions to reduce prevalence.
@@
 Figure~\ref{fig:r0-bias} displays the bias as a function of the relative sampling fraction ignoring the final term (i.e., $S_t / S_{t-1} \approx 1$).
+Section~\ref{app:cori_rt} in the supplementary materials presents an alternative estimator of the instantaneous effective reproductive number and shows similar bias~\cite{Cori20113}.
 

### 2022-02-21 write-up/covid-writeup.tex
Updated paper - but needs more!

@@
 
+\begin{remark}[An evolving pandemic]
+
+\end{remark}
+
  % \subsection{Outline}
@@
 
- Here we provide the necessary background to understand COVID-19 diagnostic testing, its scientific use in managing the pandemic, and the data streams considered in this manuscript.
+Here we provide the necessary background to understand COVID-19 diagnostic and serological testing, its scientific use in managing the pandemic, and the data streams considered in this manuscript.
+
+\subsection{Purposes of testing}
 
@@
 Rapid and frequent molecular testing is cited as a key component~\citep{OECD2021} in effective strategies to identify active infections and prevent systemic outbreaks.
+
+\emph{Serological tests} look for an immunological response to the virus.
+A week or so after an individual is infected with SARS-CoV-2, the individual will start producing antibodies.
+At this point, a serological test with perfect sensitivity will come back positive.
+This test provides evidence of past infection while the molecular test provides evidence of an \emph{active} infection.
+
+To better estimate SARS-CoV-2 immunity in a population, seroprevalence studies that generate a probabilistic population sample and perform serological tests on the sample can be collected.
+These studies are useful for disease surveillance.
+To date, the CDC has conducted ten large-scale geographic serological surveys with three rounds.
+While population-based sampling strategies provide a more representative nationwide sample, they are very time intensive and expensive.
 

### 2022-02-21 write-up/letters/response-ref-1-v1.tex
Eff R_t now all set
Intro updated

@@
 {\bf Response to Reviewer 1}
+\vspace{5mm}
+
 I am incredibly grateful for this review. (PRAISE OUT THE WAZZOOO).
@@
 \vspace{5mm}
-Section XX now discusses the purposes of testing.
+In the introduction, we now connect random testing to its purpose as a screening tool and non-random testing which generates the case count data as a diagnostic tool.
+Given the infrequent nature of random testing and desire for ongoing prevalence estimates, we motivate the second objective of our paper by the question of whether randomized screening tests can be used to address selection bias in the non-random diagnostic testing to estimate prevalence or incidence over time.
 \vspace{5mm}
@@
 Figure 3b shows the reproduction number estimates and potential biases under SEIR dynamics which shows the same general conclusions hold.
+For completeness, we also present the instantaneous reproduction number estimates using the proposal of Cori et.
+al.
+(2014) in Section I of the supplementary materials, which shows the same general conclusions.
 \end{itemize}

### 2022-02-21 write-up/covid-writeup.tex
Intro updated

@@
 
+Selection bias in case-count data is primarily due to it being a \emph{diagnostic tool}, i.e., individuals who are symptomatic or have a known/suspected exposure are more likely to present for diagnostic testing.
+Case-count data arising from non-random testing means it cannot provide valid prevalence or incidence estimates due to the significant proportion of asymptomatic and pauci-symptomatic cases.
+Random testing, on the other hand, is used for \emph{screening purposes}, i.e., is an appropriate tool for reconstructing prevalence/incidence estimates.
 Due to monetary and time constraints, however, random testing is performed infrequently.
@@
 Therefore, while random testing may be preferable in theory, in practice governments, researchers, and policy makers continue to use coronavirus case-counts to understand the impact of COVID-19 on the population and make data-informed decisions.
+
+% Given the infrequent nature of random testing, however, a question is whether randomized screening tests can be used to address selection bias in the non-random diagnostic testing.
 
@@
 
-Here we provide the necessary background to understand COVID-19 diagnostic and serological testing, its scientific use in managing the pandemic, and the data streams considered in this manuscript.
-
-\subsection{Purposes of testing}
+Here we provide the necessary background to understand COVID-19 diagnostic testing, its scientific use in managing the pandemic, and the data streams considered in this manuscript.
 

### 2022-02-25 write-up/letters/response-ref-1-v1.tex
Update response-ref-1-v1.tex
Section 1.1 cleaning up

@@
 	\vspace{5mm}
-	We now connect with XX
+
+	We now connect with this literature on potential sources of biases in Section 1.1.
+We highlight that our work adds a new perspective on this critical issue, from the error decomposition perspective of Meng et al.
+(2018).
 	\vspace{5mm}
+
 	\item {\it For the second goal of the paper, it would be useful to discuss how the proposed method differs from the more traditional epidemiologic method of obtaining a validation sample.
@@
 	\vspace{5mm}
-	Remark within validation subsection to contrast with validation sample.
+
+	Remark within validation subsection to contrast with validation sample. We state that this is a fairly novel
 	\vspace{5mm}

### 2022-02-25 write-up/covid-writeup.tex
Section 1.1 cleaning up

@@
  This article discusses the relationship between three statistical concepts: selection bias, measurement error, and population size.
-While the impact of measurement error~\citep{Smeden2019} and selection bias~\citep{Keiding2016} on estimation are both well-studied topics, here we provide a new perspective by building on the work of \cite{Meng2018} who studied an error decomposition to understand the relationship between selection bias and population size.
+Potential biases in observational studies of COVID-19 have been identified elsewhere in the literature~\cite{Kahn2021,Accorsi2020}.
+While the impact of measurement error~\citep{Smeden2019} and selection bias~\citep{Keiding2016} on estimation are both well-studied topics in general, here we provide a new perspective by building on the work of \cite{Meng2018} who studied an error decomposition to understand the relationship between selection bias and population size.
 Specifically, we quantify the interaction between measurement-error and selection bias on statistical error, showing how the sign and magnitude can change drastically.
- % the impact on the effective sample size can therefore be quite large.
+We then discuss this relationship in the context of observational COVID-19 case-count data, showing the impact on the effective sample size can be quite large.
 

### 2022-02-26 write-up/letters/response-ref-1-v1.tex
COVID writeup improving!

@@
 
-	Remark within validation subsection to contrast with validation sample. We state that this is a fairly novel
+	We now clarify this point in Section 1.1.
+We make it clear that our goal is to use a probability sample to correct for selection bias.
+We connect this with validation samples to build auxiliary data to address selection bias.
+
 	\vspace{5mm}
@@
 	\vspace{5mm}
-	Validation subsection
+
+	We now do this at the end of Section 1.1 in order to help statisticians thinking about validity connect our arguments to this literature.
 	\vspace{5mm}
@@
 
-	We now connect with XX
+	We now clarify this distinction between incubation and latent period.
 	\vspace{5mm}
@@
 
-	We clarify that here we are interested in all individuals who .
+	We clarify that we define active infection as individuals infected with SARS-CoV-2 who have yet to fall below a threshold for detectable viral load levels by RT-PCR testing.
 
@@
 
-	We clarify the means. These were individuals who report to the exact same set of locations as in the non-random sample.  So they are representative.
-	Note, we do not use the outcome data in the method for validation, only for adjustment for self-selection in the diagnostic tests.
+	We clarify that, to the best of our knowledge, both Indiana Department of Health's (IDOH) molecular testing and the random state-wide sample relied primarily on nasopharyngeal swab using the same set of RT-PCR tests.
+This was primarily due to the random sampling being done by a scientific team working closely with the IDOH.
 	\vspace{5mm}
 \end{itemize}
-\item {\bf Section 2.1}
+\item {\bf Section 2.2: Publicly available data}
 \begin{itemize}

### 2022-02-26 write-up/covid-writeup.tex
COVID writeup improving!

@@
 This article discusses the relationship between three statistical concepts: selection bias, measurement error, and population size.
-Potential biases in observational studies of COVID-19 have been identified elsewhere in the literature~\cite{Kahn2021,Accorsi2020}.
+Potential biases in observational studies of COVID-19 have been identified elsewhere in the literature~\citep{Kahn2021,Accorsi2020}.
 While the impact of measurement error~\citep{Smeden2019} and selection bias~\citep{Keiding2016} on estimation are both well-studied topics in general, here we provide a new perspective by building on the work of \cite{Meng2018} who studied an error decomposition to understand the relationship between selection bias and population size.
@@
 We then provide an extension of the statistical error decomposition and discuss the trade-offs inherent in such a weighting approach.
+The proposed approach is distinct from the use of validation studies~\citep{Fox2020} -- a traditional epidemiologic method in which investigators compare the accuracy with gold standard measure to mitigate bias.
+Here, the \emph{gold standard} of a different test is replaced by a less biased selection mechanism.
 
@@
 In this paper, rather than focusing on construction of a study sample, we ask whether one can collect auxiliary information to address selection bias in the observed case count data directly.
+This question is related recent work on \emph{target validity}~\citep{Westreich2018}, in which the issues of internal and external validity are jointly addressed  with resepct to a specific population of interest.
+This article is a concrete attempt to address both types of validity and extends the conversation on target validity within the context of analysis of observational studies of COVID-19.
 
@@
 
- Upon infection with SARS-CoV-2, an incubation period starts and lasts approximately five days~\citep{Lauer2020} before the viral load is high enough to be detected.
-A \emph{molecular test} refers to diagnostic tests that aim to detect increased viral load (e.g., RT-PCR or antigen tests).
-If someone has an active infection then after the incubation period, a molecular test with perfect sensitivity will yield a positive result for the next few weeks.
+ Upon infection with SARS-CoV-2, an incubation period (time to symptom onset) starts and lasts approximately five days~\citep{Lauer2020}.
+The viral load will be detectable by at least the end of the latent period (time to infectiousness), which for SARS-CoV-2 occurs before the end of the incubation period.
+A \emph{molecular test} refers to diagnostic tests that aim to detect viral load above a certain threshold (e.g., RT-PCR or antigen tests).
+If someone has an active infection -- here defined as individuals infected with SARS-CoV-2 who's viral load has yet to fall below detectable levels by molecular test -- then after the incubation period, a molecular test with perfect sensitivity will yield a positive result for the next few weeks.
 After that, the viral load will decrease and a molecular test with perfect specificity will come back negative.
@@
 Systematic reviews suggest that 87\% sensitivity and 97.6\% specificity are reasonable estimates for current RT-PCR tests~\citep{Arevalo2020, Woloshin2020,Cohen2020}.
+\textcolor{red}{To the best of our knowledge, both Indiana Department of Health's molecular testing and the random state-wide sample relied primarily on nasopharyngeal swab using the same set of RT-PCR tests.}
 
@@
 These studies are useful for disease surveillance.
-To date, the CDC has conducted ten large-scale geographic serological surveys with three rounds.
+By June 2021, the CDC has conducted ten large-scale geographic serological surveys with three rounds.
 While population-based sampling strategies provide a more representative nationwide sample, they are very time intensive and expensive.
@@
 Random samples over time to supplement this data with auxiliary information is desperately needed.
-We acknowledge that this proposal will lead to data privacy concenrs that will need to be addressed.
+We acknowledge that this proposal will lead to data privacy concerns that will need to be addressed.
 
@@
 Such data collection and dissemination is critical to address selection bias.
+
+ \textcolor{red}{I want to say that if i have many more random samples, then we should bias towards the random samples and not only smooth the means. But this is future work.}
 

### 2022-03-09 write-up/letters/response-ref-1-v1.tex
table1 construction

@@
 
-	Clarify that Indiana released a lot of aggregate information.
-Two distinct goals: (1) what can we say with additional surveys but not additional testing, (2) guide health departments who have access to
+	We have added Remark 3 in Section 2.2 discussing public versus government data.
+Specifically, we clarify that data privacy restrictions enforce publicly released testing and case count data be aggregated by strata-level covariates and have potentially important missing information (i.e., symptom status and )
+	While these choices impact our data analytic plan, we clarify that the data analytic framework is expressly designed to guide health departments and policy makers to assess the pandemic and guide future resposnes.
+
 	\vspace{5mm}
@@
 
-	We have a timeline
+	Section 2.2.1 has a reconstructed timeline of restrictions, which was based on reported restrictions in the local Indiana news.
+Between May 12th and April 28th, only high-risk individuals (symptomatic or having one of several important risk factors) were able to test.
+After April 28th, suspected exposure was included as a reason for testing.
+We could not find any article mention travel history as a factor for testing in the public reports.
 	\vspace{5mm}

### 2022-03-09 write-up/covid-writeup.tex
Update covid-writeup.tex
table1 construction

@@
 To minimize the impact of reporting, testing and case-count data are aggregated at the weekly level for analysis.
+ \end{remark}
+
+\begin{remark}[Public versus government datasets]
+In September 2020, the author submitted a data request to the Indiana Data Hub (IDH).
+Over the next three months, we worked with Indiana's COVID-19 response team to refine the request.
+In December 2020, the request was officially approved and data streams were made publicly available on the IDH.
+In this manuscript, we thus focus on this \emph{publicly-available data}, i.e., data  that anyone can download directly from the IDH.
+While the scientific community has contributed through independent COVID-19 observational and clinical studies, a significant component of public policy guidance has relied on testing and case count data, e.g., CDC and state guidelines based on test positivity rates and relative changes in the case counts over time.
+Addressing selection bias and measurement-error in these public datasets is imperative in better informing public policy debate.
+
+Note that the COVID-19 response team (and health departments in general) have access to \emph{government data} which may not be publicly-available and is likely stored at the individual-level.
+Due to data privacy and legal concerns, some collected covariates may not be reported publicly and others are aggregated and not reported jointly.
+While data analytic decisions made in Sections XX--XX were made due to access of publicly-available data, the overall data analysis framework is designed to be a tool for health departments to assess the pandemic and guide responses, e.g.,  account for individual-level covariate information, rather than publicly-reported strata-level information, that may include important covariates such as symptom status and recent COVID-19 contact.
  \end{remark}

### 2022-03-09 visualizations/fb_table1.R

@@ -1,0 +1,25 @@
+fb_data = read.csv("../data/fb_indiana_data.csv")
+
+
+## Gender Summary
+## 1 = MALE, 2 = Female, 3 = Non-binary, 4 = Self-Describe, 5 = Prefer not to answer
+agg_gender = aggregate(weight ~ gender, fb_data, sum)
+weights_gender = agg_gender$weight[is.element(agg_gender$gender,c(1,2))]
+weights_gender/sum(weights_gender)
+
+## Age Summary
+#  18-24 years (1), 25-34 years (2), 35-44 years (3), 45-54 years (4), 
+#  55-64 years (5), 65-74 years (6), 75 years or older (7)
+agg_age = aggregate(weight ~ age, fb_data, sum)
+agg_age_strata = c(sum(agg_age$weight[1:2]) + agg_age$weight[3]/2, 
+  agg_age$weight[3]/2 + agg_age$weight[4] +  agg_age$weight[5]/2,
+  agg_age$weight[5]/2 + sum(agg_age$weight[6:7]))
+agg_age_strata/sum(agg_age_strata)
+
+## Fever (Only  )
+fb_data$date 
+agg_fever = aggregate(weight ~ fever, subset(, sum)
+agg_age_strata = c(sum(agg_age$weight[1:2]) + agg_age$weight[3]/2, 
+                   agg_age$weight[3]/2 + agg_age$weight[4] +  agg_age$weight[5]/2,
+                   agg_age$weight[5]/2 + sum(agg_age$weight[6:7]))
+agg_age_strata/sum(agg_age_strata)

### 2022-03-09 write-up/covid-writeup.tex
finishing table 1

@@
  \begin{table}[th]
- \begin{tabular}{c | c | r r | r | r r }
- & & \multicolumn{2}{c}{Total Tests (\%)} & US & \multicolumn{2}{c}{Positive Test Rate (\%)}\\
- \cline{3-4} \cline{6-7}
- & & NonRandom & Random & Census & NonRandom & Random \\ \hline
- \multirow{2}{*}{Sex} & Female & 58.2 & 55.0 & 50.7 & 21.7 (11.2) & 1.4 \\
- & Male & 41.8 & 45.0 & 49.3 & 24.2 (12.4) & 2.1 \\ \hline
- \multirow{3}{*}{Age} & $<40$ & 39.4 & 28.0 & 52.7 & 29.7 (15.0) & 1.7 \\
- & $40-59$ & 41.1 & 36.0 & 25.2 & 24.9 (12.5) & 2.1 \\
- & $\geq 60$ & 19.5 & 36.0 & 22.1 & 6.7 (3.4) & 0.9 \\ \hline
- \multirow{2}{*}{Race} & White & 23.1 & 92.0 & 86.9 & 19.5 (9.6) & 1.5 \\
- & Nonwhite & 76.9 & 8.0 & 13.1 & 25.0 (12.3) & 3.4 \\ \hline
+ \begin{tabular}{c | c | r r r | r | r r }
+ & & \multicolumn{3}{c}{Total Tests (\%)} & US & \multicolumn{2}{c}{Positive Test Rate (\%)}\\
+ \cline{3-5} \cline{7-8}
+ & & NonRandom & Random & Facebook & Census & NonRandom & Random \\ \hline
+ \multirow{2}{*}{Sex} & Female & 58.2 & 55.0 & 54.0 & 50.7 & 21.7 (11.2) & 1.4 \\
+ & Male & 41.8 & 45.0  & 46.1 & 49.3 & 24.2 (12.4) & 2.1 \\ \hline
+ \multirow{3}{*}{Age} & $<40$ & 39.4 & 28.0  & 36.2 & 52.7
+ & 29.7 (15.0) & 1.7 \\
+ & $40-59$ & 41.1 & 36.0  & 34.3 & 25.2 & 24.9 (12.5) & 2.1 \\
+ & $\geq 60$ & 19.5 & 36.0  & 29.5 & 22.1 & 6.7 (3.4) & 0.9 \\ \hline
+ \multirow{2}{*}{Race} & White & 23.1 & 92.0  & - & 86.9 & 19.5 (9.6) & 1.5 \\
+ & Nonwhite & 76.9 & 8.0  & - & 13.1 & 25.0 (12.3) & 3.4 \\ \hline
  % Hispanic & Hispanic & 44.1 & 2.0 & 7.1 & 37.6 & 6.9 \\
  % Origin & Non-Hispanic & 55.9 & 98.0 & 92.9 & 13.0 & 1.3 \\ \hline
- \multirow{2}{*}{Fever} & Yes & 17.0 & 1.8 & - & 66.4 (32.1) & 4.5 \\
- & No & 83.0 & 98.2 & - & 15.6 (7.5) & 1.3 \\ \hline
- Household & Yes & 10.8 & 1.4 & - & 46.1 (22.4) & 29.4 \\
- $+$ Case & No & 89.2 & 98.6 & - & 21.6 (10.4) & 1.3 \\ \hline
- Prior $+$ & Yes & 6.1 & 1.4 & - & 39.2 (20.2) & 24.4 \\
- Test & No & 93.9 & 98.6 & - & 21.6 (11.1) & 1.3 \\ \hline
+ \multirow{2}{*}{Fever} & Yes & 17.0 & 1.8  & - & - & 66.4 (32.1) & 4.5 \\
+ & No & 83.0 & 98.2  & - & - & 15.6 (7.5) & 1.3 \\ \hline
+ Household & Yes & 10.8 & 1.4  & - & - & 46.1 (22.4) & 29.4 \\
+ $+$ Case & No & 89.2 & 98.6  & - & - & 21.6 (10.4) & 1.3 \\ \hline
+ Prior $+$ & Yes & 6.1 & 1.4  & - & - & 39.2 (20.2) & 24.4 \\
+ Test & No & 93.9 & 98.6  & - & - & 21.6 (11.1) & 1.3 \\ \hline
  \end{tabular}

### 2022-03-09 visualizations/fb_table1.R

@@ -1,5 +1,7 @@
-fb_data = read.csv("../data/fb_indiana_data.csv")
-
+work_dir = "/home/wdem/Dropbox/Fb-Data/"
+home_dir = "../data/"
+fb_data = read.csv(paste(work_dir, "fb_indiana_data.csv", sep = ""))
+fb_data = read.csv(paste(home_dir, "fb_indiana_data.csv", sep = ""))
 
 ## Gender Summary
 ## 1 = MALE, 2 = Female, 3 = Non-binary, 4 = Self-Describe, 5 = Prefer not to answer
@@ -16,8 +18,10 @@ agg_age_strata = c(sum(agg_age$weight[1:2]) + agg_age$weight[3]/2,
   agg_age$weight[5]/2 + sum(agg_age$weight[6:7]))
 agg_age_strata/sum(agg_age_strata)
 
-## Fever (Only  )
-fb_data$date 
+## Fever (Only April 25th to 29th)
+library(lubridate)
+fb_data$date  = ymd(fb_data$date )
+window = fb_data$date >= ymd("2020-04-25") & fb_data$date <= ymd("2020-04-29") 
 agg_fever = aggregate(weight ~ fever, subset(, sum)
 agg_age_strata = c(sum(agg_age$weight[1:2]) + agg_age$weight[3]/2, 
                    agg_age$weight[3]/2 + agg_age$weight[4] +  agg_age$weight[5]/2,

### 2022-03-10 write-up/covid-supplement-writeup.tex
Adding FB to table 1

@@
  & Lower CI & Estimate & Upper CI & Estimate \\ \hline
- Fever &  0.017 & 0.020 & 0.024 & 0.018 \\
- Cough &  0.142 & 0.148 & 0.154 & 0.149 \\
- Shortness & 0.058 & 0.060 & 0.062 & 0.062 \\ \hline
+ Fever &  0.073 & 0.097 & 0.012 & 0.018 \\
+ Cough &  0.140 & 0.150 & 0.159 & 0.149 \\
+ Shortness & 0.055 & 0.062 & 0.068 & 0.062 \\ \hline
  \end{tabular}

### 2022-03-10 write-up/covid-writeup.tex
Adding FB to table 1

@@
  \multirow{2}{*}{Sex} & Female & 58.2 & 55.0 & 54.0 & 50.7 & 21.7 (11.2) & 1.4 \\
- & Male & 41.8 & 45.0  & 46.1 & 49.3 & 24.2 (12.4) & 2.1 \\ \hline
+ & Male & 41.8 & 45.0  & 46.0 & 49.3 & 24.2 (12.4) & 2.1 \\ \hline
  \multirow{3}{*}{Age} & $<40$ & 39.4 & 28.0  & 36.2 & 52.7
@@
  % Origin & Non-Hispanic & 55.9 & 98.0 & 92.9 & 13.0 & 1.3 \\ \hline
- \multirow{2}{*}{Fever} & Yes & 17.0 & 1.8  & - & - & 66.4 (32.1) & 4.5 \\
- & No & 83.0 & 98.2  & - & - & 15.6 (7.5) & 1.3 \\ \hline
- Household & Yes & 10.8 & 1.4  & - & - & 46.1 (22.4) & 29.4 \\
- $+$ Case & No & 89.2 & 98.6  & - & - & 21.6 (10.4) & 1.3 \\ \hline
+ \multirow{2}{*}{Fever} & Yes & 17.0 & 1.8  & 1.0 & - & 66.4 (32.1) & 4.5 \\
+ & No & 83.0 & 98.2  & 99.0 & - & 15.6 (7.5) & 1.3 \\ \hline
+ Household & Yes & 10.8 & 1.4  & 1.8 & - & 46.1 (22.4) & 29.4 \\
+ $+$ Case & No & 89.2 & 98.6  & 98.2 & - & 21.6 (10.4) & 1.3 \\ \hline
  Prior $+$ & Yes & 6.1 & 1.4  & - & - & 39.2 (20.2) & 24.4 \\

### 2022-03-10 visualizations/fb_table1.R

@@ -23,13 +23,33 @@ library(lubridate)
 fb_data$date  = ymd(fb_data$date )
 window = fb_data$date >= ymd("2020-04-25") & fb_data$date <= ymd("2020-04-29") 
 agg_fever = aggregate(weight ~ fever, subset(fb_data, window), sum)
-agg_fever$weight/sum(agg_fever$weight)
+fever_proportions = (agg_fever$weight/sum(agg_fever$weight))
+print(fever_proportions)
+r = fever_proportions[2]
+sub_fb_data = subset(fb_data, window)
+var_r = sum((sub_fb_data$fever - r)^2 * sub_fb_data$weight^2, na.rm = T)/ sum(sub_fb_data$weight)^2
+c(r - 1.96*sqrt(var_r), r, r + 1.96*sqrt(var_r))
 
 agg_cough = aggregate(weight ~ cough, subset(fb_data, window), sum)
-agg_cough$weight/sum(agg_cough$weight)
+cough_proportions = agg_cough$weight/sum(agg_cough$weight)
+print(cough_proportions)
+r = cough_proportions[2]
+var_r = sum((sub_fb_data$cough - r)^2 * sub_fb_data$weight^2, na.rm = T)/ sum(sub_fb_data$weight)^2
+c(r - 1.96*sqrt(var_r), r, r + 1.96*sqrt(var_r))
 
 agg_shortness = aggregate(weight ~ shortness, subset(fb_data, window), sum)
 agg_shortness$weight/sum(agg_shortness$weight)
+shortness_proportions = agg_shortness$weight/sum(agg_shortness$weight)
+print(shortness_proportions)
+r = shortness_proportions[2]
+var_r = sum((sub_fb_data$shortness - r)^2 * sub_fb_data$weight^2, na.rm = T)/ sum(sub_fb_data$weight)^2
+c(r - 1.96*sqrt(var_r), r, r + 1.96*sqrt(var_r))
 
 agg_household = aggregate(weight ~ contact, subset(fb_data, window), sum)
 agg_household$weight/sum(agg_household$weight)
+household_proportions = agg_household$weight/sum(agg_household$weight)
+print(household_proportions)
+r = household_proportions[1]
+var_r = sum((sub_fb_data$contact - r)^2 * sub_fb_data$weight^2, na.rm = T)/ sum(sub_fb_data$weight)^2
+c(r - 1.96*sqrt(var_r), r, r + 1.96*sqrt(var_r))
+

### 2022-03-14 write-up/letters/response-ref-1-v1.tex
updated paper looking good. close!

@@
 \usepackage{amsthm, amsfonts}
+\usepackage{xcolor}
 \usepackage{hyperref}
@@
 \item {\it The AE also mentions a few comments as well:}
-\begin{itemize}
+\begin{enumerate}
 	\item {\it Title: As noted by one of the reviewers, the “paradox” is never discussed/defined in the manuscript.
@@
 Much of the results appear to rely on the SRS and it ignores the potential for nonresponse bias.}
-\end{itemize}
+\end{enumerate}
 
@@
 \item {\bf Section 1.1}
-\begin{itemize}
+\begin{enumerate}
 	\item {\it For the first goal of the paper, it might be helpful to link the biases with those outlined in a set of papers from the Harvard Center for Disease Dynamics: \url{https://doi.org/10.1007/s10654-021-00727-7} and DOI:\url{10.1093/aje/kwaa188}}
@@
 	\vspace{5mm}
-\end{itemize}
+\end{enumerate}
 \item {\bf Section 2.1: Diagnostic testing}
-\begin{itemize}
+\begin{enumerate}
 	\item {\it Please clarify in the first paragraph the distinction between incubation period (time to symptom onset) and latent period (time to infectiousness).
@@
 	\vspace{5mm}
-\end{itemize}
+\end{enumerate}
 \item {\bf Section 2.2: Publicly available data}
-\begin{itemize}
+\begin{enumerate}
 	\item {\it This section could benefit from a few remarks on why it is important / useful to focus on publicly available testing data.
@@
 	\vspace{5mm}
-\end{itemize}
+\end{enumerate}
 \item {\bf Section 2.3: Probabilistic Samples}
-\begin{itemize}
+\begin{enumerate}
 	\item {\it The $\approx 30\%$ response rate in the supposedly random sample requires some remarks.
@@
 
-	We can discuss the non-response weights used in their paper to adjust.  And that this is a different and solvable issue (we can just weight by the )
+	We now remark on the response rate in Section 5.
+Specifically, we cite how Yiannoutsos et al.
+(2021) handle non-response in prevalence estimation.
+We acknowledge the potential for a difference in characteristics between intended and final sample.
+We clarify that the Census data is Indiana specific and is presented as a reference to demonstrate the potential difference between intended and final sample.
+
 	\vspace{5mm}
@@
 
-	We include an age breakdown of the data.
-Note that weights are used to adjust and we are using these in the calculations.
-We now include a table in the appendix a breakdown of the demographic information for each sample.
+	The difference in age breakdown of sampled individuals in the Facebook / CMU survey, now referred to as Delphi's COVID-19 Trends and Impact Survey (CTIS), from the general population is acknowledged in Section 5.
+Table 1 now presents estimates of all demographic and covariate information using Delphi CTIS's weighting procedure.
+These weights are used in computing probabilities of selection as shown in Equation (4.2) in Section 4.1.1.
+Estimates for 'Fever' and 'Household + Case' only use data collected between April 25th and 29th to match the timing of the random survey.
+Note that individual-level racial demographics were not provided by CTIS due to data privacy concerns and there was no corresponding question on percent of tests from individuals with a prior positive test.
+
 	\vspace{5mm}
-\end{itemize}
+\end{enumerate}
 \item {\bf Section 3: Analysis of case count data}
-\begin{itemize}
+\begin{enumerate}
 	\item {\it Please clarify the definition of ‘active infection’ (any infection, vs infectious, vs symptomatic, vs detectable by molecular testing?).
@@
 
-	We clarify our choice of notation by adding an appendix with notation.
+	We clarify that this notation was chosen to match standard survey statistics notation: $Y_i$ is the true, fixed outcome value and $y_i$ is the  population-level .
+To help the reader, we now include an appendix with all the notation described in detail.
 	\vspace{5mm}
@@
 
-	This is now stated explicitly
-\end{itemize}
+	This is now stated explicitly in Section 3, paragraph 2.
+\end{enumerate}
 \item {\bf Section 3.1: Imperfect testing}
-\begin{itemize}
+\begin{enumerate}
 	\item {\it Here, the value of distinguishing between test status and infection status becomes clearer.
@@
 
-	We now include a discussion of the underlying continuous cycle (with proper credit).
-The binary variable of interest can depend on infectious is what is often most important.
-So we highlight that most but acknowledge the binary variable can be defined....
+	Remark 4 is now included to discuss that the binary testing data is generated from an underlying continuous cycle threshold and that the issue of measurement error remains even in the absence of selection bias.
 	\vspace{5mm}
 
-	\item {\it The estimator $y = (yn-FP) /(1-(FP+FN))$ is a standard estimator for correcting for measurement error using FP \& FN estimated from a validation sample, where yn is the mis-measured average \& y is the corrected average.
+	\item {\it The estimator $y = (\bar y_n-FP) /(1-(FP+FN))$ is a standard estimator for correcting for measurement error using FP \& FN estimated from a validation sample, where yn is the mis-measured average \& y is the corrected average.
 The use of this estimator to iteratively estimate the point prevalence is a good contribution of this paper, but the failure to distinguish between infection \& test status in the variable names somewhat obscures that this is what the author is doing.}
@@
 
-	We now more carefully distinguish.
+	We now more carefully distinguish this and connect with the epidemiologic terminology to help readers from epidemiology.
 And yes, this is the standard estimator.
-We have simply derived it from a new perspective (the decomposition perspective which didn't require any Bayesian machinery).
+We clarify, as you stated above, that the contribution of this paper was to derive this estimator from a new perspective (i.e., the decomposition perspective which didn't require any Bayesian machinery).
 	\vspace{5mm}
@@
 
-	I now give an explicit example.
-The test characteristics may change based on the nurse (cite) and/or what exact test is used.
-For example, a center that moves from nasal swab to spit to increase testing capacity will see a ...(cite).
-\end{itemize}
+	We clarify that increasing testing capacity may impact FP and FN rates if altering changes the sample collection quality, e.g., a testing center switches from nasal swab to oropharyngeal swabs or saliva specimen to increase testing capacity will change measurement error due to the different test characteristic.
+\end{enumerate}
 \item {\bf Section 3.2: Regrettable Rates}
-\begin{itemize}
+\begin{enumerate}
 	\item {\it Here, and throughout the theoretical example, the model is an SIR model.
@@
 	\vspace{5mm}
-\end{itemize}
+\end{enumerate}
 \item {\bf Section 3.3: Effective Reproduction Number}
-\begin{itemize}
+\begin{enumerate}
 	\item {\it Here again, an SEIR model would be valuable.
@@
 (2014) in Section I of the supplementary materials, which shows the same general conclusions.
-\end{itemize}
+\end{enumerate}
 \item {\bf Section 3.4: Rate comparisons}
-\begin{itemize}
+\begin{enumerate}
 	\item {\it This section would also be more relevant with an SEIR model}
@@
 	\vspace{5mm}
-\end{itemize}
+\end{enumerate}
 \item {\bf Section 4.1: Selection propensity estimation}
-\begin{itemize}
+\begin{enumerate}
 	\item {\it The use of weights to correct for sampling bias is an established practice in epidemiology.
@@
 
-	Inverse probability of selection weights describe by Cole and Stuart are exactly what we are building.  However, how to build them is the question.
-	As far as we are aware, the \href{approaches}{https://academic.oup.com/aje/article/172/1/107/82724} require knowledge of the selection mechanism which is not available in our setting.
-Instead, we require auxiliary information.
-Here, (cite elliott stuff).
-A key different is the time-varying nature which we highlight.
+	Remark 5 in Section 4.1.1 discusses how the proposal is tied closely to the use of weights to correct for sampling bias.
+Specifically, in epidemiology literature, the survey sampling weights and transportability weights require knowledge of the selection mechanism, which is not available in our setting.
+Instead, we require auxiliary information in the form of a random sample to estimate the probability of selection.
+A key difference from prior approaches is the need to estimate time-varying probabilities of selection, e.g., Figure 6 shows how the selection propensity of young individuals changes dramatically over the observation window.
 	\vspace{5mm}
-	\item {\it This section, and the sections below, also become somewhat confusing to follow because of the multiple different variables which are all denoted Ij.
-Since a chief goal is to describe and correct for errors in the sampling process, it would be much more useful for the reader if there were different notation for each of the different sampling mechanisms being discussed / used (eg, inclusion in the random sample, inclusion in the non-random validation sample, inclusion in the original diagnostic testing data, inclusion in ??
+	\item {\it This section, and the sections below, also become somewhat confusing to follow because of the multiple different variables which are all denoted $I_j$.
+Since a chief goal is to describe and correct for errors in the sampling process, it would be much more useful for the reader if there were different notation for each of the different sampling mechanisms being discussed/used (eg, inclusion in the random sample, inclusion in the non-random validation sample, inclusion in the original diagnostic testing data, inclusion in ??
 on day x).}
@@
 
-	We have no built a notation table and made the notation different.  Apologies for the confusion. We use $I_j^{R}$ and $I_j^{NR}$.
+	We have no built a notation table and made the notation different.  Apologies for the confusion. We use $I_j^{R}$ and $I_j^{NR}$.  As stated in our answer to question set XX (subquestion XX) we have
 	\vspace{5mm}
@@
 
-	Since we normalizing the weights, this is just to write it in simple form but basically $\sum w y$ is the estimate and we subtract FP.  So you are right!
+	Since weights are normalized, these are equivalent, i.e., $\sum w (x_i) FP / \sum w(x_i) = FP$.
 	\vspace{5mm}
@@
 
-	Necessary for smoothing.
+	Section 4.1.3 allows researchers to choose to go beyond moving windowed averages (i.e., simply take weekly averages).
+The kernel function allows the user to choose weighted averages that may be more robust than the uniform kernel (which is equivalent to the moving averages).
 	\vspace{5mm}
-\end{itemize}
+\end{enumerate}
 \item {\bf Section 4.2: model-based estimation}
-\begin{itemize}
+\begin{enumerate}
 	\item {\it The SEIR model described in the text and in the supplement differ in a way that I find confusing – the code and equations describe a simple SEIR model but the written text describes an SEmInR model with stratum-specific covariates which appear to apply to at least the I compartments and possibly the E compartments.
@@
 
-	Prior is Dirichlet (uniform) but not time-varying.  This allows us to learn the links.
+	We thank the reviewer for catching this issue.
+We now clarify that the model proposed does not assume equal mixing.
+Rather due to limited data, we assume a these mixing probabilities are not time varying and that the  prior probability of mixing is chosen from a symmetric Dirichlet distribution (uniform) time-varying.
+We clarify this in Section 5.2 and comment on the code to clarify this more carefully in Appendix H.
+	\textcolor{red}{Need To Do.}
 	\vspace{5mm}
-\end{itemize}
+\end{enumerate}
 \item {\bf Section 4.3: Doubly Robust Estimation}
-\begin{itemize}
+\begin{enumerate}
 	\item {\it The term $\mu(x_j)$ in the unnumbered equation does not seem to be defined anywhere – this is presumably coming from the SEIR model given that the rest of the equation is essentially the IPW-based estimator.
@@
 
-	The SEIR model is used to generate the number of new weekly cases which allows us to calculate the active infection rate.  $\mu(x_j)$ is the model-based estimate.
+	We clarify this notation in Section 4.2.
+Specifically, the SEIR model is used to generate the number of new weekly cases which allows us to calculate the number of individuals with an active infection within each strata (i.e.~$I_{k,t}$ for the $k$th strata).
+We then can use this to estimate the active infection rate per strata,~$\mu_k$.
+	\textcolor{red}{Need To Do.}
 	\vspace{5mm}
-\end{itemize}
+\end{enumerate}
 \item {\bf Section 5: Applied analysis}
-\begin{itemize}
+\begin{enumerate}
 	\item {\it The applied example relies heavily on the idea that the random sample is a true random sample of the population of Indiana eligible to be infected with COVID, and that the non-random sample and the Facebook / CMU sample similarly provide representative measures of covariate data.
@@
 	\vspace{5mm}
-\end{itemize}
+\end{enumerate}
 \item {\bf Section 5.1.1. Disease Prevalence}
-\begin{itemize}
+\begin{enumerate}
 	\item {\it This section seems to assume that individuals can only be tested once but that assumption is unlikely to be true.
@@
 	\vspace{5mm}
-\end{itemize}
+\end{enumerate}
 \item {\bf Section 5.1.2 Time varying estimate}
-\begin{itemize}
+\begin{enumerate}
 	\item {\it 5.1.2.
@@
 	\vspace{5mm}
-\end{itemize}
+\end{enumerate}
 \item {\bf 5.2 SEIR model}
-\begin{itemize}
+\begin{enumerate}
 	\item {\it As stated above, some description of the assumptions related to contact patterns and transmission between and among age strata is needed.
@@
 	\vspace{5mm}
-\end{itemize}
+\end{enumerate}
 \end{enumerate}
@@
 
-\begin{itemize}
+\begin{enumerate}
 \item {\it The paper states some facts about the SARS-CoV-2 virus that are outdated, for example “Upon infection with SARS-CoV-2, an incubation period starts and lasts approximately five days [Lauer et al., 2020] before the viral load is high enough to be detected.” (Page 5).
@@
 	\end{enumerate}
-\end{itemize}
+\end{enumerate}
 

### 2022-03-14 write-up/covid-writeup.tex
updated paper looking good. close!

@@
 Suppose that $n$ tests are performed and we observe the values $y_1, \ldots, y_n \in \{0,1\}$.
-Then a natural candidate for AIR is the proportion of positive tests $\bar y = \frac{1}{n} \sum_{i=1}^n y_i$, and a natural candidate for overall active cases is $N \times \bar y$.
+Then a natural candidate for AIR is the proportion of positive tests $\bar y = \frac{1}{n} \sum_{i=1}^n y_i$ -- commonly referred to as the test positive rate -- and a natural candidate for overall active cases is $N \times \bar y$.
  Under simple random sampling (SRS) or any other epsem\footnote{equal probability of selection method} design, the above are unbiased estimators of the population-level quantities of interest.
@@
 In Appendix~\ref{app:memestimator}, assuming sensitivity and specificity are known a priori, a novel iterative procedure is used to construct the estimator $\tilde y_n = (\bar y_n - FP)/(1-(FP+FN))$, which is unbiased under simple random sampling (SRS); see Appendix~\ref{app:modelbased} for a discussion of the connection to model-based estimators.
+In the language of the epidemiology literature, the estimator $\tilde y_n$ is the standard estimator for correcting for measurement error using false positive and negative rates that have been estimated from a validation sample, where $\bar y_n$ is the mis-measured average \& $\tilde y_n$ is the corrected average.
 To understand the impact of selection bias and imperfect testing, we derive the following statistical decomposition of the error between $\tilde y_n$ and $\bar Y$:
@@
 
+ \begin{remark}
+ \label{rmk:testbinary}
+ Note that test positivity is based on an underlying continuous cycle threshold and therefore is not strictly binary.
+Therefore, while we focus on the interplay of measurement error and selection bias in the context of binary testing due to the binary nature of the COVID-19 data, it should be noted that measurement error can lead to bias even in the absence of selection bias.
+ \end{remark}
+
  Figure~\ref{fig:heatmap} shows that $D_M$, as a function of the relative frequency ($f_1/f_0$) and log odds ratio, can be both positive and negative as well as a range of magnitudes~\citep{Beesley2020,Beesley2019,Smeden2019}.
@@
 Thus the effective sample size even in optimistic scenarios is equivalent to a moderate random sample from the population.
-Moreover, increased testing capacity may also come at the cost of increases in false positive and negative rates.
+Moreover, increased testing capacity may alter false positive and negative rates due to changes in sample collection quality, e.g., a testing center switches from nasal swab to oropharyngeal swabs or saliva specimen to speed up data collection.
 Consider the case where $f$ increasing from $0.003$ to $0.01$ and is associated with the false negative rate rising from $13\%$ to $20$\%.
@@
 Solving~\eqref{eq:auxinfoprob} is done by iteratively re-weighted least squares (IRLS); see Section~\ref{app:irls} in the supplementary materials for details.
+
+ \begin{remark}
+ Weights built from selection propensities~$\{\pi(x_i; \theta)\}_{i=1}^n$ are common practice in the epidemiology literature.
+Survey sampling and transportability weights require knowledge of the selection mechanism~\citep{10.1093/aje/kwx164,ColeStuart2010}, which is not available in our setting.
+Instead, we require auxiliary information in the form of a random sample to estimate the probability of selection.
+A key difference from prior approaches is the need to estimate time-varying probabilities of selection, e.g., Figure 6 shows how the selection propensity of young individuals changes dramatically over the observation window.
+ \end{remark}
 

### 2022-03-16 write-up/letters/response-ref-1-v1.tex
Updating paper, almost there

@@
 
-	We have no built a notation table and made the notation different.  Apologies for the confusion. We use $I_j^{R}$ and $I_j^{NR}$.  As stated in our answer to question set XX (subquestion XX) we have
+	We apologize for the confusion.
+We now include distinct notation for each of the sampling mechanisms.
+To strike a balance and keep notation at a minimum, we decided to use superscripts $R$ and $NR$ to distinguish between the probability and non-probability samples.
+To facilitate comprehension, we have added a notation glossary in Section B of the supplementary materials.
 	\vspace{5mm}
@@
 
-	Section 4.1.3 allows researchers to choose to go beyond moving windowed averages (i.e., simply take weekly averages).
+	Section 4.1.3 allows researchers to choose to go beyond simpler moving window averages (i.e., weekly averages).
 The kernel function allows the user to choose weighted averages that may be more robust than the uniform kernel (which is equivalent to the moving averages).
+This approach helps smooth the estimated propensity over time; moreover, the flexibility of the approach for government agencies who may observed random samples at different frequencies is of practical importance.
 	\vspace{5mm}

### 2022-03-16 write-up/covid-supplement-writeup.tex
updated paper looking good. close!
Updating paper, almost there

@@
 All relevant code can be found at \url{https://github.com/wdempsey/covid-umich}.
+
+\section{Notation Glossary}
+\label{app:notation}
+
+\paragraph{Notation for Section~\ref{section:casecount} (non-temporal setting)}
+\begin{itemize}
+\item[$Y_j$] \quad Binary COVID-19 status of individual $j$ in the population.
+\item[$N$] \quad Population size
+\item[$\bar Y$] \quad Population average, $\bar Y = N^{-1} \sum_{j=1}^N Y_j$
+\item[$I_j$] \quad Selection indicator of individual $j$ in the population into the sample
+\item[$y_j$] \quad Binary COVID-19 status of sample individual $j=1,\ldots,n$.
+\item[$\bar y_n$] \quad Sample average, $\bar y_n = n^{-1} \sum_{j=1}^N I_j Y_j = n^{-1} \sum_{j=1}^n y_j$
+\item[$\rho_{I,Y}$] \quad \emph{Data quality}, i.e., correlation between selection indicator and population outcome
+\item[$f$] \quad \emph{Data quantity}, i.e., sampling fraction $f = n/N$
+\item[$\sigma_Y$] \quad \emph{Problem Difficulty}, i.e., population variance $\sigma_Y^2 = (N)^{-1} \sum_{j=1}^N (Y_j - \bar Y)^2$.
+\item[$FP$] \quad False Positive Rate
+\item[$FN$] \quad False Negative Rate
+\item[$\tilde y_n$] \quad Sample average adjusted for false negative and positive rates, i.e., $\tilde y_n = (1-FP-FN)^{-1} (\bar y_n - FP)$
+\item[$f_0, f_1$] \quad Sampling fraction among COVID-19 negative and positive individuals respectively, i.e., $f_1 := \sum_{j=1}^N I_j Y_j / \sum_{j=1}^N Y_j$
+and $f_0 := \sum_{j=1}^N I_j (1-Y_j) / \sum_{j=1}^N (1-Y_j)$.
+\item[$\Delta, M$] \quad Sampling rate differential on additive -- $\Delta = f_1 - f_0$ -- and multiplicative -- $M = f_1/f_0$ -- scales.
+\end{itemize}
+
+\paragraph{\bf Notation for Section~\ref{section:improvedcasecount} (temporal setting)}
+\begin{itemize}
+\item[$Y_{j,t}$] \quad Binary COVID-19 status of individual $j$ in the population at time~$t$.
+\item[$I^{NR}_{j,t}$] \quad Selection indicator of individual $j$ in the population into the non-probability sample.
+\item[$X_{j,t}$] \quad Feature vector for individual $j$ in the population at time~$t$ that impact self-selection into the non-probability sample.
+\item[$I_{j,t}^{R}$] \quad Selection indicator of individual $j$ in the population into the probability sample.
+\item[$W_{j,t}^{R}$] \quad Weight of individual $j$ in the population in the probability sample (i.e., inverse-probability of selection)
+\item[$\pi (X_{j,t}; \theta)$] \quad Self-selection propensity of individual~$j$ into the non-probability sample given feature vector~$X_{j,t}$ and Probability of selection model evaluated indicator of selection of the
+\item[$w (X_{j,t})$] \quad Weight of indivdual~$j$ in the population in the non-probability sample, i.e., $w(x) = 1/\pi(x; \theta)$.
+\item[$\mu(X_{j,t})$] \quad The indicator of selection of the
+\end{itemize}
 

### 2022-03-16 write-up/covid-writeup.tex
Updating paper, almost there

@@
  %, which is scaled by the factor $(1 - FP - FN)^{-1}$.
+ For ease of comprehension, a notation glossary is provided in Section~\ref{app:notation} in the supplementary materials.
 
@@
 For simplicity, the dynamic nature of the outbreak and recoverability of individuals is ignored for now.
-Suppose the selection indicator $I_j$ is a Bernoulli random variable that depends only on these covariates, i.e., $P(I_j = 1 \mid X_j = x) = \pi (x; \theta)$.
+Let~$I_j^{NR}$ denote the selection indicator for individual~$j$ in the population into the non-probability sample.
+Assume it is a Bernoulli random variable that depends only on these covariates, i.e., $P(I^{NR}_j = 1 \mid X_j = x) = \pi (x; \theta)$.
 Here, we focus on logistic regression models, i.e., $\pi(x; \theta) = \text{expit} \left( x^\top \theta \right)$.
@@
  \label{eq:propensity}
- \sum_{j=1}^N I_j \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi (X_j; \theta) \right).
+ \sum_{j=1}^N I_j^{NR} \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi (X_j; \theta) \right).
  \end{equation}
@@
  Here we assume access to a probability sample measuring the same set of covariates.
-Let $\tilde I_j$ be an indicator that the individual was included in the probability sample and $\tilde W_j$ be the probability of inclusion for the $j$th individual.
+Let $I_j^{R}$ be an indicator that the individual was included in the probability sample and $W_j^R$ be the probability of inclusion for the $j$th individual.
 \cite{Chen2019} use a probability sample to construct a design-unbiased estimator of the second term
@@
  \label{eq:auxinfoprob}
- \sum_{j=1}^N I_j \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
+ \sum_{j=1}^N I_j^R \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N I_j^R W_j^R \log ( 1 - \pi (X_j; \theta)).
  \end{equation}
@@
 Solving~\eqref{eq:auxinfoprob} is done by iteratively re-weighted least squares (IRLS); see Section~\ref{app:irls} in the supplementary materials for details.
+For ease of comprehension, a notation glossary is provided in Section~\ref{app:notation} in the the supplementary materials.
 
@@
 
- Given a selection propensity, define the inverse probability weight~$w(x) = \pi (x; \hat \theta)^{-1}$. Then the IPW estimator adjusted for measurement error is given by
+ Given a selection propensity, define the inverse probability weight~$w^{NR}(x) = \pi (x; \hat \theta)^{-1}$. Then the IPW estimator adjusted for measurement error is given by
  \begin{equation}
@@
  \bar y_n^\star
- = \frac{1}{1-FP-FN} \cdot \frac{\sum_{i=1}^n w(x_i) (y_i - FP)}{\sum_{i=1}^n w (x_i)}
+ = \frac{1}{1-FP-FN} \cdot \frac{\sum_{i=1}^n w^{NR} (x_i) (y_i - FP)}{\sum_{i=1}^n w^{NR} (x_i)}
  \stackrel{(2)}{=} \frac{1}{1-FP-FN} \sum_{k=1}^K \frac{d_k w_k}{w} (\bar y_k - FP),
@@
 
- Let $\tilde I_j (X_j) = I_j  \cdot W(X_j)$ for $j=1,\ldots,N$.  Then the error when comparing weighted estimator~$\bar y_n^\star$ to the true prevalence $\bar Y$ can be expressed as:
+ Let $I_j^{NR} (X_j) = I_j^{NR}  \cdot w^{NR}(X_j)$ for $j=1,\ldots,N$.  Then the error when comparing weighted estimator~$\bar y_n^\star$ to the true prevalence $\bar Y$ can be expressed as:
  \begin{equation}
  \label{eq:statdecomp2}
- \rho_{\tilde I (X), Y} \times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y} \times \underbrace{\left[ 1 - \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{\tilde D_M}
+ \rho_{I^{NR} (X), Y} \times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y} \times \underbrace{\left[ 1 - \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{\tilde D_M}
  \end{equation}
- where $CV_W$ is the coefficient of variation (i.e., standard deviation/mean) of $W_J$ given $I_J = 1$, $\rho_{\tilde I(X), Y}$ is the empirical correlation which here depends on covariate distribution, $\tilde f_k = E[ W_J I_J \mid Y_j = k]$ for $k=0,1$, and $\tilde \Delta = \tilde f_1 - \tilde f_0$.
+ where $CV_W$ is the coefficient of variation (i.e., standard deviation/mean) of $w^{NR} (X_J)$ given $I_J = 1$, $\rho_{I^{NR} (X), Y}$ is the empirical correlation which here depends on covariate distribution, $\tilde f_k = E[ w^{NR} (X_J) I_J \mid Y_J = k]$ for $k=0,1$, and $\tilde \Delta = \tilde f_1 - \tilde f_0$.
 See Appendix~\ref{app:ipwderivation} for the derivation.
@@
  $\sqrt{1 + \frac{CV_W^2}{1-f}} \geq 1$.
-Hence, if the data quality does not increase (i.e., $| \rho_{\tilde I (X), Y} | = | \rho_{I,Y}|$ ) then weighting increases the error magnitude.
+Hence, if the data quality does not increase (i.e., $| \rho_{I^{NR} (X), Y} | = | \rho_{I,Y}|$ ) then weighting increases the error magnitude.
 Second, the relative error increase depends on the fraction of population sampled $f$, implying that for large samples there is a larger potential increase in the error if the weights do not improve data quality.
@@
 
- As demonstrated below in Lemma~\ref{lemma:ipw}, if the propensity model is correctly specified then the $E_{\bf R} [ \rho_{\tilde I (X), Y} ] = 0$ and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then the data quality index is unlikely to inversely scale with population size.
+ As demonstrated below in Lemma~\ref{lemma:ipw}, if the propensity model is correctly specified then the $E_{\bf R} [ \rho_{I^{NR} (X), Y} ] = 0$ and therefore $\rho_{I^{NR} (X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then the data quality index is unlikely to inversely scale with population size.
  % Indeed, Horvitz-Thompson estimators are sensitive to errors in the estimated weights.
@@
  \label{eq:tvpropensity}
- \sum_{t=1}^T \left[ \sum_{j=1}^N I_{j,t} \log \left( \frac{\pi_t (X_{j,t}; \theta)}{1-\pi_t(X_{j,t}; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi_t (X_{j,t}; \theta) \right) \right]
+ \sum_{t=1}^T \left[ \sum_{j=1}^N I^{NR}_{j,t} \log \left( \frac{\pi_t (X_{j,t}; \theta)}{1-\pi_t(X_{j,t}; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi_t (X_{j,t}; \theta) \right) \right]
  \end{equation}
- where $t=1,\ldots,T$ are the days when case-count data is reported.
-Here, $I_{j,t}$ denotes self-selection into testing on day $t$, which is highly correlated with prior testing and results.
+ where $t=1,\ldots,T$ are the days when case-count data is reported, and $I^{NR}_{j,t}$ denotes self-selection into testing on day $t$, which is highly correlated with prior testing and results.
 For example, an individual who tests positive may be unlikely to seek testing in the subsequent few days/weeks.
@@
  $$
- \sum_{t^\prime=1}^T K_h(|t^\prime - t|) \left[ \sum_{j=1}^N I_{j,t^\prime} \log \left( \frac{\pi_t (X_{j,t^\prime}; \theta)}{1-\pi_t(X_{j,t^\prime}; \theta)} \right) + \sum_{j=1}^N \tilde W_{j,t^\prime} \tilde I_{j,t^\prime} \log \left( 1 - \pi_t (X_{j,t^\prime}; \theta) \right) \right]
+ \sum_{t^\prime=1}^T K_h(|t^\prime - t|) \left[ \sum_{j=1}^N I_{j,t^\prime}^{NR} \log \left( \frac{\pi_t (X_{j,t^\prime}; \theta)}{1-\pi_t(X_{j,t^\prime}; \theta)} \right) + \sum_{j=1}^N W^{R}_{j,t^\prime} I^R_{j,t^\prime} \log \left( 1 - \pi_t (X_{j,t^\prime}; \theta) \right) \right]
  $$
@@
  \frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t}{\pi_{j,t}} \\
- \frac{1}{N \times \sum_{l = 1}^L K_h \left(|t - t_l^\prime| \right)} \sum_{l = 1}^L K_h \left(|t - t_l^\prime| \right) \left[ \sum_{j=1}^N I_{j,t_l^\prime} X_{j,t_l^\prime} - \sum_{j = 1}^N \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime}  \pi_{j,t_l^\prime} X_{j,t_l^\prime}  \right] \\
+ \frac{1}{N \times \sum_{l = 1}^L K_h \left(|t - t_l^\prime| \right)} \sum_{l = 1}^L K_h \left(|t - t_l^\prime| \right) \left[ \sum_{j=1}^N I^{NR}_{j,t_l^\prime} X_{j,t_l^\prime} - \sum_{j = 1}^N I^{R}_{j,t_l^\prime} W^{R}_{j,t_l^\prime}  \pi_{j,t_l^\prime} X_{j,t_l^\prime}  \right] \\
  \frac{1}{n_{FP}} \sum_{i=1}^{n_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\

### 2022-03-17 write-up/letters/response-ref-1-v1.tex
So close

@@
 
-	We clarify this notation in Section 4.2.
+	We clarify this notation in Section 4.3.
 Specifically, the SEIR model is used to generate the number of new weekly cases which allows us to calculate the number of individuals with an active infection within each strata (i.e.~$I_{k,t}$ for the $k$th strata).
-We then can use this to estimate the active infection rate per strata,~$\mu_k$.
-	\textcolor{red}{Need To Do.}
+We then can use this to calculate the posterior mean for the active infection rate per strata,~$\mu_k$.
 	\vspace{5mm}
@@
 
-	We provide further emphasis and evidence.  A key point is that this data is used to illustrate the 'what to do' point that was raised in the first round.
+	We added Remark 6 to emphasize these potential limitations.
+A key point is that the case study is to demonstrate our framework for answering the \emph{what to do with this knowledge of selection bias and measurement error} point that was raised in the first round of submission.
+In Remark 6, we urge health policy experts and government officials to assess these issues when applying this framework in their own future work.
 	\vspace{5mm}
@@
 
-	We acknowledge this limitation.  As an alternative we assume the risk set is always total pop.  This presents a range.
+	Section 5.1.1 studies testing between April 25th to 29th 2020 during which testing was limited.
+This led us to the assumption that the individuals who already tested were unlikely to be available for testing.
+We continue to calculate the sampling fraction as number of tests divided by the total population minus prior tests, but note at the end of Section 5.1.1 that using the total population in the denominator yields a very similar sampling fraction and the calculations have negligible change.
 	\vspace{5mm}
@@
 
-	Cite PNAS for measurement error and response rate cleaning.
+	See answers to Quesitons 5a and 14a above.  The prevalence estimate comes from Yiannoutsos et al. (2021) who detail their adjustments for response rates in estimation of active prevalence.
 	\vspace{5mm}
@@
 
-	We have added Remark XX to discuss the bias.
-In Remark XX, we state that availability of symptom status (which we know the state collects but has told us is unavailable due to data privacy / confidentiality issues) is important to use.
+	We thank the reviewer for raising this issue.
+We have added a remark in sction 5.1.2 that raises this as a limitation.
+We restart here that the need for the approaches is due to the lack of availability of symptom status (which we know the state collects but has told us is unavailable due to data privacy / confidentiality issues).
+Therefore, as a framework for health policy experts and government officials who have access to the complete government datasets (i.e., not just the public data), these steps will not be necessary.
 	\vspace{5mm}
@@
 
-	We acknowledge this shortcoming.  We now include versions of Figure 6 for non-white fr
+	We acknowledge this shortcoming.  We now include versions of Figure 6 for non-white groups in Section XX of the supplementary materials.
+	\textcolor{red}{NEED TO DO}
 	\vspace{5mm}
@@
 
-	We now describe the model in more detail.
+	See answers to Question XX above.
+In Section 5.2, we now clarify how we calculate the infection-fatality rate.
+Similar to Johndrow (2021), the model only uses COVID-19 related deaths.
+Deaths to non-COVID causes are not modeled but would appear in the R component.
 	\vspace{5mm}

### 2022-03-17 write-up/covid-writeup.tex
So close

@@
 
-\title[Statistical paradoxes in coronavirus case-counts]{Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic} %
+\title[Addressing selection bias and measurement error in COVID-19 case counts]{Addressing selection bias and measurement error in COVID-19 case count data using auxiliary information} %
 
@@
 We then combine these selection propensities with epidemiological forecast models to construct a \emph{doubly robust} estimation method that accounts for both measurement-error and selection bias.
-This method is then applied to estimate Indiana's prevalence using case-count, hospitalization, and death data with demographic information, a statewide random molecular sample collected from April 25--29th, and Facebook's COVID-19 symptom survey.
+This method is then applied to estimate Indiana's prevalence using case-count, hospitalization, and death data with demographic information, a statewide random molecular sample collected from April 25--29th, and Delphi's COVID-19 Trends and Impact Survey.
 We end with a series of recommendations based on the proposed methodology.
@@
 
- \subsubsection{Facebook/CMU symptom survey.}
+ \subsubsection{Delphi's COVID-19 Trends and Impact Survey}
  \label{subsection:fbsymptom}
- Since April 2020, Facebook has started randomly showing an opt-in symptom survey to its users~\citep{delphisurvey}.
+ Since April 2020, in collaboration with Facebook, the Delphi group at Carnegie Mellon University has conducted the COVID-19 Trends and Impact Survey (CTIS) to monitor the spread and impact of the COVID-19 pandemic in the United States.
+The survey is advertised through Facebook, who automatically select a random sample of its users to see the advertisement~\citep{doi:10.1073/pnas.2111454118}.
 Data collected includes basic demographic information and if the respondent has symptoms such as fever, coughing, shortness of breath, or loss of smell which are associated with COVID-19.
@@
 Figure~\ref{fig:fbsymptoms} displays smoothed estimates of the fraction of individuals who report COVID-like symptoms within the past 24-hours by age and gender.
-The Facebook symptom survey is used as the main source of auxiliary information on time-varying characteristics of the population of Indiana, e.g., displaying COVID-like symptoms or in contact with COVID-19 positive individuals.
+The Delphi's COVID-19 Trends and Impact Survey (CTIS) is used as the main source of auxiliary information on time-varying characteristics of the population of Indiana, e.g., displaying COVID-like symptoms or in contact with COVID-19 positive individuals.
 
@@
  \caption{Rate of reported COVID-19 symptoms per strata.
-Daily rates were estimated using weighted method suggested by~\citep{delphisurvey} on each day separately and then smoothed over time using local-linear nonparametric regression.}
+Daily rates were estimated using weighted method suggested by~\citep{doi:10.1073/pnas.2111454118} on each day separately and then smoothed over time using local-linear nonparametric regression.}
  \label{fig:fbsymptoms}
@@
  Rather than relying solely on IPW or epidemiological forecasts, here we combine forecasting and inverse-probability weighting by extending recent work by~\cite{Chen2019} to account time-varying propensities and measurement-error.
+ Let~$\mu(x)$ denote the posterior mean of the active infection rate individuals with covariate value~$x$ (i.e., age strata) based on the SEIR model described in SEction~\ref{section:modelbased}.
 In this context, the doubly-robust estimator is given by
@@
 
- \item {\bf Facebook symptom survey}: From the daily symptom surveys completed on Facebook (see Section~\ref{subsection:fbsymptom} for details), we extract survey responses for individuals who identify as living in Indiana.
-We collect age, sex, and demographic information as well as COVID-19 related symptoms, which Facebook defines as having (1) a fever and cough, (2) shortness of breath, \emph{or} (3) difficulty breathing.
-Table~\ref{tab:comparison} in Section~\ref{app:in_add_details} of the supplementary materials shows minimal bias with respect to symptom distributions when comparing to Indiana's random sample to the Facebook symptom survey data collected from April 25th to April 29th.
+ \item {\bf Delphi's COVID-19 Trends and Impact Survey}: From the daily symptom surveys completed as part of Delphi's CTIS (see Section~\ref{subsection:fbsymptom} for details), we extract survey responses for individuals who identify as living in Indiana.
+We collect age, sex, and demographic information as well as COVID-19 related symptoms, which Delphi defines as having (1) a fever and cough, (2) shortness of breath, \emph{or} (3) difficulty breathing.
+Table~\ref{tab:comparison} in Section~\ref{app:in_add_details} of the supplementary materials shows minimal bias with respect to symptom distributions when comparing to Indiana's random sample to the CTIS data collected from April 25th to April 29th.
 
@@
  \cline{3-5} \cline{7-8}
- & & NonRandom & Random & Facebook & Census & NonRandom & Random \\ \hline
+ & & NonRandom & Random & CTIS & Census & NonRandom & Random \\ \hline
  \multirow{2}{*}{Sex} & Female & 58.2 & 55.0 & 54.0 & 50.7 & 21.7 (11.2) & 1.4 \\
@@
  \end{tabular}
- \caption{Estimated point prevalence of active infection with SARS-CoV-2 by demographics in Indiana~\citep{Yiannoutsos2021}.}
+ \caption{Estimated total tests ($\%$) and point prevalence of active infection with SARS-CoV-2 by demographics in Indiana~\citep{Yiannoutsos2021}.}
  \label{tab:indiana}
  \end{table}
+
+\begin{remark}
+ We acknowledge several limitations with respect to the representativeness.
+First, the random sample had significant missingness ($3,658/15,495 \approx 24\%$ response rate).
+See~\cite{Yiannoutsos2021} for a discussion of how the scientific team handled missing data.
+Second, the hit rate on randomly selected Facebook users is likely low and Facebook's Indiana user population may differ from the Indiana population of interest.
+To address these concerns, the Delphi team provided respondent weights calculated by Facebook.
+See~\cite{Barkay2020} for a description of the weights and corresponding methodology.
+Further adjustments are beyond the scope of this paper but are considered important future work.
+We urge health policy experts and government officials to assess these issues when applying this framework in their own work.
+ \end{remark}
 
@@
 Subtracting off the $95,879$ tests that had already performed yields a sampling fraction of $f = 2.96 \times 10^{-3}$.
-Here we consider estimation of data quality $E_{\I} \left[ \rho_{\tilde I(X),Y} \right]$.
+Here we consider estimation of data quality $E_{\I} \left[ \rho_{I^{NR} (X),Y} \right]$.
 In \cite{Meng2018}, estimation relied on observing the true outcome (i.e., election vote totals).
@@
 This leads to a range for $M$ of $(5.60, 6.44)$ for the unweighted analysis, and $(3.63,4.07)$ and $(3.10, 3.45)$ for the weighted analyses.
+Note that in calculating the sampling fraction~$f$, we assumed individuals who tested recently are unlikely to test in this time window.
+If we instead do not subtract off these tests, the sampling fraction is $f = 2.92 \times 10^{-3}$ and the above calculations have negligible change.
 
@@
 We use these models and the case-to-hospitalization rates by demographic strata to perform mean imputation of the number of positive and negative tests given demographic strata, fever status, and COVID-19 contact status.
-Inverse probability weights are then computed based on the resulting dataset (termed IPW2).
+The use of hospitalization data to estimate the probability of fever may be subject to survivor bias as well as biases driven by differential care.
+AGain   Inverse probability weights are then computed based on the resulting dataset (termed IPW2).
 

### 2022-03-20 write-up/letters/response-ref-1-v1.tex
covid paper getting close

@@
 \vspace{5mm}
-We thank the editor-in-chief as this is an excellent point.
-We now acknowledge that both topics are well known and cite relevant literature in Section 1.1; however, while infectious disease researchers and survey statisticians have been aware of the perils of self-selection and measurement error, the interplay in this context has been less well studied.
-By extending the statistical decomposition of Meng (2018), we help shed light on how measurement error does not act to simply alter the magnitude of error but may in fact change the sign as well.
-This decomposition motivates the new inverse-probability and doubly robust methodologies in the revised manuscript.
+We thank the editor-in-chief.  We have addressed the two reviewers specific comments.  Responses to their concerns are detailed below and have significantly improved the clarify of this manuscript.
 \vspace{5mm}
@@
 \vspace{5mm}
-x
+Remark 1 has been added at the end of the introduction to contextualize the manuscript.
+Specifically we clarify that our manuscript focuses on COVID-19 case count, testing, and death data collected from April through February 2021 and therefore the numbers presented on disease dynamics are based on the original strain.
+While using the framework to account for selection bias and measurement error for data arising from the delta and omicron strains would require additional work, the framework would remain an appropriate tool for this task.
 \vspace{5mm}
@@
 Consider changing to “Statistical estimation of Coronavirus Case Counts: Measurement…” or some alternative.}
+	\vspace{5mm}
+
+	We agree with the reviewer.
+This was not updated in the prior submission simply to ensure continuity of presentation (i.e., didn’t want to change from first submission).
+Here we replace with the title “Addressing selection bias and measurement error in COVID-19 case count data using auxiliary information”.
+
 	\item {\it Organization: The figures and tables are not located anywhere near much of their introduction/discussion.
@@
 Consider the reorganization such that Figures and Tables are located near their discussion.}
+	\vspace{5mm}
+
+	Tables and figures
+	\textcolor{red}{Need to do!}
 	\item {\it Discussion on the Indiana Random Sample: On page 22, table 1, it appears that the random sample is not similar to the census estimates, potentially indicating non-response bias.}
+
+    We comment on raking techniques
+
+    Yian
+
+    See our answer to Question 5a from Reviewer \#1.
+    We now remark on the response rate in Section 5.
+Specifically, we cite how Yiannoutsos et al.
+(2021) handle non-response in prevalence estimation.
+We acknowledge the potential for a difference in characteristics between intended and final sample.
+We clarify that the Census data is Indiana specific and is presented as a reference to demonstrate the potential difference between intended and final sample.
+
 	\item {\it Additionally, there may be significant potentially non-ignorable variables (e.g., political opinions and rural status) that drive non-response and are not presented.
@@
 Much of the results appear to rely on the SRS and it ignores the potential for nonresponse bias.}
+
+	We added remark XX to discuss the issue of ran
+	Non-response bias due to unobserved missing
+
+	we see randomization as less a ``gold-standard'' and more as a starting point.
+
 \end{enumerate}
@@
 
-	See answers to Quesitons 5a and 14a above.  The prevalence estimate comes from Yiannoutsos et al. (2021) who detail their adjustments for response rates in estimation of active prevalence.
+	See answers to Questions 5a and 14a above.  The prevalence estimate comes from Yiannoutsos et al. (2021) who detail their adjustments for response rates in estimation of active prevalence.
 	\vspace{5mm}
@@
 
-	See answers to Question XX above.
-In Section 5.2, we now clarify how we calculate the infection-fatality rate.
+	See our answer to Question 12 above.
+In Section 5.2, we now clarify how we calculate the infection-fatality rate from the compartments of the model.
 Similar to Johndrow (2021), the model only uses COVID-19 related deaths.
-Deaths to non-COVID causes are not modeled but would appear in the R component.
+Deaths to non-COVID causes are not directly modeled but would appear as part of the R component in the global SEIR model.
 	\vspace{5mm}
@@
 
-    We have now acknowledged this along with the fact that IFR is time-varying.
-So this shows the IFR is likely biased (or was only accurate early).
-We performed sensitivity analysis with time-varying IFR based on estimates to improve estimation.
+    We contextualize the discussion of IFR in Remark 7 in Section 5.2.
+We acknowledge that these IFR estimates are likely biased (or potentially accurate in very different circumstances).
+A sensitivity analysis focused on the issue of choosing an appropriate IFR given the potential biases in available IFR estimates is now performed.
+    \textcolor{red}{To Do!}
 	\vspace{5mm}
@@
 This was not updated in the prior submission simply to ensure continuity of presentation (i.e., didn't want to change from first submission).
-Here we replace with the title ``Good Title''.
+Here we replace with the title ``Addressing selection bias and measurement error in COVID-19 case count data using auxiliary information''.
 We next address the reviewers specific comments.
@@
 
-We have added remark XX and more carefully put this paper within proper context.
+Remark 1 has been added at the end of the introduction to contextualize the manuscript.
+Specifically we clarify that our manuscript focuses on COVID-19 case count, testing, and death data collected from April through February 2021 and therefore the numbers presented on disease dynamics are based on the original strain.
+While using the framework to account for selection bias and measurement error for data arising from the delta and omicron strains would require additional work, the framework would remain an appropriate tool for this task.
+
 \vspace{5mm}
-\item {\it The author states: “If someone has an active infection then after the incubation period, a molecular test with perfect sensitivity will yield a positive result for the next few weeks.” (Page 5) The author states “A molecular test refers to diagnostic tests that aim to detect increased viral load” (page 5) - I think a more precise deinition is needed, for example aim to detect viral load above a certain threshold.}
+\item {\it The author states: “If someone has an active infection then after the incubation period, a molecular test with perfect sensitivity will yield a positive result for the next few weeks.” (Page 5) The author states “A molecular test refers to diagnostic tests that aim to detect increased viral load” (page 5) - I think a more precise definition is needed, for example aim to detect viral load above a certain threshold.}
 \vspace{5mm}
 
-We use Figure XX to make precise the importance of viral threshold.
+We thank the reviewer for this point. A more precise definition is provided in Section 2.1.  We also point readers to Mina et al. (2020) where the issue of testing thresholds is discussed in detail.
+
 \vspace{5mm}
@@
 
-We thank the reviewer and now match this more careful language to be more exact in our description.
+We thank the reviewer and now match this more careful language to be more exact in our description found in Section 2.
 \vspace{5mm}
@@
 \vspace{5mm}
-\item {\it XX}
 
-\vspace{5mm}
-WXX
-\vspace{5mm}
 \item {\it I’m not sure this statement “The primary goal of molecular tests is diagnosis of active infections in the population.” (page 5) is entirely true.
@@
 \vspace{5mm}
-WXX
+We apologize for the confusion.
+See answer to Question 3b where we clarify the definition of active infection.
+An active infection was meant to refer to individuals who were infected with SARS-CoV-2 who have yet to fall below a threshold for detectable viral load levels by RT-PCR testing.
+This includes the infectious period as well as a pre-infectious and post-infectious periods of time.
+
 \vspace{5mm}
@@
 \vspace{5mm}
-WXX
+We clarify in Section 3.2 that this is the peak in the fraction of infected individuals in the population which is presented in Figure 4a.
 \vspace{5mm}
@@
 \vspace{5mm}
-WXX
+See our answer to Question 17b of Reviewer \#1.
+We acknowledge that these IFR estimates are likely biased and time-varying due to introduction of vaccines in 2020.
+A sensitivity analysis focused on the issue of choosing an appropriate IFR and allowing for the IFR to be time-varying is now performed.
+\textcolor{red}{To Do!}
+
 \vspace{5mm}
-\item {\it I found Figure 3 in the supplement conusing.
+\item {\it I found Figure 3 in the supplement [confusing].
 Figure 3A shows the profile for a 35-44 white male with symptoms and COVID-19 contact and Figure 3B shows the profile for a 65-74 white female without symptoms nor COVID-19 contact – I would have expected the covariates between these (age, ethnicity, race, gender) to be the same rather than differ with the only thing differing being symptoms 1/ COVID-19 contact?
@@
 \vspace{5mm}
-WXX
+We have fixed Figure 3b to match on covariates and have matching y-axes.  See our answer to Question 16b from Reviewer \#1 on additional figures we are including in the revision.
+\textcolor{red}{Need to do.}
 \vspace{5mm}
@@
 \vspace{5mm}
-WXX
+See our answer to Question 17b of Reviewer #1 and Question 7 above. A sensitivity analysis focused on the issue of choosing an appropriate IFR is now performed.
+\textcolor{red}{To Do!}
 \vspace{5mm}

### 2022-03-20 write-up/covid-writeup.tex
covid paper getting close

@@
 \begin{remark}[An evolving pandemic]
-
+This manuscript focuses on COVID-19 case count, testing, and death data collected from April through February 2021.
+Numbers presented on disease dynamics are therefore based on the original strain.
+Selection bias and measurement error persist in data arising from the delta and omicron strains and will likely persist for future variants.
+While not discussed in this manuscript, the framework presented will remain an appropriate tool for addressing selection bias and measurement error in these settings.
 \end{remark}
@@
 The viral load will be detectable by at least the end of the latent period (time to infectiousness), which for SARS-CoV-2 occurs before the end of the incubation period.
-A \emph{molecular test} refers to diagnostic tests that aim to detect viral load above a certain threshold (e.g., RT-PCR or antigen tests).
-If someone has an active infection -- here defined as individuals infected with SARS-CoV-2 who's viral load has yet to fall below detectable levels by molecular test -- then after the incubation period, a molecular test with perfect sensitivity will yield a positive result for the next few weeks.
+A \emph{molecular test} refers to diagnostic tests that aim to detect viral load above a certain threshold (e.g., RT-PCR or antigen tests); see~\cite{Mina2020} for a detailed discussion of cycle thresholds.
+If someone has an active infection -- \textcolor{red}{here defined as individuals infected with SARS-CoV-2 who's viral load has yet to fall below detectable levels by molecular test} -- then after the incubation period, a molecular test with perfect sensitivity will yield a positive result for the next few weeks.
 After that, the viral load will decrease and a molecular test with perfect specificity will come back negative.
@@
 The SEIR model has been used extensively as a model for SARS-CoV-2 dynamics~\cite{Song2020}.
-In terms of bias, the rate is overestimated prior to the peak and underestimated afterwards; the bias increases dramatically when the relative fraction exceeds $2$.
+In terms of bias, the rate is overestimated prior to the peak in the fraction infected and underestimated afterwards; the bias increases dramatically when the relative fraction exceeds $2$.
 Such biases may impact policy making.
@@
 
- \subsection{Estimation of effective reproduction number}
- \label{section:r0-estimation}
- Many epidemiologists argue that tracking the effective reproduction number is the only way to manage through the crisis~\citep{Gabriel2020}.
-Here, we study the instantaneous reproduction number~\citep{Cori20113,Fraser2007}, denoted $R_t$, which is the average number of secondary cases that each infected individual would infect if the conditions remained as they were at time~$t$.
-This is distinct from the case reproduction number,~$R_t^c$, which is the average number of secondary cases that a case infected at time step $t$ will eventually infect~\citep{Wallinga2004}.
-The case reproduction number accounts for potential changes to contact rates and transmissibility, which include impact of control measures.
-The instantaneous reproduction number~$R_t$ is the only reproduction number easily estimated in real time, and therefore has been a key focal point in the COVID-19 pandemic.
-
-Under a Poisson likelihood, a simple relation between the trajectory of new cases and the instantaneous reproduction number can be derived \citep{Bettencourt2008}.
-In particular, under an SIR model the number of case counts on day $t$, denoted $K_t$, is Poisson distributed with rate $K_{t-1} \exp \left( \gamma (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval, which is approximately $7$ days for COVID-19~\citep{Sanche2020}.
-
-\cite{Heng2020} derive an approximate formula for the basic reproduction number under SEIR dynamics such that for $S_0 \approx 1$ and $I_0 \ll 1$, we have the approximate realtionship, which ``corresponds to [...] the solution for the SIR model assuming an infectious period of $1/\gamma + 1/\sigma$.
-Using this connection, a moment-based estimator is given by
-$$
-R_t \approx 1 + \left( \frac{1}{\gamma} + \frac{1}{\sigma} \right) \log \left( \frac{K_t}{K_{t-1}} \right).
-$$
-This approximation ignores a term that is quadratic in the epidemic growth rate and inversely linear in $\sigma \times \gamma$; for realistic values of these quantities, however, the term is relatively negligible and therefore ignored.
-Of course, we do not observe $K_t$ and $K_{t-1}$.
-Under SRS of new cases among those susceptible on day $t$, the natural estimator is $S_t \tilde y_t$.
-Unfortunately the number of susceptible individuals on day $t$ is unknown.
-Here, we study the estimator $\hat R_t = 1 + \frac{1}{\gamma} \log \left( \tilde y_t / \tilde y_{t-1} \right)$.
-We can again express the statistical error of $\hat R_t - R_t$ in useful terms as follows
- $$
- \begin{aligned}
- \left( \frac{1}{\gamma} + \frac{1}{\sigma} \right)\log &\bigg( 1 + \bigg[ \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t)  -\rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (K_{t-1}) \bigg] \\
- &\times \left[ 1 - \rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (K_{t-1}) \right] \bigg) - \left( \frac{1}{\gamma} + \frac{1}{\sigma} \right) \log \left( \frac{S_t}{S_{t-1}} \right).
- \end{aligned}
- $$
- This implies a similar trade-off as before but on the logarithmic scale.
-The error is no longer scaled by $\bar Y_t/\bar Y_{t-1}$ but by the serial interval and does not depend on prevalence but on the fraction of new cases out of those susceptible.
-This leads to differences in when the bias is most pronounced.
-Figure~\ref{fig:r0-bias} displays the bias as a function of the relative sampling fraction ignoring the final term (i.e., $S_t / S_{t-1} \approx 1$).
-Section~\ref{app:cori_rt} in the supplementary materials presents an alternative estimator of the instantaneous effective reproductive number and shows similar bias~\cite{Cori20113}.
-
- \begin{figure}
+ \begin{figure}[!th]
  \centering
@@
  \end{figure}
+
+ \subsection{Estimation of effective reproduction number}
+ \label{section:r0-estimation}
+ Many epidemiologists argue that tracking the effective reproduction number is the only way to manage through the crisis~\citep{Gabriel2020}.
+Here, we study the instantaneous reproduction number~\citep{Cori20113,Fraser2007}, denoted $R_t$, which is the average number of secondary cases that each infected individual would infect if the conditions remained as they were at time~$t$.
+This is distinct from the case reproduction number,~$R_t^c$, which is the average number of secondary cases that a case infected at time step $t$ will eventually infect~\citep{Wallinga2004}.
+The case reproduction number accounts for potential changes to contact rates and transmissibility, which include impact of control measures.
+The instantaneous reproduction number~$R_t$ is the only reproduction number easily estimated in real time, and therefore has been a key focal point in the COVID-19 pandemic.
+
+Under a Poisson likelihood, a simple relation between the trajectory of new cases and the instantaneous reproduction number can be derived \citep{Bettencourt2008}.
+In particular, under an SIR model the number of case counts on day $t$, denoted $K_t$, is Poisson distributed with rate $K_{t-1} \exp \left( \gamma (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval, which is approximately $7$ days for COVID-19~\citep{Sanche2020}.
+
+\cite{Heng2020} derive an approximate formula for the basic reproduction number under SEIR dynamics such that for $S_0 \approx 1$ and $I_0 \ll 1$, we have the approximate realtionship, which ``corresponds to [...] the solution for the SIR model assuming an infectious period of $1/\gamma + 1/\sigma$.
+Using this connection, a moment-based estimator is given by
+$$
+R_t \approx 1 + \left( \frac{1}{\gamma} + \frac{1}{\sigma} \right) \log \left( \frac{K_t}{K_{t-1}} \right).
+$$
+This approximation ignores a term that is quadratic in the epidemic growth rate and inversely linear in $\sigma \times \gamma$; for realistic values of these quantities, however, the term is relatively negligible and therefore ignored.
+Of course, we do not observe $K_t$ and $K_{t-1}$.
+Under SRS of new cases among those susceptible on day $t$, the natural estimator is $S_t \tilde y_t$.
+Unfortunately the number of susceptible individuals on day $t$ is unknown.
+Here, we study the estimator $\hat R_t = 1 + \frac{1}{\gamma} \log \left( \tilde y_t / \tilde y_{t-1} \right)$.
+We can again express the statistical error of $\hat R_t - R_t$ in useful terms as follows
+ $$
+ \begin{aligned}
+ \left( \frac{1}{\gamma} + \frac{1}{\sigma} \right)\log &\bigg( 1 + \bigg[ \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t)  -\rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (K_{t-1}) \bigg] \\
+ &\times \left[ 1 - \rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (K_{t-1}) \right] \bigg) - \left( \frac{1}{\gamma} + \frac{1}{\sigma} \right) \log \left( \frac{S_t}{S_{t-1}} \right).
+ \end{aligned}
+ $$
+ This implies a similar trade-off as before but on the logarithmic scale.
+The error is no longer scaled by $\bar Y_t/\bar Y_{t-1}$ but by the serial interval and does not depend on prevalence but on the fraction of new cases out of those susceptible.
+This leads to differences in when the bias is most pronounced.
+Figure~\ref{fig:r0-bias} displays the bias as a function of the relative sampling fraction ignoring the final term (i.e., $S_t / S_{t-1} \approx 1$).
+Section~\ref{app:cori_rt} in the supplementary materials presents an alternative estimator of the instantaneous effective reproductive number and shows similar bias~\cite{Cori20113}.
 
@@
 
+ \begin{remark}[Sensitivity to IFR]
+
+ \end{remark}
+
   \begin{figure}

### 2022-03-21 write-up/covid-supplement-writeup.tex
New Figures for Supplement

@@
 \includegraphics[width=.9\linewidth]{../figs/tvprop_fig2.png}
- \caption{Testing Propensity}
+ \caption{Testing Likelihood Given Fever and COVID-19 Contact}
  \label{fig:testinglik2}
@@
 We see that the likelihood is time-varying and depends heavily on both fever status.
+
+\subsubsection{Propensity plots}
+
+Test
+
+\begin{figure}[!th]
+\centering
+\begin{subfigure}{.5\textwidth}
+ \centering
+ \includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig1_supp4.png}
+ \caption{Symptom Likelihood Given Negative Test}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+ \centering
+\includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig2_supp4.png}
+ \caption{Symptom Likelihood Given Positive Test}
+\end{subfigure}
+\caption{Likelihood of COVID-19 contact for Non-Hispanic, White Females}
+\end{figure}
+
+\begin{figure}[!th]
+\centering
+\begin{subfigure}{.5\textwidth}
+ \centering
+ \includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig1_supp1.png}
+ \caption{Symptom Likelihood Given Negative Test}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+ \centering
+\includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig2_supp1.png}
+ \caption{Symptom Likelihood Given Positive Test}
+\end{subfigure}
+\caption{Likelihood of COVID-19 contact for Non-Hispanic, White Females}
+\end{figure}
+
+\begin{figure}[!th]
+\centering
+\begin{subfigure}{.5\textwidth}
+ \centering
+ \includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig1_supp2.png}
+ \caption{Symptom Likelihood Given Negative Test}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+ \centering
+\includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig2_supp2.png}
+ \caption{Symptom Likelihood Given Positive Test}
+\end{subfigure}
+\caption{Likelihood of COVID-19 contact for Non-Hispanic, White Females}
+\end{figure}
+
+\begin{figure}[!th]
+\centering
+\begin{subfigure}{.5\textwidth}
+ \centering
+ \includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig1_supp3.png}
+ \caption{Symptom Likelihood Given Negative Test}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+ \centering
+\includegraphics[width=.9\linewidth]{../figs/tvprop_alt_fig2_supp3.png}
+ \caption{Symptom Likelihood Given Positive Test}
+\end{subfigure}
+\caption{Likelihood of COVID-19 contact for Non-Hispanic, White Females}
+\end{figure}
+
+\newpage
+\subsubsection{Confidence Intervals}
 

### 2022-03-21 visualizations/tv_propensities.R

@@ -2,7 +2,7 @@ library(lubridate)
 library(MMWRweek)
 library(ggplot2)
 
-propensities = readRDS("../data/smoothedpropensities_071521.RDS")
+propensities = readRDS("../data/smoothedpropensities_082621.RDS")
 
 propensities$date = MMWRweek::MMWRweek2Date(MMWRyear = propensities$year,
                                      MMWRweek = propensities$week,
@@ -24,8 +24,8 @@ ggplot(data = propensities, aes(x = date, y = probs)) +
 
 dev.off()
 
-## Example propensity: Not Fever + Not Contact + Not H + White + Male + 65to74
-x = as.matrix(c(1,0,0,1,1,0,0,0,1,0,0,1,0,0), ncol = 1)
+## Example propensity: Not Fever + Not Contact + Not H + White + Male + 35to44
+x = as.matrix(c(1,0,0,0,1,0,0,1,0,1,0,0,0,0), ncol = 1)
 propensities$probs = as.vector(1/(1+exp(-as.matrix(propensities[,3:16])%*%x)))
 
 
@@ -35,8 +35,8 @@ png(filename = "../figs/tvprop_fig2.png",
 ggplot(data = propensities, aes(x = date, y = probs)) +
   geom_point(size = 5) +
   labs(x = "Date",
-       y = "Testing Propensity",
-       title = "65-74, non-Hispanic, White Female without symptoms nor COVID-19 Contact") +
+       y = "COVID-19 Testing Propensity",
+       title = "35-44, non-Hispanic, White Male without symptoms nor COVID-19 Contact") +
   theme(text = element_text(size=15))
 
 dev.off()

### 2022-03-21 visualizations/tv_propensities_alt.R

@@ -143,3 +143,391 @@ dev.off()
 
 
 propensities_long_fever$propensity/propensities_long_withoutfever$propensity
+
+## APPENDIX VERSIONS OF MAIN PAPER PLOT
+
+## NotHoL, AFrican American
+
+## Fig1
+x_age1 = as.matrix(c(0,1,0,0,1,0,0,1,0,0,0,0,0), ncol = 1)
+x_age2 = as.matrix(c(0,1,0,0,1,0,0,0,1,0,0,0,0), ncol = 1)
+x_age3 = as.matrix(c(0,1,0,0,1,0,0,0,0,1,0,0,0), ncol = 1)
+x_age4 = as.matrix(c(0,1,0,0,1,0,0,0,0,0,1,0,0), ncol = 1)
+x_age5 = as.matrix(c(0,1,0,0,1,0,0,0,0,0,0,1,0), ncol = 1)
+x_age6 = as.matrix(c(0,1,0,0,1,0,0,0,0,0,0,0,1), ncol = 1)
+propensities$probs1 = as.vector(1/(1+exp(-as.matrix(propensities[,3:15])%*%x_age1)))
+propensities$probs2 = as.vector(1/(1+exp(-as.matrix(propensities[,3:15])%*%x_age2)))
+propensities$probs3 = as.vector(1/(1+exp(-as.matrix(propensities[,3:15])%*%x_age3)))
+propensities$probs4 = as.vector(1/(1+exp(-as.matrix(propensities[,3:15])%*%x_age4)))
+propensities$probs5 = as.vector(1/(1+exp(-as.matrix(propensities[,3:15])%*%x_age5)))
+propensities$probs6 = as.vector(1/(1+exp(-as.matrix(propensities[,3:15])%*%x_age6)))
+
+propensities_long = rbind(propensities, propensities, propensities,
+                          propensities, propensities, propensities)
+
+propensities_long$age = c(rep("25to34", nrow(propensities)),
+                          rep("35to44", nrow(propensities)),
+                          rep("45to54", nrow(propensities)),
+                          rep("55to64", nrow(propensities)),
+                          rep("65to74", nrow(propensities)),
+                          rep("75plus", nrow(propensities))
+)
+
+propensities_long$propensity = propensities_long$probs1 * (propensities_long$age == "25to34") +
+  propensities_long$probs2 * (propensities_long$age == "35to44") +
+  propensities_long$probs3 * (propensities_long$age == "45to54") +
+  propensities_long$probs4 * (propensities_long$age == "55to64") +
[…]

### 2022-03-21 write-up/letters/response-ref-1-v1.tex
only seir fixes to go

@@
 
-	Tables and figures
+	We apologize.
+We kept figures/tables related to the case study in Section 5 (the case study section) but can now see this led to confusion as we discuss them early in the paper.
+Tables and figures have been moved to be located much closer to their introduction.
 \textcolor{red}{Need to do!}
+
 	\item {\it Discussion on the Indiana Random Sample: On page 22, table 1, it appears that the random sample is not similar to the census estimates, potentially indicating non-response bias.}
-
-    We comment on raking techniques
-
-    Yian
 
@@
 We acknowledge the potential for a difference in characteristics between intended and final sample.
-We clarify that the Census data is Indiana specific and is presented as a reference to demonstrate the potential difference between intended and final sample.
 
@@
 
-	We added remark XX to discuss the issue of ran
-	Non-response bias due to unobserved missing
-
-	we see randomization as less a ``gold-standard'' and more as a starting point.
-
+	We do not see random sampling as a ``gold-standard'' nor a panacea that will fix every issue.
+Instead, we see it as a means to remove one very important source of potential bias in the estimation of population quantities of interest.
+A key takeaway of our paper is that with access to \emph{only} a non-probabilistic sample, we know there is potential bias but have little hope of reducing said bias.
+With a probability sample, we have means to reduce said bias.
+Statisticians, epidemioligists, and government officials, and health policy experts can then focus on practical issues such as non-ignorable missingness.
+If we thought, for example, political opinions and rural status are important predictors of missingness, then these should be collected as part of the random sampling protocol.
+Note, even collecting strata information that is correlates with testing propensity (e.g., symptom status) in the non-probabilistic sample is insufficient to address selection bias.
+This is not true for probability samples.
+We have added this important discussion to Section 6.
 \end{enumerate}
 
-\vspace{5mm}
-Thank you for raising this issue.
-We discuss reporting delays and the relevant literature in Remark 1 in Section 2.2.
-Given currently available public data, one cannot correct at the daily level for reporting delays.
-Instead, we analyze COVID-19 data at the weekly level to account for potential reporting delays.
-While not a perfect solution, it avoids clear reporting variation across day-of-week which can be seen in Figure 5.
-\vspace{5mm}
 \end{enumerate}
@@
 
-	We acknowledge this shortcoming.  We now include versions of Figure 6 for non-white groups in Section XX of the supplementary materials.
-	\textcolor{red}{NEED TO DO}
+	We acknowledge this shortcoming.
+We now include versions of Figure 6 for other groups in Section H.2.1 of the supplementary materials with a discussion of the relative propensities.
+Specifically, as the reviewer states above, individuals who identify as Hispanic, Male, and select `Some other race' were twice as likely to present for testing as non-Hispanic, white men -- empirically confirming that our data captures these systematic differences.
 	\vspace{5mm}
@@
 \vspace{5mm}
-See our answer to Question 17b of Reviewer #1 and Question 7 above. A sensitivity analysis focused on the issue of choosing an appropriate IFR is now performed.
+See our answer to Question 17b of Reviewer \#1 and Question 7 above. A sensitivity analysis focused on the issue of choosing an appropriate IFR is now performed.
 \textcolor{red}{To Do!}

### 2022-03-21 write-up/covid-supplement-writeup.tex
only seir fixes to go

@@
 \subsubsection{Propensity plots}
-
-Test
+\label{section:propplots}
+
+Here we present the testing propensity for four strata: (1) non-Hispanic, white female, (2) non-Hispanic, African American male, (3) Hispanic, White male.
+Strata (1) presented in Figure~\ref{fig:nonh-white-female} demonstrates a small gender difference in testing propensities.
+Strata (2) and (3) presented in Figures~\ref{fig:nonh-aa-male} and~\ref{fig:h-white-male} demonstrates a lower rate of testing among African Americans and Hispanic white men compared to non-Hispanic, white men respectively (approximately 2 times lower).
+Strata (4) presented in Figure~\ref{fig:h-other-male} demonstrates a much higher testing rate for Hispanic men who select ``Some Other Race''.
 
@@
 \caption{Likelihood of COVID-19 contact for Non-Hispanic, White Females}
+\label{fig:nonh-white-female}
 \end{figure}
@@
 \end{subfigure}
-\caption{Likelihood of COVID-19 contact for Non-Hispanic, White Females}
+\caption{Likelihood of COVID-19 contact for Non-Hispanic, African American Males}
+\label{fig:nonh-aa-male}
 \end{figure}
@@
 \end{subfigure}
-\caption{Likelihood of COVID-19 contact for Non-Hispanic, White Females}
+\caption{Likelihood of COVID-19 contact for Hispanic, White Males}
+\label{fig:h-white-male}
 \end{figure}
@@
 \end{subfigure}
-\caption{Likelihood of COVID-19 contact for Non-Hispanic, White Females}
+\caption{Likelihood of COVID-19 contact for Hispanic, Males who select `Some Other Race'.}
+\label{fig:h-other-male}
 \end{figure}

### 2022-03-21 write-up/covid-writeup.tex
only seir fixes to go

@@
 \begin{remark}[An evolving pandemic]
-This manuscript focuses on COVID-19 case count, testing, and death data collected from April through February 2021.
+This paper focuses on COVID-19 case count, testing, and death data collected from April through February 2021.
 Numbers presented on disease dynamics are therefore based on the original strain.
 Selection bias and measurement error persist in data arising from the delta and omicron strains and will likely persist for future variants.
-While not discussed in this manuscript, the framework presented will remain an appropriate tool for addressing selection bias and measurement error in these settings.
+While not discussed in this paper, the framework presented will remain an appropriate tool for addressing selection bias and measurement error in these settings.
 \end{remark}
@@
 
-Here we provide the necessary background to understand COVID-19 diagnostic testing, its scientific use in managing the pandemic, and the data streams considered in this manuscript.
+Here we provide the necessary background to understand COVID-19 diagnostic testing, its scientific use in managing the pandemic, and the data streams considered in this paper.
 
@@
 
- The primary goal of this manuscript is to produce accurate estimates of the population-level active infection rates over time using \emph{publicly available} viral testing data.
+ The primary goal of this paper is to produce accurate estimates of the population-level active infection rates over time using \emph{publicly available} viral testing data.
 Secondary goals include estimation of rates of change and the effective reproduction number which characterize disease trajectory.
@@
 In December 2020, the request was officially approved and data streams were made publicly available on the IDH.
-In this manuscript, we thus focus on this \emph{publicly-available data}, i.e., data  that anyone can download directly from the IDH.
+In this paper, we thus focus on this \emph{publicly-available data}, i.e., data  that anyone can download directly from the IDH.
 While the scientific community has contributed through independent COVID-19 observational and clinical studies, a significant component of public policy guidance has relied on testing and case count data, e.g., CDC and state guidelines based on test positivity rates and relative changes in the case counts over time.
@@
  $$
- \frac{E_{\I} \left[ (\bar y_n - \bar Y)^2 \right]}{\sqrt{V_{SRS} (\bar Y)}} = (N-1) E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right]
- $$
- implying the error relative to SRS increases as a function of population size~\citep{Meng2018}.
+ \frac{E_{\I} \left[ (\bar y_n - \bar Y)^2 \right]}{\sqrt{V_{SRS} (\bar Y)}} = (N-1) E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right].
+ $$
+ % implying the error relative to SRS increases as a function of population size~\citep{Meng2018}.
 That two countries with the same testing strategy can yield wildly different estimates due to population size is too often ignored in current communications of case-count data.
- % Thus, conclusions drawn from observed case-counts may not just be wrong, but very wrong.
-
+ % % Thus, conclusions drawn from observed case-counts may not just be wrong, but very wrong.
  A key question is ``What is the (effective) sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' In Appendix~\ref{app:effss}, we show the effective sample size $n_{eff}$ can be bounded by
@@
 Then the effective sample size is $863$, representing a $5.1$ factor increase rather than the expected $6.1$ factor increase.
-See Section~\ref{section:effss} in the Supplementary Materials for additional effective sample size calculations.
+See Section~\ref{section:effss} in the Supplementary Materials for additional effective sample size calculations which show that while~\citep{Meng2018} argues the relative error increases as a function of population size, our calculations indicate this is not true for fixed relative frequency~$f_1/f_0$.
  % \begin{wrapfigure}{r}{0.5\textwidth}
@@
 Note that the ratio of testing propensity within strata across fever status is time-varying, starting 50 times more likely in early April and dropping to ten times more likely by end of 2020.
+Similar plots for other strata and a discussion of their relative testing propensities can be found in Section~\ref{section:propplots} of the supplementary materials.
 
@@
 
+\textcolor{red}{
+ While this paper emphasizes the importance of random sampling, we do not naively view it as a panacea.
+With access to \emph{only} a non-probabilistic sample, we have no means to reduce self-selection bias.
+Adding a probability sample gives us the necessary auxiliary information to do so.
+This leaves the data analyst more time to focus on statistical issues related to the random sample such as how to handle non-response bias.
+Covariates that the scientific team thinks are correlated with non-response (e.g., political opinion) should be collected as part of the random sampling protocol and corrected for in the data analysis.
+On the other hand, collecting covariate information that correlates with testing propensity (e.g., symptom status) in the non-probabilistic sample is insufficient to address selection bias.
+}
+
 % \subsection*{Higher quality information on prevalence.}
@@
 
- \textcolor{red}{I want to say that if i have many more random samples, then we should bias towards the random samples and not only smooth the means. But this is future work.}
+ % \textcolor{red}{I want to say that if i have many more random samples, then we should bias towards the random samples and not only smooth the means. But this is future work.}
 

### 2022-03-21 write-up/letters/response-ref-1-v1.tex
Almost

@@
 Remark 1 has been added at the end of the introduction to contextualize the manuscript.
-Specifically we clarify that our manuscript focuses on COVID-19 case count, testing, and death data collected from April through February 2021 and therefore the numbers presented on disease dynamics are based on the original strain.
+Specifically we clarify that our manuscript focuses on COVID-19 case count, testing, and death data collected from April 2020 through February 2021 and therefore the numbers presented on disease dynamics are based on the original strain.
 While using the framework to account for selection bias and measurement error for data arising from the delta and omicron strains would require additional work, the framework would remain an appropriate tool for this task.
@@
 
-	We agree with the reviewer.
+	We agree with the AE.
 This was not updated in the prior submission simply to ensure continuity of presentation (i.e., didn’t want to change from first submission).
-Here we replace with the title “Addressing selection bias and measurement error in COVID-19 case count data using auxiliary information”.
+Here we replace with the title ``Addressing selection bias and measurement error in COVID-19 case count data using auxiliary information.''
+	\vspace{5mm}
 
@@
 Tables and figures have been moved to be located much closer to their introduction.
-\textcolor{red}{Need to do!}
+	\vspace{5mm}
 
 	\item {\it Discussion on the Indiana Random Sample: On page 22, table 1, it appears that the random sample is not similar to the census estimates, potentially indicating non-response bias.}
+	\vspace{5mm}
 
@@
 We acknowledge the potential for a difference in characteristics between intended and final sample.
+    \vspace{5mm}
 
@@
 Much of the results appear to rely on the SRS and it ignores the potential for nonresponse bias.}
+	\vspace{5mm}
 
@@
 If we thought, for example, political opinions and rural status are important predictors of missingness, then these should be collected as part of the random sampling protocol.
-Note, even collecting strata information that is correlates with testing propensity (e.g., symptom status) in the non-probabilistic sample is insufficient to address selection bias.
+Note, even collecting strata information that correlates with testing propensity (e.g., symptom status) in the non-probabilistic sample is insufficient to address selection bias.
 This is not true for probability samples.
@@
 
-I am incredibly grateful for this review. (PRAISE OUT THE WAZZOOO).
+I am very grateful for this review.  The detailed suggestions and comments helped improve the paper in numerous ways.  See below for our responses to each particular reviewer comment.
 
@@
 
-	We clarify that we define active infection as individuals infected with SARS-CoV-2 who have yet to fall below a threshold for detectable viral load levels by RT-PCR testing.
+	\textcolor{red}{We clarify that we define active infection as individuals infected with SARS-CoV-2 who have yet to fall below a threshold for detectable viral load levels by RT-PCR testing.}
 
@@
 	We have added Remark 3 in Section 2.2 discussing public versus government data.
-Specifically, we clarify that data privacy restrictions enforce publicly released testing and case count data be aggregated by strata-level covariates and have potentially important missing information (i.e., symptom status and )
-	While these choices impact our data analytic plan, we clarify that the data analytic framework is expressly designed to guide health departments and policy makers to assess the pandemic and guide future resposnes.
+Specifically, we clarify that data privacy restrictions enforce publicly released testing and case count data be aggregated to strata-level covariates and have potentially important missing information (i.e., symptom status and indicator of contact with COVID-19 positive individual).
+	While these choices impact our data analytic plan, we clarify that the data analytic framework is expressly designed to guide health departments and policy makers to assess the pandemic and guide future responses.
 
@@
 
-	Section 2.2.1 has a reconstructed timeline of restrictions, which was based on reported restrictions in the local Indiana news.
-Between May 12th and April 28th, only high-risk individuals (symptomatic or having one of several important risk factors) were able to test.
-After April 28th, suspected exposure was included as a reason for testing.
-We could not find any article mention travel history as a factor for testing in the public reports.
+	Section 2.2.1 has a reconstructed timeline of restrictions, which was based on news reports of state-level testing restrictions.
 	\vspace{5mm}
@@
 
-	XXXXX
+	\textcolor{red}{We clarify that we define active infection as individuals infected with SARS-CoV-2 who have yet to fall below a threshold for detectable viral load levels by RT-PCR testing.}
 	\vspace{5mm}
@@
 
-	We clarify that this notation was chosen to match standard survey statistics notation: $Y_i$ is the true, fixed outcome value and $y_i$ is the  population-level .
-To help the reader, we now include an appendix with all the notation described in detail.
+	We clarify that this notation was chosen to match standard survey statistics notation: $Y_i$ is the true, fixed value for individual $i$ from the population, while $y_i$ is the observed value for individual $i$ from the sample.
+To help the reader, we now include a notation glossary in Appendix B.
 	\vspace{5mm}
@@
 	This is now stated explicitly in Section 3, paragraph 2.
+	\vspace{5mm}
 \end{enumerate}
@@
 	The adjustment is exactly related to the Bayes rule PPV/NPV relationship with FP, FN, and $\bar Y$.
+	\textcolor{red}{Got to here.}
 	\vspace{5mm}

### 2022-03-21 write-up/covid-writeup.tex
Almost

@@
 
+\begin{remark}[An evolving pandemic]
+This paper focuses on COVID-19 case count, testing, and death data collected from April 2020 through February 2021.
+Numbers presented on disease dynamics are therefore based on the original strain.
+Selection bias and measurement error persist in data arising from the delta and omicron strains and will likely persist for future variants.
+While not discussed in this paper, the framework presented will remain an appropriate tool for addressing selection bias and measurement error in these settings.
+\end{remark}
+
 \subsection{Related work}
@@
 The proposed approach is distinct from the use of validation studies~\citep{Fox2020} -- a traditional epidemiologic method in which investigators compare the accuracy with gold standard measure to mitigate bias.
-Here, the \emph{gold standard} of a different test is replaced by a less biased selection mechanism.
+Here, the \emph{gold standard} of a different test is replaced by a sample with a less biased selection mechanism.
 
@@
 In this paper, rather than focusing on construction of a study sample, we ask whether one can collect auxiliary information to address selection bias in the observed case count data directly.
-This question is related recent work on \emph{target validity}~\citep{Westreich2018}, in which the issues of internal and external validity are jointly addressed  with resepct to a specific population of interest.
-This article is a concrete attempt to address both types of validity and extends the conversation on target validity within the context of analysis of observational studies of COVID-19.
-
-\begin{remark}[An evolving pandemic]
-This paper focuses on COVID-19 case count, testing, and death data collected from April through February 2021.
-Numbers presented on disease dynamics are therefore based on the original strain.
-Selection bias and measurement error persist in data arising from the delta and omicron strains and will likely persist for future variants.
-While not discussed in this paper, the framework presented will remain an appropriate tool for addressing selection bias and measurement error in these settings.
-\end{remark}
+This question is related to recent work on \emph{target validity}~\citep{Westreich2018}, in which the issues of internal and external validity are jointly addressed with respect to a specific population of interest.
+This article is a concrete attempt to address both types of validity and extends the conversation on target validity within the context of analysis of observational COVID-19 studies.
 
@@
 Systematic reviews suggest that 87\% sensitivity and 97.6\% specificity are reasonable estimates for current RT-PCR tests~\citep{Arevalo2020, Woloshin2020,Cohen2020}.
-\textcolor{red}{To the best of our knowledge, both Indiana Department of Health's molecular testing and the random state-wide sample relied primarily on nasopharyngeal swab using the same set of RT-PCR tests.}
+To the best of our knowledge, both Indiana Department of Health's molecular testing and the random state-wide sample relied primarily on nasopharyngeal swab using the same set of RT-PCR tests.
 
@@
 
- Public databases most often, however, only contain aggregate information.
-The Johns Hopkins dashboard, for example, provides demographic breakdown of case counts as well as the total confirmed cases and deaths by county.
-Aggregate time series of case count and deaths can also be extracted.
-Unfortunately, most dashboards do not provide demographic information on who requested a test nor on who tested positive for SARS-CoV-2 \emph{over time}.
-Working closely with the State of Indiana, we were able to obtain access to COVID-19 total tests, positive tests, and related deaths per day broken out marginally by age, gender, ethnicity, and race~\citep{IndianaData2021}.
-These granular datasets are now publicly available and motivate the proposed approach.
- % https://www.regenstrief.org/covid-dashboard/
- %
-
-\begin{remark}[Reporting Delays]
-Figure~\ref{fig:in-data} shows clear reduced testing and case count reporting on weekends compared to weekdays.
-Moreover, COVID-19 tests are reported on the day they were administered, while case counts are reported based on the date the positive test was reported to and confirmed by the Indiana Department of Health system.
-To minimize the impact of reporting, testing and case-count data are aggregated at the weekly level for analysis.
- \end{remark}
-
-\begin{remark}[Public versus government datasets]
-In September 2020, the author submitted a data request to the Indiana Data Hub (IDH).
-Over the next three months, we worked with Indiana's COVID-19 response team to refine the request.
-In December 2020, the request was officially approved and data streams were made publicly available on the IDH.
-In this paper, we thus focus on this \emph{publicly-available data}, i.e., data  that anyone can download directly from the IDH.
-While the scientific community has contributed through independent COVID-19 observational and clinical studies, a significant component of public policy guidance has relied on testing and case count data, e.g., CDC and state guidelines based on test positivity rates and relative changes in the case counts over time.
-Addressing selection bias and measurement-error in these public datasets is imperative in better informing public policy debate.
-
-Note that the COVID-19 response team (and health departments in general) have access to \emph{government data} which may not be publicly-available and is likely stored at the individual-level.
-Due to data privacy and legal concerns, some collected covariates may not be reported publicly and others are aggregated and not reported jointly.
-While data analytic decisions made in Sections XX--XX were made due to access of publicly-available data, the overall data analysis framework is designed to be a tool for health departments to assess the pandemic and guide responses, e.g.,  account for individual-level covariate information, rather than publicly-reported strata-level information, that may include important covariates such as symptom status and recent COVID-19 contact.
- \end{remark}
-
- %https://www.npr.org/sections/goatsandsoda/2020/05/31/865932474/should-i-get-tested-for-coronavirus-just-for-the-heck-of-it
-
- \subsubsection{Testing restrictions and public health policy in Indiana}
- \label{section:publicpolicyindiana}
-
- Due to limited testing capacity, many US states instituted testing restrictions early on in the pandemic.
- % on who can request a molecular test at a given time.
- %https://www.coronavirus.in.gov/
- % Early on, limited testing meant that only those at high risk were allowed to be tested (i.e., symptomatic or high risk populations); as testing capacity expanded, these restrictions were gradually lifted.
- Here, we reconstruct the testing restriction history for the state of Indiana.
-As of May 12th, 2020, the Indiana State Department of Health (ISDH) laboratory was testing high-risk individuals for COVID-19.
-Any symptomatic individual could receive a test.
-Moreover there were key factors for people who can be tested regardless of symptom status: over the age of 65, diabetic, high-blood pressure, obesity, pregnancy, and minority populations at greater risk of severe illness~\citep{indystar2020}.
- % https://www.indystar.com/story/news/health/2020/05/12/coronavirus-testing-indiana-who-should-get-tested/3110592001/
- On April 28th, 2020, the criteria expanded to include any Indiana resident with virus symptoms, people in close contact with those who have tested positive and residents of congregant communities~\citep{wishtv2020}.
- % https://www.wishtv.com/news/medical/indiana-opens-up-covid-19-testing-to-more-hoosiers/
- On June 15th, 2020, ISDH lifted all restrictions and testing expanded to include anyone in Indiana who wants to be tested for coronavirus~\citep{indystar2020v2}.
- % https://www.indystar.com/story/news/health/2020/06/12/indiana-says-anyone-who-wants-coronavirus-test-can-get-one/3179151001/
- Testing restrictions impact the propensity of an individual to receive a test and therefore, if COVID-19 positive, contribute to the case count.
-In this paper, testing restrictions are addressed by fitting time-varying testing propensities that depend on relevant covariate information.
-
- On March 23rd, 2020, Indiana's governor issued a \emph{stay at home} order effective March 26th through April 5th.
-The order was extended until April 30th.
-On May 1st, 2020, a five-stage plan for gradual reopening was announced by Governon Holcomb~\citep{fivestageplan}.
-Such policies target reduction in active infection rates.
-In this paper, Indiana's public health policy is incorporated in our construction of epidemiological forecasts in Section~\ref{section:modelbased}.
-
- \subsection{Probabilistic samples in Indiana}
-
- Due to testing restrictions and other potential selection biases, publicly reported COVID-19 case count data may not be sufficient to understand active infection rates or the disease trajectory.
-Here, we discuss two random samples that provide auxiliary information and may help address selection bias.
-
- \subsubsection{Random statewide testing}
-
- Between April 25--29, 2020, Indiana conducted statewide random molecular testing of persons ages $\geq 12$ years to assess prevalence of active infection to SARS-CoV-2~\citep{Yiannoutsos2021}.
-A stratified random sampling design was conducted using Indiana’s 10 public health preparedness districts as sampling strata.
-15,495 participants were contacted resulting in a final sample size of 3,658.
-Demographic data was collected (e.g., summary statistics on age, sex, and race) as well as data on whether they experienced any COVID-19 compatible symptoms during the past 2 weeks or had shared a household with someone who had a positive test result for SARS-CoV-2.
-During May 2--3, 2020, an additional non-random sample of 898 individuals was also collected.
- Table~\ref{tab:indiana} summarizes the data.
- % In this paper, the nonrandom sample is used to assess whether the cumulative summary statistics are appropriate in constructing the inverse-probability weights as well as in constructing appropriate weights using symptom information.
-
- \subsubsection{Delphi's COVID-19 Trends and Impact Survey}
- \label{subsection:fbsymptom}
- Since April 2020, in collaboration with Facebook, the Delphi group at Carnegie Mellon University has conducted the COVID-19 Trends and Impact Survey (CTIS) to monitor the spread and impact of the COVID-19 pandemic in the United States.
-The survey is advertised through Facebook, who automatically select a random sample of its users to see the advertisement~\citep{doi:10.1073/pnas.2111454118}.
-Data collected includes basic demographic information and if the respondent has symptoms such as fever, coughing, shortness of breath, or loss of smell which are associated with COVID-19.
-The survey defines an individual as displaying \emph{COVID-like symptoms} if they exhibit a fever along with a cough, or shortness of breath, or difficulty breathing.
-Figure~\ref{fig:fbsymptoms} displays smoothed estimates of the fraction of individuals who report COVID-like symptoms within the past 24-hours by age and gender.
-The Delphi's COVID-19 Trends and Impact Survey (CTIS) is used as the main source of auxiliary information on time-varying characteristics of the population of Indiana, e.g., displaying COVID-like symptoms or in contact with COVID-19 positive individuals.
-
  \begin{figure}[!th]
- \centering
- \includegraphics[width = 0.9\textwidth]{../figs/fbcovid19symptoms.png}
- \caption{Rate of reported COVID-19 symptoms per strata.
-Daily rates were estimated using weighted method suggested by~\citep{doi:10.1073/pnas.2111454118} on each day separately and then smoothed over time using local-linear nonparametric regression.}
- \label{fig:fbsymptoms}
- \vspace{-0.3cm}
- \end{figure}
-
- %% GOOD THROUGH HERE
-
- \section{Analysis of case-count data}
- \label{section:casecount}
-
- We start with some simple notation.
-Let $N$ denote the population size.
-At a given time, let $Y_j$ denote COVID-19 status for the $j$th individual in the population, $j=1,\ldots, N$.
-Here, like in survey methodology~\citep{Cochran77}, we treat COVID-19 status as a fixed but unknown quantity of interest.
-For simplicity, we start by ignoring the dynamic nature of the outbreak and recoverability of individuals.
-We assume either individual $j$ is COVID-19 positive and $Y_j=1$ or is COVID-19 negative and $Y_j=0$.
-We also let $I_j \in \{0,1\}$ be an indicator that the individual was tested ($I_j = 1$) or not ($I_j=0$).
-
- To start, we assume the overall number of active COVID-19 cases and/or active infection rate (AIR) are of primary interest.
-That is, we are interested in either the population total $Y = \sum_{j=1}^N Y_j$ or the population average $\bar Y = Y/N$.
-Suppose that $n$ tests are performed and we observe the values $y_1, \ldots, y_n \in \{0,1\}$.
-Then a natural candidate for AIR is the proportion of positive tests $\bar y = \frac{1}{n} \sum_{i=1}^n y_i$ -- commonly referred to as the test positive rate -- and a natural candidate for overall active cases is $N \times \bar y$.
- Under simple random sampling (SRS) or any other epsem\footnote{equal probability of selection method} design, the above are unbiased estimators of the population-level quantities of interest.
-Under SRS, the variance of the estimator can be expressed as $\frac{1}{N-1} \times \frac{1-f}{f} \times \sigma_Y^2$ where $f = n/N$ is the sampling fraction and $\sigma_Y^2 = \frac{1}{N} \sum_{j=1}^N (Y_j - \bar Y)^2 = \bar Y (1- \bar Y)$.
-
- These random selection mechanisms are independent of the outcome of interest.
-When this is not the case, selection effects may cause bias.
-To better understand this issue, \cite{Meng2018} recently provided the following intuitive and powerful statistical decomposition of the error between $\bar y$ and the true proportion $\bar Y$
- $$
- \bar y_n - \bar Y =  \rho_{I, Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_Y.
- $$
- The first term represents \emph{data quality}, the second \emph{data quantity}, and the third \emph{problem difficulty}.
-The term $\rho_{I,Y}$ is the empirical correlation between the population values~$\{ Y_j \}_{j=1}^N$ and the selection values $\{ I_j \}_{j=1}^N$.
-Under simple random sampling, $E_{\I} [ \rho_{I,Y} ] = 0$, where the expectation is with respect to the selection mechanism~$\I$, so there is no bias.
-The SRS variance formula above shows that $E_{\I} [ \rho_{I,Y}^2 ]  = 1/(N-1)$.
-
- % The key issue with selective testing is that $E_{\I} [ \rho_{I,Y} ] \neq 0$.
-Meng identified this as the fundamental issue that can lead to paradoxes in the analysis of big data.
-The remainder of this paper aims to build upon this fundamental work by extending the decomposition in two directions: accounting for measurement error and the temporal nature of the pandemic.
-We then discuss effective sampling methods and their importance in the current pandemic.
-
- \subsection{Imperfect testing}
- \label{section:imperfecttesting}
-
- Tests are imperfect.
-COVID-19 testing is no exception.
-Here we investigate the interplay between imperfect testing and selection bias.
-Researchers often assume measurement error leads to parameter attenuation.
-When paired with selection bias, however, the two sources become entangled, and resulting errors in magnified, muted, or even switch signs.
-
- First we require some additional notation.
-Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the binary outcome and $0$ when we observe the true binary outcome.
-We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.
-If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^{\star} = Y_j(1-P_j) + (1-Y_j) P_j$.
-The attentive data analyst will recognize the estimator $\bar y_n$ is now biased for simple random samples.
-In Appendix~\ref{app:memestimator}, assuming sensitivity and specificity are known a priori, a novel iterative procedure is used to construct the estimator $\tilde y_n = (\bar y_n - FP)/(1-(FP+FN))$, which is unbiased under simple random sampling (SRS); see Appendix~\ref{app:modelbased} for a discussion of the connection to model-based estimators.
-In the language of the epidemiology literature, the estimator $\tilde y_n$ is the standard estimator for correcting for measurement error using false positive and negative rates that have been estimated from a validation sample, where $\bar y_n$ is the mis-measured average \& $\tilde y_n$ is the corrected average.
-To understand the impact of selection bias and imperfect testing, we derive the following statistical decomposition of the error between $\tilde y_n$ and $\bar Y$:
- \begin{equation}
- \label{eq:statdecomp}
- \rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
- \times \underbrace{\left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{D_M},
- \end{equation}
- where $f=n/N$ is the sampling fraction, $f_1$ and $f_0$ are sampling fractions for COVID-19 positive and negative individuals respectively, and $\Delta = f_1 - f_0$ is the sampling rate differential.
-See Appendix~\ref{app:memestimator} for the derivation.
-This extends work by \cite{Meng2018} to account for imperfect testing.
-The first three terms continue to represent \emph{data quality}, \emph{data quantity}, and \emph{problem difficulty} respectively.
-The new term~$D_M$ represents the \emph{imperfect testing adjustment} which is a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence.
- %, which is scaled by the factor $(1 - FP - FN)^{-1}$.
- For ease of comprehension, a notation glossary is provided in Section~\ref{app:notation} in the supplementary materials.
-
- \begin{remark}
- \label{rmk:testbinary}
- Note that test positivity is based on an underlying continuous cycle threshold and therefore is not strictly binary.
-Therefore, while we focus on the interplay of measurement error and selection bias in the context of binary testing due to the binary nature of the COVID-19 data, it should be noted that measurement error can lead to bias even in the absence of selection bias.
- \end{remark}
-
- Figure~\ref{fig:heatmap} shows that $D_M$, as a function of the relative frequency ($f_1/f_0$) and log odds ratio, can be both positive and negative as well as a range of magnitudes~\citep{Beesley2020,Beesley2019,Smeden2019}.
-Assuming no measurement error, $D_M = 1$ so the relation between estimation and selection bias is simple, e.g., if COVID-19 positive individuals were more likely to receive test then this implies upward bias in prevalence estimates.
-Under random testing (i.e., $f_0 = f_1$), $D_M = (1-FP-FN)^{-1}$ so measurement error simply magnifies this error.
-When tests are imperfect and selection bias exists, this simple relationship no longer holds.
-
- \begin{figure}[!th]
- \centering
- \includegraphics[width = 0.7\textwidth]{../figs/mem_heatmap.png}
- \caption{Imperfect testing adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.024$ and $FN=0.13$. Color scaled so blue = $-10$, white = $0$, and red = $10$.}
- %\textcolor{red}{NEED TO DO FOR FP = 0.05 and FN = 0.3}}
- \label{fig:heatmap}
- \vspace{-0.3cm}
- \end{figure}
-
- Comparing the mean-squared error (MSE) under a selection mechanism $\I$ with imperfect testing and SRS with perfect testing, we see that
- $$
- \frac{E_{\I} \left[ (\bar y_n - \bar Y)^2 \right]}{\sqrt{V_{SRS} (\bar Y)}} = (N-1) E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right].
- $$
- % implying the error relative to SRS increases as a function of population size~\citep{Meng2018}.
-That two countries with the same testing strategy can yield wildly different estimates due to population size is too often ignored in current communications of case-count data.
- % % Thus, conclusions drawn from observed case-counts may not just be wrong, but very wrong.
- A key question is ``What is the (effective) sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' In Appendix~\ref{app:effss}, we show the effective sample size $n_{eff}$ can be bounded by
- $$
- \frac{f}{1-f} \times \frac{1}{E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right]}.
- $$
- Between April 25th to 29th 2020, Indiana performed $95,879$ tests.
-Indiana's population is roughly $6.732$ million, so $f = 0.003$.
-The active infection rate was estimated to be~$1.81\%$~\citep{Yiannoutsos2021} in this time interval.
-Recent studies have suggested 87\% sensitivity~\citep{Arevalo2020} and 97.6\% specificity~\citep{Cohen2020} are reasonable measurement error rates for RT-PCR tests.
-Supposing COVID-19 positive individuals are $1.5$ times more likely to get tested, then the effective sample size is $168$.
-Recent proposals~\citep{Siddarth2020} have argued for increased testing capacity, which may likely reduce the relative sampling rate.
-Even if the relative sampling rate drops to $1.2$ and $f$ increases to $0.01$ then the effective sample size will increase to $1025$.
-Thus the effective sample size even in optimistic scenarios is equivalent to a moderate random sample from the population.
-Moreover, increased testing capacity may alter false positive and negative rates due to changes in sample collection quality, e.g., a testing center switches from nasal swab to oropharyngeal swabs or saliva specimen to speed up data collection.
-Consider the case where $f$ increasing from $0.003$ to $0.01$ and is associated with the false negative rate rising from $13\%$ to $20$\%.
-Then the effective sample size is $863$, representing a $5.1$ factor increase rather than the expected $6.1$ factor increase.
-See Section~\ref{section:effss} in the Supplementary Materials for additional effective sample size calculations which show that while~\citep{Meng2018} argues the relative error increases as a function of population size, our calculations indicate this is not true for fixed relative frequency~$f_1/f_0$.
- % \begin{wrapfigure}{r}{0.5\textwidth}
-
- % {\bf Figure of individual going through the states? Can discuss exposure length as selection more likely.}
-
- \subsection{Regrettable rates: complex biases resulting from self-selection}
- \label{section:rates}
-
- The prior analysis demonstrates the potentially limited information regarding COVID-19 prevalence in observational case-count data.
-Scientists, however, may claim that daily observed case-counts simply undercount daily total cases by a constant multiple over time (i.e., undercounting).
-If true then the ratio of case-counts at consecutive times may be a good estimate of the true change in prevalence, helping scientists understand the disease trajectory.
-We next demonstrate how selection bias and imperfect testing impact such estimates.
-
- Let $\bar Y_{t-1}$ and $\bar Y_{t}$ denote the prevalence on two consecutive days and consider the estimator $r = \tilde y_t / \tilde y_{t-1}$.
-Using a second-order Taylor series approximation, the error between ${\tilde y_t}/{\tilde y_{t-1}}$ and ${\bar Y_{t}}/{\bar Y_{t-1}}$ can be expressed approximately as
- $$
- \begin{aligned}
- \frac{\bar Y_t}{\bar Y_{t-1}} &\times \bigg[ \rho_{I_t,Y_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (Y_t)  -\rho_{I_{t-1},Y_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \bigg] \\
- &\times \left[ 1 - \rho_{I_{t-1},Y_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \right]
- \end{aligned}
- $$
- where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement error adjustment, and $CV(Y_j) = \sigma_{Y_j}/ \bar Y_j$ is the coefficient of variation on day $j$.
-See Appendix~\ref{app:ratio} for the derivation.
-The error magnitude depends on the true rate $\bar Y_{t} / \bar Y_{t-1}$ so a large decrease will have a small error relative to a large increase.
-The second term represents potential \emph{cancellation} which can  occur when data quality, sampling fraction, measurement error, and prevalence are constant across time.
-
- Figure~\ref{fig:ratio-bias} displays the trajectory of the true ratio and the potential biased estimators under a susceptible-exposed-infected-recovered (SEIR) model~\citep{Pastor2001,Newman2002,Parshani2010} for the epidemic dynamics, with state evolution given by
- \begin{align}
- \frac{\partial s_t}{\partial t} &= - \beta s_t i_t; \quad \frac{\partial e_t}{\partial t} = \beta s_t i_t - \sigma e_t; \quad \label{eq:seir} \\
- \frac{\partial i_t}{\partial t} &= \sigma e_t - \gamma i_t; \quad
- \frac{\partial r_t}{\partial t} = - \gamma i_t. \nonumber
-\end{align}
- % \begin{equation}
- % \label{eq:sir}
- % \frac{\partial s_t}{\partial t} = - \beta s_t i_t; \quad
- % \frac{\partial i_t}{\partial t} = \beta s_t i_t - \gamma i_t; \quad
- % \frac{\partial r_t}{\partial t} = - \gamma i_t
- % \end{equation}
- where $s_t, e_t, i_t$ and $r_t$ are the fraction of susceptible, exposed infected, and removed (recovered or deceased) individuals in the population at time $t$ respectively.
-The SEIR model has been used extensively as a model for SARS-CoV-2 dynamics~\cite{Song2020}.
-In terms of bias, the rate is overestimated prior to the peak in the fraction infected and underestimated afterwards; the bias increases dramatically when the relative fraction exceeds $2$.
-Such biases may impact policy making.
-Overestimation pre-peak may give policy makers more leverage in proposing aggressive actions to reduce prevalence.
-Underestimation post-peak puts pressure on policy makers to prematurely relax social distancing measures.
-Estimates at the peak time appear to have minimal bias.
- %Note, however, that the standard errors will be large due to small effective sample sizes.
-
- \begin{figure}[!th]
- \centering
- \begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{../figs/seir_ratio.png}
-  \caption{Ratio estimator}
-  \label{fig:ratio-bias}
- \end{subfigure}%
- \begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{../figs/seir_rt.png}
-  \caption{Effective reproductive rate estimator}
-  \label{fig:r0-bias}
- \end{subfigure}
- \caption{Potential bias in ratio and effective reproductive rate estimators under an SEIR model with $\beta = 1.2$, $\gamma = 0.15$, and $\sigma = 0.3$.
-Here, $f = 0.02$, $FP = 0.024$, $FN = 0.13$, and a range of relative sampling fractions $M = f_1/f_0$ are considered.}
- \label{fig:rates}
- \end{figure}
-
- \subsection{Estimation of effective reproduction number}
- \label{section:r0-estimation}
- Many epidemiologists argue that tracking the effective reproduction number is the only way to manage through the crisis~\citep{Gabriel2020}.
-Here, we study the instantaneous reproduction number~\citep{Cori20113,Fraser2007}, denoted $R_t$, which is the average number of secondary cases that each infected individual would infect if the conditions remained as they were at time~$t$.
-This is distinct from the case reproduction number,~$R_t^c$, which is the average number of secondary cases that a case infected at time step $t$ will eventually infect~\citep{Wallinga2004}.
-The case reproduction number accounts for potential changes to contact rates and transmissibility, which include impact of control measures.
-The instantaneous reproduction number~$R_t$ is the only reproduction number easily estimated in real time, and therefore has been a key focal point in the COVID-19 pandemic.
-
-Under a Poisson likelihood, a simple relation between the trajectory of new cases and the instantaneous reproduction number can be derived \citep{Bettencourt2008}.
-In particular, under an SIR model the number of case counts on day $t$, denoted $K_t$, is Poisson distributed with rate $K_{t-1} \exp \left( \gamma (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval, which is approximately $7$ days for COVID-19~\citep{Sanche2020}.
-
-\cite{Heng2020} derive an approximate formula for the basic reproduction number under SEIR dynamics such that for $S_0 \approx 1$ and $I_0 \ll 1$, we have the approximate realtionship, which ``corresponds to [...] the solution for the SIR model assuming an infectious period of $1/\gamma + 1/\sigma$.
-Using this connection, a moment-based estimator is given by
-$$
-R_t \approx 1 + \left( \frac{1}{\gamma} + \frac{1}{\sigma} \right) \log \left( \frac{K_t}{K_{t-1}} \right).
-$$
-This approximation ignores a term that is quadratic in the epidemic growth rate and inversely linear in $\sigma \times \gamma$; for realistic values of these quantities, however, the term is relatively negligible and therefore ignored.
-Of course, we do not observe $K_t$ and $K_{t-1}$.
-Under SRS of new cases among those susceptible on day $t$, the natural estimator is $S_t \tilde y_t$.
-Unfortunately the number of susceptible individuals on day $t$ is unknown.
-Here, we study the estimator $\hat R_t = 1 + \frac{1}{\gamma} \log \left( \tilde y_t / \tilde y_{t-1} \right)$.
-We can again express the statistical error of $\hat R_t - R_t$ in useful terms as follows
- $$
- \begin{aligned}
- \left( \frac{1}{\gamma} + \frac{1}{\sigma} \right)\log &\bigg( 1 + \bigg[ \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t)  -\rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (K_{t-1}) \bigg] \\
- &\times \left[ 1 - \rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (K_{t-1}) \right] \bigg) - \left( \frac{1}{\gamma} + \frac{1}{\sigma} \right) \log \left( \frac{S_t}{S_{t-1}} \right).
- \end{aligned}
- $$
- This implies a similar trade-off as before but on the logarithmic scale.
-The error is no longer scaled by $\bar Y_t/\bar Y_{t-1}$ but by the serial interval and does not depend on prevalence but on the fraction of new cases out of those susceptible.
-This leads to differences in when the bias is most pronounced.
-Figure~\ref{fig:r0-bias} displays the bias as a function of the relative sampling fraction ignoring the final term (i.e., $S_t / S_{t-1} \approx 1$).
-Section~\ref{app:cori_rt} in the supplementary materials presents an alternative estimator of the instantaneous effective reproductive number and shows similar bias~\cite{Cori20113}.
-
- \subsection{Rate comparisons}
-
- So far we have focused on understanding the limitations of using case-count data to understand population quantities of interest for a \emph{single} population.
-Many are interested in cross-population comparisons to contrast the impact of countries' mitigation policies.
-Here, for simplicity, we focus on comparing the estimated effective reproductive rate.
-We assume the two time-series are aligned so that $t=0$ is the time of first case in each population respectively.
- % This negates alignment issues and is common in practice.
-
- % While the above issues on $Z$-scores and effective sample size are important, here we highlight a separate issue.
- %Here we demonstrate the interaction of biases in estimation of the two trajectories when the peak infection times differ slightly.
- %First, we can write the difference in the rates as $R_{t1} - R_{t2} = \frac{1}{\gamma} \log \left( \frac{1 + e_{1t}}{1+e_{2t}} \right)$ where $e_{jt}$ is the error associated given in Section \ref{section:r0-estimation}.
- %Recall that the error allowed for over-estimation prior to the peak and then under-estimation post-peak.  Here, these errors can mingle in interesting ways.
- Consider two countries (A and B) in which the peak occurs 2 weeks prior for country A than country B.
-Figure~\ref{fig:comparison} presents such a comparison where each country's disease trajectory follows an SEIR model (A=black and B=grey).
-Figure~\ref{fig:eff} shows how biases interact in complex ways.
-At first, the difference is correctly estimated; then the gap is over-estimated as country A sees a rapid rise in cases; then the magnitude of over-estimation increases as country A sees declining case-count while country B sees rapidly increasing case-count; then country A's rate is correctly estimated while country B's rate is under-estimated as it sees declining case-count; finally, the gap disappears.
-While this may not always be the case, the analysis demonstrates how estimates can tell a more complex story than the truth (i.e., country A's peak is 2 weeks prior to country B's peak).
-
- \begin{figure}
- \centering
- \begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{../figs/seir.png}
-  \caption{Fraction of new cases in population}
-  \label{fig:fracpop}
- \end{subfigure}%
- \begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{../figs/seir_rt_comparison.png}
-  \caption{Effective reproduction rate estimators}
-  \label{fig:eff}
- \end{subfigure}
- \caption{Left: fraction infected in two SEIR models with $\beta = 1.2$ and $0.9$ respectively, $\sigma = 0.3$, and $\gamma = 0.15$ with same initial conditions.
-Right: comparison of $\hat R_t$ across time with $FN = 0.30$, $FP = 0.024$, and $M = 4$.}
- \label{fig:comparison}
- \end{figure}
-
- \section{Potential improvements to prevalence estimation}
- \label{section:improvedcasecount}
-
- The prior section presented negative consequences of selection-bias and measurement error when estimating infection prevalence, rates of change, and the effective reproduction number from observed case-count data.
-In this section, we consider two directions to improve upon these estimators.
- The proposed methods and statistical error decompositions guide our recommendations in Section~\ref{section:discussion}.
-
- \subsection{Selection propensity estimation}
-
- With non-probability samples, bias can be reduced by modelling the self-selection propensity and using inverse probability weighting (IPW)~\citep{Elliott2017} to adjust for selection bias.
- % In causal inference~\citep{Hernan2020}, under certain assumptions, the observational data permits direct estimation of treatment propensity.
- In the current context, however, one does not observe those who are not tested.  To estimate selection propensities, auxiliary information is needed.
-
- To see this mathematically, let $X_j$ denote a vector of covariates for the $j$th individual in the population, $j=1,\ldots,N$.
-For simplicity, the dynamic nature of the outbreak and recoverability of individuals is ignored for now.
-Let~$I_j^{NR}$ denote the selection indicator for individual~$j$ in the population into the non-probability sample.
-Assume it is a Bernoulli random variable that depends only on these covariates, i.e., $P(I^{NR}_j = 1 \mid X_j = x) = \pi (x; \theta)$.
-Here, we focus on logistic regression models, i.e., $\pi(x; \theta) = \text{expit} \left( x^\top \theta \right)$.
-Maximum likelihood estimation follows by maximizing
- \begin{equation}
- \label{eq:propensity}
- \sum_{j=1}^N I_j^{NR} \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi (X_j; \theta) \right).
- \end{equation}
- The first sum only involves individuals observed in the nonprobability sample.
- The second sum is over the entire population.
-Maximum likelihood estimation therefore requires knowledge of covariate information for every individual in the population.
-Typically, this is not possible.
-Here we present a method that uses auxiliary information obtained from probability samples.
-
- \subsubsection{Auxiliary information through probability samples}
- \label{subsec:auxprob}
-
- Here we assume access to a probability sample measuring the same set of covariates.
-Let $I_j^{R}$ be an indicator that the individual was included in the probability sample and $W_j^R$ be the probability of inclusion for the $j$th individual.
-\cite{Chen2019} use a probability sample to construct a design-unbiased estimator of the second term
- \begin{equation}
- \label{eq:auxinfoprob}
- \sum_{j=1}^N I_j^R \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N I_j^R W_j^R \log ( 1 - \pi (X_j; \theta)).
- \end{equation}
- Expectation of~\ref{eq:auxinfoprob} with respect to the sampling design yields~\eqref{eq:propensity}.
-Solving~\eqref{eq:auxinfoprob} is done by iteratively re-weighted least squares (IRLS); see Section~\ref{app:irls} in the supplementary materials for details.
-For ease of comprehension, a notation glossary is provided in Section~\ref{app:notation} in the the supplementary materials.
-
- \begin{remark}
- Weights built from selection propensities~$\{\pi(x_i; \theta)\}_{i=1}^n$ are common practice in the epidemiology literature.
-Survey sampling and transportability weights require knowledge of the selection mechanism~\citep{10.1093/aje/kwx164,ColeStuart2010}, which is not available in our setting.
-Instead, we require auxiliary information in the form of a random sample to estimate the probability of selection.
-A key difference from prior approaches is the need to estimate time-varying probabilities of selection, e.g., Figure 6 shows how the selection propensity of young individuals changes dramatically over the observation window.
- \end{remark}
-
- % Under simple random sampling and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the $\hat \pi_k = (d_k / \tilde d_k) (n / N)$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.
-
- \subsubsection{An IPW estimator and statistical error decomposition}
- \label{section:IPWerrordecomp}
-
- Given a selection propensity, define the inverse probability weight~$w^{NR}(x) = \pi (x; \hat \theta)^{-1}$. Then the IPW estimator adjusted for measurement error is given by
- \begin{equation}
- \label{eq:ipwest}
- \bar y_n^\star
- = \frac{1}{1-FP-FN} \cdot \frac{\sum_{i=1}^n w^{NR} (x_i) (y_i - FP)}{\sum_{i=1}^n w^{NR} (x_i)}
- \stackrel{(2)}{=} \frac{1}{1-FP-FN} \sum_{k=1}^K \frac{d_k w_k}{w} (\bar y_k - FP),
- \end{equation}
- where equality (2) is under the assumption that $x_i$ is a stratification variable with $k$ indexing the strata, $w_k$ is the weight and $d_k$ is the number of samples in strata $k$, and $w = \sum_{k=1}^K d_k w_k$.
- % Under simple random sampling with one strata~$\hat \pi_k = n/N$ and we recover the estimator $\tilde y_n$ from Section~\ref{section:imperfecttesting}.
-
- Let $I_j^{NR} (X_j) = I_j^{NR}  \cdot w^{NR}(X_j)$ for $j=1,\ldots,N$.  Then the error when comparing weighted estimator~$\bar y_n^\star$ to the true prevalence $\bar Y$ can be expressed as:
- \begin{equation}
- \label{eq:statdecomp2}
- \rho_{I^{NR} (X), Y} \times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y} \times \underbrace{\left[ 1 - \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{\tilde D_M}
- \end{equation}
- where $CV_W$ is the coefficient of variation (i.e., standard deviation/mean) of $w^{NR} (X_J)$ given $I_J = 1$, $\rho_{I^{NR} (X), Y}$ is the empirical correlation which here depends on covariate distribution, $\tilde f_k = E[ w^{NR} (X_J) I_J \mid Y_J = k]$ for $k=0,1$, and $\tilde \Delta = \tilde f_1 - \tilde f_0$.
-See Appendix~\ref{app:ipwderivation} for the derivation.
-
- Comparing~\eqref{eq:statdecomp2} to~\eqref{eq:statdecomp} shows that weighting impacts the estimation error in three ways.
-First, there is a negative impact on the data quantity component; taking the ratio of these quantities yields
- $\sqrt{1 + \frac{CV_W^2}{1-f}} \geq 1$.
-Hence, if the data quality does not increase (i.e., $| \rho_{I^{NR} (X), Y} | = | \rho_{I,Y}|$ ) then weighting increases the error magnitude.
-Second, the relative error increase depends on the fraction of population sampled $f$, implying that for large samples there is a larger potential increase in the error if the weights do not improve data quality.
-Third, the impact of measurement-error on data quality is changed when considering a weighted estimand.
-In particular, weighting may result in $\text{sgn}(\Delta) \neq \text{sgn} (\tilde \Delta)$ which implies the impact of measurement-error may be in a different direction.
-
- As demonstrated below in Lemma~\ref{lemma:ipw}, if the propensity model is correctly specified then the $E_{\bf R} [ \rho_{I^{NR} (X), Y} ] = 0$ and therefore $\rho_{I^{NR} (X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then the data quality index is unlikely to inversely scale with population size.
- % Indeed, Horvitz-Thompson estimators are sensitive to errors in the estimated weights.
-Low selection propensity imply large weights which can have deleterious impacts by certain observations dominating estimation.
- Similar to~\cite{Meng2018}, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from an IPW estimator are likely to put too much confidence in the sheer data size.
-
- \subsubsection{Time-varying propensities}
-
- Here we extend the IPW approach to the temporal setting to account for the dynamic nature of the outbreak by considering the joint likelihood
- \begin{equation}
- \label{eq:tvpropensity}
- \sum_{t=1}^T \left[ \sum_{j=1}^N I^{NR}_{j,t} \log \left( \frac{\pi_t (X_{j,t}; \theta)}{1-\pi_t(X_{j,t}; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi_t (X_{j,t}; \theta) \right) \right]
- \end{equation}
- where $t=1,\ldots,T$ are the days when case-count data is reported, and $I^{NR}_{j,t}$ denotes self-selection into testing on day $t$, which is highly correlated with prior testing and results.
-For example, an individual who tests positive may be unlikely to seek testing in the subsequent few days/weeks.
-Moreover, an individual in a high prevalence area may be more likely to seek out testing.
-Here, we assume that the covariate vector $X_{j,t}$ contains all features of the past relevant for selection.
-
- If sufficiently large random samples are collected at each time $t =1,\ldots,T$, then the pseudo-likelihood can be re-written as in~\eqref{eq:auxinfoprob} and propensities estimated separately per time point.
-Unfortunately, large probabilistic samples are not available at every time within a given region.
-To address this, here we consider a non-parametric kernel-based approach where the selection propensity at time~$t$, denoted $\hat \theta_t$, maximizes the smoothed pseudo-likelihood
- $$
- \sum_{t^\prime=1}^T K_h(|t^\prime - t|) \left[ \sum_{j=1}^N I_{j,t^\prime}^{NR} \log \left( \frac{\pi_t (X_{j,t^\prime}; \theta)}{1-\pi_t(X_{j,t^\prime}; \theta)} \right) + \sum_{j=1}^N W^{R}_{j,t^\prime} I^R_{j,t^\prime} \log \left( 1 - \pi_t (X_{j,t^\prime}; \theta) \right) \right]
- $$
- where $K_h$ is a kernel function with tunable parameter $h$.
-Given $\pi (x;\hat \theta_t)$, the prevalence estimator $\bar y_{n,t}^\star$ is given by~\ref{eq:ipwest} using case-count data observed on day~$t$.
-
- \subsubsection{Asymptotics}
- \label{section:asymptotics}
-
- Here we suppose there is a sequence of finite populations of size $N_{\nu}$ indexed by $\nu$ and that there are $L_\nu$ non-probability samples and probability samples of size $n$ drawn at equally spaced times $\{ t^\prime_l \}_{l=1}^{L_{\nu}}$ over the study window~$[0,T]$.
-Assuming correct selection model specification, then under the probability sample design and nonprobability sample propensities, Lemma~\ref{lemma:ipw} shows that the IPW estimator at a time~$t \in \{ t^\prime_l \}$ is consistent as $\nu$ goes to infinity (i.e., $N_{\nu}, L_{\nu} \to \infty$) and calculates the estimator's variance.
-In Lemma~\ref{lemma:ipw}, sensitivity/specificity are unknown and estimated using a pseudo-likelihood on two random samples whose sizes (denoted~$n_{FP}$ and $n_{FN}$) reflect uncertainty in these quantities.
-See Appendix~\ref{app:asympderivations} for additional details.
-
- \begin{lemma}[Variance of IPW estimator] \normalfont
- \label{lemma:ipw}
- % \textcolor{red}{Must check.}
-The estimates~$\hat \mu_t := \bar y_{n,t}^\star$, $\hat \pi_{j,t}$, $\hat FP$, and $\hat FN$ at a time~$t \in \{t_l^\prime\}$ are solutions to the following set of estimating equations
- $$
- \Phi_n (\eta_t) =
- \left (
- \begin{array}{c}
- \frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t}{\pi_{j,t}} \\
- \frac{1}{N \times \sum_{l = 1}^L K_h \left(|t - t_l^\prime| \right)} \sum_{l = 1}^L K_h \left(|t - t_l^\prime| \right) \left[ \sum_{j=1}^N I^{NR}_{j,t_l^\prime} X_{j,t_l^\prime} - \sum_{j = 1}^N I^{R}_{j,t_l^\prime} W^{R}_{j,t_l^\prime}  \pi_{j,t_l^\prime} X_{j,t_l^\prime}  \right] \\
- \frac{1}{n_{FP}} \sum_{i=1}^{n_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
- \frac{1}{n_{FN}} \sum_{i=1}^{n_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
- \end{array}
- \right ) = {\bf 0}
- $$
- where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
-Under certain regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (\bar n^{-1/2})$  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (\bar n^{-1})$ where $\bar n = n \sum_{l=1}^L K_{h} (|t - t_l^\prime|)$ and $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$ where $\phi_n = \frac{\partial \Phi (\eta)}{\partial \eta}$ with $E[\cdot]$ and $\text{Var} (\cdot)$ are under the joint randomization of propensity score and sampling designs.
- \end{lemma}
-
- \subsection{Model-based estimation}
- \label{section:modelbased}
-
- Up to this point, the primary focus has been selection bias in coronavirus case-count data from a survey sampling perspective.
-Here, we consider compartmental model approaches from infectious disease epidemiology.
-Our primary objective is a model-based forecast of strata-level prevalence $\bar y_k$.
-To do this, a probabilistic extension of a standard epidemiological state-space model -- the susceptible, exposed, infected, and removed (recovered and death) model, or SEIR model -- is presented.
-A probabilistic SIR model was originally proposed by~\cite{Osthus2017} with only one-dimensional time series of infected proportions; this formulation was extended by~\cite{Song2020} to model coronavirus case-counts.
-
- Let $s_t$, $e_t$, $i_t$, and $r_t$ denote the proportion of survivors, exposed, infected, and removed cases (i.e., including both recovered cases and deaths) at time $t$.
-The population-level SEIR dynamics are given by the set of differential equations in~\eqref{eq:seir}.
-% \begin{align}
-%  \frac{\partial s_t}{\partial t} &= - \beta s_t i_t; \quad \frac{\partial e_t}{\partial t} = \beta s_t i_t - \sigma e_t; \quad \label{eq:seir} \\
-%  \frac{\partial i_t}{\partial t} &= \sigma e_t - \gamma i_t; \quad
-%  \frac{\partial r_t}{\partial t} = - \gamma i_t. \nonumber
-% \end{align}
-%Following~\cite{Song2020}, we employ Runge-Kutta (RK4) approximations for discretization.
-Here, we consider covariate information that takes the form of a stratification variable with $K$ strata.
-Let $I_{k,t}$ to denote the number of new infections on day $t$ in the $k$th strata.
-The joint distribution of strata-specific new infection totals follows a multinomial distribution with
- $$
- ( I_{1,t}, \ldots, I_{K,t} ) \sim \text{Multinomial} \left( I_t, (p_{1,t}, \ldots, p_{K, t}) \right).
- $$
- where $( {\bf p}_t )_{t=1}^T = ( (p_{1,t}, \ldots, p_{K, t}) )_{t=1}^T$ are a sequence of parameters on the simplex, i.e., $\sum_{k=1}^K p_{k,t} = 1$.
-
- Selection bias is addressed by analyzing COVID-19 death data rather than case counts.  Let $D_{k,r}$ denote the number of individuals who pass away on day $r$ from strata $k$.  Then
- $$
- D_{k,r} \mid p, \theta, \nu \sim \text{Poisson} \left( \sum_{t=1}^r \text{IFR}_{k} \cdot I_{k,t} \theta_{(r-t)} \right)
- $$
- where $\text{IFR}_{k}$ is the infection fatality rate for the $k$th strata and $\theta_{j}$ is a discrete-time distribution for time from infection to death.
-This extends prior analysis of death data~\cite{Johndrow2020} by allowing the infection fatality rate to depend on a stratification variable, which is important as IFR depends heavily on age~\citep{Levin2020}.
-See Section~\ref{section:application_of_modelbased} for discussion of parameter choices.
-%Additional details on inference are discussed in Appendix~\ref{app:modelbased}.
-
- % 0-17: 20/1000000 * 100 -> 0.002 (at age 9)
- % 18-49: 500/1000000 * 100 -> 0.05 (at age 33)
- % 50–64 years old: 6,000/1000000 * 100 -> 0.6
- % 65+ years old: 90,000/1000000 * 100 -> 9.0
- %
-
- \subsection{Doubly robust estimation.}
- Rather than relying solely on IPW or epidemiological forecasts, here we combine forecasting and inverse-probability weighting by extending recent work by~\cite{Chen2019} to account time-varying propensities and measurement-error.
- Let~$\mu(x)$ denote the posterior mean of the active infection rate individuals with covariate value~$x$ (i.e., age strata) based on the SEIR model described in SEction~\ref{section:modelbased}.
-In this context, the doubly-robust estimator is given by
- $$
- \bar y_{n}^{(DR)} = \frac{1}{N} \sum_{j=1}^N \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \hat \mu(x_j) \right).
- $$
- where $\hat \mu(x_j)$ is not corrected for measurement-error as it estimates true case counts.
-This estimator is called ``doubly-robust'' because it is consistent if either the model-based forecasts or the time-varying propensities are correctly specified.
-A statistical error decomposition can be derived
- \begin{align}
- \label{eq:statdecomp3}
- \begin{split}
- \rho_{\tilde I (X), Y-\mu(X)} &\times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y-\mu(X)}  \\
- &\times \underbrace{\left[ 1 - \frac{\rho_{\tilde I(X),Y} \sigma_{Y}}{\rho_{\tilde I(X),Y-\mu(X)} \sigma_{Y-\mu(X)}} \times \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right]}_{\tilde D_M}.
- \end{split}
- \end{align}
- See Appendix~\ref{app:DRderivation} for the derivation.
- % Comparing~\eqref{eq:statdecomp3} to~\eqref{eq:statdecomp2} shows that the addition of the model-based forecast impacts the estimation error in two ways.
- If the model is adequate, then one may expect a reduction in the problem difficulty and (potentially) in the data quality components.
- % Note that the model-based forecast does not impact the data quantity component.
-% Second, if the model-based forecast is correct then $E_{\bf R} [ \rho_{\tilde I (X), Y-\mu(X)} ] = 0$  and therefore $\rho_{\tilde I(X), Y-\mu(X)} = O(N^{-1})$; however, if the forecasts are not correctly specified then the data quality index is unlikely to inversely scale with population size.
- Interestingly, the impact of measurement-error on data quality now depends on a relative comparison of the data quality and problem difficulty of the weighted estimator and the doubly-robust estimator.
- % For decent models, one would expect this ratio to be greater than one.
-Therefore, $\tilde \Delta < 0$ implies an expected increase in impact of measurement-error and $\tilde \Delta > 0$ implies the magnitude could shrink toward zero or change sign and have an even large magnitude in the opposite direction.
-Again, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from a DR estimator are likely to put too much confidence in the sheer size of the data.
-
- Unfortunately, the first term of the doubly robust estimator cannot be computed as covariate information is not collected on the entire population.
-Here, we suppose the same asymptotic regime as in Section~\ref{section:asymptotics} and use the probability sample at the time~$t \in \{t_l^\prime\}$ to estimate this term:
- $$
- \frac{1}{\sum_{j=1}^N \tilde I_{j,t} \tilde W_{j,t}} \sum_{j=1}^N  \tilde I_{j,t} \tilde W_{j,t} \hat \mu_{t} (X_{j,t}) + \frac{1}{\sum_{j=1}^N I_{j,t} w (X_{j,t})} \sum_{j=1}^N I_{j,t} w(X_{j,t}) \left( \frac{Y_{j,t} - FP}{1 - FP - FN} - \hat \mu_t(X_{j,t}) \right).
- $$
- Assuming correct selection propensity model specification, then under the probability sample design and nonprobability sample design, Lemma~\ref{lemma:dr} shows that the doubly robust estimator is consistent and calculates the estimator's variance.
-See Appendix~\ref{app:asympderivations} for additional details.
-
- \begin{lemma}[Variance of doubly-robust estimator] \normalfont
- \label{lemma:dr}
- The estimates~$\mu_t := \bar y_{n,t}^\star$ and $\hat \pi_{j,t}$
- are solutions to the following set of estimating equations
- $$
- \Phi_n (\eta_t) =
- \left (
- \begin{array}{c}
- \frac{1}{N} \left[ \sum_{j=1}^N \frac{I_{j,t}}{\pi_{j,t}} \left( \left( \frac{Y_{j,t} - FP}{1-FP-FN} - \hat \mu_{j,t}  \right)  - \mu_t \right) + \frac{\sum_{l=1}^L K_{h} (|t - t_l^\prime|) \sum_{j=1}^N \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime} \hat \mu_{j,t_l^\prime}}{\sum_{l=1}^L K_{h} (|t - t_l^\prime|) \sum_{j=1}^N \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime}} \right] \\
- \frac{1}{N \times \sum_{l=1}^L K_{h} (|t - t_l^\prime|)} \sum_{l=1}^L K_{h} (|t - t_l^\prime|) \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
- \frac{1}{n_{FP}} \sum_{j=1}^{n_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
- \frac{1}{n_{FN}} \sum_{j=1}^{n_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
- \end{array}
- \right ) = {\bf 0}
- $$
- where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
-Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (\bar n^{-1/2})$ and $\text{var} (\hat y_{n,t}) = V_{t}^{(DR)} + o (\bar n^{-1})$ where $\bar n = n \sum_{l=1}^L K_h(|t-t_l|)$ and $V_t^{(DR)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$ where $\phi_n = \frac{\partial \Phi (\eta)}{\partial \eta}$ with $E[\cdot]$ and $\text{Var} (\cdot)$ are under the joint randomization of propensity score and sampling designs.
- \end{lemma}
-
- \section{COVID-19 active infection prevalence in Indiana}
- \label{section:applications}
-
- % For COVID-19 case-count data, the only covariate information provided by state and local governments are summary statistics for those factors discussed in Section~\ref{section:testinginfo} -- age, sex,  and race.
-These are discrete factors and therefore define a stratification of the population.
-Currently, US testing does not collect additional covariate information such as symptom information; however, this need not be the case moving forward.
-Recommendation XX below states that additional covariate information should be collected from nonprobabilistic samples.
-
- We next consider estimation of the active infection rate in Indiana using unweighted, IPW, model-based, and doubly robust estimates. To start, we recap the data sources used in estimation:
- \begin{itemize}[leftmargin=*]
- \item {\bf Testing data}:  The number of daily tests performed and number of daily positive tests are reported.
-These counts are broken out jointly by age, gender, and racial demographic information.
-Figure~\ref{fig:in-cases} and~\ref{fig:in-tests} plots daily COVID-19 positive cases and tests by age strata.
-
- \begin{figure}
  \centering
@@
 
- \item {\bf Random/Nonrandom statewide sample}: Table~\ref{tab:indiana} summarizes data from a random sample from April 25--29th as well as a nonrandom sample obtained between May 2--3 in racial/ethnic minority communities~\citep{Yiannoutsos2021}.
-The nonrandom sample is not a random subsample of the overall case-counts; however, this data provides important supplementary covariate information.
- % Namely, the samples collected included covariate information on (1) prior symptoms, (2) COVID-19 positive individual in household, and (3) prior COVID-19 positive test.
- To account for the nonrandom sample being from high risk areas, we adjust estimates to match the statewide positivity rate of 11.7\% on August 30th.
-These adjusted rates are presented in parentheses in Table~\ref{tab:indiana} along with the relevant covariate information.
-
- \item {\bf Delphi's COVID-19 Trends and Impact Survey}: From the daily symptom surveys completed as part of Delphi's CTIS (see Section~\ref{subsection:fbsymptom} for details), we extract survey responses for individuals who identify as living in Indiana.
-We collect age, sex, and demographic information as well as COVID-19 related symptoms, which Delphi defines as having (1) a fever and cough, (2) shortness of breath, \emph{or} (3) difficulty breathing.
-Table~\ref{tab:comparison} in Section~\ref{app:in_add_details} of the supplementary materials shows minimal bias with respect to symptom distributions when comparing to Indiana's random sample to the CTIS data collected from April 25th to April 29th.
-
- % \begin{table}[!th]
- % \centering
- % \begin{tabular}{c | c c c | c}
- % & \multicolumn{3}{c}{Facebook} & Random \\ \cline{2-4}
- % & Lower CI & Estimate & Upper CI & Estimate \\ \hline
- % Fever &  0.017 & 0.020 & 0.024 & 0.018 \\
- % Cough &  0.142 & 0.148 & 0.154 & 0.149 \\
- % Shortness & 0.058 & 0.060 & 0.062 & 0.062 \\ \hline
- % \end{tabular}
- % \caption{Comparison of Facebook survey and random sample}
- % \label{tab:comparison}
- % \end{table}
-
- % \item {\bf Hospitalization data}: On each day, we observe the number of individuals who are COVID-19 positive that visited a hospital on that day.
-Unfortunately, hospitalization data is not broken out by age, gender, and racial demographic information.
-Instead, we rely on combining Facebook data to calculate the fraction of those who sought the hospital per week in each strata.
-We then use this to figure out how many individuals per strata were hospitalized and then back out how many
-
- \item {\bf Death data}: Daily COVID-19 related deaths are observed by age, gender, and racial demographics.
-Figure~\ref{fig:in-deaths} plots COVID-19 related deaths over time by age group.
-While most tests and positive cases are within the younger age strata, most deaths are within the 70+ age strata.
- \end{itemize}
-
- \begin{table}[th]
+ Public databases most often, however, only contain aggregate information.
+The Johns Hopkins dashboard, for example, provides demographic breakdown of case counts as well as the total confirmed cases and deaths by county.
+Aggregate time series of case count and deaths can also be extracted.
+Unfortunately, most dashboards do not provide demographic information on who requested a test nor on who tested positive for SARS-CoV-2 \emph{over time}.
+Working closely with the State of Indiana, we were able to obtain access to COVID-19 total tests, positive tests, and related deaths per day broken out marginally by age, gender, ethnicity, and race~\citep{IndianaData2021}.
+These granular datasets are now publicly available and motivate the proposed approach.
+ % https://www.regenstrief.org/covid-dashboard/
+ %
+
+\begin{remark}[Reporting Delays]
+Figure~\ref{fig:in-data} shows clear reduced testing and case count reporting on weekends compared to weekdays.
+Moreover, COVID-19 tests are reported on the day they were administered, while case counts are reported based on the date the positive test was reported to and confirmed by the Indiana Department of Health system.
+To minimize the impact of reporting, testing and case-count data are aggregated at the weekly level for analysis.
+ \end{remark}
+
+\begin{remark}[Public versus government datasets]
+In September 2020, the author submitted a data request to the Indiana Data Hub (IDH).
+Over the next three months, we worked with Indiana's COVID-19 response team to refine the request.
+In December 2020, the request was officially approved and data streams were made publicly available on the IDH.
+In this paper, we thus focus on this \emph{publicly-available data}, i.e., data  that anyone can download directly from the IDH.
+While the scientific community has contributed through independent COVID-19 observational and clinical studies, a significant component of public policy guidance has relied on testing and case count data, e.g., CDC and state guidelines based on test positivity rates and relative changes in the case counts over time.
+Addressing selection bias and measurement-error in these public datasets is imperative in better informing public policy debate.
+
+Note that the COVID-19 response team (and health departments in general) have access to \emph{government data} which may not be publicly-available and is likely stored at the individual-level.
+Due to data privacy and legal concerns, some collected covariates may not be reported publicly and others are aggregated and not reported jointly.
+While data analytic decisions made in Section~\ref{section:applications} were made due to access of publicly-available data, the overall data analysis framework is designed to be a tool for health departments to assess the pandemic and guide responses, e.g.,  account for individual-level covariate information, rather than publicly-reported strata-level information, that may include important covariates such as symptom status and recent COVID-19 contact.
+ \end{remark}
+
+ %https://www.npr.org/sections/goatsandsoda/2020/05/31/865932474/should-i-get-tested-for-coronavirus-just-for-the-heck-of-it
+
+ \subsubsection{Testing restrictions and public health policy in Indiana}
+ \label{section:publicpolicyindiana}
+
+ Due to limited testing capacity, many US states instituted testing restrictions early on in the pandemic.
+ % on who can request a molecular test at a given time.
+ %https://www.coronavirus.in.gov/
+ % Early on, limited testing meant that only those at high risk were allowed to be tested (i.e., symptomatic or high risk populations); as testing capacity expanded, these restrictions were gradually lifted.
+ Here, we reconstruct the testing restriction history for the state of Indiana.
+ On March 6th, Indiana State Health Department of Health confirmed the first case of COVID-19 in Indiana.
+From early March 2020 until April 28th, 2020, only symptomatic essential workers and their households, symptomatic high risk individuals were eligible, and individuals who have returned recently from overseas travel were eligible.
+An individual was considered high risk if they were over the age of 65, diabetic, obese, pregnant, a member of a minority population at greater risk of severe illness, or had high blood pressure.
+ On April 28th, 2020, the criteria expanded to include any symptomatic Indiana resident, people in close contact with those who have tested positive, and residents of congregant communities~\citep{wishtv2020}.
+ % https://www.wishtv.com/news/medical/indiana-opens-up-covid-19-testing-to-more-hoosiers/
+ As of May 12th, 2020, testing expanded to include any high risk individual regardless of symptom status~\citep{indystar2020}
+ % https://www.indystar.com/story/news/health/2020/05/12/coronavirus-testing-indiana-who-should-get-tested/3110592001/
+ On June 15th, 2020, Indiana State Department of Health (ISDH) lifted all testing restrictions~\citep{indystar2020v2}.
+ % https://www.indystar.com/story/news/health/2020/06/12/indiana-says-anyone-who-wants-coronavirus-test-can-get-one/3179151001/
+ Testing restrictions impact the propensity of an individual to receive a test and therefore, if COVID-19 positive, contribute to the case count.
+In this paper, testing restrictions are addressed by fitting time-varying testing propensities that depend on relevant covariate information.
+
+ On March 23rd, 2020, Indiana's governor issued a \emph{stay at home} order effective March 26th through April 5th.
+The order was extended until April 30th.
+On May 1st, 2020, a five-stage plan for gradual reopening was announced by Governon Holcomb~\citep{fivestageplan}.
+Such policies target reduction in active infection rates.
+In this paper, Indiana's public health policy is incorporated in our construction of epidemiological forecasts in Section~\ref{section:modelbased}.
+
+ \subsection{Probabilistic samples in Indiana}
+
+ Due to testing restrictions and other potential selection biases, publicly reported COVID-19 case count data may not be sufficient to understand active infection rates or the disease trajectory.
+Here, we discuss two random samples that provide auxiliary information and may help address selection bias.
+
+ \subsubsection{Random statewide testing}
+
+ Between April 25--29, 2020, Indiana conducted statewide random molecular testing of persons ages $\geq 12$ years to assess prevalence of active infection to SARS-CoV-2~\citep{Yiannoutsos2021}.
+A stratified random sampling design was conducted using Indiana’s 10 public health preparedness districts as sampling strata.
+15,495 participants were contacted resulting in a final sample size of 3,658.
+Demographic data was collected (e.g., summary statistics on age, sex, and race) as well as data on whether they experienced any COVID-19 compatible symptoms during the past 2 weeks or had shared a household with someone who had a positive test result for SARS-CoV-2.
+During May 2--3, 2020, an additional non-random sample of 898 individuals was also collected.
+ Table~\ref{tab:indiana} summarizes the data.
+ % In this paper, the nonrandom sample is used to assess whether the cumulative summary statistics are appropriate in constructing the inverse-probability weights as well as in constructing appropriate weights using symptom information.
+
+ \begin{table}[!th]
  \begin{tabular}{c | c | r r r | r | r r }
- & & \multicolumn{3}{c}{Total Tests (\%)} & US & \multicolumn{2}{c}{Positive Test Rate (\%)}\\
+ & & \multicolumn{3}{c}{Total Tests (\%)} & IN & \multicolumn{2}{c}{Positive Test Rate (\%)}\\
  \cline{3-5} \cline{7-8}
@@
  \end{tabular}
- \caption{Estimated total tests ($\%$) and point prevalence of active infection with SARS-CoV-2 by demographics in Indiana~\citep{Yiannoutsos2021}.}
+ \caption{Estimated total tests ($\%$) and point prevalence of active infection with SARS-CoV-2 by demographics in Indiana~\citep{Yiannoutsos2021,doi:10.1073/pnas.2111454118}.}
  \label{tab:indiana}
  \end{table}
+
+ \subsubsection{Delphi's COVID-19 Trends and Impact Survey}
+ \label{subsection:fbsymptom}
+ Since April 2020, in collaboration with Facebook, the Delphi group at Carnegie Mellon University has conducted the COVID-19 Trends and Impact Survey (CTIS) to monitor the spread and impact of the COVID-19 pandemic in the United States.
+The survey is advertised through Facebook, who automatically select a random sample of its users to see the advertisement~\citep{doi:10.1073/pnas.2111454118}.
+Data collected includes basic demographic information and if the respondent has symptoms such as fever, coughing, shortness of breath, or loss of smell which are associated with COVID-19.
+The survey defines an individual as displaying \emph{COVID-like symptoms} if they exhibit a fever along with a cough, or shortness of breath, or difficulty breathing.
+Figure~\ref{fig:fbsymptoms} displays smoothed estimates of the fraction of individuals who report COVID-like symptoms within the past 24-hours by age and gender.
+The Delphi's COVID-19 Trends and Impact Survey (CTIS) is used as the main source of auxiliary information on time-varying characteristics of the population of Indiana, e.g., displaying COVID-like symptoms or in contact with COVID-19 positive individuals.
+
+ \begin{figure}[!th]
+ \centering
+ \includegraphics[width = 0.9\textwidth]{../figs/fbcovid19symptoms.png}
+ \caption{Rate of reported COVID-19 symptoms per strata.
+Daily rates were estimated using weighted method suggested by~\citep{doi:10.1073/pnas.2111454118} on each day separately and then smoothed over time using local-linear nonparametric regression.}
+ \label{fig:fbsymptoms}
+ \vspace{-0.3cm}
+ \end{figure}
+
+ %% GOOD THROUGH HERE
+
+ \section{Analysis of case-count data}
+ \label{section:casecount}
+
+ We start with some simple notation.
+Let $N$ denote the population size.
+At a given time, let $Y_j$ denote COVID-19 status for the $j$th individual in the population, $j=1,\ldots, N$.
+Here, like in survey methodology~\citep{Cochran77}, we treat COVID-19 status as a fixed but unknown quantity of interest.
+For simplicity, we start by ignoring the dynamic nature of the outbreak and recoverability of individuals.
+We assume either individual $j$ is COVID-19 positive and $Y_j=1$ or is COVID-19 negative and $Y_j=0$.
+We also let $I_j \in \{0,1\}$ be an indicator that the individual was tested ($I_j = 1$) or not ($I_j=0$).
+
+ To start, we assume the overall number of active COVID-19 cases and/or active infection rate (AIR) are of primary interest.
+That is, we are interested in either the population total $Y = \sum_{j=1}^N Y_j$ or the population average $\bar Y = Y/N$.
+Suppose that $n$ tests are performed and we observe the values $y_1, \ldots, y_n \in \{0,1\}$.
+Then a natural candidate for AIR is the proportion of positive tests $\bar y = \frac{1}{n} \sum_{i=1}^n y_i$ -- commonly referred to as the test positivity rate -- and a natural candidate for overall active cases is $N \times \bar y$.
+ Under simple random sampling (SRS) or any other epsem\footnote{equal probability of selection method} design, the above are unbiased estimators of the population-level quantities of interest.
+Under SRS, the variance of the estimator can be expressed as $\frac{1}{N-1} \times \frac{1-f}{f} \times \sigma_Y^2$ where $f = n/N$ is the sampling fraction and $\sigma_Y^2 = \frac{1}{N} \sum_{j=1}^N (Y_j - \bar Y)^2 = \bar Y (1- \bar Y)$.
+
+ These random selection mechanisms are independent of the outcome of interest.
+When this is not the case, selection effects may cause bias.
+To better understand this issue, \cite{Meng2018} recently provided the following intuitive and powerful statistical decomposition of the error between $\bar y$ and the true proportion $\bar Y$
+ $$
+ \bar y_n - \bar Y =  \rho_{I, Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_Y.
+ $$
+ The first term represents \emph{data quality}, the second \emph{data quantity}, and the third \emph{problem difficulty}.
+The term $\rho_{I,Y}$ is the empirical correlation between the population values~$\{ Y_j \}_{j=1}^N$ and the selection values $\{ I_j \}_{j=1}^N$.
+Under simple random sampling, $E_{\I} [ \rho_{I,Y} ] = 0$, where the expectation is with respect to the selection mechanism~$\I$, so there is no bias.
+The SRS variance formula above shows that $E_{\I} [ \rho_{I,Y}^2 ]  = 1/(N-1)$.
+
+ % The key issue with selective testing is that $E_{\I} [ \rho_{I,Y} ] \neq 0$.
+Meng identified this as the fundamental issue that can lead to paradoxes in the analysis of big data.
+The remainder of this paper aims to build upon this fundamental work by extending the decomposition in two directions: accounting for measurement error and the temporal nature of the pandemic.
+We then discuss effective sampling methods and their importance in the current pandemic.
+
+ \subsection{Imperfect testing}
+ \label{section:imperfecttesting}
+
+ Tests are imperfect.
+COVID-19 testing is no exception.
+Here we investigate the interplay between imperfect testing and selection bias.
+Researchers often assume measurement error leads to parameter attenuation.
+When paired with selection bias, however, the two sources become entangled, and resulting errors in magnified, muted, or even switch signs.
+
+ First we require some additional notation.
+Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the binary outcome and $0$ when we observe the true binary outcome.
+We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.
+If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^{\star} = Y_j(1-P_j) + (1-Y_j) P_j$.
+The attentive data analyst will recognize the estimator $\bar y_n$ is now biased for simple random samples.
+In Appendix~\ref{app:memestimator}, assuming sensitivity and specificity are known a priori, a novel iterative procedure is used to construct the estimator $\tilde y_n = (\bar y_n - FP)/(1-(FP+FN))$, which is unbiased under simple random sampling (SRS); see Appendix~\ref{app:modelbased} for a discussion of the connection to model-based estimators.
+In the language of the epidemiology literature, the estimator $\tilde y_n$ is the standard estimator for correcting for measurement error using false positive and negative rates that have been estimated from a validation sample, where $\bar y_n$ is the mis-measured average \& $\tilde y_n$ is the corrected average.
+To understand the impact of selection bias and imperfect testing, we derive the following statistical decomposition of the error between $\tilde y_n$ and $\bar Y$:
+ \begin{equation}
+ \label{eq:statdecomp}
+ \rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
+ \times \underbrace{\left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{D_M},
+ \end{equation}
+ where $f=n/N$ is the sampling fraction, $f_1$ and $f_0$ are sampling fractions for COVID-19 positive and negative individuals respectively, and $\Delta = f_1 - f_0$ is the sampling rate differential.
+See Appendix~\ref{app:memestimator} for the derivation.
+This extends work by \cite{Meng2018} to account for imperfect testing.
+The first three terms continue to represent \emph{data quality}, \emph{data quantity}, and \emph{problem difficulty} respectively.
+The new term~$D_M$ represents the \emph{imperfect testing adjustment} which is a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence.
+ %, which is scaled by the factor $(1 - FP - FN)^{-1}$.
+ For ease of comprehension, a notation glossary is provided in Section~\ref{app:notation} in the supplementary materials.
+
+ \begin{remark}
+ \label{rmk:testbinary}
+ Note that test positivity is based on an underlying continuous cycle threshold and therefore not strictly binary.
+While we focus on the interplay of measurement error and selection bias in the context of binary outcomes due to the dichotomous nature of the COVID-19 testing data, it should be noted that measurement error can lead to bias even in the absence of selection bias.
+ \end{remark}
+
+ Figure~\ref{fig:heatmap} shows that $D_M$, as a function of the relative frequency ($f_1/f_0$) and log odds ratio, can be both positive and negative as well as a range of magnitudes~\citep{Beesley2020,Beesley2019,Smeden2019}.
+Assuming no measurement error, $D_M = 1$ so the relation between estimation and selection bias is simple, e.g., if COVID-19 positive individuals were more likely to receive test then this implies upward bias in prevalence estimates.
+Under random testing (i.e., $f_0 = f_1$), $D_M = (1-FP-FN)^{-1}$ so measurement error simply magnifies this error.
+When tests are imperfect and selection bias exists, this simple relationship no longer holds.
+
+ \begin{figure}[!th]
+ \centering
+ \includegraphics[width = 0.7\textwidth]{../figs/mem_heatmap.png}
+ \caption{Imperfect testing adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.024$ and $FN=0.13$. Color scaled so blue = $-10$, white = $0$, and red = $10$.}
+ %\textcolor{red}{NEED TO DO FOR FP = 0.05 and FN = 0.3}}
+ \label{fig:heatmap}
+ \vspace{-0.3cm}
+ \end{figure}
+
+ Comparing the mean-squared error (MSE) under a selection mechanism $\I$ with imperfect testing and SRS with perfect testing, we see that
+ $$
+ \frac{E_{\I} \left[ (\bar y_n - \bar Y)^2 \right]}{\sqrt{V_{SRS} (\bar Y)}} = (N-1) E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right].
+ $$
+ % implying the error relative to SRS increases as a function of population size~\citep{Meng2018}.
+That two countries with the same testing strategy can yield wildly different estimates due to population size is too often ignored in current communications of case-count data.
+ % % Thus, conclusions drawn from observed case-counts may not just be wrong, but very wrong.
+ A key question is ``What is the (effective) sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' In Appendix~\ref{app:effss}, we show the effective sample size $n_{eff}$ can be bounded by
+ $$
+ \frac{f}{1-f} \times \frac{1}{E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right]}.
+ $$
+ Between April 25th to 29th 2020, Indiana performed $95,879$ tests.
+Indiana's population is roughly $6.732$ million, so $f = 0.003$.
+The active infection rate was estimated to be~$1.81\%$~\citep{Yiannoutsos2021} in this time interval.
+Recent studies have suggested 87\% sensitivity~\citep{Arevalo2020} and 97.6\% specificity~\citep{Cohen2020} are reasonable measurement error rates for RT-PCR tests.
+Supposing COVID-19 positive individuals are $1.5$ times more likely to get tested, then the effective sample size is $168$.
+Recent proposals~\citep{Siddarth2020} have argued for increased testing capacity, which may likely reduce the relative sampling rate.
+Even if the relative sampling rate drops to $1.2$ and $f$ increases to $0.01$ then the effective sample size will increase to $1025$.
+Thus the effective sample size even in optimistic scenarios is equivalent to a moderate random sample from the population.
+Moreover, increased testing capacity may alter false positive and negative rates due to changes in sample collection quality, e.g., a testing center switches from nasal swab to oropharyngeal swabs or saliva specimen to speed up data collection.
+Consider the case where $f$ increasing from $0.003$ to $0.01$ and is associated with the false negative rate rising from $13\%$ to $20$\%.
+Then the effective sample size is $863$, representing a $5.1$ factor increase rather than the expected $6.1$ factor increase.
+See Section~\ref{section:effss} in the Supplementary Materials for additional effective sample size calculations which show that while~\citep{Meng2018} argues the relative error increases as a function of population size, our calculations indicate this is not true for fixed relative frequency~$f_1/f_0$.
+ % \begin{wrapfigure}{r}{0.5\textwidth}
+
+ % {\bf Figure of individual going through the states? Can discuss exposure length as selection more likely.}
+
+ \subsection{Regrettable rates: complex biases resulting from self-selection}
+ \label{section:rates}
+
+ The prior analysis demonstrates the potentially limited information regarding COVID-19 prevalence in observational case-count data.
+Scientists, however, may claim that daily observed case-counts simply undercount daily total cases by a constant multiple over time (i.e., undercounting).
+If true then the ratio of case-counts at consecutive times may be a good estimate of the true change in prevalence, helping scientists understand the disease trajectory.
+We next demonstrate how selection bias and imperfect testing impact such estimates.
+
+ Let $\bar Y_{t-1}$ and $\bar Y_{t}$ denote the prevalence on two consecutive days and consider the estimator $r = \tilde y_t / \tilde y_{t-1}$.
+Using a second-order Taylor series approximation, the error between ${\tilde y_t}/{\tilde y_{t-1}}$ and ${\bar Y_{t}}/{\bar Y_{t-1}}$ can be expressed approximately as
+ $$
+ \begin{aligned}
+ \frac{\bar Y_t}{\bar Y_{t-1}} &\times \bigg[ \rho_{I_t,Y_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (Y_t)  -\rho_{I_{t-1},Y_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \bigg] \\
+ &\times \left[ 1 - \rho_{I_{t-1},Y_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \right]
+ \end{aligned}
+ $$
+ where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement error adjustment, and $CV(Y_j) = \sigma_{Y_j}/ \bar Y_j$ is the coefficient of variation on day $j$.
+See Appendix~\ref{app:ratio} for the derivation.
+The error magnitude depends on the true rate $\bar Y_{t} / \bar Y_{t-1}$ so a large decrease will have a small error relative to a large increase.
+The second term represents potential \emph{cancellation} which can  occur when data quality, sampling fraction, measurement error, and prevalence are constant across time.
+
+ Figure~\ref{fig:ratio-bias} displays the trajectory of the true ratio and the potential biased estimators under a susceptible-exposed-infected-recovered (SEIR) model~\citep{Pastor2001,Newman2002,Parshani2010} for the epidemic dynamics, with state evolution given by
+ \begin{align}
+ \frac{\partial s_t}{\partial t} &= - \beta s_t i_t; \quad \frac{\partial e_t}{\partial t} = \beta s_t i_t - \sigma e_t; \quad \label{eq:seir} \\
+ \frac{\partial i_t}{\partial t} &= \sigma e_t - \gamma i_t; \quad
+ \frac{\partial r_t}{\partial t} = - \gamma i_t. \nonumber
+\end{align}
+ % \begin{equation}
+ % \label{eq:sir}
+ % \frac{\partial s_t}{\partial t} = - \beta s_t i_t; \quad
+ % \frac{\partial i_t}{\partial t} = \beta s_t i_t - \gamma i_t; \quad
+ % \frac{\partial r_t}{\partial t} = - \gamma i_t
+ % \end{equation}
+ where $s_t, e_t, i_t$ and $r_t$ are the fraction of susceptible, exposed infected, and removed (recovered or deceased) individuals in the population at time $t$ respectively.
+The SEIR model has been used extensively as a model for SARS-CoV-2 dynamics~\cite{Song2020}.
+In terms of bias, the rate is overestimated prior to the peak in the fraction infected and underestimated afterwards; the bias increases dramatically when the relative fraction exceeds $2$.
+Such biases may impact policy making.
+Overestimation pre-peak may give policy makers more leverage in proposing aggressive actions to reduce prevalence.
+Underestimation post-peak puts pressure on policy makers to prematurely relax social distancing measures.
+Estimates at the peak time appear to have minimal bias.
+ %Note, however, that the standard errors will be large due to small effective sample sizes.
+
+ \begin{figure}[!th]
+ \centering
+ \begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{../figs/seir_ratio.png}
+  \caption{Ratio estimator}
+  \label{fig:ratio-bias}
+ \end{subfigure}%
+ \begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{../figs/seir_rt.png}
+  \caption{Effective reproductive rate estimator}
+  \label{fig:r0-bias}
+ \end{subfigure}
+ \caption{Potential bias in ratio and effective reproductive rate estimators under an SEIR model with $\beta = 1.2$, $\gamma = 0.15$, and $\sigma = 0.3$.
+Here, $f = 0.02$, $FP = 0.024$, $FN = 0.13$, and a range of relative sampling fractions $M = f_1/f_0$ are considered.}
+ \label{fig:rates}
+ \end{figure}
+
+ \subsection{Estimation of effective reproduction number}
+ \label{section:r0-estimation}
+ Many epidemiologists argue that tracking the effective reproduction number is the only way to manage through the crisis~\citep{Gabriel2020}.
+Here, we study the instantaneous reproduction number~\citep{Cori20113,Fraser2007}, denoted $R_t$, which is the average number of secondary cases that each infected individual would infect if the conditions remained as they were at time~$t$.
+This is distinct from the case reproduction number,~$R_t^c$, which is the average number of secondary cases that a case infected at time step $t$ will eventually infect~\citep{Wallinga2004}.
+The case reproduction number accounts for potential changes to contact rates and transmissibility, which include impact of control measures.
+The instantaneous reproduction number~$R_t$ is the only reproduction number easily estimated in real time, and therefore has been a key focal point in the COVID-19 pandemic.
+
+Under a Poisson likelihood, a simple relation between the trajectory of new cases and the instantaneous reproduction number can be derived \citep{Bettencourt2008}.
+In particular, under an SIR model the number of case counts on day $t$, denoted $K_t$, is Poisson distributed with rate $K_{t-1} \exp \left( \gamma (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval, which is approximately $7$ days for COVID-19~\citep{Sanche2020}.
+
+\cite{Heng2020} derive an approximate formula for the basic reproduction number under SEIR dynamics such that for $S_0 \approx 1$ and $I_0 \ll 1$, we have the approximate realtionship, which ``corresponds to [...] the solution for the SIR model assuming an infectious period of $1/\gamma + 1/\sigma$.
+Using this connection, a moment-based estimator is given by
+$$
+R_t \approx 1 + \left( \frac{1}{\gamma} + \frac{1}{\sigma} \right) \log \left( \frac{K_t}{K_{t-1}} \right).
+$$
+This approximation ignores a term that is quadratic in the epidemic growth rate and inversely linear in $\sigma \times \gamma$; for realistic values of these quantities, however, the term is relatively negligible and therefore ignored.
+Of course, we do not observe $K_t$ and $K_{t-1}$.
+Under SRS of new cases among those susceptible on day $t$, the natural estimator is $S_t \tilde y_t$.
+Unfortunately the number of susceptible individuals on day $t$ is unknown.
+Here, we study the estimator $\hat R_t = 1 + \frac{1}{\gamma} \log \left( \tilde y_t / \tilde y_{t-1} \right)$.
+We can again express the statistical error of $\hat R_t - R_t$ in useful terms as follows
+ $$
+ \begin{aligned}
+ \left( \frac{1}{\gamma} + \frac{1}{\sigma} \right)\log &\bigg( 1 + \bigg[ \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t)  -\rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (K_{t-1}) \bigg] \\
+ &\times \left[ 1 - \rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (K_{t-1}) \right] \bigg) - \left( \frac{1}{\gamma} + \frac{1}{\sigma} \right) \log \left( \frac{S_t}{S_{t-1}} \right).
+ \end{aligned}
+ $$
+ This implies a similar trade-off as before but on the logarithmic scale.
+The error is no longer scaled by $\bar Y_t/\bar Y_{t-1}$ but by the serial interval and does not depend on prevalence but on the fraction of new cases out of those susceptible.
+This leads to differences in when the bias is most pronounced.
+Figure~\ref{fig:r0-bias} displays the bias as a function of the relative sampling fraction ignoring the final term (i.e., $S_t / S_{t-1} \approx 1$).
+Section~\ref{app:cori_rt} in the supplementary materials presents an alternative estimator of the instantaneous effective reproductive number and shows similar bias~\cite{Cori20113}.
+
+ \subsection{Rate comparisons}
+
+ \begin{figure}[!th]
+ \centering
+ \begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{../figs/seir.png}
+  \caption{Fraction of new cases in population}
+  \label{fig:fracpop}
+ \end{subfigure}%
+ \begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{../figs/seir_rt_comparison.png}
+  \caption{Effective reproduction rate estimators}
+  \label{fig:eff}
+ \end{subfigure}
+ \caption{Left: fraction infected in two SEIR models with $\beta = 1.2$ and $0.9$ respectively, $\sigma = 0.3$, and $\gamma = 0.15$ with same initial conditions.
+Right: comparison of $\hat R_t$ across time with $FN = 0.30$, $FP = 0.024$, and $M = 4$.}
+ \label{fig:comparison}
+ \end{figure}
+
+ So far we have focused on understanding the limitations of using case-count data to understand population quantities of interest for a \emph{single} population.
+Many are interested in cross-population comparisons to contrast the impact of countries' mitigation policies.
+Here, for simplicity, we focus on comparing the estimated effective reproductive rate.
+We assume the two time-series are aligned so that $t=0$ is the time of first case in each population respectively.
+ % This negates alignment issues and is common in practice.
+
+ % While the above issues on $Z$-scores and effective sample size are important, here we highlight a separate issue.
+ %Here we demonstrate the interaction of biases in estimation of the two trajectories when the peak infection times differ slightly.
+ %First, we can write the difference in the rates as $R_{t1} - R_{t2} = \frac{1}{\gamma} \log \left( \frac{1 + e_{1t}}{1+e_{2t}} \right)$ where $e_{jt}$ is the error associated given in Section \ref{section:r0-estimation}.
+ %Recall that the error allowed for over-estimation prior to the peak and then under-estimation post-peak.  Here, these errors can mingle in interesting ways.
+ Consider two countries (A and B) in which the peak occurs 2 weeks prior for country A than country B.
+Figure~\ref{fig:comparison} presents such a comparison where each country's disease trajectory follows an SEIR model (A=black and B=grey).
+Figure~\ref{fig:eff} shows how biases interact in complex ways.
+At first, the difference is correctly estimated; then the gap is over-estimated as country A sees a rapid rise in cases; then the magnitude of over-estimation increases as country A sees declining case-count while country B sees rapidly increasing case-count; then country A's rate is correctly estimated while country B's rate is under-estimated as it sees declining case-count; finally, the gap disappears.
+While this may not always be the case, the analysis demonstrates how estimates can tell a more complex story than the truth (i.e., country A's peak is 2 weeks prior to country B's peak).
+
+ \section{Potential improvements to prevalence estimation}
+ \label{section:improvedcasecount}
+
+ The prior section presented negative consequences of selection-bias and measurement error when estimating infection prevalence, rates of change, and the effective reproduction number from observed case-count data.
+In this section, we consider two directions to improve upon these estimators.
+ The proposed methods and statistical error decompositions guide our recommendations in Section~\ref{section:discussion}.
+
+ \subsection{Selection propensity estimation}
+
+ With non-probability samples, bias can be reduced by modelling the self-selection propensity and using inverse probability weighting (IPW)~\citep{Elliott2017} to adjust for selection bias.
+ % In causal inference~\citep{Hernan2020}, under certain assumptions, the observational data permits direct estimation of treatment propensity.
+ In the current context, however, one does not observe those who are not tested.  To estimate selection propensities, auxiliary information is needed.
+
+ To see this mathematically, let $X_j$ denote a vector of covariates for the $j$th individual in the population, $j=1,\ldots,N$.
+For simplicity, the dynamic nature of the outbreak and recoverability of individuals is ignored for now.
+Let~$I_j^{NR}$ denote the selection indicator for individual~$j$ in the population into the non-probability sample.
+Assume it is a Bernoulli random variable that depends only on these covariates, i.e., $P(I^{NR}_j = 1 \mid X_j = x) = \pi (x; \theta)$.
+Here, we focus on logistic regression models, i.e., $\pi(x; \theta) = \text{expit} \left( x^\top \theta \right)$.
+Maximum likelihood estimation follows by maximizing
+ \begin{equation}
+ \label{eq:propensity}
+ \sum_{j=1}^N I_j^{NR} \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi (X_j; \theta) \right).
+ \end{equation}
+ The first sum only involves individuals observed in the nonprobability sample.
+ The second sum is over the entire population.
+Maximum likelihood estimation therefore requires knowledge of covariate information for every individual in the population.
+Typically, this is not possible.
+Here we present a method that uses auxiliary information obtained from probability samples.
+
+ \subsubsection{Auxiliary information through probability samples}
+ \label{subsec:auxprob}
+
+ Here we assume access to a probability sample measuring the same set of covariates.
+Let $I_j^{R}$ be an indicator that the individual was included in the probability sample and $W_j^R$ be the probability of inclusion for the $j$th individual.
+\cite{Chen2019} use a probability sample to construct a design-unbiased estimator of the second term
+ \begin{equation}
+ \label{eq:auxinfoprob}
+ \sum_{j=1}^N I_j^R \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N I_j^R W_j^R \log ( 1 - \pi (X_j; \theta)).
+ \end{equation}
+ Expectation of~\ref{eq:auxinfoprob} with respect to the sampling design yields~\eqref{eq:propensity}.
+Solving~\eqref{eq:auxinfoprob} is done by iteratively re-weighted least squares (IRLS); see Section~\ref{app:irls} in the supplementary materials for details.
+For ease of comprehension, a notation glossary is provided in Section~\ref{app:notation} in the the supplementary materials.
+
+ \begin{remark}
+ Weights built from selection propensities~$\{\pi(x_i; \theta)\}_{i=1}^n$ are common practice in the epidemiology literature.
+Survey sampling and transportability weights require knowledge of the selection mechanism~\citep{10.1093/aje/kwx164,ColeStuart2010}, which is not available in our setting.
+Instead, we require auxiliary information in the form of a random sample to estimate the probability of selection.
+A key difference from prior approaches is the need to estimate time-varying probabilities of selection, e.g., Figure 6 shows how the selection propensity of young individuals changes dramatically over the observation window.
+ \end{remark}
+
+ % Under simple random sampling and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the $\hat \pi_k = (d_k / \tilde d_k) (n / N)$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.
+
+ \subsubsection{An IPW estimator and statistical error decomposition}
+ \label{section:IPWerrordecomp}
+
+ Given a selection propensity, define the inverse probability weight~$w^{NR}(x) = \pi (x; \hat \theta)^{-1}$. Then the IPW estimator adjusted for measurement error is given by
+ \begin{equation}
+ \label{eq:ipwest}
+ \bar y_n^\star
+ = \frac{1}{1-FP-FN} \cdot \frac{\sum_{i=1}^n w^{NR} (x_i) (y_i - FP)}{\sum_{i=1}^n w^{NR} (x_i)}
+ \stackrel{(2)}{=} \frac{1}{1-FP-FN} \sum_{k=1}^K \frac{d_k w_k}{w} (\bar y_k - FP),
+ \end{equation}
+ where equality (2) is under the assumption that $x_i$ is a stratification variable with $k$ indexing the strata, $w_k$ is the weight and $d_k$ is the number of samples in strata $k$, and $w = \sum_{k=1}^K d_k w_k$.
+ % Under simple random sampling with one strata~$\hat \pi_k = n/N$ and we recover the estimator $\tilde y_n$ from Section~\ref{section:imperfecttesting}.
+
+ Let $I_j^{NR} (X_j) = I_j^{NR}  \cdot w^{NR}(X_j)$ for $j=1,\ldots,N$.  Then the error when comparing weighted estimator~$\bar y_n^\star$ to the true prevalence $\bar Y$ can be expressed as:
+ \begin{equation}
+ \label{eq:statdecomp2}
+ \rho_{I^{NR} (X), Y} \times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y} \times \underbrace{\left[ 1 - \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{\tilde D_M}
+ \end{equation}
+ where $CV_W$ is the coefficient of variation (i.e., standard deviation/mean) of $w^{NR} (X_J)$ given $I_J = 1$, $\rho_{I^{NR} (X), Y}$ is the empirical correlation which here depends on covariate distribution, $\tilde f_k = E[ w^{NR} (X_J) I_J \mid Y_J = k]$ for $k=0,1$, and $\tilde \Delta = \tilde f_1 - \tilde f_0$.
+See Appendix~\ref{app:ipwderivation} for the derivation.
+
+ Comparing~\eqref{eq:statdecomp2} to~\eqref{eq:statdecomp} shows that weighting impacts the estimation error in three ways.
+First, there is a negative impact on the data quantity component; taking the ratio of these quantities yields
+ $\sqrt{1 + \frac{CV_W^2}{1-f}} \geq 1$.
+Hence, if the data quality does not increase (i.e., $| \rho_{I^{NR} (X), Y} | = | \rho_{I,Y}|$ ) then weighting increases the error magnitude.
+Second, the relative error increase depends on the fraction of population sampled $f$, implying that for large samples there is a larger potential increase in the error if the weights do not improve data quality.
+Third, the impact of measurement-error on data quality is changed when considering a weighted estimand.
+In particular, weighting may result in $\text{sgn}(\Delta) \neq \text{sgn} (\tilde \Delta)$ which implies the impact of measurement-error may be in a different direction.
+
+ As demonstrated below in Lemma~\ref{lemma:ipw}, if the propensity model is correctly specified then the $E_{\bf R} [ \rho_{I^{NR} (X), Y} ] = 0$ and therefore $\rho_{I^{NR} (X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then the data quality index is unlikely to inversely scale with population size.
+ % Indeed, Horvitz-Thompson estimators are sensitive to errors in the estimated weights.
+Low selection propensity imply large weights which can have deleterious impacts by certain observations dominating estimation.
+ Similar to~\cite{Meng2018}, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from an IPW estimator are likely to put too much confidence in the sheer data size.
+
+ \subsubsection{Time-varying propensities}
+
+ Here we extend the IPW approach to the temporal setting to account for the dynamic nature of the outbreak by considering the joint likelihood
+ \begin{equation}
+ \label{eq:tvpropensity}
+ \sum_{t=1}^T \left[ \sum_{j=1}^N I^{NR}_{j,t} \log \left( \frac{\pi_t (X_{j,t}; \theta)}{1-\pi_t(X_{j,t}; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi_t (X_{j,t}; \theta) \right) \right]
+ \end{equation}
+ where $t=1,\ldots,T$ are the days when case-count data is reported, and $I^{NR}_{j,t}$ denotes self-selection into testing on day $t$, which is highly correlated with prior testing and results.
+For example, an individual who tests positive may be unlikely to seek testing in the subsequent few days/weeks.
+Moreover, an individual in a high prevalence area may be more likely to seek out testing.
+Here, we assume that the covariate vector $X_{j,t}$ contains all features of the past relevant for selection.
+
+ If sufficiently large random samples are collected at each time $t =1,\ldots,T$, then the pseudo-likelihood can be re-written as in~\eqref{eq:auxinfoprob} and propensities estimated separately per time point.
+Unfortunately, large probabilistic samples are not available at every time within a given region.
+To address this, here we consider a non-parametric kernel-based approach where the selection propensity at time~$t$, denoted $\hat \theta_t$, maximizes the smoothed pseudo-likelihood
+ $$
+ \sum_{t^\prime=1}^T K_h(|t^\prime - t|) \left[ \sum_{j=1}^N I_{j,t^\prime}^{NR} \log \left( \frac{\pi_t (X_{j,t^\prime}; \theta)}{1-\pi_t(X_{j,t^\prime}; \theta)} \right) + \sum_{j=1}^N W^{R}_{j,t^\prime} I^R_{j,t^\prime} \log \left( 1 - \pi_t (X_{j,t^\prime}; \theta) \right) \right]
+ $$
+ where $K_h$ is a kernel function with tunable parameter $h$.
+Given $\pi (x;\hat \theta_t)$, the prevalence estimator $\bar y_{n,t}^\star$ is given by~\ref{eq:ipwest} using case-count data observed on day~$t$.
+
+ \subsubsection{Asymptotics}
+ \label{section:asymptotics}
+
+ Here we suppose there is a sequence of finite populations of size $N_{\nu}$ indexed by $\nu$ and that there are $L_\nu$ non-probability samples and probability samples of size $n$ drawn at equally spaced times $\{ t^\prime_l \}_{l=1}^{L_{\nu}}$ over the study window~$[0,T]$.
+Assuming correct selection model specification, then under the probability sample design and nonprobability sample propensities, Lemma~\ref{lemma:ipw} shows that the IPW estimator at a time~$t \in \{ t^\prime_l \}$ is consistent as $\nu$ goes to infinity (i.e., $N_{\nu}, L_{\nu} \to \infty$) and calculates the estimator's variance.
+In Lemma~\ref{lemma:ipw}, sensitivity/specificity are unknown and estimated using a pseudo-likelihood on two random samples whose sizes (denoted~$n_{FP}$ and $n_{FN}$) reflect uncertainty in these quantities.
+See Appendix~\ref{app:asympderivations} for additional details.
+
+ \begin{lemma}[Variance of IPW estimator] \normalfont
+ \label{lemma:ipw}
+ % \textcolor{red}{Must check.}
+The estimates~$\hat \mu_t := \bar y_{n,t}^\star$, $\hat \pi_{j,t}$, $\hat FP$, and $\hat FN$ at a time~$t \in \{t_l^\prime\}$ are solutions to the following set of estimating equations
+ $$
+ \Phi_n (\eta_t) =
+ \left (
+ \begin{array}{c}
+ \frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t}{\pi_{j,t}} \\
+ \frac{1}{N \times \sum_{l = 1}^L K_h \left(|t - t_l^\prime| \right)} \sum_{l = 1}^L K_h \left(|t - t_l^\prime| \right) \left[ \sum_{j=1}^N I^{NR}_{j,t_l^\prime} X_{j,t_l^\prime} - \sum_{j = 1}^N I^{R}_{j,t_l^\prime} W^{R}_{j,t_l^\prime}  \pi_{j,t_l^\prime} X_{j,t_l^\prime}  \right] \\
+ \frac{1}{n_{FP}} \sum_{i=1}^{n_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
+ \frac{1}{n_{FN}} \sum_{i=1}^{n_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
+ \end{array}
+ \right ) = {\bf 0}
+ $$
+ where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
+Under certain regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (\bar n^{-1/2})$  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (\bar n^{-1})$ where $\bar n = n \sum_{l=1}^L K_{h} (|t - t_l^\prime|)$ and $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$ where $\phi_n = \frac{\partial \Phi (\eta)}{\partial \eta}$ with $E[\cdot]$ and $\text{Var} (\cdot)$ are under the joint randomization of propensity score and sampling designs.
+ \end{lemma}
+
+ \subsection{Model-based estimation}
+ \label{section:modelbased}
+
+ Up to this point, the primary focus has been selection bias in coronavirus case-count data from a survey sampling perspective.
+Here, we consider compartmental model approaches from infectious disease epidemiology.
+Our primary objective is a model-based forecast of strata-level prevalence $\bar y_k$.
+To do this, a probabilistic extension of a standard epidemiological state-space model -- the susceptible, exposed, infected, and removed (recovered and death) model, or SEIR model -- is presented.
+A probabilistic SIR model was originally proposed by~\cite{Osthus2017} with only one-dimensional time series of infected proportions; this formulation was extended by~\cite{Song2020} to model coronavirus case-counts.
+
+ Let $s_t$, $e_t$, $i_t$, and $r_t$ denote the proportion of survivors, exposed, infected, and removed cases (i.e., including both recovered cases and deaths) at time $t$.
+The population-level SEIR dynamics are given by the set of differential equations in~\eqref{eq:seir}.
+% \begin{align}
+%  \frac{\partial s_t}{\partial t} &= - \beta s_t i_t; \quad \frac{\partial e_t}{\partial t} = \beta s_t i_t - \sigma e_t; \quad \label{eq:seir} \\
+%  \frac{\partial i_t}{\partial t} &= \sigma e_t - \gamma i_t; \quad
+%  \frac{\partial r_t}{\partial t} = - \gamma i_t. \nonumber
+% \end{align}
+%Following~\cite{Song2020}, we employ Runge-Kutta (RK4) approximations for discretization.
+Here, we consider covariate information that takes the form of a stratification variable with $K$ strata.
+Let $I_{k,t}$ to denote the number of new infections on day $t$ in the $k$th strata.
+The joint distribution of strata-specific new infection totals follows a multinomial distribution with
+ $$
+ ( I_{1,t}, \ldots, I_{K,t} ) \sim \text{Multinomial} \left( I_t, (p_{1,t}, \ldots, p_{K, t}) \right).
+ $$
+ where $( {\bf p}_t )_{t=1}^T = ( (p_{1,t}, \ldots, p_{K, t}) )_{t=1}^T$ are a sequence of parameters on the simplex, i.e., $\sum_{k=1}^K p_{k,t} = 1$.
+
+ Selection bias is addressed by analyzing COVID-19 death data rather than case counts.  Let $D_{k,r}$ denote the number of individuals who pass away on day $r$ from strata $k$.  Then
+ $$
+ D_{k,r} \mid p, \theta, \nu \sim \text{Poisson} \left( \sum_{t=1}^r \text{IFR}_{k} \cdot I_{k,t} \theta_{(r-t)} \right)
+ $$
+ where $\text{IFR}_{k}$ is the infection fatality rate for the $k$th strata and $\theta_{j}$ is a discrete-time distribution for time from infection to death.
+This extends prior analysis of death data~\cite{Johndrow2020} by allowing the infection fatality rate to depend on a stratification variable, which is important as IFR depends heavily on age~\citep{Levin2020}.
+See Section~\ref{section:application_of_modelbased} for discussion of parameter choices.
+%Additional details on inference are discussed in Appendix~\ref{app:modelbased}.
+
+ % 0-17: 20/1000000 * 100 -> 0.002 (at age 9)
+ % 18-49: 500/1000000 * 100 -> 0.05 (at age 33)
+ % 50–64 years old: 6,000/1000000 * 100 -> 0.6
+ % 65+ years old: 90,000/1000000 * 100 -> 9.0
+ %
+
+ \subsection{Doubly robust estimation.}
+ Rather than relying solely on IPW or epidemiological forecasts, here we combine forecasting and inverse-probability weighting by extending recent work by~\cite{Chen2019} to account time-varying propensities and measurement-error.
+ Let~$\mu(x)$ denote the posterior mean of the active infection rate individuals with covariate value~$x$ (i.e., age strata) based on the SEIR model described in SEction~\ref{section:modelbased}.
+In this context, the doubly-robust estimator is given by
+ $$
+ \bar y_{n}^{(DR)} = \frac{1}{N} \sum_{j=1}^N \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \hat \mu(x_j) \right).
+ $$
+ where $\hat \mu(x_j)$ is not corrected for measurement-error as it estimates true case counts.
+This estimator is called ``doubly-robust'' because it is consistent if either the model-based forecasts or the time-varying propensities are correctly specified.
+A statistical error decomposition can be derived
+ \begin{align}
+ \label{eq:statdecomp3}
+ \begin{split}
+ \rho_{\tilde I (X), Y-\mu(X)} &\times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y-\mu(X)}  \\
+ &\times \underbrace{\left[ 1 - \frac{\rho_{\tilde I(X),Y} \sigma_{Y}}{\rho_{\tilde I(X),Y-\mu(X)} \sigma_{Y-\mu(X)}} \times \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right]}_{\tilde D_M}.
+ \end{split}
+ \end{align}
+ See Appendix~\ref{app:DRderivation} for the derivation.
+ % Comparing~\eqref{eq:statdecomp3} to~\eqref{eq:statdecomp2} shows that the addition of the model-based forecast impacts the estimation error in two ways.
+ If the model is adequate, then one may expect a reduction in the problem difficulty and (potentially) in the data quality components.
+ % Note that the model-based forecast does not impact the data quantity component.
+% Second, if the model-based forecast is correct then $E_{\bf R} [ \rho_{\tilde I (X), Y-\mu(X)} ] = 0$  and therefore $\rho_{\tilde I(X), Y-\mu(X)} = O(N^{-1})$; however, if the forecasts are not correctly specified then the data quality index is unlikely to inversely scale with population size.
+ Interestingly, the impact of measurement-error on data quality now depends on a relative comparison of the data quality and problem difficulty of the weighted estimator and the doubly-robust estimator.
+ % For decent models, one would expect this ratio to be greater than one.
+Therefore, $\tilde \Delta < 0$ implies an expected increase in impact of measurement-error and $\tilde \Delta > 0$ implies the magnitude could shrink toward zero or change sign and have an even large magnitude in the opposite direction.
+Again, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from a DR estimator are likely to put too much confidence in the sheer size of the data.
+
+ Unfortunately, the first term of the doubly robust estimator cannot be computed as covariate information is not collected on the entire population.
+Here, we suppose the same asymptotic regime as in Section~\ref{section:asymptotics} and use the probability sample at the time~$t \in \{t_l^\prime\}$ to estimate this term:
+ $$
+ \frac{1}{\sum_{j=1}^N \tilde I_{j,t} \tilde W_{j,t}} \sum_{j=1}^N  \tilde I_{j,t} \tilde W_{j,t} \hat \mu_{t} (X_{j,t}) + \frac{1}{\sum_{j=1}^N I_{j,t} w (X_{j,t})} \sum_{j=1}^N I_{j,t} w(X_{j,t}) \left( \frac{Y_{j,t} - FP}{1 - FP - FN} - \hat \mu_t(X_{j,t}) \right).
+ $$
+ Assuming correct selection propensity model specification, then under the probability sample design and nonprobability sample design, Lemma~\ref{lemma:dr} shows that the doubly robust estimator is consistent and calculates the estimator's variance.
+See Appendix~\ref{app:asympderivations} for additional details.
+
+ \begin{lemma}[Variance of doubly-robust estimator] \normalfont
+ \label{lemma:dr}
+ The estimates~$\mu_t := \bar y_{n,t}^\star$ and $\hat \pi_{j,t}$
+ are solutions to the following set of estimating equations
+ $$
+ \Phi_n (\eta_t) =
+ \left (
+ \begin{array}{c}
+ \frac{1}{N} \left[ \sum_{j=1}^N \frac{I_{j,t}}{\pi_{j,t}} \left( \left( \frac{Y_{j,t} - FP}{1-FP-FN} - \hat \mu_{j,t}  \right)  - \mu_t \right) + \frac{\sum_{l=1}^L K_{h} (|t - t_l^\prime|) \sum_{j=1}^N \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime} \hat \mu_{j,t_l^\prime}}{\sum_{l=1}^L K_{h} (|t - t_l^\prime|) \sum_{j=1}^N \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime}} \right] \\
+ \frac{1}{N \times \sum_{l=1}^L K_{h} (|t - t_l^\prime|)} \sum_{l=1}^L K_{h} (|t - t_l^\prime|) \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
+ \frac{1}{n_{FP}} \sum_{j=1}^{n_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
+ \frac{1}{n_{FN}} \sum_{j=1}^{n_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
+ \end{array}
+ \right ) = {\bf 0}
+ $$
+ where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
+Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (\bar n^{-1/2})$ and $\text{var} (\hat y_{n,t}) = V_{t}^{(DR)} + o (\bar n^{-1})$ where $\bar n = n \sum_{l=1}^L K_h(|t-t_l|)$ and $V_t^{(DR)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$ where $\phi_n = \frac{\partial \Phi (\eta)}{\partial \eta}$ with $E[\cdot]$ and $\text{Var} (\cdot)$ are under the joint randomization of propensity score and sampling designs.
+ \end{lemma}
+
+ \section{COVID-19 active infection prevalence in Indiana}
+ \label{section:applications}
+
+ % For COVID-19 case-count data, the only covariate information provided by state and local governments are summary statistics for those factors discussed in Section~\ref{section:testinginfo} -- age, sex,  and race.
+These are discrete factors and therefore define a stratification of the population.
+Currently, US testing does not collect additional covariate information such as symptom information; however, this need not be the case moving forward.
+Recommendation XX below states that additional covariate information should be collected from nonprobabilistic samples.
+
+ We next consider estimation of the active infection rate in Indiana using unweighted, IPW, model-based, and doubly robust estimates. To start, we recap the data sources used in estimation:
+ \begin{itemize}[leftmargin=*]
+ \item {\bf Testing data}:  The number of daily tests performed and number of daily positive tests are reported.
+These counts are broken out jointly by age, gender, and racial demographic information.
+Figure~\ref{fig:in-cases} and~\ref{fig:in-tests} plots daily COVID-19 positive cases and tests by age strata.
+
+ \item {\bf Random/Nonrandom statewide sample}: Table~\ref{tab:indiana} summarizes data from a random sample from April 25--29th as well as a nonrandom sample obtained between May 2--3 in racial/ethnic minority communities~\citep{Yiannoutsos2021}.
+The nonrandom sample is not a random subsample of the overall case-counts; however, this data provides important supplementary covariate information.
+ % Namely, the samples collected included covariate information on (1) prior symptoms, (2) COVID-19 positive individual in household, and (3) prior COVID-19 positive test.
+ To account for the nonrandom sample being from high risk areas, we adjust estimates to match the statewide positivity rate of 11.7\% on August 30th.
+These adjusted rates are presented in parentheses in Table~\ref{tab:indiana} along with the relevant covariate information.
+
+ \item {\bf Delphi's COVID-19 Trends and Impact Survey}: From the daily symptom surveys completed as part of Delphi's CTIS (see Section~\ref{subsection:fbsymptom} for details), we extract survey responses for individuals who identify as living in Indiana.
+We collect age, sex, and demographic information as well as COVID-19 related symptoms, which Delphi defines as having (1) a fever and cough, (2) shortness of breath, \emph{or} (3) difficulty breathing.
+Table~\ref{tab:comparison} in Section~\ref{app:in_add_details} of the supplementary materials shows minimal bias with respect to symptom distributions when comparing to Indiana's random sample to the CTIS data collected from April 25th to April 29th.
+
+ % \begin{table}[!th]
+ % \centering
+ % \begin{tabular}{c | c c c | c}
+ % & \multicolumn{3}{c}{Facebook} & Random \\ \cline{2-4}
+ % & Lower CI & Estimate & Upper CI & Estimate \\ \hline
+ % Fever &  0.017 & 0.020 & 0.024 & 0.018 \\
+ % Cough &  0.142 & 0.148 & 0.154 & 0.149 \\
+ % Shortness & 0.058 & 0.060 & 0.062 & 0.062 \\ \hline
+ % \end{tabular}
+ % \caption{Comparison of Facebook survey and random sample}
+ % \label{tab:comparison}
+ % \end{table}
+
+ % \item {\bf Hospitalization data}: On each day, we observe the number of individuals who are COVID-19 positive that visited a hospital on that day.
+Unfortunately, hospitalization data is not broken out by age, gender, and racial demographic information.
+Instead, we rely on combining Facebook data to calculate the fraction of those who sought the hospital per week in each strata.
+We then use this to figure out how many individuals per strata were hospitalized and then back out how many
+
+ \item {\bf Death data}: Daily COVID-19 related deaths are observed by age, gender, and racial demographics.
+Figure~\ref{fig:in-deaths} plots COVID-19 related deaths over time by age group.
+While most tests and positive cases are within the younger age strata, most deaths are within the 70+ age strata.
+ \end{itemize}
 
@@
 \textcolor{red}{
- While this paper emphasizes the importance of random sampling, we do not naively view it as a panacea.
+ While this paper emphasizes the importance of random sampling, we do not view its use as a panacea.
 With access to \emph{only} a non-probabilistic sample, we have no means to reduce self-selection bias.

### 2022-03-22 write-up/letters/response-ref-1-v1.tex
updated paper

@@
 	The adjustment is exactly related to the Bayes rule PPV/NPV relationship with FP, FN, and $\bar Y$.
-	\textcolor{red}{Got to here.}
+	\textcolor{red}{I'm still a bit confused by this point.  Will review further.}
 	\vspace{5mm}
@@
 
-	We clarify that increasing testing capacity may impact FP and FN rates if altering changes the sample collection quality, e.g., a testing center switches from nasal swab to oropharyngeal swabs or saliva specimen to increase testing capacity will change measurement error due to the different test characteristic.
+	We clarify that increasing testing capacity may impact FP and FN rates if altering changes the sample collection quality, e.g., a testing center switches from nasal swab to oropharyngeal swabs or saliva specimen to increase testing capacity will alter measurement error due to differences in the test characteristics.
 \end{enumerate}
@@
 We highlight that the SEIR model has been shown to better represent the SARS-CoV-2 dynamics.
-Figure 3a presents updated versions of the potnetial bias in ratio estimators under an SEIR model.
+Figure 3a presents updated versions of the potential bias in ratio estimators under an SEIR model.
 The conclusion remains largely the same.
-The only difference from the prior manuscript is that the peak in bias prior to the ratio going less than 1 is less pronounced (was up to 5, now up to 2).
+The only difference from the prior manuscript is that the peak in bias prior to the ratio going less than 1 is less pronounced (prior submission with SIR model has an approximate peak of 5 while current manuscript has approximate peak of 2).
 	\vspace{5mm}
@@
 	, this is an approximate solution which lacks the term $\Lambda^2 \frac{1}{\gamma \sigma}$.
-``It corresponds to the case of an exponentially distributed generation time with mean duration $1/\gamma + 1/\sigma$, which is the same as the solution for the SIR model assuming an infectious period of $1/\gamma + 1/\sigma$.
+It corresponds to the case of an exponentially distributed generation time with mean duration $1/\gamma + 1/\sigma$, which is the same as the solution for the SIR model assuming an infectious period of $1/\gamma + 1/\sigma$.
 For realistic growth rates and choices of $\sigma$ and $\gamma$ for the COVID-19 pandemic, this term is negligible and therefore we work with the approximate formula.
@@
 	\noindent Note that for the SIR model, the basic reproduction number is~$R_0 = 1 + \frac{1}{\gamma} \Lambda$.
-So estimate of the instantaneous reproduction number is obatined by replacing $\Lambda$ by an estimate of the  instantaneous growth rate $\hat \Lambda_t = \log (K_t / K_{t-1})$.
+So estimate of the instantaneous reproduction number is obtained by replacing $\Lambda$ by an estimate of the  instantaneous growth rate $\hat \Lambda_t = \log (K_t / K_{t-1})$.
 Then using the approximate relationship between the SEIR model and an SIR model with infectious period $1/\gamma + 1/\sigma$, we arrive at the new moment-based estimator
@@
 
-	Section 4.1.3 allows researchers to choose to go beyond simpler moving window averages (i.e., weekly averages).
-The kernel function allows the user to choose weighted averages that may be more robust than the uniform kernel (which is equivalent to the moving averages).
-This approach helps smooth the estimated propensity over time; moreover, the flexibility of the approach for government agencies who may observed random samples at different frequencies is of practical importance.
+	Section 4.1.3 allows researchers to choose to go beyond simpler moving window averages (i.e., daily/weekly averages).
+The kernel function allows the user to choose weighted averages that may be more robust than the uniform kernel (which is equivalent to the a moving average over some fixed window of time).
+The proposed approach helps smooth the estimated propensity over time; moreover, the flexibility of the approach for government agencies who may observed random samples at different frequencies is of practical importance.
 	\vspace{5mm}

### 2022-03-22 write-up/covid-supplement-writeup.tex
updated paper

@@
 
-Table~\ref{tab:comparison} demonstrates minimal bias in the Facebook symptom survey with respect to symptom distributions.
+Table~\ref{tab:comparison} demonstrates minimal bias in the Delphi’s COVID-19 Trends and Impact Survey (CTIS) with respect to symptom distributions.
 
@@
  \begin{tabular}{c | c c c | c}
- & \multicolumn{3}{c}{Facebook} & Random \\ \cline{2-4}
+ & \multicolumn{3}{c}{CTIS} & Random \\ \cline{2-4}
  & Lower CI & Estimate & Upper CI & Estimate \\ \hline
@@
  \end{tabular}
- \caption{Comparison of Facebook survey and random sample}
+ \caption{Comparison of CTIS and random sample restricting CTIS to surveys collected from April 25--29th, 2020.}
  \label{tab:comparison}
@@
 
-Here we present the estimate of~\cite{Cori20113}.
-Let~$I_t$ detnoe the total infectiousness of infected individuals at time~$t$.
+Here we present estimates of the instantaneous reproductive number using the approach of~\cite{Cori20113} under the SEIR model from Section~\ref{section:r0-estimation}.
+Let~$I_t$ denote the total infectiousness of infected individuals at time~$t$.
 Then~$E[I_t] = R_t \sum_{s=1}^{t} I_{t-s} w_s$ where $w_{u}$ is the infectivity function.
@@
 Therefore, the bias can be readily obtained from the associated Taylor series decomposition, with the terms related to time~$t-1$ replaced by a weighted version.
+Figure~\ref{fig:cori-bias} presents the bias.
 

### 2022-03-22 write-up/covid-writeup.tex
updated paper

@@
 Researchers often assume measurement error leads to parameter attenuation.
-When paired with selection bias, however, the two sources become entangled, and resulting errors in magnified, muted, or even switch signs.
+When paired with selection bias, however, the two sources become entangled, and resulting errors can be magnified, muted, or even switch signs.
 
@@
  Note that test positivity is based on an underlying continuous cycle threshold and therefore not strictly binary.
-While we focus on the interplay of measurement error and selection bias in the context of binary outcomes due to the dichotomous nature of the COVID-19 testing data, it should be noted that measurement error can lead to bias even in the absence of selection bias.
+We focus on the interplay of measurement error and selection bias in the context of binary outcomes due to the dichotomous nature of the COVID-19 testing data.
+Note that measurement error can lead to bias even in the absence of selection bias.
  \end{remark}
@@
  \includegraphics[width = 0.7\textwidth]{../figs/mem_heatmap.png}
- \caption{Imperfect testing adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.024$ and $FN=0.13$. Color scaled so blue = $-10$, white = $0$, and red = $10$.}
+ \caption{Imperfect testing adjustment~($D_M$) contour plot as a function of relative frequency $f_1/f_0$ (x-axis) and odds ratio (y-axis) for $FP=0.024$ and $FN=0.13$.}
  %\textcolor{red}{NEED TO DO FOR FP = 0.05 and FN = 0.3}}
@@
 Then the effective sample size is $863$, representing a $5.1$ factor increase rather than the expected $6.1$ factor increase.
-See Section~\ref{section:effss} in the Supplementary Materials for additional effective sample size calculations which show that while~\citep{Meng2018} argues the relative error increases as a function of population size, our calculations indicate this is not true for fixed relative frequency~$f_1/f_0$.
+See Section~\ref{section:effss} in the Supplementary Materials for additional effective sample size calculations.
+Note that while~\citep{Meng2018} argues the relative error increases as a function of population size, our calculations indicate this is not true when relative frequency~$f_1/f_0$ is held fixed.
  % \begin{wrapfigure}{r}{0.5\textwidth}
@@
  % \end{equation}
- where $s_t, e_t, i_t$ and $r_t$ are the fraction of susceptible, exposed infected, and removed (recovered or deceased) individuals in the population at time $t$ respectively.
+ where $s_t, e_t, i_t$ and $r_t$ are the fraction of susceptible, exposed, infected, and removed (recovered or deceased) individuals in the population at time $t$ respectively.
 The SEIR model has been used extensively as a model for SARS-CoV-2 dynamics~\cite{Song2020}.
@@
  Expectation of~\ref{eq:auxinfoprob} with respect to the sampling design yields~\eqref{eq:propensity}.
-Solving~\eqref{eq:auxinfoprob} is done by iteratively re-weighted least squares (IRLS); see Section~\ref{app:irls} in the supplementary materials for details.
-For ease of comprehension, a notation glossary is provided in Section~\ref{app:notation} in the the supplementary materials.
+Solving~\eqref{eq:auxinfoprob} is done by iteratively re-weighted least squares (IRLS); see Section~\ref{app:irls} and Section~\ref{app:notation} in the supplementary materials for details and a notation glossary respectively.
 
@@
 Survey sampling and transportability weights require knowledge of the selection mechanism~\citep{10.1093/aje/kwx164,ColeStuart2010}, which is not available in our setting.
-Instead, we require auxiliary information in the form of a random sample to estimate the probability of selection.
-A key difference from prior approaches is the need to estimate time-varying probabilities of selection, e.g., Figure 6 shows how the selection propensity of young individuals changes dramatically over the observation window.
+Instead, here we require auxiliary information in the form of a random sample to estimate the probability of selection.
  \end{remark}
@@
  % \textcolor{red}{Must check.}
-The estimates~$\hat \mu_t := \bar y_{n,t}^\star$, $\hat \pi_{j,t}$, $\hat FP$, and $\hat FN$ at a time~$t \in \{t_l^\prime\}$ are solutions to the following set of estimating equations
+The estimates~$\hat \mu_t := \bar y_{n,t}^\star$, $\hat \pi_{j,t}$, $\hat{FP}$, and $\hat{FN}$ at a time~$t \in \{t_l^\prime\}$ are solutions to the following set of estimating equations
  $$

### 2022-03-22 write-up/covid-writeup.tex
Updated DR part

@@
  Rather than relying solely on IPW or epidemiological forecasts, here we combine forecasting and inverse-probability weighting by extending recent work by~\cite{Chen2019} to account time-varying propensities and measurement-error.
- Let~$\mu(x)$ denote the posterior mean of the active infection rate individuals with covariate value~$x$ (i.e., age strata) based on the SEIR model described in SEction~\ref{section:modelbased}.
+ Let~$\mu(x)$ denote the posterior mean of the active infection rate individuals with covariate value~$x$ (i.e., age strata) based on the SEIR model described in Section~\ref{section:modelbased}.
 In this context, the doubly-robust estimator is given by

### 2022-03-23 write-up/covid-writeup.tex
Updated DR part
updated paper

@@
  $$
- \bar y_{n}^{(DR)} = \frac{1}{N} \sum_{j=1}^N \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \hat \mu(x_j) \right).
+ \bar y_{n}^{(DR)} = \frac{1}{N} \sum_{j=1}^N \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j^{NR} w (x_j)} \sum_{j=1}^N I_j^{NR} w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \hat \mu(x_j) \right).
  $$
@@
  \begin{split}
- \rho_{\tilde I (X), Y-\mu(X)} &\times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y-\mu(X)}  \\
- &\times \underbrace{\left[ 1 - \frac{\rho_{\tilde I(X),Y} \sigma_{Y}}{\rho_{\tilde I(X),Y-\mu(X)} \sigma_{Y-\mu(X)}} \times \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right]}_{\tilde D_M}.
+ \rho_{I^{NR} (X), Y-\mu(X)} &\times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y-\mu(X)}  \\
+ &\times \underbrace{\left[ 1 - \frac{\rho_{I^{NR}(X),Y} \sigma_{Y}}{\rho_{I^{NR} (X),Y-\mu(X)} \sigma_{Y-\mu(X)}} \times \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right]}_{\tilde D_M}.
  \end{split}
@@
  $$
- \frac{1}{\sum_{j=1}^N \tilde I_{j,t} \tilde W_{j,t}} \sum_{j=1}^N  \tilde I_{j,t} \tilde W_{j,t} \hat \mu_{t} (X_{j,t}) + \frac{1}{\sum_{j=1}^N I_{j,t} w (X_{j,t})} \sum_{j=1}^N I_{j,t} w(X_{j,t}) \left( \frac{Y_{j,t} - FP}{1 - FP - FN} - \hat \mu_t(X_{j,t}) \right).
+ \frac{1}{\sum_{j=1}^N I^{R}_{j,t} W^R_{j,t}} \sum_{j=1}^N  I_{j,t}^{R} W_{j,t}^{R} \hat \mu_{t} (X_{j,t}) + \frac{1}{\sum_{j=1}^N I_{j,t}^{NR} w (X_{j,t})} \sum_{j=1}^N I_{j,t}^{NR} w(X_{j,t}) \left( \frac{Y_{j,t} - FP}{1 - FP - FN} - \hat \mu_t(X_{j,t}) \right).
  $$

### 2022-03-24 write-up/letters/response-ref-1-v1.tex
Updated DR part
updated paper
Just fixing SEIR model.

@@
 
-	\textcolor{red}{We clarify that we define active infection as individuals infected with SARS-CoV-2 who have yet to fall below a threshold for detectable viral load levels by RT-PCR testing.}
+	\textcolor{red}{We now clarify that we define active infection as individuals infected with SARS-CoV-2 who have yet to fall below a threshold for detectable viral load levels by RT-PCR testing.
+This includes the infectious period as well as a pre-infectious and post-infectious periods of time.}
 
@@
 
-	\textcolor{red}{We clarify that we define active infection as individuals infected with SARS-CoV-2 who have yet to fall below a threshold for detectable viral load levels by RT-PCR testing.}
+	\textcolor{red}{As stated in answer to Question 3a, we now clarify that we define active infection as individuals infected with SARS-CoV-2 who have yet to fall below a threshold for detectable viral load levels by RT-PCR testing.
+As stated here, this includes the infectious period as well as a pre-infectious and post-infectious periods of time.
+We discuss how this impacts the compartmental model and calculation of the AIR in answer to Questions 6a and 17a.}
 	\vspace{5mm}
@@
 We clarify this in Section 5.2 and comment on the code to clarify this more carefully in Appendix H.
-	\textcolor{red}{Need To Do.}
+	\textcolor{red}{Need To Do.  Keep model just sensitivity on IFR and smooth over infectious.}
 	\vspace{5mm}
@@
 
-	We added Remark 6 to emphasize these potential limitations.
-A key point is that the case study is to demonstrate our framework for answering the \emph{what to do with this knowledge of selection bias and measurement error} point that was raised in the first round of submission.
+	We added Remark 6 to emphasize these potential limitations and highlight how the scientific teams that collected these data sources aimed to address these important issues.
+A key point is that the case study is to demonstrate our framework for answering the \emph{what to do with this knowledge of selection bias and measurement error} point that was raised by reviewers during the first submission round.
 In Remark 6, we urge health policy experts and government officials to assess these issues when applying this framework in their own future work.
@@
 	Section 5.1.1 studies testing between April 25th to 29th 2020 during which testing was limited.
-This led us to the assumption that the individuals who already tested were unlikely to be available for testing.
-We continue to calculate the sampling fraction as number of tests divided by the total population minus prior tests, but note at the end of Section 5.1.1 that using the total population in the denominator yields a very similar sampling fraction and the calculations have negligible change.
+This led us to the assumption that the individuals who already tested were unlikely to be available for testing as it had only been available since March 2020 and quite limited in supply.
+We continue to calculate the sampling fraction as number of tests divided by the total population minus prior tests.
+To address the reviewers concern, we note at the end of Section 5.1.1 that using the total population in the denominator yields a very similar sampling fraction and the calculations change by a negligible amount.
 	\vspace{5mm}
@@
 
-	See answers to Questions 5a and 14a above.  The prevalence estimate comes from Yiannoutsos et al. (2021) who detail their adjustments for response rates in estimation of active prevalence.
+	See answers to Questions 5a and 14a above.
+The prevalence estimate comes from Yiannoutsos et al.
+(2021) who detail their adjustments for response rates and measurement error in estimation of active prevalence.
+We do not attempt to correct it further, and simply note that (unlike in Meng's case where the ``truth'' is observed) this calculation is approximate and meant to help us understand what the relative sampling rates may be in our setting.
 	\vspace{5mm}
@@
 	We thank the reviewer for raising this issue.
-We have added a remark in sction 5.1.2 that raises this as a limitation.
-We restart here that the need for the approaches is due to the lack of availability of symptom status (which we know the state collects but has told us is unavailable due to data privacy / confidentiality issues).
-Therefore, as a framework for health policy experts and government officials who have access to the complete government datasets (i.e., not just the public data), these steps will not be necessary.
+We have added a sentence in section 5.1.2 that raises this as a limitation.
+We restate here that the need for the imputation approaches is due to the lack of publicly available symptom status information.
+We have confirmed that this data is collected by the state, but were informed that the data could not be released due to data privacy / confidentiality issues.
+Therefore, as a framework for health policy experts and government officials who have access to the complete government datasets (i.e., not just the public data), these steps would not be necessary.
 	\vspace{5mm}
@@
 
-	See our answer to Question 12 above.
+	\textcolor{red}{See our answer to Question 12 above.
 In Section 5.2, we now clarify how we calculate the infection-fatality rate from the compartments of the model.
 Similar to Johndrow (2021), the model only uses COVID-19 related deaths.
-Deaths to non-COVID causes are not directly modeled but would appear as part of the R component in the global SEIR model.
+Deaths to non-COVID causes are not directly modeled but would appear as part of the R component in the global SEIR model.}
 	\vspace{5mm}
@@
 
-    We contextualize the discussion of IFR in Remark 7 in Section 5.2.
+    \textcolor{red}{We contextualize the discussion of IFR in Remark 7 in Section 5.2.
 We acknowledge that these IFR estimates are likely biased (or potentially accurate in very different circumstances).
-A sensitivity analysis focused on the issue of choosing an appropriate IFR given the potential biases in available IFR estimates is now performed.
+A sensitivity analysis focused on the issue of choosing an appropriate IFR given the potential biases in available IFR estimates is now performed.}
     \textcolor{red}{To Do!}
@@
 We now make this distinction much clearer.
+The section focuses primarily on RT-PCR tests given they are the primary molecular test offered by the Indiana's State Department of Health (ISDH) and thus the basis for the case count and testing data in this paper.
 \vspace{5mm}
@@
 \vspace{5mm}
-We apologize for the confusion.
+\textcolor{red}{We apologize for the confusion.
 See answer to Question 3b where we clarify the definition of active infection.
 An active infection was meant to refer to individuals who were infected with SARS-CoV-2 who have yet to fall below a threshold for detectable viral load levels by RT-PCR testing.
-This includes the infectious period as well as a pre-infectious and post-infectious periods of time.
+This includes the infectious period as well as a pre-infectious and post-infectious periods of time.}
 
@@
 \vspace{5mm}
-See our answer to Question 17b of Reviewer \#1.
+\textcolor{red}{See our answer to Question 17b of Reviewer \#1.
 We acknowledge that these IFR estimates are likely biased and time-varying due to introduction of vaccines in 2020.
-A sensitivity analysis focused on the issue of choosing an appropriate IFR and allowing for the IFR to be time-varying is now performed.
+A sensitivity analysis focused on the issue of choosing an appropriate IFR and allowing for the IFR to be time-varying is now performed.}
 \textcolor{red}{To Do!}
@@
 \vspace{5mm}
-We have fixed Figure 3b to match on covariates and have matching y-axes.  See our answer to Question 16b from Reviewer \#1 on additional figures we are including in the revision.
-\textcolor{red}{Need to do.}
+We have fixed Figure 3b to match on covariates and have matching y-axes.
+See our answer to Question 16b from Reviewer \#1 on additional figures we are including in the revision.
+The figure now shows that testing propensity is time-varying, increasing from near 0 in the first few months to a peak of 0.01 around December.
+This peak is 8 times lower than the peak in December for the same strata given fever and COVID-19 contact.
 \vspace{5mm}
@@
 \vspace{5mm}
-See our answer to Question 17b of Reviewer \#1 and Question 7 above. A sensitivity analysis focused on the issue of choosing an appropriate IFR is now performed.
+\textcolor{red}{See our answer to Question 17b of Reviewer \#1 and Question 7 above. A sensitivity analysis focused on the issue of choosing an appropriate IFR is now performed.}
 \textcolor{red}{To Do!}
@@
 
-		We have removed this and simply state the SEIR parameter values ($\beta = 1.2, \gamma = 0.14, \sigma = 0.3$) underlying the Figure.
+		We have removed this and simply state the SEIR parameter values ($\beta = 1.2, \gamma = 0.15, \sigma = 0.3$) underlying the Figure.
 
@@
 This also claims to show the likelihood for all age ranges - I believe it is only for males.}
+		\vspace{5mm}
+
+		We have fixed this reference.
+
 	\end{enumerate}

### 2022-03-24 write-up/covid-supplement-writeup.tex
Just fixing SEIR model.

@@
  \includegraphics[width=.9\linewidth]{../figs/tvprop_fig1.png}
- \caption{Testing Likelihood Given Fever and COVID-19 Contact}
+ \caption{Testing Likelihood Given Fever and \\COVID-19 Contact}
  \label{fig:testinglik1}
@@
 \includegraphics[width=.9\linewidth]{../figs/tvprop_fig2.png}
- \caption{Testing Likelihood Given Fever and COVID-19 Contact}
+ \caption{Testing Likelihood Given no Fever nor \\  COVID-19 Contact}
  \label{fig:testinglik2}
@@
 Based on mean-imputation of COVID-19 fever using hospitalization records indicators, we can use the proposed pseudo-likelihood approach to compute the likelihood of getting tested for COVID-19 given age, ethnicity, race, gender, and fever status.
-Figure~\ref{fig:testinglik1_mainpaper} and~\ref{fig:testinglik1_mainpaper} in Section~\ref{section:tvipw} of the manuscript plots the likelihood of testing given fever and no fever respectfully for all age ranges.
+Figure~\ref{fig:testinglik1_mainpaper} and~\ref{fig:testinglik2_mainpaper} in Section~\ref{section:tvipw} of the manuscript plots the likelihood of testing given fever and no fever respectfully for all age ranges.
 We see that the likelihood is time-varying and depends heavily on both fever status.

### 2022-03-24 write-up/covid-writeup.tex
Just fixing SEIR model.

@@
 A \emph{molecular test} refers to diagnostic tests that aim to detect viral load above a certain threshold (e.g., RT-PCR or antigen tests); see~\cite{Mina2020} for a detailed discussion of cycle thresholds.
-If someone has an active infection -- \textcolor{red}{here defined as individuals infected with SARS-CoV-2 who's viral load has yet to fall below detectable levels by molecular test} -- then after the incubation period, a molecular test with perfect sensitivity will yield a positive result for the next few weeks.
-After that, the viral load will decrease and a molecular test with perfect specificity will come back negative.
+If someone has an active infection -- \textcolor{red}{here defined as individuals infected with SARS-CoV-2 who's viral load has yet to fall below detectable levels by RT-PCR testing} -- then after the incubation period, a molecular test with perfect sensitivity will yield a positive result while the patient has a viral load above the threshold of detection.
+After that, the viral load will decrease below that threshold and a molecular test with perfect specificity will come back negative.
 While an infected individual may yield a positive molecular test for several weeks, they will likely stop transmitting the disease within a few days of infection, meaning a positive molecular test does not imply transmissibility.
@@
  \begin{array}{c}
- \frac{1}{N} \left[ \sum_{j=1}^N \frac{I_{j,t}}{\pi_{j,t}} \left( \left( \frac{Y_{j,t} - FP}{1-FP-FN} - \hat \mu_{j,t}  \right)  - \mu_t \right) + \frac{\sum_{l=1}^L K_{h} (|t - t_l^\prime|) \sum_{j=1}^N \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime} \hat \mu_{j,t_l^\prime}}{\sum_{l=1}^L K_{h} (|t - t_l^\prime|) \sum_{j=1}^N \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime}} \right] \\
- \frac{1}{N \times \sum_{l=1}^L K_{h} (|t - t_l^\prime|)} \sum_{l=1}^L K_{h} (|t - t_l^\prime|) \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
+ \frac{1}{N} \left[ \sum_{j=1}^N \frac{I^{NR}_{j,t}}{\pi_{j,t}} \left( \left( \frac{Y_{j,t} - FP}{1-FP-FN} - \hat \mu_{j,t}  \right)  - \mu_t \right) + \frac{\sum_{l=1}^L K_{h} (|t - t_l^\prime|) \sum_{j=1}^N I^{R}_{j,t_l^\prime} W^{R}_{j,t_l^\prime} \hat \mu_{j,t_l^\prime}}{\sum_{l=1}^L K_{h} (|t - t_l^\prime|) \sum_{j=1}^N I^{R}_{j,t_l^\prime} W^{R}_{j,t_l^\prime}} \right] \\
+ \frac{1}{N \times \sum_{l=1}^L K_{h} (|t - t_l^\prime|)} \sum_{l=1}^L K_{h} (|t - t_l^\prime|) \sum_{j=1}^N I^{NR}_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N I^{R}_{j,t^\prime} W^{R}_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
  \frac{1}{n_{FP}} \sum_{j=1}^{n_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
@@
 \begin{remark}
- We acknowledge several limitations with respect to the representativeness.
+ We acknowledge several limitations with respect to representativeness of these samples.
 First, the random sample had significant missingness ($3,658/15,495 \approx 24\%$ response rate).
@@
 This leads to a range for $M$ of $(5.60, 6.44)$ for the unweighted analysis, and $(3.63,4.07)$ and $(3.10, 3.45)$ for the weighted analyses.
-Note that in calculating the sampling fraction~$f$, we assumed individuals who tested recently are unlikely to test in this time window.
-If we instead do not subtract off these tests, the sampling fraction is $f = 2.92 \times 10^{-3}$ and the above calculations have negligible change.
+Note that in calculating the sampling fraction~$f$, we assumed individuals who tested recently are unlikely to test again in this time window.
+If we instead do not subtract off these tests, the sampling fraction is $f = 2.92 \times 10^{-3}$ and the above calculations change by a negligible amount.
 
@@
 Inverse probability weights are computed based on the resulting dataset (termed IPW1).
+
+ In the second approach, we leverage hospitalization information.
+We first estimate the probability of fever given demographic and COVID-19 test status.
+When the COVID-19 test status is positive, the probability of fever is allowed to depend on whether the individual was hospitalized.
+Figure~\ref{fig:symptomlik1_model2} and~\ref{fig:symptomlik2_model2} in Section~\ref{app:in_add_details} in the supplementary materials presents the estimated likelihood of reporting a fever with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.
+We see that hospitalization significantly increases the likelihood of fever as expected.
+We use these models and the case-to-hospitalization rates by demographic strata to perform mean imputation of the number of positive and negative tests given demographic strata, fever status, and COVID-19 contact status.
+The use of hospitalization data to estimate the probability of fever may be subject to survivor bias as well as biases driven by differential care.
+Inverse probability weights are then computed based on the resulting dataset (termed IPW2).
 
@@
  \end{figure}
-
- In the second approach, we leverage hospitalization information.
-We first estimate the probability of fever given demographic and COVID-19 test status.
-When the COVID-19 test status is positive, the probability of fever is allowed to depend on whether the individual was hospitalized.
-Figure~\ref{fig:symptomlik1_model2} and~\ref{fig:symptomlik2_model2} in Section~\ref{app:in_add_details} in the supplementary materials presents the estimated likelihood of reporting a fever with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.
-We see that hospitalization significantly increases the likelihood of fever as expected.
-We use these models and the case-to-hospitalization rates by demographic strata to perform mean imputation of the number of positive and negative tests given demographic strata, fever status, and COVID-19 contact status.
-The use of hospitalization data to estimate the probability of fever may be subject to survivor bias as well as biases driven by differential care.
-AGain   Inverse probability weights are then computed based on the resulting dataset (termed IPW2).
 

### 2022-03-28 write-up/letters/response-ref-1-v1.tex
All plots finalized

@@
 
+	\textcolor{red}{
 	We thank the reviewer for catching this issue.
 We now clarify that the model proposed does not assume equal mixing.
-Rather due to limited data, we assume a these mixing probabilities are not time varying and that the  prior probability of mixing is chosen from a symmetric Dirichlet distribution (uniform) time-varying.
+Due to limited death data, we consider a global SEIR model and only generate strata-specific exposure counts~$E_{t,k}$ via a Multinomial distribution with a parameter~$\{ p_{t,k}  \} $ that has a symmetric Dirichlet prior.
+These strata-specific exposure counts are then linked to the observed strata-level death counts via the conditional Poisson model.
 We clarify this in Section 5.2 and comment on the code to clarify this more carefully in Appendix H.
-	\textcolor{red}{Need To Do.  Keep model just sensitivity on IFR and smooth over infectious.}
+	}
 	\vspace{5mm}
@@
 
-	\textcolor{red}{See our answer to Question 12 above.
-In Section 5.2, we now clarify how we calculate the infection-fatality rate from the compartments of the model.
+	\textcolor{red}{See our answer to Question 12a above.
+In Section 5.2, we now clarify hwo the the infection-fatality rate is used in our model.
 Similar to Johndrow (2021), the model only uses COVID-19 related deaths.
@@
 \textcolor{red}{See our answer to Question 17b of Reviewer \#1.
-We acknowledge that these IFR estimates are likely biased and time-varying due to introduction of vaccines in 2020.
-A sensitivity analysis focused on the issue of choosing an appropriate IFR and allowing for the IFR to be time-varying is now performed.}
-\textcolor{red}{To Do!}
+We acknowledge that these IFR estimates may be biased and/or time-varying.
+Note the study period ends January 2021 so the bias due to vaccines may be limited.
+We now include a sensitivity analysis focused on the issue of choosing an appropriate IFR.}
+% \textcolor{red}{To Do!}
 
@@
 \vspace{5mm}
-\textcolor{red}{See our answer to Question 17b of Reviewer \#1 and Question 7 above. A sensitivity analysis focused on the issue of choosing an appropriate IFR is now performed.}
-\textcolor{red}{To Do!}
+\textcolor{red}{See our answer to Question 17b of Reviewer \#1 and Question 7 above.
+A sensitivity analysis focused on the issue of choosing an appropriate IFR is now performed.
+We also note that the active infection estimate being very different for the model-based method in the prior manuscript was a result of how we decided to define the active infection rate using the SEIR model.
+With our more precise definition (see answer to Question XX), we now have estimates that are more in line with the IPW estimates.
+So while there is likely bias, this bias is likely less pronounced than one woul dhave thought using the calculations from our prior submission.}
+% \textcolor{red}{To Do!}
 \vspace{5mm}

### 2022-03-28 write-up/covid-supplement-writeup.tex
All plots finalized

@@
 Following~\cite{Song2020}, we employ Runge-Kutta (RK4) approximations for discretization.
+\textcolor{red}{Sub strata via kernel weighting due to limited data.}
 
@@
  \centering
- \includegraphics[width=.6\linewidth]{../figs/undercounting.png}
+ \includegraphics[width=.6\linewidth]{../figs/cumulative_undercounting.png}
+ \caption{Cumulative undercount based on SEIR model}
+ \label{fig:undercounting}
+\end{figure}
+
+Figure~\ref{fig:tv_air_sens} below is a sensitivity analysis of the active infection rate under SEIR model with 10\% increase and decrease in the average IFR.
+We see that the impact on the doubly robust estimates is minimal.
+
+\begin{figure}[!th]
+ \centering
+ \includegraphics[width=.6\linewidth]{../figs/tv_air_sensitivity.png}
  \caption{Undercount based on SEIR model}
- \label{fig:undercounting}
+ \label{fig:tv_air_sens}
 \end{figure}

### 2022-03-28 write-up/covid-writeup.tex
All plots finalized

@@
 \def\code#1{\texttt{#1}}
+\def\new{\text{new}}
 
@@
 Here, we consider covariate information that takes the form of a stratification variable with $K$ strata.
-Let $I_{k,t}$ to denote the number of new infections on day $t$ in the $k$th strata.
-The joint distribution of strata-specific new infection totals follows a multinomial distribution with
- $$
- ( I_{1,t}, \ldots, I_{K,t} ) \sim \text{Multinomial} \left( I_t, (p_{1,t}, \ldots, p_{K, t}) \right).
+Based on these population-level dynamics, we can compute the number of \emph{new exposures} at time~$t$, i.e., $E^\new_{t} := - N \cdot (e_{t+1} - e_{t} + s_{t+1} - s_t)$ to denote the number of new exposures on day $t$ in the $k$th strata.
+The joint distribution of strata-specific new exposures~$ \{ E^{\new}_{t,k} \}_{k=1}^K$ follows a multinomial distribution with
+ $$
+ ( E^\new_{1,t}, \ldots, E^\new_{K,t} ) \sim \text{Multinomial} \left( E^\new_t, (p_{1,t}, \ldots, p_{K, t}) \right).
  $$
@@
  $$
- D_{k,r} \mid p, \theta, \nu \sim \text{Poisson} \left( \sum_{t=1}^r \text{IFR}_{k} \cdot I_{k,t} \theta_{(r-t)} \right)
+ D_{k,r} \mid p, \theta, \nu \sim \text{Poisson} \left( \sum_{t=1}^r \text{IFR}_{k} \cdot E^\new_{k,t} \theta_{(r-t)} \right)
  $$
@@
  \begin{remark}[Sensitivity to IFR]
-
+To account for potential misspecification of the marginal IFR, we perform the same analysis with a 10\% increase and decrease in the marginal IFR under the log-linear model.
+Figure~\ref{fig:tv_air_sens} in the supplementary materials presents the model-based and doubly robust estimates in these scenarios.
 \end{remark}
@@
   \centering
-  \includegraphics[width=.9\linewidth]{../figs/cases_ppc.png}
- \caption{Posterior on daily case counts.}
+  \includegraphics[width=.9\linewidth]{../figs/exposures_ppc.png}
+ \caption{Posterior on daily new exposures.}
  \label{fig:casesppc}
@@
 Figure~\ref{fig:deathsppc} shows the results which suggests chosen SEIR model fits the observed data well.
-Figure~\ref{fig:casesppc} presents the posterior distribution of total new cases.
-Figure~\ref{fig:undercounting} in the supplementary materials computes the weekly undercount factor of observed case counts.
+Figure~\ref{fig:casesppc} presents the posterior distribution of total new exposures.
+As the active infection rate is defined as all infected individuals who’s viral load has yet to fall below detectable levels by RT-PCR testing, parameters underlying the population-level dynamics are combines with model-estimates of the number of newly exposed individuals to estimate the number of individuals with an active infection on each day.
+To compare with case count data, Figure~\ref{fig:undercounting} in the supplementary materials computes the weekly cumulative undercount factor of observed case counts.
 Posterior distributions for the age-specific case counts are used as part of the doubly-robust estimator of the active infection rate.

### 2022-03-28 methods/9_dr_estimation.R

@@ -18,15 +18,15 @@ propensities = readRDS("../data/smoothedpropensities_alt_08262021.RDS")
 # air_flags = readRDS("../data/modelbased_flags_2022_24_03.RDS")
 
 ### MODEL BASED ESTIMATES: LOWER IFR
-aggregate_air = readRDS("../data/aggregate_air_lowerifr_2022_24_03.RDS")
-strata_air = readRDS("../data/modelbased_air_bystrata_lowerifr_2022_24_03.RDS")
-air_flags = readRDS("../data/modelbased_flags_lowerifr_2022_24_03.RDS")
+# aggregate_air = readRDS("../data/aggregate_air_lowerifr_2022_24_03.RDS")
+# strata_air = readRDS("../data/modelbased_air_bystrata_lowerifr_2022_24_03.RDS")
+# air_flags = readRDS("../data/modelbased_flags_lowerifr_2022_24_03.RDS")
 
 ### MODEL BASED ESTIMATES: HIGHER IFR
-# aggregate_air = readRDS("../data/aggregate_air_2022_24_03.RDS")
-# strata_air = readRDS("../data/modelbased_air_bystrata_2022_24_03.RDS")
-# air_flags = readRDS("../data/modelbased_flags_2022_24_03.RDS")
-
+aggregate_air = readRDS("../data/aggregate_air_upperifr_2022_24_03.RDS")
+strata_air = readRDS("../data/modelbased_air_bystrata_upperifr_2022_24_03.RDS")
+air_flags = readRDS("../data/modelbased_flags_upperifr_2022_24_03.RDS")
+ 
 
 ### FB Data
 fb_data = readRDS("../data/fb_weeklycomplete.RDS")
@@ -146,6 +146,6 @@ for(i in 1:length(weeks)) {
 }
 
 # saveRDS(results,"../data/drestimates_alt_2022_24_03.RDS")
-saveRDS(results,"../data/drestimates_alt_lowerifr_2022_24_03.RDS")
-# saveRDS(results,"../data/drestimates_alt_upperifr_2022_24_03.RDS")
+# saveRDS(results,"../data/drestimates_alt_lowerifr_2022_24_03.RDS")
+saveRDS(results,"../data/drestimates_alt_upperifr_2022_24_03.RDS")
 

### 2022-03-28 preprocessing/8_modelbased_preproc.R

@@ -55,8 +55,8 @@ df_coviddeath_agg <- aggregate(covid_deaths ~ date, data = df_coviddeath, FUN = 
 dates = ymd(df_coviddeath_agg$date)[-length(df_coviddeath_agg$date)]
 
 # fit_forcing = readRDS("../data/fit_forcing_byage_090121.RDS")
-fit_forcing = readRDS("../data/fit_forcing_byage_lowerifr_032522.RDS")
-# fit_forcing = readRDS("../data/fit_forcing_byage_upperifr_032522.RDS")
+# fit_forcing = readRDS("../data/fit_forcing_byage_lowerifr_032522.RDS")
+fit_forcing = readRDS("../data/fit_forcing_byage_upperifr_032522.RDS")
 ## USE a in model to see how long exposed/infected
 a = summary(fit_forcing, pars = "a", probs = c(0.025, 0.05, 0.1, 0.5, 0.9, 0.95, 0.975))$summary
 a_mean = a[1]
@@ -239,14 +239,14 @@ for(i in 1:nrow(allcombinations)) {
 }
 
 # saveRDS(air_results, "../data/modelbased_air_bystrata_2022_24_03.RDS")
-saveRDS(air_results, "../data/modelbased_air_bystrata_lowerifr_2022_24_03.RDS")
-# saveRDS(air_results, "../data/modelbased_air_bystrata_upperifr_2022_24_03.RDS")
+# saveRDS(air_results, "../data/modelbased_air_bystrata_lowerifr_2022_24_03.RDS")
+saveRDS(air_results, "../data/modelbased_air_bystrata_upperifr_2022_24_03.RDS")
 
 allcombinations$flagged = flagged_user_aggregate_air
 
 # saveRDS(allcombinations, "../data/modelbased_flags_2022_24_03.RDS")
-saveRDS(allcombinations, "../data/modelbased_flags_lowerifr_2022_24_03.RDS")
-# saveRDS(allcombinations, "../data/modelbased_flags_upperifr_2022_24_03.RDS")
+# saveRDS(allcombinations, "../data/modelbased_flags_lowerifr_2022_24_03.RDS")
+saveRDS(allcombinations, "../data/modelbased_flags_upperifr_2022_24_03.RDS")
 
 
 ### AGGREGATE 
[…]

### 2022-03-28 visualizations/model_based.R

@@ -84,7 +84,7 @@ agg_cases$X5. = aggregate(X5. ~ timepoints, data = summary_fit, FUN = sum)$X5.
 agg_cases$X50. = aggregate(X50. ~ timepoints, data = summary_fit, FUN = sum)$X50.
 agg_cases$X95. = aggregate(X95. ~ timepoints, data = summary_fit, FUN = sum)$X95.
 
-png(filename = "../figs/cases_ppc.png",
+png(filename = "../figs/exposures_ppc.png",
     width = 6.5, height = 4, units = "in", res = 1200, pointsize = 12)
 par(mar = c(2,2,1,1)+0.1)
 
@@ -99,7 +99,7 @@ ggplot(smr_pred, mapping = aes(x = date)) +
   geom_ribbon(aes(ymin = X5., ymax = X95.), fill = c_posterior, alpha=0.35) +
   #geom_ribbon(aes(ymin = X10., ymax = X90.), fill = c_light) +
   geom_line(mapping = aes(x = date, y = X50.), color = c_posterior) +
-  labs(x = "Day", y = "Cases") +
+  labs(x = "Day", y = "New Exposures") +
   scale_x_date(date_labels = "%b %Y")
 
 dev.off()
@@ -121,16 +121,15 @@ undercount = data.frame(week = weeks_for_undercount, year = years_for_undercount
                         undercountfactor.X95 = rep(0, number_of_dates))
 
 for(i in 1:nrow(undercount)) {
-  observed_count = sum(df_case_counts_total$covid_count[week(df_case_counts_total$date) == weeks_for_undercount[i] &
-                                                      year(df_case_counts_total$date) == years_for_undercount[i]
-                                                    ], na.rm = TRUE)
-  mean_latent_count = sum(smr_pred$mean[week(smr_pred$date) == weeks_for_undercount[i] &
-                                      year(smr_pred$date) == years_for_undercount[i]], na.rm = TRUE)
-  X5_latent_count = sum(smr_pred$X5.[week(smr_pred$date) == weeks_for_undercount[i] &
-                                          year(smr_pred$date) == years_for_undercount[i]], na.rm = TRUE)
-  X95_latent_count = sum(smr_pred$X95.[week(smr_pred$date) == weeks_for_undercount[i] &
-                                          year(smr_pred$date) == years_for_undercount[i]], na.rm = TRUE)
-  
[…]

### 2022-03-28 visualizations/tv_prevalence_sensitivity.R

@@ -7,7 +7,7 @@ library(tidybayes)
 library(gridExtra)
 library(rstan)
 
-my_palette <- brewer.pal(name="Greys",n=9)[seq(1,9,2)]
+my_palette <- brewer.pal(name="Greys",n=9)[seq(3,9,3)]
 prevalence_long = data.frame(date = rep(0,0), Method = rep(0,0), estimate = rep(0,0))
 ratio_long = data.frame(date = rep(0,0), Method = rep(0,0), estimate = rep(0,0))
 
@@ -45,39 +45,43 @@ prevalence_temp = data.frame(date = prevalence_temp$date, Method = rep("Doubly R
                              estimate = prevalence_temp[,5])
 prevalence_long = rbind(prevalence_long, prevalence_temp)
 
+## ADD IN MODEL-BASED: UPPER
+prevalence_temp = readRDS("../data/aggregate_air_upperifr_2022_24_03.RDS")
+prevalence_temp = data.frame(date = prevalence_temp$date, Method = rep("Model-based, High", nrow(prevalence_temp)), 
+                             estimate = prevalence_temp$estimate)
+date = prevalence_temp$date[2:length(prevalence_temp$estimate)]
+prevalence_long = rbind(prevalence_long, prevalence_temp)
+
+## ADD IN DOUBLY ROBUST: UPPER
+prevalence_temp = readRDS("../data/drestimates_alt_upperifr_2022_24_03.RDS")
+prevalence_temp = data.frame(prevalence_temp)
+prevalence_temp$date = MMWRweek::MMWRweek2Date(MMWRyear = prevalence_temp[,2],
+                                               MMWRweek = prevalence_temp[,1],
+                                               MMWRday = 1)
+prevalence_temp = data.frame(date = prevalence_temp$date, Method = rep("Doubly Robust, High", nrow(prevalence_temp)), 
+                             estimate = prevalence_temp[,5])
+prevalence_long = rbind(prevalence_long, prevalence_temp)
+
+
 ## FINAL FIGURES
-# png(filename = "../figs/tv_air_sensitivity.png",
-# width = 960, height = 480, units = "px", pointsize = 25)
+png(filename = "../figs/tv_air_sensitivity.png",
+width = 960, height = 480, units = "px", pointsize = 25)
 
 ggplot(data = prevalence_long, aes(x = date, y = estimate, col = Method)) +
   geom_line(size = 2) +
   labs(x = "Date",
[…]

### 2022-03-28 write-up/covid-supplement-writeup.tex
fixed new exp to new inf and now all good

@@
 Following~\cite{Song2020}, we employ Runge-Kutta (RK4) approximations for discretization.
-\textcolor{red}{Sub strata via kernel weighting due to limited data.}
-
-Figure~\ref{fig:undercounting} below is a plot of undercount based on the SEIR model.  We see the undercount goes away by January; however, this points to potential misspecification of the IFR.
+Due to the low number of deaths per strata, i.e., based on Race, Age, Sex and Ethnicity, we employ a kernel smoothing of the new infections per age strata into these sub-strata using relative number of observed deaths.
+This ensured a suitably complex model that could fit the data while generating strata-specific active infection rates that can be used in the doubly-robust estimation procedure.
 
@@
 
-Figure~\ref{fig:tv_air_sens} below is a sensitivity analysis of the active infection rate under SEIR model with 10\% increase and decrease in the average IFR.
+Figure~\ref{fig:undercounting} is a plot of undercount based on the SEIR model.
+We plot relative undercound on a cumulative basis since January due to the case counts reflecting active infections.
+Figure~\ref{fig:tv_air_sens} is a sensitivity analysis of the active infection rate under SEIR model with 10\% increase and decrease in the average IFR.
 We see that the impact on the doubly robust estimates is minimal.

### 2022-03-28 write-up/covid-writeup.tex
fixed new exp to new inf and now all good

--- 2022-03-24 write-up/covid-writeup.tex
@@
 Here, we consider compartmental model approaches from infectious disease epidemiology.
-Our primary objective is a model-based forecast of strata-level prevalence $\bar y_k$.
+Our primary objective is a model-based forecast of strata-level active infection rate~$\bar y_k$.
 To do this, a probabilistic extension of a standard epidemiological state-space model -- the susceptible, exposed, infected, and removed (recovered and death) model, or SEIR model -- is presented.
@@
 Here, we consider covariate information that takes the form of a stratification variable with $K$ strata.
-Let $I_{k,t}$ to denote the number of new infections on day $t$ in the $k$th strata.
-The joint distribution of strata-specific new infection totals follows a multinomial distribution with
- $$
- ( I_{1,t}, \ldots, I_{K,t} ) \sim \text{Multinomial} \left( I_t, (p_{1,t}, \ldots, p_{K, t}) \right).
+Based on these population-level dynamics, we can compute the number of \emph{new infections} at time~$t$, i.e., $I^\new_{t} := - N \cdot (e_{t+1} - e_{t} + s_{t+1} - s_t)$ to denote the number of new exposures on day $t$ in the $k$th strata.
+The joint distribution of strata-specific new exposures~$ \{ I^{\new}_{t,k} \}_{k=1}^K$ follows a multinomial distribution with
+ $$
+ ( I^\new_{1,t}, \ldots, I^\new_{K,t} ) \sim \text{Multinomial} \left( I^\new_t, (p_{1,t}, \ldots, p_{K, t}) \right).
  $$
@@
  $$
- D_{k,r} \mid p, \theta, \nu \sim \text{Poisson} \left( \sum_{t=1}^r \text{IFR}_{k} \cdot I_{k,t} \theta_{(r-t)} \right)
+ D_{k,r} \mid p, \theta, \nu \sim \text{Poisson} \left( \sum_{t=1}^r \text{IFR}_{k} \cdot I^\new_{k,t} \theta_{(r-t)} \right)
  $$
@@
   \centering
-  \includegraphics[width=.9\linewidth]{../figs/cases_ppc.png}
- \caption{Posterior on daily case counts.}
+  \includegraphics[width=.9\linewidth]{../figs/infections_ppc.png}
+ \caption{Posterior on daily new infections.}
  \label{fig:casesppc}
  \end{subfigure}
- \caption{Posterior predictive check (PPC) on death counts and posterior distribution of case counts.}
+ \caption{Posterior predictive check (PPC) on death counts and posterior distribution of infections.}
  \label{fig:ppcs}
@@
 Figure~\ref{fig:deathsppc} shows the results which suggests chosen SEIR model fits the observed data well.
-Figure~\ref{fig:casesppc} presents the posterior distribution of total new cases.
-Figure~\ref{fig:undercounting} in the supplementary materials computes the weekly undercount factor of observed case counts.
+Figure~\ref{fig:casesppc} presents the posterior distribution of total new infections.
+As the active infection rate is defined as all infected individuals who’s viral load has yet to fall below detectable levels by RT-PCR testing, to construct the active infection estimates we exponentially discount the number of newly infected individuals to estimate the number of individuals with an active infection on each day.
+In our analysis, the exponential discounting is set to $20$ days to match with prior evidence that ``30\% to 40\% of people will still test positive at three weeks''~\citep{wlocneg}.
+To compare with case count data, Figure~\ref{fig:undercounting} in the supplementary materials computes the cumulative undercount factor of observed case counts as compared to new infections.
 Posterior distributions for the age-specific case counts are used as part of the doubly-robust estimator of the active infection rate.

### 2022-03-28 methods/9_dr_estimation.R

@@ -13,19 +13,19 @@ indiana_data$startdate = mdy(indiana_data$startdate)
 propensities = readRDS("../data/smoothedpropensities_alt_08262021.RDS")
 
 ### MODEL BASED ESTIMATES
-# aggregate_air = readRDS("../data/aggregate_air_2022_24_03.RDS")
-# strata_air = readRDS("../data/modelbased_air_bystrata_2022_24_03.RDS")
-# air_flags = readRDS("../data/modelbased_flags_2022_24_03.RDS")
+# aggregate_air = readRDS("../data/aggregate_air_2022_28_03.RDS")
+# strata_air = readRDS("../data/modelbased_air_bystrata_2022_28_03.RDS")
+# air_flags = readRDS("../data/modelbased_flags_2022_28_03.RDS")
 
 ### MODEL BASED ESTIMATES: LOWER IFR
-# aggregate_air = readRDS("../data/aggregate_air_lowerifr_2022_24_03.RDS")
-# strata_air = readRDS("../data/modelbased_air_bystrata_lowerifr_2022_24_03.RDS")
-# air_flags = readRDS("../data/modelbased_flags_lowerifr_2022_24_03.RDS")
+aggregate_air = readRDS("../data/aggregate_air_lowerifr_2022_28_03.RDS")
+strata_air = readRDS("../data/modelbased_air_bystrata_lowerifr_2022_28_03.RDS")
+air_flags = readRDS("../data/modelbased_flags_lowerifr_2022_28_03.RDS")
 
 ### MODEL BASED ESTIMATES: HIGHER IFR
-aggregate_air = readRDS("../data/aggregate_air_upperifr_2022_24_03.RDS")
-strata_air = readRDS("../data/modelbased_air_bystrata_upperifr_2022_24_03.RDS")
-air_flags = readRDS("../data/modelbased_flags_upperifr_2022_24_03.RDS")
+# aggregate_air = readRDS("../data/aggregate_air_upperifr_2022_28_03.RDS")
+# strata_air = readRDS("../data/modelbased_air_bystrata_upperifr_2022_28_03.RDS")
+# air_flags = readRDS("../data/modelbased_flags_upperifr_2022_28_03.RDS")
  
 
 ### FB Data
@@ -145,7 +145,7 @@ for(i in 1:length(weeks)) {
   results[i,5] = fb_air + (results[i,4] - invweighted_air)
 }
 
[…]

### 2022-03-28 preprocessing/8_modelbased_preproc.R

@@ -57,11 +57,14 @@ dates = ymd(df_coviddeath_agg$date)[-length(df_coviddeath_agg$date)]
 # fit_forcing = readRDS("../data/fit_forcing_byage_090121.RDS")
 # fit_forcing = readRDS("../data/fit_forcing_byage_lowerifr_032522.RDS")
 fit_forcing = readRDS("../data/fit_forcing_byage_upperifr_032522.RDS")
+
+# HOW LONG AS ACTIVE INFECTION
 ## USE a in model to see how long exposed/infected
-a = summary(fit_forcing, pars = "a", probs = c(0.025, 0.05, 0.1, 0.5, 0.9, 0.95, 0.975))$summary
-a_mean = a[1]
+# a = summary(fit_forcing, pars = "a", probs = c(0.025, 0.05, 0.1, 0.5, 0.9, 0.95, 0.975))$summary
+# a_mean = a[1]
 ## Gamma much larger so 
-gamma = summary(fit_forcing, pars = "gamma", probs = c(0.025, 0.05, 0.1, 0.5, 0.9, 0.95, 0.975))$summary
+# gamma = summary(fit_forcing,pars = "gamma", probs = c(0.025, 0.05, 0.1, 0.5, 0.9, 0.95, 0.975))$summary
+a_mean = -log(0.35)/21
 
 summary_fit = summary(fit_forcing, pars = "pred_cases_per_agegroup", probs = c(0.025, 0.05, 0.1, 0.5, 0.9, 0.95, 0.975))$summary
 n_days = nrow(summary_fit)/6
@@ -238,15 +241,15 @@ for(i in 1:nrow(allcombinations)) {
   }
 }
 
-# saveRDS(air_results, "../data/modelbased_air_bystrata_2022_24_03.RDS")
-# saveRDS(air_results, "../data/modelbased_air_bystrata_lowerifr_2022_24_03.RDS")
-saveRDS(air_results, "../data/modelbased_air_bystrata_upperifr_2022_24_03.RDS")
+# saveRDS(air_results, "../data/modelbased_air_bystrata_2022_28_03.RDS")
+# saveRDS(air_results, "../data/modelbased_air_bystrata_lowerifr_2022_28_03.RDS")
+saveRDS(air_results, "../data/modelbased_air_bystrata_upperifr_2022_28_03.RDS")
 
 allcombinations$flagged = flagged_user_aggregate_air
 
[…]

### 2022-03-28 visualizations/model_based.R

@@ -84,7 +84,7 @@ agg_cases$X5. = aggregate(X5. ~ timepoints, data = summary_fit, FUN = sum)$X5.
 agg_cases$X50. = aggregate(X50. ~ timepoints, data = summary_fit, FUN = sum)$X50.
 agg_cases$X95. = aggregate(X95. ~ timepoints, data = summary_fit, FUN = sum)$X95.
 
-png(filename = "../figs/exposures_ppc.png",
+png(filename = "../figs/infections_ppc.png",
     width = 6.5, height = 4, units = "in", res = 1200, pointsize = 12)
 par(mar = c(2,2,1,1)+0.1)
 
@@ -99,7 +99,7 @@ ggplot(smr_pred, mapping = aes(x = date)) +
   geom_ribbon(aes(ymin = X5., ymax = X95.), fill = c_posterior, alpha=0.35) +
   #geom_ribbon(aes(ymin = X10., ymax = X90.), fill = c_light) +
   geom_line(mapping = aes(x = date, y = X50.), color = c_posterior) +
-  labs(x = "Day", y = "New Exposures") +
+  labs(x = "Day", y = "New Infections") +
   scale_x_date(date_labels = "%b %Y")
 
 dev.off()

### 2022-03-28 visualizations/tv_prevalence.R

@@ -55,7 +55,7 @@ for(method in c("Unweighted", "IPW1", "IPW2")) {
 
 
 ## ADD IN MODEL-BASED
-prevalence_temp = readRDS("../data/aggregate_air_2022_24_03.RDS")
+prevalence_temp = readRDS("../data/aggregate_air_2022_28_03.RDS")
 prevalence_temp = data.frame(date = prevalence_temp$date, Method = rep("Model-based", nrow(prevalence_temp)), 
                              estimate = prevalence_temp$estimate)
 ratio = prevalence_temp$estimate[2:length(prevalence_temp$estimate)]/prevalence_temp$estimate[1:(length(prevalence_temp$estimate)-1)]
@@ -69,7 +69,7 @@ ratio_long = rbind(ratio_long, ratio_temp)
 prevalence_long = rbind(prevalence_long, prevalence_temp)
 
 ## ADD IN DOUBLY ROBUST
-prevalence_temp = readRDS("../data/drestimates_alt_2022_24_03.RDS")
+prevalence_temp = readRDS("../data/drestimates_alt_2022_28_03.RDS")
 prevalence_temp = data.frame(date = dates, Method = rep("Doubly Robust", nrow(prevalence_temp)), 
                              estimate = prevalence_temp[,5])
 ratio = prevalence_temp$estimate[2:length(prevalence_temp$estimate)]/prevalence_temp$estimate[1:(length(prevalence_temp$estimate)-1)]
@@ -108,8 +108,8 @@ prevalence_alt_cis$date = MMWRweek::MMWRweek2Date(MMWRyear = prevalence_alt_cis$
 
 pd <- position_dodge(0.1) # move them .05 to the left and right
 
-# png(filename = "../figs/tv_air_cis.png",
-    # width = 960, height = 480, units = "px", pointsize = 25)
+png(filename = "../figs/tv_air_cis.png",
+    width = 960, height = 480, units = "px", pointsize = 25)
 
 ggplot(data = prevalence_alt_cis, aes(x = date, y = estimate)) +
   geom_point(size = 2, color = my_palette) +
@@ -117,4 +117,4 @@ ggplot(data = prevalence_alt_cis, aes(x = date, y = estimate)) +
        y = "Active Infection Rate Estimate") + 
   theme(text = element_text(size=25))
 
-# dev.off()
+dev.off()

### 2022-03-28 visualizations/tv_prevalence_sensitivity.R

@@ -12,14 +12,14 @@ prevalence_long = data.frame(date = rep(0,0), Method = rep(0,0), estimate = rep(
 ratio_long = data.frame(date = rep(0,0), Method = rep(0,0), estimate = rep(0,0))
 
 ## ADD IN MODEL-BASED
-prevalence_temp = readRDS("../data/aggregate_air_2022_24_03.RDS")
+prevalence_temp = readRDS("../data/aggregate_air_2022_28_03.RDS")
 prevalence_temp = data.frame(date = prevalence_temp$date, Method = rep("Model-based", nrow(prevalence_temp)), 
                              estimate = prevalence_temp$estimate)
 date = prevalence_temp$date[2:length(prevalence_temp$estimate)]
 prevalence_long = rbind(prevalence_long, prevalence_temp)
 
 ## ADD IN DOUBLY ROBUST
-prevalence_temp = readRDS("../data/drestimates_alt_2022_24_03.RDS")
+prevalence_temp = readRDS("../data/drestimates_alt_2022_28_03.RDS")
 prevalence_temp = data.frame(prevalence_temp)
 prevalence_temp$date = MMWRweek::MMWRweek2Date(MMWRyear = prevalence_temp[,2],
                                           MMWRweek = prevalence_temp[,1],
@@ -29,14 +29,14 @@ prevalence_temp = data.frame(date = prevalence_temp$date, Method = rep("Doubly R
 prevalence_long = rbind(prevalence_long, prevalence_temp)
 
 ## ADD IN MODEL-BASED: LOWER
-prevalence_temp = readRDS("../data/aggregate_air_lowerifr_2022_24_03.RDS")
+prevalence_temp = readRDS("../data/aggregate_air_lowerifr_2022_28_03.RDS")
 prevalence_temp = data.frame(date = prevalence_temp$date, Method = rep("Model-based, Low", nrow(prevalence_temp)), 
                              estimate = prevalence_temp$estimate)
 date = prevalence_temp$date[2:length(prevalence_temp$estimate)]
 prevalence_long = rbind(prevalence_long, prevalence_temp)
 
 ## ADD IN DOUBLY ROBUST: LOWER
-prevalence_temp = readRDS("../data/drestimates_alt_lowerifr_2022_24_03.RDS")
+prevalence_temp = readRDS("../data/drestimates_alt_lowerifr_2022_28_03.RDS")
 prevalence_temp = data.frame(prevalence_temp)
[…]

### 2022-03-30 write-up/covid-supplement-writeup.tex
paper is ready; tg

@@
 \item[$W_{j,t}^{R}$] \quad Weight of individual $j$ in the population in the probability sample (i.e., inverse-probability of selection)
-\item[$\pi (X_{j,t}; \theta)$] \quad Self-selection propensity of individual~$j$ into the non-probability sample given feature vector~$X_{j,t}$ and Probability of selection model evaluated indicator of selection of the
-\item[$w (X_{j,t})$] \quad Weight of indivdual~$j$ in the population in the non-probability sample, i.e., $w(x) = 1/\pi(x; \theta)$.
-\item[$\mu(X_{j,t})$] \quad The indicator of selection of the
+\item[$\pi (X_{j,t}; \theta)$] \quad Self-selection propensity of individual~$j$ into the non-probability sample given feature vector~$X_{j,t}$ and parameter~$\theta$.
+\item[$w (X_{j,t})$] \quad Weight of individual~$j$ in the population in the non-probability sample, i.e., $w(x) = 1/\pi(x; \theta)$.
+\item[$\hat \mu(X_{j,t})$] \quad Model-based estimate of the active infection rate at time~$t$ for individuals with feature vector~$X_{j,t}$
 \end{itemize}
@@
 
-Here we present the testing propensity for four strata: (1) non-Hispanic, white female, (2) non-Hispanic, African American male, (3) Hispanic, White male.
+Here we present the testing propensity for four strata: (1) non-Hispanic, white female, (2) non-Hispanic, African American male, (3) Hispanic, White male, and (4) Hispanic male who selects `Some other Race'.
 Strata (1) presented in Figure~\ref{fig:nonh-white-female} demonstrates a small gender difference in testing propensities.
@@
 Figure~\ref{fig:undercounting} is a plot of undercount based on the SEIR model.
-We plot relative undercound on a cumulative basis since January due to the case counts reflecting active infections.
+We plot relative undercount on a cumulative basis since January due to the case counts reflecting active infections.
 Figure~\ref{fig:tv_air_sens} is a sensitivity analysis of the active infection rate under SEIR model with 10\% increase and decrease in the average IFR.

### 2022-03-30 write-up/covid-writeup.tex
paper is ready; tg

@@
 We then combine these selection propensities with epidemiological forecast models to construct a \emph{doubly robust} estimation method that accounts for both measurement-error and selection bias.
-This method is then applied to estimate Indiana's prevalence using case-count, hospitalization, and death data with demographic information, a statewide random molecular sample collected from April 25--29th, and Delphi's COVID-19 Trends and Impact Survey.
+This method is then applied to estimate Indiana's active infection prevalence using case-count, hospitalization, and death data with demographic information, a statewide random molecular sample collected from April 25--29th, and Delphi's COVID-19 Trends and Impact Survey.
 We end with a series of recommendations based on the proposed methodology.
@@
 \cite{Valliant2011} consider a weighted logistic regression procedure using the pooled probability and nonprobability samples.
-\cite{Chen2019} consider a pseudo-log-likelihood approach that uses the random samples as a proxy for a term in the log-likelihood.
+\cite{Chen2019} consider a pseudo-likelihood approach that uses the random samples as a proxy for a term in the log-likelihood.
 Here, we extend this approach to account for measurement-error as well as observing random samples at multiple times.
 We then provide an extension of the statistical error decomposition and discuss the trade-offs inherent in such a weighting approach.
-The proposed approach is distinct from the use of validation studies~\citep{Fox2020} -- a traditional epidemiologic method in which investigators compare the accuracy with gold standard measure to mitigate bias.
+The proposed approach is distinct from validation studies~\citep{Fox2020} -- a traditional epidemiological method in which investigators compare measurement accuracy with a gold standard measure to mitigate bias.
 Here, the \emph{gold standard} of a different test is replaced by a sample with a less biased selection mechanism.
@@
 Here, we combine the compartmental model of \cite{Song2020} but instead, as in \cite{Johndrow2020}, focus on COVID-19 confirmed death count data.
-We generate epidemiological forecasts for point prevalence of each population strata.
-We then demonstrate how to combine these forecasts with the IPW approach to construct doubly-robust estimates of point prevalence.
+We generate epidemiological forecasts for active infection rates within each population strata.
+We then demonstrate how to combine these forecasts with the IPW approach to construct doubly-robust estimates of active infection rates.
 A derived statistical error decomposition guides this discussion.
@@
 Recent work by \cite{Zhao2021} pointed out that estimation of key epidemiological parameters such as the incubation time using standard epidemiological models can suffer from severe bias due to issues beyond selection bias and measurement error.
-Namely, right truncation and epidemic growth leading to patients ``being more likely to be infected towards the end of their exposure period''~\citep[pp.
+Right truncation and epidemic growth lead to patients ``being more likely to be infected towards the end of their exposure period''~\citep[pp.
 3]{Zhao2021}.
-Their approach carefully constructs the study sample and statistical model to account for these issues.
-In this paper, rather than focusing on construction of a study sample, we ask whether one can collect auxiliary information to address selection bias in the observed case count data directly.
-This question is related to recent work on \emph{target validity}~\citep{Westreich2018}, in which the issues of internal and external validity are jointly addressed with respect to a specific population of interest.
-This article is a concrete attempt to address both types of validity and extends the conversation on target validity within the context of analysis of observational COVID-19 studies.
+Their approach constructs a study sample and statistical model to account for these issues.
+In this paper, rather than focusing on sample construction, we ask whether one can collect auxiliary information to address selection bias in the observed case count data directly.
+Our approach is related to the concept of \emph{target validity}~\citep{Westreich2018}, in which the issues of internal and external validity are jointly addressed with respect to a specific population of interest.
+This article is a concrete attempt to address both types of validity and extend the conversation on target validity within the context of analysis of observational COVID-19 studies.
 
@@
 A \emph{molecular test} refers to diagnostic tests that aim to detect viral load above a certain threshold (e.g., RT-PCR or antigen tests); see~\cite{Mina2020} for a detailed discussion of cycle thresholds.
-If someone has an active infection -- \textcolor{red}{here defined as individuals infected with SARS-CoV-2 who's viral load has yet to fall below detectable levels by RT-PCR testing} -- then after the incubation period, a molecular test with perfect sensitivity will yield a positive result while the patient has a viral load above the threshold of detection.
+If someone has an active infection -- here defined as being infected with SARS-CoV-2 and having a viral load that has yet to fall below detectable levels by RT-PCR testing -- then after the incubation period, a molecular test with perfect sensitivity will yield a positive result while the patient has a viral load above the threshold of detection.
 After that, the viral load will decrease below that threshold and a molecular test with perfect specificity will come back negative.
@@
 Systematic reviews suggest that 87\% sensitivity and 97.6\% specificity are reasonable estimates for current RT-PCR tests~\citep{Arevalo2020, Woloshin2020,Cohen2020}.
-To the best of our knowledge, both Indiana Department of Health's molecular testing and the random state-wide sample relied primarily on nasopharyngeal swab using the same set of RT-PCR tests.
+To the best of our knowledge, both Indiana Department of Health's molecular testing and the random state-wide RT-PCR tests were primarily collected via nasopharyngeal swab.
 
@@
 \begin{remark}[Public versus government datasets]
-In September 2020, the author submitted a data request to the Indiana Data Hub (IDH).
+% In September 2020, the author submitted a data request to the Indiana Data Hub (IDH).
 Over the next three months, we worked with Indiana's COVID-19 response team to refine the request.
 In December 2020, the request was officially approved and data streams were made publicly available on the IDH.
-In this paper, we thus focus on this \emph{publicly-available data}, i.e., data  that anyone can download directly from the IDH.
+In this paper, we focus on \emph{publicly-available data}, i.e., data that anyone can download directly from official government data portals such as the Indiana Data Hub (IDH).
 While the scientific community has contributed through independent COVID-19 observational and clinical studies, a significant component of public policy guidance has relied on testing and case count data, e.g., CDC and state guidelines based on test positivity rates and relative changes in the case counts over time.
-Addressing selection bias and measurement-error in these public datasets is imperative in better informing public policy debate.
-
-Note that the COVID-19 response team (and health departments in general) have access to \emph{government data} which may not be publicly-available and is likely stored at the individual-level.
-Due to data privacy and legal concerns, some collected covariates may not be reported publicly and others are aggregated and not reported jointly.
-While data analytic decisions made in Section~\ref{section:applications} were made due to access of publicly-available data, the overall data analysis framework is designed to be a tool for health departments to assess the pandemic and guide responses, e.g.,  account for individual-level covariate information, rather than publicly-reported strata-level information, that may include important covariates such as symptom status and recent COVID-19 contact.
+Addressing selection bias and measurement-error in these public datasets is imperative for better informed public policy debates.
+
+Note that Indiana's COVID-19 response team has access to official \emph{government data} which may not be publicly-available and is likely stored at the individual-level.
+Due to data privacy and legal concerns, some collected covariates may not be reported publicly and others are aggregated and reported marginally.
+While data analytic decisions discussed in Section~\ref{section:applications} were made due to access of publicly-available data, the overall data analysis framework is designed to be a tool for health departments to assess the pandemic and guide responses, e.g.,  account for individual-level covariate information, rather than publicly-reported strata-level information, that may include important covariates such as symptom status and recent COVID-19 contact.
  \end{remark}
@@
  On March 6th, Indiana State Health Department of Health confirmed the first case of COVID-19 in Indiana.
-From early March 2020 until April 28th, 2020, only symptomatic essential workers and their households, symptomatic high risk individuals were eligible, and individuals who have returned recently from overseas travel were eligible.
+From early March 2020 until April 28th, 2020, only symptomatic essential workers and their households, symptomatic high risk individuals, and individuals who had returned recently from overseas travel were eligible.
 An individual was considered high risk if they were over the age of 65, diabetic, obese, pregnant, a member of a minority population at greater risk of severe illness, or had high blood pressure.
- On April 28th, 2020, the criteria expanded to include any symptomatic Indiana resident, people in close contact with those who have tested positive, and residents of congregant communities~\citep{wishtv2020}.
+ On April 28th, 2020, the criteria expanded to include any symptomatic Indiana resident, people in close contact with those who had tested positive, and residents of congregant communities~\citep{wishtv2020}.
  % https://www.wishtv.com/news/medical/indiana-opens-up-covid-19-testing-to-more-hoosiers/
@@
 The order was extended until April 30th.
-On May 1st, 2020, a five-stage plan for gradual reopening was announced by Governon Holcomb~\citep{fivestageplan}.
+On May 1st, 2020, a five-stage plan for gradual reopening was announced by Governor Holcomb~\citep{fivestageplan}.
 Such policies target reduction in active infection rates.
@@
 
- We start with some simple notation.
+ % We start with some simple notation.
  Let $N$ denote the population size.
@@
 The term $\rho_{I,Y}$ is the empirical correlation between the population values~$\{ Y_j \}_{j=1}^N$ and the selection values $\{ I_j \}_{j=1}^N$.
-Under simple random sampling, $E_{\I} [ \rho_{I,Y} ] = 0$, where the expectation is with respect to the selection mechanism~$\I$, so there is no bias.
-The SRS variance formula above shows that $E_{\I} [ \rho_{I,Y}^2 ]  = 1/(N-1)$.
+Under simple random sampling, $E_{\I} [ \rho_{I,Y} ] = 0$,
+ %where the expectation is with respect to the selection mechanism~$\I$,
+ so there is no bias.
+ % The SRS variance formula above shows that $E_{\I} [ \rho_{I,Y}^2 ]  = 1/(N-1)$.
 
@@
 
- First we require some additional notation.
-Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the binary outcome and $0$ when we observe the true binary outcome.
+ % First we require some additional notation.
+ Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the binary outcome and $0$ otherwise.
 We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.
@@
  For ease of comprehension, a notation glossary is provided in Section~\ref{app:notation} in the supplementary materials.
-
+ \vspace{-0.25cm}
  \begin{remark}
@@
 We focus on the interplay of measurement error and selection bias in the context of binary outcomes due to the dichotomous nature of the COVID-19 testing data.
-Note that measurement error can lead to bias even in the absence of selection bias.
+Also note that measurement error can lead to bias even in the absence of selection bias.
  \end{remark}
-
+ \vspace{-0.25cm}
  Figure~\ref{fig:heatmap} shows that $D_M$, as a function of the relative frequency ($f_1/f_0$) and log odds ratio, can be both positive and negative as well as a range of magnitudes~\citep{Beesley2020,Beesley2019,Smeden2019}.
@@
  % % Thus, conclusions drawn from observed case-counts may not just be wrong, but very wrong.
- A key question is ``What is the (effective) sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' In Appendix~\ref{app:effss}, we show the effective sample size $n_{eff}$ can be bounded by
- $$
- \frac{f}{1-f} \times \frac{1}{E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right]}.
- $$
+ A key question is ``What is the (effective) sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' In Appendix~\ref{app:effss}, we show the effective sample size $n_{eff}$ can be bounded by $\frac{f}{1-f} \times \frac{1}{E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right]}$.
 Between April 25th to 29th 2020, Indiana performed $95,879$ tests.
@@
 See Section~\ref{section:effss} in the Supplementary Materials for additional effective sample size calculations.
-Note that while~\citep{Meng2018} argues the relative error increases as a function of population size, our calculations indicate this is not true when relative frequency~$f_1/f_0$ is held fixed.
+Note that while~\citep{Meng2018} argues the relative error increases as a function of population size, calculations in our setting indicate this is not true when relative frequency~$f_1/f_0$ is held fixed.
  % \begin{wrapfigure}{r}{0.5\textwidth}
@@
  The prior analysis demonstrates the potentially limited information regarding COVID-19 prevalence in observational case-count data.
-Scientists, however, may claim that daily observed case-counts simply undercount daily total cases by a constant multiple over time (i.e., undercounting).
+Analysts may claim that daily observed case-counts simply undercount daily total cases by a constant multiple over time (i.e., undercounting).
 If true then the ratio of case-counts at consecutive times may be a good estimate of the true change in prevalence, helping scientists understand the disease trajectory.
@@
 
-\cite{Heng2020} derive an approximate formula for the basic reproduction number under SEIR dynamics such that for $S_0 \approx 1$ and $I_0 \ll 1$, we have the approximate realtionship, which ``corresponds to [...] the solution for the SIR model assuming an infectious period of $1/\gamma + 1/\sigma$.
+\cite{Heng2020} derive an approximate formula for the basic reproduction number under SEIR dynamics, which corresponds to the solution for the SIR model assuming an infectious period of $1/\gamma + 1/\sigma$ for $s_0 \approx 1$ and $i_0 \ll 1$.
 Using this connection, a moment-based estimator is given by
@@
  \label{eq:auxinfoprob}
- \sum_{j=1}^N I_j^R \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N I_j^R W_j^R \log ( 1 - \pi (X_j; \theta)).
+ \sum_{j=1}^N I_j^{NR} \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N I_j^R W_j^R \log ( 1 - \pi (X_j; \theta)).
  \end{equation}
@@
 Solving~\eqref{eq:auxinfoprob} is done by iteratively re-weighted least squares (IRLS); see Section~\ref{app:irls} and Section~\ref{app:notation} in the supplementary materials for details and a notation glossary respectively.
-
+ \vspace{-0.5cm}
  \begin{remark}
  Weights built from selection propensities~$\{\pi(x_i; \theta)\}_{i=1}^n$ are common practice in the epidemiology literature.
-Survey sampling and transportability weights require knowledge of the selection mechanism~\citep{10.1093/aje/kwx164,ColeStuart2010}, which is not available in our setting.
-Instead, here we require auxiliary information in the form of a random sample to estimate the probability of selection.
+Survey sampling and transportability weights require knowledge of the selection mechanism~\citep{10.1093/aje/kwx164,ColeStuart2010}.
+Here auxiliary information is required to estimate the selection propensities.
  \end{remark}
-
+ \vspace{-0.5cm}
  % Under simple random sampling and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the $\hat \pi_k = (d_k / \tilde d_k) (n / N)$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.
@@
 
- As demonstrated below in Lemma~\ref{lemma:ipw}, if the propensity model is correctly specified then the $E_{\bf R} [ \rho_{I^{NR} (X), Y} ] = 0$ and therefore $\rho_{I^{NR} (X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then the data quality index is unlikely to inversely scale with population size.
+ As demonstrated below in Lemma~\ref{lemma:ipw}, if the propensity model is correctly specified then the $E [ \rho_{I^{NR} (X), Y} ] = 0$ and therefore $\rho_{I^{NR} (X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then the data quality index is unlikely to inversely scale with population size.
  % Indeed, Horvitz-Thompson estimators are sensitive to errors in the estimated weights.
@@
  \begin{array}{c}
- \frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t}{\pi_{j,t}} \\
+ \frac{1}{N} \sum_{j=1}^N I^{NR}_{j,t} \frac{Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t}{\pi_{j,t}} \\
  \frac{1}{N \times \sum_{l = 1}^L K_h \left(|t - t_l^\prime| \right)} \sum_{l = 1}^L K_h \left(|t - t_l^\prime| \right) \left[ \sum_{j=1}^N I^{NR}_{j,t_l^\prime} X_{j,t_l^\prime} - \sum_{j = 1}^N I^{R}_{j,t_l^\prime} W^{R}_{j,t_l^\prime}  \pi_{j,t_l^\prime} X_{j,t_l^\prime}  \right] \\
@@
 Here, we consider covariate information that takes the form of a stratification variable with $K$ strata.
-Based on these population-level dynamics, we can compute the number of \emph{new infections} at time~$t$, i.e., $I^\new_{t} := - N \cdot (e_{t+1} - e_{t} + s_{t+1} - s_t)$ to denote the number of new exposures on day $t$ in the $k$th strata.
-The joint distribution of strata-specific new exposures~$ \{ I^{\new}_{t,k} \}_{k=1}^K$ follows a multinomial distribution with
+Based on these population-level dynamics, we can compute the number of \emph{new infections} at time~$t$, i.e., $I^\new_{t} := - N \cdot (e_{t+1} - e_{t} + s_{t+1} - s_t)$ to denote the number of new infections on day $t$ in the $k$th strata.
+The joint distribution of strata-specific new infections~$ \{ I^{\new}_{t,k} \}_{k=1}^K$ follows a multinomial distribution with
  $$
@@
  Rather than relying solely on IPW or epidemiological forecasts, here we combine forecasting and inverse-probability weighting by extending recent work by~\cite{Chen2019} to account time-varying propensities and measurement-error.
- Let~$\mu(x)$ denote the posterior mean of the active infection rate individuals with covariate value~$x$ (i.e., age strata) based on the SEIR model described in Section~\ref{section:modelbased}.
+ Let~$\hat \mu(x)$ denote the posterior mean of the active infection rate individuals with covariate value~$x$ based on the SEIR model described in Section~\ref{section:modelbased}.
 In this context, the doubly-robust estimator is given by
@@
  $$
- where $\hat \mu(x_j)$ is not corrected for measurement-error as it estimates true case counts.
+ where $\hat \mu(x_j)$ is not corrected for measurement-error as it estimates true active infections.
 This estimator is called ``doubly-robust'' because it is consistent if either the model-based forecasts or the time-varying propensities are correctly specified.
@@
 First, the random sample had significant missingness ($3,658/15,495 \approx 24\%$ response rate).
-See~\cite{Yiannoutsos2021} for a discussion of how the scientific team handled missing data.
+See~\cite{Yiannoutsos2021} for how the scientific team handled missing data in prevalence estimation.
 Second, the hit rate on randomly selected Facebook users is likely low and Facebook's Indiana user population may differ from the Indiana population of interest.
@@
  We start by using Table~\ref{tab:indiana} to compute IPW weights for end of April, early May.
-The random sample is $n=3,658$ and IPW weights are computed per strata which depend on Gender, Age, Race, Fever, Positive Case in Household, and Prior Positive Test.
+The random sample is $n=3,658$.
+IPW weights are computed per strata using Indiana Census and random survey data, and are allowed to depend on Gender, Age, Race, Fever, Positive Case in Household, and Prior Positive Test.
 The weight for the strata defined as Male, 40--59, White, has a Fever, no household cases, and no prior tests, for example, is proportional to:
@@
  Using the constructed weights and the adjusted strata-level prevalence estimates, the IPW estimate under no measurement-error is $7.7\%$, a decrease of four percentage points from the observed prevalence of $11.7\%$.
-We also make use of Facebook's symptom survey over that window of time to construct a second IPW estimate conditional on demographics and fever but ignoring the household and prior testing information, which under no measurement-error is $6.5$\%.
+We also make use of CTIS sample over that window of time to construct a second IPW estimate conditional on demographics and fever but ignoring the household and prior testing information, which under no measurement-error is $6.5$\%.
 
  The Indiana study did not report sensitivity and specificity; therefore, we take the suggested measures from~\cite{Arevalo2020} which report 87\% sensitivity is a reasonable estimates and~\cite{Cohen2020} which report 97.6\% specificity.
-This corresponds to a false negative rate of $13$\% and false positive rate of $2.4\%$, resulting in estimates of $6.2\%$ and $4.9\%$ respectively.
+This corresponds to a false negative rate of $13$\% and false positive rate of $2.4\%$, resulting in estimates of $6.2\%$ and $4.9\%$ using the random and CTIS samples respectively.
 
@@
 In \cite{Meng2018}, estimation relied on observing the true outcome (i.e., election vote totals).
-Here, we use the random sample which estimates the true prevalence of active COVID-19 disease at 1.81\%~\cite{Yiannoutsos2021}.
+Here, we rely on~\cite{Yiannoutsos2021} who use the random sample to estimate the true prevalence of active COVID-19 disease at 1.81\% after accounting for non-response and measurement error.
 Our goal is to build sensitivity analyses to understand the amount of information in observational case-count data.
@@
  \end{equation*}
- using the Indiana non-random sample.  For the IPW estimate using Facebook data, $\tilde M = 3.3$. Therefore, bias is high using unweighted data and remains moderate using the weighted estimates.
+ using the Indiana non-random sample.  For the IPW estimate using CTIS data, $\tilde M = 3.3$. Therefore, bias is high using unweighted data and remains moderate using the weighted estimates.
 
@@
  The IPW analysis is next extended to the time-varying setting.
-To do so, we make use of the Facebook symptom survey, hospitalization, and testing data.
+To do so, we make use of the CTIS, hospitalization, and testing data.
 Recall the COVID-19 testing and positive case counts are known by age, gender, and racial demographics.
-Symptom status (e.g., fever, coughing, shortness of breath) and related covariate information (COVID-19 positive contact), however, are currently unavailable\footnote{Whether this is due to data privacy considerations or lack of data collection efforts is unclear.}.
-As COVID-19 symptom status is likely to alter the testing propensity and likelihood of active infection, for illustrative purposes Facebook symptom survey and hospitalization data are used in two ways to impute symptom status per strata.
+Symptom status (e.g., fever, coughing, shortness of breath) and related covariate information (COVID-19 positive contact), however, are currently unavailable, which is likely due to data privacy considerations.
+As COVID-19 symptom status is likely to alter the testing propensity and likelihood of active infection, for illustrative purposes CTIS and hospitalization data are used in two ways to impute symptom status per strata.
 We outline the imputation methods below; see Section~\ref{section:discussion} for additional discussion.
 
- In the first approach, we estimate two logistic regressions using weighted pseudo-likelihoods with the Facebook survey symptom data.
+ In the first approach, we estimate two logistic regressions using weighted pseudo-likelihoods with the CTIS data.
 The first estimates the probability of contact with a COVID-19 positive individual given demographic and COVID-19 test status (positive or negative COVID-19 test).
@@
 Strata for ages less than $40$ are collapsed due to the limited number of COVID-19 related deaths in this age range.
-For similar reasons, the multinomial parameter~${\bf p}_t$ is assumed to be constant in time with a Dirichlet prior.
-
+For similar reasons, the multinomial parameter~${\bf p}_t$ is assumed to be constant in time with a symmetric Dirichlet prior.
+ \vspace{-0.3cm}
  \begin{remark}[Sensitivity to IFR]
@@
  \end{remark}
-
-  \begin{figure}
+\vspace{-0.3cm}
+
+ The SEIR model in equation~\eqref{eq:seir} is extended to allow for time-varying transmission rates~$\beta_t$ to account for public policy changes discussed in Section~\ref{section:publicpolicyindiana} over the time period considered.
+Here, we insert three change points $(t_1,t_2,t_3)$ on March 23rd, June 15th, and October 1st of 2020 respectively.
+Given sensitivity to these choices, instead of considering a simple multiplicative structure, we allow for delayed implementation and slow change in the transmission rate by using weights:
+ $$
+ \beta_t = \sum_{k=1}^{3} \left(\beta_{k-1} (1-w^{(k)}_t ) + \beta_{k} w_t\right),\quad w^{(k)}_t = (1 + \exp \left( -\xi (t - t_k - \nu) \right))^{-1} {\bf 1}[t \geq t_k]
+ $$
+ where~$(\beta_0,\beta_1, \beta_2, \beta_3)$ is the vector of transmission rate parameters, $\nu \geq 0$ is the potential delay, $\xi$ is the rate of change, and ${\bf 1}[\cdot]$ is an indicator function.
+Each weight~$w_t^{(k)} = 0$ for $t < t_k$ and $w_t^{(k)} \to 1$ as $t \to \infty$.
+See Section~\ref{app:prior_modelbased} in the supplementary materials for details on prior specification.
+
+ \begin{figure}[!th]
  \centering
@@
 
- The SEIR model in equation~\eqref{eq:seir} is extended to allow for time-varying transmission rates~$\beta_t$ to account for public policy changes discussed in Section~\ref{section:publicpolicyindiana} over the time period considered.
-Here, we insert three change points $(t_1,t_2,t_3)$ on March 23rd, June 15th, and October 1st of 2020 respectively.
-Given sensitivity to these choices, instead of considering a simple multiplicative structure, we allow for delayed implementation and slow change in the transmission rate by using weights:
- $$
- \beta_t = \sum_{k=1}^{3} \left(\beta_{k-1} (1-w^{(k)}_t ) + \beta_{k} w_t\right),\quad w^{(k)}_t = (1 + \exp \left( -\xi (t - t_k - \nu) \right))^{-1} {\bf 1}[t \geq t_k]
- $$
- where~$(\beta_0,\beta_1, \beta_2, \beta_3)$ is the vector of transmission rate parameters, $\nu \geq 0$ is the potential delay, $\xi$ is the rate of change, and ${\bf 1}[\cdot]$ is an indicator function.
-Each weight~$w_t^{(k)} = 0$ for $t < t_k$ and $w_t^{(k)} \to 1$ as $t \to \infty$.
-See Section~\ref{app:prior_modelbased} in the supplementary materials for details on prior specification.
-
  To check model fit, a posterior predictive check on aggregate death counts was performed.
 Figure~\ref{fig:deathsppc} shows the results which suggests chosen SEIR model fits the observed data well.
-Figure~\ref{fig:casesppc} presents the posterior distribution of total new infections.
-As the active infection rate is defined as all infected individuals who’s viral load has yet to fall below detectable levels by RT-PCR testing, to construct the active infection estimates we exponentially discount the number of newly infected individuals to estimate the number of individuals with an active infection on each day.
-In our analysis, the exponential discounting is set to $20$ days to match with prior evidence that ``30\% to 40\% of people will still test positive at three weeks''~\citep{wlocneg}.
+Figure~\ref{fig:casesppc} presents the posterior distribution of daily new infections.
+As the active infection rate is defined as all infected individuals who’s viral load has yet to fall below detectable levels by RT-PCR testing, to construct the active infection estimates we exponentially discount the number of newly infected individuals to estimate the number of these newly infected individuals who still have an active infection on future days.
+In our analysis, the exponential discounting parameter is set to $20$ days to match with prior evidence that ``30\% to 40\% of people will still test positive at three weeks''~\citep{wlocneg}.
 To compare with case count data, Figure~\ref{fig:undercounting} in the supplementary materials computes the cumulative undercount factor of observed case counts as compared to new infections.
-Posterior distributions for the age-specific case counts are used as part of the doubly-robust estimator of the active infection rate.
+Posterior distributions for the age-specific new infections are used as part of the doubly-robust estimator of the active infection rate.
 
@@
 The inverse-probability weighting methods does reduce bias compared to the unweighted estimates.
-Bias, however, does remain which we conjecture is due to the limited availability of symptom and other important covariate information to estimate strata-specific infection rates.
+Bias likely remains which we conjecture is due to the limited availability of symptom and other important covariate information to estimate strata-specific infection rates.
 This suggests better data collection in the future may strongly improve performance.
 
- While the model-based estimates appear more reasonable outside the summer months, note that the estimate on the last week of April is $0.30\%$ which is a severe underestimate compared to the estimated prevalence using the stratified random sample of $1.81\%$.
-Caution is therefore warranted when interpreting these estimates.
-Infection fatality rates are likely time-varying and may vary by other factors such as quality of healthcare.
+ Model based estimates appear more reasonable.
+The estimate on the last week of April is $1.17\%$ which is an underestimate when compared to the estimated prevalence using the stratified random sample of $1.81\%$.
+Caution is warranted when interpreting these estimates.
+Infection fatality rates are likely time-varying and may vary by other factors such as quality of healthcare and access to vaccines which were made available starting November 2020.
 Therefore, improvements could be made by using more accurate IFR estimates, but there is currently minimal publicly available data to do so.
 Moreover, the (thankfully) low number of deaths per strata make uncertainty in these estimates quite high.
-Finally, the doubly robust estimate appears similar to the IPW estimate.
+Finally, the doubly robust estimate appears similar to the IPW estimate, with the largest differences occurring in November and December 2020.
 
@@
 
-\textcolor{red}{
- While this paper emphasizes the importance of random sampling, we do not view its use as a panacea.
-With access to \emph{only} a non-probabilistic sample, we have no means to reduce self-selection bias.
-Adding a probability sample gives us the necessary auxiliary information to do so.
-This leaves the data analyst more time to focus on statistical issues related to the random sample such as how to handle non-response bias.
-Covariates that the scientific team thinks are correlated with non-response (e.g., political opinion) should be collected as part of the random sampling protocol and corrected for in the data analysis.
-On the other hand, collecting covariate information that correlates with testing propensity (e.g., symptom status) in the non-probabilistic sample is insufficient to address selection bias.
-}
+% \textcolor{red}{
+ While we emphasize the importance of random sampling, we do not view it as a panacea.
+With access to \emph{only} a non-probabilistic sample, we cannot address self-selection bias.
+Probability samples provide necessary auxiliary information to do so.
+Data analysts can then focus on statistical issues related to the random samples such as how to handle non-response bias.
+Covariates that the scientific team think are correlated with non-response (e.g., political affiliation, rural/urban location) should be collected as part of the random sampling protocol and corrected for in the data analysis.
+On the other hand, collecting covariate information that correlates with testing propensity (e.g., symptom status) in the non-probabilistic sample is insufficient.
+% }
 
@@
 
- The proposed methodology relied on access to two critical datasets: (1) COVID-19 testing, case, hospitalization, and death data by demographic strata and (2) Facebook symptom survey.
+ The proposed methodology relied on access to two critical datasets: (1) COVID-19 testing, case, hospitalization, and death data by demographic strata and (2) CTIS data.
 Public access to (1) required submission of data requests to the state of Indiana.

### 2022-03-30 write-up/letters/cover letter-covid.tex
paper is ready; tg

--- 2021-05-14 write-up/cover letter-covid.tex
@@
 
-\documentclass[10pt]{letter} % Default font size of the document, change to 10pt to fit more text
+\documentclass[11pt]{letter} % Default font size of the document, change to 10pt to fit more text
 
@@
 
-Kindly find enclosed the Report titled ``Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic,'' by Walter Dempsey.
+Kindly find enclosed the revised version of the manuscript number AOAS2006-048 titled ``Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic'' now titled ``Addressing selection bias and measurement error in COVID-19 case count data using auxiliary information.''  We have revised the article to address the referees’ requests about exposition and other points of clarification.
+The reports called for a major revision to make the manuscript ``forward compatible'', to re-structure the manuscript to present figures and tables near their corresponding discussion, to discuss potential non-response bias in the motivating studies, to include a more careful discussion of COVID-19 testing and disease dynamics, to include a more precise definition of the active infection rate and how it impacts model-based estimation, and to more carefully tie the proposal to related parts of the epidemiological literature.
+The new version complies with these requests.
 
-The report contains novel empirical findings and an investigation of statistical paradoxes arising from the analysis of reported coronavirus case-counts that are being used by governments to guide policies and decision making.
-We have discovered an estimator and statistical decomposition of the associated error that shows the 21.5 million tests conducted in the US provides information on COVID-19 prevalence equivalent to a simple random sample of size 92.
-Analogous decompositions for the rate of change in prevalence and the effective reproduction number are used to demonstrate how the magnitude and sign of estimation bias varies over time, a bias which has strong policy implications.
-Overall, our findings demonstrate the statistical limitations of increased testing in the presence of selection bias and measurement error as well as how easy it is for researchers analysing reported coronavirus case-counts to not only fool themselves but be confident in their incorrect conclusions.
-We believe the work presents an important advance in the statistical understanding of the COVID-19 pandemic.
+We believe the work presents an important advance in how researchers should think about addressing time-varying selection bias and measurement error to better understanding the COVID-19 pandemic trajectory.
+Attached please find a detailed response to the referees outlining how the changes to the manuscript address the concerns raised in their reports.
 

### 2022-03-30 write-up/letters/response-ref-1-v1.tex
updated paper; ready

@@
 \vspace{5mm}
-We thank the editor-in-chief.  We have addressed the two reviewers specific comments.  Responses to their concerns are detailed below and have significantly improved the clarify of this manuscript.
+We thank the editor-in-chief.  We have addressed the two reviewers' specific comments.  Responses to their concerns are detailed below and have significantly improved the clarify of this manuscript.
 \vspace{5mm}
@@
 We acknowledge the potential for a difference in characteristics between intended and final sample.
+In Section 5.1, we clarify that Indiana Census and random survey data are combined when calculating the IPW weights.
     \vspace{5mm}
@@
 Instead, we see it as a means to remove one very important source of potential bias in the estimation of population quantities of interest.
-A key takeaway of our paper is that with access to \emph{only} a non-probabilistic sample, we know there is potential bias but have little hope of reducing said bias.
-With a probability sample, we have means to reduce said bias.
-Statisticians, epidemioligists, and government officials, and health policy experts can then focus on practical issues such as non-ignorable missingness.
+A key takeaway of our paper is that with access to \emph{only} a non-probabilistic sample, we know there is potential bias but have no way to reduce said bias.
+With a probability sample, we have a means to do so.
+Statisticians, epidemioligists, government officials, and health policy experts can then focus on practical issues such as addressing non-ignorable missingness.
 If we thought, for example, political opinions and rural status are important predictors of missingness, then these should be collected as part of the random sampling protocol.
@@
 
-	\textcolor{red}{We now clarify that we define active infection as individuals infected with SARS-CoV-2 who have yet to fall below a threshold for detectable viral load levels by RT-PCR testing.
-This includes the infectious period as well as a pre-infectious and post-infectious periods of time.}
+	We now clarify that we define active infection as individuals infected with SARS-CoV-2 who have yet to fall below a threshold for detectable viral load levels by RT-PCR testing.
+	% This includes the infectious period as well as a pre-infectious and post-infectious periods of time.
 
@@
 
-	Section 2.2.1 has a reconstructed timeline of restrictions, which was based on news reports of state-level testing restrictions.
+	Section 2.2.1 now has a complete reconstructed timeline of restrictions, which was based on news reports of state-level testing restrictions.
 	\vspace{5mm}
@@
 We clarify that the Census data is Indiana specific and is presented as a reference to demonstrate the potential difference between intended and final sample.
+In Section 5.1, we clarify that Indiana Census and random survey data are combined when calculating the IPW weights.
 
@@
 These weights are used in computing probabilities of selection as shown in Equation (4.2) in Section 4.1.1.
-Estimates for 'Fever' and 'Household + Case' only use data collected between April 25th and 29th to match the timing of the random survey.
+Estimates for `Fever' and `Household + Case' only use data collected between April 25th and 29th to match the timing of the random survey.
 Note that individual-level racial demographics were not provided by CTIS due to data privacy concerns and there was no corresponding question on percent of tests from individuals with a prior positive test.
@@
 
-	\textcolor{red}{As stated in answer to Question 3a, we now clarify that we define active infection as individuals infected with SARS-CoV-2 who have yet to fall below a threshold for detectable viral load levels by RT-PCR testing.
-As stated here, this includes the infectious period as well as a pre-infectious and post-infectious periods of time.
-We discuss how this impacts the compartmental model and calculation of the AIR in answer to Questions 6a and 17a.}
+	As stated in answer to Question 3a, we now clarify that we define active infection as individuals infected with SARS-CoV-2 who have yet to fall below a threshold for detectable viral load levels by RT-PCR testing.
+We discuss how this impacts the compartmental model and calculation of the AIR in answer to Question 9 from Reviewer \#2.
+
 	\vspace{5mm}
@@
 
-	The adjustment is exactly related to the Bayes rule PPV/NPV relationship with FP, FN, and $\bar Y$.
-	\textcolor{red}{I'm still a bit confused by this point.  Will review further.}
+	As stated above, the approach benefits from the fact that FP and FN do not depend on $Y$ or $y$ and are only affected by test characteristics.
+The edited version of Section 3.1 tries to highlight this a bit more clearly.
+	% \textcolor{red}{I'm still a bit confused by this point.  Will review further.}
 	\vspace{5mm}
@@
 
-	Since weights are normalized, these are equivalent, i.e., $\sum w (x_i) FP / \sum w(x_i) = FP$.
+	Since weights are normalized, these are equivalent, i.e., $\sum w (x_i) FP / \sum w(x_i) = FP$.  We continue to write it this way to keep the formulas compact.
 	\vspace{5mm}
@@
 
-	\textcolor{red}{
 	We thank the reviewer for catching this issue.
 We now clarify that the model proposed does not assume equal mixing.
-Due to limited death data, we consider a global SEIR model and only generate strata-specific exposure counts~$E_{t,k}$ via a Multinomial distribution with a parameter~$\{ p_{t,k}  \} $ that has a symmetric Dirichlet prior.
-These strata-specific exposure counts are then linked to the observed strata-level death counts via the conditional Poisson model.
+Due to limited death data, we consider a population-level SEIR model and extract new infection counts at each time~$\{ I^{new}_t \}$.
+We then generate strata-specific new infection counts at each time~$\{ I^{new}_{t,k} \}$ via a Multinomial distribution given the new infection counts~$\{ I^{new}_t \}$ and parameters~$\{ p_{t,k} \} $ that has a symmetric Dirichlet prior.
+These strata-specific new infection counts are then linked to the observed strata-level death counts via the conditional Poisson model using the infection fatality rate.
 We clarify this in Section 5.2 and comment on the code to clarify this more carefully in Appendix H.
-	}
 	\vspace{5mm}
@@
 	We clarify this notation in Section 4.3.
-Specifically, the SEIR model is used to generate the number of new weekly cases which allows us to calculate the number of individuals with an active infection within each strata (i.e.~$I_{k,t}$ for the $k$th strata).
-We then can use this to calculate the posterior mean for the active infection rate per strata,~$\mu_k$.
+Specifically, the SEIR model is used to generate the number of new daily infections which are then used to  calculate the number of individuals with an active infection within each strata.
 	\vspace{5mm}
@@
 We continue to calculate the sampling fraction as number of tests divided by the total population minus prior tests.
-To address the reviewers concern, we note at the end of Section 5.1.1 that using the total population in the denominator yields a very similar sampling fraction and the calculations change by a negligible amount.
+To address the reviewer's concern, we note at the end of Section 5.1.1 that using the total population in the denominator yields a very similar sampling fraction and the calculations change by a negligible amount.
 	\vspace{5mm}
@@
 
-	\textcolor{red}{See our answer to Question 12a above.
-In Section 5.2, we now clarify hwo the the infection-fatality rate is used in our model.
+	See our answer to Question 12a above.
+In Section 5.2, we now clarify how the infection-fatality rate is used in our model.
 Similar to Johndrow (2021), the model only uses COVID-19 related deaths.
-Deaths to non-COVID causes are not directly modeled but would appear as part of the R component in the global SEIR model.}
+Deaths to non-COVID causes are not directly modeled but would appear as part of the R component in the population-level SEIR model.
+Moreover, since the active infection rate is defined as all infected individuals who’s viral load has yet to fall below detectable levels by RT-PCR testing, to construct the active infection estimates we exponentially discount the number of newly infected individuals to estimate the number of individuals with an active infection on each day per strata.
+In our analysis, the exponential discounting is set to 20 days to match with prior evidence that 30\% to 40\% of people will still test positive at three weeks.
+
 	\vspace{5mm}
@@
 
-    \textcolor{red}{We contextualize the discussion of IFR in Remark 7 in Section 5.2.
-We acknowledge that these IFR estimates are likely biased (or potentially accurate in very different circumstances).
-A sensitivity analysis focused on the issue of choosing an appropriate IFR given the potential biases in available IFR estimates is now performed.}
-    \textcolor{red}{To Do!}
+   	We contextualize the discussion of IFR in Remark 7 in Section 5.2.
+We acknowledge that these IFR estimates may be biased.
+A sensitivity analysis focused on the issue of choosing an appropriate IFR given the potential biases in available IFR estimates is now performed and presented in Appendix I in the supplementary materials.
+
 	\vspace{5mm}
@@
 \vspace{5mm}
-\textcolor{red}{We apologize for the confusion.
-See answer to Question 3b where we clarify the definition of active infection.
+We apologize for the confusion.
+See our answer to Question 3b from Reviewer \#1 where we clarify the definition of active infection.
 An active infection was meant to refer to individuals who were infected with SARS-CoV-2 who have yet to fall below a threshold for detectable viral load levels by RT-PCR testing.
-This includes the infectious period as well as a pre-infectious and post-infectious periods of time.}
+% This includes the infectious period as well as a pre-infectious and post-infectious periods of time.}
 
@@
 \vspace{5mm}
-\textcolor{red}{See our answer to Question 17b of Reviewer \#1.
+See our answer to Question 17b of Reviewer \#1.
 We acknowledge that these IFR estimates may be biased and/or time-varying.
 Note the study period ends January 2021 so the bias due to vaccines may be limited.
-We now include a sensitivity analysis focused on the issue of choosing an appropriate IFR.}
+We now include a sensitivity analysis focused on the issue of choosing an appropriate IFR in Appendix I in the supplementary materials.
 % \textcolor{red}{To Do!}
@@
 \vspace{5mm}
-\textcolor{red}{See our answer to Question 17b of Reviewer \#1 and Question 7 above.
-A sensitivity analysis focused on the issue of choosing an appropriate IFR is now performed.
+See our answer to Question 17b of Reviewer \#1 and Question 7 above.
+A sensitivity analysis focused on the issue of choosing an appropriate IFR is now performed in Appendix I in the supplementary materials.
 We also note that the active infection estimate being very different for the model-based method in the prior manuscript was a result of how we decided to define the active infection rate using the SEIR model.
-With our more precise definition (see answer to Question XX), we now have estimates that are more in line with the IPW estimates.
-So while there is likely bias, this bias is likely less pronounced than one woul dhave thought using the calculations from our prior submission.}
+With our more precise definition (see answer to Question 5 above), we now have estimates that are more in line with the IPW estimates.
+So while there is likely bias, this bias is likely less pronounced than one would have thought using the calculations from our prior submission.
 % \textcolor{red}{To Do!}

### 2022-03-30 write-up/covid-supplement-writeup.tex
updated paper; ready

@@
 Due to the low number of deaths per strata, i.e., based on Race, Age, Sex and Ethnicity, we employ a kernel smoothing of the new infections per age strata into these sub-strata using relative number of observed deaths.
-This ensured a suitably complex model that could fit the data while generating strata-specific active infection rates that can be used in the doubly-robust estimation procedure.
+This ensured a suitably complex model that fits the observed death data well while generating strata-specific active infection rates that can be used in the doubly-robust estimation procedure.
 
@@
  \includegraphics[width=.6\linewidth]{../figs/tv_air_sensitivity.png}
- \caption{Undercount based on SEIR model}
+ \caption{Model-based and Doubly Robust active infection rate estimates under~\cite{Ironse2103272118} marginal IFR estimate of $0.84$\%, under a 10\% higher marginal IFR of $0.924$\%, and under a 10\% lower marginal IFR of $0.756$\%.}
  \label{fig:tv_air_sens}

### 2022-03-30 write-up/covid-writeup.tex
updated paper; ready

@@
 Figure~\ref{fig:casesppc} presents the posterior distribution of daily new infections.
-As the active infection rate is defined as all infected individuals who’s viral load has yet to fall below detectable levels by RT-PCR testing, to construct the active infection estimates we exponentially discount the number of newly infected individuals to estimate the number of these newly infected individuals who still have an active infection on future days.
+As the active infection rate is defined as all infected individuals whose viral load has yet to fall below detectable levels by RT-PCR testing, to construct the active infection estimates we exponentially discount the number of newly infected individuals to estimate the number of these newly infected individuals who still have an active infection on future days.
+That is, the number of active infections on day~$t$ is given by~$\sum_{s=0}^t I_t^{\new} e^{-\lambda (t-s)}$.
+Strata-specific active infections are calculated similarly.
 In our analysis, the exponential discounting parameter is set to $20$ days to match with prior evidence that ``30\% to 40\% of people will still test positive at three weeks''~\citep{wlocneg}.

### 2022-09-12 write-up/letters/response-ref-1-v1.tex
updating letters and sensitivity analysis

--- 2022-03-24 write-up/letters/response-ref-1-v1.tex
@@
 
-{\bf Response to Referees}
+{\bf Response to Referee}
 
-Let us start by saying that these are the most thorough and helpful reviews we have ever received on a submitted manuscript.
-We appreciate all of the very helpful comments from the editor, associate editor, and the two referees which has helped improve the paper substantially.
+We appreciate the very helpful comments from the referee which has helped improve the paper substantially.
 We have addressed the concerns as best as possible in this revision.
@@
 Any comments in {\it italics} indicate a statement from the referee report.
-
-{\bf Response to Editor and Associate Editor}
 \begin{enumerate}
-\item {\it The referees have many specific concerns they are asking to address, and their expertise and viewpoints are important to revise this manuscript to improve its clarity and impact.}
+\item {\it I appreciate the addition of the remark that states that the dynamics mentioned in this paper refer to the original strain.
+There are still lines (for example: “Upon infection with SARS-CoV-2, an incubation period (time to symptom onset) starts and lasts approximately five days”) that are more definitive than I would prefer.
+Perhaps adding in “Upon infection with the original SARS-CoV-2 variant” would be preferable.}
 
 \vspace{5mm}
-We thank the editor-in-chief.  We have addressed the two reviewers specific comments.  Responses to their concerns are detailed below and have significantly improved the clarify of this manuscript.
+We agree with this point and have gone through the manuscript to reduce the chances a reader miscontrues our explanations.
+We aim for clarity in these edits by avoiding definitive statements while not making statements that are too vauge or ambiguous.
 \vspace{5mm}
 
-\item {\it Some of these deal with the fast-evolving nature of the pandemic with new information emerging every week and changing variants with different characteristics also changing our understanding of the disease course and spread and timing, that affect the content in the paper.
-I understand that it is not feasible for a peer-reviewed paper like this to include all the latest information in a fast-changing environment like this in which understanding can change quickly, but it is important in this revision to update the KEY points as much as possible based on updated knowledge, but otherwise revise the language to make it “forward compatible” in terms of relevance as things continue to change, e.g.
-by acknowledging some details of spread dynamics may change with future variants, but presenting the current work based on a particular SARS-CoV-2 variant as illustrative of the issues that arise with each new variant and other viruses, as well.
-That is, recognizing it may not be feasible to keep updating based on emerging knowledge from each new strain, but to lock down the discussion context but add enough explanation so the reader can see its relevance for new strains.}
+\item {\it The reliance on the random sample with such a low response rate may be subject to unmeasured confounding - perhaps a sensitivity analysis showing the impact of a potential unmeasured confounder could strengthen this point.}
 
 \vspace{5mm}
-Remark 1 has been added at the end of the introduction to contextualize the manuscript.
-Specifically we clarify that our manuscript focuses on COVID-19 case count, testing, and death data collected from April 2020 through February 2021 and therefore the numbers presented on disease dynamics are based on the original strain.
-While using the framework to account for selection bias and measurement error for data arising from the delta and omicron strains would require additional work, the framework would remain an appropriate tool for this task.
+We have added a sensitivity analysis to the supplementary materials where we consider an unmeasured binary confounder~$Z_{j,t} \in \{0,1\}$.
+In this setting, {\it prevalence}~$pr(Z=1|X,Y)$ and {\it propensity} $\pi(X, Z; (\theta,\theta_z))$ both impact the sensitivity analysis.
+Imputation-based sensitivity analyses will thus depend heavily on the choice of prevalence and $\theta_z$.
+Therefore, we consider a Taylor series expansion of the IPW and DR-estimators at the estimated propesity which is equivalent to~$\theta_z = 0$.denote the coefficient in the selection propensity model and
+$$
+\hat \mu_{IPW} + \theta_z \frac{1}{\sum_{j=1}^N I^{NR}_j w_j (X_{j,t}; \hat \theta_t)} \left[ \sum_{j=1}^N I^{NR}_j \frac{1-\pi_j(X_{j,t}; \hat \theta_t)}{\pi_j(X_{j,t}; \hat \theta_t)} E[Z_j] (Y_j - \hat \mu_{IPW}) \right]
+$$
+Thus bias depends on the correlation of the error $Y_j - \hat \mu_{IPW}$ to the odds ratio times the expected value of the unmeasured binary confounder.  We use this to calculate the
+
 \vspace{5mm}
 
-\item {\it The AE also mentions a few comments as well:}
-\begin{enumerate}
-	\item {\it Title: As noted by one of the reviewers, the “paradox” is never discussed/defined in the manuscript.
-Consider changing to “Statistical estimation of Coronavirus Case Counts: Measurement…” or some alternative.}
+\item {\it On the plots that show the active infection rate estimates, perhaps including a line that is the case counts divided by the population as a floor would be worthwhile (at least in the supplement) to demonstrate that the model-based estimates are at least above this.}
+
+\vspace{5mm}
+We have added this this floor .  We clarify that ``case counts'' in the comment above should be a measure of recent case counts to be a relevant  active infection rate
 \vspace{5mm}
 
-	We agree with the AE.
-This was not updated in the prior submission simply to ensure continuity of presentation (i.e., didn’t want to change from first submission).
-Here we replace with the title ``Addressing selection bias and measurement error in COVID-19 case count data using auxiliary information.''
+\item {\it I found the grey hard to differentiate for the figures that have more than two categories – perhaps varying the shapes would help if trying to keep these grey-scale?}
+
+\vspace{5mm}
+We have added shapes to help differentiate figures while keeping the grey-scale.
 \vspace{5mm}
 
-	\item {\it Organization: The figures and tables are not located anywhere near much of their introduction/discussion.
-For example, Table 1 was discussed on page 7 and is itself located on page 22.
-Consider the reorganization such that Figures and Tables are located near their discussion.}
-	\vspace{5mm}
-
-	We apologize.
-We kept figures/tables related to the case study in Section 5 (the case study section) but can now see this led to confusion as we discuss them early in the paper.
-Tables and figures have been moved to be located much closer to their introduction.
-	\vspace{5mm}
-
-	\item {\it Discussion on the Indiana Random Sample: On page 22, table 1, it appears that the random sample is not similar to the census estimates, potentially indicating non-response bias.}
-	\vspace{5mm}
-
-    See our answer to Question 5a from Reviewer \#1.
-We now remark on the response rate in Section 5.
-Specifically, we cite how Yiannoutsos et al.
-(2021) handle non-response in prevalence estimation.
-We acknowledge the potential for a difference in characteristics between intended and final sample.
-    \vspace{5mm}
-
-	\item {\it Additionally, there may be significant potentially non-ignorable variables (e.g., political opinions and rural status) that drive non-response and are not presented.
-Is there any comment/discussion on the success of the randomization to be used as the “gold-standard” in this example?
-Much of the results appear to rely on the SRS and it ignores the potential for nonresponse bias.}
-	\vspace{5mm}
-
-	We do not see random sampling as a ``gold-standard'' nor a panacea that will fix every issue.
-Instead, we see it as a means to remove one very important source of potential bias in the estimation of population quantities of interest.
-A key takeaway of our paper is that with access to \emph{only} a non-probabilistic sample, we know there is potential bias but have little hope of reducing said bias.
-With a probability sample, we have means to reduce said bias.
-Statisticians, epidemioligists, and government officials, and health policy experts can then focus on practical issues such as non-ignorable missingness.
-If we thought, for example, political opinions and rural status are important predictors of missingness, then these should be collected as part of the random sampling protocol.
-Note, even collecting strata information that correlates with testing propensity (e.g., symptom status) in the non-probabilistic sample is insufficient to address selection bias.
-This is not true for probability samples.
-We have added this important discussion to Section 6.
-\end{enumerate}
-
-\end{enumerate}
-\newpage
-{\bf Response to Reviewer 1}
-\vspace{5mm}
-
-I am very grateful for this review.  The detailed suggestions and comments helped improve the paper in numerous ways.  See below for our responses to each particular reviewer comment.
-
-\begin{enumerate}
-\item {\bf Section 1: Introduction.} {\it It may be useful in the introduction to briefly discuss the various purposes of testing for diseases – for COVID we have largely focused on diagnostic testing of people who are symptomatic or have a known or suspect exposure.
-Diagnostic testing will never provide valid prevalence or incidence estimates, especially for a disease where a significant proportion of cases are asymptomatic or pauci-symptomatic and thus unlikely to present for diagnostic testing.
-Screening tests are more appropriate as a potential tool for reconstructing prevalence or incidence tests, although these are seldom conducted at random.
-The random and non-random testing samples appear to represent screening and diagnostic testing goals, respectively, and this is part of why they can be expected not to deliver the same prevalence estimates.}
-
-\vspace{5mm}
-In the introduction, we now connect random testing to its purpose as a screening tool and non-random testing which generates the case count data as a diagnostic tool.
-Given the infrequent nature of random testing and desire for ongoing prevalence estimates, we motivate the second objective of our paper by the question of whether randomized screening tests can be used to address selection bias in the non-random diagnostic testing to estimate prevalence or incidence over time.
-\vspace{5mm}
-
-\item {\bf Section 1.1}
-\begin{enumerate}
-	\item {\it For the first goal of the paper, it might be helpful to link the biases with those outlined in a set of papers from the Harvard Center for Disease Dynamics: \url{https://doi.org/10.1007/s10654-021-00727-7} and DOI:\url{10.1093/aje/kwaa188}}
-	\vspace{5mm}
-
-	We now connect with this literature on potential sources of biases in Section 1.1.
-We highlight that our work adds a new perspective on this critical issue, from the error decomposition perspective of Meng et al.
-(2018).
-	\vspace{5mm}
-
-	\item {\it For the second goal of the paper, it would be useful to discuss how the proposed method differs from the more traditional epidemiologic method of obtaining a validation sample.
-As I see it, the chief distinction is that here the “gold standard” is not a different test (although one could argue that the gold standard being used here is a screening test), but rather it is a less biased selection mechanism.
-Clarifying the links with validation studies could be helpful for epidemiologists \& statisticians alike, as the use of a validation study to correct for selection bias is a fairly novel suggestion.}
-	\vspace{5mm}
-
-	We now clarify this point in Section 1.1.
-We make it clear that our goal is to use a probability sample to correct for selection bias.
-We connect this with validation samples to build auxiliary data to address selection bias.
-
-	\vspace{5mm}
-	\item {\it Finally, consider linking the arguments you are making with recent work on ‘target validity’ (see: \url{https://doi.org/10.1093/aje/kwy228}).
-Literature on target validity has largely been focused on specifying the problem of tackling internal and external validity together, but the method you propose here is a concrete attempt to address both types of validity and extends this conversation nicely.}
-	\vspace{5mm}
-
-	We now do this at the end of Section 1.1 in order to help statisticians thinking about validity connect our arguments to this literature.
-	\vspace{5mm}
-\end{enumerate}
-\item {\bf Section 2.1: Diagnostic testing}
-\begin{enumerate}
-	\item {\it Please clarify in the first paragraph the distinction between incubation period (time to symptom onset) and latent period (time to infectiousness).
-The viral load will be detectable by at least the end of the latent period which for SARS-CoV-2 occurs before the end of the incubation period.
-However, individuals are unlikely to present for diagnostic testing until after the end of the incubation period.}
-	\vspace{5mm}
-
-	We now clarify this distinction between incubation and latent period.
-	\vspace{5mm}
-	\item {\it Please also clarify the terminology ‘active infection’ which in some instances appears to refer to individuals in the symptomatic period of infection, and in other instances appears to refer to individuals in the infectious period of infection.
-We know that these are not identical in SARS-CoV-2 infection and that difference is a key reason why screening and diagnostic tests are not easily comparable.}
-	\vspace{5mm}
-
-	\textcolor{red}{We now clarify that we define active infection as individuals infected with SARS-CoV-2 who have yet to fall below a threshold for detectable viral load levels by RT-PCR testing.
-This includes the infectious period as well as a pre-infectious and post-infectious periods of time.}
-
-	\vspace{5mm}
-	\item {\it Consider adding some additional information on screening tests, particular anything relevant to the random sample conducted in Indiana which is used as the validation set.
-Were the screening tests conducted using nasopharyngeal swabs (in many places screening tests use nasal swabs instead and often self-collected samples)?
-Were the molecular tests used to assay the samples the same for screening and diagnostic tests?
-Differences in sample collection and testing processes could undermine the utility of this random sample as a validation set.}
-	\vspace{5mm}
-
-	We clarify that, to the best of our knowledge, both Indiana Department of Health's (IDOH) molecular testing and the random state-wide sample relied primarily on nasopharyngeal swab using the same set of RT-PCR tests.
-This was primarily due to the random sampling being done by a scientific team working closely with the IDOH.
-	\vspace{5mm}
-\end{enumerate}
-\item {\bf Section 2.2: Publicly available data}
-\begin{enumerate}
-	\item {\it This section could benefit from a few remarks on why it is important / useful to focus on publicly available testing data.
-There are certainly distinct statistical issues, if public data are aggregated, but are the selection bias and measurement error issues different between public and government datasets?
-If so, how do the solutions presented here apply to government datasets?
-Can similar or the same approaches be used by health departments to assess the pandemic and guide responses?}
-	\vspace{5mm}
-
-	We have added Remark 3 in Section 2.2 discussing public versus government data.
-Specifically, we clarify that data privacy restrictions enforce publicly released testing and case count data be aggregated to strata-level covariates and have potentially important missing information (i.e., symptom status and indicator of contact with COVID-19 positive individual).
-	While these choices impact our data analytic plan, we clarify that the data analytic framework is expressly designed to guide health departments and policy makers to assess the pandemic and guide future responses.
-
-	\vspace{5mm}
-	\item {\it A timeline of restrictions and testing eligibility would be a useful addition to Section 2.2.1.
-Also, what there no time period where testing in Indiana was restricted on the basis of travel history or suspected exposure?}
-	\vspace{5mm}
-
-	Section 2.2.1 has a reconstructed timeline of restrictions, which was based on news reports of state-level testing restrictions.
-	\vspace{5mm}
-\end{enumerate}
-\item {\bf Section 2.3: Probabilistic Samples}
-\begin{enumerate}
-	\item {\it The $\approx 30\%$ response rate in the supposedly random sample requires some remarks.
-Even if the original list of eligible individuals was random, it seems unlikely that the final sample remains random with such low response.
-Is it possible to describe any characteristics that may differ between the intended sample and the final sample?
-Certainly the final random sample differs in many ways from the Census data presented in Table 1.
-Also, please clarify if the Census data is US total or Indiana specific.
-}
-	\vspace{5mm}
-
-	We now remark on the response rate in Section 5.
-Specifically, we cite how Yiannoutsos et al.
-(2021) handle non-response in prevalence estimation.
-We acknowledge the potential for a difference in characteristics between intended and final sample.
-We clarify that the Census data is Indiana specific and is presented as a reference to demonstrate the potential difference between intended and final sample.
-
-	\vspace{5mm}
-	\item {\it For the Facebook / CMU survey sample, please present the age breakdown of sampled individuals.
-It is well-known that the age-distribution of Facebook users differs from the general population, so this is unlikely to be a representative sample of the Indiana population.
-Other demographic information on this sample would be valuable too.
-(Why is it not included in Table 1?)}
-	\vspace{5mm}
-
-	The difference in age breakdown of sampled individuals in the Facebook / CMU survey, now referred to as Delphi's COVID-19 Trends and Impact Survey (CTIS), from the general population is acknowledged in Section 5.
-Table 1 now presents estimates of all demographic and covariate information using Delphi CTIS's weighting procedure.
-These weights are used in computing probabilities of selection as shown in Equation (4.2) in Section 4.1.1.
-Estimates for 'Fever' and 'Household + Case' only use data collected between April 25th and 29th to match the timing of the random survey.
-Note that individual-level racial demographics were not provided by CTIS due to data privacy concerns and there was no corresponding question on percent of tests from individuals with a prior positive test.
-
-	\vspace{5mm}
-\end{enumerate}
-\item {\bf Section 3: Analysis of case count data}
-\begin{enumerate}
-	\item {\it Please clarify the definition of ‘active infection’ (any infection, vs infectious, vs symptomatic, vs detectable by molecular testing?).
-This has important implications for the construction of the compartmental model and for the definition of your period prevalence (AIR).}
-	\vspace{5mm}
-
-	\textcolor{red}{As stated in answer to Question 3a, we now clarify that we define active infection as individuals infected with SARS-CoV-2 who have yet to fall below a threshold for detectable viral load levels by RT-PCR testing.
-As stated here, this includes the infectious period as well as a pre-infectious and post-infectious periods of time.
-We discuss how this impacts the compartmental model and calculation of the AIR in answer to Questions 6a and 17a.}
-	\vspace{5mm}
-	\item {\it Throughout, there is a general issue of too few distinct variable identifiers – for example, using Y to refer to infection status and y to refer to test status will likely lead to confusion.}
-	\vspace{5mm}
-
-	We clarify that this notation was chosen to match standard survey statistics notation: $Y_i$ is the true, fixed value for individual $i$ from the population, while $y_i$ is the observed value for individual $i$ from the sample.
-To help the reader, we now include a notation glossary in Appendix B.
-	\vspace{5mm}
-
-	\item {\it The “natural candidate for AIR” presented is the quantity that has more commonly been referred to as the “test positivity rate” (assuming that yi does in fact indicate test status rather than infection status which is obviously unknown).
-It would be helpful to mention this.}
-	\vspace{5mm}
-
-	This is now stated explicitly in Section 3, paragraph 2.
-	\vspace{5mm}
-\end{enumerate}
-\item {\bf Section 3.1: Imperfect testing}
-\begin{enumerate}
-	\item {\it Here, the value of distinguishing between test status and infection status becomes clearer.
-By using $P_j$ to mean ‘error status of testing’, the derivations in this section lead to a focus on false positives and false negatives.
-However, the real quantity of interest both for individuals and for determining prevalence is surely the positive predictive value and the negative predictive value – that is, Pr($Y_j$=1|positive test result) and Pr($Y_j$=0|negative test result).
-Unlike the FP and FN which (because they are strictly functions of sensitivity and specificity) are only affected by test characteristics, the PPV and NPV are affected also by prevalence.
-Acknowledging this leads more directly to the benefit of an iterative method for validating the test results (since FP \& FN should not depend on Y or y \& thus not need an iterative method).}
-	\vspace{5mm}
-
-	The adjustment is exactly related to the Bayes rule PPV/NPV relationship with FP, FN, and $\bar Y$.
-	\textcolor{red}{I'm still a bit confused by this point.  Will review further.}
-	\vspace{5mm}
-	\item {\it Also worth noting that test positivity is not strictly binary (since it’s generally based on an underlying continuous cycle threshold), and so measurement error could potentially lead towards bias away from, or across, the null even in the absence of selection bias.}
-	\vspace{5mm}
-
-	Remark 4 is now included to discuss that the binary testing data is generated from an underlying continuous cycle threshold and that the issue of measurement error remains even in the absence of selection bias.
-	\vspace{5mm}
-
-	\item {\it The estimator $y = (\bar y_n-FP) /(1-(FP+FN))$ is a standard estimator for correcting for measurement error using FP \& FN estimated from a validation sample, where yn is the mis-measured average \& y is the corrected average.
-The use of this estimator to iteratively estimate the point prevalence is a good contribution of this paper, but the failure to distinguish between infection \& test status in the variable names somewhat obscures that this is what the author is doing.}
-	\vspace{5mm}
-
-	We now more carefully distinguish this and connect with the epidemiologic terminology to help readers from epidemiology.
-And yes, this is the standard estimator.
-We clarify, as you stated above, that the contribution of this paper was to derive this estimator from a new perspective (i.e., the decomposition perspective which didn't require any Bayesian machinery).
-	\vspace{5mm}
-
-	\item {\it Finally, in the section on effective sample size, please clarify why the FP and FN would be dependent on testing capacity.
-False positive and false negative rates, as well as sensitivity and specificity, are strictly characteristics of a given test – if increased test capacity or burden would alter the sample collection quality then FP \& FN could change.
-But if increased test capacity simply changes the underlying prevalence in the group tested (because of opening up testing to uninfected individuals), then the FP \& FN should not change, but the PPV and the NPV will change since these are dependent on prevalence.}
-	\vspace{5mm}
-
-	We clarify that increasing testing capacity may impact FP and FN rates if altering changes the sample collection quality, e.g., a testing center switches from nasal swab to oropharyngeal swabs or saliva specimen to increase testing capacity will alter measurement error due to differences in the test characteristics.
-\end{enumerate}
-\item {\bf Section 3.2: Regrettable Rates}
-\begin{enumerate}
-	\item {\it Here, and throughout the theoretical example, the model is an SIR model.
-However, we know that SARS-CoV-2 is better represented with an SEIR model, and indeed the applied example later in the manuscript uses an SEIR approach.
-Since this paper is very particularly addressing an issue with COVID data it would be beneficial to use an SEIR model throughout rather than the simpler SIR model.}
-	\vspace{5mm}
-
-	The SIR model in Section 3.2 is replaced by the SEIR model.
-We highlight that the SEIR model has been shown to better represent the SARS-CoV-2 dynamics.
-Figure 3a presents updated versions of the potential bias in ratio estimators under an SEIR model.
-The conclusion remains largely the same.
-The only difference from the prior manuscript is that the peak in bias prior to the ratio going less than 1 is less pronounced (prior submission with SIR model has an approximate peak of 5 while current manuscript has approximate peak of 2).
-	\vspace{5mm}
-\end{enumerate}
-\item {\bf Section 3.3: Effective Reproduction Number}
-\begin{enumerate}
-	\item {\it Here again, an SEIR model would be valuable.
-It is also important to clarify that the effective reproductive number can be defined in a number of subtly different ways, and that these impact both the method of estimation and the interpretation of the effect.
-See for example Cori et al; Am J Epi, 178(9):1505-12.
-2013 and Wallinga \& Tuenis Am J Epi; 160(6):509-16.
-2004 for methods that, while similar, estimate subtly different quantities.
-It would be helpful to clarify exactly what effective reproductive number is being estimated here – from the formulae, I suspect it is the type described by Cori et al rather than Wallinga \& Tuenis.}
-	\vspace{5mm}
-
-	We clarify the subtle differences between the definitions of the instantaneous and case reproductive numbers.
-We establish that our approach aims to estimate the instantaneous reproduction number, i.e., is closely related to the proposal by Cori et al.
-(2013)).
-
-	\vspace{5mm}
-	\item {\it The Rt estimator will differ if an SEIR model is used, and it is not clear (to me) that the same intuition about the error will hold.}
-	\vspace{5mm}
-
-	%https://academic.oup.com/aje/article/189/11/1438/5864525?searchresult=1, plot SIR
-	%https://academic.oup.com/aje/article/178/9/1505/89262?searchresult=1, CORI
-	%https://www.nature.com/articles/s41598-020-76563-8.pdf?origin=ppub, Althaus approximate
-
-	This estimator is originally motivated by the SIR model and is exact under the poisson likelihood; however, the moment-based estimator has been proposed as a general tool for assessing epidemic trends in real time (Xinhua Yu, Assessing Epidemic Trends in Real Time With a Simple Ratio Plot, Am J of Epid, 189(11), 2020).
-The estimator is closely related to the instantaneous reproduction numbers proposed by Cori et.
-al.
-(2013), but we now rely on recent work (Heng and Althaus, The approximately universal shapes of epidemic curves in the SEIR model, Sci Reports, 2020) to propose an approximate moment-based estimator in the SEIR setting.
-Specifically, Heng and Althaus (2020) derive the following approximate formula for the basic reproduction number under SEIR dynamics:
-	$$
-	R_0 = \frac{1 + \Lambda \left(\frac{1}{\sigma} + \frac{1}{\gamma} \right)}{S_0 + I_0 (1 + \gamma/\sigma)}.
-	$$
-	where~$\Lambda$ is the epidemic growth rate, and $S_0$/$I_0$ are the initial fractions of the population that are susceptible and infectious respectively.
-For $S_0 \approx 1$ and $I_0 \ll 1$, we have the approximate relationship $R_0 \approx 1 + (\frac{1}{\gamma} + \frac{1}{\sigma})\Lambda$.
-	, this is an approximate solution which lacks the term $\Lambda^2 \frac{1}{\gamma \sigma}$.
-It corresponds to the case of an exponentially distributed generation time with mean duration $1/\gamma + 1/\sigma$, which is the same as the solution for the SIR model assuming an infectious period of $1/\gamma + 1/\sigma$.
-For realistic growth rates and choices of $\sigma$ and $\gamma$ for the COVID-19 pandemic, this term is negligible and therefore we work with the approximate formula.
-	\vspace{5mm}
-
-	\noindent Note that for the SIR model, the basic reproduction number is~$R_0 = 1 + \frac{1}{\gamma} \Lambda$.
-So estimate of the instantaneous reproduction number is obtained by replacing $\Lambda$ by an estimate of the  instantaneous growth rate $\hat \Lambda_t = \log (K_t / K_{t-1})$.
-Then using the approximate relationship between the SEIR model and an SIR model with infectious period $1/\gamma + 1/\sigma$, we arrive at the new moment-based estimator
-	$$
-    R_t \approx 1 + \left( \frac{1}{\gamma} + \frac{1}{\sigma} \right) \log \left( \frac{K_t}{K_{t-1}} \right).
-	$$
-	The updated $R_t$ estimator shows that the same intuition about the error holds when we are considering an SEIR rather than an SIR model.
-Figure 3b shows the reproduction number estimates and potential biases under SEIR dynamics which shows the same general conclusions hold.
-For completeness, we also present the instantaneous reproduction number estimates using the proposal of Cori et.
-al.
-(2014) in Section I of the supplementary materials, which shows the same general conclusions.
-\end{enumerate}
-\item {\bf Section 3.4: Rate comparisons}
-\begin{enumerate}
-	\item {\it This section would also be more relevant with an SEIR model}
-	\vspace{5mm}
-
-	This section has been also updated to consider an SEIR model.  All conclusions remain the same.
-	\vspace{5mm}
-\end{enumerate}
-\item {\bf Section 4.1: Selection propensity estimation}
-\begin{enumerate}
-	\item {\it The use of weights to correct for sampling bias is an established practice in epidemiology.
-It would be helpful to discuss how the proposed method differs from survey sampling weights or transportability weights (see Am J Epidemiol 2017 Oct 15;186(8):1010-1014.
-doi: 10.1093/aje/kwx164.).}
-	\vspace{5mm}
-
-	Remark 5 in Section 4.1.1 discusses how the proposal is tied closely to the use of weights to correct for sampling bias.
-Specifically, in epidemiology literature, the survey sampling weights and transportability weights require knowledge of the selection mechanism, which is not available in our setting.
-Instead, we require auxiliary information in the form of a random sample to estimate the probability of selection.
-A key difference from prior approaches is the need to estimate time-varying probabilities of selection, e.g., Figure 6 shows how the selection propensity of young individuals changes dramatically over the observation window.
-	\vspace{5mm}
-	\item {\it This section, and the sections below, also become somewhat confusing to follow because of the multiple different variables which are all denoted $I_j$.
-Since a chief goal is to describe and correct for errors in the sampling process, it would be much more useful for the reader if there were different notation for each of the different sampling mechanisms being discussed/used (eg, inclusion in the random sample, inclusion in the non-random validation sample, inclusion in the original diagnostic testing data, inclusion in ??
-on day x).}
-	\vspace{5mm}
-
-	We apologize for the confusion.
-We now include distinct notation for each of the sampling mechanisms.
-To strike a balance and keep notation at a minimum, we decided to use superscripts $R$ and $NR$ to distinguish between the probability and non-probability samples.
-To facilitate comprehension, we have added a notation glossary in Section B of the supplementary materials.
-	\vspace{5mm}
-	\item {\it In the final formula (eq 4.3), I am not clear on why the weights are applied to both yk and FP (ie why sum(w*(yk-FP)) and not sum(w*yk) – FP)?
-This is very likely just me not understanding the derivation in Appendix C, but some explanation of the intuition would be helpful.}
-	\vspace{5mm}
-
-	Since weights are normalized, these are equivalent, i.e., $\sum w (x_i) FP / \sum w(x_i) = FP$.
-	\vspace{5mm}
-	\item {\it The time-varying section seems somewhat unnecessary since it is simply taking the expectation of the data over some time period – the daily data also does this, but where the time-period is 1 day.
-}
-	\vspace{5mm}
-
-	Section 4.1.3 allows researchers to choose to go beyond simpler moving window averages (i.e., daily/weekly averages).
-The kernel function allows the user to choose weighted averages that may be more robust than the uniform kernel (which is equivalent to the a moving average over some fixed window of time).
-The proposed approach helps smooth the estimated propensity over time; moreover, the flexibility of the approach for government agencies who may observed random samples at different frequencies is of practical importance.
-	\vspace{5mm}
-\end{enumerate}
-\item {\bf Section 4.2: model-based estimation}
-\begin{enumerate}
-	\item {\it The SEIR model described in the text and in the supplement differ in a way that I find confusing – the code and equations describe a simple SEIR model but the written text describes an SEmInR model with stratum-specific covariates which appear to apply to at least the I compartments and possibly the E compartments.
-From the code and equations, presumably the assumption is that all strata mix with identical probability and all strata experience the same probabilities of infection, exposure, and transmission.
-However, these assumptions are (1) unlikely to hold and (2) make it hard to understand why the model includes strata at all.
-}
-	\vspace{5mm}
-
-	We thank the reviewer for catching this issue.
-We now clarify that the model proposed does not assume equal mixing.
-Rather due to limited data, we assume a these mixing probabilities are not time varying and that the  prior probability of mixing is chosen from a symmetric Dirichlet distribution (uniform) time-varying.
-We clarify this in Section 5.2 and comment on the code to clarify this more carefully in Appendix H.
-	\textcolor{red}{Need To Do.  Keep model just sensitivity on IFR and smooth over infectious.}
-	\vspace{5mm}
-\end{enumerate}
-\item {\bf Section 4.3: Doubly Robust Estimation}
-\begin{enumerate}
-	\item {\it The term $\mu(x_j)$ in the unnumbered equation does not seem to be defined anywhere – this is presumably coming from the SEIR model given that the rest of the equation is essentially the IPW-based estimator.
-However, it is unclear to me what output from the SEIR this term includes.}
-	\vspace{5mm}
-
-	We clarify this notation in Section 4.3.
-Specifically, the SEIR model is used to generate the number of new weekly cases which allows us to calculate the number of individuals with an active infection within each strata (i.e.~$I_{k,t}$ for the $k$th strata).
-We then can use this to calculate the posterior mean for the active infection rate per strata,~$\mu_k$.
-	\vspace{5mm}
-\end{enumerate}
-\item {\bf Section 5: Applied analysis}
-\begin{enumerate}
-	\item {\it The applied example relies heavily on the idea that the random sample is a true random sample of the population of Indiana eligible to be infected with COVID, and that the non-random sample and the Facebook / CMU sample similarly provide representative measures of covariate data.
-This is not well-established by the short data descriptions given and deserves further emphasis.}
-	\vspace{5mm}
-
-	We added Remark 6 to emphasize these potential limitations and highlight how the scientific teams that collected these data sources aimed to address these important issues.
-A key point is that the case study is to demonstrate our framework for answering the \emph{what to do with this knowledge of selection bias and measurement error} point that was raised by reviewers during the first submission round.
-In Remark 6, we urge health policy experts and government officials to assess these issues when applying this framework in their own future work.
-	\vspace{5mm}
-\end{enumerate}
-\item {\bf Section 5.1.1. Disease Prevalence}
-\begin{enumerate}
-	\item {\it This section seems to assume that individuals can only be tested once but that assumption is unlikely to be true.
-In particular, people with higher exposures will get tested more frequently than people with lower exposures, and similarly people at higher risk of severe outcomes may be tested more frequently than people with lower risk.
-Calculating the sampling fraction as the (number of tests)/(total pop – previously tested) thus seems likely to provide a biased sampling fraction estimate.}
-	\vspace{5mm}
-
-	Section 5.1.1 studies testing between April 25th to 29th 2020 during which testing was limited.
-This led us to the assumption that the individuals who already tested were unlikely to be available for testing as it had only been available since March 2020 and quite limited in supply.
-We continue to calculate the sampling fraction as number of tests divided by the total population minus prior tests.
-To address the reviewers concern, we note at the end of Section 5.1.1 that using the total population in the denominator yields a very similar sampling fraction and the calculations change by a negligible amount.
-	\vspace{5mm}
-	\item {\it Also, it would be useful to clarify whether the random sample, which is being used as the gold standard, was corrected for measurement error, as well as how the issue of 30\% response rate was dealt with in obtaining the prevalence estimate.
-Given these issues, the random sample estimate could easily also be biased.}
-	\vspace{5mm}
-
-	See answers to Questions 5a and 14a above.
-The prevalence estimate comes from Yiannoutsos et al.
-(2021) who detail their adjustments for response rates and measurement error in estimation of active prevalence.
-We do not attempt to correct it further, and simply note that (unlike in Meng's case where the ``truth'' is observed) this calculation is approximate and meant to help us understand what the relative sampling rates may be in our setting.
-	\vspace{5mm}
-\end{enumerate}
-\item {\bf Section 5.1.2 Time varying estimate}
-\begin{enumerate}
-	\item {\it 5.1.2.
-In the second approach, using hospitalization data to attempt to back-predict fever status seems highly likely to be subject to survivor bias as well as biases driven by differential access to care.
-Some discussion of the issues with this approach is warranted.}
-	\vspace{5mm}
-
-	We thank the reviewer for raising this issue.
-We have added a sentence in section 5.1.2 that raises this as a limitation.
-We restate here that the need for the imputation approaches is due to the lack of publicly available symptom status information.
-We have confirmed that this data is collected by the state, but were informed that the data could not be released due to data privacy / confidentiality issues.
-Therefore, as a framework for health policy experts and government officials who have access to the complete government datasets (i.e., not just the public data), these steps would not be necessary.
-	\vspace{5mm}
-	\item {\it Also, focusing on the non-Hispanic, white male groups seems potentially to limit understanding of the scope of error.
-Although Indiana has a very high percent of white individuals, the individuals who present for testing (as well as those who test positive, are exposed in the workplace and home, and hospitalized or die) are disproportionately non-white.
-It is important to show the performance of the method in both white and non-white individuals because non-white individuals are the ones who are experiencing higher rates of illness and death and therefore will likely be more negatively impacted by biased estimates of disease prevalence.}
-	\vspace{5mm}
-
-	We acknowledge this shortcoming.
-We now include versions of Figure 6 for other groups in Section H.2.1 of the supplementary materials with a discussion of the relative propensities.
-Specifically, as the reviewer states above, individuals who identify as Hispanic, Male, and select `Some other race' were twice as likely to present for testing as non-Hispanic, white men -- empirically confirming that our data captures these systematic differences.
-	\vspace{5mm}
-\end{enumerate}
-\item {\bf 5.2 SEIR model}
-\begin{enumerate}
-	\item {\it As stated above, some description of the assumptions related to contact patterns and transmission between and among age strata is needed.
-In addition, please clarify which compartments of the model the infection-fatality rate applies to (E \& I or I only), as well as how the model handles death from non-COVID causes from these and other model compartments.}
-	\vspace{5mm}
-
-	\textcolor{red}{See our answer to Question 12 above.
-In Section 5.2, we now clarify how we calculate the infection-fatality rate from the compartments of the model.
-Similar to Johndrow (2021), the model only uses COVID-19 related deaths.
-Deaths to non-COVID causes are not directly modeled but would appear as part of the R component in the global SEIR model.}
-	\vspace{5mm}
-	\item {\it Also, the discussion of IFR would benefit greatly from being contextualized to the rest of the paper – the goal of the manuscript is specifically to provide a better estimate of who is infected.
-Such an estimate would therefore be expected to provide a more valid infection fatality rate, and this in turn implies the IFR used here must be biased.
-This should be discussed.}
-	\vspace{5mm}
-
-    \textcolor{red}{We contextualize the discussion of IFR in Remark 7 in Section 5.2.
-We acknowledge that these IFR estimates are likely biased (or potentially accurate in very different circumstances).
-A sensitivity analysis focused on the issue of choosing an appropriate IFR given the potential biases in available IFR estimates is now performed.}
-    \textcolor{red}{To Do!}
-	\vspace{5mm}
-\end{enumerate}
-\end{enumerate}
-\newpage
-
-{\bf Response to Reviewer \#2}
-
-I am incredibly grateful for this review.
-I agree that the paper has evolved beyond the current title.
-This was not updated in the prior submission simply to ensure continuity of presentation (i.e., didn't want to change from first submission).
-Here we replace with the title ``Addressing selection bias and measurement error in COVID-19 case count data using auxiliary information''.
-We next address the reviewers specific comments.
-
-\begin{enumerate}
-\item {\it The paper states some facts about the SARS-CoV-2 virus that are outdated, for example “Upon infection with SARS-CoV-2, an incubation period starts and lasts approximately five days [Lauer et al., 2020] before the viral load is high enough to be detected.” (Page 5).
-Lauer was based on the original
-strain, we see different disease dynamics with the delta strain.
-Perhaps to be more precise, the author could clarify that this is an example of the original SARS-CoV-2 incubation rather than definitative as there will likely be future variants that will change this further.
-Similarly, on page 13, the serial
-interval of 7 is stated as fact – this is again based on the original strain rather than the current disease dynamics.
-Stating these numbers as an example rather than definitive would be preferable, I believe.}
-\vspace{5mm}
-
-Remark 1 has been added at the end of the introduction to contextualize the manuscript.
-Specifically we clarify that our manuscript focuses on COVID-19 case count, testing, and death data collected from April through February 2021 and therefore the numbers presented on disease dynamics are based on the original strain.
-While using the framework to account for selection bias and measurement error for data arising from the delta and omicron strains would require additional work, the framework would remain an appropriate tool for this task.
-
-\vspace{5mm}
-\item {\it The author states: “If someone has an active infection then after the incubation period, a molecular test with perfect sensitivity will yield a positive result for the next few weeks.” (Page 5) The author states “A molecular test refers to diagnostic tests that aim to detect increased viral load” (page 5) - I think a more precise definition is needed, for example aim to detect viral load above a certain threshold.}
-\vspace{5mm}
-
-We thank the reviewer for this point. A more precise definition is provided in Section 2.1.  We also point readers to Mina et al. (2020) where the issue of testing thresholds is discussed in detail.
-
-\vspace{5mm}
-\item {\it The author states “If someone has an active infection then after the incubation period, a molecular test with perfect sensitivity will yield a positive result for the next few weeks.” (page 5) I’m not sure it is accurate that the patient will have a positive result for the next few weeks, perhaps “If someone has an active infection then after the incubation period, a molecular test with perfect sensitivity will yield a positive result while the patient has a viral load above the threshold of detection”.}
-\vspace{5mm}
-
-We thank the reviewer and now match this more careful language to be more exact in our description found in Section 2.
-\vspace{5mm}
-\item {\it The author lumps RT-PCR and antigen tests in together and discusses that a patients will likely have positive tests for several weeks while not infectious (able to transmit) – for the most part this is the case for RT-PCR tests, however antigen tests give a clearer picture of infectiousness as the viral threshold is higher and often correlates with the likelihood to transmit.}
-
-\vspace{5mm}
-We now make this distinction much clearer.
-The section focuses primarily on RT-PCR tests given they are the primary molecular test offered by the Indiana's State Department of Health (ISDH) and thus the basis for the case count and testing data in this paper.
-\vspace{5mm}
-
-\item {\it I’m not sure this statement “The primary goal of molecular tests is diagnosis of active infections in the population.” (page 5) is entirely true.
-I definitely think this is a worthwhile goal, but many in the medical community use molecular tests to determine the diagnosis of an individual patient, regardless if the infection is “active” or not (i.e.
-whether they are currently infectious rather than infected which is what I assume the author means by “active”).}
-
-\vspace{5mm}
-\textcolor{red}{We apologize for the confusion.
-See answer to Question 3b where we clarify the definition of active infection.
-An active infection was meant to refer to individuals who were infected with SARS-CoV-2 who have yet to fall below a threshold for detectable viral load levels by RT-PCR testing.
-This includes the infectious period as well as a pre-infectious and post-infectious periods of time.}
-
-\vspace{5mm}
-\item {\it On page 13, the author discusses the bias relative to the “peak” – is this the peak in cases? I believe so, but it would be nice to define a bit more clearly.}
-
-\vspace{5mm}
-We clarify in Section 3.2 that this is the peak in the fraction of infected individuals in the population which is presented in Figure 4a.
-\vspace{5mm}
-\item {\it We would expect the IFR to dramatically change after the introductions of vaccines in December 2020, making this method less applicable without introducing a time-varying component.}
-
-\vspace{5mm}
-\textcolor{red}{See our answer to Question 17b of Reviewer \#1.
-We acknowledge that these IFR estimates are likely biased and time-varying due to introduction of vaccines in 2020.
-A sensitivity analysis focused on the issue of choosing an appropriate IFR and allowing for the IFR to be time-varying is now performed.}
-\textcolor{red}{To Do!}
-
-\vspace{5mm}
-\item {\it I found Figure 3 in the supplement [confusing].
-Figure 3A shows the profile for a 35-44 white male with symptoms and COVID-19 contact and Figure 3B shows the profile for a 65-74 white female without symptoms nor COVID-19 contact – I would have expected the covariates between these (age, ethnicity, race, gender) to be the same rather than differ with the only thing differing being symptoms 1/ COVID-19 contact?
-As a minor note, the y-axis labels do not match for this plot.}
-
-\vspace{5mm}
-We have fixed Figure 3b to match on covariates and have matching y-axes.
-See our answer to Question 16b from Reviewer \#1 on additional figures we are including in the revision.
-The figure now shows that testing propensity is time-varying, increasing from near 0 in the first few months to a peak of 0.01 around December.
-This peak is 8 times lower than the peak in December for the same strata given fever and COVID-19 contact.
-\vspace{5mm}
-\item {\it The model based estimates in Figure 8 seem so different from the other estimates that it is potentially implausible and suggests misspecification of the IFR.
-Is this potentially due to the age distribution of cases and the IFR collapsing choices?
-Perhaps as a sensitivity analysis different IFR values should be examined}
-
-\vspace{5mm}
-\textcolor{red}{See our answer to Question 17b of Reviewer \#1 and Question 7 above. A sensitivity analysis focused on the issue of choosing an appropriate IFR is now performed.}
-\textcolor{red}{To Do!}
-\vspace{5mm}
-\item {\bf Minor comments:}
-	\begin{enumerate}
-		\item {\it Figure 3 mention “$\beta = 1.2$ (black)” – I’m not sure what “black” refers to?}
-		\vspace{5mm}
-
-		We have removed this and simply state the SEIR parameter values ($\beta = 1.2, \gamma = 0.15, \sigma = 0.3$) underlying the Figure.
-
-		\item {\it In the supplement, page 21, the author refers to “Figure 6a and 6a” – I believe they mean “Figure 6a and 6b”.
-This also claims to show the likelihood for all age ranges - I believe it is only for males.}
-		\vspace{5mm}
-
-		We have fixed this reference.
-
-	\end{enumerate}
 \end{enumerate}

### 2022-09-12 write-up/letters/round2/response-ref-1-v1.tex
updating letters and sensitivity analysis

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Plain Cover Letter
% LaTeX Template
%
% This template has been downloaded from:
% http://www.latextemplates.com
%
% Original author:
% Rensselaer Polytechnic Institute (http://www.rpi.edu/dept/arc/training/latex/resumes/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[11pt]{letter} % Default font size of the document, change to 10pt to fit more text

\usepackage{newcent} % Default font is the New Century Schoolbook PostScript font
%\usepackage{helvet} % Uncomment this (while commenting the above line) to use the Helvetica font

% Margins
\topmargin=-1in % Moves the top of the document 1 inch above the default
\textheight=8.5in % Total height of the text on the page before text goes on to the next page, this can be increased in a longer letter
\oddsidemargin=-10pt % Position of the left margin, can be negative or positive if you want more or less room
\textwidth=6.5in % Total width of the text, increase this if the left margin was decreased and vice-versa

\let\raggedleft\raggedright % Pushes the date (at the top) to the left, comment this line to have the date on the right
\usepackage{amsthm, amsfonts}
\usepackage{xcolor}
\usepackage{hyperref}

\def\Y{{\bf Y}}

\begin{document}

%----------------------------------------------------------------------------------------
%	ADDRESSEE SECTION
%----------------------------------------------------------------------------------------

\begin{letter}{Professor
	Peter Glynn\\
	Editor, {\em Journal of Applied Probability}}

%----------------------------------------------------------------------------------------
%	YOUR NAME & ADDRESS SECTION
%----------------------------------------------------------------------------------------

%\begin{center}
%\large
%\end{center}
%\vfill

\signature{Walter Dempsey\\
University of Michigan\\
Department of Biostatistics\\
1415 Washington Heights\\
Ann Arbor, MI 48103} % Your name for the signature at the bottom

%----------------------------------------------------------------------------------------
%	LETTER CONTENT SECTION
%----------------------------------------------------------------------------------------

\vspace{5mm}

\newpage

{\bf Response to Referees}

Let us start by saying that these are the most thorough and helpful reviews we have ever received on a submitted manuscript.  We appreciate all of the very helpful comments from the editor, associate editor, and the two referees which has helped improve the paper substantially. We have addressed the concerns as best as possible in this revision. We provide a further point-by-point explanation below. Any comments in {\it italics} indicate a statement from the referee report.

{\bf Response to Editor and Associate Editor}
\begin{enumerate}
\item {\it The referees have many specific concerns they are asking to address, and their expertise and viewpoints are important to revise this manuscript to improve its clarity and impact.}

\vspace{5mm}
We thank the editor-in-chief.  We have addressed the two reviewers' specific comments.  Responses to their concerns are detailed below and have significantly improved the clarify of this manuscript.
\vspace{5mm}

\item {\it Some of these deal with the fast-evolving nature of the pandemic with new information emerging every week and changing variants with different characteristics also changing our understanding of the disease course and spread and timing, that affect the content in the paper. I understand that it is not feasible for a peer-reviewed paper like this to include all the latest information in a fast-changing environment like this in which understanding can change quickly, but it is important in this revision to update the KEY points as much as possible based on updated knowledge, but otherwise revise the language to make it “forward compatible” in terms of relevance as things continue to change, e.g. by acknowledging some details of spread dynamics may change with future variants, but presenting the current work based on a particular SARS-CoV-2 variant as illustrative of the issues that arise with each new variant and other viruses, as well. That is, recognizing it may not be feasible to keep updating based on emerging knowledge from each new strain, but to lock down the discussion context but add enough explanation so the reader can see its relevance for new strains.}

\vspace{5mm}
Remark 1 has been added at the end of the introduction to contextualize the manuscript. Specifically we clarify that our manuscript focuses on COVID-19 case count, testing, and death data collected from April 2020 through February 2021 and therefore the numbers presented on disease dynamics are based on the original strain. While using the framework to account for selection bias and measurement error for data arising from the delta and omicron strains would require additional work, the framework would remain an appropriate tool for this task.
\vspace{5mm}

\item {\it The AE also mentions a few comments as well:}
\begin{enumerate}
	\item {\it Title: As noted by one of the reviewers, the “paradox” is never discussed/defined in the manuscript. Consider changing to “Statistical estimation of Coronavirus Case Counts: Measurement…” or some alternative.}
	\vspace{5mm}

	We agree with the AE. This was not updated in the prior submission simply to ensure continuity of presentation (i.e., didn’t want to change from first submission). Here we replace with the title ``Addressing selection bias and measurement error in COVID-19 case count data using auxiliary information.''
	\vspace{5mm}

	\item {\it Organization: The figures and tables are not located anywhere near much of their introduction/discussion. For example, Table 1 was discussed on page 7 and is itself located on page 22. Consider the reorganization such that Figures and Tables are located near their discussion.}
	\vspace{5mm}

	We apologize.  We kept figures/tables related to the case study in Section 5 (the case study section) but can now see this led to confusion as we discuss them early in the paper. Tables and figures have been moved to be located much closer to their introduction.
	\vspace{5mm}

	\item {\it Discussion on the Indiana Random Sample: On page 22, table 1, it appears that the random sample is not similar to the census estimates, potentially indicating non-response bias.}
	\vspace{5mm}

    See our answer to Question 5a from Reviewer \#1. We now remark on the response rate in Section 5.  Specifically, we cite how Yiannoutsos et al. (2021) handle non-response in prevalence estimation.  We acknowledge the potential for a difference in characteristics between intended and final sample.
    \vspace{5mm}

	\item {\it Additionally, there may be significant potentially non-ignorable variables (e.g., political opinions and rural status) that drive non-response and are not presented. Is there any comment/discussion on the success of the randomization to be used as the “gold-standard” in this example? Much of the results appear to rely on the SRS and it ignores the potential for nonresponse bias.}
	\vspace{5mm}

	We do not see random sampling as a ``gold-standard'' nor a panacea that will fix every issue.  Instead, we see it as a means to remove one very important source of potential bias in the estimation of population quantities of interest.  A key takeaway of our paper is that with access to \emph{only} a non-probabilistic sample, we know there is potential bias but have no way to reduce said bias.  With a probability sample, we have a means to do so.  Statisticians, epidemioligists, government officials, and health policy experts can then focus on practical issues such as addressing non-ignorable missingness.  If we thought, for example, political opinions and rural status are important predictors of missingness, then these should be collected as part of the random sampling protocol.  Note, even collecting strata information that correlates with testing propensity (e.g., symptom status) in the non-probabilistic sample is insufficient to address selection bias.  This is not true for probability samples.  We have added this important discussion to Section 6.
\end{enumerate}

\end{enumerate}
\newpage
{\bf Response to Reviewer 1}
\vspace{5mm}

I am very grateful for this review.  The detailed suggestions and comments helped improve the paper in numerous ways.  See below for our responses to each particular reviewer comment.

\begin{enumerate}
\item {\bf Section 1: Introduction.} {\it It may be useful in the introduction to briefly discuss the various purposes of testing for diseases – for COVID we have largely focused on diagnostic testing of people who are symptomatic or have a known or suspect exposure. Diagnostic testing will never provide valid prevalence or incidence estimates, especially for a disease where a significant proportion of cases are asymptomatic or pauci-symptomatic and thus unlikely to present for diagnostic testing. Screening tests are more appropriate as a potential tool for reconstructing prevalence or incidence tests, although these are seldom conducted at random. The random and non-random testing samples appear to represent screening and diagnostic testing goals, respectively, and this is part of why they can be expected not to deliver the same prevalence estimates.}

\vspace{5mm}
In the introduction, we now connect random testing to its purpose as a screening tool and non-random testing which generates the case count data as a diagnostic tool.  Given the infrequent nature of random testing and desire for ongoing prevalence estimates, we motivate the second objective of our paper by the question of whether randomized screening tests can be used to address selection bias in the non-random diagnostic testing to estimate prevalence or incidence over time.
\vspace{5mm}

\item {\bf Section 1.1}
\begin{enumerate}
	\item {\it For the first goal of the paper, it might be helpful to link the biases with those outlined in a set of papers from the Harvard Center for Disease Dynamics: \url{https://doi.org/10.1007/s10654-021-00727-7} and DOI:\url{10.1093/aje/kwaa188}}
	\vspace{5mm}

	We now connect with this literature on potential sources of biases in Section 1.1.  We highlight that our work adds a new perspective on this critical issue, from the error decomposition perspective of Meng et al. (2018).
	\vspace{5mm}

	\item {\it For the second goal of the paper, it would be useful to discuss how the proposed method differs from the more traditional epidemiologic method of obtaining a validation sample. As I see it, the chief distinction is that here the “gold standard” is not a different test (although one could argue that the gold standard being used here is a screening test), but rather it is a less biased selection mechanism. Clarifying the links with validation studies could be helpful for epidemiologists \& statisticians alike, as the use of a validation study to correct for selection bias is a fairly novel suggestion.}
	\vspace{5mm}

	We now clarify this point in Section 1.1.  We make it clear that our goal is to use a probability sample to correct for selection bias. We connect this with validation samples to build auxiliary data to address selection bias.

	\vspace{5mm}
	\item {\it Finally, consider linking the arguments you are making with recent work on ‘target validity’ (see: \url{https://doi.org/10.1093/aje/kwy228}). Literature on target validity has largely been focused on specifying the problem of tackling internal and external validity together, but the method you propose here is a concrete attempt to address both types of validity and extends this conversation nicely.}
	\vspace{5mm}

	We now do this at the end of Section 1.1 in order to help statisticians thinking about validity connect our arguments to this literature.
	\vspace{5mm}
\end{enumerate}
\item {\bf Section 2.1: Diagnostic testing}
\begin{enumerate}
	\item {\it Please clarify in the first paragraph the distinction between incubation period (time to symptom onset) and latent period (time to infectiousness). The viral load will be detectable by at least the end of the latent period which for SARS-CoV-2 occurs before the end of the incubation period. However, individuals are unlikely to present for diagnostic testing until after the end of the incubation period.}
	\vspace{5mm}

	We now clarify this distinction between incubation and latent period.
	\vspace{5mm}
	\item {\it Please also clarify the terminology ‘active infection’ which in some instances appears to refer to individuals in the symptomatic period of infection, and in other instances appears to refer to individuals in the infectious period of infection. We know that these are not identical in SARS-CoV-2 infection and that difference is a key reason why screening and diagnostic tests are not easily comparable.}
	\vspace{5mm}

	We now clarify that we define active infection as individuals infected with SARS-CoV-2 who have yet to fall below a threshold for detectable viral load levels by RT-PCR testing.
	% This includes the infectious period as well as a pre-infectious and post-infectious periods of time.

	\vspace{5mm}
	\item {\it Consider adding some additional information on screening tests, particular anything relevant to the random sample conducted in Indiana which is used as the validation set. Were the screening tests conducted using nasopharyngeal swabs (in many places screening tests use nasal swabs instead and often self-collected samples)? Were the molecular tests used to assay the samples the same for screening and diagnostic tests? Differences in sample collection and testing processes could undermine the utility of this random sample as a validation set.}
	\vspace{5mm}

	We clarify that, to the best of our knowledge, both Indiana Department of Health's (IDOH) molecular testing and the random state-wide sample relied primarily on nasopharyngeal swab using the same set of RT-PCR tests. This was primarily due to the random sampling being done by a scientific team working closely with the IDOH.
	\vspace{5mm}
\end{enumerate}
\item {\bf Section 2.2: Publicly available data}
\begin{enumerate}
	\item {\it This section could benefit from a few remarks on why it is important / useful to focus on publicly available testing data. There are certainly distinct statistical issues, if public data are aggregated, but are the selection bias and measurement error issues different between public and government datasets? If so, how do the solutions presented here apply to government datasets? Can similar or the same approaches be used by health departments to assess the pandemic and guide responses?}
	\vspace{5mm}

	We have added Remark 3 in Section 2.2 discussing public versus government data.  Specifically, we clarify that data privacy restrictions enforce publicly released testing and case count data be aggregated to strata-level covariates and have potentially important missing information (i.e., symptom status and indicator of contact with COVID-19 positive individual).
	While these choices impact our data analytic plan, we clarify that the data analytic framework is expressly designed to guide health departments and policy makers to assess the pandemic and guide future responses.

	\vspace{5mm}
	\item {\it A timeline of restrictions and testing eligibility would be a useful addition to Section 2.2.1. Also, what there no time period where testing in Indiana was restricted on the basis of travel history or suspected exposure?}
	\vspace{5mm}

	Section 2.2.1 now has a complete reconstructed timeline of restrictions, which was based on news reports of state-level testing restrictions.
	\vspace{5mm}
\end{enumerate}
\item {\bf Section 2.3: Probabilistic Samples}
\begin{enumerate}
	\item {\it The $\approx 30\%$ response rate in the supposedly random sample requires some remarks. Even if the original list of eligible individuals was random, it seems unlikely that the final sample remains random with such low response. Is it possible to describe any characteristics that may differ between the intended sample and the final sample? Certainly the final random sample differs in many ways from the Census data presented in Table 1. Also, please clarify if the Census data is US total or Indiana specific. }
	\vspace{5mm}

	We now remark on the response rate in Section 5.  Specifically, we cite how Yiannoutsos et al. (2021) handle non-response in prevalence estimation.  We acknowledge the potential for a difference in characteristics between intended and final sample.  We clarify that the Census data is Indiana specific and is presented as a reference to demonstrate the potential difference between intended and final sample.

	\vspace{5mm}
	\item {\it For the Facebook / CMU survey sample, please present the age breakdown of sampled individuals. It is well-known that the age-distribution of Facebook users differs from the general population, so this is unlikely to be a representative sample of the Indiana population. Other demographic information on this sample would be valuable too. (Why is it not included in Table 1?)}
	\vspace{5mm}

	The difference in age breakdown of sampled individuals in the Facebook / CMU survey, now referred to as Delphi's COVID-19 Trends and Impact Survey (CTIS), from the general population is acknowledged in Section 5.  Table 1 now presents estimates of all demographic and covariate information using Delphi CTIS's weighting procedure.  These weights are used in computing probabilities of selection as shown in Equation (4.2) in Section 4.1.1. Estimates for `Fever' and `Household + Case' only use data collected between April 25th and 29th to match the timing of the random survey.  Note that individual-level racial demographics were not provided by CTIS due to data privacy concerns and there was no corresponding question on percent of tests from individuals with a prior positive test.

	\vspace{5mm}
\end{enumerate}
\item {\bf Section 3: Analysis of case count data}
\begin{enumerate}
	\item {\it Please clarify the definition of ‘active infection’ (any infection, vs infectious, vs symptomatic, vs detectable by molecular testing?). This has important implications for the construction of the compartmental model and for the definition of your period prevalence (AIR).}
	\vspace{5mm}

	As stated in answer to Question 3a, we now clarify that we define active infection as individuals infected with SARS-CoV-2 who have yet to fall below a threshold for detectable viral load levels by RT-PCR testing. We discuss how this impacts the compartmental model and calculation of the AIR in answer to Question 9 from Reviewer \#2.

	\vspace{5mm}
	\item {\it Throughout, there is a general issue of too few distinct variable identifiers – for example, using Y to refer to infection status and y to refer to test status will likely lead to confusion.}
	\vspace{5mm}

	We clarify that this notation was chosen to match standard survey statistics notation: $Y_i$ is the true, fixed value for individual $i$ from the population, while $y_i$ is the observed value for individual $i$ from the sample. To help the reader, we now include a notation glossary in Appendix B.
	\vspace{5mm}

	\item {\it The “natural candidate for AIR” presented is the quantity that has more commonly been referred to as the “test positivity rate” (assuming that yi does in fact indicate test status rather than infection status which is obviously unknown). It would be helpful to mention this.}
	\vspace{5mm}

	This is now stated explicitly in Section 3, paragraph 2.
	\vspace{5mm}
\end{enumerate}
\item {\bf Section 3.1: Imperfect testing}
\begin{enumerate}
	\item {\it Here, the value of distinguishing between test status and infection status becomes clearer. By using $P_j$ to mean ‘error status of testing’, the derivations in this section lead to a focus on false positives and false negatives. However, the real quantity of interest both for individuals and for determining prevalence is surely the positive predictive value and the negative predictive value – that is, Pr($Y_j$=1|positive test result) and Pr($Y_j$=0|negative test result). Unlike the FP and FN which (because they are strictly functions of sensitivity and specificity) are only affected by test characteristics, the PPV and NPV are affected also by prevalence. Acknowledging this leads more directly to the benefit of an iterative method for validating the test results (since FP \& FN should not depend on Y or y \& thus not need an iterative method).}
	\vspace{5mm}

	As stated above, the approach benefits from the fact that FP and FN do not depend on $Y$ or $y$ and are only affected by test characteristics. The edited version of Section 3.1 tries to highlight this a bit more clearly.
	% \textcolor{red}{I'm still a bit confused by this point.  Will review further.}
	\vspace{5mm}
	\item {\it Also worth noting that test positivity is not strictly binary (since it’s generally based on an underlying continuous cycle threshold), and so measurement error could potentially lead towards bias away from, or across, the null even in the absence of selection bias.}
	\vspace{5mm}

	Remark 4 is now included to discuss that the binary testing data is generated from an underlying continuous cycle threshold and that the issue of measurement error remains even in the absence of selection bias.
	\vspace{5mm}

	\item {\it The estimator $y = (\bar y_n-FP) /(1-(FP+FN))$ is a standard estimator for correcting for measurement error using FP \& FN estimated from a validation sample, where yn is the mis-measured average \& y is the corrected average. The use of this estimator to iteratively estimate the point prevalence is a good contribution of this paper, but the failure to distinguish between infection \& test status in the variable names somewhat obscures that this is what the author is doing.}
	\vspace{5mm}

	We now more carefully distinguish this and connect with the epidemiologic terminology to help readers from epidemiology.  And yes, this is the standard estimator.  We clarify, as you stated above, that the contribution of this paper was to derive this estimator from a new perspective (i.e., the decomposition perspective which didn't require any Bayesian machinery).
	\vspace{5mm}

	\item {\it Finally, in the section on effective sample size, please clarify why the FP and FN would be dependent on testing capacity. False positive and false negative rates, as well as sensitivity and specificity, are strictly characteristics of a given test – if increased test capacity or burden would alter the sample collection quality then FP \& FN could change. But if increased test capacity simply changes the underlying prevalence in the group tested (because of opening up testing to uninfected individuals), then the FP \& FN should not change, but the PPV and the NPV will change since these are dependent on prevalence.}
	\vspace{5mm}

	We clarify that increasing testing capacity may impact FP and FN rates if altering changes the sample collection quality, e.g., a testing center switches from nasal swab to oropharyngeal swabs or saliva specimen to increase testing capacity will alter measurement error due to differences in the test characteristics.
\end{enumerate}
\item {\bf Section 3.2: Regrettable Rates}
\begin{enumerate}
	\item {\it Here, and throughout the theoretical example, the model is an SIR model. However, we know that SARS-CoV-2 is better represented with an SEIR model, and indeed the applied example later in the manuscript uses an SEIR approach. Since this paper is very particularly addressing an issue with COVID data it would be beneficial to use an SEIR model throughout rather than the simpler SIR model.}
	\vspace{5mm}

	The SIR model in Section 3.2 is replaced by the SEIR model.  We highlight that the SEIR model has been shown to better represent the SARS-CoV-2 dynamics.  Figure 3a presents updated versions of the potential bias in ratio estimators under an SEIR model.   The conclusion remains largely the same. The only difference from the prior manuscript is that the peak in bias prior to the ratio going less than 1 is less pronounced (prior submission with SIR model has an approximate peak of 5 while current manuscript has approximate peak of 2).
	\vspace{5mm}
\end{enumerate}
\item {\bf Section 3.3: Effective Reproduction Number}
\begin{enumerate}
	\item {\it Here again, an SEIR model would be valuable. It is also important to clarify that the effective reproductive number can be defined in a number of subtly different ways, and that these impact both the method of estimation and the interpretation of the effect. See for example Cori et al; Am J Epi, 178(9):1505-12. 2013 and Wallinga \& Tuenis Am J Epi; 160(6):509-16. 2004 for methods that, while similar, estimate subtly different quantities. It would be helpful to clarify exactly what effective reproductive number is being estimated here – from the formulae, I suspect it is the type described by Cori et al rather than Wallinga \& Tuenis.}
	\vspace{5mm}

	We clarify the subtle differences between the definitions of the instantaneous and case reproductive numbers.  We establish that our approach aims to estimate the instantaneous reproduction number, i.e., is closely related to the proposal by Cori et al. (2013)).

	\vspace{5mm}
	\item {\it The Rt estimator will differ if an SEIR model is used, and it is not clear (to me) that the same intuition about the error will hold.}
	\vspace{5mm}

	%https://academic.oup.com/aje/article/189/11/1438/5864525?searchresult=1, plot SIR
	%https://academic.oup.com/aje/article/178/9/1505/89262?searchresult=1, CORI
	%https://www.nature.com/articles/s41598-020-76563-8.pdf?origin=ppub, Althaus approximate

	This estimator is originally motivated by the SIR model and is exact under the poisson likelihood; however, the moment-based estimator has been proposed as a general tool for assessing epidemic trends in real time (Xinhua Yu, Assessing Epidemic Trends in Real Time With a Simple Ratio Plot, Am J of Epid, 189(11), 2020).  The estimator is closely related to the instantaneous reproduction numbers proposed by Cori et. al. (2013), but we now rely on recent work (Heng and Althaus, The approximately universal shapes of epidemic curves in the SEIR model, Sci Reports, 2020) to propose an approximate moment-based estimator in the SEIR setting. Specifically, Heng and Althaus (2020) derive the following approximate formula for the basic reproduction number under SEIR dynamics:
	$$
	R_0 = \frac{1 + \Lambda \left(\frac{1}{\sigma} + \frac{1}{\gamma} \right)}{S_0 + I_0 (1 + \gamma/\sigma)}.
	$$
	where~$\Lambda$ is the epidemic growth rate, and $S_0$/$I_0$ are the initial fractions of the population that are susceptible and infectious respectively.  For $S_0 \approx 1$ and $I_0 \ll 1$, we have the approximate relationship $R_0 \approx 1 + (\frac{1}{\gamma} + \frac{1}{\sigma})\Lambda$.
	, this is an approximate solution which lacks the term $\Lambda^2 \frac{1}{\gamma \sigma}$.  It corresponds to the case of an exponentially distributed generation time with mean duration $1/\gamma + 1/\sigma$, which is the same as the solution for the SIR model assuming an infectious period of $1/\gamma + 1/\sigma$. For realistic growth rates and choices of $\sigma$ and $\gamma$ for the COVID-19 pandemic, this term is negligible and therefore we work with the approximate formula.
	\vspace{5mm}

	\noindent Note that for the SIR model, the basic reproduction number is~$R_0 = 1 + \frac{1}{\gamma} \Lambda$. So estimate of the instantaneous reproduction number is obtained by replacing $\Lambda$ by an estimate of the  instantaneous growth rate $\hat \Lambda_t = \log (K_t / K_{t-1})$.  Then using the approximate relationship between the SEIR model and an SIR model with infectious period $1/\gamma + 1/\sigma$, we arrive at the new moment-based estimator
	$$
    R_t \approx 1 + \left( \frac{1}{\gamma} + \frac{1}{\sigma} \right) \log \left( \frac{K_t}{K_{t-1}} \right).
	$$
	The updated $R_t$ estimator shows that the same intuition about the error holds when we are considering an SEIR rather than an SIR model.  Figure 3b shows the reproduction number estimates and potential biases under SEIR dynamics which shows the same general conclusions hold.  For completeness, we also present the instantaneous reproduction number estimates using the proposal of Cori et. al. (2014) in Section I of the supplementary materials, which shows the same general conclusions.
\end{enumerate}
\item {\bf Section 3.4: Rate comparisons}
\begin{enumerate}
	\item {\it This section would also be more relevant with an SEIR model}
	\vspace{5mm}

	This section has been also updated to consider an SEIR model.  All conclusions remain the same.
	\vspace{5mm}
\end{enumerate}
\item {\bf Section 4.1: Selection propensity estimation}
\begin{enumerate}
	\item {\it The use of weights to correct for sampling bias is an established practice in epidemiology. It would be helpful to discuss how the proposed method differs from survey sampling weights or transportability weights (see Am J Epidemiol 2017 Oct 15;186(8):1010-1014. doi: 10.1093/aje/kwx164.).}
	\vspace{5mm}

	Remark 5 in Section 4.1.1 discusses how the proposal is tied closely to the use of weights to correct for sampling bias.  Specifically, in epidemiology literature, the survey sampling weights and transportability weights require knowledge of the selection mechanism, which is not available in our setting.  Instead, we require auxiliary information in the form of a random sample to estimate the probability of selection. A key difference from prior approaches is the need to estimate time-varying probabilities of selection, e.g., Figure 6 shows how the selection propensity of young individuals changes dramatically over the observation window.
	\vspace{5mm}
	\item {\it This section, and the sections below, also become somewhat confusing to follow because of the multiple different variables which are all denoted $I_j$. Since a chief goal is to describe and correct for errors in the sampling process, it would be much more useful for the reader if there were different notation for each of the different sampling mechanisms being discussed/used (eg, inclusion in the random sample, inclusion in the non-random validation sample, inclusion in the original diagnostic testing data, inclusion in ?? on day x).}
	\vspace{5mm}

	We apologize for the confusion.  We now include distinct notation for each of the sampling mechanisms.  To strike a balance and keep notation at a minimum, we decided to use superscripts $R$ and $NR$ to distinguish between the probability and non-probability samples.  To facilitate comprehension, we have added a notation glossary in Section B of the supplementary materials.
	\vspace{5mm}
	\item {\it In the final formula (eq 4.3), I am not clear on why the weights are applied to both yk and FP (ie why sum(w*(yk-FP)) and not sum(w*yk) – FP)? This is very likely just me not understanding the derivation in Appendix C, but some explanation of the intuition would be helpful.}
	\vspace{5mm}

	Since weights are normalized, these are equivalent, i.e., $\sum w (x_i) FP / \sum w(x_i) = FP$.  We continue to write it this way to keep the formulas compact.
	\vspace{5mm}
	\item {\it The time-varying section seems somewhat unnecessary since it is simply taking the expectation of the data over some time period – the daily data also does this, but where the time-period is 1 day. }
	\vspace{5mm}

	Section 4.1.3 allows researchers to choose to go beyond simpler moving window averages (i.e., daily/weekly averages).  The kernel function allows the user to choose weighted averages that may be more robust than the uniform kernel (which is equivalent to the a moving average over some fixed window of time).  The proposed approach helps smooth the estimated propensity over time; moreover, the flexibility of the approach for government agencies who may observed random samples at different frequencies is of practical importance.
	\vspace{5mm}
\end{enumerate}
\item {\bf Section 4.2: model-based estimation}
\begin{enumerate}
	\item {\it The SEIR model described in the text and in the supplement differ in a way that I find confusing – the code and equations describe a simple SEIR model but the written text describes an SEmInR model with stratum-specific covariates which appear to apply to at least the I compartments and possibly the E compartments. From the code and equations, presumably the assumption is that all strata mix with identical probability and all strata experience the same probabilities of infection, exposure, and transmission. However, these assumptions are (1) unlikely to hold and (2) make it hard to understand why the model includes strata at all. }
	\vspace{5mm}

	We thank the reviewer for catching this issue.  We now clarify that the model proposed does not assume equal mixing.  Due to limited death data, we consider a population-level SEIR model and extract new infection counts at each time~$\{ I^{new}_t \}$. We then generate strata-specific new infection counts at each time~$\{ I^{new}_{t,k} \}$ via a Multinomial distribution given the new infection counts~$\{ I^{new}_t \}$ and parameters~$\{ p_{t,k} \} $ that has a symmetric Dirichlet prior.  These strata-specific new infection counts are then linked to the observed strata-level death counts via the conditional Poisson model using the infection fatality rate.  We clarify this in Section 5.2 and comment on the code to clarify this more carefully in Appendix H.
	\vspace{5mm}
\end{enumerate}
\item {\bf Section 4.3: Doubly Robust Estimation}
\begin{enumerate}
	\item {\it The term $\mu(x_j)$ in the unnumbered equation does not seem to be defined anywhere – this is presumably coming from the SEIR model given that the rest of the equation is essentially the IPW-based estimator. However, it is unclear to me what output from the SEIR this term includes.}
	\vspace{5mm}

	We clarify this notation in Section 4.3.  Specifically, the SEIR model is used to generate the number of new daily infections which are then used to  calculate the number of individuals with an active infection within each strata.
	\vspace{5mm}
\end{enumerate}
\item {\bf Section 5: Applied analysis}
\begin{enumerate}
	\item {\it The applied example relies heavily on the idea that the random sample is a true random sample of the population of Indiana eligible to be infected with COVID, and that the non-random sample and the Facebook / CMU sample similarly provide representative measures of covariate data. This is not well-established by the short data descriptions given and deserves further emphasis.}
	\vspace{5mm}

	We added Remark 6 to emphasize these potential limitations and highlight how the scientific teams that collected these data sources aimed to address these important issues. A key point is that the case study is to demonstrate our framework for answering the \emph{what to do with this knowledge of selection bias and measurement error} point that was raised by reviewers during the first submission round.  In Remark 6, we urge health policy experts and government officials to assess these issues when applying this framework in their own future work.
	\vspace{5mm}
\end{enumerate}
\item {\bf Section 5.1.1. Disease Prevalence}
\begin{enumerate}
	\item {\it This section seems to assume that individuals can only be tested once but that assumption is unlikely to be true. In particular, people with higher exposures will get tested more frequently than people with lower exposures, and similarly people at higher risk of severe outcomes may be tested more frequently than people with lower risk. Calculating the sampling fraction as the (number of tests)/(total pop – previously tested) thus seems likely to provide a biased sampling fraction estimate.}
	\vspace{5mm}

	Section 5.1.1 studies testing between April 25th to 29th 2020 during which testing was limited.  This led us to the assumption that the individuals who already tested were unlikely to be available for testing as it had only been available since March 2020 and quite limited in supply.  We continue to calculate the sampling fraction as number of tests divided by the total population minus prior tests.  To address the reviewer's concern, we note at the end of Section 5.1.1 that using the total population in the denominator yields a very similar sampling fraction and the calculations change by a negligible amount.
	\vspace{5mm}
	\item {\it Also, it would be useful to clarify whether the random sample, which is being used as the gold standard, was corrected for measurement error, as well as how the issue of 30\% response rate was dealt with in obtaining the prevalence estimate. Given these issues, the random sample estimate could easily also be biased.}
	\vspace{5mm}

	See answers to Questions 5a and 14a above.  The prevalence estimate comes from Yiannoutsos et al. (2021) who detail their adjustments for response rates and measurement error in estimation of active prevalence.  We do not attempt to correct it further, and simply note that (unlike in Meng's case where the ``truth'' is observed) this calculation is approximate and meant to help us understand what the relative sampling rates may be in our setting.
	\vspace{5mm}
\end{enumerate}
\item {\bf Section 5.1.2 Time varying estimate}
\begin{enumerate}
	\item {\it 5.1.2. In the second approach, using hospitalization data to attempt to back-predict fever status seems highly likely to be subject to survivor bias as well as biases driven by differential access to care. Some discussion of the issues with this approach is warranted.}
	\vspace{5mm}

	We thank the reviewer for raising this issue.  We have added a sentence in section 5.1.2 that raises this as a limitation.  We restate here that the need for the imputation approaches is due to the lack of publicly available symptom status information.  We have confirmed that this data is collected by the state, but were informed that the data could not be released due to data privacy / confidentiality issues.  Therefore, as a framework for health policy experts and government officials who have access to the complete government datasets (i.e., not just the public data), these steps would not be necessary.
	\vspace{5mm}
	\item {\it Also, focusing on the non-Hispanic, white male groups seems potentially to limit understanding of the scope of error. Although Indiana has a very high percent of white individuals, the individuals who present for testing (as well as those who test positive, are exposed in the workplace and home, and hospitalized or die) are disproportionately non-white. It is important to show the performance of the method in both white and non-white individuals because non-white individuals are the ones who are experiencing higher rates of illness and death and therefore will likely be more negatively impacted by biased estimates of disease prevalence.}
	\vspace{5mm}

	We acknowledge this shortcoming.  We now include versions of Figure 6 for other groups in Section H.2.1 of the supplementary materials with a discussion of the relative propensities.  Specifically, as the reviewer states above, individuals who identify as Hispanic, Male, and select `Some other race' were twice as likely to present for testing as non-Hispanic, white men -- empirically confirming that our data captures these systematic differences.
	\vspace{5mm}
\end{enumerate}
\item {\bf 5.2 SEIR model}
\begin{enumerate}
	\item {\it As stated above, some description of the assumptions related to contact patterns and transmission between and among age strata is needed. In addition, please clarify which compartments of the model the infection-fatality rate applies to (E \& I or I only), as well as how the model handles death from non-COVID causes from these and other model compartments.}
	\vspace{5mm}

	See our answer to Question 12a above.  In Section 5.2, we now clarify how the infection-fatality rate is used in our model. Similar to Johndrow (2021), the model only uses COVID-19 related deaths.  Deaths to non-COVID causes are not directly modeled but would appear as part of the R component in the population-level SEIR model. Moreover, since the active infection rate is defined as all infected individuals who’s viral load has yet to fall below detectable levels by RT-PCR testing, to construct the active infection estimates we exponentially discount the number of newly infected individuals to estimate the number of individuals with an active infection on each day per strata. In our analysis, the exponential discounting is set to 20 days to match with prior evidence that 30\% to 40\% of people will still test positive at three weeks.

	\vspace{5mm}
	\item {\it Also, the discussion of IFR would benefit greatly from being contextualized to the rest of the paper – the goal of the manuscript is specifically to provide a better estimate of who is infected. Such an estimate would therefore be expected to provide a more valid infection fatality rate, and this in turn implies the IFR used here must be biased. This should be discussed.}
	\vspace{5mm}

   	We contextualize the discussion of IFR in Remark 7 in Section 5.2. We acknowledge that these IFR estimates may be biased.  A sensitivity analysis focused on the issue of choosing an appropriate IFR given the potential biases in available IFR estimates is now performed and presented in Appendix I in the supplementary materials.

	\vspace{5mm}
\end{enumerate}
\end{enumerate}
\newpage

{\bf Response to Reviewer \#2}

I am incredibly grateful for this review.  I agree that the paper has evolved beyond the current title.  This was not updated in the prior submission simply to ensure continuity of presentation (i.e., didn't want to change from first submission).  Here we replace with the title ``Addressing selection bias and measurement error in COVID-19 case count data using auxiliary information''.  We next address the reviewers specific comments.

\begin{enumerate}
\item {\it The paper states some facts about the SARS-CoV-2 virus that are outdated, for example “Upon infection with SARS-CoV-2, an incubation period starts and lasts approximately five days [Lauer et al., 2020] before the viral load is high enough to be detected.” (Page 5). Lauer was based on the original
strain, we see different disease dynamics with the delta strain. Perhaps to be more precise, the author could clarify that this is an example of the original SARS-CoV-2 incubation rather than definitative as there will likely be future variants that will change this further. Similarly, on page 13, the serial
interval of 7 is stated as fact – this is again based on the original strain rather than the current disease dynamics. Stating these numbers as an example rather than definitive would be preferable, I believe.}
\vspace{5mm}

Remark 1 has been added at the end of the introduction to contextualize the manuscript.  Specifically we clarify that our manuscript focuses on COVID-19 case count, testing, and death data collected from April through February 2021 and therefore the numbers presented on disease dynamics are based on the original strain.  While using the framework to account for selection bias and measurement error for data arising from the delta and omicron strains would require additional work, the framework would remain an appropriate tool for this task.

\vspace{5mm}
\item {\it The author states: “If someone has an active infection then after the incubation period, a molecular test with perfect sensitivity will yield a positive result for the next few weeks.” (Page 5) The author states “A molecular test refers to diagnostic tests that aim to detect increased viral load” (page 5) - I think a more precise definition is needed, for example aim to detect viral load above a certain threshold.}
\vspace{5mm}

We thank the reviewer for this point. A more precise definition is provided in Section 2.1.  We also point readers to Mina et al. (2020) where the issue of testing thresholds is discussed in detail.

\vspace{5mm}
\item {\it The author states “If someone has an active infection then after the incubation period, a molecular test with perfect sensitivity will yield a positive result for the next few weeks.” (page 5) I’m not sure it is accurate that the patient will have a positive result for the next few weeks, perhaps “If someone has an active infection then after the incubation period, a molecular test with perfect sensitivity will yield a positive result while the patient has a viral load above the threshold of detection”.}
\vspace{5mm}

We thank the reviewer and now match this more careful language to be more exact in our description found in Section 2.
\vspace{5mm}
\item {\it The author lumps RT-PCR and antigen tests in together and discusses that a patients will likely have positive tests for several weeks while not infectious (able to transmit) – for the most part this is the case for RT-PCR tests, however antigen tests give a clearer picture of infectiousness as the viral threshold is higher and often correlates with the likelihood to transmit.}

\vspace{5mm}
We now make this distinction much clearer.  The section focuses primarily on RT-PCR tests given they are the primary molecular test offered by the Indiana's State Department of Health (ISDH) and thus the basis for the case count and testing data in this paper.
\vspace{5mm}

\item {\it I’m not sure this statement “The primary goal of molecular tests is diagnosis of active infections in the population.” (page 5) is entirely true. I definitely think this is a worthwhile goal, but many in the medical community use molecular tests to determine the diagnosis of an individual patient, regardless if the infection is “active” or not (i.e. whether they are currently infectious rather than infected which is what I assume the author means by “active”).}

\vspace{5mm}
We apologize for the confusion.  See our answer to Question 3b from Reviewer \#1 where we clarify the definition of active infection.  An active infection was meant to refer to individuals who were infected with SARS-CoV-2 who have yet to fall below a threshold for detectable viral load levels by RT-PCR testing.
% This includes the infectious period as well as a pre-infectious and post-infectious periods of time.}

\vspace{5mm}
\item {\it On page 13, the author discusses the bias relative to the “peak” – is this the peak in cases? I believe so, but it would be nice to define a bit more clearly.}

\vspace{5mm}
We clarify in Section 3.2 that this is the peak in the fraction of infected individuals in the population which is presented in Figure 4a.
\vspace{5mm}
\item {\it We would expect the IFR to dramatically change after the introductions of vaccines in December 2020, making this method less applicable without introducing a time-varying component.}

\vspace{5mm}
See our answer to Question 17b of Reviewer \#1. We acknowledge that these IFR estimates may be biased and/or time-varying.  Note the study period ends January 2021 so the bias due to vaccines may be limited.  We now include a sensitivity analysis focused on the issue of choosing an appropriate IFR in Appendix I in the supplementary materials.
% \textcolor{red}{To Do!}

\vspace{5mm}
\item {\it I found Figure 3 in the supplement [confusing]. Figure 3A shows the profile for a 35-44 white male with symptoms and COVID-19 contact and Figure 3B shows the profile for a 65-74 white female without symptoms nor COVID-19 contact – I would have expected the covariates between these (age, ethnicity, race, gender) to be the same rather than differ with the only thing differing being symptoms 1/ COVID-19 contact? As a minor note, the y-axis labels do not match for this plot.}

\vspace{5mm}
We have fixed Figure 3b to match on covariates and have matching y-axes.  See our answer to Question 16b from Reviewer \#1 on additional figures we are including in the revision.  The figure now shows that testing propensity is time-varying, increasing from near 0 in the first few months to a peak of 0.01 around December.  This peak is 8 times lower than the peak in December for the same strata given fever and COVID-19 contact.
\vspace{5mm}
\item {\it The model based estimates in Figure 8 seem so different from the other estimates that it is potentially implausible and suggests misspecification of the IFR. Is this potentially due to the age distribution of cases and the IFR collapsing choices? Perhaps as a sensitivity analysis different IFR values should be examined}

\vspace{5mm}
See our answer to Question 17b of Reviewer \#1 and Question 7 above. A sensitivity analysis focused on the issue of choosing an appropriate IFR is now performed in Appendix I in the supplementary materials.  We also note that the active infection estimate being very different for the model-based method in the prior manuscript was a result of how we decided to define the active infection rate using the SEIR model.  With our more precise definition (see answer to Question 5 above), we now have estimates that are more in line with the IPW estimates.  So while there is likely bias, this bias is likely less pronounced than one would have thought using the calculations from our prior submission.
% \textcolor{red}{To Do!}
\vspace{5mm}
\item {\bf Minor comments:}
	\begin{enumerate}
		\item {\it Figure 3 mention “$\beta = 1.2$ (black)” – I’m not sure what “black” refers to?}
		\vspace{5mm}

		We have removed this and simply state the SEIR parameter values ($\beta = 1.2, \gamma = 0.15, \sigma = 0.3$) underlying the Figure.

		\item {\it In the supplement, page 21, the author refers to “Figure 6a and 6a” – I believe they mean “Figure 6a and 6b”. This also claims to show the likelihood for all age ranges - I believe it is only for males.}
		\vspace{5mm}

		We have fixed this reference.

	\end{enumerate}
\end{enumerate}

\end{letter}

\end{document}

### 2022-09-12 write-up/covid-supplement-writeup.tex
updating letters and sensitivity analysis

--- 2022-03-30 write-up/covid-supplement-writeup.tex
@@
 
+\def\logit{\text{logit}}
 \def\pr{\text{pr}}
@@
 \def\code#1{\texttt{#1}}
+\def\EE{\mathbb{E}}
+\newcommand{\indep}{\perp \!\!\! \perp}
 
@@
 
+\section{Sensitivity Analysis}
+
+The proposed method relies on the assumption $I_{j,t}^{(NR)} \indep Y_{j,t} \, | \, X_{j,t}$, i.e., the sampling indicator in the non-probabilistic survey is conditionally independent of the outcome given the covariates.
+However, estimates of population quantities based on inverse-weighting will be biased in the presence of `unobserved confounding'.
+We may hope that the bias
+
+The purpose of this section is to introduce XXX,
+
+We consider a
+\begin{align*}
+\logit \pi \left( I_{t} = 1  | x_{t}, u_{t} \right) &= h_t(x_t) + \alpha u \\
+E[ Y | i_t, x_t, u_t] = l(i_t, x_t)
+\end{align*}
+for some function~$h_t$, $l_t$ that are allowed to depend on time. Re-arranging the terms we see that .  This can be
+$$
+E[ Y | i_t x_t, u_t ] = l(t,x)  + \frac{\delta}{\alpha} \left( \logit \pi \left( I_{t} = 1  | x_{t}, u_{t} \right) - h_t (x_t) \right)
+$$
+The key idea in XXXX is that positing a distribution on $P(I_t = 1 | x_y, u_t)$ directly allows one to circumvent the need to specify a distribution for~$U_t$, a highly non-trivial task.
+
+The logit-linear model does not lead to a tractable sensitivy analysis.  Instead, we define a sensitivity model as:
+\begin{equation}
+\begin{aligned}
+\label{eq:sensmodel}
+\pi_t(X_t, U_t) &\sim \text{Beta} \left( \pi_t (X_t) (1/\alpha_t - 1), (1-\pi_t(X_t)) (1/\alpha_t - 1) \right) \\
+I_t | X_t, U_t &\sim \text{Bern} (\pi (X_t,U_t)) \\
+\rho_t(X_t, U_t) &= Q_t(X_t,1) + \delta_t \left( \logit \pi_t(X_t,U_t) - \EE \left[ \logit \pi_t(X_t,U_t) | X_t, I_t = 1 \right] \right) \\
+Y_t | I_t, X_t, U_t &\sim \text{Bern}(\rho_t(X_t, U_t))
+\end{aligned}
+\end{equation}
+The time-varying sensitivy parameter~$\alpha_t \in (0,1)$  controls the influence of the unobserved confounder~$U_t$ on selection propensity.
+In particular, $\alpha_t$ measure the change in belief of how likely an individual self-selects into the non-probabilistic sample:
+$$
+\alpha_t = E[ \pi_t (X_t,U_t) | I = 1] - E[ \pi_t (X_t,U_t) | I = 0].
+$$
+The sensitivity model satisfies the requirement that the self-selection propensity and conditional expectated outcome model can match the observed data:
+\begin{align*}
+P(I_t = 1 | X_t ) &= E[ E[ I_t | X_t, U_t ] | X_t ] = E[ \pi_t(X_t, U_t) | X_t ] = \pi_t (X_t) \\
+E[Y_t = 1 | X_t, I_t = 1 ] &= E[ E[ Y_t | X_t, U_t, I_t =1 ] | X_t, I_t = 1 ] = Q(1,X) \\
+\end{align*}
+The sensitivity model is based on XXXX, and is adapted here to the non-probabilistic survey setting with potential time-varying confounding and a binary outcome.
+
+\subsection{Bias}
+
+By assumption, $X_t$ and $U_t$ together lead to consistent estimation via the IPW estimator:
+\begin{align*}
+&\EE\left[\pi^{-1}_t (X_{J,t}, U_{J,t}) I_{J,t} Y_{J,t} \right] \\
+= &\EE[ E[ Y_{J,t} | X_{J,t}, U_{J,t}] ] \\
+= &\EE \left[ Q(1,X_{J,t}) + \delta \left( \logit \pi_t (X_{J,t}, U_{J,t}) - \EE \left[ \logit \pi_t (X_{J,t}, U_{J,t}) | X_{J,t}, I_{J,t} = 1 \right] \right) \right]
+\end{align*}
+
+\noindent {\bf IPW estimator:} Investigating the IPW estimator, we have:
+\begin{align*}
+&\EE\left[\pi^{-1}_t (X_{J,t}) I_{J,t} Y_{J,t} \right] \\
+=&\EE\left[\frac{\pi_t (X_{J,t}, U_{J,t})}{\pi_t (X_{J,t})} \pi^{-1} (X_{J,t}, U_{J,t}) I_{J,t} Y_{J,t} \right] \\
+= &\EE\left[ \frac{\pi_t (X_{J,t}, U_{J,t})}{\pi_t (X_{J,t})} \EE[ Y_{J,t} | X_{J,t}, U_{J,t}] \right].
+\end{align*}
+Using the fact that $\EE \left[ \pi_t (X_{J,t}, U_{J,t}) | X_{J,t} \right] = \pi_{t} (X_{J,t})$, we can see that the bias is given by
+$$
+\delta \EE \left[ \logit \pi_t (X_t, U_t) - \frac{\pi_t (X_{J,t}, U_{J,t})}{\pi_t (X_{J,t})} \logit \pi_t (X_t, U_t) \right].
+$$
+Using the fact that~$Z \sim Beta(\alpha, \beta)$ then $E[ ln(Z) ] = \psi(\alpha) - \psi(\alpha + \beta)$, $E[ ln(1-Z) ] = \psi(\beta) - \psi(\alpha + \beta)$ then~$E[ \logit (Z) ] = \psi(\alpha) - \psi(\beta)$ where~$\psi$ is the digamma function.
+Moreover,~$\EE[ Z ln Z ] = \frac{\alpha}{\alpha + \beta} \left[ \psi(\alpha + 1) - \psi(\alpha + \beta +1) \right]$ and
+\begin{align*}
+\EE \left[ (1-Z) ln(1-Z) \right] &= \frac{\beta}{\alpha + \beta} \left[ \psi(\beta + 1) - \psi(\alpha + \beta +1) \right] \\
+\EE \left[ ln(1-Z) \right] &= \left[ \psi(\beta) - \psi(\alpha + \beta) \right] \\
+\Rightarrow - \EE \left[ Z ln(1-Z) \right] &= \frac{\beta}{\alpha + \beta} \left[ \psi(\beta + 1) - \psi(\alpha + \beta +1) \right] - \left[ \psi(\beta) - \psi(\alpha + \beta) \right].
+\end{align*}
+and thus
+\begin{align*}
+\EE \left[ Z \logit(Z) \right] =& \frac{\alpha}{\alpha + \beta} \left[ \psi(\alpha + 1) - \psi(\alpha + \beta +1) \right] + \frac{\beta}{\alpha + \beta} \left[ \psi(\beta + 1) - \psi(\alpha + \beta +1) \right] \\
+&- \left[ \psi(\beta) - \psi(\alpha + \beta) \right] \\
+=& \psi(\alpha + \beta) - \psi(\alpha + \beta -1) + \frac{\alpha}{\alpha + \beta} \left[ \psi(\alpha + 1) - \psi(\beta+1) \right] + \frac{1}{\beta} \\
+=& \frac{\alpha}{\alpha + \beta} \left[ \psi(\alpha + 1) - \psi(\beta+1) \right] + \left[ \frac{1}{\beta} - \frac{1}{\alpha + \beta} \right].
+\end{align*}
+Plugging in $\alpha = \pi_t (X_t) (1/\alpha_t -1)$ and $\beta = (1-\pi_t (X_t))(1/\alpha_t - 1)$ yields:
+$$
+\pi_t (X_t) \left[ \psi \left(\pi_t (X_t) \left(\frac{1-\alpha_t}{\alpha_t}\right) + 1\right) - \psi \left((1-\pi_t (X_t)) \left(\frac{1-\alpha_t}{\alpha_t}\right)+1\right) \right] + \frac{\alpha_t}{1-\alpha_t}\frac{\pi_t(X_t)}{1-\pi_t(X_t)}.
+$$
+Since $\pi_t(X_t)$ cancels with the denominator term, the first term will match the form of the $E[ \logit (Z)]$ and therefore using the fact that $\psi(x+1) = \psi(x) + 1/x$, we can write the bias simply as
+$$
+\delta \EE \left[\frac{\alpha_t}{1-\alpha_t}\left( \frac{2}{1-\pi_t(X_t)} - \frac{1}{\pi_t(X_t)} \right) \right].
+$$
+Note that this is an expectation over the population-level distribution of the covariate~$X$.  Thus, computation requires use of the
+
+\noindent {\bf Model-based estimator:} Investigating the model-based estimator, we have:
+$$
+\EE\left[ E[ Y_{J,t} | X_{J,t}, I_{J,t} = 1] \right] = \EE\left[ Q(X_{J,t}, 1) \right]
+$$
+and so the bias is given by
+$$
+\delta \EE \left[ \left( \logit \pi_t (X_{J,t}, U_{J,t}) - \EE \left[ \logit \pi_t (X_{J,t}, U_{J,t}) | X_{J,t}, I_{J,t} = 1 \right] \right) \right]
+$$
+Using the above argument and the Beta-Bernoulli conjugacy we can re-write the bias as
+\begin{align*}
+\delta \EE \bigg[ &\psi \left( \frac{1-\alpha_t}{\alpha_t} \pi_t(X_t) \right) - \psi \left( \frac{1-\alpha_t}{\alpha_t} \left( 1 - \pi_t(X_t) \right) \right) \\
+&- \EE \bigg[ \psi \left( \frac{1-\alpha_t}{\alpha_t} \pi_t(X_t) + 1 \right) - \psi \left(\frac{1-\alpha_t}{\alpha_t} \left( 1 - \pi_t(X_t) \right) \right) \big | X_t, I_t = 1 \bigg] \bigg]
+\end{align*}
+Note that the 2nd expectation is over the distribution of $X$ conditional on response~$I=1$ while the first is not conditional.
+Therefore, the first expectation requires use of the random sample for computation while the second expectation requires use of the non-probabilistic sample.
+If the distribution of $X$ were independent of $I$ then the bias would simplify to
+$$
+-\delta_t \frac{1-\alpha_t}{\alpha_t} \EE \left[ \frac{1}{\pi_t (X_t)} \right].
+$$
+So we see that the bias is equal to the 2nd term in the bias under the IPW estimator.
+
+\subsection{Reparametrization}
+
+Following XX and XX, we reexpress the outcome-confounder strength in terms of the partial coefficient of determination
+$$
+R_Y^2 (\alpha, \delta) = \frac{\EE [( Y - Q(1,X) )^2 | I=1 ] - \EE[ (Y - \EE[ Y | X, U ])^2]}{E[(Y - Q(1,X))^2 | I = 1]}
+$$
+in terms of $\delta^2$.  Note the  To do so, write
+\begin{align*}
+\EE [ (Y - \EE[ Y | X, U] )^2 | I = 1 ] =& \EE [ (Y - Q(1,X))^2 | I = 1 ] \\
+&- 2 \delta \EE \left[ (Y - Q(1,X)) (\logit \pi(X,U) - \EE[ \logit \pi (X,U) | X, I=1 ]) | I = 1 \right] \\
+&+ \delta^2 \EE \left[ (\logit \pi(X,U) - \EE[ \logit \pi (X,U) | X, I=1 ])^2 | I = 1 \right] \\
+&= \EE [ (Y - Q(1,X))^2 | I = 1 ] - \delta^2 \EE \left[ \text{var} \left( \logit \pi(X,U) \right) | X, I = 1 \right]
+\end{align*}
+By Beta-Bernoulli conjugacy, the second term is the variance of the logit-transformed Beta distribution which has an analytic expression:
+$$
+\text{var} \left( \logit \pi(X,U) | X, I = 1 \right) = \psi_1 ( \pi(X) (\alpha - 1) + 1 ) + \psi_1 ( (1- \pi(X)) (1/\alpha - 1))
+$$
+where $\psi_1$ is the trigamma function.
+
+By Theorem 5 in XXX, under the sensitivity model
+$$
+\alpha = 1 - \frac{\EE [ ]}{\EE[ ] }
+$$
+
+\subsection{Estimating sensitivity parameters}
+
+Based on equations XX and XX, we can estimate the time-varying sensitivity parameters.
+
+With these estimates
+
 \bibliographystyle{plainnat}

### 2022-09-12 write-up/letters/response-ref-1-v1.tex

--- 2022-03-30 write-up/letters/response-ref-1-v1.tex
@@
 Any comments in {\it italics} indicate a statement from the referee report.
-
-{\bf Response to Editor and Associate Editor}
 \begin{enumerate}
-\item {\it The referees have many specific concerns they are asking to address, and their expertise and viewpoints are important to revise this manuscript to improve its clarity and impact.}
+\item {\it I appreciate the addition of the remark that states that the dynamics mentioned in this paper refer to the original strain.
+There are still lines (for example: “Upon infection with SARS-CoV-2, an incubation period (time to symptom onset) starts and lasts approximately five days”) that are more definitive than I would prefer.
+Perhaps adding in “Upon infection with the original SARS-CoV-2 variant” would be preferable.}
 
 \vspace{5mm}
-We thank the editor-in-chief.  We have addressed the two reviewers' specific comments.  Responses to their concerns are detailed below and have significantly improved the clarify of this manuscript.
+We agree with this point and have gone through the manuscript to reduce the chances a reader miscontrues our explanations.
+We aim for clarity in these edits by avoiding definitive statements while not making statements that are too vauge or ambiguous.
 \vspace{5mm}
 
-\item {\it Some of these deal with the fast-evolving nature of the pandemic with new information emerging every week and changing variants with different characteristics also changing our understanding of the disease course and spread and timing, that affect the content in the paper.
-I understand that it is not feasible for a peer-reviewed paper like this to include all the latest information in a fast-changing environment like this in which understanding can change quickly, but it is important in this revision to update the KEY points as much as possible based on updated knowledge, but otherwise revise the language to make it “forward compatible” in terms of relevance as things continue to change, e.g.
-by acknowledging some details of spread dynamics may change with future variants, but presenting the current work based on a particular SARS-CoV-2 variant as illustrative of the issues that arise with each new variant and other viruses, as well.
-That is, recognizing it may not be feasible to keep updating based on emerging knowledge from each new strain, but to lock down the discussion context but add enough explanation so the reader can see its relevance for new strains.}
+\item {\it The reliance on the random sample with such a low response rate may be subject to unmeasured confounding - perhaps a sensitivity analysis showing the impact of a potential unmeasured confounder could strengthen this point.}
 
 \vspace{5mm}
-Remark 1 has been added at the end of the introduction to contextualize the manuscript.
-Specifically we clarify that our manuscript focuses on COVID-19 case count, testing, and death data collected from April 2020 through February 2021 and therefore the numbers presented on disease dynamics are based on the original strain.
-While using the framework to account for selection bias and measurement error for data arising from the delta and omicron strains would require additional work, the framework would remain an appropriate tool for this task.
+We have added a sensitivity analysis to the supplementary materials where we consider an unmeasured binary confounder~$Z_{j,t} \in \{0,1\}$.
+In this setting, {\it prevalence}~$pr(Z=1|X,Y)$ and {\it propensity} $\pi(X, Z; (\theta,\theta_z))$ both impact the sensitivity analysis.
+Imputation-based sensitivity analyses will thus depend heavily on the choice of prevalence and $\theta_z$.
+Therefore, we consider a Taylor series expansion of the IPW and DR-estimators at the estimated propesity which is equivalent to~$\theta_z = 0$.denote the coefficient in the selection propensity model and
+$$
+\hat \mu_{IPW} + \theta_z \frac{1}{\sum_{j=1}^N I^{NR}_j w_j (X_{j,t}; \hat \theta_t)} \left[ \sum_{j=1}^N I^{NR}_j \frac{1-\pi_j(X_{j,t}; \hat \theta_t)}{\pi_j(X_{j,t}; \hat \theta_t)} E[Z_j] (Y_j - \hat \mu_{IPW}) \right]
+$$
+Thus bias depends on the correlation of the error $Y_j - \hat \mu_{IPW}$ to the odds ratio times the expected value of the unmeasured binary confounder.  We use this to calculate the
+
 \vspace{5mm}
 
-\item {\it The AE also mentions a few comments as well:}
-\begin{enumerate}
-	\item {\it Title: As noted by one of the reviewers, the “paradox” is never discussed/defined in the manuscript.
-Consider changing to “Statistical estimation of Coronavirus Case Counts: Measurement…” or some alternative.}
+\item {\it On the plots that show the active infection rate estimates, perhaps including a line that is the case counts divided by the population as a floor would be worthwhile (at least in the supplement) to demonstrate that the model-based estimates are at least above this.}
+
+\vspace{5mm}
+We have added this this floor .  We clarify that ``case counts'' in the comment above should be a measure of recent case counts to be a relevant  active infection rate
 \vspace{5mm}
 
-	We agree with the AE.
-This was not updated in the prior submission simply to ensure continuity of presentation (i.e., didn’t want to change from first submission).
-Here we replace with the title ``Addressing selection bias and measurement error in COVID-19 case count data using auxiliary information.''
+\item {\it I found the grey hard to differentiate for the figures that have more than two categories – perhaps varying the shapes would help if trying to keep these grey-scale?}
+
 \vspace{5mm}
-
-	\item {\it Organization: The figures and tables are not located anywhere near much of their introduction/discussion.
-For example, Table 1 was discussed on page 7 and is itself located on page 22.
-Consider the reorganization such that Figures and Tables are located near their discussion.}
-	\vspace{5mm}
-
-	We apologize.
-We kept figures/tables related to the case study in Section 5 (the case study section) but can now see this led to confusion as we discuss them early in the paper.
-Tables and figures have been moved to be located much closer to their introduction.
+We have added shapes to help differentiate figures while keeping the grey-scale.
 \vspace{5mm}

### 2022-09-12 write-up/covid-supplement-writeup.tex
Update covid-supplement-writeup.tex

@@
 The proposed method relies on the assumption $I_{j,t}^{(NR)} \indep Y_{j,t} \, | \, X_{j,t}$, i.e., the sampling indicator in the non-probabilistic survey is conditionally independent of the outcome given the covariates.
-However, estimates of population quantities based on inverse-weighting will be biased in the presence of `unobserved confounding'.
-We may hope that the bias
-
-The purpose of this section is to introduce XXX,
-
-We consider a
+However, estimates of population quantities based on inverse-weighting and/or model-based estimates will be biased in the presence of `unobserved confounding'.
+The purpose of this section is to introduce a sensitivity analysis.
+To this, we consider the logit-linear model by XXXX adjusted to self-selection rather than causal inference.
+That is,
 \begin{align*}
 \logit \pi \left( I_{t} = 1  | x_{t}, u_{t} \right) &= h_t(x_t) + \alpha u \\
-E[ Y | i_t, x_t, u_t] = l(i_t, x_t)
+E[ Y | i_t, x_t, u_t] &= l(i_t, x_t) + \delta u
 \end{align*}

### 2022-09-12 write-up/covid-supplement-writeup.tex
updated supplement

@@
 \def\EE{\mathbb{E}}
-\newcommand{\indep}{\perp \!\!\! \perp}
+%\newcommand{\indep}{\perp \!\!\! \perp}
 
@@
 
-The proposed method relies on the assumption $I_{j,t}^{(NR)} \indep Y_{j,t} \, | \, X_{j,t}$, i.e., the sampling indicator in the non-probabilistic survey is conditionally independent of the outcome given the covariates.
+The proposed approcah relies on the assumption $I_{j,t}^{(NR)} \indep Y_{j,t} \, | \, X_{j,t}$, i.e., the sampling indicator in the non-probabilistic survey is conditionally independent of the outcome given the covariates.
 However, estimates of population quantities based on inverse-weighting and/or model-based estimates will be biased in the presence of `unobserved confounding'.
 The purpose of this section is to introduce a sensitivity analysis.
-To this, we consider the logit-linear model by XXXX adjusted to self-selection rather than causal inference.
-That is,
+To this, we consider the logit-linear model by~\cite{Imbens2003} adjusted to self-selection in a time-varying setting rather than causal inference in a single time-point setting.
+That is, consider the model
 \begin{align*}
-\logit \pi \left( I_{t} = 1  | x_{t}, u_{t} \right) &= h_t(x_t) + \alpha u \\
-E[ Y | i_t, x_t, u_t] &= l(i_t, x_t) + \delta u
+\logit \pi \left( I_{t} = 1  | x_{t}, u_{t} \right) &= h_t(x_t) + \alpha_t u_t \\
+E[ Y_t | i_t, x_t, u_t] &= l_t(i_t, x_t) + \delta_t u_t
 \end{align*}
-for some function~$h_t$, $l_t$ that are allowed to depend on time. Re-arranging the terms we see that .  This can be
-$$
-E[ Y | i_t x_t, u_t ] = l(t,x)  + \frac{\delta}{\alpha} \left( \logit \pi \left( I_{t} = 1  | x_{t}, u_{t} \right) - h_t (x_t) \right)
-$$
-The key idea in XXXX is that positing a distribution on $P(I_t = 1 | x_y, u_t)$ directly allows one to circumvent the need to specify a distribution for~$U_t$, a highly non-trivial task.
-
-The logit-linear model does not lead to a tractable sensitivy analysis.  Instead, we define a sensitivity model as:
+for some functions~$h_t$ and $l_t$ that are allowed to depend on time. Re-arranging the terms we see that
+$$
+E[ Y | i_t x_t, u_t ] = l_t(i_t,x)  + \frac{\delta_t}{\alpha_t} \left( \logit \pi \left( I_{t} = 1  | x_{t}, u_{t} \right) - h_t (x_t) \right)
+$$
+The key idea in~\cite{Imbens2003} is that positing a distribution on $P(I_t = 1 | x_y, u_t)$ directly allows one to circumvent the need to specify a distribution for~$U_t$, a highly non-trivial task.
+
+The logit-linear model above unfortunately does not lead to a tractable sensitivity analysis.
+Instead, we consider a sensitivity model based on work by~\cite{Veitch2020}, again adjusted to deal with the current setting of self-selection, given by:
 \begin{equation}
@@
 \end{equation}
-The time-varying sensitivy parameter~$\alpha_t \in (0,1)$  controls the influence of the unobserved confounder~$U_t$ on selection propensity.
-In particular, $\alpha_t$ measure the change in belief of how likely an individual self-selects into the non-probabilistic sample:
+The time-varying sensitivity parameter~$\alpha_t \in (0,1)$  controls the influence of the unobserved confounder~$U_t$ on selection propensity.
+In particular, $\alpha_t$ measures the change in belief of how likely an individual self-selects into the non-probabilistic sample at time~$t$:
 $$
@@
 P(I_t = 1 | X_t ) &= E[ E[ I_t | X_t, U_t ] | X_t ] = E[ \pi_t(X_t, U_t) | X_t ] = \pi_t (X_t) \\
-E[Y_t = 1 | X_t, I_t = 1 ] &= E[ E[ Y_t | X_t, U_t, I_t =1 ] | X_t, I_t = 1 ] = Q(1,X) \\
+E[Y_t = 1 | X_t, I_t = 1 ] &= E[ E[ Y_t | X_t, U_t, I_t =1 ] | X_t, I_t = 1 ] = Q(1,X).
 \end{align*}
-The sensitivity model is based on XXXX, and is adapted here to the non-probabilistic survey setting with potential time-varying confounding and a binary outcome.
-
-\subsection{Bias}
-
-By assumption, $X_t$ and $U_t$ together lead to consistent estimation via the IPW estimator:
+Moreover, the model satisfies that $Y_t$ does not depend on $I_t$ given $(X_t, U_t)$.
+This is a key adaptation to the non-probabilistic survey setting with potential time-varying confounding and a binary outcome.
+
+\noindent {\bf IPW estimator:} By assumption, observing $X_t$ and $U_t$ together leads to consistent estimation via the estimator:
 \begin{align*}
-&\EE\left[\pi^{-1}_t (X_{J,t}, U_{J,t}) I_{J,t} Y_{J,t} \right] \\
+\EE\left[\pi^{-1}_t (X_{J,t}, U_{J,t}) I_{J,t} Y_{J,t} \right]
 = &\EE[ E[ Y_{J,t} | X_{J,t}, U_{J,t}] ] \\
@@
 \end{align*}
-
-\noindent {\bf IPW estimator:} Investigating the IPW estimator, we have:
+\noindent {\bf IPW-based estimator:} Investigating the IPW estimator using only~$X_t$, we have:
 \begin{align*}
-&\EE\left[\pi^{-1}_t (X_{J,t}) I_{J,t} Y_{J,t} \right] \\
+\EE\left[\pi^{-1}_t (X_{J,t}) I_{J,t} Y_{J,t} \right]
 =&\EE\left[\frac{\pi_t (X_{J,t}, U_{J,t})}{\pi_t (X_{J,t})} \pi^{-1} (X_{J,t}, U_{J,t}) I_{J,t} Y_{J,t} \right] \\
@@
 \end{align*}
-Using the fact that $\EE \left[ \pi_t (X_{J,t}, U_{J,t}) | X_{J,t} \right] = \pi_{t} (X_{J,t})$, we can see that the bias is given by
-$$
-\delta \EE \left[ \logit \pi_t (X_t, U_t) - \frac{\pi_t (X_{J,t}, U_{J,t})}{\pi_t (X_{J,t})} \logit \pi_t (X_t, U_t) \right].
+Using the fact that $\EE \left[ \pi_t (X_{J,t}, U_{J,t}) | X_{J,t} \right] = \pi_{t} (X_{J,t})$, the \emph{bias} is
+$$
+\delta \EE \left[ \logit \pi_t (X_{t,J}, U_{t,J}) - \frac{\pi_t (X_{J,t}, U_{J,t})}{\pi_t (X_{J,t})} \logit \pi_t (X_{t,J}, U_{t,J}) \right].
 $$
@@
 $$
-\pi_t (X_t) \left[ \psi \left(\pi_t (X_t) \left(\frac{1-\alpha_t}{\alpha_t}\right) + 1\right) - \psi \left((1-\pi_t (X_t)) \left(\frac{1-\alpha_t}{\alpha_t}\right)+1\right) \right] + \frac{\alpha_t}{1-\alpha_t}\frac{\pi_t(X_t)}{1-\pi_t(X_t)}.
+\pi_t (X_{t,J}) \left[ \psi \left(\pi_t (X_{t,J}) \left(\frac{1-\alpha_t}{\alpha_t}\right) + 1\right) - \psi \left((1-\pi_t (X_{t,J})) \left(\frac{1-\alpha_t}{\alpha_t}\right)+1\right) \right] + \frac{\alpha_t}{1-\alpha_t}\frac{\pi_t(X_{t,J})}{1-\pi_t(X_{t,J})}.
 $$
@@
 $$
-\delta \EE \left[\frac{\alpha_t}{1-\alpha_t}\left( \frac{2}{1-\pi_t(X_t)} - \frac{1}{\pi_t(X_t)} \right) \right].
-$$
-Note that this is an expectation over the population-level distribution of the covariate~$X$.  Thus, computation requires use of the
-
-\noindent {\bf Model-based estimator:} Investigating the model-based estimator, we have:
-$$
-\EE\left[ E[ Y_{J,t} | X_{J,t}, I_{J,t} = 1] \right] = \EE\left[ Q(X_{J,t}, 1) \right]
-$$
-and so the bias is given by
-$$
-\delta \EE \left[ \left( \logit \pi_t (X_{J,t}, U_{J,t}) - \EE \left[ \logit \pi_t (X_{J,t}, U_{J,t}) | X_{J,t}, I_{J,t} = 1 \right] \right) \right]
-$$
-Using the above argument and the Beta-Bernoulli conjugacy we can re-write the bias as
-\begin{align*}
-\delta \EE \bigg[ &\psi \left( \frac{1-\alpha_t}{\alpha_t} \pi_t(X_t) \right) - \psi \left( \frac{1-\alpha_t}{\alpha_t} \left( 1 - \pi_t(X_t) \right) \right) \\
-&- \EE \bigg[ \psi \left( \frac{1-\alpha_t}{\alpha_t} \pi_t(X_t) + 1 \right) - \psi \left(\frac{1-\alpha_t}{\alpha_t} \left( 1 - \pi_t(X_t) \right) \right) \big | X_t, I_t = 1 \bigg] \bigg]
-\end{align*}
-Note that the 2nd expectation is over the distribution of $X$ conditional on response~$I=1$ while the first is not conditional.
+-\delta_t \frac{\alpha_t}{1-\alpha_t} \EE \left[ \frac{1}{\pi_t(X_{t,J})} \right]
+$$
+Note that this is an expectation over the population-level distribution of the covariate~$X$.  Thus, computation requires use of the probabilistic sample in order to estimate the potential bias.
+
+% \noindent {\bf Model-based estimator:} Investigating the model-based estimator, we have:
+% $$
+% \EE\left[ E[ Y_{J,t} | X_{J,t}, I_{J,t} = 1] \right] = \EE\left[ Q(X_{J,t}, 1) \right]
+% $$
+% and so the bias is given by
+% $$
+% \delta \EE \left[ \left( \logit \pi_t (X_{J,t}, U_{J,t}) - \EE \left[ \logit \pi_t (X_{J,t}, U_{J,t}) | X_{J,t}, I_{J,t} = 1 \right] \right) \right]
+% $$
+% Using the above argument and the Beta-Bernoulli conjugacy we can re-write the bias as
+% \begin{align*}
+% \delta \EE \bigg[ &\psi \left( \frac{1-\alpha_t}{\alpha_t} \pi_t(X_{J,t}) \right) - \psi \left( \frac{1-\alpha_t}{\alpha_t} \left( 1 - \pi_t(X_{J,t}) \right) \right) \\
+% &- \EE \bigg[ \psi \left( \frac{1-\alpha_t}{\alpha_t} \pi_t(X_{J,t}) + 1 \right) - \psi \left(\frac{1-\alpha_t}{\alpha_t} \left( 1 - \pi_t(X_{J,t}) \right) \right) \big | X_{J,t}, I_{J,t} = 1 \bigg] \bigg]
+% \end{align*}
+% Note that the 2nd expectation is over the distribution of $X$ conditional on response~$I=1$ while the first is not conditional.
 Therefore, the first expectation requires use of the random sample for computation while the second expectation requires use of the non-probabilistic sample.
 If the distribution of $X$ were independent of $I$ then the bias would simplify to
-$$
--\delta_t \frac{1-\alpha_t}{\alpha_t} \EE \left[ \frac{1}{\pi_t (X_t)} \right].
-$$
-So we see that the bias is equal to the 2nd term in the bias under the IPW estimator.
+% $$
+% -\delta_t \frac{1-\alpha_t}{\alpha_t} \EE \left[ \frac{1}{\pi_t (X_t)} \right].
+% $$
+% which is exactly equal to the IPW estimator bias.  In settings where $X$ strongly depends on $I$ this is not the case.
 
@@
 
-Following XX and XX, we reexpress the outcome-confounder strength in terms of the partial coefficient of determination
+Following~\cite{Veitch2020}, we re-express the outcome-confounder strength in terms of the partial coefficient of determination
 $$
@@
 $$
-in terms of $\delta^2$.  Note the  To do so, write
-\begin{align*}
+in terms of $\delta^2$.  To do so, writ
+em\begin{align*}
 \EE [ (Y - \EE[ Y | X, U] )^2 | I = 1 ] =& \EE [ (Y - Q(1,X))^2 | I = 1 ] \\
@@
 $$
-where $\psi_1$ is the trigamma function.
-
+where $\psi_1$ is the trigamma function. This implies the outcome influence is
+$$
+R_Y^2 (\alpha, \delta) = \delta^2 \frac{\EE [\psi_1 ( \pi(X) (\alpha - 1) + 1 ) + \psi_1 ( (1- \pi(X)) (1/\alpha - 1))] }{E [(Y- Q(1,X))^2 | I = 1]}
+$$
+Note this expectation is over those who
 By Theorem 5 in XXX, under the sensitivity model

### 2022-09-13 write-up/covid-supplement-writeup.tex
Updated the appendix, almost there

@@
 
-The proposed approcah relies on the assumption $I_{j,t}^{(NR)} \indep Y_{j,t} \, | \, X_{j,t}$, i.e., the sampling indicator in the non-probabilistic survey is conditionally independent of the outcome given the covariates.
+The proposed approach relies on the assumption $I_{j,t}^{(NR)} \indep Y_{j,t} \, | \, X_{j,t}$, i.e., the sampling indicator in the non-probabilistic survey is conditionally independent of the outcome given the covariates.
 However, estimates of population quantities based on inverse-weighting and/or model-based estimates will be biased in the presence of `unobserved confounding'.
-The purpose of this section is to introduce a sensitivity analysis.
-To this, we consider the logit-linear model by~\cite{Imbens2003} adjusted to self-selection in a time-varying setting rather than causal inference in a single time-point setting.
+The purpose of this section is to introduce a sensitivity analysis to address this concern.
+To this end, we consider the logit-linear model by~\cite{Imbens2003}, adjusted to the time-varying, self-selection setting from the single time-point, causal inference setting.
 That is, consider the model
@@
 $$
-E[ Y | i_t x_t, u_t ] = l_t(i_t,x)  + \frac{\delta_t}{\alpha_t} \left( \logit \pi \left( I_{t} = 1  | x_{t}, u_{t} \right) - h_t (x_t) \right)
-$$
-The key idea in~\cite{Imbens2003} is that positing a distribution on $P(I_t = 1 | x_y, u_t)$ directly allows one to circumvent the need to specify a distribution for~$U_t$, a highly non-trivial task.
+E[ Y_t | i_t x_t, u_t ] = l_t(i_t,x)  + \frac{\delta_t}{\alpha_t} \left( \logit \pi \left( I_{t} = 1  | x_{t}, u_{t} \right) - h_t (x_t) \right)
+$$
+The key idea in~\cite{Imbens2003} is that positing a distribution on $\pi(I_t = 1 | x_y, u_t)$ directly allows one to circumvent the need to specify a distribution for~$U_t$, a highly non-trivial task.
 
@@
 P(I_t = 1 | X_t ) &= E[ E[ I_t | X_t, U_t ] | X_t ] = E[ \pi_t(X_t, U_t) | X_t ] = \pi_t (X_t) \\
-E[Y_t = 1 | X_t, I_t = 1 ] &= E[ E[ Y_t | X_t, U_t, I_t =1 ] | X_t, I_t = 1 ] = Q(1,X).
+E[Y_t = 1 | X_t, I_t = 1 ] &= E[ E[ Y_t | X_t, U_t, I_t =1 ] | X_t, I_t = 1 ] =: Q_t(1,X_t).
 \end{align*}
@@
 &- \left[ \psi(\beta) - \psi(\alpha + \beta) \right] \\
-=& \psi(\alpha + \beta) - \psi(\alpha + \beta -1) + \frac{\alpha}{\alpha + \beta} \left[ \psi(\alpha + 1) - \psi(\beta+1) \right] + \frac{1}{\beta} \\
+=& \psi(\alpha + \beta) - \psi(\alpha + \beta +1) + \frac{\alpha}{\alpha + \beta} \left[ \psi(\alpha + 1) - \psi(\beta+1) \right] + \frac{1}{\beta} \\
 =& \frac{\alpha}{\alpha + \beta} \left[ \psi(\alpha + 1) - \psi(\beta+1) \right] + \left[ \frac{1}{\beta} - \frac{1}{\alpha + \beta} \right].
@@
 Note this expectation is over those who
-By Theorem 5 in XXX, under the sensitivity model
-$$
-\alpha = 1 - \frac{\EE [ ]}{\EE[ ] }
+
+Following from~\cite[Theorem 5]{Veitch2020}, the parameter~$\alpha$ can be reexpressed in a more convenient form:
+$$
+\alpha = 1 - \frac{\EE \left[ \pi (X_{t,J},U_{t,J}) \left( 1 - \pi (X_{t,J},U_{t,J}) \right)  \right]}{\EE\left[ \pi (X_{t,J}) \left( 1 - \pi (X_{t,J}) \right) \right] }
 $$

### 2022-09-14 write-up/covid-supplement-writeup.tex
Update covid-supplement-writeup.tex

@@
 $$
-The sensitivity model satisfies the requirement that the self-selection propensity and conditional expectated outcome model can match the observed data:
+The sensitivity model satisfies the requirement that the self-selection propensity and conditional expectation outcome model can match the observed data:
 \begin{align*}

### 2022-09-14 write-up/covid-supplement-writeup.tex
updated supplement - now need to fit models for austen plots

@@
 However, estimates of population quantities based on inverse-weighting and/or model-based estimates will be biased in the presence of `unobserved confounding'.
-The purpose of this section is to introduce a sensitivity analysis to address this concern.
+Here we introduce a sensitivity analysis to address this concern.
 To this end, we consider the logit-linear model by~\cite{Imbens2003}, adjusted to the time-varying, self-selection setting from the single time-point, causal inference setting.
@@
 \end{align*}
-for some functions~$h_t$ and $l_t$ that are allowed to depend on time. Re-arranging the terms we see that
+for functions~$h_t$ and $l_t$ that depend on time. Re-arranging terms we have
 $$
@@
 P(I_t = 1 | X_t ) &= E[ E[ I_t | X_t, U_t ] | X_t ] = E[ \pi_t(X_t, U_t) | X_t ] = \pi_t (X_t) \\
-E[Y_t = 1 | X_t, I_t = 1 ] &= E[ E[ Y_t | X_t, U_t, I_t =1 ] | X_t, I_t = 1 ] =: Q_t(1,X_t).
+E[Y_t | X_t, I_t = 1 ] &= E[ E[ Y_t | X_t, U_t, I_t =1 ] | X_t, I_t = 1 ] =: Q_t(1,X_t).
 \end{align*}
@@
 $$
-Note that this is an expectation over the population-level distribution of the covariate~$X$.  Thus, computation requires use of the probabilistic sample in order to estimate the potential bias.
+Note that this is an expectation over the population.  Thus, estimation requires use of the probabilistic sample in order to estimate the potential bias.
+
+% In the main manuscript, we consider a stabilized version of the IPW estimator which can be expressed as
+% $$
+% \frac{E[ I_{J,t} \pi^{-1}_{j,t} (X_{J,t}) Y_{J,t}]}{E[ I_{J,t} \pi^{-1}_{j,t} (X_{J,t})]}.
+% $$
+% The denominator satisfies
+% \begin{align*}
+% E[ I_{J,t} \pi^{-1}_{j,t} (X_{J,t})] &= E \left[ E[ I_{J,t} | X_{J,t}, U_{J,t}] \pi^{-1}_{j,t} (X_{J,t}) \right] \\
+% &= E \left[ \frac{E[ \pi_{j,t} (X_{J,t}, U_{J,t}) | X_{J,t}]}{\pi_{j,t} (X_{J,t})} \right] \\
+% &= E \left[ \frac{\pi_{j,t} (X_{J,t})}{\pi_{j,t} (X_{J,t})} \right] = 1.
+% \end{align*}
+% Therefore both versions under~\eqref{eq:sensmodel} have the same bias calculation for large $N$.
 
@@
 $$
-R_Y^2 (\alpha, \delta) = \frac{\EE [( Y - Q(1,X) )^2 | I=1 ] - \EE[ (Y - \EE[ Y | X, U ])^2]}{E[(Y - Q(1,X))^2 | I = 1]}
-$$
-in terms of $\delta^2$.  To do so, writ
-em\begin{align*}
-\EE [ (Y - \EE[ Y | X, U] )^2 | I = 1 ] =& \EE [ (Y - Q(1,X))^2 | I = 1 ] \\
-&- 2 \delta \EE \left[ (Y - Q(1,X)) (\logit \pi(X,U) - \EE[ \logit \pi (X,U) | X, I=1 ]) | I = 1 \right] \\
-&+ \delta^2 \EE \left[ (\logit \pi(X,U) - \EE[ \logit \pi (X,U) | X, I=1 ])^2 | I = 1 \right] \\
-&= \EE [ (Y - Q(1,X))^2 | I = 1 ] - \delta^2 \EE \left[ \text{var} \left( \logit \pi(X,U) \right) | X, I = 1 \right]
+R_t^2 (\alpha_t, \delta_t) = \frac{\EE [( Y_{J,t} - Q_t(1,X_{J,t}) )^2 | I_{J,t}=1 ] - \EE[ (Y_{J,t} - \EE[ Y_{J,t} | X_{J,t}, U_{J,t} ])^2 | I_{J,t} = 1]}{E[(Y_{J,t} - Q_t(1,X_{J,t}))^2 | I = 1]}
+$$
+in terms of $\delta_t^2$. To do so, write
+\begin{align*}
+&\EE [ (Y_{J,t} - \EE[ Y_{J,t} | X_{J,t}, U_{J,t}] )^2 | I_{J,t} = 1 ] \\
+= &\EE [ (Y_{J,t} - Q_t(1,X_{J,t}))^2 | I_{J,t} = 1 ] \\
+&- 2 \delta \EE \left[ (Y_{J,t} - Q_t(1,X_{J,t})) (\logit \pi(X_{J,t},U_{J,t}) - \EE[ \logit \pi (X_{J,t},U_{J,t}) | X_{J,t}, I_{J,t}=1 ]) | I_{J,t} = 1 \right] \\
+&+ \delta^2 \EE \left[ (\logit \pi_t(X_{J,t},U_{J,t}) - \EE[ \logit \pi_t (X_{J,t},U_{J,t}) | X_{J,t}, I_{J,t}=1 ])^2 | I_{J,t} = 1 \right] \\
+= &\EE [ (Y_{J,t} - Q_t(1,X_{J,t}))^2 | I_{J,t} = 1 ] - \delta^2 \EE \left[ \text{var} \left( \logit \pi_t(X_{J,t},U_{J,t}) \right) | X_{J,t}, I_{J,t} = 1 \right]
 \end{align*}
@@
 $$
-\text{var} \left( \logit \pi(X,U) | X, I = 1 \right) = \psi_1 ( \pi(X) (\alpha - 1) + 1 ) + \psi_1 ( (1- \pi(X)) (1/\alpha - 1))
-$$
-where $\psi_1$ is the trigamma function. This implies the outcome influence is
-$$
-R_Y^2 (\alpha, \delta) = \delta^2 \frac{\EE [\psi_1 ( \pi(X) (\alpha - 1) + 1 ) + \psi_1 ( (1- \pi(X)) (1/\alpha - 1))] }{E [(Y- Q(1,X))^2 | I = 1]}
-$$
-Note this expectation is over those who
-
-Following from~\cite[Theorem 5]{Veitch2020}, the parameter~$\alpha$ can be reexpressed in a more convenient form:
-$$
-\alpha = 1 - \frac{\EE \left[ \pi (X_{t,J},U_{t,J}) \left( 1 - \pi (X_{t,J},U_{t,J}) \right)  \right]}{\EE\left[ \pi (X_{t,J}) \left( 1 - \pi (X_{t,J}) \right) \right] }
-$$
-
-\subsection{Estimating sensitivity parameters}
-
-Based on equations XX and XX, we can estimate the time-varying sensitivity parameters.
-
-With these estimates
+\text{var} \left( \logit \pi_t(X_{J,t},U_{J,t}) | X_{J,t}, I_{J,t} = 1 \right) = \psi_1 ( \pi(X_{J,t}) (\alpha_t - 1) + 1 ) + \psi_1 ( (1- \pi(X_{J,t})) (1/\alpha_t - 1))
+$$
+where $\psi_1$ is the trigamma function. This implies the following relationship:
+\begin{equation}
+\label{eq:partialcoef}
+R_t^2 (\alpha_t, \delta_t) = \delta_t^2 \frac{\EE [\psi_1 ( \pi(X_{J,t}) (\alpha_t - 1) + 1 ) + \psi_1 ( (1- \pi(X_{J,t})) (1/\alpha_t - 1)) | I_{J,t} = 1 ] }{E [(Y_{J,t}- Q_t(1,X_{J,t}))^2 | I_{J,t} = 1]}
+\end{equation}
+Note that unlike the prior expecation, these are expressed conditional on self-selection, i.e., $I_{J,t} = 1$.
+This was done since the binary outcomes are only measured in the non-probabilistic sample.
+
+Following from~\cite[Theorem 5]{Veitch2020}, the parameter~$\alpha_t$ can be reexpressed as
+\begin{equation}
+\label{eq:alphaest}
+\alpha_t = 1 - \frac{\EE \left[ \pi (X_{t,J},U_{t,J}) \left( 1 - \pi (X_{t,J},U_{t,J}) \right)  \right]}{\EE\left[ \pi (X_{t,J}) \left( 1 - \pi (X_{t,J}) \right) \right] },
+\end{equation}
+which is a more convenient form when trying to estimate the parameter from observed data.
+
+\subsection{Calibration of sensitivity parameters}
+
+Based on~\eqref{eq:partialcoef} and~\eqref{eq:alphaest}, one can use the probability and nonprobability samples to calibrate the sensitivity parameters to be in line with the observed covariate influence on treatment and outcome.
+That is, for a given observed covariate~$Z_{J,t}$, we wish to measure the degree of influence it has on self-selection and outcome given the other observed covariates~$X_{J,t} \backslash Z_{J,t}$.
+Note that this dependence is likely time-varying, which is why it is important estimate~$\alpha_t$ and~$R_t^2$ as functions of time~$t$.
+
+For the outcome, we can measure the partial coefficient of determination as:
+$$
+R_{t, Z}^2 := \frac{\frac{1}{n} \sum_{i=1}^n (y_{i,t} - \hat Q_{t,Z} (1, x_{i,t} \backslash z_{i,t}) )^2 - \frac{1}{n} \sum_{i=1}^n (y_{i,t} - \hat Q_{t,Z} (1, x_{i,t}) )^2}{\frac{1}{n} \sum_{i=1}^n (y_{i,t} - \hat Q_{t,Z} (1, x_{i,t} \backslash z_{i,t}) )^2}.
+$$
+We can measure influence of $Z$ on self-selection propensity given $X_{J,t} \backslash Z_{J,t}$ by
+$$
+\hat \alpha_{t, Z} := 1 - \frac{\sum_{j=1}^N I_{j,t}^R W_{j,t}^R \cdot \hat \pi(X_{J,t}) (1- \hat \pi(X_{J,t}))}{\sum_{j=1}^N I_{j,t}^R W_{j,t}^R \cdot \hat \pi(X_{J,t} \backslash Z_{J,t}) (1- \hat \pi(X_{J,t} \backslash Z_{J,t}))}
+$$
+where~$\hat \pi_t (\cdot)$ is the fitted self-selection probability based on~\eqref{eq:auxinfoprob} in Section~\ref{subsec:auxprob} using the appropriately chosen set of observed covariates.
+
+\subsection{IPW Sensitivity Analysis of COVID-19 Active Infection Prevalence in Indiana}
+
+Here we perform a sensitivity analysis of the IPW estimator of COVID-19 active infection prevalence in indiana.
+While~\cite{Veitch2020} use Austen plots, we do not have such a simple visualization due to time-dependence.
+To handle this, for each observed covariate~$Z$, we plot $(\alpha_{t,Z}, R^2_{t,Z})$ in the $[0,1]^2$.
+These ``calibration curves'' help us in selection of a range of values for~$(\alpha_t, R^2_t)$ in our sensitivity analysis.
+Here, we consider $Z$ to be separately (a) age, (b) gender, (c) race, and (d) symptom status.
+The four plots are presented in Figure XX.
 

### 2022-09-15 write-up/letters/response-ref-1-v1.tex
updating sensitivity model and making all relevant changes for final referee comments

--- 2022-09-12 write-up/letters/response-ref-1-v1.tex
@@
 
-We appreciate the very helpful comments from the referee which has helped improve the paper substantially.
+We appreciate the comments from the referee which has helped improve the paper.
 We have addressed the concerns as best as possible in this revision.
@@
 Any comments in {\it italics} indicate a statement from the referee report.
+
 \begin{enumerate}
@@
 \vspace{5mm}
-We have added a sensitivity analysis to the supplementary materials where we consider an unmeasured binary confounder~$Z_{j,t} \in \{0,1\}$.
-In this setting, {\it prevalence}~$pr(Z=1|X,Y)$ and {\it propensity} $\pi(X, Z; (\theta,\theta_z))$ both impact the sensitivity analysis.
-Imputation-based sensitivity analyses will thus depend heavily on the choice of prevalence and $\theta_z$.
-Therefore, we consider a Taylor series expansion of the IPW and DR-estimators at the estimated propesity which is equivalent to~$\theta_z = 0$.denote the coefficient in the selection propensity model and
-$$
-\hat \mu_{IPW} + \theta_z \frac{1}{\sum_{j=1}^N I^{NR}_j w_j (X_{j,t}; \hat \theta_t)} \left[ \sum_{j=1}^N I^{NR}_j \frac{1-\pi_j(X_{j,t}; \hat \theta_t)}{\pi_j(X_{j,t}; \hat \theta_t)} E[Z_j] (Y_j - \hat \mu_{IPW}) \right]
-$$
-Thus bias depends on the correlation of the error $Y_j - \hat \mu_{IPW}$ to the odds ratio times the expected value of the unmeasured binary confounder.  We use this to calculate the
-
+We have added a sensitivity analysis in Section K of the supplementary materials.
+Note the difficulty of such an analysis comes from needing to specify a joint distribution of the measured covariates~$X_t$, the potential unmeasured and \emph{time-varying} confounder~$U_t$, the self-selection indicator~$I_t$ and the outcome~$Y_t$.
+Here, we extend recent work in the causal inference literature on sensitivity analysis to the self-selection setting in order to avoid direct specification of these distributions.
+Instead, we specify a conditional distribution of the self-selection propensity~$\pi_t(X_t, U_t)$ given the estimated propensity~$\pi_t (X_t)$ and use this as the basis for a sensitivity analysis to derive a simple expression for potential bias.
+The bias depends on two sensitivity parameters: (1) $\alpha_t$ which measures the confounder-selection relationship strength, and (2) $\delta_t$ which measures the confounder-outcome relationship strength.
+We calibrate the selected sensitivity parameter values by analyzing the observed data as if a particular observed covariate were an unobserved confounder.
+A complete set of derivations and associated discussion are presented in Section K of the supplementary materials, which we briefly discuss in Section 5.3 of the main manuscript.
 \vspace{5mm}
@@
 \vspace{5mm}
-We have added this this floor .  We clarify that ``case counts'' in the comment above should be a measure of recent case counts to be a relevant  active infection rate
+We have added the floor of weekly case counts divided by population size to the figure of active infection rate estimates.
+We interpret ``case counts'' in the referee comment above as ``recent case counts'' so that the floor calculation is relevant for estimating the active infection rate.
 \vspace{5mm}
@@
 \vspace{5mm}
-We have added shapes to help differentiate figures while keeping the grey-scale.
+We have added shapes in each figure with more than two categories to help differentiate each curve while keeping them grey-scale to match AOAS figure guidelines.
 \vspace{5mm}

### 2022-09-15 write-up/covid-writeup.tex
updated supplement - now need to fit models for austen plots
updating sensitivity model and making all relevant changes for final referee comments

@@
 
- Upon infection with SARS-CoV-2, an incubation period (time to symptom onset) starts and lasts approximately five days~\citep{Lauer2020}.
-The viral load will be detectable by at least the end of the latent period (time to infectiousness), which for SARS-CoV-2 occurs before the end of the incubation period.
+ Upon infection with the original SARS-CoV-2 variant, an incubation period (time to symptom onset) starts and lasts approximately five days~\citep{Lauer2020}.
+The viral load will be detectable by at least the end of the latent period (time to infectiousness), which for the original SARS-CoV-2 variant occurs before the end of the incubation period.
 A \emph{molecular test} refers to diagnostic tests that aim to detect viral load above a certain threshold (e.g., RT-PCR or antigen tests); see~\cite{Mina2020} for a detailed discussion of cycle thresholds.
@@
 \begin{remark}
+\label{rmk:limitations}
  We acknowledge several limitations with respect to representativeness of these samples.
@@
 Figure~\ref{fig:air_cis} in the supplementary materials presents the confidence intervals per time point for the IPW2 estimator.
-The confidence interval length decreases substantially over time, reflecting the increased testing capacity.
+The confidence interval length decreases substantially over time, reflecting the increased testing capacity over this window of time.
 Due to the number of surveys per week, there is minimal uncertainty in the parameter estimates.
@@
 This points to the importance of the statistical decomposition~\eqref{eq:statdecomp2} and the discussion in Section~\ref{section:IPWerrordecomp}.
-
+Finally, given the reliance on a probabilistic sample with low response rate (see Remark~\ref{rmk:limitations}), a sensitivity analysis showing the impact of a potential unmeasured confounder is presented in Section~\ref{sec:sensitivity_confounders} of the supplementary materials.
   % We also construct ratio estimates as discussed in Section~\ref{section:rates}.

### 2022-09-15 methods/12_propensity_sensitivity.R

@@ -1,0 +1,171 @@
+## Bring in libraries
+library(lubridate)
+
+## Define Functions
+indiana_distance <- function(current_week, current_year) {
+  ind_dist <- function(row) {
+    rowyear = year(row[1])
+    rowweek = week(row[1])
+    if (rowyear == current_year){
+      dis = abs(rowweek - current_week)
+    } else {
+      if (rowyear == 2020) {
+        dis = (53 - rowweek) + current_week
+      } else {
+        dis = (rowweek) + (53-current_week)
+      }
+    }
+    return(dis)
+  }
+  return(ind_dist)
+}
+
+fb_distance <-  function(current_week, current_year) {
+  fb_dist <- function(row) {
+    rowyear = as.numeric(row[7])
+    rowweek = as.numeric(row[6])
+    if (rowyear == current_year){
+      dis = abs(rowweek - current_week)
+    } else {
+      if (rowyear == 2020) {
+        dis = (53 - rowweek) + current_week
+      } else {
+        dis = (rowweek) + (53-current_week)
+      }
+    }
+    return(dis)
+  }
+  return(fb_dist)
+}
+
+kernelweight <- function(distance) {
+  kw2 <- function(row) {
+    dis = distance(row)
+    h = sqrt(-1/(2*log(0.7))) # Set a deterministic bandwidth for now
+    kernel = exp(-dis^2/(2*h^2))
+    return(kernel)
+    }
+  return(kw2)
+}
+
+compute_probs <- function(X, theta) {
+  return(1/(1+exp(-X%*%theta)))
+}
+
+irls <- function(current_week, current_year, fb_X, indiana_X, 
+                 indiana_data, fb_data, max.iter = 100, min.tol = 1e-20) {
+
+  ind_dist = indiana_distance(current_week, current_year)
+  weights = apply(indiana_data, 1, kernelweight(ind_dist))
+  weighted_tests = weights*indiana_data$covid_tests
+  first_term = t(indiana_X)%*%weighted_tests
+  
+  fb_dist = fb_distance(current_week, current_year)
+  fbweights = apply(fb_data, 1, kernelweight(fb_dist))
+  fbweighted_tests = fbweights*fb_data$weight
+  
+  base_rate = sum(weighted_tests)/sum(fbweighted_tests)
+  theta_0 = -log(1/base_rate-1)
+  
+  current_theta = c(theta_0,rep(0,ncol(fb_X)-1))
+  # current_theta = rep(0,ncol(fb_X))
[…]

### 2022-09-19 write-up/covid-supplement-writeup.tex
updating sensitivity model and making all relevant changes for final referee comments
Sensitivity analysis complete :)

@@
 
-\section{Sensitivity Analysis}
+\section{Sensitivity Analysis to Unmeasured Confounding}
+\label{sec:sensitivity_confounders}
 
@@
 $$
-E[ Y_t | i_t x_t, u_t ] = l_t(i_t,x)  + \frac{\delta_t}{\alpha_t} \left( \logit \pi \left( I_{t} = 1  | x_{t}, u_{t} \right) - h_t (x_t) \right)
+E[ Y_t | i_t, x_t, u_t ] = l_t(i_t,x)  + \frac{\delta_t}{\alpha_t} \left( \logit \pi \left( I_{t} = 1  | x_{t}, u_{t} \right) - h_t (x_t) \right)
 $$
@@
 $$
-\alpha_t = E[ \pi_t (X_t,U_t) | I = 1] - E[ \pi_t (X_t,U_t) | I = 0].
+\alpha_t = E[ \pi_t (X_t,U_t) | I_t = 1] - E[ \pi_t (X_t,U_t) | I_t = 0].
 $$
@@
 $$
-R_t^2 (\alpha_t, \delta_t) = \frac{\EE [( Y_{J,t} - Q_t(1,X_{J,t}) )^2 | I_{J,t}=1 ] - \EE[ (Y_{J,t} - \EE[ Y_{J,t} | X_{J,t}, U_{J,t} ])^2 | I_{J,t} = 1]}{E[(Y_{J,t} - Q_t(1,X_{J,t}))^2 | I = 1]}
+R_t^2 (\alpha_t, \delta_t) = \frac{\EE [( Y_{J,t} - Q_t(1,X_{J,t}) )^2 | I_{J,t}=1 ] - \EE[ (Y_{J,t} - \EE[ Y_{J,t} | X_{J,t}, U_{J,t} ])^2 | I_{J,t} = 1]}{E[(Y_{J,t} - Q_t(1,X_{J,t}))^2 | I_t = 1]}
 $$
@@
 That is, for a given observed covariate~$Z_{J,t}$, we wish to measure the degree of influence it has on self-selection and outcome given the other observed covariates~$X_{J,t} \backslash Z_{J,t}$.
-Note that this dependence is likely time-varying, which is why it is important estimate~$\alpha_t$ and~$R_t^2$ as functions of time~$t$.
+Note that this dependence is likely time-varying, which is why it is important that~$\alpha_t$ and~$R_t^2$ are functions of time~$t$.
 
@@
 $$
-We can measure influence of $Z$ on self-selection propensity given $X_{J,t} \backslash Z_{J,t}$ by
+among the non-probabilistic sample.  We can measure influence of $Z$ on self-selection propensity given $X_{J,t} \backslash Z_{J,t}$ by
 $$
@@
 While~\cite{Veitch2020} use Austen plots, we do not have such a simple visualization due to time-dependence.
-To handle this, for each observed covariate~$Z$, we plot $(\alpha_{t,Z}, R^2_{t,Z})$ in the $[0,1]^2$.
-These ``calibration curves'' help us in selection of a range of values for~$(\alpha_t, R^2_t)$ in our sensitivity analysis.
-Here, we consider $Z$ to be separately (a) age, (b) gender, (c) race, and (d) symptom status.
-The four plots are presented in Figure XX.
+Here, we plot  $\alpha_{t,Z}$ and $R^2_{t,Z}$ as functions of time for three covariates that demonstrate different patterns: (1) Symptoms, (2) Race, and (3) Ethnicity.
+Figures~\ref{fig:alpha_sensitivity} and~\ref{fig:rsq_sensitivity} present these ``calibration curves.'' Figure~\ref{fig:bias_sensitivity} presents the bias under these three curves.
+Here, we see that if an unobserved confounder has similar strength in selection-confounder and outcome-confounder relationships as ``Symptom Status'', then bias would be high even into 2022.
+However, bias under an unobserved confounder with similar strength to ``Ethnicitiy'' dissipates significantly by late 2021.
+
+\begin{figure}[!th]
+ \centering
+ \begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{../figs/alpha_sensitivity_fig.png}
+  \caption{Calibration curves for~$\alpha$ per week}
+  \label{fig:alpha_sensitivity}
+ \end{subfigure}
+ \begin{subfigure}{.45\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{../figs/rsq_sensitivity_fig.png}
+ \caption{Calibration curves for~$R^2$ per week}
+ \label{fig:rsq_sensitivity}
+ \end{subfigure} \\ [1ex]
+ \begin{subfigure}{\linewidth}
+ \centering
+ \includegraphics[width=.5\linewidth]{../figs/bias_sensitivity_fig.png}
+  \caption{Bias as a function of~$\alpha$ and~$R^2$ calibration curves}
+  \label{fig:bias_sensitivity}
+ \end{subfigure}
+ \caption{Sensitivity analysis for time-varying active infection rate estimates}
+ \label{fig:sensitivity}
+ \end{figure}
 

### 2022-09-19 methods/13_propvariance_sensitivity.R

@@ -95,11 +95,13 @@ png(filename = "../figs/alpha_sensitivity_fig.png",
     width = 960, height = 480, units = "px", pointsize = 25)
 
 ggplot(data = subset_alpha_plot_df, aes(x = date, y = alpha_t, group = Method)) +
-  ylim(0,0.08)+ geom_point(size = 2, aes(color = Method, shape = Method)) +
+  ylim(0,0.08)+ geom_point(size = 5, aes(color = Method, shape = Method)) +
   labs(x = "Date",
        y = expression(alpha)) + 
-  theme(text = element_text(size=25)) +
-  scale_color_manual(values=my_palette) + theme_minimal() + geom_smooth(method = "loess", aes(color = Method))
+  # theme(text = element_text(size=25)) +
+  scale_color_manual(values=my_palette) + theme_minimal() + 
+  geom_smooth(method = "loess", aes(color = Method)) + 
+  theme(text = element_text(size=20))
 
 dev.off()
 

### 2022-09-19 methods/14_modelfit_sensitivity.R

@@ -91,11 +91,13 @@ png(filename = "../figs/rsq_sensitivity_fig.png",
     width = 960, height = 480, units = "px", pointsize = 25)
 
 ggplot(data = subset_rsq_plot_df, aes(x = date, y = rsq_t, group = Method)) +
-  ylim(0,0.125)+ geom_point(size = 2, aes(color = Method, shape = Method)) +
+  ylim(0,0.125)+ geom_point(size = 5, aes(color = Method, shape = Method)) +
   labs(x = "Date",
        y = expression(R^2)) + 
-  theme(text = element_text(size=25)) +
-  scale_color_manual(values=my_palette) + theme_minimal() + geom_smooth(method = "loess", aes(color = Method))
+  # theme(text = element_text(size=25)) +
+  scale_color_manual(values=my_palette) + theme_minimal() + 
+  geom_smooth(method = "loess", aes(color = Method)) + 
+  theme(text = element_text(size=20))
 
 dev.off()
 

### 2022-09-19 methods/15_finalsensitivity_analysis.R

@@ -103,10 +103,12 @@ png(filename = "../figs/bias_sensitivity_fig.png",
     width = 960, height = 480, units = "px", pointsize = 25)
 
 ggplot(data = bias_plot_df, aes(x = date, y = bias, group = Method)) +
-  ylim(0,0.2)+ geom_point(aes(shape = Method, color = Method), size = 2) +
+  ylim(0,0.2)+ geom_point(aes(shape = Method, color = Method), size = 5) +
   labs(x = "Date",
        y = "Bias") + 
-  theme(text = element_text(size=25)) +
-  scale_color_manual(values=my_palette) + theme_minimal() #+ geom_smooth(method = "loess")
+  # theme(text = element_text(size=25)) +
+  scale_color_manual(values=my_palette) + 
+  theme_minimal() + 
+  theme(text = element_text(size=20)) # + geom_smooth(method = "loess")
 
 dev.off()

### 2022-09-19 write-up/letters/response-ref-1-v1.tex
White background on figures with many categories to enhance visuals

@@
 \vspace{5mm}
-We have added shapes in each figure with more than two categories to help differentiate each curve while keeping them grey-scale to match AOAS figure guidelines.
+We tested a variety of means to differentiate among categories in figures with more than two categories.
+We found varied shapes made the plots much harder to understand in settings with significant overlap.
+Therefore, we opted to simply remove the gray background.
+The white background helps with differentiation while keeping the figures in grey-scale to match AOAS figure guidelines.
 \vspace{5mm}

### 2022-09-19 visualizations/tv_propensities_alt.R

@@ -68,11 +68,12 @@ png(filename = "../figs/tvprop_alt_fig1_mainpaper.png",
 
 ggplot(data = propensities_long, aes(x = date, y = propensity, col = age)) +
   geom_point(size = 5) +
+  theme_classic() +
   labs(x = "Date",
        y = "COVID-19 Testing Propensity",
        title = "Non-Hispanic, White Male with Symptoms") +
   theme(text = element_text(size=25)) +
-  scale_color_manual(values=my_palette)
+  scale_color_manual(values=my_palette) 
 
 dev.off()
 
@@ -133,11 +134,12 @@ png(filename = "../figs/tvprop_alt_fig2_mainpaper.png",
 
 ggplot(data = propensities_long, aes(x = date, y = propensity, col = age)) +
   geom_point(size = 5) +
+  theme_classic() +
   labs(x = "Date",
        y = "COVID-19 Testing Propensity",
        title = "Non-Hispanic, White Male without Symptoms") +
   theme(text = element_text(size=25)) +
-  scale_color_manual(values=my_palette)
+  scale_color_manual(values=my_palette) 
 
 dev.off()
 
@@ -189,6 +191,7 @@ png(filename = "../figs/tvprop_alt_fig1_supp1.png",
 
 ggplot(data = propensities_long, aes(x = date, y = propensity, col = age)) +
   geom_point(size = 5) +
+  theme_classic() + 
   labs(x = "Date",
        y = "COVID-19 Testing Propensity",
        title = "Non-Hispanic, African American Male with Symptoms") +
@@ -236,6 +239,7 @@ png(filename = "../figs/tvprop_alt_fig2_supp1.png",
 
 ggplot(data = propensities_long, aes(x = date, y = propensity, col = age)) +
   geom_point(size = 5) +
+  theme_classic() +
   labs(x = "Date",
        y = "COVID-19 Testing Propensity",
        title = "Non-Hispanic, African American Male without Symptoms") +
@@ -285,6 +289,7 @@ png(filename = "../figs/tvprop_alt_fig1_supp2.png",
 
 ggplot(data = propensities_long, aes(x = date, y = propensity, col = age)) +
   geom_point(size = 5) +
+  theme_classic() +
   labs(x = "Date",
        y = "COVID-19 Testing Propensity",
        title = "Hispanic, White Male with Symptoms") +
@@ -332,6 +337,7 @@ png(filename = "../figs/tvprop_alt_fig2_supp2.png",
 
[…]

### 2022-09-19 write-up/letters/response-ref-1-v1.tex

@@
 \vspace{5mm}
-We have added the floor of weekly case counts divided by population size to the figure of active infection rate estimates.
-We interpret ``case counts'' in the referee comment above as ``recent case counts'' so that the floor calculation is relevant for estimating the active infection rate.
+We have added the floor of weekly case counts divided by population size to Figure 11 in the supplementary materials.
+We interpret ``case counts'' in the referee comment above as ``weekly case counts'' so that the floor calculation is relevant for estimating the active infection rate.
+Note all estimates are above the floor with the doubly robust being closest to the floor in September.
 \vspace{5mm}

### 2022-09-20 write-up/letters/cover letter-covid.tex
Cov Letter added
Updated papers - almost there

@@
 
-\begin{letter}{Professor Karen Kafadar \\
+\begin{letter}{Professor Ji Zhu \\
     Editor-In-Chief, {\em Annals of Applied Statistics}
@@
 
-\opening{Dear Dr. Kafadar:}
+\opening{Dear Dr. Zhu:}
 
-Kindly find enclosed the revised version of the manuscript number AOAS2006-048 titled ``Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic'' now titled ``Addressing selection bias and measurement error in COVID-19 case count data using auxiliary information.''  We have revised the article to address the referees’ requests about exposition and other points of clarification.
-The reports called for a major revision to make the manuscript ``forward compatible'', to re-structure the manuscript to present figures and tables near their corresponding discussion, to discuss potential non-response bias in the motivating studies, to include a more careful discussion of COVID-19 testing and disease dynamics, to include a more precise definition of the active infection rate and how it impacts model-based estimation, and to more carefully tie the proposal to related parts of the epidemiological literature.
-The new version complies with these requests.
+Kindly find enclosed the revised version of the manuscript number AOAS2006-048 titled ``Addressing selection bias and measurement error in COVID-19 case count data using auxiliary information.''  We are happy that the two reviewers, associate editor, and editor have all found the revision much improved.
+We have revised the article to address the few remaining concerns that one of the reviewers raised but are glad that they ``appreciate the paper and [are] very supportive of the pragmatic approach.''  We have edited the remark, performed a sensitivy analysis to assess sensitivity due to potential confounding, and adjusted the figures to assist in comprehension.
 
-We believe the work presents an important advance in how researchers should think about addressing time-varying selection bias and measurement error to better understanding the COVID-19 pandemic trajectory.
-Attached please find a detailed response to the referees outlining how the changes to the manuscript address the concerns raised in their reports.
+We continue to believe the work presents an important advance in how researchers should think about addressing time-varying selection bias and measurement error to better understanding the COVID-19 pandemic trajectory.
+Attached please find a detailed response to the referee outlining how the changes to the manuscript address the concerns raised in their report.
 

### 2022-09-20 write-up/covid-supplement-writeup.tex
Update covid-supplement-writeup.tex
Updated papers - almost there
supplementary materials, final pass

@@
 
-\title[Statistical paradoxes in coronavirus case-counts]{Supplementary Materials to ``Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic''} %
+\title[Addressing selection bias and measurement error in COVID-19 case counts]{Supplementary Materials to ``Addressing selection bias and measurement error in COVID-19 case count data using auxiliary information''} %
 
@@
 \begin{abstract}
-This is the supplementary materials to the manuscript titled ``Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic.''  Section~\ref{app:imperfect} derives the statistical decompositions for the measurement-error corrected estimator and provides the connection to model-based estimators.
+This is the supplementary materials to the manuscript titled ``Addressing selection bias and measurement error in COVID-19 case count data using auxiliary information.''  Section~\ref{app:imperfect} derives the statistical decompositions for the measurement-error corrected estimator and provides the connection to model-based estimators.
 Sections~\ref{app:ipwderivation} and~\ref{app:DRderivation} derives the statistical decomposition for the inverse-probability weighted and doubly robust estimators respectively.
@@
 We see that the impact on the doubly robust estimates is minimal.
+A floor based on weekly case count divided by the Indian population size is also presented.
 
@@
 Therefore, the bias can be readily obtained from the associated Taylor series decomposition, with the terms related to time~$t-1$ replaced by a weighted version.
-Figure~\ref{fig:cori-bias} presents the bias.
+% Figure~\ref{fig:cori-bias} presents the bias.
 

### 2022-09-21 write-up/covid-supplement-writeup.tex
Update covid-supplement-writeup.tex

@@
 Section~\ref{app:irls} outlines the iteratively re-weighted least squares algorithm.
-Section~\ref{app:in_add_details} provides additional detals on the COVID-19 analysis of Indiana's case-count data.
+Section~\ref{app:in_add_details} provides additional details on the COVID-19 analysis of Indiana's case-count data.
 Section~\ref{app:prior_modelbased} provides prior specifications for the SEIR model.
+Section~\ref{app:cori_rt} provides an alternative estimator of the instantaneous reproductive number and associated analysis of bias under an SEIR model.
+Section~\ref{sec:sensitivity_confounders} provides a sensitivity analysis to unmeasured confounders.
 \end{abstract}
@@
  \centering
-  \includegraphics[width=.9\linewidth]{../figs/seir_cori_rt.png}
+  \includegraphics[width=.75\linewidth]{../figs/seir_cori_rt.png}
   \caption{Effective reproductive rate estimator}
@@
 \logit \pi \left( I_{t} = 1  | x_{t}, u_{t} \right) &= h_t(x_t) + \alpha_t u_t \\
-E[ Y_t | i_t, x_t, u_t] &= l_t(i_t, x_t) + \delta_t u_t
+E[ Y_t | i_t, x_t, u_t] &= l_t(x_t) + \delta_t u_t
 \end{align*}
@@
 $$
-E[ Y_t | i_t, x_t, u_t ] = l_t(i_t,x)  + \frac{\delta_t}{\alpha_t} \left( \logit \pi \left( I_{t} = 1  | x_{t}, u_{t} \right) - h_t (x_t) \right)
+E[ Y_t | i_t, x_t, u_t ] = l_t(x_t)  + \frac{\delta_t}{\alpha_t} \left( \logit \pi \left( I_{t} = 1  | x_{t}, u_{t} \right) - h_t (x_t) \right)
 $$
@@
 The logit-linear model above unfortunately does not lead to a tractable sensitivity analysis.
-Instead, we consider a sensitivity model based on work by~\cite{Veitch2020}, again adjusted to deal with the current setting of self-selection, given by:
+Instead, we consider a sensitivity model based on work by~\cite{Veitch2020}, again adjusted to deal with the current setting of self-selection, given by
 \begin{equation}
@@
 \rho_t(X_t, U_t) &= Q_t(X_t,1) + \delta_t \left( \logit \pi_t(X_t,U_t) - \EE \left[ \logit \pi_t(X_t,U_t) | X_t, I_t = 1 \right] \right) \\
-Y_t | I_t, X_t, U_t &\sim \text{Bern}(\rho_t(X_t, U_t))
+Y_t | I_t, X_t, U_t &\sim \text{Bern}(\rho_t(X_t, U_t)),
 \end{aligned}
 \end{equation}
+where~$Q_t(X_t,1)$ is the conditional expectation of the outcome given covariates~$X_t$ and self-selection into the nonprobability sample, i.e.,~$I_t = 1$.
 The time-varying sensitivity parameter~$\alpha_t \in (0,1)$  controls the influence of the unobserved confounder~$U_t$ on selection propensity.
@@
 = &\EE[ E[ Y_{J,t} | X_{J,t}, U_{J,t}] ] \\
-= &\EE \left[ Q(1,X_{J,t}) + \delta \left( \logit \pi_t (X_{J,t}, U_{J,t}) - \EE \left[ \logit \pi_t (X_{J,t}, U_{J,t}) | X_{J,t}, I_{J,t} = 1 \right] \right) \right]
+= &\EE \left[ Q(1,X_{J,t}) + \delta_t \left( \logit \pi_t (X_{J,t}, U_{J,t}) - \EE \left[ \logit \pi_t (X_{J,t}, U_{J,t}) | X_{J,t}, I_{J,t} = 1 \right] \right) \right]
 \end{align*}
@@
 $$
-\delta \EE \left[ \logit \pi_t (X_{t,J}, U_{t,J}) - \frac{\pi_t (X_{J,t}, U_{J,t})}{\pi_t (X_{J,t})} \logit \pi_t (X_{t,J}, U_{t,J}) \right].
+\delta_t \EE \left[ \logit \pi_t (X_{t,J}, U_{t,J}) - \frac{\pi_t (X_{J,t}, U_{J,t})}{\pi_t (X_{J,t})} \logit \pi_t (X_{t,J}, U_{t,J}) \right].
 $$
@@
 \end{equation}
-Note that unlike the prior expecation, these are expressed conditional on self-selection, i.e., $I_{J,t} = 1$.
+Note that unlike the prior expectation, these are expressed conditional on self-selection, i.e., $I_{J,t} = 1$.
 This was done since the binary outcomes are only measured in the non-probabilistic sample.
 
-Following from~\cite[Theorem 5]{Veitch2020}, the parameter~$\alpha_t$ can be reexpressed as
+Following from~\cite[Theorem 5]{Veitch2020}, the parameter~$\alpha_t$ can be re-expressed as
 \begin{equation}
@@
 $$
-R_{t, Z}^2 := \frac{\frac{1}{n} \sum_{i=1}^n (y_{i,t} - \hat Q_{t,Z} (1, x_{i,t} \backslash z_{i,t}) )^2 - \frac{1}{n} \sum_{i=1}^n (y_{i,t} - \hat Q_{t,Z} (1, x_{i,t}) )^2}{\frac{1}{n} \sum_{i=1}^n (y_{i,t} - \hat Q_{t,Z} (1, x_{i,t} \backslash z_{i,t}) )^2}.
-$$
-among the non-probabilistic sample.  We can measure influence of $Z$ on self-selection propensity given $X_{J,t} \backslash Z_{J,t}$ by
+R_{t, Z}^2 := \frac{\frac{1}{n} \sum_{i=1}^n (y_{i,t} - \hat Q_{t,Z} (1, x_{i,t} \backslash z_{i,t}) )^2 - \frac{1}{n} \sum_{i=1}^n (y_{i,t} - \hat Q_{t} (1, x_{i,t}) )^2}{\frac{1}{n} \sum_{i=1}^n (y_{i,t} - \hat Q_{t,Z} (1, x_{i,t} \backslash z_{i,t}) )^2},
+$$
+where~$\hat Q_t$ is the conditional expectation given all covariates and~$\hat Q_{t,Z}$ is the conditional expectation given all covariates except~$Z$.
+Note these are sums averages the non-probabilistic sample.
+We can measure influence of $Z$ on self-selection propensity given $X_{J,t} \backslash Z_{J,t}$ by
 $$
@@
 
-Here we perform a sensitivity analysis of the IPW estimator of COVID-19 active infection prevalence in indiana.
+Here we perform a sensitivity analysis of the IPW estimator of COVID-19 active infection prevalence in Indiana.
 While~\cite{Veitch2020} use Austen plots, we do not have such a simple visualization due to time-dependence.
@@
 Figures~\ref{fig:alpha_sensitivity} and~\ref{fig:rsq_sensitivity} present these ``calibration curves.'' Figure~\ref{fig:bias_sensitivity} presents the bias under these three curves.
-Here, we see that if an unobserved confounder has similar strength in selection-confounder and outcome-confounder relationships as ``Symptom Status'', then bias would be high even into 2022.
-However, bias under an unobserved confounder with similar strength to ``Ethnicitiy'' dissipates significantly by late 2021.
+Here, we see that if an unobserved confounder has similar strength in selection-confounder and outcome-confounder relationships as ``Symptom Status'', then bias would be high even into early 2021.
+However, bias under an unobserved confounder with similar strength to ``Ethnicity'' dissipates significantly by late 2020.
 

### 2023-01-04 write-up/covid-supplement-writeup.tex
updating for final round

@@
 The asymptotic variance for the IPW estimator is then the first diagonal element of the matrix $E \left[ \phi_n (\eta_0) \right]^{-1} \times \left[ V_{1,t} + V_{2,t} + V_{3,t} + V_{4,t} \right] E \left[ \phi_n (\eta_0) \right]^{-1}$.
+
+We next state Lemma~\ref{lemma:dr} for the doubly-robust estimator presented in Section XX.  The proof follows the same derivation as above.
+
+\begin{lemma}[Variance of doubly-robust estimator] \normalfont
+ \label{lemma:dr}
+ The estimates~$\mu_t := \bar y_{n,t}^\star$ and $\hat \pi_{j,t}$
+ are solutions to the following set of estimating equations
+ $$
+ \Phi_n (\eta_t) =
+ \left (
+ \begin{array}{c}
+ \frac{1}{N} \left[ \sum_{j=1}^N \frac{I^{NR}_{j,t}}{\pi_{j,t}} \left( \left( \frac{Y_{j,t} - FP}{1-FP-FN} - \hat \mu_{j,t}  \right)  - \mu_t \right) + \frac{\sum_{l=1}^L K_{h} (|t - t_l^\prime|) \sum_{j=1}^N I^{R}_{j,t_l^\prime} W^{R}_{j,t_l^\prime} \hat \mu_{j,t_l^\prime}}{\sum_{l=1}^L K_{h} (|t - t_l^\prime|) \sum_{j=1}^N I^{R}_{j,t_l^\prime} W^{R}_{j,t_l^\prime}} \right] \\
+ \frac{1}{N \times \sum_{l=1}^L K_{h} (|t - t_l^\prime|)} \sum_{l=1}^L K_{h} (|t - t_l^\prime|) \sum_{j=1}^N I^{NR}_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N I^{R}_{j,t^\prime} W^{R}_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
+ \frac{1}{n_{FP}} \sum_{j=1}^{n_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
+ \frac{1}{n_{FN}} \sum_{j=1}^{n_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
+ \end{array}
+ \right ) = {\bf 0}
+ $$
+ where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
+Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (\bar n^{-1/2})$ and $\text{var} (\hat y_{n,t}) = V_{t}^{(DR)} + o (\bar n^{-1})$ where $\bar n = n \sum_{l=1}^L K_h(|t-t_l|)$ and $V_t^{(DR)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$ where $\phi_n = \frac{\partial \Phi (\eta)}{\partial \eta}$ with $E[\cdot]$ and $\text{Var} (\cdot)$ are under the joint randomization of propensity score and sampling designs.
+ \end{lemma}
+
 \section{IRLS}
@@
  \centering
+ \begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{../figs/deaths_ppc.png}
+ \caption{PPC on aggregate daily death counts.}
+ \label{fig:deathsppc}
+ \end{subfigure}%
+ \begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{../figs/infections_ppc.png}
+ \caption{Posterior on daily new infections.}
+ \label{fig:casesppc}
+ \end{subfigure}
+ \caption{Posterior predictive check (PPC) on death counts and posterior distribution of infections.}
+ \label{fig:ppcs}
+ \end{figure}
+
+ To check model fit, a posterior predictive check on aggregate death counts was performed.
+Figure~\ref{fig:deathsppc} shows the results which suggests chosen SEIR model fits the observed data well.
+Figure~\ref{fig:casesppc} presents the posterior distribution of daily new infections.
+As the active infection rate is defined as all infected individuals whose viral load has yet to fall below detectable levels by RT-PCR testing, to construct the active infection estimates we exponentially discount the number of newly infected individuals to estimate the number of these newly infected individuals who still have an active infection on future days.
+That is, the number of active infections on day~$t$ is given by~$\sum_{s=0}^t I_t^{\new} e^{-\lambda (t-s)}$.
+Strata-specific active infections are calculated similarly.
+In our analysis, the exponential discounting parameter is set to $20$ days to match with prior evidence that ``30\% to 40\% of people will still test positive at three weeks''~\citep{wlocneg}.
+To compare with case count data, Figure~\ref{fig:undercounting} in the supplementary materials computes the cumulative undercount factor of observed case counts as compared to new infections.
+Posterior distributions for the age-specific new infections are used as part of the doubly-robust estimator of the active infection rate.
+
+\begin{figure}[!th]
+ \centering
  \includegraphics[width=.6\linewidth]{../figs/cumulative_undercounting.png}

### 2023-01-04 write-up/covid-writeup.tex
updating for final round

@@
 
-\newpage
-
 \section{Introduction}
@@
 After that, the viral load will decrease below that threshold and a molecular test with perfect specificity will come back negative.
-While an infected individual may yield a positive molecular test for several weeks, they will likely stop transmitting the disease within a few days of infection, meaning a positive molecular test does not imply transmissibility.
+While an individual infected with the original SARS-CoV-2 strain may yield a positive molecular test for several weeks, they will likely stop transmitting the disease within a few days of infection, meaning a positive molecular test does not imply transmissibility.
 
@@
 In the incubation phase, most molecular tests will return a false negative result.
-Molecular tests are most sensitive when the viral loads are highest which occurs during the first few days of transmissibility~\citep{Mina2020}.
+Molecular tests are most sensitive when the viral loads are highest which for the original strain occurs during the first few days of transmissibility~\citep{Mina2020}.
 Moreover, most molecular tests are performed via nasopharyngeal swab.
 Specimen collection by swab is known to impact false negative/positive rates regardless of test timing.
-Systematic reviews suggest that 87\% sensitivity and 97.6\% specificity are reasonable estimates for current RT-PCR tests~\citep{Arevalo2020, Woloshin2020,Cohen2020}.
+Systematic reviews suggest that 87\% sensitivity and 97.6\% specificity are reasonable estimates for RT-PCR tests performed during the time window under consideration in this paper~\citep{Arevalo2020, Woloshin2020,Cohen2020}.
 To the best of our knowledge, both Indiana Department of Health's molecular testing and the random state-wide RT-PCR tests were primarily collected via nasopharyngeal swab.
@@
 Unfortunately, most dashboards do not provide demographic information on who requested a test nor on who tested positive for SARS-CoV-2 \emph{over time}.
-Working closely with the State of Indiana, we were able to obtain access to COVID-19 total tests, positive tests, and related deaths per day broken out marginally by age, gender, ethnicity, and race~\citep{IndianaData2021}.
+Working closely with the State of Indiana, we were able to access COVID-19 total tests, positive tests, and related deaths per day broken out marginally by age, gender, ethnicity, and race~\citep{IndianaData2021}.
 These granular datasets are now publicly available and motivate the proposed approach.
@@
  \end{tabular}
- \caption{Estimated total tests ($\%$) and point prevalence of active infection with SARS-CoV-2 by demographics in Indiana~\citep{Yiannoutsos2021,doi:10.1073/pnas.2111454118}.}
+ \caption{Estimated total tests ($\%$) and point prevalence of active infection with SARS-CoV-2 by demographics in Indiana~\citep{Yiannoutsos2021,doi:10.1073/pnas.2111454118}.
+NonRandom positive test rates in parentheses are adjusted rates to match the statewide rate of $11.7\%$ on August 30th.}
  \label{tab:indiana}
@@
  where $s_t, e_t, i_t$ and $r_t$ are the fraction of susceptible, exposed, infected, and removed (recovered or deceased) individuals in the population at time $t$ respectively.
-The SEIR model has been used extensively as a model for SARS-CoV-2 dynamics~\cite{Song2020}.
+The SEIR model has been used extensively as a model for SARS-CoV-2 dynamics~\citep{Song2020}.
 In terms of bias, the rate is overestimated prior to the peak in the fraction infected and underestimated afterwards; the bias increases dramatically when the relative fraction exceeds $2$.
@@
 
-Under a Poisson likelihood, a simple relation between the trajectory of new cases and the instantaneous reproduction number can be derived \citep{Bettencourt2008}.
+% Under a Poisson likelihood, a simple relation between the trajectory of new cases and the instantaneous reproduction number can be derived \citep{Bettencourt2008}.
 In particular, under an SIR model the number of case counts on day $t$, denoted $K_t$, is Poisson distributed with rate $K_{t-1} \exp \left( \gamma (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval, which is approximately $7$ days for COVID-19~\citep{Sanche2020}.
@@
 Unfortunately the number of susceptible individuals on day $t$ is unknown.
-Here, we study the estimator $\hat R_t = 1 + \frac{1}{\gamma} \log \left( \tilde y_t / \tilde y_{t-1} \right)$.
+Here, we study the estimator $\hat R_t = 1 + \left( \frac{1}{\gamma} + \frac{1}{\sigma}\right) \log \left( \tilde y_t / \tilde y_{t-1} \right)$.
 We can again express the statistical error of $\hat R_t - R_t$ in useful terms as follows
@@
 
- \subsection{Rate comparisons}
-
- \begin{figure}[!th]
- \centering
- \begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{../figs/seir.png}
-  \caption{Fraction of new cases in population}
-  \label{fig:fracpop}
- \end{subfigure}%
- \begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{../figs/seir_rt_comparison.png}
-  \caption{Effective reproduction rate estimators}
-  \label{fig:eff}
- \end{subfigure}
- \caption{Left: fraction infected in two SEIR models with $\beta = 1.2$ and $0.9$ respectively, $\sigma = 0.3$, and $\gamma = 0.15$ with same initial conditions.
+ % \subsection{Rate comparisons}
+
+ % \begin{figure}[!th]
+ % \centering
+ % \begin{subfigure}{.5\textwidth}
+ %  \centering
+ %  \includegraphics[width=.9\linewidth]{../figs/seir.png}
+ %  \caption{Fraction of new cases in population}
+ %  \label{fig:fracpop}
+ % \end{subfigure}%
+ % \begin{subfigure}{.5\textwidth}
+ %  \centering
+ %  \includegraphics[width=.9\linewidth]{../figs/seir_rt_comparison.png}
+ %  \caption{Effective reproduction rate estimators}
+ %  \label{fig:eff}
+ % \end{subfigure}
+ % \caption{Left: fraction infected in two SEIR models with $\beta = 1.2$ and $0.9$ respectively, $\sigma = 0.3$, and $\gamma = 0.15$ with same initial conditions.
 Right: comparison of $\hat R_t$ across time with $FN = 0.30$, $FP = 0.024$, and $M = 4$.}
- \label{fig:comparison}
- \end{figure}
-
- So far we have focused on understanding the limitations of using case-count data to understand population quantities of interest for a \emph{single} population.
+ % \label{fig:comparison}
+ % \end{figure}
+
+ % So far we have focused on understanding the limitations of using case-count data to understand population quantities of interest for a \emph{single} population.
 Many are interested in cross-population comparisons to contrast the impact of countries' mitigation policies.
@@
 We assume the two time-series are aligned so that $t=0$ is the time of first case in each population respectively.
- % This negates alignment issues and is common in practice.
-
- % While the above issues on $Z$-scores and effective sample size are important, here we highlight a separate issue.
- %Here we demonstrate the interaction of biases in estimation of the two trajectories when the peak infection times differ slightly.
- %First, we can write the difference in the rates as $R_{t1} - R_{t2} = \frac{1}{\gamma} \log \left( \frac{1 + e_{1t}}{1+e_{2t}} \right)$ where $e_{jt}$ is the error associated given in Section \ref{section:r0-estimation}.
- %Recall that the error allowed for over-estimation prior to the peak and then under-estimation post-peak.  Here, these errors can mingle in interesting ways.
- Consider two countries (A and B) in which the peak occurs 2 weeks prior for country A than country B.
+ % % This negates alignment issues and is common in practice.
+
+ % % While the above issues on $Z$-scores and effective sample size are important, here we highlight a separate issue.
+ % %Here we demonstrate the interaction of biases in estimation of the two trajectories when the peak infection times differ slightly.
+ % %First, we can write the difference in the rates as $R_{t1} - R_{t2} = \frac{1}{\gamma} \log \left( \frac{1 + e_{1t}}{1+e_{2t}} \right)$ where $e_{jt}$ is the error associated given in Section \ref{section:r0-estimation}.
+ % %Recall that the error allowed for over-estimation prior to the peak and then under-estimation post-peak.  Here, these errors can mingle in interesting ways.
+ % Consider two countries (A and B) in which the peak occurs 2 weeks prior for country A than country B.
 Figure~\ref{fig:comparison} presents such a comparison where each country's disease trajectory follows an SEIR model (A=black and B=grey).
@@
  % \textcolor{red}{Must check.}
-The estimates~$\hat \mu_t := \bar y_{n,t}^\star$, $\hat \pi_{j,t}$, $\hat{FP}$, and $\hat{FN}$ at a time~$t \in \{t_l^\prime\}$ are solutions to the following set of estimating equations
- $$
- \Phi_n (\eta_t) =
- \left (
- \begin{array}{c}
- \frac{1}{N} \sum_{j=1}^N I^{NR}_{j,t} \frac{Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t}{\pi_{j,t}} \\
- \frac{1}{N \times \sum_{l = 1}^L K_h \left(|t - t_l^\prime| \right)} \sum_{l = 1}^L K_h \left(|t - t_l^\prime| \right) \left[ \sum_{j=1}^N I^{NR}_{j,t_l^\prime} X_{j,t_l^\prime} - \sum_{j = 1}^N I^{R}_{j,t_l^\prime} W^{R}_{j,t_l^\prime}  \pi_{j,t_l^\prime} X_{j,t_l^\prime}  \right] \\
- \frac{1}{n_{FP}} \sum_{i=1}^{n_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
- \frac{1}{n_{FN}} \sum_{i=1}^{n_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
- \end{array}
- \right ) = {\bf 0}
- $$
+The estimates~$\hat \mu_t := \bar y_{n,t}^\star$, $\hat \pi_{j,t}$, $\hat{FP}$, and $\hat{FN}$ at a time~$t \in \{t_l^\prime\}$ are solutions to the set of estimating equations~$\Phi_n(\eta_t)$ given in Appendix~\ref{app:asympderivations},
+ % $$
+ % \Phi_n (\eta_t) =
+ % \left (
+ % \begin{array}{c}
+ % \frac{1}{N} \sum_{j=1}^N I^{NR}_{j,t} \frac{Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t}{\pi_{j,t}} \\
+ % \frac{1}{N \times \sum_{l = 1}^L K_h \left(|t - t_l^\prime| \right)} \sum_{l = 1}^L K_h \left(|t - t_l^\prime| \right) \left[ \sum_{j=1}^N I^{NR}_{j,t_l^\prime} X_{j,t_l^\prime} - \sum_{j = 1}^N I^{R}_{j,t_l^\prime} W^{R}_{j,t_l^\prime}  \pi_{j,t_l^\prime} X_{j,t_l^\prime}  \right] \\
+ % \frac{1}{n_{FP}} \sum_{i=1}^{n_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
+ % \frac{1}{n_{FN}} \sum_{i=1}^{n_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
+ % \end{array}
+ % \right ) = {\bf 0}
+ % $$
  where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
@@
  $$
- Assuming correct selection propensity model specification, then under the probability sample design and nonprobability sample design, Lemma~\ref{lemma:dr} shows that the doubly robust estimator is consistent and calculates the estimator's variance.
-See Appendix~\ref{app:asympderivations} for additional details.
-
- \begin{lemma}[Variance of doubly-robust estimator] \normalfont
- \label{lemma:dr}
- The estimates~$\mu_t := \bar y_{n,t}^\star$ and $\hat \pi_{j,t}$
- are solutions to the following set of estimating equations
- $$
- \Phi_n (\eta_t) =
- \left (
- \begin{array}{c}
- \frac{1}{N} \left[ \sum_{j=1}^N \frac{I^{NR}_{j,t}}{\pi_{j,t}} \left( \left( \frac{Y_{j,t} - FP}{1-FP-FN} - \hat \mu_{j,t}  \right)  - \mu_t \right) + \frac{\sum_{l=1}^L K_{h} (|t - t_l^\prime|) \sum_{j=1}^N I^{R}_{j,t_l^\prime} W^{R}_{j,t_l^\prime} \hat \mu_{j,t_l^\prime}}{\sum_{l=1}^L K_{h} (|t - t_l^\prime|) \sum_{j=1}^N I^{R}_{j,t_l^\prime} W^{R}_{j,t_l^\prime}} \right] \\
- \frac{1}{N \times \sum_{l=1}^L K_{h} (|t - t_l^\prime|)} \sum_{l=1}^L K_{h} (|t - t_l^\prime|) \sum_{j=1}^N I^{NR}_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N I^{R}_{j,t^\prime} W^{R}_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
- \frac{1}{n_{FP}} \sum_{j=1}^{n_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
- \frac{1}{n_{FN}} \sum_{j=1}^{n_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
- \end{array}
- \right ) = {\bf 0}
- $$
- where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
-Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (\bar n^{-1/2})$ and $\text{var} (\hat y_{n,t}) = V_{t}^{(DR)} + o (\bar n^{-1})$ where $\bar n = n \sum_{l=1}^L K_h(|t-t_l|)$ and $V_t^{(DR)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$ where $\phi_n = \frac{\partial \Phi (\eta)}{\partial \eta}$ with $E[\cdot]$ and $\text{Var} (\cdot)$ are under the joint randomization of propensity score and sampling designs.
- \end{lemma}
+ Assuming correct selection propensity model specification, then under the probability sample design and nonprobability sample design, Lemma~\ref{lemma:dr} in Appendix~\ref{app:asympderivations} shows that the doubly robust estimator is consistent and calculates the estimator's variance.
 
@@
 Each weight~$w_t^{(k)} = 0$ for $t < t_k$ and $w_t^{(k)} \to 1$ as $t \to \infty$.
-See Section~\ref{app:prior_modelbased} in the supplementary materials for details on prior specification.
-
- \begin{figure}[!th]
- \centering
- \begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{../figs/deaths_ppc.png}
- \caption{PPC on aggregate daily death counts.}
- \label{fig:deathsppc}
- \end{subfigure}%
- \begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{../figs/infections_ppc.png}
- \caption{Posterior on daily new infections.}
- \label{fig:casesppc}
- \end{subfigure}
- \caption{Posterior predictive check (PPC) on death counts and posterior distribution of infections.}
- \label{fig:ppcs}
- \end{figure}
-
- To check model fit, a posterior predictive check on aggregate death counts was performed.
+See Section~\ref{app:prior_modelbased} in the supplementary materials for details on prior specification, posterior predict checks, and discussion on case undercounting.
+
+ % \begin{figure}[!th]
+ % \centering
+ % \begin{subfigure}{.5\textwidth}
+ %  \centering
+ %  \includegraphics[width=.9\linewidth]{../figs/deaths_ppc.png}
+ % \caption{PPC on aggregate daily death counts.}
+ % \label{fig:deathsppc}
+ % \end{subfigure}%
+ % \begin{subfigure}{.5\textwidth}
+ %  \centering
+ %  \includegraphics[width=.9\linewidth]{../figs/infections_ppc.png}
+ % \caption{Posterior on daily new infections.}
+ % \label{fig:casesppc}
+ % \end{subfigure}
+ % \caption{Posterior predictive check (PPC) on death counts and posterior distribution of infections.}
+ % \label{fig:ppcs}
+ % \end{figure}
+
+ % To check model fit, a posterior predictive check on aggregate death counts was performed.
 Figure~\ref{fig:deathsppc} shows the results which suggests chosen SEIR model fits the observed data well.

### 2023-01-05 write-up/covid-supplement-writeup.tex
Shortening paper a bit more. almost there

@@
 \def\EE{\mathbb{E}}
+\def\new{\text{new}}
 %\newcommand{\indep}{\perp \!\!\! \perp}
@@
 
+% \subsubsection{Asymptotics}
+%  \label{section:asymptotics}
+
+%  Here we suppose there is a sequence of finite populations of size $N_{\nu}$ indexed by $\nu$ and that there are $L_\nu$ non-probability samples and probability samples of size $n$ drawn at equally spaced times $\{ t^\prime_l \}_{l=1}^{L_{\nu}}$ over the study window~$[0,T]$.
+Assuming correct selection model specification, then under the probability sample design and nonprobability sample propensities, Lemma~\ref{lemma:ipw} shows that the IPW estimator at a time~$t \in \{ t^\prime_l \}$ is consistent as $\nu$ goes to infinity (i.e., $N_{\nu}, L_{\nu} \to \infty$) and calculates the estimator's variance.
+In Lemma~\ref{lemma:ipw}, sensitivity/specificity are unknown and estimated using a pseudo-likelihood on two random samples whose sizes (denoted~$n_{FP}$ and $n_{FN}$) reflect uncertainty in these quantities.
+See Appendix~\ref{app:asympderivations} for additional details.
+
+ \begin{lemma}[Variance of IPW estimator] \normalfont
+ \label{lemma:ipw}
+ % \textcolor{red}{Must check.}
+The estimates~$\hat \mu_t := \bar y_{n,t}^\star$, $\hat \pi_{j,t}$, $\hat{FP}$, and $\hat{FN}$ at a time~$t \in \{t_l^\prime\}$ are solutions to the set of estimating equations~$\Phi_n(\eta_t)$ given in Appendix~\ref{app:asympderivations},
+ % $$
+ % \Phi_n (\eta_t) =
+ % \left (
+ % \begin{array}{c}
+ % \frac{1}{N} \sum_{j=1}^N I^{NR}_{j,t} \frac{Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t}{\pi_{j,t}} \\
+ % \frac{1}{N \times \sum_{l = 1}^L K_h \left(|t - t_l^\prime| \right)} \sum_{l = 1}^L K_h \left(|t - t_l^\prime| \right) \left[ \sum_{j=1}^N I^{NR}_{j,t_l^\prime} X_{j,t_l^\prime} - \sum_{j = 1}^N I^{R}_{j,t_l^\prime} W^{R}_{j,t_l^\prime}  \pi_{j,t_l^\prime} X_{j,t_l^\prime}  \right] \\
+ % \frac{1}{n_{FP}} \sum_{i=1}^{n_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
+ % \frac{1}{n_{FN}} \sum_{i=1}^{n_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
+ % \end{array}
+ % \right ) = {\bf 0}
+ % $$
+ where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
+Under certain regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (\bar n^{-1/2})$  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (\bar n^{-1})$ where $\bar n = n \sum_{l=1}^L K_{h} (|t - t_l^\prime|)$ and $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$ where $\phi_n = \frac{\partial \Phi (\eta)}{\partial \eta}$ with $E[\cdot]$ and $\text{Var} (\cdot)$ are under the joint randomization of propensity score and sampling designs.
+ \end{lemma}
+
 To prove consistency, first consider the setting where a single random sample of size $n$ is observed at time $t$.
@@
 
-We next state Lemma~\ref{lemma:dr} for the doubly-robust estimator presented in Section XX.  The proof follows the same derivation as above.
+\subsection{Doubly robust estimation.}
+\label{app:drestimation}
+ Rather than relying solely on IPW or epidemiological forecasts, here we combine forecasting and inverse-probability weighting by extending recent work by~\cite{Chen2019} to account time-varying propensities and measurement-error.
+ Let~$\hat \mu(x)$ denote the posterior mean of the active infection rate individuals with covariate value~$x$ based on the SEIR model described in Section~\ref{section:modelbased}.
+In this context, the doubly-robust estimator is given by
+ $$
+ \bar y_{n}^{(DR)} = \frac{1}{N} \sum_{j=1}^N \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j^{NR} w (x_j)} \sum_{j=1}^N I_j^{NR} w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \hat \mu(x_j) \right).
+ $$
+ where $\hat \mu(x_j)$ is not corrected for measurement-error as it estimates true active infections.
+This estimator is called ``doubly-robust'' because it is consistent if either the model-based forecasts or the time-varying propensities are correctly specified.
+A statistical error decomposition can be derived
+ \begin{align}
+ \label{eq:statdecomp3}
+ \begin{split}
+ \rho_{I^{NR} (X), Y-\mu(X)} &\times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y-\mu(X)}  \\
+ &\times \underbrace{\left[ 1 - \frac{\rho_{I^{NR}(X),Y} \sigma_{Y}}{\rho_{I^{NR} (X),Y-\mu(X)} \sigma_{Y-\mu(X)}} \times \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right]}_{\tilde D_M}.
+ \end{split}
+ \end{align}
+ See Appendix~\ref{app:DRderivation} for the derivation.
+ % Comparing~\eqref{eq:statdecomp3} to~\eqref{eq:statdecomp2} shows that the addition of the model-based forecast impacts the estimation error in two ways.
+ If the model is adequate, then one may expect a reduction in the problem difficulty and (potentially) in the data quality components.
+ % Note that the model-based forecast does not impact the data quantity component.
+% Second, if the model-based forecast is correct then $E_{\bf R} [ \rho_{\tilde I (X), Y-\mu(X)} ] = 0$  and therefore $\rho_{\tilde I(X), Y-\mu(X)} = O(N^{-1})$; however, if the forecasts are not correctly specified then the data quality index is unlikely to inversely scale with population size.
+ Interestingly, the impact of measurement-error on data quality now depends on a relative comparison of the data quality and problem difficulty of the weighted estimator and the doubly-robust estimator.
+ % For decent models, one would expect this ratio to be greater than one.
+Therefore, $\tilde \Delta < 0$ implies an expected increase in impact of measurement-error and $\tilde \Delta > 0$ implies the magnitude could shrink toward zero or change sign and have an even large magnitude in the opposite direction.
+Again, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from a DR estimator are likely to put too much confidence in the sheer size of the data.
+
+ Unfortunately, the first term of the doubly robust estimator cannot be computed as covariate information is not collected on the entire population.
+Here, we suppose the same asymptotic regime as in Section~\ref{section:asymptotics} and use the probability sample at the time~$t \in \{t_l^\prime\}$ to estimate this term:
+ $$
+ \frac{1}{\sum_{j=1}^N I^{R}_{j,t} W^R_{j,t}} \sum_{j=1}^N  I_{j,t}^{R} W_{j,t}^{R} \hat \mu_{t} (X_{j,t}) + \frac{1}{\sum_{j=1}^N I_{j,t}^{NR} w (X_{j,t})} \sum_{j=1}^N I_{j,t}^{NR} w(X_{j,t}) \left( \frac{Y_{j,t} - FP}{1 - FP - FN} - \hat \mu_t(X_{j,t}) \right).
+ $$
+ Assuming correct selection propensity model specification, then under the probability sample design and nonprobability sample design, Lemma~\ref{lemma:dr} shows that the doubly robust estimator is consistent and calculates the estimator's variance.
+The proof follows the same derivation as above.
 
@@
 \label{app:in_add_details}
+
+Table~\ref{tab:indiana} summarizes the statewide random molecular tests conducted in Indiana between April 25--29, 2020 as well as an additional non-random sample of 898 individuals during May 2--3, 2020.
+
+\begin{table}[!th]
+ \begin{tabular}{c | c | r r r | r | r r }
+ & & \multicolumn{3}{c}{Total Tests (\%)} & IN & \multicolumn{2}{c}{Positive Test Rate (\%)}\\
+ \cline{3-5} \cline{7-8}
+ & & NonRandom & Random & CTIS & Census & NonRandom & Random \\ \hline
+ \multirow{2}{*}{Sex} & Female & 58.2 & 55.0 & 54.0 & 50.7 & 21.7 (11.2) & 1.4 \\
+ & Male & 41.8 & 45.0  & 46.0 & 49.3 & 24.2 (12.4) & 2.1 \\ \hline
+ \multirow{3}{*}{Age} & $<40$ & 39.4 & 28.0  & 36.2 & 52.7
+ & 29.7 (15.0) & 1.7 \\
+ & $40-59$ & 41.1 & 36.0  & 34.3 & 25.2 & 24.9 (12.5) & 2.1 \\
+ & $\geq 60$ & 19.5 & 36.0  & 29.5 & 22.1 & 6.7 (3.4) & 0.9 \\ \hline
+ \multirow{2}{*}{Race} & White & 23.1 & 92.0  & - & 86.9 & 19.5 (9.6) & 1.5 \\
+ & Nonwhite & 76.9 & 8.0  & - & 13.1 & 25.0 (12.3) & 3.4 \\ \hline
+ % Hispanic & Hispanic & 44.1 & 2.0 & 7.1 & 37.6 & 6.9 \\
+ % Origin & Non-Hispanic & 55.9 & 98.0 & 92.9 & 13.0 & 1.3 \\ \hline
+ \multirow{2}{*}{Fever} & Yes & 17.0 & 1.8  & 1.0 & - & 66.4 (32.1) & 4.5 \\
+ & No & 83.0 & 98.2  & 99.0 & - & 15.6 (7.5) & 1.3 \\ \hline
+ Household & Yes & 10.8 & 1.4  & 1.8 & - & 46.1 (22.4) & 29.4 \\
+ $+$ Case & No & 89.2 & 98.6  & 98.2 & - & 21.6 (10.4) & 1.3 \\ \hline
+ Prior $+$ & Yes & 6.1 & 1.4  & - & - & 39.2 (20.2) & 24.4 \\
+ Test & No & 93.9 & 98.6  & - & - & 21.6 (11.1) & 1.3 \\ \hline
+ \end{tabular}
+ \caption{Estimated total tests ($\%$) and point prevalence of active infection with SARS-CoV-2 by demographics in Indiana~\citep{Yiannoutsos2021,doi:10.1073/pnas.2111454118}.
+NonRandom positive test rates in parentheses are adjusted rates to match the statewide rate of $11.7\%$ on August 30th.}
+ \label{tab:indiana}
+ \end{table}
 
@@
 
+ Figure~\ref{fig:fbsymptoms} displays smoothed estimates of the fraction of individuals who report COVID-like symptoms within the past 24-hours by age and gender.
+
+ \begin{figure}[!th]
+ \centering
+ \includegraphics[width = 0.9\textwidth]{../figs/fbcovid19symptoms.png}
+ \caption{Rate of reported COVID-19 symptoms per strata.
+Daily rates were estimated using weighted method suggested by~\citep{doi:10.1073/pnas.2111454118} on each day separately and then smoothed over time using local-linear nonparametric regression.}
+ \label{fig:fbsymptoms}
+ \vspace{-0.3cm}
+ \end{figure}
+
 Next, we describe the two mean-imputation methods.
 \subsection{Model 1}
+
+% In the first approach, we estimate two logistic regressions using weighted pseudo-likelihoods with the CTIS data.
+The first estimates the probability of contact with a COVID-19 positive individual given demographic and COVID-19 test status (positive or negative COVID-19 test).
+Figure~\ref{fig:contactlik1} and~\ref{fig:contactlik2} in Section~\ref{app:in_add_details} in the supplementary materials presents the estimated likelihood of contact with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.
+The second estimates the probability of fever given demographic, COVID-19 test status, \emph{and} whether the individual has had contact with a COVID-19 positive individual.
+Figure~\ref{fig:symptomlik1} and~\ref{fig:symptomlik2} in Section~\ref{app:in_add_details} in the supplementary materials presents the estimated likelihood of reporting a fever with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.
+We see that likelihood of fever varies greatly based on whether the individual also reported contact with a COVID-19 positive individual.
+Using these two models, mean imputation is used to calculate the number of positive and negative tests given demographic strata, fever status, and COVID-19 contact status.
+Inverse probability weights are computed based on the resulting dataset (termed IPW1).
 

### 2023-01-05 write-up/covid-writeup.tex
Shortening paper a bit more. almost there

@@
 
-\begin{remark}[Public versus government datasets]
-% In September 2020, the author submitted a data request to the Indiana Data Hub (IDH).
+% \begin{remark}[Public versus government datasets]
+% % In September 2020, the author submitted a data request to the Indiana Data Hub (IDH).
 Over the next three months, we worked with Indiana's COVID-19 response team to refine the request.
 In December 2020, the request was officially approved and data streams were made publicly available on the IDH.
-In this paper, we focus on \emph{publicly-available data}, i.e., data that anyone can download directly from official government data portals such as the Indiana Data Hub (IDH).
+% In this paper, we focus on \emph{publicly-available data}, i.e., data that anyone can download directly from official government data portals such as the Indiana Data Hub (IDH).
 While the scientific community has contributed through independent COVID-19 observational and clinical studies, a significant component of public policy guidance has relied on testing and case count data, e.g., CDC and state guidelines based on test positivity rates and relative changes in the case counts over time.
@@
 
-Note that Indiana's COVID-19 response team has access to official \emph{government data} which may not be publicly-available and is likely stored at the individual-level.
+% Note that Indiana's COVID-19 response team has access to official \emph{government data} which may not be publicly-available and is likely stored at the individual-level.
 Due to data privacy and legal concerns, some collected covariates may not be reported publicly and others are aggregated and reported marginally.
 While data analytic decisions discussed in Section~\ref{section:applications} were made due to access of publicly-available data, the overall data analysis framework is designed to be a tool for health departments to assess the pandemic and guide responses, e.g.,  account for individual-level covariate information, rather than publicly-reported strata-level information, that may include important covariates such as symptom status and recent COVID-19 contact.
- \end{remark}
+%  \end{remark}
 
@@
 During May 2--3, 2020, an additional non-random sample of 898 individuals was also collected.
- Table~\ref{tab:indiana} summarizes the data.
+Table~\ref{tab:indiana} in Appendix~\ref{app:in_add_details} summarizes the data.
  % In this paper, the nonrandom sample is used to assess whether the cumulative summary statistics are appropriate in constructing the inverse-probability weights as well as in constructing appropriate weights using symptom information.
 
- \begin{table}[!th]
- \begin{tabular}{c | c | r r r | r | r r }
- & & \multicolumn{3}{c}{Total Tests (\%)} & IN & \multicolumn{2}{c}{Positive Test Rate (\%)}\\
- \cline{3-5} \cline{7-8}
- & & NonRandom & Random & CTIS & Census & NonRandom & Random \\ \hline
- \multirow{2}{*}{Sex} & Female & 58.2 & 55.0 & 54.0 & 50.7 & 21.7 (11.2) & 1.4 \\
- & Male & 41.8 & 45.0  & 46.0 & 49.3 & 24.2 (12.4) & 2.1 \\ \hline
- \multirow{3}{*}{Age} & $<40$ & 39.4 & 28.0  & 36.2 & 52.7
- & 29.7 (15.0) & 1.7 \\
- & $40-59$ & 41.1 & 36.0  & 34.3 & 25.2 & 24.9 (12.5) & 2.1 \\
- & $\geq 60$ & 19.5 & 36.0  & 29.5 & 22.1 & 6.7 (3.4) & 0.9 \\ \hline
- \multirow{2}{*}{Race} & White & 23.1 & 92.0  & - & 86.9 & 19.5 (9.6) & 1.5 \\
- & Nonwhite & 76.9 & 8.0  & - & 13.1 & 25.0 (12.3) & 3.4 \\ \hline
- % Hispanic & Hispanic & 44.1 & 2.0 & 7.1 & 37.6 & 6.9 \\
- % Origin & Non-Hispanic & 55.9 & 98.0 & 92.9 & 13.0 & 1.3 \\ \hline
- \multirow{2}{*}{Fever} & Yes & 17.0 & 1.8  & 1.0 & - & 66.4 (32.1) & 4.5 \\
- & No & 83.0 & 98.2  & 99.0 & - & 15.6 (7.5) & 1.3 \\ \hline
- Household & Yes & 10.8 & 1.4  & 1.8 & - & 46.1 (22.4) & 29.4 \\
- $+$ Case & No & 89.2 & 98.6  & 98.2 & - & 21.6 (10.4) & 1.3 \\ \hline
- Prior $+$ & Yes & 6.1 & 1.4  & - & - & 39.2 (20.2) & 24.4 \\
- Test & No & 93.9 & 98.6  & - & - & 21.6 (11.1) & 1.3 \\ \hline
- \end{tabular}
- \caption{Estimated total tests ($\%$) and point prevalence of active infection with SARS-CoV-2 by demographics in Indiana~\citep{Yiannoutsos2021,doi:10.1073/pnas.2111454118}.
+ % \begin{table}[!th]
+ % \begin{tabular}{c | c | r r r | r | r r }
+ % & & \multicolumn{3}{c}{Total Tests (\%)} & IN & \multicolumn{2}{c}{Positive Test Rate (\%)}\\
+ % \cline{3-5} \cline{7-8}
+ % & & NonRandom & Random & CTIS & Census & NonRandom & Random \\ \hline
+ % \multirow{2}{*}{Sex} & Female & 58.2 & 55.0 & 54.0 & 50.7 & 21.7 (11.2) & 1.4 \\
+ % & Male & 41.8 & 45.0  & 46.0 & 49.3 & 24.2 (12.4) & 2.1 \\ \hline
+ % \multirow{3}{*}{Age} & $<40$ & 39.4 & 28.0  & 36.2 & 52.7
+ % & 29.7 (15.0) & 1.7 \\
+ % & $40-59$ & 41.1 & 36.0  & 34.3 & 25.2 & 24.9 (12.5) & 2.1 \\
+ % & $\geq 60$ & 19.5 & 36.0  & 29.5 & 22.1 & 6.7 (3.4) & 0.9 \\ \hline
+ % \multirow{2}{*}{Race} & White & 23.1 & 92.0  & - & 86.9 & 19.5 (9.6) & 1.5 \\
+ % & Nonwhite & 76.9 & 8.0  & - & 13.1 & 25.0 (12.3) & 3.4 \\ \hline
+ % % Hispanic & Hispanic & 44.1 & 2.0 & 7.1 & 37.6 & 6.9 \\
+ % % Origin & Non-Hispanic & 55.9 & 98.0 & 92.9 & 13.0 & 1.3 \\ \hline
+ % \multirow{2}{*}{Fever} & Yes & 17.0 & 1.8  & 1.0 & - & 66.4 (32.1) & 4.5 \\
+ % & No & 83.0 & 98.2  & 99.0 & - & 15.6 (7.5) & 1.3 \\ \hline
+ % Household & Yes & 10.8 & 1.4  & 1.8 & - & 46.1 (22.4) & 29.4 \\
+ % $+$ Case & No & 89.2 & 98.6  & 98.2 & - & 21.6 (10.4) & 1.3 \\ \hline
+ % Prior $+$ & Yes & 6.1 & 1.4  & - & - & 39.2 (20.2) & 24.4 \\
+ % Test & No & 93.9 & 98.6  & - & - & 21.6 (11.1) & 1.3 \\ \hline
+ % \end{tabular}
+ % \caption{Estimated total tests ($\%$) and point prevalence of active infection with SARS-CoV-2 by demographics in Indiana~\citep{Yiannoutsos2021,doi:10.1073/pnas.2111454118}.
 NonRandom positive test rates in parentheses are adjusted rates to match the statewide rate of $11.7\%$ on August 30th.}
- \label{tab:indiana}
- \end{table}
+ % \label{tab:indiana}
+ % \end{table}
 
@@
 The survey defines an individual as displaying \emph{COVID-like symptoms} if they exhibit a fever along with a cough, or shortness of breath, or difficulty breathing.
-Figure~\ref{fig:fbsymptoms} displays smoothed estimates of the fraction of individuals who report COVID-like symptoms within the past 24-hours by age and gender.
+Figure~\ref{fig:fbsymptoms} in Appendix~\ref{app:in_add_details} displays smoothed estimates of the fraction of individuals who report COVID-like symptoms within the past 24-hours by age and gender.
 The Delphi's COVID-19 Trends and Impact Survey (CTIS) is used as the main source of auxiliary information on time-varying characteristics of the population of Indiana, e.g., displaying COVID-like symptoms or in contact with COVID-19 positive individuals.
 
- \begin{figure}[!th]
- \centering
- \includegraphics[width = 0.9\textwidth]{../figs/fbcovid19symptoms.png}
- \caption{Rate of reported COVID-19 symptoms per strata.
+ % \begin{figure}[!th]
+ % \centering
+ % \includegraphics[width = 0.9\textwidth]{../figs/fbcovid19symptoms.png}
+ % \caption{Rate of reported COVID-19 symptoms per strata.
 Daily rates were estimated using weighted method suggested by~\citep{doi:10.1073/pnas.2111454118} on each day separately and then smoothed over time using local-linear nonparametric regression.}
- \label{fig:fbsymptoms}
- \vspace{-0.3cm}
- \end{figure}
+ % \label{fig:fbsymptoms}
+ % \vspace{-0.3cm}
+ % \end{figure}
 
@@
 
- \subsubsection{Asymptotics}
- \label{section:asymptotics}
-
- Here we suppose there is a sequence of finite populations of size $N_{\nu}$ indexed by $\nu$ and that there are $L_\nu$ non-probability samples and probability samples of size $n$ drawn at equally spaced times $\{ t^\prime_l \}_{l=1}^{L_{\nu}}$ over the study window~$[0,T]$.
+%  \subsubsection{Asymptotics}
+%  \label{section:asymptotics}
+
+%  Here we suppose there is a sequence of finite populations of size $N_{\nu}$ indexed by $\nu$ and that there are $L_\nu$ non-probability samples and probability samples of size $n$ drawn at equally spaced times $\{ t^\prime_l \}_{l=1}^{L_{\nu}}$ over the study window~$[0,T]$.
 Assuming correct selection model specification, then under the probability sample design and nonprobability sample propensities, Lemma~\ref{lemma:ipw} shows that the IPW estimator at a time~$t \in \{ t^\prime_l \}$ is consistent as $\nu$ goes to infinity (i.e., $N_{\nu}, L_{\nu} \to \infty$) and calculates the estimator's variance.
@@
 
- \begin{lemma}[Variance of IPW estimator] \normalfont
- \label{lemma:ipw}
- % \textcolor{red}{Must check.}
-The estimates~$\hat \mu_t := \bar y_{n,t}^\star$, $\hat \pi_{j,t}$, $\hat{FP}$, and $\hat{FN}$ at a time~$t \in \{t_l^\prime\}$ are solutions to the set of estimating equations~$\Phi_n(\eta_t)$ given in Appendix~\ref{app:asympderivations},
- % $$
- % \Phi_n (\eta_t) =
- % \left (
- % \begin{array}{c}
- % \frac{1}{N} \sum_{j=1}^N I^{NR}_{j,t} \frac{Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t}{\pi_{j,t}} \\
- % \frac{1}{N \times \sum_{l = 1}^L K_h \left(|t - t_l^\prime| \right)} \sum_{l = 1}^L K_h \left(|t - t_l^\prime| \right) \left[ \sum_{j=1}^N I^{NR}_{j,t_l^\prime} X_{j,t_l^\prime} - \sum_{j = 1}^N I^{R}_{j,t_l^\prime} W^{R}_{j,t_l^\prime}  \pi_{j,t_l^\prime} X_{j,t_l^\prime}  \right] \\
- % \frac{1}{n_{FP}} \sum_{i=1}^{n_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
- % \frac{1}{n_{FN}} \sum_{i=1}^{n_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
- % \end{array}
- % \right ) = {\bf 0}
- % $$
- where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
+%  \begin{lemma}[Variance of IPW estimator] \normalfont
+%  \label{lemma:ipw}
+%  % \textcolor{red}{Must check.}
+% The estimates~$\hat \mu_t := \bar y_{n,t}^\star$, $\hat \pi_{j,t}$, $\hat{FP}$, and $\hat{FN}$ at a time~$t \in \{t_l^\prime\}$ are solutions to the set of estimating equations~$\Phi_n(\eta_t)$ given in Appendix~\ref{app:asympderivations},
+%  % $$
+%  % \Phi_n (\eta_t) =
+%  % \left (
+%  % \begin{array}{c}
+%  % \frac{1}{N} \sum_{j=1}^N I^{NR}_{j,t} \frac{Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t}{\pi_{j,t}} \\
+%  % \frac{1}{N \times \sum_{l = 1}^L K_h \left(|t - t_l^\prime| \right)} \sum_{l = 1}^L K_h \left(|t - t_l^\prime| \right) \left[ \sum_{j=1}^N I^{NR}_{j,t_l^\prime} X_{j,t_l^\prime} - \sum_{j = 1}^N I^{R}_{j,t_l^\prime} W^{R}_{j,t_l^\prime}  \pi_{j,t_l^\prime} X_{j,t_l^\prime}  \right] \\
+%  % \frac{1}{n_{FP}} \sum_{i=1}^{n_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
+%  % \frac{1}{n_{FN}} \sum_{i=1}^{n_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
+%  % \end{array}
+%  % \right ) = {\bf 0}
+%  % $$
+%  where $\eta_t = (\mu_t, \pi_t, FP, FN)$.
 Under certain regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (\bar n^{-1/2})$  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (\bar n^{-1})$ where $\bar n = n \sum_{l=1}^L K_{h} (|t - t_l^\prime|)$ and $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$ where $\phi_n = \frac{\partial \Phi (\eta)}{\partial \eta}$ with $E[\cdot]$ and $\text{Var} (\cdot)$ are under the joint randomization of propensity score and sampling designs.
- \end{lemma}
+%  \end{lemma}
 
@@
 
- \subsection{Doubly robust estimation.}
- Rather than relying solely on IPW or epidemiological forecasts, here we combine forecasting and inverse-probability weighting by extending recent work by~\cite{Chen2019} to account time-varying propensities and measurement-error.
- Let~$\hat \mu(x)$ denote the posterior mean of the active infection rate individuals with covariate value~$x$ based on the SEIR model described in Section~\ref{section:modelbased}.
+ \begin{remark}[Doubly robust estimation]
+Rather than relying solely on IPW or epidemiological forecasts, we combine forecasting and inverse-probability weighting by extending recent work by~\cite{Chen2019} to account time-varying propensities and measurement-error.
+The doubly-robust estimator is presented in Appendix~\ref{app:drestimation}.
+ \end{remark}
+
+%  \subsection{Doubly robust estimation.}
+%  Rather than relying solely on IPW or epidemiological forecasts, here we combine forecasting and inverse-probability weighting by extending recent work by~\cite{Chen2019} to account time-varying propensities and measurement-error.
+%  Let~$\hat \mu(x)$ denote the posterior mean of the active infection rate individuals with covariate value~$x$ based on the SEIR model described in Section~\ref{section:modelbased}.
 In this context, the doubly-robust estimator is given by
- $$
- \bar y_{n}^{(DR)} = \frac{1}{N} \sum_{j=1}^N \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j^{NR} w (x_j)} \sum_{j=1}^N I_j^{NR} w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \hat \mu(x_j) \right).
- $$
- where $\hat \mu(x_j)$ is not corrected for measurement-error as it estimates true active infections.
+%  $$
+%  \bar y_{n}^{(DR)} = \frac{1}{N} \sum_{j=1}^N \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j^{NR} w (x_j)} \sum_{j=1}^N I_j^{NR} w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \hat \mu(x_j) \right).
+%  $$
+%  where $\hat \mu(x_j)$ is not corrected for measurement-error as it estimates true active infections.
 This estimator is called ``doubly-robust'' because it is consistent if either the model-based forecasts or the time-varying propensities are correctly specified.
 A statistical error decomposition can be derived
- \begin{align}
- \label{eq:statdecomp3}
- \begin{split}
- \rho_{I^{NR} (X), Y-\mu(X)} &\times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y-\mu(X)}  \\
- &\times \underbrace{\left[ 1 - \frac{\rho_{I^{NR}(X),Y} \sigma_{Y}}{\rho_{I^{NR} (X),Y-\mu(X)} \sigma_{Y-\mu(X)}} \times \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right]}_{\tilde D_M}.
- \end{split}
- \end{align}
- See Appendix~\ref{app:DRderivation} for the derivation.
- % Comparing~\eqref{eq:statdecomp3} to~\eqref{eq:statdecomp2} shows that the addition of the model-based forecast impacts the estimation error in two ways.
- If the model is adequate, then one may expect a reduction in the problem difficulty and (potentially) in the data quality components.
- % Note that the model-based forecast does not impact the data quantity component.
-% Second, if the model-based forecast is correct then $E_{\bf R} [ \rho_{\tilde I (X), Y-\mu(X)} ] = 0$  and therefore $\rho_{\tilde I(X), Y-\mu(X)} = O(N^{-1})$; however, if the forecasts are not correctly specified then the data quality index is unlikely to inversely scale with population size.
- Interestingly, the impact of measurement-error on data quality now depends on a relative comparison of the data quality and problem difficulty of the weighted estimator and the doubly-robust estimator.
- % For decent models, one would expect this ratio to be greater than one.
+%  \begin{align}
+%  \label{eq:statdecomp3}
+%  \begin{split}
+%  \rho_{I^{NR} (X), Y-\mu(X)} &\times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y-\mu(X)}  \\
+%  &\times \underbrace{\left[ 1 - \frac{\rho_{I^{NR}(X),Y} \sigma_{Y}}{\rho_{I^{NR} (X),Y-\mu(X)} \sigma_{Y-\mu(X)}} \times \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right]}_{\tilde D_M}.
+%  \end{split}
+%  \end{align}
+%  See Appendix~\ref{app:DRderivation} for the derivation.
+%  % Comparing~\eqref{eq:statdecomp3} to~\eqref{eq:statdecomp2} shows that the addition of the model-based forecast impacts the estimation error in two ways.
+%  If the model is adequate, then one may expect a reduction in the problem difficulty and (potentially) in the data quality components.
+%  % Note that the model-based forecast does not impact the data quantity component.
+% % Second, if the model-based forecast is correct then $E_{\bf R} [ \rho_{\tilde I (X), Y-\mu(X)} ] = 0$  and therefore $\rho_{\tilde I(X), Y-\mu(X)} = O(N^{-1})$; however, if the forecasts are not correctly specified then the data quality index is unlikely to inversely scale with population size.
+%  Interestingly, the impact of measurement-error on data quality now depends on a relative comparison of the data quality and problem difficulty of the weighted estimator and the doubly-robust estimator.
+%  % For decent models, one would expect this ratio to be greater than one.
 Therefore, $\tilde \Delta < 0$ implies an expected increase in impact of measurement-error and $\tilde \Delta > 0$ implies the magnitude could shrink toward zero or change sign and have an even large magnitude in the opposite direction.
@@
 
- Unfortunately, the first term of the doubly robust estimator cannot be computed as covariate information is not collected on the entire population.
+%  Unfortunately, the first term of the doubly robust estimator cannot be computed as covariate information is not collected on the entire population.
 Here, we suppose the same asymptotic regime as in Section~\ref{section:asymptotics} and use the probability sample at the time~$t \in \{t_l^\prime\}$ to estimate this term:
- $$
- \frac{1}{\sum_{j=1}^N I^{R}_{j,t} W^R_{j,t}} \sum_{j=1}^N  I_{j,t}^{R} W_{j,t}^{R} \hat \mu_{t} (X_{j,t}) + \frac{1}{\sum_{j=1}^N I_{j,t}^{NR} w (X_{j,t})} \sum_{j=1}^N I_{j,t}^{NR} w(X_{j,t}) \left( \frac{Y_{j,t} - FP}{1 - FP - FN} - \hat \mu_t(X_{j,t}) \right).
- $$
- Assuming correct selection propensity model specification, then under the probability sample design and nonprobability sample design, Lemma~\ref{lemma:dr} in Appendix~\ref{app:asympderivations} shows that the doubly robust estimator is consistent and calculates the estimator's variance.
+%  $$
+%  \frac{1}{\sum_{j=1}^N I^{R}_{j,t} W^R_{j,t}} \sum_{j=1}^N  I_{j,t}^{R} W_{j,t}^{R} \hat \mu_{t} (X_{j,t}) + \frac{1}{\sum_{j=1}^N I_{j,t}^{NR} w (X_{j,t})} \sum_{j=1}^N I_{j,t}^{NR} w(X_{j,t}) \left( \frac{Y_{j,t} - FP}{1 - FP - FN} - \hat \mu_t(X_{j,t}) \right).
+%  $$
+%  Assuming correct selection propensity model specification, then under the probability sample design and nonprobability sample design, Lemma~\ref{lemma:dr} in Appendix~\ref{app:asympderivations} shows that the doubly robust estimator is consistent and calculates the estimator's variance.
 
@@
 
- We next consider estimation of the active infection rate in Indiana using unweighted, IPW, model-based, and doubly robust estimates. To start, we recap the data sources used in estimation:
- \begin{itemize}[leftmargin=*]
- \item {\bf Testing data}:  The number of daily tests performed and number of daily positive tests are reported.
+ We next consider estimation of the active infection rate in Indiana using unweighted, IPW, model-based, and doubly robust estimates based on the combination of testing data, random/nonrandom statewide samples, CTIS survey data, and hospitalization records previously discussed in Section~\ref{section:data}.
+ % To start, we recap the data sources used in estimation:
+ % \begin{itemize}[leftmargin=*]
+ % \item {\bf Testing data}:  The number of daily tests performed and number of daily positive tests are reported.
 These counts are broken out jointly by age, gender, and racial demographic information.
@@
 
- \item {\bf Random/Nonrandom statewide sample}: Table~\ref{tab:indiana} summarizes data from a random sample from April 25--29th as well as a nonrandom sample obtained between May 2--3 in racial/ethnic minority communities~\citep{Yiannoutsos2021}.
+ % \item {\bf Random/Nonrandom statewide sample}: Table~\ref{tab:indiana} summarizes data from a random sample from April 25--29th as well as a nonrandom sample obtained between May 2--3 in racial/ethnic minority communities~\citep{Yiannoutsos2021}.
 The nonrandom sample is not a random subsample of the overall case-counts; however, this data provides important supplementary covariate information.
- % Namely, the samples collected included covariate information on (1) prior symptoms, (2) COVID-19 positive individual in household, and (3) prior COVID-19 positive test.
- To account for the nonrandom sample being from high risk areas, we adjust estimates to match the statewide positivity rate of 11.7\% on August 30th.
+ % % Namely, the samples collected included covariate information on (1) prior symptoms, (2) COVID-19 positive individual in household, and (3) prior COVID-19 positive test.
+ % To account for the nonrandom sample being from high risk areas, we adjust estimates to match the statewide positivity rate of 11.7\% on August 30th.
 These adjusted rates are presented in parentheses in Table~\ref{tab:indiana} along with the relevant covariate information.
 
- \item {\bf Delphi's COVID-19 Trends and Impact Survey}: From the daily symptom surveys completed as part of Delphi's CTIS (see Section~\ref{subsection:fbsymptom} for details), we extract survey responses for individuals who identify as living in Indiana.
+ % \item {\bf Delphi's COVID-19 Trends and Impact Survey}: From the daily symptom surveys completed as part of Delphi's CTIS (see Section~\ref{subsection:fbsymptom} for details), we extract survey responses for individuals who identify as living in Indiana.
 We collect age, sex, and demographic information as well as COVID-19 related symptoms, which Delphi defines as having (1) a fever and cough, (2) shortness of breath, \emph{or} (3) difficulty breathing.
@@
 
- % \begin{table}[!th]
- % \centering
- % \begin{tabular}{c | c c c | c}
- % & \multicolumn{3}{c}{Facebook} & Random \\ \cline{2-4}
- % & Lower CI & Estimate & Upper CI & Estimate \\ \hline
- % Fever &  0.017 & 0.020 & 0.024 & 0.018 \\
- % Cough &  0.142 & 0.148 & 0.154 & 0.149 \\
- % Shortness & 0.058 & 0.060 & 0.062 & 0.062 \\ \hline
- % \end{tabular}
- % \caption{Comparison of Facebook survey and random sample}
- % \label{tab:comparison}
- % \end{table}
-
- % \item {\bf Hospitalization data}: On each day, we observe the number of individuals who are COVID-19 positive that visited a hospital on that day.
+ % % \begin{table}[!th]
+ % % \centering
+ % % \begin{tabular}{c | c c c | c}
+ % % & \multicolumn{3}{c}{Facebook} & Random \\ \cline{2-4}
+ % % & Lower CI & Estimate & Upper CI & Estimate \\ \hline
+ % % Fever &  0.017 & 0.020 & 0.024 & 0.018 \\
+ % % Cough &  0.142 & 0.148 & 0.154 & 0.149 \\
+ % % Shortness & 0.058 & 0.060 & 0.062 & 0.062 \\ \hline
+ % % \end{tabular}
+ % % \caption{Comparison of Facebook survey and random sample}
+ % % \label{tab:comparison}
+ % % \end{table}
+
+ % % \item {\bf Hospitalization data}: On each day, we observe the number of individuals who are COVID-19 positive that visited a hospital on that day.
 Unfortunately, hospitalization data is not broken out by age, gender, and racial demographic information.
@@
 
- \item {\bf Death data}: Daily COVID-19 related deaths are observed by age, gender, and racial demographics.
+ % \item {\bf Death data}: Daily COVID-19 related deaths are observed by age, gender, and racial demographics.
 Figure~\ref{fig:in-deaths} plots COVID-19 related deaths over time by age group.
 While most tests and positive cases are within the younger age strata, most deaths are within the 70+ age strata.
- \end{itemize}
+ % \end{itemize}
 
@@
 Symptom status (e.g., fever, coughing, shortness of breath) and related covariate information (COVID-19 positive contact), however, are currently unavailable, which is likely due to data privacy considerations.
-As COVID-19 symptom status is likely to alter the testing propensity and likelihood of active infection, for illustrative purposes CTIS and hospitalization data are used in two ways to impute symptom status per strata.
-We outline the imputation methods below; see Section~\ref{section:discussion} for additional discussion.
-
- In the first approach, we estimate two logistic regressions using weighted pseudo-likelihoods with the CTIS data.
+As COVID-19 symptom status is likely to alter the testing propensity and likelihood of active infection, for illustrative purposes CTIS and hospitalization data are used in two ways (termed IPW1 and IPW2 respectively) to impute symptom status per strata.
+We present IPW2 below; see Section~\ref{section:discussion} for details on IPW1 and additional discussion.
+
+ % In the first approach, we estimate two logistic regressions using weighted pseudo-likelihoods with the CTIS data.
 The first estimates the probability of contact with a COVID-19 positive individual given demographic and COVID-19 test status (positive or negative COVID-19 test).
@@
 
- In the second approach, we leverage hospitalization information.
-We first estimate the probability of fever given demographic and COVID-19 test status.
+ IPW2 starts by estimating the probability of fever given demographic and COVID-19 test status.
 When the COVID-19 test status is positive, the probability of fever is allowed to depend on whether the individual was hospitalized.
@@
 For example, symptom status and COVID-19 contact are clearly relevant factors in testing selection.
-Future work may consider how to include/exclude factors over time that are not significant to limit citizen reporting burden.
+Future work may consider how to include/exclude factors over time that are not significant to limit reporting burden.
 Moreover, equation~\eqref{eq:auxinfoprob} presupposes covariate information is collected for all individuals in the nonprobability sample, which is often not feasible for state/local governments where rapid sample collection is prioritized.

### 2023-01-09 write-up/covid-supplement-writeup.tex
fixing links

@@
 
-Table~\ref{tab:indiana} summarizes the statewide random molecular tests conducted in Indiana between April 25--29, 2020 as well as an additional non-random sample of 898 individuals during May 2--3, 2020.
+Table~\ref{app:indianatable} summarizes the statewide random molecular tests conducted in Indiana between April 25--29, 2020 as well as an additional non-random sample of 898 individuals during May 2--3, 2020.
 
@@
 NonRandom positive test rates in parentheses are adjusted rates to match the statewide rate of $11.7\%$ on August 30th.}
- \label{tab:indiana}
+ \label{app:indianatable}
  \end{table}

### 2023-01-09 write-up/covid-writeup.tex
fixing links

@@
  Due to testing restrictions and other potential selection biases, publicly reported COVID-19 case count data may not be sufficient to understand active infection rates or the disease trajectory.
-Here, we discuss two random samples that provide auxiliary information and may help address selection bias.
+Here, we discuss two samples that provide auxiliary information and may help address selection bias.
 
@@
 During May 2--3, 2020, an additional non-random sample of 898 individuals was also collected.
-Table~\ref{tab:indiana} in Appendix~\ref{app:in_add_details} summarizes the data.
+Table 1 in Appendix~\ref{app:in_add_details} summarizes the data.
  % In this paper, the nonrandom sample is used to assess whether the cumulative summary statistics are appropriate in constructing the inverse-probability weights as well as in constructing appropriate weights using symptom information.
@@
 
- We start by using Table~\ref{tab:indiana} to compute IPW weights for end of April, early May.
+ We start by using Table~\ref{app:indianatable} to compute IPW weights for end of April, early May.
 The random sample is $n=3,658$.
@@
 Precision in reported case-count data gives the illusion of information when what what is needed is quantification of uncertainty.
-Extensions of recent statistical error decompositions~\citep{Meng2018} demonstrate how selection bias leads data analysts to feel certain about incorrect conclusions.
+%Extensions of recent statistical error decompositions~\citep{Meng2018} demonstrate how selection bias leads data analysts to feel certain about incorrect conclusions.
 As case-count data is routinely used for public health policy making, we presented an inverse-probability weighting method and a doubly-robust estimation method that leverage auxiliary information collected through random samples to overcome these issues.
@@
 % \textcolor{red}{
- While we emphasize the importance of random sampling, we do not view it as a panacea.
+% While we emphasize the importance of random sampling, we do not view it as a panacea.
 With access to \emph{only} a non-probabilistic sample, we cannot address self-selection bias.
@@
 Absent some type of random sampling, the best route forward for all data analyses is careful associated sensitivity analyses and humility in data-driven conclusions.
-\cite{Ironse2103272118}, for example, use the Indiana seroprevalence random survey and to anchor their analysis; however, infrequent random surveys imply time-varying aspects such as the IFR may not be resolved by use of a single anchor.
+%\cite{Ironse2103272118}, for example, use the Indiana seroprevalence random survey and to anchor their analysis; however, infrequent random surveys imply time-varying aspects such as the IFR may not be resolved by use of a single anchor.
 
@@
 Random samples over time to supplement this data with auxiliary information is desperately needed.
-We acknowledge that this proposal will lead to data privacy concerns that will need to be addressed.
+We acknowledge that this proposal will lead to data privacy concerns as well as adjustments to the proposed method to account for the time-consuming and costly nature of data collection will both need to be addressed.
 
@@
 
- A valid criticism of the proposed approach is that recording covariates on every tested individual is time-consuming and costly.
+% A valid criticism of the proposed approach is that recording covariates on every tested individual is time-consuming and costly.
 We argue that the list of relevant factors is of reasonable length.
@@
 One can easily extend~\eqref{eq:auxinfoprob} by randomly sampling a subset of the nonprobability sample on which to collect the covariate information.
- This would balance goals of rapid testing and auxiliary data collection.
+% This would balance goals of rapid testing and auxiliary data collection.
 

### 2023-01-09 write-up/covid-supplement-writeup.tex
making changes and removing old commented out parts

@@
  Unfortunately, the first term of the doubly robust estimator cannot be computed as covariate information is not collected on the entire population.
-Here, we suppose the same asymptotic regime as in Section~\ref{section:asymptotics} and use the probability sample at the time~$t \in \{t_l^\prime\}$ to estimate this term:
+Here, we suppose the same asymptotic regime as in Section~\ref{app:asympderivations} and use the probability sample at the time~$t \in \{t_l^\prime\}$ to estimate this term:
  $$
@@
 Next, we describe the two mean-imputation methods.
-\subsection{Model 1}
+\subsection{Model 1 (IPW1)}
 
@@
 
-\subsection{Model 2 (Hospitalization)}
+\subsection{Model 2 (IPW2)}
 

### 2023-01-09 write-up/covid-writeup.tex
making changes and removing old commented out parts

@@
 During May 2--3, 2020, an additional non-random sample of 898 individuals was also collected.
-Table 1 in Appendix~\ref{app:in_add_details} summarizes the data.
+Table~\ref{app:indianatable} in Section~\ref{app:in_add_details} of the supplementary materials summarizes the data.
  % In this paper, the nonrandom sample is used to assess whether the cumulative summary statistics are appropriate in constructing the inverse-probability weights as well as in constructing appropriate weights using symptom information.
@@
 The survey defines an individual as displaying \emph{COVID-like symptoms} if they exhibit a fever along with a cough, or shortness of breath, or difficulty breathing.
-Figure~\ref{fig:fbsymptoms} in Appendix~\ref{app:in_add_details} displays smoothed estimates of the fraction of individuals who report COVID-like symptoms within the past 24-hours by age and gender.
+Figure~\ref{fig:fbsymptoms} in Section~\ref{app:in_add_details} of the supplementary materials displays smoothed estimates of the fraction of individuals who report COVID-like symptoms within the past 24-hours by age and gender.
 The Delphi's COVID-19 Trends and Impact Survey (CTIS) is used as the main source of auxiliary information on time-varying characteristics of the population of Indiana, e.g., displaying COVID-like symptoms or in contact with COVID-19 positive individuals.
@@
 The attentive data analyst will recognize the estimator $\bar y_n$ is now biased for simple random samples.
-In Appendix~\ref{app:memestimator}, assuming sensitivity and specificity are known a priori, a novel iterative procedure is used to construct the estimator $\tilde y_n = (\bar y_n - FP)/(1-(FP+FN))$, which is unbiased under simple random sampling (SRS); see Appendix~\ref{app:modelbased} for a discussion of the connection to model-based estimators.
+In Section~\ref{app:memestimator} of the supplementary materials, assuming sensitivity and specificity are known a priori, a novel iterative procedure is used to construct the estimator $\tilde y_n = (\bar y_n - FP)/(1-(FP+FN))$, which is unbiased under simple random sampling (SRS); see Section~\ref{app:modelbased} in the Supplementary Materials for a discussion of the connection to model-based estimators.
 In the language of the epidemiology literature, the estimator $\tilde y_n$ is the standard estimator for correcting for measurement error using false positive and negative rates that have been estimated from a validation sample, where $\bar y_n$ is the mis-measured average \& $\tilde y_n$ is the corrected average.
@@
  where $f=n/N$ is the sampling fraction, $f_1$ and $f_0$ are sampling fractions for COVID-19 positive and negative individuals respectively, and $\Delta = f_1 - f_0$ is the sampling rate differential.
-See Appendix~\ref{app:memestimator} for the derivation.
+See Section~\ref{app:memestimator} of the Supplementary Materials for the derivation.
 This extends work by \cite{Meng2018} to account for imperfect testing.
@@
  % % Thus, conclusions drawn from observed case-counts may not just be wrong, but very wrong.
- A key question is ``What is the (effective) sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' In Appendix~\ref{app:effss}, we show the effective sample size $n_{eff}$ can be bounded by $\frac{f}{1-f} \times \frac{1}{E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right]}$.
+ A key question is ``What is the (effective) sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' In Section~\ref{app:effss} of the Supplementary Materials, we show the effective sample size $n_{eff}$ can be bounded by $\frac{f}{1-f} \times \frac{1}{E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right]}$.
 Between April 25th to 29th 2020, Indiana performed $95,879$ tests.
@@
  where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement error adjustment, and $CV(Y_j) = \sigma_{Y_j}/ \bar Y_j$ is the coefficient of variation on day $j$.
-See Appendix~\ref{app:ratio} for the derivation.
+See Section~\ref{app:ratio} of the Supplementary Materials for the derivation.
 The error magnitude depends on the true rate $\bar Y_{t} / \bar Y_{t-1}$ so a large decrease will have a small error relative to a large increase.
@@
  where $CV_W$ is the coefficient of variation (i.e., standard deviation/mean) of $w^{NR} (X_J)$ given $I_J = 1$, $\rho_{I^{NR} (X), Y}$ is the empirical correlation which here depends on covariate distribution, $\tilde f_k = E[ w^{NR} (X_J) I_J \mid Y_J = k]$ for $k=0,1$, and $\tilde \Delta = \tilde f_1 - \tilde f_0$.
-See Appendix~\ref{app:ipwderivation} for the derivation.
+See Section~\ref{app:ipwderivation} of the Supplementary Materials for the derivation.
 
@@
 
- As demonstrated below in Lemma~\ref{lemma:ipw}, if the propensity model is correctly specified then the $E [ \rho_{I^{NR} (X), Y} ] = 0$ and therefore $\rho_{I^{NR} (X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then the data quality index is unlikely to inversely scale with population size.
+ If the propensity model is correctly specified then the $E [ \rho_{I^{NR} (X), Y} ] = 0$ and therefore $\rho_{I^{NR} (X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then the data quality index is unlikely to inversely scale with population size.
  % Indeed, Horvitz-Thompson estimators are sensitive to errors in the estimated weights.
@@
 Rather than relying solely on IPW or epidemiological forecasts, we combine forecasting and inverse-probability weighting by extending recent work by~\cite{Chen2019} to account time-varying propensities and measurement-error.
-The doubly-robust estimator is presented in Appendix~\ref{app:drestimation}.
+Given the overlap with Section~\ref{section:IPWerrordecomp}, for conciseness the doubly-robust estimator is presented in detail in Section~\ref{app:drestimation} of the Supplementary Materials.
  \end{remark}
@@
 As COVID-19 symptom status is likely to alter the testing propensity and likelihood of active infection, for illustrative purposes CTIS and hospitalization data are used in two ways (termed IPW1 and IPW2 respectively) to impute symptom status per strata.
-We present IPW2 below; see Section~\ref{section:discussion} for details on IPW1 and additional discussion.
+We present IPW2 below; see Section~\ref{app:in_add_details} in  of the Supplementary Materials for details on IPW1 and additional discussion.
 

### 2023-01-09 write-up/covid-writeup.tex
ready

@@
 Each weight~$w_t^{(k)} = 0$ for $t < t_k$ and $w_t^{(k)} \to 1$ as $t \to \infty$.
-See Section~\ref{app:prior_modelbased} in the supplementary materials for details on prior specification, posterior predict checks, and discussion on case undercounting.
+See Section~\ref{app:prior_modelbased} in the supplementary materials for details on prior specification, posterior predictive checks, and discussion on case undercounting.
 
@@
 Random samples over time to supplement this data with auxiliary information is desperately needed.
-We acknowledge that this proposal will lead to data privacy concerns as well as adjustments to the proposed method to account for the time-consuming and costly nature of data collection will both need to be addressed.
+We acknowledge that this proposal will lead to data privacy concerns that will need to be addressed as well as adjustments to the proposed method to account for the time-consuming and costly nature of sample collection.
 

### 2023-01-10 write-up/covid-writeup.tex
Finishing paper once and for all...ha

@@
 During May 2--3, 2020, an additional non-random sample of 898 individuals was also collected.
-Table~\ref{app:indianatable} in Section~\ref{app:in_add_details} of the supplementary materials summarizes the data.
+Table~\ref{app:indianatable} in Section~\ref{app:in_add_details} of the Supplementary Materials summarizes the data.
  % In this paper, the nonrandom sample is used to assess whether the cumulative summary statistics are appropriate in constructing the inverse-probability weights as well as in constructing appropriate weights using symptom information.
@@
 The survey defines an individual as displaying \emph{COVID-like symptoms} if they exhibit a fever along with a cough, or shortness of breath, or difficulty breathing.
-Figure~\ref{fig:fbsymptoms} in Section~\ref{app:in_add_details} of the supplementary materials displays smoothed estimates of the fraction of individuals who report COVID-like symptoms within the past 24-hours by age and gender.
+Figure~\ref{fig:fbsymptoms} in Section~\ref{app:in_add_details} of the Supplementary Materials displays smoothed estimates of the fraction of individuals who report COVID-like symptoms within the past 24-hours by age and gender.
 The Delphi's COVID-19 Trends and Impact Survey (CTIS) is used as the main source of auxiliary information on time-varying characteristics of the population of Indiana, e.g., displaying COVID-like symptoms or in contact with COVID-19 positive individuals.
@@
 The attentive data analyst will recognize the estimator $\bar y_n$ is now biased for simple random samples.
-In Section~\ref{app:memestimator} of the supplementary materials, assuming sensitivity and specificity are known a priori, a novel iterative procedure is used to construct the estimator $\tilde y_n = (\bar y_n - FP)/(1-(FP+FN))$, which is unbiased under simple random sampling (SRS); see Section~\ref{app:modelbased} in the Supplementary Materials for a discussion of the connection to model-based estimators.
+In Section~\ref{app:memestimator} of the Supplementary Materials, assuming sensitivity and specificity are known a priori, a novel iterative procedure is used to construct the estimator $\tilde y_n = (\bar y_n - FP)/(1-(FP+FN))$, which is unbiased under simple random sampling (SRS); see Section~\ref{app:modelbased} in the Supplementary Materials for a discussion of the connection to model-based estimators.
 In the language of the epidemiology literature, the estimator $\tilde y_n$ is the standard estimator for correcting for measurement error using false positive and negative rates that have been estimated from a validation sample, where $\bar y_n$ is the mis-measured average \& $\tilde y_n$ is the corrected average.
@@
  %, which is scaled by the factor $(1 - FP - FN)^{-1}$.
- For ease of comprehension, a notation glossary is provided in Section~\ref{app:notation} in the supplementary materials.
+ For ease of comprehension, a notation glossary is provided in Section~\ref{app:notation} in the Supplementary Materials.
  \vspace{-0.25cm}
@@
 Figure~\ref{fig:r0-bias} displays the bias as a function of the relative sampling fraction ignoring the final term (i.e., $S_t / S_{t-1} \approx 1$).
-Section~\ref{app:cori_rt} in the supplementary materials presents an alternative estimator of the instantaneous effective reproductive number and shows similar bias~\cite{Cori20113}.
+Section~\ref{app:cori_rt} in the Supplementary Materials presents an alternative estimator of the instantaneous effective reproductive number and shows similar bias~\cite{Cori20113}.
 
@@
  Expectation of~\ref{eq:auxinfoprob} with respect to the sampling design yields~\eqref{eq:propensity}.
-Solving~\eqref{eq:auxinfoprob} is done by iteratively re-weighted least squares (IRLS); see Section~\ref{app:irls} and Section~\ref{app:notation} in the supplementary materials for details and a notation glossary respectively.
+Solving~\eqref{eq:auxinfoprob} is done by iteratively re-weighted least squares (IRLS); see Section~\ref{app:irls} and Section~\ref{app:notation} in the Supplementary Materials for details and a notation glossary respectively.
  \vspace{-0.5cm}
@@
 Rather than relying solely on IPW or epidemiological forecasts, we combine forecasting and inverse-probability weighting by extending recent work by~\cite{Chen2019} to account time-varying propensities and measurement-error.
-Given the overlap with Section~\ref{section:IPWerrordecomp}, for conciseness the doubly-robust estimator is presented in detail in Section~\ref{app:drestimation} of the Supplementary Materials.
+Given the overlap with Section~\ref{section:IPWerrordecomp}, for conciseness the doubly-robust estimator is presented in Section~\ref{app:drestimation} of the Supplementary Materials.
  \end{remark}
@@
 We collect age, sex, and demographic information as well as COVID-19 related symptoms, which Delphi defines as having (1) a fever and cough, (2) shortness of breath, \emph{or} (3) difficulty breathing.
-Table~\ref{tab:comparison} in Section~\ref{app:in_add_details} of the supplementary materials shows minimal bias with respect to symptom distributions when comparing to Indiana's random sample to the CTIS data collected from April 25th to April 29th.
+Table~\ref{tab:comparison} in Section~\ref{app:in_add_details} of the Supplementary Materials shows minimal bias with respect to symptom distributions when comparing to Indiana's random sample to the CTIS data collected from April 25th to April 29th.
 
@@
 As COVID-19 symptom status is likely to alter the testing propensity and likelihood of active infection, for illustrative purposes CTIS and hospitalization data are used in two ways (termed IPW1 and IPW2 respectively) to impute symptom status per strata.
-We present IPW2 below; see Section~\ref{app:in_add_details} in  of the Supplementary Materials for details on IPW1 and additional discussion.
+We present IPW2 below; see Section~\ref{app:in_add_details} of the Supplementary Materials for details on IPW1 and additional discussion.
 
@@
 The first estimates the probability of contact with a COVID-19 positive individual given demographic and COVID-19 test status (positive or negative COVID-19 test).
-Figure~\ref{fig:contactlik1} and~\ref{fig:contactlik2} in Section~\ref{app:in_add_details} in the supplementary materials presents the estimated likelihood of contact with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.
+Figure~\ref{fig:contactlik1} and~\ref{fig:contactlik2} in Section~\ref{app:in_add_details} in the Supplementary Materials presents the estimated likelihood of contact with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.
 The second estimates the probability of fever given demographic, COVID-19 test status, \emph{and} whether the individual has had contact with a COVID-19 positive individual.
-Figure~\ref{fig:symptomlik1} and~\ref{fig:symptomlik2} in Section~\ref{app:in_add_details} in the supplementary materials presents the estimated likelihood of reporting a fever with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.
+Figure~\ref{fig:symptomlik1} and~\ref{fig:symptomlik2} in Section~\ref{app:in_add_details} in the Supplementary Materials presents the estimated likelihood of reporting a fever with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.
 We see that likelihood of fever varies greatly based on whether the individual also reported contact with a COVID-19 positive individual.
@@
 When the COVID-19 test status is positive, the probability of fever is allowed to depend on whether the individual was hospitalized.
-Figure~\ref{fig:symptomlik1_model2} and~\ref{fig:symptomlik2_model2} in Section~\ref{app:in_add_details} in the supplementary materials presents the estimated likelihood of reporting a fever with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.
+Figure~\ref{fig:symptomlik1_model2} and~\ref{fig:symptomlik2_model2} in Section~\ref{app:in_add_details} in the Supplementary Materials presents the estimated likelihood of reporting a fever with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.
 We see that hospitalization significantly increases the likelihood of fever as expected.
@@
 Note that the ratio of testing propensity within strata across fever status is time-varying, starting 50 times more likely in early April and dropping to ten times more likely by end of 2020.
-Similar plots for other strata and a discussion of their relative testing propensities can be found in Section~\ref{section:propplots} of the supplementary materials.
+Similar plots for other strata and a discussion of their relative testing propensities can be found in Section~\ref{section:propplots} of the Supplementary Materials.
 
@@
  To account for potential misspecification of the marginal IFR, we perform the same analysis with a 10\% increase and decrease in the marginal IFR under the log-linear model.
-Figure~\ref{fig:tv_air_sens} in the supplementary materials presents the model-based and doubly robust estimates in these scenarios.
+Figure~\ref{fig:tv_air_sens} in the Supplementary Materials presents the model-based and doubly robust estimates in these scenarios.
  \end{remark}
@@
 Each weight~$w_t^{(k)} = 0$ for $t < t_k$ and $w_t^{(k)} \to 1$ as $t \to \infty$.
-See Section~\ref{app:prior_modelbased} in the supplementary materials for details on prior specification, posterior predictive checks, and discussion on case undercounting.
+See Section~\ref{app:prior_modelbased} in the Supplementary Materials for details on prior specification, posterior predictive checks, and discussion on case undercounting.
 
@@
 In our analysis, the exponential discounting parameter is set to $20$ days to match with prior evidence that ``30\% to 40\% of people will still test positive at three weeks''~\citep{wlocneg}.
-To compare with case count data, Figure~\ref{fig:undercounting} in the supplementary materials computes the cumulative undercount factor of observed case counts as compared to new infections.
+To compare with case count data, Figure~\ref{fig:undercounting} in the Supplementary Materials computes the cumulative undercount factor of observed case counts as compared to new infections.
 Posterior distributions for the age-specific new infections are used as part of the doubly-robust estimator of the active infection rate.
@@
 
-Figure~\ref{fig:air_cis} in the supplementary materials presents the confidence intervals per time point for the IPW2 estimator.
+Figure~\ref{fig:air_cis} in the Supplementary Materials presents the confidence intervals per time point for the IPW2 estimator.
 The confidence interval length decreases substantially over time, reflecting the increased testing capacity over this window of time.
@@
 This points to the importance of the statistical decomposition~\eqref{eq:statdecomp2} and the discussion in Section~\ref{section:IPWerrordecomp}.
-Finally, given the reliance on a probabilistic sample with low response rate (see Remark~\ref{rmk:limitations}), a sensitivity analysis showing the impact of a potential unmeasured confounder is presented in Section~\ref{sec:sensitivity_confounders} of the supplementary materials.
+Finally, given the reliance on a probabilistic sample with low response rate (see Remark~\ref{rmk:limitations}), a sensitivity analysis showing the impact of a potential unmeasured confounder is presented in Section~\ref{sec:sensitivity_confounders} of the Supplementary Materials.
   % We also construct ratio estimates as discussed in Section~\ref{section:rates}.

### 2023-01-10 write-up/letters/cover letter-covid.tex
Finishing paper once and for all...ha

@@
 
-Kindly find enclosed the revised version of the manuscript number AOAS2006-048 titled ``Addressing selection bias and measurement error in COVID-19 case count data using auxiliary information.''  We are happy that the two reviewers, associate editor, and editor have all found the revision much improved.
-We have revised the article to address the few remaining concerns that one of the reviewers raised but are glad that they ``appreciate the paper and [are] very supportive of the pragmatic approach.''  We have edited the remark, performed a sensitivy analysis to assess sensitivity due to potential confounding, and adjusted the figures to assist in comprehension.
+Kindly find enclosed the revised version of the manuscript number AOAS2006-048 titled ``Addressing selection bias and measurement error in COVID-19 case count data using auxiliary information.''  We are happy that the revised manuscript was found to address the remaining concerns raised by the reviewers.
+At the Editor's request, we have shortened the article to 26 pages without references and included 5 figures.
+The main paper still retains all of the main points while the supplement contains important but secondary analyses and discussions that were removed from the main paper in this editing process.
 
 We continue to believe the work presents an important advance in how researchers should think about addressing time-varying selection bias and measurement error to better understanding the COVID-19 pandemic trajectory.
-Attached please find a detailed response to the referee outlining how the changes to the manuscript address the concerns raised in their report.
 
